### 2017-01-06 CO750-ApproximationAlgorithmsTopics/lectures/lecture_01.tex
Starting courses repo

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{mathtools}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-01-03}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\section{Introduction}
\paragraph{}
The central task we are interested in is solving $NP$-hard optimization problems. Finding optimal solutions to such problems in polynomial time is unlikely (unless $P = NP$).

\paragraph{}
Given that we are faced with an $NP$-hard problem there are a couple possible approaches:
\begin{enumerate}
\item Exact algorithms: Find optimal solution slowly for large input.
\item Heuristic algorithms: Find efficiently a feasible solution, but with no guarantee of optimality.
\item Approximation Algorithms: Find efficiently a feasible solution and give a certified bound on how close the solution is to optimal.
\end{enumerate}
The goal of this course is to study the third approach, and learn successful methods of designing approximation algorithms. We will draw heavily from network design for examples.

\paragraph{}
What we will see in this course:
\begin{enumerate}
\item Combinatorial techniques: Those which exploit the structure of feasible solutions.
\item Linear Programming techniques:
\begin{enumerate}
\item Primal-Dual methods: Those which exploit Duality Theory.
\item Rounding methods: Those which round fractional variables.
\end{enumerate}
\item Recent Results and Open Problems!
\end{enumerate}
\section{Vertex Cover Problem}
\paragraph{}
In an instance of {\it Vertex Cover} (VC) the input is a graph $G=(V,E)$. We may assume without loss of generality that $G$ is connected. If $G$ is not connected we may apply an algorithm which solves VC for connected graphs to each component of $G$ and union our solutions to obtain a solution for $G$. The output (solution) to an instance of VC is a subset of vertices $C \subseteq V$ of minimum cardinality satisfying $$\forall e \in E, e \cap C \neq \emptyset.$$
\paragraph{}
A possible exact algorithm would list all subsets of $V$ and return the minimum cardinality subset among those which are covers. The running time of this algorithm would be $O(2^n)$ where $n = |V|$. This is inefficient, ie. not polynomial in the size of the input graph.
\paragraph{}
An instance of a problem has a given size $m \in \Z$. The {\it running time} of the an algorithm is the number of elementary operations it performs ($+$, $-$, $\times$, $==$, $<$, $>$, $\dots$) during its execution. We measure the running time as a function of the input size $m$. An algorithm is said to be {\it efficient} if its running time is a polynomial function of $m$, ie. $O(m^k)$ for some constant $k$. For problems on graphs $n =|V|$ provides a rough estimate of the input size.
\paragraph{}
Let $\alpha \geq 1$ and let $\Pi$ be a minimization problem. We say $\cA$ is an {\it $\alpha$-approximation algorithm} for $\Pi$ provided $\cA$ is efficient on all instances of $\Pi$ and for every instance $\pi$ of $\Pi$, the solution returned by $\cA$ on $\pi$ is feasible and has objective value at most $\alpha$ times the optimal value of a solution for $\pi$. 
\paragraph{}
If instead $\Pi$ is a maximation problem then $\alpha \leq 1$ and $\cA$ returns a solution at least $\alpha$ times the optimal value.
\paragraph{}
Typical values of $\alpha$ are $2$, $O(1)$, $\log n$, or $(1 + \epsilon)$ for instance.
\paragraph{}
Returning our attention to VC, let $C^*$ be an optimal solution of a given instance of $VC$. Our goal is to find a feasible solution $C$ such that $|C| \leq \alpha |C^*|$. But we run into some difficulty since we do not know $|C^*|$. Our idea to get around this is to obtain a lower bound on $|C^*|$ and compare $|C|$ with our lower bound.
\paragraph{}
Let $M$ be a maximum (we only need maximal in full generality) matching on $G$. It is not too hard to see that $|M| \leq |C^*|$. Indeed if $|C^*| < |M|$ then there exists two edges in $M$ which are covered by the same vertex in $C^*$ (from pigeonhole principle), but this contradicts that $M$ is a matching. Note that any matching provides a lower bound on $|C^*|$, but a maximum matching yields the tightest lower bound, a desirable property.
\paragraph{}
Via the Blossom Algorithm of Edmunds a maximum cardinality (in fact maximum weight for arbitrary weights) matching can be computed efficiently.
\paragraph{Approximation Algorithm for VC}
Consider the algorithm:
\begin{enumerate}
\item Compute a maximum cardinality matching $M$
\item Output $C = \bigcup_{e \in M} e$.
\end{enumerate}
\begin{theorem}
Our Approximation Algorithm for VC is a $2$-approximation for VC.
\end{theorem}
\begin{proof}
Since maximum matching runs in polynomial time, our algorithm runs in polynomial time. To see that $C$ is feasible suppose that there exists $e \in E$ such that $e \cap C = \emptyset$. The $M \cup e$ is a matching greater cardinality than $M$. Therefore $C$ is a cover. Furthermore $$|C| = 2|M| \leq 2|C^*|$$ as desired.
\end{proof}
\paragraph{}
To see that our analysis is tight consider a graph consisting of a single edge. In this case $|C| = 2 = 2|C*|$ and hence our analysis is tight.
\section{Weighted Vertex Cover}
\paragraph{}
In an instance of {\it Weighted Vertex Cover} (WVC)the input is a connected graph $G = (V,E)$ and weights $w_v$ for all $v \in V$. The output is again a cover, but now of minimum weight, not cardinality. This problem generalizes VC whose instances can be recovered by assigned all weights the value $1$.
\paragraph{}
Our previous algorithm may fail. To see this consider a star graph with very large weight on the center vertex and very low weights on the outer vertices. The reader is encouraged to verify that the previous algorithm fails under such conditions. 
\paragraph{}
For WVC we will use linear programming (LP) to give lower bounds. Often optimization problems can be naturally formulated as integer programs (IP). Consider the following IP for WVC:
\begin{align*}
\min &\sum_{v \in V} w_v x_v \\
x_u + x_v &\geq 1 &\forall uv \in E \\
x_u& \in \{0,1\} &\forall u \in V.
\end{align*}
By relaxing the IP (change constraints $x_u \in \{0,1\}$ to $0 \leq x_u \leq 1$) we obtain an LP that can be solved efficiently. Since this LP has $O(|V|^2)$ constraints we can solve it in time polynomial in the size of our input graph. This LP gives a lower bound on $OPT$, the optimal value of our WVC instance.
\paragraph{Approximation Algorithm for WVC}
Consider the algorithm:
\begin{enumerate}
\item Solve the LP relaxation. Let $x^*$ be the optimal solution of the LP.
\item (Round up all variables of value at least $\frac{1}{2}$) Output: $C = \{v \in V: x^*_v \geq \frac{1}{2}\}$.
\end{enumerate}
\begin{theorem}
Our Approximation Algorithm for WVC is a $2$-approximation for WVC.
\end{theorem}
\begin{proof}
Algorithm is polynomial since compact linear programs can be solved in polynomial time. Now the weight of $C$ is
$$w(C) = \sum_{v \in C} w_v = 2\sum_{v \in C} w_v \frac{1}{2} \leq 2 \sum_{v \in C}w_v x^*_v \leq 2 \sum_{v \in V} w_v x^*_v \leq 2\cdot OPT.$$
It remains to verify that $C$ is feasible. Indeed for each edge $uv \in E$, either $x^*_u \geq \frac{1}{2}$ or $x^*_v \geq \frac{1}{2}$ by the constraint
$$x_u + x_v \geq 1$$
and thus either $u \in C$ or $v \in C$.
\end{proof}
\end{document}

### 2017-01-06 CO750-ApproximationAlgorithmsTopics/lectures/lecture_02.tex
Starting courses repo

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\U}{\mathbb{U}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-01-05}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\paragraph{}
A natural idea when it comes to constructing feasible solutions for algorithmic problems is by making greedy choices. That is, by making at each iteration a choice which appears locally to be the best possible choice. In this lecture we will study some problems for which we can construct greedy approximation algorithms.
\section{Set Cover}
\paragraph{}
In the Set Cover problem (SC) we are given a set of elements $\U = \{e_1, \dots, e_n\}$ and a collection of sets $S_1, \dots, S_m \subseteq \U$ with costs $c(S_i)$ for $i \in [1,m] \cap \Z$. We are to find a sub-collection of sets of minimum cost covering $\U$. More formally $opt$ is equal to
$$\min_{I \subseteq \{1,\dots,m\}} \{ \sum_{i \in I} c(S_i) : \bigcup_{i \in I} S_i = \U \}.$$
\paragraph{}
Consider the following greedy algorithm for SC:
\begin{enumerate}
\item Set $I = \emptyset$
\item While $\bigcup_{i \in I} S_i \neq \U$
	\begin{enumerate}
	\item Compute $$price(S_i) = \frac{c(S_i)}{|S_i \backslash \cup_{j\in I} S_j|}$$ for all $i \in [1,m]\cap \Z$
	\item Let $S_{i^*}$ be the set of minimum price.
	\item Set $I = I \cup \{i^*\}$
	\end{enumerate}
\item Return $I$.
\end{enumerate}
\begin{theorem}
The greedy algorithm for SC is an $O(\log n)$-approximation algorithm (recall $n = |\U|$).
\end{theorem}
\begin{proof}
Reindex the elements of $\U$ in the order they are covered by the greedy algorithm. So then $e_1$ is the first element covered by the greedy algorithm, and $e_n$ is the last. We say $S_i$ newly covers $e_j$ if $S_i$ is the first set taken by the greedy algorithm for which $e_j \in S_i$. Fix $price(S_i)$ as computed when taken by the greedy algorithm, and define
$$price(e_j) = price(S_i)$$
where $e_j$ is newly coved by $S_i$.
\paragraph{}
Let $N(i)$ be the number of elements newly covered by $S_i$. Observe that the fixed $price(S_i) = \frac{c(S_i)}{N(i)}$. Now we have that
$$c(I) = \sum_{i \in I} c(S_i) = \sum_{i \in I} price(S_i)N(i) = \sum_{j=1}^n price(e_j).$$
\paragraph{Claim 1}
Let $a_1, \dots, a_\ell$ and $b_1, \dots, b_\ell$ be positive integers. Then $$\min_{i = 1,\dots,\ell} \frac{a_i}{b_i} \leq \frac{\sum_{i=1}^\ell a_i}{\sum_{i=1}^\ell b_i}.$$
\paragraph{Proof of Claim 1}
Let $\frac{a_k}{b_k} = \min_{i = 1,\dots,\ell} \frac{a_i}{b_i}$. Then
$$ \frac{\sum_{i=1}^\ell a_i}{\sum_{i=1}^\ell b_i} = \frac{\sum_{i\neq k} a_i}{\sum_{i=1}^\ell b_i} + \frac{a_k}{\sum_{i=1}^\ell b_i} \geq \frac{a_k}{\sum_{i=1}^\ell b_i} \geq \frac{a_k}{b_k}.$$
\paragraph{Claim 2}
The inequality $price(e_j) \leq \frac{opt}{n-j+1}$ for any $j \in [1,n]\cap \Z$.
\paragraph{Proof of Claim 2}
Consider the iteration where $e_j$ is covered. At the beginning of this iteration there are at least $n-j+1$ elements uncovered. At the cost of $opt$ they can all be covered. Let $I^*$ be the indices of the subsets of the optimal solution covering $e_j, \dots, e_n$. Then:
$$price(e_j) \leq \min_{i \in I^*} \{ price(S_i)\} \leq \frac{ \sum_{i \in I^*} c(S_i)}{\sum_{i \in I^*} N(i)} \leq \frac{opt}{n-j+1}$$
with the second inequality following from Claim $1$.
\paragraph{Finishing Main Result}
We have the following
$$c(I) \leq \sum_{j=1}^n price(e_j) \leq \sum_{j=1}^n \frac{opt}{n-j+1} = \sum_{j=1}^n \frac{opt}{j} = H_n \cdot opt = O(\log n) opt$$
where $H_n = 1 + \frac{1}{2} + \dots + \frac{1}{n}$ is the harmonic series.
\end{proof}
\paragraph{}
Is this analysis tight? The answer is yes. Consider the following example. Let $\U = \{e_1, e_2, \dots, e_n\}$ and let $S_i = \{e_i\}$ for $i = 1 \dots n$ and let $S_{n+1} = \U$. Set $c(S_i) = \frac{1}{i}$ for $i = 1 \dots n$ and set $c(S_{n+1}) = 1 + \epsilon$ for some $\epsilon > 0$. Then the greedy algorithm will choose the sets $S_n, S_{n-1}, \dots, S_1$ in that order with cost $H_n$, while $S_{n+1}$ covers $\U$ and is the optimal solution for appropriate $\epsilon$ with $c(\sum_{i=1}^n S_i) = H_n \leq O(\log n) (1+\epsilon)$.
\paragraph{}
In fact, this is the best approximation factor we can obtain (Feige '98) unless $P = NP$. Later we will see a method of proving such a hardness result.
\section{k-Center}
\paragraph{}
In the $k$-Center problem we are given a graph $G = (V,E)$ with edge ``lengths" $l_e \geq 0$ for all $e \in E$. We are also given a number $k \in \N$. We are to find $k$ centers $F \subseteq V$ that minimize the maximum distance from any $u \in V$ to the nearest center. Formally  for all $u,v \in V$ we let $\ell(u,v) := $ length of a shortest $uv$-path. For $F \subseteq V$, $v \in V$ let $\ell (u,F) = \min_{v \in F} \ell(u,v)$. Our optimal value is given as
$$ opt = \min_{F \subseteq V} \max_{v \in V} \ell (v,F).$$
\paragraph{}
Consider the following greedy algorithm for $k$-Center:
\begin{enumerate}
\item Pick $i \in V$ arbitrarily. Set $F = \{i\}$.
\item While $|F| < k$ do:
	\begin{enumerate}
	\item Let $j = argmax_{j \in V} \ell(j,F)$
	\item Set $F:= F \cup \{j\}$
	\end{enumerate}
\item Return $F$.
\end{enumerate}
\begin{theorem}
The greedy algorithm for $k$-Center is a $2$-approximation algorithm.
\end{theorem}
\begin{proof}
Let $F^* = \{u_1^*, \dots, u_k^*\}$ be the optimal solution. Partition the vertices of $V$ as $V_1^*, \dots, V_k^*$ where $V_i^*$ contains the vertices of $V$ closest to $u_i^*$ versus any $u_j^*$ for $j \neq i$. We delineate two cases:
\begin{enumerate}
\item $F$ contains two vertices $w,z$ from some $V_i^*$
\item $F$ contains one vertex $f_i$ from each $V_i^*$.
\end{enumerate}
\paragraph{Case 1}
By our greedy choice $$\max_{v \in V} \ell(v,F) \leq \ell (w,z).$$
Hence by shortest path we observe:
 $$\max_{v \in V} \ell(v,F) \leq \ell (w,z) \leq \ell(w, u_i^*) + \ell(u_i^*, f_i) \leq 2opt.$$
\paragraph{Case 2}
By shortest path, for all $v \in V_i$,
$$\ell(v,f_i) \leq \ell(v,u_i^*) + \ell(u_i^*,f_i) \leq 2opt.$$
Hence
$$\max_{v \in V} \ell(v,F) \leq 2opt.$$
\end{proof}
\paragraph{}
As usual at this point we ask ourselves if our analysis is tight. The answer is again yes, but instead of constructing a worst case example, this time we will demonstrate something stronger: that our algorithm is the best possible unless $P = NP$.
\begin{theorem}
Unless $P=NP$, for all $\epsilon > 0$ there does not exist a $(2-\epsilon)$-approximation algorithm for $k$-Center. 
\end{theorem}
\paragraph{}
To prove this theorem we will perform a gap-producing reduction from Dominating Set (which is $NP$-complete). The general strategy for such a reduction given some problem $A$ is as follows:
\begin{enumerate}
\item Take $NP$-complete problem $B$
\item Decide on a constant $a$, and factor $\alpha$ for which $opt$, the optimal value for $A$, does not lie in $(a, \alpha a]$.
\item Give a polynomial time reduction from $B$ to $A$ such that
\begin{enumerate}
\item A yes instance of $B$ is mapped to instances of $A$ with $opt \leq a$
\item A no instance of $B$ is mapped to instances of $A$ with $opt > \alpha a$.
\end{enumerate}
\end{enumerate}
Then there can be no $\alpha$-approximation algorithm for $A$, as otherwise we could use it to decide $B$. We will see how this plays out more concretely in the proof of our above theorem.\\ \\
\begin{proof}
The Dominating Set problem gives a graph $G = (V,E)$ and a number $k$, and asks one to decide if $\exists U \subseteq V$ with $|U| \leq k$ such that $\forall v \in V\backslash U$, $\exists u \in U$, $uv \in E$. Consider an instance of Dominating Set, $(G,k)$. Construct an instance of $k$-Center $(G', \ell)$ by setting $G' = G$ and $\ell_e =1$ for all $e \in E$. Observe that there exists a dominating set of size $k$ on $G$ if and only if there exists a $k$-Center solution on $G'$ of cost $1$. Therefore if we have a $(2-\epsilon)$-approximation algorithm for $k$-Center, say $\cA$ we could run $\cA$ on $G'$ and consider the returned value $c$. If $c < 2$ then $opt \leq 1$ since our optimal values are integral by construction. Thus our $G$ is a yes instance of Dominating Set. If $c \geq 2$ then $G$ is a no instance of Dominating Set. Observe that $c \leq (2-\epsilon) opt$ and thus
$$2 \leq (2-\epsilon)opt$$
which implies $opt$ is greater than $1$. Note that our proof is analogous with the general strategy when $a = 1$ and $\alpha = 2-\epsilon$.
\end{proof}
\paragraph{}
We conclude by observing that if $\alpha = 1$ our gap-producing reductions reduce to standard $NP$-completeness reductions.
\end{document}

### 2017-01-06 CO750-ApproximationAlgorithmsTopics/lectures/lecture_03.tex
Starting courses repo

--- 2017-01-06 CO750-ApproximationAlgorithmsTopics/lectures/lecture_02.tex
@@
 \newcommand{\Q}{\mathbb{Q}}
-\newcommand{\U}{\mathbb{U}}
 
@@
 
-\lhead{2017-01-05}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-01-10}
+\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+\section{Steiner Trees}
 \paragraph{}
-A natural idea when it comes to constructing feasible solutions for algorithmic problems is by making greedy choices.
-That is, by making at each iteration a choice which appears locally to be the best possible choice.
-In this lecture we will study some problems for which we can construct greedy approximation algorithms.
-\section{Set Cover}
+In the Steiner Tree (ST) problem we are given a graph $G=(V,E)$ and edge lengths (costs) $c_e \geq 0$, and a set $R \subseteq V$ called the set of terminals.
+We are to find a minimum cost tree $T$ connecting all the terminals.
 \paragraph{}
-In the Set Cover problem (SC) we are given a set of elements $\U = \{e_1, \dots, e_n\}$ and a collection of sets $S_1, \dots, S_m \subseteq \U$ with costs $c(S_i)$ for $i \in [1,m] \cap \Z$.
-We are to find a sub-collection of sets of minimum cost covering $\U$.
-More formally $opt$ is equal to
-$$\min_{I \subseteq \{1,\dots,m\}} \{ \sum_{i \in I} c(S_i) : \bigcup_{i \in I} S_i = \U \}.$$
+Note that if $R = V$ we have a spanning tree instance (thus we can solve in polynomial time).
+Also if $|R| = 1$ we have a shortest path instance (and again we can solve in polynomial time).
+But for general $|R| = k$ the problem in $NP$-hard.
 \paragraph{}
-Consider the following greedy algorithm for SC:
+A function $d : V \times V \rightarrow \R_+$ is a {\it metric} if 
 \begin{enumerate}
-\item Set $I = \emptyset$
-\item While $\bigcup_{i \in I} S_i \neq \U$
-	\begin{enumerate}
-	\item Compute $$price(S_i) = \frac{c(S_i)}{|S_i \backslash \cup_{j\in I} S_j|}$$ for all $i \in [1,m]\cap \Z$
-	\item Let $S_{i^*}$ be the set of minimum price.
-	\item Set $I = I \cup \{i^*\}$
+\item $d(u,u) = 0$
+\item $d(u,v) = d(v,u)$ for all $u,v \in V$
+\item $d(u,v) \leq d(u,z) + d(z,v)$ for all $u,v,z \in V$.
 \end{enumerate}
-\item Return $I$.
-\end{enumerate}
-\begin{theorem}
-The greedy algorithm for SC is an $O(\log n)$-approximation algorithm (recall $n = |\U|$).
-\end{theorem}
-\begin{proof}
-Reindex the elements of $\U$ in the order they are covered by the greedy algorithm.
-So then $e_1$ is the first element covered by the greedy algorithm, and $e_n$ is the last.
-We say $S_i$ newly covers $e_j$ if $S_i$ is the first set taken by the greedy algorithm for which $e_j \in S_i$.
-Fix $price(S_i)$ as computed when taken by the greedy algorithm, and define
-$$price(e_j) = price(S_i)$$
-where $e_j$ is newly coved by $S_i$.
-\paragraph{}
-Let $N(i)$ be the number of elements newly covered by $S_i$. Observe that the fixed $price(S_i) = \frac{c(S_i)}{N(i)}$. Now we have that
-$$c(I) = \sum_{i \in I} c(S_i) = \sum_{i \in I} price(S_i)N(i) = \sum_{j=1}^n price(e_j).$$
-\paragraph{Claim 1}
-Let $a_1, \dots, a_\ell$ and $b_1, \dots, b_\ell$ be positive integers. Then $$\min_{i = 1,\dots,\ell} \frac{a_i}{b_i} \leq \frac{\sum_{i=1}^\ell a_i}{\sum_{i=1}^\ell b_i}.$$
-\paragraph{Proof of Claim 1}
-Let $\frac{a_k}{b_k} = \min_{i = 1,\dots,\ell} \frac{a_i}{b_i}$. Then
-$$ \frac{\sum_{i=1}^\ell a_i}{\sum_{i=1}^\ell b_i} = \frac{\sum_{i\neq k} a_i}{\sum_{i=1}^\ell b_i} + \frac{a_k}{\sum_{i=1}^\ell b_i} \geq \frac{a_k}{\sum_{i=1}^\ell b_i} \geq \frac{a_k}{b_k}.$$
-\paragraph{Claim 2}
-The inequality $price(e_j) \leq \frac{opt}{n-j+1}$ for any $j \in [1,n]\cap \Z$.
-\paragraph{Proof of Claim 2}
-Consider the iteration where $e_j$ is covered.
-At the beginning of this iteration there are at least $n-j+1$ elements uncovered.
-At the cost of $opt$ they can all be covered.
-Let $I^*$ be the indices of the subsets of the optimal solution covering $e_j, \dots, e_n$.
-Then:
-$$price(e_j) \leq \min_{i \in I^*} \{ price(S_i)\} \leq \frac{ \sum_{i \in I^*} c(S_i)}{\sum_{i \in I^*} N(i)} \leq \frac{opt}{n-j+1}$$
-with the second inequality following from Claim $1$.
-\paragraph{Finishing Main Result}
-We have the following
-$$c(I) \leq \sum_{j=1}^n price(e_j) \leq \sum_{j=1}^n \frac{opt}{n-j+1} = \sum_{j=1}^n \frac{opt}{j} = H_n \cdot opt = O(\log n) opt$$
-where $H_n = 1 + \frac{1}{2} + \dots + \frac{1}{n}$ is the harmonic series.
-\end{proof}
-\paragraph{}
-Is this analysis tight?
-The answer is yes.
-Consider the following example.
-Let $\U = \{e_1, e_2, \dots, e_n\}$ and let $S_i = \{e_i\}$ for $i = 1 \dots n$ and let $S_{n+1} = \U$.
-Set $c(S_i) = \frac{1}{i}$ for $i = 1 \dots n$ and set $c(S_{n+1}) = 1 + \epsilon$ for some $\epsilon > 0$.
-Then the greedy algorithm will choose the sets $S_n, S_{n-1}, \dots, S_1$ in that order with cost $H_n$, while $S_{n+1}$ covers $\U$ and is the optimal solution for appropriate $\epsilon$ with $c(\sum_{i=1}^n S_i) = H_n \leq O(\log n) (1+\epsilon)$.
-\paragraph{}
-In fact, this is the best approximation factor we can obtain (Feige '98) unless $P = NP$. Later we will see a method of proving such a hardness result.
-\section{k-Center}
-\paragraph{}
-In the $k$-Center problem we are given a graph $G = (V,E)$ with edge ``lengths" $l_e \geq 0$ for all $e \in E$.
-We are also given a number $k \in \N$.
-We are to find $k$ centers $F \subseteq V$ that minimize the maximum distance from any $u \in V$ to the nearest center.
-Formally  for all $u,v \in V$ we let $\ell(u,v) := $ length of a shortest $uv$-path.
-For $F \subseteq V$, $v \in V$ let $\ell (u,F) = \min_{v \in F} \ell(u,v)$.
-Our optimal value is given as
-$$ opt = \min_{F \subseteq V} \max_{v \in V} \ell (v,F).$$
-\paragraph{}
-Consider the following greedy algorithm for $k$-Center:
-\begin{enumerate}
-\item Pick $i \in V$ arbitrarily. Set $F = \{i\}$.
-\item While $|F| < k$ do:
-	\begin{enumerate}
-	\item Let $j = argmax_{j \in V} \ell(j,F)$
-	\item Set $F:= F \cup \{j\}$
-	\end{enumerate}
-\item Return $F$.
-\end{enumerate}
-\begin{theorem}
-The greedy algorithm for $k$-Center is a $2$-approximation algorithm.
-\end{theorem}
-\begin{proof}
-Let $F^* = \{u_1^*, \dots, u_k^*\}$ be the optimal solution.
-Partition the vertices of $V$ as $V_1^*, \dots, V_k^*$ where $V_i^*$ contains the vertices of $V$ closest to $u_i^*$ versus any $u_j^*$ for $j \neq i$.
-We delineate two cases:
-\begin{enumerate}
-\item $F$ contains two vertices $w,z$ from some $V_i^*$
-\item $F$ contains one vertex $f_i$ from each $V_i^*$.
-\end{enumerate}
-\paragraph{Case 1}
-By our greedy choice $$\max_{v \in V} \ell(v,F) \leq \ell (w,z).$$
-Hence by shortest path we observe:
- $$\max_{v \in V} \ell(v,F) \leq \ell (w,z) \leq \ell(w, u_i^*) + \ell(u_i^*, f_i) \leq 2opt.$$
-\paragraph{Case 2}
-By shortest path, for all $v \in V_i$,
-$$\ell(v,f_i) \leq \ell(v,u_i^*) + \ell(u_i^*,f_i) \leq 2opt.$$
-Hence
-$$\max_{v \in V} \ell(v,F) \leq 2opt.$$
-\end{proof}
-\paragraph{}
-As usual at this point we ask ourselves if our analysis is tight.
-The answer is again yes, but instead of constructing a worst case example, this time we will demonstrate something stronger: that our algorithm is the best possible unless $P = NP$.
-\begin{theorem}
-Unless $P=NP$, for all $\epsilon > 0$ there does not exist a $(2-\epsilon)$-approximation algorithm for $k$-Center. 
-\end{theorem}
-\paragraph{}
-To prove this theorem we will perform a gap-producing reduction from Dominating Set (which is $NP$-complete). The general strategy for such a reduction given some problem $A$ is as follows:
-\begin{enumerate}
-\item Take $NP$-complete problem $B$
-\item Decide on a constant $a$, and factor $\alpha$ for which $opt$, the optimal value for $A$, does not lie in $(a, \alpha a]$.
-\item Give a polynomial time reduction from $B$ to $A$ such that
-\begin{enumerate}
-\item A yes instance of $B$ is mapped to instances of $A$ with $opt \leq a$
-\item A no instance of $B$ is mapped to instances of $A$ with $opt > \alpha a$.
-\end{enumerate}
-\end{enumerate}
-Then there can be no $\alpha$-approximation algorithm for $A$, as otherwise we could use it to decide $B$. We will see how this plays out more concretely in the proof of our above theorem.\\ \\
-\begin{proof}
-The Dominating Set problem gives a graph $G = (V,E)$ and a number $k$, and asks one to decide if $\exists U \subseteq V$ with $|U| \leq k$ such that $\forall v \in V\backslash U$, $\exists u \in U$, $uv \in E$.
-Consider an instance of Dominating Set, $(G,k)$.
-Construct an instance of $k$-Center $(G', \ell)$ by setting $G' = G$ and $\ell_e =1$ for all $e \in E$.
-Observe that there exists a dominating set of size $k$ on $G$ if and only if there exists a $k$-Center solution on $G'$ of cost $1$.
-Therefore if we have a $(2-\epsilon)$-approximation algorithm for $k$-Center, say $\cA$ we could run $\cA$ on $G'$ and consider the returned value $c$.
-If $c < 2$ then $opt \leq 1$ since our optimal values are integral by construction.
-Thus our $G$ is a yes instance of Dominating Set.
-If $c \geq 2$ then $G$ is a no instance of Dominating Set.
-Observe that $c \leq (2-\epsilon) opt$ and thus
-$$2 \leq (2-\epsilon)opt$$
-which implies $opt$ is greater than $1$. Note that our proof is analogous with the general strategy when $a = 1$ and $\alpha = 2-\epsilon$.
-\end{proof}
-\paragraph{}
-We conclude by observing that if $\alpha = 1$ our gap-producing reductions reduce to standard $NP$-completeness reductions.
+An edge weighted graph is called a {\it metric graph} if its weight function induces a metric on the edges.
 \end{document}

### 2017-01-10 CO750-ApproximationAlgorithmsTopics/lectures/lecture_03.tex
Working on scribing lecture 3

@@
 \end{enumerate}
-An edge weighted graph is called a {\it metric graph} if its weight function induces a metric on the edges.
+An edge weighted graph is called a {\it metric graph} if its weight function induces a metric on $V \times V$. A metric graph is necessarily complete in order to make sense.
+\paragraph{}
+Let $G=(V,E)$ be a graph. Let $c : E \rightarrow \R_+$ be a weight function. The {\it metric arising from} $G$ is the function $d$ defined on $V \times V$ where
+$$d(u,v) := \text{ the length of a shortest path from } u \text{ to } v \text{ on } G \text{ according to } c.$$ Given $(G,c)$ we can construct the {\it metric completion} of $G$.
+That is, a complete graph $G^M := (V, E^M)$ with edge cost $d(u,v)  = $ length of shortest $uv$-path in $G$.
+\begin{lemma}
+There exists a Steiner Tree $T$ connecting $R\subseteq V$ of optimal cost $z$ if and only if there exists a Steiner Tree $\bar{T}$ connecting $R$ on $G^M$ of optimal cost $z$.
+Furthermore there is a polynomial time map between $T$ and $\bar{T}$.
+\end{lemma}
+\begin{proof}
+First we observe that the cost of the path between any two vertices $u,v$ in $T$ is at most the cost of the shortest $uv$-path in $G$, as otherwise we could replace the $uv$-path in $T$ with the shortest $uv$-path in $G$, and break cycles arbitrarily to obtain a new tree $T'$ of strictly lower cost than $T$.
+Hence the distance between any two vertices in $T$ is a shortest path distance.
+Thus by definition of $G^M$, both $T$ and $\bar{T}$ have same optimal cost.
+\paragraph{}
+To map from $\bar{T}$ to $T$ replace each edge that is not an edge of $G$ with a shortest path in $G$ (reverse this operation to go from $T$ to $\bar{T}$). If cycles form, break them arbitrarily.
+\end{proof}
+\paragraph{}
+Due to the previous lemma we may assume our input (ST) instances is a metric graph. Consider the following algorithm for (ST) given a metric graph $G$, weights $d$, and terminals $R \subseteq V(G)$.
+\begin{enumerate}
+\item Find a Minimum Spanning Tree $\bar{T}$ on $G[R]$
+\item Return $\bar{T}$.
+\end{enumerate}
+\begin{theorem}
+The above algorithm is a $2$-approximation for (ST).
+\end{theorem}
 \end{document}

### 2017-01-10 CO750-ApproximationAlgorithmsTopics/lectures/lecture_03.tex
Scribing lecture 3

--- 2017-01-06 CO750-ApproximationAlgorithmsTopics/lectures/lecture_03.tex
@@
 \end{enumerate}
-An edge weighted graph is called a {\it metric graph} if its weight function induces a metric on the edges.
+An edge weighted graph is called a {\it metric graph} if its weight function induces a metric on $V \times V$. A metric graph is necessarily complete in order to make sense.
+\paragraph{}
+Let $G=(V,E)$ be a graph. Let $c : E \rightarrow \R_+$ be a weight function. The {\it metric arising from} $G$ is the function $d$ defined on $V \times V$ where
+$$d(u,v) := \text{ the length of a shortest path from } u \text{ to } v \text{ on } G \text{ according to } c.$$ Given $(G,c)$ we can construct the {\it metric completion} of $G$.
+That is, a complete graph $G^M := (V, E^M)$ with edge cost $d(u,v)  = $ length of shortest $uv$-path in $G$.
+\begin{lemma}
+There exists a Steiner Tree $T$ connecting $R\subseteq V$ of optimal cost $z$ if and only if there exists a Steiner Tree $\bar{T}$ connecting $R$ on $G^M$ of optimal cost $z$.
+Furthermore there is a polynomial time map between $T$ and $\bar{T}$.
+\end{lemma}
+\begin{proof}
+First we observe that the cost of the path between any two vertices $u,v$ in $T$ is at most the cost of the shortest $uv$-path in $G$, as otherwise we could replace the $uv$-path in $T$ with the shortest $uv$-path in $G$, and break cycles arbitrarily to obtain a new tree $T'$ of strictly lower cost than $T$.
+Hence the distance between any two vertices in $T$ is a shortest path distance.
+Thus by definition of $G^M$, both $T$ and $\bar{T}$ have same optimal cost.
+\paragraph{}
+To map from $\bar{T}$ to $T$ replace each edge that is not an edge of $G$ with a shortest path in $G$ (reverse this operation to go from $T$ to $\bar{T}$). If cycles form, break them arbitrarily.
+\end{proof}
+\paragraph{}
+Due to the previous lemma we may assume our input (ST) instances is a metric graph. Consider the following algorithm for (ST) given a metric graph $G$, weights $d$, and terminals $R \subseteq V(G)$.
+\begin{enumerate}
+\item Find a Minimum Spanning Tree $\bar{T}$ on $G[R]$
+\item Return $\bar{T}$.
+\end{enumerate}
+\begin{theorem}
+The above algorithm is a $2$-approximation for (ST).
+\end{theorem}
+\begin{proof}
+Let $T^*$ be an optimal solution to a problem instance of (ST).
+Create an Eulerian multigraph from $T^*$ by replacing each edge with two parallel edges.
+The cost of this multigraph is is at most $2d(T^*)$.
+Since this multigraph is Eulerian there exists an Eulerian tour $P$.
+Let $r \in R$.
+Traverse $P$ starting at $r$ and shortcut it to obtain a Hamiltonian cycle $C$ on $G[R]$.
+To be more precise order the vertices of $R$ as $r_1, \dots, r_{|R|}$ in the order they are traversed starting from $r$ along $P$.
+Form cycle $C$ by taking $C = \{r_1r_2, \dots, r_{|R|-1}r_{|R|}\}$.
+By the triangle inequality $d(C) \leq d(P) \leq 2d(T^*)$.
+Further by removing any edge of $C$ we can obtain a spanning tree on $G[R]$, hence $d(\bar{T}) \leq d(C)$.
+Thus we have
+$$d(\bar{T}) \leq d(C) \leq 2d(T^*).$$
+Since computing Minimum Spanning Trees can be done in polynomial time, this completes our result.
+\end{proof}
+\paragraph{}
+To see the analysis is tight consider the graph $K_{n+1}$.
+Label a vertex $v$ and set $d(v,w) = 1$ for all $w \in V\backslash\{v\}$.
+Set $d(w,u) = 2$ for all $u,w \in V\backslash\{v\}$.
+It is not hard to verify this is a metric graph.
+Set $R = V\backslash\{v\}$.
+Then a minimum spanning tree on $G[R]$ has cost $2(n-1)$ while an optimal solution to (ST) has cost $n$.
+Hence the approximation ratio $\alpha$ is
+$$\alpha = \frac{2(n-1)}{2n} = 2 - \frac{1}{n}.$$
+QUESTION: Why is this a ``tight" example?
+\paragraph{}
+For several optimization problems, optimizing on $G$ or $G^M$ is equivalent.
+\paragraph{}
+A {\it tree metric} $(V,d^T)$ is a metric that arises from a weighted tree $T$.
+Many optimization problems are easy on trees.
+For example minimum Steiner Tree is very easy, taking only the nodes needed to connect the terminals.
+So then, given a generic metric $(V,d)$ can we find a tree metric $(V,d^T)$ such that
+$$d(u,v) \leq d^T(u,v) \leq \alpha d(u,v)$$
+for any $u,v \in V$ for some small {\it distortion factor} $\alpha$?
+\paragraph{}
+Some bad news: For a metric arising from an $n$-cycle (simplest imaginable example), any tree-metric must have distortion $\Omega(n)$.
+This is easy to see for spanning tree distance (any removed edge has their distance increase from $1$ to $n-1$ in unit-weighted case), but the observation holds in general.
+So then, what if we take a spanning tree randomly?
+In other words, do the following:
+\begin{enumerate}
+\item Take uniformly at random one edge from the cycle
+\item Output tree $T$ obtained by deleting chosen edge.
+\end{enumerate}
+We will now attempt to bound $E[d^T(u,v)]$. Let $u,v \in V$ and assume $d(u,v) = k$. Then (consider $k$ out of $n$ edges break shortest path, $n-k$ don't)
+$$E[d^T(u,v)] = \frac{k}{n}(n-k) + (1-\frac{k}{n})k \leq 2k$$
+and our expected distortion factor is thus $\alpha = 2$.
+\begin{theorem}
+Given any metric $(V,d)$ one can randomly find a tree $T$ such that
+\begin{enumerate}
+\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in V$
+\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
+\end{enumerate}
+\end{theorem}
+\begin{proof}
+We may assume $\max_{u,v \in V} d(u,v) = 2^\delta$ for some $\delta \in \N$ and $d(u,v) > 1$ for all $u,v\in V$. Let $\cA$ be the algorithm given by
+\begin{enumerate}
+\item Choose randomly a permutation $\pi$ of $V$.
+\item Choose $\beta \in [0,1]$ uniformly at random.
+\item Set $D_\delta = \{V\}$.
+\item For $i = \delta -1$ to $0$ do:
+\begin{enumerate}
+\item Assign every node to the first node according to $\pi$ that has distance at most $2^\beta \cdot 2^{i-1}$.
+\item All nodes assigned to the same node that are currently in the same cluster will form a new cluster in $D_i$.
+\end{enumerate}
+\end{enumerate}
+(TO BE CONTINUED)
+\end{proof}
 \end{document}

### 2017-01-11 CS798-ConvexityAndOptimization/reading/reading_01.tex
Notes from reading in CS798

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-11-01}
\rhead{William Justin Toth CS798-Convexity and Optimization Reading 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\section{Mathematical Background}
This section covers Appendix A  of Convex Optimization by Boyd and Vandenberghe. It is not a complete transcription but notes on particular things that I didn't already feel I understood strongly.
\subsection{Norms}
\paragraph{}
The {\it Cauchy-Schwartz inequality} states that $x^Ty \leq ||x||_2||y||_2$ for any $x,y,\in \R^n$. Such an inequality is useful for proving that the Euclidean norms satisfies the triangle inequality:
$$ ||x + y||_2^2 = (x+y)^T(x+y) = ||x||_2^2 + ||y||_2^2 + 2x^Ty \leq ||x||_2^2 + ||y||_2^2 + 2||x||_2||y||_2 = (||x||_2 + ||y||_2)^2$$
where the inequality follows from Cauchy-Schwartz.
\paragraph{}
The {\it standard inner product} for matrices, $\langle X,Y\rangle$, is defined as
$$\langle X, Y \rangle = trace(X^TY).$$
\paragraph{}
Suppose that $||\cdot ||_a$ and $||\cdot  ||_b$ are norms on $\R^n$. A fundamental result in analysis shows that there exists positive constants $\alpha$ and $\beta$ such that, for all $x \in \R^n$,
$$\alpha ||x||_a \leq ||x||_b \leq \beta ||x||_a.$$
This result implies that norms are {\it equivalent}, in an analytical sense, since they (for instance) define the same set of open subsets, the same set of convergent sequences, etc. An important caveat: this result works on finite dimensional vector spaces only. We can strengthen this result using convex analysis: If $||\cdot ||$ is any norm on $\R^n$, then there exists a quadratic norm $|| \cdot ||_P$ for which
$$||x||_P \leq ||x|| \leq \sqrt{n} ||x||_P,$$
for all $x$.
\subsection{Derivatives}
\paragraph{}
Let $f: \R^n \rightarrow \R^m$ and $x \in \textbf{int dom } f$. The function $f$ is differentiable at $x$ if there exists a matrix $Df(x) \in \R^{m \times n}$ that satisfies
$$ \lim_{z \in \textbf{dom} f, z\neq x, z\rightarrow x} \frac{||f(z) - f(x) - Df(x)(z-x)||_2}{||z-x||_2} = 0.$$
In such a case we refer to $Df(x)$ as the {\it derivative} (aka {\it Jacobian}) of $f$ at $x$. There can be at most one such matrix. The function $f$ is {\it differentiable } if $\textbf{dom} f$ is open, and it is differentiable at every point in its domain. The affine function of $z$ given by
$$f(x) + Df(x)(z-x)$$
is called the {\it first-order approximation} of $f$ at (or near) $x$.  The derivative can be found by deriving the first-order approximation of the function $f$ at $x$ or from partial derivatives as:
$$Df(x)_{ij} = \frac{\partial f_i(x)}{\partial x_j}, \quad i=1,\dots,m,\ \ \ j=1,\dots,n.$$
\paragraph{}
When $f$ is real-valued, the derivative $Df(x)$ is a$1\times n$ matrix. Its transpose in $\R^n$ is called the {\it gradient} of the function:
$$\nabla f(x) = Df(x)^T.$$
Its components are the partial derivatives of $f$.
\paragraph{}
Suppose $f: \R^n \rightarrow \R^m$ is differentiable at $x \in \textbf{int dom} f$ and $g: \R^m \rightarrow \R^p$ is differentiable at $f(x) \in \textbf{int dom}g$. Let $h(z) = g(f(z))$ define the composition $h: \R^n \rightarrow \R^p$. Then $h$ is differentiable at $x$, with derivative
$$Dh(x) = Dg(f(x))(Df(x)).$$
\paragraph{}
Let $f: \R^n \rightarrow \R$ be a real-valued function. The second derivative or {\it Hessian matrix} of $f$ at $x$ denoted $\nabla^2 f(x) \in \R^{n \times n}$ is given by,
$$\nabla^2f(x)_{ij} = \frac{\partial^2f(x)}{\partial x_i \partial x_j},$$
provided $f$ is twice differentiable at $x$. The {\it second-order approximation} of $f$, at or near $x$, is the quadratic function of $z$ defined by
$$\hat{f}(z) = f(x) + \nabla f(x)^T(z-x) + \frac{1}{2}(z-x)^T\nabla^2 f(x)(z-x).$$
This can be interpreted as the derivative of the gradient mapping.
\section{Operations that Preserve Convexity}
\paragraph{}
The following is a list of operations which preserve convexity:
\begin{enumerate}
\item The intersection of convex sets is a convex set.
\item The image (and inverse image) of a convex set under an affine map is a convex set. For instance, scaling, translations, and projection.
\item The image of a convex set under a {\it linear fractional } map (to be defined below) is a convex set.
\end{enumerate}
\subsection{Linear Fractional and Perspective Functions}
\paragraph{}
The {\it perspective function} $P: \R^{n+1} \rightarrow \R^n$ with domain $\textbf{dom} P = \R^n \times \{x\in\R : x > 0 \}$, is defined as
$$P(z,t) = z/t.$$
The perspective functions scales vectors so the last component is one then drops the last component. If $C \subseteq \textbf{dom} P$ is convex then $P(C)$ is convex. We will show that $P$ maps line segments to line segments, and the result will follow. Let $x = (\tilde{x}, x_{n+1}), y = (\tilde{y}, y_{n+1}) \in \R^{n+1}$ with $x_{n+1} > 0$. Then for $0 \leq \theta \leq 1$,
$$P(\theta x + (1-\theta)y) = \frac{\theta \tilde{x} + (1-\theta)\tilde{y}}{\theta x_{n+1} + (1-\theta)y_{n+1}} = \mu P(x) + (1-\mu)P(y),$$
where
$$\mu = \frac{\theta x_{n+1}}{\theta x_{n+1} + (1-\theta)y_{n+1}} \in [0,1].$$
\paragraph{}
A {\it linear-fractional function} is formed by composing the perspective function with an affine function. Suppose $g: \R^n \rightarrow \R^{m+1}$ is affine, i.e.,
$$g(x) = \begin{bmatrix}
A \\ c^T
\end{bmatrix} x + \begin{bmatrix}
b \\ d \end{bmatrix},$$
the function $f : \R^n \rightarrow \R^m$ given by $f = P \circ g$, i.e.,
$$fg(x) = \frac{Ax + b}{c^Tx + d}$$
is called a {\it linear-fractional} (or projective) function. If $c = 0$ and $d>0$ then $f$ is an affine function. The fact that linear-fractional functions preserve convexity follows since affine functions and perspective functions preserve convexity.
\end{document}

### 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_04.tex
partial lecture 5 scribe

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-01-17}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 4} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
Recall from Lecture $3$:
\begin{theorem}
Given any metric $(V,d)$ one can randomly find a tree $T$ such that
\begin{enumerate}
\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in V$
\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
\end{enumerate}
\end{theorem}
\begin{proof}
We may assume $\max_{u,v \in V} d(u,v) = 2^\delta$ for some $\delta \in \N$ and $d(u,v) > 1$ for all $u,v\in V$. Construct a tree $T$ by:
\begin{enumerate}
\item Choose randomly a permutation $\pi$ of $V$.
\item Choose $\beta \in [0,1]$ uniformly at random.
\item Set $D_\delta = \{V\}$.
\item For $i = \delta -1$ to $0$ do:
\begin{enumerate}
\item Assign every node to the first node according to $\pi$ that has distance at most $2^\beta \cdot 2^{i-1}$.
\item All nodes assigned to the same node that are currently in the same cluster will form a new cluster in $D_i$.
\end{enumerate}
\item Return tree $T$ where $V(T) = \{s \in D_i : \forall i = 0, \dots, \delta\}$ and $E(T) = \{s_is_{i-1}  \in D_{i}\times D_{i-1}: s_{i-1} \subseteq s_i, \forall i = 0,\dots,\delta\}$, and $d(s_is_{i-1}) = 2^i$.
\end{enumerate}
Note that the clusters of $D_0$ are singletons because the bound on the maximum distance is $2^\beta 2^0 \leq 1$. Hence the clusters of $D_0$ can be identified with nodes in $V$. 
\paragraph{Proof of 1.}
Suppose $u,v$ are in the same cluster at level $D_i$ and are separated at level $D_{i-1}$. Let $w$ be the node identified with $u$ and $v$'s cluster $S_w$ at level $D_i$. Then $d(u,v) \leq d(u,w) + d(w,v) \leq 2^\beta 2^{i-1} + 2^\beta 2^{i-1} \leq 2^{i+1}$. On the other hand $d^T(u, S_w) \geq 2^i$ and $d^T(S_w, v) \geq 2^i$ so $d^T(u,v) \geq 2^i + 2^i = 2^{i+1}$. This proves part $1.$ of the theorem.
\paragraph{Proof of 2.}
Observe that if $u,v$ are separated first at level $D_i$ then $d^T(u,v) \leq 2^{i+2}$. Let $S_i$ be the cluster in $D_i$ containing $u,v$. Let $S^u_{i-1}$ be descended cluster of $S_i$ from which $u$ descends and similarly define $S^v_{i-1}$. Then 
\begin{align*}
d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_, S^u_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
&= 2^i + 2^i +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
&\leq 2^i + 2^i + 2\sum_{j = 1}^{i-1} 2^j \\
&= 2^{i+1} + 2^{i+1} \\
&= 2^{i+2}.
\end{align*}
Now we way that node $c$ {\it settles} $(u,v)$ at level $D_i$ if $c$ is the first node with respect to $\pi$ to which one of $u$ or $v$ gets assigned at level $D_i$. We say that node $c$ {\it cuts} $(u,v)$ at level $D_i$ if exactly one of $u,v$ is at distance at most $2^\beta 2^i$ from $c$. If $u,v$ are separated at level $D_i$ then there exists some $c$ which both settles and cuts $u,v$ at level $D_i$.
\paragraph{}
Note that $$E[d^T(u,v)] \leq \sum_{i=0}^{\delta - 1} \sum_{c \in V} Pr [ \text{$c$ settles (A) and cuts (B) $u,v$ at $D_i$}] 2^{i+2}.$$ The bound simply follows from our observation, and the prior definitions. Recall that $Pr[A \cap B] = P[A\mid B] Pr[B]$.
\paragraph{}
We first analyze $Pr[B]$. Without loss we may assume $d(c,u) \leq d(c,v)$. For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition. Thus
$$Pr[B] = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{|[2^{i-1},2^i]|} = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}}$$.
Now we need $Pr[A \mid B]$. Let us order the nodes of $V$ according to their distance to $u,v$. That is, $V = \{c_1, \dots, c_n\}$ where $\min\{d(c_i, v), d(c_i, u)\} \leq \min \{d(c_{i+1}, v), d(c_{i+1}, u)\}$ for all $i$. If $c= c_k$ then the probability that $c$ comes before $c_1, \dots, c_{k-1}$ in $\pi$ is at most $\frac{1}{k}$. In order for $c$ (given that $c$ cuts $u,v$) to settle $u,v$ at least that has to happen. Hence
$$Pr[A \mid B] \leq \frac{1}{k}.$$
\paragraph{}
Putting it all together we obtain the following expectation:
\begin{align*}
E[d^T(u,v)] &\leq \sum_{i=0}^{\delta -1} \sum_{k=1}^n Pr [A \mid B]Pr[B] 2^{i+2} \\
&\leq \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  2^{i+2}\frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}} \\
&= 8 \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  | [2^{i-1},2^i] \cap [d(c,u),d(c,v)]| \\
&\leq 8 d(u,v) \sum_{k=1}^n \frac{1}{k} \\
&= 8 H_n d(u,v) \\
&= O(log n) d(u,v).
\end{align*}
Thus the second claim and therefore the theorem holds.
\end{proof}
\begin{theorem}
There exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
\end{theorem}
\begin{proof}
Pick $v \in V$ with a parent $w \not \in V$ and contract edge $vw$ in $T$. Repeat until all nodes of $T$ are in $V$. Multiply by $4$ the weights on all remaining edges. Call the resulting tree $T'$. Note that clearly $d^{T'}(u,v) \leq 4d^T(u,v)$ as contracting only shortens the distances. Furthermore, in the worst case $d^{T'}(u,v) = 2^i \cdot 4$ if contraction results in contracting $u$ and $v$ right up to the cluster where they were separated (that is, contracting preserves the largest edge distance on the $uv$-path in $T$). But 
$$2^i \cdot 4 = 2^i + 2^i + 2^i + 2^i \geq d(S^u_{i-1}, S_i) + d(S^v_{i-1}, S_i) +  d(S^u_{i-1}, u) + d(S^v_{i-1}, v) = d^T(u,v).$$
Hence the theorem holds.
\end{proof}
\section{Generalized Steiner Network}
\paragraph{}
Here we seek to given an application of the previous theory. In the Generalized Steiner Network problem (GSN), we are given a graph $G=(V,E)$, a distance function $d: E \rightarrow \R_+$, and requirements $r(u,v) \in \N$ for all $u,v \in V$. We are to find an integer capacity function $x : E \rightarrow \N$ of minimum cost $\sum_{e \in E} d(e) x_e$ such that capacities $x$ allow for routing a flow of value $r(u,v)$ from $u$ to $v$.
\paragraph{Proposition 1} Optimizing on $G$ is equivalent to optimizing on $G^M$ for $GSN$ problems. This proposition is immediate from the same arguments as was used for ST.
\paragraph{Proposition 2} GSN is solvable in polynomial time on trees. Since each edge is a cut in the tree, we must set $x_{uv} = r(u,v)$ and this is the best we can do.
\paragraph{Algorithm} Consider the following simple algorithm for GSN:
\begin{enumerate}
\item Find a tree $T$ of $G$ according to previous section's metric embedding algorithms.
\item Solve GSN on $T$ and return the solution.
\end{enumerate}
\begin{theorem}
Algorithm above is a randomized $O(log n)$-approximation algorithm for GSN.
\end{theorem}
\begin{proof}
Let $\bar{x}$ be the solution return by our algorithm. Let $x^*$ be OPT solution. Then $\sum_{uv} d(u,v) \bar{x}_{uv} \leq \sum_{uv} d^T(u,v)\bar{x}_{uv} \leq \sum_{uv} d^T(u,v) x^*_{uv}$. The first inequality follows since $d(u,v) \leq d^T(u,v)$. The second inequality follows since $x^*$ is a feasible solution for $T$. Taking expectation we see:
$$E[\sum_{uv} d(u,v) \bar{x}_{uv}] \leq E[\sum_{uv} d^T(u,v)] \leq O(log n) \sum_{uv} d(u,v) x^*_{uv}$$
with the last inequality following from our theorem.
\end{proof}
\end{document}

### 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
partial lecture 5 scribe

--- 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_04.tex
@@
 
-\lhead{2017-01-17}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 4} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-01-19}
+\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 5} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-Recall from Lecture $3$:
+\paragraph{}
+Before moving on to the next topic, we just want to mention an important theorem that results from the theory built in the previous lectures.
+It tells us that we can have deterministic approximation algorithms that give just as good bounds as the randomized ones studied previously.
 \begin{theorem}
-Given any metric $(V,d)$ one can randomly find a tree $T$ such that
-\begin{enumerate}
-\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in V$
-\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
-\end{enumerate}
+Let $G=(V,E)$ be a graph.
+Let $d: E \rightarrow \R_+$ be a metric function.
+Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$.
+There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
 \end{theorem}
 \begin{proof}
-We may assume $\max_{u,v \in V} d(u,v) = 2^\delta$ for some $\delta \in \N$ and $d(u,v) > 1$ for all $u,v\in V$. Construct a tree $T$ by:
+Omitted. May be in a talk at the end of course via some students project. Involves derandomization via the method of conditional expectation.
+\end{proof}
+\section{Primal - Dual Methods}
+\paragraph{}
+Primal-Dual methods are a common paradigm in algorithm design.
+We will begin by showing how they are used to given exact algorithms, and then explain relaxed complementary slackness which allows us to extend the paradigm to give approximation algorithms.
+\paragraph{}
+Consider the {\it Arborescence Problem} (ARB).
+In this problem we are given a directed graph $D= (V,E)$, cost function $c: E \rightarrow \R_+$, and a root vertex $r \in V$.
+We are to find an arborescence $T$ rooted at $r$ which spans $V$ and minimizes $\sum_{e\in E(T)} c_e$.
+\paragraph{}
+As the name of this method might suggest we will used integrally the linear programming relaxation of this problem:
+\begin{align*}
+\min \sum_{e \in E} c_e x_e  &\ \\
+\text{s.t.} \sum_{e \in \delta^+(S)} x_e &\geq 1 &\forall S \subseteq V\backslash\{r\} \\
+x_e &\geq 0 &\forall e \in E,
+\end{align*}
+and its dual program:
+\begin{align*}
+\max \sum_{S \subseteq V\backslash\{r\}} y_S &\ \\
+\text{s.t.} \sum_{S : e \in \delta^+(S)} y_S &\leq c_e &\forall e \in E \\
+y_S &\geq 0 &\forall S \subseteq V\backslash \{r\}.
+\end{align*}
+Here $\delta^+(S) = \{(u,v) \in E : u \in S, v\not\in S\}$ denotes set of arcs outgoing from $S$.
+\paragraph{}
+Our primal-dual algorithm for (ARB) operates as follows (start by setting $x=0$, $y=0$, for implementation track only non-negative $y$ to avoid exponential runtime): 
 \begin{enumerate}
-\item Choose randomly a permutation $\pi$ of $V$.
-\item Choose $\beta \in [0,1]$ uniformly at random.
-\item Set $D_\delta = \{V\}$.
-\item For $i = \delta -1$ to $0$ do:
+\item While $x$ is not feasible for the primal:
 	\begin{enumerate}
-\item Assign every node to the first node according to $\pi$ that has distance at most $2^\beta \cdot 2^{i-1}$.
-\item All nodes assigned to the same node that are currently in the same cluster will form a new cluster in $D_i$.
+	\item Increase $y_S$ uniformly for all ``strongly connected components" (with respect to $x$) $S$ where $x(\delta^+(S)) = 0$ until some constraints for an edge $e$ becomes tight.
+	\item Set $x_e=1$ for tight edge $e$ in previous step.
 	\end{enumerate}
-\item Return tree $T$ where $V(T) = \{s \in D_i : \forall i = 0, \dots, \delta\}$ and $E(T) = \{s_is_{i-1}  \in D_{i}\times D_{i-1}: s_{i-1} \subseteq s_i, \forall i = 0,\dots,\delta\}$, and $d(s_is_{i-1}) = 2^i$.
+\item Deletion Step: Consider each edge $e$ where $x_e = 1$ in reverse order from the order they were set to $1$ and set $x_e = 0$ as long as this does not cause the resulting $x$ to become infeasible.
+\item Return $x$.
 \end{enumerate}
-Note that the clusters of $D_0$ are singletons because the bound on the maximum distance is $2^\beta 2^0 \leq 1$. Hence the clusters of $D_0$ can be identified with nodes in $V$. 
-\paragraph{Proof of 1.}
-Suppose $u,v$ are in the same cluster at level $D_i$ and are separated at level $D_{i-1}$.
-Let $w$ be the node identified with $u$ and $v$'s cluster $S_w$ at level $D_i$.
-Then $d(u,v) \leq d(u,w) + d(w,v) \leq 2^\beta 2^{i-1} + 2^\beta 2^{i-1} \leq 2^{i+1}$.
-On the other hand $d^T(u, S_w) \geq 2^i$ and $d^T(S_w, v) \geq 2^i$ so $d^T(u,v) \geq 2^i + 2^i = 2^{i+1}$.
-This proves part $1.$ of the theorem.
-\paragraph{Proof of 2.}
-Observe that if $u,v$ are separated first at level $D_i$ then $d^T(u,v) \leq 2^{i+2}$.
-Let $S_i$ be the cluster in $D_i$ containing $u,v$.
-Let $S^u_{i-1}$ be descended cluster of $S_i$ from which $u$ descends and similarly define $S^v_{i-1}$.
-Then
-\begin{align*}
-d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_, S^u_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
-&= 2^i + 2^i +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
-&\leq 2^i + 2^i + 2\sum_{j = 1}^{i-1} 2^j \\
-&= 2^{i+1} + 2^{i+1} \\
-&= 2^{i+2}.
-\end{align*}
-Now we way that node $c$ {\it settles} $(u,v)$ at level $D_i$ if $c$ is the first node with respect to $\pi$ to which one of $u$ or $v$ gets assigned at level $D_i$.
-We say that node $c$ {\it cuts} $(u,v)$ at level $D_i$ if exactly one of $u,v$ is at distance at most $2^\beta 2^i$ from $c$.
-If $u,v$ are separated at level $D_i$ then there exists some $c$ which both settles and cuts $u,v$ at level $D_i$.
-\paragraph{}
-Note that $$E[d^T(u,v)] \leq \sum_{i=0}^{\delta - 1} \sum_{c \in V} Pr [ \text{$c$ settles (A) and cuts (B) $u,v$ at $D_i$}] 2^{i+2}.$$ The bound simply follows from our observation, and the prior definitions.
-Recall that $Pr[A \cap B] = P[A\mid B] Pr[B]$.
-\paragraph{}
-We first analyze $Pr[B]$. Without loss we may assume $d(c,u) \leq d(c,v)$. For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition. Thus
-$$Pr[B] = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{|[2^{i-1},2^i]|} = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}}$$.
-Now we need $Pr[A \mid B]$.
-Let us order the nodes of $V$ according to their distance to $u,v$.
-That is, $V = \{c_1, \dots, c_n\}$ where $\min\{d(c_i, v), d(c_i, u)\} \leq \min \{d(c_{i+1}, v), d(c_{i+1}, u)\}$ for all $i$.
-If $c= c_k$ then the probability that $c$ comes before $c_1, \dots, c_{k-1}$ in $\pi$ is at most $\frac{1}{k}$.
-In order for $c$ (given that $c$ cuts $u,v$) to settle $u,v$ at least that has to happen.
-Hence
-$$Pr[A \mid B] \leq \frac{1}{k}.$$
-\paragraph{}
-Putting it all together we obtain the following expectation:
-\begin{align*}
-E[d^T(u,v)] &\leq \sum_{i=0}^{\delta -1} \sum_{k=1}^n Pr [A \mid B]Pr[B] 2^{i+2} \\
-&\leq \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  2^{i+2}\frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}} \\
-&= 8 \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  | [2^{i-1},2^i] \cap [d(c,u),d(c,v)]| \\
-&\leq 8 d(u,v) \sum_{k=1}^n \frac{1}{k} \\
-&= 8 H_n d(u,v) \\
-&= O(log n) d(u,v).
-\end{align*}
-Thus the second claim and therefore the theorem holds.
-\end{proof}
 \begin{theorem}
-There exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
+The previous algorithm is an exact polynomial time algorithm for (ARB).
 \end{theorem}
 \begin{proof}
-Pick $v \in V$ with a parent $w \not \in V$ and contract edge $vw$ in $T$.
-Repeat until all nodes of $T$ are in $V$.
-Multiply by $4$ the weights on all remaining edges.
-Call the resulting tree $T'$.
-Note that clearly $d^{T'}(u,v) \leq 4d^T(u,v)$ as contracting only shortens the distances.
-Furthermore, in the worst case $d^{T'}(u,v) = 2^i \cdot 4$ if contraction results in contracting $u$ and $v$ right up to the cluster where they were separated (that is, contracting preserves the largest edge distance on the $uv$-path in $T$).
-But
-$$2^i \cdot 4 = 2^i + 2^i + 2^i + 2^i \geq d(S^u_{i-1}, S_i) + d(S^v_{i-1}, S_i) +  d(S^u_{i-1}, u) + d(S^v_{i-1}, v) = d^T(u,v).$$
-Hence the theorem holds.
-\end{proof}
-\section{Generalized Steiner Network}
+When the algorithm terminates we have that if $x_e > 0$ then $x_e = 1$  and so the dual constraint is tight, that is:
+$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$
+This is immediate from our construction of $x$. Hence the primal complementary slackness conditions hold.
 \paragraph{}
-Here we seek to given an application of the previous theory.
-In the Generalized Steiner Network problem (GSN), we are given a graph $G=(V,E)$, a distance function $d: E \rightarrow \R_+$, and requirements $r(u,v) \in \N$ for all $u,v \in V$.
-We are to find an integer capacity function $x : E \rightarrow \N$ of minimum cost $\sum_{e \in E} d(e) x_e$ such that capacities $x$ allow for routing a flow of value $r(u,v)$ from $u$ to $v$.
-\paragraph{Proposition 1} Optimizing on $G$ is equivalent to optimizing on $G^M$ for $GSN$ problems. This proposition is immediate from the same arguments as was used for ST.
-\paragraph{Proposition 2} GSN is solvable in polynomial time on trees. Since each edge is a cut in the tree, we must set $x_{uv} = r(u,v)$ and this is the best we can do.
-\paragraph{Algorithm} Consider the following simple algorithm for GSN:
-\begin{enumerate}
-\item Find a tree $T$ of $G$ according to previous section's metric embedding algorithms.
-\item Solve GSN on $T$ and return the solution.
-\end{enumerate}
-\begin{theorem}
-Algorithm above is a randomized $O(log n)$-approximation algorithm for GSN.
-\end{theorem}
-\begin{proof}
-Let $\bar{x}$ be the solution return by our algorithm.
-Let $x^*$ be OPT solution.
-Then $\sum_{uv} d(u,v) \bar{x}_{uv} \leq \sum_{uv} d^T(u,v)\bar{x}_{uv} \leq \sum_{uv} d^T(u,v) x^*_{uv}$.
-The first inequality follows since $d(u,v) \leq d^T(u,v)$.
-The second inequality follows since $x^*$ is a feasible solution for $T$.
-Taking expectation we see:
-$$E[\sum_{uv} d(u,v) \bar{x}_{uv}] \leq E[\sum_{uv} d^T(u,v)] \leq O(log n) \sum_{uv} d(u,v) x^*_{uv}$$
-with the last inequality following from our theorem.
+For the dual complementary slackness conditions, suppose there exists $S$ such that $y_S > 0$, yet $\sum_{e\in \delta^+(S)} x_e \geq 2$.
+Let $e, \bar{e} \in \delta^+(S)$ and suppose without loss of generality that $\bar{e}$ was added before $e$ by the algorithm.
+Since we did not delete $e$, there exists a node $v \in S$ that uses $e$ to reach $r$.
+But since $S$ is strongly connected, no nodes in $S$ need $\bar{e}$ to reach $r$ (they can go $v$ then eventually through $e$), so we could have deleted $\bar{e}$ in the Deletion Step.
+A contradiction.
 \end{proof}

### 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
Partial scribe of lecture 5

@@
 When the algorithm terminates we have that if $x_e > 0$ then $x_e = 1$  and so the dual constraint is tight, that is:
-$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$
+$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$$
 This is immediate from our construction of $x$. Hence the primal complementary slackness conditions hold.
@@
 \end{proof}
+\paragraph{}
+To extend this method to one which yields approximation algorithms, we need to be able to relax our complementary slackness conditions.
+In (ARB) our relaxed complementary slackness conditions are:  $x$ (with corresponding dual solution $y$) is an $\alpha \beta$-approximate solution if and only if
+\begin{enumerate}
+\item If $x_e > 0$ then $\sum_{S: e \in \delta^+(S)} y_S \geq \frac{1}{\alpha} c_e$.
+\item If $y_S > 0$ then $\sum_{e\in \delta^+(S)} x_e \leq \beta \cdot 1$.
+\end{enumerate}
+To see it observe that
+$$\sum_{e} c_e x_e \leq \sum_e x_e(\alpha \sum_{S: e \in \delta^+(S)} y_S) = \alpha \sum_{e}x_e(\sum_{S: e \in \delta^+(S)} y_S) \leq \alpha \beta \sum_S y_S.$$
+\subsection{Primal-Dual Algorithm for General Network Design Problems}
+\paragraph{}
+In a {\it general network design problem} (GND) we are given a graph $G = (V,E)$, a cost function $c : E \rightarrow \R_+$, and a function $f: 2^S \rightarrow \R_+$.
+We are to solve the integer program:
+\begin{align*}
+min \sum_{e \in E} c_e x_e  &\ \\
+\text{s.t.} \sum_{e \in \delta(S)} x_e &\geq f(S) &\forall S \subseteq V\\
+x_e &\geq 0 &\forall e \in E \\
+x \in \Z^E.
+\end{align*}
+This problem's linear programming relaxation is:
+\begin{align*}
+min \sum_{e \in E} c_e x_e  &\ \\
+\text{s.t.} \sum_{e \in \delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+x_e &\geq 0 &\forall e \in E 
+\end{align*}
+with corresponding dual program:
+\begin{align*}
+\max \sum_{S} f(S)y_S &\ \\
+\text{s.t.} \sum_{S : e \in \delta(S)} y_S &\leq c_e &\forall e \in E \\
+y_S &\geq 0 &\forall S \subseteq V.
+\end{align*}
+We will focus our attention on the special case where $f$ is a $01$-proper function. A function $f$ is called {\it proper} if it satisfies:
+\begin{enumerate}
+\item $f(V) = 0$
+\item $f(S) = f(V\backslash S)$, $\forall S\subseteq V$ (Symmetry property)
+\item If $A \cap B = \emptyset$ then $f(A \cup B) \leq \max \{f(A), f(B)\}$ (Maximality property).
+\end{enumerate}
+Our algorithm for (GSN) where $f$ is $01$-proper is as follows:
+\begin{enumerate}
+\item Start with ($x=0,y=0$) and $F = \emptyset$.
+\item While $F$ is not feasible:
+	\begin{enumerate}
+	\item Increase $y_S$ uniformly for "minimal violated cuts" (with respect to $F$) until a constraint corresponding to some edge $e$ becomes tight in the dual program.
+	\item For such $e$ set $x_e = 1$ and $F = F \cup \{e\}$.
+	\end{enumerate}
+\item Deletion Step: Consider all edges in $F$ in reverse order they were added in Step $2$. If $F\backslash\{e\}$ is feasible then set $F = F\backslash\{e\}$ and set $x_e = 0$.
+\item Return $F$.
+\end{enumerate}
+\begin{lemma}
+Let $f$ be a $01$ function satisfying the Maximality property. Then
+\begin{enumerate}
+\item $F \subseteq E$ is feasible if and only if every connected component $C$ of $(V,F)$ satisfies $f(C) = 0$.
+\item The minimal violated sets (cuts) are connected components $C$ that have $f(C) = 1$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+Let $S$ be a minimal violated set with $f(S) = 1$.
+If $(V,S)$ contains some connected component $A$ and $S\backslash A := B$ is not empty then by Maximality either $f(A) = 1$ or $f(B) = 1$.
+But then either $A$ or $B$ contradicts the minimality of $S$.
+Hence $S$ is a connected component and the lemma follows.
+\end{proof}
+\begin{theorem}
+Our algorithm above is a $2$-approximation algorithm for (GND) when $f$ is a $01$-proper function.
+\end{theorem}
+\begin{proof}
+From the lemma it should not be hard to see that our algorithm runs in polynomial time.
+\end{proof}
 \end{document}

### 2017-01-23 optimization-w2017/smpolytope-2017-01-25.tex
Drafting of a helper document to give to attendees of Wednesay's talk

--- 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
@@
 
-\lhead{2017-01-19}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 5} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-01-25}
+\rhead{W. Justin Toth: Optimization Seminar Helper Document} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+\section*{Notation}
 \paragraph{}
-Before moving on to the next topic, we just want to mention an important theorem that results from the theory built in the previous lectures.
-It tells us that we can have deterministic approximation algorithms that give just as good bounds as the randomized ones studied previously.
-\begin{theorem}
-Let $G=(V,E)$ be a graph.
-Let $d: E \rightarrow \R_+$ be a metric function.
-Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$.
-There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
-\end{theorem}
-\begin{proof}
-Omitted. May be in a talk at the end of course via some students project. Involves derandomization via the method of conditional expectation.
-\end{proof}
-\section{Primal - Dual Methods}
+$G=(A\cup B, E)$ will denote a bipartite graph with vertex partitions $A$ and $B$. The vertex set of $G$ is denoted $V(G)$, and the edge set $E(G)$.
 \paragraph{}
-Primal-Dual methods are a common paradigm in algorithm design.
-We will begin by showing how they are used to given exact algorithms, and then explain relaxed complementary slackness which allows us to extend the paradigm to give approximation algorithms.
+$\delta(v) = \{e \in E(G) : v \in e\}$ for all $v \in V(G)$.
 \paragraph{}
-Consider the {\it Arborescence Problem} (ARB).
-In this problem we are given a directed graph $D= (V,E)$, cost function $c: E \rightarrow \R_+$, and a root vertex $r \in V$.
-We are to find an arborescence $T$ rooted at $r$ which spans $V$ and minimizes $\sum_{e\in E(T)} c_e$.
+$N(v) = \{w \in V(G): vw \in E(G)\}$ for all $V\in V(G)$.
 \paragraph{}
-As the name of this method might suggest we will used integrally the linear programming relaxation of this problem:
+$>_v$ denotes the strict total preference order of $v$ over $N(v)$. By convention $v$ prefers being matched to unmatched.
+\paragraph{}
+$\delta^{>w}(v) = \{vu \in \delta(v) : u >_v w\}$ for all $v,w \in V(G)$.
+\paragraph{}
+$N_{max}(v) = w$ if and only if $w \geq_v u$ for all $u \in N(v)$. Similarly $N_{min}(v) = w$ if and only if $w \leq_v u$ for all $u \in N(v)$.
+\paragraph{}
+Commonly $M \subseteq E$ such that $|\delta(v) \cap M| \leq 1$ for all $v \in V$ denotes a matching, and $M(v) = w $ if and only if $vw \in M$.
+For matchings $M_1$ and $M_2$, $M_1 \Delta M_2 = (M_1 \cup M_2) \backslash (M_1 \cap M_2)$ denotes their symmetric difference
+\paragraph{}
+$P(G)$ denotes the set of $x \in \R^{E(G)}$ satisfying
 \begin{align*}
-\min \sum_{e \in E} c_e x_e  &\ \\
-\text{s.t.} \sum_{e \in \delta^+(S)} x_e &\geq 1 &\forall S \subseteq V\backslash\{r\} \\
-x_e &\geq 0 &\forall e \in E,
+ x(\delta(v)) &\leq 1 &\forall v \in V(G) \numberthis \\
+x(\delta^{>b}(a)) + x(\delta^{>a}(b)) + x_{ab} &\geq 1 &\forall ab \in E(G) \numberthis \\
+x &\geq 0\numberthis.
 \end{align*}
-and its dual program:
-\begin{align*}
-\max \sum_{S \subseteq V\backslash\{r\}} y_S &\ \\
-\text{s.t.} \sum_{S : e \in \delta^+(S)} y_S &\leq c_e &\forall e \in E \\
-y_S &\geq 0 &\forall S \subseteq V\backslash \{r\}.
-\end{align*}
-Here $\delta^+(S) = \{(u,v) \in E : u \in S, v\not\in S\}$ denotes set of arcs outgoing from $S$.
+Where $x(M) = \sum_{e \in M} x_e$ for $M \subseteq E(G)$ and $x \in \R^{E(G)}$.
 \paragraph{}
-Our primal-dual algorithm for (ARB) operates as follows (start by setting $x=0$, $y=0$, for implementation track only non-negative $y$ to avoid exponential runtime): 
+We let $\chi(M) \in \{0,1\}^{E(G)}$ be the characteristic vector of $M$, defined by $\chi(M)_e = 1$ if $e\in M$ and $\chi(M)_e = 0$ otherwise.
+\paragraph{}
+For $x \in P(G)$ we let $V_x$ denote the set of vertices corresponding to constraints of the form $(1)$ satisfied at equality by $x$.
+Similarly $E_x$ denotes the set of constraints of the form $(2)$ satisfied at equality by $x$.
+\section*{Outline (of proof $P(G)$ is integral)}
 \begin{enumerate}
-\item While $x$ is not feasible for the primal:
-	\begin{enumerate}
-	\item Increase $y_S$ uniformly for all ``strongly connected components" (with respect to $x$) $S$ where $x(\delta^+(S)) = 0$ until some constraints for an edge $e$ becomes tight.
-	\item Set $x_e=1$ for tight edge $e$ in previous step.
+\item Show there exists $e \in E$ for which $x_e \in \{0,1\}$ where $x$ is an extreme point of $P(G)$.
+\item Proceed by induction on $|E(G)|$
+\item Case $x_e = 0$ using some Stable Matching basics and geometry of polytopes.
+\item Case $x_e = 1$ reducing to case $x_e = 0$ or simply induction.
 \end{enumerate}
-\item Deletion Step: Consider each edge $e$ where $x_e = 1$ in reverse order from the order they were set to $1$ and set $x_e = 0$ as long as this does not cause the resulting $x$ to become infeasible.
-\item Return $x$.
-\end{enumerate}
-\begin{theorem}
-The previous algorithm is an exact polynomial time algorithm for (ARB).
-\end{theorem}
-\begin{proof}
-When the algorithm terminates we have that if $x_e > 0$ then $x_e = 1$  and so the dual constraint is tight, that is:
-$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$
-This is immediate from our construction of $x$. Hence the primal complementary slackness conditions hold.
-\paragraph{}
-For the dual complementary slackness conditions, suppose there exists $S$ such that $y_S > 0$, yet $\sum_{e\in \delta^+(S)} x_e \geq 2$.
-Let $e, \bar{e} \in \delta^+(S)$ and suppose without loss of generality that $\bar{e}$ was added before $e$ by the algorithm.
-Since we did not delete $e$, there exists a node $v \in S$ that uses $e$ to reach $r$.
-But since $S$ is strongly connected, no nodes in $S$ need $\bar{e}$ to reach $r$ (they can go $v$ then eventually through $e$), so we could have deleted $\bar{e}$ in the Deletion Step.
-A contradiction.
-\end{proof}
 \end{document}

### 2017-01-24 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
lecture 5 scribe
Scribe lecture 6

@@
 \begin{proof}
-From the lemma it should not be hard to see that our algorithm runs in polynomial time.
+Polynomial run time follows from our lemma, because it allows us to find minimal violated cuts in polynomial time by checking the value of $f$ on connected components.
+\paragraph{}
+For the relaxed complementary slackness conditions observe that if $e \in F$ then $\sum_{S : e \in \delta(S)} y_S = c_e$ by construction.
+It remains to show that for $y_S > 0$ we have $\sum_{e\in\delta(S)}  x_e = |\delta(S) \cap F| \leq 2$.
+We will show that this bound holds on average.
+That is that
+$$ \sum_{e \in F} c_e = \sum_{e \in F} (\sum_{S : e \in \delta(S)} y_S) = (a) = \sum_S |\delta_F(S)| y_S \leq 2 \sum_S y_S = (b) $$
+Let $\Delta_i$ be the amount by which all minimum violated sets get increased in iteration $i$. Call such sets ACTIVE in iteration $i$ (denoted $A(i)$). Then
+$$(a) = \sum_i \sum_{S \in A(i)} |\delta_F(S)| \Delta_i \quad (b) = \sum_i \sum_{S\in A(i)} \Delta_i.$$
+We will show that for every iteration $i$, $\sum_i |\delta_F(S)| \leq 2 \cdot |A(i)|.$
+\paragraph{}
+So fix some $i$. Consider $(V,F)$ (which is a forest as cycles can be removed in Deletion Step), and contract all sets $S$ in $A(i)$ into one node. Call the resulting graph $H$.
+\paragraph{Claim 1}
+$H$ is a forest.
+\paragraph{Proof of Claim 1}
+Suppose the contraction of one set $S$ creates a cycle.
+But then in the reverse deletion step, one the two edges in $\delta_F(S)$ could be removed without violating feasibility, a contradiction.
+$\blacksquare$
+\paragraph{Claim 2}
+All leaves of $H$ correspond to active nodes.
+\paragraph{Proof of Claim 2}
+Let $c$ be a node of a component $\bar{H}$ of $H$.
+Suppose $f(c) = 0$.
+Note that $f(\bar{H}) = 0$ by our previous lemma.
+By symmetry $f(V\backslash \bar{H}) = 0$.
+By maximality $f((V\backslash \bar{H})\cup \{c\}) = 0$.
+By symmetry $f(\bar{H}\backslash \{c\}) = 0$.
+This implies that the edge connecting $c$ to $\bar{H}$ could have been removed in the deletion step, a contradiction.$\blacksquare$
+\paragraph{}
+Now our claims imply the following. Let $W$ be the set of shrunk nodes in $H$, and $\bar{W}$ the remains nodes. Discard isolated nodes in $H$. Then
+$$\sum_{v\in W} d_H(v) + \sum_{v\in\bar{W}}d_H(v) \leq 2|W \cup \bar{W}|$$
+since $H$ is a forest. But then $\sum_{v\in W} |\delta_H(v)| \leq 2|W|$, which is exactly equivalent to that which we set out to show. 
 \end{proof}

### 2017-01-24 CO750-ApproximationAlgorithmsTopics/lectures/lecture_06.tex
Scribe lecture 6

--- 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
@@
 
-\lhead{2017-01-19}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 5} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-01-24}
+\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 6} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+\section{Generalization}
 \paragraph{}
-Before moving on to the next topic, we just want to mention an important theorem that results from the theory built in the previous lectures.
-It tells us that we can have deterministic approximation algorithms that give just as good bounds as the randomized ones studied previously.
+A function $f: 2^V \rightarrow \{0,1\}$ is called {\it uncrossable} if
+\begin{enumerate}
+\item $f(V) = 0$
+\item If $f(A) = f(B) = 1$ for some $A,B \subseteq V$, then either
+\begin{itemize}
+\item$f(A\cup B) = f(A \cap B) = 1$
+\item or $f(A\backslash B) = f(B\backslash A) = 1$.
+\end{itemize}
+\end{enumerate}
+Observe that if $f$ is $01$-proper then $f$ is uncrossable (this is left as an exercise).
 \begin{theorem}
-Let $G=(V,E)$ be a graph.
-Let $d: E \rightarrow \R_+$ be a metric function.
-Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$.
-There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
+The Primal-Dual algorithm of the previous lecture is a $2$-approximation algorithm for uncrossable functions provided that you have a polynomial time oracle to find minimum violated sets.
 \end{theorem}
 \begin{proof}
-Omitted. May be in a talk at the end of course via some students project. Involves derandomization via the method of conditional expectation.
+Omitted.
 \end{proof}
-\section{Primal - Dual Methods}
+\section{General Proper Functions}
 \paragraph{}
-Primal-Dual methods are a common paradigm in algorithm design.
-We will begin by showing how they are used to given exact algorithms, and then explain relaxed complementary slackness which allows us to extend the paradigm to give approximation algorithms.
+We now consider a proper function $f : 2^V \rightarrow \N$. Consider the following IP for proper function $f$:
+\begin{align*}
+\min &\sum_e c_e x_e \\
+\text{s.t.} \sum_{e\in \delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+1 \geq &x_e \geq 0 &\forall e \in E \\
+x \in \Z^E.
+\end{align*}
+Our goal is to design a primal-dual approximation algorithm for solving this IP with approximation factor $O(\log f_{max})$ where $f_{max}$ is $\max_S\{f(S)\}$.
+Note that there exists a $2$-approximation algorithm for this IP using iterative rounding (to be seen in later lectures).
+It is an open problem to design a $2$-approximation algorithm for this IP using primal-dual methods.
 \paragraph{}
-Consider the {\it Arborescence Problem} (ARB).
-In this problem we are given a directed graph $D= (V,E)$, cost function $c: E \rightarrow \R_+$, and a root vertex $r \in V$.
-We are to find an arborescence $T$ rooted at $r$ which spans $V$ and minimizes $\sum_{e\in E(T)} c_e$.
-\paragraph{}
-As the name of this method might suggest we will used integrally the linear programming relaxation of this problem:
-\begin{align*}
-\min \sum_{e \in E} c_e x_e  &\ \\
-\text{s.t.} \sum_{e \in \delta^+(S)} x_e &\geq 1 &\forall S \subseteq V\backslash\{r\} \\
-x_e &\geq 0 &\forall e \in E,
-\end{align*}
-and its dual program:
-\begin{align*}
-\max \sum_{S \subseteq V\backslash\{r\}} y_S &\ \\
-\text{s.t.} \sum_{S : e \in \delta^+(S)} y_S &\leq c_e &\forall e \in E \\
-y_S &\geq 0 &\forall S \subseteq V\backslash \{r\}.
-\end{align*}
-Here $\delta^+(S) = \{(u,v) \in E : u \in S, v\not\in S\}$ denotes set of arcs outgoing from $S$.
-\paragraph{}
-Our primal-dual algorithm for (ARB) operates as follows (start by setting $x=0$, $y=0$, for implementation track only non-negative $y$ to avoid exponential runtime): 
+Consider the following Primal-Dual Algorithm:
 \begin{enumerate}
-\item While $x$ is not feasible for the primal:
+\item $F_0 = \emptyset$
+\item For phase $p = 1, \dots, f_{max}$
 	\begin{enumerate}
-	\item Increase $y_S$ uniformly for all ``strongly connected components" (with respect to $x$) $S$ where $x(\delta^+(S)) = 0$ until some constraints for an edge $e$ becomes tight.
-	\item Set $x_e=1$ for tight edge $e$ in previous step.
+	\item Set $h_p(S) = 1$ if $f(S) - |\delta_{F_{p-1}(S)}| = f_{max} - p + 1$, and set $h_p(S) = 0$ otherwise.
+	\item Set $E_p = E - F_{p-1}$
+	\item Apply $2$-approximation algorithm using primal-dual for $(V,E_p)$ with function $h_p$, and let $A$ be the edges returned by the algorithm.
+	\item Set $F_p = F_{p-1} \cup A$.
 	\end{enumerate}
-\item Deletion Step: Consider each edge $e$ where $x_e = 1$ in reverse order from the order they were set to $1$ and set $x_e = 0$ as long as this does not cause the resulting $x$ to become infeasible.
-\item Return $x$.
+\item Return $F_{f_{max}}$.
 \end{enumerate}
+\begin{lemma}
+For all iterations $p$, the function $h_p$ is uncrossable.
+\end{lemma}
+\begin{proof}
+(Exercise).
+\end{proof}
 \begin{theorem}
-The previous algorithm is an exact polynomial time algorithm for (ARB).
+The primal-dual algorithm given above is an $O(\log f_{max})$-approximation algorithm provided that you have a $2$-approximation primal-dual algorithm for $h_p$.
 \end{theorem}
 \begin{proof}
-When the algorithm terminates we have that if $x_e > 0$ then $x_e = 1$  and so the dual constraint is tight, that is:
-$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$
-This is immediate from our construction of $x$. Hence the primal complementary slackness conditions hold.
-\paragraph{}
-For the dual complementary slackness conditions, suppose there exists $S$ such that $y_S > 0$, yet $\sum_{e\in \delta^+(S)} x_e \geq 2$.
-Let $e, \bar{e} \in \delta^+(S)$ and suppose without loss of generality that $\bar{e}$ was added before $e$ by the algorithm.
-Since we did not delete $e$, there exists a node $v \in S$ that uses $e$ to reach $r$.
-But since $S$ is strongly connected, no nodes in $S$ need $\bar{e}$ to reach $r$ (they can go $v$ then eventually through $e$), so we could have deleted $\bar{e}$ in the Deletion Step.
-A contradiction.
+Let $y^p$ be the dual solution constructed in each iteration $p$.
+Set $z_e^p = 0$ for all $e \in E_p$.
+Set $z_e^p = \sum_{S: e\in \delta(S)} y_S^p$ otherwise.
+Then $(y^p, z^p)$ is dual feasible for the original problem.
+We compute the bound on its optimal value $D^*$:
+\begin{align*}
+D^* &\geq \sum_S f(S) y_S^p - \sum_{e\in E} z_e^p\\
+& = \sum_S f(S) y_S^p - \sum_{e\not\in E_p}\sum_{S : e\in\delta(S)}y_S^p\\
+&= \sum_S f(S) y_S^p - \sum_S |\delta_{F_{p-1}}(S)| y_S \\
+&= \sum_S (f(S) - |\delta_{F_{p-1}}(S)|)y_S^p \\
+&= (f_{max} - p + 1)\sum_S y_S^p.
+\end{align*}
+This implies that
+$$\sum_{e \in F_{max}^p} c_e \leq 2 \sum_{p=1}^{f_{max}} y_S^p \leq 2 \sum_{p=1}^{f_{max}} \frac{D^*}{f_{max} - p + 1} = O(\log f_{max}) D^* .$$
 \end{proof}

### 2017-01-27 CO750-ApproximationAlgorithmsTopics/lectures/lecture_07.tex
Reading

--- 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
@@
 
-\lhead{2017-01-19}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 5} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-01-26}
+\rhead{William Justin Toth CO750-Approximation Algorithms Lecture n} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+\section*{Integrality Gap}
 \paragraph{}
-Before moving on to the next topic, we just want to mention an important theorem that results from the theory built in the previous lectures.
-It tells us that we can have deterministic approximation algorithms that give just as good bounds as the randomized ones studied previously.
+Let $P$ be an optimization problem, and suppose that there exists an integer programming formulation for $P$.
+Denote by $OPT(I)$ the objective function value of an optimal solution to an instance $I$ of $P$.
+Denote by $OPT_f(I)$ the objective function value of the Linear Programming relaxation of the IP formulation of $I$.
+\paragraph{}
+The integrality gap of the formulation is
+$$\inf_{I \in P} \frac{OPT(I)}{OPT_f(I)}$$
+for maximization problems, and
+$$\sup_{I \in P} \frac{OPT(I)}{OPT_f(I)}$$
+for minimization problems.
+\section{Iterative Rounding}
+\paragraph{}  Recall the linear programming formulation $LP$ for Generalized Steiner Network (GSN):
+\begin{align*}
+\min &\sum_e c_e x_e \\
+\text{s.t.} x(\delta(S)) &\geq f(S) &\forall S \subseteq V\\
+x&\geq 0
+\end{align*}
+where $f(S) = \max_{u \in S, v \not\in S} r(u,v).$
+\paragraph{}
+A function $f: 2^V \rightarrow \Z$ is called weakly-supermodular (wsm) if
+\begin{enumerate}
+\item $f(V) = 0$ and
+\item For all $A,B \subseteq V$ at least one of the following holds
+\begin{enumerate}
+\item $f(A) + f(B) \leq f(A\backslash B) + f(B \backslash A)$ or
+\item $f(A) + f(B) \leq f(A\cup B) + f(A\cap B).$
+\end{enumerate}
+\end{enumerate}
+\begin{proposition}
+If $f$ is a proper function then $f$ is wsm.
+\end{proposition}
+\paragraph{}
+Recall that an extreme point solution (xps) to an LP is a feasible solution which cannot be written as a convex combination of other feasible solutions.
+\begin{proposition}
+An optimal xps of $LP$ for GSN can be computed in polynomial time.
+\end{proposition}
 \begin{theorem}
-Let $G=(V,E)$ be a graph.
-Let $d: E \rightarrow \R_+$ be a metric function.
-Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$.
-There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
+For any wsm function $f$, any xps to $LP$ satisfies $x_e \geq \frac{1}{2}$ for at least one $e \in E$.
+\end{theorem}
+\paragraph{}
+The previous theorem yields a natural algorithm for GSN:
+\begin{enumerate}
+\item Set $H= \emptyset$, and $f' = f$.
+\item While $f' \neq 0$
+\begin{enumerate}
+\item Find an optimal xps for $LP$ with $f'$
+\item For all $e$ such that $x_e \geq \frac{1}{2}$, include $\ceil{x_e}$ copies of $e$ in $H$
+\item Update $f'$: for all $S\subseteq V$, set $f(S) = \max \{f(S)-|\delta_H(S)|, 0\}$.
+\end{enumerate}
+\item Output $H$.
+\end{enumerate}
+\begin{lemma}
+Let $H$ be a (multi)subgraph of $G$. If $f: 2^V \rightarrow \Z_+$ is wsm then $f'(S) = \max \{ f(S) - |\delta_H(S)|, 0\}$ is wsm.
+\end{lemma}
+\begin{lemma}
+We can solve the $LP$ relaxation for GSN in polynomial time at any iteration of the algorithm for instances of GSN.
+\end{lemma}
+\begin{theorem}
+The solution $H$ output by the algorithm has cost at most $2$ times the cost of the optimal solution to the $LP$ relaxation.
 \end{theorem}
 \begin{proof}
-Omitted. May be in a talk at the end of course via some students project. Involves derandomization via the method of conditional expectation.
+Proceed by induction on the number of iterations.
+If the algorithm only requires one iteration then the result is immediate.
+Otherwise let $x$ be an optimal xps in the first iteration.
+Let $\hat{x}$ be defined as
+$$ \hat{x}_e = \begin{cases} 1, &\forall e, x_e \geq \frac{1}{2} \\
+0, &\text{otherwise}\end{cases}$$
+Let $H^1$ be the set of edges taken in the first iteration.
+Then $c(H^1) \leq 2c(\hat{x})$.
+Let $f'$ be the requirement function after the first iteration and let $H'$ be the set of edges picked by the algorithm in all subsequent iterations.
+\paragraph{}
+The key observation: $(x-\hat{x})$ is a feasible solution for $f'$. So by induction $c(H') \leq 2c(x-\hat{x})$. In total
+$$c(H^1 \cup H') \leq 2c(\hat{x}) + 2c(x-\hat{x}) \leq 2c(x).$$
 \end{proof}
-\section{Primal - Dual Methods}
 \paragraph{}
-Primal-Dual methods are a common paradigm in algorithm design.
-We will begin by showing how they are used to given exact algorithms, and then explain relaxed complementary slackness which allows us to extend the paradigm to give approximation algorithms.
+An xps $x$ of $LP$ satisfies $m = |E|$ linearly independent inequalities with equality (tight).
+In our analysis to follow we will make the simplifying assumption that $x_e > 0$ for all $e\in E$.
+We may assume this by removing all $e$ such that $x_e = 0$.
+For a set $S$ we will denote by $\chi(S)$ the incidence vector of $S$.
 \paragraph{}
-Consider the {\it Arborescence Problem} (ARB).
-In this problem we are given a directed graph $D= (V,E)$, cost function $c: E \rightarrow \R_+$, and a root vertex $r \in V$.
-We are to find an arborescence $T$ rooted at $r$ which spans $V$ and minimizes $\sum_{e\in E(T)} c_e$.
-\paragraph{}
-As the name of this method might suggest we will used integrally the linear programming relaxation of this problem:
-\begin{align*}
-\min \sum_{e \in E} c_e x_e  &\ \\
-\text{s.t.} \sum_{e \in \delta^+(S)} x_e &\geq 1 &\forall S \subseteq V\backslash\{r\} \\
-x_e &\geq 0 &\forall e \in E,
-\end{align*}
-and its dual program:
-\begin{align*}
-\max \sum_{S \subseteq V\backslash\{r\}} y_S &\ \\
-\text{s.t.} \sum_{S : e \in \delta^+(S)} y_S &\leq c_e &\forall e \in E \\
-y_S &\geq 0 &\forall S \subseteq V\backslash \{r\}.
-\end{align*}
-Here $\delta^+(S) = \{(u,v) \in E : u \in S, v\not\in S\}$ denotes set of arcs outgoing from $S$.
-\paragraph{}
-Our primal-dual algorithm for (ARB) operates as follows (start by setting $x=0$, $y=0$, for implementation track only non-negative $y$ to avoid exponential runtime): 
-\begin{enumerate}
-\item While $x$ is not feasible for the primal:
-	\begin{enumerate}
-	\item Increase $y_S$ uniformly for all ``strongly connected components" (with respect to $x$) $S$ where $x(\delta^+(S)) = 0$ until some constraints for an edge $e$ becomes tight.
-	\item Set $x_e=1$ for tight edge $e$ in previous step.
-	\end{enumerate}
-\item Deletion Step: Consider each edge $e$ where $x_e = 1$ in reverse order from the order they were set to $1$ and set $x_e = 0$ as long as this does not cause the resulting $x$ to become infeasible.
-\item Return $x$.
-\end{enumerate}
+A collection $\cL$ of subsets of $V$ is a \textit{laminar} family if no two sets in this collection ``crosses".
+By crosses we mean for all $A, B \in \cL$, we have $A\cap B \neq \emptyset$ and either $A\backslash B \neq \emptyset$ or $B\backslash A \neq \emptyset$.
 \begin{theorem}
-The previous algorithm is an exact polynomial time algorithm for (ARB).
+For any xps $x$ to $LP$ there is a collection of $m$ tight sets such that their incidence vectors are linearly independent, and the collection is a laminar family.
 \end{theorem}
 \begin{proof}
-When the algorithm terminates we have that if $x_e > 0$ then $x_e = 1$  and so the dual constraint is tight, that is:
-$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$
-This is immediate from our construction of $x$. Hence the primal complementary slackness conditions hold.
-\paragraph{}
-For the dual complementary slackness conditions, suppose there exists $S$ such that $y_S > 0$, yet $\sum_{e\in \delta^+(S)} x_e \geq 2$.
-Let $e, \bar{e} \in \delta^+(S)$ and suppose without loss of generality that $\bar{e}$ was added before $e$ by the algorithm.
-Since we did not delete $e$, there exists a node $v \in S$ that uses $e$ to reach $r$.
-But since $S$ is strongly connected, no nodes in $S$ need $\bar{e}$ to reach $r$ (they can go $v$ then eventually through $e$), so we could have deleted $\bar{e}$ in the Deletion Step.
-A contradiction.
+To follow in next lecture.
 \end{proof}

### 2017-01-27 CS798-ConvexityAndOptimization/reading/reading_02.tex
Reading

--- 2017-01-27 CO750-ApproximationAlgorithmsTopics/lectures/lecture_07.tex
@@
 
-\lhead{2017-01-26}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture n} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-01-27}
+\rhead{William Justin Toth CS798-Convexity and Optimization Reading 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-\section*{Integrality Gap}
 \paragraph{}
-Let $P$ be an optimization problem, and suppose that there exists an integer programming formulation for $P$.
-Denote by $OPT(I)$ the objective function value of an optimal solution to an instance $I$ of $P$.
-Denote by $OPT_f(I)$ the objective function value of the Linear Programming relaxation of the IP formulation of $I$.
+We cover Sections 5.8 and 5.9 of Convex Optimization by Boyd and Vandenberghe.
+\section*{5.8 - Theorems of  Alternatives}
+\subsection*{Weak Alternatives via the dual}
 \paragraph{}
-The integrality gap of the formulation is
-$$\inf_{I \in P} \frac{OPT(I)}{OPT_f(I)}$$
-for maximization problems, and
-$$\sup_{I \in P} \frac{OPT(I)}{OPT_f(I)}$$
-for minimization problems.
-\section{Iterative Rounding}
-\paragraph{}  Recall the linear programming formulation $LP$ for Generalized Steiner Network (GSN):
+We study the use of Lagrangian Duality Theory to the problem of determining feasibility for a set of inequality and equalities:
+$$f_i(x) \leq 0,\ \ i=1,\dots,m \quad h_i(x) = 0, \ \ i=1,\dots,p $$
+We assume the domain $\cD = \cap_{i=1}^m \textbf{dom} f_i \cap \cap_{i=1}^p \textbf{dom}h_i$ is nonempty.
+We can think of this as a standard minimization problem with objective function $0$ whose optimal value if $0$ if the system is feasible and $\infty$ otherwise.
+\paragraph{}
+We may associate with our inequality system a dual function
+$$g(\lambda, \nu) = \inf_{x \in \cD} (\sum_{i=1}^m \lambda_i f_i(x) + \sum_{i=1}^p \nu_i h_i(x))$$
+The associated dual problem to our associated primal problem is to maximize $g(\lambda, \nu)$ subject to $\lambda \geq 0$. Since $f_0 = 0$ this dual as optimal value
+$$ d^* = \begin{cases}
+\infty, &\lambda \geq 0 g(\lambda,\nu) > 0 \text{ is feasible} \\
+0, &\text{otherwise}.
+\end{cases}
+$$
+Hence at most one of our primal and dual systems is feasible since the dual lower bounds the primal.
+The two systems are called $\textit{weak alternatives}$ since they satisfy this property.
+Notice the primal system need not be convex, but the dual system always is.
+\subsection*{Strong Alternatives}
+\paragraph{}
+When the original inequality system is convex, and some type of constraint qualification holds, we can obtain that the systems are $\textit{strong alternatives}$ which means that exactly one of them always holds.
+\paragraph{}
+Consider the system
+$$f_i(x) \leq 0,\ \ i=1,\dots,m \quad Ax=b $$
+and its alternative
+$$\lambda \geq 0 \quad g(\lambda, \nu) > 0$$
+Provided there exists an $x \in \textbf{relint} \cD$ with $Ax = b$ and optimal value $p^*$ (the minimum slack scalar optimization problem) that , then the two systems above are strong alternatives.
+With these assumptions we have strong duality $p^*=d^*$.
+Now suppose that our primal system is infeasible.
+Then $p^* > 0$.
+Then the dual optimal point $(\lambda^*, \nu^*)$ satisfies the alternative system.
+\section*{5.9 - Generalized Inequalities}
+\paragraph{}
+The theory above associated with Lagrangian Duality can be extended to problems with generalized inequality constraints:
 \begin{align*}
-\min &\sum_e c_e x_e \\
-\text{s.t.} x(\delta(S)) &\geq f(S) &\forall S \subseteq V\\
-x&\geq 0
+\min &f_0(x) \\
+\text{s.t.} f_i(x) &\preccurlyeq_{K_i} 0, &i=1,\dots,m \\
+h_i(x) & = 0, &i=1,\dots,p.
 \end{align*}
-where $f(S) = \max_{u \in S, v \not\in S} r(u,v).$
-\paragraph{}
-A function $f: 2^V \rightarrow \Z$ is called weakly-supermodular (wsm) if
-\begin{enumerate}
-\item $f(V) = 0$ and
-\item For all $A,B \subseteq V$ at least one of the following holds
-\begin{enumerate}
-\item $f(A) + f(B) \leq f(A\backslash B) + f(B \backslash A)$ or
-\item $f(A) + f(B) \leq f(A\cup B) + f(A\cap B).$
-\end{enumerate}
-\end{enumerate}
-\begin{proposition}
-If $f$ is a proper function then $f$ is wsm.
-\end{proposition}
-\paragraph{}
-Recall that an extreme point solution (xps) to an LP is a feasible solution which cannot be written as a convex combination of other feasible solutions.
-\begin{proposition}
-An optimal xps of $LP$ for GSN can be computed in polynomial time.
-\end{proposition}
-\begin{theorem}
-For any wsm function $f$, any xps to $LP$ satisfies $x_e \geq \frac{1}{2}$ for at least one $e \in E$.
-\end{theorem}
-\paragraph{}
-The previous theorem yields a natural algorithm for GSN:
-\begin{enumerate}
-\item Set $H= \emptyset$, and $f' = f$.
-\item While $f' \neq 0$
-\begin{enumerate}
-\item Find an optimal xps for $LP$ with $f'$
-\item For all $e$ such that $x_e \geq \frac{1}{2}$, include $\ceil{x_e}$ copies of $e$ in $H$
-\item Update $f'$: for all $S\subseteq V$, set $f(S) = \max \{f(S)-|\delta_H(S)|, 0\}$.
-\end{enumerate}
-\item Output $H$.
-\end{enumerate}
-\begin{lemma}
-Let $H$ be a (multi)subgraph of $G$. If $f: 2^V \rightarrow \Z_+$ is wsm then $f'(S) = \max \{ f(S) - |\delta_H(S)|, 0\}$ is wsm.
-\end{lemma}
-\begin{lemma}
-We can solve the $LP$ relaxation for GSN in polynomial time at any iteration of the algorithm for instances of GSN.
-\end{lemma}
-\begin{theorem}
-The solution $H$ output by the algorithm has cost at most $2$ times the cost of the optimal solution to the $LP$ relaxation.
-\end{theorem}
-\begin{proof}
-Proceed by induction on the number of iterations.
-If the algorithm only requires one iteration then the result is immediate.
-Otherwise let $x$ be an optimal xps in the first iteration.
-Let $\hat{x}$ be defined as
-$$ \hat{x}_e = \begin{cases} 1, &\forall e, x_e \geq \frac{1}{2} \\
-0, &\text{otherwise}\end{cases}$$
-Let $H^1$ be the set of edges taken in the first iteration.
-Then $c(H^1) \leq 2c(\hat{x})$.
-Let $f'$ be the requirement function after the first iteration and let $H'$ be the set of edges picked by the algorithm in all subsequent iterations.
-\paragraph{}
-The key observation: $(x-\hat{x})$ is a feasible solution for $f'$. So by induction $c(H') \leq 2c(x-\hat{x})$. In total
-$$c(H^1 \cup H') \leq 2c(\hat{x}) + 2c(x-\hat{x}) \leq 2c(x).$$
-\end{proof}
-\paragraph{}
-An xps $x$ of $LP$ satisfies $m = |E|$ linearly independent inequalities with equality (tight).
-In our analysis to follow we will make the simplifying assumption that $x_e > 0$ for all $e\in E$.
-We may assume this by removing all $e$ such that $x_e = 0$.
-For a set $S$ we will denote by $\chi(S)$ the incidence vector of $S$.
-\paragraph{}
-A collection $\cL$ of subsets of $V$ is a \textit{laminar} family if no two sets in this collection ``crosses".
-By crosses we mean for all $A, B \in \cL$, we have $A\cap B \neq \emptyset$ and either $A\backslash B \neq \emptyset$ or $B\backslash A \neq \emptyset$.
-\begin{theorem}
-For any xps $x$ to $LP$ there is a collection of $m$ tight sets such that their incidence vectors are linearly independent, and the collection is a laminar family.
-\end{theorem}
-\begin{proof}
-To follow in next lecture.
-\end{proof}
+where $K_i$ are proper cones (that is pointed closed convex cones).
+We need not assume the problem is convex, and as before we assume $\cD \neq \emptyset$.
+The order associated with cone $K_i$, denoted $\preccurlyeq_{K_i}$ is given as $A \preccurlyeq_{K_i} B$ if and only if $B-A \in K_i$.
 \end{document}

### 2017-01-27 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Assignment templating

--- 2017-01-27 CS798-ConvexityAndOptimization/reading/reading_02.tex
@@
 
-\lhead{2017-01-27}
-\rhead{William Justin Toth CS798-Convexity and Optimization Reading 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-02-09}
+\rhead{William Justin Toth CO750-Approximation Algorithms Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-\paragraph{}
-We cover Sections 5.8 and 5.9 of Convex Optimization by Boyd and Vandenberghe.
-\section*{5.8 - Theorems of  Alternatives}
-\subsection*{Weak Alternatives via the dual}
-\paragraph{}
-We study the use of Lagrangian Duality Theory to the problem of determining feasibility for a set of inequality and equalities:
-$$f_i(x) \leq 0,\ \ i=1,\dots,m \quad h_i(x) = 0, \ \ i=1,\dots,p $$
-We assume the domain $\cD = \cap_{i=1}^m \textbf{dom} f_i \cap \cap_{i=1}^p \textbf{dom}h_i$ is nonempty.
-We can think of this as a standard minimization problem with objective function $0$ whose optimal value if $0$ if the system is feasible and $\infty$ otherwise.
-\paragraph{}
-We may associate with our inequality system a dual function
-$$g(\lambda, \nu) = \inf_{x \in \cD} (\sum_{i=1}^m \lambda_i f_i(x) + \sum_{i=1}^p \nu_i h_i(x))$$
-The associated dual problem to our associated primal problem is to maximize $g(\lambda, \nu)$ subject to $\lambda \geq 0$. Since $f_0 = 0$ this dual as optimal value
-$$ d^* = \begin{cases}
-\infty, &\lambda \geq 0 g(\lambda,\nu) > 0 \text{ is feasible} \\
-0, &\text{otherwise}.
-\end{cases}
-$$
-Hence at most one of our primal and dual systems is feasible since the dual lower bounds the primal.
-The two systems are called $\textit{weak alternatives}$ since they satisfy this property.
-Notice the primal system need not be convex, but the dual system always is.
-\subsection*{Strong Alternatives}
-\paragraph{}
-When the original inequality system is convex, and some type of constraint qualification holds, we can obtain that the systems are $\textit{strong alternatives}$ which means that exactly one of them always holds.
-\paragraph{}
-Consider the system
-$$f_i(x) \leq 0,\ \ i=1,\dots,m \quad Ax=b $$
-and its alternative
-$$\lambda \geq 0 \quad g(\lambda, \nu) > 0$$
-Provided there exists an $x \in \textbf{relint} \cD$ with $Ax = b$ and optimal value $p^*$ (the minimum slack scalar optimization problem) that , then the two systems above are strong alternatives.
-With these assumptions we have strong duality $p^*=d^*$.
-Now suppose that our primal system is infeasible.
-Then $p^* > 0$.
-Then the dual optimal point $(\lambda^*, \nu^*)$ satisfies the alternative system.
-\section*{5.9 - Generalized Inequalities}
-\paragraph{}
-The theory above associated with Lagrangian Duality can be extended to problems with generalized inequality constraints:
-\begin{align*}
-\min &f_0(x) \\
-\text{s.t.} f_i(x) &\preccurlyeq_{K_i} 0, &i=1,\dots,m \\
-h_i(x) & = 0, &i=1,\dots,p.
-\end{align*}
-where $K_i$ are proper cones (that is pointed closed convex cones).
-We need not assume the problem is convex, and as before we assume $\cD \neq \emptyset$.
-The order associated with cone $K_i$, denoted $\preccurlyeq_{K_i}$ is given as $A \preccurlyeq_{K_i} B$ if and only if $B-A \in K_i$.
+\section{}
+
 \end{document}

### 2017-01-28 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
1a apxalgs, 1,2a convopt

@@
 \section{}
-
+\paragraph{}
+Let $E$ be the edge set of a graph $G$.
+Let $\cS$ be the collection of all sets $S \subseteq E$ that are matchings of $G$.
+Let $w: E \rightarrow \R_+$ be a weight function on the edges.
+For a subset $\bar{E} \subseteq E$ we denote by $G[\bar{E}]$ the subgraph of $G$ induced by the edges in $\bar{E}$.
+\subsection{a}
+\paragraph{}
+Let $\bar{E} \subseteq E$.
+Suppose that $M \subseteq \bar{E}$ is an inclusion-wise maximal matching of $G[\bar{E}]$.
+Let $\bar{M}$ be a matching of $G[\bar{E}]$.
+Let $f : M \rightarrow \{ \{e_1,e_2\} | e_1, e_2 \in \bar{M}\}$ be defined by
+$$ f(uv) = \{e_1, e_2\}$$
+where $e_1 \in \bar{M}$ covers $u$ and $e_2 \in \bar{M}$ covers $v$, for each $uv \in M$.
+\paragraph{}
+We claim that for all $uv \in \bar{M}$, there exists $P \in Im(f)$ (where $Im(f)$ denotes the image of $f$) such that $uv \in P$.
+To see this suppose for a contradiction that there exists some $uv \in \bar{M}$ where each $P \in Im(f)$ does not contain $uv$.
+Then by the definition of $f$, neither $u$ nor $v$ are covered by $M$.
+But this violates that $M$ is inclusion-wise maximal since $uv \not\in M$ and $M \cup \{uv\}$ is a matching.
+Hence the claim holds.
+\paragraph{}
+From the previous claim we observe $|\bar{M}| \leq | \cup_{P \in Im(f)} P|$. But this yields a string of inequalities which complete the proof:
+\begin{align*}
+|\bar{M}| &\leq |\cup_{P \in Im(f)} P| \\
+&\leq \sum_{P \in Im(f)} |P| \\
+&\leq \sum_{P \in Im(f)} 2 \\
+&\leq 2|Im(f)| \\
+&\leq 2|M|. 
+\end{align*}
+With the last inequality following since $M$ is the domain of $f$. Hence we have $|M| \geq \frac{1}{2}|\bar{M}|$ as desired. $\blacksquare$
+\subsection{b}
+\paragraph{}
+Let $e$
 \end{document}

### 2017-01-28 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
Assignment templating
1a apxalgs, 1,2a convopt

--- 2017-01-27 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
@@
 
-\lhead{2017-02-09}
-\rhead{William Justin Toth CO750-Approximation Algorithms Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-02-15}
+\rhead{William Justin Toth CS798-Convexity and Optimization Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+%Q1%
 \section{}
-
+\paragraph{}
+Let $q > p > 0$ and let $x_1, \dots, x_n \in \R_{>0}$.
+Let $f : \R_{>0} \rightarrow \R_{>0}$ be the function given by $f(x) = x^\frac{q}{p}$.
+We claim that $f$ is convex.
+To see this we will use the second order condition ($f'' \geq 0$).
+We have by calculus that
+$$f' (x)= \frac{q}{p} x^{\frac{q}{p}-1}$$
+and thus 
+$$f''(x) = \frac{q}{p}(\frac{q}{p} -1) x^{\frac{q}{p} - 2}.$$
+Now since $q > p$, we have $\frac{q}{p} > 1$.
+So then $\frac{q}{p} - 1$ is non-negative, and further $x^{\frac{q}{p} - 2} > 0$ since $x >0$.
+Hence $f''(x) \geq 0$ on the domain of $f$.
+Therefore by the second order condition, $f$ is convex.
+\paragraph{}
+Now observe by convexity that
+$$f(\frac{1}{n}\sum_{i=1}^n x_i^p) \leq \frac{1}{n} \sum_{i=1}^n f(x_i^p) = \frac{1}{n} \sum_{i=1}^n x_i^q.$$
+On the other hand, by applying $f$ directly we see that
+$$f(\frac{1}{n}\sum_{i=1}^n x_i^p) = \frac{1}{n}^\frac{q}{p}(\sum_{i=1}^n x_i^p)^\frac{q}{p}.$$
+Combining the previous two expressions yields the inequality
+$$\frac{1}{n}^\frac{q}{p}(\sum_{i=1}^n x_i^p)^\frac{q}{p} \leq \frac{1}{n} \sum_{i=1}^n x_i^q.$$
+By raising both sides to the exponent $\frac{1}{q}$ (which preserves order) we obtain that
+$$(\frac{1}{n}\sum_{i=1}^n x_i^p)^\frac{1}{p} \leq (\frac{1}{n} \sum_{i=1}^n x_i^q)^\frac{1}{q}$$
+as desired. $\blacksquare$
+\section{}
+%Pa
+\subsection{a}
+\paragraph{}
+Consider the function
+$$f(x) = (\sum_{i=1}^n x_i^p)^\frac{1}{p}$$
+with $\text{dom} f = \R^n_{>0}$. We will compute the Hessian matrix. First observe that
+$$\frac{\partial f}{\partial x_i} f = x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p}$$
+for $i = 1, \dots, n$. Now for $j = 1, \dots, n$
+\begin{align*}
+\frac{\partial f}{\partial x_j}\frac{\partial f}{\partial x_i} f &= \frac{\partial f}{\partial x_j}   x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p} \\
+&= \begin{cases}
+0, &\text{ if } i=j \\
+(p-1)x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p}[1- \frac{x_i^{p-1}}{\sum_{i=1}^n x_i^p}], &\text{if } i \neq j.
+\end{cases}
+\end{align*}
+Observe that if $p \geq 1$ each factor in the second case is non-negative when $x \in \R^n_{>0}$.
+Hence $\frac{\partial f}{\partial x_i} f  \geq 0$ for all $x \in \text{dom f}$ for all $i,j = 1, \dots, n$.
+Thus each entry of the Hessian is non-negative, and so the Hessian is PSD.
+Thus by the second order condition $f$ is convex.
+\paragraph{}
+Now when $p < 1$, by a similar calculation we observe
+$$\frac{\partial f}{\partial x_i} -f = \begin{cases}
+0, &\text{ if } i=j \\
+(1-p)x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p}[1- \frac{x_i^{p-1}}{\sum_{i=1}^n x_i^p}], &\text{if } i \neq j.
+\end{cases}$$
+Again each factor is non-negative, and hence $-f$ is convex by the second order condition. Therefore $f$ is concave. $\blacksquare$
+\paragraph{}
+Next we show that
+$$||x||_p = (\sum_{i=1}^n |x_i|^p)^\frac{1}{p}$$
+is a norm for $p\geq 1$.
+It is easy to see that $||x||_p$ is non-negative since it is a sum of absolute values.
+Similarly if any $x_i > 0$ then $||x||_p > 0$, so $||x||_p = 0$ if and only if $x = 0$.
+Now consider some scalar $\alpha$.
+We have
+$$||\alpha x||_p = (\sum_{i=1}^n |\alpha x_i|^p)^\frac{1}{p} = (\sum_{i=1}^n |\alpha|^p|\cdot |x_i|^p)^\frac{1}{p} = |\alpha|(\sum_{i=1}^n |x_i|^p)^\frac{1}{p} = |\alpha|\cdot ||x||_p.$$
+\paragraph{}
+Lastly it remains to verify the triangle inequality. Let $x, y \in \R^n$. If either $x$ or $y$ is $0$, say without loss of generality $y=0$, then 
+$$||x + y||_p = ||x||_p + 0 = ||x||_p + ||y||_p$$
+immediately. So we may assume $x$ and $y$ are non-zero. Thus it suffices to prove
+$$f(x+y) \leq f(x) + f(y).$$
+To show this we exploit convexity of $f$ as follows
+$$f(x+y) = f(\frac{1}{2}(2x+2y)) \leq \frac{1}{2}(f(2x) + f(2y)) = \frac{1}{2} ((2^p\sum_{i=1}^n x^p)^\frac{1}{p} +(2^p\sum_{i=1}^n y^p)^\frac{1}{p}) = f(x) + f(y).$$
+Therefore $||x + y||_p \leq ||x||_p + ||y||_p$, and hence $||x||_p$ is a norm. $\blacksquare$
+\paragraph{}
+Finally we give a counter-example showing $||x||_p$ is not a norm when $0 < p < 1$.
+Consider $p =\frac{1}{2}$ and $e_1, e_2 \in \R^2$ the standard basis vectors of $\R^2$.
+We prove $||x||_p$ violates the triangle inequality.
+Observe
+$$||e_1 + e_2 ||_p = (1 + 1)^2 = 4$$
+while 
+$$||e_1|| = (1+0)^2 = 1 = ||e_2||$$
+hence 
+$$||e_1 + e_2||  = 4 < 2 = ||e_1|| + ||e_2||.$$
+Thus the triangle inequality is violated, and so $||x||_p$ is not a norm.
+%Pb
+\subsection{b}
 \end{document}

### 2017-01-30 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
3b of apxalgs

@@
 as desired. $\blacksquare$
-\section{}
 %Pa
@@
 \subsection{b}
+%Pc
+\subsection{c}
+%Q2
+\section{}
+
+%Q3
+\section{}
+\paragraph{}
+Let $G = (V,E)$ be a graph. For any $S, T \subseteq V$ we denote by $E(S,T)$ the set of edges between $S$ and $T$. Formally $E(S,T) = \{\{s,t\} \in E: s \in S, t \in T\}$.
+%Pa
+\subsection{a}
+
+%Pb
+\subsection{b}
+\paragraph{}
+Our goal is to find a maximum $k$-Cut of $G$. That is, we seek to find a partition of $V$ into $V_1, \dots, V_k$ maximizing
+$$\sum_{i=1}^k \sum_{j = 1}^{i-1} |E(V_i, V_j)| := c(\{V_1, \dots, V_k\}).$$
+Intuitively we are maximizing the edges between the selected cuts.
+Let $\cB$ denote the $2$-approximation algorithm for max cut given in problem $3a$.
+Consider the following algorithm, which we will denote by $\cA$:
+\begin{enumerate}
+\item Set $V_1 = S$ and $V_2 = V\backslash S$ where $S$ is the cut returned by $\cB$ when run on $G$.
+\item Set $V_i = \emptyset$ for $i = 3, \dots, k$.
+\item For $i = 2, \dots, k-1$
+\begin{enumerate}
+\item For $j = 1, \dots, i$ set $S_j$ to the cut returned by $\cB$ when run on $G[V_j]$.
+\item Set $j^* = \text{argmax}_{j=1,\dots, i} \{|E(S_j, V_j\backslash S_j)|\}$.
+\item Set $V_{i+1} = S_{j^*}$ and set $V_{j^*} = V_{j^*} \backslash S_{j^*}$.
+\end{enumerate}
+\item Return $V_1, \dots, V_k$.
+\end{enumerate}
+The idea behind the operation of $\cA$ is to start with a $2$-cut of $G$, and in every successive iteration take the current partition and $2$-cut one of the parts in an approximately optimal way until we reach a $k$-cut.
+\begin{lemma}\label{lemma:3b1}
+Let $\cO = \{V^*_1, \dots, V^*_k\}$.
+Consider an arbitrary iteration $i = 1, \dots, k-1$ (iteration $1$ is original $2$-cut).
+Let $\cS = \{V_1, \dots, V_{i+1}\}$ be the set of partitions chosen so far by $\cA$.
+Suppose that at the end of iteration $i$, the algorithm has split $V_j$ into $V_j'$ and $V_{i+1}$.
+That is the solution after iteration $i$ is
+$$\cS' = \cS \backslash V_j \cup \{V_j', V_{i+1}\}.$$
+Then 
+$$2k(c(\cS') - c(\cS)) \geq c(\cO) - c(\cS).$$
+\end{lemma}
+\begin{proof}
+The difference between $c(\cS')$ and $c(\cS)$ is precisely $|E(V_j', V_{i+1})|$ (the number of edges gained by splitting $V_j$).
+Further the difference $c(\cO) - c(\cS)$ is upper-bounded by the edges counted in $c(\cO)$ that are not counted in $c(\cS)$.
+Call such edges $E'$.
+The edges $E'$ are necessarily internal to a part of $\cS$ (that is, both endpoints lie in some $V_k \in \cS$).
+Otherwise they are counted in $c(\cS)$ since their endpoints lie in distinct parts of $\cS$.
+Thus if we consider the $k$-cut induced by $\cO$ on $G[E']$ it consists of at most $k$ splits ($2$-cuts) of parts $V_k \in \cS$.
+\paragraph{}
+Let $V'_k$ denote the best possible $2$-cut of $V_k \in \cS$ for all $k = 1, \dots, i$. Now let $$k^* = \text{argmax}_{k=1,\dots,i}  \{|E(V'_k, V_k \backslash V'_k)|\}.$$ We have
+$$ |E'| \leq k |E(V'_{k^*}, V_k \backslash V'_{k^*})|$$
+by our previous discussion, since $E'$ edges are formed by a $k$-cut on edges internal to parts of $\cS$, and  $|E(V'_{k^*}, V_k \backslash V'_{k^*}|$ is largest $2$-cut to be formed by edges internal to a part of $\cS$.
+Now the cut found by our algorithm $\cA$ in this iteration in step $3(c)$ is a $2$-approximation of the best possible split cut.
+That is,
+$$|E(V'_{k^*}, V_k \backslash V'_{k^*})| \leq 2|E(V_j', V_{i+1})| .$$
+Hence, combining these inequalities, we see that
+$$|E'| \leq 2k |E(V_j', V_{i+1})|.$$
+Therefore $2k(c(\cS') - c(\cS)) \geq c(\cO) - c(\cS)$ as desired.
+\end{proof}
+\paragraph{}
+Using the previous lemma, we can demonstrate that $\cA$ is a $\alpha := 1-(1-\frac{1}{2k})^k$-approximation algorithm for the max $k$-cut problem.
+\paragraph{Main Proof}
+Since $\cB$ can be run in polynomial time, and $k$ is at most $|V|$, it is clear that $\cA$ runs in polynomial time.
+At termination the algorithm returns a partition of $V$ into $k$ parts so it returns a feasible solution.
+\paragraph{} 
+Now to see the approximation factor holds let $\cS^i$ denote the solution maintained by $\cA$ at the end of iteration $i$.
+Then the returned solution is $\cS^{k-1}$.
+Let $\cO$ be an optimal solution.
+Observe from Lemma \ref{lemma:3b1} that $c(\cS^{i}) \geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{i-1})$.
+We will apply this inequality inductively to achieve the bound:
+\begin{align*}
+c(\cS^{k-1} &\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-2}) \\
+		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(\frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-3})) \\
+		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(1 + (1-\frac{1}{2k}) + (1-\frac{1}{2k})^2 + \dots + (1-\frac{1}{2k})^{k-1}) \\
+		&= \frac{c(\cO)}{2k} \cdot \frac{1-(1-\frac{1}{2k})^k}{1-(1-\frac{1}{2k})} \\
+		&=\frac{c(\cO)}{2k} \cdot \frac{1-(1-\frac{1}{2k})^k}{\frac{1}{2k}}\\
+		&= 1-(1-\frac{1}{2k})^kc(\cO).
+\end{align*}
+Thus the approximation factor holds as desired. $\blacksquare$
 \end{document}

### 2017-01-30 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Fixed 3b of apx algs a1

@@
 \begin{document}
+%Q1
 \section{}
@@
 For a subset $\bar{E} \subseteq E$ we denote by $G[\bar{E}]$ the subgraph of $G$ induced by the edges in $\bar{E}$.
+	%Pa
 \subsection{a}
@@
 With the last inequality following since $M$ is the domain of $f$. Hence we have $|M| \geq \frac{1}{2}|\bar{M}|$ as desired. $\blacksquare$
+	%Pb
 \subsection{b}
@@
 Let $e$
+
+%Q2
+\section{}
+
+%Q3
+\section{}
+\paragraph{}
+Let $G = (V,E)$ be a graph. For any $S, T \subseteq V$ we denote by $E(S,T)$ the set of edges between $S$ and $T$. Formally $E(S,T) = \{\{s,t\} \in E: s \in S, t \in T\}$.
+%Pa
+\subsection{a}
+
+%Pb
+\subsection{b}
+\paragraph{}
+Our goal is to find a maximum $k$-Cut of $G$. That is, we seek to find a partition of $V$ into $V_1, \dots, V_k$ maximizing
+$$\sum_{i=1}^k \sum_{j = 1}^{i-1} |E(V_i, V_j)| := c(\{V_1, \dots, V_k\}).$$
+Intuitively we are maximizing the edges between the selected cuts.
+Let $\cB$ denote the $2$-approximation algorithm for max cut given in problem $3a$.
+Consider the following algorithm, which we will denote by $\cA$:
+\begin{enumerate}
+\item Set $V_1 = S$ and $V_2 = V\backslash S$ where $S$ is the cut returned by $\cB$ when run on $G$.
+\item Set $V_i = \emptyset$ for $i = 3, \dots, k$.
+\item For $i = 2, \dots, k-1$
+\begin{enumerate}
+\item For $j = 1, \dots, i$ set $S_j$ to the cut returned by $\cB$ when run on $G[V_j]$.
+\item Set $j^* = \text{argmax}_{j=1,\dots, i} \{|E(S_j, V_j\backslash S_j)|\}$.
+\item Set $V_{i+1} = S_{j^*}$ and set $V_{j^*} = V_{j^*} \backslash S_{j^*}$.
+\end{enumerate}
+\item Return $V_1, \dots, V_k$.
+\end{enumerate}
+The idea behind the operation of $\cA$ is to start with a $2$-cut of $G$, and in every successive iteration take the current partition and $2$-cut one of the parts in an approximately optimal way until we reach a $k$-cut.
+\begin{lemma}\label{lemma:3b1}
+Let $\cO = \{V^*_1, \dots, V^*_k\}$.
+Consider an arbitrary iteration $i = 1, \dots, k-1$ (iteration $1$ is original $2$-cut).
+Let $\cS = \{V_1, \dots, V_{i+1}\}$ be the set of partitions chosen so far by $\cA$.
+Suppose that at the end of iteration $i$, the algorithm has split $V_j$ into $V_j'$ and $V_{i+1}$.
+That is the solution after iteration $i$ is
+$$\cS' = \cS \backslash V_j \cup \{V_j', V_{i+1}\}.$$
+Then 
+$$2k(c(\cS') - c(\cS)) \geq c(\cO) - c(\cS).$$
+\end{lemma}
+\begin{proof}
+The difference between $c(\cS')$ and $c(\cS)$ is precisely $|E(V_j', V_{i+1})|$ (the number of edges gained by splitting $V_j$).
+Further the difference $c(\cO) - c(\cS)$ is upper-bounded by the edges counted in $c(\cO)$ that are not counted in $c(\cS)$.
+Call such edges $E'$.
+The edges $E'$ are necessarily internal to a part of $\cS$ (that is, both endpoints lie in some $V_k \in \cS$).
+Otherwise they are counted in $c(\cS)$ since their endpoints lie in distinct parts of $\cS$.
+Thus if we consider the $k$-cut induced by $\cO$ on $G[E']$ it consists of at most $k$ splits ($2$-cuts) of parts $V_k \in \cS$.
+\paragraph{}
+Let $V'_k$ denote the best possible $2$-cut of $V_k \in \cS$ for all $k = 1, \dots, i$. Now let $$k^* = \text{argmax}_{k=1,\dots,i}  \{|E(V'_k, V_k \backslash V'_k)|\}.$$ We have
+$$ |E'| \leq k |E(V'_{k^*}, V_k \backslash V'_{k^*})|$$
+by our previous discussion, since $E'$ edges are formed by a $k$-cut on edges internal to parts of $\cS$, and  $|E(V'_{k^*}, V_k \backslash V'_{k^*}|$ is largest $2$-cut to be formed by edges internal to a part of $\cS$.
+Now the cut found by our algorithm $\cA$ in this iteration in step $3(c)$ is a $2$-approximation of the best possible split cut.
+That is,
+$$|E(V'_{k^*}, V_k \backslash V'_{k^*})| \leq 2|E(V_j', V_{i+1})| .$$
+Hence, combining these inequalities, we see that
+$$|E'| \leq 2k |E(V_j', V_{i+1})|.$$
+Therefore $2k(c(\cS') - c(\cS)) \geq c(\cO) - c(\cS)$ as desired.
+\end{proof}
+\paragraph{}
+Using the previous lemma, we can demonstrate that $\cA$ is a $\alpha := 1-(1-\frac{1}{2k})^k$-approximation algorithm for the max $k$-cut problem.
+\paragraph{Main Proof}
+Since $\cB$ can be run in polynomial time, and $k$ is at most $|V|$, it is clear that $\cA$ runs in polynomial time.
+At termination the algorithm returns a partition of $V$ into $k$ parts so it returns a feasible solution.
+\paragraph{} 
+Now to see the approximation factor holds let $\cS^i$ denote the solution maintained by $\cA$ at the end of iteration $i$.
+Then the returned solution is $\cS^{k-1}$.
+Let $\cO$ be an optimal solution.
+Observe from Lemma \ref{lemma:3b1} that $c(\cS^{i}) \geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{i-1})$.
+We will apply this inequality inductively to achieve the bound:
+\begin{align*}
+c(\cS^{k-1} &\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-2}) \\
+		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(\frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-3})) \\
+		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(1 + (1-\frac{1}{2k}) + (1-\frac{1}{2k})^2 + \dots + (1-\frac{1}{2k})^{k-1}) \\
+		&= \frac{c(\cO)}{2k} \cdot \frac{1-(1-\frac{1}{2k})^k}{1-(1-\frac{1}{2k})} \\
+		&=\frac{c(\cO)}{2k} \cdot \frac{1-(1-\frac{1}{2k})^k}{\frac{1}{2k}}\\
+		&= 1-(1-\frac{1}{2k})^kc(\cO).
+\end{align*}
+Thus the approximation factor holds as desired. $\blacksquare$
 \end{document}

### 2017-01-30 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
3a

@@
 \subsection{a}
-
+\paragraph{}
+The fact that the Greedy Algorithm runs in polynomial time is obvious.
+$|E(v,S)|$ can be computed in $O(|E|)$ time and the while loop runs for $O(|V|)$ iterations.
+Further it clearly returns a cut.
+It remains to verify the approximation factor of $\frac{1}{2}$ holds.
+\paragraph{}
+We show that the approximation factor is maintained throughout operation of the algorithm on the graph $G[V\backslash \bar{V}]$ induced by vertices considered so far.
+After the first iteration only one vertex has been considered and the approximation factor holds trivially on the graph induced by that one vertex.
+Now for induction let $V' = V \backslash \bar{V}$ be the set of vertices considered so far by the Greedy Algorithm, and at the start of the next iteration the vertex $v \in V$ is being considered.
+Let $S \subseteq V$ be the cut set considered so far, and let $S^*$ be the optimal cut set on $G[V']$.
+Let $\bar{S} = V'\backslash S$ and $\bar{S^*} = V' \backslash S^*$.
+We may assume without loss of generality that the Greedy Algorithm puts $v \in S$ and the optimal solution puts $v \in S^*$ after this iteration (if this does not hold simply relabel $S$ with $\bar{S}$ below, or $S^*$ with $\bar{S^*}$ with respect to which assumption does not hold).
+\paragraph{}
+So after this iteration the optimal solution has value
+$$|E(S^* \cup \{v\}, \bar{S^*})| =  \sum_{u \in S^* \cup \{v\}} |E(u, \bar{S^*})| = |E(v,\bar{S^*})| + \sum_{u \in S^*} |E(u, \bar{S^*})| \leq |E(v,\bar{S^*})| + 2\sum_{u \in S} |E(u, \bar{S})|$$
+with the inequality following by induction. Hence the approximation factor will hold as desired provided that
+$$|E(v,\bar{S^*})| \leq 2|E(v,\bar{S})|.$$
+Suppose for a contradiction that
+$$|E(v,\bar{S^*})| > 2|E(v,\bar{S})|.$$
+We observe that
+$$|E(v, \bar{S^*})| = |E(v,\bar{S^*}\backslash S)| + |E(v, \bar{S^*}\cap S)| \leq  |E(v, \bar{S})| + |E(v, S)|.$$
+The inequality follows since $\bar{S^*}\backslash S \subseteq V' \backslash S = \bar{S}$, and $\bar{S^*} \cap S \subseteq S$. Combining this inequality with the contradiction assumption we see
+$$2|E(v,\bar{S})| <|E(v, \bar{S})| + |E(v, S)|$$
+and subtracting $|E(v, \bar{S})|$ from both sides yields
+$$ |E(v,\bar{S})| < |E(v,S)|.$$
+This contradicts our Greedy choice $v \in S$ since such choice implies
+$$ |E(v,\bar{S})| \geq |E(v,S)|.$$
+Hence we have
+$$|E(v,\bar{S^*})| \leq 2|E(v,\bar{S})|.$$
+Thus after this iteration the optimal solution has value:
+$$|E(S^* \cup \{v\}, \bar{S^*})|  \leq |E(v,\bar{S^*})| + 2\sum_{u \in S} |E(u, \bar{S})| \leq 2|E(v,\bar{S})| + 2\sum_{u \in S} |E(u, \bar{S})| = 2 |E(S \cup \{v\}, \bar{S})|.$$
+Therefore $|E(S \cup \{v\}, \bar{S})| \geq \frac{1}{2} |E(S^* \cup \{v\}, \bar{S^*})|$ as desired. 
+\paragraph{}
+Now we observe that, from the invariant we just demonstrated, upon termination of the algorithm the approximation factor holds for the greedy solution versus the optimal solution on $G[V \backslash \emptyset] = G$, and hence the Greedy Algorithm is $\frac{1}{2}$-approximation algorithm.$\blacksquare$
 %Pb

### 2017-01-30 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Written up 1-3 of A1 in apx algs

@@
 \subsection{b}
+(This argument follows the exact same lines as the argument for problem $2$)
 \paragraph{}
-Let $e$
-
+We can decided in polynomial time if $M\cup \{e\}$ is a matching in $\cS$ by iterating over $V$, counting the number of edges in $M\cup\{e\}$ which cover each vertex in $V$, and returning ``No" is the count is greater than $1$ for any vertex and ``Yes" otherwise.
+Thus the algorithm clearly runs in polynomial time and returns a feasible set.
+\paragraph{}
+Let $S$ be the set of edges returned by the greedy algorithm.
+Order the edges as $S = \{e_1, \dots, e_n\}$ in the order they were taken during the operation of $\cA$.
+Let $F_i := E\backslash \tilde{E}$ be the edges removed from $E$ at the edge of iteration $i$ of $\cA$ (the iteration where $e_i$ is added to $S$).
+Let $S^*$ be the optimal solution and partition $S^* = S^*_1 \cup \dots \cup S^*_n$ where $S^*_i = S^* \cap F_i$.
+\paragraph{}
+We want that $|S^*_i| \leq 2$ to hold, but we may need to redistribute some edges to ensure that this happens (and we will do this without effecting our ability to bound the optimal solution).
+Run the following algorithm, denoted $\cB$, to redistribute the edges in our partition of $S^*$:
+\begin{enumerate}
+\item For $i = 1, \dots, n$
+\begin{enumerate}
+\item While $|S^*_i| > 2$: take an edge $e$ from $S^*_i$ and redistribute $e$ to some $S^*_j$ for $j<i$ with $|S^*_j| < 2$.
+\end{enumerate}
+\end{enumerate}
+\paragraph{Claim 1}
+After the operation of $\cB$, for all $i = 1, \dots, n$, $|S^*_i| \leq 2$ and $w_{e_i} \geq w_e$ for all $e \in S^*_i$.
+\paragraph{Proof of Claim 1}
+We proceed by induction.
+For $e_1$ $\{e_1\}$ is inclusion-wise maximal on $\cS[F_1]$ (by definition of $F_1$), and $S_1^*$ is feasible on $\cS[F_1]$ since $S^*_1 \subseteq F_1$ and $S^* \in \cS$.
+Thus we have $|\{e_1\}| \geq \frac{1}{2} |S_1^*|$, and hence
+$$|S^*_1| \leq \alpha.$$
+Further by our greedy choice of $e_1$, $w_{e_1} \geq w_e$ for all $e \in F_1$ (and hence for all $e \in S^*_1$).
+\paragraph{}
+Suppose that the claim holds for $S^*_1 ,\dots, S^*_i$.
+Now by our greedy choice, and induction, $w_{e_{i+1}} \geq w_{e_j} \geq w_e$ for all $e \in S^*_j$ for all $j = 1,\dots, i$.
+So when $\cB$ redistributes edges of $S^*_{i+1}$ the cost property on the earlier sets is not violated.
+Also by our greedy choice, any edges that were not moved from $S^*_{i+1}$ by $\cB$ have cost at most $w_{e_{i+1}}$.
+To finish the proof of the claim, we demonstrate that while $|S_{i+1}^*| > \alpha$ there exists some $S^*_j$ with $j\leq i$ and $|S^*_j| < 2$.
+Observe that $\{e_1, \dots, e_{i+1}\}$ is inclusion-wise maximal on $\cS[F_1 \cup \dots \cup F_{i+1}]$ by definition of $F_1 \cup \dots \cup F_{i+1}$.
+Also $\bigcup_{j=1}^{i+1} S^*_j \in \cS[F_1 \cup \dots \cup F_{i+1}]$ so we have that
+$$|\{e_1, \dots, e_{i+1}\}| \geq \frac{1}{2} |\bigcup_{j=1}^{i+1} S^*_j | = \frac{1}{2} \sum_{j=1}^{i+1} |S^*_j|.$$
+Thus we see that
+$$ \frac{1}{i+1}\sum_{j=1}^{i+1} |S^*_j| \leq 2.$$
+Now since the desired bound holds on average, if $|S^*_{i+1}| > 2$ there exists some $S^*_j$ with $j \leq i$ and $|S^*_j| < 2$ (otherwise the bound does not hold).
+Thus we have proven Claim $1$.
+$\blacksquare$
+\paragraph{}
+With Claim $1$ in hand we are prepared to compute the desired approximation factor:
+\begin{align*}
+w(S^*) &= \sum_{i=1}^n c(S^*_i) \\
+&\leq \sum_{i=1}^n w_{e_i} |S^*_i| \\
+&\leq \sum_{i=1}^n w_{e_i} 2 \\
+&= 2 w(S).
+\end{align*}
+Thus $w(S) \geq \frac{1}{2} w(S^*)$ and hence the approximation factor holds as desired. $\blacksquare$
+	%Pc
+	\subsection{c}
+	\paragraph{}
+	Suppose for a contradiction that the Greedy Algorithm is in fact a $(\frac{1}{2} + \epsilon)$-approximation algorithm for some $\epsilon \in (0, \frac{1}{2}]$.
+Then consider the following example.
+Let $G$ be the path graph with vertex set $V(G) = \{a,b,c,d\}$ and edge set $E(G) = \{ab, bc,cd\}$.
+Define the weight function $w : E(G) \rightarrow \R_+$ by
+	$$w_{ab} = w_{cd} = 1 \quad\text{and}\quad w_{bc} = 1+\epsilon.$$
+	Then the Greedy algorithm would choose the edge $bc$ as its solution.
+This solution has cost $1+\epsilon$.
+The optimal solution is to take edges $ab$ and $cd$ with cost $2$.
+The ratio of these solutions is
+	$$ \frac{1+\epsilon}{2} = \frac{1}{2} + \frac{\epsilon}{2} < \frac{1}{2} + \epsilon$$
+	Contradicting that the Greedy Algorithm is a $(\frac{1}{2} + \epsilon)$-approximation algorithm. $\blacksquare$
 %Q2
 \section{}
-
+\paragraph{}
+Let $E$ be a set of $m$ elements and $\cS$ a collection of sets $S \subseteq E$.
+Let $w: E\rightarrow \R_+$ be a weight function.
+Let $\cS[\bar{E}] := \{ S \in \cS : S \subseteq \bar{E}\}$.
+Suppose that $\cS$ satisfies the following for a given $\alpha \geq 1$:
+\begin{enumerate}
+\item $S \in \cS$ implies $S' \in \cS$ for any $S' \subseteq S$
+\item For all $\bar{E} \subseteq E$, if $S$ is an inclusion-wise maximal set in $\cS[\bar{E}]$ then $|S| \geq \frac{1}{\alpha} |\bar{S}|$ for all $\bar{S} \in \cS[\bar{E}]$.
+\end{enumerate}
+We will show the Greedy Algorithm, denoted $\cA$, is a $\frac{1}{\alpha}$-approximation algorithm for the problem of finding a maximum weight $S \in \cS$.
+\paragraph{}
+Since we have a polynomial time oracle for membership in $\cS$ the algorithm clearly runs in polynomial time and returns a feasible set.
+\paragraph{}
+Let $S$ be the set of edges returned by the greedy algorithm.
+Order the edges as $S = \{e_1, \dots, e_n\}$ in the order they were taken during the operation of $\cA$.
+Let $F_i := E\backslash \tilde{E}$ be the edges removed from $E$ at the edge of iteration $i$ of $\cA$ (the iteration where $e_i$ is added to $S$).
+Let $S^*$ be the optimal solution and partition $S^* = S^*_1 \cup \dots \cup S^*_n$ where $S^*_i = S^* \cap F_i$.
+\paragraph{}
+We want that $|S^*_i| \leq \alpha$ to hold, but we may need to redistribute some edges to ensure that this happens (and we will do this without effecting our ability to bound the optimal solution).
+Run the following algorithm, denoted $\cB$, to redistribute the edges in our partition of $S^*$:
+\begin{enumerate}
+\item For $i = 1, \dots, n$
+\begin{enumerate}
+\item While $|S^*_i| > \alpha$: take an edge $e$ from $S^*_i$ and redistribute $e$ to some $S^*_j$ for $j<i$ with $|S^*_j| < \alpha$.
+\end{enumerate}
+\end{enumerate}
+\paragraph{Claim 1}
+After the operation of $\cB$, for all $i = 1, \dots, n$, $|S^*_i| \leq \alpha$ and $w_{e_i} \geq w_e$ for all $e \in S^*_i$.
+\paragraph{Proof of Claim 1}
+We proceed by induction.
+For $e_1$ $\{e_1\}$ is inclusion-wise maximal on $\cS[F_1]$ (by definition of $F_1$), and $S_1^*$ is feasible on $\cS[F_1]$ since $S^*_1 \subseteq F_1$ and $S^* \in \cS$.
+Thus we have $|\{e_1\}| \geq \frac{1}{\alpha} |S_1^*|$, and hence
+$$|S^*_1| \leq \alpha.$$
+Further by our greedy choice of $e_1$, $w_{e_1} \geq w_e$ for all $e \in F_1$ (and hence for all $e \in S^*_1$).
+\paragraph{}
+Suppose that the claim holds for $S^*_1 ,\dots, S^*_i$.
+Now by our greedy choice, and induction, $w_{e_{i+1}} \geq w_{e_j} \geq w_e$ for all $e \in S^*_j$ for all $j = 1,\dots, i$.
+So when $\cB$ redistributes edges of $S^*_{i+1}$ the cost property on the earlier sets is not violated.
+Also by our greedy choice, any edges that were not moved from $S^*_{i+1}$ by $\cB$ have cost at most $w_{e_{i+1}}$.
+To finish the proof of the claim, we demonstrate that while $|S_{i+1}^*| > \alpha$ there exists some $S^*_j$ with $j\leq i$ and $|S^*_j| < \alpha$.
+Observe that $\{e_1, \dots, e_{i+1}\}$ is inclusion-wise maximal on $\cS[F_1 \cup \dots \cup F_{i+1}]$ by definition of $F_1 \cup \dots \cup F_{i+1}$.
+Also $\bigcup_{j=1}^{i+1} S^*_j \in \cS[F_1 \cup \dots \cup F_{i+1}]$ so we have that
+$$|\{e_1, \dots, e_{i+1}\}| \geq \frac{1}{\alpha} |\bigcup_{j=1}^{i+1} S^*_j | = \frac{1}{\alpha} \sum_{j=1}^{i+1} |S^*_j|.$$
+Thus we see that
+$$ \frac{1}{i+1}\sum_{j=1}^{i+1} |S^*_j| \leq \alpha.$$
+Now since the desired bound holds on average, if $|S^*_{i+1}| > \alpha$ there exists some $S^*_j$ with $j \leq i$ and $|S^*_j| < \alpha$ (otherwise the bound does not hold).
+Thus we have proven Claim $1$.
+$\blacksquare$
+\paragraph{}
+With Claim $1$ in hand we are prepared to compute the desired approximation factor:
+\begin{align*}
+w(S^*) &= \sum_{i=1}^n c(S^*_i) \\
+&\leq \sum_{i=1}^n w_{e_i} |S^*_i| \\
+&\leq \sum_{i=1}^n w_{e_i} \alpha \\
+&= \alpha w(S).
+\end{align*}
+Thus $w(S) \geq \frac{1}{\alpha} w(S^*)$ and hence the approximation factor holds as desired. $\blacksquare$
 %Q3

### 2017-01-31 CO750-ApproximationAlgorithmsTopics/lectures/lecture_08.tex
Scribe lecture 7

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-mm-DD}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 8} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\paragraph{}
This lecture is a continuation of the previous lecture. We are proving the following theorem:
\begin{theorem}
For any xps $x$ to $LP$ there is a collection of $m$ tight sets such that their incidence vectors are linearly independent, and the collection is a laminar family.
\end{theorem}
\begin{proof}
Let $\cL$ be a laminar family of tight sets whose incidence vectors are linearly independent. We will prove through the following lemmas that if $span(\cL) \in \R^m$ then there exists a tight set we can add to $\cL$, and if this set violates laminarity then we can ``uncross" and recover a larger laminar set than $\cL$ as desired.
\end{proof}
\begin{lemma}
Let $A$ and $B$ be two crossing tight sets. The one of the following holds:
\begin{enumerate}
\item $A\backslash B$ and $B \backslash A$ are both tight and $\chi(A) + \chi(B) = \chi(A\backslash B) + \chi(B\backslash A)$.
\item $A\cup B$ and $A\cap B$ are both tight and $\chi(A) + \chi(B) = \chi(A\cup B) + \chi(B\cap A)$.
\end{enumerate}
\end{lemma}
\begin{proof}
The proof follows by case distinction on $f$. Since $f$ is weakly supermodular (wsm) either:
\begin{enumerate}
\item $f(A) + f(B) \leq f(A\backslash B) + f(B\backslash A)$ or
\item $f(A) + f(B) \leq f(A\cup B) + f(A \cap B)$.
\end{enumerate}
We will prove the first case, and the second case follows similarly.
\paragraph{}
Since $A$ and $B$ are tight
$$x(\delta(A)) + x(\delta(B)) = f(A) + f(B).$$
By feasibility
$$x(\delta(A\backslash B)) + x(\delta(B\backslash A)) \geq f(A\backslash B) + f(B\backslash A).$$
However $\delta(A) \cup \delta(B) \supseteq \delta(A\backslash B) + \delta(B \backslash A)$, so
$$x(delta(A)) + x(\delta(B)) \geq x(\delta(A\backslash B)) + x(\delta(B\backslash A).$$
Combining inequalities we see
$$f(A\backslash B) + f(B\backslash A) \leq x(\delta(A\backslash B)) + x(\delta(B\backslash A) \leq f(A\backslash B) + f(B\backslash A) $$
Hence $A\backslash B$ and $B \backslash A$ are tight.
\paragraph{}
Further by our chain of inequalities holding as equality we have
$$x(delta(A)) + x(\delta(B)) = x(\delta(A\backslash B)) + x(\delta(B\backslash A)$$
and since all $e$ have $x_e > 0$, we know that there are no edges from $A\cap B$ to $V\backslash (A\cup B)$, otherwise the above equality would be violated. Hence
$$\chi(A) + \chi(B) = \chi(A\backslash B) + \chi(B\backslash A).$$
\end{proof}
\paragraph{}
For a set $S \subseteq V$ we define the $\textit{crossing number}$ of $S$ with respect to $\cL$ to be the number of sets in $\cL$ that cross with $S$.
\begin{lemma}
Let $S$ be a tight set crossing with $T \in \cL$. Then each of the sets $S\backslash T$, $T\backslash S$. $S\cup T$, and $S\cap T$ have a crossing number smaller than that of $S$.
\end{lemma}
\begin{proof}
Let $T'$ be any set in $\cL$ crossing with any of the four sets. One checks that $T'$ crosses $S$ too. This implies that each of those four sets do not cross with $T$.
\end{proof}
\begin{lemma}
Let $S$ be a tight set such that $\chi(S)\not\in span(\cL)$, and $S$ crosses with some set in $\cL$. Then there is a tight set $S'$ that has a smaller crossing number than $S$ and $\chi(S')\not\in span(\cL)$.
\end{lemma}
\begin{proof}
We again proceed by cases on $f$ and only prove the first one, with the second following similarly.
\begin{enumerate}
\item $f(A) + f(B) \leq f(A\backslash B) + f(B\backslash A)$ or
\item $f(A) + f(B) \leq f(A\cup B) + f(A \cap B)$.
\end{enumerate}
\paragraph{}
By prior lemmas $S\backslash T$ and $T\backslash S$ have smaller crossing numbers than $S$, and further $S\backslash T$ and $T\backslash S$ are tight with
$$\chi(S) + \chi(T) = \chi(S\backslash T) + \chi(T\backslash S).$$
Hence one of  $\chi(T\backslash S)$ or $\chi(S\backslash T)$ are not in $span(\cL)$, as otherwise $\chi(S) \in span(\cL)$.
\end{proof}
\paragraph{}
Together these lemmas yield the theorem. Now with this theorem in hand we are ready to prove final unproven result from the previous lecture:
\begin{theorem}
For any wsm function $f$, any xps to $LP$ satisfies $x_e \geq \frac{1}{2}$ for at least one $e \in E$.
\end{theorem}
\begin{proof}
We have $m =|E|$ with $x_e >0$ for all $e\in E$, and $m = |\cL|$. Suppose for a contradiction that $x_e \leq \frac{1}{2}$ for all $e \in E$. We proceed by a token argument.
\paragraph{}
Assign $1$ token to every $e \in E$. For each $uv \in E$ distribute $x_{uv}$ units to each of $u$ and $v$, and distribute $1-2x_{uv}$ units to $uv$. Now reassign these values to sets in $\cL$ as follows. Let $S \in \cL$ and let $S_1,\dots, S_n \in \cL$ be $S$'s children. Ie sets strictly contained in $S$:
\begin{itemize}
\item if $v \in S\backslash \cup_i S_i$ give $S$ the value held by $v$.
\item if $uv \in E(S) \backslash \cup_i E(S_i)$ give $S$ the value held by $uv$.
\end{itemize}
Our goal is to show that after this process $S$ gets value at least one. We need some definitions:
\begin{itemize}
\item Let $C_0$ be set of edges with both endpoints in $S\backslash \cup_i S_i$.
\item Let $C_1$ be the set of edges with one endpoint in $S$ and one in $V\backslash S$.
\item Let $C_2$ be the set of edges with one endpoint in $S\backslash \cup_i S_i$ and one endpoint in some $S_i$.
\item Let $C_3$ be the set of edges with one endpoint in some $S_i$ and the other in some $S_j$ for $i\neq j$.
\end{itemize}
Observe that if $C_0 \neq \emptyset$ then $S$ trivially obtains value at least one. So suppose that $C_0 = \emptyset$. Now each $S$ gets value
$$z_S \geq \sum_{e\in C_1} x_e + \sum_{e \in C_2}(x_e + (1-2x_e)) + \sum_{e \in C_3} (1-2x_e) = x(C_1) + |C_2| -x(C_2) + |C_3| - 2x(C_3).$$
Note that since each $x_e \leq \frac{1}{2}$ this value $z_S > 0$ unless $C_1 = C_2 = C_3 = \emptyset$ which cannot happen as this would imply that $\chi(S) = \sum_i \chi(S_i)$ which contradicts linear independence.
\paragraph{}
To complete the proof we need only demonstrate that $z_S$ is integral. To see this observe by tightness of the sets in $\cL$ that
$$x(C_1) -x(C_2) -2(C_3) = x(\delta(S)) - \sum_i x(\delta(S_i)) = f(S) - \sum_i f(S_i)$$
and since $f$ is integer valued the result holds. Therefore $z_S \geq 1$ for all $S$. Now observe that no set in $\cL$ will receive value from $uv \in \delta(S)$ where $S\in \cL$ is maximal with respect to $\cL$. This implies $|E| > m$ a contradiction.
\end{proof}
\end{document}

### 2017-02-01 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
partial question 4 of apx algs

@@
 Thus the approximation factor holds as desired. $\blacksquare$
+%Q4
+\section{}
+\paragraph{}
+Let $G = (V,E)$ be a graph with edge costs $c: E \rightarrow \R_+$.
+Let $r \in V$ be a root vertex.
+Let $R \subseteq V \backslash \{r\}$ be a set of terminals.
+Let $d_v \in \{2^0, 2^1, \dots, 2^\delta\}$ denote the demand of $v$ for each $v \in R$.
+We assume we have an $O(1)$-approximation algorithm for the Steiner Tree problem.
+Then each iteration of step $2$ of the given algorithm runs in polynomial time using the assumed approximation algorithm.
+There are $\delta+1$ iterations of step $2$ which is $O(\log \max_v d_v)$ and hence polynomial in the number of bits needed to specify the input to the problem.
+Thus the entirety of step $2$ runs in polynomial time and hence the given algorithm runs in polynomial time.
+It is clear the tree returned spans $R \cup \{r\}$ since the partial solution at each iteration $i$ spans $R_\delta \cup \dots \cupR_i \cup \{r\}$.
+Further at each iteration $i$ the paths in the final solution connecting $v \in R_i$ to $r$ are assigned costs $2^i$ on their edges and hence a feasible flow of value $d_v$ from $r$ to $v$ is possible.
+Thus the algorithm returns a feasible solution in polynomial time.
+\paragraph{}
+It remains to verify the approximation factor of $O(1)$. We do so by induction on the number of iterations the algorithm needs to run to return a solution.
 \end{document}

### 2017-02-01 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
p4

@@
 Thus the entirety of step $2$ runs in polynomial time and hence the given algorithm runs in polynomial time.
-It is clear the tree returned spans $R \cup \{r\}$ since the partial solution at each iteration $i$ spans $R_\delta \cup \dots \cupR_i \cup \{r\}$.
+It is clear the tree returned spans $R \cup \{r\}$ since the partial solution at each iteration $i$ spans $R_\delta \cup \dots \cup R_i \cup \{r\}$.
 Further at each iteration $i$ the paths in the final solution connecting $v \in R_i$ to $r$ are assigned costs $2^i$ on their edges and hence a feasible flow of value $d_v$ from $r$ to $v$ is possible.
@@
 \paragraph{}
-It remains to verify the approximation factor of $O(1)$. We do so by induction on the number of iterations the algorithm needs to run to return a solution.
+It remains to verify the approximation factor of $O(1)$.
+We do so by induction on the number of iterations the algorithm needs to run to return a solution.
+In the base case, that is when $\delta = 0$, the algorithm returns an $O(1)$-approximate Steiner Tree connecting $R \cup \{r\}$ and installs capacity of $1$ on each edge.
+By observing that the optimal solution in this case is a Steiner Tree connecting $R \cup \{r\}$ with capacity $1$ on each edge we see that the approximation factor holds in this case.
+\paragraph{}
+Suppose that $\delta > 0$ and for any problem instance with maximum demand $2^{\alpha}$ with $0 \leq \alpha < \delta$ the algorithm returns an $O(1)$ approximate solution.
+We will describe solutions by an ordered pair $(T, x)$ where $T$ is a tree and $x : T \rightarrow \R_+$ maps edges of $T$ with their corresponding capacity.
+Then the objective function, which we denote by $f$, in our problem maps feasible solutions $(T,x)$ as follows:
+$$f(T,x) = \sum_{e \in T} c_e x_e.$$
+\paragraph{}
+Let $(T,x)$ be the solution returned by running our algorithm, and let $(T^*, x^*)$ be the optimal solution.
+Let $T_{\delta}$ be the $O(1)$-approximate Steiner Tree connecting $R_\delta \cup \{r\}$ in the first iteration (iteration $\delta$) of our algorithm.
+Let $\bar{T}$ denote the solution returned by our algorithm running on the residual problem with input graph $G_{\delta-1}$
 \end{document}

### 2017-02-02 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
p4 of apx algs

@@
 Let $T_{\delta}$ be the $O(1)$-approximate Steiner Tree connecting $R_\delta \cup \{r\}$ in the first iteration (iteration $\delta$) of our algorithm.
-Let $\bar{T}$ denote the solution returned by our algorithm running on the residual problem with input graph $G_{\delta-1}$
+Let $(\bar{T}, x)$ denote the solution returned by our algorithm running on the residual problem with input graph $G_{\delta-1}$ (the resulting graph after contracting $T_{\delta}$ into $r$).
+Then $T = T_\delta \cup \bar{T}$ and
+$$f(T,x) = f(T_\delta, x) + f(\bar{T},x).$$
+If we let $(\bar{T}^*, \bar{x}^*)$ denote the optimal solution for the residual problem with graph $G_{\delta - 1}$ then by our inductive hypothesis
+$$f(T,x) \leq f(T_\delta,x) + O(1)f(\bar{T}^*, \bar{x}^*).$$
+Now we observe that $T^*$ contains a Steiner Tree $T'$ with terminals $R_\delta \cup \{r\}$.
+Since $T_\delta$ is an $O(1)$-approximate optimal Steiner tree with terminals $R_\delta \cup \{r\}$ we have $c(T_\delta) \leq O(1) c(T')$.
+Since $T'$ necessarily installs capacities $2^\delta$ on all of its edges, $x_e^* = 2^\delta$ for all $e \in T'$.
+To see that a capacity of $2^\delta$ is necessary, suppose some edge $e$ in $T'$ has $x^*_e < 2^\delta$.
+Since $T'$ is a Steiner Tree, some terminal $v \in R_\delta$ uses $e$ on its unique path to $r$.
+But then an $rv$-flow along this path has value at most $x^*_e < 2^\delta$ contradicting feasibility of $(T^*, x^*)$.
+Thus $x_e^* = 2^\delta$ for all $e \in T'$.
+Hence we have
+$$f(T_\delta, x) = \sum_{e \in T_\delta} c_e x_e = 2^\delta c(T_\delta) \leq O(1)2^\delta c(T') = O(1)f(T', x^*) \leq O(1) f(T^*, x^*).$$
+Thus we see that
+$$f(T,x) \leq O(1) f(T^*,x^*) + O(1) f(\bar{T}^*, \bar{x}^*).$$
+Now consider the graph that results from contracting all edges of $T_\delta \cap T^*$ into $r$ in the tree $T^*$.
+If any cycles form then do the following: while there still exists a cycle $C$ take the edge $e = \text{argmin}_{e\in C} x^*_e$ and remove it.
+Call the resulting graph $S^*$.
+\paragraph{}
+We claim that $(S^*, x^*)$ is feasible for the residual problem on $G_{\delta-1}$.
+Indeed $S^*$ is a tree since we trimmed one edge from all cycles.
+Since trimming cycles preserves connectivity, and we contracted edges into $r$, all $rv$-paths in $T^*$ are preserved in $S^*$.
+Hence for all $v \in R\backslash R_\delta$ there exists an $rv$-path in $S^*$.
+Now we claim that the capacities on any $rv$-path still admit feasible flows after converting from $T^*$ to $S^*$.
+Observe that contraction will not decrease the minimum capacity.
+Further observe that deleted edges had minimum capacity on a cycle, so any flow that used a deleted edge $uv$ could send the same flow through the $uv$-path remaining from the cycle $uv$ was deleted from.
+Hence the flows on $S^*$ under capacities $x^*$ can still meet all demands in $R\backslash R_\delta$.
+Therefore $(S^*,x^*)$ (with $x^*$ restricted to $E(G_{\delta-1})$) is feasible for $G_{\delta -1}$.
+Hence, since $(\bar{T}^*, \bar{x}^*)$ is optimal for $G_{\delta -1}$:
+$$f(\bar{T}^*, \bar{x}^*) \leq f(S^*, x^*) \leq f(T^*, x^*).$$
+Therefore
+\begin{align*}
+f(T,x) &\leq O(1)f(T^*,x^*) + O(1)f(\bar{T}^*,\bar{x}^*) \\
+&\leq O(1)f(T^*,x^*) + O(1)f(T^*,x^*) \\
+&= 2O(1)f(T^*,x^*)\\
+&= O(1)f(T^*,x^*).
+\end{align*}
+So the approximation factor $O(1)$ holds in the inductive case. Thus the algorithm given is a $O(1)$-approximation algorithm.$\blacksquare$
 \end{document}

### 2017-02-02 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Retyped 3b of apx algs

@@
 \paragraph{}
-Our goal is to find a maximum $k$-Cut of $G$. That is, we seek to find a partition of $V$ into $V_1, \dots, V_k$ maximizing
-$$\sum_{i=1}^k \sum_{j = 1}^{i-1} |E(V_i, V_j)| := c(\{V_1, \dots, V_k\}).$$
-Intuitively we are maximizing the edges between the selected cuts.
-Let $\cB$ denote the $2$-approximation algorithm for max cut given in problem $3a$.
-Consider the following algorithm, which we will denote by $\cA$:
+Our goal is to find $k$ cuts of $G$ $V_1, \dots, V_k \subseteq V$ maximizing:
+$$|\bigcup_{i=1}^k \delta(V_i)|.$$
+Let $\cB$ denote the $2$-approximation algorithm for max cut given in problem $3a$. Consider the following algorithm, which we will denote by $\cA$:
 \begin{enumerate}
-\item Set $V_1 = S$ and $V_2 = V\backslash S$ where $S$ is the cut returned by $\cB$ when run on $G$.
-\item Set $V_i = \emptyset$ for $i = 3, \dots, k$.
-\item For $i = 2, \dots, k-1$
+\item Set $V_1 = \dots = V_k = \emptyset$. Set $G_1 = G$
+\item For $i = 1, \dots, k$
 \begin{enumerate}
-\item For $j = 1, \dots, i$ set $S_j$ to the cut returned by $\cB$ when run on $G[V_j]$.
-\item Set $j^* = \text{argmax}_{j=1,\dots, i} \{|E(S_j, V_j\backslash S_j)|\}$.
-\item Set $V_{i+1} = S_{j^*}$ and set $V_{j^*} = V_{j^*} \backslash S_{j^*}$.
+\item Let $S$ be the cut returned by $\cB$ run on $G_i$.
+\item Set $V_i = S$. Set $G_{i+1} = G_i \backslash \delta(S).$
 \end{enumerate}
@@
 \end{enumerate}
-The idea behind the operation of $\cA$ is to start with a $2$-cut of $G$, and in every successive iteration take the current partition and $2$-cut one of the parts in an approximately optimal way until we reach a $k$-cut.
+The idea behind the operation of $\cA$ is to find an optimal cut, then remove the edges of the cut and iterate.
 \begin{lemma}\label{lemma:3b1}
-Let $\cO = \{V^*_1, \dots, V^*_k\}$.
-Consider an arbitrary iteration $i = 1, \dots, k-1$ (iteration $1$ is original $2$-cut).
-Let $\cS = \{V_1, \dots, V_{i+1}\}$ be the set of partitions chosen so far by $\cA$.
-Suppose that at the end of iteration $i$, the algorithm has split $V_j$ into $V_j'$ and $V_{i+1}$.
-That is the solution after iteration $i$ is
-$$\cS' = \cS \backslash V_j \cup \{V_j', V_{i+1}\}.$$
+Let $\cO = \{V^*_1, \dots, V^*_k\}$ denote the optimal solution.
+For any $i=1,\dots, k$ let $\cS^i = \{V_1, \dots, V_{i}\}$ be the set of partitions chosen so far at the end of iteration $i$ by $\cA$.
 Then
-$$2k(c(\cS') - c(\cS)) \geq c(\cO) - c(\cS).$$
+$$2k(c(\cS^i) - c(\cS^{i-1})) \geq c(\cO) - c(\cS^{i-1}).$$
+With $S^0 = \emptyset$.
 \end{lemma}
 \begin{proof}
-The difference between $c(\cS')$ and $c(\cS)$ is precisely $|E(V_j', V_{i+1})|$ (the number of edges gained by splitting $V_j$).
-Further the difference $c(\cO) - c(\cS)$ is upper-bounded by the edges counted in $c(\cO)$ that are not counted in $c(\cS)$.
-Call such edges $E'$.
-The edges $E'$ are necessarily internal to a part of $\cS$ (that is, both endpoints lie in some $V_k \in \cS$).
-Otherwise they are counted in $c(\cS)$ since their endpoints lie in distinct parts of $\cS$.
-Thus if we consider the $k$-cut induced by $\cO$ on $G[E']$ it consists of at most $k$ splits ($2$-cuts) of parts $V_k \in \cS$.
-\paragraph{}
-Let $V'_k$ denote the best possible $2$-cut of $V_k \in \cS$ for all $k = 1, \dots, i$. Now let $$k^* = \text{argmax}_{k=1,\dots,i}  \{|E(V'_k, V_k \backslash V'_k)|\}.$$ We have
-$$ |E'| \leq k |E(V'_{k^*}, V_k \backslash V'_{k^*})|$$
-by our previous discussion, since $E'$ edges are formed by a $k$-cut on edges internal to parts of $\cS$, and  $|E(V'_{k^*}, V_k \backslash V'_{k^*}|$ is largest $2$-cut to be formed by edges internal to a part of $\cS$.
-Now the cut found by our algorithm $\cA$ in this iteration in step $3(c)$ is a $2$-approximation of the best possible split cut.
-That is,
-$$|E(V'_{k^*}, V_k \backslash V'_{k^*})| \leq 2|E(V_j', V_{i+1})| .$$
-Hence, combining these inequalities, we see that
-$$|E'| \leq 2k |E(V_j', V_{i+1})|.$$
-Therefore $2k(c(\cS') - c(\cS)) \geq c(\cO) - c(\cS)$ as desired.
+For $i = 1,\dots, k$ let $S_i^* = V_i^k \backslash \bigcup_{S\in \cS^{i-1}} S$. Then
+$$c(\cO) - c(\cS^{i-1}) = c(\{S^*_1,\dots, S^*_l\}) = |\bigcup_{i=1}^k \delta(S^*_i)|.$$
+Let $$S^* = \text{arg max}_{S^*_i : i=1,\dots, k} |\delta(S^*_i)|.$$
+Then 
+$$c(\cO) - c(\cS^{i-1}) \leq k|\delta(S^*)|.$$
+Now observe that for all $i = 1,\dots, k$, $\delta(S^*_i) \subseteq E(G_i)$, hence in particular $\delta(S^*) \subseteq E(G_i)$.
+Thus $|\delta(S^*)|$ has size at most that of an optimal cut in $G_i$.
+Now by our choice of $V_i$, using $2$-approximation algorithm $\cB$, $2|\delta(V_i)|$ is at the size of an optimal cut in $G_i$.
+Hence we have
+$$|\delta(S^*)| \leq 2|\delta(V_i)|.$$
+Thus, combining inequalities, we observe
+$$c(\cO) - c(\cS^{i-1}) \leq 2k |\delta(V_i)|.$$
+But $|\delta(V_i)| = c(\cS^i) - c(\cS^{i-1})$ and thus
+$$c(\cO) - c(\cS^{i-1}) \leq 2k (c(\cS^i) - c(\cS^{i-1}))$$
+as desired.
 \end{proof}
@@
 \paragraph{Main Proof}
-Since $\cB$ can be run in polynomial time, and $k$ is at most $|V|$, it is clear that $\cA$ runs in polynomial time.
-At termination the algorithm returns a partition of $V$ into $k$ parts so it returns a feasible solution.
+Since $\cB$ can be run in polynomial time, and $k$ is at most $|V|$, it is clear that step $2$ of $\cA$, and hence $\cA$, runs in polynomial time.
+At termination the algorithm returns $k$ subsets of $V$, so the algorithm returns a feasible solution.
 \paragraph{} 
 Now to see the approximation factor holds let $\cS^i$ denote the solution maintained by $\cA$ at the end of iteration $i$.
-Then the returned solution is $\cS^{k-1}$.
+Then the returned solution is $\cS^{k}$.
 Let $\cO$ be an optimal solution.
@@
 \begin{align*}
-c(\cS^{k-1} &\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-2}) \\
-		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(\frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-3})) \\
+c(\cS^{k}) &\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-1}) \\
+		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(\frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{k-2})) \\
 		&\geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})(1 + (1-\frac{1}{2k}) + (1-\frac{1}{2k})^2 + \dots + (1-\frac{1}{2k})^{k-1}) \\
@@
 		&=\frac{c(\cO)}{2k} \cdot \frac{1-(1-\frac{1}{2k})^k}{\frac{1}{2k}}\\
-		&= 1-(1-\frac{1}{2k})^kc(\cO).
+		&= (1-(1-\frac{1}{2k})^k)\cdot c(\cO).
 \end{align*}

### 2017-02-02 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
Fixed 3b of apx algs a1
conv opt 2b of a1

--- 2017-01-28 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
@@
 \subsection{b}
+\paragraph{}
+Let $n \in \N$ and consider the function $f: S^n_{++} \rightarrow \R$ given by
+$$f(x) = (\det X)^\frac{1}{n}.$$
+We will demonstrate that $f$ is concave. It suffices to show that the function 
+$$g(t) = (\det (Z + t V))^\frac{1}{n} $$
+is concave for arbitrary $Z \in S^n_{++}$ and $V \in S^n$. Note that $g$ is the restriction of $f$ to some arbitrary line. The domain of $g$ is $\{t \in \R: Z + tV \in S^n_{++}\}$.
+\paragraph{}
+Since $Z \in S^n_{++}$ we can decompose $Z$ as $Z=Z^\frac{1}{2}Z^\frac{1}{2}$ with $Z^\frac{1}{2}$ non-singular.
+Let $\lambda_1, \dots, \lambda_n$ denote the eigenvalues of $Z^\frac{-1}{2}VZ^\frac{-1}{2}$.
+Then we observe that
+$$g(t) = (\det(Z+tV))^\frac{1}{n} = (\det(Z^\frac{1}{2}(I + tZ^\frac{-1}{2}VZ^\frac{-1}{2})Z^\frac{1}{2}))^\frac{1}{n}.$$
+Using that $\det(\cdot)$ distributes over matrix products and that $\det(A)$ is equal to the product of the eigenvalues of $A$ we have
+$$g(t) = \det(Z^\frac{1}{2})^\frac{2}{n}\det(1 + tZ^\frac{-1}{2}VZ^\frac{-1}{2}) = \det(Z^\frac{1}{2})^\frac{2}{n}(\Pi_{i=1}^n 1+t\lambda_i)^\frac{1}{n}.$$
+Now we will show that $g$ is a composition of affine and concave functions, and hence that $g$ is concave. Let $h : \R \rightarrow \R^n$ be the affine function given by
+$$h(t)_i = 1+t\lambda_i$$
+for $i = 1\dots n$. Let $ y : \R^n \rightarrow \R$ be the geometric mean function
+$$y(x) = (\Pi_{i=1}^n x_i)^\frac{1}{n}.$$
+We know from the notes that $y$ is concave. Finally, let $z: \R\rightarrow \R$ be the affine (linear actually) function given by
+$$z(x) = \det(Z^\frac{1}{2})^\frac{2}{n}$$
+Then we see that
+$$g(t) = z\circ y\circ h(t).$$
+Thus $g$ is a composition of affine functions with a concave function and hence $g$ is concave.$\blacksquare$
 \end{document}

### 2017-02-06 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
starting q5

@@
 So the approximation factor $O(1)$ holds in the inductive case. Thus the algorithm given is a $O(1)$-approximation algorithm.$\blacksquare$
+
+%Q5
+\section{}
+\paragraph{}
+Let $G=(V,E)$ be a metric graph with edge lengths $d: E \rightarrow \N_+$ and let $f : 2^V \rightarrow \N_+$  be a a function satisfying:
+\begin{enumerate}
+\item (Symmetric) $f(S) = f(V\backslash S)$ for all $S \subseteq V$
+\item (Subadditive) $f(A \cup B) = f(A) + f(B)$ for all $A,B \subseteq V$.
+\end{enumerate}
+Consider the following integer program, to be labelled $IP$:
+\begin{align*}
+\text{min} \sum_{e \in E} d(e) x_e\\
+\text{s.t.} \sum_{e\in \delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+x &\geq 0, \text{ integer}.
+\end{align*}
+\paragraph{Solving $IP$ of Trees}
+We first observe that this $IP$ is easy to solve on trees.
+Indeed let $T$ be a tree, and for every $e \in E(T)$ let $S^e$ denote the set of vertices in one of the two connected components of $T-e$.
+By definition, $\delta(S^e) = \{e\}$.
+We define the following algorithm $\cA$ to solve $IP$ on a tree $T$:
+\begin{enumerate}
+\item For all $e \in E(T)$ set $x_e = f(S^e)$.
+\item Return $x$.
+\end{enumerate}
+To see that $\cA$ solves $IP$ on trees, we will show that $(1)$ the solution returned is feasible, and $(2)$ that every feasible $x$ has necessarily $x_e \geq f(S^e)$, implying that the $x$ returned by $\cA$ is optimal.
+Let $x$ be the solution returned by $\cA$.
+To see $(1)$, it is immediate from the definition of $f$ that $x \geq 0$.
+It remains to verify the cut inequalities.
+Let $S \subseteq V$.
+Enumerate $\delta(S)$ as $\delta(S) = \{e_1, \dots, e_k \}$.
 \end{document}

### 2017-02-07 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
working on a1

@@
 Let $x$ be the solution returned by $\cA$.
+\paragraph{}
 To see $(1)$, it is immediate from the definition of $f$ that $x \geq 0$.
@@
 Enumerate $\delta(S)$ as $\delta(S) = \{e_1, \dots, e_k \}$.
+For each $i = 1, \dots, k$ define
+$$S_i = \{ v \in S : v \text{ can reach an endpoint of } e_i \text{ via a path internal to } S\}.$$
+Since $T$ is connected, each node $v \in S$ can reach each node in $V \backslash S$, and must use an edge $e_i \in \delta(S)$ to do so.
+The vertex $v$ is assigned to $S_i$ where $e_i$ is the first edge in $\delta(S)$ on any such path across the cut.
+Hence $S = \cup_i S_i.$ Further $S_i \subseteq S^{e_i}$ or $S_i \subseteq V\backslash S^{e_i}$.
+To see this, observe that if it were to fail then there would exist $u, v \in S_i$ with $u \in S^{e_i}$ and $v \in V\backslash S^{e_i}$.
+But then $u$ and $v$ are connected internally within $S$ (since they both can internally reach an endpoint of $e_i$) and so removing $e_i$ does not separate them, violating that $u \in S^{e_i}$ and $v \in V\backslash S^{e_i}$.
+Thus
 \end{document}

### 2017-02-07 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Partial Scribe of 5

--- 2017-02-02 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
@@
 So the approximation factor $O(1)$ holds in the inductive case. Thus the algorithm given is a $O(1)$-approximation algorithm.$\blacksquare$
+
+%Q5
+\section{}
+\paragraph{}
+Let $G=(V,E)$ be a metric graph with edge lengths $d: E \rightarrow \N_+$ and let $f : 2^V \rightarrow \N_+$  be a a function satisfying:
+\begin{enumerate}
+\item (Symmetric) $f(S) = f(V\backslash S)$ for all $S \subseteq V$
+\item (Subadditive) $f(A \cup B) = f(A) + f(B)$ for all $A,B \subseteq V$.
+\end{enumerate}
+Consider the following integer program, to be labelled $IP$:
+\begin{align*}
+\text{min} \sum_{e \in E} d(e) x_e\\
+\text{s.t.} \sum_{e\in \delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+x &\geq 0, \text{ integer}.
+\end{align*}
+\paragraph{Solving $IP$ of Trees}
+We first observe that this $IP$ is easy to solve on trees.
+Indeed let $T$ be a tree, and for every $uv \in E(T)$ let $S^{uv}_u$ denote the set of vertices in connected component of $T-uv$ containing $u$, and define $S^{uv}_v$ similarly.
+By definition, $\delta(S^{uv}_u) =  \{e\} = \delta(S^{uv}_v)$.
+We define the following algorithm $\cA$ to solve $IP$ on a tree $T$:
+\begin{enumerate}
+\item For all $uv \in E(T)$ set $x_e = f(S^{uv}_u)$.
+\item Return $x$.
+\end{enumerate}
+To see that $\cA$ solves $IP$ on trees, we will show that $(1)$ the solution returned is feasible, and $(2)$ that every feasible $x$ has necessarily $x_{uv} \geq f(S^{uv}_u)$, implying that the $x$ returned by $\cA$ is optimal.
+Let $x$ be the solution returned by $\cA$.
+\paragraph{}
+To see $(1)$, it is immediate from the definition of $f$ that $x \geq 0$.
+It remains to verify the cut inequalities.
+Let $S \subseteq V$.
+Enumerate $\delta(S)$ as $\delta(S) = \{e_1, \dots, e_k \}$, where each $e_i := u_iv_i$ and $u_i \in S$.
+Let $S_1, \dots, S_\ell$ be the partition of $S$ induced by the connected components of $G-\delta(S)$ contained in $S$.
+Now let $D(S_i) = \{ e_j \in \delta(S) : u_j \in S_i \}$ be the set of edges in $\delta(S)$ incident with $S_i$.
+Notice that the set of $D(S_i)$ partition $\delta(S)$.
+\paragraph{}
+We claim that $\bigcap_{e_j \in D(S_i)} S^{e_j}_{u_j} = S_i.$ If $v \in S_i$ then for all $e_j \in D(S_i)$ there exists a $vu_j$-path that does not use $e_j$, since $v$ and $u_j$ are in the same connected component $S_i$.
+Thus $v \in \bigcap_{e_j \in D(S_i)}S^{e_j}_{u_j}$.
+If $v \not\in S_i$ then either $v \in S_{k}$ for $k \neq i$ or $v \in V\backslash S$.
+In either case, $v$ is connected to any vertex in $S_i$ via a path using some $e_j \in D(S_i)$.
+But then $v \in S^{e_j}_{v_j} = V \backslash S^{e_j}_{u_j}$.
+Therefore $v \not\in \bigcap_{e_j \in D(S_i)}S^{e_j}_{u_j}$.
+Thus the claim holds.
+\paragraph{}
+So we have $$S = \bigcup_{i=1}^\ell S_i = \bigcup_{i=1}^\ell \bigcap_{e_j \in D(S_i)}S^{e_j}_{u_j}.$$ Now observe that
+\begin{align*}
+f(S) &\leq \sum_{i=1}^\ell f(\bigcap_{e_j \in D(S_i)}S^{e_j}_{u_j}) &\text{by subadditivity} \\
+&= \sum_{i=1}^\ell f(V\backslash \bigcap_{e_j \in D(S_i)} S^{e_j}_{u_j}) &\text{by symmetry} \\
+&= \sum_{i=1}^\ell f(\bigcup_{e_j\in D(S_i)} V\backslash S^{e_j}_{u_j}). \\
+&\leq \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(V\backslash S^{e_j}_{u_j}) &\text{by subadditivity} \\
+&= \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(S^{e_j}_{u_j}) &\text{by symmetry} \\
+&= \sum_{e_j \in \delta(S)} f(S^{e_j}_{u_j}) \\
+&= \sum_{e_j \i \delta(S)} x_{e_j}.
+\end{align*}
+Hence $x$ satisfies the cut constraint for $S$. Therefore $x$ is feasible and $(1)$ holds.
+\paragraph{}
+Now to demonstrate $(2)$, let $uv \in E(T)$. Consider the cut $S^{uv}_u$. We have $\delta(S^{uv}_u) = \{uv\}$ and hence by the cut constraint for $S^{uv}_u$, all feasible $x$ satisfy
+$$x_{uv} \geq f(S^{uv}_u).$$
+Now the $x$ we construct is tight for all the above, and since $d \geq 0$ we can do no better without violating feasibility.
+Thus $(2)$ holds and our $x$ is optimal.
+So we conclude that we can solve $IP$ on trees easily.
+\paragraph{An Approximation Algorithm for $IP$}
+We now give an algorithm that produces an integral solution to $IP$ for a general metric graph $G$ which approximates the optimal $LP$-relaxation of $IP$ within a factor of $O(\log n)$.
+Our algorithm will use the following theorem from class:
+\begin{theorem} \label{th:metric-tree}
+Let $G=(V,E)$ be a graph.
+Let $d: E \rightarrow \R_+$ be a metric function.
+Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$.
+There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
+\end{theorem}
+Consider the algorithm $\cB$:
+\begin{enumerate}
+\item Let $x^*$ be the optimal solution to the linear programming relaxation.
+\item Use Theorem \ref{th:metric-tree} with $\lambda^* = x^*$ to obtain a weighted tree $T$.
+\item Solve $IP$ on $T$ as discussed above and return the solution $\bar{x}$.
+\end{enumerate}
+
 \end{document}

### 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
q5 a1

@@
 So we conclude that we can solve $IP$ on trees easily.
-\paragraph{An Approximation Algorithm for $IP$}
+\paragraph{Integrality Gap for $IP$}
 We now give an algorithm that produces an integral solution to $IP$ for a general metric graph $G$ which approximates the optimal $LP$-relaxation of $IP$ within a factor of $O(\log n)$.
@@
 \end{enumerate}
-
+\paragraph{}
+We claim that the solution $\bar{x}$ returned by $\cB$ is feasible for $IP$.
+To see this, observe that completing $T$ to the graph $G$ does not remove any edges across any cut set $S$, hence the bound $\sum_{e \in \delta(S)} \bar{x}_e \geq f(S)$ still holds, since additional edges contribute non-negative amounts to $\sum_{e \in \delta(S)}$.
+This implies that $\bar{x}$ is feasible for $IP$ on $G$.
+\paragraph{}
+Now we analyze the cost. We see that
+$$\sum_{u,v \in V} d(u,v) \bar{x}_{uv} \leq \sum_{u,v \in V} d^T(u,v) \bar{x}_{uv}$$
+from Theorem \ref{th:metric-tree}.
 \end{document}

### 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
starting to write up 6

@@
 &= \sum_{i=1}^\ell f(\bigcup_{e_j\in D(S_i)} V\backslash S^{e_j}_{u_j}). \\
-&\leq \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(V\backslash S^{e_j}_{u_j}) &\text{by subadditivity} \\
+&\leq \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(V\backslash S^{e_j}_{u_j}) &\text{by subadditivity} \\ t
 &= \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(S^{e_j}_{u_j}) &\text{by symmetry} \\
@@
 So we conclude that we can solve $IP$ on trees easily.
+Further this proof shows that $x$ is optimal for the linear programming relaxation of $IP$ on $T$, and since $x$ is integral this implies the equivalence of the linear programming relaxation and the integer program itself on $T$.
 \paragraph{Integrality Gap for $IP$}
@@
 Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$.
-There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
+There exists a polynomial time algorithm which constructs a weighted tree $T$ on $V$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
 \end{theorem}
@@
 \item Use Theorem \ref{th:metric-tree} with $\lambda^* = x^*$ to obtain a weighted tree $T$.
-\item Solve $IP$ on $T$ as discussed above and return the solution $\bar{x}$.
+\item Solve $IP$ (equivalently its $LP$ relaxation) on $T$ as discussed above and return the solution $\bar{x}$.
 \end{enumerate}
@@
 from Theorem \ref{th:metric-tree}.
+Now we convert $x^*$, the optimal solution for the linear programming relaxation of $IP$ on $G$, to a feasible solution on $T$.
+We will call such a solution $x'$.
+Our transformation will  preserve that $\sum_{u,v \in V} d^T(u,v) x^*_{uv} = \sum_{u,v \in V} d^T(u,v) x'_{uv}$.
+\begin{enumerate}
+\item Begin with $x'_{uv} = x^*_{uv}$ for all $uv \in E(T)$.
+\item Now for all $u,v \in V$ such that $uv \not\in E(T)$ do:
+\begin{enumerate}
+\item Let $P$ be the $uv$-path in $T$.
+\item For each $e \in P$ add $x^*_{uv}$ to the current value of $x'_e$.
+\end{enumerate}
+\end{enumerate}
+First we show that $x'$ is feasible on $T$.
+Let $\delta_G(S)$ denote $\delta(S)$ with respect to $G$ and similarly define $\delta_T(S)$ as $\delta(S)$ with respect to $T$.
+Let $P_{uv}$ denote the $uv$-path in $T$.
+Then for any cut set $S$ we have
+$$f(S) \leq \sum_{e \in \delta_G(S)} x^*_e = \sum_{e\in \delta_T(S)} x^*_e + \sum_{e \in \delta_G(S)\backslash \delta_T(S)} x^*_e \leq \sum_{e \in \delta_T(S)} x'_e.$$
+The second inequality follows from the observation that for every edge $e \in \delta_G(S) \backslash \delta_T(S)$ there exists some $e' \in P_{e} \cap \delta_T(S)$ and $$x'_{e'} = x^*_e + \sum_{uv \in E(G) \backslash E(T) : e' \in P_{uv}} x^*_{uv}.$$
+The fact that $\sum_{u,v \in V} d^T(u,v) x^*_{uv} = \sum_{u,v \in V} d^T(u,v) x'_{uv}$ is immediate from the definitions:
+\begin{align*}
+\sum_{u,v \in V} d^T(u,v) x^*_{uv} &= \sum_{uv \in E(T)} d^T(u,v)x^*_{uv} + \sum_{e \in E(G) \backslash E(T)} \sum_{uv \in P_{e}} d^T(u,v) x^*_{e}\\
+&= \sum_{uv \in E(T)} d^T(u,v)( x^*_{uv} + \sum_{e \in E(G)\backslash E(T): uv \in P_{e}} x^*{e}) \\
+&= \sum_{uv \in E(T)} d^T(u,v) x'_{uv} \\
+&= \sum_{u,v \in V} d^T(u,v) x'_{uv}.
+\end{align*}
+More intuitively, the distance between two vertices $u,v$ in $T$ is the sum of the distances on the edges in the path that joins them.
+Thus if we transfer the allocation $x^*_{u,v}$ to each edge in $P_{uv}$ we lose nothing in the objection function as $d^T(u,v) x^*_{u,v} = \sum_{ab \in P_{uv}} d^T(a,b) x^*_{uv}$.
+\paragraph{}
+Now we have that $x'$ is feasible on $T$, so by the optimality of $\bar{x}$ on $T$ we have
+$$\sum_{u,v \in V} d^T(u,v) \bar{x}_{uv} \leq \sum_{u,v \in V} d^T(u,v) x'_{uv} = \sum_{u,v \in V} d^T(u,v) x^*_{uv}.$$
+Thus combining inequalities we see that
+$$\sum_{u,v \in V} d(u,v) \bar{x}_{uv} \leq \sum_{u,v \in V} d^T(u,v) x^*_{uv}.$$
+Now by Theorem \ref{th:metric-tree} and our choice of $\lambda^* =x^*$ we see that 
+$$\sum_{u,v \in V} d^T(u,v) x^*_{uv} \leq O(\log n) \sum_{u,v \in V} d(u,v) x^*_{uv}$$
+and hence
+$$\sum_{u,v \in V} d(u,v) \bar{x}_{uv} \leq O(\log n) \sum_{u,v \in V} d(u,v) x^*_{uv}.$$
+Now since $\bar{x}$ is feasible for the $IP$ on $G$, this inequality tell us that the ratio of the optimal value of $IP$ to the the optimal value of its linear programming relaxation is bounded above by $O(\log n)$.
+That is, the integrality gap of $IP$ is $O(\log n)$.
+$\blacksquare$
+
+%Q^6
+\section{}
+%Pa
+\subsection{}
+\paragraph{}
+We want to show the following theorem:
+\begin{theorem}
+Let $G = (V,E)$ be a metric graph with edge lengths $d:E\rightarrow \R_+$ and $W \subset V$. One can randomly find a tree $T$ such that
+\begin{enumerate}
+\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in V$
+\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
+\end{enumerate}
+\end{theorem}
+\begin{proof}
+We may assume $\max_{u,v \in V} d(u,v) = 2^\delta$ for some $\delta \in \N$ and $d(u,v) > 1$ for all $u,v\in V$. Construct a tree $T$ by:
+\begin{enumerate}
+\item Choose randomly a permutation $\pi$ of $V$.
+\item Choose $\beta \in [0,1]$ uniformly at random.
+\item Set $D_\delta = \{V\}$.
+\item For $i = \delta -1$ to $0$ do:
+\begin{enumerate}
+\item Assign every node to the first node according to $\pi$ that has distance at most $2^\beta \cdot 2^{i-1}$.
+\item All nodes assigned to the same node that are currently in the same cluster will form a new cluster in $D_i$.
+\end{enumerate}
+\item Return tree $T$ where $V(T) = \{s \in D_i : \forall i = 0, \dots, \delta\}$ and $E(T) = \{s_is_{i-1}  \in D_{i}\times D_{i-1}: s_{i-1} \subseteq s_i, \forall i = 0,\dots,\delta\}$, and $d(s_is_{i-1}) = 2^i$.
+\end{enumerate}
+Note that the clusters of $D_0$ are singletons because the bound on the maximum distance is $2^\beta 2^0 \leq 1$. Hence the clusters of $D_0$ can be identified with nodes in $V$. 
+\paragraph{Proof of 1.}
+Suppose $u,v$ are in the same cluster at level $D_i$ and are separated at level $D_{i-1}$.
+Let $w$ be the node identified with $u$ and $v$'s cluster $S_w$ at level $D_i$.
+Then $d(u,v) \leq d(u,w) + d(w,v) \leq 2^\beta 2^{i-1} + 2^\beta 2^{i-1} \leq 2^{i+1}$.
+On the other hand $d^T(u, S_w) \geq 2^i$ and $d^T(S_w, v) \geq 2^i$ so $d^T(u,v) \geq 2^i + 2^i = 2^{i+1}$.
+This proves part $1.$ of the theorem.
+\paragraph{Proof of 2.}
+Observe that if $u,v$ are separated first at level $D_i$ then $d^T(u,v) \leq 2^{i+2}$.
+Let $S_i$ be the cluster in $D_i$ containing $u,v$.
+Let $S^u_{i-1}$ be descended cluster of $S_i$ from which $u$ descends and similarly define $S^v_{i-1}$.
+Then
+\begin{align*}
+d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_, S^u_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
+&= 2^i + 2^i +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
+&\leq 2^i + 2^i + 2\sum_{j = 1}^{i-1} 2^j \\
+&= 2^{i+1} + 2^{i+1} \\
+&= 2^{i+2}.
+\end{align*}
+Now we way that node $c$ {\it settles} $(u,v)$ at level $D_i$ if $c$ is the first node with respect to $\pi$ to which one of $u$ or $v$ gets assigned at level $D_i$.
+We say that node $c$ {\it cuts} $(u,v)$ at level $D_i$ if exactly one of $u,v$ is at distance at most $2^\beta 2^i$ from $c$.
+If $u,v$ are separated at level $D_i$ then there exists some $c$ which both settles and cuts $u,v$ at level $D_i$.
+\paragraph{}
+Note that $$E[d^T(u,v)] \leq \sum_{i=0}^{\delta - 1} \sum_{c \in V} Pr [ \text{$c$ settles (A) and cuts (B) $u,v$ at $D_i$}] 2^{i+2}.$$ The bound simply follows from our observation, and the prior definitions.
+Recall that $Pr[A \cap B] = P[A\mid B] Pr[B]$.
+\paragraph{}
+We first analyze $Pr[B]$. Without loss we may assume $d(c,u) \leq d(c,v)$. For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition. Thus
+$$Pr[B] = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{|[2^{i-1},2^i]|} = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}}$$.
+Now we need $Pr[A \mid B]$.
+Let us order the nodes of $V$ according to their distance to $u,v$.
+That is, $V = \{c_1, \dots, c_n\}$ where $\min\{d(c_i, v), d(c_i, u)\} \leq \min \{d(c_{i+1}, v), d(c_{i+1}, u)\}$ for all $i$.
+If $c= c_k$ then the probability that $c$ comes before $c_1, \dots, c_{k-1}$ in $\pi$ is at most $\frac{1}{k}$.
+In order for $c$ (given that $c$ cuts $u,v$) to settle $u,v$ at least that has to happen.
+Hence
+$$Pr[A \mid B] \leq \frac{1}{k}.$$
+\paragraph{}
+Putting it all together we obtain the following expectation:
+\begin{align*}
+E[d^T(u,v)] &\leq \sum_{i=0}^{\delta -1} \sum_{k=1}^n Pr [A \mid B]Pr[B] 2^{i+2} \\
+&\leq \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  2^{i+2}\frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}} \\
+&= 8 \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  | [2^{i-1},2^i] \cap [d(c,u),d(c,v)]| \\
+&\leq 8 d(u,v) \sum_{k=1}^n \frac{1}{k} \\
+&= 8 H_n d(u,v) \\
+&= O(log n) d(u,v).
+\end{align*}
+Thus the second claim and therefore the theorem holds.
+\end{proof}
+\begin{theorem}
+There exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
+\end{theorem}
+\begin{proof}
+Pick $v \in V$ with a parent $w \not \in V$ and contract edge $vw$ in $T$.
+Repeat until all nodes of $T$ are in $V$.
+Multiply by $4$ the weights on all remaining edges.
+Call the resulting tree $T'$.
+Note that clearly $d^{T'}(u,v) \leq 4d^T(u,v)$ as contracting only shortens the distances.
+Furthermore, in the worst case $d^{T'}(u,v) = 2^i \cdot 4$ if contraction results in contracting $u$ and $v$ right up to the cluster where they were separated (that is, contracting preserves the largest edge distance on the $uv$-path in $T$).
+But
+$$2^i \cdot 4 = 2^i + 2^i + 2^i + 2^i \geq d(S^u_{i-1}, S_i) + d(S^v_{i-1}, S_i) +  d(S^u_{i-1}, u) + d(S^v_{i-1}, v) = d^T(u,v).$$
+Hence the theorem holds.
+\end{proof}
 \end{document}

### 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Typed up 6a

@@
 %Pa
-\subsection{}
+\subsection{a}
 \paragraph{}
@@
 \begin{enumerate}
-\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in V$
+\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in W$
 \item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
@@
 \begin{enumerate}
-\item Choose randomly a permutation $\pi$ of $V$.
+\item Choose randomly a permutation $\pi$ of $W$.
 \item Choose $\beta \in [0,1]$ uniformly at random.
@@
 \item Assign every node to the first node according to $\pi$ that has distance at most $2^\beta \cdot 2^{i-1}$.
+\item For each node such that its distance from each vertex in $W$ is greater that $2^\beta \cdot 2^{i-1}$ assign it its own cluster.
 \item All nodes assigned to the same node that are currently in the same cluster will form a new cluster in $D_i$.
+\item Add a node to $T$ for each new cluster in $D_i$ and add edges to $T$ from each such node to the parent node in $T$ corresponding to their common cluster in $D_{i+1}$.
+\item For the clusters formed in Step $(b)$ assign distance $0$ to the new edge in $T$. For all others (ie those from $(a)$) assign distance $2^{i+1}$.
 \end{enumerate}
-\item Return tree $T$ where $V(T) = \{s \in D_i : \forall i = 0, \dots, \delta\}$ and $E(T) = \{s_is_{i-1}  \in D_{i}\times D_{i-1}: s_{i-1} \subseteq s_i, \forall i = 0,\dots,\delta\}$, and $d(s_is_{i-1}) = 2^i$.
+\item Return tree $T$.
 \end{enumerate}
-Note that the clusters of $D_0$ are singletons because the bound on the maximum distance is $2^\beta 2^0 \leq 1$. Hence the clusters of $D_0$ can be identified with nodes in $V$. 
+Note that the clusters of $D_0$ are singletons because the bound on the maximum distance is $2^\beta 2^0 \leq 1$.
+Hence all nodes of $V$ are invariably separated at some level by the end of operation of the algorithm.
 \paragraph{Proof of 1.}
-Suppose $u,v$ are in the same cluster at level $D_i$ and are separated at level $D_{i-1}$.
-Let $w$ be the node identified with $u$ and $v$'s cluster $S_w$ at level $D_i$.
-Then $d(u,v) \leq d(u,w) + d(w,v) \leq 2^\beta 2^{i-1} + 2^\beta 2^{i-1} \leq 2^{i+1}$.
-On the other hand $d^T(u, S_w) \geq 2^i$ and $d^T(S_w, v) \geq 2^i$ so $d^T(u,v) \geq 2^i + 2^i = 2^{i+1}$.
+Suppose $u,v \in W$ are in the same cluster at level $D_i$ and are separated at level $D_{i-1}$. Let $w$ be the node identified with $u$ and $v$'s cluster $S_w$ at level $D_i$. Then
+$$d(u,v) \leq d(u,w) + d(w,v) \leq 2^\beta 2^{i-1} + 2^\beta 2^{i-1} \leq 2^{i+1}.$$ The first inequality follows from the triangle inequality.
+The second inequality follows since the distance from $u$ (similarly $v$) to the next node on the $uw$-path is $2^{i-1}$ and the distance of the rest of the path is at most $\sum_{k=0}^{i-2} 2^k \leq 2^{i-1}$.
+On the other hand $d^T(u, S_w) \geq 2^i$ and $d^T(S_w, v) \geq 2^i$ so $d^T(u,v) = d^T(u, S_w) + d^T(S_w, v) \geq 2^i + 2^i = 2^{i+1}$.
 This proves part $1.$ of the theorem.
@@
 d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_, S^u_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
-&= 2^i + 2^i +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
+&\leq \max\{2^i,0\} + \max\{2^i,0\} +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
 &\leq 2^i + 2^i + 2\sum_{j = 1}^{i-1} 2^j \\
@@
 \end{align*}
-Now we way that node $c$ {\it settles} $(u,v)$ at level $D_i$ if $c$ is the first node with respect to $\pi$ to which one of $u$ or $v$ gets assigned at level $D_i$.
+Now we say that node $c$ {\it settles} $(u,v)$ at level $D_i$ if $c$ is the first node with respect to $\pi$ to which one of $u$ or $v$ gets assigned at level $D_i$.
 We say that node $c$ {\it cuts} $(u,v)$ at level $D_i$ if exactly one of $u,v$ is at distance at most $2^\beta 2^i$ from $c$.
-If $u,v$ are separated at level $D_i$ then there exists some $c$ which both settles and cuts $u,v$ at level $D_i$.
+If $u,v$ are separated at level $D_i$ then either there exists some $c \in W$ which both settles and cuts $u,v$ at level $D_i$ (in which case they were separated by Step $4(a)$), or every $c \in W$ does not settle $u,v$ (in which case they were separated by Step $4(b)$).
 \paragraph{}
-Note that $$E[d^T(u,v)] \leq \sum_{i=0}^{\delta - 1} \sum_{c \in V} Pr [ \text{$c$ settles (A) and cuts (B) $u,v$ at $D_i$}] 2^{i+2}.$$ The bound simply follows from our observation, and the prior definitions.
-Recall that $Pr[A \cap B] = P[A\mid B] Pr[B]$.
+Note that in the first case the distance between $u$ and $v$ is at most $2^{i+2}$ by our previous observation.
+In the second case both $u$ and $v$ were assigned their own cluster in $D_i$ and given distance $0$ to their common ancestor.
+Hence $d^T(u,v) = 0$ in this second case.
+So we have the following bound on $E[d^T(u,v)]$:
+ \begin{align*}E[d^T(u,v)] &\leq \sum_{i=0}^{\delta - 1} (\sum_{c \in W} Pr [ \text{$c$ settles $u,v$ and  $c$ cuts $u,v$ at $D_i$}] 2^{i+2} + Pr[\text{all $c \in W$ do not settle $u,v$}]\cdot 0) \\&= \sum_{i=0}^{\delta - 1} \sum_{c \in W} Pr [ \text{$c$ settles $u,v$ and  $c$ cuts $u,v$ at $D_i$}] 2^{i+2}.
+ \end{align*}
+ For $c \in W$, $u,v \in V$, and level $D_i$,  let $A$ be the event that $c$ settles $u,v$ and let $B$ be the event that $c$ cuts $u,v$ at $D_i$.
+We are interested in $Pr[A \cap B]$, or equivalently, $Pr[A \mid B] \cdot Pr[B]$.
  \paragraph{}
-We first analyze $Pr[B]$. Without loss we may assume $d(c,u) \leq d(c,v)$. For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition. Thus
-$$Pr[B] = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{|[2^{i-1},2^i]|} = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}}$$.
+We first analyze $Pr[B]$.
+Without loss we may assume $d(c,u) \leq d(c,v)$.
+For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition.
+Since we sample $2^\beta$ uniformly at random, the probability of this happening is simply $\frac{1}{2^{i-1}}$ times the length of the interval intersecting $[2^{i-1},2^i]$ with $[d(c,u), d(c,v)]$.
+More formally:
+$$Pr[B] = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{|[2^{i-1},2^i]|} = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}}.$$
+\paragraph{}
 Now we need $Pr[A \mid B]$.
-Let us order the nodes of $V$ according to their distance to $u,v$.
-That is, $V = \{c_1, \dots, c_n\}$ where $\min\{d(c_i, v), d(c_i, u)\} \leq \min \{d(c_{i+1}, v), d(c_{i+1}, u)\}$ for all $i$.
+Let us order the nodes of $W$ according to their distance to $u,v$.
+That is, $W = \{c_1, \dots, c_n\}$ where $\min\{d(c_i, v), d(c_i, u)\} \leq \min \{d(c_{i+1}, v), d(c_{i+1}, u)\}$ for all $i$.
 If $c= c_k$ then the probability that $c$ comes before $c_1, \dots, c_{k-1}$ in $\pi$ is at most $\frac{1}{k}$.
@@
 \end{proof}
+
+%Pb
+\subsection{b}
 \begin{theorem}

### 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
6a written up

@@
 \subsection{b}
-\begin{theorem}
-There exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
-\end{theorem}
-\begin{proof}
-Pick $v \in V$ with a parent $w \not \in V$ and contract edge $vw$ in $T$.
-Repeat until all nodes of $T$ are in $V$.
-Multiply by $4$ the weights on all remaining edges.
-Call the resulting tree $T'$.
-Note that clearly $d^{T'}(u,v) \leq 4d^T(u,v)$ as contracting only shortens the distances.
-Furthermore, in the worst case $d^{T'}(u,v) = 2^i \cdot 4$ if contraction results in contracting $u$ and $v$ right up to the cluster where they were separated (that is, contracting preserves the largest edge distance on the $uv$-path in $T$).
-But
-$$2^i \cdot 4 = 2^i + 2^i + 2^i + 2^i \geq d(S^u_{i-1}, S_i) + d(S^v_{i-1}, S_i) +  d(S^u_{i-1}, u) + d(S^v_{i-1}, v) = d^T(u,v).$$
-Hence the theorem holds.
-\end{proof}
+
+\paragraph{}
+Let $G = (V,E)$ be a metric graph with edge costs $c:E \rightarrow \R_+$ and requirements $r(u,v) \in \N$ for all $u,v \in V$.
+Consider the following algorithm, $\cA$, for the Generalized Steiner Network (GSN) problem:
+\begin{enumerate}
+
+\end{enumerate}
+
 \end{document}

### 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
writing 6b

--- 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
@@
 We want to show the following theorem:
-\begin{theorem}
+\begin{theorem}\label{th:tree-metric}
 Let $G = (V,E)$ be a metric graph with edge lengths $d:E\rightarrow \R_+$ and $W \subset V$. One can randomly find a tree $T$ such that
@@
 \subsection{b}
+\paragraph{}
+Let $G = (V,E)$ be a metric graph with edge costs $c:E \rightarrow \R_+$ and requirements $r(u,v) \in \N$ for all $u,v \in V$.
+We wish to operate on a tree with the same vertex set at $G$.
+The following theorem will let us do so.
 \begin{theorem}
-There exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
+Let $T$ be a tree obtained by Theorem \ref{th:tree-metric}. Then there exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
 \end{theorem}
@@
 \end{proof}
+\paragraph{}
+Combining the above with Theorem \ref{th:tree-metric} and hiding the factor of $4$ in the $O$-notation we have the important corollary
+\begin{corollary}\label{cor:tree-metric}
+Let $G = (V,E)$ be a metric graph with edge lengths $d:E\rightarrow \R_+$ and $W \subset V$. One can randomly find a tree $T$ with vertex set $V$ such that
+\begin{enumerate}
+\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in W$
+\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
+\end{enumerate}
+\end{corollary}
+\paragraph{}
+Consider the following algorithm, $\cA$, for GSN on $G$:
+\begin{enumerate}
+\item Let $W = \{v \in V: \exists u \in V, r(u,v) > 0\}$.
+\item Use Corollary \ref{cor:tree-metric} to compute a tree metric $T$ which approximates $G$ with $W$.
+\item Solve GSN on $T$ and let $x$ be the resulting solution.
+\item Return $x$.
+\end{enumerate}
+\paragraph{}
+We want to show that $\cA$ is an $O(log |W|)$-approximation algorithm for GSN on $G$. Thus we need to show a few things:
+\begin{enumerate}
+\item That we can solve GSN on trees in polynomial time.
+This implies that $\cA$ runs in polynomial time since the tree of Theorem \ref{th:tree-metric} can be found in polynomial time, and the contractions necessary to yield the tree in Corollary \ref{cor:tree-metric} can be done in polynomial time.
+\item The $x$ returned is feasible on $G$.
+\item The approximation factor holds.
+\end{enumerate}
+\paragraph{Proof of $1$}
+
+\paragraph{Proof of $2$}
+
+\paragraph{Proof of $3$}
+
 \end{document}

### 2017-02-09 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Draft of A1 of apx algs done

--- 2017-02-08 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
@@
 \subsection{b}
-
 \paragraph{}
 Let $G = (V,E)$ be a metric graph with edge costs $c:E \rightarrow \R_+$ and requirements $r(u,v) \in \N$ for all $u,v \in V$.
-Consider the following algorithm, $\cA$, for the Generalized Steiner Network (GSN) problem:
+We wish to operate on a tree with the same vertex set at $G$.
+The following theorem will let us do so.
+\begin{theorem}
+Let $T$ be a tree obtained by the algorithm of Theorem \ref{th:tree-metric}.
+Then there exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v),$ and every $v \in V\backslash W$ is a leaf of $T'$.
+\end{theorem}
+\begin{proof}
+Pick $v \in W$ with a parent $w \not \in V$ and contract edge $vw$ in $T$.
+Repeat until all nodes of $T$ are in $V$.
+Multiply by $4$ the weights on all remaining edges.
+Call the resulting tree $T'$.
+To see this algorithm works observe that every node not in $V$ is a associated with some cluster $S_w$ corresponding to $w$.
+Descendant of the node $S_w$ in $T$ is the node associated with $w$.
+The path from $S_w$ to $w$ contains only cluster nodes not in $V$ or nodes in $W$.
+Thus we can find some cluster node with child in $W$ by following the path from $S_w$ to $w$ until we reach the first node in $W$ along this path.
+Its parent is a cluster node.
+The contraction operation preserves this structure of the tree, and hence we can repeat until all cluster nodes have been contracted into nodes in $W$.
+Since each $v \in V\backslash W$ began with its associated vertex in $T$ as a leaf, and was never contracted it remains a leaf vertex in the resulting tree $T'$.
+\paragraph{}
+Note that clearly $d^{T'}(u,v) \leq 4d^T(u,v)$ as contracting only shortens the distances, but then edge distances were multiplied by a factor of $4$ at the end of operation.
+Furthermore, in the worst case $d^{T'}(u,v) = 2^i \cdot 4$ if contraction results in contracting $u$ and $v$ in $W$ right up to the cluster where they were separated (that is, contracting preserves the largest edge distance on the $uv$-path in $T$).
+But then
+$$d^{T'}(u,v) \geq 2^i \cdot 4 = 2^i + 2^i + 2^i + 2^i \geq d(S^u_{i-1}, S_i) + d(S^v_{i-1}, S_i) +  d(S^u_{i-1}, u) + d(S^v_{i-1}, v) = d^T(u,v).$$
+Hence the theorem holds.
+\end{proof}
+\paragraph{}
+Combining the above with Theorem \ref{th:tree-metric} and hiding the factor of $4$ in the $O$-notation we have the important corollary
+\begin{corollary}\label{cor:tree-metric}
+Let $G = (V,E)$ be a metric graph with edge lengths $d:E\rightarrow \R_+$ and $W \subset V$.
+One can randomly find a tree $T$ with vertex set $V$ such that every $v \in V\backslash W$ is a leaf of $T$ and
 \begin{enumerate}
-
+\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in W$
+\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
 \end{enumerate}
-
+\end{corollary}
+\paragraph{}
+Consider the following algorithm, $\cA$, for GSN on $G$:
+\begin{enumerate}
+\item Let $W = \{v \in V: \exists u \in V, r(u,v) > 0\}$.
+\item Use Corollary \ref{cor:tree-metric} to compute a tree metric $T$ which approximates $G$ with $W$.
+\item Solve GSN on $T$ and let $x$ be the resulting solution.
+\item Return $x$.
+\end{enumerate}
+\paragraph{}
+We want to show that $\cA$ is an $O(log |W|)$-approximation algorithm for GSN on $G$. Thus we need to show the following things:
+\begin{enumerate}
+\item That we can solve GSN on trees in polynomial time.
+This implies that $\cA$ runs in polynomial time since the tree of Theorem \ref{th:tree-metric} can be found in polynomial time, and the contractions necessary to yield the tree in Corollary \ref{cor:tree-metric} can be done in polynomial time.
+\item The $x$ returned is feasible on $G$.
+\item The approximation factor holds.
+\end{enumerate}
+\paragraph{Proof of $1$}
+Observe that every edge $ab \in E(T)$ is a bridge.
+Thus is we let $T_a$ denote the connected component of $T-ab$ containing $a$ and $T_b$ be the connected component of $T-ab$ containing $b$ then any feasible solution $x$ to GSN on $T$ necessarily has
+$$x_{ab} \geq \max_{u \in V(T_a), v \in V(T_b)} \{r(u,v)\}$$
+since any $uv$-flow from $u\in V(T_a)$ to $v \in V(T_b)$ routes its entire value across $ab$. We claim that setting
+$$x_{ab} = \max_{u \in V(T_a), v\in V(T_b)} \{r(u,v)\}$$
+is actually feasible, and hence optimal.
+Let $u,v \in V$.
+Let $ab$ be any edge along the $uv$-path in $T$.
+Then deleting $ab$ separates $u$ from $v$.
+That is, without loss  of generality we have $u \in T_a$ and $v \in T_b$.
+So,
+$$x_{ab} = \max_{u \in V(T_a), v\in V(T_b)} \{r(u,v)\} \geq r(u,v).$$
+Thus we can route a flow of value of $r(u,v)$ from $u$ to $v$ along the path from $u$ to $v$ in $T$.
+Thus we can solve GSN on trees in $O(|E| |V|^2)$ time by iterating over all edges, and then iterating over all pairs of vertices to compute the optimal $x$ discussed above.
+\paragraph{Proof of $2$}
+To see that $x$ is feasible for $G$ observe that since $G$ is a metric graph it is necessarily complete.
+So $T$ is a subgraph of $G$.
+Since $T$ spans $G$ every demand $r(u,v) \in V(G)$ can be satisfied by installing capacities $x_e$ on the edges in $e\in E(T) \subseteq E(G)$.
+Hence $x$ is feasible for GSN on $G$.
+\paragraph{Proof of $3$}
+It remains to verify the desired approximation factor.
+To do so we first make an interesting observation about the structure of our tree $T$ and the resulting solution $x$ on $T$.
+Let $v \in V\backslash W$.
+Then $v$ is a leaf of $T$.
+Let $w \in W$ be the adjacent vertex of $v$ in $T$.
+Then
+$$x_{vw} = \max_{a \in T_v, b \in T_w} \{r(a,b)\} = \max_{b \in T_w} \{r(v, b)\} = 0$$
+with the last inequality following since $v \not\in W$ and by the definition of $W$ this implies $r(v,b) = 0$ for all $b \in V$ (otherwise there would exist some vertex $b$ where $r(v,b) > 0$ and that would force $v \in W$).
+Therefore
+$$\sum_{u,v \in V} d(u,v) x_{uv} = \sum_{u,v \in W} d(u,v) x_{uv} .$$
+Now we use that $d(u,v) \leq d^T(u,v)$ for all $u,w \in W$ to see that
+$$\sum_{u,v \in V} d(u,v) x_{uv} \leq \sum_{u,v \in W} d^T(u,v) x_{uv} = \sum_{u,v \in V} d^T(u,v) x_{uv}.$$
+Let $x^*$ be the optimal solution to GSN on $G$. We translate $x^*$ to a feasible solution on $T$ as follows. Start with $x' = 0$. For each $uv \in E(G)$ for each $ab \in P_{uv}$ update
+$$x'_{ab} = x'_{ab} + x^*_{uv}.$$
+That is we send each $uv$'s capacity on an edge of $G$ to each edge on the $uv$-path in $T$. Since $d^T(u,v) = \sum_{ab \in P_{uv}} d^T(ab)$ we have
+$$\sum_{u,v \in V} d^T(u,v) x^*_{uv} = \sum_{u,v \in V} d^T(u,v) x'_{uv}.$$
+Now we claim that $x'$ is feasible for $T$.
+Observe that since $x^*$ is feasible on $G$ every $uv$ pair has their demand satisfied by $G$.
+On $T$ the same demand can be satisfied by rerouting any flow on an edge no longer present in $T$ by routing along the $uv$-path using the capacity transferred to that path from $x^*_{uv}$.
+Thus $x'$ can route flows satisfying all demands using only edges of $T$.
+Hence $x'$ is feasible on $T$.
+Then,
+$$\sum_{u,v\in V} d^T(u,v) x_{uv} \leq \sum_{u,v in V} d^T(u,v) x'_{uv} = \sum_{u,v \in V} d^T(u,v) x^*_{uv}$$
+with the inequality following from $x$ being optimal on $T$ while $x'$ is feasible on $T$. So combining our inequalities so far we have
+$$ \sum_{u,v \in V} d(u,v) x_{uv} \leq \sum_{u,v \in V} d^T(u,v) x^*_{uv}.$$
+Taking Expectation of both sides and applying property $2.$ of Corollary \ref{cor:tree-metric} we see that
+$$E[\sum_{u,v \in V} d(u,v) x_{uv}] \leq O(\log |W|) E[\sum_{u,v\in V} d(u,v) x^*_{uv}].$$
+Therefore, in expectation, the cost of the solution returned by our algorithm is at most $O(\log |W|)$ times the cost of the optimal solution.
+Hence the algorithm $\cA$ is an $O(\log k)$-approximation algorithm for GSN where $k = |W| = |\{ v \in V: \exists u \in V, r(u,v) > 0 \}|.\blacksquare$
 \end{document}

### 2017-02-10 CO750-ApproximationAlgorithmsTopics/lectures/lecture_10.tex
Scribe lecture 10

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-02-07}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 10} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\paragraph{}
We focus on the proof of the result that there exists a $1.39$ approximation algorithm for Steiner Tree.
Recall that $$\sum_t E(c(C^t)] \leq \sum_{t} E[\sum_j \frac{x_j^t}{M} c(C_j)] \leq \frac{1}{M} \sum_t E[opt^t_{k,f}] \leq \frac{1}{M} \sum_t E[opt^t_k].$$
Let $S^*$ be the optimal $k$-restricted Steiner Tree. Each time we sample $C^t$ we delete a subset of edges of $S^*$. Let $S^* = S^0 \supseteq S^1 \supseteq S^2 \supseteq \dots$ be the subgraphs obtained in this manner. We will show:
\begin{enumerate}
\item $S^t$ plus the sampled components until iteration $t$ span the terminals.
\item Each edge in $S^*$ is deleted after an (expected) number of at most $\ln (4) \cdot M$ iterations.
\end{enumerate}
$(1) + (2)$ implies the existence of a $\ln(4)$-approximation algorithm for $k$-restricted Steiner Tree.
\paragraph{Proof of $(1)$}
We begin with demonstrating $(1)$. How do we delete edges from $S^*$? By the following random process:
\begin{enumerate}
\item Convert $S^*$ to a binary tree by adding $0$-cost edges if necessary. Note that the height of the tree is at most $|R|-1$ at this stage/
\item For $u,v \in R$ let $P_{uv}$ denote the unique $uv$-path on $S^*$.
\item For each Steiner node, choose uniformly at random one of the two edges going to its children. Call the set of all such edges $\tilde{B}$.
\item Clearly $Pr[e \in \tilde{B}] = \frac{1}{2}$ for all $e \in S^*$.
\item Set $W = \{ uv \in {R\choose 2} : |P_{uv} \cap \tilde{B}| = 1\}$.
\item For each $e \in S^*$, let $W(e) = \{uv \in W: e \in P_{uv}\}$.
\item For a given component $C$ we let the set of candidate bridges be $\cB(C) = \{B \subseteq W: |B| = |R(C)| - 1, (W\backslash B) \cup C \text{ is connected} \}$.
\end{enumerate}
\paragraph{Deletion Process}
Let $C^t$ be the component sampled in iteration $t$. Delete from $W$ a random subset $\bar{B_{r_w}}(C^t) \in \cB_w (C^t)$ according to a probability distribution: $\omega_{C^t} : \cB_w (C^t) \rightarrow [0,1].$ Delete $e \in S^*$ as soon as all edges in $W(e)$ are deleted.
\paragraph{Lemma 1}
The graph $S^t \cup \bigcup_{t' = 1} ^t C^{t'}$ spans $R$.
\paragraph{Proof of Lemma 1}
Let $W^t \subseteq W$ be the set of edges of $W$ that are not deleted at iteration $t$. By the definition of $\cB_w$, $W^t \cup \bigcup_{t' = 1} ^t C^{t'}$ spans $R$. For all $e = uv \in W^t$, $P_{uv} \in W(e)$, all edges in $P_{uv}$ are not deleted. Thus $uv$ are connected in $S^t$. $\blacksquare$.
\paragraph{Proof of $(2)$}
We now consider part $(2)$. We need the following lemma.
\paragraph{Lemma $2$}
There exist probability distributions $\omega_c: \cB_w(C) \rightarrow [0,1]$ such that, in each iteration $t$, $e \in W$ gets deleted with probability at least $\frac{1}{M}$.
\paragraph{Proof of Lemma $2$}
We claim that there exists $\omega_c$ such that
\begin{equation}\label{eq:star}\sum_C \sum_{B \in \cB(C) : e\in B} x_C \cdot \omega_C (B) \geq 1\end{equation}
for all $e \in W$. If equation \ref{eq:star} holds, then $Pr[e \in \bar{B_{r_w}} (C^t)] = \sum_{C} \sum_{B\in \cB_w(C) : e \in B} \frac{x_C}{M} \cdot \omega_C(B) \geq \frac{1}{M}$ as desired. If equation \ref{eq:star} does not hold then the following linear program has no solution:
\begin{align*}
\sum_{C}\sum{B \in \cB_w(C): e\in B}x_C \cdot \omega_C(B) &\geq 1, &\forall e \in E\\
\sum_{B \in \cB(C)} \omega_C(B) &\leq 1 &\forall C \\
\omega_C(B) \geq 0.
\end{align*}
By Farkas Lemma, the following system has a solution:
\begin{align*}
y_C &\geq \sum_{e \in B} c_e x_C &\forall C, \forall B \in \cB_W(C) \\
\sum_C y_C &< \sum_{e \in W} c_e\\
(y,c) &\geq 0.
\end{align*}
Let us interpret the values of $c$ variables as edge costs. Then $y_C \geq x_C \sum_{e \in B} c_e$ for all $B$. So $y_C \geq x_C \cdot \max\{c(B) : B \in \cB_W(C)\} = x_C \cdot c(\cB_{r_w}(C))$ (bridges with $c$ as cost function). Then $\sum_C x_C\cdot \text{cost}(\cB_{r_w}(C)) \leq \sum_{C} y_C < \sum_{e\in W} c_e = \text{cost}(W)$. This contradicts bridge lemma (should be $\geq$ by lemma). $\blacksquare$.
\paragraph{}
Summarize: there exists probability distributions $\omega_C$ for all $C$ such that each edge $e \in W$ has a ``uniform" probability to be deleted in each iteration. Thus $\frac{1}{M}$. Now to more forward against $(2)$ what is the probability that an edge $e \in S^*$ is deleted? In other words, what is the expected number of iterations until $e$ is deleted?
\end{document}

### 2017-02-10 CO750-ApproximationAlgorithmsTopics/lectures/lecture_11.tex
Scribe lecture 10

--- 2017-01-27 CO750-ApproximationAlgorithmsTopics/lectures/lecture_07.tex
@@
 
-\lhead{2017-01-26}
+\lhead{2017-mm-DD}
 \rhead{William Justin Toth CO750-Approximation Algorithms Lecture n} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
@@
 \begin{document}
-\section*{Integrality Gap}
-\paragraph{}
-Let $P$ be an optimization problem, and suppose that there exists an integer programming formulation for $P$.
-Denote by $OPT(I)$ the objective function value of an optimal solution to an instance $I$ of $P$.
-Denote by $OPT_f(I)$ the objective function value of the Linear Programming relaxation of the IP formulation of $I$.
-\paragraph{}
-The integrality gap of the formulation is
-$$\inf_{I \in P} \frac{OPT(I)}{OPT_f(I)}$$
-for maximization problems, and
-$$\sup_{I \in P} \frac{OPT(I)}{OPT_f(I)}$$
-for minimization problems.
-\section{Iterative Rounding}
-\paragraph{}  Recall the linear programming formulation $LP$ for Generalized Steiner Network (GSN):
-\begin{align*}
-\min &\sum_e c_e x_e \\
-\text{s.t.} x(\delta(S)) &\geq f(S) &\forall S \subseteq V\\
-x&\geq 0
-\end{align*}
-where $f(S) = \max_{u \in S, v \not\in S} r(u,v).$
-\paragraph{}
-A function $f: 2^V \rightarrow \Z$ is called weakly-supermodular (wsm) if
-\begin{enumerate}
-\item $f(V) = 0$ and
-\item For all $A,B \subseteq V$ at least one of the following holds
-\begin{enumerate}
-\item $f(A) + f(B) \leq f(A\backslash B) + f(B \backslash A)$ or
-\item $f(A) + f(B) \leq f(A\cup B) + f(A\cap B).$
-\end{enumerate}
-\end{enumerate}
-\begin{proposition}
-If $f$ is a proper function then $f$ is wsm.
-\end{proposition}
-\paragraph{}
-Recall that an extreme point solution (xps) to an LP is a feasible solution which cannot be written as a convex combination of other feasible solutions.
-\begin{proposition}
-An optimal xps of $LP$ for GSN can be computed in polynomial time.
-\end{proposition}
-\begin{theorem}
-For any wsm function $f$, any xps to $LP$ satisfies $x_e \geq \frac{1}{2}$ for at least one $e \in E$.
-\end{theorem}
-\paragraph{}
-The previous theorem yields a natural algorithm for GSN:
-\begin{enumerate}
-\item Set $H= \emptyset$, and $f' = f$.
-\item While $f' \neq 0$
-\begin{enumerate}
-\item Find an optimal xps for $LP$ with $f'$
-\item For all $e$ such that $x_e \geq \frac{1}{2}$, include $\ceil{x_e}$ copies of $e$ in $H$
-\item Update $f'$: for all $S\subseteq V$, set $f(S) = \max \{f(S)-|\delta_H(S)|, 0\}$.
-\end{enumerate}
-\item Output $H$.
-\end{enumerate}
-\begin{lemma}
-Let $H$ be a (multi)subgraph of $G$. If $f: 2^V \rightarrow \Z_+$ is wsm then $f'(S) = \max \{ f(S) - |\delta_H(S)|, 0\}$ is wsm.
-\end{lemma}
-\begin{lemma}
-We can solve the $LP$ relaxation for GSN in polynomial time at any iteration of the algorithm for instances of GSN.
-\end{lemma}
-\begin{theorem}
-The solution $H$ output by the algorithm has cost at most $2$ times the cost of the optimal solution to the $LP$ relaxation.
-\end{theorem}
-\begin{proof}
-Proceed by induction on the number of iterations.
-If the algorithm only requires one iteration then the result is immediate.
-Otherwise let $x$ be an optimal xps in the first iteration.
-Let $\hat{x}$ be defined as
-$$ \hat{x}_e = \begin{cases} 1, &\forall e, x_e \geq \frac{1}{2} \\
-0, &\text{otherwise}\end{cases}$$
-Let $H^1$ be the set of edges taken in the first iteration.
-Then $c(H^1) \leq 2c(\hat{x})$.
-Let $f'$ be the requirement function after the first iteration and let $H'$ be the set of edges picked by the algorithm in all subsequent iterations.
-\paragraph{}
-The key observation: $(x-\hat{x})$ is a feasible solution for $f'$. So by induction $c(H') \leq 2c(x-\hat{x})$. In total
-$$c(H^1 \cup H') \leq 2c(\hat{x}) + 2c(x-\hat{x}) \leq 2c(x).$$
-\end{proof}
-\paragraph{}
-An xps $x$ of $LP$ satisfies $m = |E|$ linearly independent inequalities with equality (tight).
-In our analysis to follow we will make the simplifying assumption that $x_e > 0$ for all $e\in E$.
-We may assume this by removing all $e$ such that $x_e = 0$.
-For a set $S$ we will denote by $\chi(S)$ the incidence vector of $S$.
-\paragraph{}
-A collection $\cL$ of subsets of $V$ is a \textit{laminar} family if no two sets in this collection ``crosses".
-By crosses we mean for all $A, B \in \cL$, we have $A\cap B \neq \emptyset$ and either $A\backslash B \neq \emptyset$ or $B\backslash A \neq \emptyset$.
-\begin{theorem}
-For any xps $x$ to $LP$ there is a collection of $m$ tight sets such that their incidence vectors are linearly independent, and the collection is a laminar family.
-\end{theorem}
-\begin{proof}
-To follow in next lecture.
-\end{proof}
+
 \end{document}

### 2017-02-13 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
Proofreading A1 of apxalgs

@@
 Order the edges as $S = \{e_1, \dots, e_n\}$ in the order they were taken during the operation of $\cA$.
-Let $F_i := E\backslash \tilde{E}$ be the edges removed from $E$ at the edge of iteration $i$ of $\cA$ (the iteration where $e_i$ is added to $S$).
+Let $F_i := E\backslash \tilde{E}$ be the edges removed from $E$ at the end of iteration $i$ of $\cA$ (the iteration where $e_i$ is added to $S$).
 Let $S^*$ be the optimal solution and partition $S^* = S^*_1 \cup \dots \cup S^*_n$ where $S^*_i = S^* \cap F_i$.
@@
 We proceed by induction.
-For $e_1$ $\{e_1\}$ is inclusion-wise maximal on $\cS[F_1]$ (by definition of $F_1$), and $S_1^*$ is feasible on $\cS[F_1]$ since $S^*_1 \subseteq F_1$ and $S^* \in \cS$.
+For the base case $e_1$, $\{e_1\}$ is inclusion-wise maximal on $\cS[F_1]$ (by definition of $F_1$), and $S_1^*$ is feasible on $\cS[F_1]$ since $S^*_1 \subseteq F_1$ and $S^* \in \cS$.
 Thus we have $|\{e_1\}| \geq \frac{1}{2} |S_1^*|$, and hence
-$$|S^*_1| \leq \alpha.$$
+$$|S^*_1| \leq 2.$$
 Further by our greedy choice of $e_1$, $w_{e_1} \geq w_e$ for all $e \in F_1$ (and hence for all $e \in S^*_1$).
@@
 Also by our greedy choice, any edges that were not moved from $S^*_{i+1}$ by $\cB$ have cost at most $w_{e_{i+1}}$.
-To finish the proof of the claim, we demonstrate that while $|S_{i+1}^*| > \alpha$ there exists some $S^*_j$ with $j\leq i$ and $|S^*_j| < 2$.
+To finish the proof of the claim, we demonstrate that while $|S_{i+1}^*| > 2$ there exists some $S^*_j$ with $j\leq i$ and $|S^*_j| < 2$.
 Observe that $\{e_1, \dots, e_{i+1}\}$ is inclusion-wise maximal on $\cS[F_1 \cup \dots \cup F_{i+1}]$ by definition of $F_1 \cup \dots \cup F_{i+1}$.
@@
 We proceed by induction.
-For $e_1$ $\{e_1\}$ is inclusion-wise maximal on $\cS[F_1]$ (by definition of $F_1$), and $S_1^*$ is feasible on $\cS[F_1]$ since $S^*_1 \subseteq F_1$ and $S^* \in \cS$.
+For the base case $e_1$, $\{e_1\}$ is inclusion-wise maximal on $\cS[F_1]$ (by definition of $F_1$), and $S_1^*$ is feasible on $\cS[F_1]$ since $S^*_1 \subseteq F_1$ and $S^* \in \cS$.
 Thus we have $|\{e_1\}| \geq \frac{1}{\alpha} |S_1^*|$, and hence
@@
 \paragraph{}
-The fact that the Greedy Algorithm runs in polynomial time is obvious.
-$|E(v,S)|$ can be computed in $O(|E|)$ time and the while loop runs for $O(|V|)$ iterations.
-Further it clearly returns a cut.
+To see that the Greedy Algorithm runs in polynomial time observe that $|E(v,S)|$ can be computed in $O(|E|)$ time and the while loop runs for $O(|V|)$ iterations.
+Further the algorithm returns a feasible solution by definition.
 It remains to verify the approximation factor of $\frac{1}{2}$ holds.
@@
 Let $\bar{S} = V'\backslash S$ and $\bar{S^*} = V' \backslash S^*$.
-We may assume without loss of generality that the Greedy Algorithm puts $v \in S$ and the optimal solution puts $v \in S^*$ after this iteration (if this does not hold simply relabel $S$ with $\bar{S}$ below, or $S^*$ with $\bar{S^*}$ with respect to which assumption does not hold).
+We may assume without loss of generality that the Greedy Algorithm puts $v \in S$ and the optimal solution puts $v \in S^*$ after this iteration (if this does not hold simply interchange $S$ with $\bar{S}$ below, or $S^*$ with $\bar{S^*}$ with respect to which assumption does not hold).
 \paragraph{}
@@
 Thus $|\delta(S^*)|$ has size at most that of an optimal cut in $G_i$.
-Now by our choice of $V_i$, using $2$-approximation algorithm $\cB$, $2|\delta(V_i)|$ is at the size of an optimal cut in $G_i$.
+Now by our choice of $V_i$, using $2$-approximation algorithm $\cB$, $2|\delta(V_i)|$ is at most the size of an optimal cut in $G_i$.
 Hence we have
@@
 Let $\cO$ be an optimal solution.
-Observe from Lemma \ref{lemma:3b1} that $c(\cS^{i}) \geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{i-1})$.
+Observe from rewriting Lemma \ref{lemma:3b1} that $c(\cS^{i}) \geq \frac{1}{2k} c(\cO) + (1- \frac{1}{2k})c(S^{i-1})$.
 We will apply this inequality inductively to achieve the bound:
@@
 We first observe that this $IP$ is easy to solve on trees.
-Indeed let $T$ be a tree, and for every $uv \in E(T)$ let $S^{uv}_u$ denote the set of vertices in connected component of $T-uv$ containing $u$, and define $S^{uv}_v$ similarly.
+Indeed let $T$ be a tree, and for every $uv \in E(T)$ let $S^{uv}_u$ denote the set of vertices in the connected component of $T-uv$ containing $u$, and define $S^{uv}_v$ similarly.
 By definition, $\delta(S^{uv}_u) =  \{e\} = \delta(S^{uv}_v)$.
@@
 Now let $D(S_i) = \{ e_j \in \delta(S) : u_j \in S_i \}$ be the set of edges in $\delta(S)$ incident with $S_i$.
-Notice that the set of $D(S_i)$ partition $\delta(S)$.
+Notice that $\{D(S_i): i = 1\dots \ell\}$ is a partition of $\delta(S)$.
 \paragraph{}
@@
 &= \sum_{i=1}^\ell f(\bigcup_{e_j\in D(S_i)} V\backslash S^{e_j}_{u_j}). \\
-&\leq \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(V\backslash S^{e_j}_{u_j}) &\text{by subadditivity} \\ t
+&\leq \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(V\backslash S^{e_j}_{u_j}) &\text{by subadditivity} \\ 
 &= \sum_{i=1}^\ell \sum_{e_j \in D(S_i)} f(S^{e_j}_{u_j}) &\text{by symmetry} \\
@@
 \begin{align*}
-d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_, S^u_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
+d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_i, S^v_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
 &\leq \max\{2^i,0\} + \max\{2^i,0\} +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
@@
 For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition.
-Since we sample $2^\beta$ uniformly at random, the probability of this happening is simply $\frac{1}{2^{i-1}}$ times the length of the interval intersecting $[2^{i-1},2^i]$ with $[d(c,u), d(c,v)]$.
+Since we sample $2^\beta$ uniformly at random, the probability of this happening is simply $$\frac{1}{2^{i-1}}$$ times the length of the interval intersecting $[2^{i-1},2^i]$ with $[d(c,u), d(c,v)]$.
 More formally:

### 2017-02-13 CO750-ApproximationAlgorithmsTopics/lectures/lecture_11.tex
conv opt a1 p3

@@
 
-\lhead{2017-mm-DD}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture n} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-02-09}
+\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 11} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-
+(This lecture is a continuation of Lecture 10)
+\paragraph{Recall}
+Given an optimal Steiner Tree $S^*$ we construct witness tree $W$ by the following process:
+\begin{enumerate}
+\item Turn $S^*$ into a binary tree by adding $0$ cost edges.
+\item Choose one edge incident to each Steiner Node with probability $\frac{1}{2}$ and call the set of edges chosen $\tilde{B}$.
+\item Set $W = \{ uv \in {R \choose 2} : |P_{uv} \cap \tilde{B}| = 1\}$.
+\item For each $e \in S^*$ set $w(e) = \{ uv \in W : e \in P_{uv}\}$
+\end{enumerate}
+The Deletion Process is a follows: When a component $C^*$ is sampled, we delete some edges from $W$ such that $Pr[e \in W \text{ is deleted in each iteration}] \geq \frac{1}{M}$.
+When $w(e)$ are deleted, delete $e \in S^*$.
+\paragraph{}
+We hope to answer the question: How many iterations until $e \in S^*$ is deleted?
+\paragraph{Lemma $3$} For any $e \in S^*$ at level $k_e \leq |R| - 1$ from the root of the binary tree, one has
+$$Pr[|w(e)| = q] = \begin{cases}
+\frac{1}{2^q}, &1\leq q < k_e \\
+\frac{2}{2^q}, &q=k_e \\
+0, &\text{otherwise}.
+\end{cases}$$
+\paragraph{Proof of Lemma $3$}
+Let $e=v_0v_1$.
+If $(v_{q-1}, v_q)$ is the first edge of $\tilde{B}$ on the path $v_0$ to $r$ then $|w(e)| = q$.
+This happens with probability $\frac{1}{2^q}$.
+If $(v_{k_e - 1}, v_{k_e})$ is the first edge of $\tilde{B}$ on this path, OR if no edges of $\tilde{B}$ are on this path then $|w(e)| = k_e$.
+Each happens with probability $\frac{1}{2^{k_e}}$, so in total we have probability $\frac{1}{2^{k_e}} + \frac{1}{2^{k_e}} = \frac{2}{2^{k_e}}$.$\blacksquare$
+\paragraph{Lemma $4$}
+Let $\tilde{W} \subseteq W$. The the expected number of iterations until all edges of $\tilde{W}$ are removed is at most $H_{|\tilde{W}|} \cdot M$.
+\paragraph{Proof Idea Sketch for Lemma $4$}
+This situation is analogous to the coupon collector's problem: You have $n$ coupons, and you sample one in each iteration (with repetition).
+How many iterations $T$ do we wait (in expectation) until we collect all our coupons?
+You can define a random variable $t_i := $ time needed to collect $i$-th coupon after you have collected $i-1$ coupons.
+Then
+$$Pr[\text{collect new coupon } i] = \frac{n -(i-1)}{n}.$$
+One proves that $E[T] = \sum_{i=1}^n E[t_i] = \frac{n}{n} + \frac{n}{n-1} + \dots + \frac{n}{1} = n H_n$.
+\paragraph{Proof of Lemma $4$}
+We proceed by induction $q$ where $m_q$ is the number of iterations until $\tilde{W}$ is deleted ($q = |\tilde{W}|$).
+In the Base Case: $q=1$.
+Each edge is deleted with probability $\geq \frac{1}{M}$ which implies that $m_1 \leq M$.
+\paragraph{}
+Inductive Case: If we condition on the event that $i \in \{0,\dots,q\}$ edges are deleted in the first iteration then we need in expectation at most $m_{q-i}$ more iterations to delete the rest.
+Let $\lambda_i$ be the probability that at least $i$ edges are deleted in the first iteration.
+We obtain:
+\begin{align*}
+m_q &\leq 1 + \sum_{i=0}^q Pr[i \text{ edge are deleted in 1st iteration}] m_{q-i} \\
+&\leq 1 + \sum_{i=0}^q (\lambda_i - \lambda_{i+1}) \cdot m_q &(\lambda_0 = 1, \lambda_{q+1} = 0, m_0 = 0) \\
+&\leq 1 + M \sum_{i=1}^q \lambda_i(H_{q-i} - H_{q-i + 1}) +\lambda_1H_qM +(1-\lambda_1)m_q \\
+&\leq 1 - \frac{1}{q}M\sum_{i=1}^q\lambda_i + \lambda_1H_qM(1-\lambda_1)m_q \\
+&\leq \lambda_1 H_q M + 1-\lambda_1 m_q.
+\end{align*}
+Thus we have $m_q \leq H_q M.$ $\blacksquare$
+\paragraph{Theorem}
+For every fixed $k$ there exists a $\ln(4)$-approximation algorithm for $k$-restricted Steiner Tree.
+\paragraph{Proof}
+For each $e \in S^*$, let $D(e) = \max\{t | e \in S^t\}$. Then we have
+\begin{align*}
+E[D(e)] &= \sum_{q=1}^{k_e} Pr[|w(e)|=q] \cdot E[D(e) | |w(e)|=q] \\
+&\leq \sum_{q=1}^{k_e} Pr[|w(e)|=q] \cdot H_q \cdot M \\
+&\leq \sum_{q=1}^{k_e-1}\frac{1}{2^q}H_qM  \frac{2}{2^{k_e}}H_{k_e}M \\
+&\leq \sum_{q\geq 1} \frac{1}{2^q}H_q M \\
+&= \ln(4) M.
+\end{align*}
+Finally we see:
+\begin{align*}
+E[\sum_{t\geq 1} c(C^t)] &\leq \frac{1}{M} \sum_{t\geq 1} E[opt_k^t] \\
+&\leq \frac{1}{M} \sum_{t \geq 1} E[c(S^t)] \\
+&\leq \frac{1}{M} \sum_{e\in S^*} E[D(e)]c_e \\
+&=\ln(4) \cdot c(S^*) \\
+&= \ln(4) opt_k.\blacksquare
+\end{align*}
+\paragraph{Corollary}
+There exists a $\ln(4)+epsilon \leq 1.39$-approximation algorithm for Steiner Tree.
+\paragraph{Examples}
+Quasi-bipartite instances: instances where Steiner nodes are not adjacent to each other.
+Consider $S^*$ in this case.
+Each component of $S^*$ is a star graph centred at a Steiner Node.
+For each component of $S^*$ we take uniformly at random all but one edge in $\tilde{B}$.
+Let $\bar{k}$ be the number of terminals in one such component.
+Let $e$ be an edge of this component.
+Then
+$$|w(e)| = \begin{cases}
+1, &\text{with probability }1-\frac{1}{\bar{k}} \\
+\bar{k}-1, &\text{with probability } \frac{1}{\bar{k}}.
+\end{cases}$$
+So $E[D(e)] \leq \frac{1}{\bar{k}}H_{\bar{k}-1}M + (1-\frac{1}{\bar{k}} H_1 M \leq \frac{73}{60}M$.
+\paragraph{Final Remarks}
+The algorithm can be derandomized.
+There is a theorem of Goemans, Olver, Rothvoss, and Zenklusen from $2012$ which states the integrality gap is at most $1.39$ for the Directed Cut Relaxation Linear Program.
+Open question: Can one prove $<2$ bound on the integrality gap of the Bidirected Cut Relaxation?
+\paragraph{Theorem}
+(Feldman, Koenemann, Olver, S' '14) BCR and DCR are equivalent for all instances such that Steiner Nodes induce a graph of maximum degree $2$.
 \end{document}

### 2017-02-13 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
conv opt a1 p3

@@
 as desired. $\blacksquare$
+%Q2
 \section{}
@@
 Thus $g$ is a composition of affine functions with a concave function and hence $g$ is concave.$\blacksquare$
+%Q3
+\section{}
+\paragraph{}
+Let $f : \R^n \rightarrow \R$ be a convex function and let $g : \R^n \rightarrow \R$ be a concave function with $\text{dom } f = \text{dom } g = \R^n$.
+Suppose that $g(x) < f(x)$ for all $x$.
+Let $e(f) = \{(x,t) \in \R^n \times \R: t \geq f(x) \}$ be the epigraph of $f$.
+We know from our notes that $e(f)$ is convex.
+Let $b(f) = \{(x,t) \in \R^n \times \R: t \leq g(x)\}$ denote the ``below" graph (sorry I don't know the technical name of this object) of $g$.
+We claim that $b(g)$ is convex.
+\paragraph{}
+Let $(x_1, t_1), (x_2, t_2) \in b(g)$. Then we have
+\begin{align*}
+-g(\frac{1}{2}(x_1 + x_2)) &\leq \frac{1}{2}(-g(x_1) - g(x_2)) &\text{ since $-g$ is convex} \\
+&=\frac{-1}{2}(g(x_1) + g(x_2)).
+\end{align*}
+Multiplying by $(-1)$, then using that $x_1, t_1), (x_2, t_2) \in b(g)$ we obtain
+$$g(\frac{1}{2}(x_1 + x_2)) \geq \frac{1}{2}(g(x_1) + g(x_2)) \geq \frac{1}{2}(t_1 + t_2).$$
+Thus $\frac{1}{2}[(x_1,t_1) + (x_2,t_2)] \in b(g)$, that is to say $b(g)$ is convex.
+\paragraph{}
+Now we claim that $e(f) \cap b(g) = \emptyset$. Suppose for a contradiction that there exists $(x,t) \in e(f) \cap b(g)$. Thus we have
+$$t \geq f(x) \quad \text{and}\quad t \leq g(x).$$
+But then $f(x) \leq g(x)$, contradicting that $f(x) > g(x)$ for all $x$. Thus $e(f)$ and $b(g)$ are disjoint.
+\paragraph{}
+Using the Separating Hyperplane Theorem as written in section $2.5.1$ of Boyd and Vandenberghe, on $e(f)$ and $b(g)$ we observe that there exists $(y,a) \in \R^n \times \R$ and $b \in \R$, a hyperplane, which separates $e(f)$ and $b(g)$:
+\begin{align}
+y^Tx + at &\geq b &\forall (x,t) \in e(f) \label{eq:3.1}\\
+y^Tx + at &\leq b &\forall (x,t) \in b(g) \label{eq:3.2}.
+\end{align}
+Now we verify that $a\geq 0$. Let $x \in \R^n$. Then applying $\ref{eq:3.1}$ to $(x,f(x))$ and $\ref{eq:3.2}$ to $(x,g(x))$ and applying transitivity through $b$ we have
+$$y^Tx + af(x) \geq y^Tx + ag(x)$$
+Thus
+$$a(f(x) - g(x)) \geq 0.$$
+Since $f(x) > g(x)$ this implies
+$$a \geq 0.$$
+Now observe that
+$$y^Tx + af(x) \geq b$$
+implies
+$$ f(x) \geq \frac{b}{a} - \frac{y^Tx}{a}$$
+and
+$$ y^Tx + ag(x) \leq b$$
+implies 
+$$g(x) \leq \frac{b}{a} - \frac{y^Tx}{a}.$$
+Let $h(x): \R^n \rightarrow \R$ be the affine function given by
+$$h(x) =\frac{-1}{a}(y^Tx - b)$$
+Then we have immediately
+$$g(x) \leq h(x) \leq f(x)$$
+for all $x \in \R^n$ as desired. $\blacksquare$
 \end{document}

### 2017-02-14 CO750-ApproximationAlgorithmsTopics/assignments/assignment_01.tex
conv opt a1 p3
fixed 4

@@
 \lhead{2017-02-09}
-\rhead{William Justin Toth CO750-Approximation Algorithms Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CO750-Approximation Algorithms Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \paragraph{}
-It remains to verify the approximation factor of $O(1)$.
-We do so by induction on the number of iterations the algorithm needs to run to return a solution.
-In the base case, that is when $\delta = 0$, the algorithm returns an $O(1)$-approximate Steiner Tree connecting $R \cup \{r\}$ and installs capacity of $1$ on each edge.
-By observing that the optimal solution in this case is a Steiner Tree connecting $R \cup \{r\}$ with capacity $1$ on each edge we see that the approximation factor holds in this case.
-\paragraph{}
-Suppose that $\delta > 0$ and for any problem instance with maximum demand $2^{\alpha}$ with $0 \leq \alpha < \delta$ the algorithm returns an $O(1)$ approximate solution.
-We will describe solutions by an ordered pair $(T, x)$ where $T$ is a tree and $x : T \rightarrow \R_+$ maps edges of $T$ with their corresponding capacity.
+It remains to verify the approximation factor of $O(1)$.We will describe solutions by an ordered pair $(T, x)$ where $T$ is a tree and $x : T \rightarrow \R_+$ maps edges of $T$ with their corresponding capacity.
 Then the objective function, which we denote by $f$, in our problem maps feasible solutions $(T,x)$ as follows:
@@
 Let $(T,x)$ be the solution returned by running our algorithm, and let $(T^*, x^*)$ be the optimal solution.
-Let $T_{\delta}$ be the $O(1)$-approximate Steiner Tree connecting $R_\delta \cup \{r\}$ in the first iteration (iteration $\delta$) of our algorithm.
-Let $(\bar{T}, x)$ denote the solution returned by our algorithm running on the residual problem with input graph $G_{\delta-1}$ (the resulting graph after contracting $T_{\delta}$ into $r$).
-Then $T = T_\delta \cup \bar{T}$ and
-$$f(T,x) = f(T_\delta, x) + f(\bar{T},x).$$
-If we let $(\bar{T}^*, \bar{x}^*)$ denote the optimal solution for the residual problem with graph $G_{\delta - 1}$ then by our inductive hypothesis
-$$f(T,x) \leq f(T_\delta,x) + O(1)f(\bar{T}^*, \bar{x}^*).$$
-Now we observe that $T^*$ contains a Steiner Tree $T'$ with terminals $R_\delta \cup \{r\}$.
-Since $T_\delta$ is an $O(1)$-approximate optimal Steiner tree with terminals $R_\delta \cup \{r\}$ we have $c(T_\delta) \leq O(1) c(T')$.
-Since $T'$ necessarily installs capacities $2^\delta$ on all of its edges, $x_e^* = 2^\delta$ for all $e \in T'$.
-To see that a capacity of $2^\delta$ is necessary, suppose some edge $e$ in $T'$ has $x^*_e < 2^\delta$.
-Since $T'$ is a Steiner Tree, some terminal $v \in R_\delta$ uses $e$ on its unique path to $r$.
-But then an $rv$-flow along this path has value at most $x^*_e < 2^\delta$ contradicting feasibility of $(T^*, x^*)$.
-Thus $x_e^* = 2^\delta$ for all $e \in T'$.
-Hence we have
-$$f(T_\delta, x) = \sum_{e \in T_\delta} c_e x_e = 2^\delta c(T_\delta) \leq O(1)2^\delta c(T') = O(1)f(T', x^*) \leq O(1) f(T^*, x^*).$$
-Thus we see that
-$$f(T,x) \leq O(1) f(T^*,x^*) + O(1) f(\bar{T}^*, \bar{x}^*).$$
-Now consider the graph that results from contracting all edges of $T_\delta \cap T^*$ into $r$ in the tree $T^*$.
-If any cycles form then do the following: while there still exists a cycle $C$ take the edge $e = \text{argmin}_{e\in C} x^*_e$ and remove it.
-Call the resulting graph $S^*$.
+Let $T_{i}$ be the $O(1)$-approximate Steiner Tree connecting $R_i \cup \{r\}$ in iteration $i$ of our algorithm.
+Similarly let $T^*_i$ denote the sub-tree of $T^*$ which spans $\{r\} \cup R_i$ after contracting $T^*_{i+1} \dots T^*_\delta$.
+Observe that the capacities on edges, $e$ of $T_i$ or $T^*_i$ are necessarily $2^i$, as otherwise a flow of value $2^i$ could not be routed along that edge to a vertex, $v$, in $R_i$ which $e$ is on the unique path to in $T_i$ or $T^*_i$.
 \paragraph{}
-We claim that $(S^*, x^*)$ is feasible for the residual problem on $G_{\delta-1}$.
-Indeed $S^*$ is a tree since we trimmed one edge from all cycles.
-Since trimming cycles preserves connectivity, and we contracted edges into $r$, all $rv$-paths in $T^*$ are preserved in $S^*$.
-Hence for all $v \in R\backslash R_\delta$ there exists an $rv$-path in $S^*$.
-Now we claim that the capacities on any $rv$-path still admit feasible flows after converting from $T^*$ to $S^*$.
+Thus we observe that $f(T_i,x) = 2^i |T_i|$ and $f(T^*_i, x^*) = 2^i |T^*_i|$. Now in the first iteration (iteration $\delta$), $T_\delta$ is computed as an $O(1)$-approximation of $T^*_\delta$. Thus
+$$2^\delta |T_\delta| = f(T_\delta, x) \leq O(1) f(T^*_\delta, x^*) = 2^\delta O(1) |T^*_\delta|.$$
+Now consider the sub-tree that forms by taking the iterations of $$\bigcup_{j=\delta}^i T^*_j$$
+and contracting $T_\delta \dots T_{i+1}$ into $r$.
+If any cycles are formed then do the following: while there still exists a cycle $C$ take the edge $e = \text{argmin}_{e\in C} x^*_e$ and remove it.
+Call the resulting tree $S^*$.
+\paragraph{}
+We claim that $(S^*,x^*)$ is feasible on $G_i$ for the nodes in $V(G_i) \cap R_i$, the graph which results from contracting $T_\delta, \dots, T_{i+1}$ into $r$.
+Since trimming cycles preserves connectivity, and we contracted edges into $r$, all $rv$-paths in $V(G_i) \cap R_i$ are preserved in $S^*$.
+Hence for all $v \in V(G_i) \cap R_i$ there exists an $rv$-path in $S^*$.
+Now we claim that the capacities on any $rv$-path still admit a feasible flow after converting to $S^*$.
 Observe that contraction will not decrease the minimum capacity.
-Further observe that deleted edges had minimum capacity on a cycle, so any flow that used a deleted edge $uv$ could send the same flow through the $uv$-path remaining from the cycle $uv$ was deleted from.
-Hence the flows on $S^*$ under capacities $x^*$ can still meet all demands in $R\backslash R_\delta$.
-Therefore $(S^*,x^*)$ (with $x^*$ restricted to $E(G_{\delta-1})$) is feasible for $G_{\delta -1}$.
-Hence, since $(\bar{T}^*, \bar{x}^*)$ is optimal for $G_{\delta -1}$:
-$$f(\bar{T}^*, \bar{x}^*) \leq f(S^*, x^*) \leq f(T^*, x^*).$$
-Therefore
+Further observe that deleted edges had minimum capacity on a cycle, so any flow that used a deleted edge $uv$ could send the same flow through the $uv$-path remaining in $S^*$ instead.
+Hence the flows on $S^*$ under capacities $x^*$ still meet the demands in $v \in V(G_i) \cap R_i$.
+Therefore $(S^*, x^*)$ feasible on $G_i$.
+Thus by our choice of $T_i$,
+$$f(T_i, x) \leq O(1) f(S^*,x^*).$$
+That is $$2^i |T_i| \leq 2^i O(1)|S^*| \leq 2^iO(1) (\sum_{j=\delta}^i |T^*_j|).$$
+\paragraph{}
+Summing these inequalities over $i$ we obtain:
 \begin{align*}
-f(T,x) &\leq O(1)f(T^*,x^*) + O(1)f(\bar{T}^*,\bar{x}^*) \\
-&\leq O(1)f(T^*,x^*) + O(1)f(T^*,x^*) \\
-&= 2O(1)f(T^*,x^*)\\
-&= O(1)f(T^*,x^*).
+f(T,x) &= \sum_{i=1}^\delta 2^i|T_i| \\
+&\leq \sum_{i=1}^\delta 2^i O(1)(\sum_{j=\delta}^i |T^*_j|)\\
+&= O(1) \sum_{i=1}^\delta \sum_{j=i}^\delta 2^i|T^*_j| \\
+&= O(1) \sum_{i=1}^\delta (\sum_{j=1}^{i}(2^j) |T^*_j| &\text{reindexing, and collecting coefficients on each $T^*_j$} \\
+&= O(1) \sum_{i=1}^\delta 2^{i+1} |T^*_i| \\
+&= 2O(1) \sum_{i=1}^\delta 2^i |T^*_i| \\
+&= 2O(1) \sum_{i=1}^\delta f(T^*_i, x^*) \\
+&= O(1) f(T^*, x^*).
 \end{align*}
-So the approximation factor $O(1)$ holds in the inductive case. Thus the algorithm given is a $O(1)$-approximation algorithm.$\blacksquare$
-
+Thus the $O(1)$ approximation factor holds, and hence the given algorithm is an $O(1)$-approximation algorithm.$\blacksquare$
 %Q5

### 2017-02-14 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
Conv Opt A1 q4

@@
 for all $x \in \R^n$ as desired. $\blacksquare$
+
+%Q4
+\section{}
+\paragraph{}
+Let $P = \{ x| Ax \leq b\}$ be a polytope, where $A \in \R^{m\times n}$, $b \in \R^m_{>0}$.
+Let $a_i$ denote the $i$-th row of $A$.
+Let $[m] = \{1,\dots, m\}$.
+Let $C = \text{conv} \{\frac{a_i}{b_i} : i\in [m]\}$ be the convex hull of $\frac{a_i}{b_i}$'s.
+We want to show that $P^* = C$, where $P^*$ denotes the polar set of $P$.
+\paragraph{}
+We will use the theorem from class that $P^{**} = P$.
+With this theorem it is equivalent to show that $P = C^*$.
+First we will rewrite $P$ as $P = \{x| a_i^Tx \leq b_i \forall i\in[m]\}$.
+Observe that
+$$C^* := \{ x: x^Ty \leq 1\forall y \in C\}.$$
+Observe that $\forall y \in C$ there exists $\lambda_i$ for $i\in [m]$ such that $y = \sum_{i\in[m]} \lambda_i \frac{a_i^T}{b_i}$, $\sum_{i\in [m]}\lambda_i = 1$, and $\lambda \geq 0$.
+This is the definition of convex hull.
+Hence substituting into our definition of $C^*$ we see
+$$C^* = \{x : x^T(\sum_{i} \lambda_i \frac{a_i^T}{b_i}) \leq 1,\ \forall \lambda \geq 0, \sum_i \lambda_i = 1\}.$$
+Rewriting slightly this is equivalent to
+$$C^* = \{x : \sum_{i}\lambda_i \frac{a_i}{b_i}x \leq 1,\ \forall \lambda \geq 0, \sum_i\lambda_i = 1\}.$$
+It is not hard to see that such $C^* = P$. We will write it out formally below for the sake of completeness.
+\paragraph{}
+Let $x\in C^*$. Then if we for each $i \in [m]$ define the vector $\lambda^i$ as
+$$\lambda^i_j = \begin{cases}
+1, &i=j \\
+0, &i\neq j.
+\end{cases}$$
+Then $\lambda^i \geq 0$ and $\sum_j \lambda^i_j = \lambda^i_i = 1$. Since $x \in C$ we have
+$$\sum_{j}\lambda^i_j \frac{a_j}{b_j} x \leq 1$$
+and hence
+$$\frac{a_i}{b_i} x \leq 1.$$
+Thus for all $i \in [m]$, $a_i x\leq b_i$ ($b_i > 0$ implies inequality does not change order) and therefore $x \in P$.
+Now if we let $x\in P$, and let $\lambda \geq 0$ such that $\sum_i \lambda_i = 1$ then we need to show that $\sum_i \lambda_i \frac{a_i}{b_i} x \leq 1$.
+Since $x \in P$, $a_i x \leq b_i$ for all $i \in [m]$.
+Dividing by $b_i$ and multiplying by $\lambda_i \geq 0$ we obtain
+$$\lambda_i\frac{a_i}{b_i} x \leq \lambda_i \quad \forall i \in [m].$$
+Summing these inequalities over $i\in[m]$ obtains the desired bound:
+$$\sum_i \lambda_i \frac{a_i}{b_i}x \leq \sum_i \lambda_i = 1.$$
+Thus $P= C^*$ and by our theorem, $P^* = C^{**} = C$. $\blacksquare$
 \end{document}

### 2017-02-14 CO750-ApproximationAlgorithmsTopics/lectures/lecture_12.tex
scribe lecture 12

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-02-14}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 12} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\section{TSP}
\paragraph{}
In the Traveling Salesman Problem (TSP) we are given an undirected graph $G = (V,E)$ and a cost function $c: E \rightarrow \R_+$. We are to find a tour (Hamiltonian Cycle) visiting all vertices of $V$, of minimum cost.
\paragraph{Proposition}
TSP in general is ``in-approximable" (unless $P=NP$).
\paragraph{Proof}
Reduction from Hamiltonian Cycle ($NP$-hard). Suppose there exists an $\alpha(n)$-approximation algorithm for TSP. Given a graph $G=(V,E)$ we set $c_e = 1$ for all $e\in E$ and set $c_e = n\alpha(n) + 1$ for all $e \in \bar{E}$ (the complement of $E$). Run our TSP algorithm on $(V,E\cup \bar{E})$. The solution has cost at most $n\alpha(n)$ if and only if $G$ has a Hamiltonian Cycle.$\blacksquare$
\section{Metric TSP}
Now we suppose we are given a metric graph instead. Going forward we will implicitly assume this without writing it.
\begin{theorem} (Christofides '76) There is a $\frac{3}{2}$-approximation algorithm for TSP.
\end{theorem}

\begin{proof}
Consider the following algorithm for TSP:
\begin{enumerate}
\item Find a minimum cost spanning tree $F$ of $G$.
\item Let $T\subseteq V$ be the set of odd degree vertices in in $F$. It is easy to see by handshaking that $|T|$ is even.
\item Add minimum-cost $T$-join $H$.
\item Shortcut $F\cup H$ (without increasing cost, by triangle inequality) into a tour.
\end{enumerate}
Recall that a $T$-join is a set of edges such that only the vertices in $T$ have odd degree, and $T$-joins can be found in polynomial time. We have $c(F) \leq OPT$ since OPT spans $G$. We also have $H \leq OPT/2$. If we start at some vertex in $T$ and follow the cycle given by $OPT$ to the next vertex in $T$ we obtain a path. Traveling further to the next vertex in $T$, obtaining another a path, and proceeding this way yield two $T$-joins depending on which way around the cycle you decided to go when starting this process. Hence $OPT$ contains two feasible $T$-joins and the claimed bound on $H$ holds. Therefore $c(F\cup H) \leq c(F) + c(H) \leq \frac{3}{2} OPT$.
\end{proof}
\paragraph{Big Open Question}
Can one improve over $\frac{3}{2}$? There is an $LP$-relaxation for TSP that seems to be promising: The Held-Karp relaxation $(HK)$:
\begin{align*}
\min &\sum_e c_e x_e \\
\text{s.t.} x(\delta(v)) &= 2 &\forall v \in V \\
x(\delta(S)) &\geq 2 &\forall S\subseteq V \\
x&\geq 0.
\end{align*}
\begin{lemma}
The integrality gap of $(HK)$ is $\leq \frac{3}{2}$ and $\geq \frac{4}{3}$.
\end{lemma}
\begin{proof}
For the lower bound consider graph formed by taking two $K_3$ and connecting each vertex in one $K_3$ to exactly one vertex in the other $K_3$ via a path of length $n$ with all edge weights one. Assigning $x_e = \frac{1}{2}$ for the $K_3$ edges and $x_e = 1$ otherwise is clearly feasible for $(HK)$. Any integral tour must use each $n$ length path at least once, and in particular must cross one of them twice. Thus an integral tour has cost $\frac{4}{3}n-2$.
\paragraph{}
To prove the upper bound, we'll use the solution output by Christofides algorithm: $F \cup H$. Recall (Edmonds '70) The convex hull of incidence vectors of spanning trees of $G = (V,E)$ is given by $(MST):$
\begin{align*}
x(E) &n-1 \\
\sum_{e \in E(U)} x_e &\leq |U| - 1 &\forall \emptyset\neq U \subseteq V\\
x&\geq 0.
\end{align*}
Also observe that if $x$ is feasible for $(HK)$ then $\bar{x} = \frac{n-1}{n}x$ is feasible for $(MST)$. This is an immediate and easy calculation.
\paragraph{}
Thus if we let $x^*$ be optimal for $(HK)$ then $$c(F) \leq c(x^*).$$
\paragraph{}
Now we use (Edmonds and Johnson '73) which says that given a graph $G=(V,E)$ with costs $c:E\rightarrow \R_+$ and $T\subseteq V$ that the cost of a minimum cost $T$-join on $G$ is equal to the optimal solution to $(T-JOIN)$:
\begin{align*}
\min &\sum_e c_ex_e \\
\text{s.t.} x(\delta(U)) &\geq 1 &\forall U \subseteq V : |U\cap T| \text{ is odd}\\
x&\geq 0.
\end{align*}
Observe that if $x$ is feasible for $(HK)$ then $\frac{x}{2}$ is feasible for $(T-JOIN)$ (trivially). Thus we see that
$$c(H) \leq \frac{c(x^*)}{2}.$$
Therefore $c(F\cup H) \leq c(F) + c(H) \leq \frac{3}{2}c(x^*).$
\end{proof}
\paragraph{Another Big Open Question}
Conjecture: The integrality gap of $(HK)$ is $\leq \frac{4}{3}$.
\paragraph{A related proposition on $T$-joins}
(Edmonds and Johnson '73) The convex hull of incidence vectors of $T$-joins is given by:
$$|A| - x(A) + x(\delta(U\backslash A)) \geq 1 \quad \forall U\subseteq V, A\subseteq \delta(U) : |U\cap T| + |A| \text{ is odd}.$$
\section{Special Case: Graphic TSP}
\paragraph{}
Given an undirected graph $G=(V,E)$, a graphic TSP instance is the TSP instance defined on the metric closure of $G$ (edge weights of $G$ are all $1$). We can reformulate this problem as: 
\paragraph{}
Given an unweight $G=(V,E)$, find an Eulerian spanning (multi) subgraph of $G$ with minimum number of edges.
\paragraph{Improvements over Christofides for graphic TSP}
\begin{itemize}
\item(Oveis, Gharan,  Saber, Singh '11) A tiny improvement. Strategy was to solve $(HK)$ obtaining $x^*$, then express $\frac{n-1}{n}x^*$ as a convex combination of spanning trees, then sample a tree according to the maximum entropy distribution, then add a minimum cost $T$-join as usual.
\item They conjecture that this above algorithm should give a $<\frac{3}{2}$-approximation for the general case of metric TSP.
\item (Monke and Svensson '11) Removable Pairs
\begin{itemize}
\item (Mucha '12) A better analysis of same algorithm leading for $\frac{13}{9}$-approximation 
\end{itemize}
\item (Sebo, Vygen '12) Ear Decomposition plus Removable Pairs leads to a $\frac{7}{5}$-approximation.
\end{itemize}
\end{document}

### 2017-02-16 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
partial progress

@@
 with $\text{dom} f = \R^n_{>0}$. We will compute the Hessian matrix. First observe that
-$$\frac{\partial f}{\partial x_i} f = x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p}$$
+$$\frac{\partial f}{\partial x_i} f = x_i^{p-1}(\sum_{k=1}^n x_k^p)^\frac{1-p}{p}$$
 for $i = 1, \dots, n$. Now for $j = 1, \dots, n$
@@
 \frac{\partial f}{\partial x_j}\frac{\partial f}{\partial x_i} f &= \frac{\partial f}{\partial x_j}   x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p} \\
-&= \begin{cases}
-0, &\text{ if } i=j \\
-(p-1)x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p}[1- \frac{x_i^{p-1}}{\sum_{i=1}^n x_i^p}], &\text{if } i \neq j.
+&=(p-1)(\sum_{k=1}^n x_k^p)^\frac{1-2p}{p}\begin{cases}
+-x_i^{p-1}x_j^{p-1}, &i\neq j \\
+x_i^{p-2}\sum_{k=1}^n x_k^p  -x_i^{p-1}x_j^{p-1}, &i=j.
 \end{cases}
 \end{align*}
-Observe that if $p \geq 1$ each factor in the second case is non-negative when $x \in \R^n_{>0}$.
-Hence $\frac{\partial f}{\partial x_i} f  \geq 0$ for all $x \in \text{dom f}$ for all $i,j = 1, \dots, n$.
-Thus each entry of the Hessian is non-negative, and so the Hessian is PSD.
-Thus by the second order condition $f$ is convex.
+We claim that the Hessian is positive semidefinite. We prove this directly from the definition, let $y \in \R^n$:
+\begin{align*}
+y^T\nabla^2f(x)y &= \sum_{i=1}^n \sum_{j=1}^n \nabla^2f(x)_{ij} y_i y_j \\
+&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[\sum_{i=1}^nx_i^{p-2}(\sum_{k=1}^nx_k^p)y_i^2 - \sum_{i=1}^n\sum_{j=1}^n x_i^{p-1}x_j^{p-1}y_i y_j ] \\
+&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(\sum_{i=1}^nx_i^{p-2}y_i^2) - (\sum_{i=1}^n x_i^{p-1}y_i)(\sum_{i=1}^n x_i^{p-1} y_i) ] \\
+&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(\sum_{i=1}^nx_i^{p-2}py_i^2) - (\sum_{i=1}^n x_i^\frac{p}{2}x_i^\frac{p-2}{2}y_i)(\sum_{i=1}^n x_i^\frac{p}{2}x_i^\frac{p-2}{2}y_i)]\\
+&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(||x^\frac{p}{2}||^2||x^\frac{p-2}{2}y||^2 - \langle x^\frac{p}{2}, x^\frac{p-2}{2}y\rangle^2]
+\geq 0.
+\end{align*}
+The inequality follows since $p-1 \geq 1$ (as $p \geq 1$), $(\sum_{k=1}^nx^p)^\frac{1-2p}{p} \geq 0$ (as $x \in \R^n_{>0}$), and 
+$$||x^\frac{p}{2}||^2||x^\frac{p-2}{2}y||^2 \geq \langle x^\frac{p}{2}, x^\frac{p-2}{2}y\rangle^2$$
+by Cauchy-Schwarz. Hence $\nabla^2f(x)$ is positive semidefinite and thus by the Second Order Condition, (since $\R^n_{>0}$ is open) $f$ is convex.
 \paragraph{}
 Now when $p < 1$, by a similar calculation we observe
-$$\frac{\partial f}{\partial x_i} -f = \begin{cases}
-0, &\text{ if } i=j \\
-(1-p)x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p}[1- \frac{x_i^{p-1}}{\sum_{i=1}^n x_i^p}], &\text{if } i \neq j.
-\end{cases}$$
-Again each factor is non-negative, and hence $-f$ is convex by the second order condition. Therefore $f$ is concave. $\blacksquare$
+\begin{align*}
+\frac{\partial -f}{\partial x_j}\frac{\partial f}{\partial x_i} f &=(1-p)(\sum_{k=1}^n x_k^p)^\frac{1-2p}{p}\begin{cases}
+-x_i^{p-1}x_j^{p-1}, &i\neq j \\
+x_i^{p-2}\sum_{k=1}^n x_k^p  -x_i^{p-1}x_j^{p-1}, &i=j.
+\end{cases}
+\end{align*}
+So again calculating the quadratic form $y^T\nabla^2-f(x)y$ we have
+\begin{align*}
+y^T\nabla^2-f(x)y &=(1-p) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(||x^\frac{p}{2}||^2||x^\frac{p-2}{2}y||^2 - \langle x^\frac{p}{2}, x^\frac{p-2}{2}y\rangle^2]\\
+&\geq 0.
+\end{align*}
+Since $p < 1$, $1-p >0$, and so the inequality follows as before. Therefore, by the Second Order Condition $-f$ is convex, and hence $f$ is concave. $\blacksquare$
 \paragraph{}
@@
 \paragraph{}
-We will use the theorem from class that $P^{**} = P$.
+We will use the theorem from class that $P^{**} = P$ when $0 \in P$.
 With this theorem it is equivalent to show that $P = C^*$.
@@
 $$\sum_i \lambda_i \frac{a_i}{b_i}x \leq \sum_i \lambda_i = 1.$$
-Thus $P= C^*$ and by our theorem, $P^* = C^{**} = C$. $\blacksquare$
+Thus $P= C^*$ and by our theorem, $P^* = C^{**} = C$. For this theorem to apply it remains to verify that $0 \in C$.
+\paragraph{}
+Suppose for a contradiction that $0 \not\in C$.
+Since $C$ is a convex hull of a finite set it is closed and bounded.
+Hence, by the separating hyperplane theorem there exists a (non-zero) hyperplane $h \in \R^n$ such that
+$$h^T0 > h^Tx \quad\forall x \in C.$$
+We claim that $h$ is a direction in $P$.
+To see this let $\alpha \in \R$ be a scalar.
+Let $x \in P$.
+We will show that $x + \alpha h$ is feasible, implying that $h$ is a direction in $P$.
+Let $ i \in [m]$ and consider constraint $a_i$ with $b_i$.
+Then
+$$\frac{a_i}{b_i} (x + \alpha h) = \frac{a_i}{b_i}x + \alpha \frac{a_i}{b_i}h < \frac{a_i}{b_i}x \leq 1.$$
+The first inequality follows since $\frac{a_i}{b_i} \in C$ (and thus $\frac{a_i}{b_i} h < h^T0 = 0)$ and the second inequality follows since $x \in P$.
+Thus multiplying through by $b_i$, which will not reverse the order of inequalities since $b_i > 0$, we obtain
+$$\frac{a_i}(x + \alpha h) < b_i.$$
+Since this holds for all $i \in [m]$ this implies that $x + \alpha h \in P$. Hence $h$ is a direction, contradicting that $P$ is bounded. Therefore $0\in C$ as desired. $\blacksquare$
+
+%Q5
+\section{}
+Let $Q \in S^n_{++}$. We consider the quadratic norm $||x||_Q := (x^TQx)^\frac{1}{2}.$
+%pa
+\subsection{a}
+\paragraph{}
+We show that $||x||_Q$ is a norm. First since $Q\in S^n_{++}$ there exists non-singular, symmetric $V$ such that $Q = V^TV$. Observe that
+$$||x||_Q = (x^TQx)^\frac{1}{2} = (x^TV^TVx)^\frac{1}{2} =  ((Vx)^T(Vx))^\frac{1}{2} = ||Vx||_2.$$
+We will show that $||x||_Q$ is non-negative and equal to $0$ only when $x = 0$, that $||t x||_Q = |t|\cdot ||x||_Q$, and that triangle inequality holds.
+\paragraph{Non-negativity}
+Since $||\cdot||_2$ is a norm, $||x||_Q = ||Vx||_2 \geq 0$ and further $||Vx||_2 = ||x||_Q = 0$ only if $Vx = 0$, but then $x = V^{-1}0 = 0$.
+\paragraph{Scaling}
+Observe that $$||tx||_Q = ||V(tx)||_2 = |t|\cdot ||Vx||_2 = |t|\cdot ||x||_Q.$$
+\paragraph{Triangle Inequality}
+Let $x,y \in \R^n$. Then
+$$||x+y||_Q = ||V(x+y)||_2 = ||Vx + Vy||_2 \leq ||Vx||_2 + ||Vy||_2 = ||x||_Q + ||y||_Q.$$
+Therefore $||\cdot||_Q$ is a norm. $\blacksquare$
+%pb
+\subsection{b}
+\paragraph{}
+Let $x \in \R^n$. Then observe that
+\begin{align*}
+||x||_{Q^*} = \sup \{ x^Ty : ||y||_Q \leq 1\} \\
+&= \sup \{x^Ty : ||Vy||_2 \leq 1\} \\
+&= \sup \{x^TV^{-1}Vy : ||Vy||_2 \leq 1\} \\
+&= \sup\{x^TV^{-1}z : ||z||_2 \leq 1\} &\text{assigning $z=Vy$}. 
+\end{align*}
+The supremum is realized when $z = (x^TV^{-1})^T = V^{-1}x$. Therefore
+$$||x||_{Q^*} = x^TV^{-1}V^{-1}x = x^TQ^{-1}x = ||x||_{Q^{-1}}.$$
+Thus our nice formula is $||x||_{Q^*} = ||x||_{Q^{-1}}$.$\blacksquare$
+%pc
+\subsection{c}
+\paragraph{}
+Let $\cE = \{x : x^TQx \leq 1\}.$  Rewriting we have $\cE = \{x : ||x||_Q \leq 1\} = \cB$ where $\cB$ denote the norm ball of $||\cdot||_Q$. Then we have
+$$\cE^* = \{y : y^Tx \leq 1, \forall x \in \cE\} = \{y : y^Tx \leq 1, \forall x \in \cB\}  = \cB^* = \cB_*,$$
+where $\cB_*$ is the norm ball of $||\cdot||_{Q^*}$, and the last equality follows from the notes.
+Rewriting using part $(b)$ we see that $\cB_* = \{x: x^TQ^{-1}x \leq 1\}$.
+That is $\cE^*$ is ellipsoid given by $Q^{-1}$.
 \end{document}

### 2017-02-17 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
p6

@@
 That is $\cE^*$ is ellipsoid given by $Q^{-1}$.
+\paragraph{}
+So suppose $\cE$ is a $2D$-ellipsoid with semi-axes $v_1$ and $v_2$ with corresponding side lengths $a_1$ and $a_2$. Then the spectral decomposition of $Q$ is
+$$Q = \begin{bmatrix} v_1 & v_2 \end{bmatrix} \begin{bmatrix} a_1^{-1} & 0 \\ 0 & a_2^{-1} \end{bmatrix} \begin{bmatrix} v_1^T \\v_2^T\end{bmatrix}.$$
+And further
+$$Q^{-1} =  \begin{bmatrix} v_1 & v_2 \end{bmatrix} \begin{bmatrix} a_1 & 0 \\ 0 & a_2 \end{bmatrix} \begin{bmatrix} v_1^T \\v_2^T\end{bmatrix}$$
+using that $ \begin{bmatrix} v_1 & v_2 \end{bmatrix}^T =  \begin{bmatrix} v_1 & v_2 \end{bmatrix}^{-1}$.
+Hence the polar ellipsoid $\cE^*$ has the same semi-axes as $\cE$, but with reciprocal side lengths.
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}
+\draw (0,0) ellipse (0.5cm and 2cm);
+\draw (0,0) ellipse (2cm and 0.5cm);
+\end{tikzpicture}
+\caption{Ellipsoid and its polar where $v_1 = e_1$ and $v_2 = e_2$ (standard basis vectors), and $a_1 = 2$ while $a_2 = \frac{1}{2}$.
+The polar scales $v_1$ by $\frac{1}{a_1} = \frac{1}{2}$ and scales $v_2$ by $\frac{1}{a_2} = 2$.}
+\end{figure}
+
+%Q6
+\section{}
+%Pa
+\subsection{a}
+\paragraph{}
+Let $\cE_1:= \{P_1 u + q_1 : ||u||_2 \leq 1\}$ and $\cE_2:= \{P_2 u + q_2 : ||u||_2 \leq 1\}$ be given ellipsoids. We want to find a separating hyperplane $(a,b)$ for these ellipsoids, that is:
+\begin{align*}
+a^Tx + b &\geq 0 &\forall x \in \cE_1 \\
+a^Tx+b &\leq 0 &\forall x \in \cE_2.
+\end{align*}
+This can be formulated as an optimization feasibility problem as
+\begin{align*}
+\min\ 0\\
+\text{s.t.}\ a^Tx + b &\geq 0 &\forall x \in \cE_1\\
+a^Tx + b &\leq 0 &\forall x \in \cE_2.
+\end{align*}
+This problem is equivalent to
+\begin{align*}
+\min\ 0\\
+\text{s.t.}\ a^T(-x)&\leq b &\forall x \in \cE_1\\
+a^Tx &\leq -b &\forall x \in \cE_2.
+\end{align*}
+Which is equivalent to
+\begin{align*}
+\min\ 0\\
+\text{s.t.}\  
+\sup\{ a^T(-x) : x \in \cE_1\}&\leq b\\
+\sup\{a^Tx : x\in \cE_2\} &\leq -b.
+\end{align*}
+Equivalently
+\begin{align*}
+\min\ 0\\
+\text{s.t.}\  
+\sup\{ a^T(-P_1 u - q_1) : ||u||_1 \leq 1\}&\leq b\\
+\sup\{a^T(P_2 u + q_2) : ||u||_2 \leq 1\} &\leq -b.
+\end{align*}
+Which is equivalent to
+\begin{align*}
+\min\ 0\\
+\text{s.t.}\  
+a^T(-q_1) + ||P_1a||_2 &\leq b\\
+a^Tq_2 + ||P_2a||_2 &\leq -b.
+\end{align*}
+Setting $A_1 = P_1$, $b_1 = 0$, $c_1=q_1$, $d_1 = b$, and $A_2 = P_2$, $b_2 = 0$, $c_2 = -q_2$, and $d_2 = -b$ we obtain the exact form of a Second Order Conic Program as described in the general formulation of such problems.
+$\blacksquare$
+\subsection{b}
+\paragraph{}
+We will write the Second-Order Cone Program (SOCP):
+\begin{align*}
+\min &f^Tx \\
+\text{s.t.} ||A_ix + b_i||_2 &\leq c^T_i x + d_i, &1\leq i \leq m \\
+Fx &= g,
+\end{align*}
+as a Semidefinite Program (SDP) in inequality form (see section $4.6.2$ of Boyd and Vandenberghe) :
+\begin{align*}
+\min\ &f^Tx \\
+\text{s.t.}\ G + \sum_{i=1}^n x_i F_i &\preccurlyeq 0 \\
+Fx &=g.
+\end{align*}
+Observe that it is enough to show that a constraint of the form $||Ax + b|| \leq c^Tx + d$ can be written as $G + \sum_{i=1}^n x_i F_i \preccurlyeq 0$ as then we could write each constraint $||A_ix + b_i||_2 \leq c^T_i x + d_i$ for $i\in [m]$ as $G^i + \sum_{j=1} x_j F^i_j$, and stack them as blocks as
+$$G = \begin{bmatrix} G^1 & & \\ & \ddots & \\ & & G^m \end{bmatrix} \quad F_i = \begin{bmatrix} F^1_i & & \\ & \ddots & \\ & & F^m_i \end{bmatrix}$$
+to obtain an equivalent single semidefinite constraint in place of $m$ semidefinite constraints.
+\paragraph{Lemma 6b}
+We will need the following lemma using Schur's complement: For any $x\in \R^n$ and $c\in \R$ we have $||x||_2 \leq c$ if and only if
+$$X = \begin{bmatrix} c &x^T \\ x & cI\end{bmatrix} \succcurlyeq 0.$$
+\paragraph{Proof of Lemma 6b}
+If $c = 0$ this is trivial, so we may assume that $c\neq 0$. In this case we invoke Schur's Complement Lemma and thus
+$$X \succcurlyeq 0$$
+if and only if $c \geq 0$ and $tI - \frac{1}{c}xx^T \succcurlyeq 0$.
+Which holds if and only if $c^2I - xx^T \succcurlyeq 0$.
+Now the eigenvalues of $c^2I - xx^T$ are non-negative if and only if $c^2 - ||x||^2_2 \geq 0$ (since $c^2$ is the least eigenvalue of $c^2I$ and $||x^2_2$ is the largest eigenvalue of $xx^T$), which holds if and only if $||x||_2 \leq c$ as desired.
+$\blacksquare$
+\paragraph{}
+Now applying Lemma $6b$ to our SOCP cone constraint we have $||Ax +b||_2 \leq c^Tx + d$ (suppose $x\in\R^n$ and $A \in \R^{r\times n}$) if and only if
+$$\begin{bmatrix} c^Tx + d & (Ax+b)^T \\ Ax + b &(c^T+d)xI \end{bmatrix} \succcurlyeq 0.$$
+Let $a^1, \dots, a^r$ be the rows of $A$. Separating out we obtain:
+$$\begin{bmatrix} d & b^T \\ b & dI \end{bmatrix} + \sum_{i=1}^n x_i \begin{bmatrix} c_i & a^1_i & \dots & a^r_i \\ a^1_i \\ \vdots & &c_iI\\ a^r_i \end{bmatrix} \succcurlyeq 0.$$
+Thus $$G = \begin{bmatrix} d & b^T \\ b & dI \end{bmatrix}$$
+and 
+$$F_i = \begin{bmatrix} c_i & a^1_i & \dots & a^r_i \\ a^1_i \\ \vdots & &c_iI\\ a^r_i \end{bmatrix}$$
+we have
+$$G + \sum_{i=1}^n F_i \succcurlyeq 0$$
+if and only if
+$$||Ax+b||_2 \leq c^Tx + d.$$
+Therefore we have shown that any SOCP can be written as an SDP. $\blacksquare$
 \end{document}

### 2017-02-17 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
p7

@@
 $$\begin{bmatrix} d & b^T \\ b & dI \end{bmatrix} + \sum_{i=1}^n x_i \begin{bmatrix} c_i & a^1_i & \dots & a^r_i \\ a^1_i \\ \vdots & &c_iI\\ a^r_i \end{bmatrix} \succcurlyeq 0.$$
-Thus $$G = \begin{bmatrix} d & b^T \\ b & dI \end{bmatrix}$$
+Thus setting $$G = -\begin{bmatrix} d & b^T \\ b & dI \end{bmatrix}$$
 and 
-$$F_i = \begin{bmatrix} c_i & a^1_i & \dots & a^r_i \\ a^1_i \\ \vdots & &c_iI\\ a^r_i \end{bmatrix}$$
+$$F_i = -\begin{bmatrix} c_i & a^1_i & \dots & a^r_i \\ a^1_i \\ \vdots & &c_iI\\ a^r_i \end{bmatrix}$$
 we have
-$$G + \sum_{i=1}^n F_i \succcurlyeq 0$$
+$$G + \sum_{i=1}^n F_i \preccurlyeq 0$$
 if and only if
@@
 Therefore we have shown that any SOCP can be written as an SDP. $\blacksquare$
+
+%Q7
+\section{}
+\paragraph{}
+Consider the following optimization problem $(P)$:
+\begin{align*}
+\min\ &e^{-x} \\
+\text{s.t.}\ \frac{x^2}{y} &\leq 0,
+\end{align*}
+with domain $\cD = \{(x,y) : y > 0 \}$. The Lagrangian function $g(\lambda)$ is
+$$ g(\lambda) := \inf_{x,y} \{ e^{-x} + \lambda \frac{x^2}{y}\}.$$
+We will study the behaviour of $g$ to understand the value of the dual problem.
+\paragraph{}
+The dual problem $(D)$ is:
+\begin{align*}
+\max\ &g(\lambda) \\
+\lambda &\geq 0.
+\end{align*}
+We consider two cases: when $\lambda = 0$ and when $\lambda > 0$.
+\paragraph{}
+When $\lambda = 0$, 
+$$g(\lambda) = \inf_x e^{-x} = 0.$$
+Now when $\lambda > 0$ observe that $y = \lambda x^4 > 0$ for $x > 0$ and so
+$$\lim_{x \rightarrow \infty, y=x^4} (e^{-x} + \lambda\frac{x^2}{y}) = \lim_{x\rightarrow \infty} (e^{-x} + \lambda\frac{x^2}{\lambda x^4}) = \lim_{x\rightarrow \infty}( e^{-x} + \frac{1}{x^2}) = 0.$$
+Since $g(\lambda) \geq 0$ this implies $g(\lambda) = 0$. Therefore in either case $g(\lambda) = 0$. Thus $(D)$ is equal to:
+\begin{align*}
+\max\ &0 \\
+\lambda &\geq 0.
+\end{align*}
+And thus if we let$d^*$ be the optimal value of $(D)$, then $d^*=0$.
+\paragraph{}
+Now we return our attention to $(P)$. Since $x^2 \geq 0$ for all $x$ and $y > 0$ we observe that
+$$\frac{x^2}{y} \leq 0 \iff x = 0.$$
+Thus all feasible $(x,y)$ for $(P)$ have objective value $e^{-0} = 1$. So if we let $p^*$ be the optimal value of $(P)$ then $p^* = 1$.
+\paragraph{}
+Therefore $p^* = 1 > 0 = d^*$, so strong duality does not hold for this problem. $\blacksquare$
 \end{document}

### 2017-02-28 CO750-ApproximationAlgorithmsTopics/lectures/lecture_13.tex
Scribing lectures

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}
\usepackage{enumitem}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-02-17}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 13} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\section{Graphic TSP}
\paragraph{} We consider the setting of Graphic TSP. We are given an unweighted graph $G=(V,E)$, and we are to find an Eulerian Spanning (multi)-subgraph of $G$ with the minimum number of edges.
\section{Removable Pairs}
\paragraph{}
In this section we study the method of Removable Pairs from (Momke and Svensonn 2011).
\paragraph{}
Given a $2$-vertex connected graph $G=(V,E)$, a $\textit{removable pairing}$ is a pair $(R, \cP)$ with the following properties:
\begin{enumerate}[label=(\alph*)]
\item $R \subseteq E$, $\cP \subseteq E\times E$.
\item For all $P \in \cP$ there exists $v \in V$ and $3$ edges $e_1$, $e_2$, $e_3$ incident to $v$ such that $P = \{e_1, e_2\}$
\item The elements of $\cP$ are pairwise disjoint
\item For all $F\subseteq R$ with $|F\cap P| \leq $ for all $P \in \cP$, the graph $(V, E\backslash F)$ is connected.
\end{enumerate}
\paragraph{Questions}
How do we find a ``good" $(R,\cP)$. and how do we use it?
\section{Finding a ``good" $(R,\cP)$}
\begin{lemma}
Compute a DFS tree in $G$, rooted at $r$ and call in $(V,S)$. For all $e =vw \in E\backslash S$ let $v$ be on the $rw$-path in $S$, and $v'$ denotes the successor of $V$ on this path.
\begin{itemize}
\item Add $e$ to $R$
\item If $|\delta(v)| \geq 3$ and $e' = vv' \not\in R$ add $e' $ to $R$ and $\{e,e'\}$ to $\cP$.
\end{itemize}
The resulting $(R,\cP)$ is a removable pairing.
\end{lemma}
\begin{proof}
\paragraph{} It is trivial from our construction to see that $(a),(b),(c)$ hold. It remains to verify $(d)$.
\paragraph{}
Let $F \subseteq R$ such that $|F\cap P |\leq 1$ for all $P \in \cP$. For each $v$ let $W_v$ b the nodes in the subtree of $S$ rooted at $v$. We proceed by induction on $|W_v|$ to argue that $W_v$ is connected in $(V, E\backslash F)$. This is an easy exercise.
\end{proof}
\begin{theorem}\label{th:1}
Let $G$ be a $2$-connected graph, $c: E \rightarrow \R_+$ and $(R,\cP)$ be a removable pairing in $G$. Then one can find a tour in $G$ of cost at most $$\frac{4}{3}c(E) - \frac{2}{3}c(R)$$ in polynomial time.
\end{theorem}
Before proving this theorem we will show its power in giving an approximation for graphic TSP
\begin{theorem}
There exists a $\frac{4}{3}$-approximation algorithm for graphic TSP in subcubic graphs (graphs where all nodes have degree at most $3$)
\end{theorem}
\begin{proof}
We construct $(R,\cP)$ as in the previous Lemma. Observe that
$$|R| \geq 2(|E|-|S|) - 1$$
since all non-tree edges (except possibly one at the root) can be be paired with tree edges. From Theorem \ref{th:1} we get
$$\frac{4}{3}|E| - \frac{2}{3}|R| \leq \frac{4}{3} |E| + \frac{4}{3}|E| + \frac{4}{3}|S| + \frac{2}{3} = \frac{4}{3}n - \frac{2}{3}.$$
\end{proof}
\paragraph{}
We now present the proof of Theorem \ref{th:1}
\paragraph{}
\begin{proof}
Let $T$ be the set of vertices of odd degree in $G$. Set $c'(e) = c(e)$ for all $E\backslash R$, $c'(e) = -c(e)$ for all $e \in R$. Note that for any $T$-join $J$ in $G$ that intersects each pair at most once, we can construct a tour on $G$ of cost at most $c(E) + c(J)$.
\paragraph{}
We can do that by doubling the edges of $J\backslash R$ and deleting the edges of $J\cap R$. Notice that the resulting graph is Eulerian, and it is connected since you remove only edges in $R$ that intersect each pair at most once.
\paragraph{}
We that there exists a $T$-join $J$ of $c'$ cost at most 
$$\frac{1}{3} c'(E) - \frac{1}{3}c(E) - \frac{2}{3}c(R)$$
intersecting each pair at most once. If we had such a $T$-join the proof would be complete.
\paragraph{}
We construct an auxiliary subgraph $G'$ as follows. For all $P=\{vw,v'w\} \in \cP$ we introduce one vertex $v_P$ connecting $v_P$ to $v$ with edge cost $0$, and keep edges $v_Pw, v_Pw'$ with their cost $c'$. Note that $G'$ will stay $2$-connected (WMA that the root has degree $3$), Let $T'$ be the set of odd vertices in $G'$. Then each $v_P \in T'$.
\paragraph{}
We show that there exists a $T'$-join in $G'$ of cost at most $\frac{1}{3}c'(E)$ (this naturally corresponds to a $T'$-join in $G$ with the same property) that intersects each pair at most once.
\paragraph{}
Consider $x'_e = \frac{1}{3}$ for all $e \in E(G')$. We will show that this is feasible for the $T'$-join polytope. Recall the constraints for the $T'$-join polytope:
$$|A| - x(A) + x(\delta(U)\backslash A) \geq 1 \quad \forall U \subseteq V, A \subseteq \delta(U): |U \cap T'| + |A|\text{ is odd}$$
Consider some $U\subseteq V$ and $A\subseteq \delta(U)$. Proceed by cases on $|A|$:
\begin{enumerate}
\item If $|A| \geq 2$ then $|A| - x'(A) = 2- \frac{2}{3} \geq 1$.
\item If $|A| = 1$ then $x'(\delta(U)\backslash A) \geq \frac{1}{3}$ and $|A|-x'(A) = \frac{2}{3}$.
\item If $A = \emptyset$ then there are at least $3$ edges crossing the cut, otherwise there exists a graph(by contracting all vertices on other side of cut) with an odd number of odd vertices. Thus $x'(\delta(U)) = 3\frac{1}{3} = 1$.
\end{enumerate}
So $x'$ is a fractional $T'$-join of cost $\frac{1}{3}c'(E)$ with the property $x'(\delta(v_P)) = 1$. So by polyhedral theory there exists an integral $T'$-join of cost at most $\frac{1}{3}c'(E)$ such that it intersect each pair at most once. This completes the proof.
\end{proof}
\end{document}

### 2017-02-28 CO750-ApproximationAlgorithmsTopics/lectures/lecture_14.tex
Scribing lectures

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-02-28}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 14} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\section{Unsplittable Flow}
\paragraph{}
In the Unsplittable Flow problem (UF) we are given a directed graph $G = (V,E)$, edge capacities $c : E \rightarrow \R_+$, a source $s \in V$, and $k$ terminals $t_i \in V$, each with demand $d_i \in \R_+$. We are to find a path $P_i$ for all $i \in [k]$ along which we can route $d_i$ from $s$ to $t_i$, while respecting capacities $c$.
\paragraph{Observation}
A necessary condition for the existence of a solution is that there exists a fractional (splittable) flow. A fractional flow exists if and only if the cut condition holds:
$$\sum_{e\in \delta(S)} c(e) \geq \sum_{i:t_i \in S} d_i \quad \forall S \subseteq V\backslash \{s\}.$$
\paragraph{Proposition}
Deciding whether an unsplittable flow exists is $NP$-complete.
\paragraph{Proof Idea}
Reduction from PARTITION: Given a (multi)-set of integers $a_1,\dots, a_n \in \N$ find a partition into $S_1$ and $S_2$ such that $\sum_{i \in S_1} a_i = \sum_{i\in S_2} a_i$. To perform the reduction we construct a graph with source $s$, a vertex $t$, and terminals $t_1, \dots, t_n$. The terminals correspond to each $a_i$ and have demand $d_i = a_i$. There is an arc from $t$ to $t_i$ for each $i$ with capacity $a_i$. There are two arcs from $s$ to $t$ each with capacity $\sum_{i} a_i /2$. Then it is not hard to see that the decision of which of the two arcs to take in path $P_i$ (top or bottom) indicates which partition to put $a_i$ into should a solution exist.
\begin{theorem}
(Dinitz, Garg, Goemans, '99) Given a fractional flow $f$ satisfying demands $d_i$ without violating $c$, we can construct an unsplittable flow satisfying demands such that the total flow on each edge is at most $c(e) + d_{max}$ where $d_{max} = \max_i \{d_i\}$.
\end{theorem}
\paragraph{}
Note that we may assume without loss of generality that $G$ is acyclic (we remove arcs with $f=0$) and $c(e) = f(e)$ for all $e$ (this is worst case, we can replace flow with one of this form as desired). The algorithm we present will:
\begin{itemize}
\item modify the flow
\item move terminals (until they reach $s$)
\item eliminate edges $e$ with $f(e) = 0$.
\end{itemize}
The algorithm will maintain the important invariant that the flow at each iteration satisfies the current demands.
\paragraph{Definitions}
A terminal $t_i$ at a vertex $v$ is called $\textit{regular}$ if $d_i > f(e)$ for all $e$ entering $v$, and is called $\textit{irregular}$ otherwise. An edge $(u,v)$ is called $\textit{singular}$ if $v$ and all vertices reachable from $v$ have out-degree at most $1$.
\paragraph{Algorithm}
Preliminary Phase:
\begin{enumerate}
\item Let $t_i$ be an irregular terminal at $v$, and $(u,v)$ be s.t. $f(u,v) \geq d_i$. Move $t_i$ to $u$. Set $f(u,v) = f(u,v) - d_i$.
\item If $f(u,v) = 0$, eliminate $(u,v)$. If $t_i = s$, eliminate $t_i$.
\item End phase when no more irregular terminals.
\end{enumerate}
Iteration(s):
\begin{enumerate}
\item Find an ``alternating cycle"
\item ``Augment" the flow along with cycle.
\item ``Move Terminals".
\item STOP when all terminals have reached $s$.
\end{enumerate}
Each step will be explained in turn.
\paragraph{1. Alternating Cycle}
Take an arbitrary vertex $v$. Follow the outgoing edges forming a forward path until we reach a node with no outgoing edges. Since $G$ is acyclic, this node is a terminal $t_i$. Use singular edges to construct a backward path from $t_i$ until we find a vertex $u$ with an outgoing edge. Start a forward path from $u$. Repeat this process until we find a vertex we have previously visited.
\paragraph{}
We can view this cycle $C$ as a union of alternating forward and backward paths. Hence the name.
\paragraph{2. Augment}
We augment the flow on $C$ by increasing flow on backward paths and decreasing the flow on forward paths by $\epsilon$.
$$\epsilon := \min\{\epsilon_1, \epsilon_2\}$$
$$\epsilon_1 := \min\{f(e): e \text{ is on a forward path in } C\}$$
$$\epsilon_2 := \min\{d_i - f(e): e=(v',v'') \text{ for } e \text{ on a backward path to } t_i \text{ with } d_i > f(e) \}.$$
\paragraph{3. Move Terminals}
Move $t_i$ along $e=(v,t_i)$ (as in Preliminary phase) whenever either
\begin{itemize}
\item $e$ is singular and $f(e) = d_i$, or,
\item $e$ is not singular and $f(e) \geq d_i$.
\end{itemize}
\subsection{Analysis}
\paragraph{Observation}
Whenever an edge becomes singular it stays singular because we do not add edges.
\paragraph{Lemma 1}
If, at the beginning of an iteration, $v$ contains an irregular terminal $t_j$, then:
\begin{enumerate}
\item The out-degree of $v$ is $0$.
\item $v$ contains no other irregular terminals.
\item $v$ contains another (regular) terminal.
\end{enumerate}
\paragraph{Proof}
Since $t_j$ is irregular, there exists $e=(u,v)$ such that $d_j < f(e)$ (if $d_j = f(e)$ we would have moved $t_j$ in previous iteration). We claim that $t_j$ became irregular exactly when moved to $v$. To see this observe that we increase flow only on backwards edges, but we would not exceed $d_j$ by our choice of $\epsilon_2$, so if $t_j$ was not irregular when first moved to $v$ it will not become irregular in later iterations.
\paragraph{}
Consider when $t_j$ was moved from (say) $w$ to $v$. Then $(u,v)$ is singular, which implies $(v,w)$ disappears proving $1.$. Then note that by $1.$ $v$ has to contain another terminal. But this terminal cannot be irregular, since when we move an irregular terminal, the out-degree becomes $0$ and we cannot move another such terminal to $v$. This proves $2.$ and $3.$ $\blacksquare$
\paragraph{Corollary}
At the beginning of any iteration, the in-degree of any vertex containing one or more terminals is at least $2$.
\paragraph{Lemma 2}
Unless all terminals reach sources, the algorithm will find an alternating cycle.
\paragraph{Proof}
When constructing a forward path, we stop at a vertex of out-degree $0$, hence it has a terminal. So by corollary we can start constructing a backward path. When constructing a backward path we will either find a previously visited node (node that flows must join source), or one with out-degree $2$. $\blacksquare$
\paragraph{Proof of Theorem}
Before an edge becomes singular it never increases flow value. After, we move at most $1$ terminal along the edge. Thus the increase in demand is at most that of the demand of one terminal which is at most $d_{max}$.$\blacksquare$
\end{document}

### 2017-03-04 CO750-ApproximationAlgorithmsTopics/lectures/lecture_15.tex
lecture 15 scribe

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-03-02}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 15} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\paragraph{Recall}
Given directed graph $G=(V,E)$, source $s \in V$,and terminals $t_1,\dots,t_k$ with demands $d_1, \dots, d_k$, if $f$ is a fractional flow realizing the demands then there exists an unsplittable flow $\bar{f}$ realizing demands such that $\bar{f}(e) \leq f(e) + d_{max}$ for all $e \in E$.
\paragraph{Note}
The bound can be tight. Consider a graphs where source $s$ is adjacent to $q$ terminals, $t_1,\dots,t_q$ with arc capacities $1$ and demands $d_1=\dots=d_q=1-\frac{1}{q}$, and those $q$ terminals are all adjacent to terminal $t_{q+1}$ with capacity $\frac{1}{q}$ and demand $d_{q+1} = 1$. It is a good exercise to verify this is a tight example.
\paragraph{}
Now suppose we consider the weighted cases. If each edge has weight $w(e)\geq 0$ then we say our flow $f$ has weight $w(f) = \sum_e w(e) f(e)$.
\paragraph{Conjecture}(Goemans '2000) There exists an unspittable flow $\bar{f}$ realizing the demands such that $\bar{f}(e) \leq f(e) + d_{max}$ and $w(\bar{f}) \leq w(f)$. (It is a good exercise to verify that our algorithm in the unweighted case does not generalize to the weighted case: that is, we can make the difference in weight arbitrarily large).
\paragraph{}(Skutella '02) There exists an unsplittable flow $\bar{f}$ with $w(\bar{f})\leq w(f)$ and $\bar{f}(e) \leq 2f(e) + d_{max}$. We will prove this later but first we need to review some basic theory about flows.
\paragraph{Theorem} Let $G$ be a digraph, $s \in V$, $t_1,\dots, t_k \in V$, $d_1,\dots, d_k$ as before. If $d_1,\dots,d_k \in \alpha \N$ for some $\alpha \in \R$ and each capacity $c(e) \in \alpha \N$ such that $\sum_{e\in\delta(s)} c(e) \geq \sum_{i:t_I \in S} d_i$ for all $S \subseteq V\backslash\{s\}$, then there exists a fractional flow such that $f(e) \in \alpha \N$ for all $e\in E$. This follows quickly from the theorem from $650$ when $alpha =1 $.
\paragraph{}
We say $d_i \mid d_j$ if there exists $p \in \Z$ such that $d_j = p d_i$.
\paragraph{}
Order our demands as $d_{min} = d+1 \leq d_2 \leq \dots \leq d_k = d_{max}$.
\begin{theorem}\label{th:div}
If $\forall i,j \in [k]$ with $i<j$ we have $d_i \mid d_j$ then there exists an unsplittable flow $\bar{f}$ such that $\bar{f}(e) \leq f(e) + d_{max}$ and $w(\bar{f}) \leq w(f)$.
\end{theorem}
\begin{proof}
Let $\delta_1\leq \dots \leq \delta_\ell$ be the values of the different demands. Consider the algorithm:

\begin{enumerate}
\item Set $f_0 = f$, $j=1$
\item While $j\leq \ell$ do:
	\begin{enumerate}
	\item Set $u^j(e) = f_{j-1}(e)$ rounded up to the nearest multiple of $\delta_j$
	\item Compute $\delta_j$-integral flow  $f_j$ that satisfies all demands and such that $w(f_j) \leq w(f_{j-1})$.
	\item For all terminals $t_i$ with $d_i = \delta_j$, select an arbitrary path $P_i$ carrying flow $d_i$ from $s$ to $t_i$.
	\item Decrease $f_j$ along $P_i$ by $d_j$ for all $i$ and remove edges with flow $0$. Remove such terminals from our instance.
	\end{enumerate}
\end{enumerate}
Let $\bar{f}$ be the unsplittable flow realized by the paths, $P_1, \dots, P_k$, in our algorithm. Trivially by inductino $w(\bar{f}) \leq w(f)$. Now it remains to bound $\bar{f}(e)$.
\paragraph{}
At the end of each iteration $j$ of our algorithm we have:
\begin{align*}
\bar{f}(e) &\leq f_j(e) + \sum_{i:d_i\leq \delta_j, e\in P_i} d_i \\
&= f_j(e) + \sum_{i:d_i = \delta_j, e\in P_i} d_i + \sum_{i:< \delta_j, e\in P_i} d_i \\
&\leq u^j(e) + \sum_{i:d_i < \delta_j, e\in P_i} d_i. 
\end{align*}
\paragraph{}Note that $f_{j-1}$ is $\delta_{j-1}$-integral, and $\delta_{j-1}\mid \delta_j$ for $j > 1$. Therefore $u^j(e) \leq f_{j-1}(e) + \delta_j - \delta_{j-1}$.
Setting $\delta_0 = 0$ we obtain:
$$\bar{f}(e) \leq f_{j-1}(e) + \delta_j - \delta_{j-1} + \sum_{i:d_i < \delta_j, e\in P_i} d_i.$$
Applying the above iteratively we obtain
$$f_j(e) + \sum_{i:d_i\leq \delta_j, e\in P_i} d_i \leq f_0(e) + \delta_j - \delta_0.$$
In particular in the last iteration this implies:

$$\bar{f}(e) \leq f_0(e) + \delta_\ell - \delta_0 = f(e) + d_{max} - 0.$$

\end{proof}
\paragraph{Corollary}
For all $e\in E$ let $i_e$ be the index of a terminal using $e$, that is $e \in P_i$, and has maximum demand among all such terminals. Then $\sum_{i\neq i_e : e \in P_i} d_i \leq f(e)$. This follows easily by rearranging the inequality obtained in the previous theorem.
\paragraph{General case}
We now use the above theorem to prove the general case. Consider the following algorithm:
\begin{enumerate}
\item Round as $\bar{d}_i = d_{min}\cdot 2^{\floor{\log \frac{d_i}{d_{min}}}}$
\item Modify $f$: Consider all terminals $t_i$ with $d_i \neq \bar{d}_i$ and:
\begin{enumerate}
\item Iteratively reduce flow on the most expensive $st_i$-path (deleting edges with $0$ flow). We obtain $\tilde{f}$ satisfying $\bar{d}$.
\end{enumerate}
\item Apply previous algorithm to $\tilde{f}$ with demands $\bar{d}$, returning unsplittable flow $\bar{f}$ satisfying $w(\bar{f}) \leq w(\tilde{f})$ and $\bar{f}(e) \leq \tilde{f}(e) + d_{max}$.
\item Add $d_i - \bar{d}_i$ amount of flow on each path $P_i$ specified by $\bar{f}$. We obtain an unsplittable flow $\bar{\bar{f}}$ satisfying $d_1,\dots,d_k$.
\end{enumerate}
We have $$w(\bar{\bar{f}}) = \sum_{i=1}^k w(P_i)(d_i -\bar{d}_i) + w(\tilde{f}) \leq w(f) - w(\tilde{f}) + w(\tilde{f}) = w(f).$$
Furthermore, $d_i \leq 2 \bar{d}_i$ so with this, and the corollary, we get 
$$\sum_{i: e\in P_i} d_i \leq d_{i_e} + 2\sum_{i\neq i_e : e\in P_i} \bar{d}_i \leq d_{i_e} + 2f(e)$$
as desired.$\blacksquare$
\end{document}

### 2017-03-05 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
typing 1

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-03-14}
\rhead{W. Justin Toth CO750-Approximation Algorithms Assignment 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}

%Question 1
\section{}
\paragraph{}
Let $G = (V,E)$ be an undirected graph with edge cost $c: E\rightarrow \R_+$. Consider the problem $(IP)$:
\begin{align*}
\min\ \sum_{e\in E} c(e) x_e&\\
\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
x&\geq 0\\
x&\in \Z,
\end{align*}
where $f : S \rightarrow \{0,1\}$ satisfies the maximality property:
$$\text{ for all disjoint } A,B\subseteq V, \text{ we have } f(A \cup B ) \leq \max\{f(A), f(B)\},$$
and $f$ satisfies that there does not exist $S \subseteq V$ and $A, B \subseteq S$ such that
\begin{equation}\label{eq:1}
A\cap B = \emptyset,\ f(S)=f(A)=f(B)=0 \text{ and } f(S\backslash A) = f(S\backslash B) = 1.
\end{equation}
We may assume that $f(V') = 0$, for the every connected component vertex set $V'$, otherwise $(IP)$ is infeasible. Let $(LP)$ denote the linear programming relaxation of $(IP)$ and let $(D)$ denote the dual of $(LP)$:
\begin{align*}
\max\ \sum_{S\subseteq V} f(S) y_S&\\
\text{s.t.}\ \sum_{S : e \in \delta(S)} y_S &\leq c(e) &\forall e \in E \\
y&\geq 0. 
\end{align*}
\paragraph{Algorithm}
Consider the primal-dual algorithm, $\cA$, for this problem:
\begin{enumerate}
\item Start with $x=0$, $y=0$, and $F \rightarrow \emptyset$.
\item While $F$ is not feasible do:
	\begin{enumerate}
	\item Increase $y_S$ uniformly for ``minimal violated cuts" with respect to $F$ until some dual constraint corresponding to an edge $e$ becomes tight.
	\item Set $x_e = 1$, $F = F\cup \{e\}$.
	\end{enumerate}
\item Reverse Deletion: Consider all edges $e$ in $F$ in reverse of the order they were added to $F$ and if $F\backslash \{e\}$ is feasible then set $F = F\backslash\{e\}$ and $x_e = 0$.
\end{enumerate}
\paragraph{}
We intend to show that $\cA$ is a polynomial time $2$-approximation algorithm for $(IP)$. We will need the following lemma from class:
\begin{lemma}\label{lemma:maximality}
Let $f$ be a $01$-function satisfying maximality. Let $F\subseteq E$. Then
\begin{enumerate}
\item $F$ is feasible for $(IP)$ with $f$ if and only if every connected component $C$ of $(V,F)$ satisfies $f(C) = 0$.
\item The minimal violated cuts are connected components $C$ that have $f(C) = 1$.
\end{enumerate}
\end{lemma}
This lemma allows us to show that $\cA$ terminates in polynomial time.
\begin{lemma}\label{lemma:poly1}
Suppose we have a polynomial time oracle for $f$. Then $\cA$ returns a feasible solution for $(IP)$ in polynomial time
\end{lemma}
\begin{proof}
By definition of $\cA$ if $\cA$ terminates then the result is feasible. To see that $\cA$ terminates observe that in every iteration of step $2$ some edge $e \in |E|$ is added to $F$. Since $E$ is feasible, the while loop terminates after $O(|E|)$ iterations. Hence $\cA$ terminates, and returns a feasible solution.
\paragraph{} 
It remains to show that $\cA$ runs in polynomial time. Since we have a polynomial time oracle for $f$ it is easy to see that step $3$ runs in polynomial time. Further we need only maintain $y_S$ for non-zero $y_S$, of which we will show there are a polynomial number. We need to verify step $2(a)$ can be done in polynomial time. By lemma \ref{lemma:maximality} we need only maintain connected components of $F$ since the connected components $C$ with $f(C) = 1$ are precisely the minimal violated cuts. There are a polynomial number of connected components at each iteration (hence a polynomial number of non-zero $y_S$), and since we have a polynomial time oracle for $f$ this implies step $2(a)$ can be done in polynomial time.
\end{proof}
\paragraph{}
Now we need to check the approximation factor. We present the following reduction from class as a lemma:
\begin{lemma}\label{lemma:degree}
Let $w_i$ denote the number of minimally violated sets at iteration $i$.  Let $\cS_i$ denote the set of minimal violated cuts at iteration $i$. If $\sum_{S\in \cS_i} |\delta_F(S)| \leq 2w_i$ for all $i$ then $\cA$ is a $2$-approximation algorithm for $(IP)$ (for any $01$-function $f$).
\end{lemma}
\begin{proof}
It is immediate from our construction of $x$ and $y$ that if $e \in F$ then
$$\sum_{S : e\in\delta(S)} y_S = c(e).$$
Thus we have the objection function value on $F$ is
$$\sum_{e\in F} c(e) = \sum_{e\in F} \sum_{S : e\in\delta(S)} y_S = \sum_{S\subseteq V} |\delta_F(S)|y_S.$$
Let $\Delta_i$ denote the amount by which all minimal violated cuts get increased in iteration $i$. Then
$$\sum_{S\subseteq V} |\delta_F(S)|y_S = \sum_{i} \sum_{S\in \cS_i} |\delta_F(S)|\Delta_i \leq \sum_{i} 2w_i \Delta_i = 2\sum_{S\subseteq V} f(S) y_S.$$
The inequality follows from our hypothesis, and the last equality follows since $f$ is a $01$-function for which non-zero $y_S$ (minimally violated sets) have $f(S) = 1$. 
\end{proof}

\end{document}

### 2017-03-05 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
writing up 1

@@
 \end{proof}
+\paragraph{}
+Now we intend to show that the hypothesis of the previous lemma holds.
+Consider some iteration $i$ of $\cA$, and let $A$ be the of edges chosen so far at the end of $i$.
+Let $B = F \cap A$.
+Construct a graph $H$ by taking the graph $(V,F)$ and contracting the connected components of $(V,B)$ to vertices.
+Then $E(H) = F\backslash B$.
+Let $S_v$ denote the set of vertices in $G$ corresponding to vertex $v \in V(H)$.
+It is not hard to see that $H$ is a forest.
+Indeed if $H$ has some cycle $C$ then we can remove any edge in $E(C)$ from $F$ without increasing the number of connected components.
 
+\begin{lemma}\label{lemma:leaves}
+$\sum_{v\in W} d_v \leq 2|W|$.
+\end{lemma}
+\begin{proof}
+First we claim that for every connected component $C$ of $H$, at most one leaf of $C$ satisfies $f(S_v) = 0$.
+Suppose for a contradiction there exists some connected component $C$ of $H$ that has two leaves $v$ and $w$ such that $f(S_v) = f(S_w) = 0$.
+Let $S_C = \cup_{u\in C} S_u$.
+Let $e_v, e_w \in F$ be the edges incident to $v$ and $w$ respectively in $H$.
+By minimality via the reverse deletion step, $F\backslash \{e_v\}$ and $F\backslash \{e_w\}$ are infeasible.
+But then $f(S_C\backslash S_v) =1 $ and $f(S_C\backslash S_w) = 1$.
+Since $C$ is a connected component of $H$ we have $f(S_C) = 0$.
+Hence
+$$f(S_v) = f(S_w) = f(S_C) = 0 \quad\text{and}\quad f(S_C\backslash S_v) = f(S_C \backslash S_w) = 1$$
+but $S_v, S_w \subseteq S_C$ and $S_v \cap S_w = \emptyset$. This contradictions property \ref{eq:1}.
+\end{proof}
 \end{document}

### 2017-03-06 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
Typed 1

@@
 It is not hard to see that $H$ is a forest.
-Indeed if $H$ has some cycle $C$ then we can remove any edge in $E(C)$ from $F$ without increasing the number of connected components.
-
+Indeed if $H$ has a cycle $C$ then, since for all $v \in V(C)$, $S_v$ is a connected component in $G$, we can find a cycle $C'$ in $G$ by expanding each $v \in V(C)$ and joining its vertices which lie on $C$ by a path in $S_v$.
+But $F$ is a forest since it is necessary that any edge $e$ be a bridge to survive the reverse deletion step as $f$ is a $01$-function (observe that for all $S$ $|\delta(S) \backslash \{e\}| \geq 1 \geq f(S)$ if $e$ is not a bridge).
+\paragraph{}
+Now let $W\subseteq V(H)$ be the set of vertices of $H$ such that for all $w \in W$, $f(S_w) = 1$. Then $\{S_w : w\in W\} = \cS_i$. Let $d_v$ denote the degree of vertex $v$ in $H$. So the condition
+$$\sum_{S \in \cS_i} |\delta_F(S)| \leq 2w_i$$
+is equivalent to
+$$\sum_{w \in W} d_w \leq 2|W|.$$
 \begin{lemma}\label{lemma:leaves}
-$\sum_{v\in W} d_v \leq 2|W|$.
+$\sum_{w\in W} d_w \leq 2|W|$.
 \end{lemma}
@@
 By minimality via the reverse deletion step, $F\backslash \{e_v\}$ and $F\backslash \{e_w\}$ are infeasible.
-But then $f(S_C\backslash S_v) =1 $ and $f(S_C\backslash S_w) = 1$.
-Since $C$ is a connected component of $H$ we have $f(S_C) = 0$.
+But then $f(S_C\backslash S_v) =1 $ and $f(S_C\backslash S_w) = 1$ as these are precisely the cuts which would have no incident edges when $e_v$ and $e_w$ are removed respectively (we already know $f(S_v) = f(S_w) = 0$).
+Since $C$ is a connected component of $H$ we have $f(S_C) = 0$ by the feasibility of $F$.
 Hence
 $$f(S_v) = f(S_w) = f(S_C) = 0 \quad\text{and}\quad f(S_C\backslash S_v) = f(S_C \backslash S_w) = 1$$
-but $S_v, S_w \subseteq S_C$ and $S_v \cap S_w = \emptyset$. This contradictions property \ref{eq:1}.
+but $S_v, S_w \subseteq S_C$ and $S_v \cap S_w = \emptyset$. This contradiction property (\ref{eq:1}).
+\paragraph{}
+Now we proceed to demonstrate the inequality. First discard isolated vertices of $H$ since they affect the count in no way. Let $c$ denote the number of connected components of $H$. Now we have,
+\begin{align*}
+\sum_{w\in W} d_w &= \sum_{v \in H} d_v - \sum_{v\not\in W} d_v \\
+&= 2|H| -2c - \sum_{v\not\in W} d_v &\text{since $H$ is a forest} \\
+&\leq 2|H| - 2c - 2(|H| - |W| - c) \\
+&= 2|W|.
+\end{align*}
+The inequality follows from our claim which implies all vertices not in $W$ have degree at least $2$ except for one leaf per connected component of $H$, which has degree $1$.
 \end{proof}
+\paragraph{}
+Hence combining lemma \ref{lemma:degree} and lemma \ref{lemma:leaves} with our discussion about $H$ we have shown that $\cA$ returns a solution $F$ with $$c(F) \leq 2\sum_{S\subseteq V} f(S) y_S \leq 2\cdot OPT,$$
+where $OPT$ is the optimal solution of $(IP)$. Therefore we have shown that
+\begin{theorem}
+$\cA$ is a $2$-approximation algorithm for $(IP)$.
+\end{theorem}
 \end{document}

### 2017-03-06 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
typed 2

@@
 \end{theorem}
+
+%Q2
+\section{}
+\paragraph{}
+Let $G = (V,E)$ be a metric graph with edge costs $c: E\rightarrow \R_+$.
+Consider the problem of finding a minimum cost subset of edges $F\subseteq E$ such that each connected component of $(V,F)$ is an eulerian subgraph with at least $100$ vertices.
+We will denote this problem $(E100)$.
+\paragraph{}
+Define the function $f:2^V \rightarrow \{0,2\}$ by:
+$$f(S) = \begin{cases}
+2, &\text{if } 1\leq |S| < 100 \\
+0, &\text{otherwise.}
+\end{cases}$$
+Then we can define a linear programming relaxation $(LP)$ for $(E100)$:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+x&\geq 0.
+\end{align*}
+To justify this relaxation consider a feasible solution $F$ to $(E100)$.
+Let $x^F$ be the incidence vector of $F$.
+Then $x^F$ is feasible for $(LP)$.
+Let $S \subseteq V$ such that $1 \leq |S| < 100$.
+Since $|S| < 100$, $S$ does not contain any single connected component of $(V,F)$.
+Since $1\leq |S|$, $S$ has non-trivial intersection with some connected component $C$ of $(V,F)$ (i.e.
+$S\cap C \neq \emptyset$ and $S\cap C \supset C$).
+Since $F$ is feasible, $C$ is eulerian.
+Hence the degree of each vertex in $C$ (with respect to $F$) is even.
+Thus $|E(C\cap S, C\backslash S)| \geq 2$.
+Therefore
+$$\sum_{e \in \delta(S)} x_e \geq \sum_{e \in E(C\cap S, C\backslash S)} x_e \geq 2 \geq f(S),$$
+and hence $x$ is feasible for $(LP)$.
+\paragraph{}
+We now define another function $g: 2^V \rightarrow \{0,1\}$, given by
+$$g(S) = f(S)/2 = \begin{cases}
+1, &\text{if } 1\leq |S| < 100 \\
+0, &\text{otherwise.}
+\end{cases}$$  
+The integer program $(IP)$ for $g$ given by:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq g(S) &\forall S \subseteq V \\
+x&\geq 0 \\
+x\in \Z
+\end{align*}
+describes the problem of finding a forest where each connected component is a tree with at least $100$ vertices.
+From our choice of $g$ it is clear that the connected components of any feasible solution must have at least $100$ vertices, as the cut defined by any smaller component does not satisfy the inequality for $g$.
+Optimal solutions to $(IP)$ do not have cycles, as we can remove any single edge present in a cycle and preserve connectivity.
+\paragraph{}
+Now we observe that $g$ satisfies property (\ref{eq:1}) from question $1$.
+Let $S \subseteq V$ and let $A, B \subseteq S$ such that $A \cap B = \emptyset$.
+Suppose that $g(S)=g(A) = g(B) = 0$.
+If $g(S\backslash A) = 1$ then $1 \leq |S\backslash A| < 100$.
+But $B \subseteq S\backslash A$, and hence $|B| \leq |S\backslash A|$.
+Since $g(B) = 0$ this means $|B| = 0$, that is $B = \emptyset$.
+Thus $g(S\backslash B) = g(S) = 0$.
+Similarly we can show if $g(S\backslash B) = 1$ then $g(S\backslash A) = 0$.
+Hence property $(\ref{eq:1})$ is satisfied by $g$.
+\paragraph{}
+Consider the following algorithm, $\cA$, for $(E100)$:
+\begin{enumerate}
+\item Set $S = \emptyset$.
+\item Solve $(IP)$ for $g$ using the algorithm of question $1$, obtaining a forest $F$.
+\item For each connected component $T$ of $(V,F)$:
+	\begin{enumerate}
+	\item Compute a matching $M$ on the odd vertices of $T$.
+	\item Shortcut $T\cup M$ obtaining an Euler Tour $A$.
+	\item Set $S = S \cup A$.
+	\end{enumerate}
+\item Return $S$.
+\end{enumerate}
+\paragraph{}
+We claim that $\cA$ is a $2$-approximation algorithm for $(E100)$.
+In question $1$ we showed that step $2$ can be done in polynomial time.
+In the lecture notes we have shown that the shortcutting process of each iteration of step $3$ can be done in polynomial time.
+Further there are a polynomial number of components the shortcutting is done on, so step $3$ also runs in polynomial time.
+Therefore $\cA$ returns a solution in polynomial time.
+The solution is feasible since each component is a cycle (hence is eulerian) on at least $100$ vertices.
+\paragraph{}
+We now demonstrate the approximation factor holds.
+We know from lecture that the shortcutting process of step $3$ produces a set of cycles of cost at most twice that of the original trees in a metric graph.
+That is,
+$$c(A)\leq 2c(F).$$
+Let $y$ be the optimal dual solution produces by the primal-dual algorithm of step $1$. We have from question $1$ that
+$$c(F) \leq 2\sum_{S\subseteq V} g(S) y_S = \sum_{S\subseteq V} f(S) y_S \leq OPT.$$
+where $OPT$ is the optimal solution to $(LP)$.
+This follows since the dual to $(LP)$ (with respect to $(E100)$) has the same feasible region as the dual to the linear programming relaxation of $(IP)$.
+Combining inequalities we see that
+$$c(A) \leq 2c(F) \leq 2 \cdot OPT.$$
+Hence $\cA$ is a $2$-approximation algorithm for $(E100)$.$\blacksquare$
 \end{document}

### 2017-03-12 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
starting to write p3

@@
 x&\geq 0\\
-x&\in \Z,
+x&\in \Z^{|E|},
 \end{align*}
@@
 x&\geq 0 \\
-x\in \Z
+x\in \Z^{|E|}
 \end{align*}
@@
 Hence $\cA$ is a $2$-approximation algorithm for $(E100)$.$\blacksquare$
+
+%Q3
+\section{}
+\paragraph{}
+Consider again the problem $(IP)$, associated with a graph $G=(V,E)$ with edges costs $c: E \rightarrow \R_+$ given by:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+x&\geq 0 \\
+x&\in \Z^{|E|}.
+\end{align*}
+Suppose that $f:2^V \rightarrow \N$ is symmetric, meaning for all $S\subseteq V$:
+$$f(S) = f(V\backslash S)$$
+and that $f$ is subadditive, meaning for all $A,B \subseteq V$:
+$$f(A\cup B) \leq f(A) + f(B).$$
+We also suppose that the graph $G$ is a cycle on $n$ vertices.
+%3a
+\subsection*{a}
+\paragraph{}
+We begin by working towards a compact formulation for $(IP)$ on $G$.
+Since $G$ is a cycle we may label the vertices of $G$ as $V(G) = \{v_1,\dots, v_n\}$ such that $v_iv_j \in E(G)$ if and only if $$j \equiv i\pm1\mod n$$.
+For $i, \ell \in [n]$ ($[n] := \{1,\dots, n\}$) we define the path from $i$ of length $\ell$ as
+$$P(i, \ell) := \{v_i, v_{i+1}, \dots, v_{i + \ell} \}$$
+where addition is taken modulo $n$. We use the convention that when $\ell = n$ we have the entire cycle, and when $\ell = 0$ we have just $v_i$. We define the family of paths on $G$ to as
+$$\cP := \{P(i,\ell) : i, \ell \in [n]\}.$$
+Observe that the size of $\cP$ is polynomial in $n$, in particular $|\cP| = O(n^2)$.
+\begin{lemma}\label{lemma:compact}
+The problem $(IP)$ on $G$ is equivalent to the compact integer program $(\overline{IP})$ given by:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
+x&\geq 0 \\
+x&\in \Z^{|E|}.
+\end{align*}
+\end{lemma}
+\begin{proof}
+
+\end{proof}
+%3b
+\subsection*{b}
 \end{document}

### 2017-03-12 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
3a

--- 2017-03-06 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
@@
 Hence $\cA$ is a $2$-approximation algorithm for $(E100)$.$\blacksquare$
+
+%Q3
+\section{}
+\paragraph{}
+Consider again the problem $(IP)$, associated with a graph $G=(V,E)$ with edges costs $c: E \rightarrow \R_+$ given by:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
+x&\geq 0 \\
+x&\in \Z^{|E|}.
+\end{align*}
+Suppose that $f:2^V \rightarrow \N$ is symmetric, meaning for all $S\subseteq V$:
+$$f(S) = f(V\backslash S)$$
+and that $f$ is subadditive, meaning for all $A,B \subseteq V$:
+$$f(A\cup B) \leq f(A) + f(B).$$
+We also suppose that the graph $G$ is a cycle on $n$ vertices.
+%3a
+\subsection*{a}
+\paragraph{}
+We begin by working towards a compact formulation for $(IP)$ on $G$.
+Since $G$ is a cycle we may label the vertices of $G$ as $V(G) = \{v_1,\dots, v_n\}$ such that $v_iv_j \in E(G)$ if and only if $$j \equiv i\pm1\mod n$$.
+For $i, \ell \in [n]$ ($[n] := \{1,\dots, n\}$) we define the path from $i$ of length $\ell$ as
+$$P(i, \ell) := \{v_i, v_{i+1}, \dots, v_{i + \ell} \}$$
+where addition is taken modulo $n$. We use the convention that when $\ell = n$ we have the entire cycle, and when $\ell = 0$ we have just $v_i$. We define the family of all paths on $G$ as
+$$\cP := \{P(i,\ell) : i, \ell \in [n]\}.$$
+Observe that the size of $\cP$ is polynomial in $n$, in particular $|\cP| = O(n^2)$ (since $n$ choices for $i$ and $n$ choices for $\ell$ covers all possible $P(i,\ell)$ with some overlap).
+\begin{lemma}\label{lemma:compact}
+The problem $(IP)$ on $G$ is equivalent to the compact integer program $(\overline{IP})$ given by:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
+x&\geq 0 \\
+x&\in \Z^{|E|}.
+\end{align*}
+\end{lemma}
+\begin{proof}
+Since the objective functions are the same, we need only demonstrate that the feasible regions of $(IP)$ and $(\overline{IP})$ are equal.
+Since each $P \in \cP$ is a subset of $V$, it is clear that the feasible region of $(IP)$ is a subset of the feasible region of $(\overline{IP})$.
+So we just have to show the reverse containment.
+\paragraph{}
+Let $x$ be a feasible solution for $(\overline{IP})$.
+Let $S \subseteq V$.
+If we consider the connected components of $G[S]$ we observe that they are all paths.
+Hence there exist disjoint $P_1, \dots, P_k \in \cP$ such that
+$$S = P_1 \dot\cup \dots \dot\cup P_k$$
+where each $P_i$ corresponds to a connected component of $G[S]$. Now we see that
+\begin{align*}
+f(S) &= f(P_1 \dot\cup \dots \dot \cup P_k) \\
+&\leq f(P_1) + \dots + f(P_k) &\text{by subadditivity} \\
+&\leq x(\delta(P_1)) + \dots x(\delta(P_k)) &\text{by feasibility of $x$}.
+\end{align*}
+We claim that the edges incident to each path are disjoint.
+To see this, suppose that we have $P_i$ and $P_j$ with $\delta(P_i) \cap \delta(P_j) \neq \emptyset$.
+Let $v_iv_j \in \delta(P_i) \cap \delta(P_j)$.
+Since $P_i$ and $P_j$ are disjoint we may assume $v_i \in P_i \backslash P_j$ and $v_j \in P_j \backslash P_i$.
+But then $P_i$ and $P_j$ are connected by $v_iv_j$ violating that they are distinct connected components.
+So the claim holds.
+Thus we observe
+\begin{align*}
+f(S) &\leq x(\delta(P_1)) + \dots x(\delta(P_k)) \\
+&= x(\bigcup_{i=1}^k \delta(P_i)) &\text{by disjointness}\\
+&= x(\delta(S)) &\text{by disjointness}.
+\end{align*}
+Therefore $x$ satisfies the cut constraint for all $S$, and hence $x$ is feasible for $(IP)$.
+\end{proof}
+\paragraph{}
+Hence by lemma \ref{lemma:compact} it is equivalent for us to solve $(\overline{IP})$. Now we define a linear programming relaxation for $(\overline{IP})$, which we will denote $(\overline{LP})$:
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
+x&\geq 0.
+\end{align*}
+Since $(\overline{LP})$ has a polynomial number of constraints and variables it can be solved in polynomial time.
+We now define our rounding-based approximation algorithm, $\cA$, for $(\overline{IP})$:
+\begin{enumerate}
+\item Solve $(\overline{LP})$ on $G$  returning solution $\bar{x}$.
+\item Let $\bar{x}_I \in \Z^{|E|}$ denote the integral part of $\bar{x}$. Let $\bar{x}_F \in [0,1)^{|E|}$ denote the fractional part of $\bar{x}$. Thus $\bar{x} = \bar{x}_I + \bar{x}_F.$
+\item Construct solution $x \in \Z^{|E|}$ by rounding. For each $e \in E$ set:
+$$x_e = \begin{cases}
+\bar{x}_I + 1, &\text{if } x_F \geq 0.5 \\
+\bar{x}_I, &\text{otherwise.}
+\end{cases}$$
+\item Return solution $x$.
+\end{enumerate}
+\begin{lemma}\label{lemma:feasible-rounding}
+If $\bar{x}$ is a feasible solution to $(\overline{LP})$ then the $x$ obtained from rounding $\bar{x}$ in step $3$ of $\cA$ is a feasible solution to $(\overline{IP})$.
+\end{lemma}
+\begin{proof}
+It is immediate from the description of step $3$ that $x$ is integral and non-negative.
+We need only verify the cut constraints.
+Let $P \in \cP$.
+Let $v_i, v_j \in P$ be the start and end vertices of $P$.
+That is,
+$$\delta(P) = \{v_{i-1}v_i,  v_jv_{j+1}\}.$$
+For convenience let $e_i = v_{i-1}v_i$ and let $e_j = v_jv_{j+1}$. Since $\bar{x}$ is feasible for $(\overline{LP})$:
+$$ f(P)\leq \bar{x}(e_i) + \bar{x}(e_j) = \bar{x}_I(e_i) + \bar{x}_F(e_i)  + \bar{x}_I(e_j) + \bar{x}_F(e_j).$$
+Since $f$ maps to $\N$ we have:
+$$f(P) \leq \floor{\bar{x}_I(e_i) + \bar{x}_F(e_i)  + \bar{x}_I(e_j) + \bar{x}_F(e_j)}.$$
+We consider the following cases: Case 1: $\bar{x}_F(e_i) + \bar{x}_F(e_j) < 1$ and Case 2: $\bar{x}_F(e_i) + \bar{x}_F(e_j) \geq 1$.
+\paragraph{}
+In Case 1 the fractional parts vanish during the floor-ing operation and hence
+$$ f(P) \leq \bar{x}_I(e_i) + \bar{x}_I(e_j)  \leq x(e_i) + x(e_j).$$
+In Case 2 we observe that $x_F(e_i) + x_F(e_j) = 1 + \alpha$ for some $\alpha \in [0,1)$. Thus
+$$f(P) \leq \floor{\bar{x}_I(e_i) + \bar{x}_F(e_i)  + \bar{x}_I(e_j) + \bar{x}_F(e_j)} = \bar{x}_I(e_i) + \bar{x}_I(e_j)  + 1.$$
+Since $\bar{x}_F(e_i) + \bar{x}_F(e_j) \geq 1$ at least one of $\bar{x}_F(e_i) ,\bar{x}_F(e_j) $ is greater than or equal to $0.5$.
+Assume without loss of generality $\bar{x}_F(e_i) \geq 0.5$.
+Then $x_(e_i) = \bar{x}_I(e_i) + 1$.
+So
+$$f(P) \leq \bar{x}_I(e_i) + \bar{x}_I(e_j)  + 1 = x(e_i) +\bar{x}(e_j) \leq x(e_i) + x(e_j).$$
+Therefore in either case $x$ satisfies the cut constraint for $P$. Therefore $x$ is feasible for $(\overline{IP})$.
+\end{proof}
+\begin{theorem}
+The algorithm $\cA$ is a $2$-approximation algorithm for $(\overline{IP})$.
+\end{theorem}
+\begin{proof}
+\paragraph{}
+As we have argued above, $(\overline{LP})$ can be solved in polynomial time, so step $1$ can be done in polynomial time.
+Step $3$ can be done in $O(|E|)$ time.
+Hence $\cA$ terminates in polynomial time.
+From lemma \ref{lemma:feasible-rounding} we see that $x$ is feasible for $(\overline{IP})$.
+It remains to check the approximation factor.
+\paragraph{}
+Let $e \in E$.
+If $\bar{x}(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq 2\bar{x}(e)$.
+It remains to consider the pertinent case when $\bar{x}(e) \geq 0.5$.
+Therein we have
+\begin{align*}
+\frac{x(e)}{\bar{x}(e)} &= \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) + \bar{x}_F(e)} \\
+&\leq  \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) +0.5} \\
+&= \frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}.
+\end{align*}
+Since $\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}$ is a non-increasing function, it is maximized when $\bar{x}_I(e) = 0$. Hence
+$$\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1} \leq \frac{2\cdot 0 + 2}{2\cdot 0 + 1} = 2.$$
+Thus $x(e) \leq 2\bar{x}(e)$ as desired. Therefore
+$$\sum_{e\in E}c(e) x(e) \leq \sum_{e \in E}c(e)2\bar{x}(e) = 2\sum_{e\in E}\bar{x}(e) \leq 2\cdot OPT,$$
+where $OPT$ is the cost of an optimal solution to $(\overline{IP})$ (the last inequality follows immediately from $\bar{x}$ being optimal for $(\overline{LP})$.
+So $\cA$ is a $2$-approximation algorithm.
+\end{proof}
+%3b
+\subsection*{b}
 \end{document}

### 2017-03-12 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
writing up 3b

@@
 \subsection*{b}
+\paragraph{}
+We make an observation about optimal solutions to $(\overline{IP})$ which contain a $0$-valued edge. This will lead us to an approximation algorithm with a better bound than the one found in $3(a)$.
+\begin{lemma}\label{lemma:0edge}
+Let $x$ be an optimal solution $(\bar{IP})$. If there exists $v_iv_{i+1} \in E(G)$ such that $x(v_iv_{i+1}) = 0$ then for all $v_jv_{j+1} \in E(G)$
+$$x(v_jv_{j+1}) = f(P(i+1, \ell)),$$
+where $\ell = j-(i+1)$. Recall that $P(i+1, \ell)$ is the $\ell$ length path starting at $i+1$. In this case $P(i+1,\ell)$ is the path joining $v_{i+1}$ to $v_j$ avoiding $v_i$ and $v_{j+1}$. 
+\end{lemma}
+\begin{proof}
+For each $v_jv_{j+1} \neq v_iv_{i+1} \in E(G)$ let $P(v_jv_{j+1}) = P(i+1, \ell)$ where $\ell = j-(i+1)$. Let $e_j \in E(G)$.  Since $x$ is feasible and $x(v_iv_{i+1}) = 0$,
+$$f(P(e_j)) \leq x(\delta(P(e_j))) = x(e_j) + x(v_iv_{i+1}) = x(e_j).$$
+Suppose for a contradiction there exists $e_j = v_{j}v_{j+1} \in E(G)$ such that
+$$f(P(e_j)) < x(e_j).$$
+ Let $x' \in \Z^{|E|}$ we given by
+$$x'(e) = \begin{cases}
+x(e), &\text{ if $e\neq e_j$} \\
+f(P(e_j)), &\text{if $e=e_j$}. 
+\end{cases}$$
+We claim that $x'$ is feasible for $(\overline{IP})$.
+By the feasibility of $x$, we need only check cut constraints involving $e_j$.
+Let $P \in \cP$ such that $e_j \in \delta(P)$.
+Let $e$ be the other edge in $\delta(P)$.
+If $e = v_iv_{i+1}$ then $P = P(i+1, \ell)$ or $P = V\backslash P(i+1, \ell)$.
+In either case $f(P) = x(e_j)$.
+So we may assume $e \neq v_iv_{i+1}$.
+Then either $v_i, v_{i+1} \in P$ or $v_i, v_{i+1} \in V\backslash P$.
+By the symmetry of $f$ we may assume $v_i, v_{i+1} \in P$ (otherwise proceed by replacing $P$ with $V\backslash P$).
+Thus since $v_i,v_{i+1} \in P$ and $e_j \in \delta(P)$, we have either $P(e_j) \subseteq P$ or $(V\backslash P(e_j)) \subseteq P$.
+Again by symmetry we may assume $P(e_j) \subseteq P$.
+Similarly $(V\backslash P(e)) \subseteq P$.
+In fact,
+$$P = P(e_j) \cup (V\backslash P(e)).$$
+Hence we have
+\begin{align*}
+f(P) &\leq f(P(e_j)) + f(V\backslash P(e)) &\text{by subadditivity}
+\\
+&= f(P(e_j)) + f(P(e)) &\text{ by symmetry} \\
+&\leq f(P(e_j)) + x(\delta(P(e))) &\text{by feasibility of x} \\
+&= f(P(e_j)) + x(e) + x(v_iv_{i+1}) \\
+&= f(P(e_j)) + x(e) \\
+&= x'(e_j) + x'(e) &\text{by definition of $x'$} \\
+&= x'(\delta(P)).
+\end{align*}
+Thus $x'$ is feasible. But 
+\begin{align*}
+\sum_{e\in E} c(e) x(e) &= c(e_j)x(e_j) + \sum_{e\in E\backslash\{e_j\}} c(e)x(e) \\
+&> c(e_j)x'(e_j) +  \sum_{e\in E\backslash\{e_j\}} c(e)x(e) &\text{by contradiction hypothesis and our choice of $x'$} \\
+&= c(e_j)x'(e_j) + \sum_{e\in E\backslash\{e_j\}} c(e)x'(e) &\text{by our choice of $x'$} \\
+&= \sum_{e\in E} c(e) x'(e),
+\end{align*}
+contradicting the optimality of $x$.
+\end{proof}
+\paragraph{}
+This lemma tells us precisely how to construct optimal solutions to $(\overline{IP})$ in the presence of $0$-valued edges.
 \end{document}

### 2017-03-13 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
improved notation, and thus clarity in 3b

--- 2017-03-12 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
@@
 Since $G$ is a cycle we may label the vertices of $G$ as $V(G) = \{v_1,\dots, v_n\}$ such that $v_iv_j \in E(G)$ if and only if $$j \equiv i\pm1\mod n$$.
-For $i, \ell \in [n]$ ($[n] := \{1,\dots, n\}$) we define the path from $i$ of length $\ell$ as
-$$P(i, \ell) := \{v_i, v_{i+1}, \dots, v_{i + \ell} \}$$
-where addition is taken modulo $n$. We use the convention that when $\ell = n$ we have the entire cycle, and when $\ell = 0$ we have just $v_i$. We define the family of all paths on $G$ as
-$$\cP := \{P(i,\ell) : i, \ell \in [n]\}.$$
-Observe that the size of $\cP$ is polynomial in $n$, in particular $|\cP| = O(n^2)$ (since $n$ choices for $i$ and $n$ choices for $\ell$ covers all possible $P(i,\ell)$ with some overlap).
+For $v_i, v_j \in V(G)$ we define the path from $v_i$ to $v_j$ as
+$$P(i, \ell) := \{v_i, v_{i+1}, \dots, v_{j} \}$$
+where addition is taken modulo $n$. We define the family of all paths on $G$ as
+$$\cP := \{P(v_i, v_j) : v_i, v_j \in V(G)\}.$$
+Observe that the size of $\cP$ is polynomial in $n$, in particular $|\cP| = O(n^2)$ (since there are $n$ choices for $v_i$ and $n$ choices for $v_j$).
 \begin{lemma}\label{lemma:compact}
@@
 Let $S \subseteq V$.
+If $S = V$ then $f(S)= f(V)=0\leq x(\delta(S)) $ (otherwise $(\overline{IP})$ and $(IP)$ are infeasible).
+By symmetry of $f$ the same holds for $S =\emptyset$.
+Thus we may assume $\emptyset \subset S \subset V$.
 If we consider the connected components of $G[S]$ we observe that they are all paths.
@@
 \subsection*{b}
+\paragraph{}
+We make an observation about optimal solutions to $(\overline{IP})$ which contain a $0$-valued edge. This will lead us to an approximation algorithm with a better bound than the one found in $3(a)$.
+\begin{lemma}\label{lemma:0edge}
+Let $x$ be an optimal solution $(\bar{IP})$.
+Suppose there exists $v_iv_{i+1} \in E(G)$ such that $x(v_iv_{i+1}) = 0$.
+Reindex (by subtracting $i$ from each index) $V(G)$ so that $i = 0$.
+That is $x(v_0v_1)= 0$.
+Then for all $v_jv_{j+1} \in E(G)$, $j \neq 0$,
+$$x(v_jv_{j+1}) = f(P(v_1, v_j)).$$
+\end{lemma}
+\begin{proof}
+For each $v_jv_{j+1} \neq v_0v_{1} \in E(G)$. Since $x$ is feasible and $x(v_0v_{1}) = 0$,
+$$f(P(v_1, v_j)) \leq x(\delta(P(v_1,v_j))) = x(v_jv_{j+1}) + x(v_0v_{1}) = x(v_jv_{j+1}).$$
+Suppose for a contradiction there exists $e_j = v_{j}v_{j+1} \in E(G)$ such that
+$$f(P(v_1,v_j)) < x(v_jv_{j+1}).$$
+Since $f(V) = 0$, $j\neq 0$. Let $x' \in \Z^{|E|}$ we given by
+$$x'(e) = \begin{cases}
+x(e), &\text{ if $e\neq v_jv_{j+1}$} \\
+f(P(v_1,v_{j})), &\text{if $e=v_jv_{j+1}$}. 
+\end{cases}$$
+We claim that $x'$ is feasible for $(\overline{IP})$.
+By the feasibility of $x$, we need only check cut constraints involving $v_jv_{j+1}$.
+Let $v_i \in V(G)$, such that $i \neq j-1$.
+We consider $P(v_i,v_j)\in \cP$.
+\paragraph{}
+First consider the case where $0 < i \leq j$. In this case $v0,v_1 \in P(v_{j+1}, v_{i-1})$. We compute:
+\begin{align*}
+f(P(v_i,v_j)) &= f(P(v_{j+1}, v_{i-1})) &\text{by symmetry} \\
+&\leq f(P(v_{j+1}, v_0)) + f(P(v_{1},v_{i-1})) &\text{by subadditivity} \\
+&= f(P(v_1, v_j)) + f(P(v_1, v_{i-1})) &\text{by symmetry} \\
+&= x'(v_jv_{j+1}) + f(P(v_1, v_{i-1})) &\text{by definition of $x'$} \\
+&\leq x'(v_jv_{j+1}) + x(v_1v_{i-1}) +x(v_0v_1) &\text{by feasibility of $x$} \\
+&= x'(v_jv_{j+1}) + x'(v_1v_{i-1}) &\text{by definition of $x'$, and $x(v_0v_1)=0$} \\
+&=x'(\delta(P(v_i,v_j))).
+\end{align*}
+Hence $x'$ satisfies the cut constraint.
+\paragraph{}
+Now consider the case where $i > j$. In this case $v_0,v_1 \in P(v_i,v_j)$. We compute (similarly):
+\begin{align*}
+f(P(v_i,v_j))&\leq f(P(v_{i}, v_0)) + f(P(v_{1},v_{j})) &\text{by subadditivity} \\
+&= f(P(v_1, v_{i-1})) + f(P(v_1, v_{j})) &\text{by symmetry} \\
+&= f(P(v_1,v_{i-1})) + x'(v_jv_{j+1}) &\text{by definition of $x'$} \\
+&\leq x(v_1v_{i-1}) + x(v_0v_1) + x'(v_jv_{j+1}) &\text{by feasibility of $x$} \\
+&= x'(v_jv_{j+1}) + x'(v_1v_{i-1}) &\text{by definition of $x'$, and $x(v_0v_1)=0$} \\
+&=x'(\delta(P(v_i,v_j))).
+\end{align*}
+Therefore in either case $x'$ is feasible. But
+\begin{align*}
+\sum_{e\in E} c(e) x(e) &= c(v_jv_{j+1})x(v_jv_{j+1}) + \sum_{e\in E\backslash\{v_jv_{j+1}\}} c(e)x(e) \\
+&> c(v_jv_{j+1})x'(v_jv_{j+1}) +  \sum_{e\in E\backslash\{v_jv_{j+1}\}} c(e)x(e) &\text{since $x'(v_jv_{j+1}) = f(P(v_1,v_j) < x(v_jv_{j+1})$} \\
+&= c(v_jv_{j+1})x'(v_jv_{j+1}) + \sum_{e\in E\backslash\{v_jv_{j+1}\}} c(e)x'(e) &\text{by our choice of $x'$} \\
+&= \sum_{e\in E} c(e) x'(e),
+\end{align*}
+contradicting the optimality of $x$.
+\end{proof}
+\paragraph{}
+This lemma tells us precisely how to construct optimal solutions to $(\overline{IP})$ in the presence of $0$-valued edges.
 \end{document}

### 2017-03-13 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
3b

@@
 This lemma tells us precisely how to construct optimal solutions to $(\overline{IP})$ in the presence of $0$-valued edges.
+The idea for our algorithm is to construct from this lemma each potential solution with a $0$-valued edge, as well as an approximate solution with no $0$-valued edges, and return the best possible of those $|E|+1$ solutions.
+\paragraph{}
+In our algorithm we will use the following linear program, which is $(\overline{LP})$ with the constraint $x \geq 1$ instead of $x\geq 0$. We define the linear program $(\tilde{LP})$ as
+\begin{align*}
+\min\ \sum_{e\in E} c(e) x_e&\\
+\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
+x&\geq 1.
+\end{align*}
+\paragraph{}
+Consider the following improved algorithm for solving $(\overline{IP})$ which we will denote by $\cA'$:
+\begin{enumerate}
+\item Let $E' = \emptyset$. For each $e=v_iv_{i+1} \in E$:
+	\begin{enumerate}
+	\item Construct the solution $x^{e}$ as follows.
+	\item Set $x^{e}(e)=0$.
+	\item For each $v_jv_{j+1} \neq e$ Set $x^{e}(v_jv_{j+1}) = f(P(v_{i+1},v_j))$.
+	\item If $x^{e}$ is feasible for $(\overline{IP})$ add $e$ to $E'$.
+	\end{enumerate} 
+\item Solve $(\tilde{LP})$ on $G$  returning solution $\bar{x}$.
+\item Let $\bar{x}_I \in \Z^{|E|}$ denote the integral part of $\bar{x}$. Let $\bar{x}_F \in [0,1)^{|E|}$ denote the fractional part of $\bar{x}$. Thus $\bar{x} = \bar{x}_I + \bar{x}_F.$
+\item Construct solution $x \in \Z^{|E|}$ by rounding. For each $e \in E$ set:
+$$x_e = \begin{cases}
+\bar{x}_I + 1, &\text{if } x_F \geq 0.5 \\
+\bar{x}_I, &\text{otherwise.}
+\end{cases}$$
+\item Return $\arg\min\{c^Tx : x \in \{x^e : e \in E'\} \cup \{x\}\}$.
+\end{enumerate}
+\begin{theorem}
+The algorithm $\cA'$ is a $\frac{4}{3}$-approximation algorithm for $(\overline{IP})$.
+\end{theorem}
+\begin{proof}
+Step $1$ can be done in $O(|E|)$ iterations, each using $O(|E|)$ calls to the (we may assume polynomial time) oracle for $f$, and hence runs in polynomial time.
+Step $2$ can be done in polynomial time via the same argument used for $(\overline{LP})$.
+Step $4$ can be done in $O(|E|)$ time, as can step $5$.
+Therefore $\cA'$ runs in polynomial time.
+\paragraph{}
+By definition, for each $e \in E'$, $x^{e}$ is feasible.
+Since the feasible region of $(\tilde{LP})$ is a subset of the feasible region of $(\overline{LP})$, $\bar{x}$ is feasible for $(\overline{LP})$ and therefore by lemma \ref{lemma:feasible-rounding} $x$ is feasible for $(\overline{IP})$ as Step $4$ of $\cA'$ is the same rounding procedure as step $3$ of $\cA$.
+Hence $\cA'$ returns a feasible solution for $(\overline{IP})$.
+\paragraph{}
+If there exists an optimal solution $x^*$ for $(\overline{IP})$ and $e \in E$ such that $x^*(e) = 0$ then $x^*(e) = x^{e}$ by lemma \ref{lemma:0edge} and $e \in E'$.
+Thus in step $5$ $\cA'$ returns a solution of cost at most that of $x^{e} = x^*$ which is therefore optimal.
+Trivially such a solution is of cost at most $\frac{4}{3}$ that of optimal.
+\paragraph{}
+It remains to consider the case where no optimal solution $x^*$ for $(\overline{IP})$ has $e\in E$ such that $x^*(e) = 0$.
+Since $x^*$ is integral and no component is zero optimal solutions for $(\overline{IP})$ satisfy the constraint $x\geq 1$.
+Therefore $(\tilde{LP})$ is a relaxation of $(\overline{IP})$.
+\paragraph{}
+Now it suffices to show that the returned solution $x$ has cost at most $\frac{4}{3}$ that of $\bar{x}$.
+Let $e \in E$.
+If $\bar{x}(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq \frac{4}{3}\bar{x}(e)$.
+It remains to consider the pertinent case when $\bar{x}(e) \geq 0.5$.
+Therein we have
+\begin{align*}
+\frac{x(e)}{\bar{x}(e)} &= \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) + \bar{x}_F(e)} \\
+&\leq  \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) +0.5} \\
+&= \frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}.
+\end{align*}
+Since $\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}$ is a non-increasing function, it is maximized when $\bar{x}_I(e) = 1$ (since $\bar{x}_I(e) \geq 1$). Hence
+$$\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1} \leq \frac{2\cdot 1 + 2}{2\cdot 1 + 1} = \frac{4}{3}.$$
+Thus $x(e) \leq \frac{4}{3}\bar{x}(e)$ as desired. 
+\end{proof}
 \end{document}

### 2017-03-13 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
starting 4

@@
 \end{proof}
+
+%Q4
+\section{}
+\paragraph{}
+Consider the packing problem $(P)$: Let $U$ be a set of elements.
+Let $\cS$ be a non-empty family of sets $S\subseteq U$.
+The goal is to find a subcollection of $\cS$ of maximum cardinality, subject to the constraint that each element is contained in at most $1$ set.
+\paragraph{}
+We can write a linear programming relaxation for $(P)$, denoted $(LP)$ as
+\begin{align*}
+\max \sum_{S\in\cS} x_S\ & \\
+\text{s.t.} \sum_{S:e\in S} x_S &\leq 1 &\forall e\in U \\
+x&\geq 0.
+\end{align*}
+Let $k = \max_{S\in \cS} \{ |S|\}$.
+%pa
+\subsection*{a}
+
 \end{document}

### 2017-03-13 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
problem 4

--- 2017-03-13 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
@@
 \end{proof}
+
+%Q4
+\section{}
+\paragraph{}
+Consider the packing problem $(P)$: Let $U$ be a set of elements.
+Let $\cS$ be a non-empty family of sets $S\subseteq U$.
+The goal is to find a subcollection of $\cS$ of maximum cardinality, subject to the constraint that each element is contained in at most $1$ set.
+\paragraph{}
+We can write a linear programming relaxation for $(P)$, denoted $(LP)$ as
+\begin{align*}
+\max \sum_{S\in\cS} x_S\ & \\
+\text{s.t.} \sum_{S:e\in S} x_S &\leq 1 &\forall e\in U \\
+x&\geq 0.
+\end{align*}
+Let $k = \max_{S\in \cS} \{ |S|\}$.
+%pa
+\subsection*{a}
+\paragraph{}
+Let $x$ be a basic optimal solution to $(LP)$.
+Suppose for a contradiction that for all $S \in \cS$, $x_S < \frac{1}{k}$.
+We will contradict the fact that the number of tight constraints with respect to $x$ is at least the number of variables, that is $|\cS|$.
+We will proceed by a token argument.
+Give a token to each $S \in \cS$.
+Redistribute the tokens according to the following rules, for each $S \in \cS$:
+\begin{enumerate}
+\item If $x_S = 0$ distribute the token to the non-negativity constraint corresponding to $S$.
+\item Otherwise, for each $e \in S$ distribute $x_S$ token value to $e$.
+\end{enumerate}
+To achieve the desired contradiction, we need to show three things.
+First, that each $S \in \cS$ gives away at most $1$ token.
+Second, that each tight constraint receives at least $1$ token.
+Third, that some $S \in \cS$ keeps some fraction of their token.
+Together these propositions imply that $|\cS|$ is greater than the number of tight constraints at $x$.
+We complete the proof by showing these three porpositions.
+\paragraph{}
+We begin with the first proposition. If $x_S = 0$ then $S$ gives away at most one token by rule $1$. If $x_S > 0$ then $S$ gives away $\sum_{e: e\in S} x_S$ token value. Observe that
+$$\sum_{e: e\in S} x_S < \sum_{e: e\in S} \frac{1}{k} = \frac{|S|}{k} \leq 1.$$
+Thus $S$ gives away strictly less than one token.
+Thus to show proposition three we need only to show that there exists some $S$ with $x_S > 0$.
+This follows since $x$ is optimal.
+If all $S$ had $x_S = 0$, since $\cS \neq\emptyset$ we could simply set $x_S = 1$ for one $S \in \cS$ and increase cardinality, contradicting the optimality of $x$.
+Hence some $S$ exists with $x_S > 0$.
+Therefore we have show propositions one and three.
+\paragraph{}
+It remains to verify proposition two: that each tight constraint receives at least $1$ token.
+It is easy to see by rule $1$ that each tight non-negativity constraint receives $1$ token.
+Now consider a tight constraint corresponding to some $e \in U$.
+By rule two $e$ receives
+$$\sum_{S : e\in S, x_S > 0} x_S$$
+token value. But this is equal to
+$$\sum_{S: e\in S} x_S$$
+which in turn, since $e$ corresponds to a tight constraint satisfies
+$$\sum_{S: e\in S} x_S = 1.$$
+That is, $e$ receives $1$ token. Therefore every tight constraint receives $1$ token and this completes the proof. $\blacksquare$
+%pb
+\subsection*{b}
+\paragraph{}
+Consider the following iterative rounding algorithm, $\cA$, for packing problem $(P)$:
+\begin{enumerate}
+\item Set $H = \emptyset$, $\cS' = \cS$, and $U' = U$.
+\item While $\cS' \neq \emptyset$ do:
+	\begin{enumerate}
+	\item Let $x$ be a basic optimal solution to $(LP)$ on $\cS'$ and $U'$.
+	\item Let $S \subseteq \cS'$ such that $x_S \geq \frac{1}{k}$ where $k = \max_{S' \in \cS'}\{|S'|\}$ .
+	\item Set $H = H \cup \{S\}$.
+	\item For each $S' \in \cS'$ such that $S' \cap S \neq \emptyset$, remove $S'$ from $\cS'$.
+	\item For each $e \in S$, remove $e$ from $U'$. 
+	\end{enumerate}
+\item Return $H$.
+\end{enumerate}
+\begin{theorem}
+The algorithm $\cA$ is a $\frac{1}{k-1+\frac{1}{k}}$-approximation algorithm for $(P)$.
+\end{theorem}
+\begin{proof}
+\paragraph{}
+To see that $\cA$ terminates observe that by problem $3(a)$, a set $S \subseteq \cS'$ is always found in step $2(b)$, and hence $\cS'$ is eventually empty.
+To verify that $\cA$ runs in polynomial time, observe that $(LP)$ has a constraint for every $e \in U$ and a variable for every $S \in \cS$.
+Thus the size of $(LP)$ is polynomial in the size of the input to the problem.
+Each subsequent $(LP)$ is smaller than the previous, thus $2(a)$ runs in polynomial time.
+Step $2(b)$ can be done in $O(|\cS|)$ time, $2(c)$ in constant time, $2(d)$ in $O(|\cS|^2)$ time, and step $2(e)$ in $O(k|U|)$ time.
+These are all very pessimistic estimates,  but in any case they show that each iteration of step $2$ runs in polynomial time.
+Step $2$ has at most $|\cS|$ iterations and hence runs in polynomial time.
+Thus $\cA$ terminates in polynomial time.
+\paragraph{}
+Now we verify that the solution returned is feasible for $(P)$.
+We claim that $\cA$ maintains the invariant that after each iteration of step $2$, for each $S \in \cS'$ and each $T \in H$, $S\cap T = \emptyset$.
+It is easy to see that this invariant holds before step $2$.
+Suppose it holds inductively at the beginning of some iteration of step $2$.
+Let $S$ be the set chosen to enter $H$ in step $2(c)$.
+By step $2(d)$ each $S' \in \cS'$ that intersects $S$ is removed from$ \cS'$.
+Since no set in $\cS'$ intersects sets in $H$ by induction, after step $2(d)$ no set in $\cS'$ intersects $H \cup \{S\}$.
+Therefore the invariant holds.
+Using this invariant it is easy to see that at each iteration of step $2$ a set is added to $H$ which intersects no other set in $H$.
+Therefore at the termination of step $2$ the set $H$ returned is feasible for $(P)$.
+\paragraph{}
+It remains to demonstrate the approximation factor.
+We will bound the increase in $H$ versus the decrease in the fractional solution at each iteration of step $2$.
+Let $\Delta_i H$ denote the change in $|H|$ from the beginning of iteration $i$ to the end of iteration $i$..
+Exactly one set is added to $H$ in each iteration so $\Delta_i H = 1$ for all $i$.
+Let $x^i$ be the fractional solution to $(LP)$ found at the beginning of iteration $i$.
+Let $\Delta_i x = \sum_{S \in \cS}x^{i}_S - \sum_{e \in \cS}x^{i+1}_S$, where we take $x^{n+1} = 0$ if step $2$ terminates in $n$ iterations, and take $x^{i}_S = 0$ if $S \not\in \cS'$ during iteration $i$.
+\paragraph{}
+Consider an arbitrary iteration of step $2$ of $\cA$.
+Let $x^i$ be the solution to $(LP)$ found in step $2(a)$.
+Let $S$ be the set chosen in step $2(b)$.
+For each $e \in S$, let $\cS^e = \{S' \in \cS'\backslash\{S\} : e \in S'\}$.
+Let the collection of deleted sets from step $2(d)$ be $D: = \bigcup_{e\in S} \cS^e$.
+Then
+$$x(D) \leq \sum_{e \in S} x(\cS^{e}) = \sum_{e\in S} \sum_{S' \in \cS'\backslash\{S\} : e\in S'} x_{S'} \leq \sum_{e \in S} (1 - x_S) = |S|(1-x_S),$$
+with the last inequality following from the corresponding packing constraints of $(LP)$ for each $e$.
+Now we observe that $x^{i}$ restricted to the variables of $\cS' \backslash (\{S\} \cup D)$ is feasible for $(LP)$ during iteration $i+1$.
+Hence $$\Delta_i x \leq \sum_{S'\in \cS'} x^i_{S'} - \sum_{S' \in \cS'\backslash (\{S\} \cup D)\}} x^i_{S'} = x_S + x(D).$$
+Thus by our bound on $x(D)$ we have
+\begin{align*}
+\Delta_i x &\leq x_S + |S|(1-x_S) \\
+&= 1 - (1-x_S) + |S|(1-x_S) \\
+&= 1 + (|S| -1)(1-x_S) \\
+&\leq 1+(k - 1)(1-x_S) &\text{by our choice of $k$ in step $2(b)$} \\
+&\leq 1 + (k-1)(1-\frac{1}{k}) &\text{by our choice of $x_S$} \\
+&= 1 + k - 1 - 1 + \frac{1}{k} \\
+&= k-1 + \frac{1}{k}.
+\end{align*}
+So we observe that $$\Delta_i H - \frac{1}{k-1+\frac{1}{k}}\Delta_i x \geq 0.$$
+Summing over each iteration $i$ we obtain
+$$\sum_i \Delta_i H - \frac{1}{k-1+\frac{1}{k}} \sum_i \Delta_i x \geq 0.$$
+Since $\sum_i \Delta_i H = |H|$, and $\sum_i \Delta_i x = \sum_i (x^i - x^{i+1}) = x^1$, we obtain
+$$|H| - \frac{1}{k-1+\frac{1}{k}}x^1 \geq 0.$$
+Since $x^1$ is an optimal solution to $(LP)$ for our original problem $(P)$ this implies that $|H|$ is at least $\frac{1}{k-1+\frac{1}{k}}$ times the cardinality of an optimal solution to $(P)$.
+\end{proof}
 \end{document}

### 2017-03-14 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
q5 partial

@@
 So we observe that $$\Delta_i H - \frac{1}{k-1+\frac{1}{k}}\Delta_i x \geq 0.$$
-Summing over each iteration $i$ we obtain
+Summing over each iteration $i$ (and taking $k$ to be the initial $k$ from the first iteration, since $k$ is non-increasing as the algorithm proceeds) we obtain
 $$\sum_i \Delta_i H - \frac{1}{k-1+\frac{1}{k}} \sum_i \Delta_i x \geq 0.$$
@@
 \end{proof}
+
+%Q5
+\section{}
+\paragraph{}
+Academic Honesty Disclaimer: In this problem I rely heavily on her paper Dr. Sanita gave us on Learn: Steiner Tree Approximation via Iterative Randomized Rounding.
+\paragraph{}
+Given a metric graph $G=(V,E)$ with edge costs $c:E\rightarrow \R_+$ and a set of terminals $R\subseteq V$, the $k$-restricted Steiner Tree problem asks for a tree of minimum cost spanning $R$ whose full components each contain at most $k$ terminals.
+From class the following algorithm was shown to be a $\ln(4)$-approximation algorithm for this problem.
+\paragraph{}
+Let $\cC$ denote the set of all directed components which can be obtained in the following way.
+Let $R' \subseteq R$ such that $|R'| \leq k$.
+and let $r' \in R'$.
+Let $C$ be the minimum cost Steiner tree on $R'$, and direct all edges of $C$ towards $r'$.
+Then $r'$ denotes the unique sink terminal in this case.
+A component $C \in \cC$ ``crosses" $U\subseteq R$ if $C$ has at least one source in $U$ and its unique sink in $R\backslash U$.
+The set $\delta^+(U)$ will denote the set of $C \in \cC$ which cross $U$.
+Fix an arbitrary root terminal $r \in R$.
+The directed cut relaxation $(k-DCR)$ is
+\begin{align*}
+\min \sum_{C\in \cC} c(C)x_C& \\
+\text{s.t.} \sum_{C\in\delta^+(U)} x_C &\geq 1 &\forall U\subseteq R\backslash \{r\}, U\neq \emptyset \\
+x&\geq 0.
+\end{align*}
+As argued in class this is a relaxation of the $k$-restricted Steiner tree problem. The iterative randomized rounding algorithm, $\cA$, for this problem is:
+\begin{enumerate}
+\item For $t =1,2,\dots$
+\begin{enumerate}
+\item Solve $(k-DCR)$ on $G$ to obtain $x^t$.
+\item Sample one component $C^t$ where $C^t=C$ with probability $x^t_C/\sum_{C\in\cC} x^t_C$. Contract $C^t$ into its sink in $G$.
+\item If a single terminal remains, return $\bigcup_{i=1}^t C^i$.
+\end{enumerate}
+\end{enumerate}
+\paragraph{}
+Since we already know from class that $\cA$ is a $\ln(4)$-approximation algorithm for $k$-restricted Steiner Tree, it only remains to verify the new approximation factor of $\frac{5}{4}$ when we fix $k=4$.
+Let $S^*$ be an optimal $4$-restricted Steiner Tree with undirected full components $Z_1, \dots, Z_\ell$.
+We will start by studying the structure of each $Z_i$.
+The first observation, made in the paper, is that we may assume Steiner Nodes of $S^*$ have degree at least $3$.
+This follows since $G$ is a metric graph we may shortcut any Steiner Node of degree two by connecting its endpoints and removing it.
+By the triangle inequality this shortcutting operation does not increase the cost of the solution.
+\paragraph{}
+We now move to observations about each $Z_i$.
+\begin{lemma}\label{lemma:dk}
+If $Z_i$ is a $k$-component for any $k\leq 4$, then the degree of any Steiner Node in $Z_i$ is at most $k$.
+\end{lemma}
+\begin{proof}
+Suppose for a contradiction that $Z_i$ has a Steiner Node $s$ of degree greater than $k$.
+Then $s$ has at least $k+1$ neighbours: $v_1, \dots, v_{k+1}$.
+Since $Z_i$ is a tree, there is a unique path from $s$ to each of the $k$ terminals of $Z_i$, $r_1, \dots, r_k$.
+Each path $P_{s,r_j}$ can use only one of $v_1,\dots, v_{k+1}$.
+If they use more, say first $v_i$ and then next $v_j$ then the subpath $r,v_i,\dots,v_j$ with the edge $v_js$ forms a cycle in $Z_i$ contradicting that $Z_i$ is a tree.
+But then one of $v_1, \dots, v_{k+1}$ is not used on any path from $s$ to any of $r_1, \dots, r_k$.
+The edge from $s$ to the unused $v_j$ can then be removed from $Z_i$ without disconnecting the terminals, contradicting optimality.
+\end{proof}
 \end{document}

### 2017-03-14 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
q5

@@
 \begin{lemma}\label{lemma:dk}
-If $Z_i$ is a $k$-component for any $k\leq 4$, then the degree of any Steiner Node in $Z_i$ is at most $k$.
+If $Z_i$ is a $k$-restricted component for any $k$, then the degree of any Steiner Node in $Z_i$ is at most $k$.
 \end{lemma}
@@
 \end{proof}
+\paragraph{}
+We now consider the classes of graphs these components appear as.
+It is an easy observation that $1$-restricted components only appear in the trivial case where $|R| = 1$.
+It is also easy to see that $2$-restricted components arise as a single edge connecting two terminals since Steiner Nodes have degree at least three.
+Further we see that $k$-restricted components with exactly one Steiner Node are star graphs centred at the Steiner Node.
+By lemma \ref{lemma:dk} $3$-restricted components have a Steiner Nodes with degree exactly three, and thus they can only have one Steiner Node.
+This implies that $3$-restricted components are star graphs.
+Things get slightly more interesting with $4$-restricted components.
+\begin{lemma}\label{lemma:four}
+A $4$-restricted component $Z_i$ is either a star graph or consists of exactly two adjacent degree three Steiner Nodes each adjacent to precisely two distinct terminals.
+\end{lemma}
+\begin{proof}
+Suppose $Z_i$ is not a star graph.
+Then $Z_i$ has more than one Steiner Node.
+We claim that $Z_i$ has at most two adjacent Steiner Nodes.
+Suppose for a contradiction $Z_i$ has at least three adjacent Steiner Nodes.
+Say $s_1,s_2,s_3$ is a path of Steiner Nodes in $Z_i$.
+If the removal of edge $s_1s_2$ separates one terminal in the component containing $s_1$ then $s_1$ is adjacent to a redundant edge since its degree is at least three.
+So the edge removal separates at least two terminals in the component containing $s_1$.
+Similarly, by removing $s_2s_3$ we see that there are two terminals in the component containing $s_3$.
+But then, since $s_2$ of degree at least three, any third component it connects to (that is, not containing $s_1$ or $s_3$) is redundant to connect the terminals.
+This violates optimality.
+Hence $Z_i$ contains exactly two Steiner Nodes, and we see by the above argument that the removal of their adjacent edge separates two terminals into each connected component.
+\end{proof}
+\paragraph{}
+We are prepared to analyze the algorithm.
+Following the notation in the paper we let $D(e) = \max\{t : e \in S^t\}$ be the iteration at which $e$ is deleted.
+We construct the witness tree via the following procedure:
+\begin{enumerate}
+\item Start with $\tilde{B} = \emptyset$
+\item For each $i = 1,\dots, \ell$
+	\begin{enumerate}
+	\item If $Z_i$ is a $2$-restricted component add its edge to $\tilde{B}$.
+	\item If $Z_i$ is a $3$-restricted component or $4$-restricted star graph component add all the edges of $Z_i$ but one chosen uniformly at random.
+	\item If $Z_i$ is a $4$-restricted component with two adjacent Steiner Nodes $s_1,s_2$ and terminals $r_1, r_2$ connected to $s_1$ and terminals $r_3,r_4$ connected to $s_2$ then:
+	\begin{enumerate}
+		 \item add edge $s_1s_2$ to $\tilde{B}$
+		 \item add one edge of $r_1s_1$, $r_2s_1$ to $\tilde{B}$ chosen uniformly at random
+		 \item add one edge of $r_3s_2$, $r_4s_2$ to $\tilde{B}$ chosen uniformly at random
+\end{enumerate}
+	\end{enumerate}
+\item Return Witness Tree $W= \{\{u,v\} in {R\choose 2} : |P_uv \cap \tilde{B}| = 1\}$.
+\end{enumerate}
+\paragraph{}
+By lemma \ref{lemma:four} and the preceding discussion, this procedure covers all possible forms the components $Z_i$ can take. Further one can verify that $W$ is a terminal spanning tree. 
+\paragraph{}
+For any $e \in E(S^*)$ we define $W(e) = \{ uv \in E(W): e \in P_{uv}\}$.
+When the edges of $W(e)$ are $\textit{marked}$ (see paper) the edge $e$ is deleted.
+We have the following lemma from the paper/class (rephrased in terms of $W(e)$ for simplicity):
+\begin{lemma}\label{lemma:20}
+Let $e \in E(S^*)$. Then the expected number of iterations until all the edges of $W(e)$ are $\textit{marked}$ satisfies $$E[X(W(e))] \leq H_{|W(e)|} \cdot M,$$
+where $M$ is an upperbound on $|\cC|$ polynomial in $n$.
+\end{lemma}
+\begin{lemma}\label{lemma:de}
+For any edge $e \in E(S^*)$, the expected number of iterations until $e$ is deleted is
+$$E[D(e)] \leq \frac{5}{4} M.$$
+\end{lemma}
+\begin{proof}
+Let $Z_i$ be the undirected full component containing $e$. We proceed by case analysis on the class of graph $Z_i$ belongs to. We have the following cases:
+\begin{enumerate}
+\item $Z_i$ is $2$-restricted,
+\item $Z_i$ is $3$ or $4$-restricted star,
+\item $Z_i$ is a $4$-restricted component with two degree three Steiner Nodes.
+\end{enumerate}
+\paragraph{}
+We begin with Case $1$. This case is trivial: with probability $1$, $e$ is the edge connecting the terminals. Thus by lemma \ref{lemma:20} $E[D(e)] = 1\cdot H_1 \cdot M = M \leq \frac{5}{4} M.$
+\paragraph{}
+Now we consider Case $2$.
+Suppose $Z_i$ is a $k$-restricted star where $k \in \{3,4\}$.
+Then with probability $\frac{1}{k}$ (when $e$ is the edge not in $\tilde{B}$), $|W(e)| = k-1$.
+With probability $1-\frac{1}{k}$), $|W(e)| = 1$.
+Thus by lemma \ref{lemma:20} we have
+\begin{align*}E[D(e)] &\leq \frac{1}{k} H_{k-1} M + (1-\frac{1}{k})H_1 M
+\\&=\begin{cases}
+(\frac{1}{3}H_2 + \frac{2}{3})M, &\text{if $k = 3$} \\
+(\frac{1}{4}H_3 + \frac{3}{4})M, &\text{if $k=4$}.
+\end{cases} \\
+&=\begin{cases}
+\frac{7}{6}M, &\text{if $k = 3$} \\
+(\frac{29}{24})M, &\text{if $k=4$}.
+\end{cases} \\
+&\leq \frac{5}{4} M.
+\end{align*}
+\paragraph{}
+Finally we consider Case $3$.
+By lemma \ref{lemma:four} we have the following structure.Let $s_1,s_2$ be the two adjacent Steiner Nodes of $Z_i$ and let $r_1,r_2$ be the terminals adjacent to $s_1$, and let $r_3,r_4$ be the terminals adjacent to $s_2$.
+Suppose without loss of generality that $s_1r_1, s_2r_4 \in \tilde{B}$.
+The the probability that $|W(e)| = 1$ is $\frac{3}{5}$ (such edges are $s_1r_1, s_2r_4$ and $s_1s_2$).
+The probability that $|W(e)| = 2$ is $\frac{2}{5}$ (such edges are $s_1r_2$ and $s_2r_3$).
+Thus by lemma \ref{lemma:20} we have
+\begin{align*}
+E[D(e)] &\leq \frac{3}{5}H_1M + \frac{2}{5}H_2M \\
+&= (\frac{3}{5} + \frac{2}{10}) M \\
+&= (\frac{8}{10})M \\
+&= \frac{4}{5}M.
+\end{align*} 
+Therefore the expectation holds in each case as desired.
+\end{proof}
+\paragraph{}
+We now finish by computing the expected cost of the solution returned by $\cA$. Let $opt^t_f$ denote the optimal fractional cost to $(k-DCR)$ at iteration $t$. Let $opt$ denote the cost of $S^*$.
+\begin{theorem}
+The algorithm $\cA$ is $\frac{5}{4}$-approximation algorithm for $4$-restricted Steiner Tree.
+\end{theorem}
+\begin{proof}
+We observe that
+\begin{align*}
+E[\sum_{t\geq 1} c(C^t)] &\leq \sum_{t\geq 1} \frac{1}{M}E[opt^t_f] \\
+&\leq \frac{1}{M} \sum_{t\geq 1} E[c(S^t)] \\
+&=\frac{1}{M} \sum_{t\geq 1} E[D(e)]c(e) \\
+&\leq \frac{1}{M} \sum_{t\geq 1} \frac{5}{4}M\cdot c(e) &\text{by lemma \ref{lemma:de}} \\
+&=\frac{5}{4} \sum_{t \geq 1} c(e) \\
+&= \frac{5}{4} opt.
+\end{align*}
+\end{proof}
+
+%Q6
+\section{}
 \end{document}

### 2017-03-14 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
finished typing 6

@@
 Then with probability $\frac{1}{k}$ (when $e$ is the edge not in $\tilde{B}$), $|W(e)| = k-1$.
-With probability $1-\frac{1}{k}$), $|W(e)| = 1$.
+With probability $(1-\frac{1}{k}$), $|W(e)| = 1$.
 Thus by lemma \ref{lemma:20} we have
@@
 E[D(e)] &\leq \frac{3}{5}H_1M + \frac{2}{5}H_2M \\
-&= (\frac{3}{5} + \frac{2}{10}) M \\
-&= (\frac{8}{10})M \\
-&= \frac{4}{5}M.
+&= (\frac{3}{5} + \frac{3}{5}) M \\
+&= (\frac{6}{5})M \\
+&\leq \frac{5}{4}M.
 \end{align*} 
@@
 \section{}
+\paragraph{}
+In the metric Travelling Salesman Problem $(TSP)$ we are given a metric graph $G=(V,E)$ with edge costs $c: E\rightarrow \R_+$.
+The goal is to find a minimum cost Hamilton Cycle visiting each $v \in V$.
+The Held-Karp relaxation for $(TSP)$, denoted $(HK)$ is
+\begin{align*}
+\min \sum_e c(e) x_e& \\
+\text{s.t.} x(\delta(v)) &= 2 &\forall v \in V \\
+x(\delta(S)) &\geq 2 &\forall S \subseteq V \\
+x &\geq 0.
+\end{align*}
+Let $n = |V|$.
+Let $x$ be a basic solution to $(HK)$.
+Our goal is to show that $x$ has $O(n)$ non-zero variables.
+The number of variables in $x$ correspond to a maximum size set of linearly independent tight constraints.
+Therefore the non-zero variables correspond to a collection $\cL \subseteq 2^V$ such that $x(\delta(S) = 2$ for all $S \in \cL$, and $\{\chi(S): S \in \cL\}$ is maximal linearly independent and $\chi(S)$ denotes the incidence vector of $S$.
+Hence we need to show that $|\cL| = O(n)$.
+\paragraph{}
+The constant function $f: 2^V \rightarrow \{2\}$ is trivially weakly supermodular. Therefore we may invoke a theorem from class saying we may choose $\cL$ to be a laminar family.
+\paragraph{}
+To show $|\cL| = O(n)$ we proceed by induction on $|V|$, the size of the ground set.
+The claim is trivial for $|V| = 1$: $|\cL| \leq |V|  = O(n)$.
+Suppose $|V| >1$.
+For the inductive hypothesis suppose for all $V' \subseteq V$, any laminar family on ground set $V'$ has size $O(|V'|)$.
+Let $S \in \cL\backslash\{V\}$ such that $|S|$ is maximal.
+We may assume such $S$ exists, as otherwise $|\cL| = 1 = O(n)$ as desired.
+Since $\cL$ is laminar, for all $S' \in \cL \backslash \{V\}$ either $S' \subseteq S$ or $S' \cap S = \emptyset$.
+Let $\cL' = \{ S' \in \cL\backslash\{V\} : S' \subseteq S\}$ and let $\cL'' = \{S' \in \cL\backslash \{V\} : S'\cap S \neq \emptyset\}$.
+Since $\cL'$ and $\cL''$ are subsets of $\cL$ they are both laminar families.
+The ground set of $\cL'$ is $S$, and the ground set of $\cL''$ is $V\backslash S$.
+Therefore by the inductive hypothesis $|\cL'| = O(|S|)$ and $|\cL''| = O(|V\backslash S|)$.
+Also, $\cL \subseteq \{V\} \cup \cL' \cup \cL''$.
+Then
+$$|\cL| \leq |\{V\}| + |\cL'| + |\cL''| = 1 + O(|S|) + O(|V\backslash S|) = O(|V\backslash S| + |S| + 1) = O(|V|).$$
+Hence by induction $|\cL| = O(n)$ as desired. $\blacksquare$
 \end{document}

### 2017-03-16 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
Proofreading

@@
 Now we intend to show that the hypothesis of the previous lemma holds.
-Consider some iteration $i$ of $\cA$, and let $A$ be the of edges chosen so far at the end of $i$.
+Consider some iteration $i$ step $2$ of $\cA$, and let $A$ be the of edges chosen so far at the end of $i$.
 Let $B = F \cap A$.
@@
 We begin by working towards a compact formulation for $(IP)$ on $G$.
-Since $G$ is a cycle we may label the vertices of $G$ as $V(G) = \{v_1,\dots, v_n\}$ such that $v_iv_j \in E(G)$ if and only if $$j \equiv i\pm1\mod n$$.
-For $v_i, v_j \in V(G)$ we define the path from $v_i$ to $v_j$ as
+Since $G$ is a cycle we may label the vertices of $G$ as $V(G) = \{v_1,\dots, v_n\}$ such that $v_iv_j \in E(G)$ if and only if $$j \equiv i\pm1\mod n.$$ For $v_i, v_j \in V(G)$ we define the path from $v_i$ to $v_j$ as
 $$P(i, \ell) := \{v_i, v_{i+1}, \dots, v_{j} \}$$
@@
 Let $e \in E$.
-If $\bar{x}(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq 2\bar{x}(e)$.
-It remains to consider the pertinent case when $\bar{x}(e) \geq 0.5$.
+If $\bar{x}_F(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq 2\bar{x}(e)$.
+It remains to consider the pertinent case when $\bar{x}_F(e) \geq 0.5$.
 Therein we have
@@
 Thus $x(e) \leq 2\bar{x}(e)$ as desired. Therefore
-$$\sum_{e\in E}c(e) x(e) \leq \sum_{e \in E}c(e)2\bar{x}(e) = 2\sum_{e\in E}\bar{x}(e) \leq 2\cdot OPT,$$
+$$\sum_{e\in E}c(e) x(e) \leq \sum_{e \in E}c(e)2\bar{x}(e) = 2\sum_{e\in E}c(e)\bar{x}(e) \leq 2\cdot OPT,$$
 where $OPT$ is the cost of an optimal solution to $(\overline{IP})$ (the last inequality follows immediately from $\bar{x}$ being optimal for $(\overline{LP})$.
@@
 \paragraph{}
-If there exists an optimal solution $x^*$ for $(\overline{IP})$ and $e \in E$ such that $x^*(e) = 0$ then $x^*(e) = x^{e}$ by lemma \ref{lemma:0edge} and $e \in E'$.
-Thus in step $5$ $\cA'$ returns a solution of cost at most that of $x^{e} = x^*$ which is therefore optimal.
+If there exists an optimal solution $x^*$ for $(\overline{IP})$ and $e \in E$ such that $x^*(e) = 0$ then $x^* = x^{e}$ by lemma \ref{lemma:0edge} and $e \in E'$.
+Thus in step $5$, $\cA'$ returns a solution of cost at most that of $x^{e} = x^*$ which is therefore optimal.
 Trivially such a solution is of cost at most $\frac{4}{3}$ that of optimal.
@@
 Since $x^*$ is integral and no component is zero optimal solutions for $(\overline{IP})$ satisfy the constraint $x\geq 1$.
-Therefore $(\tilde{LP})$ is a relaxation of $(\overline{IP})$.
+Therefore $(\tilde{LP})$ is a relaxation of $(\overline{IP})$ in this case.
 \paragraph{}
@@
 Let $e \in E$.
-If $\bar{x}(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq \frac{4}{3}\bar{x}(e)$.
-It remains to consider the pertinent case when $\bar{x}(e) \geq 0.5$.
+If $\bar{x}_F(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq \frac{4}{3}\bar{x}(e)$.
+It remains to consider the pertinent case when $\bar{x}_F(e) \geq 0.5$.
 Therein we have
@@
 $$\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1} \leq \frac{2\cdot 1 + 2}{2\cdot 1 + 1} = \frac{4}{3}.$$
-Thus $x(e) \leq \frac{4}{3}\bar{x}(e)$ as desired. 
+Thus $x(e) \leq \frac{4}{3}\bar{x}(e)$ as desired. Therefore
+$$\sum_{e\in E}c(e) x(e) \leq \sum_{e \in E}c(e)\frac{4}{3}\bar{x}(e) = \frac{4}{3}\sum_{e\in E}c(e)\bar{x}(e)$$
+and the approximation factor follows.
 \end{proof}
@@
 $$\sum_{S: e\in S} x_S$$
-which in turn, since $e$ corresponds to a tight constraint satisfies
+which in turn, since $e$ corresponds to a tight constraint, satisfies
 $$\sum_{S: e\in S} x_S = 1.$$
@@
 We will bound the increase in $H$ versus the decrease in the fractional solution at each iteration of step $2$.
-Let $\Delta_i H$ denote the change in $|H|$ from the beginning of iteration $i$ to the end of iteration $i$..
+Let $\Delta_i H$ denote the change in $|H|$ from the beginning of iteration $i$ to the end of iteration $i$.
 Exactly one set is added to $H$ in each iteration so $\Delta_i H = 1$ for all $i$.
@@
 \paragraph{}
-Academic Honesty Disclaimer: In this problem I rely heavily on her paper Dr. Sanita gave us on Learn: Steiner Tree Approximation via Iterative Randomized Rounding.
+When I refer to ``the paper" below, I mean Dr. Sanita's paper she gave us on Learn: Steiner Tree Approximation via Iterative Randomized Rounding.
 \paragraph{}
@@
 	\end{enumerate}
-\item Return Witness Tree $W= \{\{u,v\} in {R\choose 2} : |P_uv \cap \tilde{B}| = 1\}$.
+\item Return Witness Tree $W= \{\{u,v\} \in {R\choose 2} : |P_{uv} \cap \tilde{B}| = 1\}$.
 \end{enumerate}
 \paragraph{}
-By lemma \ref{lemma:four} and the preceding discussion, this procedure covers all possible forms the components $Z_i$ can take. Further one can verify that $W$ is a terminal spanning tree. 
+By lemma \ref{lemma:four} and the preceding discussion, this procedure covers all possible forms the components $Z_i$ can take.
+Further it is immediate from our choice of $\tilde{B}$ that $W$ is a terminal spanning tree.
+A simple inspection of the edges taken in $W$ with respect to the terminals of each $Z_i$ verifies this.
 \paragraph{}
 For any $e \in E(S^*)$ we define $W(e) = \{ uv \in E(W): e \in P_{uv}\}$.
-When the edges of $W(e)$ are $\textit{marked}$ (see paper) the edge $e$ is deleted.
+When the edges of $W(e)$ are $\textit{marked}$ (see the paper) the edge $e$ is deleted.
 We have the following lemma from the paper/class (rephrased in terms of $W(e)$ for simplicity):

### 2017-03-25 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
Starting presentation

\documentclass[10pt]{beamer}

\usetheme[progressbar=frametitle]{metropolis}

\usepackage{booktabs}
\usepackage[scale=2]{ccicons}

\usepackage{pgfplots}
\usepgfplotslibrary{dateplot}

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}

\usepackage{tikz}
\usetikzlibrary[topaths]

\usepackage{wasysym}

\title{Popularity, Mixed Matchings, and Self-duality}
\subtitle{Chien-Chung Huang and Telikepalli Kavitha}
\date{\today}
\author{Justin Toth}
\institute{University of Waterloo}
%\titlegraphic{\hfill\includegraphics[height=1.5cm]{logo}}

%Macros
\newcommand{\A}{\mathbb{A}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
 \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS

\begin{document}

\maketitle

\section{Introduction}

\begin{frame}
\frametitle{Voting}
Consider a graph $\alert{G = (V,E)}$, and a matching $\alert{M}$ on $\alert{G}$.

We define the vote of $\alert{u}$ for $\alert{v}$ versus $\alert{v'}$ as
$$vote_u(v,v') = \begin{cases}
1, &\text{if $u$ prefers $v$ to $v'$} \\
-1, &\text{if $u$ prefers $v'$ to $v$} \\
0, &\text{otherwise}
\end{cases}
$$
We can aggregate the margin of victory in an election between two matchings $\alert{M}$ and $\alert{M'}$ by
$$\Delta(M,M') := \sum_{u \in V(G)} vote_u(M(u), M'(u)) $$
where $\alert{M(u)}$ denotes the partner of $\alert{u}$ in $\alert{M}$.
\end{frame}

\begin{frame}
\frametitle{Popular Matchings}
We call $\text{\alert{M}}$ a $\text{\alert{popular}}$ matching if it never loses elections:
$$\forall \text{matchings } M', \quad \Delta(M,M') \geq 0.$$

The max utility popular matching problem $\alert{\text{(MaxPM)}}$ asks to find a popular matching $\alert{M}$ on $\alert{G}$ which maximizes
$$\sum_{e\in M} w(e)$$
for some weights $\alert{ w: E(G) \rightarrow \Q_+}$.
\end{frame}

\begin{frame}
\frametitle{Fractional Matchings}
We can extend this to the fractional matching case:
$$FM_{G} = \{x \in \R^m_+ : x(\delta(u)) \leq 1, \forall u \in V(G)\}$$
By defining for all $\alert{x, y \in FM_{G}}$:
$$\Delta(x,y) = \sum_{u \in V(G)} \sum_{uv, uv' \in \delta(u)} x_{uv}y_{uv'} vote_u(v,v').$$
The polytope of all fractional popular matchings is 
$$P_G := \{x \in FM_{G}: \Delta(x,y) \geq 0, \forall y \in FM_G\}.$$
\end{frame}

\begin{frame}
\frametitle{Mixed Popular Matching Example}
In bipartite graphs, each $\alert{x \in FM_G}$ is a convex combination of matchings. But observe that each $x\in P_G$ need not be a convex combination of popular matchings.
\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}
\begin{figure}
\begin{tikzpicture} 
\node[] (a0) at (0,5) {$a_0$};
    \node[] (a1) at (0,4) {$a_1$};
    \node[] (a2) at (0,3) {$a_2$};

    \node[] (b1) at (2,4) {$b_1$};
    \node[] (b2) at (2,3) {$b_2$};

    \path[dashed] (a0) edge (b1);
    \path[dashed] (a0) edge (b2);
    \path[-] (a1) edge (b1);
    \path[-] (a2) edge (b2);
    \path[-,red] (a1) edge (b2);
    \path[-,red] (a2) edge (b1);
\end{tikzpicture}
\end{figure}
\end{column}%
\hfill%
\begin{column}{.48\textwidth}
$$ $$
Say each $\alert{a_i}$ has preferences $\alert{b_1 > b_2}$, and each $\alert{b_i}$ has preferences $\alert{a_1 > a_2 > a_0}$.
\end{column}%
\end{columns}

Solid black edges $\alert{S}$ is the only popular matching. If we let $\alert{M}$ be the solid red edges then $$\Pi= \{(S,\frac{1}{2}), (M,\frac{1}{2})\}$$ is a fractional popular matching.
\end{frame}
\section{Integrality of $P_G$ in a special case}

\section{Half-Integrality of $P_G$ in any bipartite instance}
%maybe omit this as next section generalizes and uses same ideas
\section{Half-Integrality of $P_G$ in a roommates instance}

\section{Hardness and Inapproximability}

\section{Open Problems}

\section{Introduction to Stable Matching}
\begin{frame}
\frametitle{Stable Matching Problem}
\begin{overprint}
\onslide<1-3>
Introduced by Gale and Shapley in 1962
\onslide<4-9>
How to solve? Gale Shapley Algorithm:
\onslide<10->
Not very fair to women. Can we do better?
\end{overprint}
\begin{figure}
\centering
\begin{tikzpicture}
   \node[] (m1) at (0,4) {Adam};
    \node[] (m2) at (0,3) {Bob};
    \node[] (m3) at (0,2) {Chad};
    \node[] (m4) at (0,1) {Dave};

    \node[] (w1) at (2,4) {Alice};
    \node[] (w2) at (2,3) {Brooke};
    \node[] (w3) at (2,2) {Carly};
    \node[] (w4) at (2,1) {Daisy};

     \node[] () at (-2,4){$\alert<10>{A}>B>C>D\quad:$};
    \node[] () at (-2,3){$B>\alert<10>{C}>A>D\quad:$};
    \node[] () at (-2,2){$A>\alert<10>{B}>C\quad:$};
    \node[] () at (-2,1){$\alert<10>{D}>C>B>A\quad:$};

    \node[] () at (4,4){$:\quad \alert<10>{A}>B>C$};
    \node[] () at (4,3){$:\quad D>A>\alert<10>{C}>B$};
    \node[] () at (4,2){$:\quad D>C>\alert<10>{B}>A$};
    \node[] () at (4,1){$:\quad A>B>C>\alert<10>{D}$};

    \path[-]<2-3> (m1) edge (w1);
    \path[-]<2-3> (m2) edge (w2);
    \path[-]<2-3> (m3) edge (w3);
    \path[-]<2-3> (m4) edge (w4);

    \path[-]<4-> (m1) edge (w1);
    \path[-]<5-7> (m2) edge (w2);
    \path[-]<6> (m3) edge (w1);
    \path[-]<7-> (m3) edge (w2);
    \path[-]<8-> (m2) edge (w3);
    \path[-]<9-> (m4) edge (w4);

    \path[-,decoration={zigzag,segment length=4,amplitude=.9,
  post=lineto,post length=2pt},color=red]<3> (m3) edge[decorate] (w2);

\end{tikzpicture}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Weighted Stable Matching}
Add weights to pairs to incentivize equitable pairings.\\
Example:
$$c_{\text{Adam,Alice}} = 8$$ since Adam and Alice really like each other,
$$c_{\text{Adam, Carly}} = 2$$ since Adam and Carly don't like each other much.\\
New Goal:
$$\max_{\{\text{Stable Matchings $M$}\}} \sum_{\text{$(m,w)$ Pairs in $M$}} c_{m,w}$$
\end{frame}

\section{Introduction to Linear Programs}

\begin{frame}
\frametitle{What is a Linear Program?}
Let $c, a^1, \dots, a^m \in \mathbb{R}^n$. Let $b \in \mathbb{R}^m$. Find $x \in \mathbb{R}^n$ which satisfies:
\begin{align*}
\max \sum_{i=1}^n c_i x_i&\  &\text{\alert{The Objective Function}}\\
\text{subject to } \sum_{i=1}^n a_i^1 x_i &\geq b_1\\
&\vdots &\text{\alert{The Feasible Region}}\\
\sum_{i=1}^n a_i^m x_i &\geq b_m \\
x &\geq 0
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Polytopes - Geometry of the Feasible Region}
\begin{center}
\begin{tikzpicture}
\draw[->] (0,0) node[left]{$0$} -- (0,5)node[above]{$x_2$};
\draw[->] (0,0) -- (6,0) node[right]{$x_1$};

\path[dashed]<1-> (-1,0) edge node[left]{$\langle a^1 x \rangle \geq b_1$} (3.2,5.6);
\path[dashed]<2-> (-1.5,5.5) edge node[right]{$\langle a^2x\rangle \geq b_2$} (7.5,0.5);
\path[dashed]<3-> (2.5,-1) edge node[right]{$\langle a^3 x \rangle \geq b_3$} (7,2);

\path[fill=lightgray]<4-> (0,0) -- (0,1.3) -- (1.75,3.65) -- (5.95,1.35) -- (4,0) -- (0,0);

\path[->] <5-> (2,1) edge node[right]{$c$} (2.25,2);
\draw<6>[] (1.75,3.65) circle (2mm) node[right]{\ \ Extreme Point Solution} ;
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Why Are Extreme Points Interesting?}
\begin{itemize}
\item We have very efficient algorithms for solving Linear Programs (\alert{Simplex}, \alert{Ellipsoid})
\item These algorithms always give \alert{extreme point} solutions
\item It can be shown that when we can find \alert{optimal solutions} we can always find them at \alert{extreme points}
\end{itemize}
\end{frame}
\section{Formulating Stable Matching as a Linear Program}

\begin{frame}
\frametitle{Incidence Vectors}
We can translate between Matchings and Vectors using \alert{Incidence Vectors} in $\mathbb{R}^E$ indexed by set of all acceptable pairs $E$
\begin{align*}
&\underline{Matching} &\underline{Real\ Vector} \\
&\text{(Adam,Alice)} &x_\text{Adam,Alice} = 1 \\
&\text{(Bob,Carly)} &x_\text{Bob,Carly} = 1\\
&\text{(Chad, Brooke)} &x_\text{Chad, Brooke} = 1\\
&\text{(Dave, Daisy)} &x_\text{Dave, Daisy} = 1 \\
\end{align*}
$x_{m,w} = 0$ for all other pairs $(m,w)$ not in Matching
\end{frame}

\begin{frame}
\frametitle{Rothblum's Formulation}
\begin{align*}
\onslide<1,2->{\max \sum_{\text{Acceptable Pairs } (m,w)} c_{m,w} x_{m,w}}\\
\onslide<1,3->{\text{subject to } \sum_{w \text{ acceptable to } m} x_{m,w} &\leq 1 &\text{For all Men $m$}}\\
\onslide<1,4->{\sum_{m\text{ acceptable to }w} x_{m,w} &\leq 1 &\text{For all Women $w$}} \\
\onslide<1,5->{x_{m,w} + \sum_{\substack{w' \text{ preferred to $w$}\\ \text{by $m$}}} x_{m, w'} + \sum_{\substack{\text{$m'$ preferred to $m$}\\\text{ by $w$}}} x_{m',w} &\geq 1 &\text{For all pairs $(m,w)$}\\}
\onslide<1,6->{1 \geq x &\geq 0}
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Are Extreme Points Incidence Vectors?}
\begin{itemize}
\item We know $0$-$1$ solutions to our Linear Program are Stable Matchings
\item But what if our LP Solver gives an extreme point solution with \alert{$X_{m,w} = \frac{1}{2}$}?
\item We need to show that this cannot happen
\end{itemize}
\end{frame}

\section{Locality Technique}
\begin{frame}
\frametitle{Locality Proof Strategy}
We argue by contradiction:
\begin{enumerate}
\item Start with a minimal counterexample
\item \alert<2>{Show that your fractional extreme point has a $0$ or a $1$}
\item Drop the $0$ or $1$ variable and use minimality to find a contradiction in the smaller problem
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Proof Highlights - Finding a $0$ or $1$}
\alert{Rank Lemma:} ``For extreme point $x$, $0 < x < 1$, the maximum number of linearly independent constraints satisfied at equality equals the number of variables."\\
Will contradict with \alert{Token Argument:}
\begin{figure}
\centering
\begin{tikzpicture}
\node[]() at (0,5){\underline{Acceptable Pairs}};
\node[](a1) at (0,4){(Adam,Alice)};
\node<2>[]() at (1.2,4){$\CIRCLE$};
\node[](a2)at(0,3) {$\vdots$};
\node[](a3) at (0,2){(Dave,Daisy)};
\node<2>[]() at (1.2,2){$\CIRCLE$};

\node[]() at (3.5,5){\underline{Redistribute Tokens}};

\node[]() at (7,5){\underline{Tight Constraints}};
\node[](b1) at (7,4){$\sum_{w} x_{Bob,w} = 1$};
\node[](b2) at (7,3){$\vdots$};
\node[text width = 3cm](b3) at (7,2){\small Charles,Carly blocking pair constraint $=1$};

\path[->] (a1) edge (b3);
\path[->] (a2) edge  (b2);
\path[->] (a3) edge (b1);

\node<3>[]() at (3.5, 3.5){$\CIRCLE$$\CIRCLE$};

\node<4->[]() at (8.4, 4){$\CIRCLE$};
\node<4->[]() at (8.4, 2){$\CIRCLE$};

\node<5>[]() at (-1, 3.4){(Bob,Brooke)};
\node<5>[]() at (-1,2.9){$\LEFTCIRCLE$};
\end{tikzpicture}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Proof Highlights - Token Rules We Used}
Any given acceptable pair $(m,w)$ redistributes its token as follows:
\begin{enumerate}
\item \small If the constraint $$x_{m,w} + \sum_{\substack{w' \text{ preferred to $w$}\\ \text{by $m$}}} x_{m, w'} + \sum_{\substack{\text{$m'$ preferred to $m$}\\\text{ by $w$}}} x_{m',w} \geq 1$$
is tight, give it the token.
\item \small Otherwise
\begin{itemize}
\item If the constraint
$$\sum_{w \text{ acceptable to } m} x_{m,w} \leq 1 $$
is tight, give it $\frac{1}{2}$ a token.
\item \small If the constraint 
$$ \sum_{m \text{ acceptable to } w} x_{m,w} \leq 1$$
is tight, give it $\frac{1}{2}$ a token.
\end{itemize}
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Locality Proof Strategy}
We argue by contradiction:
\begin{enumerate}
\item Start with a minimal counterexample
\item \alert<1>{Show that your fractional extreme point has a $0$ or a $1$}
\item \alert<2>{Drop the $0$ or $1$ variable and use minimality to find a contradiction in the smaller problem}
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Proof Highlights - Dropping $0$ Variable}

\begin{overprint}
\onslide<1-3>
\alert{Case:} We have $(m,w)$ with $x_{m,w} = 0$ \\
LP for smaller problem:
\onslide<4->
\alert{Lemma:} Assigning all men (or women) favourite partner among two Stable Matchings gives a Stable Matching!
\end{overprint}
\begin{figure}[h]
\begin{tikzpicture}

\draw[fill=lightgray] (0,0) -- (0,4) -- (1.5,4)node[above, shape=circle, draw=black]{$M_1$} --node[right]{$\textbf{x} = \frac{1}{2} M_1 + \frac{1}{2} M_2$} (5,1.5)node[right, shape=circle, draw=black]{$M_2$} -- (5,0) -- (0,0);
\node<1>[]() at (7,5){\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ };
\node<2-4>[]() at (7,5){Minimality: $M_1$ and $M_2$ are Stable Matchings};

\node<3-4>[]() at (6.78,4.5){$M^*$: Give each man favourite from $M_1$, $M_2$};
\node<3-4>[]() at (7,4){$W^*$: Give each woman favourite from $M_1$, $M_2$};

\node<4->[shape=circle, draw=black](M) at (1.5,2){$M^*$};
\node<4->[shape=circle, draw=black](W) at (4,3.4){$W^*$};
\path<4->[dashed] (M) edge (W);

\node[]() at (3.1,2.9){$\CIRCLE$};

\node<5>[]() at (7,5){But $W^*$ should lie in polytope (shaded region)};
\node<5>[]() at (7,4.5){since it is incidence vector of a Stable Matching?!};
\end{tikzpicture}
\caption{Stable Matching Polytope}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{References}
\begin{enumerate}
\item Jochen Könemann, Kanstantsin Pashkovich, and Justin Toth. \alert{"The Power of Locality: An Elementary Integrality Proof of Rothblum's Stable Matching Formulation."} $\textit{arXiv preprint}$ arXiv:1605.04427 (2016).
\item Uriel G. Rothblum. \alert{"Characterization of stable matchings as extreme points of a polytope."} $\textit{Mathematical Programming}$ 54.1-3 (1992): 57-67.
\item Lap Chi Lau, Ramamoorthi Ravi, and Mohit Singh. \alert{Iterative methods in combinatorial optimization}. Vol. 46. $\textit{Cambridge University Press}$, 2011
\end{enumerate}
\end{frame}

\end{document}

### 2017-03-26 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
working on slides

@@
 We can extend this to the fractional matching case:
-$$FM_{G} = \{x \in \R^m_+ : x(\delta(u)) \leq 1, \forall u \in V(G)\}$$
-By defining for all $\alert{x, y \in FM_{G}}$:
+$$FM_{G} = \{x \in \R^m_+ : x(\delta(u)) \leq 1, \forall u \in V(G)\}.$$
+We extend the voting function by defining for all $\alert{x \in FM_G}$, for all $\alert{u,v' \in V(G)}$:
+$$vote_u(x,v) = \sum_{v\in V(G)} x_{uv} vote_u(v,v').$$ 
+We extend the vote aggregation defining for all $\alert{x, y \in FM_{G}}$:
 $$\Delta(x,y) = \sum_{u \in V(G)} \sum_{uv, uv' \in \delta(u)} x_{uv}y_{uv'} vote_u(v,v').$$
@@
 \frametitle{Mixed Popular Matching Example}
-In bipartite graphs, each $\alert{x \in FM_G}$ is a convex combination of matchings. But observe that each $x\in P_G$ need not be a convex combination of popular matchings.
+In bipartite graphs, each $\alert{x \in FM_G}$ is a convex combination of matchings. But observe that each $\alert{x\in P_G}$ need not be a convex combination of popular matchings.
 \begin{columns}[T] % align columns
@@
 \end{columns}
+Solid black edges $\alert{S}$ is the only popular matching. If we let $\alert{M}$ be the solid red edges then $$\Pi= \{(S,\frac{1}{2}), (M,\frac{1}{2})\}$$ is a fractional popular matching.
+\end{frame}
 
-Solid black edges $\alert{S}$ is the only popular matching. If we let $\alert{M}$ be the solid red edges then $$\Pi= \{(S,\frac{1}{2}), (M,\frac{1}{2})\}$$ is a fractional popular matching.
+\begin{frame}
+\frametitle{Extended Formulation}
+
 \end{frame}

### 2017-03-27 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
Writing up slides

@@
 \begin{document}
-
+\setbeamercolor{background canvas}{bg=white}
 \maketitle
@@
 \frametitle{Extended Formulation}
+Fix a popular fractional matching $\alert{y}$.
+Then $\alert{\Delta(x,y)}$ is a linear function in $\alert{x}$.
+Since $\alert{y}$ is popular $\alert{\Delta(x,y) \leq 0}$ with equality when $\alert{x}$ is popular.
+\begin{columns}[T]
+\begin{column}{.48\textwidth}
+\begin{align*}
+\max &\Delta(x,y) \\
+\text{s.t.} \sum_{e \in \delta(u)} x_e &= 1 &\forall u \in V(G) \\
+x_e &\geq 0 &\forall e \in E
+\end{align*}
+\end{column}
+\begin{column}{.48\textwidth}
+\begin{align*}
+\min &\sum_{u \in V(G)} \alpha_u  \\
+\text{s.t.} \alpha_a + \alpha_b &\geq vote_a(b,y) \\&+ vote_b(a,y) &\forall ab \in E(G) \\
+\alpha_u &\geq vote_u(\ell(u), y) &\forall u \in V(G).
+\end{align*}
+\end{column}
+\end{columns}
+Where $\alert{\ell(u)}$ is a last choice vertex for each $\alert{u}$ so that each vertex is fully matched in any fractional matching.
 
+We call $\alert{\alpha}$ such that $\alert{\sum_{u} \alpha_u = 0}$ a witness to the popularity of $\alert{y}$.
+\end{frame}
+
+\begin{frame}
+\frametitle{Extended Formulation}
+From the primal-dual pair we obtain the extended formulation $\alert{(LP)}$:
+\begin{align*}
+\min &\sum_{u \in V(G)} \alpha_u \\
+\text{s.t.} \alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E(G) \\
+\alpha_u &\geq vote_u(\ell(u),x) &\forall u \in V(G) \\
+\sum_{e \in \delta(u)} x_e &=1 &\forall u \in V(G) \\
+x_e &\geq 0 &\forall e \in E.
+\end{align*}
+Optimal value is $0$. The optimal solutions are $\text{\alert{popular matchings}}$ and their $\text{\alert{witnesses}}$ (see Popular Mixed Matchings by Kavitha et al. for details).
+\end{frame}
+
+\begin{frame}
+\frametitle{Extended Formulation}
+Let $\alert{P'_G}$ be the optimal solutions to $\alert{(LP)}$. Then $\alert{P'_G}$ is described by the set of $\alert{(x,\alpha)}$ such that:
+\begin{align*}
+\sum_{u \in V(G)} \alpha_u &= 0 \\
+\text{s.t.} \alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E(G) \\
+\alpha_u &\geq vote_u(\ell(u),x) &\forall u \in V(G) \\
+\sum_{e \in \delta(u)} x_e &=1 &\forall u \in V(G) \\
+x_e &\geq 0 &\forall e \in E.
+\end{align*}
+\end{frame}
+
+\begin{frame}
+\frametitle{Self-Duality}
+It is straightforward to see that $\alert{(LP)}$ is self-dual (exercise).
+This implies, by $\text{\alert{complementary slackness}}$, that the covering constraint for $\alert{ab}$ is tight when $\alert{x_{ab} > 0}$ for any popular fractional matching $\alert{x}$.
+
+$\textbf{Lemma 1}$: For any $\alert{(x,\alpha) \in P'_G}$, if $\alert{x_{ab} > 0}$ then:
+$$\alpha_a + \alpha_b = vote_a(b,x) + vote_b(a,x).$$ 
 \end{frame}
 \section{Integrality of $P_G$ in a special case}
+\begin{frame}
+\frametitle{Objective}
+$\textbf{Theorem 1:}$ If $\alert{G =(A\cup B, E)}$ is $\text{\alert{bipartite}}$ and admits a $\text{\alert{perfect stable matching}}$ then $\alert{P_G}$ is integral.
 
+From prior talks: $\text{\alert{stable matchings}}$ are $\text{\alert{minimum size popular matchings}}$.
+
+So the hypothesis says $\text{\alert{popular matching}}$ is perfect and, in the fractional setting, for each $\alert{u\in A \cup B}$, $\alert{x_{u\ell(u)} = 0}$.
+
+To prove $\textbf{Theorem 1}$ we will write $\alert{(x,\alpha^x) \in P_G'}$ as a convex combination of $\text{\alert{popular matchings}}$.
+\end{frame}
+
+\begin{frame}
+\frametitle{Bounding $\alpha^x$}
+$\textbf{Lemma 2:}$ $\alert{\alpha^x \in [-1,1]}$.
+
+$\textbf{Proof:}$ First
+$$\alpha^x_u \geq vote_u(\ell(u), x) = -1 \text{(since }x_{u,\ell(u)} = 0\text{).}$$
+Now let $\alert{v}$ be the least preferred neighbour of $\alert{u}$ such that $\alert{x_{uv} >0}$. Then
+$$vote_u(v,x) = -vote_u(x,v) = -(1-x_{uv}).$$
+Also
+$$vote_v(u,x) = \sum_{u':u' <_v u} x_{u'v} - \sum_{u':u' >_v u} x_{u'v} = 1-x_{uv}.$$
+So by $\textbf{Lemma 1}$
+$$\alpha_u + \alpha_v = vote_u(v,x) + vote_v(u,x) \leq -(1-x_{uv}) + (1-x_{uv}) = 0.$$
+Since $\alert{\alpha_v \geq -1}$ this implies $\alert{\alpha_u \leq 1}$.$\blacksquare$
+\end{frame}
+
+\begin{frame}
+\frametitle{Constructing the Table: $a \in A$}
+\centering
+$r_a\cdot 1 + (1-r_a)(-1) = \alpha^x_a$ 
+\includegraphics[scale=0.25]{Xa}
+
+For each $\alert{ab \in \delta(a)}$ create a cell of length $\alert{x_{ab}}$ (omit $0$-length cells) and arrange cells in increasing order (w.r.t. $\alert{a}$) to form $\alert{X_a}$.
+\end{frame}
+
+\begin{frame}
+\frametitle{Constructing the Table: $b \in B$}
+\centering
+$r_b(-1) + (1-r_b)\cdot 1 = \alpha^x_b$
+\includegraphics[scale=0.25]{Xb}
+
+For each $\alert{ab \in \delta(b)}$ create a cell of length $\alert{x_{ab}}$ (omit $0$-length cells) and arrange cells in decreasing order (w.r.t. $\alert{b}$) to form $\alert{X_b}$.
+\end{frame}
+
+\begin{frame}
+\frametitle{Constructing the Table: Finding Matchings}
+We form the table $\alert{T}$ whose rows are the arrays $\alert{X'_u}$ for each $\alert{u \in V(G)}$.
+
+For any $\alert{t \in [0,1)}$ form $\alert{M_t\subseteq E}$ as follows:
+\begin{enumerate}
+\item Draw a vertical line at distance $\alert{t}$ from left end of $\alert{T}$.
+\item For each $\alert{u \in V(G)}$ define $\alert{c_u(t)}$ to be the cell in $\alert{X'_u}$ which the line intersects (or touches the left boundary of).
+\item Set $\alert{M_t = \{ uv \in E(G) : u \in V(G) \text{and} v \in c_u(t)\}}$.
+\end{enumerate}
+
+We claim that each $\alert{M_t}$ is a $\text{\alert{popular matching}}$.
+\end{frame}
+
+\begin{frame}
+\frametitle{Brief detour to pick up a Lemma}
+$\textbf{Lemma 3}$ For any $\alert{ab \in E(G)}$,
+$$ x_{ab} + \sum_{b' <_a b} x_{ab'} + \sum_{a' <_b a} x_{a'b} \leq r_a + (1-r_b).$$
+Equality holds if $\alert{x_{ab} > 0}$.
+
+$\textbf{Proof:}$
+From the covering constraint for $\alert{P'_G}$:
+$$\alpha^x_a + \alpha^x_b \geq \sum_{b' <_a b} x_{ab'} - \sum_{b' >_a b} x_{ab'} + \sum_{a' <_b a} x_{a'b} - \sum_{a' >_b a} x_{a'b}.$$
+Substitute $\alert{\sum_{b' >_a b} x_{ab'} = 1- x_{ab} - \sum_{b'<_a b} x_{ab'}}$, and similar for $\alert{\sum_{a' >_b a}x_{a'b}}$.
+Substitute $\alert{\alpha^x_a = 2r_a -1}$ and $\alert{\alpha^x_b = 1-2r_b}$.
+
+The result follows. $\textbf{Lemma 1}$ yields equality when $\alert{x_{ab}} > 0$.$\blacksquare$
+\end{frame}
+
+\begin{frame}
+\frametitle{$M_t$ is a matching}
+%May take multiple frames
+\end{frame}
+
+\begin{frame}
+\frametitle{$M_t$ is popular}
+%May take multiple frames
+\end{frame}
 \section{Half-Integrality of $P_G$ in any bipartite instance}

### 2017-03-27 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
Writing slides

@@
 \frametitle{$M_t$ is a matching}
-%May take multiple frames
+$\textbf{Strategy:}$
+\begin{enumerate}
+\item Consider some edge $\alert{ab}$ such that $\alert{x_{ab} > 0}$.
+\item Want to show that $\alert{b \in c_t(a) \iff a \in c_t(b)}$ for any $\alert{t \in [0,1)}$.
+\item Will show that the $\alert{ab}$ cell in rows $\alert{X'_a}$ and $\alert{X'_b}$ is perfectly aligned in table $\alert{T}$.
+\end{enumerate}
+
+$\textbf{Recall:}$
+\begin{itemize}
+\item In $\alert{X'_a}$: $\alert{a}$'s increasing order start from $\textit{left}$ of blue subarray (wraps around).
+\item In $\alert{X'_b}$: $\alert{b}$'s increasing order starts from $\textit{right}$ of blue subarray (wraps around).
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{$M_t$ is a matching}
+$\textbf{Case 1: }$ $\alert{\sum_{b' <_a b} x_{ab'} \geq r_a}$
+\begin{columns}[T] % align columns
+\begin{column}{.48\textwidth}
+\includegraphics[scale=0.23]{geqRa}
+\end{column}
+\begin{column}{.48\textwidth}
+$$ $$
+$\quad\quad d:= \sum_{b' <_a b} x_{ab'} -r_a$
+
+$\quad\quad d':=\sum_{a'<_b a} x_{a'b}$
+
+$\quad\quad$By $\textbf{Lemma 3}$:
+$$d + x_{ab} + d' = 1-r_b.$$
+\end{column}
+\end{columns}
+The $\textbf{case}$ where $\alert{\sum_{a'<_b a} x_{a'b} \geq (1-r_b)}$ is symmetric.
+\end{frame}
+
+\begin{frame}
+\frametitle{$M_t$ is a matching}
+$\textbf{Case 2: }$ $\alert{\sum_{b' <_a b} x_{ab'} < r_a}$ and $\alert{\sum_{a' <_b a} x_{a'b} < 1 - r_b}$.
+
+In this case, the $\alert{ab}$ cell has been split in $\alert{X'_a}$ and $\alert{X'_b}$.
+\begin{columns}[T] % align columns
+\begin{column}{.48\textwidth}
+\includegraphics[scale=0.23]{lessRa}
+\end{column}
+\begin{column}{.48\textwidth}
+$$ $$
+$\quad\quad d_0:= \sum_{b' <_a b} x_{ab'} $
+
+$\quad\quad d_1:=\sum_{a'<_b a} x_{a'b}$
+
+$\quad\quad x^0_{ab} := \text{hatched blue of }X'_a.$
+
+$\quad\quad x^1_{ab} := \text{hatched blue of }X'_b.$
+
+$\quad\quad$By $\textbf{Lemma 3}$:
+$$d_0 + x_{ab} + d_1 = r_a + 1-r_b.$$
+\end{column}
+\end{columns}
+\centering
+Hence $\alert{x_{ab} + r_a - x^0_{ab}  + (1-r_b) - x^1_{ab} = r_a + (1-r_b)}$, and so $\alert{x_{ab} = x^0_{ab} + x^1_{ab}}$.
+\end{frame}
+\begin{frame}
+\frametitle{$M_t$ is popular}
+$\textbf{Strategy:}$
+\begin{enumerate}
+\item Exhibit a witness vector $\alert{\alpha^t}$ which satisfies $\alert{P'_G}$.
+\item For each $\alert{u \in A\cup B}$ if vertical line intersects $\alert{c_t(u)}$ in blue (positive) subarray set $\alert{\alpha^t_u = 1}$
+\item Otherwise, in red (negative) subarray, set $\alert{\alpha^t_u = -1}$.
+\end{enumerate}
+From previous proof, for any $\alert{ab \in M_t}$:
+
+$\quad\alert{c_t(a)}$ is positive if and only if $\alert{c_t(b)}$ is negative.
+
+So then $\alert{\sum_{u \in A\cup B} \alpha^t_u = 0}$. Here we used that $\alert{M_t}$ is perfect. Remains to verify covering constraints.
 \end{frame}
@@
 \frametitle{$M_t$ is popular}
-%May take multiple frames
+Want to show $\alert{\forall ab \in E(G)}$, $\alert{\alpha^t_a + \alpha^t_b \geq vote_a(b,M_t(a)) + vote_b(a, M_t(b))}.$
+We do a case analysis on the ``colours" of $\alert{c_t(a)}$ and $\alert{c_t(b)}$:
+\begin{columns}[T] % align columns
+\begin{column}{.48\textwidth}
+\begin{enumerate}
+\item If exactly one of $\alert{c_t(a)}$ or $\alert{c_t(b)}$ is positive then either $\alert{ab \in M_t}$ or at least one of $\alert{a,b}$ is matched to someone better.
+\item If both $\alert{c_t(a)}$ and $\alert{c_t(b)}$ are negative then both $\alert{a}$ and $\alert{b}$ get matched to someone better.
+\item If both $\alert{c_t(a)}$ and $\alert{c_t(b)}$ are positive their coverage is trivial (RHS $\leq 2$).
+\end{enumerate}
+\end{column}
+\begin{column}{.48\textwidth}
+\includegraphics[scale=0.2]{covering}
+\end{column}
+\end{columns}
 \end{frame}
-\section{Half-Integrality of $P_G$ in any bipartite instance}
-%maybe omit this as next section generalizes and uses same ideas
-\section{Half-Integrality of $P_G$ in a roommates instance}
+
+\begin{frame}
+\frametitle{Sweeping Procedure}
+We write $\alert{x}$ as a convex combination of $\text{\alert{popular matchings}}$:
+\begin{enumerate}
+\item Initialize $\alert{i=0}$.
+\item Sweep a vertical line from left to right across $\alert{T}$. Denote its distance by $\alert{t}$. 
+\item Whenever a new cell is encountered fix matching $\alert{M^i = M_t}$ and fix $\alert{t_i = t}$. 
+\end{enumerate}
+Say we find $\alert{k}$ such matchings. We can then write $\alert{x}$ as:
+$$x = t_1\chi(M^0) + (t_2 - t_1)\chi(M^1) + \dots + (1 - t_{k-1}) \chi(M^{k-1}).$$
+This constructive procedure runs in $\alert{O(|E| + |V|)}$ time.
+\end{frame}
+\section{Half-Integrality of $P_G$ in a general instance}
 

### 2017-03-27 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
draft of slides

@@
 
-The max utility popular matching problem $\alert{\text{(MaxPM)}}$ asks to find a popular matching $\alert{M}$ on $\alert{G}$ which maximizes
+The max utility popular matching problem $\text{\alert{(MaxPM)}}$ asks to find a popular matching $\alert{M}$ on $\alert{G}$ which maximizes
 $$\sum_{e\in M} w(e)$$
@@
 
-\section{Hardness and Inapproximability}
+\begin{frame}
+\frametitle{Constructing Bipartite Instance}
+\begin{center}
+From a graph $\alert{G}$ we can construct its bipartite graph $\alert{H}$:
+\begin{columns}[T]
+\begin{column}{.48\textwidth}
+\centering
+Preferences corresponding to $\textbf{a:} \alert{b} > \alert{c}$ in $\alert{H}$:
+$$\textbf{a:} \alert{b'} > \alert{c'} > \alert{a'}$$
+$$\textbf{a':} \alert{b}>\alert{c} > \alert{a}.$$
+\end{column}
+\begin{column}{.48\textwidth}
+\includegraphics[scale=0.125]{bipartite}
+\end{column}
+\end{columns}
+\end{center}
+$\textbf{Claim:}$ $\alert{H}$ admits a perfect stable matching.
 
-\section{Open Problems}
+$\textbf{Proof:}$ Say stable matching $\alert{S}$ exposes $\alert{u}$.
 
-\section{Introduction to Stable Matching}
-\begin{frame}
-\frametitle{Stable Matching Problem}
-\begin{overprint}
-\onslide<1-3>
-Introduced by Gale and Shapley in 1962
-\onslide<4-9>
-How to solve? Gale Shapley Algorithm:
-\onslide<10->
-Not very fair to women. Can we do better?
-\end{overprint}
-\begin{figure}
-\centering
-\begin{tikzpicture}
-   \node[] (m1) at (0,4) {Adam};
-    \node[] (m2) at (0,3) {Bob};
-    \node[] (m3) at (0,2) {Chad};
-    \node[] (m4) at (0,1) {Dave};
+Then $\alert{S' = \{e' : e\in S\} \cup \{e : e' \in S\}}$ is a stable matching that exposes $\alert{u'}$.
 
-    \node[] (w1) at (2,4) {Alice};
-    \node[] (w2) at (2,3) {Brooke};
-    \node[] (w3) at (2,2) {Carly};
-    \node[] (w4) at (2,1) {Daisy};
+Since stable matchings covers the same vertices, $\alert{S}$ does not cover $\alert{u'}$.
 
-     \node[] () at (-2,4){$\alert<10>{A}>B>C>D\quad:$};
-    \node[] () at (-2,3){$B>\alert<10>{C}>A>D\quad:$};
-    \node[] () at (-2,2){$A>\alert<10>{B}>C\quad:$};
-    \node[] () at (-2,1){$\alert<10>{D}>C>B>A\quad:$};
-
-    \node[] () at (4,4){$:\quad \alert<10>{A}>B>C$};
-    \node[] () at (4,3){$:\quad D>A>\alert<10>{C}>B$};
-    \node[] () at (4,2){$:\quad D>C>\alert<10>{B}>A$};
-    \node[] () at (4,1){$:\quad A>B>C>\alert<10>{D}$};
-
-    \path[-]<2-3> (m1) edge (w1);
-    \path[-]<2-3> (m2) edge (w2);
-    \path[-]<2-3> (m3) edge (w3);
-    \path[-]<2-3> (m4) edge (w4);
-
-    \path[-]<4-> (m1) edge (w1);
-    \path[-]<5-7> (m2) edge (w2);
-    \path[-]<6> (m3) edge (w1);
-    \path[-]<7-> (m3) edge (w2);
-    \path[-]<8-> (m2) edge (w3);
-    \path[-]<9-> (m4) edge (w4);
-
-    \path[-,decoration={zigzag,segment length=4,amplitude=.9,
-  post=lineto,post length=2pt},color=red]<3> (m3) edge[decorate] (w2);
-
-\end{tikzpicture}
-\end{figure}
+Thus $\alert{uu'}$ blocks $\alert{S}$. $\blacksquare$
 \end{frame}
@@
 \begin{frame}
-\frametitle{Weighted Stable Matching}
-Add weights to pairs to incentivize equitable pairings.\\
-Example:
-$$c_{\text{Adam,Alice}} = 8$$ since Adam and Alice really like each other,
-$$c_{\text{Adam, Carly}} = 2$$ since Adam and Carly don't like each other much.\\
-New Goal:
-$$\max_{\{\text{Stable Matchings $M$}\}} \sum_{\text{$(m,w)$ Pairs in $M$}} c_{m,w}$$
+\frametitle{$\frac{1}{2}$-integrality of $P_G$}
+By $\textbf{Theorem 1}$, $\alert{P_H}$ is integral.
+
+Let $\alert{(x, \alpha) \in P'_G}$. Define $\alert{f: P_G \rightarrow P_H}$ by
+$$f(x) = z \text{ where } z_{uv'}=z_{u'v} =x_{uv} \text{ and } z_{uu'} = x_{u\ell(u)}.$$
+With vector $\alert{\beta}$ such that $\alert{\beta_u = \beta_{u'} = \alpha_u}$ it is easy to check $\alert{Im(f) \subseteq P_H}$.
+Define $\alert{h: P_H \rightarrow P_G}$ by 
+$$h(z) = y \text{ where } y_{uv} = (z_{uv'} + z_{u'v})/2 \text{ and } y_{u\ell(u)} = z_{uu'}.$$
+With witness $\alert{\alpha_u = (\beta_u + \beta_u')/2}$ it is easy to check $\alert{Im(h) \subseteq P_G}$.
+$$x = h(f(x)) = \sum_{i=0}^{k-1} h(\lambda_i \chi(M^i)) = \sum_{i=0}^{k-1} \lambda_i h(\chi(M^i)).$$
+Each $\alert{h(\chi(M^i))}$ is $\frac{1}{2}$-integral. $\blacksquare$
 \end{frame}
-
-\section{Introduction to Linear Programs}
+\section{Conclusions}
 
 \begin{frame}
-\frametitle{What is a Linear Program?}
-Let $c, a^1, \dots, a^m \in \mathbb{R}^n$. Let $b \in \mathbb{R}^m$. Find $x \in \mathbb{R}^n$ which satisfies:
-\begin{align*}
-\max \sum_{i=1}^n c_i x_i&\  &\text{\alert{The Objective Function}}\\
-\text{subject to } \sum_{i=1}^n a_i^1 x_i &\geq b_1\\
-&\vdots &\text{\alert{The Feasible Region}}\\
-\sum_{i=1}^n a_i^m x_i &\geq b_m \\
-x &\geq 0
-\end{align*}
-\end{frame}
+\frametitle{Optimization Problems and Complexity}
+A $\text{\alert{max utility popular matching}}$ can be found in poly-time on $\text{\alert{bipartite graphs}}$ which admit a $\text{\alert{perfect stable matching}}$
 
-\begin{frame}
-\frametitle{Polytopes - Geometry of the Feasible Region}
-\begin{center}
-\begin{tikzpicture}
-\draw[->] (0,0) node[left]{$0$} -- (0,5)node[above]{$x_2$};
-\draw[->] (0,0) -- (6,0) node[right]{$x_1$};
+Also a $\alert{\frac{1}{2}}$$\text{\alert{-integral fractional popular matching}}$ of max utility can be found on any graph.
 
-\path[dashed]<1-> (-1,0) edge node[left]{$\langle a^1 x \rangle \geq b_1$} (3.2,5.6);
-\path[dashed]<2-> (-1.5,5.5) edge node[right]{$\langle a^2x\rangle \geq b_2$} (7.5,0.5);
-\path[dashed]<3-> (2.5,-1) edge node[right]{$\langle a^3 x \rangle \geq b_3$} (7,2);
-
-\path[fill=lightgray]<4-> (0,0) -- (0,1.3) -- (1.75,3.65) -- (5.95,1.35) -- (4,0) -- (0,0);
-
-\path[->] <5-> (2,1) edge node[right]{$c$} (2.25,2);
-\draw<6>[] (1.75,3.65) circle (2mm) node[right]{\ \ Extreme Point Solution} ;
-\end{tikzpicture}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Why Are Extreme Points Interesting?}
-\begin{itemize}
-\item We have very efficient algorithms for solving Linear Programs (\alert{Simplex}, \alert{Ellipsoid})
-\item These algorithms always give \alert{extreme point} solutions
-\item It can be shown that when we can find \alert{optimal solutions} we can always find them at \alert{extreme points}
-\end{itemize}
-\end{frame}
-\section{Formulating Stable Matching as a Linear Program}
-
-\begin{frame}
-\frametitle{Incidence Vectors}
-We can translate between Matchings and Vectors using \alert{Incidence Vectors} in $\mathbb{R}^E$ indexed by set of all acceptable pairs $E$
-\begin{align*}
-&\underline{Matching} &\underline{Real\ Vector} \\
-&\text{(Adam,Alice)} &x_\text{Adam,Alice} = 1 \\
-&\text{(Bob,Carly)} &x_\text{Bob,Carly} = 1\\
-&\text{(Chad, Brooke)} &x_\text{Chad, Brooke} = 1\\
-&\text{(Dave, Daisy)} &x_\text{Dave, Daisy} = 1 \\
-\end{align*}
-$x_{m,w} = 0$ for all other pairs $(m,w)$ not in Matching
-\end{frame}
-
-\begin{frame}
-\frametitle{Rothblum's Formulation}
-\begin{align*}
-\onslide<1,2->{\max \sum_{\text{Acceptable Pairs } (m,w)} c_{m,w} x_{m,w}}\\
-\onslide<1,3->{\text{subject to } \sum_{w \text{ acceptable to } m} x_{m,w} &\leq 1 &\text{For all Men $m$}}\\
-\onslide<1,4->{\sum_{m\text{ acceptable to }w} x_{m,w} &\leq 1 &\text{For all Women $w$}} \\
-\onslide<1,5->{x_{m,w} + \sum_{\substack{w' \text{ preferred to $w$}\\ \text{by $m$}}} x_{m, w'} + \sum_{\substack{\text{$m'$ preferred to $m$}\\\text{ by $w$}}} x_{m',w} &\geq 1 &\text{For all pairs $(m,w)$}\\}
-\onslide<1,6->{1 \geq x &\geq 0}
-\end{align*}
-\end{frame}
-
-\begin{frame}
-\frametitle{Are Extreme Points Incidence Vectors?}
-\begin{itemize}
-\item We know $0$-$1$ solutions to our Linear Program are Stable Matchings
-\item But what if our LP Solver gives an extreme point solution with \alert{$X_{m,w} = \frac{1}{2}$}?
-\item We need to show that this cannot happen
-\end{itemize}
-\end{frame}
-
-\section{Locality Technique}
-\begin{frame}
-\frametitle{Locality Proof Strategy}
-We argue by contradiction:
+By reducing from $\text{\alert{Vertex Cover}}$ they show:
 \begin{enumerate}
-\item Start with a minimal counterexample
-\item \alert<2>{Show that your fractional extreme point has a $0$ or a $1$}
-\item Drop the $0$ or $1$ variable and use minimality to find a contradiction in the smaller problem
+\item $\text{\alert{(MaxPM)}}$ is $NP$-hard, even when preferences are $\text{\alert{strict}}$, $\text{\alert{complete}}$, and $\alert{w: E \rightarrow \{1,2\}}$, and $\alert{G}$ admits a $\text{\alert{stable matching}}$.
+\item Under UGC there is no $O(1)$-approximation algorithm for $\text{\alert{(MaxPM)}}$.
 \end{enumerate}
@@
 \begin{frame}
-\frametitle{Proof Highlights - Finding a $0$ or $1$}
-\alert{Rank Lemma:} ``For extreme point $x$, $0 < x < 1$, the maximum number of linearly independent constraints satisfied at equality equals the number of variables."\\
-Will contradict with \alert{Token Argument:}
-\begin{figure}
-\centering
-\begin{tikzpicture}
-\node[]() at (0,5){\underline{Acceptable Pairs}};
-\node[](a1) at (0,4){(Adam,Alice)};
-\node<2>[]() at (1.2,4){$\CIRCLE$};
-\node[](a2)at(0,3) {$\vdots$};
-\node[](a3) at (0,2){(Dave,Daisy)};
-\node<2>[]() at (1.2,2){$\CIRCLE$};
+\frametitle{Open Questions}
 
-\node[]() at (3.5,5){\underline{Redistribute Tokens}};
+Can we settle the complexity of $\text{\alert{(MaxPM)}}$ on $\text{\alert{bipartite}}$ graphs?
 
-\node[]() at (7,5){\underline{Tight Constraints}};
-\node[](b1) at (7,4){$\sum_{w} x_{Bob,w} = 1$};
-\node[](b2) at (7,3){$\vdots$};
-\node[text width = 3cm](b3) at (7,2){\small Charles,Carly blocking pair constraint $=1$};
+$$ $$
 
-\path[->] (a1) edge (b3);
-\path[->] (a2) edge  (b2);
-\path[->] (a3) edge (b1);
-
-\node<3>[]() at (3.5, 3.5){$\CIRCLE$$\CIRCLE$};
-
-\node<4->[]() at (8.4, 4){$\CIRCLE$};
-\node<4->[]() at (8.4, 2){$\CIRCLE$};
-
-\node<5>[]() at (-1, 3.4){(Bob,Brooke)};
-\node<5>[]() at (-1,2.9){$\LEFTCIRCLE$};
-\end{tikzpicture}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Proof Highlights - Token Rules We Used}
-Any given acceptable pair $(m,w)$ redistributes its token as follows:
-\begin{enumerate}
-\item \small If the constraint $$x_{m,w} + \sum_{\substack{w' \text{ preferred to $w$}\\ \text{by $m$}}} x_{m, w'} + \sum_{\substack{\text{$m'$ preferred to $m$}\\\text{ by $w$}}} x_{m',w} \geq 1$$
-is tight, give it the token.
-\item \small Otherwise
-\begin{itemize}
-\item If the constraint
-$$\sum_{w \text{ acceptable to } m} x_{m,w} \leq 1 $$
-is tight, give it $\frac{1}{2}$ a token.
-\item \small If the constraint 
-$$ \sum_{m \text{ acceptable to } w} x_{m,w} \leq 1$$
-is tight, give it $\frac{1}{2}$ a token.
-\end{itemize}
-\end{enumerate}
-\end{frame}
-
-\begin{frame}
-\frametitle{Locality Proof Strategy}
-We argue by contradiction:
-\begin{enumerate}
-\item Start with a minimal counterexample
-\item \alert<1>{Show that your fractional extreme point has a $0$ or a $1$}
-\item \alert<2>{Drop the $0$ or $1$ variable and use minimality to find a contradiction in the smaller problem}
-\end{enumerate}
-\end{frame}
-
-\begin{frame}
-\frametitle{Proof Highlights - Dropping $0$ Variable}
-
-\begin{overprint}
-\onslide<1-3>
-\alert{Case:} We have $(m,w)$ with $x_{m,w} = 0$ \\
-LP for smaller problem:
-\onslide<4->
-\alert{Lemma:} Assigning all men (or women) favourite partner among two Stable Matchings gives a Stable Matching!
-\end{overprint}
-\begin{figure}[h]
-\begin{tikzpicture}
-
-\draw[fill=lightgray] (0,0) -- (0,4) -- (1.5,4)node[above, shape=circle, draw=black]{$M_1$} --node[right]{$\textbf{x} = \frac{1}{2} M_1 + \frac{1}{2} M_2$} (5,1.5)node[right, shape=circle, draw=black]{$M_2$} -- (5,0) -- (0,0);
-\node<1>[]() at (7,5){\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ };
-\node<2-4>[]() at (7,5){Minimality: $M_1$ and $M_2$ are Stable Matchings};
-
-\node<3-4>[]() at (6.78,4.5){$M^*$: Give each man favourite from $M_1$, $M_2$};
-\node<3-4>[]() at (7,4){$W^*$: Give each woman favourite from $M_1$, $M_2$};
-
-\node<4->[shape=circle, draw=black](M) at (1.5,2){$M^*$};
-\node<4->[shape=circle, draw=black](W) at (4,3.4){$W^*$};
-\path<4->[dashed] (M) edge (W);
-
-\node[]() at (3.1,2.9){$\CIRCLE$};
-
-\node<5>[]() at (7,5){But $W^*$ should lie in polytope (shaded region)};
-\node<5>[]() at (7,4.5){since it is incidence vector of a Stable Matching?!};
-\end{tikzpicture}
-\caption{Stable Matching Polytope}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{References}
-\begin{enumerate}
-\item Jochen Könemann, Kanstantsin Pashkovich, and Justin Toth.
-\alert{"The Power of Locality: An Elementary Integrality Proof of Rothblum's Stable Matching Formulation."} $\textit{arXiv preprint}$ arXiv:1605.04427 (2016).
-\item Uriel G. Rothblum. \alert{"Characterization of stable matchings as extreme points of a polytope."} $\textit{Mathematical Programming}$ 54.1-3 (1992): 57-67.
-\item Lap Chi Lau, Ramamoorthi Ravi, and Mohit Singh. \alert{Iterative methods in combinatorial optimization}. Vol. 46. $\textit{Cambridge University Press}$, 2011
-\end{enumerate}
+Is there an algorithm for deciding if a general graph admits a $\text{\alert{popular matching}}$ under $\text{\alert{strict}}$ preferences?
 \end{frame}

### 2017-03-27 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
proofread slides

@@
 
-The max utility popular matching problem $\text{\alert{(MaxPM)}}$ asks to find a popular matching $\alert{M}$ on $\alert{G}$ which maximizes
+The max utility popular matching problem $\text{\alert{(MaxPM)}}$ asks to find a $\text{\alert{popular}}$ matching $\alert{M}$ on $\alert{G}$ which maximizes
 $$\sum_{e\in M} w(e)$$
@@
 \frametitle{Fractional Matchings}
-We can extend this to the fractional matching case:
+We can extend this to the $\text{\alert{fractional matching}}$ case:
 $$FM_{G} = \{x \in \R^m_+ : x(\delta(u)) \leq 1, \forall u \in V(G)\}.$$
@@
 $$\Delta(x,y) = \sum_{u \in V(G)} \sum_{uv, uv' \in \delta(u)} x_{uv}y_{uv'} vote_u(v,v').$$
-The polytope of all fractional popular matchings is 
+The polytope of all $\text{\alert{fractional popular matchings}}$ is 
 $$P_G := \{x \in FM_{G}: \Delta(x,y) \geq 0, \forall y \in FM_G\}.$$
@@
 \frametitle{Mixed Popular Matching Example}
-In bipartite graphs, each $\alert{x \in FM_G}$ is a convex combination of matchings. But observe that each $\alert{x\in P_G}$ need not be a convex combination of popular matchings.
+In $\text{\alert{bipartite}}$ graphs, each $\alert{x \in FM_G}$ is a convex combination of matchings.
+But observe that each $\alert{x\in P_G}$ need not be a convex combination of $\text{\alert{popular}}$ matchings.
 \begin{columns}[T] % align columns
@@
 \end{columns}
-Solid black edges $\alert{S}$ is the only popular matching. If we let $\alert{M}$ be the solid red edges then $$\Pi= \{(S,\frac{1}{2}), (M,\frac{1}{2})\}$$ is a fractional popular matching.
+Solid black edges $\alert{S}$ is the only $\text{\alert{popular}}$ matching.
+If we let $\alert{M}$ be the solid red edges then $$\Pi= \{(S,\frac{1}{2}), (M,\frac{1}{2})\}$$ is a $\text{\alert{fractional popular matching}}$.
 \end{frame}
@@
 \sum_{u \in V(G)} \alpha_u &= 0 \\
-\text{s.t.} \alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E(G) \\
+ \alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E(G) \\
 \alpha_u &\geq vote_u(\ell(u),x) &\forall u \in V(G) \\
@@
 
-So the hypothesis says $\text{\alert{popular matching}}$ is perfect and, in the fractional setting, for each $\alert{u\in A \cup B}$, $\alert{x_{u\ell(u)} = 0}$.
+So the hypothesis says each $\text{\alert{popular matching}}$ is perfect and, in the fractional setting, for each $\alert{u\in A \cup B}$, $\alert{x_{u\ell(u)} = 0}$.
 
@@
 \item For each $\alert{u \in V(G)}$ define $\alert{c_u(t)}$ to be the cell in $\alert{X'_u}$ which the line intersects (or touches the left boundary of).
-\item Set $\alert{M_t = \{ uv \in E(G) : u \in V(G) \text{and} v \in c_u(t)\}}$.
+\item Set $\alert{M_t = \{ uv \in E(G) : u \in V(G) \text{ and } v \in c_u(t)\}}$.
 \end{enumerate}
@@
 $$f(x) = z \text{ where } z_{uv'}=z_{u'v} =x_{uv} \text{ and } z_{uu'} = x_{u\ell(u)}.$$
-With vector $\alert{\beta}$ such that $\alert{\beta_u = \beta_{u'} = \alpha_u}$ it is easy to check $\alert{Im(f) \subseteq P_H}$.
+Verified by witness $\alert{\beta}$ such that $\alert{\beta_u = \beta_{u'} = \alpha_u}$.
 Define $\alert{h: P_H \rightarrow P_G}$ by 
 $$h(z) = y \text{ where } y_{uv} = (z_{uv'} + z_{u'v})/2 \text{ and } y_{u\ell(u)} = z_{uu'}.$$
-With witness $\alert{\alpha_u = (\beta_u + \beta_u')/2}$ it is easy to check $\alert{Im(h) \subseteq P_G}$.
+Verified by witness $\alert{\alpha_u = (\beta_u + \beta_u')/2}$.
 $$x = h(f(x)) = \sum_{i=0}^{k-1} h(\lambda_i \chi(M^i)) = \sum_{i=0}^{k-1} \lambda_i h(\chi(M^i)).$$

### 2017-03-28 CS798-ConvexityAndOptimization/assignments/assignment_02.tex
writeup q3

--- 2017-03-14 CO750-ApproximationAlgorithmsTopics/assignments/assignment_02.tex
@@
 
-\lhead{2017-03-14}
-\rhead{W. Justin Toth CO750-Approximation Algorithms Assignment 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-03-27}
+\rhead{William Justin Toth CS798-Convexity and Optimization Assignment 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \section{}
+
+%Question 2
+\section{}
+
+%Question 3
+\section{}
 \paragraph{}
-Let $G = (V,E)$ be an undirected graph with edge cost $c: E\rightarrow \R_+$. Consider the problem $(IP)$:
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
-x&\geq 0\\
-x&\in \Z^{|E|},
-\end{align*}
-where $f : S \rightarrow \{0,1\}$ satisfies the maximality property:
-$$\text{ for all disjoint } A,B\subseteq V, \text{ we have } f(A \cup B ) \leq \max\{f(A), f(B)\},$$
-and $f$ satisfies that there does not exist $S \subseteq V$ and $A, B \subseteq S$ such that
-\begin{equation}\label{eq:1}
-A\cap B = \emptyset,\ f(S)=f(A)=f(B)=0 \text{ and } f(S\backslash A) = f(S\backslash B) = 1.
-\end{equation}
-We may assume that $f(V') = 0$, for the every connected component vertex set $V'$, otherwise $(IP)$ is infeasible.
-Let $(LP)$ denote the linear programming relaxation of $(IP)$ and let $(D)$ denote the dual of $(LP)$:
-\begin{align*}
-\max\ \sum_{S\subseteq V} f(S) y_S&\\
-\text{s.t.}\ \sum_{S : e \in \delta(S)} y_S &\leq c(e) &\forall e \in E \\
-y&\geq 0. 
-\end{align*}
-\paragraph{Algorithm}
-Consider the primal-dual algorithm, $\cA$, for this problem:
+I collaborated with Sharat on this problem.
+\paragraph{}
+Let $\cA$ denote our black-box algorithm. Let $f$ be at $\beta$-smooth function. Given an initial point $x_0$ and desired accuracy $\epsilon > 0$ our algorithm does:
 \begin{enumerate}
-\item Start with $x=0$, $y=0$, and $F \rightarrow \emptyset$.
-\item While $F$ is not feasible do:
-	\begin{enumerate}
-	\item Increase $y_S$ uniformly for ``minimal violated cuts" with respect to $F$ until some dual constraint corresponding to an edge $e$ becomes tight.
-	\item Set $x_e = 1$, $F = F\cup \{e\}$.
+\item If $||x_0 - x^*||_2^2 = 0$ return $x_0$.
+\item Let $\epsilon' = \epsilon/2$ and set $\alpha = \frac{\epsilon}{||x_0 - x^*||_2^2}$.
+\item Let $g$ be the function given by $g(x) := f(x) + \frac{\alpha}{2}||x - x_0||_2^2$.
+\item Run $\cA$ on $g$ with accuracy $\epsilon'$ and starting point $x_0$ returning a point $x'$.
+\item Return solution $x'$.
 \end{enumerate} 
-\item Reverse Deletion: Consider all edges $e$ in $F$ in reverse of the order they were added to $F$ and if $F\backslash \{e\}$ is feasible then set $F = F\backslash\{e\}$ and $x_e = 0$.
-\end{enumerate}
-\paragraph{}
-We intend to show that $\cA$ is a polynomial time $2$-approximation algorithm for $(IP)$. We will need the following lemma from class:
-\begin{lemma}\label{lemma:maximality}
-Let $f$ be a $01$-function satisfying maximality. Let $F\subseteq E$. Then
-\begin{enumerate}
-\item $F$ is feasible for $(IP)$ with $f$ if and only if every connected component $C$ of $(V,F)$ satisfies $f(C) = 0$.
-\item The minimal violated cuts are connected components $C$ that have $f(C) = 1$.
-\end{enumerate}
-\end{lemma}
-This lemma allows us to show that $\cA$ terminates in polynomial time.
-\begin{lemma}\label{lemma:poly1}
-Suppose we have a polynomial time oracle for $f$. Then $\cA$ returns a feasible solution for $(IP)$ in polynomial time
+Our algorithm assumes that the value of $||x_0 - x^*||_2$ is known in advance. In the trivial case where we know $x^*$ the algorithm returns it in step $1$. So we may assume that $x_0 \neq x^*$.
+\begin{lemma}\label{lemma:convex-smooth}
+The function $g$ is $\alpha$-strongly convex and $(\beta + \alpha)$-smooth.
 \end{lemma}
 \begin{proof}
-By definition of $\cA$ if $\cA$ terminates then the result is feasible.
-To see that $\cA$ terminates observe that in every iteration of step $2$ some edge $e \in |E|$ is added to $F$.
-Since $E$ is feasible, the while loop terminates after $O(|E|)$ iterations.
-Hence $\cA$ terminates, and returns a feasible solution.
-\paragraph{} 
-It remains to show that $\cA$ runs in polynomial time.
-Since we have a polynomial time oracle for $f$ it is easy to see that step $3$ runs in polynomial time.
-Further we need only maintain $y_S$ for non-zero $y_S$, of which we will show there are a polynomial number.
-We need to verify step $2(a)$ can be done in polynomial time.
-By lemma \ref{lemma:maximality} we need only maintain connected components of $F$ since the connected components $C$ with $f(C) = 1$ are precisely the minimal violated cuts.
-There are a polynomial number of connected components at each iteration (hence a polynomial number of non-zero $y_S$), and since we have a polynomial time oracle for $f$ this implies step $2(a)$ can be done in polynomial time.
+The Hessian of $g$ is:
+$$\nabla^2 g = \nabla^2 f + \alpha I.$$
+Since we know $f$ is $\beta$-smooth we have
+$$0I \preccurlyeq \nabla^2 f \preccurlyeq \beta I.$$
+Since adding scalar multiples of $I$ only shifts eigenvalues we observe
+$$\alpha I \preccurlyeq \nabla^2 g \preccurlyeq (\beta + \alpha) I.$$
+Hence the lemma holds as desired. 
 \end{proof}
 \paragraph{}
-Now we need to check the approximation factor. We present the following reduction from class as a lemma:
-\begin{lemma}\label{lemma:degree}
-Let $w_i$ denote the number of minimally violated sets at iteration $i$.
-Let $\cS_i$ denote the set of minimal violated cuts at iteration $i$.
-If $\sum_{S\in \cS_i} |\delta_F(S)| \leq 2w_i$ for all $i$ then $\cA$ is a $2$-approximation algorithm for $(IP)$ (for any $01$-function $f$).
+By lemma \ref{lemma:convex-smooth} we know that $\cA$ can process our function $g$. It remains to verify the accuracy of the point returned and to bound the running time.
+\begin{lemma}\label{lemma:q3-accuracy}
+The point $x'$ returned by our algorithm satisfies
+$$f(x') - f(x^*) \leq \epsilon.$$
 \end{lemma}
 \begin{proof}
-It is immediate from our construction of $x$ and $y$ that if $e \in F$ then
-$$\sum_{S : e\in\delta(S)} y_S = c(e).$$
-Thus we have the objection function value on $F$ is
-$$\sum_{e\in F} c(e) = \sum_{e\in F} \sum_{S : e\in\delta(S)} y_S = \sum_{S\subseteq V} |\delta_F(S)|y_S.$$
-Let $\Delta_i$ denote the amount by which all minimal violated cuts get increased in iteration $i$. Then
-$$\sum_{S\subseteq V} |\delta_F(S)|y_S = \sum_{i} \sum_{S\in \cS_i} |\delta_F(S)|\Delta_i \leq \sum_{i} 2w_i \Delta_i = 2\sum_{S\subseteq V} f(S) y_S.$$
-The inequality follows from our hypothesis, and the last equality follows since $f$ is a $01$-function for which non-zero $y_S$ (minimally violated sets) have $f(S) = 1$. 
+Let $\tilde{x}$ be the optimal solution for $g$, By our guarantees on $\cA$:
+$$g(x') - g(\tilde{x}) \leq \epsilon'.$$
+Now observe that
+\begin{align*}
+f(x') - f(x^*) &= g(x') - \frac{\alpha}{2}||x'-x_0||^2_2 - g(x^*) +\frac{\alpha}{2}||x^*-x_0||_2^2 \\
+&\leq g(x') - g(x^*) + \frac{\alpha}{2}||x^*-x_0||_2^2 &\text{since $\frac{\alpha}{2}||x'-x_0||^2_2\geq 0$} \\
+&= g(x') - g(x^*) + \frac{\epsilon}{2} &\text{by our choice of $\alpha$} \\
+&\leq g(\tilde{x}) + \epsilon' -g(x^*) + \frac{\epsilon}{2} \\
+&\leq \epsilon' +\frac{\epsilon}{2} &\text{by optimality of $\tilde{x}$} \\
+&= \epsilon &\text{by our choice of $\epsilon'$}.
+\end{align*}
 \end{proof}
 \paragraph{}
-Now we intend to show that the hypothesis of the previous lemma holds.
-Consider some iteration $i$ of $\cA$, and let $A$ be the of edges chosen so far at the end of $i$.
-Let $B = F \cap A$.
-Construct a graph $H$ by taking the graph $(V,F)$ and contracting the connected components of $(V,B)$ to vertices.
-Then $E(H) = F\backslash B$.
-Let $S_v$ denote the set of vertices in $G$ corresponding to vertex $v \in V(H)$.
-It is not hard to see that $H$ is a forest.
-Indeed if $H$ has a cycle $C$ then, since for all $v \in V(C)$, $S_v$ is a connected component in $G$, we can find a cycle $C'$ in $G$ by expanding each $v \in V(C)$ and joining its vertices which lie on $C$ by a path in $S_v$.
-But $F$ is a forest since it is necessary that any edge $e$ be a bridge to survive the reverse deletion step as $f$ is a $01$-function (observe that for all $S$ $|\delta(S) \backslash \{e\}| \geq 1 \geq f(S)$ if $e$ is not a bridge).
-\paragraph{}
-Now let $W\subseteq V(H)$ be the set of vertices of $H$ such that for all $w \in W$, $f(S_w) = 1$. Then $\{S_w : w\in W\} = \cS_i$. Let $d_v$ denote the degree of vertex $v$ in $H$. So the condition
-$$\sum_{S \in \cS_i} |\delta_F(S)| \leq 2w_i$$
-is equivalent to
-$$\sum_{w \in W} d_w \leq 2|W|.$$
-\begin{lemma}\label{lemma:leaves}
-$\sum_{w\in W} d_w \leq 2|W|$.
-\end{lemma}
-\begin{proof}
-First we claim that for every connected component $C$ of $H$, at most one leaf of $C$ satisfies $f(S_v) = 0$.
-Suppose for a contradiction there exists some connected component $C$ of $H$ that has two leaves $v$ and $w$ such that $f(S_v) = f(S_w) = 0$.
-Let $S_C = \cup_{u\in C} S_u$.
-Let $e_v, e_w \in F$ be the edges incident to $v$ and $w$ respectively in $H$.
-By minimality via the reverse deletion step, $F\backslash \{e_v\}$ and $F\backslash \{e_w\}$ are infeasible.
-But then $f(S_C\backslash S_v) =1 $ and $f(S_C\backslash S_w) = 1$ as these are precisely the cuts which would have no incident edges when $e_v$ and $e_w$ are removed respectively (we already know $f(S_v) = f(S_w) = 0$).
-Since $C$ is a connected component of $H$ we have $f(S_C) = 0$ by the feasibility of $F$.
-Hence
-$$f(S_v) = f(S_w) = f(S_C) = 0 \quad\text{and}\quad f(S_C\backslash S_v) = f(S_C \backslash S_w) = 1$$
-but $S_v, S_w \subseteq S_C$ and $S_v \cap S_w = \emptyset$. This contradiction property (\ref{eq:1}).
-\paragraph{}
-Now we proceed to demonstrate the inequality. First discard isolated vertices of $H$ since they affect the count in no way. Let $c$ denote the number of connected components of $H$. Now we have,
+Therefore our algorithm operates properly and returns a point within the desired accuracy.
+It remains to verify the running time of our algorithm.
+This running time is entirely bounded by the operation of $\cA$ on $g$ with $\epsilon'$.
+By the guarantee on $\cA$ this gives a bound of:
 \begin{align*}
-\sum_{w\in W} d_w &= \sum_{v \in H} d_v - \sum_{v\not\in W} d_v \\
-&= 2|H| -2c - \sum_{v\not\in W} d_v &\text{since $H$ is a forest} \\
-&\leq 2|H| - 2c - 2(|H| - |W| - c) \\
-&= 2|W|.
+O(\sqrt{\frac{\beta}{\alpha} + 1}\log (\frac{g(x_0)-g(\tilde{x})}{\epsilon'})) &= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{g(x_0)-g(\tilde{x})}{\epsilon})) &\text{dropping constants}\\
+&= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{f(x_0)-f(\tilde{x}) - \frac{\alpha}{2}||x_0-x^*||^2_2}{\epsilon})) &\text{expanding $g$}\\
+&= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{f(x_0)-f(\tilde{x})}{\epsilon})) &\text{dropping non-positive term} \\
+&= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{f(x_0)-f(x^*)}{\epsilon})) &\text{by optimality of $x^*$} \\
+&= O(\sqrt{\frac{\beta}{\epsilon}} ||x_0-x^*||_2\log (\frac{f(x_0)-f(x^*)}{\epsilon}))  &\text{expanding $\alpha$}.
 \end{align*}
-The inequality follows from our claim which implies all vertices not in $W$ have degree at least $2$ except for one leaf per connected component of $H$, which has degree $1$.
-\end{proof}
-\paragraph{}
-Hence combining lemma \ref{lemma:degree} and lemma \ref{lemma:leaves} with our discussion about $H$ we have shown that $\cA$ returns a solution $F$ with $$c(F) \leq 2\sum_{S\subseteq V} f(S) y_S \leq 2\cdot OPT,$$
-where $OPT$ is the optimal solution of $(IP)$. Therefore we have shown that
-\begin{theorem}
-$\cA$ is a $2$-approximation algorithm for $(IP)$.
-\end{theorem}
-
-%Q2
-\section{}
-\paragraph{}
-Let $G = (V,E)$ be a metric graph with edge costs $c: E\rightarrow \R_+$.
-Consider the problem of finding a minimum cost subset of edges $F\subseteq E$ such that each connected component of $(V,F)$ is an eulerian subgraph with at least $100$ vertices.
-We will denote this problem $(E100)$.
-\paragraph{}
-Define the function $f:2^V \rightarrow \{0,2\}$ by:
-$$f(S) = \begin{cases}
-2, &\text{if } 1\leq |S| < 100 \\
-0, &\text{otherwise.}
-\end{cases}$$
-Then we can define a linear programming relaxation $(LP)$ for $(E100)$:
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
-x&\geq 0.
-\end{align*}
-To justify this relaxation consider a feasible solution $F$ to $(E100)$.
-Let $x^F$ be the incidence vector of $F$.
-Then $x^F$ is feasible for $(LP)$.
-Let $S \subseteq V$ such that $1 \leq |S| < 100$.
-Since $|S| < 100$, $S$ does not contain any single connected component of $(V,F)$.
-Since $1\leq |S|$, $S$ has non-trivial intersection with some connected component $C$ of $(V,F)$ (i.e.
-$S\cap C \neq \emptyset$ and $S\cap C \supset C$).
-Since $F$ is feasible, $C$ is eulerian.
-Hence the degree of each vertex in $C$ (with respect to $F$) is even.
-Thus $|E(C\cap S, C\backslash S)| \geq 2$.
-Therefore
-$$\sum_{e \in \delta(S)} x_e \geq \sum_{e \in E(C\cap S, C\backslash S)} x_e \geq 2 \geq f(S),$$
-and hence $x$ is feasible for $(LP)$.
-\paragraph{}
-We now define another function $g: 2^V \rightarrow \{0,1\}$, given by
-$$g(S) = f(S)/2 = \begin{cases}
-1, &\text{if } 1\leq |S| < 100 \\
-0, &\text{otherwise.}
-\end{cases}$$  
-The integer program $(IP)$ for $g$ given by:
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq g(S) &\forall S \subseteq V \\
-x&\geq 0 \\
-x\in \Z^{|E|}
-\end{align*}
-describes the problem of finding a forest where each connected component is a tree with at least $100$ vertices.
-From our choice of $g$ it is clear that the connected components of any feasible solution must have at least $100$ vertices, as the cut defined by any smaller component does not satisfy the inequality for $g$.
-Optimal solutions to $(IP)$ do not have cycles, as we can remove any single edge present in a cycle and preserve connectivity.
-\paragraph{}
-Now we observe that $g$ satisfies property (\ref{eq:1}) from question $1$.
-Let $S \subseteq V$ and let $A, B \subseteq S$ such that $A \cap B = \emptyset$.
-Suppose that $g(S)=g(A) = g(B) = 0$.
-If $g(S\backslash A) = 1$ then $1 \leq |S\backslash A| < 100$.
-But $B \subseteq S\backslash A$, and hence $|B| \leq |S\backslash A|$.
-Since $g(B) = 0$ this means $|B| = 0$, that is $B = \emptyset$.
-Thus $g(S\backslash B) = g(S) = 0$.
-Similarly we can show if $g(S\backslash B) = 1$ then $g(S\backslash A) = 0$.
-Hence property $(\ref{eq:1})$ is satisfied by $g$.
-\paragraph{}
-Consider the following algorithm, $\cA$, for $(E100)$:
-\begin{enumerate}
-\item Set $S = \emptyset$.
-\item Solve $(IP)$ for $g$ using the algorithm of question $1$, obtaining a forest $F$.
-\item For each connected component $T$ of $(V,F)$:
-	\begin{enumerate}
-	\item Compute a matching $M$ on the odd vertices of $T$.
-	\item Shortcut $T\cup M$ obtaining an Euler Tour $A$.
-	\item Set $S = S \cup A$.
-	\end{enumerate}
-\item Return $S$.
-\end{enumerate}
-\paragraph{}
-We claim that $\cA$ is a $2$-approximation algorithm for $(E100)$.
-In question $1$ we showed that step $2$ can be done in polynomial time.
-In the lecture notes we have shown that the shortcutting process of each iteration of step $3$ can be done in polynomial time.
-Further there are a polynomial number of components the shortcutting is done on, so step $3$ also runs in polynomial time.
-Therefore $\cA$ returns a solution in polynomial time.
-The solution is feasible since each component is a cycle (hence is eulerian) on at least $100$ vertices.
-\paragraph{}
-We now demonstrate the approximation factor holds.
-We know from lecture that the shortcutting process of step $3$ produces a set of cycles of cost at most twice that of the original trees in a metric graph.
-That is,
-$$c(A)\leq 2c(F).$$
-Let $y$ be the optimal dual solution produces by the primal-dual algorithm of step $1$. We have from question $1$ that
-$$c(F) \leq 2\sum_{S\subseteq V} g(S) y_S = \sum_{S\subseteq V} f(S) y_S \leq OPT.$$
-where $OPT$ is the optimal solution to $(LP)$.
-This follows since the dual to $(LP)$ (with respect to $(E100)$) has the same feasible region as the dual to the linear programming relaxation of $(IP)$.
-Combining inequalities we see that
-$$c(A) \leq 2c(F) \leq 2 \cdot OPT.$$
-Hence $\cA$ is a $2$-approximation algorithm for $(E100)$.$\blacksquare$
-
-%Q3
-\section{}
-\paragraph{}
-Consider again the problem $(IP)$, associated with a graph $G=(V,E)$ with edges costs $c: E \rightarrow \R_+$ given by:
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(S)} x_e &\geq f(S) &\forall S \subseteq V \\
-x&\geq 0 \\
-x&\in \Z^{|E|}.
-\end{align*}
-Suppose that $f:2^V \rightarrow \N$ is symmetric, meaning for all $S\subseteq V$:
-$$f(S) = f(V\backslash S)$$
-and that $f$ is subadditive, meaning for all $A,B \subseteq V$:
-$$f(A\cup B) \leq f(A) + f(B).$$
-We also suppose that the graph $G$ is a cycle on $n$ vertices.
-%3a
-\subsection*{a}
-\paragraph{}
-We begin by working towards a compact formulation for $(IP)$ on $G$.
-Since $G$ is a cycle we may label the vertices of $G$ as $V(G) = \{v_1,\dots, v_n\}$ such that $v_iv_j \in E(G)$ if and only if $$j \equiv i\pm1\mod n$$.
-For $v_i, v_j \in V(G)$ we define the path from $v_i$ to $v_j$ as
-$$P(i, \ell) := \{v_i, v_{i+1}, \dots, v_{j} \}$$
-where addition is taken modulo $n$. We define the family of all paths on $G$ as
-$$\cP := \{P(v_i, v_j) : v_i, v_j \in V(G)\}.$$
-Observe that the size of $\cP$ is polynomial in $n$, in particular $|\cP| = O(n^2)$ (since there are $n$ choices for $v_i$ and $n$ choices for $v_j$).
-\begin{lemma}\label{lemma:compact}
-The problem $(IP)$ on $G$ is equivalent to the compact integer program $(\overline{IP})$ given by:
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
-x&\geq 0 \\
-x&\in \Z^{|E|}.
-\end{align*}
-\end{lemma}
-\begin{proof}
-Since the objective functions are the same, we need only demonstrate that the feasible regions of $(IP)$ and $(\overline{IP})$ are equal.
-Since each $P \in \cP$ is a subset of $V$, it is clear that the feasible region of $(IP)$ is a subset of the feasible region of $(\overline{IP})$.
-So we just have to show the reverse containment.
-\paragraph{}
-Let $x$ be a feasible solution for $(\overline{IP})$.
-Let $S \subseteq V$.
-If $S = V$ then $f(S)= f(V)=0\leq x(\delta(S)) $ (otherwise $(\overline{IP})$ and $(IP)$ are infeasible).
-By symmetry of $f$ the same holds for $S =\emptyset$.
-Thus we may assume $\emptyset \subset S \subset V$.
-If we consider the connected components of $G[S]$ we observe that they are all paths.
-Hence there exist disjoint $P_1, \dots, P_k \in \cP$ such that
-$$S = P_1 \dot\cup \dots \dot\cup P_k$$
-where each $P_i$ corresponds to a connected component of $G[S]$. Now we see that
-\begin{align*}
-f(S) &= f(P_1 \dot\cup \dots \dot \cup P_k) \\
-&\leq f(P_1) + \dots + f(P_k) &\text{by subadditivity} \\
-&\leq x(\delta(P_1)) + \dots x(\delta(P_k)) &\text{by feasibility of $x$}.
-\end{align*}
-We claim that the edges incident to each path are disjoint.
-To see this, suppose that we have $P_i$ and $P_j$ with $\delta(P_i) \cap \delta(P_j) \neq \emptyset$.
-Let $v_iv_j \in \delta(P_i) \cap \delta(P_j)$.
-Since $P_i$ and $P_j$ are disjoint we may assume $v_i \in P_i \backslash P_j$ and $v_j \in P_j \backslash P_i$.
-But then $P_i$ and $P_j$ are connected by $v_iv_j$ violating that they are distinct connected components.
-So the claim holds.
-Thus we observe
-\begin{align*}
-f(S) &\leq x(\delta(P_1)) + \dots x(\delta(P_k)) \\
-&= x(\bigcup_{i=1}^k \delta(P_i)) &\text{by disjointness}\\
-&= x(\delta(S)) &\text{by disjointness}.
-\end{align*}
-Therefore $x$ satisfies the cut constraint for all $S$, and hence $x$ is feasible for $(IP)$.
-\end{proof}
-\paragraph{}
-Hence by lemma \ref{lemma:compact} it is equivalent for us to solve $(\overline{IP})$. Now we define a linear programming relaxation for $(\overline{IP})$, which we will denote $(\overline{LP})$:
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
-x&\geq 0.
-\end{align*}
-Since $(\overline{LP})$ has a polynomial number of constraints and variables it can be solved in polynomial time.
-We now define our rounding-based approximation algorithm, $\cA$, for $(\overline{IP})$:
-\begin{enumerate}
-\item Solve $(\overline{LP})$ on $G$  returning solution $\bar{x}$.
-\item Let $\bar{x}_I \in \Z^{|E|}$ denote the integral part of $\bar{x}$. Let $\bar{x}_F \in [0,1)^{|E|}$ denote the fractional part of $\bar{x}$. Thus $\bar{x} = \bar{x}_I + \bar{x}_F.$
-\item Construct solution $x \in \Z^{|E|}$ by rounding. For each $e \in E$ set:
-$$x_e = \begin{cases}
-\bar{x}_I + 1, &\text{if } x_F \geq 0.5 \\
-\bar{x}_I, &\text{otherwise.}
-\end{cases}$$
-\item Return solution $x$.
-\end{enumerate}
-\begin{lemma}\label{lemma:feasible-rounding}
-If $\bar{x}$ is a feasible solution to $(\overline{LP})$ then the $x$ obtained from rounding $\bar{x}$ in step $3$ of $\cA$ is a feasible solution to $(\overline{IP})$.
-\end{lemma}
-\begin{proof}
-It is immediate from the description of step $3$ that $x$ is integral and non-negative.
-We need only verify the cut constraints.
-Let $P \in \cP$.
-Let $v_i, v_j \in P$ be the start and end vertices of $P$.
-That is,
-$$\delta(P) = \{v_{i-1}v_i,  v_jv_{j+1}\}.$$
-For convenience let $e_i = v_{i-1}v_i$ and let $e_j = v_jv_{j+1}$. Since $\bar{x}$ is feasible for $(\overline{LP})$:
-$$ f(P)\leq \bar{x}(e_i) + \bar{x}(e_j) = \bar{x}_I(e_i) + \bar{x}_F(e_i)  + \bar{x}_I(e_j) + \bar{x}_F(e_j).$$
-Since $f$ maps to $\N$ we have:
-$$f(P) \leq \floor{\bar{x}_I(e_i) + \bar{x}_F(e_i)  + \bar{x}_I(e_j) + \bar{x}_F(e_j)}.$$
-We consider the following cases: Case 1: $\bar{x}_F(e_i) + \bar{x}_F(e_j) < 1$ and Case 2: $\bar{x}_F(e_i) + \bar{x}_F(e_j) \geq 1$.
-\paragraph{}
-In Case 1 the fractional parts vanish during the floor-ing operation and hence
-$$ f(P) \leq \bar{x}_I(e_i) + \bar{x}_I(e_j)  \leq x(e_i) + x(e_j).$$
-In Case 2 we observe that $x_F(e_i) + x_F(e_j) = 1 + \alpha$ for some $\alpha \in [0,1)$. Thus
-$$f(P) \leq \floor{\bar{x}_I(e_i) + \bar{x}_F(e_i)  + \bar{x}_I(e_j) + \bar{x}_F(e_j)} = \bar{x}_I(e_i) + \bar{x}_I(e_j)  + 1.$$
-Since $\bar{x}_F(e_i) + \bar{x}_F(e_j) \geq 1$ at least one of $\bar{x}_F(e_i) ,\bar{x}_F(e_j) $ is greater than or equal to $0.5$.
-Assume without loss of generality $\bar{x}_F(e_i) \geq 0.5$.
-Then $x_(e_i) = \bar{x}_I(e_i) + 1$.
-So
-$$f(P) \leq \bar{x}_I(e_i) + \bar{x}_I(e_j)  + 1 = x(e_i) +\bar{x}(e_j) \leq x(e_i) + x(e_j).$$
-Therefore in either case $x$ satisfies the cut constraint for $P$. Therefore $x$ is feasible for $(\overline{IP})$.
-\end{proof}
-\begin{theorem}
-The algorithm $\cA$ is a $2$-approximation algorithm for $(\overline{IP})$.
-\end{theorem}
-\begin{proof}
-\paragraph{}
-As we have argued above, $(\overline{LP})$ can be solved in polynomial time, so step $1$ can be done in polynomial time.
-Step $3$ can be done in $O(|E|)$ time.
-Hence $\cA$ terminates in polynomial time.
-From lemma \ref{lemma:feasible-rounding} we see that $x$ is feasible for $(\overline{IP})$.
-It remains to check the approximation factor.
-\paragraph{}
-Let $e \in E$.
-If $\bar{x}(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq 2\bar{x}(e)$.
-It remains to consider the pertinent case when $\bar{x}(e) \geq 0.5$.
-Therein we have
-\begin{align*}
-\frac{x(e)}{\bar{x}(e)} &= \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) + \bar{x}_F(e)} \\
-&\leq  \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) +0.5} \\
-&= \frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}.
-\end{align*}
-Since $\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}$ is a non-increasing function, it is maximized when $\bar{x}_I(e) = 0$. Hence
-$$\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1} \leq \frac{2\cdot 0 + 2}{2\cdot 0 + 1} = 2.$$
-Thus $x(e) \leq 2\bar{x}(e)$ as desired. Therefore
-$$\sum_{e\in E}c(e) x(e) \leq \sum_{e \in E}c(e)2\bar{x}(e) = 2\sum_{e\in E}\bar{x}(e) \leq 2\cdot OPT,$$
-where $OPT$ is the cost of an optimal solution to $(\overline{IP})$ (the last inequality follows immediately from $\bar{x}$ being optimal for $(\overline{LP})$.
-So $\cA$ is a $2$-approximation algorithm.
-\end{proof}
-%3b
-\subsection*{b}
-\paragraph{}
-We make an observation about optimal solutions to $(\overline{IP})$ which contain a $0$-valued edge. This will lead us to an approximation algorithm with a better bound than the one found in $3(a)$.
-\begin{lemma}\label{lemma:0edge}
-Let $x$ be an optimal solution $(\bar{IP})$.
-Suppose there exists $v_iv_{i+1} \in E(G)$ such that $x(v_iv_{i+1}) = 0$.
-Reindex (by subtracting $i$ from each index) $V(G)$ so that $i = 0$.
-That is $x(v_0v_1)= 0$.
-Then for all $v_jv_{j+1} \in E(G)$, $j \neq 0$,
-$$x(v_jv_{j+1}) = f(P(v_1, v_j)).$$
-\end{lemma}
-\begin{proof}
-For each $v_jv_{j+1} \neq v_0v_{1} \in E(G)$. Since $x$ is feasible and $x(v_0v_{1}) = 0$,
-$$f(P(v_1, v_j)) \leq x(\delta(P(v_1,v_j))) = x(v_jv_{j+1}) + x(v_0v_{1}) = x(v_jv_{j+1}).$$
-Suppose for a contradiction there exists $e_j = v_{j}v_{j+1} \in E(G)$ such that
-$$f(P(v_1,v_j)) < x(v_jv_{j+1}).$$
-Since $f(V) = 0$, $j\neq 0$. Let $x' \in \Z^{|E|}$ we given by
-$$x'(e) = \begin{cases}
-x(e), &\text{ if $e\neq v_jv_{j+1}$} \\
-f(P(v_1,v_{j})), &\text{if $e=v_jv_{j+1}$}. 
-\end{cases}$$
-We claim that $x'$ is feasible for $(\overline{IP})$.
-By the feasibility of $x$, we need only check cut constraints involving $v_jv_{j+1}$.
-Let $v_i \in V(G)$, such that $i \neq j-1$.
-We consider $P(v_i,v_j)\in \cP$.
-\paragraph{}
-First consider the case where $0 < i \leq j$. In this case $v0,v_1 \in P(v_{j+1}, v_{i-1})$. We compute:
-\begin{align*}
-f(P(v_i,v_j)) &= f(P(v_{j+1}, v_{i-1})) &\text{by symmetry} \\
-&\leq f(P(v_{j+1}, v_0)) + f(P(v_{1},v_{i-1})) &\text{by subadditivity} \\
-&= f(P(v_1, v_j)) + f(P(v_1, v_{i-1})) &\text{by symmetry} \\
-&= x'(v_jv_{j+1}) + f(P(v_1, v_{i-1})) &\text{by definition of $x'$} \\
-&\leq x'(v_jv_{j+1}) + x(v_1v_{i-1}) +x(v_0v_1) &\text{by feasibility of $x$} \\
-&= x'(v_jv_{j+1}) + x'(v_1v_{i-1}) &\text{by definition of $x'$, and $x(v_0v_1)=0$} \\
-&=x'(\delta(P(v_i,v_j))).
-\end{align*}
-Hence $x'$ satisfies the cut constraint.
-\paragraph{}
-Now consider the case where $i > j$. In this case $v_0,v_1 \in P(v_i,v_j)$. We compute (similarly):
-\begin{align*}
-f(P(v_i,v_j))&\leq f(P(v_{i}, v_0)) + f(P(v_{1},v_{j})) &\text{by subadditivity} \\
-&= f(P(v_1, v_{i-1})) + f(P(v_1, v_{j})) &\text{by symmetry} \\
-&= f(P(v_1,v_{i-1})) + x'(v_jv_{j+1}) &\text{by definition of $x'$} \\
-&\leq x(v_1v_{i-1}) + x(v_0v_1) + x'(v_jv_{j+1}) &\text{by feasibility of $x$} \\
-&= x'(v_jv_{j+1}) + x'(v_1v_{i-1}) &\text{by definition of $x'$, and $x(v_0v_1)=0$} \\
-&=x'(\delta(P(v_i,v_j))).
-\end{align*}
-Therefore in either case $x'$ is feasible. But
-\begin{align*}
-\sum_{e\in E} c(e) x(e) &= c(v_jv_{j+1})x(v_jv_{j+1}) + \sum_{e\in E\backslash\{v_jv_{j+1}\}} c(e)x(e) \\
-&> c(v_jv_{j+1})x'(v_jv_{j+1}) +  \sum_{e\in E\backslash\{v_jv_{j+1}\}} c(e)x(e) &\text{since $x'(v_jv_{j+1}) = f(P(v_1,v_j) < x(v_jv_{j+1})$} \\
-&= c(v_jv_{j+1})x'(v_jv_{j+1}) + \sum_{e\in E\backslash\{v_jv_{j+1}\}} c(e)x'(e) &\text{by our choice of $x'$} \\
-&= \sum_{e\in E} c(e) x'(e),
-\end{align*}
-contradicting the optimality of $x$.
-\end{proof}
-\paragraph{}
-This lemma tells us precisely how to construct optimal solutions to $(\overline{IP})$ in the presence of $0$-valued edges.
-The idea for our algorithm is to construct from this lemma each potential solution with a $0$-valued edge, as well as an approximate solution with no $0$-valued edges, and return the best possible of those $|E|+1$ solutions.
-\paragraph{}
-In our algorithm we will use the following linear program, which is $(\overline{LP})$ with the constraint $x \geq 1$ instead of $x\geq 0$. We define the linear program $(\tilde{LP})$ as
-\begin{align*}
-\min\ \sum_{e\in E} c(e) x_e&\\
-\text{s.t.}\  \sum_{e\in\delta(P)} x_e &\geq f(P) &\forall P \in \cP \\
-x&\geq 1.
-\end{align*}
-\paragraph{}
-Consider the following improved algorithm for solving $(\overline{IP})$ which we will denote by $\cA'$:
-\begin{enumerate}
-\item Let $E' = \emptyset$. For each $e=v_iv_{i+1} \in E$:
-	\begin{enumerate}
-	\item Construct the solution $x^{e}$ as follows.
-	\item Set $x^{e}(e)=0$.
-	\item For each $v_jv_{j+1} \neq e$ Set $x^{e}(v_jv_{j+1}) = f(P(v_{i+1},v_j))$.
-	\item If $x^{e}$ is feasible for $(\overline{IP})$ add $e$ to $E'$.
-	\end{enumerate} 
-\item Solve $(\tilde{LP})$ on $G$  returning solution $\bar{x}$.
-\item Let $\bar{x}_I \in \Z^{|E|}$ denote the integral part of $\bar{x}$. Let $\bar{x}_F \in [0,1)^{|E|}$ denote the fractional part of $\bar{x}$. Thus $\bar{x} = \bar{x}_I + \bar{x}_F.$
-\item Construct solution $x \in \Z^{|E|}$ by rounding. For each $e \in E$ set:
-$$x_e = \begin{cases}
-\bar{x}_I + 1, &\text{if } x_F \geq 0.5 \\
-\bar{x}_I, &\text{otherwise.}
-\end{cases}$$
-\item Return $\arg\min\{c^Tx : x \in \{x^e : e \in E'\} \cup \{x\}\}$.
-\end{enumerate}
-\begin{theorem}
-The algorithm $\cA'$ is a $\frac{4}{3}$-approximation algorithm for $(\overline{IP})$.
-\end{theorem}
-\begin{proof}
-Step $1$ can be done in $O(|E|)$ iterations, each using $O(|E|)$ calls to the (we may assume polynomial time) oracle for $f$, and hence runs in polynomial time.
-Step $2$ can be done in polynomial time via the same argument used for $(\overline{LP})$.
-Step $4$ can be done in $O(|E|)$ time, as can step $5$.
-Therefore $\cA'$ runs in polynomial time.
-\paragraph{}
-By definition, for each $e \in E'$, $x^{e}$ is feasible.
-Since the feasible region of $(\tilde{LP})$ is a subset of the feasible region of $(\overline{LP})$, $\bar{x}$ is feasible for $(\overline{LP})$ and therefore by lemma \ref{lemma:feasible-rounding} $x$ is feasible for $(\overline{IP})$ as Step $4$ of $\cA'$ is the same rounding procedure as step $3$ of $\cA$.
-Hence $\cA'$ returns a feasible solution for $(\overline{IP})$.
-\paragraph{}
-If there exists an optimal solution $x^*$ for $(\overline{IP})$ and $e \in E$ such that $x^*(e) = 0$ then $x^*(e) = x^{e}$ by lemma \ref{lemma:0edge} and $e \in E'$.
-Thus in step $5$ $\cA'$ returns a solution of cost at most that of $x^{e} = x^*$ which is therefore optimal.
-Trivially such a solution is of cost at most $\frac{4}{3}$ that of optimal.
-\paragraph{}
-It remains to consider the case where no optimal solution $x^*$ for $(\overline{IP})$ has $e\in E$ such that $x^*(e) = 0$.
-Since $x^*$ is integral and no component is zero optimal solutions for $(\overline{IP})$ satisfy the constraint $x\geq 1$.
-Therefore $(\tilde{LP})$ is a relaxation of $(\overline{IP})$.
-\paragraph{}
-Now it suffices to show that the returned solution $x$ has cost at most $\frac{4}{3}$ that of $\bar{x}$.
-Let $e \in E$.
-If $\bar{x}(e) < 0.5$ then we have immediately $x(e) \leq \bar{x}(e)$ and so trivially we have $x(e) \leq \frac{4}{3}\bar{x}(e)$.
-It remains to consider the pertinent case when $\bar{x}(e) \geq 0.5$.
-Therein we have
-\begin{align*}
-\frac{x(e)}{\bar{x}(e)} &= \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) + \bar{x}_F(e)} \\
-&\leq  \frac{\bar{x}_I(e) + 1}{\bar{x}_I(e) +0.5} \\
-&= \frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}.
-\end{align*}
-Since $\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1}$ is a non-increasing function, it is maximized when $\bar{x}_I(e) = 1$ (since $\bar{x}_I(e) \geq 1$). Hence
-$$\frac{2\bar{x}_I(e) + 2}{2\bar{x}_I(e) + 1} \leq \frac{2\cdot 1 + 2}{2\cdot 1 + 1} = \frac{4}{3}.$$
-Thus $x(e) \leq \frac{4}{3}\bar{x}(e)$ as desired. 
-\end{proof}
-
-%Q4
-\section{}
-\paragraph{}
-Consider the packing problem $(P)$: Let $U$ be a set of elements.
-Let $\cS$ be a non-empty family of sets $S\subseteq U$.
-The goal is to find a subcollection of $\cS$ of maximum cardinality, subject to the constraint that each element is contained in at most $1$ set.
-\paragraph{}
-We can write a linear programming relaxation for $(P)$, denoted $(LP)$ as
-\begin{align*}
-\max \sum_{S\in\cS} x_S\ & \\
-\text{s.t.} \sum_{S:e\in S} x_S &\leq 1 &\forall e\in U \\
-x&\geq 0.
-\end{align*}
-Let $k = \max_{S\in \cS} \{ |S|\}$.
-%pa
-\subsection*{a}
-\paragraph{}
-Let $x$ be a basic optimal solution to $(LP)$.
-Suppose for a contradiction that for all $S \in \cS$, $x_S < \frac{1}{k}$.
-We will contradict the fact that the number of tight constraints with respect to $x$ is at least the number of variables, that is $|\cS|$.
-We will proceed by a token argument.
-Give a token to each $S \in \cS$.
-Redistribute the tokens according to the following rules, for each $S \in \cS$:
-\begin{enumerate}
-\item If $x_S = 0$ distribute the token to the non-negativity constraint corresponding to $S$.
-\item Otherwise, for each $e \in S$ distribute $x_S$ token value to $e$.
-\end{enumerate}
-To achieve the desired contradiction, we need to show three things.
-First, that each $S \in \cS$ gives away at most $1$ token.
-Second, that each tight constraint receives at least $1$ token.
-Third, that some $S \in \cS$ keeps some fraction of their token.
-Together these propositions imply that $|\cS|$ is greater than the number of tight constraints at $x$.
-We complete the proof by showing these three porpositions.
-\paragraph{}
-We begin with the first proposition. If $x_S = 0$ then $S$ gives away at most one token by rule $1$. If $x_S > 0$ then $S$ gives away $\sum_{e: e\in S} x_S$ token value. Observe that
-$$\sum_{e: e\in S} x_S < \sum_{e: e\in S} \frac{1}{k} = \frac{|S|}{k} \leq 1.$$
-Thus $S$ gives away strictly less than one token.
-Thus to show proposition three we need only to show that there exists some $S$ with $x_S > 0$.
-This follows since $x$ is optimal.
-If all $S$ had $x_S = 0$, since $\cS \neq\emptyset$ we could simply set $x_S = 1$ for one $S \in \cS$ and increase cardinality, contradicting the optimality of $x$.
-Hence some $S$ exists with $x_S > 0$.
-Therefore we have show propositions one and three.
-\paragraph{}
-It remains to verify proposition two: that each tight constraint receives at least $1$ token.
-It is easy to see by rule $1$ that each tight non-negativity constraint receives $1$ token.
-Now consider a tight constraint corresponding to some $e \in U$.
-By rule two $e$ receives
-$$\sum_{S : e\in S, x_S > 0} x_S$$
-token value. But this is equal to
-$$\sum_{S: e\in S} x_S$$
-which in turn, since $e$ corresponds to a tight constraint satisfies
-$$\sum_{S: e\in S} x_S = 1.$$
-That is, $e$ receives $1$ token. Therefore every tight constraint receives $1$ token and this completes the proof. $\blacksquare$
-%pb
-\subsection*{b}
-\paragraph{}
-Consider the following iterative rounding algorithm, $\cA$, for packing problem $(P)$:
-\begin{enumerate}
-\item Set $H = \emptyset$, $\cS' = \cS$, and $U' = U$.
-\item While $\cS' \neq \emptyset$ do:
-	\begin{enumerate}
-	\item Let $x$ be a basic optimal solution to $(LP)$ on $\cS'$ and $U'$.
-	\item Let $S \subseteq \cS'$ such that $x_S \geq \frac{1}{k}$ where $k = \max_{S' \in \cS'}\{|S'|\}$ .
-	\item Set $H = H \cup \{S\}$.
-	\item For each $S' \in \cS'$ such that $S' \cap S \neq \emptyset$, remove $S'$ from $\cS'$.
-	\item For each $e \in S$, remove $e$ from $U'$. 
-	\end{enumerate}
-\item Return $H$.
-\end{enumerate}
-\begin{theorem}
-The algorithm $\cA$ is a $\frac{1}{k-1+\frac{1}{k}}$-approximation algorithm for $(P)$.
-\end{theorem}
-\begin{proof}
-\paragraph{}
-To see that $\cA$ terminates observe that by problem $3(a)$, a set $S \subseteq \cS'$ is always found in step $2(b)$, and hence $\cS'$ is eventually empty.
-To verify that $\cA$ runs in polynomial time, observe that $(LP)$ has a constraint for every $e \in U$ and a variable for every $S \in \cS$.
-Thus the size of $(LP)$ is polynomial in the size of the input to the problem.
-Each subsequent $(LP)$ is smaller than the previous, thus $2(a)$ runs in polynomial time.
-Step $2(b)$ can be done in $O(|\cS|)$ time, $2(c)$ in constant time, $2(d)$ in $O(|\cS|^2)$ time, and step $2(e)$ in $O(k|U|)$ time.
-These are all very pessimistic estimates,  but in any case they show that each iteration of step $2$ runs in polynomial time.
-Step $2$ has at most $|\cS|$ iterations and hence runs in polynomial time.
-Thus $\cA$ terminates in polynomial time.
-\paragraph{}
-Now we verify that the solution returned is feasible for $(P)$.
-We claim that $\cA$ maintains the invariant that after each iteration of step $2$, for each $S \in \cS'$ and each $T \in H$, $S\cap T = \emptyset$.
-It is easy to see that this invariant holds before step $2$.
-Suppose it holds inductively at the beginning of some iteration of step $2$.
-Let $S$ be the set chosen to enter $H$ in step $2(c)$.
-By step $2(d)$ each $S' \in \cS'$ that intersects $S$ is removed from$ \cS'$.
-Since no set in $\cS'$ intersects sets in $H$ by induction, after step $2(d)$ no set in $\cS'$ intersects $H \cup \{S\}$.
-Therefore the invariant holds.
-Using this invariant it is easy to see that at each iteration of step $2$ a set is added to $H$ which intersects no other set in $H$.
-Therefore at the termination of step $2$ the set $H$ returned is feasible for $(P)$.
-\paragraph{}
-It remains to demonstrate the approximation factor.
-We will bound the increase in $H$ versus the decrease in the fractional solution at each iteration of step $2$.
-Let $\Delta_i H$ denote the change in $|H|$ from the beginning of iteration $i$ to the end of iteration $i$..
-Exactly one set is added to $H$ in each iteration so $\Delta_i H = 1$ for all $i$.
-Let $x^i$ be the fractional solution to $(LP)$ found at the beginning of iteration $i$.
-Let $\Delta_i x = \sum_{S \in \cS}x^{i}_S - \sum_{e \in \cS}x^{i+1}_S$, where we take $x^{n+1} = 0$ if step $2$ terminates in $n$ iterations, and take $x^{i}_S = 0$ if $S \not\in \cS'$ during iteration $i$.
-\paragraph{}
-Consider an arbitrary iteration of step $2$ of $\cA$.
-Let $x^i$ be the solution to $(LP)$ found in step $2(a)$.
-Let $S$ be the set chosen in step $2(b)$.
-For each $e \in S$, let $\cS^e = \{S' \in \cS'\backslash\{S\} : e \in S'\}$.
-Let the collection of deleted sets from step $2(d)$ be $D: = \bigcup_{e\in S} \cS^e$.
-Then
-$$x(D) \leq \sum_{e \in S} x(\cS^{e}) = \sum_{e\in S} \sum_{S' \in \cS'\backslash\{S\} : e\in S'} x_{S'} \leq \sum_{e \in S} (1 - x_S) = |S|(1-x_S),$$
-with the last inequality following from the corresponding packing constraints of $(LP)$ for each $e$.
-Now we observe that $x^{i}$ restricted to the variables of $\cS' \backslash (\{S\} \cup D)$ is feasible for $(LP)$ during iteration $i+1$.
-Hence $$\Delta_i x \leq \sum_{S'\in \cS'} x^i_{S'} - \sum_{S' \in \cS'\backslash (\{S\} \cup D)\}} x^i_{S'} = x_S + x(D).$$
-Thus by our bound on $x(D)$ we have
-\begin{align*}
-\Delta_i x &\leq x_S + |S|(1-x_S) \\
-&= 1 - (1-x_S) + |S|(1-x_S) \\
-&= 1 + (|S| -1)(1-x_S) \\
-&\leq 1+(k - 1)(1-x_S) &\text{by our choice of $k$ in step $2(b)$} \\
-&\leq 1 + (k-1)(1-\frac{1}{k}) &\text{by our choice of $x_S$} \\
-&= 1 + k - 1 - 1 + \frac{1}{k} \\
-&= k-1 + \frac{1}{k}.
-\end{align*}
-So we observe that $$\Delta_i H - \frac{1}{k-1+\frac{1}{k}}\Delta_i x \geq 0.$$
-Summing over each iteration $i$ (and taking $k$ to be the initial $k$ from the first iteration, since $k$ is non-increasing as the algorithm proceeds) we obtain
-$$\sum_i \Delta_i H - \frac{1}{k-1+\frac{1}{k}} \sum_i \Delta_i x \geq 0.$$
-Since $\sum_i \Delta_i H = |H|$, and $\sum_i \Delta_i x = \sum_i (x^i - x^{i+1}) = x^1$, we obtain
-$$|H| - \frac{1}{k-1+\frac{1}{k}}x^1 \geq 0.$$
-Since $x^1$ is an optimal solution to $(LP)$ for our original problem $(P)$ this implies that $|H|$ is at least $\frac{1}{k-1+\frac{1}{k}}$ times the cardinality of an optimal solution to $(P)$.
-\end{proof}
-
-%Q5
-\section{}
-\paragraph{}
-Academic Honesty Disclaimer: In this problem I rely heavily on her paper Dr. Sanita gave us on Learn: Steiner Tree Approximation via Iterative Randomized Rounding.
-\paragraph{}
-Given a metric graph $G=(V,E)$ with edge costs $c:E\rightarrow \R_+$ and a set of terminals $R\subseteq V$, the $k$-restricted Steiner Tree problem asks for a tree of minimum cost spanning $R$ whose full components each contain at most $k$ terminals.
-From class the following algorithm was shown to be a $\ln(4)$-approximation algorithm for this problem.
-\paragraph{}
-Let $\cC$ denote the set of all directed components which can be obtained in the following way.
-Let $R' \subseteq R$ such that $|R'| \leq k$.
-and let $r' \in R'$.
-Let $C$ be the minimum cost Steiner tree on $R'$, and direct all edges of $C$ towards $r'$.
-Then $r'$ denotes the unique sink terminal in this case.
-A component $C \in \cC$ ``crosses" $U\subseteq R$ if $C$ has at least one source in $U$ and its unique sink in $R\backslash U$.
-The set $\delta^+(U)$ will denote the set of $C \in \cC$ which cross $U$.
-Fix an arbitrary root terminal $r \in R$.
-The directed cut relaxation $(k-DCR)$ is
-\begin{align*}
-\min \sum_{C\in \cC} c(C)x_C& \\
-\text{s.t.} \sum_{C\in\delta^+(U)} x_C &\geq 1 &\forall U\subseteq R\backslash \{r\}, U\neq \emptyset \\
-x&\geq 0.
-\end{align*}
-As argued in class this is a relaxation of the $k$-restricted Steiner tree problem. The iterative randomized rounding algorithm, $\cA$, for this problem is:
-\begin{enumerate}
-\item For $t =1,2,\dots$
-\begin{enumerate}
-\item Solve $(k-DCR)$ on $G$ to obtain $x^t$.
-\item Sample one component $C^t$ where $C^t=C$ with probability $x^t_C/\sum_{C\in\cC} x^t_C$. Contract $C^t$ into its sink in $G$.
-\item If a single terminal remains, return $\bigcup_{i=1}^t C^i$.
-\end{enumerate}
-\end{enumerate}
-\paragraph{}
-Since we already know from class that $\cA$ is a $\ln(4)$-approximation algorithm for $k$-restricted Steiner Tree, it only remains to verify the new approximation factor of $\frac{5}{4}$ when we fix $k=4$.
-Let $S^*$ be an optimal $4$-restricted Steiner Tree with undirected full components $Z_1, \dots, Z_\ell$.
-We will start by studying the structure of each $Z_i$.
-The first observation, made in the paper, is that we may assume Steiner Nodes of $S^*$ have degree at least $3$.
-This follows since $G$ is a metric graph we may shortcut any Steiner Node of degree two by connecting its endpoints and removing it.
-By the triangle inequality this shortcutting operation does not increase the cost of the solution.
-\paragraph{}
-We now move to observations about each $Z_i$.
-\begin{lemma}\label{lemma:dk}
-If $Z_i$ is a $k$-restricted component for any $k$, then the degree of any Steiner Node in $Z_i$ is at most $k$.
-\end{lemma}
-\begin{proof}
-Suppose for a contradiction that $Z_i$ has a Steiner Node $s$ of degree greater than $k$.
-Then $s$ has at least $k+1$ neighbours: $v_1, \dots, v_{k+1}$.
-Since $Z_i$ is a tree, there is a unique path from $s$ to each of the $k$ terminals of $Z_i$, $r_1, \dots, r_k$.
-Each path $P_{s,r_j}$ can use only one of $v_1,\dots, v_{k+1}$.
-If they use more, say first $v_i$ and then next $v_j$ then the subpath $r,v_i,\dots,v_j$ with the edge $v_js$ forms a cycle in $Z_i$ contradicting that $Z_i$ is a tree.
-But then one of $v_1, \dots, v_{k+1}$ is not used on any path from $s$ to any of $r_1, \dots, r_k$.
-The edge from $s$ to the unused $v_j$ can then be removed from $Z_i$ without disconnecting the terminals, contradicting optimality.
-\end{proof}
-\paragraph{}
-We now consider the classes of graphs these components appear as.
-It is an easy observation that $1$-restricted components only appear in the trivial case where $|R| = 1$.
-It is also easy to see that $2$-restricted components arise as a single edge connecting two terminals since Steiner Nodes have degree at least three.
-Further we see that $k$-restricted components with exactly one Steiner Node are star graphs centred at the Steiner Node.
-By lemma \ref{lemma:dk} $3$-restricted components have a Steiner Nodes with degree exactly three, and thus they can only have one Steiner Node.
-This implies that $3$-restricted components are star graphs.
-Things get slightly more interesting with $4$-restricted components.
-\begin{lemma}\label{lemma:four}
-A $4$-restricted component $Z_i$ is either a star graph or consists of exactly two adjacent degree three Steiner Nodes each adjacent to precisely two distinct terminals.
-\end{lemma}
-\begin{proof}
-Suppose $Z_i$ is not a star graph.
-Then $Z_i$ has more than one Steiner Node.
-We claim that $Z_i$ has at most two adjacent Steiner Nodes.
-Suppose for a contradiction $Z_i$ has at least three adjacent Steiner Nodes.
-Say $s_1,s_2,s_3$ is a path of Steiner Nodes in $Z_i$.
-If the removal of edge $s_1s_2$ separates one terminal in the component containing $s_1$ then $s_1$ is adjacent to a redundant edge since its degree is at least three.
-So the edge removal separates at least two terminals in the component containing $s_1$.
-Similarly, by removing $s_2s_3$ we see that there are two terminals in the component containing $s_3$.
-But then, since $s_2$ of degree at least three, any third component it connects to (that is, not containing $s_1$ or $s_3$) is redundant to connect the terminals.
-This violates optimality.
-Hence $Z_i$ contains exactly two Steiner Nodes, and we see by the above argument that the removal of their adjacent edge separates two terminals into each connected component.
-\end{proof}
-\paragraph{}
-We are prepared to analyze the algorithm.
-Following the notation in the paper we let $D(e) = \max\{t : e \in S^t\}$ be the iteration at which $e$ is deleted.
-We construct the witness tree via the following procedure:
-\begin{enumerate}
-\item Start with $\tilde{B} = \emptyset$
-\item For each $i = 1,\dots, \ell$
-	\begin{enumerate}
-	\item If $Z_i$ is a $2$-restricted component add its edge to $\tilde{B}$.
-	\item If $Z_i$ is a $3$-restricted component or $4$-restricted star graph component add all the edges of $Z_i$ but one chosen uniformly at random.
-	\item If $Z_i$ is a $4$-restricted component with two adjacent Steiner Nodes $s_1,s_2$ and terminals $r_1, r_2$ connected to $s_1$ and terminals $r_3,r_4$ connected to $s_2$ then:
-	\begin{enumerate}
-		 \item add edge $s_1s_2$ to $\tilde{B}$
-		 \item add one edge of $r_1s_1$, $r_2s_1$ to $\tilde{B}$ chosen uniformly at random
-		 \item add one edge of $r_3s_2$, $r_4s_2$ to $\tilde{B}$ chosen uniformly at random
-\end{enumerate}
-	\end{enumerate}
-\item Return Witness Tree $W= \{\{u,v\} in {R\choose 2} : |P_uv \cap \tilde{B}| = 1\}$.
-\end{enumerate}
-\paragraph{}
-By lemma \ref{lemma:four} and the preceding discussion, this procedure covers all possible forms the components $Z_i$ can take. Further one can verify that $W$ is a terminal spanning tree. 
-\paragraph{}
-For any $e \in E(S^*)$ we define $W(e) = \{ uv \in E(W): e \in P_{uv}\}$.
-When the edges of $W(e)$ are $\textit{marked}$ (see paper) the edge $e$ is deleted.
-We have the following lemma from the paper/class (rephrased in terms of $W(e)$ for simplicity):
-\begin{lemma}\label{lemma:20}
-Let $e \in E(S^*)$. Then the expected number of iterations until all the edges of $W(e)$ are $\textit{marked}$ satisfies $$E[X(W(e))] \leq H_{|W(e)|} \cdot M,$$
-where $M$ is an upperbound on $|\cC|$ polynomial in $n$.
-\end{lemma}
-\begin{lemma}\label{lemma:de}
-For any edge $e \in E(S^*)$, the expected number of iterations until $e$ is deleted is
-$$E[D(e)] \leq \frac{5}{4} M.$$
-\end{lemma}
-\begin{proof}
-Let $Z_i$ be the undirected full component containing $e$. We proceed by case analysis on the class of graph $Z_i$ belongs to. We have the following cases:
-\begin{enumerate}
-\item $Z_i$ is $2$-restricted,
-\item $Z_i$ is $3$ or $4$-restricted star,
-\item $Z_i$ is a $4$-restricted component with two degree three Steiner Nodes.
-\end{enumerate}
-\paragraph{}
-We begin with Case $1$. This case is trivial: with probability $1$, $e$ is the edge connecting the terminals. Thus by lemma \ref{lemma:20} $E[D(e)] = 1\cdot H_1 \cdot M = M \leq \frac{5}{4} M.$
-\paragraph{}
-Now we consider Case $2$.
-Suppose $Z_i$ is a $k$-restricted star where $k \in \{3,4\}$.
-Then with probability $\frac{1}{k}$ (when $e$ is the edge not in $\tilde{B}$), $|W(e)| = k-1$.
-With probability $1-\frac{1}{k}$), $|W(e)| = 1$.
-Thus by lemma \ref{lemma:20} we have
-\begin{align*}E[D(e)] &\leq \frac{1}{k} H_{k-1} M + (1-\frac{1}{k})H_1 M
-\\&=\begin{cases}
-(\frac{1}{3}H_2 + \frac{2}{3})M, &\text{if $k = 3$} \\
-(\frac{1}{4}H_3 + \frac{3}{4})M, &\text{if $k=4$}.
-\end{cases} \\
-&=\begin{cases}
-\frac{7}{6}M, &\text{if $k = 3$} \\
-(\frac{29}{24})M, &\text{if $k=4$}.
-\end{cases} \\
-&\leq \frac{5}{4} M.
-\end{align*}
-\paragraph{}
-Finally we consider Case $3$.
-By lemma \ref{lemma:four} we have the following structure.Let $s_1,s_2$ be the two adjacent Steiner Nodes of $Z_i$ and let $r_1,r_2$ be the terminals adjacent to $s_1$, and let $r_3,r_4$ be the terminals adjacent to $s_2$.
-Suppose without loss of generality that $s_1r_1, s_2r_4 \in \tilde{B}$.
-The the probability that $|W(e)| = 1$ is $\frac{3}{5}$ (such edges are $s_1r_1, s_2r_4$ and $s_1s_2$).
-The probability that $|W(e)| = 2$ is $\frac{2}{5}$ (such edges are $s_1r_2$ and $s_2r_3$).
-Thus by lemma \ref{lemma:20} we have
-\begin{align*}
-E[D(e)] &\leq \frac{3}{5}H_1M + \frac{2}{5}H_2M \\
-&= (\frac{3}{5} + \frac{2}{10}) M \\
-&= (\frac{8}{10})M \\
-&= \frac{4}{5}M.
-\end{align*} 
-Therefore the expectation holds in each case as desired.
-\end{proof}
-\paragraph{}
-We now finish by computing the expected cost of the solution returned by $\cA$. Let $opt^t_f$ denote the optimal fractional cost to $(k-DCR)$ at iteration $t$. Let $opt$ denote the cost of $S^*$.
-\begin{theorem}
-The algorithm $\cA$ is $\frac{5}{4}$-approximation algorithm for $4$-restricted Steiner Tree.
-\end{theorem}
-\begin{proof}
-We observe that
-\begin{align*}
-E[\sum_{t\geq 1} c(C^t)] &\leq \sum_{t\geq 1} \frac{1}{M}E[opt^t_f] \\
-&\leq \frac{1}{M} \sum_{t\geq 1} E[c(S^t)] \\
-&=\frac{1}{M} \sum_{t\geq 1} E[D(e)]c(e) \\
-&\leq \frac{1}{M} \sum_{t\geq 1} \frac{5}{4}M\cdot c(e) &\text{by lemma \ref{lemma:de}} \\
-&=\frac{5}{4} \sum_{t \geq 1} c(e) \\
-&= \frac{5}{4} opt.
-\end{align*}
-\end{proof}
-
-%Q6
-\section{}
+Thus we obtain the desired running time. Together with lemma \ref{lemma:q3-accuracy} this completes the problem.
 \end{document}

### 2017-03-28 CS798-ConvexityAndOptimization/assignments/project_outline.tex
writeup q3

--- 2017-03-28 CS798-ConvexityAndOptimization/assignments/assignment_02.tex
@@
 
-\lhead{2017-03-27}
-\rhead{William Justin Toth CS798-Convexity and Optimization Assignment 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-03-03}
+\rhead{William Justin Toth CS798-Convexity and Optimization Project Outline} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+\paragraph{Project Title:} Applications of the Multiplicative Weight Update Method to Game Theory.
+\paragraph{Goal:} The goal of this report is to study problems in algorithmic game theory where algorithms featuring the Multiplicative Weight Update Method have found use.
+\paragraph{Review}
+We will begin by reviewing the multiplicative weight update method studied in class \cite{arora2012multiplicative} and covering its analysis.
+This will give time to set up some basic notation and make the report more self-contained.
+\paragraph{Two-Player Zero Sum Games} We will define Two-Player Zero Sum Games, and show how multiplicative weights can be used to provide approximately optimal (up to a an additive error) mixed strategies.
+This result first appeared in \cite{freund1999adaptive}
+\paragraph{A Generalized Multiplicative Weight Update Method}
+In \cite{bhalgat2013optimal} a generalization of the multiplicative weight update method previously discussed is presented.
+Specifically they are generalizing the application to solving linear inequality feasibility presented in class.
+Their generalization allows different linear constraints in each round, and is now able to bound the average violation of a constraint in any given round, not just the total at the end.
+They omit their proofs, claiming they follow simply from the original analysis of \cite{arora2012multiplicative}.
+I intend to fill in this analysis in my report.
+\paragraph{Designing Multi-Unit Bayesian Auctions}
+The motivation for the generalization above \cite{bhalgat2013optimal} is to use multiplicative weights at the heart of an algorithm for designing approximately optimal multi-unit auction mechanisms.
+I will define all the necessary terms to understand what they are trying to do, present their algorithm, and its analysis.
 
-%Question 1
-\section{}
-
-%Question 2
-\section{}
-
-%Question 3
-\section{}
-\paragraph{}
-I collaborated with Sharat on this problem.
-\paragraph{}
-Let $\cA$ denote our black-box algorithm. Let $f$ be at $\beta$-smooth function. Given an initial point $x_0$ and desired accuracy $\epsilon > 0$ our algorithm does:
-\begin{enumerate}
-\item If $||x_0 - x^*||_2^2 = 0$ return $x_0$.
-\item Let $\epsilon' = \epsilon/2$ and set $\alpha = \frac{\epsilon}{||x_0 - x^*||_2^2}$.
-\item Let $g$ be the function given by $g(x) := f(x) + \frac{\alpha}{2}||x - x_0||_2^2$.
-\item Run $\cA$ on $g$ with accuracy $\epsilon'$ and starting point $x_0$ returning a point $x'$.
-\item Return solution $x'$.
-\end{enumerate} 
-Our algorithm assumes that the value of $||x_0 - x^*||_2$ is known in advance. In the trivial case where we know $x^*$ the algorithm returns it in step $1$. So we may assume that $x_0 \neq x^*$.
-\begin{lemma}\label{lemma:convex-smooth}
-The function $g$ is $\alpha$-strongly convex and $(\beta + \alpha)$-smooth.
-\end{lemma}
-\begin{proof}
-The Hessian of $g$ is:
-$$\nabla^2 g = \nabla^2 f + \alpha I.$$
-Since we know $f$ is $\beta$-smooth we have
-$$0I \preccurlyeq \nabla^2 f \preccurlyeq \beta I.$$
-Since adding scalar multiples of $I$ only shifts eigenvalues we observe
-$$\alpha I \preccurlyeq \nabla^2 g \preccurlyeq (\beta + \alpha) I.$$
-Hence the lemma holds as desired. 
-\end{proof}
-\paragraph{}
-By lemma \ref{lemma:convex-smooth} we know that $\cA$ can process our function $g$. It remains to verify the accuracy of the point returned and to bound the running time.
-\begin{lemma}\label{lemma:q3-accuracy}
-The point $x'$ returned by our algorithm satisfies
-$$f(x') - f(x^*) \leq \epsilon.$$
-\end{lemma}
-\begin{proof}
-Let $\tilde{x}$ be the optimal solution for $g$, By our guarantees on $\cA$:
-$$g(x') - g(\tilde{x}) \leq \epsilon'.$$
-Now observe that
-\begin{align*}
-f(x') - f(x^*) &= g(x') - \frac{\alpha}{2}||x'-x_0||^2_2 - g(x^*) +\frac{\alpha}{2}||x^*-x_0||_2^2 \\
-&\leq g(x') - g(x^*) + \frac{\alpha}{2}||x^*-x_0||_2^2 &\text{since $\frac{\alpha}{2}||x'-x_0||^2_2\geq 0$} \\
-&= g(x') - g(x^*) + \frac{\epsilon}{2} &\text{by our choice of $\alpha$} \\
-&\leq g(\tilde{x}) + \epsilon' -g(x^*) + \frac{\epsilon}{2} \\
-&\leq \epsilon' +\frac{\epsilon}{2} &\text{by optimality of $\tilde{x}$} \\
-&= \epsilon &\text{by our choice of $\epsilon'$}.
-\end{align*}
-\end{proof}
-\paragraph{}
-Therefore our algorithm operates properly and returns a point within the desired accuracy.
-It remains to verify the running time of our algorithm.
-This running time is entirely bounded by the operation of $\cA$ on $g$ with $\epsilon'$.
-By the guarantee on $\cA$ this gives a bound of:
-\begin{align*}
-O(\sqrt{\frac{\beta}{\alpha} + 1}\log (\frac{g(x_0)-g(\tilde{x})}{\epsilon'})) &= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{g(x_0)-g(\tilde{x})}{\epsilon})) &\text{dropping constants}\\
-&= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{f(x_0)-f(\tilde{x}) - \frac{\alpha}{2}||x_0-x^*||^2_2}{\epsilon})) &\text{expanding $g$}\\
-&= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{f(x_0)-f(\tilde{x})}{\epsilon})) &\text{dropping non-positive term} \\
-&= O(\sqrt{\frac{\beta}{\alpha}}\log (\frac{f(x_0)-f(x^*)}{\epsilon})) &\text{by optimality of $x^*$} \\
-&= O(\sqrt{\frac{\beta}{\epsilon}} ||x_0-x^*||_2\log (\frac{f(x_0)-f(x^*)}{\epsilon}))  &\text{expanding $\alpha$}.
-\end{align*}
-Thus we obtain the desired running time. Together with lemma \ref{lemma:q3-accuracy} this completes the problem.
+\paragraph{Competitive Online Algorithms via Primal-Dual Methods}
+In \cite{buchbinder2009design} a framework for designing online algorithms which are competitive (approximate the optimal solution of the offline version up to a given factor) is presented.
+The framework is based on the primal-dual method.
+Multiplicative weights comes in as the scheme for updating the dual variables, and thus our bounds arising from multiplicative weights can be used in the analysis of these algorithms.
+We provide the necessary background and study the general framework.
+\paragraph{Competitive Ad Auctions}
+We study the application of the Primal-Dual method in the design of competitive ad auctions as presented in \cite{buchbinder2007online}.
+We give an online primal dual algorithm for maximizing revenue and analyze its competitiveness.
+An elegant use of Strong Duality arises when we extend this to multi-slot auctions.
+\paragraph{Connections}
+We will conclude by trying to see if we can observe any connections between the two mechanism design algorithms for different auctions presented in the report.
+If we are lucky perhaps we can combine ideas to solve an auction design problem that shares features of both applications by plugging the generalized multiplicative weight update method into the primal-dual framework.
+\bibliography{references}
+\bibliographystyle{plain}
 \end{document}

### 2017-03-28 CO750-ApproximationAlgorithmsTopics/assignments/presentation.tex
last minute fixes

@@
 We extend the voting function by defining for all $\alert{x \in FM_G}$, for all $\alert{u,v' \in V(G)}$:
-$$vote_u(x,v) = \sum_{v\in V(G)} x_{uv} vote_u(v,v').$$ 
+$$vote_u(x,v) = \sum_{uv'\in \delta(u)} x_{uv'} vote_u(v',v) \text{ and } vote_u(v,x) = -vote_u(x,v).$$ 
 We extend the vote aggregation defining for all $\alert{x, y \in FM_{G}}$:
-$$\Delta(x,y) = \sum_{u \in V(G)} \sum_{uv, uv' \in \delta(u)} x_{uv}y_{uv'} vote_u(v,v').$$
+$$\Delta(x,y) = \sum_{u \in V(G)} \sum_{uv \in \delta(u)} x_{uv}vote_u(v,y).$$
 The polytope of all $\text{\alert{fractional popular matchings}}$ is 
@@
 In $\text{\alert{bipartite}}$ graphs, each $\alert{x \in FM_G}$ is a convex combination of matchings.
-But observe that each $\alert{x\in P_G}$ need not be a convex combination of $\text{\alert{popular}}$ matchings.
+But each $\alert{x\in P_G}$ need not be a convex combination of $\text{\alert{popular}}$ matchings.
 \begin{columns}[T] % align columns
@@
 \end{columns}
-Where $\alert{\ell(u)}$ is a last choice vertex for each $\alert{u}$ so that each vertex is fully matched in any fractional matching.
+Where $\alert{\ell(u)}$ is an appended last choice vertex for each $\alert{u}$.
 
@@
 It is straightforward to see that $\alert{(LP)}$ is self-dual (exercise).
-This implies, by $\text{\alert{complementary slackness}}$, that the covering constraint for $\alert{ab}$ is tight when $\alert{x_{ab} > 0}$ for any popular fractional matching $\alert{x}$.
+By $\text{\alert{complementary slackness}}$: the covering constraint for $\alert{ab}$ is tight when $\alert{x_{ab} > 0}$ for any popular fractional matching $\alert{x}$.
 
@@
 
-So the hypothesis says each $\text{\alert{popular matching}}$ is perfect and, in the fractional setting, for each $\alert{u\in A \cup B}$, $\alert{x_{u\ell(u)} = 0}$.
+Hypothesis says each $\text{\alert{popular matching}}$ is perfect and, in fractional setting, for each $\alert{u\in A \cup B}$, $\alert{x_{u\ell(u)} = 0}$.
 
@@
 $$x = t_1\chi(M^0) + (t_2 - t_1)\chi(M^1) + \dots + (1 - t_{k-1}) \chi(M^{k-1}).$$
-This constructive procedure runs in $\alert{O(|E| + |V|)}$ time.
 \end{frame}
@@
 Is there an algorithm for deciding if a general graph admits a $\text{\alert{popular matching}}$ under $\text{\alert{strict}}$ preferences?
+
+$$ $$
+Can we say anything better for $\text{\alert{roommates}}$ instances with a $\text{\alert{perfect stable matching}}$?
 \end{frame}

### 2017-03-28 CS798-ConvexityAndOptimization/assignments/assignment_02.tex
Problem 1

@@
 \section{}
-
+\paragraph{}
+Let $A \in \{0,1\}^{m\times n}$ be a matrix. Let $\mathbf{1}$ denote the all-ones vector where dimension is clear from context. Consider the following $(LP)$:
+\begin{align*}
+\min\ &\mathbf{1}^T x \\
+\text{s.t. } Ax &\geq \mathbf{1} \\
+x &\geq 0.
+\end{align*}
+\paragraph{}
+We intend to use the multiplicative weight update method to solve this linear program.
+So we will need an oracle function.
+Thinking ahead to our intention to use multiplicative weights within a binary search framework we will design our oracle with a parameter representing the intended upper bound on our linear program.
+Let $\mu \in [0,n]$ and define the oracle $\cO(\mu)$ as follows:
+\begin{enumerate}
+\item Suppose $p\in \R^m_+$ represents the probability distribution passed to the oracle.
+\item Let $a_j$ denote the $j$-th column of $A$. Compute
+$$j^* = \text{argmax}_{j \in [n]} \{p^Ta_j\}.$$
+\item If $p^Ta_{j^*} = 0$ return infeasible.
+\item Let $r = \frac{1}{p^Ta_{j^*}}$. If $r  > \mu$ return infeasible.
+\item Set $x_{j^*} = r$ and set $x_j = 0$ for all $j \neq j^*$.
+\item Return $x$.
+\end{enumerate}
+\begin{lemma}\label{lemma:oracle}
+Let $(LP)(p)$ be the linear program
+$$\min\{\mathbf{1}^Tx : p^TAx \geq 1 \text{ and } x \geq 0\}.$$
+If $\cO(\mu)$ returns infeasible then there is no feasible solution to $(LP)(p)$ of cost at most $\mu$.
+Otherwise $\cO(\mu)$ returns an optimal solution to $(LP)(p)$.
+The width of $\cO(\mu)$ is at most $n$.
+\end{lemma}
+\begin{proof}
+If $\cO(\mu)$ returns infeasible in step $3$ then $p^TA = 0$ and hence $(LP)(p)$ is infeasible.
+We will show that $\cO(\mu)$ returns an optimal solution to $(LP)(p)$ of cost $r$, and thus if $\cO(\mu)$ returns infeasible in step $4$ then by optimality there is no feasible solution to $(LP)(p)$ of cost  at most $\mu$.
+\paragraph{}
+To see the optimality of $x$ observe the dual of $(LP)(p)$:
+$$\max\{y : yp^Ta_j \leq 1 \forall j \in [n] \text{ and } y \geq 0\}.$$
+By our choice of $j^*$, setting $y = r = \frac{1}{p^Ta_{j^*}}$ satisfies all constraints of the dual to $(LP)(p)$ and thus by strong duality, $x$ is an optimal solution to $(LP)(p)$.
+\paragraph{}
+It remains to verify the width of $\cO(\mu)$. Consider of row $A_i$ of $A$. If $A_{ij^*} = 0$ then $|A_i x - 1 | = 1 \leq n$. If $A_{ij^*} = 1$ then
+$$|A_i x - 1| = |r - 1| \leq \mu -1 \leq n.$$
+Thus the width of $\cO(\mu)$ is at most $n$.
+\end{proof}
+We will use a version the procedure given in Lecture $9$ for solving linear programs via multiplicative weight update. Let $\cM(\mu)$ denote the procedure:
+\begin{enumerate}
+\item Let $p^{(1)} = \frac{1}{m} \mathbf{1}.$
+\item For $t = 1, \dots, T$
+	\begin{enumerate}
+	\item Use $\cO(\mu)$ with $p^{(t)}$ to find $x^{(t)}$ such that $(p^{(t)})^TA x^{(t)} \geq 1.$
+	\item For all $i$ set $m_i^{(t)} = \frac{A_ix^{(t)} - b_i}{n}.$
+	\item Use the multiplicative weight update method with $m_i^{(t)}$ to compute $p^(t+1)$.
+	\end{enumerate}
+\item Return $x = \frac{1}{T}\sum_{t=1}^T x^{(t)}.$
+\end{enumerate}
+We have the following Theorem from class, which by Lemma \ref{lemma:oracle} we can modify slightly with respect to upper bound $\mu$.
+\begin{theorem}\label{th:mwu}
+Let $\epsilon > 0$ be a given error parameter.
+Then $\cM(\mu)$ either finds a solution $x$ such that $A_ix\geq b_i - \epsilon$ for all $i$, or correctly concludes that the system has no feasible solution of cost at most $\mu$.
+The algorithm $\cM(\mu)$ is called at most $O(n^2\log (m) / \epsilon^2)$ times with $O(m)$ additional processing per call.
+\end{theorem}
+The theorem as stated in Lecture $9$ said that the oracle correctly concludes the system is infeasible, but by lemma \ref{lemma:oracle} this needs to be changed to infeasibility for $x$ of cost at most $\mu$.
+We note that feasibility is not affected by ignoring constraints $x\geq 0$ since our oracle always gives non-negative solutions and thus there average will be non-negative.
+\paragraph{}
+Now we want to use $\cM(\cdot)$ in a bineary search procedure to find a solution to $(LP)$ with additive error $\epsilon>0$. Consider the following algorithm denoted $\cA$:
+\begin{enumerate}
+\item Fix $\ell = 0$ and $u = n$. Set $x = null$ (denoting infeasible)
+\item While $\ell \neq u$ do:
+	\begin{enumerate}
+	\item Compute $\mu = \floor{\frac{\ell + u}{2}}$
+	\item Run $\cM(\mu)$ with error parameter $\epsilon$.
+	\item If $\cM(\mu)$ returns infeasible set $\ell = \mu$.
+	\item Otherwise let $x$ be the solution returned by $\cM(\mu)$ and set $u = \mu$.
+	\end{enumerate}
+\item Return $x$.
+\end{enumerate}
+\paragraph{}
+The binary search procedure $\cA$ will find the optimal $\mu$ for which $\cM(\mu)$ returns a feasible solution.
+Therefore $\cA$ returns an optimal solution to $(LP)$ and by Theorem \ref{th:mwu} the solution returned is infeasible by at most an additive error of $\epsilon$.
+The running time of binary search is $O(\log(n))$.
+Our oracle uses an additional $O(m)$ processing to find $j^*$.
+Therefore, along with the running time of $\cM(\mu)$, our algorithm runs in time
+$$O(\frac{m^2n^2\log(m)\log(n)}{\epsilon^2}).$$
 %Question 2
@@
 \paragraph{}
-Let $\cA$ denote our black-box algorithm. Let $f$ be at $\beta$-smooth function. Given an initial point $x_0$ and desired accuracy $\epsilon > 0$ our algorithm does:
+Let $\cA$ denote our black-box algorithm. Let $f$ be a $\beta$-smooth function. Given an initial point $x_0$ and desired accuracy $\epsilon > 0$ our algorithm does:
 \begin{enumerate}

### 2017-03-29 CS798-ConvexityAndOptimization/assignments/assignment_02.tex
q2 template

@@
 \section{}
+\paragraph{}
+Let $A \in \R^{m\times n}$ and let $i$-th row of $A$ be denoted by $a_i$.
+Let $b \in \R^m$.
+We assume that each entry of $A$ and $b$ has absolute value at most $c$.
+To solve the system of linear inequalities we can reformulate as
+$$\min_{x \in \R^n} f(x)$$
+where $f(x) :=\max_{1\leq i \leq m} (b_i - a_ix)$. We attempt to approximate $f(x)$ with the $0$-convex soft-max function (for some parameter $K$):
+$$g(x) := \frac{1}{K}\log \sum_{i=1}^m \exp (K(b_i - a_i x)).$$
+
+%TODO compute the Hessian and determine for which \beta is g \beta-smooth
+
+%TODO figure out which values of K and epilson' are needed so that we can find a point x with approximates f(x)-f(x^*) \leq \epsilon
+
+%TODO propose algorithm: run gradient descent on g with K and \epsilon'
+
+%TODO verify correctness and running time using theorem about gradient descent on \beta-smooth functions
 

### 2017-03-29 CS798-ConvexityAndOptimization/assignments/assignment_02.tex
A2 done

@@
 
-\lhead{2017-03-27}
+\lhead{2017-03-29}
 \rhead{William Justin Toth CS798-Convexity and Optimization Assignment 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
@@
 We assume that each entry of $A$ and $b$ has absolute value at most $c$.
+Let $a_0$ be a row of zeroes and let $b_0$ be zero.
 To solve the system of linear inequalities we can reformulate as
 $$\min_{x \in \R^n} f(x)$$
-where $f(x) :=\max_{1\leq i \leq m} (b_i - a_ix)$. We attempt to approximate $f(x)$ with the $0$-convex soft-max function (for some parameter $K$):
-$$g(x) := \frac{1}{K}\log \sum_{i=1}^m \exp (K(b_i - a_i x)).$$
-
+where $f(x) :=\max_{0 \leq i \leq m} (b_i - a_ix)$.
+We added $b_0 - a_0 x = 0$ to avoid the situation where the polytope described by $Ax \geq b$ is unbounded.
+In such a situation attempting this optimization would lead to something unbounded, which is undesirable.
+Now $f(x) \geq 0$, and for any feasible point $x^*$, $f(x^*) = 0$.
+Thus any feasible point is optimal for this choice of function $f$.
+We attempt to approximate $f(x)$ with the $0$-convex soft-max function (for some parameter $K$):
+$$g(x) := \frac{1}{K}\log (1 + \sum_{i=1}^m \exp (K(b_i - a_i x))).$$
+Using the inequalities for log-sum-exp we saw in lecture $2$ we see this function has the property that
+$$f(x) \leq g(x) \leq f(x) + \frac{\log(m + 1)}{k}.$$
+We let $\cA(\epsilon, \beta)$ denote the algorithm of running standard gradient descent on $g(x)$ with accuracy $\epsilon$ and step size $\eta = \frac{1}{\beta}$.
+We will use the theorem from class lecture $7$ on the convergence rate of gradient descent on $\beta$-smooth functions:
+\begin{theorem}\label{th:gd}
+If $f$ is $\beta$-smooth then $\cA(\epsilon, \beta)$ returns a point $x$ such that $f(x) -f(x^*) \leq \epsilon$, where $x^*$ minimizes $f$,  in running time $O(\frac{\beta}{\epsilon})$.
+\end{theorem}
+\paragraph{}
+We need to determine two things: first what $K$ and $\epsilon$ to choose so that optimizing $g$ gives us the desired accuracy with respect to $f$, and second what is an upper bound on $\beta$.
+\paragraph{}
+Let $\epsilon >0$ be our desired accuracy for approximating $Ax \geq b$.
+That is we want $x$ such that $Ax \geq b - \epsilon \textbf{1}$.
+Equivalently we want $x$ such that $f(x) - f(x^*) \leq \epsilon$ where $x^*$ minimizes $f$ (that is, $x^*$ is feasible for $Ax \geq b$).
+Suppose we run $\cA(\frac{\epsilon}{2}, \beta)$ on $g$ with some choice of $K$ and $\beta$ obtaining a point $x$.
+Say $\tilde{x}$ minimizes $g$.
+Then is we set $K = \frac{2\log(m+1)}{\epsilon}$ we obtain:
+$$f(x) -f(x^*) \leq g(x) -g(x^*) + \frac{\log(m+1)}{K} \leq g(x) - g(\tilde{x}) + \frac{\log(m+1)}{K} \leq \frac{\epsilon}{2} + \frac{\log(m+1)}{\frac{2\log(m+1)}{\epsilon}} = \epsilon.$$
+\paragraph{}
+Thus if we run $\cA(\frac{\epsilon}{2}, \beta)$, the $x$ returned will achieve our desired accuracy for the system of linear inequalities.
+It remains to bound $\beta$ so we can calculate the running time.
+We compute a generic entry of the gradient of $g$ (and then later an entry of the hessian).
+For ease of notation we let $z(x) \in \R^m$ we defined by $z_i(x) = \exp(b_i-a_ix)$ and let $S(x) = \textbf{1}^Tz(x) +1$.
+$$\nabla g(x)_j = \frac{\partial}{\partial x_j} g(x) = \frac{1}{S(x)} \sum_{i=1}^m z_i(x)a_{ij}.$$
+From this we compute a generic entry of the hessian:
+$$\nabla^2g(x)_{jk} = \frac{\partial^2}{\partial x_k \partial x_j} g(x) = \frac{K}{S(x)^2} [(\sum_{i=1}^m z_i(x)a_{ik})(z_i(x)a_{ij}) - S(x)\sum_{i=1}^m z_i(x)a_{ij}a_{ik}].$$ 
+We want to obtain a bound on the largest eigenvalue, $\lambda_1$, of $\nabla^2g(x)$. To do this we use the loose bound
+$$\lambda_1 \leq n \max_{j,k} |\nabla^2g(x)_{jk}|.$$
+Observe that
+\begin{align*}
+|\nabla^2g(x)_{jk}| &= |\frac{K}{S(x)^2} [(\sum_{i=1}^m z_i(x)a_{ik})(z_i(x)a_{ij}) - S(x)\sum_{i=1}^m z_i(x)a_{ij}a_{ik}]| \\
+&\leq \frac{K}{S(x)^2}[ |\sum_{i=1}^m z_i(x)a_{ik}| \cdot |\sum_{i=1}^m z_i(x)a_{ij}| + S(x) |\sum_{i=1}^m z_i(x)a_{ij}a_{ik}| ]&\text{note that $S(x) \geq 0$} \\
+&\leq \frac{K}{S(x)^2}[ c|\sum_{i=1}^m z_i(x)| \cdot c|\sum_{i=1}^m z_i(x)| + S(x) c^2|\sum_{i=1}^m z_i(x)| ] &\text{since $|a_{ij}| \leq c$}\\
+&\leq \frac{Kc^2}{S(x)^2}[2S(x)^2 ] \\
+&= 4c^2\log(m) &\text{by our choice of $K$}.
+\end{align*}
+Therefore $\lambda_1 \leq 4c^2 n\log(m)$. Hence we can run our algorithm with $\beta = 4c^2n\log(m)$. So the final running time is
+$$O(\frac{4c^2n\log(m)}{\frac{\epsilon}{2}}) = O(\frac{c^2n\log(m)}{\epsilon}).$$
 %TODO compute the Hessian and determine for which \beta is g \beta-smooth

### 2017-03-31 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
wrote up problem 2

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-04-07}
\rhead{W. Justin Toth CO750-Approximation Algorithms Assignment 3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}

%Question 1
\section{}

%Question 2
\section{}
\paragraph{}
Let $M$ be a set of machines and let $J$ be a set of jobs. For each $j \in J$ let $M_j \subseteq M$ denote the subset of machines $j$ can be processed on. Further for each $j \in J$, let $p_j \in \N$ denote the processing time of job $j$ on any machine in $M_j$. We consider the scheduling problem, which we will denote $(SP)$, that asks us to find a schedule minimizing the makespan: the maximum completion time of any machine provided we process all jobs. We will formalize this objective momentarily.
\paragraph{}
The order we process jobs on a machine is irrelevant in this problem, and we cannot split jobs in pieces to give to different machines. So a solution to $(SP)$ is an assignment $a: J \rightarrow M$. Let $A$ be the set of all feasible assignments. That is $A = \{ a : \forall j \in J, a(j) \in M_j\}.$ Then $(SP)$ can be stated as
$$\min_{a \in A} \max_{m \in M} \sum_{j : a(j) = m} p_j.$$
\paragraph{}
Given an instance of $(SP)$ and a tentative makespan value $s \in \N$ we can construct an instance of Unsplittable Flow, which we'll denote $(UF)$, which we can use to decide if $(SP)$ has an assignment yielding a makespan of value $s$. The vertex set will consist of a source $r$, a node $v_m$ for every $m \in M$, and a terminal node $t_j$ for every $j\in J$. The demands for each $t_j$ are the processing times $p_j$. The arc set consists of an arc $(r,v_m)$ for every $m \in M$. The capacity of each $(r,v_m)$ will be the makespan $s$. The arc set will also contain an arc $(v_m, t_j)$ for each $j \in J$ and $m \in M_j$. The capacity of each $(v_m, t_j)$ will be $p_j$. We will let $c$ denote the capacity function. Let $D$ be the resulting directed graph.
\begin{lemma}\label{lemma:sp-uf}
An instance of $(SP)$ has an assignment of makespan at most $s$ if and only if the corresponding instance of $(UF)$ has a solution. Moreover the translation between the two problems can be done in polynomial time.
\end{lemma}
\begin{proof}
\paragraph{$(\implies)$} Let $a$ be the assignment which yields a makespan of value at most $s$. For each $j \in J$ define $P_j = \{ (r,a(j)),(a(j),t_j)\}$. That is, $P_j$ is the path in $D$ (the directed graph representing $(UF)$) upon which we will route a flow of $p_j$. Each arc $(a(j), t_j)$ for $j \in J$ receives total flow $p_j$, hence respecting its capacity. Each arc $(r,m)$ for $m \in M$ receives $\sum_{j : a(j) = m}p_j$ total flow. Since $\sum_{j : a(j) = m} p_j \leq s$ the capacity of $(r,m)$ is respected. Thus we have a $(UF)$ solution.
\paragraph{$(\impliedby)$}
Let $\{P_j : j \in J\}$ be the set of paths given in the solution of $(UF)$ corresponding an instance of $(SP)$ and value $s$. We will construct an assignment $a$ for $(SP)$ that has makespan $s$. For each $j \in J$ set $a(j) = m$ if $(m,j) \in P_j$. Such an $m$ must exist, and is unique, by our construction of $D$. Since arcs incoming to $j$ are precisely those in $M_j$ this assignment is feasible. Now we consider the makespan of this assignment. Since the capacities on each arc $(r,m)$ for all $m \in M$ is respected by flow we have for all $m \in M$:
$$\sum_{j: a(j) = m} p_j = \sum_{j: (m,j) \in P_j} p_j = f(r,m) \leq c(r,m) = s.$$ 
Thus $a$ has makespan at most $s$. Clearly the translation given between the solutions of $(UF)$ and $(SP)$ can be done in polynomial time.
\end{proof}
\paragraph{}
In our algorithm we will also use the following critical lemma from class.
\begin{lemma}\label{lemma:ff-uf}
Given an instance of $(UF)$ with demands $d_1, \dots, d_k$, if $f$ is a fractional flow realizing the demands then there exists an unsplittable flow $\bar{f}$ realizing the demands such that $\bar{f}(e) \leq f(e) + d_{max}$ for all $e \in E$, where $d_{max} = \max_{i\in [k]} d_i$. Moreover this unsplittable flow can be found in polynomial time.
\end{lemma}
\paragraph{}
We give an approximation algorithm for $(SP)$. The algorithm performs a binary search to find the optimal makespan value. Let $\cA$ be the algorithm for $(SP)$ that follows:
\begin{enumerate}
\item Let $\ell = 0$ and let $u = \sum_{j \in J} p_j$.
\item While $\ell \neq u$ do:
	\begin{enumerate}
	\item Construct the corresponding $(UF)$ instance with value $s=\floor{\frac{\ell + u}{2}}$.
	\item Use a standard max flow algorithm to find a fractional flow $f$ on the $(UF)$ instance.
	\item If $f$ satisfies the demands for all $t_j$ then set $\ell = s$
	\item Otherwise set $u = s$.
	\end{enumerate}
\item Using lemma \ref{lemma:ff-uf} compute $\bar{f}$ from $f$.
\item Using lemma \ref{lemma:sp-uf} compute an assignment $a$ from $\bar{f}$ (with makespan at most $2s$). Return $a$.
\end{enumerate}
\begin{theorem}
The algorithm $\cA$ is a $2$-approximation algorithm for $(SP)$.
\end{theorem}
\begin{proof}
From lemma \ref{lemma:sp-uf} we know that $\cA$ will return a feasible solution. Each inner iteration of the While loop runs in polynomial time since we have polynomial time algorithms for max flow. Further we know from lemma \ref{lemma:ff-uf} that step $3$ can be done in polynomial time, and from lemma \ref{lemma:sp-uf} that step $4$ can be done in polynomial time. It remains to verify that the binary search procedure runs in polynomial time. For our choice of $\ell$ and $u$ the while loop will run in $O(\log(\sum_{j \in J} p_j))$ iterations. The question is whether this is polynomial in the number of bits used to represent our input.
\paragraph{}
We know each $p_j$ uses approximately $\log(p_j)$ bits to be represented in memory. Thus $\sum_{j \in J} \log(p_j)$ is a polynomial in the input size. But we may assume for the purpose of counting input size that each $p_j \geq 2$. We may do so because each $p_j$ uses at least one bit of memory to represent itself, yet $\log(1) = 0$. Now, if each $p_j \geq 2$ then $\Pi_{j\in J} p_j \geq \sum_{j \in J} p_j$. Thus $\sum_{j \in J} \log(p_j) \geq \log(\sum_{j \in J} p_j)$, and hence the binary search runs in polynomial time.
\paragraph{}
We have verified that $\cA$ runs in polynomial time returning a feasible solution. It remains to verify the approximation factor of $2$. Let $s^*$ be the makespan value of an optimal solution to $(SP)$. The instance of $(UF)$ constructed in the last iteration of the binary search has the maximal $s$ such that a fractional flow is present on $(UF)$ satisfying all the demands. Since the existence of an unsplittable flow implies the existence of a fractional flow satisfying the demands, by lemma \ref{lemma:sp-uf} using $s^*$, $$s \leq s^*.$$
\paragraph{}
By lemma \ref{lemma:ff-uf}, $\cA$ returns a solution of makespan $2s$. This follows since the maximum capacity of an arc in our instance of $(UF)$ is $s$. Hence we return a solution of value
$$2s \leq 2s^*.$$
Therefore the approximation factor holds.
\end{proof}
%Question 3
\section{}

%Question 4
\section{}
\end{document}

### 2017-04-01 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
q1 write up

@@
 \section{}
-
+\paragraph{}
+Let $G=(V,E)$ be a metric graph with costs $c: E \rightarrow \R_+$.
+Let $T \subseteq V$ be a subset of nodes.
+Let $\equiv_2$ denote equivalence modulo $2$ (that is $a \equiv_2 b$ if and only if $\exists k \in \Z$ such that $a = b + 2k$).
+Suppose that $|T| \equiv_2 0$.
+\paragraph{}
+The minimum cost connected $T$-Join problem asks us to find a minimum cost multiset $F$ of edges such that the set of nodes with odd degree in the multigraph $H=(V,F)$ is exactly $T$ and $H$ is connected.
+\paragraph{}
+Let $\cA$ denote the following algorithm for solving minimum cost connected $T$-Join on $G$:
+\begin{enumerate}
+\item Compute a minimum cost spanning tree on $G$. Call it $S$.
+\item Let $A \subseteq T$ be the set of nodes in $T$ of even degree in $S$.
+\item Let $B \subseteq V\backslash T$ be the set of nodes not in $T$ of odd degree in $S$.
+\item Let $T' = A \cup B$. Compute a minimum cost $T'$-join in $G$. Call it $J$.
+\item Return $S \cup J$.
+\end{enumerate}
+\begin{theorem}
+The algorithm $\cA$ is a $\frac{5}{3}$-approximation algorithm for the minium cost connected $T$-join problem.
+\end{theorem}
+\begin{proof}
+Standard algorithms for minimum cost spanning tree and $T$-join run in polynomial time.
+Hence $\cA$ is a polynomial time algorithm.
+The multigraph $(V, S\cup J)$ is connected since it contains a spanning tree $S$.
+Further by our choice of $A$ and $B$, the nodes in $T$ are precisely the nodes of odd degree in $(V, S \cup J)$.
+With respect to correctness of $\cA$, the only question which remains is that $|T'| \equiv_2 0$ so that a $T'$-join can be computed.
+\paragraph{}
+Let any $C \subseteq V$ let $C^o := \{ v \in C: d_S(v) \equiv_2 1 \}$ where $d_S$ denotes $|\delta(v) \cap S|$. Then by our choice of $A$ and $B$: 
+$$|T| = |A| + |T^o| = |A| + |V^o\backslash B| = |A| + |V^o| - |B|.$$
+Since $|T| \equiv_2 0$ and $|V^o| \equiv_2 0$ (it is the set of odd degree vertices in $(V,S)$ so its cardinality is even):
+$$|A| \equiv_2 |B|,$$
+Thus $$|T'| \equiv_2 |A| + |B| \equiv_2 2|A| \equiv_2 0.$$
+So we have verified that $|T'|$ is even. Therefore the algorithm $\cA$ terminates in polynomial time returning a feasible solution to minimum cost connected $T$-Join.
+\paragraph{}
+We now verify the approximation factor. Let $J^*$ denote an optimal minimum cost connected $T$-join. Since $(V,J^*)$ is connected, $J^*$ contains a spanning tree. Therefore 
+$$c(S) \leq c(J^*).$$
+Now since $(V,S)$ is connected it contains a $T'$-Join.
+Let $J_1$ be such a $T'$-Join.
+Also since $(V,J^*)$ is connected it contains a $T'$-Join.
+Let $J_2$ be such a $T'$-Join.
+We claim that $J_3 := S\backslash J_1 \cup J^* \backslash J_2$ is a $T'$-join.
+To verify this consider a vertex $v \in T'$.
+We will study the parity of $d(v)$ with respect various sets mentioned above.
+Since $J_1$ and $J_2$ are $T'$-Joins we have
+$$d_{J_1}(v) \equiv_2 d_{J_2}(v) \equiv_2 0.$$
+Thus
+$$d_{J_3}(v) \equiv_2 d_S(v) + d_{J^*}(v).$$
+Since $J^*$ is a $T$-Join we have
+$$d_{J^*}(v) \equiv_2 \begin{cases}
+1, &\text{if } v \in V\backslash T \\
+0, &\text{if } v \in T.
+\end{cases}$$
+By our choice of $T'$, $v$ is either in $A\subseteq T$ or $B\subseteq V\backslash T$. So the case breakdown for $d_{J^*}(v)$ is more precisely:
+ $$d_{J^*}(v) \equiv_2 \begin{cases}
+1, &\text{if } v \in B \\
+0, &\text{if } v \in A.
+\end{cases}$$
+By our choice $A$ and $B$ we have
+$$d_{S}(v) \equiv_2 \begin{cases}
+1, &\text{if } v \in B \\
+0, &\text{if } v \in A.
+\end{cases}$$
+Now observe that in either case ($v \in A$ or $v\in B$)
+$$d_{J_3}(v) = d_S(v) + d_{J^*}(v) = 0.$$
+Thus $J_3$ is a $T'$-Join as claimed.
+\paragraph{}
+Since $J_1$, $J_2$, and $J_3$ are $T'$-Joins, and $J$ is a minimum cost $T'$-Join,
+$$3c(J) \leq c(J_1) + c(J_2) + c(J_3).$$
+By our choices of $J_1$, $J_2$, and $J_3$ their union of edges is a subset of $S \cup J^*$. So we have
+$$c(J_1) + c(J_2) + c(J_3) \leq c(S) + c(J^*) \leq 2c(J^*).$$
+Combining inequalities we see that
+$$3c(J) \leq 2c(J^*).$$
+Therefore 
+$$c(S) + c(J) \leq c(J^*) + \frac{2}{3}c(J^*) = \frac{5}{3}c(J^*)$$
+and so the approximation factor holds as desired.
+\end{proof}
 %Question 2

### 2017-04-02 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
3a

@@
 \section{}
+\paragraph{}
+In an instance of $(WTAP)$ we are given a tree $G=(V,E)$ and a set $L \subseteq V\times V$, called ``links".
+We are also given a cost function $c: L \rightarrow \R_+$.
+The goal is to find a minimum cost subset $\bar{L} \subseteq L$ such that $(V, E \cup \bar{L})$ is $2$-edge-connected.
+%Part (a) of Q3
+\subsection{a}
+We are given the following algorithm for $(WTAP)$, as usual we will denote it $\cA$:
+\begin{enumerate}
+\item Select an arbitrary $r \in V$ as the root. Orient $E$ towards $r$, and set the $w(a) = 0$ for all the resulting arcs $a$.
+\item For each $uv \in L$ do:
+	\begin{enumerate}
+	\item Let $p$ be the lowest-common-ancestor of $u$ and $v$ on $G$.
+Add to $G$ the arcs $(p,u)$ and $(p,v)$ with $w(p,u) = w(p,v) = c(uv)$.
+Note that if $p=u$ or $p=v$ we only add one arc).
+We say the arcs $(p,u)$ and $(p,v)$ ``correspond" to the link $uv$.
+	\end{enumerate}
+\item Compute a minimum weight $out$-arborescence $R$ rooted at $r$ in the resulting digraph from steps $1$ and $2$.
+\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that correponds to $\ell$.
+\end{enumerate}
 
+\begin{lemma}\label{lemma:3a-poly}
+The algorithm $\cA$ returns a feasible solution for $(WTAP)$ on $G$ with links $L$ in polynomial time.
+\end{lemma}
+\begin{proof}
+First we check that the algorithm terminates in polynomial time.
+Step $1$ can be done in $O(|E|)$ time.
+Step $2$ uses $O(|L|)$ iterations, and each iteration can use a depth-first-search to find the lowest-common ancestor in polynomial time.
+Hence Step $2$ runs in polynomial time.
+We have seen an exact polynomial time algorithm in class for computing a minimum weight arborescence.
+Hence Step $3$ runs in polynomial time.
+Step $4$ only necessitates iterating over links of $L$ and arcs of $R$ and thus runs in polynomial time.
+Hence $\cA$ terminates in polynomial time.
+\paragraph{}
+Consider the solution $\bar{L}$ returned by the algorithm.
+Let $up \in E$ where $p$ is the parent of $u$ with respect to the orientation in Step $1$.
+Let $G_u$ be the connected component of $G-up$ which contains $u$ and let $G_p$ be the connected component of $G-up$ which contains $p$.
+To verify that $(V, E \cup \bar{L})$ is $2$-edge-connected we will show that there exists $\ell \in \bar{L}$ with one endpoint in $V(G_u)$ and one in $V(G_p)$.
+\paragraph{}
+Consider the directed path $P$ from $r$ to $u$ in the arborescence $R$ computed in Step $3$.
+Let $g$ be the last ancestor of $u$ on $P$ before reaching $u$.
+Then $(g,c) \in P$ where $c$ is either a child of $u$ or $c=u$.
+By our construction, since $(g,c)$ travels from a vertex closer to the root to a vertex farther from the root, $(g,c)$ corresponds to some link $\ell \in \bar{L}$.
+The endpoint $c$ of $\ell$ lies $G_u$.
+The other endpoint, say $v$, has their lowest common ancestor with $c$ being $g$.
+Thus $v \not \in V(G_u)$ and therefore $v \in V(G_p)$.
+Therefore $\ell$ is a link connected $G_u$ to $G_p$, and thus $(V, E \cup \bar{L})$ is $2$-edge-connected.
+\end{proof}
+\begin{theorem}
+The algorithm $\cA$ is $2$-approximation algorithm for $(WTAP)$.
+\end{theorem}
+\begin{proof}
+By lemma \ref{lemma:3a-poly} it remains to verify the approximation factor.
+Let $L^* \subseteq L$ be an optimal solution to $(WTAP)$ on $G$.
+Let $D$ denote the directed graph which results from the construction in Steps $1$ and $2$ of $\cA$.
+We will construct from $L^*$ an arc set $R^*$ of cost
+$$c(R^*)=2c(L^*)$$
+which contains at $out$-arborescence rooted at $r$ on $D$.
+Then, since $R$ is a minimum weight $out$-arborescence on $G$ with $c(\bar{L}) = c(R)$ this will suffice to prove the approximation factor as $c(R) \leq c(R^*) = 2c(L^*)$.
+\paragraph{}
+For each $\ell \in L^*$ add the arcs which correspoind to $\ell$ to $R^*$.
+Now we have $c(L^*) = 2c(R^*)$.
+It remains to fix connectivity in $R^*$ so that there is a directed path from $r$ to each vertex.
+We will fix it by adding $0$ cost arcs.
+For each $u \in V\backslash \{r\}$ do the following.
+Let $p$ be the parent of $u$ with respect to the orientation of step $1$.
+Let $G_u$ be the connected component of $G-up$ which contains $u$ and let $G_p$ be the connected component of $G-up$ which contains $p$.
+Since $L^*$ is feasible for $(WTAP)$ there exists some $ab \in L^*$ such that $a \in V(G_u)$ and $b \in V(G_p)$.
+Now $a$ is a descendant of $u$.
+Thus there exists a $0$-cost directed path from $a$ to $u$.
+Add this path to $R^*$.
+Note this does not change $c(R^*)$.
+\paragraph{}
+We claim that after doing this for each vertex, the connectivity requirement is satisfied.
+Let $u \in V$.
+We proceed by induction on the distance from $u$ to the root.
+In the base case $r=u$ and so there trivially is a path from $r$ to $u$ in $R^*$.
+Now suppose $u$ is at least at distance $1$ from $r$, and for all ancestors $u$ there a directed path from $r$ to $u$'s ancestor using arcs of $R^*$.
+Let $p$, $a$, and $b$ be as in the previous paragraph with respect to $u$.
+Since $a$ is in $G_u$ and $b$ is in $G_p$ the lowest common ancestor of $a$ and $b$ is some vertex $g$ that is an ancestor of $u$.
+By induction there is a directed path $P_{rg}$ from $r$ to $g$ using arcs of $R^*$.
+By construction the arc $(g,a)$ is in $R^*$.
+So $P_{rg}$, $(g,a)$, and the $0$-cost path from $a$ to $u$ added to $R^*$ in the previous paragraph forms a directed path from $r$ to $u$ in $R^*$.
+Hence the claim holds.
+\paragraph{}
+So $R^*$ contains a directed path from $r$ to every vertex of $D$. Therefore $R^*$ contains an $out$-arborescence, and as argued in the first paragraph this suffices to prove the approximation factor.
+\end{proof}
+%Part b of Q3
+\subsection{b}
+\paragraph{}
+Now we change $(WTAP)$ so that the requirement is $(V, E\cup \bar{L})$ is $2$-vertex-connected.
 %Question 4

### 2017-04-03 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
trying 2b

@@
 \item Compute a minimum weight $out$-arborescence $R$ rooted at $r$ in the resulting digraph from steps $1$ and $2$.
-\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that correponds to $\ell$.
+\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that corresponds to $\ell$.
 \end{enumerate}
@@
 \paragraph{}
-For each $\ell \in L^*$ add the arcs which correspoind to $\ell$ to $R^*$.
+For each $\ell \in L^*$ add the arcs which correspond to $\ell$ to $R^*$.
 Now we have $c(L^*) = 2c(R^*)$.
@@
 Now we change $(WTAP)$ so that the requirement is $(V, E\cup \bar{L})$ is $2$-vertex-connected.
+Since links between adjacent vertices do not help $2$-vertex-connectivity we may assume there are no such links, otherwise just discard them before starting the algorithm.
+We will modify the algorithm $\cA$ as follows, the big changes are that we take a leaf to be the root, and we ``shorten" the downward arcs:
+\begin{enumerate}
+\item Select an arbitrary leaf $r \in V$ as the root. Orient $E$ towards $r$, and set the $w(a) = 0$ for all the resulting arcs $a$.
+\item For each $uv \in L$ do:
+	\begin{enumerate}
+	\item Let $p$ be the lowest-common-ancestor of $u$ and $v$ on $G$.
+Add to $G$ the arcs $(p,u)$ and $(p,v)$ with $w(p,u) = w(p,v) = c(uv)$.
+Note that if $p=u$ or $p=v$ we only add one arc).
+We say the arcs $(p,u)$ and $(p,v)$ ``correspond" to the link $uv$.
+	\item For each downward arc $(p,u)$ do: if $p$ is not a leaf let $c_p$ be the child of $p$  that succeeds $p$ on the path to $u$ (otherwise let $c_p = p$).
+If $u$ is not a leaf let $p_u$ be the parent of $u$ (otherwise let $p_u = u$).
+Replace the arc $(p,u)$ with the arc $(c_p, p_u)$ with the same weight, and same corresponding link as $(p,u)$.
+	\end{enumerate}
+\item Compute a minimum weight $out$-arborescence $R$ rooted at $r$ in the resulting digraph from steps $1$ and $2$.
+\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that corresponds to $\ell$.
+\end{enumerate}
+\begin{lemma}\label{lemma:3b-poly}
+The algorithm $\cA$ returns a feasible solution for $(WTAP)$ on $G$ with links $L$ in polynomial time.
+\end{lemma}
+\begin{proof}
+As argued in lemma \ref{lemma:3a-poly} the algorithm $\cA$ runs in polynomial time, it is clear that the additional processing runs in polynomial time.
+To verify feasibility suppose for a contradiction that the graph $H := (V, E \cup \bar{L})$ has a vertex $p$ whose deletion disconnects the graph.
+Let $g$ be the parent of $p$.
+The vertex $g$ exists since we choose $r$ to be a leaf and hence $p \neq r$.
+Let $u_1, \dots, u_k$ be the children of $p$.
+For each $v \in V$, let $C(v)$ denote the connected component of $H-p$ containing vertex $v$.
+Let $A := \{u \in \{u_1, \dots, u_k\} : u \in C(g)\}$ be the set of children of $p$ which are connected to $g$ after removing $p$.
+We claim that $A \neq \emptyset$.
+\paragraph{}
+Clearly $r \in C(g)$, as are all ancestors of $p$.
+Now since $R$ is an $out$-arborescence there exists a directed path from $r$ to $p \cup \bigcup_{i=1}^k u_i$.
+The arc such a path uses to enter $p \cup \bigcup_{i=1}^k u_i$ is a downward arc from some ancestor of $p$, $a$, to either $p$ or a descendant of $p$, $d$.
+If the arc is of the form $(a,d)$ then the corresponding link connects an ancestor of $p$ to some descendant of a child $u_i$.
+Hence $C(g)$ is connected to $C(u_i)$.
+That is $u_i \in A$ and hence $A \neq \emptyset$.
+In the case the arc is of the form $(a,p)$, then by our construction of $(a,p)$ in step $2(b)$ there is a link in $\bar{L}$ connecting an ancestor of $p$ to some child of $p$, $u_i$.
+Thus $C(g)$ is connected to $C(u_i)$.
+Therefore our claim that $A$ is not empty holds.
+Let $\bar{A} = \{u_1, \dots, u_k\} \backslash A$.
+By our contradiction assumption, $\bar{A}\neq \emptyset$ as otherwise removing $p$ does not disconnect $H$.
+\paragraph{}
+Now, again since $R$ is an $out$-arborescence there exists a directed path from $r$ to $\bigcup_{u \in \bar{A}} C(u)$.
+The directed arc which enters $\bigcup_{u \in \bar{A}} C(u)$ is either a downward ar
+\paragraph{} 
+\end{proof}
 %Question 4

### 2017-04-03 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
3b starting correct argument

--- 2017-04-02 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
@@
 Now we change $(WTAP)$ so that the requirement is $(V, E\cup \bar{L})$ is $2$-vertex-connected.
+Since links between adjacent vertices do not help $2$-vertex-connectivity we may assume there are no such links, otherwise just discard them before starting the algorithm.
+We will modify the algorithm $\cA$ as follows, the big changes are that we take a leaf to be the root, and we ``shorten" the downward arcs, and we add ``cross" arcs:
+\begin{enumerate}
+\item Select an arbitrary leaf $r \in V$ as the root. Orient $E$ towards $r$, and set the $w(a) = 0$ for all the resulting arcs $a$.
+\item For each $uv \in L$ do:
+	\begin{enumerate}
+	\item Let $p$ be the lowest-common-ancestor of $u$ and $v$ on $G$.
+Add to $G$ the arcs $(p,u)$ and $(p,v)$ with $w(p,u) = w(p,v) = c(uv)$.
+Note that if $p=u$ or $p=v$ we only add one arc).
+We say the arcs $(p,u)$ and $(p,v)$ ``correspond" to the link $uv$.
+	\item If neither $u$ nor $v$ are each others' ancestor then  add to $G$ the arcs $(u,v)$ and $(v,u)$.
+Set $w(u,v) = w(v,u) = c(uv)$.
+We also say the arcs $(u,v)$ and $(v,u)$ ``correspond to the link $uv$.
+Note that at most four arcs are generated between this step and the previous step for each link.
+	\item For each downward arc ($p$ is an ancestor of $u$) $(p,u)$ do: if $p$ is not a leaf let $c_p$ be the child of $p$  that succeeds $p$ on the path to $u$ (otherwise let $c_p = p$).
+Replace the arc $(p,u)$ with the arc $(c_p, u)$ with the same weight, and same corresponding link as $(p,u)$.
+	\end{enumerate}
+\item Compute a minimum weight $out$-arborescence $R$ rooted at $r$ in the resulting digraph from steps $1$ and $2$.
+\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that corresponds to $\ell$.
+\end{enumerate}
+\begin{lemma}\label{lemma:3b-poly}
+The algorithm $\cA$ returns a feasible solution for $(WTAP)$ on $G$ with links $L$ in polynomial time.
+\end{lemma}
+\begin{proof}
+As argued in lemma \ref{lemma:3a-poly} the algorithm $\cA$ runs in polynomial time, it is clear that the additional processing runs in polynomial time.
+To verify feasibility suppose for a contradiction that the graph $H := (V, E \cup \bar{L})$ has a vertex $p$ whose deletion disconnects the graph.
+Let $g$ be the parent of $p$.
+The vertex $g$ exists since we choose $r$ to be a leaf and hence $p \neq r$.
+Let $u_1, \dots, u_k$ be the children of $p$.
+For each $v \in V$, let $C(v,p)$ denote the connected component of $H-p$ containing vertex $v$.
+Let $A := \{u \in \{u_1, \dots, u_k\} : u \in C(g,p)\}$ be the set of children of $p$ which are connected to $g$ after removing $p$.
+We claim that $A \neq \emptyset$.
+\paragraph{}
+Clearly $r \in C(g,p)$, as are all ancestors of $p$.
+Now since $R$ is an $out$-arborescence there exists a directed path from $r$ to $\bigcup_{i=1}^k u_i$.
+The arc such a path uses to enter $\bigcup_{i=1}^k u_i$ is a downward arc, from either $p$ or  some ancestor of $p$, $a$, to a descendant of $p$, $d$.
+Let $u$ be a child of $p$ such that $d \in C(u,p)$.
+If the arc is of the form $(a,d)$ then the corresponding link  in $\bar{L}$ connects a vertex of $C(g,p)$ to some descendant of a child $u$.
+Hence $C(g,p)$ is connected to $C(u)$.
+That is $u \in A$ and hence $A \neq \emptyset$.
+In the case the arc is of the form $(p,d)$, then by step $2(c)$, the arc $(p,d)$ arose from a downward arc $(g,d)$.
+Thus there is a link in $\bar{L}$ connecting a vertex of $C(g,p)$ to $d$.
+Thus $C(g)$ is connected to $C(u)$.
+Therefore our claim that $A$ is not empty holds.
+\paragraph{}
+ Let $\bar{A} = \{u_1, \dots, u_k\} \backslash A$.
+By our contradiction assumption, $\bar{A}\neq \emptyset$ as otherwise removing $p$ does not disconnect $H$.
+Now, again since $R$ is an $out$-arborescence there exists a directed path from $r$ to $\bigcup_{u \in \bar{A}} C(u)$.
+The directed arc which enters $\bigcup_{u \in \bar{A}} C(u)$ is either a downward arc or a cross arc arising from step $2(b)$.
+In the case it is a downward arc, by the same argument as the previous paragraph the corresponding link in $\bar{L}$ connects some $C(v, p)$ for $v \in \bar{A}$ to some $C(s,p)$ for $s \in A \cup \{g\}$.
+But this contradicts the definition of $\bar{A}$.
+In the case of a cross arc, the head is necessarily in $C(g,p)$ and the tail in some $C(v,p)$ for $v \in \bar{A}$.
+Thus, as in the previous case, the corresponding link in $\bar{L}$ connects some $C(v, p)$ for $v \in \bar{A}$ to some $C(s,p)$ for $s \in A \cup \{g\}$.
+In either case a contradiction is reached since the corresponding components to vertices in $\bar{A}$ cannot be connected to $C(g,p)$ by definition.
+Therefore no vertex exists whose deletion disconnects the graph.
+\paragraph{} 
+\end{proof}
 %Question 4

### 2017-04-03 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
Wrote up 3b

@@
 We will construct from $L^*$ an arc set $R^*$ of cost
-$$c(R^*)=2c(L^*)$$
+$$c(R^*)\leq 2c(L^*)$$
 which contains at $out$-arborescence rooted at $r$ on $D$.
-Then, since $R$ is a minimum weight $out$-arborescence on $G$ with $c(\bar{L}) = c(R)$ this will suffice to prove the approximation factor as $c(R) \leq c(R^*) = 2c(L^*)$.
+Then, since $R$ is a minimum weight $out$-arborescence on $G$ with $c(\bar{L}) = c(R)$ this will suffice to prove the approximation factor as $c(R) \leq c(R^*) \leq 2c(L^*)$.
 \paragraph{}
@@
 Now we change $(WTAP)$ so that the requirement is $(V, E\cup \bar{L})$ is $2$-vertex-connected.
-Since links between adjacent vertices do not help $2$-vertex-connectivity we may assume there are no such links, otherwise just discard them before starting the algorithm.
+We may assume we are not in the degenerate case where we have a graph consisting of only two vertices connected by an edge.
+Such a tree is $2$-vertex-connected by default.
 We will modify the algorithm $\cA$ as follows, the big changes are that we take a leaf to be the root, and we ``shorten" the downward arcs, and we add ``cross" arcs:
@@
 Therefore no vertex exists whose deletion disconnects the graph.
+\end{proof}
+\begin{theorem}
+The algorithm $\cA$ is $2$-approximation algorithm for $(WTAP)$.
+\end{theorem}
+\begin{proof}
+By lemma \ref{lemma:3b-poly} it remains to verify the approximation factor.
+Let $L^* \subseteq L$ be an optimal solution to $(WTAP)$ on $G$.
+Let $D$ denote the directed graph which results from the construction in Steps $1$ and $2$ of $\cA$.
+We will construct from $L^*$ an arc set $A^*$ of cost
+$$c(A^*)\leq 4c(L^*)$$
+which contains at $out$-arborescence $R^*$ rooted at $r$ on $D$ with cost $c(R^*) \leq 2c(L^*)$.
+Then, since $R$ is a minimum weight $out$-arborescence on $G$ with $c(\bar{L}) = c(R)$ this will suffice to prove the approximation factor as $c(R) \leq c(R^*) \leq 2c(L^*)$.
 \paragraph{}
+For each $\ell \in L^*$ add the arcs which correspond to $\ell$ (with respect to Step $2$) to $A^*$.
+At most four arcs are added so we have $c(A^*) \leq 4c(L^*)$.
+Suppose we also add to $A^*$ the $0$-cost arcs formed in Step $1$ of $\cA$.
+We want to show that every vertex is reachable from $r$ by a directed path in $A^*$.
+Then we can compute an arborescence on $A^*$ and use that to bound the optimal solution.
+\paragraph{}
+We claim that for every $r' \in V$, $r'$ can reach, by a directed path using arcs of $A^*$, every vertex in the subtree rooted at $r'$.
+We proceed by induction on the number of vertices of the subtree rooted at $r'$.
+In the base case the subtree is of size $1$, and hence $r'$ is a leaf and the claim is immediate.
+Suppose the claim holds for all subtrees with less vertices that the subtree rooted at $r'$.
+Let $d$ be a descendant of $r'$.
+We have two cases: either $r' = r$ or not.
+In the case $r'\neq r$, the deletion of $r'$ disconnects $G$.
+Let $u_1, \dots, u_k$ be the children of $r'$.
+Since $r'$ is not the root $r$, $r'$ has some parent $p$.
+Let $C_{u_1}, \dots, C_{u_k}, C_p$ denote the respective connected components of $G$ after deleting $r'$.
+Suppose without loss of generality that $d \in C_{u_1}$.
+Since $(V, E \cup L^*)$ is $2$-vertex-connected, if we think of $C_{u_1}, \dots, C_{u_k}, C_p$ as vertices, there is a subset of links $T \subseteq L^*$ which forms a spanning tree on $C_{u_1}, \dots, C_{u_k}, C_p$.
+There exists a path from $C_{u_1}$ to $C_g$  using links of $T$.
+We may assume without loss generality that this path is of the form $C_{u_1}, C_{u_2}, \dots, C_{u_k}, C_p$ since we may simply ignore connected components which do not join this path (decreasing $k$) and relabelling the children.
+\paragraph{}
+Now, for $i \in [k-1]$ the link between $C_{u_{i}}$ and $C_{u_{i+1}}$ induces a directed arc in $A^*$ from $C_{u_{i+1}}$ to $C_{u_i}$ by Step $2(b)$.
+Call this arc $t_i$, The link between $C_{u_k}$ and $C_p$ induces a directed arc from an ancestor of $u_k$ to a vertex of $C_{u_k}$ by Step $2(a)$ and $2(c)$.
+Call this arc $t_k$.
+Now we describe the directed path to $d$.
+From $r'$ take the up-arcs of $0$-cost to the head of $t_k$ (note the head of $t_k$ may be $r'$, in which case take an empty path).
+Then follow $t_k$ to a vertex of $C_{u_k}$.
+We use the following process to traverse $C_k$: Use the $0$-cost arcs to reach $u_k$.
+Use the inductive hypothesis to reach the head of $t_{k-1}$, and follow $t_{k-1}$ to a vertex of $C_{k-1}$.
+Repeat this process for each $C_i$ from $C_{k-1}$ to $C_1$.
+After doing so the directed path has reached the tail of $t_1$.
+Again follow the $0$-cost arcs up to $u_1$, and use induction to find a directed path from $u_1$ to $d$.
+All together this process yields a directed path from $r'$ to $d$.
+\paragraph{}
+In the induction step the case where $r' =r$ is simpler.
+Let $u$ be the child of $r'$.
+Then $d$ is a descendant of $u$ or equal to $u$.
+In either case, since $L^*$ is feasible there exists some link from $r$ to a descendant of $u$.
+This link induces an arc from $r$ to a descendant $d'$ of $u$ in $A^*$.
+Since $r$ is a leaf, the head of this arc is not-moved from $r$ in step $2(c)$.
+To reach $d$ from $r$, simply traverse the arc from $r$ to $d'$ then traverse the $0$-cost path from $d'$ up to $u$ and use the inductive hypothesis to find a directed path from $u$ to $d$.
+The resulting concatenated directed path connects $r$ to $d$.
+Hence by induction, the root $r$ is connected to every vertex of $D$ by a directed path using only arcs of $A^*$.
+\paragraph{}
+Therefore we can form a arborescence on $D$, say $R^*$ which is a subset of $A^*$.
+Choose the minimum cost such $R^*$.
+We claim that $R^*$ uses at most two arcs corresponding to any given link $uv \in L^*$.
+Suppose for a contradiction $R^*$ uses at least three arcs corresponding to $uv$.
+If $R^*$ uses both $(u,v)$ and $(v,u)$ then those two arcs form a directed cycle in $R^*$, contradicting that $R^*$ is acyclic.
+Hence $R^*$ uses one of $(u,v)$ or $(v,u)$, as well as the two downward arcs formed.
+Say without loss of generality $R^*$ uses $(u,v)$.
+Let $p$ be the lowest common ancestor of $u$ and $v$.
+Let $p_u$ and $p_v$ be the respective children of $p$ on the path to $u$ and $v$ respectively.
+So $R^*$ uses the arcs $(p_u,u)$ and $(p_v,v)$ as well as $(u,v)$ since $R^*$ uses at least three corresponding arcs.
+But then $R^*$ has two directed paths from $r$ to $v$.
+The first from $r$ to $p_u$, and then along $(p_u,u)$ and $(u,v)$ to reach $v$.
+The second along the directed path from $r$ to $p_v$ then along $(p_v, v)$ to reach $v$.
+So arc $(u,v)$ is not needed for connectivity in $R^*$, contradicting the optimality of $R^*$.
+\paragraph{}
+Therefore we have formed an arborescence $R^*$ such that $c(R^*) \leq 2c(L^*)$. Since $R$ is a minimum cost arborescence and $c(\bar{L}) = c(R)$ we have
+$$c(\bar{L}) = c(R) \leq c(R^*) \leq 2c(L^*).$$
+So the approximation factor holds as desired.
 \end{proof}

### 2017-04-03 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
starting 4

@@
 \section{}
+\paragraph{}
+We are given a bipartite graph $G=(A\cup B, E)$ where $A$ denotes the set of ``men" and $B$ denotes the set of ``women".
+Each vertex gives a preference order over their neighbouring vertices.
+The preference orders are strict for men, but may contain ties for women.
+We let $>_u$ denote the preference order for vertex $u$.
+That is, we say $w >_u v $ if $u$ prefers neighbour $w$ to neighbour $v$.
+We will use $\geq_u$, $=_u$, $<_u$, and $\leq_u$ in a similar manner.
+Let $N(u)$ denote the neighbourhood of vertex $u$.
+Let the first choice of vertex $u$ with be denoted $f(u)$.
+That is $f(u) \geq_u w$ for all $w \in N(u)$.
+Let $(SM)$ denote the problem of finding a maximum cardinality stable matching in $G$.
+\paragraph{}
+We assume that any vertex prefers to be matched to unmatched.
+Given a matching $M$, we let $M(u) := v$ iff $uv \in M$.
+If $u$ is unmatched in $M$ we let $M(u)$ denote some empty element which indicates that $u$ is unmatched.
+Let $V(M):= \{ u \in V(G): \exists v \in V(G), uv \in M\}$ be the set of matched vertices.
+We are given the following algorithm, which we will denote $\cA$:
+\begin{enumerate}
+\item Let $M = \emptyset$. Let $G' = G$. No $m \in A$ is a bachelor to start.
+\item While there exists an unmatched $m \in A$ with some women on his list do:
+	\begin{enumerate}
+	\item $m$ proposes to $w = f(m)$ with respect to $G'$. Let $m' = M(w)$.
+	\item If $m >_w m'$ then set $M = M\backslash \{m'w\} \cup \{mw\}$. Set $G' = G' - m'w$ (remove $w$ from $m'$'s list).
+	\item If $m =_w m'$ and $m$ is a bachelor then set $M = M\backslash \{m'w\} \cup \{mw\}$. Set $G' = G'-m'w$.
+	\item Else (both previous conditions fail) $w$ rejects $m$ and set $G' = G' -mw$ (remove $w$ from $m$'s list). 
+	\item If $N(m) = \emptyset$ for the first time, label $m$ a bachelor and add all edges in $\delta(m)$ with respect to $G$ back to $G'$.
+	\end{enumerate}
+\item Return $M$.
+\end{enumerate}
+\begin{lemma}\label{lemma:4poly}
+The algorithm $\cA$ returns a stable matching in polynomial time.
+\end{lemma}
+\begin{proof}
+The algorithm iterates over each man's list at most twice, and each iteration takes polynomial time, so $\cA$ runs in polynomial time.
+So it remains to verify that $\cA$ returns a stable matching.
+It is easy to see that $M$ is a matching since each vertex has at most one partner at any time in the operation of $\cA$.
+Suppose for a contradiction there exists some blocking pair $ab$ with $a \in A$ and $b \in B$.
+Since $b >_a M(a)$, whenever $a$ proposes to $b$, $a$ is rejected.
+Thus during the iterations $a$ proposes to $b$, $M(b) \geq_b a$.
+Since $b$ never releases a man for one she likes less, the quality of $b$'s partner from her perspective is non-decreasing each iteration.
+So since $M(b) \geq_b a$ when $a$ proposed to $b$ for the last time, this also holds at the end of $\cA$.
+But this contradicts $a>_b M(a)$ which comes from $ab$ being a blocking pair.
+Hence the matching $M$ returned is stable.
+\end{proof}
+\begin{theorem}
+The algorithm $\cA$ is a $\frac{3}{2}$-approximation algorithm for $(SM)$.
+\end{theorem}
+\begin{proof}
+By lemma \ref{lemma:4poly} it only remains to show the approximation factor holds.
+Let $M^*$ be an optimal stable matching solution to $(SM)$.
+Let $H = (V, M \Delta M^*)$.
+By standard matching theory, $H$ is a composition of paths and cycles.
+If we can show there are no $M$-exposing paths of lengths at most $3$ then the approximation ratio holds.
+This follows since we would have $$\frac{|M|}{|M^*} \leq \frac{|P \cap M|}{|P\cap M^*|} \leq \frac{2}{3},$$
+where $P$ is the shortest $M$-exposing path in $H$. The second inequality follows since $M$-exposing paths are of odd length in $H$, and so $|P| \geq 5$.
+\paragraph{}
+To complete the proof we show there are no $M$-exposing paths of length at most $3$.
+An $M$-exposing path of length $1$ would be a blocking pair against $M$, so by stability such a path does not exist.
+So suppose for a contradiction that there is an $M$-exposing path of length $3$.
+Say the path is $a_1b_1, b_1a_2,a_2b_2$ where $a_1b_1,a_2b_2 \in M^*$ and $b_1a_2 \in M$.
+Since $M^*$ is stable $a_1 \geq_{b_1} a_2$ or $b_2 >_{a_2} b_1$.
+\end{proof}
 \end{document}

### 2017-04-04 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
drafted A3

@@
 This follows since we would have $$\frac{|M|}{|M^*} \leq \frac{|P \cap M|}{|P\cap M^*|} \leq \frac{2}{3},$$
-where $P$ is the shortest $M$-exposing path in $H$. The second inequality follows since $M$-exposing paths are of odd length in $H$, and so $|P| \geq 5$.
+where $P$ is the shortest $M$-augmenting path in $H$. The second inequality follows since $M$-augmenting paths are of odd length in $H$, and so $|P| \geq 5$.
 \paragraph{}
-To complete the proof we show there are no $M$-exposing paths of length at most $3$.
-An $M$-exposing path of length $1$ would be a blocking pair against $M$, so by stability such a path does not exist.
-So suppose for a contradiction that there is an $M$-exposing path of length $3$.
-Say the path is $a_1b_1, b_1a_2,a_2b_2$ where $a_1b_1,a_2b_2 \in M^*$ and $b_1a_2 \in M$.
+To complete the proof we show there are no $M$-augmenting paths of length at most $3$.
+An $M$-augmenting path of length $1$ would be a blocking pair against $M$, so by stability such a path does not exist.
+So suppose for a contradiction that there is an $M$-augmenting path of length $3$.
+Say the path is $a_1b_1, b_1a_2,a_2b_2$ where $a_1b_1,a_2b_2 \in M^*$ and $b_1a_2 \in M$ (with $a_1,a_2 \in A$ and $b_1,b_2 \in B$).
 Since $M^*$ is stable $a_1 \geq_{b_1} a_2$ or $b_2 >_{a_2} b_1$.
+\paragraph{}
+If $b_2 >_{a_2} b_1$ then $a_2b_2$ blocks $M$, since $b_2$ is unmatched in $M$ and prefers to be matched.
+Similarly if $a_1 >_{b_1} a_2$ then $a_1b_1$ blocks $M$.
+So $a_1 =_{b_1} a_2$, and $b_1 >_{a_2} b_2$.
+We claim that $a_2$ proposes to $b_1$ at most once during $\cA$.
+Once a woman becomes matched in $\cA$ she never becomes unmatched.
+Thus since $b_2$ is unmatched at the end of $\cA$ and $b_2$ is acceptable to $a_2$, $a_2$ never ran through his whole proposal list during $\cA$.
+Therefore $a_2$ is accepted by $b_1$ in their first proposal round.
+Since we know women never free partners for someone worse during $\cA$ whoever $b_1$ is paired with at any stage of the algorithm she likes at most as much as $a_1$ and $a_2$.
+Since $b_1$ would reject $a_2$ otherwise, this implies that $b_1$ is matched to $a_2$ during the first stage of the algorithm, and in particular during $a_1$'s first proposal.
+Since $a_1$ is unmatched at the end of $\cA$, we know $a_1$ proposed to $b_1$ twice, and she rejected him twice.
+But during the second round $a_1$ became a bachelor such that $a_1 =_{b_1} a_2$, and so $\cA$ would match $a_1b_1$ during $a_1$'s second proposal to $b_1$.
+But this is a contradiction.
+Therefore $M$-augmenting paths of length at most three do not exist.
 \end{proof}

### 2017-04-04 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
Preliminaries written up

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{2017-04-04}
\rhead{W. Justin Toth CO750-Approximation Algorithms Report} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\paragraph{Paper Title:} Popularity, Mixed Matchings, and Self-duality
\paragraph{Authors:} Chien-Chun Huang and Telikepalli Kavitha

\section{Preliminaries}
\paragraph{}
We will begin by introducing the definitions of the main objects of study, and some useful notation. Then we will state the theorems to be shown in the later sections of this report.
\paragraph{}
As in most problems in the stable matching family we are given a graph $G=(V,E)$ where each vertex has ranking over its neighbours. We will assume throughout that these preference orders are strict, and a vertex prefers any neighbour to being unmatched. We will denote by $N(v)$ the set of neighbours of vertex $v$, and by $\delta(v)$ the set of edges incident with $v$. We will use the notation $>_v$ to describe $v$'s preference order over $N(v)$. Precisely for $u, w \in N(v)$, we write $u >_v w$ if $u$ is preferred by $v$ to $w$. We will use $\geq_v$, $\leq_v$, $=_v$, and $<_v$ analogously.
\paragraph{}
One way to think of $\textit{popular}$ matchings is as matchings which never lose an election to any other matching. Since we will speak often about partners in matchings, we introduce the notation: if $M$ is a matching then $M(u) = v$ provided $uv \in M$. For any $u \in V$ and $v,v' \in N(v)$ we define $vote_u(v,v')$ as
$$vote_u(v,v') := \begin{cases}
1, \text{ if } v >_u v' \\
-1, \text{ if } v <_u v' \\
0, \text{ otherwise (ie. } v = v').
\end{cases}$$
To decide the winner of the election we some the votes. The aggregate voting between two matchings $M$ and $M'$ is
$$\Delta(M, M') = \sum_{u \in V} vote_u(M(u), M'(u)).$$
Observe that $\Delta(M,M') \geq 0$ if at least as many vertices prefer their partner in $M$ to their partner in $M'$. A $\textit{popular}$ matching $M$ is  a matching where
$$\Delta(M, M') \geq 0$$
for any matching $M'$.
\paragraph{}
We are interested in the $\textit{max weight popular matching}$ problem in this report. That is, given a weight function $w: E \rightarrow \R_+$ find a popular matching $M$ on $G$ which maximizes $w(M) := \sum_{e \in M} w(e).$ A standard approach to such problems is to attempt to formulate them as linear programs. This will motivate our definitions of fractional popular matchings to follow. In this approach the major question is whether or not your linear programming formulation gives back integral solutions which can be used to solve the original combinatorial problem. In this paper we will prove that there is an integral formulation for a special case, and a $\frac{1}{2}$-integral formulation in general.
\paragraph{}
The fractional matching polytope $FM_{G}$ is defined as
$$FM_{G} := \{x \in \R_{\geq 0}^{E} : x(\delta(v)) \leq 1, \forall v \in V \}.$$
This object has been well-studied in combinatorial optimization. We want to study the analogous $\textit{popular fractional matching polytope}$ which we will call $P_G$. To formally define $P_G$ we will need to extend our notions related to voting as used in defining popular matchings. Let $x \in FM_G$. We define
$$vote_u(x, v') := \sum_{v \in V(G)} x_{uv} vote_u(v,v') = x(\delta^{>v'}(u)) - x(\delta^{<v'}(u)),$$
where $\delta^{>v'}(u) := \{vu \in \delta(u) : v >_u v'\}$ denotes the set of edges $u$ prefers to $uv'$, and $\delta^{<v'}(u)$ is defined analogously.  We define $vote_u(v',x) = -vote_u(x,v')$. Now we have a means of voting between two fractional matchings $x,y \in FM_G$:
$$vote_u(x,y) := \sum_{uv' \in \delta(u)} y_{uv'}vote(x,v'), $$
and the total votes:
$$\Delta(x,y) := \sum_{u \in V} vote_u(x,y).$$
Now we can formally describe $P_G$ as 
$$P_G = \{ x\in FM_G: \Delta(x,y) \geq 0, \forall y \in FM_G\}.$$
\paragraph{}
We finally have sufficient definitions to describe the results of this paper. In section \ref{sec:formulation} we describe a useful extended formulation which describes $P_G$ that will come into play when studying popular fractional matchings throughout this work. In section \ref{sec:special} we build up the following theorem
\begin{theorem}\label{th:special}
Let $G = (A \cup B, E)$ be a bipartite graph with strict preferences. If $G$ admits a perfect stable matching the $P_G$ is integral.
\end{theorem}
Building upon this theorem we show that $\frac{1}{2}$-integrality holds in general in section \ref{sec:general}.
\begin{theorem}
Let $G=(V,E)$ be an arbitrary graph with strict preferences. Then the extreme points of $P_G$ lie in $\{0,\frac{1}{2}, 1\}^E$.
\end{theorem}
In section \ref{sec:complexity} we discuss some results showing that max weight popular matching is $NP$-hard and under the Unique Games Conjecture is difficult to approximate. We conclude with a section discussing open problems.
\section{Extended Formulation}\label{sec:formulation}

\section{Integrality of $P_G$ in a special case}\label{sec:special}

\section{$\frac{1}{2}$-Integrality of $P_G$ in general}\label{sec:general}

\section{Complexity Results}\label{sec:complexity}

\section{Conclusions}
\end{document}

### 2017-04-05 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
bibliographies
citations

@@
 \paragraph{Authors:} Chien-Chun Huang and Telikepalli Kavitha
+\paragraph{Reference:} \cite{huang2017popularity}
 
@@
 
+\paragraph{}
+Here we present the extended formulation $P'_G$ which we will use to study the popular fractional matching polytope.
+The properties of $P'_G$ were first studied in \cite{kavitha2011popular}.
+For a more rigorous treatment of its connection to $P_G$ refer to that paper.
+Here we try to present ideas about how $P'_G$ was derived.
+
+\paragraph{}
+We begin by augmenting our graph $G$ with a set of last resort neighbours $\ell(u)$ for every $u \in V$.
+The vertices $\ell(u)$ are only connected to their respective vertex $u$.
+The preference lists of each $u$ are modified so that $\ell(u)$ is their last choice.
+That is for all $v \in N(u)$, $v \geq_u \ell(u)$.
+
 \section{Integrality of $P_G$ in a special case}\label{sec:special}
@@
 \section{Conclusions}
+
+\bibliography{references}
+\bibliographystyle{plain}
 \end{document}

### 2017-04-05 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
working on bipartite case

@@
 We define
-$$vote_u(x, v') := \sum_{v \in V(G)} x_{uv} vote_u(v,v') = x(\delta^{>v'}(u)) - x(\delta^{<v'}(u)),$$
+$$vote_u(x, v') := \sum_{uv \in \delta(u)} x_{uv} vote_u(v,v') = x(\delta^{>v'}(u)) - x(\delta^{<v'}(u)),$$
 where $\delta^{>v'}(u) := \{vu \in \delta(u) : v >_u v'\}$ denotes the set of edges $u$ prefers to $uv'$, and $\delta^{<v'}(u)$ is defined analogously.
@@
 Now we have a means of voting between two fractional matchings $x,y \in FM_G$:
-$$vote_u(x,y) := \sum_{uv' \in \delta(u)} y_{uv'}vote(x,v'), $$
+$$vote_u(x,y) := \sum_{uv' \in \delta(u)} y_{uv'}vote_u(x,v'), $$
 and the total votes:
@@
 That is for all $v \in N(u)$, $v \geq_u \ell(u)$.
+The vertex set and edge set now look like
+$$V(G) = V \cup \bigcup_{u \in V} \ell(u) \quad\text{and}\quad E(G) = E \cup \bigcup_{u \in V} u\ell(u).$$  With the addition of the last resort vertices we may now assume that every $v \in V$ (the original vertices) is fully matched in any popular fractional matching.
 
+\paragraph{}
+Fix a fractional matching $x \in FM_G$.
+We seek a linear program which characterizes when $x$ is popular.
+Let $w: E(G) \rightarrow \R_+$ be the weight function given by assigning  $$w(ab) := vote_a(b,x) + vote_b(a,x) = x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b))$$
+for any $ab \in E$, and assigning
+$$w(u\ell(u)) := vote_u(\ell(u),x) = -x(\delta^{>\ell(u)}(u)).$$
+Observe that if $y \in FM_G$ fully matches each $u \in V$ then $\sum_{ab \in E(G)} w(ab)y_{ab} = \Delta(y,x).$ So the fractional matching $x$ is popular if and only if the optimal value of the following LP is $0$ (it cannot be less than $0$ as $x$ is a feasible solution, and $\Delta(x,x) = 0$) :
+\begin{align*}
+\max\  &\sum_{e \in E(G)} w(e) y_{e} \\
+\text{s.t.}\ y(\delta(u)) &= 1 &\forall u \in V \\
+y_e &\geq 0 &\forall e \in E(G).
+\end{align*}
+Consider the dual to previous linear program, to be denoted $(LP1)$:
+\begin{align*}
+\min\ &\sum_{u \in V} \alpha_u \\
+\text{s.t.}\ \alpha_a + \alpha_b &\geq w(ab) &\forall ab \in E \\
+\alpha_u &\geq w(u\ell(u)) &\forall u \in V.
+\end{align*}
+If $x$ is popular then the optimal value of $(LP1)$ is $0$.
+Also note that if $x$ is not popular then the optimal value is greater than $0$.
+So one means of finding popular fractional matchings would be to let $x$ vary as a fractional matching and minimize $\sum_{u} \alpha_u$.
+We call that linear program $(LP2)$:
+\begin{align*}
+\min\ &\sum_{u \in V} \alpha_u \\
+\text{s.t.}\ \alpha_a + \alpha_b &\geq w(ab) &\forall ab \in E \\
+\alpha_u &\geq w(u\ell(u)) &\forall u \in V \\
+x(\delta(u)) &= 1 &\forall u \in V \\
+x_e &\geq 0 &\forall e \in E(G).
+\end{align*}
+As discussed above, the optimal value of $(LP2)$ is $0$, and so a corresponding optimal solution for $(LP2)$ is a vector $(x,\alpha) \in \R^{E(G)}\times \R^{V}$ where $x$ is a popular fractional matching and $\sum_{u \in V} \alpha_u = 0$.
+Such $\alpha$ is called a $\textit{witness}$ to the popularity of $x$.
+The polytope of optimal solutions of $(LP2)$ is what we will use as our extended formulation for the popular fractional matching polytope.
+We can describe it as $P'_G$, the set of solutions to the set of linear inequalities:
+\begin{align*}
+\sum_{u \in V} \alpha_u &= 0\\
+\alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E \\
+\alpha_u &\geq vote_u(\ell(u),x) &\forall u \in V \\
+x(\delta(u)) &= 1 &\forall u \in V \\
+x_e &\geq 0 &\forall e \in E(G).
+\end{align*}
+Thus $P'_G$ involves $O(|E| + |V|)$ constraints, and hence provides a polynomial size description of popular fractional matchings.
+The following lemma demonstrates a very interesting property of $(LP2)$
+\begin{lemma}\label{lemma:self-dual}
+The linear program $(LP2)$ is its own dual program program. That is $(LP2)$ is self-dual.
+\end{lemma}
+\begin{proof}
+Consider the dual to $(LP2)$ in variables $y \in \R^{E(G)}$ and $\beta \in \R^V$:
+\begin{align*}
+\max\ &\sum_{u \in V} \beta_u \\
+\text{s.t.} \beta_a + \beta_b + y(\delta^{<b}(a))  - y(\delta^{>b}(a)) + y(\delta^{<a}(b)) - y(\delta^{>a}(b)) &\leq 0 &\forall ab \in E \\
+\beta_u - y(\delta^{>\ell(u)}(u)) &\leq 0 &\forall u \in V \\
+y(\delta(u)) &= 1 &\forall u \in V \\
+y_e &\geq 0 &\forall e \in E(G).
+\end{align*}
+If we perform the substitutions $\alpha_u = -\beta_u$ for each $u \in V$ and $x_e = y_e$ for each $e \in E(G)$ then the above dual LP becomes exactly the same as $(LP2)$.
+\end{proof}
+\paragraph{}
+This lemma leads to the following very useful lemma, which follows immediately from the previous result via complementary slackness.
+\begin{lemma}\label{lemma:self-dual-cs}
+Let $(x, \alpha^x) \in P'_G$.  For every $ab \in E$, if $x_{ab} > 0$ then the ``covering constraint" in $P'_G$ for $ab$ is tight. Formally we have
+$$\alpha_a^x + \alpha_b^x = x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b)).$$
+\end{lemma}
 \section{Integrality of $P_G$ in a special case}\label{sec:special}
+\paragraph{}
+In this section we turn our attention to theorem \ref{th:special}.
+So suppose that our graph $G$ is bipartite with bipartition $A \cup B$.
+Also suppose that $G$ admits a perfect stable matching.
+This condition actually implies that every popular matching is perfect.
+This follows since stable matchings always match the same set of vertices, and stable matchings are minimum size popular matchings \cite{huang2011popular}.
+It is not hard to see that this extends to popular fractional matchings, meaning every popular fractional matching fully matches each $u \in V$ to its true neighbours (not last resort vertices).
+\paragraph{}
+Let $(x, \alpha^x) \in P'_G$. We want to write $x$ as a convex combination of popular matchings. First we want bounds on the values the witness vector $\alpha^x$ takes
+\begin{lemma}
+For every $u \in V$, $\alpha_u^x \in [-1,1]$.
+\end{lemma}
+\begin{proof}First
+$$\alpha^x_u \geq vote_u(\ell(u), x) = -1 \quad\text{(since }x_{u,\ell(u)} = 0\text{).}$$
+Now let $v$ be the least preferred neighbour of $u$ such that $x_{uv} >0$. Then
+$$vote_u(v,x) = -(1-x_{uv}).$$
+Also
+$$vote_v(u,x) = x(\delta^{<u}(v))-x(\delta^{>u}(v))= 1-x_{uv}.$$
+So by lemma \ref{lemma:self-dual-cs},
+$$\alpha_u + \alpha_v = vote_u(v,x) + vote_v(u,x) = -(1-x_{uv}) + (1-x_{uv}) = 0.$$
+Since $\alpha_v \geq -1$ this implies $\alpha_u \leq 1$.
+\end{proof}
+\paragraph{}
+Thinking ahead, we want to write $x$ as a convex combination of popular matchings.
+In doing this we will use a witness vector in $\{-1,1\}^V$ to certify the popularity of each chosen matching.
+From our previous lemma we can think of each $\alpha^x_u$ as a convex combination of $-1$ and $1$.
+For each $a \in A$ let $r_a$ denote the fraction of popular matchings in the convex combination forming $x$ whose $\alpha_a^x$ value is assigned $1$.
+Similarly define for each $b \in B$ the value $r_b$ to be the fraction assigning $\alpha^x_b$ to $-1$.
+Precisely we have $2r_a -1 = \alpha^x_a$ and $1-2r_b = \alpha^x_b$.
+\paragraph{}
+Let $X_a$ be the array for the assignment of $a\in A$ in $x$.
+Each cell of $X_a$ corresponds to some $b \in N(a)$ and the length of the cell is $x_{ab} > 0$ (we omit $0$-length cells).
+The cells in $X_a$ are arranged in increasing order with respect to $a$'s preferences.
+The first $r_a$ fraction of $X_a$ is called the $\textit{positive}$ subarray and the remaing the $\textit{negative}$ subarray.
+We split and reorder the arrays as shown in the following figure to obtain $X'_a$
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.25]{Xa}
+\end{figure}
+We do the same thing for each $b \in B$, creating $X_b$.
+But the only difference is that we arrange $X_b$'s cell in decreasing order of $b$'s preferences.
+Also notice that since $r_b$ corresponds to the negative $\alpha_b$ cells, the array begins with the negative subarray.
+We again split and reorder to obtain $X'_b$ as shown
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.25]{Xb}
+\end{figure}
+One interesting thing about our choice of the $r_a$'s and $r_b$'s is that they allow us to rewrite our covering constraints for $ab$ in a simpler form.
+\begin{lemma}\label{lemma:covering}
+For any $ab \in E$ we have
+$$x_{ab} + x(\delta^{<b}(a)) + x(\delta^{<a}(b)) \leq r_a + (1-r_b),$$
+and if $x_{ab} > 0$ then the above is tight.
+\end{lemma}
+\begin{proof}
+Take the constraint
+$$ \alpha^x_a + \alpha^x_b \geq x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b)) $$
+and rewrite using $\alpha^x_a = 2r_a-1$ and $\alpha^x_b = 1-2r_b$.
+Further substitute $-x(\delta^{>b}(a)) = -(1 -x_{ab} - x(\delta^{<b}(a)))$ and $-x(\delta^{>a}(b)) = -(1- x_{ab} - x(\delta^{<a}(b)))$.
+After simplifying the inequality in the lemma statement is obtained.
+By lemma \ref{lemma:self-dual-cs} this inequality is tight when $x_{ab} > 0$.
+\end{proof}
 
+\paragraph{}
+Now we show how to obtain popular matchings whose convex hull contains $x$.
+Form a table $T$ by stacking each $X'_u$ for $u \in V$ in rows.
+The width of $T$ is $1$, so for any $t \in [0,1)$ we can define a set of edges $M_t$ as follows:
+\begin{enumerate}
+\item Form a vertical line at distance $t$ from the left end of $T$
+\item For each $u \in V$ the vertical line intersects (or touches the left boundary of) a cell in $X'_u$. We will denote this cell by $c_u(t)$.
+\item Form the set $M_t$ as $M_t := \{uv : u \in V \text{ and } uv \text{ corresponds to cell } c_u(t)\}$.
+\end{enumerate}
+\paragraph{}
+We need to verify two things: the first is that $M_t$ is a matching, and the second is that $M_t$ is indeed popular.
+This will be done in the following two subsections.
+Supposing we are successful in the promised arguments to follow, we will show how this allows us to find a set of popular matchings for which $x$ is a convex combination.
+Execute the following algorithm:
+\begin{enumerate}
+\item Initialize $i=0$.
+\item Sweep a vertical line from left to right across $T$. Denote its distance from the left by $t$. 
+\item Whenever a new cell is encountered fix matching $M^i = M_t$ and fix $t_i = t$. 
+\end{enumerate}
+Say we find $k$ such matchings. We can then write $x$ as:
+$$x = t_1\chi(M^0) + (t_2 - t_1)\chi(M^1) + \dots + (1 - t_{k-1}) \chi(M^{k-1})$$
+where $\chi(\cdot)$ if the function mapping sets to incidence vectors.
+
+\subsection{$M_t$ is a matching}
+
+\subsection{$M_t$ is popular}
 \section{$\frac{1}{2}$-Integrality of $P_G$ in general}\label{sec:general}

### 2017-04-06 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
integrality

@@
 \subsection{$M_t$ is a matching}
+\paragraph{}
+Let $ab$ be an edge with $x_{ab} > 0$.
+To show that $M_t$ is a matching it suffices to show that the cell corresponding to $ab$ is perfectly aligned in its position in $X'_a$ with its position in $X'_b$.
+Then when a vertical line is dropped through the $ab$-cell in $X'_a$ it also dropped through the $ab$-cell in $X'_b$.
+Thus each vertex is matched to exactly one partner in $M_t$.
 
+\begin{proof} There are two cases to consider: When $x(\delta^{<b}(a)) \geq r_a$ (the case $x(\delta^{<a}(b)) \geq 1-r_b$ is absolutely symmetric) and the case when both $x(\delta^{<b}(a)) < r_a$ and $x(\delta^{<a}(b)) < 1-r_b$.
+\paragraph{Case 1}
+Suppose that $x(\delta^{<b}(a)) \geq r_a$.
+Then $x_{ab}$ is in $a$'s negative subarray.
+Let $d = x(\delta^{<b}(a)) - r_a$ be the distance from the left $X'_a$ to $x_{ab}$.
+Let $d' = x(\delta^{<a}(b))$ be the distance to $x_{ab}$ from the right of the positive subarray of $X'_b$ (see the figure).
+From lemma \ref{lemma:covering} we have
+$$d + x_{ab} + d' = 1-r_b.$$
+Thus $x_{ab}$ is perfectly aligned as desired. Further since $d \geq 0$, $x_{ab}$ lies in $X'_b$'s positive subarray.
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.25]{geqRa}
+\caption{Case 1}
+\end{figure}
+\paragraph{Case 2}
+Suppose $x(\delta^{<b}(a)) < r_a$ and $x(\delta^{<a}(b)) < 1-r_b$.
+In this case we have separated the cell corresponding to $x_{ab}$.
+We will describe the proof below, but as before it is more clear from the figure
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.25]{lessRa}
+\caption{Case 2}
+\end{figure}
+Let $d_0 = x(\delta^{<b}(a))$ be the distance from $X'_a$'s positive subarray to $x_{ab}$.
+Let $d_1 = x(\delta^{<a}(b))$ be the distance from the right of $X'_b$'s positive subarray to the right of $x_{ab}$.
+Let $x^0_{ab} = r_a-d_0$ be the length of the portion of the $x_{ab}$ cell on the right end of $X'_a$.
+Let $x^1_{ab} = 1-r_b - d_1$ be the length of the portion of the $x_{ab}$ cell on the left end of $X'_b$.
+By lemma \ref{lemma:covering} we know that
+$$x_{ab} + d_0 + d_1 = r_a + (1-r_b)$$
+and thus $x_{ab} = x^0_{ab} + x^1_{ab}$.
+Therefore as in the previous case the cells for $x_{ab}$ are perfectly aligned.
+Again observe that the cell portions in the positive subarray of $X'_a$ are aligned in the negative subarray of $X'_b$, and vice versa.
+This additional observation will be useful when we move to popularity.
+\end{proof}
 \subsection{$M_t$ is popular}
+\paragraph{}
+Our plan is to exhibit a witness vector $\alpha \in \{-1,1\}^V$ such that $(\chi(M_t), \alpha) \in P'_G$.
+Then $M_t$ is popular.
+We construct $\alpha$ as follows, for every $u \in A\cup B$: set $\alpha_u = -1$  if the vertical line intersect the negative subarray of $X'_u$, else set $\alpha_u = 1$ (the vertical line intersects or touches the boundary of the positive subarray).
+\paragraph{}
+Let $ab \in M_t$.
+Then, as an artifact of the proof that $M_t$ is a matching, the vertical line intersects the positive subarray of $X'_a$ if and only if it doesn't intersect the positive subarray for $X'_b$.
+Thus $\alpha_a + \alpha_b = 0$.
+Since $M_t$ is perfect this implies that
+$$\sum_{u \in A\cup B} \alpha_u = 0.$$
+Since $\alpha_u \geq -1 = vote_u(\ell(u), M_t)$ as $M_t$ is perfect, it only remains to verify the covering constraints. That is for all $ab \in E$
+$$\alpha_a + \alpha_b \geq vote_a(b,M_t) + vote_b(a,M_t).$$
+
+\begin{proof} Recall that we use $c_u(t)$ to denote the cell of $X'_u$ which the vertical line at distance $t$ intersects.
+Then the cell $c_u(t)$ has a correspondence with $uM_t(u)$.
+We delineate the following three cases: $1)$ both $c_a(t)$ and $c_b(t)$ are lie in their respective positive subarrays, $2)$ one of $c_a(t)$ or $c_b(t)$ is positive and the other negative, $3)$ or both $c_a(t)$ and $c_b(t)$ are negative.
+\paragraph{Case 1}
+Suppose that both $c_a(t)$ is positive and $c_b(t)$ is positive. Then $\alpha_a = \alpha_b = 1$. Since $vote_a(b,M_t) + vote_b(a,M_t) \leq 2$ this constraint is easily satisfied.
+\paragraph{Case 2}
+Suppose that one of $c_a(t), c_b(t)$ is positive and the other is negative. Then $\alpha_a + \alpha_b = 0$. From lemma \ref{lemma:covering} we know
+$$x(\delta^{<b}(a)) + x_{ab} + x(\delta^{<a}(b)) \leq r_a + (1-r_b)$$
+Hence for $a$ and $b$ to be matched to worse partners than each other, we need $c_t(a)$ and $c_t(b)$ to be in their respective positive subarrays.
+That cannot happen in this case, so either $a$ and $b$ matched, or at least one of them prefers the other to their partner.
+Hence
+$$vote_a(b,M_t) + vote_b(a,M_t) \geq 0$$
+as desired.
+\paragraph{Case 3}
+Suppose that both $c_a(t)$ and $c_b(t)$ are negative. Then $\alpha_a + \alpha_b = -2$. Again using the inequality 
+$$x(\delta^{<b}(a)) + x_{ab} + x(\delta^{<a}(b)) \leq r_a + (1-r_b)$$
+we see that both $a$ and $b$ are matched to better partners than each other. Therefore $vote_a(b,M_t) + vote_b(a,M_t) = -2$. This is clear from the following picture:
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.25]{covering}
+\caption{In Case 3, $c_a(t)$ and $c_b(t)$ are in the common red (negative) area for both. Since $x_{ab}$ lies in the sum length of $r_a + 1-r_b$, both $a$ and $b$ prefer their match to each other.}
+\end{figure}
+\end{proof}
 \section{$\frac{1}{2}$-Integrality of $P_G$ in general}\label{sec:general}

### 2017-04-06 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
finished report

@@
 Building upon this theorem we show that $\frac{1}{2}$-integrality holds in general in section \ref{sec:general}.
-\begin{theorem}
+\begin{theorem}\label{th:general}
 Let $G=(V,E)$ be an arbitrary graph with strict preferences. Then the extreme points of $P_G$ lie in $\{0,\frac{1}{2}, 1\}^E$.
@@
 \section{$\frac{1}{2}$-Integrality of $P_G$ in general}\label{sec:general}
+\paragraph{}
+We now return our attention to general, ie.
+non-bipartite, graphs $G = (V,E)$ and do not assume the presence of a perfect stable matching.
+In this case we are interested in proving theorem \ref{th:general}, which says that $P_G$ is $\frac{1}{2}$-integral in these instances.
+Our strategy is to map $G$ to a bipartite graph with a perfect stable matching, use the previous result, and map what we get back.
+\paragraph{}
+We begin by describing the bipartite graph $H$ obtained from $G$.
+The vertex set is $V(H) := V \cup V'$, where $V' = \{v' : v \in V\}$, and so $H$ has two vertices $v$ and $v'$ for each vertex in $G$.
+The edge set is $E(H) := \{uv', vu': uv \in E\} \cup \{uu' : u \in V(G)\}$.
+Hence $V$, $V'$ forms a bipartition of $V(H)$.
+Now we describe the preference lists.
+For each $v \in V$, if $u >_v w$ then $u'>_v w'$ in $H$ and $u >_{v'} w$ in $H$.
+In $H$ each vertices last choice is their ``twin" vertex.
+So $u' \geq_v v'$ for all $u' \in N(v)$, and similarly $u \geq_{v'} v$ for all $u \in N(v')$.
+\begin{lemma}
+$H$ admits a stable matching.
+\end{lemma}
+\begin{proof}
+Let $S$ be a stable matching in $H$.
+Suppose without loss of generality that $S$ exposes $u \in V(H)$.
+Let $S' := \{uv' :u'v \in S\} \cup \{u'v: uv' \in S\}$ be the ``twin" matching of $S$.
+By symmetry of the preference orders, since $S$ is stable so is $S'$.
+By our construction of $S'$, $S'$ exposes $u'$.
+But further since each stable matching matches the same vertex set, this implies that $S$ also exposes $u'$.
+But then the edge $uu'$ blocks $S$, a contradiction.
+\end{proof}
 
+\paragraph{}
+Thus by theorem \ref{th:special}, $P_H$ is integral. Now we describe the mapping alluded to an the outset of this section. Let $f: P_G \rightarrow P_H$ be defined as follows.
+$$f(x) = z \text{ where } z_{uv'}=z_{u'v} =x_{uv} \text{ and } z_{uu'} = x_{u\ell(u)}.$$
+We need to verify that $f$ indeed maps popular fractional matchings to popular fractional matchings. It is clear that $z$ is a fractional matching on $H$. We need to check that $z$ is popular.
+\begin{lemma}
+Let $x \in P_G$. Then $z = f(x) \in P_H$.
+\end{lemma}
+\begin{proof}
+We construct a witness $\beta$ to the popularity of $z$.
+Let $\alpha$ be a witness to the popularity of $x$.
+For each $u \in V$ set $\beta_u = \beta_{u'} = \alpha_u$.
+Now $\sum_{v\in V(H)} \beta_v = \sum_{v \in V} \beta_v + \sum_{v' \in V'} \beta_v' = 2 \sum_{u \in V} \alpha_u = 0$.
+Now we need to check the covering constraints.
+Let $ab \in E(G)$.
+Since $(x,\alpha)$ satisfies the $ab$-covering constraint for $P'_G$, it immediate that $(z,\beta)$ satisfies the covering constraints for $ab'$ and $a'b$ in $P'_H$.
+\paragraph{}
+It remains to consider covering constraints for $uu' \in E(H)$.
+Consider $u\ell(u) \in E(G)$.
+We have $\alpha_u \geq x_{u\ell(u)} - 1$ from the description of $P'_G$.
+Now observe that the right-hand side of the covering constraint for $uu'$ in $P'_H$ is
+$$vote_u(u', z) + vote_u'(u,z) = 2(x_{u\ell}-1).$$
+The left-hand side of the covering constraint for $uu'$ in $P'_H$ is
+$$\beta_u + \beta_u' = 2\alpha_u \geq 2(x_{u\ell(u)}-1)$$
+and hence the covering constraint for $uu'$ is satisfied. It is trivial to see that the last-resort covering constraints in $P'_H$ are satisfied. Hence $z \in P_H$.
+\end{proof}
+\paragraph{}
+Now we need a mapping that will take us back from $P_H$ to $P_G$.
+We want this mapping to be the inverse of $f$ when restricted to fractional matchings in the image of $f$.
+Let $h: P_H \rightarrow P_G$.
+We want that $h \circ f (x) = x$ for all $x \in P_G$.
+To satisfy this we define $h$ as follows:
+$$h(z) = y \text{ where } y_{uv} = (z_{uv'} + z_{u'v})/2 \text{ and } y_{u\ell(u)} = z_{uu'}.$$
+Verifying that $h(z)$ is popular is a routine calculation as in the previous lemma.
+The witness vector chosen is  $\alpha$ where $\alpha_u = (\beta_u + \beta_u')/2$ and $\beta$ is a witness to the popularity of $z$.
+\paragraph{}
+Now we can show how to write $x \in P_G$ as a convex combination of $\frac{1}{2}$-integral popular matchings.
+Let $z=f(x)$.
+Then $z \in P_H$.
+By theorem \ref{th:special}, that there exist $M_1, \dots, M_k$ and $\lambda \in \R_+^k$ such that $\sum_{i=}^k \lambda_i = 1$ and
+$$z = \sum_{i=1}^k \lambda_i \chi(M_i).$$
+Now we have $x = h(z)$, and from the definition of $h$, $h$ distributes over scalar multiplication so we have
+$$x = h(z) = \sum_{i=1}^k \lambda_i h(\chi(M_i)).$$
+Now from the definition of $h$, $h(\chi(M_i)) \in \{0,\frac{1}{2}, 1\}^{E(G)}$, so we have shown that $x$ is a convex combination of $\frac{1}{2}$-integral popular matchings.$\blacksquare$
 \section{Complexity Results}\label{sec:complexity}
+\paragraph{}
+The previous section showed that we can find max-weight popular $\frac{1}{2}$-integral matchings in polynomial time.
+This section seeks to study the hardness of max-weight popular matching.
+We will show that not only is this problem $NP$-hard, but hard to approximate under the Unique Games Conjecture.
+\paragraph{}
+We give a reduction from VERTEX COVER $(VC)$.
+Let $H = (V_H, E_H)$ be an instance of $(VC)$ with $|V(H)| = n$.
+From $H$ we will construct a graph $G=(V,E)$ with preferences, and a weight function $w: E \rightarrow \{1,2\}$.
+The graph $G$ will be the complete graph on $4n$ vertices.
+For each $v \in V(H)$ we have $4$ vertices $v_0, v_1, v_2, v_3 \in V(G)$.
+Below we present the preference lists of $v_0,v_1,v_2,v_3$.
+In these lists we use $\pi(u_0: u \in N(v))$ to denote an arbitrary permutation on  $\{u_0: u \in N(v) \text{ with respect to $H$}\}$, and we use $\pi(\dots)$ to denote an arbitrary permutation of the vertices not mentioned so far in the particular list.
+We use these notations to indicate that our reductions are indifferent to how these parts of the preference lists are arranged.
+\begin{align*}
+v_0 &: v_1 > \pi(u_0: u \in N(v)) > v_2 > v_3 >\pi(\dots) \\
+v_1 &: v_0 > v_2 > v_3 > \pi(\dots) \\
+v_2 &: v_1 > v_0 > v_3 > \pi(\dots) \\
+v_3 &: v_1 > v_2 > v_0 > \pi(\dots).
+\end{align*}
+The weight function $w$ is described by $w(v_0v_2) = w(v_1v_3) = 2$ for each $v \in V(H)$ and $w(e) = 1$ for all other edges $e\in E(G)$. 
+\paragraph{}
+In the reductions we rely critically on the following characterization of popular matchings from \cite{huang2011popular}.
+\begin{lemma}\label{lemma:char}
+\cite{huang2011popular} Let $M$ be a perfect matching in $G$.
+For all $uv \in E\backslash M$ label $uv$ with $(vote_u(v,M(u)), vote_v(u,M(v)))$.
+Then $(1,1)$ edges are blocking edges in the stable matching sense.
+Let $G_M$ be the graph obtained from $G$ by deleting $(-1,-1)$ edges (doesn't hurt popularity).
+Then $M$ is popular if and only if in $G_M$:
+\begin{enumerate}
+\item There is no $M$-alternating cycle containing a blocking edge, and
+\item there is not $M$-alternating path with two or more blocking edges.
+\end{enumerate}
+\end{lemma}
 
+\begin{lemma}\label{lemma:firstreduc}
+Let $C$ be a vertex cover of size $k$ in $H$. Then there exists a popular matching $M$ of $G$ with $w(M) = 4n-2k$. This gives one direction of the desired reduction.
+\end{lemma}
+\begin{proof}
+Construct matching $M$ by adding two edges for each $v \in V(H)$:
+$$\text{ if $v \in C$ then add $v_0v_1, v_2v_3$ to $M$, otherwise add $v_0v_2, v_1v_3$ to $M$}.$$
+Then $w(M) = (1+1)k + (2+2)(n-k) = 4n - 2k$ as desired.
+We need only verify popularity of $M$.
+This is done through checking the two conditions of lemma \ref{lemma:char} on $G_M$.
+One can observe there are four classes of edges in $G_M$:
+\begin{enumerate}
+\item Edges of $M$.
+\item Edges $v_0v_2, v_1v_2, v_1v_3$ for all $v \in C$. The label on these edges is $(-1,1)$.
+\item Edges $v_0v_1, v_1v_2$ for all $v \not\in C$. The label on these edges is $(1,1)$, that is they are blocking edges.
+\item The edge $v_0u_0$ where $vu \in E(H)$ and exactly one of $u,v \in C$. The label on this edge is either $(1,-1)$ or $(-1,1)$.
+\end{enumerate}
+It is not hard to verify the first two conditions of lemma \ref{lemma:char} from this point.
+To see the alternating cycle condition, one observes that the edges containing any $v_3$ are either in $M$ or $(1,1)$ edges, hence the degree of $v_3$ is one.
+Since blocking edges are either $v_0v_1$ of $v_1v_2$ and their partner is $v_3$ there cannot be an alternating cycle containing a blocking edge.
+\paragraph{}
+Suppose $P$ is an $M$-alternating path containing either $v_0v_1$ or $v_1v_2$ for $v \not\in C$.
+It cannot contain both as there is no such alternating path in $G_M$.
+Then the path $P$ needs to use a class $4$ edge $v_0u_0$ to get out of the ``$v$" edges and find another blocking edge.
+Since $v \not\in C$, such $u$ is in $C$.
+Thus $M(u_0) = u_1$.
+Now the neighbours of $u_1$ are $u_2$ and $u_3$ only.
+Since $u_2$ and $u_3$ are not incident to any blocking edges, $P$ cannot find a second blocking edge.
+\end{proof}
+\begin{lemma}\label{lemma:6.2}
+Let $M$ be a popular matching in $G$. Then there does not exist $v_su_j \in M$ with $v\neq u$ and $s,t \in \{0,1,2,3\}$.
+\end{lemma}
+\begin{proof}
+The proof is by case analysis using the characterization from lemma \ref{lemma:char}.
+They show that allowing such an edge contradicts the characterization.
+The computations are routine in popular matchings, so we'll enumerate the cases and spare the details.
+The cases are
+\begin{enumerate}
+\item $s,t \in \{0,1\}$
+\item $s, \in \{0,1\}$ and $t=2$
+\item $s \in \{0,1\}$ and $t=3$
+\item $s,t \in\{2,3\}$.
+\end{enumerate}
+\end{proof}
+\begin{lemma}\label{lemma:6.3}
+Suppose $M$ is a popular matching in $G$. Then 
+\begin{enumerate}
+\item For all $v_0,v_1,v_2,v_3 \in V(G)$, either $v_0v_1,v_2v_3 \in M$ or $v_0v_2v_1v_3 \in M$
+\item If $vu \in E(H)$ then either $v_0v_1 v_2v_3 \in M$ or $u_0u_1, u_2u_3 \in M$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+The previous lemma (lemma \ref{lemma:6.2}) tells us that $v_0,v_1,v_2,v_3$ are necessarily matched together.
+But if $v_0v_3, v_1v_2 \in M$ then $v_0,v_3,v_2,v_1,v_0$ forms an $M$-alternating cycle in $G_M$ containing blocking edge $v_0v_1$.
+Hence the options stated in $1.$ are the only possibilities.
+For the second part one can similarly find an alternating path containing two blocking edges.
+\end{proof}
+\begin{lemma}\label{lemma:secondreduc}
+If $M$ is a popular matching in $G$ with $w(M) \geq 4n-2k$ then there is a vertex cover $C$ in $H$ with $|C| \leq k$. This gives the other direction of the desired reduction.
+\end{lemma}
+\begin{proof}
+From part $1)$ of lemma \ref{lemma:6.3}, for all $v \in V(H)$ either $v_0v_1, v_2v_3 \in M$ or $v_0v_2,v_1v_3 \in M$. Define $C$ as
+$$C := \{ v \in V(H): v_0v_1, v_2v_3 \in M\}.$$
+Lemma \ref{lemma:6.3} part $2)$ implies that $H$ is a vertex cover. Now we have the equation
+$$w(M) = (1+1)|C| + (2+2)(n-|C|).$$
+Rearranging, using $w(M) \geq 4n-2k$,  we obtain that $|C| \leq k$.
+\end{proof}
+\paragraph{}
+Thus we have shown the following result. Since we used a stricter version of max-weight popular matching we have a stronger $NP$-hardness result
+\begin{theorem}
+The max-weight popular matching problem on graph $G$ is $NP$-hard, even in the special case that $w: E(G)\rightarrow \{1,2\}$, preferences are strict, $G$ is a complete graph, and $G$ admits a stable matching.
+\end{theorem}
+\paragraph{}
+We can obtain an inapproximability result by slightly modifying our reduction from $H$ to $G$.
+We change the weight function $w$ so that $w(v_0v_2) = w(v_1v_3) = 1$ for all $v \in V(H)$, and $w(e) = 0$ for all other edges of $G$.
+Now our previous lemmas of this section still hold, but the size-weight connection is slightly different: there is a vertex cover of $H$ of size $\leq k$ if and only if there is a popular matching of $G$ of weight $\geq 2n-2k$ (and this vertex cover can be found in polynomial time).
+\begin{theorem}
+There does not exist an $O(1)$-approximation algorithm for max-weight popular matching with non-negative weights under Unique Games Conjecture.
+\end{theorem}
+\begin{proof}
+Let $C^*$ be an optimal vertex cover on $H$.
+Since $H$ is hardest to approximate when it contains a perfect matching, we may assume it does so.
+Thus $|C^*| \geq \frac{n}{2}$.
+Suppose for a contradiction that there exists an $\epsilon$-approximation algorithm $\cA$ for max-weight popular matching on $G$ where $\epsilon  \in (0,1)$.
+Compute a popular matching $M$ on $G$ using $\cA$.
+We split into two cases on $|C^*|$.
+\paragraph{Case 1}
+Suppose that $|C^*| \leq (1+\epsilon)\frac{n}{2}$.
+In this case $w(M) \geq \epsilon(1-\epsilon) n$.
+Using our reduction we can find a vertex cover $C$ with $|C| \leq (2-\epsilon+\epsilon^2)\frac{n}{2}$.
+Since $\frac{n}{2} \leq |C^*|$ this implies
+$$|C| \leq (2-\epsilon +\epsilon^2)|C^*|.$$
+\paragraph{Case 2}
+In this case we obtain $C$ as above. The approximation factor is $\frac{n}{(1+\epsilon)n/2}$ which is at most $2(1-\epsilon + \epsilon^2)$.
+\paragraph{}
+So we can find a vertex cover $C$ with $|C| \leq (2-\epsilon + \epsilon^2)|C^*|$ in either case. Since $\epsilon \in (0,1)$ this contradiction Unique Games Conjecture.
+\end{proof}
 \section{Conclusions}
-
+\paragraph{}
+We have shown that given a bipartite graph with strict preferences, if the graph admits a perfect stable matching then we can solve the max-weight popular matching problem in polynomial time using linear programming.
+We have also shown that for arbitrary graphs we can find a max-weight $\frac{1}{2}$-integral popular matching in polynomial time.
+In the bipartite case, since the fractional matching polytope is integral, this implies we can actually find a mixed matching (ie a convex combination of matchings) using at most two matchings.
+\paragraph{}
+We have demonstrated hardness results saying that on general graphs the max-weight popular matching problem is not only $NP$-hard but also inapproximable unless Unique Games Conjecture does not hold.
+\paragraph{}
+The major open problems here are two: $1)$ settle the complexity of the max-weight popular matching problem on bipartite graphs, and $2)$ to resolve the popular matching problem on general graphs.
+The latter problem is simply the decision problem resolving existence of a popular matching on graphs with strict preferences.
 \bibliography{references}

### 2017-04-10 CS798-ConvexityAndOptimization/assignments/project.tex
started project file

--- 2017-04-05 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
@@
 
-\lhead{2017-04-04}
-\rhead{W. Justin Toth CO750-Approximation Algorithms Report} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{2017-04-09}
+\rhead{William Justin Toth CS798-Convexity and Optimization Project} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-\paragraph{Paper Title:} Popularity, Mixed Matchings, and Self-duality
-\paragraph{Authors:} Chien-Chun Huang and Telikepalli Kavitha
-\paragraph{Reference:} \cite{huang2017popularity}
+\paragraph{}
+The multiplicative weights update method studied in Lecture $9$ is a powerful and ubiquitous tool in algorithm design \cite{arora2012multiplicative}.
+This method is the workhorse behind the primal-dual paradigm in designing competitive online algorithms \cite{buchbinder2009design}.
+One major area of application for such algorithms is the design of auctions for selling ad space in search query results.
+\paragraph{}
+We will begin by introducing the framework in which competitive online algorithms are studied.
+General packing/covering problems with be defined, and a we will give an online algorithm using the primal-dual paradigm for such problems.
+We will observe how a multiplicative weights update procedure is used to update the primal variables in each iteration.
+\paragraph{}
+Afterwards we will see how this general approach can be applied to applications in implementing ad auctions.
+We will start with the simple single-unit case, then move on to the more complex multi-slot case.
+In the multi-slot case strong duality of linear programs will be of great utility.
+\section{Competitive Online Algorithms via Primal-Dual}
+\subsection{Online Packing and Covering}
+\paragraph{}
 
-\section{Preliminaries}
-\paragraph{}
-We will begin by introducing the definitions of the main objects of study, and some useful notation. Then we will state the theorems to be shown in the later sections of this report.
-\paragraph{}
-As in most problems in the stable matching family we are given a graph $G=(V,E)$ where each vertex has ranking over its neighbours.
-We will assume throughout that these preference orders are strict, and a vertex prefers any neighbour to being unmatched.
-We will denote by $N(v)$ the set of neighbours of vertex $v$, and by $\delta(v)$ the set of edges incident with $v$.
-We will use the notation $>_v$ to describe $v$'s preference order over $N(v)$.
-Precisely for $u, w \in N(v)$, we write $u >_v w$ if $u$ is preferred by $v$ to $w$.
-We will use $\geq_v$, $\leq_v$, $=_v$, and $<_v$ analogously.
-\paragraph{}
-One way to think of $\textit{popular}$ matchings is as matchings which never lose an election to any other matching.
-Since we will speak often about partners in matchings, we introduce the notation: if $M$ is a matching then $M(u) = v$ provided $uv \in M$.
-For any $u \in V$ and $v,v' \in N(v)$ we define $vote_u(v,v')$ as
-$$vote_u(v,v') := \begin{cases}
-1, \text{ if } v >_u v' \\
--1, \text{ if } v <_u v' \\
-0, \text{ otherwise (ie. } v = v').
-\end{cases}$$
-To decide the winner of the election we some the votes. The aggregate voting between two matchings $M$ and $M'$ is
-$$\Delta(M, M') = \sum_{u \in V} vote_u(M(u), M'(u)).$$
-Observe that $\Delta(M,M') \geq 0$ if at least as many vertices prefer their partner in $M$ to their partner in $M'$. A $\textit{popular}$ matching $M$ is  a matching where
-$$\Delta(M, M') \geq 0$$
-for any matching $M'$.
-\paragraph{}
-We are interested in the $\textit{max weight popular matching}$ problem in this report.
-That is, given a weight function $w: E \rightarrow \R_+$ find a popular matching $M$ on $G$ which maximizes $w(M) := \sum_{e \in M} w(e).$ A standard approach to such problems is to attempt to formulate them as linear programs.
-This will motivate our definitions of fractional popular matchings to follow.
-In this approach the major question is whether or not your linear programming formulation gives back integral solutions which can be used to solve the original combinatorial problem.
-In this paper we will prove that there is an integral formulation for a special case, and a $\frac{1}{2}$-integral formulation in general.
-\paragraph{}
-The fractional matching polytope $FM_{G}$ is defined as
-$$FM_{G} := \{x \in \R_{\geq 0}^{E} : x(\delta(v)) \leq 1, \forall v \in V \}.$$
-This object has been well-studied in combinatorial optimization.
-We want to study the analogous $\textit{popular fractional matching polytope}$ which we will call $P_G$.
-To formally define $P_G$ we will need to extend our notions related to voting as used in defining popular matchings.
-Let $x \in FM_G$.
-We define
-$$vote_u(x, v') := \sum_{uv \in \delta(u)} x_{uv} vote_u(v,v') = x(\delta^{>v'}(u)) - x(\delta^{<v'}(u)),$$
-where $\delta^{>v'}(u) := \{vu \in \delta(u) : v >_u v'\}$ denotes the set of edges $u$ prefers to $uv'$, and $\delta^{<v'}(u)$ is defined analogously.
-We define $vote_u(v',x) = -vote_u(x,v')$.
-Now we have a means of voting between two fractional matchings $x,y \in FM_G$:
-$$vote_u(x,y) := \sum_{uv' \in \delta(u)} y_{uv'}vote_u(x,v'), $$
-and the total votes:
-$$\Delta(x,y) := \sum_{u \in V} vote_u(x,y).$$
-Now we can formally describe $P_G$ as 
-$$P_G = \{ x\in FM_G: \Delta(x,y) \geq 0, \forall y \in FM_G\}.$$
-\paragraph{}
-We finally have sufficient definitions to describe the results of this paper.
-In section \ref{sec:formulation} we describe a useful extended formulation which describes $P_G$ that will come into play when studying popular fractional matchings throughout this work.
-In section \ref{sec:special} we build up the following theorem
-\begin{theorem}\label{th:special}
-Let $G = (A \cup B, E)$ be a bipartite graph with strict preferences. If $G$ admits a perfect stable matching the $P_G$ is integral.
-\end{theorem}
-Building upon this theorem we show that $\frac{1}{2}$-integrality holds in general in section \ref{sec:general}.
-\begin{theorem}
-Let $G=(V,E)$ be an arbitrary graph with strict preferences. Then the extreme points of $P_G$ lie in $\{0,\frac{1}{2}, 1\}^E$.
-\end{theorem}
-In section \ref{sec:complexity} we discuss some results showing that max weight popular matching is $NP$-hard and under the Unique Games Conjecture is difficult to approximate.
-We conclude with a section discussing open problems.
-\section{Extended Formulation}\label{sec:formulation}
+\subsection{General Algorithm}
 
-\paragraph{}
-Here we present the extended formulation $P'_G$ which we will use to study the popular fractional matching polytope.
-The properties of $P'_G$ were first studied in \cite{kavitha2011popular}.
-For a more rigorous treatment of its connection to $P_G$ refer to that paper.
-Here we try to present ideas about how $P'_G$ was derived.
+\section{Ad Auctions}
+\subsection{Single Slot}
 
-\paragraph{}
-We begin by augmenting our graph $G$ with a set of last resort neighbours $\ell(u)$ for every $u \in V$.
-The vertices $\ell(u)$ are only connected to their respective vertex $u$.
-The preference lists of each $u$ are modified so that $\ell(u)$ is their last choice.
-That is for all $v \in N(u)$, $v \geq_u \ell(u)$.
-The vertex set and edge set now look like
-$$V(G) = V \cup \bigcup_{u \in V} \ell(u) \quad\text{and}\quad E(G) = E \cup \bigcup_{u \in V} u\ell(u).$$  With the addition of the last resort vertices we may now assume that every $v \in V$ (the original vertices) is fully matched in any popular fractional matching.
-
-\paragraph{}
-Fix a fractional matching $x \in FM_G$.
-We seek a linear program which characterizes when $x$ is popular.
-Let $w: E(G) \rightarrow \R_+$ be the weight function given by assigning  $$w(ab) := vote_a(b,x) + vote_b(a,x) = x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b))$$
-for any $ab \in E$, and assigning
-$$w(u\ell(u)) := vote_u(\ell(u),x) = -x(\delta^{>\ell(u)}(u)).$$
-Observe that if $y \in FM_G$ fully matches each $u \in V$ then $\sum_{ab \in E(G)} w(ab)y_{ab} = \Delta(y,x).$ So the fractional matching $x$ is popular if and only if the optimal value of the following LP is $0$ (it cannot be less than $0$ as $x$ is a feasible solution, and $\Delta(x,x) = 0$) :
-\begin{align*}
-\max\  &\sum_{e \in E(G)} w(e) y_{e} \\
-\text{s.t.}\ y(\delta(u)) &= 1 &\forall u \in V \\
-y_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-Consider the dual to previous linear program, to be denoted $(LP1)$:
-\begin{align*}
-\min\ &\sum_{u \in V} \alpha_u \\
-\text{s.t.}\ \alpha_a + \alpha_b &\geq w(ab) &\forall ab \in E \\
-\alpha_u &\geq w(u\ell(u)) &\forall u \in V.
-\end{align*}
-If $x$ is popular then the optimal value of $(LP1)$ is $0$.
-Also note that if $x$ is not popular then the optimal value is greater than $0$.
-So one means of finding popular fractional matchings would be to let $x$ vary as a fractional matching and minimize $\sum_{u} \alpha_u$.
-We call that linear program $(LP2)$:
-\begin{align*}
-\min\ &\sum_{u \in V} \alpha_u \\
-\text{s.t.}\ \alpha_a + \alpha_b &\geq w(ab) &\forall ab \in E \\
-\alpha_u &\geq w(u\ell(u)) &\forall u \in V \\
-x(\delta(u)) &= 1 &\forall u \in V \\
-x_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-As discussed above, the optimal value of $(LP2)$ is $0$, and so a corresponding optimal solution for $(LP2)$ is a vector $(x,\alpha) \in \R^{E(G)}\times \R^{V}$ where $x$ is a popular fractional matching and $\sum_{u \in V} \alpha_u = 0$.
-Such $\alpha$ is called a $\textit{witness}$ to the popularity of $x$.
-The polytope of optimal solutions of $(LP2)$ is what we will use as our extended formulation for the popular fractional matching polytope.
-We can describe it as $P'_G$, the set of solutions to the set of linear inequalities:
-\begin{align*}
-\sum_{u \in V} \alpha_u &= 0\\
-\alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E \\
-\alpha_u &\geq vote_u(\ell(u),x) &\forall u \in V \\
-x(\delta(u)) &= 1 &\forall u \in V \\
-x_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-Thus $P'_G$ involves $O(|E| + |V|)$ constraints, and hence provides a polynomial size description of popular fractional matchings.
-The following lemma demonstrates a very interesting property of $(LP2)$
-\begin{lemma}\label{lemma:self-dual}
-The linear program $(LP2)$ is its own dual program program. That is $(LP2)$ is self-dual.
-\end{lemma}
-\begin{proof}
-Consider the dual to $(LP2)$ in variables $y \in \R^{E(G)}$ and $\beta \in \R^V$:
-\begin{align*}
-\max\ &\sum_{u \in V} \beta_u \\
-\text{s.t.} \beta_a + \beta_b + y(\delta^{<b}(a))  - y(\delta^{>b}(a)) + y(\delta^{<a}(b)) - y(\delta^{>a}(b)) &\leq 0 &\forall ab \in E \\
-\beta_u - y(\delta^{>\ell(u)}(u)) &\leq 0 &\forall u \in V \\
-y(\delta(u)) &= 1 &\forall u \in V \\
-y_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-If we perform the substitutions $\alpha_u = -\beta_u$ for each $u \in V$ and $x_e = y_e$ for each $e \in E(G)$ then the above dual LP becomes exactly the same as $(LP2)$.
-\end{proof}
-\paragraph{}
-This lemma leads to the following very useful lemma, which follows immediately from the previous result via complementary slackness.
-\begin{lemma}\label{lemma:self-dual-cs}
-Let $(x, \alpha^x) \in P'_G$.  For every $ab \in E$, if $x_{ab} > 0$ then the ``covering constraint" in $P'_G$ for $ab$ is tight. Formally we have
-$$\alpha_a^x + \alpha_b^x = x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b)).$$
-\end{lemma}
-\section{Integrality of $P_G$ in a special case}\label{sec:special}
-\paragraph{}
-In this section we turn our attention to theorem \ref{th:special}.
-So suppose that our graph $G$ is bipartite with bipartition $A \cup B$.
-Also suppose that $G$ admits a perfect stable matching.
-This condition actually implies that every popular matching is perfect.
-This follows since stable matchings always match the same set of vertices, and stable matchings are minimum size popular matchings \cite{huang2011popular}.
-It is not hard to see that this extends to popular fractional matchings, meaning every popular fractional matching fully matches each $u \in V$ to its true neighbours (not last resort vertices).
-\paragraph{}
-Let $(x, \alpha^x) \in P'_G$. We want to write $x$ as a convex combination of popular matchings. First we want bounds on the values the witness vector $\alpha^x$ takes
-\begin{lemma}
-For every $u \in V$, $\alpha_u^x \in [-1,1]$.
-\end{lemma}
-\begin{proof}First
-$$\alpha^x_u \geq vote_u(\ell(u), x) = -1 \quad\text{(since }x_{u,\ell(u)} = 0\text{).}$$
-Now let $v$ be the least preferred neighbour of $u$ such that $x_{uv} >0$. Then
-$$vote_u(v,x) = -(1-x_{uv}).$$
-Also
-$$vote_v(u,x) = x(\delta^{<u}(v))-x(\delta^{>u}(v))= 1-x_{uv}.$$
-So by lemma \ref{lemma:self-dual-cs},
-$$\alpha_u + \alpha_v = vote_u(v,x) + vote_v(u,x) = -(1-x_{uv}) + (1-x_{uv}) = 0.$$
-Since $\alpha_v \geq -1$ this implies $\alpha_u \leq 1$.
-\end{proof}
-\paragraph{}
-Thinking ahead, we want to write $x$ as a convex combination of popular matchings.
-In doing this we will use a witness vector in $\{-1,1\}^V$ to certify the popularity of each chosen matching.
-From our previous lemma we can think of each $\alpha^x_u$ as a convex combination of $-1$ and $1$.
-For each $a \in A$ let $r_a$ denote the fraction of popular matchings in the convex combination forming $x$ whose $\alpha_a^x$ value is assigned $1$.
-Similarly define for each $b \in B$ the value $r_b$ to be the fraction assigning $\alpha^x_b$ to $-1$.
-Precisely we have $2r_a -1 = \alpha^x_a$ and $1-2r_b = \alpha^x_b$.
-\paragraph{}
-Let $X_a$ be the array for the assignment of $a\in A$ in $x$.
-Each cell of $X_a$ corresponds to some $b \in N(a)$ and the length of the cell is $x_{ab} > 0$ (we omit $0$-length cells).
-The cells in $X_a$ are arranged in increasing order with respect to $a$'s preferences.
-The first $r_a$ fraction of $X_a$ is called the $\textit{positive}$ subarray and the remaing the $\textit{negative}$ subarray.
-We split and reorder the arrays as shown in the following figure to obtain $X'_a$
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.25]{Xa}
-\end{figure}
-We do the same thing for each $b \in B$, creating $X_b$.
-But the only difference is that we arrange $X_b$'s cell in decreasing order of $b$'s preferences.
-Also notice that since $r_b$ corresponds to the negative $\alpha_b$ cells, the array begins with the negative subarray.
-We again split and reorder to obtain $X'_b$ as shown
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.25]{Xb}
-\end{figure}
-One interesting thing about our choice of the $r_a$'s and $r_b$'s is that they allow us to rewrite our covering constraints for $ab$ in a simpler form.
-\begin{lemma}\label{lemma:covering}
-For any $ab \in E$ we have
-$$x_{ab} + x(\delta^{<b}(a)) + x(\delta^{<a}(b)) \leq r_a + (1-r_b),$$
-and if $x_{ab} > 0$ then the above is tight.
-\end{lemma}
-\begin{proof}
-Take the constraint
-$$ \alpha^x_a + \alpha^x_b \geq x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b)) $$
-and rewrite using $\alpha^x_a = 2r_a-1$ and $\alpha^x_b = 1-2r_b$.
-Further substitute $-x(\delta^{>b}(a)) = -(1 -x_{ab} - x(\delta^{<b}(a)))$ and $-x(\delta^{>a}(b)) = -(1- x_{ab} - x(\delta^{<a}(b)))$.
-After simplifying the inequality in the lemma statement is obtained.
-By lemma \ref{lemma:self-dual-cs} this inequality is tight when $x_{ab} > 0$.
-\end{proof}
-
-\paragraph{}
-Now we show how to obtain popular matchings whose convex hull contains $x$.
-Form a table $T$ by stacking each $X'_u$ for $u \in V$ in rows.
-The width of $T$ is $1$, so for any $t \in [0,1)$ we can define a set of edges $M_t$ as follows:
-\begin{enumerate}
-\item Form a vertical line at distance $t$ from the left end of $T$
-\item For each $u \in V$ the vertical line intersects (or touches the left boundary of) a cell in $X'_u$. We will denote this cell by $c_u(t)$.
-\item Form the set $M_t$ as $M_t := \{uv : u \in V \text{ and } uv \text{ corresponds to cell } c_u(t)\}$.
-\end{enumerate}
-\paragraph{}
-We need to verify two things: the first is that $M_t$ is a matching, and the second is that $M_t$ is indeed popular.
-This will be done in the following two subsections.
-Supposing we are successful in the promised arguments to follow, we will show how this allows us to find a set of popular matchings for which $x$ is a convex combination.
-Execute the following algorithm:
-\begin{enumerate}
-\item Initialize $i=0$.
-\item Sweep a vertical line from left to right across $T$. Denote its distance from the left by $t$. 
-\item Whenever a new cell is encountered fix matching $M^i = M_t$ and fix $t_i = t$. 
-\end{enumerate}
-Say we find $k$ such matchings. We can then write $x$ as:
-$$x = t_1\chi(M^0) + (t_2 - t_1)\chi(M^1) + \dots + (1 - t_{k-1}) \chi(M^{k-1})$$
-where $\chi(\cdot)$ if the function mapping sets to incidence vectors.
-
-\subsection{$M_t$ is a matching}
-
-\subsection{$M_t$ is popular}
-\section{$\frac{1}{2}$-Integrality of $P_G$ in general}\label{sec:general}
-
-\section{Complexity Results}\label{sec:complexity}
-
-\section{Conclusions}
-
+\subsection{Multiple Slots}
 \bibliography{references}

### 2017-04-11 CS798-ConvexityAndOptimization/assignments/project.tex
current writeup

@@
 \paragraph{}
-Afterwards we will see how this general approach can be applied to applications in implementing ad auctions.
+Afterwards we will see how ideas from this general approach can be applied to applications in implementing ad auctions.
 We will start with the simple single-unit case, then move on to the more complex multi-slot case.
@@
 \paragraph{}
+We will start by discussing packing/covering problems in the standard, offline setting then move from there to the online model.
+The $\textit{covering}$ problem is specified by a linear program $(P)$ of the form:
+\begin{align*}
+\min &\sum_{i=1}^n c_i x_i \\
+\text{s.t.} \sum_{i \in S(j)} x_i &\geq 1 &\forall j \in [m]\\
+x_i &\geq 0 &\forall  i \in [n].
+\end{align*}
+There is some notation here to unpack.
+First for any $k \in \N$ we have $[k] := \{1, \dots, k\}$.
+The vector $x \in \R^n$ is the variable set.
+The vector $c \in \R^n$ is the objective function, which is assumed to be non-negative.
+The sets $S(1), \dots, S(m)$ are subsets of $[n]$.
+We note here that this model is not the most general form of covering problem imaginable, but it is somewhat simpler while still being general enough to capture the main ideas we are striving for.
+\paragraph{}
+We are discussing $\textit{fractional}$ covering here, but if we were to restrict our attention to $0-1$ integral solutions we would have nice a combinatorial interpretations for this problem.
+In this context, covering asks you to find a minimum cost subset of $[n]$ whose intersection with each $S(j)$ is non-empty.
+\paragraph{}
+The $\textit{packing}$ problem $(D)$ is defined as the dual to a corresponding covering problem:
+\begin{align*}
+\max &\sum_{j=1}^m y_j \\
+\text{s.t.} \sum_{j : i \in S(j)} y_j &\leq c_i &\forall i \in [n] \\
+y_j &\geq 0 &\forall j \in [m].
+\end{align*}
+\paragraph{}
+In the $\textit{online covering problem}$ we are solving the covering problem, but we do not have all the information in advance.
+The cost function $c$ is known to us, as well as the size of the variable space.
+The entire space of constraints is not known.
+The constraints are given in some sequence unknown to the algorithm.
+At each point in the sequence the algorithm receives knowledge of one constraint, and must decide on a feasible solution to the covering problem with the constraints at hand.
+Any variables the algorithm decides to increase to maintain feasibility of the current sub-instance may not be decreased at future points in the input sequence.
+\paragraph{}
+One can observe that at any point in the operation of the algorithm in the online model, the constraints revealed so far form a sub-instance of the covering problem which is itself a covering problem.
+We define the online packing problem in such a way that its sub-instances are the duals to the covering problem sub-instances.
+That is, we reveal the variables one-at-a-time in the online packing problem, as opposed to the constraints as we did for online covering.
+\paragraph{}
+In the $\textit{online packing problem}$ we are solving the packing problem, but again we do not have all the information in advance.
+The values of $c$ are all known to us in advance, however the number of variables and the precise way they must be packed is not known.
+At each point in the sequence for which the input is given, a new variable can be introduced to the algorithm.
+If $y_j$ is the variable introduced, the algorithm is at this point made aware of precisely which $i \in [n]$ satisfy $i \in S(j)$.
+So the constraints are revealed over time.
+The algorithm must decide on the value of $y_j$ at the point it is introduced and may not change it in subsequent iterations.
+Under this definition, for any online covering problem, there is a corresponding online packing problem that, if run simultaneously, have their sub-instances at each point in the sequence the input is given acting as primal-dual pairs.
+\subsection{General Algorithm}
+\paragraph{}
+Consider the following algorithm for simultaneously solving a corresponding pair of online packing/covering problems of the form $(P)$ and $(D)$ from the previous subsection.
+By scaling, we may assume that each $c_i \geq 1$.
+We denote the algorithm by $\cA$:
+\begin{enumerate}
+\item When a new constraint for $(P)$ of the form $\sum_{i \in S(j)} x_i \geq 1$ arrives with corresponding dual variable $y_j$ do:
+\item While $\sum_{i \in S(j)} x_i < 1$:
+	\begin{enumerate}
+	\item For each $i \in S(j)$: $x_i \leftarrow x_i(1+\frac{1}{c_i}) + 1(|S(j)|\cdot c_i)$.
+	\item $y_j \leftarrow y_j +1$.
+	\end{enumerate}
+\end{enumerate}
+\paragraph{}
+We now proceed to analyze this algorithm.
+The critical notion we will use here is that of a $\textit{competitive online algorithm}$.
+An online algorithm $\cA$ (or its solution) is said to be $c$-competitive if for every instance of the minimization problem $\cA$ solves, the cost of the solution produced by $\cA$ is at most $c\cdot OPT + \alpha$ where $OPT$ is the optimal value of solving the ``offline" version of the problem, and $\alpha$ is an additive factor independent of the sequence in which the input is presented to $\cA$.
+If $\cA$ instead solves a maximization problem, to be $c$-competitive the cost of the solution produced instead needs to be at least $\frac{1}{c} OPT -\alpha$.
+\begin{theorem}\label{th:1}
+Let $d = \max_{j \in [m]} |S_(j)|$. Then $\cA$ produces two things:
+\begin{enumerate}
+\item An $O(\log d)$-competitive fractional covering solution,
+\item and a $2$-competitive integral packing solution which violates each packing constraint by at most $O(\log d)$.
+\end{enumerate}
+Note: We can rescale our packing solution by $O(\log d)$ to obtain a feasible fractional packing if desired.
+\end{theorem}
+\begin{proof}
+The dual variables start at $0$ and increase by $1$ whenever they are changed.
+Thus the packing solution returned is integral.
+By an iteration of $\cA$ we mean a complete execution of step $2$.
+Let $x$ be the solution for $(P)$ returned by $\cA$.
+Let $y$ be the solution for $(D)$ returned by $\cA$.
+We want to show three things:
+\begin{enumerate}
+\item $x$ is feasible for $(P)$.
+\item In each iteration of $\cA$ the increase in $c^Tx$ is at most twice the increase in $\sum_{j=1}^m y_j$.
+\item for all $i \in [n]$, $\sum_{j : i \in S(j)} y_j \leq c_i + O(\log d)$.
+\end{enumerate}
 
-\subsection{General Algorithm}
-
+\end{proof}
 \section{Ad Auctions}

### 2017-04-11 CS798-ConvexityAndOptimization/assignments/project.tex
proof for first alg

@@
 \end{enumerate}
-
+\paragraph{}
+Part $1.$ is quite clear.
+Whenever a primal constraint is introduced, the While loop of Step $2$ does not terminate until the constraint is satisfied.
+Subsequent iterations do not decrease the variables of $x$, so the constraint remains satisfied.
+\paragraph{}
+To demonstrate $2.$ we will study a single iteration of the While loop.
+In an iteration of Step $2$ the change in $\sum_{j=1}^m y_j$ is precisely $1$.
+The change in the primal objective value $c^Tx$ is (from the update in Step $2a$):
+$$\sum_{i \in S(j)} c_i (\frac{x_i}{c_i} + \frac{1}{|S(j)| c_i}) = \sum_{i \in S(j)}(x_i + \frac{1}{|S(j)|}) \leq 2$$
+since $\sum_{i \in S(j)} x_i \leq 1$ at the time Step $2a$. Therefore $2.$ holds.
+\paragraph{}
+Let $i \in [n]$.
+We consider the dual constraint corresponding to $i$.
+Whenever we increase a dual variable $y_j$ such that $i \in S(j)$ by $1$ in Step $2b$ we had just increased $x_i$ in Step $2a$.
+We claim that
+$$x_i \geq \frac{1}{d}((1+\frac{1}{c_i})^{\sum_{j : i \in S(j)} y_j} - 1).$$
+We will proceed by induction on the iterations of $\cA$.
+In the base case $x_i = 0$, and $\sum_{j : i \in S(j)} y_j = 0$ so the claim holds.
+Now consider an iteration in which some dual variable $y_k$ with $k \in S(j)$ is increased.
+Let $x_i'$ denote the value of $x_i$ after this iteration (and $x_i$ denote the value before).
+\begin{align*}
+x'_i &= x_i(1 + \frac{1}{c_i}) + \frac{1}{|S(j)|c_i} \\
+&\geq x_i(1+\frac{1}{c_i}) + \frac{1}{d\cdot c_i} \\
+&\geq \frac{1}{d}((1+\frac{1}{c_i})^{\sum_{j : i \in S(j)\backslash\{k\}}y_j} - 1)(1 + \frac{1}{c_i})^y_k + \frac{1}{d\cdot c_i} &\text{by the inductive hypothesis on $x_i$} \\
+&= \frac{1}{d}((1+\frac{1}{c_i})^{\sum_{j :i \in S(j)} y_j} -1).
+\end{align*}
+So the claim holds by induction. Now we compute an upper bound on $x'_i$. By the condition of the While loop, $x_i <1$. We also have $c_i \geq 1$ and $d\geq 1$, so
+$$x'_i =  x_i(1 + \frac{1}{c_i}) + \frac{1}{|S(j)|c_i} \leq 1(1+1) + 1 = 3.$$
+Combining inequalities we have
+$$ \frac{1}{d}((1+\frac{1}{c_i})^{\sum_{j :i \in S(j)} y_j} -1) \leq 3.$$
+Using once more that $c_i \geq 1$ and rearranging for $\sum_{j : in \in S(j)} y_j$ we obtain
+$$\sum_{j : i \in S(j)} y_j \leq c_i \log(3d+1) = c_i \cdot O(\log d)$$
+as desired.
+\paragraph{}
+It remains to use these claims to verify the competitive factors.
+This is done via duality of linear programs.
+Let $p^*$ be the optimal value of $(P)$.
+Let $f^*$ be the optimal value of $(D)$.
+By claim $2$ we have
+$$c^Tx \leq 2 \sum_{j} y_j.$$
+Since $x$ is primal feasible, $c^Tx \geq p^*$. So 
+$$2\sum_j y_j \geq c^Tx \geq p^* = f^*.$$
+Thus $y$ is $2$-competitive. By claim $3$, $y' \in \R^m$ where $y'_j = y_j/O(\log d)$ is feasible for $(D)$.
+So similarly,
+$$c^Tx \leq 2 O(\log d) \sum_j y_j \leq O(\log d) f^* = O(\log d) p^*.$$
+Therefore $x$ is $O(\log d)$-competitive.
 \end{proof}

### 2017-04-11 CS798-ConvexityAndOptimization/assignments/project.tex
packing-covering section

@@
 The multiplicative weights update method studied in Lecture $9$ is a powerful and ubiquitous tool in algorithm design \cite{arora2012multiplicative}.
-This method is the workhorse behind the primal-dual paradigm in designing competitive online algorithms \cite{buchbinder2009design}.
+This method is the inspiration behind the critical step in the primal-dual paradigm for designing competitive online algorithms \cite{buchbinder2009design}.
 One major area of application for such algorithms is the design of auctions for selling ad space in search query results.
@@
 \item In each iteration of $\cA$ the increase in $c^Tx$ is at most twice the increase in $\sum_{j=1}^m y_j$.
-\item for all $i \in [n]$, $\sum_{j : i \in S(j)} y_j \leq c_i + O(\log d)$.
+\item for all $i \in [n]$, $\sum_{j : i \in S(j)} y_j \leq c_i \cdot O(\log d)$.
 \end{enumerate}
@@
 \end{proof}
+\paragraph{Connection to Multiplicative Weights}
+If we looks at Step $2a$ of $\cA$, we observe multiplicative update rule very similar to those used in the study of the Multiplicative Weights Update Method of Lecture $9$:
+$$x_i \leftarrow x_i(1+\frac{1}{c_i}) + 1(|S(j)|\cdot c_i).$$
+So each $x_i$ involved in constraint $j$ is increased multiplicatively, while $y_j$ increases additively during the operation of the While loop in Step $2$.
+Notice that during the proof for the competitive factor we used an induction obtaining that $x_i$ is approximately an exponential function of the $y_j$'s:
+$$x_i = \frac{1}{d}((1+\frac{1}{c_i})^{\sum_{j : i \in S(j)} y_j} -1),$$
+similar to how we observe an exponential relationship between the potential function (in the proof for the performance guarantee in multiplicative weights update) and the variables of the multiplicative weights update procedure.
+In both scenarios these exponentials that arise from the respective multiplicative update rules allow us to derive logarithmic performance factors.
+The difference being that factor is additive in multiplicative update method, and is multiplicative in primal-dual.
 \section{Ad Auctions}

### 2017-04-12 CS798-ConvexityAndOptimization/assignments/project.tex
starting in on describing ad auctions

@@
 \paragraph{Connection to Multiplicative Weights}
-If we looks at Step $2a$ of $\cA$, we observe multiplicative update rule very similar to those used in the study of the Multiplicative Weights Update Method of Lecture $9$:
+If we looks at Step $2a$ of $\cA$, we observe a multiplicative update rule very similar to those used in the study of the Multiplicative Weights Update Method of Lecture $9$:
 $$x_i \leftarrow x_i(1+\frac{1}{c_i}) + 1(|S(j)|\cdot c_i).$$
@@
 \section{Ad Auctions}
+\paragraph{}
+The business of displaying ads in search engine results is a natural and important application for online algorithms.
+Search provides like Google and Microsoft will sell ad space to advertisers in the search results corresponding with some users given keyword query.
+The way they sell the space is through auctions to buyers interested in displaying ads related to the given query.
+
 \subsection{Single Slot}
-
+\paragraph{}
+In the $\textit{ad-auctions problem}$ we are given a set $I$ of $n$ buyers, each with a daily budget of $B(i)$.
+We are also given a set $M$ of items with $|M| = m$.
+The items arrive in an online manner.
+Upon the arrival of some item, say the $j$-th item for $j \in [m]$, each buyer will submit a bid for said item, denoted $b(i,j)$.
+The algorithm can decide to allocate the item to any of the buyer upon its arrival.
+The online constraint is that they cannot reallocate items once they have decided who they are to be allocated to.
+\paragraph{}
+Allocations can be either $\textit{fractional}$ or $\textit{integral}$.
+In integral allocation, each item is allocated to a single buyer.
+In fractional allocations the items can be subdivided and allocated to multiple buyers.
+The algorithm we present will give an integral allocation, but the linear programming formulation we base it on will be fractional.
+Clearly, integral allocations suffice as fractional allocations, so we can naturally come our integral solutions to the optimal values of the fractional linear programs.
+\paragraph{}
+The objective is to maximize the total revenue of the seller.
+The revenue received from each buyer is the minimum of the sum total of bids corresponding to items allocated to that buyer and their budget.
+It is defined as such since we cannot overcharge buyers.
 \subsection{Multiple Slots}

### 2017-04-12 CS798-ConvexityAndOptimization/assignments/project.tex
Wrote up single slot

@@
 
-\lhead{2017-04-09}
+\lhead{\today}
 \rhead{William Justin Toth CS798-Convexity and Optimization Project} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
@@
 It is defined as such since we cannot overcharge buyers.
+\paragraph{}
+We will denote by $(D)$ the $\textit{fractional ad-allocation}$ linear program, defined as follows:
+\begin{align*}
+\max \sum_{j =1}^m \sum_{i=1}^n &b(i,j) y(i,j) \\
+\text{s.t. } \sum_{i=1}^n y(i,j) &\leq 1 &\forall j \in [m] \\
+\sum_{j=1}^m b(i,j)y(i,j) &\leq B(i) &\forall i \in [n] \\
+y(i,j)&\geq 0 &\forall i \in [n], \forall j \in [m]. 
+\end{align*}
+The variables of $(D)$ are the values $y(i,j)$.
+The meaning of $y(i,j)$ is the fraction of item $j$ that is allocated to buyer $i$.
+The objective function maximizes our revenue, assuming that we do not overcharge any buyer.
+The second collection of constraints ensures just that: that we do not charge any buyer more than their budget $B(i)$.
+The first set of constraints ensures that each item is not over-allocated.
+That is, you can allocated at most one of item $j$.
+We observe that $(D)$ is a packing problem, similar to the simplified packing problem studied in the previous section.
+\paragraph{}
+Since $(D)$ is a packing problem, its dual is a covering linear program which we will denote $(P)$. The linear program $(P)$ is as follows:
+\begin{align*}
+\min \sum_{i=1}^nB(i)x(i) &+ \sum_{j=1}^m z(j) \\
+\text{s.t. } b(i,j)x(i) + z(j) &\geq b(i,j) &\forall i \in [n], \forall j \in [m] \\
+x(i) &\geq 0 &\forall i \in [n] \\
+z(j) &\geq 0 &\forall j \in [m].
+\end{align*}
+We describe the following primal-dual algorithm, $\cA$, for ad-allocation (initially all variables are set to $0$):
+\begin{enumerate}
+\item When a new item $j$ is given, allocate $j$ to buyer $i$ which maximizes $b(i,j)(1-x(i))$.
+\item If $x(i) < 1$ then do
+	\begin{enumerate}
+	\item Charge buyer $i$ the minimum of $b(i,j)$ and the remainder of $B(i)$ not yet charged. Set $y(i,j) = 1$
+	\item Set $z(j) = b(i,j)(1-x(i))$.
+	\item Set $x(i) = x(i)\frac{1+b(i,j)}{B(i)} + \frac{b(i,j)}{(c-1)B(i)}$.
+	\end{enumerate} 
+\end{enumerate}
+The value of $c$ above is taken to be $(1+R_{max})^\frac{1}{R_{max}}$ where $R_{max} = \max_{i \in I, j\in M} \frac{b(i,j)}{B(i)}$.
+\begin{theorem}
+The algorithm $\cA$ is $(1-\frac{1}{c})(1-R_{max})$-competitive for the ad-allocation problem. Note that when $R_{max} \rightarrow 0$, $\cA$ tends to being $(1-\frac{1}{e})$-competitive.
+\end{theorem}
+\begin{proof}
+As in the proof of theorem \ref{th:1} we prove three claims about feasibility and the relative change in objective value for $(P)$ and $(D)$ in each iteration:
+\begin{enumerate}
+\item The solution returned by $\cA$ for $(P)$ is feasible.
+\item The increase in the solution for $(P)$ in each iteration is at most $(1+\frac{1}{c-1})$ times in the increase in the solution for $(D)$ during that iteration.
+\item The solution returned for $(D)$ is ``nearly feasible". We will derive the precise bounds when we reach this step of the proof. 
+\end{enumerate}
+\paragraph{}
+Consider a constraint of $(P$) for buyer $i$ and item $j$:
+$$b(i,j)x(i) + z(j) \geq b(i,j).$$
+Suppose $x$ with $z$ is the solution for $(P)$ returned by $\cA$.
+If $x(i) \geq 1$ then the constraint is satisfied immediately.
+Otherwise let $i'$ be the buyer $\cA$ allocates $j$ to during operation.
+The buyer $i'$ maximizes $b(i',j)(1-x(i'))$ by Step $1$.
+In Step $2b$ $z(j)$ is set to $b(i',j)(1-x(i')).$ It is never changed in later iterations.
+By our choice of $i'$,
+$$z(j) \geq b(i,j)(1-x(i))$$
+and hence the constraint for $i$ and $j$ is satisfied by $x$ and $z$. Therefore we have shown claim $1$.
+\paragraph{}
+In each iteration of $\cA$, the increase in value for $(D)$ is $b(i,j)$ where $i$ is the buyer the item $j$ corresponding to the given iteration is allocated to.
+This follows since we always set $y(i,j)$ to $1$.
+By step $2b$ the increase in $z$ is
+$$b(i,j) (1-x(i)).$$
+By step $2c$ the increase in $x$ is
+$$(\frac{1+b(i,j)}{B(i)}-1)x(i) + \frac{b(i,j)}{(c-1)B(i)}.$$
+So the increase in the objective value for $(P)$ is
+\begin{align*}
+&B(i)[(\frac{1+b(i,j)}{B(i)}-1)x(i) + \frac{b(i,j)}{(c-1)B(i)}] + b(i,j)(1-x(i)) \\&= x(i) +b(i,j)x(i) - B(i)x(i) +\frac{b(i,j)}{(c-1)} + b(i,j) -b(i,j)x(i)\\
+&\leq b(i,j)(1+\frac{1}{c-1}).
+\end{align*}
+Therefore claim $2$ holds.
+\paragraph{}
+We want to show that there is at most one iteration per buyer $i$ that we charge them in a manner which exceeds their budget.
+In particular during the last iteration where we charge $i$, this can happen.
+To prove this we observe that $\cA$ never increases $y(i,j)$ when $j$ has been allocated to $i$ yet $x(i) \geq 1$.
+So we want to show that
+$$\sum_j b(i,j)y(i,j) \geq B(i)$$
+implies that $x(i) \geq 1$. It suffices to show that 
+\begin{equation}\label{eq:1}
+x(i) \geq \frac{1}{c-1}(c^{\frac{\sum_j b(i,j)y(i,j)}{B(i)}} -1)
+\end{equation}
+and when $\sum_j b(i,j)y(i,j) \geq B(i)$ this implies $x(i) \geq 1$.
+We proceed by induction on the iterations in which an item is allocated to buyer $i$.
+Before any item has been allocated to buyer $i$, the right hand side of (\ref{eq:1}) is $0$ and so the claim holds.
+Now consider the inductive case: suppose we are in some iteration where item $k$ is allocated to buyer $i$.
+Let $x'$ be the value of $x$ after this iteration and $x$ will denote the prior value.
+Then
+\begin{align*}
+x'(i) &= x(i)(1+\frac{b(i,k)}{B(i)}) + \frac{b(i,k)}{(c-1)B(i)} \\
+&\geq \frac{1}{c-1}(c^{\frac{\sum_{j\neq k} b(i,j)y(i,j)}{B(i)}} -1)(1+\frac{b(i,k)}{B(i)}) + \frac{b(i,k)}{(c-1)B(i)}  &\text{by inductive hypothesis}\\
+&= \frac{1}{c-1}(c^{\frac{\sum_{j\neq k} b(i,j)y(i,j)}{B(i)}}(1+\frac{b(i,k)}{B(i)}) - 1) \\
+&\geq \frac{1}{c-1}(c^{\frac{\sum_{j\neq k} b(i,j)y(i,j)}{B(i)}}c^{\frac{b(i,k)}{B(i)}} -1) \\
+&=\frac{1}{c-1}(c^{\frac{\sum_{j} b(i,j)y(i,j)}{B(i)}} -1).
+\end{align*}
+When $\frac{b(i,k)}{B(i)} = R_{max}$, the second inequality holds with equality. From this we derived our choice of $c$.
+\paragraph{}
+Now we have shown that once the sum of charges for items allocated to any buyer exceeds their budget, we no longer charge the buyer.
+So their charge can exceed their budget by at most the cost of one bid.
+Hence for each buyer $i$:
+$$\sum_j b(i,j)y(i,j) \leq B(i) + \max_j b(i,j).$$
+This is precisely how much $(D)$ is violated by in the solution $y$ returned by $\cA$. So we have claim $3$.
+\paragraph{}
+Now the revenue we receive from buyer $i$ is at least
+$$(\sum_j b(i,j)y(i,j))) \frac{B(i)}{B(i) + \max_j b(i,j)} \geq (\sum_j b(i,j)y(i,j)) (1-R_{max}).$$
+From claim $2$ the value of $y$ for $(D)$ present above is at least $1-\frac{1}{c}$ times the value of $(P)$.
+Since $x$ and $z$ is feasible by claim $1$, the value of $(P)$ is an upper bound on the optimal value of $(D)$.
+Together this demonstrates the competitive ratio.
+\end{proof}
 \subsection{Multiple Slots}

### 2017-04-13 CS798-ConvexityAndOptimization/assignments/project.tex
multiple slot case write up

@@
 Now the revenue we receive from buyer $i$ is at least
-$$(\sum_j b(i,j)y(i,j))) \frac{B(i)}{B(i) + \max_j b(i,j)} \geq (\sum_j b(i,j)y(i,j)) (1-R_{max}).$$
+$$(\sum_j b(i,j)y(i,j)) \frac{B(i)}{B(i) + \max_j b(i,j)} \geq (\sum_j b(i,j)y(i,j)) (1-R_{max}).$$
 From claim $2$ the value of $y$ for $(D)$ present above is at least $1-\frac{1}{c}$ times the value of $(P)$.
@@
 \subsection{Multiple Slots}
+\paragraph{}
+Now we extend to a more general problem.
+As before we will have $n$ buyers and $m$ items, but now each item will have $\ell$ ``slots".
+Think of slots as different positions for an ad on a given page of search results corresponding to some keyword.
+Naturally, different positions may have different value to advertisers, so we allow different bids for different slots.
+Thus each buyer $i \in [n]$ submits a bid for each slot $k \in[\ell]$ of item $j \in [m]$: $b(i,j,k)$.
+We fractionally allocate slots just as we did before for items.
+Thus our general form for $(D)$ is
+\begin{align*}
+\max \sum_{j =1}^m \sum_{i=1}^n \sum_{k=1}^\ell&b(i,j,k) y(i,j,k) \\
+\text{s.t. } \sum_{i=1}^n y(i,j,k) &\leq 1 &\forall j \in [m], \forall k \in [\ell]. \\
+\sum_{j=1}^m \sum_{k=1}^\ell b(i,j,k)y(i,j,k) &\leq B(i) &\forall i \in [n] \\
+y(i,j,k)&\geq 0 &\forall i \in [n], \forall j \in [m], \forall k \in [\ell]. 
+\end{align*}
+The dual to $(D)$, denoted $(P)$ is now
+\begin{align*}
+\min \sum_{i=1}^nB(i)x(i) &+ \sum_{j=1}^m\sum_{k=1}^\ell z(j,k) + \sum_{i=1}^n\sum_{j=1}^m s(i,j) \\
+\text{s.t. } b(i,j,k)x(i) + z(j,k) + s(i,j) &\geq b(i,j,k) &\forall i \in [n], \forall j \in [m] , \forall k \in [\ell]\\
+x(i) &\geq 0 &\forall i \in [n] \\
+z(j,k) &\geq 0 &\forall j \in [m], \forall k \in [\ell] \\
+s(i,j) &\geq 0 &\forall i \in [n], \forall j \in [m].
+\end{align*}
+The generalized algorithm, $\cA$, is as follows:
+\begin{enumerate}
+\item When a new item $j$ is given, do:
+	\begin{enumerate}
+	\item Let $H$ be a bipartite graph with vertex set $[n] \cup [\ell]$. Each edge of $H$ is of the form $ik$ with $i \in [n]$ and $k \in [\ell]$.  The weight of edge $ik$ is $b(i,j,k)(1-x(i))$.
+	\item Find a maximum weight integral matching in $H$. Such a matching forms an assignment on the variables $y(i,j,k)$.
+	\item Charge buyer $i$ the minimum of $\sum_{k=1}^\ell b(i,j,k)y(i,j,k)$ and the remainder of $B(i)$ not yet charged.
+	\item For each $i \in [n]$, for each $k \in [\ell]$ with $y(i,j,k) > 0$ update  $$x(i) = x(i)(1+ \frac{b(i,j,k)y(i,j,k)}{B(i)}) + \frac{b(i,j,k)y(i,j,k)}{(c-1)B(i)}.$$
+	\end{enumerate} 
+\end{enumerate}
+One can observe that this algorithm is indeed a proper generalization of the previous algorithm.
+In the case where $\ell = 1$ (ie.
+each item has one slot), the multiple slot algorithm computes a maximum weight matching in a very simple graph: where there is one vertex in the item partition, and all the buyers in the other.
+It amounts to picking the max weight bidder as we did in the previous algorithm.
+Thus the operation of the multiple slot algorithm reduces to the operation of the single slot algorithm in the case where $\ell = 1$.
+\paragraph{}
+We will need the linear programs for maximum weight matching, and its dual in our analysis of $\cA$.
+Fix some item $j$ and some previously set $x \in \R^n$.
+Then the maximum weight matching linear program for $H$, denoted $(M)$ is
+\begin{align*}
+\max \sum_{i=1}^n \sum_{k=1}^\ell b(i,j,k)(1-x(i))&y(i,j,k) \\
+\text{s.t.} \sum_{i=1}^ny(i,j,k) &\leq 1 &\forall j \in [\ell] \\
+\sum_{k=1}^\ell y(i,j,k) &\leq 1 &\forall i \in [n] \\
+y(i,j,k) &\geq 0 &\forall i \in [n], \forall k \in [\ell].
+\end{align*}
+The dual of $(M)$ is a vertex cover problem $(C)$ of the form
+\begin{align*}
+\min \sum_{i=1}^n s(i,j) &+ \sum_{k=1]}^\ell z(j,k) \\
+\text{s.t.} s(i,j) + z(j,k) &\geq b(i,j,k)(1-x(i)) &\forall i \in [n], \forall k \in [\ell] \\
+z(j,k) &\geq 0 &\forall k \in [\ell] \\
+s(i,j) &\geq 0 &\forall i \in [n].
+\end{align*}
+It is a well known theorem that $(M)$ always has integral optimal solutions, and thus by strong duality the integral matching found along with dual vertex cover in $H$  satisfies:
+\begin{equation}\label{eq:sd}
+\sum_{i=1}^n \sum_{k=1}^\ell b(i,j,k)(1-x(i))y(i,j,k) = \sum_{i=1}^n s(i,j) + \sum_{k=1]}^\ell z(j,k)
+\end{equation}
+The theorem will use this connection critically in its analysis, but otherwise follows the same structure as the proof in the single slot case.
+\begin{theorem}
+The algorithm $\cA$ is $(1-\frac{1}{c})(1-R_{max})$-competitive for the online ad allocation problem with multiple slots.
+\end{theorem}
+\begin{proof}
+As in the single slot case we will show the following three claims:
+\begin{enumerate}
+\item The solution returned by $\cA$ for $(P)$ is feasible.
+\item The increase in the solution for $(P)$ in each iteration is at most $(1+\frac{1}{c-1})$ times in the increase in the solution for $(D)$ during that iteration.
+\item The solution returned for $(D)$ is ``nearly feasible". We will derive the precise bounds when we reach this step of the proof. 
+\end{enumerate}
+Then, just before we can achieve the desired competitive ratio using weak duality.
+The proof for claims $1.$ and $3.$ are directly analogous to their proofs in the single slot case.
+So we shall focus our attention on the proof of claim $2.$ where strong duality is used.
+\paragraph{}
+Consider the iteration for item $j$. The increase in $x(i)$ during this iteration is 
+$$ \sum_{k=1}^\ell (x(i)\frac{b(i,j,k)y(i,j,k)}{B(i)} + \frac{b(i,j,k)y(i,j,k)}{(c-1)B(i)})$$
+for each $i \in [n]$. So the increase in $(P)$ is
+$$\sum_{i=1}^n z(j,i) + \sum_{k=1}^\ell s(i,j) +\sum_{i=1}^n\sum_{k=1}^\ell B(i) (x(i)\frac{b(i,j,k)y(i,j,k)}{B(i)} + \frac{b(i,j,k)y(i,j,k)}{(c-1)B(i)})$$
+Using strong duality equation (\ref{eq:sd}) this is equal to
+$$\sum_{i=1}^n \sum_{k=1}^\ell b(i,j,k)(1-x(i))y(i,j,k) + \sum_{i=1}^n \sum_{k=1}^\ell (b(i,j,k)x(i)y(i,j,k) + \frac{b(i,j,k)y(i,j,k)}{c-1}) $$
+Simplifying this gives
+$$\sum_{i=1}^n \sum_{k=1}^\ell b(i,j,k)y(i,j,k)(1+\frac{1}{c-1}).$$
+Now by step $1c$ the increase in $(D)$ is precisely
+$$\sum_{i=1}^n \sum_{k=1}^\ell b(i,j,k)y(i,j,k)$$
+and so claim $2.$ holds.
+\end{proof}
 \bibliography{references}

### 2017-04-13 CS798-ConvexityAndOptimization/assignments/project.tex
drafted project

@@
 \end{proof}
+\section{Conclusions}
+\paragraph{}
+Putting things together, we can synthesize a general approach to using the primal-dual method for designing competitive online algorithms.
+First one formulates the offline version of their problem as a primal-dual pair of linear programs such that the introduction of a new part of the input in the online model corresponds to the introduction of some new constraints in the primal and some new variables to the dual.
+Then design an algorithm which additively decides the value of the new dual variables, and multiplicatively updates the primal variables in the new primal constraints until they are satisfied.
+This multiplicative update will secure a high quality competitive factor if chosen properly.
+\paragraph{}
+The analysis of such general algorithms above proceeds via three claims.
+First that the primal solution is feasible.
+Second that the increase in the primal in each iteration can be bounded by some factor of the increase in the dual.
+The tighter you get this bound, the better your competitive ratio.
+Third that the dual solution violates its constraints by some small factor.
+That factor will appear in your competitive ratio and thus should be as tight as possible.
+The three claims together can be filtered through weak duality to provide a bound on the competitive ratio with factors coming from claim $2$ and $3$.
+\paragraph{}
+For the ad-auction problems we gave algorithms for both the single-slot and multi-slot versions with a competitive ratio that tends to $(1-\frac{1}{e})$.
+These were first studied in \cite{buchbinder2007online}.
+In \cite{mehta2007adwords} they demonstrate that the ratios we give are in fact tight, showing that $(1-\frac{1}{e})$ is a lower bound for these problems.
+\paragraph{}
+Some work has been done on going beyond the lower bounds by instituting additional assumptions for ad-auctions that are reasonable in real world applications.
+These are detailed in \cite{buchbinder2009design}.
+One possibility is to assume that each bidder is likely to spend some fraction of their budget, say $g_i$.
+Such an assumption can be verified by observing past behaviour of the bidder in the real world.
+In this case, known as the stochastic information case, one can use primal-dual with some modifications to achieve a competitive ratio of $$1-\frac{1-g}{e^{1-g}}$$
+where $g = \min_{i} g_i$.
+Another assumption that allows one to get below the lower bound is the bounded degree assumption.
+Here we assume that the buyer is interested in at most $d$ of the items.
+Usually $d$ is much smaller than $m$.
+In this case a ratio of
+$$1-\frac{d-1}{d(1+\frac{1}{d-1})^{d-1}}$$
+can be achieved.
 \bibliography{references}

### 2017-04-17 CS798-ConvexityAndOptimization/assignments/assignment_03.tex
a3

--- 2017-04-03 CO750-ApproximationAlgorithmsTopics/assignments/assignment_03.tex
@@
 
-\lhead{2017-04-07}
-\rhead{W. Justin Toth CO750-Approximation Algorithms Assignment 3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{\today}
+\rhead{William Justin Toth CS798-Convexity and Optimization Assignment 3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-
 %Question 1
 \section{}
-\paragraph{}
-Let $G=(V,E)$ be a metric graph with costs $c: E \rightarrow \R_+$.
-Let $T \subseteq V$ be a subset of nodes.
-Let $\equiv_2$ denote equivalence modulo $2$ (that is $a \equiv_2 b$ if and only if $\exists k \in \Z$ such that $a = b + 2k$).
-Suppose that $|T| \equiv_2 0$.
-\paragraph{}
-The minimum cost connected $T$-Join problem asks us to find a minimum cost multiset $F$ of edges such that the set of nodes with odd degree in the multigraph $H=(V,F)$ is exactly $T$ and $H$ is connected.
-\paragraph{}
-Let $\cA$ denote the following algorithm for solving minimum cost connected $T$-Join on $G$:
-\begin{enumerate}
-\item Compute a minimum cost spanning tree on $G$. Call it $S$.
-\item Let $A \subseteq T$ be the set of nodes in $T$ of even degree in $S$.
-\item Let $B \subseteq V\backslash T$ be the set of nodes not in $T$ of odd degree in $S$.
-\item Let $T' = A \cup B$. Compute a minimum cost $T'$-join in $G$. Call it $J$.
-\item Return $S \cup J$.
-\end{enumerate}
-\begin{theorem}
-The algorithm $\cA$ is a $\frac{5}{3}$-approximation algorithm for the minium cost connected $T$-join problem.
-\end{theorem}
-\begin{proof}
-Standard algorithms for minimum cost spanning tree and $T$-join run in polynomial time.
-Hence $\cA$ is a polynomial time algorithm.
-The multigraph $(V, S\cup J)$ is connected since it contains a spanning tree $S$.
-Further by our choice of $A$ and $B$, the nodes in $T$ are precisely the nodes of odd degree in $(V, S \cup J)$.
-With respect to correctness of $\cA$, the only question which remains is that $|T'| \equiv_2 0$ so that a $T'$-join can be computed.
-\paragraph{}
-Let any $C \subseteq V$ let $C^o := \{ v \in C: d_S(v) \equiv_2 1 \}$ where $d_S$ denotes $|\delta(v) \cap S|$. Then by our choice of $A$ and $B$: 
-$$|T| = |A| + |T^o| = |A| + |V^o\backslash B| = |A| + |V^o| - |B|.$$
-Since $|T| \equiv_2 0$ and $|V^o| \equiv_2 0$ (it is the set of odd degree vertices in $(V,S)$ so its cardinality is even):
-$$|A| \equiv_2 |B|,$$
-Thus $$|T'| \equiv_2 |A| + |B| \equiv_2 2|A| \equiv_2 0.$$
-So we have verified that $|T'|$ is even. Therefore the algorithm $\cA$ terminates in polynomial time returning a feasible solution to minimum cost connected $T$-Join.
-\paragraph{}
-We now verify the approximation factor. Let $J^*$ denote an optimal minimum cost connected $T$-join. Since $(V,J^*)$ is connected, $J^*$ contains a spanning tree. Therefore 
-$$c(S) \leq c(J^*).$$
-Now since $(V,S)$ is connected it contains a $T'$-Join.
-Let $J_1$ be such a $T'$-Join.
-Also since $(V,J^*)$ is connected it contains a $T'$-Join.
-Let $J_2$ be such a $T'$-Join.
-We claim that $J_3 := S\backslash J_1 \cup J^* \backslash J_2$ is a $T'$-join.
-To verify this consider a vertex $v \in T'$.
-We will study the parity of $d(v)$ with respect various sets mentioned above.
-Since $J_1$ and $J_2$ are $T'$-Joins we have
-$$d_{J_1}(v) \equiv_2 d_{J_2}(v) \equiv_2 0.$$
-Thus
-$$d_{J_3}(v) \equiv_2 d_S(v) + d_{J^*}(v).$$
-Since $J^*$ is a $T$-Join we have
-$$d_{J^*}(v) \equiv_2 \begin{cases}
-1, &\text{if } v \in V\backslash T \\
-0, &\text{if } v \in T.
-\end{cases}$$
-By our choice of $T'$, $v$ is either in $A\subseteq T$ or $B\subseteq V\backslash T$. So the case breakdown for $d_{J^*}(v)$ is more precisely:
- $$d_{J^*}(v) \equiv_2 \begin{cases}
-1, &\text{if } v \in B \\
-0, &\text{if } v \in A.
-\end{cases}$$
-By our choice $A$ and $B$ we have
-$$d_{S}(v) \equiv_2 \begin{cases}
-1, &\text{if } v \in B \\
-0, &\text{if } v \in A.
-\end{cases}$$
-Now observe that in either case ($v \in A$ or $v\in B$)
-$$d_{J_3}(v) = d_S(v) + d_{J^*}(v) = 0.$$
-Thus $J_3$ is a $T'$-Join as claimed.
-\paragraph{}
-Since $J_1$, $J_2$, and $J_3$ are $T'$-Joins, and $J$ is a minimum cost $T'$-Join,
-$$3c(J) \leq c(J_1) + c(J_2) + c(J_3).$$
-By our choices of $J_1$, $J_2$, and $J_3$ their union of edges is a subset of $S \cup J^*$. So we have
-$$c(J_1) + c(J_2) + c(J_3) \leq c(S) + c(J^*) \leq 2c(J^*).$$
-Combining inequalities we see that
-$$3c(J) \leq 2c(J^*).$$
-Therefore 
-$$c(S) + c(J) \leq c(J^*) + \frac{2}{3}c(J^*) = \frac{5}{3}c(J^*)$$
-and so the approximation factor holds as desired.
-\end{proof}
+
 %Question 2
 \section{}
-\paragraph{}
-Let $M$ be a set of machines and let $J$ be a set of jobs.
-For each $j \in J$ let $M_j \subseteq M$ denote the subset of machines $j$ can be processed on.
-Further for each $j \in J$, let $p_j \in \N$ denote the processing time of job $j$ on any machine in $M_j$.
-We consider the scheduling problem, which we will denote $(SP)$, that asks us to find a schedule minimizing the makespan: the maximum completion time of any machine provided we process all jobs.
-We will formalize this objective momentarily.
-\paragraph{}
-The order we process jobs on a machine is irrelevant in this problem, and we cannot split jobs in pieces to give to different machines.
-So a solution to $(SP)$ is an assignment $a: J \rightarrow M$.
-Let $A$ be the set of all feasible assignments.
-That is $A = \{ a : \forall j \in J, a(j) \in M_j\}.$ Then $(SP)$ can be stated as
-$$\min_{a \in A} \max_{m \in M} \sum_{j : a(j) = m} p_j.$$
-\paragraph{}
-Given an instance of $(SP)$ and a tentative makespan value $s \in \N$ we can construct an instance of Unsplittable Flow, which we'll denote $(UF)$, which we can use to decide if $(SP)$ has an assignment yielding a makespan of value $s$.
-The vertex set will consist of a source $r$, a node $v_m$ for every $m \in M$, and a terminal node $t_j$ for every $j\in J$.
-The demands for each $t_j$ are the processing times $p_j$.
-The arc set consists of an arc $(r,v_m)$ for every $m \in M$.
-The capacity of each $(r,v_m)$ will be the makespan $s$.
-The arc set will also contain an arc $(v_m, t_j)$ for each $j \in J$ and $m \in M_j$.
-The capacity of each $(v_m, t_j)$ will be $p_j$.
-We will let $c$ denote the capacity function.
-Let $D$ be the resulting directed graph.
-\begin{lemma}\label{lemma:sp-uf}
-An instance of $(SP)$ has an assignment of makespan at most $s$ if and only if the corresponding instance of $(UF)$ has a solution.
-Moreover the translation between the two problems can be done in polynomial time.
-\end{lemma}
-\begin{proof}
-\paragraph{$(\implies)$} Let $a$ be the assignment which yields a makespan of value at most $s$.
-For each $j \in J$ define $P_j = \{ (r,a(j)),(a(j),t_j)\}$.
-That is, $P_j$ is the path in $D$ (the directed graph representing $(UF)$) upon which we will route a flow of $p_j$.
-Each arc $(a(j), t_j)$ for $j \in J$ receives total flow $p_j$, hence respecting its capacity.
-Each arc $(r,m)$ for $m \in M$ receives $\sum_{j : a(j) = m}p_j$ total flow.
-Since $\sum_{j : a(j) = m} p_j \leq s$ the capacity of $(r,m)$ is respected.
-Thus we have a $(UF)$ solution.
-\paragraph{$(\impliedby)$}
-Let $\{P_j : j \in J\}$ be the set of paths given in the solution of $(UF)$ corresponding an instance of $(SP)$ and value $s$.
-We will construct an assignment $a$ for $(SP)$ that has makespan $s$.
-For each $j \in J$ set $a(j) = m$ if $(m,j) \in P_j$.
-Such an $m$ must exist, and is unique, by our construction of $D$.
-Since arcs incoming to $j$ are precisely those in $M_j$ this assignment is feasible.
-Now we consider the makespan of this assignment.
-Since the capacities on each arc $(r,m)$ for all $m \in M$ is respected by flow we have for all $m \in M$:
-$$\sum_{j: a(j) = m} p_j = \sum_{j: (m,j) \in P_j} p_j = f(r,m) \leq c(r,m) = s.$$ 
-Thus $a$ has makespan at most $s$. Clearly the translation given between the solutions of $(UF)$ and $(SP)$ can be done in polynomial time.
-\end{proof}
-\paragraph{}
-In our algorithm we will also use the following critical lemma from class.
-\begin{lemma}\label{lemma:ff-uf}
-Given an instance of $(UF)$ with demands $d_1, \dots, d_k$, if $f$ is a fractional flow realizing the demands then there exists an unsplittable flow $\bar{f}$ realizing the demands such that $\bar{f}(e) \leq f(e) + d_{max}$ for all $e \in E$, where $d_{max} = \max_{i\in [k]} d_i$.
-Moreover this unsplittable flow can be found in polynomial time.
-\end{lemma}
-\paragraph{}
-We give an approximation algorithm for $(SP)$. The algorithm performs a binary search to find the optimal makespan value. Let $\cA$ be the algorithm for $(SP)$ that follows:
-\begin{enumerate}
-\item Let $\ell = 0$ and let $u = \sum_{j \in J} p_j$.
-\item While $\ell \neq u$ do:
-	\begin{enumerate}
-	\item Construct the corresponding $(UF)$ instance with value $s=\floor{\frac{\ell + u}{2}}$.
-	\item Use a standard max flow algorithm to find a fractional flow $f$ on the $(UF)$ instance.
-	\item If $f$ satisfies the demands for all $t_j$ then set $\ell = s$
-	\item Otherwise set $u = s$.
-	\end{enumerate}
-\item Using lemma \ref{lemma:ff-uf} compute $\bar{f}$ from $f$.
-\item Using lemma \ref{lemma:sp-uf} compute an assignment $a$ from $\bar{f}$ (with makespan at most $2s$). Return $a$.
-\end{enumerate}
-\begin{theorem}
-The algorithm $\cA$ is a $2$-approximation algorithm for $(SP)$.
-\end{theorem}
-\begin{proof}
-From lemma \ref{lemma:sp-uf} we know that $\cA$ will return a feasible solution.
-Each inner iteration of the While loop runs in polynomial time since we have polynomial time algorithms for max flow.
-Further we know from lemma \ref{lemma:ff-uf} that step $3$ can be done in polynomial time, and from lemma \ref{lemma:sp-uf} that step $4$ can be done in polynomial time.
-It remains to verify that the binary search procedure runs in polynomial time.
-For our choice of $\ell$ and $u$ the while loop will run in $O(\log(\sum_{j \in J} p_j))$ iterations.
-The question is whether this is polynomial in the number of bits used to represent our input.
-\paragraph{}
-We know each $p_j$ uses approximately $\log(p_j)$ bits to be represented in memory.
-Thus $\sum_{j \in J} \log(p_j)$ is a polynomial in the input size.
-But we may assume for the purpose of counting input size that each $p_j \geq 2$.
-We may do so because each $p_j$ uses at least one bit of memory to represent itself, yet $\log(1) = 0$.
-Now, if each $p_j \geq 2$ then $\Pi_{j\in J} p_j \geq \sum_{j \in J} p_j$.
-Thus $\sum_{j \in J} \log(p_j) \geq \log(\sum_{j \in J} p_j)$, and hence the binary search runs in polynomial time.
-\paragraph{}
-We have verified that $\cA$ runs in polynomial time returning a feasible solution.
-It remains to verify the approximation factor of $2$.
-Let $s^*$ be the makespan value of an optimal solution to $(SP)$.
-The instance of $(UF)$ constructed in the last iteration of the binary search has the maximal $s$ such that a fractional flow is present on $(UF)$ satisfying all the demands.
-Since the existence of an unsplittable flow implies the existence of a fractional flow satisfying the demands, by lemma \ref{lemma:sp-uf} using $s^*$, $$s \leq s^*.$$
-\paragraph{}
-By lemma \ref{lemma:ff-uf}, $\cA$ returns a solution of makespan $2s$. This follows since the maximum capacity of an arc in our instance of $(UF)$ is $s$. Hence we return a solution of value
-$$2s \leq 2s^*.$$
-Therefore the approximation factor holds.
-\end{proof}
+
 %Question 3
 \section{}
-\paragraph{}
-In an instance of $(WTAP)$ we are given a tree $G=(V,E)$ and a set $L \subseteq V\times V$, called ``links".
-We are also given a cost function $c: L \rightarrow \R_+$.
-The goal is to find a minimum cost subset $\bar{L} \subseteq L$ such that $(V, E \cup \bar{L})$ is $2$-edge-connected.
-%Part (a) of Q3
-\subsection{a}
-We are given the following algorithm for $(WTAP)$, as usual we will denote it $\cA$:
-\begin{enumerate}
-\item Select an arbitrary $r \in V$ as the root. Orient $E$ towards $r$, and set the $w(a) = 0$ for all the resulting arcs $a$.
-\item For each $uv \in L$ do:
-	\begin{enumerate}
-	\item Let $p$ be the lowest-common-ancestor of $u$ and $v$ on $G$.
-Add to $G$ the arcs $(p,u)$ and $(p,v)$ with $w(p,u) = w(p,v) = c(uv)$.
-Note that if $p=u$ or $p=v$ we only add one arc).
-We say the arcs $(p,u)$ and $(p,v)$ ``correspond" to the link $uv$.
-	\end{enumerate}
-\item Compute a minimum weight $out$-arborescence $R$ rooted at $r$ in the resulting digraph from steps $1$ and $2$.
-\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that corresponds to $\ell$.
-\end{enumerate}
-
-\begin{lemma}\label{lemma:3a-poly}
-The algorithm $\cA$ returns a feasible solution for $(WTAP)$ on $G$ with links $L$ in polynomial time.
-\end{lemma}
-\begin{proof}
-First we check that the algorithm terminates in polynomial time.
-Step $1$ can be done in $O(|E|)$ time.
-Step $2$ uses $O(|L|)$ iterations, and each iteration can use a depth-first-search to find the lowest-common ancestor in polynomial time.
-Hence Step $2$ runs in polynomial time.
-We have seen an exact polynomial time algorithm in class for computing a minimum weight arborescence.
-Hence Step $3$ runs in polynomial time.
-Step $4$ only necessitates iterating over links of $L$ and arcs of $R$ and thus runs in polynomial time.
-Hence $\cA$ terminates in polynomial time.
-\paragraph{}
-Consider the solution $\bar{L}$ returned by the algorithm.
-Let $up \in E$ where $p$ is the parent of $u$ with respect to the orientation in Step $1$.
-Let $G_u$ be the connected component of $G-up$ which contains $u$ and let $G_p$ be the connected component of $G-up$ which contains $p$.
-To verify that $(V, E \cup \bar{L})$ is $2$-edge-connected we will show that there exists $\ell \in \bar{L}$ with one endpoint in $V(G_u)$ and one in $V(G_p)$.
-\paragraph{}
-Consider the directed path $P$ from $r$ to $u$ in the arborescence $R$ computed in Step $3$.
-Let $g$ be the last ancestor of $u$ on $P$ before reaching $u$.
-Then $(g,c) \in P$ where $c$ is either a child of $u$ or $c=u$.
-By our construction, since $(g,c)$ travels from a vertex closer to the root to a vertex farther from the root, $(g,c)$ corresponds to some link $\ell \in \bar{L}$.
-The endpoint $c$ of $\ell$ lies $G_u$.
-The other endpoint, say $v$, has their lowest common ancestor with $c$ being $g$.
-Thus $v \not \in V(G_u)$ and therefore $v \in V(G_p)$.
-Therefore $\ell$ is a link connected $G_u$ to $G_p$, and thus $(V, E \cup \bar{L})$ is $2$-edge-connected.
-\end{proof}
-\begin{theorem}
-The algorithm $\cA$ is $2$-approximation algorithm for $(WTAP)$.
-\end{theorem}
-\begin{proof}
-By lemma \ref{lemma:3a-poly} it remains to verify the approximation factor.
-Let $L^* \subseteq L$ be an optimal solution to $(WTAP)$ on $G$.
-Let $D$ denote the directed graph which results from the construction in Steps $1$ and $2$ of $\cA$.
-We will construct from $L^*$ an arc set $R^*$ of cost
-$$c(R^*)\leq 2c(L^*)$$
-which contains at $out$-arborescence rooted at $r$ on $D$.
-Then, since $R$ is a minimum weight $out$-arborescence on $G$ with $c(\bar{L}) = c(R)$ this will suffice to prove the approximation factor as $c(R) \leq c(R^*) \leq 2c(L^*)$.
-\paragraph{}
-For each $\ell \in L^*$ add the arcs which correspond to $\ell$ to $R^*$.
-Now we have $c(L^*) = 2c(R^*)$.
-It remains to fix connectivity in $R^*$ so that there is a directed path from $r$ to each vertex.
-We will fix it by adding $0$ cost arcs.
-For each $u \in V\backslash \{r\}$ do the following.
-Let $p$ be the parent of $u$ with respect to the orientation of step $1$.
-Let $G_u$ be the connected component of $G-up$ which contains $u$ and let $G_p$ be the connected component of $G-up$ which contains $p$.
-Since $L^*$ is feasible for $(WTAP)$ there exists some $ab \in L^*$ such that $a \in V(G_u)$ and $b \in V(G_p)$.
-Now $a$ is a descendant of $u$.
-Thus there exists a $0$-cost directed path from $a$ to $u$.
-Add this path to $R^*$.
-Note this does not change $c(R^*)$.
-\paragraph{}
-We claim that after doing this for each vertex, the connectivity requirement is satisfied.
-Let $u \in V$.
-We proceed by induction on the distance from $u$ to the root.
-In the base case $r=u$ and so there trivially is a path from $r$ to $u$ in $R^*$.
-Now suppose $u$ is at least at distance $1$ from $r$, and for all ancestors $u$ there a directed path from $r$ to $u$'s ancestor using arcs of $R^*$.
-Let $p$, $a$, and $b$ be as in the previous paragraph with respect to $u$.
-Since $a$ is in $G_u$ and $b$ is in $G_p$ the lowest common ancestor of $a$ and $b$ is some vertex $g$ that is an ancestor of $u$.
-By induction there is a directed path $P_{rg}$ from $r$ to $g$ using arcs of $R^*$.
-By construction the arc $(g,a)$ is in $R^*$.
-So $P_{rg}$, $(g,a)$, and the $0$-cost path from $a$ to $u$ added to $R^*$ in the previous paragraph forms a directed path from $r$ to $u$ in $R^*$.
-Hence the claim holds.
-\paragraph{}
-So $R^*$ contains a directed path from $r$ to every vertex of $D$. Therefore $R^*$ contains an $out$-arborescence, and as argued in the first paragraph this suffices to prove the approximation factor.
-\end{proof}
-%Part b of Q3
-\subsection{b}
-\paragraph{}
-Now we change $(WTAP)$ so that the requirement is $(V, E\cup \bar{L})$ is $2$-vertex-connected.
-We may assume we are not in the degenerate case where we have a graph consisting of only two vertices connected by an edge.
-Such a tree is $2$-vertex-connected by default.
-We will modify the algorithm $\cA$ as follows, the big changes are that we take a leaf to be the root, and we ``shorten" the downward arcs, and we add ``cross" arcs:
-\begin{enumerate}
-\item Select an arbitrary leaf $r \in V$ as the root. Orient $E$ towards $r$, and set the $w(a) = 0$ for all the resulting arcs $a$.
-\item For each $uv \in L$ do:
-	\begin{enumerate}
-	\item Let $p$ be the lowest-common-ancestor of $u$ and $v$ on $G$.
-Add to $G$ the arcs $(p,u)$ and $(p,v)$ with $w(p,u) = w(p,v) = c(uv)$.
-Note that if $p=u$ or $p=v$ we only add one arc).
-We say the arcs $(p,u)$ and $(p,v)$ ``correspond" to the link $uv$.
-	\item If neither $u$ nor $v$ are each others' ancestor then  add to $G$ the arcs $(u,v)$ and $(v,u)$.
-Set $w(u,v) = w(v,u) = c(uv)$.
-We also say the arcs $(u,v)$ and $(v,u)$ ``correspond to the link $uv$.
-Note that at most four arcs are generated between this step and the previous step for each link.
-	\item For each downward arc ($p$ is an ancestor of $u$) $(p,u)$ do: if $p$ is not a leaf let $c_p$ be the child of $p$  that succeeds $p$ on the path to $u$ (otherwise let $c_p = p$).
-Replace the arc $(p,u)$ with the arc $(c_p, u)$ with the same weight, and same corresponding link as $(p,u)$.
-	\end{enumerate}
-\item Compute a minimum weight $out$-arborescence $R$ rooted at $r$ in the resulting digraph from steps $1$ and $2$.
-\item Output the set $\bar{L} \subseteq L$ such that $\ell \in \bar{L}$ iff there exists an arc in $R$ that corresponds to $\ell$.
-\end{enumerate}
-\begin{lemma}\label{lemma:3b-poly}
-The algorithm $\cA$ returns a feasible solution for $(WTAP)$ on $G$ with links $L$ in polynomial time.
-\end{lemma}
-\begin{proof}
-As argued in lemma \ref{lemma:3a-poly} the algorithm $\cA$ runs in polynomial time, it is clear that the additional processing runs in polynomial time.
-To verify feasibility suppose for a contradiction that the graph $H := (V, E \cup \bar{L})$ has a vertex $p$ whose deletion disconnects the graph.
-Let $g$ be the parent of $p$.
-The vertex $g$ exists since we choose $r$ to be a leaf and hence $p \neq r$.
-Let $u_1, \dots, u_k$ be the children of $p$.
-For each $v \in V$, let $C(v,p)$ denote the connected component of $H-p$ containing vertex $v$.
-Let $A := \{u \in \{u_1, \dots, u_k\} : u \in C(g,p)\}$ be the set of children of $p$ which are connected to $g$ after removing $p$.
-We claim that $A \neq \emptyset$.
-\paragraph{}
-Clearly $r \in C(g,p)$, as are all ancestors of $p$.
-Now since $R$ is an $out$-arborescence there exists a directed path from $r$ to $\bigcup_{i=1}^k u_i$.
-The arc such a path uses to enter $\bigcup_{i=1}^k u_i$ is a downward arc, from either $p$ or  some ancestor of $p$, $a$, to a descendant of $p$, $d$.
-Let $u$ be a child of $p$ such that $d \in C(u,p)$.
-If the arc is of the form $(a,d)$ then the corresponding link  in $\bar{L}$ connects a vertex of $C(g,p)$ to some descendant of a child $u$.
-Hence $C(g,p)$ is connected to $C(u)$.
-That is $u \in A$ and hence $A \neq \emptyset$.
-In the case the arc is of the form $(p,d)$, then by step $2(c)$, the arc $(p,d)$ arose from a downward arc $(g,d)$.
-Thus there is a link in $\bar{L}$ connecting a vertex of $C(g,p)$ to $d$.
-Thus $C(g)$ is connected to $C(u)$.
-Therefore our claim that $A$ is not empty holds.
-\paragraph{}
- Let $\bar{A} = \{u_1, \dots, u_k\} \backslash A$.
-By our contradiction assumption, $\bar{A}\neq \emptyset$ as otherwise removing $p$ does not disconnect $H$.
-Now, again since $R$ is an $out$-arborescence there exists a directed path from $r$ to $\bigcup_{u \in \bar{A}} C(u)$.
-The directed arc which enters $\bigcup_{u \in \bar{A}} C(u)$ is either a downward arc or a cross arc arising from step $2(b)$.
-In the case it is a downward arc, by the same argument as the previous paragraph the corresponding link in $\bar{L}$ connects some $C(v, p)$ for $v \in \bar{A}$ to some $C(s,p)$ for $s \in A \cup \{g\}$.
-But this contradicts the definition of $\bar{A}$.
-In the case of a cross arc, the head is necessarily in $C(g,p)$ and the tail in some $C(v,p)$ for $v \in \bar{A}$.
-Thus, as in the previous case, the corresponding link in $\bar{L}$ connects some $C(v, p)$ for $v \in \bar{A}$ to some $C(s,p)$ for $s \in A \cup \{g\}$.
-In either case a contradiction is reached since the corresponding components to vertices in $\bar{A}$ cannot be connected to $C(g,p)$ by definition.
-Therefore no vertex exists whose deletion disconnects the graph.
-\end{proof}
-\begin{theorem}
-The algorithm $\cA$ is $2$-approximation algorithm for $(WTAP)$.
-\end{theorem}
-\begin{proof}
-By lemma \ref{lemma:3b-poly} it remains to verify the approximation factor.
-Let $L^* \subseteq L$ be an optimal solution to $(WTAP)$ on $G$.
-Let $D$ denote the directed graph which results from the construction in Steps $1$ and $2$ of $\cA$.
-We will construct from $L^*$ an arc set $A^*$ of cost
-$$c(A^*)\leq 4c(L^*)$$
-which contains at $out$-arborescence $R^*$ rooted at $r$ on $D$ with cost $c(R^*) \leq 2c(L^*)$.
-Then, since $R$ is a minimum weight $out$-arborescence on $G$ with $c(\bar{L}) = c(R)$ this will suffice to prove the approximation factor as $c(R) \leq c(R^*) \leq 2c(L^*)$.
-\paragraph{}
-For each $\ell \in L^*$ add the arcs which correspond to $\ell$ (with respect to Step $2$) to $A^*$.
-At most four arcs are added so we have $c(A^*) \leq 4c(L^*)$.
-Suppose we also add to $A^*$ the $0$-cost arcs formed in Step $1$ of $\cA$.
-We want to show that every vertex is reachable from $r$ by a directed path in $A^*$.
-Then we can compute an arborescence on $A^*$ and use that to bound the optimal solution.
-\paragraph{}
-We claim that for every $r' \in V$, $r'$ can reach, by a directed path using arcs of $A^*$, every vertex in the subtree rooted at $r'$.
-We proceed by induction on the number of vertices of the subtree rooted at $r'$.
-In the base case the subtree is of size $1$, and hence $r'$ is a leaf and the claim is immediate.
-Suppose the claim holds for all subtrees with less vertices that the subtree rooted at $r'$.
-Let $d$ be a descendant of $r'$.
-We have two cases: either $r' = r$ or not.
-In the case $r'\neq r$, the deletion of $r'$ disconnects $G$.
-Let $u_1, \dots, u_k$ be the children of $r'$.
-Since $r'$ is not the root $r$, $r'$ has some parent $p$.
-Let $C_{u_1}, \dots, C_{u_k}, C_p$ denote the respective connected components of $G$ after deleting $r'$.
-Suppose without loss of generality that $d \in C_{u_1}$.
-Since $(V, E \cup L^*)$ is $2$-vertex-connected, if we think of $C_{u_1}, \dots, C_{u_k}, C_p$ as vertices, there is a subset of links $T \subseteq L^*$ which forms a spanning tree on $C_{u_1}, \dots, C_{u_k}, C_p$.
-There exists a path from $C_{u_1}$ to $C_g$  using links of $T$.
-We may assume without loss generality that this path is of the form $C_{u_1}, C_{u_2}, \dots, C_{u_k}, C_p$ since we may simply ignore connected components which do not join this path (decreasing $k$) and relabelling the children.
-\paragraph{}
-Now, for $i \in [k-1]$ the link between $C_{u_{i}}$ and $C_{u_{i+1}}$ induces a directed arc in $A^*$ from $C_{u_{i+1}}$ to $C_{u_i}$ by Step $2(b)$.
-Call this arc $t_i$, The link between $C_{u_k}$ and $C_p$ induces a directed arc from an ancestor of $u_k$ to a vertex of $C_{u_k}$ by Step $2(a)$ and $2(c)$.
-Call this arc $t_k$.
-Now we describe the directed path to $d$.
-From $r'$ take the up-arcs of $0$-cost to the head of $t_k$ (note the head of $t_k$ may be $r'$, in which case take an empty path).
-Then follow $t_k$ to a vertex of $C_{u_k}$.
-We use the following process to traverse $C_k$: Use the $0$-cost arcs to reach $u_k$.
-Use the inductive hypothesis to reach the head of $t_{k-1}$, and follow $t_{k-1}$ to a vertex of $C_{k-1}$.
-Repeat this process for each $C_i$ from $C_{k-1}$ to $C_1$.
-After doing so the directed path has reached the tail of $t_1$.
-Again follow the $0$-cost arcs up to $u_1$, and use induction to find a directed path from $u_1$ to $d$.
-All together this process yields a directed path from $r'$ to $d$.
-\paragraph{}
-In the induction step the case where $r' =r$ is simpler.
-Let $u$ be the child of $r'$.
-Then $d$ is a descendant of $u$ or equal to $u$.
-In either case, since $L^*$ is feasible there exists some link from $r$ to a descendant of $u$.
-This link induces an arc from $r$ to a descendant $d'$ of $u$ in $A^*$.
-Since $r$ is a leaf, the head of this arc is not-moved from $r$ in step $2(c)$.
-To reach $d$ from $r$, simply traverse the arc from $r$ to $d'$ then traverse the $0$-cost path from $d'$ up to $u$ and use the inductive hypothesis to find a directed path from $u$ to $d$.
-The resulting concatenated directed path connects $r$ to $d$.
-Hence by induction, the root $r$ is connected to every vertex of $D$ by a directed path using only arcs of $A^*$.
-\paragraph{}
-Therefore we can form a arborescence on $D$, say $R^*$ which is a subset of $A^*$.
-Choose the minimum cost such $R^*$.
-We claim that $R^*$ uses at most two arcs corresponding to any given link $uv \in L^*$.
-Suppose for a contradiction $R^*$ uses at least three arcs corresponding to $uv$.
-If $R^*$ uses both $(u,v)$ and $(v,u)$ then those two arcs form a directed cycle in $R^*$, contradicting that $R^*$ is acyclic.
-Hence $R^*$ uses one of $(u,v)$ or $(v,u)$, as well as the two downward arcs formed.
-Say without loss of generality $R^*$ uses $(u,v)$.
-Let $p$ be the lowest common ancestor of $u$ and $v$.
-Let $p_u$ and $p_v$ be the respective children of $p$ on the path to $u$ and $v$ respectively.
-So $R^*$ uses the arcs $(p_u,u)$ and $(p_v,v)$ as well as $(u,v)$ since $R^*$ uses at least three corresponding arcs.
-But then $R^*$ has two directed paths from $r$ to $v$.
-The first from $r$ to $p_u$, and then along $(p_u,u)$ and $(u,v)$ to reach $v$.
-The second along the directed path from $r$ to $p_v$ then along $(p_v, v)$ to reach $v$.
-So arc $(u,v)$ is not needed for connectivity in $R^*$, contradicting the optimality of $R^*$.
-\paragraph{}
-Therefore we have formed an arborescence $R^*$ such that $c(R^*) \leq 2c(L^*)$. Since $R$ is a minimum cost arborescence and $c(\bar{L}) = c(R)$ we have
-$$c(\bar{L}) = c(R) \leq c(R^*) \leq 2c(L^*).$$
-So the approximation factor holds as desired.
-\end{proof}
-%Question 4
-\section{}
 \end{document}

### 2017-04-18 CS798-ConvexityAndOptimization/assignments/assignment_03.tex
q1

@@
 \section{}
-
+Let $K \subseteq R^n$ be a convex body with $rB^n_2 \subseteq K \subseteq RB^n_2$ for some $0 < r < R$. Suppose there exists a separation oracle for $K$. Consider the following algorithm, $\cA$:
+\begin{enumerate}
+\item Start with $\cE = RB^n_2$. Throughout we use $v_1, \dots, v_n$ to denote the semi-axes of $\cE$, $\lambda_1, \dots, \lambda_n$ the respective side-lengths, and $c$ the centre of $\cE$.
+\item Repeat:
+	\begin{enumerate}
+	\item For each $p \in \{ c+ \frac{v_i}{n+1} : i \in [n]\} \cup \{c-\frac{v_i}{n+1} : i \in [n]\}$:
+		\begin{enumerate}
+		\item Call separation oracle on $p$.
+		\item If $p \not \in K$ let $v^Tx = b$ be the hyperplane returned by the oracle (with $v^Tp > b$).
+``Shrink" $\cE$ with respect to $v^Tx=b$ (to be detailed) and $\cE$ to be the new ellipsoid from shrinking.
+Continue to the next iteration of Step $2$.
+		\end{enumerate}
+	\item Return $\cE$ (each $p \in K$).
+	\end{enumerate}
+\end{enumerate}
+\paragraph{Shrinking Process}
+As in the ellipsoid method, given some ellipsoid $\cE_t$ (with $K \subseteq \cE_t$) and a hyperplane $v^Tx = b$ we want to find a new ellipsoid $\cE_{t+1}$ such that $\cE \cap \{x : v^Tx \leq b\} \subseteq \cE_{t+1}$.
+Critically we want $vol(\cE_{t+1}) \leq O(\exp(\frac{-1}{n}))  vol(\cE_t)$.
+As we have seen in Lecture $15$, via an affine change of variables this problem can be reduced to one where $\cE_t$ is the unit ball centred at the origin.
+We use the same change of variables here and detail the process in the special case below.
+Although we will get a different estimate on the volume this reduction is still valid since the critical point is that the volume proportions are invariant under an affine map.
+\paragraph{}
+As in the lecture notes, we will use $\cB = \cE_t$ to denote the origin-centered unit ball.
+We may assume that $||v|| =1$ (we use $2$-norm throughout).
+We will use $\cE$ to denote $\cE_{t+1}$.
+We will describe $\cE$ as
+$$\cE := \{x \in \R^n : (x-c)^TH^{-1}(x-c) \leq 1\}.$$
+The centre $c$ will be at $-\alpha v$ for some $\alpha \in [0,1]$. We will use $v$ as one semi-axes, while the others shall be orthogonal to $v$ of the same side length. Hence we have
+$$H^{-1} = avv^T + b(I - vv^T)$$
+where $a$ is the inverse side length square of $v$, and $b$ is the inverse side length square of the other semi-axes.
+To ensure $\cE$ satisfies the desired containment, and yet is as small as possible we want the boundary of $\cE$ to touch $-v$ and the points on the boundary of $B$ that intersect the hyperplane $v^Tx = b$.
+The first condition yields
+$$(-v+\alpha v)^T(avv^T + b(I-vv^T))(-v+\alpha v) = 1$$
+which implies
+$$a = \frac{1}{(\alpha-1)^2}.$$
+The second condition implies that for all $y$ orthogonal to $v$, with $||y + \frac{n}{n+1}v || =1$ we have
+$$(y + (\frac{n}{n+1} + \alpha)v)^T(avv^T + b(I-vv^T))(y+(\frac{n}{n+1} + \alpha)v) = 1.$$
+This condition implies
+$$ b = \frac{1-\beta^2a}{||y||^2}$$
+where $\beta = \frac{n}{n+1} + \alpha.$ Note that we may use the pythagorean theorem to find $||y||$:
+$$||y||^2 =\frac{2n+1}{(n+1)^2}.$$ 
+We hope to bound the ratio of $vol(\cE)$ to $vol(B)$. We have the equation
+$$\frac{vol(\cE)}{vol(\cB)} = \frac{1}{\sqrt{ab^{n-1}}} = \frac{||v||^{n-1}}{\sqrt{a(1-\beta^2a)^{n-1}}}.$$
+So we want to maximize
+$$f(\alpha) = a(1-\beta^2a)^{n-1}.$$
+Simplifying $1-\beta^2a$ we obtain
+$$1-\beta^2a = \frac{2n+1}{(n+1)^2}(\frac{2(n+1)}{1-\alpha} - \frac{2n+1}{(1-\alpha)^2}.$$
+Setting $h = \frac{1}{1-\alpha}$ this gives
+$$1-\beta^2a = \frac{2(n+1)}{((n+1)^2}(2(n+1)h - (2n+1)h^2).$$
+Thus $f$ may be rewritten in terms of $h$ as
+$$f(h) = (\frac{2n+1}{(n+1)^2})^{n-1}[h^2(2(n+1)h - (2n+1)h^2)^{n-1}].$$
+To maximize $f$ we take the derivate and set $f'(h) = 0$. Solving for $h$ yields
+$$h = \frac{n+1}{2n+1}.$$
+Such $h$ allows us to find $\alpha$, $a$, and $b$. Thus we have found our ellipse. Now we need to verify the volume. Plugging $h$ into $f$ yields
+$$f(h) = (\frac{1}{2n+1})^{n-1}(\frac{4n+3}{2n+1})^{n-1}.$$
+Therefore
+\begin{align*}
+\frac{f(h)}{||y||^{2(n-1)}} &= (\frac{1}{2n+1})^{n-1}(\frac{4n+3}{2n+1})^{n-1}(\frac{(n+1)^2}{2n+1})^{n-1} \\
+&=(\frac{(n+1)^2(4n+3)}{(2n+1)^3})^{n-1} \\
+&= (\frac{4m^3 - m^2}{(2m-1)^3})^{m-2} &\text{where $m=n+1$} \\
+&\geq (\frac{1}{2} - \frac{1}{8m})^{m-2} \\
+&\geq (\frac{1}{2}- \frac{1}{8(n-1)})^{n-1} \\
+&\geq exp(\frac{1}{8(n-1)}).
+\end{align*}
+Therefore $\frac{vol(\cE)}{vol(\cB)} = O(exp(\frac{-1}{n})).$
+\paragraph{Runtime of Algorithm}
+Now we bound the oracle calls used by $\cA$.
+Each iteration of Step $2$ used $O(n)$ oracle calls in Step $2a$.
+The question now is how many times Step $2$ iterates.
+Let $\cE_t$ be the current ellipsoid at iteration $t$ of Step $2$.
+By the argument above (in big-$O$),
+$$vol(\cE_{t+1}) \leq exp(\frac{-1}{n}) vol(\cE_t).$$
+Let $t$ be the last iteration of $\cA$. With $\cE_0 = RB^n_2$ we have by induction
+$$vol(\cE_t) \leq exp(\frac{-1}{n})^t vol(RB^n_2).
+$$Since $rB^n_2 \subseteq K$, $vol(rB^n_2) \leq vol(\cE_t)$. Thus we have
+$$vol(rB^n_2) \leq exp(\frac{-1}{n})^t vol(RB^n_2)$$
+Solving for $t$ we have
+$$\frac{t}{n} = \log(\frac{R}{r})^n$$
+and so 
+$$t = O(n^2\log(\frac{R}{r})).$$
+Therefore the total number of oracle calls in $\cA$ is $O(n^3\log(\frac{R}{r}))$.
+\paragraph{Correctness}
+By our shrinking procedure, $K \subseteq \cE$. When $\cA$ terminates we have that
+$c\pm \frac{v_i}{n+1} \in K$ for all $i \in [n]$.
+Let $C$ be the convex hull of $\{c\pm \frac{v_i}{n+1} : i \in [n]\}$.
+So $C \subseteq K$.
+From the definition of $C$, we have that $C$ is a symmetric convex body.
+Since the vertices of $C$ are semiaxes endpoints of $\cE' := c + \frac{1}{n+1}(\cE - c)$, we have that $\cE'$ is the minimum volume ellipsoid containing $C$.
+Thus by John's theorem, $c + \frac{1}{\sqrt{n}}(\cE'-c) \subseteq C$.
+Hence
+$$c + \frac{1}{\sqrt{n}(n+1)}(\cE  - c) \subseteq C \subseteq K \subseteq \cE$$
+as desired. $\blacksquare$
 %Question 2

### 2017-04-18 CS798-ConvexityAndOptimization/assignments/assignment_03.tex
wroteup 3

@@
 &\geq (\frac{1}{2}- \frac{1}{8(n-1)})^{n-1} \\
-&\geq exp(\frac{1}{8(n-1)}).
+&\geq \exp(\frac{1}{8(n-1)}).
 \end{align*}
@@
 By the argument above (in big-$O$),
-$$vol(\cE_{t+1}) \leq exp(\frac{-1}{n}) vol(\cE_t).$$
+$$vol(\cE_{t+1}) \leq \exp(\frac{-1}{n}) vol(\cE_t).$$
 Let $t$ be the last iteration of $\cA$. With $\cE_0 = RB^n_2$ we have by induction
-$$vol(\cE_t) \leq exp(\frac{-1}{n})^t vol(RB^n_2).
+$$vol(\cE_t) \leq \exp(\frac{-1}{n})^t vol(RB^n_2).
 $$Since $rB^n_2 \subseteq K$, $vol(rB^n_2) \leq vol(\cE_t)$. Thus we have
-$$vol(rB^n_2) \leq exp(\frac{-1}{n})^t vol(RB^n_2)$$
+$$vol(rB^n_2) \leq \exp(\frac{-1}{n})^t vol(RB^n_2)$$
 Solving for $t$ we have
@@
 \section{}
+\paragraph{}
+Let $f(x) = \exp(-h(x))$ where $h$ is a differentiable function.
+\subsection{a}
+\paragraph{}
+We want to show $f$ is log-concave (ie. $h$ is convex) if and only if $f''(x)f(x) \leq (f'(x))^2$. We compute the first and second derivatives of $f$. The first derivative is
+$$f'(x) = \exp(-h(x))\cdot (-h'(x)).$$
+The second derivative is 
+$$f''(x) = \exp(-h(x)) \cdot (h'(x))^2 + \exp(-h(x))(-h''(x)) = \exp(-h(x))((h'(x))^2 - h''(x)).$$
+So the left hand side is
+$$f''(x)f(x) = \exp(-h(x))^2((h'(x))^2 - h''(x))$$
+and the right hand side is
+$$(f'(x))^2 = \exp(-h(x))^2(h'(x))^2.$$
+So we have have $f''(x)f(x) \leq (f'(x))^2$ if and only if $h''(x) \geq 0$ for all $x$.
+But this second order condition is satisfied if and only if $h$ is convex, and therefore $f$ is log-concave.$\blacksquare$
+\subsection{b}
+\paragraph{}
+Suppose $h$ is convex. Let $F(x) = \int_{-\infty}^x f(y) dy$. Let $x \in \R$ and suppose that $h'(x) < 0$. From the fundamental theorem of calculus, $F'(x) = f(x)$. Now let 
+$$I(x) = -h'(x)F(x)$$.
+Then $$I'(x) = -h''(x)F(x) -h'(x)F'(x) \leq -h'(x)f(x)$$
+since $h''(x) \geq 0$ as $h$ is convex. Taking antidervatives we see
+$$I(x) \leq f(x).$$
+Thus
+$$F(x) \leq \frac{f(x)}{-h'(x)}$$
+since $h'(x) < 0$. So we have
+$$\int_{-\infty}^xf(y)dy \leq \frac{\exp(-h(x))}{-h'(x)}$$
+as desired. $\blacksquare$
+\paragraph{}
+Now we show that $F$ is log-concave. We have
+$$F'(x) = f(x)$$
+and
+$$F''(x) = -h'(x)f(x).$$
+Therefore
+$$F''(x)F(x) = -h'(x)f(x) F(x)$$
+and
+$$(F'(x))^2 = f(x)^2.$$
+If $h'(x) \geq 0$ then 
+$$F''(x) F(x) \leq 0$$
+and so $F''(x)F(x) \leq (F'(x))^2$ immediately. Now if $h'(x) < 0$ then we use the inequality we just demonstrated, so
+$$F''(x) F(x) = -h'(x)f(x)F(x) \leq -h'(x)f(x) \cdot \frac{f(x)}{-h'(x)} = f(x)^2.$$
+Therefore by part $3a$, $F(x)$ is log-concave. $\blacksquare$
 \end{document}

### 2017-04-20 CS798-ConvexityAndOptimization/assignments/assignment_03.tex
problem 2 solved

--- 2017-04-18 CS798-ConvexityAndOptimization/assignments/assignment_03.tex
@@
 \section{}
-
+\paragraph{}
+Let $A \in \{\pm 1\}^{m\times n}$.
+Suppose that the rank of $A$ is $r$.
+Let $a^1, \dots, a^r \in \R^m$ be a basis for the rowspace of $A$.
+Then for each row $i \in [m]$ of $A$ there exists $\lambda_1^i, \dots, \lambda^i_r$ such that the row vector can be expressed as $\sum_{j=1}^r \lambda_j^i a^j$.
+Therefore if we let $u_i^T = \begin{bmatrix} \lambda^i_1 &\dots &\lambda^i_r\end{bmatrix}$ and collect $U = \begin{bmatrix} u_1 &\dots & u_m \end{bmatrix}$, and we let $$V = \begin{bmatrix}
+(a^1)^T \\ \dots \\ (a^r)^T
+\end{bmatrix}$$ then we can write $A$ as 
+$$A = U^T V.$$
+Let $v_1, \dots, v_n$ be the columns of $V$. Since $A \in \{\pm 1\}^{m \times n}$, observe that $$|u^T_i v_j | = 1$$ for all $i \in [m], j \in [n]$. This fact will be useful later.
+\paragraph{}
+Let $K$ be the convex hull of $\{ \pm u_i : i \in [m]\}$. Then $K$ is a symmetric convex body in $\R^r$. By John's theorem there exists an ellipsoid $\cE$, centered at $0$, such that
+$$ \frac{1}{\sqrt{r}} \cE \subseteq K \subseteq \cE.$$
+There exists some positive definite matrix $H$ such that we can write $\cE = \{ x : x^TH^{-1}x \leq 1\}$.
+Then $H^\frac{1}{2} (\cE) = \cB$ where $\cB$ is the Euclidean norm unit ball.
+So $\sqrt{r}H^\frac{1}{2}(\cE) = \sqrt{r}\cB$.
+Let $\overline{U} = \sqrt{r}H^\frac{1}{2}U$, and let $\overline{V} = \frac{1}{\sqrt{r}} H^\frac{-1}{2}V$.
+Then
+$$\overline{U}^T\overline{V} = U^TH^\frac{1}{2}H^\frac{-1}{2}V = U^TV = A$$
+using that $H^\frac{1}{2}$ is symmetric.
+\paragraph{}
+For any matrix $M$ we use $col(M)$ to denote the set of column vectors of $M$. Since $col(U) \subseteq K$ and $$\sqrt{r}H^\frac{1}{2}(K) \subseteq \sqrt{r}H^\frac{1}{2} \cE = \sqrt{r} B$$
+we have $||\overline{u}|| \leq \sqrt{r}$ for all $\overline{u} \in col(\overline{U}) = col(\sqrt{r}H^\frac{1}{2}U)$. Let $\overline{v} \in col(\overline{V})$. Then 
+$$||\overline{v}||^2 = \overline{v}^T \overline{v} = \frac{1}{r}v^THv = \frac{1}{r} ||v||_H^2.$$
+Thus
+$$||\overline{v}|| = \frac{1}{\sqrt{r}} ||v||_H = ||v||_{\frac{1}{\sqrt{r}}H}.$$
+Now the norm $||\cdot ||_{\frac{1}{\sqrt{r}}H}$ is dual to the norm $||\cdot||_{\frac{1}{\sqrt{r}}H^{-1}}$ so 
+$$||\overline{v}|| = \sup \{ v^Tx : \frac{1}{\sqrt{r}} x^TH^{-1}x \leq 1\} = \sup_{x \in \frac{1}{\sqrt{r}}\cE} x^Tv.$$
+Since $\frac{1}{\sqrt{r}}\cE \subseteq K$ we have
+$$||\overline{v}|| \leq \sup_{x \in K} x^Tv.$$
+Since $K$ is the convex hull of $\{\pm u : u \in col(U)\}$, for all $x \in K$ there exists $\lambda \in \R^{2r}$ such that $\lambda \geq 0$, $\sum_{i=1}^{2r} \lambda_i = 1$ and $\sum_{i=1}^r \lambda_i u_i - \sum_{i=r+1}^{2r} \lambda_i u_i = x$.
+But, since $|u^Tv| =1$ for all $u \in col(U)$, this expression is maximized when $\lambda_{r+1} = \dots = \lambda_{2r} = 0$ and in such a case
+$$\sum_{i=1}^r \lambda_i u_iv^T \leq \sum_{i=1}^r\lambda_i\cdot 1 = 1.$$
+Hence
+$$||\overline{v}|| \leq \sup_{x \in K} x^Tv \leq 1.$$
+Therefore, for all $\overline{v} \in col(\overline{V})$, $||\overline{v}|| \leq 1\leq \sqrt{r}$, and for all $\overline{u} \in col(\overline{U})$, $||\overline{u}|| \leq \sqrt{r}$. Further
+$$A = \overline{U}^T\overline{V}$$
+as desired. $\blacksquare$
 %Question 3
 \section{}
+\paragraph{}
+Let $f(x) = \exp(-h(x))$ where $h$ is a differentiable function.
+\subsection{a}
+\paragraph{}
+We want to show $f$ is log-concave (ie. $h$ is convex) if and only if $f''(x)f(x) \leq (f'(x))^2$. We compute the first and second derivatives of $f$. The first derivative is
+$$f'(x) = \exp(-h(x))\cdot (-h'(x)).$$
+The second derivative is 
+$$f''(x) = \exp(-h(x)) \cdot (h'(x))^2 + \exp(-h(x))(-h''(x)) = \exp(-h(x))((h'(x))^2 - h''(x)).$$
+So the left hand side is
+$$f''(x)f(x) = \exp(-h(x))^2((h'(x))^2 - h''(x))$$
+and the right hand side is
+$$(f'(x))^2 = \exp(-h(x))^2(h'(x))^2.$$
+So we have have $f''(x)f(x) \leq (f'(x))^2$ if and only if $h''(x) \geq 0$ for all $x$.
+But this second order condition is satisfied if and only if $h$ is convex, and therefore $f$ is log-concave.$\blacksquare$
+\subsection{b}
+\paragraph{}
+Suppose $h$ is convex. Let $F(x) = \int_{-\infty}^x f(y) dy$. Let $x \in \R$ and suppose that $h'(x) < 0$. From the fundamental theorem of calculus, $F'(x) = f(x)$. Now let 
+$$I(x) = -h'(x)F(x)$$.
+Then $$I'(x) = -h''(x)F(x) -h'(x)F'(x) \leq -h'(x)f(x)$$
+since $h''(x) \geq 0$ as $h$ is convex. Taking antidervatives we see
+$$I(x) \leq f(x).$$
+Thus
+$$F(x) \leq \frac{f(x)}{-h'(x)}$$
+since $h'(x) < 0$. So we have
+$$\int_{-\infty}^xf(y)dy \leq \frac{\exp(-h(x))}{-h'(x)}$$
+as desired. $\blacksquare$
+\paragraph{}
+Now we show that $F$ is log-concave. We have
+$$F'(x) = f(x)$$
+and
+$$F''(x) = -h'(x)f(x).$$
+Therefore
+$$F''(x)F(x) = -h'(x)f(x) F(x)$$
+and
+$$(F'(x))^2 = f(x)^2.$$
+If $h'(x) \geq 0$ then 
+$$F''(x) F(x) \leq 0$$
+and so $F''(x)F(x) \leq (F'(x))^2$ immediately. Now if $h'(x) < 0$ then we use the inequality we just demonstrated, so
+$$F''(x) F(x) = -h'(x)f(x)F(x) \leq -h'(x)f(x) \cdot \frac{f(x)}{-h'(x)} = f(x)^2.$$
+Therefore by part $3a$, $F(x)$ is log-concave. $\blacksquare$
 \end{document}

### 2017-05-11 seminars/combopt-reading-group-s2017/shorter-tours-nicer-ears-outline.tex
starting outline

--- 2017-01-06 CO750-ApproximationAlgorithmsTopics/lectures/lecture_01.tex
@@
 \usepackage{fancyhdr}
+\usepackage{mathtools}
 \usepackage{algorithm}
 \usepackage{algpseudocode}
-\usepackage{mathtools}
 \usepackage{theorem}
@@
 
-\lhead{2017-01-03}
-\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{\today}
+\rhead{William Justin Toth - Shorter Tours by Nicer Ears - Outline} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \section{Introduction}
+\subsection{Problem Definitions}
+\paragraph{} Define $T$-tour and graph-TSP problem ($3$ equivalent views).
+The minimum $T$-tour problem, for which graph TSP is a special case.
+(Can also note path TSP is a special case).
+The $2$-edge-connected subgraph problem, which is a relaxation of graph TSP.
+\paragraph{} Give approximation results for the above problems. Will focus on graph TSP.
+\subsection{Preliminary Propositions}
 \paragraph{}
-The central task we are interested in is solving $NP$-hard optimization problems. Finding optimal solutions to such problems in polynomial time is unlikely (unless $P = NP$).
+Define $LP(G)$, the natural $LP$ relaxation for 2ECSS. Proposition $1.2$, the lower bound given by $LP(G)$.
+\paragraph{}
+For minimum $T$-tour, $LP(G)$ is not a lower bound, so we introduction the more general setting, based on cuts between multi-part partitions.
+Define $LP(G,T)$ and given lower bounds of Proposition $1.3$.
+\paragraph{}
+Proposition $1.4$, explaining how we may assume the graphs given are $2$-vertex connected.
+\section{Ear Decompositions}
 
-\paragraph{}
-Given that we are faced with an $NP$-hard problem there are a couple possible approaches:
-\begin{enumerate}
-\item Exact algorithms: Find optimal solution slowly for large input.
-\item Heuristic algorithms: Find efficiently a feasible solution, but with no guarantee of optimality.
-\item Approximation Algorithms: Find efficiently a feasible solution and give a certified bound on how close the solution is to optimal.
-\end{enumerate}
-The goal of this course is to study the third approach, and learn successful methods of designing approximation algorithms. We will draw heavily from network design for examples.
-
-\paragraph{}
-What we will see in this course:
-\begin{enumerate}
-\item Combinatorial techniques: Those which exploit the structure of feasible solutions.
-\item Linear Programming techniques:
-\begin{enumerate}
-\item Primal-Dual methods: Those which exploit Duality Theory.
-\item Rounding methods: Those which round fractional variables.
-\end{enumerate}
-\item Recent Results and Open Problems!
-\end{enumerate}
-\section{Vertex Cover Problem}
-\paragraph{}
-In an instance of {\it Vertex Cover} (VC) the input is a graph $G=(V,E)$.
-We may assume without loss of generality that $G$ is connected.
-If $G$ is not connected we may apply an algorithm which solves VC for connected graphs to each component of $G$ and union our solutions to obtain a solution for $G$.
-The output (solution) to an instance of VC is a subset of vertices $C \subseteq V$ of minimum cardinality satisfying $$\forall e \in E, e \cap C \neq \emptyset.$$
-\paragraph{}
-A possible exact algorithm would list all subsets of $V$ and return the minimum cardinality subset among those which are covers.
-The running time of this algorithm would be $O(2^n)$ where $n = |V|$.
-This is inefficient, ie.
-not polynomial in the size of the input graph.
-\paragraph{}
-An instance of a problem has a given size $m \in \Z$.
-The {\it running time} of the an algorithm is the number of elementary operations it performs ($+$, $-$, $\times$, $==$, $<$, $>$, $\dots$) during its execution.
-We measure the running time as a function of the input size $m$.
-An algorithm is said to be {\it efficient} if its running time is a polynomial function of $m$, ie.
-$O(m^k)$ for some constant $k$.
-For problems on graphs $n =|V|$ provides a rough estimate of the input size.
-\paragraph{}
-Let $\alpha \geq 1$ and let $\Pi$ be a minimization problem.
-We say $\cA$ is an {\it $\alpha$-approximation algorithm} for $\Pi$ provided $\cA$ is efficient on all instances of $\Pi$ and for every instance $\pi$ of $\Pi$, the solution returned by $\cA$ on $\pi$ is feasible and has objective value at most $\alpha$ times the optimal value of a solution for $\pi$.
-\paragraph{}
-If instead $\Pi$ is a maximation problem then $\alpha \leq 1$ and $\cA$ returns a solution at least $\alpha$ times the optimal value.
-\paragraph{}
-Typical values of $\alpha$ are $2$, $O(1)$, $\log n$, or $(1 + \epsilon)$ for instance.
-\paragraph{}
-Returning our attention to VC, let $C^*$ be an optimal solution of a given instance of $VC$.
-Our goal is to find a feasible solution $C$ such that $|C| \leq \alpha |C^*|$.
-But we run into some difficulty since we do not know $|C^*|$.
-Our idea to get around this is to obtain a lower bound on $|C^*|$ and compare $|C|$ with our lower bound.
-\paragraph{}
-Let $M$ be a maximum (we only need maximal in full generality) matching on $G$.
-It is not too hard to see that $|M| \leq |C^*|$.
-Indeed if $|C^*| < |M|$ then there exists two edges in $M$ which are covered by the same vertex in $C^*$ (from pigeonhole principle), but this contradicts that $M$ is a matching.
-Note that any matching provides a lower bound on $|C^*|$, but a maximum matching yields the tightest lower bound, a desirable property.
-\paragraph{}
-Via the Blossom Algorithm of Edmunds a maximum cardinality (in fact maximum weight for arbitrary weights) matching can be computed efficiently.
-\paragraph{Approximation Algorithm for VC}
-Consider the algorithm:
-\begin{enumerate}
-\item Compute a maximum cardinality matching $M$
-\item Output $C = \bigcup_{e \in M} e$.
-\end{enumerate}
-\begin{theorem}
-Our Approximation Algorithm for VC is a $2$-approximation for VC.
-\end{theorem}
-\begin{proof}
-Since maximum matching runs in polynomial time, our algorithm runs in polynomial time.
-To see that $C$ is feasible suppose that there exists $e \in E$ such that $e \cap C = \emptyset$.
-The $M \cup e$ is a matching greater cardinality than $M$.
-Therefore $C$ is a cover.
-Furthermore $$|C| = 2|M| \leq 2|C^*|$$ as desired.
-\end{proof}
-\paragraph{}
-To see that our analysis is tight consider a graph consisting of a single edge. In this case $|C| = 2 = 2|C*|$ and hence our analysis is tight.
-\section{Weighted Vertex Cover}
-\paragraph{}
-In an instance of {\it Weighted Vertex Cover} (WVC)the input is a connected graph $G = (V,E)$ and weights $w_v$ for all $v \in V$.
-The output is again a cover, but now of minimum weight, not cardinality.
-This problem generalizes VC whose instances can be recovered by assigned all weights the value $1$.
-\paragraph{}
-Our previous algorithm may fail.
-To see this consider a star graph with very large weight on the center vertex and very low weights on the outer vertices.
-The reader is encouraged to verify that the previous algorithm fails under such conditions.
-\paragraph{}
-For WVC we will use linear programming (LP) to give lower bounds. Often optimization problems can be naturally formulated as integer programs (IP). Consider the following IP for WVC:
-\begin{align*}
-\min &\sum_{v \in V} w_v x_v \\
-x_u + x_v &\geq 1 &\forall uv \in E \\
-x_u& \in \{0,1\} &\forall u \in V.
-\end{align*}
-By relaxing the IP (change constraints $x_u \in \{0,1\}$ to $0 \leq x_u \leq 1$) we obtain an LP that can be solved efficiently.
-Since this LP has $O(|V|^2)$ constraints we can solve it in time polynomial in the size of our input graph.
-This LP gives a lower bound on $OPT$, the optimal value of our WVC instance.
-\paragraph{Approximation Algorithm for WVC}
-Consider the algorithm:
-\begin{enumerate}
-\item Solve the LP relaxation. Let $x^*$ be the optimal solution of the LP.
-\item (Round up all variables of value at least $\frac{1}{2}$) Output: $C = \{v \in V: x^*_v \geq \frac{1}{2}\}$.
-\end{enumerate}
-\begin{theorem}
-Our Approximation Algorithm for WVC is a $2$-approximation for WVC.
-\end{theorem}
-\begin{proof}
-Algorithm is polynomial since compact linear programs can be solved in polynomial time. Now the weight of $C$ is
-$$w(C) = \sum_{v \in C} w_v = 2\sum_{v \in C} w_v \frac{1}{2} \leq 2 \sum_{v \in C}w_v x^*_v \leq 2 \sum_{v \in V} w_v x^*_v \leq 2\cdot OPT.$$
-It remains to verify that $C$ is feasible. Indeed for each edge $uv \in E$, either $x^*_u \geq \frac{1}{2}$ or $x^*_v \geq \frac{1}{2}$ by the constraint
-$$x_u + x_v \geq 1$$
-and thus either $u \in C$ or $v \in C$.
-\end{proof}
 \end{document}

### 2017-05-11 seminars/optimization-w2017/smpolytope-2017-01-25.tex
starting outline

--- 2017-01-23 optimization-w2017/smpolytope-2017-01-25.tex
@@
 \section*{Notation}
-\paragraph{}
-$G=(A\cup B, E)$ will denote a bipartite graph with vertex partitions $A$ and $B$. The vertex set of $G$ is denoted $V(G)$, and the edge set $E(G)$.
-\paragraph{}
-$\delta(v) = \{e \in E(G) : v \in e\}$ for all $v \in V(G)$.
-\paragraph{}
-$N(v) = \{w \in V(G): vw \in E(G)\}$ for all $V\in V(G)$.
-\paragraph{}
-$>_v$ denotes the strict total preference order of $v$ over $N(v)$. By convention $v$ prefers being matched to unmatched.
-\paragraph{}
-$\delta^{>w}(v) = \{vu \in \delta(v) : u >_v w\}$ for all $v,w \in V(G)$.
-\paragraph{}
-$N_{max}(v) = w$ if and only if $w \geq_v u$ for all $u \in N(v)$. Similarly $N_{min}(v) = w$ if and only if $w \leq_v u$ for all $u \in N(v)$.
-\paragraph{}
-Commonly $M \subseteq E$ such that $|\delta(v) \cap M| \leq 1$ for all $v \in V$ denotes a matching, and $M(v) = w $ if and only if $vw \in M$.
-For matchings $M_1$ and $M_2$, $M_1 \Delta M_2 = (M_1 \cup M_2) \backslash (M_1 \cap M_2)$ denotes their symmetric difference
-\paragraph{}
-$P(G)$ denotes the set of $x \in \R^{E(G)}$ satisfying
+$\mathbf{G=(A\cup B, E)}$ will denote a bipartite graph with vertex partitions $A$ and $B$. The vertex set of $G$ is denoted $\mathbf{V(G)}$, and the edge set $\mathbf{E(G)}$.
+\\
+$\mathbf{\delta(v)} = \{e \in E(G) : v \in e\}$ for all $v \in V(G)$.
+\\
+$\mathbf{N(v)} = \{w \in V(G): vw \in E(G)\}$ for all $V\in V(G)$.
+\\
+$\mathbf{>_v}$ denotes the strict total preference order of $v$ over $N(v)$. By convention $v$ prefers being matched to unmatched.
+\\
+$\mathbf{\delta^{>w}(v)} = \{vu \in \delta(v) : u >_v w\}$ for all $v,w \in V(G)$.
+\\
+$\mathbf{N_{max}(v)} = w$ if and only if $w \geq_v u$ for all $u \in N(v)$. Similarly $\mathbf{N_{min}(v)} = w$ if and only if $w \leq_v u$ for all $u \in N(v)$.
+\\
+Commonly $\mathbf{M} \subseteq E$ such that $|\delta(v) \cap M| \leq 1$ for all $v \in V$ denotes a matching, and $\mathbf{M(v)} = w $ if and only if $vw \in M$. 
+\\
+For matchings $M_1$ and $M_2$, $\mathbf{M_1 \Delta M_2} = (M_1 \cup M_2) \backslash (M_1 \cap M_2)$ denotes their symmetric difference
+\\
+$\mathbf{P(G)}$ denotes the set of $x \in \R^{E(G)}$ satisfying
 \begin{align*}
@@
 \end{align*}
-Where $x(M) = \sum_{e \in M} x_e$ for $M \subseteq E(G)$ and $x \in \R^{E(G)}$.
-\paragraph{}
-We let $\chi(M) \in \{0,1\}^{E(G)}$ be the characteristic vector of $M$, defined by $\chi(M)_e = 1$ if $e\in M$ and $\chi(M)_e = 0$ otherwise.
-\paragraph{}
-For $x \in P(G)$ we let $V_x$ denote the set of vertices corresponding to constraints of the form $(1)$ satisfied at equality by $x$.
-Similarly $E_x$ denotes the set of constraints of the form $(2)$ satisfied at equality by $x$.
+Where $\mathbf{x(M) }= \sum_{e \in M} x_e$ for $M \subseteq E(G)$ and $x \in \R^{E(G)}$.
+\\
+We let $\mathbf{\chi(M)} \in \{0,1\}^{E(G)}$ be the characteristic vector of $M$, defined by $\chi(M)_e = 1$ if $e\in M$ and $\chi(M)_e = 0$ otherwise.
+\\
+For $x \in P(G)$ we let $\mathbf{V_x}$ denote a set of vertices corresponding to constraints of the form $(1)$ satisfied at equality by $x$.
+\\Similarly $\mathbf{E_x}$ denotes a set of constraints of the form $(2)$ satisfied at equality by $x$.
+\\Together $|E_x| + |V_x| = |E|$ and the constraints corresponding to $E_x$ and $V_x$ are linearly independent.
 \section*{Outline (of proof $P(G)$ is integral)}
@@
 \item Show there exists $e \in E$ for which $x_e \in \{0,1\}$ where $x$ is an extreme point of $P(G)$.
-\item Proceed by induction on $|E(G)|$
-\item Case $x_e = 0$ using some Stable Matching basics and geometry of polytopes.
-\item Case $x_e = 1$ reducing to case $x_e = 0$ or simply induction.
+\begin{enumerate}
+\item Contradict that number of variables $|E| = |E_x| + |V_x|$ for appropriately chosen $E_x,V_x$ when $0 < x < 1$.
+\end{enumerate}
+\item Proceed by induction on $|E(G)|$: for graphs $G'$ with $|E(G')| < |E(G)|$, $P(G')$ has integral extreme points.
+\item Case $x_{ab} = 0$
+\begin{enumerate}
+\item If $x' = x\mid_{E-ab}$ is an extreme point of $P(G-ab)$ then done. 
+\item Otherwise: (Hard Case, if there's time) using Stable Matchings, and Geometry.
+ \end{enumerate}
+\item Case $x_{ab} = 1$ reducing to case $x_{ab} = 0$ or simply induction if no $0$ variables.
 \end{enumerate}

### 2017-05-12 seminars/combopt-reading-group-s2017/shorter-tours-nicer-ears-outline.tex
nearly done outline

@@
 \section{Ear Decompositions}
-
+\subsection{Definitions}
+\paragraph{}
+Define an ear decomposition, and terms such as closed, open, endpoint, internal, attached. The term ``open ear-decomposition".
+\paragraph{}
+Note the theorem of Whitney that a graph has an open ear decomposition if and only if it is 2-vertex-connected.
+\paragraph{}
+Length of ears, non-trivial ears, pendant ears. Give an example decomposition to demonstrate all these concepts.
+\subsection{M\"omke-Svensson Bound}
+\paragraph{}
+Recall definition of removable pairing. Recall theorem of M\"omke-Svensson. Show how to use this theorem and ear-decompositions to obtain a good tour when there are not many pendant ears.
+\subsection{Even, Short, and Clean Ears}
+\paragraph{}
+Define the terms even, short, and clean ears.
+Give an example, (or show them on a previous drawing).
+Explain algorithmic idea at play.
+Show how it gives workhorse lemma Lemma $2.1$.
+Connect the idea that when $\gamma(P)$ is $0$, applying this to pendant ears can give $T$-tours which pick up a $-1$ for each pendant ear, good when there are many.
+\paragraph{}
+Defer mention of Proposition $2.2$ and Theorem $1$ until needed.
+\section{Nice(r) Ear-Decompositions}
+\subsection{Nice Ear Decompositions}
+\paragraph{}
+Define a nice ear decomposition. Give an example. State Lemma that they can be found in polynomial time for $2$-vertex-connected graphs. Defer proof to end of talk.
+\subsection{Switching to Nicer Ears}
+\paragraph{}
+Lemma $2.1(b)$ indicates clean ears are expensive.
+Define eardrum as it arises for our purposes with respect to ear decompositions.
+Show how the $1$-ears connecting $V_M$ to $V(G)\backslash V_M$ effectively give us a choice in how to choose our clean ears.
+\paragraph{}
+Proposition $2.6$ tells us that we can switch paths on these $1$-ears and still have a nice ear-decomposition.
+We want to choose the paths so that the ears associated with the eardrum have as few connected components as possible.
+We are going to need to connect these so we want to minimize these.
+Notice this would give a spanning subgraph where $V_M$ vertices have even degree.
+\subsection{Earmuffs}
+\paragraph{}
+Define earmuff, it's basically what we said we were looking for before. Define maximum earmuff. Consider using Figure $2$ to communicate intent.
+\paragraph{}
+Theorem $2$ of Rado. What matroids to choose to apply this to earmuff maximization. Theorem $4$ which results.
+\section{Lower Bounds}
+\subsection{$L_\phi(G)$}
+\subsection{$L_\mu(G,M)$}
 \end{document}

### 2017-05-13 seminars/combopt-reading-group-s2017/shorter-tours-nicer-ears-outline.tex
finished outline for talk

@@
 \section{Lower Bounds}
+\subsection{$L_\mu(G,M)$}
+\paragraph{}
+We can use our previous theorem to obtain a combinatorial lower bound on $LP(G,M)$. Define $L_\mu(G,M)$ and prove Theorem $6$. Note this also lower bounds $LP(G)$.
 \subsection{$L_\phi(G)$}
-\subsection{$L_\mu(G,M)$}
+\paragraph{}
+We can also obtain a lower bound on $LP(G)$ based on $\phi(G)$, the minimum number of even ears.
+Define $L_\phi(G)$, and prove Theorem $5$.
+Reference and state Theorem $1$ here, and the theorem of Edmonds and Johnson $[12]$ and Lovasz $[22]$ on $T$-cuts.
+\section{Approximating Graph-TSP}
+\subsection{Theorem $7$}
+\paragraph{}
+Recall both statements of Lemma $2.1$ before beginning the proof.
+We prove Theorem $7$ which gives us a good tour when there are ``many" pendant ears.
+This proof uses Proposition $2.2$.
+Since we didn't prove it just transplant the proof here.
+\subsection{Theorem $10$}
+\paragraph{}
+We are finally ready to give Theorem $10$. Recall all the results needed: the combinatorial lower bounds (and their definitions), the two types of tours we can form (many and few pendant ears).
+\section{(Optional) Lemma $2.5$}
+\paragraph{}
+Verifying that we can find a nice ear decomposition in polynomial time. Proof by picture recommended.
+\section{(Optional) Tight Example}
+Present Figure $4$, an example showing the analysis of the presented algorithm is tight.
 \end{document}

### 2017-09-22 CO681-QuantumInformationProcessing/assignments/assignment_01.tex
starting a1

--- 2017-02-16 CS798-ConvexityAndOptimization/assignments/assignment_01.tex
@@
 \usepackage{tkz-berge}
+
+\usepackage[braket, qm]{qcircuit}
 
@@
 
-\lhead{2017-02-15}
-\rhead{William Justin Toth CS798-Convexity and Optimization Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{\today}
+\rhead{W. Justin Toth CO681-Quantum Information Processing A1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-%Q1%
 \section{}
 \paragraph{}
-Let $q > p > 0$ and let $x_1, \dots, x_n \in \R_{>0}$.
-Let $f : \R_{>0} \rightarrow \R_{>0}$ be the function given by $f(x) = x^\frac{q}{p}$.
-We claim that $f$ is convex.
-To see this we will use the second order condition ($f'' \geq 0$).
-We have by calculus that
-$$f' (x)= \frac{q}{p} x^{\frac{q}{p}-1}$$
-and thus 
-$$f''(x) = \frac{q}{p}(\frac{q}{p} -1) x^{\frac{q}{p} - 2}.$$
-Now since $q > p$, we have $\frac{q}{p} > 1$.
-So then $\frac{q}{p} - 1$ is non-negative, and further $x^{\frac{q}{p} - 2} > 0$ since $x >0$.
-Hence $f''(x) \geq 0$ on the domain of $f$.
-Therefore by the second order condition, $f$ is convex.
-\paragraph{}
-Now observe by convexity that
-$$f(\frac{1}{n}\sum_{i=1}^n x_i^p) \leq \frac{1}{n} \sum_{i=1}^n f(x_i^p) = \frac{1}{n} \sum_{i=1}^n x_i^q.$$
-On the other hand, by applying $f$ directly we see that
-$$f(\frac{1}{n}\sum_{i=1}^n x_i^p) = \frac{1}{n}^\frac{q}{p}(\sum_{i=1}^n x_i^p)^\frac{q}{p}.$$
-Combining the previous two expressions yields the inequality
-$$\frac{1}{n}^\frac{q}{p}(\sum_{i=1}^n x_i^p)^\frac{q}{p} \leq \frac{1}{n} \sum_{i=1}^n x_i^q.$$
-By raising both sides to the exponent $\frac{1}{q}$ (which preserves order) we obtain that
-$$(\frac{1}{n}\sum_{i=1}^n x_i^p)^\frac{1}{p} \leq (\frac{1}{n} \sum_{i=1}^n x_i^q)^\frac{1}{q}$$
-as desired. $\blacksquare$
-%Q2
+Let $X = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}$, $H = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 1 \\ 1 & -1 \end{bmatrix}$, $T = \begin{bmatrix} 1 & 0 \\ 0 & \zeta_8\end{bmatrix}$
+\begin{itemize}
+\item a) $\begin{bmatrix} 0 & 1 \\ \zeta_8 & 0 \end{bmatrix} = TX$
+\item b) $\frac{1}{\sqrt{2}}\begin{bmatrix} 1 & 1 \\ -\zeta_8 & \zeta_8 \end{bmatrix} = THX$
+\item c) $\frac{1}{\sqrt{2}}\begin{bmatrix} 1 &\zeta_8 \\ \zeta_8 & -i\end{bmatrix}=THT$
+\item d) $\frac{1}{2} \begin{bmatrix} 1+\zeta_8 & 1-\zeta_8 \\ 1-\zeta_8 & 1+\zeta_8\end{bmatrix}=HTH.$
+\end{itemize}
 \section{}
-%Pa
 \subsection{a}
 \paragraph{}
-Consider the function
-$$f(x) = (\sum_{i=1}^n x_i^p)^\frac{1}{p}$$
-with $\text{dom} f = \R^n_{>0}$. We will compute the Hessian matrix. First observe that
-$$\frac{\partial f}{\partial x_i} f = x_i^{p-1}(\sum_{k=1}^n x_k^p)^\frac{1-p}{p}$$
-for $i = 1, \dots, n$. Now for $j = 1, \dots, n$
-\begin{align*}
-\frac{\partial f}{\partial x_j}\frac{\partial f}{\partial x_i} f &= \frac{\partial f}{\partial x_j}   x_i^{p-1}(\sum_{i=1}^n x_i^p)^\frac{1-p}{p} \\
-&=(p-1)(\sum_{k=1}^n x_k^p)^\frac{1-2p}{p}\begin{cases}
--x_i^{p-1}x_j^{p-1}, &i\neq j \\
-x_i^{p-2}\sum_{k=1}^n x_k^p  -x_i^{p-1}x_j^{p-1}, &i=j.
-\end{cases}
-\end{align*}
-We claim that the Hessian is positive semidefinite. We prove this directly from the definition, let $y \in \R^n$:
-\begin{align*}
-y^T\nabla^2f(x)y &= \sum_{i=1}^n \sum_{j=1}^n \nabla^2f(x)_{ij} y_i y_j \\
-&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[\sum_{i=1}^nx_i^{p-2}(\sum_{k=1}^nx_k^p)y_i^2 - \sum_{i=1}^n\sum_{j=1}^n x_i^{p-1}x_j^{p-1}y_i y_j ] \\
-&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(\sum_{i=1}^nx_i^{p-2}y_i^2) - (\sum_{i=1}^n x_i^{p-1}y_i)(\sum_{i=1}^n x_i^{p-1} y_i) ] \\
-&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(\sum_{i=1}^nx_i^{p-2}py_i^2) - (\sum_{i=1}^n x_i^\frac{p}{2}x_i^\frac{p-2}{2}y_i)(\sum_{i=1}^n x_i^\frac{p}{2}x_i^\frac{p-2}{2}y_i)]\\
-&= (p-1) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(||x^\frac{p}{2}||^2||x^\frac{p-2}{2}y||^2 - \langle x^\frac{p}{2}, x^\frac{p-2}{2}y\rangle^2]
-\geq 0.
-\end{align*}
-The inequality follows since $p-1 \geq 1$ (as $p \geq 1$), $(\sum_{k=1}^nx^p)^\frac{1-2p}{p} \geq 0$ (as $x \in \R^n_{>0}$), and 
-$$||x^\frac{p}{2}||^2||x^\frac{p-2}{2}y||^2 \geq \langle x^\frac{p}{2}, x^\frac{p-2}{2}y\rangle^2$$
-by Cauchy-Schwarz. Hence $\nabla^2f(x)$ is positive semidefinite and thus by the Second Order Condition, (since $\R^n_{>0}$ is open) $f$ is convex.
-\paragraph{}
-Now when $p < 1$, by a similar calculation we observe
-\begin{align*}
-\frac{\partial -f}{\partial x_j}\frac{\partial f}{\partial x_i} f &=(1-p)(\sum_{k=1}^n x_k^p)^\frac{1-2p}{p}\begin{cases}
--x_i^{p-1}x_j^{p-1}, &i\neq j \\
-x_i^{p-2}\sum_{k=1}^n x_k^p  -x_i^{p-1}x_j^{p-1}, &i=j.
-\end{cases}
-\end{align*}
-So again calculating the quadratic form $y^T\nabla^2-f(x)y$ we have
-\begin{align*}
-y^T\nabla^2-f(x)y &=(1-p) (\sum_{k=1}^nx^p)^\frac{1-2p}{p}[(\sum_{i=1}^nx_i^{p})(||x^\frac{p}{2}||^2||x^\frac{p-2}{2}y||^2 - \langle x^\frac{p}{2}, x^\frac{p-2}{2}y\rangle^2]\\
-&\geq 0.
-\end{align*}
-Since $p < 1$, $1-p >0$, and so the inequality follows as before. Therefore, by the Second Order Condition $-f$ is convex, and hence $f$ is concave. $\blacksquare$
-\paragraph{}
-Next we show that
-$$||x||_p = (\sum_{i=1}^n |x_i|^p)^\frac{1}{p}$$
-is a norm for $p\geq 1$.
-It is easy to see that $||x||_p$ is non-negative since it is a sum of absolute values.
-Similarly if any $x_i > 0$ then $||x||_p > 0$, so $||x||_p = 0$ if and only if $x = 0$.
-Now consider some scalar $\alpha$.
-We have
-$$||\alpha x||_p = (\sum_{i=1}^n |\alpha x_i|^p)^\frac{1}{p} = (\sum_{i=1}^n |\alpha|^p|\cdot |x_i|^p)^\frac{1}{p} = |\alpha|(\sum_{i=1}^n |x_i|^p)^\frac{1}{p} = |\alpha|\cdot ||x||_p.$$
-\paragraph{}
-Lastly it remains to verify the triangle inequality. Let $x, y \in \R^n$. If either $x$ or $y$ is $0$, say without loss of generality $y=0$, then 
-$$||x + y||_p = ||x||_p + 0 = ||x||_p + ||y||_p$$
-immediately. So we may assume $x$ and $y$ are non-zero. Thus it suffices to prove
-$$f(x+y) \leq f(x) + f(y).$$
-To show this we exploit convexity of $f$ as follows
-$$f(x+y) = f(\frac{1}{2}(2x+2y)) \leq \frac{1}{2}(f(2x) + f(2y)) = \frac{1}{2} ((2^p\sum_{i=1}^n x^p)^\frac{1}{p} +(2^p\sum_{i=1}^n y^p)^\frac{1}{p}) = f(x) + f(y).$$
-Therefore $||x + y||_p \leq ||x||_p + ||y||_p$, and hence $||x||_p$ is a norm. $\blacksquare$
-\paragraph{}
-Finally we give a counter-example showing $||x||_p$ is not a norm when $0 < p < 1$.
-Consider $p =\frac{1}{2}$ and $e_1, e_2 \in \R^2$ the standard basis vectors of $\R^2$.
-We prove $||x||_p$ violates the triangle inequality.
-Observe
-$$||e_1 + e_2 ||_p = (1 + 1)^2 = 4$$
-while 
-$$||e_1|| = (1+0)^2 = 1 = ||e_2||$$
-hence 
-$$||e_1 + e_2||  = 4 < 2 = ||e_1|| + ||e_2||.$$
-Thus the triangle inequality is violated, and so $||x||_p$ is not a norm.
-%Pb
+Let $U$ be the unitary matrix for CNOT given by
+$$U = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0& 0 & 1 &0 \end{bmatrix}.$$
+The eigenvectors for $U$ are $\ket{00}$, $\ket{01}$, and $\ket{1+}$ all with eigenvalue $1$ and $\ket{1-}$ with eigenvalue $-1$. We have four orthonormal eigenvectors, so we've found all of them.
 \subsection{b}
 \paragraph{}
-Let $n \in \N$ and consider the function $f: S^n_{++} \rightarrow \R$ given by
-$$f(x) = (\det X)^\frac{1}{n}.$$
-We will demonstrate that $f$ is concave. It suffices to show that the function 
-$$g(t) = (\det (Z + t V))^\frac{1}{n} $$
-is concave for arbitrary $Z \in S^n_{++}$ and $V \in S^n$. Note that $g$ is the restriction of $f$ to some arbitrary line. The domain of $g$ is $\{t \in \R: Z + tV \in S^n_{++}\}$.
-\paragraph{}
-Since $Z \in S^n_{++}$ we can decompose $Z$ as $Z=Z^\frac{1}{2}Z^\frac{1}{2}$ with $Z^\frac{1}{2}$ non-singular.
-Let $\lambda_1, \dots, \lambda_n$ denote the eigenvalues of $Z^\frac{-1}{2}VZ^\frac{-1}{2}$.
-Then we observe that
-$$g(t) = (\det(Z+tV))^\frac{1}{n} = (\det(Z^\frac{1}{2}(I + tZ^\frac{-1}{2}VZ^\frac{-1}{2})Z^\frac{1}{2}))^\frac{1}{n}.$$
-Using that $\det(\cdot)$ distributes over matrix products and that $\det(A)$ is equal to the product of the eigenvalues of $A$ we have
-$$g(t) = \det(Z^\frac{1}{2})^\frac{2}{n}\det(1 + tZ^\frac{-1}{2}VZ^\frac{-1}{2}) = \det(Z^\frac{1}{2})^\frac{2}{n}(\Pi_{i=1}^n 1+t\lambda_i)^\frac{1}{n}.$$
-Now we will show that $g$ is a composition of affine and concave functions, and hence that $g$ is concave. Let $h : \R \rightarrow \R^n$ be the affine function given by
-$$h(t)_i = 1+t\lambda_i$$
-for $i = 1\dots n$. Let $ y : \R^n \rightarrow \R$ be the geometric mean function
-$$y(x) = (\Pi_{i=1}^n x_i)^\frac{1}{n}.$$
-We know from the notes that $y$ is concave. Finally, let $z: \R\rightarrow \R$ be the affine (linear actually) function given by
-$$z(x) = \det(Z^\frac{1}{2})^\frac{2}{n}$$
-Then we see that
-$$g(t) = z\circ y\circ h(t).$$
-Thus $g$ is a composition of affine functions with a concave function and hence $g$ is concave.$\blacksquare$
-%Q3
-\section{}
-\paragraph{}
-Let $f : \R^n \rightarrow \R$ be a convex function and let $g : \R^n \rightarrow \R$ be a concave function with $\text{dom } f = \text{dom } g = \R^n$.
-Suppose that $g(x) < f(x)$ for all $x$.
-Let $e(f) = \{(x,t) \in \R^n \times \R: t \geq f(x) \}$ be the epigraph of $f$.
-We know from our notes that $e(f)$ is convex.
-Let $b(f) = \{(x,t) \in \R^n \times \R: t \leq g(x)\}$ denote the ``below" graph (sorry I don't know the technical name of this object) of $g$.
-We claim that $b(g)$ is convex.
-\paragraph{}
-Let $(x_1, t_1), (x_2, t_2) \in b(g)$. Then we have
-\begin{align*}
--g(\frac{1}{2}(x_1 + x_2)) &\leq \frac{1}{2}(-g(x_1) - g(x_2)) &\text{ since $-g$ is convex} \\
-&=\frac{-1}{2}(g(x_1) + g(x_2)).
-\end{align*}
-Multiplying by $(-1)$, then using that $x_1, t_1), (x_2, t_2) \in b(g)$ we obtain
-$$g(\frac{1}{2}(x_1 + x_2)) \geq \frac{1}{2}(g(x_1) + g(x_2)) \geq \frac{1}{2}(t_1 + t_2).$$
-Thus $\frac{1}{2}[(x_1,t_1) + (x_2,t_2)] \in b(g)$, that is to say $b(g)$ is convex.
-\paragraph{}
-Now we claim that $e(f) \cap b(g) = \emptyset$. Suppose for a contradiction that there exists $(x,t) \in e(f) \cap b(g)$. Thus we have
-$$t \geq f(x) \quad \text{and}\quad t \leq g(x).$$
-But then $f(x) \leq g(x)$, contradicting that $f(x) > g(x)$ for all $x$. Thus $e(f)$ and $b(g)$ are disjoint.
-\paragraph{}
-Using the Separating Hyperplane Theorem as written in section $2.5.1$ of Boyd and Vandenberghe, on $e(f)$ and $b(g)$ we observe that there exists $(y,a) \in \R^n \times \R$ and $b \in \R$, a hyperplane, which separates $e(f)$ and $b(g)$:
-\begin{align}
-y^Tx + at &\geq b &\forall (x,t) \in e(f) \label{eq:3.1}\\
-y^Tx + at &\leq b &\forall (x,t) \in b(g) \label{eq:3.2}.
-\end{align}
-Now we verify that $a\geq 0$. Let $x \in \R^n$. Then applying $\ref{eq:3.1}$ to $(x,f(x))$ and $\ref{eq:3.2}$ to $(x,g(x))$ and applying transitivity through $b$ we have
-$$y^Tx + af(x) \geq y^Tx + ag(x)$$
-Thus
-$$a(f(x) - g(x)) \geq 0.$$
-Since $f(x) > g(x)$ this implies
-$$a \geq 0.$$
-Now observe that
-$$y^Tx + af(x) \geq b$$
-implies
-$$ f(x) \geq \frac{b}{a} - \frac{y^Tx}{a}$$
-and
-$$ y^Tx + ag(x) \leq b$$
-implies 
-$$g(x) \leq \frac{b}{a} - \frac{y^Tx}{a}.$$
-Let $h(x): \R^n \rightarrow \R$ be the affine function given by
-$$h(x) =\frac{-1}{a}(y^Tx - b)$$
-Then we have immediately
-$$g(x) \leq h(x) \leq f(x)$$
-for all $x \in \R^n$ as desired. $\blacksquare$
-
-%Q4
-\section{}
-\paragraph{}
-Let $P = \{ x| Ax \leq b\}$ be a polytope, where $A \in \R^{m\times n}$, $b \in \R^m_{>0}$.
-Let $a_i$ denote the $i$-th row of $A$.
-Let $[m] = \{1,\dots, m\}$.
-Let $C = \text{conv} \{\frac{a_i}{b_i} : i\in [m]\}$ be the convex hull of $\frac{a_i}{b_i}$'s.
-We want to show that $P^* = C$, where $P^*$ denotes the polar set of $P$.
-\paragraph{}
-We will use the theorem from class that $P^{**} = P$ when $0 \in P$.
-With this theorem it is equivalent to show that $P = C^*$.
-First we will rewrite $P$ as $P = \{x| a_i^Tx \leq b_i \forall i\in[m]\}$.
-Observe that
-$$C^* := \{ x: x^Ty \leq 1\forall y \in C\}.$$
-Observe that $\forall y \in C$ there exists $\lambda_i$ for $i\in [m]$ such that $y = \sum_{i\in[m]} \lambda_i \frac{a_i^T}{b_i}$, $\sum_{i\in [m]}\lambda_i = 1$, and $\lambda \geq 0$.
-This is the definition of convex hull.
-Hence substituting into our definition of $C^*$ we see
-$$C^* = \{x : x^T(\sum_{i} \lambda_i \frac{a_i^T}{b_i}) \leq 1,\ \forall \lambda \geq 0, \sum_i \lambda_i = 1\}.$$
-Rewriting slightly this is equivalent to
-$$C^* = \{x : \sum_{i}\lambda_i \frac{a_i}{b_i}x \leq 1,\ \forall \lambda \geq 0, \sum_i\lambda_i = 1\}.$$
-It is not hard to see that such $C^* = P$. We will write it out formally below for the sake of completeness.
-\paragraph{}
-Let $x\in C^*$. Then if we for each $i \in [m]$ define the vector $\lambda^i$ as
-$$\lambda^i_j = \begin{cases}
-1, &i=j \\
-0, &i\neq j.
-\end{cases}$$
-Then $\lambda^i \geq 0$ and $\sum_j \lambda^i_j = \lambda^i_i = 1$. Since $x \in C$ we have
-$$\sum_{j}\lambda^i_j \frac{a_j}{b_j} x \leq 1$$
-and hence
-$$\frac{a_i}{b_i} x \leq 1.$$
-Thus for all $i \in [m]$, $a_i x\leq b_i$ ($b_i > 0$ implies inequality does not change order) and therefore $x \in P$.
-Now if we let $x\in P$, and let $\lambda \geq 0$ such that $\sum_i \lambda_i = 1$ then we need to show that $\sum_i \lambda_i \frac{a_i}{b_i} x \leq 1$.
-Since $x \in P$, $a_i x \leq b_i$ for all $i \in [m]$.
-Dividing by $b_i$ and multiplying by $\lambda_i \geq 0$ we obtain
-$$\lambda_i\frac{a_i}{b_i} x \leq \lambda_i \quad \forall i \in [m].$$
-Summing these inequalities over $i\in[m]$ obtains the desired bound:
-$$\sum_i \lambda_i \frac{a_i}{b_i}x \leq \sum_i \lambda_i = 1.$$
-Thus $P= C^*$ and by our theorem, $P^* = C^{**} = C$. For this theorem to apply it remains to verify that $0 \in C$.
-\paragraph{}
-Suppose for a contradiction that $0 \not\in C$.
-Since $C$ is a convex hull of a finite set it is closed and bounded.
-Hence, by the separating hyperplane theorem there exists a (non-zero) hyperplane $h \in \R^n$ such that
-$$h^T0 > h^Tx \quad\forall x \in C.$$
-We claim that $h$ is a direction in $P$.
-To see this let $\alpha \in \R$ be a scalar.
-Let $x \in P$.
-We will show that $x + \alpha h$ is feasible, implying that $h$ is a direction in $P$.
-Let $ i \in [m]$ and consider constraint $a_i$ with $b_i$.
-Then
-$$\frac{a_i}{b_i} (x + \alpha h) = \frac{a_i}{b_i}x + \alpha \frac{a_i}{b_i}h < \frac{a_i}{b_i}x \leq 1.$$
-The first inequality follows since $\frac{a_i}{b_i} \in C$ (and thus $\frac{a_i}{b_i} h < h^T0 = 0)$ and the second inequality follows since $x \in P$.
-Thus multiplying through by $b_i$, which will not reverse the order of inequalities since $b_i > 0$, we obtain
-$$\frac{a_i}(x + \alpha h) < b_i.$$
-Since this holds for all $i \in [m]$ this implies that $x + \alpha h \in P$. Hence $h$ is a direction, contradicting that $P$ is bounded. Therefore $0\in C$ as desired. $\blacksquare$
-
-%Q5
-\section{}
-Let $Q \in S^n_{++}$. We consider the quadratic norm $||x||_Q := (x^TQx)^\frac{1}{2}.$
-%pa
-\subsection{a}
-\paragraph{}
-We show that $||x||_Q$ is a norm. First since $Q\in S^n_{++}$ there exists non-singular, symmetric $V$ such that $Q = V^TV$. Observe that
-$$||x||_Q = (x^TQx)^\frac{1}{2} = (x^TV^TVx)^\frac{1}{2} =  ((Vx)^T(Vx))^\frac{1}{2} = ||Vx||_2.$$
-We will show that $||x||_Q$ is non-negative and equal to $0$ only when $x = 0$, that $||t x||_Q = |t|\cdot ||x||_Q$, and that triangle inequality holds.
-\paragraph{Non-negativity}
-Since $||\cdot||_2$ is a norm, $||x||_Q = ||Vx||_2 \geq 0$ and further $||Vx||_2 = ||x||_Q = 0$ only if $Vx = 0$, but then $x = V^{-1}0 = 0$.
-\paragraph{Scaling}
-Observe that $$||tx||_Q = ||V(tx)||_2 = |t|\cdot ||Vx||_2 = |t|\cdot ||x||_Q.$$
-\paragraph{Triangle Inequality}
-Let $x,y \in \R^n$. Then
-$$||x+y||_Q = ||V(x+y)||_2 = ||Vx + Vy||_2 \leq ||Vx||_2 + ||Vy||_2 = ||x||_Q + ||y||_Q.$$
-Therefore $||\cdot||_Q$ is a norm. $\blacksquare$
-%pb
-\subsection{b}
-\paragraph{}
-Let $x \in \R^n$. Then observe that
-\begin{align*}
-||x||_{Q^*} = \sup \{ x^Ty : ||y||_Q \leq 1\} \\
-&= \sup \{x^Ty : ||Vy||_2 \leq 1\} \\
-&= \sup \{x^TV^{-1}Vy : ||Vy||_2 \leq 1\} \\
-&= \sup\{x^TV^{-1}z : ||z||_2 \leq 1\} &\text{assigning $z=Vy$}. 
-\end{align*}
-The supremum is realized when $z = (x^TV^{-1})^T = V^{-1}x$. Therefore
-$$||x||_{Q^*} = x^TV^{-1}V^{-1}x = x^TQ^{-1}x = ||x||_{Q^{-1}}.$$
-Thus our nice formula is $||x||_{Q^*} = ||x||_{Q^{-1}}$.$\blacksquare$
-%pc
+The circuit for $(H \otimes I)U$ is
+\[ \Qcircuit @C=1em @R=1em {
+ & \ctrl{1} & \gate{H} &  \qw \\
+ & \targ &  \qw& \qw
+}\]
+and the corresponding unitary is
+$$B:= \frac{1}{\sqrt{2}}\begin{bmatrix} 1 & 0 & 1 & 0 \\ 0 & 1 & 0 & 1 \\ 1 & 0 & -1 & 0 \\ 0 & 1 & 0 & -1 \end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0& 0 & 1 &0 \end{bmatrix} = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 0 & 0 & 1 \\ 0 & 1 & 1 & 0 \\ 1 & 0 & 0 & -1\\ 0 & 1 & -1 & 0\end{bmatrix}.$$
 \subsection{c}
 \paragraph{}
-Let $\cE = \{x : x^TQx \leq 1\}.$  Rewriting we have $\cE = \{x : ||x||_Q \leq 1\} = \cB$ where $\cB$ denote the norm ball of $||\cdot||_Q$. Then we have
-$$\cE^* = \{y : y^Tx \leq 1, \forall x \in \cE\} = \{y : y^Tx \leq 1, \forall x \in \cB\}  = \cB^* = \cB_*,$$
-where $\cB_*$ is the norm ball of $||\cdot||_{Q^*}$, and the last equality follows from the notes.
-Rewriting using part $(b)$ we see that $\cB_* = \{x: x^TQ^{-1}x \leq 1\}$.
-That is $\cE^*$ is ellipsoid given by $Q^{-1}$.
+Let $W$ be the unitary matrix for NOTC given by
+$$W = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \\ 0& 1 & 0 &0 \end{bmatrix}.$$
+The circuit for $(I \otimes H)W$ is
+\[ \Qcircuit @C=1em @R=1em {
+ & \targ & \qw &  \qw \\
+ & \ctrl{-1} & \gate{H} & \qw
+}\]
+and the corresponding unitary is
+$$C:= \frac{1}{\sqrt{2}}\begin{bmatrix} 1 & 1 & 0 & 0 \\ 1& -1 & 0 & 0 \\ 0 & 0 & 1 & 1 \\ 0& 0 & 1 &-1 \end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \\ 0& 1 & 0 &0 \end{bmatrix} = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 0 & 0 & 1 \\ 1 & 0 & 0 & -1 \\ 0 & 1 & 1 & 0\\ 0 & -1 & 1 & 0\end{bmatrix}.$$
+\subsection{d}
+\paragraph{}
+Since $(H\otimes H)U(H\otimes H) = W$ if and only if 
+$$(I \otimes H)(H \otimes H) U(H \otimes H) = (I\otimes H) W$$
+which simplifies as
+$$(H\otimes I) U(H\otimes H) = (I \otimes H) W,$$
+it suffices to show that $B (H\otimes H) = C$. Thus we compute
+\begin{align*}B(H\otimes H) &= \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 0 & 0 & 1 \\ 0 & 1 & 1 & 0 \\ 1 & 0 & 0 & -1\\ 0 & 1 & -1 & 0\end{bmatrix}\cdot \frac{1}{2}\begin{bmatrix} 1 & 1 & 1 & 1 \\ 1 & - 1 & 1 & -1 \\ 1 & 1& -1 & -1 \\ 1 & -1 & -1 &1\end{bmatrix} \\
+&=\frac{1}{\sqrt{2}}\cdot \frac{1}{2}\begin{bmatrix}2 &0 & 0 & 2 \\ 2 & 0 & 0 & -2 \\ 0 & 2 & 2 & 0 \\ 0 & -2 & 2 & 0 \end{bmatrix} \\
+& C. \blacksquare\end{align*}
 \end{document}

### 2017-09-22 CO681-QuantumInformationProcessing/assignments/circuitDemo.tex
starting a1

\documentclass[twocolumn,nofootinbib]{revtex4}
\usepackage[bookmarks = true, pdfpagemode = None, pdfstartview = FitH, colorlinks = true, urlcolor = blue]{hyperref}
\usepackage[braket, qm]{qcircuit}
\usepackage{amsmath}  
\usepackage{listings}
\renewcommand{\arraystretch}{1.5}

\begin{document}
\lstset{language=TeX}

\title{qcircuit 2.5.3 Tutorial}

\author{Original authors: Bryan Eastin, Steve T Flammia\\ Edits: Travis L Scholten}
\affiliation{Department of Physics and Astronomy, University of New
Mexico, Albuquerque, New Mexico 87131-0001, USA}

\begin{abstract}qcircuit is a list of macros that greatly simplifies the construction of 
quantum circuit diagrams (QCDs) in \LaTeX \ with the help of the \Xy-pic 
package.  This tutorial should help the reader acquire 
the skill to render arbitrary QCDs in a matter of minutes. The source code for qcircuit is available for free\footnote{The qcircuit package is distributed under the GNU public license.} on the \href{CQuIC GitHub page}{https://github.com/CQuIC/qcircuit}.
\end{abstract}

\maketitle

\section{Introduction}
\setcounter{footnote}{1}
Ever tried to use \LaTeX\ to typeset something like this?
\[
\Qcircuit @C=.5em @R=0em @!R {
& \ctrl{1} & \qw & & & \qw & \ctrl{1} & \qw & \ctrl{1} & \ctrl{2} & \qw\\
& \ctrl{1} & \qw & \push{\rule{.3em}{0em}=\rule{.3em}{0em}} & & \ctrl{1} & \targ & \ctrl{1} & \targ & \qw & \qw\\
& \gate{U} & \qw & & & \gate{V} & \qw & \gate{V^\dag} & \qw & \gate{V} & \qw
}
\]
\noindent Or maybe this?
\[
\Qcircuit @C=.7em @R=.4em @! {
\lstick{\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \control \cw\\
\lstick{\ket{0}} & \qw & \targ & \targ & \qw & \meter & \cwx\\
\lstick{\ket{0}} & \gate{H} & \ctrl{-1} & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx & \rstick{\ket{\psi}} \qw
}
\]
\noindent Or how about\footnote{Code for these circuits is given in Appendix \ref{S:code}.}
\[
\Qcircuit @C=1.3em @R=.6em {
  & & & & & & \mbox{Syndrome Measurement} & & & &
    \mbox{Recovery}\\
  & \qw & \qw & \ctrl{3} & \qw & \qw & \qw &
    \ctrl{5} & \qw & \qw &
    \multigate{2}{\ \mathcal{R}\ } & \qw\\
  & \qw & \qw & \qw & \ctrl{2} & \ctrl{3} & \qw &
    \qw & \qw & \qw & \ghost{\ \mathcal{R}\ } \qw &
    \qw\\
  & \qw & \qw & \qw & \qw & \qw & \ctrl{2} & \qw &
    \ctrl{3} & \qw & \ghost{\ \mathcal{R}\ } \qw &
    \qw\\
  & & \lstick{\ket{0}} & \targ \qw & \targ \qw &
    \qw & \qw & \qw & \qw & \measure{M_a} &
    \control \cw \cwx\\
  & & \lstick{\ket{0}} & \qw & \qw & \targ \qw &
    \targ \qw & \qw & \qw & \measure{M_b} &
    \control \cw \cwx\\
  & & \lstick{\ket{0}} & \qw & \qw & \qw & \qw &
    \targ \qw & \targ \qw & \measure{M_c}
    \gategroup{2}{2}{7}{10}{.8em}{--} &
    \control \cw \cwx
}
\]

Typesetting quantum circuit diagrams using standard \LaTeX\ graphics packages is a difficult and time consuming business.  qcircuit is a high level macro package designed to change that.  With qcircuit, drawing quantum circuit diagrams is as easy as constructing an array.  In a matter of minutes you can learn the basic syntax and start producing circuits of your own.

This tutorial teaches you to use qcircuit from the ground up.  Many readers will find that they've learned everything they need to know by the end of \S\ref{S:basics}, but plenty of material is included for those that wish to typeset more complicated circuits.

\section{Getting Started}

To install qcircuit, place the file \verb=qcircuit.sty= somewhere your \TeX\ distribution can find it and run the appropriate command to update your \TeX\ tree.  To use it, place the command
{\small \begin{verbatim}\usepackage[options]{qcircuit}\end{verbatim}}
\noindent in the preamble of your document.  \verb=qcircuit.sty= loads the \verb=amsmath= and \verb=xy= packages and implements a set of circuit commands.  If need be, you can obtain the necessary packages at \href{http://www.ctan.org/}{http://www.ctan.org/}.

qcircuit comes with two options - \verb=braket= and \verb=qm= - which provide defined commands for bras, kets, inner and outer products, matrix elements, and expectation values. By default, these options are not enabled, allowing you to define your own commands if you wish.

\section{Special Commands}
As mentioned above, qcircuit comes with predefined commands for some commonly used functions. We have chosen to use the \verb=ensuremath= command, meaning you do not need to put dollar signs around the calls to these commands.

We demonstrate the commands below along with their respective outputs:

\begin{center}
\begin{tabular}{l c l c}
\begin{lstlisting}
\ket{A}
\end{lstlisting} & \ket{A} & \begin{lstlisting}
\bra{B}
\end{lstlisting} & \bra{B}\\
\begin{lstlisting}
\ip{A}{B}
\end{lstlisting} & \ip{A}{B} & \begin{lstlisting}
\op{A}{B}
\end{lstlisting} & \op{A}{B}\\
\begin{lstlisting}
\melem{j}{B}{k}
\end{lstlisting} & \melem{j}{B}{k} &
\begin{lstlisting}
\expval{B}
\end{lstlisting} & \expval{B}\\
\end{tabular}
\end{center}

\section{Simple Quantum Circuits\label{S:basics}}

To begin, suppose the reader would like to typeset the following 
simple circuit:
\[ \Qcircuit @C=1em @R=.7em {
      & \gate{X} & \qw
}\]

This was typeset using
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
      & \gate{X} & \qw
}\end{verbatim}}

The command \verb=\Qcircuit= is simply a disguised \verb=\xymatrix= command 
with a default parameter set.  For readers unfamiliar with the \verb=xymatrix= environment, it suffices 
to know that it behaves more or less like the \verb=array= environment. 
That is, new columns are denoted by \verb=&= and new rows by \verb=\\=, 
as in the following example:
\[ \Qcircuit @C=1.4em @R=1.2em {
     a & i \\
     1 & x
} \]
which was typeset using
{\small \begin{verbatim}\Qcircuit @C=1.4em @R=1.2em {
     a & i \\
     1 & x
} \end{verbatim}}
The parameters \verb?@C=1.4em? and \verb?@R=1.2em? that appear after \verb=\Qcircuit= specify the spacing between the columns and the rows of the circuit, respectively.  They may take any length as an argument.  Additional parameters are discussed in \S\ref{S:spaces}.

\subsection{Wires and gates}

The command \verb=\qw= draws a wire between two columns of 
a QCD. The command derives its name from an abbreviation of `quantum wire'.  
\[ \Qcircuit @C=1em @R=.7em {
   & \gate{H} & \gate{Z} & \gate{H} & \qw \\
   & \qw & \gate{X} & \qw & \qw
} \]

The diagram above was drawn using 
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
   & \gate{H} & \gate{Z} & \gate{H} & \qw \\
   & \qw & \gate{X} & \qw & \qw
}\end{verbatim}}
\noindent Note that \verb=\qw= is used to connect a wire {\it towards the left}.

The \verb=\gate= command draws the argument of the function inside a 
framed box and extends a wire {\it back to the previous column}.  When using the 
\verb=\gate= and \verb=\qw= commands, make sure there is another column 
entry to the left of the current column entry in your QCD, otherwise the wire will not 
connect to anything (and you'll get an error), as in the following example code:

{\small \begin{verbatim} (**Wrong!**)
\Qcircuit @C=1em @R=.7em {
     \gate{U} & \qw \\
     \gate{U^\dag} & \qw
} \end{verbatim}}

The proper way to render this circuit would be to include space for the incoming
wires at the beginning by inserting the \verb=&= character at the start of each new line:
\[ \Qcircuit @C=1em @R=.7em {
     & \gate{U} & \qw \\
     & \gate{U^\dag} & \qw
} \]

{\small \begin{verbatim}\[ \Qcircuit @C=1em @R=.7em {
     & \gate{U} & \qw \\
     & \gate{U^\dag} & \qw
} \]\end{verbatim}}

The only difference between these two codes is that the correct code has an ampersand (\verb=&=) at the start of each new line.\\

To indicate the end of a circuit simply use the \verb=\qwa= command as the last wire.

\[ \Qcircuit @C=1em @R=.7em {
     & \gate{U} & \qwa \\
     & \gate{U^\dag} & \qwa
} \]

{\small \begin{verbatim}\[ \Qcircuit @C=1em @R=.7em {
     & \gate{U} & \qwa \\
     & \gate{U^\dag} & \qwa
} \]\end{verbatim}}

\subsection{CNOT and other controlled single qubit gates \label{S:CNOT}}

With just these few commands, one can already render a circuit with an arbitrary number of wires and single qubit gates.  In this section, we'll learn how to draw CNOT 
gates and controlled single qubit gates with an arbitrary number 
of controls.

A simple circuit with two CNOT gates in it is
\[ \Qcircuit @C=1em @R=.7em {
     & \ctrl{1} & \targ & \qw \\
     & \targ & \ctrl{-1} & \qw
}\]
which was typeset by
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
     & \ctrl{1} & \targ & \qw \\
     & \targ & \ctrl{-1} & \qw
}\end{verbatim}}
In this circuit, the command \verb=\targ= draws the target gate on the 
wire, and the \verb=\ctrl{#1}= puts a bullet down, and connects to the 
target which is \verb=#1= array elements \textit{below} the control.
Hence, to connect the second CNOT gate properly, we used -1.

A more complicated circuit with multiple controls and arbitrary gates
might look like
\[ \Qcircuit @C=1em @R=.7em {
   & \ctrl{2} & \targ & \gate{U} & \qw \\
   & \qw & \ctrl{-1} & \qw & \qw \\
   & \targ & \ctrl{-1} & \ctrl{-2} & \qw \\
   & \qw & \ctrl{-1} & \qw & \qw 
}\]
which was drawn using
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
   & \ctrl{2} & \targ & \gate{U} & \qw \\
   & \qw & \ctrl{-1} & \qw & \qw \\
   & \targ & \ctrl{-1} & \ctrl{-2} & \qw \\
   & \qw & \ctrl{-1} & \qw & \qw 
}\end{verbatim}}

In the first gate, the control bit connects to the 
target on wire 3.  In the second gate, each control connects to the object 
directly above it.  Finally, the third gate is an example of how to do 
controls on arbitrary gates; simply place the desired gate where you would
normally put a target.

\subsection{Vertical wires}

Suppose we want to typeset the following circuit:
\[ \Qcircuit @C=1em @R=1.2em {
     & \gate{U_1} & \qw \\
     & \ctrl{1} \qwx & \qw \\
     & \gate{U_2} & \qw \\
}\]
so that the middle control has to connect to more than one gate.  The way
to accomplish this is with the \verb=\qwx= command.  The command
\verb=\qwx[#1]= takes an optional input, \verb=#1=, and connects from 
the current position to a position \verb=#1= entries {\it below}
the current position.  The default argument is -1.  Thus, one way to
typeset the above diagram is with the following code:
{\small \begin{verbatim}\Qcircuit @C=1em @R=1.2em {
     & \gate{U_1} & \qw \\
     & \ctrl{-1} \qwx[1] & \qw \\
     & \gate{U_2} & \qw \\
}\end{verbatim}}
\noindent or, equivalently,
{\small \begin{verbatim}\Qcircuit @C=1em @R=1.2em {
     & \gate{U_1} & \qw \\
     & \ctrl{1} \qwx & \qw \\
     & \gate{U_2} & \qw \\
}\end{verbatim}}
\noindent which is what the author used.

Note that wire commands must not precede the gate command in an entry.
Also, remember that commands taking an optional 
argument use {\it square} braces rather than curly braces.

\subsection{Labelling input and output states \label{S:labels}}
The last element we need for simple circuits is the ability to add labels.  We'll look at input and output labels here, other kinds of labels are discussed in \S\ref{S:labels2}.

When labelling input and output qubits, one should use the \verb=\lstick= and \verb=\rstick= commands.  These commands ensure that the labels and the wires connecting to them line up correctly.  The \verb=\lstick= command is used for input labels (on the left of the diagram), and the \verb=\rstick= command is used for output labels (on the right of the diagram).  Placement rules are the same as those for gates with the exception that \verb=\lstick= and \verb=\rstick= can be inserted in the leftmost column of the array.  Here is an example circuit:
\[ \Qcircuit @C=1em @R=1em {
\lstick{\ket{1}} & \targ &  \rstick{\ket{0}} \qw \\
\lstick{\ket{1}} & \ctrl{-1} & \rstick{\ket{1}} \qw
}\]
typeset with
{\small \begin{verbatim}\Qcircuit @C=1em @R=1em {
 \lstick{\ket{1}} & \targ & \rstick{\ket{0}} \qw \\
 \lstick{\ket{1}} & \ctrl{-1} & \rstick{\ket{1}} \qw
}\end{verbatim}}

There are a few options for labelling multi-qubit input states, as well.

\[ \Qcircuit @C=1em @R=1.6em {
    \lstick{} & \qw & \qw \\
    \lstick{} & \qw & \qw \inputgroup{1}{2}{.75em}{\ket{0^k}}\\
    \lstick{} & \qw & \qw \\
    \lstick{} & \qw & \qw \inputgroupv{3}{4}{.8em}{.8em}{\ket{\psi}}\\
    \lstick{A} & \qw & \qw \\
    \lstick{B} & \qw & \qw \inputgrouph{5}{6}{.75em}{\ket{\psi}}{2.2em}
  }\]
typeset with
{\small \begin{verbatim}\Qcircuit @C=1em @R=1.6em {
    \lstick{} & \qw & \qw \\
    \lstick{} & \qw & \qw 
      \inputgroup{1}{2}{.75em}{\ket{0^k}}\\
    \lstick{} & \qw & \qw \\
    \lstick{} & \qw & \qw
      \inputgroupv{3}{4}{.8em}{.8em}{\ket{\psi}}\\
    \lstick{A} & \qw & \qw \\
    \lstick{B} & \qw & \qw 
      \inputgrouph{5}{6}{.75em}{\ket{\psi}}{2.2em}
  }\end{verbatim}}

\section{More Complicated Circuits: Multiple Qubit Gates and Beyond}

So far, we have seen how to make arbitrary QCDs involving single qubit gates and controlled gates, including CNOT.  Since this is known to be universal for computation, we could just stop here!  Of course, many circuit diagrams use more complicated structures such as multi-qubit gates, measurements, classical wires, and swaps.  We will learn how to use Q-circuit to make all of these in this section.

\subsection{Multiple qubit gates \label{S:multigate}}

Let's look at an example, and then we'll explain the code. 
\[ \Qcircuit @C=1em @R=.7em {
     & \multigate{2}{U^\dag} & \qw \\
     & \ghost{U^\dag}& \qw \\
     & \ghost{U^\dag} & \qw 
}\]
The 3-qubit gate above was typeset with
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
     & \multigate{2}{U^\dag} & \qw \\
     & \ghost{U^\dag}& \qw \\
     & \ghost{U^\dag} & \qw 
}\end{verbatim}}
First let's go over the \verb=\multigate= command.  
\verb=\multigate{#1}{#2}= is a two argument gate that takes the 
\textit{depth} of the
gate for the first argument and the \textit{label} of the gate for the
second argument.  In the above example, \verb=#1= equals 2 because the 3-qubit gate
extends two rows below the position of \verb=\multigate=.  On the other two lines, 
the \verb=\ghost= command is used to get the spacing and connections right.  \verb=\ghost= behaves like an invisible gate that allows the quantum wires on either side of your multigate to connect correctly.

The generalization to an arbitrarily large gate is now obvious.  Let's look at a 6-qubit gate. The code
{\small \begin{verbatim}\Qcircuit @C=1em @R=0em {
     & \multigate{5}{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw 
}\end{verbatim}}
\noindent yields
\[ \Qcircuit @C=1em @R=0em {
     & \multigate{5}{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw \\
     & \ghost{\mathcal{F}} & \qw 
}\]
Thus, for every entry below the top, a \verb=\ghost= command
with the label for the gate is needed.  Strictly speaking, the name of the gate is not necessary inside the \verb=\ghost= command.  Since \verb=\ghost= is just an invisible place holder, anything with the same width as the label specified in multigate will work as well.  In practice, however, it is usually easiest to use the same argument.

Note that controls to multiple qubit gates work the same as for single
qubit gates, using \verb=\ctrl= and \verb=\qwx=.

Sometimes a multi-qubit gate must be applied to qubits which are \emph{not} on adjacent rows (and as such, the \verb=\multigate= command is not suitable). To account for this, the \verb=\sgate= command can be used to ``hook together'' the application of a multi-qubit gate on non-adjacent qubits:

{\scriptsize   \begin{verbatim}\\Qcircuit @C=1em @R=.7em {
     & \multigate{1}{\mathcal{F}} & \sgate{\mathcal{G}}{2} & \qw \\
     & \ghost{\mathcal{F}} & \qw & \qw\\
     & \qw & \gate{\mathcal{G}} & \qw \\
}\end{verbatim}}
\noindent which yields
\[ \Qcircuit @C=1em @R=.7em {
     & \multigate{1}{\mathcal{F}} & \sgate{\mathcal{G}}{2} & \qw \\
     & \ghost{\mathcal{F}} & \qw & \qw\\
     & \qw & \gate{\mathcal{G}} & \qw \\
}\]

Such notation may be a bit confusing, admittedly. An alternative circuit which does use the \verb=\multigate= command would have a step where qubits 2 and 3 above were swapped, then the two-qubit gate $\mathcal{G}$ was applied, and finally the qubits were swapped back.

\subsection{Measurements and classical bits}

Measurement gates are typeset just like ordinary gates, but they typically have some sort of decoration to indicate that measurement has occurred.  At present, Q-circuit supports the following single qubit measurement gates.
{\small \begin{center}
    \begin{tabular}{l | l | l} 
        \multicolumn{1}{c}{\itshape Example} & \multicolumn{1}{c}{\itshape Command} & \multicolumn{1}{c}{\itshape Example Code }\\ \hline 
        \Qcircuit @C=1em @R=.7em {& \meter}
            & \verb=\meter= & \verb=\meter=\\
        \Qcircuit @C=1em @R=.7em {& \measure{\mbox{Basis}}}
            & \verb=\measure = & \verb=\measure{\mbox{Basis}}=\\
        \Qcircuit @C=1em @R=.7em {& \measuretab{M_{ijk}}} \hspace{.5em}
            & \verb=\measuretab= & \verb=\measuretab{M_{ijk}}=\\
        \Qcircuit @C=1em @R=.7em {& \measureD{\chi}}
            & \verb=\measureD= & \verb=\measureD{\chi}=\\\\
        \Qcircuit @C=1em @R=1.5em{ &\meterB{\ket{\xi_\pm}}}
            & \verb=\meterB= & \verb=\meterB{\ket{\xi_\pm}}=\\
        \Qcircuit @C=1em @R=1.5em{ &\metersymb}
            & \verb=\metersymb= & \verb=\metersymb=
    \end{tabular}
\end{center}}

Often we want to condition some gate on the output of a measurement.  One convenient way illustrate this is with the classical wire commands, \verb=\cw= and \verb=\cwx=.  The classical wire commands work exactly like the quantum wire commands, but they draw double instead of single lines.

Here is an example using measurement gates and classical wires and the corresponding code.
\[\Qcircuit @C=1em @R=.7em {
     & \qw & \measure{\mbox{Codebit}} \cwx[1] \\
     & \qw & \gate{\chi} & \meter & \rstick{\cdots} \cw
}\]
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
     & \qw & \measure{\mbox{Codebit}} \cwx[1] \\
     & \qw & \gate{\chi} & \meter &
        \rstick{\cdots} \cw
}\end{verbatim}}

If you are using a special basis for your measurements the \verb=\meterB= command allows you to indicate the basis.
\[\Qcircuit @C=1em @R=1.5em {
    \lstick{\ket{\psi}} & \meterB{\ket{\xi_\pm}} & \cw
}\]
{\small \begin{verbatim}\Qcircuit @C=1em @R=1.5em {
    \lstick{\ket{\psi}}  & \meterB{\ket{\xi_\pm}} & \cw
}\end{verbatim}}

Q-circuit also includes the commands \verb=\multimeasure= and \verb=\multimeasureD= for typesetting measurements on multiple qubits.  The syntax for these commands exactly parallels that of the \verb=\multigate= command (see \S\ref{S:multigate}).  An example is shown below.
\[\Qcircuit @C=1em @R=.7em {
    & \multimeasureD{1}{\text{Bell}} \\
    & \ghost{\text{Bell}}
}\]
{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
    & \multimeasureD{1}{\text{Bell}} \\
    & \ghost{\text{Bell}}
}\end{verbatim}}

 In addition, if the measurement must be done on qubits which are \emph{not} adjacent, Q-circuit provides for the \verb=\smeterB= command, which, similar to the \verb=\sgate= command, allows one to split the measurement:

 \[ \Qcircuit @C=1em @R=.7em {
     &  \smeterB{\ket{\xi_{\pm}}}{2} & \qw \\
     &  \qw & \qw\\
     & \gate{\ket{\xi_{\pm}}} & \qw \\
}\]
{\small \begin{verbatim} \Qcircuit @C=1em @R=.7em {
     &  \smeterB{\ket{\xi_{\pm}}}{2} & \qw \\
     &  \qw & \qw\\
     & \gate{\ket{\xi_{\pm}}} & \qw \\
}\end{verbatim}}

\subsection{Non-gate inserts, forcing space, and swap \label{S:inserts}}

In addition to the gates defined by Q-circuit, standard \LaTeX\ can function as a gate if enclosed in curly brackets.  By default, inputs are assumed to have zero size, so no space will be made for the resulting object and any wires connecting to it will run straight to the object's middle.  Standard \LaTeX\ entries can serve as labels or wire decorations.

To force an object to take up space, you should use the \verb=\push= command.  \verb=\push= is most useful in conjunction with the \LaTeX\ command \verb=\rule=.  Together they can be used to construct various sorts of invisible props and struts.

Q-circuit implements a gate command called \verb=\qswap= that is equivalent to the text \verb={\times} \qw=.  The effect of \verb=\qswap= is to insert half of a swap gate (that is a $\times$) which can then be connected (using \verb=\qwx=) to another instance of \verb=\qswap= to create a swap gate.

Here is a circuit that shows how to construct swap, decorate wires, and use \verb=\push= to make an invisible prop.
\[\Qcircuit @C=1em @R=.3em {
     &  & \mbox{Defective Circuit}\\
     & \qswap & \qw & \push{\rule{0em}{1em}} \qw \\
     & \qswap \qwx & \push{X} \qw & \qw \\
     & {/} \qw & \gate{H^{\otimes n}} & \qw
}\]
{\small \begin{verbatim}\Qcircuit @C=1em @R=.3em {
     &  & \mbox{Defective Circuit}\\
     & \qswap & \qw & \push{\rule{0em}{1em}} \qw \\
     & \qswap \qwx & \push{X} \qw & \qw \\
     & {/} \qw & \gate{H^{\otimes n}} & \qw
}\end{verbatim}}

To indicate a generalized circuit with $n$ iterations of something, you could use the \verb=\cds= command.
\[\Qcircuit @C=1em @R=.7em {
     & \targ & \targ & \cds{4}{\cdots} & \targ & \qw\\
     & \ctrl{-1} & \qw &\qw & \qw & \qw \\
     & \qw & \ctrl{-2} & \qw & \qw & \qw \\
     & & & & & \\
     & \qw & \qw & \qw & \ctrl{-4} & \qw \\
}\]
{\small \begin{verbatim}\Qcircuit @C=1em @R=.3em {
     & \targ & \targ & \cds{4}{\cdots} & \targ & \qw\\
     & \ctrl{-1} & \qw &\qw & \qw & \qw \\
     & \qw & \ctrl{-2} & \qw & \qw & \qw \\
     & & & & & \\
     & \qw & \qw & \qw & \ctrl{-4} & \qw \\
}\end{verbatim}}

\subsection{How to control anything}

Controlled-Z gates, wires with bends, and gates that control-on-zero can all be made using the extended family of control commands.  The complete family of control commands is \verb=\ctrl=, \verb=\cctrl=, \verb=\ctrlo=, \verb=\cctrlo=, \verb=\control=, and \verb=\controlo=.

\verb=\ctrlo= is identical to the \verb=\ctrl= command (see \S\ref{S:CNOT}) except that it draws an open bullet (indicating control-on-zero).  Both commands place a wire to the left and take one argument indicating which wire to connect to.

The commands \verb=\cctrl= and \verb=\cctrlo= are identical to the \verb=\ctrl= and \verb=\ctrlo= commands, respectively, except they use \emph{classical wires} instead of quantum ones to do the controlling. These commands may be useful for writing circuits where the gates are conditioned on classical outputs.

The commands \verb=\control= and \verb=\controlo= are isolated controls; they don't automatically connect to anything.  Isolated controls allow you to decide exactly what connections are made to your control operator, which makes them very useful for working with classical wires and rendering things like the controlled-Z.

Here is an example circuit using various controls.  
\[ \Qcircuit @C=1em @R=.7em {
    & \ctrl{2} & \ctrlo{1} & \ctrl{1} & \qw & \multigate{1}{U} & \qw \\
    & \qw & \targ & \ctrlo{2} \qw & \ctrl{1} & \ghost{U} & \qw\\
    & \control \qw & \ctrl{1} & \qw & \meter & \cctrlo{-1} &\\
    & \qw & \control \qw & \gate{H} & \meter & \cctrl{-1}
}\]

{\small \begin{verbatim}\Qcircuit @C=1em @R=.7em {
    & \ctrl{2} & \ctrlo{1} & \ctrl{1} 
        & \qw & \multigate{1}{U} & \qw \\
    & \qw & \targ & \ctrlo{2} \qw 
        & \ctrl{1} & \ghost{U} & \qw \\
    & \control \qw & \ctrl{1} & \qw 
        & \meter & \cctrlo{-1} \\
    & \qw & \control \qw & \gate{H} 
        & \meter & \cctrl{-1}
}\end{verbatim}}

Note that we, the authors, have used a pair of controls connected by a wire to denote the controlled-Z gate.  This isn't standard notation, but we feel it is a logically consistent and concise notation, and it illustrates nicely the symmetry of the controlled-Z gate.  We hope to encourage the readers to adopt this notation in their own QCDs.

\section{Bells and Whistles: Tweaking Your Diagram to Perfection}

By now, the reader should be able to quickly and easily typeset almost any QCD.  Nonetheless, it may occasionally be desirable to decorate or modify a circuit in ways not yet discussed.  This section presents additional tricks, options, and commands for putting the final polish on your QCDs.

\subsection{Spacing\label{S:spaces}}

The Q-circuit parameters \verb+@R+ and \verb+@C+ were introduced in \S\ref{S:basics}; they are examples of a family of spacing parameters that can appear between the text \verb=\Qcircuit= and the opening curly brace.  A more complete list of available parameters is given in the table below.

{\small \begin{center}
    \begin{tabular}{l | l } 
        \multicolumn{1}{c}{\itshape Parameter} & \multicolumn{1}{c}{\itshape Effect }\\ \hline 
        \verb+@R=#1+ & Sets the spacing between rows to \verb=#1=.\\
        \verb+@C=#1+ & Sets the spacing between columns to \verb=#1=.\\
        \verb+@!R+ & \parbox[t]{6cm}{Sets all rows to the height of the tallest object in the circuit.}\\
        \verb+@!C+ & \parbox[t]{6cm}{Sets all columns to the width of the widest object in the circuit.}\\
        \verb+@!+ & \parbox[t]{6cm}{Sets all entries to the size of the largest object in the circuit.}
    \end{tabular}
\end{center}}

The \verb=@R= and \verb=@C= parameters adjust the separation between elements, allowing you to dictate the compactness of your QCD.  \verb=@!R=, \verb=@!C=, and \verb=@!= force the elements of your circuit to have uniform sizes, this helps to prevent bunching that may occur when a particular row or column contains many small elements. \verb=@!R= is particularly useful for forcing wires to be evenly spaced, as in the following example.
\[ \Qcircuit @C=.7em @R=.3em @!R {
    & \qswap & \qw & \qswap & \qw\\
    & \qswap \qwx & \ctrl{1} & \qswap \qwx & \qw \\
    & \qw & \gate{T^\dag} & \qw & \qw
}\]

{\small \begin{verbatim}\Qcircuit @C=.7em @R=.3em @!R {
    & \qswap & \qw & \qswap & \qw\\
    & \qswap \qwx & \ctrl{1} & \qswap \qwx & \qw \\
    & \qw & \gate{T^\dag} & \qw & \qw
}\end{verbatim}}

\subsection{Labelling \label{S:labels2}}

A label can be placed anywhere that a gate command might normally appear.  Unlike gates, however, Q-circuit treats labels as having zero size when determining the layout of a QCD.  This prevents large labels from bending your circuit out of whack, but it also means that labels can overlap with other components.

Normally an element whose size is set to zero is drawn centered on it's entry.  This is what happens when you insert text directly using curly brackets (see \S\ref{S:inserts}).  For most labelling, however, it is more useful to have one edge of the label fixed in the center of an entry.  For this reason Q-circuit provides a set of label commands, \verb=\lstick=, \verb=\rstick=, \verb=\ustick=, and \verb=\dstick=.  The stick commands each cause their contents to ``stick out" from the center of an entry in a different direction.  \verb=\lstick=, \verb=\rstick=, \verb=\ustick=, and \verb=\dstick= produce labels that project out to the left, right, top, and bottom respectively.

Proper usage of \verb=\lstick= and \verb=\rstick= was demonstrated in \S\ref{S:labels}, so the following example focuses on \verb=\ustick= and \verb=\dstick=.
\[ \Qcircuit @C=.7em @R=.3em {
    & \ustick{a} \qw & \qw & \qw & \qw & \meter \\
    & \ustick{b} \qw & \qw & \qw & \meter \\
    & & & & \dstick{B} \cwx & \dstick{A} \cwx[-2]
}\]
\\
{\small \begin{verbatim}\Qcircuit @C=.7em @R=.3em {
    & \ustick{a} \qw & \qw & \qw & \qw & \meter \\
    & \ustick{b} \qw & \qw & \qw & \meter \\
    & & & & \dstick{B} \cwx & \dstick{A} \cwx[-2]
}\end{verbatim}}

\subsection{Grouping}

It is sometimes useful to box off sections of a circuit to indicate a subcircuit, as in the following example.
\[ \Qcircuit @C=1em @R=1em {
    & \ctrl{2} & \qw & \gate{H} & \ctrl{1} & \gate{H} & \qw \\
    & \qw & \ctrl{1} & \gate{H} & \targ & \gate{H} & \qw \\
    & \targ & \targ & \gate{Z} & \qw & \ctrl{-1} & \qw \gategroup{1}{4}{2}{6}{.7em}{--}
}\]
which was typeset using
{\small \begin{verbatim}\Qcircuit @C=1em @R=1em {
    & \ctrl{2} & \qw & \gate{H} & \ctrl{1} & 
        \gate{H} & \qw \\
    & \qw & \ctrl{1} & \gate{H} & \targ &
        \gate{H} & \qw \\
    & \targ & \targ & \gate{Z} & \qw & \ctrl{-1} &
        \qw \gategroup{1}{4}{2}{6}{.7em}{--} 
}\end{verbatim}}
The command that made the dashed box is in the last line of code and is called \verb=\gategroup=.  The \verb=\gategroup= command can be placed following any non-empty entry, but, for clarity, it is perhaps best to put it at the end.

Because it takes six arguments, \verb=\gategroup= looks intimidating, but it is actually relatively easy to use.  \verb=\gategroup{#1}{#2}{#3}{#4}{#5}{#6}= highlights the entries between rows \verb=#1= and \verb=#3= and columns \verb=#2= and \verb=#4= by adding a box or a bracket.  Argument \verb=#6= selects between various highlights, with the available options being:
\begin{center} {\small \verb=--  .  _\}  ^\}  \{  \}  _)  ^)  (  )=} \end{center}
These options produce a dashed box, a dotted box, a curly brace on the bottom, top, left, or right, and a normal brace on the bottom, top, left, or right.  Argument \verb=#5= is twice the spacing from the nearest gate to the box.

\verb=\gategroup= only checks that the gates at the four corners of the requested region are properly enclosed.  As a result, gates along the boundary that are bigger than the corner gates will tend to stick out.  This is especially unsightly when the corner entries are wires, though in that case the problem can be fixed by inserting an invisible prop of the desired height (see \S\ref{S:inserts}).

\section{Acknowledgments}

The authors would like to thank Aaron Smith, Joe Renes, and Andrew Silberfarb for useful discussions, ideas, and debugging.  Thanks to Carl Caves and Michael Nielsen for encouragement on this project.  An extra thanks to Michael Nielsen for suggesting some useful \LaTeX \ resources.

The development of Q-circuit was supported in part by the National Security Agency (NSA) and the Advanced Research and Developement Activity (ARDA) under the Army Research Office (ARO) contract numbers DAAD19-01-1-0648 and W911NF-04-1-0242.

\appendix
\section{Positioning Q-circuit diagrams in \LaTeX}
Q-circuit produces \TeX\ graphics objects.  In theory these objects should act like any symbol or character.  Thus, they can be placed in equation environments, arrays, and figures.  In practice there are a few, largely unexplained, complications.

One of these is vertical centering in a line of text.  To center the top line of a circuit, it is sufficient to invoke it in inline math mode using \verb=$=.  To center the entire circuit, place it inside an array.

Horizontal centering within figures is also problematic.  Typically this can be corrected by placing the \verb=\Qcircuit= command inside a \verb=\centerline= command, an \verb=\mbox= command, or an equation environment.  For some \LaTeX\ distributions the commands \verb=\leavevmode= and \verb=\centering= must be added to center a figure.

Finally, circuits using large labels often appear a bit off center.  This is because labels are not included when calculating the size of a circuit.  The best solution is probably to add white space (see \S\ref{S:inserts}) until the labels all fit within the boundaries of the circuit.

\section{Bugs and Future Work}

\begin{enumerate}
\item Wires often end just short of curved surfaces.
\item \verb=\gategroup= needs to check all the boundary gates when determining the highlighted area.
\item Targets look poor when the font size is set to small.
\item It would be nice if the \verb=\ghost= command could read the argument of the \verb=\multigate= command automatically.
\item Larger issues of centering within \LaTeX\ need to be addressed.
\end{enumerate}

\section{Code for the Introduction\label{S:code}}
The first QCD depicts a way of decomposing doubly controlled unitaries.
It was typeset with
{\small \begin{verbatim}\Qcircuit @C=.5em @R=0em @!R {
  & \ctrl{1} & \qw & & & \qw & \ctrl{1} & \qw &
    \ctrl{1} & \ctrl{2} & \qw\\
  & \ctrl{1} & \qw & 
    \push{\rule{.3em}{0em}=\rule{.3em}{0em}} & &
    \ctrl{1} & \targ & \ctrl{1} & \targ & \qw &
    \qw\\
  & \gate{U} & \qw & & & \gate{V} & \qw &
    \gate{V^\dag} & \qw & \gate{V} & \qw
}
\end{verbatim}}

The second QCD depicts quantum teleportation and was typeset with
{\small \begin{verbatim}\Qcircuit @C=.7em @R=.4em @! {
  \lstick{\ket{\psi}} & \qw & \qw & \ctrl{1} &
    \gate{H} & \meter & \control \cw\\
  \lstick{\ket{0}} & \qw & \targ & \targ & \qw &
    \meter & \cwx\\
  \lstick{\ket{0}} & \gate{H} & \ctrl{-1} & \qw &
    \qw & \gate{X} \cwx & \gate{Z} \cwx &
    \rstick{\ket{\psi}} \qw
}
\end{verbatim}}

The third QCD depicts quantum error correction on the bit flip code.  It was typeset with
{\small \begin{verbatim}\Qcircuit @C=1.3em @R=.6em {
  & & & & & & \mbox{Syndrome Measurement} & & & &
    \mbox{Recovery}\\
  & \qw & \qw & \ctrl{3} & \qw & \qw & \qw &
    \ctrl{5} & \qw & \qw &
    \multigate{2}{\ \mathcal{R}\ } & \qw\\
  & \qw & \qw & \qw & \ctrl{2} & \ctrl{3} & \qw &
    \qw & \qw & \qw & \ghost{\ \mathcal{R}\ } \qw &
    \qw\\
  & \qw & \qw & \qw & \qw & \qw & \ctrl{2} & \qw &
    \ctrl{3} & \qw & \ghost{\ \mathcal{R}\ } \qw &
    \qw\\
  & & \lstick{\ket{0}} & \targ \qw & \targ \qw &
    \qw & \qw & \qw & \qw & \measure{M_a} &
    \control \cw \cwx\\
  & & \lstick{\ket{0}} & \qw & \qw & \targ \qw &
    \targ \qw & \qw & \qw & \measure{M_b} &
    \control \cw \cwx\\
  & & \lstick{\ket{0}} & \qw & \qw & \qw & \qw &
    \targ \qw & \targ \qw & \measure{M_c}
    \gategroup{2}{2}{7}{10}{.8em}{--} &
    \control \cw \cwx
}
\end{verbatim}}

\pagebreak

\section{Table of Commands}

The following table is grouped according to the effect of each command.\\

{\small \begin{center}
    \begin{tabular}{l | l } 
        \multicolumn{1}{c}{\itshape Subject} & \multicolumn{1}{c}{\itshape Command }\\ \hline 
        Loading Q-circuit \hspace{.5em} & \verb=\input{Qcircuit}= \\
        Making Circuits & \verb=\Qcircuit= \\
        Spacing & \parbox[t]{6cm}{\tt
                    @C=\#1 \\
                    @R=\#1 \\
                    @!R \\
                    @!C \\
                    @! \\
                    \char92 push\{\#1\} \\
                    \char92 cds\{\#1\}\{\#2\}}\\
        Wires & \parbox[t]{6cm}{\tt
                    \char92 qw[\#1] \\
                    \char92 qwx[\#1] \\
                    \char92 qwa[\#1] \\
                    \char92 cw[\#1] \\
                    \char92 cwa[\#1] \\
                    \char92 cwx[\#1] }\\
        Gates & \parbox[t]{6cm}{\tt
                    \char92 gate\{\#1\} \\
                    \char92 targ \\
                    \char92 qswap \\
                    \char92 multigate\{\#1\}\{\#2\} \\
                    \char92 sgate\{\#1\}\{\#2\}\\
                    \char92 ghost\{\#1\} }\\
        Controls & \parbox[t]{6cm}{\tt
                    \char92 ctrl\{\#1\} \\
                    \char92 ctrlo\{\#1\} \\
                    \char92 cctrl\{\#1\} \\
                    \char92 cctrlo\{\#1\} \\
                    \char92 control \\
                    \char92 controlo } \\
        Measurements & \parbox[t]{6cm}{\tt
                    \char92 meter \\
                    \char92 meterB\{\#1\} \\
                    \char92 smeterB\{\#1\}\{\#2\} \\
                    \char92 measure\{\#1\} \\
                    \char92 measureD\{\#1\} \\
                    \char92 measuretab\{\#1\} \\
                    \char92 multimeasure\{\#1\}\{\#2\} \\
                    \char92 multimeasureD\{\#1\}\{\#2\} } \\
        Labels & \parbox[t]{6cm}{\tt
                    \char92 lstick\{\#1\} \\
                    \char92 rstick\{\#1\} \\
                    \char92 ustick\{\#1\} \\
                    \char92 dstick\{\#1\} \\
                    \char92 bra\{\#1\} \\
                    \char92 ket\{\#1\} \\
                    \char92 gategroup\{\#1\}\{\#2\}\{\#3\}\{\#4\}\{\#5\}\{\#6\}\\
                    \char92 inputgroup\{\#1\}\{\#2\}\{\#3\}\{\#4\}\\
                    \char92 inputgroupv\{\#1\}\{\#2\}\{\#3\}\{\#4\}\{\#5\}\\
                    \char92 inputgrouph\{\#1\}\{\#2\}\{\#3\}\{\#4\}\{\#5\}\\
                    }
    \end{tabular}
\end{center}}

\end{document}

### 2017-09-22 CO681-QuantumInformationProcessing/assignments/assignment_01.tex
problems 1 -3

@@
 & C. \blacksquare\end{align*}
+\section{}
+\paragraph{}
+Let $\ket{\psi} = \frac{1}{\sqrt{2}}\ket{00} + \frac{1}{2}\ket{01} + \frac{1}{2}\ket{11}.$
+\subsection{a}
+\paragraph{}
+When we measure the first qubit with respect to the computational basis ($\ket{0}, \ket{1}$) we obtain $\ket{0}$ with probability
+$$|\frac{1}{\sqrt{2}}|^2 + |\frac{1}{2}|^2 = \frac{3}{4}$$
+and the second qubit would be left in the normalized post-measurement state
+$$\sqrt{\frac{2}{3}} \ket{0} + \frac{1}{\sqrt{3}} \ket{1},$$
+and we obtain $\ket{1}$ with probability
+$$|\frac{1}{2}|^2 = \frac{1}{4}$$
+and the second qubit would be left in the normalized post-measurement state
+$$\ket{1}.$$
+\subsection{b}
+\paragraph{}
+Rewriting $\ket{\psi}$ in the $\ket{+}, \ket{-}$ basis gives
+$$\ket{\psi} = \frac{\sqrt{2} + 2}{4}\ket{++} + \frac{\sqrt{2}-2}{4}\ket{+-} + \frac{\sqrt{2}}{4}\ket{-+} + \frac{\sqrt{2}}{4}\ket{--}.$$
+So measuring the first qubit with respect to the $\ket{+}, \ket{-}$ basis gives $\ket{+}$ with probability
+$$|\frac{\sqrt{2}+2}{4}|^2+ |\frac{\sqrt{2}-2}{4}|^2 = \frac{3}{4}.$$
+and the second qubit would be left in the normalized post-measurement state
+$$ \frac{\sqrt{2} + 2}{2\sqrt{3}}\ket{+} + \frac{\sqrt{2}-2}{2\sqrt{3}}\ket{-},$$
+and gives $\ket{-}$ with probability
+$$2|\frac{\sqrt{2}}{4}|^2 = \frac{1}{4}$$
+and the second qubit would be left in the normalized post-measurement state
+$$\frac{1}{\sqrt{2}}\ket{+} + \frac{1}{\sqrt{2}}\ket{-} .$$
 \end{document}

### 2017-09-25 CO681-QuantumInformationProcessing/assignments/assignment_01.tex
solved 4

@@
 \newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}
+
+\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
+\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
 
@@
 $$\frac{1}{\sqrt{2}}\ket{+} + \frac{1}{\sqrt{2}}\ket{-} .$$
+\section{}
+\paragraph{}
+Suppose that a qubit is prepared in the state $\ket{0}$ with probability $p$ and prepared in state $\ket{+}$ with probability $q$, where $p+q=1$ and $p,q\geq 0$.
+Then a measurement in the basis $\ket{\phi_0}, \ket{\phi_1}$ identifies the state successfully with probability
+$$p\abs{\ip{\phi_0}{0}}^2 + q\abs{\ip{\phi_1}{+}}^2.$$
+Since $\ket{0}$ and $\ket{+}$ lie in $\R^2$, the optimal $\ket{\phi_0}$ and $\ket{\phi_1}$ lie in $\R^2$ so as to optimize the length of projection of $\ket{\phi_0}$ and $\ket{\phi_1}$ onto $\ket{0}$ and $\ket{+}$ respectively (ie $\ip{\phi_0}{0}$ and $\ip{\phi_1}{+}$).
+\paragraph{}
+We can describe an arbitrary orthonormal basis of $\R^2$ as
+$$\ket{\phi_0} = \begin{bmatrix} \cos\theta \\ \sin\theta \end{bmatrix}, \quad \ket{\phi_1}=\begin{bmatrix}-\sin\theta \\ \cos\theta \end{bmatrix}.$$
+Now our success probability simplifies as
+$$f(\theta) :=p\cos^2\theta + \frac{q}{2}(\cos\theta - \sin\theta)^2$$
+with first derivative
+$$f'(\theta) = -2p\sin\theta\cos\theta - q(\cos^2\theta - \sin^2\theta).$$
+\subsection{a}
+\paragraph{}
+If we set $p=q=\frac{1}{2}$ then we have
+$$f(\theta) = \frac{1}{2}\cos^2\theta + \frac{1}{4}(\cos\theta - \sin\theta)^2$$
+and
+$$f'(\theta) = -\sin\theta\cos\theta -\frac{1}{2}(\cos^2\theta-\sin^2\theta).$$
+So $f$ is periodic with period $\pi$. Setting $f'(\theta)=0$ and solving we obtain one possible optimal solution: $\theta^* = \frac{-\pi}{8}.$
+Since $f'(\frac{-2\pi}{8}) >0$ and $f'(0) < 0$, we know $\theta^*$ is a maximizer of $f$. Hence we have optimal success probability
+$$f(\frac{-\pi}{8}) = \frac{1}{4}(2 + \sqrt(2)) \approx 0.85$$
+and optimal basis
+$$\ket{\phi_0} = \begin{bmatrix} \cos(\frac{-\pi}{8}) \\ \sin(\frac{-\pi}{8}) \end{bmatrix}, \quad \ket{\phi_1}=\begin{bmatrix}-\sin(\frac{-\pi}{8}) \\ \cos(\frac{-\pi}{8}) \end{bmatrix}.$$
+\subsection{b}
+\paragraph{}
+If we set $p=\frac{3}{4}$ and $q=\frac{1}{4}$ then we have
+$$f(\theta) = \frac{3}{4}\cos^2\theta + \frac{1}{8}(\cos\theta - \sin\theta)^2$$
+and
+$$f'(\theta) = -\frac{3}{2}\sin\theta\cos\theta -\frac{1}{4}(\cos^2\theta-\sin^2\theta).$$
+So $f$ is periodic with period $\pi$. Setting $f'(\theta)=0$ and solving we obtain one possible optimal solution: $\theta^* = \arctan(3-\sqrt{10}).$
+Since $f'(\arctan(3-\sqrt{10})-\frac{\pi}{4}) >0$ and $f'(\arctan(3-\sqrt{10}) + \frac{\pi}{4}) < 0$, we know $\theta^*$ is a maximizer of $f$. Hence we have optimal success probability
+$$f(\arctan(3-\sqrt{10})) \approx 0.895$$
+and optimal basis
+$$\ket{\phi_0} = \begin{bmatrix} \cos(\arctan(3-\sqrt{10})) \\ \sin(arctan(3-\sqrt{10})) \end{bmatrix}, \quad \ket{\phi_1}=\begin{bmatrix}-\sin(\arctan(3-\sqrt{10})) \\ \cos(\arctan(3-\sqrt{10})) \end{bmatrix}.$$
+\section{}
+
 \end{document}

### 2017-09-27 CO681-QuantumInformationProcessing/assignments/assignment_01.tex
Assignment completed
edits

@@
 $$p\abs{\ip{\phi_0}{0}}^2 + q\abs{\ip{\phi_1}{+}}^2.$$
-Since $\ket{0}$ and $\ket{+}$ lie in $\R^2$, the optimal $\ket{\phi_0}$ and $\ket{\phi_1}$ lie in $\R^2$ so as to optimize the length of projection of $\ket{\phi_0}$ and $\ket{\phi_1}$ onto $\ket{0}$ and $\ket{+}$ respectively (ie $\ip{\phi_0}{0}$ and $\ip{\phi_1}{+}$).
+Since $\ket{0}$ and $\ket{+}$ lie in $\R^2$, there exist optimal $\ket{\phi_0}$ and $\ket{\phi_1}$ in $\R^2$.
+This can be seen by observing that $\abs{\ip{0}{\phi_0}}$ and $\abs{\ip{+}{\phi_1}}$ are invariant multiplications of $\ket{\phi_a}$ by some $e^{i\theta_a}$.
+With a bit of work, one can convince themselves we can always rotate an optimal solution in the real plane.
 \paragraph{}
@@
 \section{}
-
+\subsection{a}
+\paragraph{}
+\begin{align*}
+\ket{\Psi}^{12}\ket{\Psi}^{34} &= \frac{1}{2}(\ket{00}^{12} + \ket{11}^{12})(\ket{00}^{34} + \ket{11}^{34}) \\
+&= \frac{1}{2}(\ket{00}^{12}\ket{00}^{34} + \ket{00}^{12}\ket{11}^{34} + \ket{11}^{12}\ket{00}^{34} + \ket{11}^12\ket{11}^{34})\\
+&= \frac{1}{2}(\ket{00}^{23}\ket{00}^{14} + \ket{01}^{23}\ket{01}^{14} + \ket{10}^{23}\ket{10}^{34} + \ket{11}^{23}\ket{11}^{34}).
+\end{align*}
+Choose $\ket{\phi_{ab}} = \ket{\Psi_{ab}}$ for $a,b \in \{0,1\}$. Observe that 
+\begin{align*}
+&\ \ \ \ket{\Psi_{00}}^{23}\ket{\Psi_{00}}^{14} + \ket{\Psi_{11}}^{23}\ket{\Psi_{11}}^{14} \\
+&= \frac{1}{2}(\ket{00}^{23}\ket{00}^{14} + \ket{00}^{23}\ket{11}^{14} + \ket{11}^{23}\ket{00}^{14} + \ket{11}^{23}\ket{11}^{14}) \\
+&+ \frac{1}{2}(\ket{00}^{23}\ket{00}^{14} - \ket{00}^{23}\ket{11}^{14} - \ket{11}^{23}\ket{00}^{14} + \ket{11}^{23}\ket{11}^{14}) \\
+&= \ket{00}^{23}\ket{00}^{14} + \ket{11}^{23}\ket{11}^{14}.
+\end{align*}
+and similarly by symmetry
+$$\ket{\Psi_{01}}^{23}\ket{\Psi_{01}}^{14} + \ket{\Psi_{10}}^{23}\ket{\Psi_{10}}^{14} = \ket{01}^{23}\ket{01}^{14} + \ket{10}^{23}\ket{10}^{14} $$
+So using the above observations we see that
+\begin{align*}
+&\ \ \ \frac{1}{2}(\ket{\Psi_{00}}^{23}\ket{\phi_{00}}^{14} + \ket{\Psi_{01}}^{23}\ket{\phi_{01}}^{14} + \ket{\Psi_{10}}^{23}\ket{\phi_{10}}^{14} \ket{\Psi_{11}}^{23}\ket{\phi_{11}}^{14}) \\
+&= \frac{1}{2}(\ket{\Psi_{00}}^{23}\ket{\Psi_{00}}^{14} + \ket{\Psi_{01}}^{23}\ket{\Psi_{01}}^{14} + \ket{\Psi_{10}}^{23}\ket{\Psi_{10}}^{14} \ket{\Psi_{11}}^{23}\ket{\Psi_{11}}^{14}) \\
+&= \frac{1}{2}(\ket{00}^{23}\ket{00}^{14} + \ket{01}^{23}\ket{01}^{14} + \ket{10}^{23}\ket{10}^{34} + \ket{11}^{23}\ket{11}^{34})
+\end{align*}
+as desired.
+\subsection{b}
+\paragraph{}
+Performing a Bell measurement on qubits $2$ and $3$ yields state $\ket{\Psi_{ab}}$ with probability $\frac{1}{4}$ and qubits $1$ and $4$ are left in state $\ket{\Psi_{ab}}$ (for $a,b \in\{0,1\}$).
 \end{document}

### 2017-10-10 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
Q1 of A2

--- 2017-09-22 CO681-QuantumInformationProcessing/assignments/assignment_01.tex
@@
 \usepackage{tkz-berge}
-
 \usepackage[braket, qm]{qcircuit}
@@
 
+\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
+\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
 %END MACROS
@@
 \lhead{\today}
-\rhead{W. Justin Toth CO681-Quantum Information Processing A1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CO681-Quantum Information Processing A2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \section{}
-\paragraph{}
-Let $X = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}$, $H = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 1 \\ 1 & -1 \end{bmatrix}$, $T = \begin{bmatrix} 1 & 0 \\ 0 & \zeta_8\end{bmatrix}$
-\begin{itemize}
-\item a) $\begin{bmatrix} 0 & 1 \\ \zeta_8 & 0 \end{bmatrix} = TX$
-\item b) $\frac{1}{\sqrt{2}}\begin{bmatrix} 1 & 1 \\ -\zeta_8 & \zeta_8 \end{bmatrix} = THX$
-\item c) $\frac{1}{\sqrt{2}}\begin{bmatrix} 1 &\zeta_8 \\ \zeta_8 & -i\end{bmatrix}=THT$
-\item d) $\frac{1}{2} \begin{bmatrix} 1+\zeta_8 & 1-\zeta_8 \\ 1-\zeta_8 & 1+\zeta_8\end{bmatrix}=HTH.$
-\end{itemize}
-\section{}
 \subsection{a}
 \paragraph{}
-Let $U$ be the unitary matrix for CNOT given by
-$$U = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0& 0 & 1 &0 \end{bmatrix}.$$
-The eigenvectors for $U$ are $\ket{00}$, $\ket{01}$, and $\ket{1+}$ all with eigenvalue $1$ and $\ket{1-}$ with eigenvalue $-1$. We have four orthonormal eigenvectors, so we've found all of them.
+$$11+12 \mod 13 = 10$$
 \subsection{b}
 \paragraph{}
-The circuit for $(H \otimes I)U$ is
-\[ \Qcircuit @C=1em @R=1em {
- & \ctrl{1} & \gate{H} &  \qw \\
- & \targ &  \qw& \qw
-}\]
-and the corresponding unitary is
-$$B:= \frac{1}{\sqrt{2}}\begin{bmatrix} 1 & 0 & 1 & 0 \\ 0 & 1 & 0 & 1 \\ 1 & 0 & -1 & 0 \\ 0 & 1 & 0 & -1 \end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0& 0 & 1 &0 \end{bmatrix} = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 0 & 0 & 1 \\ 0 & 1 & 1 & 0 \\ 1 & 0 & 0 & -1\\ 0 & 1 & -1 & 0\end{bmatrix}.$$
+$$\Z^\times_{13} = \{2^0=1, 2^1 = 2, 2^2 = 4, 2^3 = 8, 2^4 = 3, 2^5 = 6, 2^6=12, 2^7=11, 2^8 = 9, 2^9=5, 2^{10}=10, 2^{11} = 7\}$$
 \subsection{c}
 \paragraph{}
-Let $W$ be the unitary matrix for NOTC given by
-$$W = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \\ 0& 1 & 0 &0 \end{bmatrix}.$$
-The circuit for $(I \otimes H)W$ is
-\[ \Qcircuit @C=1em @R=1em {
- & \targ & \qw &  \qw \\
- & \ctrl{-1} & \gate{H} & \qw
-}\]
-and the corresponding unitary is
-$$C:= \frac{1}{\sqrt{2}}\begin{bmatrix} 1 & 1 & 0 & 0 \\ 1& -1 & 0 & 0 \\ 0 & 0 & 1 & 1 \\ 0& 0 & 1 &-1 \end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0& 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \\ 0& 1 & 0 &0 \end{bmatrix} = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 0 & 0 & 1 \\ 1 & 0 & 0 & -1 \\ 0 & 1 & 1 & 0\\ 0 & -1 & 1 & 0\end{bmatrix}.$$
+$2^8\equiv 9 \mod 13$ so $\log_2(9)$ is $8$ in $\Z^\times_{13}$.
 \subsection{d}
 \paragraph{}
-Since $(H\otimes H)U(H\otimes H) = W$ if and only if 
-$$(I \otimes H)(H \otimes H) U(H \otimes H) = (I\otimes H) W$$
-which simplifies as
-$$(H\otimes I) U(H\otimes H) = (I \otimes H) W,$$
-it suffices to show that $B (H\otimes H) = C$. Thus we compute
-\begin{align*}B(H\otimes H) &= \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 0 & 0 & 1 \\ 0 & 1 & 1 & 0 \\ 1 & 0 & 0 & -1\\ 0 & 1 & -1 & 0\end{bmatrix}\cdot \frac{1}{2}\begin{bmatrix} 1 & 1 & 1 & 1 \\ 1 & - 1 & 1 & -1 \\ 1 & 1& -1 & -1 \\ 1 & -1 & -1 &1\end{bmatrix} \\
-&=\frac{1}{\sqrt{2}}\cdot \frac{1}{2}\begin{bmatrix}2 &0 & 0 & 2 \\ 2 & 0 & 0 & -2 \\ 0 & 2 & 2 & 0 \\ 0 & -2 & 2 & 0 \end{bmatrix} \\
-& C. \blacksquare\end{align*}
+\begin{itemize}
+\item $3^3 \equiv 1 \mod 13$ so $3$ is not a generator
+\item $4^6 \equiv 1 \mod 13$ so $4$ is not a generator
+\item $5^4 \equiv 1 \mod 13$ so $5$ is not a generator
+\item The order of $6$ is $12$ in $\Z_{13}^\times$ so is $6$ is a generator of $Z_{13}^\times$
+\item The order of $7$ is $12$ in $\Z_{13}^\times$ so is $7$ is a generator of $Z_{13}^\times$
+\item $8^4 \equiv 1 \mod 13$ so $8$ is not a generator
+\item $9^3 \equiv 1 \mod 13$ so $9$ is not a generator
+\item $10^6 \equiv 1 \mod 13$ so $10$ is not a generator
+\item The order of $11$ is $12$ in $\Z_{13}^\times$ so is $11$ is a generator of $Z_{13}^\times$
+\item $12 \equiv 1 \mod 13$ so $12$ is not a generator
+\end{itemize}
+Hence the generators of $\Z_{13}^\times$ besides $2$ are: $6,7,$ and $11$.
+
 \section{}
-\paragraph{}
-Let $\ket{\psi} = \frac{1}{\sqrt{2}}\ket{00} + \frac{1}{2}\ket{01} + \frac{1}{2}\ket{11}.$
-\subsection{a}
-\paragraph{}
-When we measure the first qubit with respect to the computational basis ($\ket{0}, \ket{1}$) we obtain $\ket{0}$ with probability
-$$|\frac{1}{\sqrt{2}}|^2 + |\frac{1}{2}|^2 = \frac{3}{4}$$
-and the second qubit would be left in the normalized post-measurement state
-$$\sqrt{\frac{2}{3}} \ket{0} + \frac{1}{\sqrt{3}} \ket{1},$$
-and we obtain $\ket{1}$ with probability
-$$|\frac{1}{2}|^2 = \frac{1}{4}$$
-and the second qubit would be left in the normalized post-measurement state
-$$\ket{1}.$$
-\subsection{b}
-\paragraph{}
-Rewriting $\ket{\psi}$ in the $\ket{+}, \ket{-}$ basis gives
-$$\ket{\psi} = \frac{\sqrt{2} + 2}{4}\ket{++} + \frac{\sqrt{2}-2}{4}\ket{+-} + \frac{\sqrt{2}}{4}\ket{-+} + \frac{\sqrt{2}}{4}\ket{--}.$$
-So measuring the first qubit with respect to the $\ket{+}, \ket{-}$ basis gives $\ket{+}$ with probability
-$$|\frac{\sqrt{2}+2}{4}|^2+ |\frac{\sqrt{2}-2}{4}|^2 = \frac{3}{4}.$$
-and the second qubit would be left in the normalized post-measurement state
-$$ \frac{\sqrt{2} + 2}{2\sqrt{3}}\ket{+} + \frac{\sqrt{2}-2}{2\sqrt{3}}\ket{-},$$
-and gives $\ket{-}$ with probability
-$$2|\frac{\sqrt{2}}{4}|^2 = \frac{1}{4}$$
-and the second qubit would be left in the normalized post-measurement state
-$$\frac{1}{\sqrt{2}}\ket{+} + \frac{1}{\sqrt{2}}\ket{-} .$$
+
 \end{document}

### 2017-10-11 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
problem 2

@@
 \section{}
-
+\subsection{a}
+\paragraph{}
+We check the action of $H$ on $X$ and $Z$.
+$$HXH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix}1 & -1 \\ 1 & 1 \end{bmatrix} =\frac{1}{2} \begin{bmatrix} 2 & 0 \\ 0 & 2 \end{bmatrix} = I.
+$$
+$$HZH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ -1 & 1\end{bmatrix} = \frac{1}{2} \begin{bmatrix} 0 & 2 \\ 2 & 0\end{bmatrix} = X.$$
+Thus $H$ is in the Clifford group on $1$ qubit.
+\subsection{b}
+\paragraph{}
+Similar to the previous subsection we consider the action of $T$ on $X$ and $Z$.
+$$TXT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & \zeta^*_8 \\ 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & \zeta_8^* \\ \zeta_8 & 0 \end{bmatrix} = -e^{\frac{3}{4}i\pi}X.$$
+$$TZT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -\zeta^*_8\end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} = Z.$$
+Thus $T$ is in the Clifford group on $1$ qubit.
+\subsection{c}
+\paragraph{}
+$$SXS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0\\ 0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & -i\\ 1 & 0 \end{bmatrix} =\begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix} = Y.$$
+$$SZS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} \begin{bmatrix} 1 & 0 \\0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} = Z.$$
+Thus $S$ is in the Clifford group on $1$ qubit.
+\subsection{d}
+\paragraph{}
+Let $C$ be the unitary matrix representing the controlled-not gate. We consider the action of $C$ on $I \otimes X$, $X\otimes I$, $I \otimes Z$. and $Z \otimes I$.
+$$C (I\otimes X) C^\dagger = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0 \\ 0 & 0 &0 & 1 \\ 0 & 0 & 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & 1& 0& 0\\ 1 & 0  & 0& 0\\ 0 & 0 & 1 & 0\\ 0 & 0 & 0 & 1\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = I \otimes X.$$
+$$ C (X \otimes I) C^\dagger =  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}  0 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \\ 1 & 0 & 0 & 0 \\ 0 &1 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}  =  \begin{bmatrix} 0 & 0 &1 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = X \otimes X$$
+\begin{align*}C (I \otimes Z) C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}1 &0 &0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 &1 & 0\\ 0& 0& 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
+ &=  \begin{bmatrix} 1 & 0 & 0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 & 0 & -1 \\ 0 &0 &1 & 0\end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\ &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 & -1 &0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 &1\end{bmatrix} = Z \otimes Z.\end{align*}
+ \begin{align*}
+ C(Z\otimes I)C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}1 &0 &0 &0 \\0 & 1 & 0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 & -1 \end{bmatrix}  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
+ &= \begin{bmatrix} 1& 0 & 0 &0 \\0 &1 & 0 &0 \\ 0 & 0 &0 &-1 \\ 0 & 0 & -1 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
+ &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 &1 & 0 &0\\ 0 & 0 &-1 & 0 \\ 0 & 0 & 0 & -1 \end{bmatrix} = Z\otimes I.
+ \end{align*}
+ Thus $C$ is in the $2$-qubit Clifford group.
+ \subsection{e}
+ Let $A$ be the Toffoli gate (CCNOT). For the sake of space we'll abbreviate the calculations. We have that
+ $$A(I\otimes X\otimes I) A^\dagger = \begin{bmatrix} X\otimes I & 0 \\ 0 & Z\otimes Z\end{bmatrix}.$$
+ This can't be written as a tensor product $\alpha V_1\otimes V_2 \otimes V_3$ with $\alpha$ scalar and $V_1,V_2, V_3 \in \{I,X,Y,Z\}$.
+To see this observe that if we take $V_1 \in  \{X, Y\}$ then the first $4x4$ diagonal block of $V_1\otimes V_2 \otimes V_3$ will be $0$, contradicting that it should be $X \otimes I$, and if we take $V_1 \in \{I,Z\}$ then the first $4x4$ diagonal block will be $\pm1$ times the second, but $X\otimes I \neq \pm Z\otimes Z$.
+Therefore no such tensor product expression for $A(I\otimes X \otimes I)A^\dagger$ is possible, and hence the Toffoli gate is not in the $3$-qubit Clifford group.
+$\blacksquare$
 \end{document}

### 2017-10-16 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
q3
Fixed

@@
 Similar to the previous subsection we consider the action of $T$ on $X$ and $Z$.
-$$TXT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & \zeta^*_8 \\ 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & \zeta_8^* \\ \zeta_8 & 0 \end{bmatrix} = -e^{\frac{3}{4}i\pi}X.$$
-$$TZT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -\zeta^*_8\end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} = Z.$$
-Thus $T$ is in the Clifford group on $1$ qubit.
+$$TXT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & \zeta^*_8 \\ 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & \zeta_8^* \\ \zeta_8 & 0 \end{bmatrix}.$$
+Such a matrix cannot be written as  $\alpha V$ for phase $\alpha$ and $V \in \{I,X,Y,Z\}$. Thus $T$ is not in Clifford group on $1$ qubit.
 \subsection{c}
@@
 $\blacksquare$
+ \section{}
+ \subsection{a}
+ \paragraph{}
+We use the same notation for gates as in question $2$. $X_4$ is given by the circuit
+ \[ \Qcircuit @C=1em @R=1em {
+ & \targ &   \qw & \qw\\
+ & \ctrl{-1} & \gate{X} & \qw
+}\]
+$Z_4$ is given by the circuit
+\[ \Qcircuit @C=1em @R=1em {
+ & \gate{S} &  \gate{S} & \qw\\
+ & \gate{S} & \qw & \qw
+}\].
+\subsection{b}
+\paragraph{}
+$X_8$ is given by the circuit (using one Toffoli gate)
+\[ \Qcircuit @C=1em @R=1em {
+ & \targ &  \qw & \qw & \qw & \qw\\
+ & \ctrl{-1} & \targ& \qw  & \qw &\qw\\
+ & \ctrl{-1} & \ctrl{-1} & \gate{X} &\qw & \qw
+}\]
+Bonus: Figure $4.9$ of Nielsen and Chuang gives a controlled single-qubit implementation of the Toffoli gate.
+Since my use of the Toffoli gate has different controls I reflect the picture, giving the following implementation of the above Toffoli gate where the first qubit is controlled by the second and third:
+
+\[ \Qcircuit @C=1em @R=1em {
+ & \gate{H} & \targ & \gate{T^\dagger} & \targ & \gate{T} & \targ & \gate{T^\dagger}& \targ & \gate{T} & \gate{H} & \qw & \qw & \qw \\
+ & \qw & \ctrl{-1}& \qw  & \qw & \qw & \ctrl{-1} & \qw & \qw & \gate{T^\dagger} & \targ & \gate{T^\dagger} & \targ & \gate{S} &\qw\\
+ & \qw & \qw & \qw & \ctrl{-2} & \qw & \qw & \qw & \ctrl{-2} & \qw & \ctrl{-1} & \qw & \ctrl{-1} & \gate{T} & \qw
+}\]
+$Z_8$ is given by the circuit
+\[ \Qcircuit @C=1em @R=1em {
+ & \gate{T} &  \gate{T} & \gate{T} & \gate{T} & \qw\\
+ & \gate{T} & \gate{T}& \qw  & \qw &\qw\\
+ & \gate{T} & \qw & \qw &\qw & \qw
+}\]
 \end{document}

### 2017-10-16 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
question 4

@@
 }\]
+\section{}
+\subsection{a}
+\paragraph{}
+We can express $F_d, F_d^{-1}, X_d$, and $Z_d$ as sums or rank one matrices as follows:
+$$F_d = \frac{1}{\sqrt{d}} \sum_{j,k \in \Z_d} \omega^{j k} \ket{j}\bra{k}$$
+$$F_d^{-1} = \frac{1}{\sqrt{d}} \sum_{j,k \in \Z_d} \omega^{- j k} \ket{k}\bra{j}$$
+$$X_d = \sum_{k \in \Z_d} \ket{k+1 \mod d} \bra{k}$$
+$$Z_d = \sum_{k \in \Z_d} \omega^k \ket{k}\bra{k}$$
+where $\omega := \exp(2\pi i / d)$.
+\paragraph{}
+Then we have
+\begin{align*}
+F_dX_dF_d^{-1} &= \frac{1}{d} \sum_{j,k \in \Z_d} \omega^{j k} \ket{j}\bra{k}\sum_{k \in \Z_d} \ket{k+1 \mod d} \bra{k}\sum_{j,k \in \Z_d} \omega^{- j k} \ket{k}\bra{j}\\
+&= \frac{1}{d} \sum_{j,k \in \Z_d} \omega^{j k} \ket{j}\bra{k} \sum_{j,k \in \Z_d} \omega^{- j k} \ket{k+1 \mod d}\bra{j} \\
+&= \frac{1}{d} \sum_{j,k,\ell \in \Z_d}\omega^{\ell(k+1\mod d)}\omega^{-jk}\ket{\ell}\bra{j} \\
+&=\frac{1}{d} \sum_{j,k,\ell \in \Z_d}\omega^{\ell(k+1)}\omega^{-jk}\ket{\ell}\bra{j}  \quad\text{(we may drop mod since $\omega^{\ell 0 } = \omega^{\ell d} = 1$)} \\
+&= \frac{1}{d} \sum_{j,k,\ell \in \Z_d}\omega^{k(\ell-j)}\omega^{\ell}\ket{\ell}\bra{j} \\
+&= \sum_{\ell \in \Z_d} \omega^\ell \ket{\ell}\bra{\ell} \quad \text{(applying Kronecker Delta summing over $k$)} \\
+&= Z_d.
+\end{align*}
+So $F_d X_d F_d^{-1} = Z_d$. Thus we have (we will show the desired form of $F_d^2$ is part $b$)
+\begin{align*}
+F_dZ_d F_d^{-1} &= F_d^2 X_d F_d^{-2} \\
+&= \sum_{k \in Z_d} \ket{-k \mod d} \bra{k} \sum_{k \in \Z_d} \ket{k+1 \mod d}\bra{k} \sum_{k \in \Z_d} \ket{k}\bra{-k\mod d} \\
+&= \sum_{k \in Z_d} \ket{-k \mod d} \bra{k} \sum_{k \in Z_d}\ket{k+1 \mod d}\bra{-k \mod d} \\
+&= \sum_{k \in Z_d} \ket{-(k+1) \mod d} \bra{- k \mod d} \\
+&= \sum_{k \in Z_d} \ket{k+1 \mod d} \bra{k \mod d} \\
+&= X_d.
+\end{align*}
+Hence $F_d Z_d F_d^{-1} = X_d$.
+\subsection{b}
+Let $p \in \Z_d$. Then we have
+\begin{align*}
+F^2_d\ket{k} &= \frac{1}{d} \sum_{j,k \in \Z_d} \omega^{j k} \ket{j}\bra{k}\sum_{j,k \in \Z_d} \omega^{j k} \ket{j}\bra{k} \ket{p} \\
+&= \frac{1}{d} \sum_{j,k \in \Z_d} \omega^{j k} \ket{j}\bra{k}\sum_{j \in \Z_d}\omega^{jp}\ket{j} \\
+&=\frac{1}{d} \sum_{j,k \in \Z_d} \omega^{k(j+ p)}\ket{j}.
+\end{align*}
+For fixed $j$, $$\frac{1}{d} \sum_{k \in \Z_d} \omega^{k(j+p)} = 1$$ if and only if $$j+p = 0 \mod d$$ and otherwise $$\frac{1}{d} \sum_{k \in \Z_d} \omega^{k(j+p)} = 0$$. Hence 
+$$F^2_d \ket{k} = \ket{-k \mod d}$$
+and so
+$$F^2_d = \sum_{k \in \Z_d} \ket{-k \mod d} \bra{k}.$$
+\subsection{c}
+Since $-0 \mod d \equiv 0$ and $-d/2 \mod d \equiv d/2$ ($d$ is even), we have that
+$$F^2_d \ket{0} = \ket{0}$$
+and $$F^2_d\ket{d/2} = \ket{d/2}$$
+so $1$ is an eigenvalue of $F^2_d$. Further, let $k \in \Z_d \backslash \{0, d/2\}$. Then
+$$F^2_d(\ket{k} + \ket{-k \mod d}) = (\ket{k} + \ket{-k \mod d})$$
+and
+$$F^2_d(\ket{k} - \ket{-k \mod d}) = -\ket{k} + \ket{-k\mod d}$$
+so $-1$ is another eigenvalue of of $F^2_d$, and we have found a basis of $\C^d$ consisting of eigenvectors of $F_d^2$ (the two eigenvectors generated for each such $k$ above, and $\ket{0}$ and $\ket{d/2}$).
+Hence we have found all eigenvalues of $F^2_d$.
+That is the eigenvalues of $F^2_d$ are precisely $\pm 1$.
+\subsection{d}
+\paragraph{}
+We check the action of $F_4$:
+\begin{align*}
+F_4(X\otimes I) F_4^{-1} &= \frac{1}{4} \sum_{j,k = 0}^3 \omega^{jk}\ket{j}\bra{k}(\ket{2}\bra{0} + \ket{3}\bra{1} + \ket{0}\bra{2} + \ket{1}\bra{3})\sum_{j,k = 0}^3 \omega^{-jk}\ket{k}\bra{j}\\
+&= \frac{1}{4} \sum_{j,k = 0}^3 \omega^{jk}\ket{j}\bra{k}(\sum_{\ell=0}^3 \ket{\ell}\bra{\ell + 2 \mod 4})\sum_{j,k=0}^3\omega^{-jk}\ket{k}\bra{j}\\
+&=  \frac{1}{4} \sum_{j,k = 0}^3 \omega^{jk}\ket{j}\bra{k}\sum_{\ell,j=0}^3 \omega^{-j(\ell+2)}\ket{\ell}\bra{j} \\
+&=\frac{1}{4} \sum_{j,k,\ell=0}^3 \omega^{jk}\omega^{-\ell(k+2)}\ket{j}\bra{\ell} \\
+&=\frac{1}{4} \sum_{j,k,\ell=0}^3\omega^{-2\ell} \omega^{k(j-\ell)} \ket{j}\bra{\ell}\\
+&=\sum_{\ell=0}^3\omega^{-2\ell} \ket{\ell}\bra{\ell}\\
+&= I \otimes Z.
+\end{align*}
+\begin{align*}
+F_4(I\otimes X) F_4^{-1} &= \frac{1}{4} \sum_{j,k = 0}^3 \omega^{jk}\ket{j}\bra{k}(\ket{1}\bra{0} + \ket{0}\bra{1} + \ket{3}\bra{2} +\ket{2}\bra{3})\sum_{j,k = 0}^3 \omega^{-jk}\ket{k}\bra{j} \\
+&= \frac{1}{4}(\sum_{j,k=0}^3(\omega^{j} + \omega^{-k} + \omega^{3j - 2k} + \omega^{2j - 3k})\ket{j}\bra{k} \\
+&= \ket{0}\bra{0} + -i\ket{3}\bra{1} - \ket{2}\bra{2} + i \ket{1}\bra{3}
+\end{align*}
+which cannot be written as a tensor product of $\{I, X, Y, Z\}$. Hence $F_4$ is not in the Clifford group.
 \end{document}

### 2017-10-16 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
temporary solution to 5

@@
 F_4(I\otimes X) F_4^{-1} &= \frac{1}{4} \sum_{j,k = 0}^3 \omega^{jk}\ket{j}\bra{k}(\ket{1}\bra{0} + \ket{0}\bra{1} + \ket{3}\bra{2} +\ket{2}\bra{3})\sum_{j,k = 0}^3 \omega^{-jk}\ket{k}\bra{j} \\
-&= \frac{1}{4}(\sum_{j,k=0}^3(\omega^{j} + \omega^{-k} + \omega^{3j - 2k} + \omega^{2j - 3k})\ket{j}\bra{k} \\
+&= \frac{1}{4}(\sum_{j,k=0}^3(\omega^{j} + \omega^{-k} + \omega^{3j - 2k} + \omega^{2j - 3k}))\ket{j}\bra{k} \\
 &= \ket{0}\bra{0} + -i\ket{3}\bra{1} - \ket{2}\bra{2} + i \ket{1}\bra{3}
@@
 which cannot be written as a tensor product of $\{I, X, Y, Z\}$. Hence $F_4$ is not in the Clifford group.
+\section{}
+\paragraph{}
+Suppose that $m$ is prime.
+Let $\lambda \in \Z_m$.
+Let $r \in \Z_m\times \Z_m \backslash \{(0,0)\}$.
+We may assume without loss of generality that $r_1 \neq 0$.
+We define $L_\lambda$ and $\ket{L_\lambda}$ as in the question preamble.
+Then we have
+\begin{align*}
+(F^{-1}_m \otimes F^{-1}_m) \ket{L_\lambda} &= \frac{1}{\sqrt{m}} \sum_{j,k \in \Z_m} \frac{1}{m} \sum_{(x_1,x_2) \in L_\lambda}\omega^{x_1j + x_2k}\ket{j}\ket{k}.
+\end{align*}
+For a fixed $j,k \in \Z_m\times \Z_m$ the $\ket{j}\ket{k}$ term in the above vanishes iff $\sum_{(x_1,x_2) \in L_\lambda}\omega^{x_1j + x_2k} = 0$.
+From the hint we know that $$\sum_{s \in \Z_m} \omega^s = 0.$$
+Hence the $\ket{j}\ket{k}$ term vanishes iff for all $s \in \Z_m$ there exists $(x_1,x_2) \in L_\lambda$ such that $x_1j + x_2k = s$ (this establishes a bijection from $L_\lambda$ to $\Z_m$, proving $\abs{L_\lambda} = m$, as we'll see below such $x$ must be unique).
+\paragraph{}
+For all $s \in \Z_m$ we want $x \in \Z_m \times \Z_m$ satisfying
+$$x_1j + x_2k = s \quad\text{and}\quad x_1r_1 + x_2r_2 = \lambda.$$
+This system has no solution provided
+$$jr_1 = kr_2 \quad \text{and} \quad s -kr_1^{-1}\lambda\neq 0 \quad(\text{mod } m)$$
+The first condition is satisfied iff $(j,k)$ is a scalar multiple of $(r_1,r_2)$.
+Given that the first condition is satisfied, the second condition is satisfied for all $s \in \Z_m$, save $s = -r^{-1}_1k\lambda$.
+Thus we have
+$$(F^{-1}_m \otimes F^{-1}_m) \ket{L_\lambda} = \frac{1}{\sqrt{m}} \sum_{\alpha \in \Z_m} \omega^{\alpha \lambda}\ket{\alpha r_1}\ket{\alpha r_2}.$$
+So measuring this state will yield a scalar multiple of $(r_1,r_2)$ each with equal probability. This does not reveal any information about $\lambda$.
 \end{document}

### 2017-10-30 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
a3 getting started

--- 2017-10-16 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
@@
 \lhead{\today}
-\rhead{W. Justin Toth CO681-Quantum Information Processing A2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CO681-Quantum Information Processing A3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \paragraph{}
-$$11+12 \mod 13 = 10$$
+The phase estimation algorithm can be thought of as running a unitary matrix $W$ which maps $\ket{0}\ket{\phi}$ to $\ket{a}\ket{\phi}$ for any eigenvector $\ket{\phi}$ of $U$ with eigenvalue $\omega^a$, and then measuring the first register.
+So if we consider $\alpha\ket{\psi} + \beta\ket{\phi}$ then running phase estimation yields
+$$W(\ket{0}(\alpha\ket{\psi} + \beta\ket{\phi})) = \alpha W(\ket{0}\ket{\psi}) + \beta W(\ket{0}\ket{\phi}) = \alpha \ket{a}\ket{\psi} + \beta\ket{b}\ket{\phi}$$
+before measuring. So measuring the first register yields $\ket{a}$ with probability $|\alpha|^2$ and $b$ with probability $|\beta|^2$.
 \subsection{b}
 \paragraph{}
-$$\Z^\times_{13} = \{2^0=1, 2^1 = 2, 2^2 = 4, 2^3 = 8, 2^4 = 3, 2^5 = 6, 2^6=12, 2^7=11, 2^8 = 9, 2^9=5, 2^{10}=10, 2^{11} = 7\}$$
-\subsection{c}
-\paragraph{}
-$2^8\equiv 9 \mod 13$ so $\log_2(9)$ is $8$ in $\Z^\times_{13}$.
-\subsection{d}
-\paragraph{}
-\begin{itemize}
-\item $3^3 \equiv 1 \mod 13$ so $3$ is not a generator
-\item $4^6 \equiv 1 \mod 13$ so $4$ is not a generator
-\item $5^4 \equiv 1 \mod 13$ so $5$ is not a generator
-\item The order of $6$ is $12$ in $\Z_{13}^\times$ so is $6$ is a generator of $Z_{13}^\times$
-\item The order of $7$ is $12$ in $\Z_{13}^\times$ so is $7$ is a generator of $Z_{13}^\times$
-\item $8^4 \equiv 1 \mod 13$ so $8$ is not a generator
-\item $9^3 \equiv 1 \mod 13$ so $9$ is not a generator
-\item $10^6 \equiv 1 \mod 13$ so $10$ is not a generator
-\item The order of $11$ is $12$ in $\Z_{13}^\times$ so is $11$ is a generator of $Z_{13}^\times$
-\item $12 \equiv 1 \mod 13$ so $12$ is not a generator
-\end{itemize}
-Hence the generators of $\Z_{13}^\times$ besides $2$ are: $6,7,$ and $11$.
 
@@
 \paragraph{}
-We check the action of $H$ on $X$ and $Z$.
-$$HXH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix}1 & -1 \\ 1 & 1 \end{bmatrix} =\frac{1}{2} \begin{bmatrix} 2 & 0 \\ 0 & 2 \end{bmatrix} = I.
-$$
-$$HZH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ -1 & 1\end{bmatrix} = \frac{1}{2} \begin{bmatrix} 0 & 2 \\ 2 & 0\end{bmatrix} = X.$$
-Thus $H$ is in the Clifford group on $1$ qubit.
-\subsection{b}
-\paragraph{}
-Similar to the previous subsection we consider the action of $T$ on $X$ and $Z$.
-$$TXT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & \zeta^*_8 \\ 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & \zeta_8^* \\ \zeta_8 & 0 \end{bmatrix}.$$
-Such a matrix cannot be written as  $\alpha V$ for phase $\alpha$ and $V \in \{I,X,Y,Z\}$. Thus $T$ is not in Clifford group on $1$ qubit.
-\subsection{c}
-\paragraph{}
-$$SXS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0\\ 0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & -i\\ 1 & 0 \end{bmatrix} =\begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix} = Y.$$
-$$SZS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} \begin{bmatrix} 1 & 0 \\0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} = Z.$$
-Thus $S$ is in the Clifford group on $1$ qubit.
-\subsection{d}
-\paragraph{}
-Let $C$ be the unitary matrix representing the controlled-not gate. We consider the action of $C$ on $I \otimes X$, $X\otimes I$, $I \otimes Z$. and $Z \otimes I$.
-$$C (I\otimes X) C^\dagger = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0 \\ 0 & 0 &0 & 1 \\ 0 & 0 & 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & 1& 0& 0\\ 1 & 0  & 0& 0\\ 0 & 0 & 1 & 0\\ 0 & 0 & 0 & 1\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = I \otimes X.$$
-$$ C (X \otimes I) C^\dagger =  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}  0 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \\ 1 & 0 & 0 & 0 \\ 0 &1 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}  =  \begin{bmatrix} 0 & 0 &1 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = X \otimes X$$
-\begin{align*}C (I \otimes Z) C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}1 &0 &0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 &1 & 0\\ 0& 0& 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
- &=  \begin{bmatrix} 1 & 0 & 0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 & 0 & -1 \\ 0 &0 &1 & 0\end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\ &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 & -1 &0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 &1\end{bmatrix} = Z \otimes Z.\end{align*}
- \begin{align*}
- C(Z\otimes I)C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}1 &0 &0 &0 \\0 & 1 & 0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 & -1 \end{bmatrix}  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
- &= \begin{bmatrix} 1& 0 & 0 &0 \\0 &1 & 0 &0 \\ 0 & 0 &0 &-1 \\ 0 & 0 & -1 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
- &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 &1 & 0 &0\\ 0 & 0 &-1 & 0 \\ 0 & 0 & 0 & -1 \end{bmatrix} = Z\otimes I.
- \end{align*}
- Thus $C$ is in the $2$-qubit Clifford group.
- \subsection{e}
- Let $A$ be the Toffoli gate (CCNOT). For the sake of space we'll abbreviate the calculations. We have that
- $$A(I\otimes X\otimes I) A^\dagger = \begin{bmatrix} X\otimes I & 0 \\ 0 & Z\otimes Z\end{bmatrix}.$$
- This can't be written as a tensor product $\alpha V_1\otimes V_2 \otimes V_3$ with $\alpha$ scalar and $V_1,V_2, V_3 \in \{I,X,Y,Z\}$.
-To see this observe that if we take $V_1 \in  \{X, Y\}$ then the first $4x4$ diagonal block of $V_1\otimes V_2 \otimes V_3$ will be $0$, contradicting that it should be $X \otimes I$, and if we take $V_1 \in \{I,Z\}$ then the first $4x4$ diagonal block will be $\pm1$ times the second, but $X\otimes I \neq \pm Z\otimes Z$.
-Therefore no such tensor product expression for $A(I\otimes X \otimes I)A^\dagger$ is possible, and hence the Toffoli gate is not in the $3$-qubit Clifford group.
-$\blacksquare$
- \section{}
- \subsection{a}
- \paragraph{}
-We use the same notation for gates as in question $2$. $X_4$ is given by the circuit
- \[ \Qcircuit @C=1em @R=1em {
- & \targ &   \qw & \qw\\
- & \ctrl{-1} & \gate{X} & \qw
-}\]
-$Z_4$ is given by the circuit
-\[ \Qcircuit @C=1em @R=1em {
- & \gate{S} &  \gate{S} & \qw\\
- & \gate{S} & \qw & \qw
-}\].
-\subsection{b}
-\paragraph{}
-$X_8$ is given by the circuit (using one Toffoli gate)
-\[ \Qcircuit @C=1em @R=1em {
- & \targ &  \qw & \qw & \qw & \qw\\
- & \ctrl{-1} & \targ& \qw  & \qw &\qw\\
- & \ctrl{-1} & \ctrl{-1} & \gate{X} &\qw & \qw
-}\]
-Bonus: Figure $4.9$ of Nielsen and Chuang gives a controlled single-qubit implementation of the Toffoli gate.
-Since my use of the Toffoli gate has different controls I reflect the picture, giving the following implementation of the above Toffoli gate where the first qubit is controlled by the second and third:
+The probability of getting outcome $j$ is
+$$Tr(A_j\ket{\psi_j}\bra{\psi_j}A_j^\dagger) = \frac{1}{2} Tr(\ket{\psi_j}\bra{\psi_j}\ket{\psi_j}\bra{\psi_j}\ket{\psi_j}\bra{\psi_j}) = \frac{1}{2}Tr(\ip{\psi}{\psi}^3) = \frac{1}{2}.$$
 
-\[ \Qcircuit @C=1em @R=1em {
- & \gate{H} & \targ & \gate{T^\dagger} & \targ & \gate{T} & \targ & \gate{T^\dagger}& \targ & \gate{T} & \gate{H} & \qw & \qw & \qw \\
- & \qw & \ctrl{-1}& \qw  & \qw & \qw & \ctrl{-1} & \qw & \qw & \gate{T^\dagger} & \targ & \gate{T^\dagger} & \targ & \gate{S} &\qw\\
- & \qw & \qw & \qw & \ctrl{-2} & \qw & \qw & \qw & \ctrl{-2} & \qw & \ctrl{-1} & \qw & \ctrl{-1} & \gate{T} & \qw
-}\]
-$Z_8$ is given by the circuit
-\[ \Qcircuit @C=1em @R=1em {
- & \gate{T} &  \gate{T} & \gate{T} & \gate{T} & \qw\\
- & \gate{T} & \gate{T}& \qw  & \qw &\qw\\
- & \gate{T} & \qw & \qw &\qw & \qw
-}\]
 \end{document}

### 2017-10-31 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
30 percent done
1b

@@
 \paragraph{}
-
+Let $\ket{\psi}$ be a normal vector in $\C^k$.
+Let $\ket{\phi^i}$ for $i=1,\dots, k$ be orthonormal eigenvectors of $U$ with corresponding eigenvalues $\lambda_1, \dots, \lambda_k$.
+Then since the set of $\ket{\phi^i}$ is a basis, there exist $\alpha_1, \dots, \alpha_k$ such that
+$$\ket{\psi} = \sum_{i=1}^k \alpha_i \ket{\phi^i}.$$
+So then $W$ acts on $\ket{\psi}$ as
+$$W\ket{\psi} = \sum_{i=1}^k \alpha_i \ket{\lambda_i}\ket{\phi^i}.$$
+Measuring the second register then yields $\ket{\phi^i}$ with probability $|\alpha_i|^2$.
+So then if we let $A_i = \ket{\phi^i}\bra{\phi^i}$ then $A_1, \dots, A_k$ forms a general measurement.
+If we let $\rho = \ket{\psi}\bra{\psi}$ then $Tr(A_i\rho A_i^\dagger) = |\alpha|^2$ and $\frac{1}{Tr(A_i\rho A_i^\dagger)} A_i \rho A_i^\dagger = \ket{\phi^i}\bra{\phi^i}$ as desired.
 \section{}
@@
 $$Tr(A_j\ket{\psi_j}\bra{\psi_j}A_j^\dagger) = \frac{1}{2} Tr(\ket{\psi_j}\bra{\psi_j}\ket{\psi_j}\bra{\psi_j}\ket{\psi_j}\bra{\psi_j}) = \frac{1}{2}Tr(\ip{\psi}{\psi}^3) = \frac{1}{2}.$$
-
+\subsection{b}
+\paragraph{}
+Let $j \in \{0,1,2,3\}$. The density matrix resulting from measuring $\ket{\psi_j}$ is
+$$\frac{1}{2}\ket{\psi_j}\bra{\psi_j} + \frac{1}{2}\ip{\psi_j}{\psi_{j+1}}^2\ket{\psi_{j+1}}\bra{\psi_{j+1}} + \frac{1}{2}\ip{\psi_j}{\psi_{j+3}}^2\ket{\psi_{j+3}}\bra{\psi_{j+3}}$$
+where addition in the subscripts is taken modulo $4$. This simplifies as
+$$\frac{1}{2}\ket{\psi_j}\bra{\psi_j} + \frac{1}{4}\ket{\psi_{j+1}}\bra{\psi_{j+1}} + \frac{1}{4}\ket{\psi_{j+3}}\bra{\psi_{j+3}}.$$
 \end{document}

### 2017-11-01 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
Completed 2

@@
 Measuring the second register then yields $\ket{\phi^i}$ with probability $|\alpha_i|^2$.
-So then if we let $A_i = \ket{\phi^i}\bra{\phi^i}$ then $A_1, \dots, A_k$ forms a general measurement.
+So then if we let $A_i = \ket{\phi^i}\bra{\phi^i}$ then $A_1, \dots, A_k$ forms a general measurement since the set $A_i$'s is a set of $k$ rank $1$ orthogonal projectors.
 If we let $\rho = \ket{\psi}\bra{\psi}$ then $Tr(A_i\rho A_i^\dagger) = |\alpha|^2$ and $\frac{1}{Tr(A_i\rho A_i^\dagger)} A_i \rho A_i^\dagger = \ket{\phi^i}\bra{\phi^i}$ as desired.
@@
 $$\frac{1}{2}\ket{\psi_j}\bra{\psi_j} + \frac{1}{4}\ket{\psi_{j+1}}\bra{\psi_{j+1}} + \frac{1}{4}\ket{\psi_{j+3}}\bra{\psi_{j+3}}.$$
+\subsection{c}
+We implement the desired unitary matrix $U$ via the following quantum circuit
+\[ \Qcircuit @C=1em @R=1em {
+ \ket{k}\ \  \ & \qw &  \gate{H} & \qw\\
+  \ket{0}\ \ \  & \gate{H} & \ctrl{-1} & \qw
+}\]
+with $k \in \{0,1\}$.
+When $k=0$ the circuit outputs the state $$\frac{1}{\sqrt{2}}\ket{00} + \frac{1}{2}\ket{01} + \frac{1}{2}\ket{11}$$ and when $k=1$ then circuit outputs the state $$\frac{1}{\sqrt{2}}\ket{10} + \frac{1}{2}\ket{01} - \frac{1}{2}\ket{11}$$
+as desired.
 \end{document}

### 2017-11-02 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
finished 3a

@@
 as desired.
+\section{}
+\subsection{a}
+\paragraph{}
+We can write any density matrix $\rho$ as
+$$\frac{1}{2}\begin{bmatrix} 1+ r_z & r_x - i r_y \\ r_x + i r_y & 1-r_z \end{bmatrix}$$
+Computing $T\rho T^{-1}$ then yields
+$$\frac{1}{2}\begin{bmatrix} 1+r_z & \frac{r_x - r_y}{\sqrt{2}} -i\frac{r_x + r_y}{\sqrt{2}} \\ \frac{r_x - r_y}{\sqrt{2}} + i\frac{r_x + r_y}{\sqrt{2}} & 1-r_z\end{bmatrix}.$$
+Then the corresponding map on $r$ sends $(r_x, r_y, r_z)$ to
+$$ (\frac{r_x - r_y}{\sqrt{2}}, \frac{r_x+r_y}{\sqrt{2}}, r_z).$$
 \end{document}

### 2017-11-02 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
3b

--- 2017-10-31 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
@@
 $$\frac{1}{2}\ket{\psi_j}\bra{\psi_j} + \frac{1}{4}\ket{\psi_{j+1}}\bra{\psi_{j+1}} + \frac{1}{4}\ket{\psi_{j+3}}\bra{\psi_{j+3}}.$$
+\subsection{c}
+We implement the desired unitary matrix $U$ via the following quantum circuit
+\[ \Qcircuit @C=1em @R=1em {
+ \ket{k}\ \  \ & \qw &  \gate{H} & \qw\\
+  \ket{0}\ \ \  & \gate{H} & \ctrl{-1} & \qw
+}\]
+with $k \in \{0,1\}$.
+When $k=0$ the circuit outputs the state $$\frac{1}{\sqrt{2}}\ket{00} + \frac{1}{2}\ket{01} + \frac{1}{2}\ket{11}$$ and when $k=1$ then circuit outputs the state $$\frac{1}{\sqrt{2}}\ket{10} + \frac{1}{2}\ket{01} - \frac{1}{2}\ket{11}$$
+as desired.
+\section{}
+\subsection{a}
+\paragraph{}
+We can write any density matrix $\rho$ as
+$$\frac{1}{2}\begin{bmatrix} 1+ r_z & r_x - i r_y \\ r_x + i r_y & 1-r_z \end{bmatrix}$$
+Computing $T\rho T^{-1}$ then yields
+$$\frac{1}{2}\begin{bmatrix} 1+r_z & \frac{r_x - r_y}{\sqrt{2}} -i\frac{r_x + r_y}{\sqrt{2}} \\ \frac{r_x - r_y}{\sqrt{2}} + i\frac{r_x + r_y}{\sqrt{2}} & 1-r_z\end{bmatrix}.$$
+Then the corresponding affine linear map on $r$ sends $(r_x, r_y, r_z)$ to
+$$ (\frac{r_x - r_y}{\sqrt{2}}, \frac{r_x+r_y}{\sqrt{2}}, r_z).$$
+\subsection{b}
+\paragraph{}
+Using the same notation for $\rho$ as the previous part, we compute 
+$$A_0\rho A_0^\dagger = \frac{1}{2}\begin{bmatrix}
+1+r_z & \sqrt{p}(r_x - i r_y) \\ \sqrt{r}(r_x + ir_y) & p(1-r_z)
+\end{bmatrix}$$
+and $$A_1 \rho A_1^\dagger = \frac{1}{2}\begin{bmatrix} (1-p)(1-r_z) & 0 \\ 0 & 0 \end{bmatrix}.$$
+So, adding the above and simplifying, the operator on $\rho$ yields
+$$\begin{bmatrix}
+1+ pr_z + 1-p & \sqrt{p}(r_x - i r_y) \\ \sqrt{p}(r_x + i r_y) & 1 - (pr_z + 1-p)
+\end{bmatrix}.$$
+Then the corresponding affine linear map on $r$ sends $(r_x, r_y, r_z)$ to
+$$ (\sqrt{p}r_x, \sqrt{p}r_y, pr_z + 1-p).$$
+\subsection{c}
 \end{document}

### 2017-11-04 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
starting 4a

@@
 \subsection{c}
+
+\section{}
+\subsection{a}
+The density matrix for $\alpha\ket{00} + \beta\ket{11}$ is
+$$\rho = |\alpha|^2\ket{00}\bra{00} + \alpha\beta^*\ket{00}\bra{11} + \alpha^*\beta\ket{11}\bra{00} + |\beta|^2\ket{11}\bra{11}.$$
+The partial trace operator $Tr_1$ is defined by matrices $A_0 = \ket{0} \otimes I$ and $A_1 = \ket{1} \otimes I$. Now we compute
+$$A_0\rho A_0^\dagger + A_1\rho A_1^\dagger = $$
 \end{document}

### 2017-11-19 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
Nearly solved 5

--- 2017-11-02 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
@@
 \subsection{c}
+
+\section{}
+\subsection{a}
+The density matrix for $\alpha\ket{00} + \beta\ket{11}$ is
+$$\rho = |\alpha|^2\ket{00}\bra{00} + \alpha\beta^*\ket{00}\bra{11} + \alpha^*\beta\ket{11}\bra{00} + |\beta|^2\ket{11}\bra{11}.$$
+Expanding in matrix notation we have
+$$\rho = \begin{bmatrix}
+|\alpha|^2 &0 & 0 &\alpha\beta^* \\
+0 & 0 & 0 & 0\\
+0 & 0 & 0 & 0\\
+\alpha\beta^* & 0 & 0 &|\beta|^2.
+\end{bmatrix}$$
+So using the explicit formula for $Tr_1(\rho)$ we have
+$$Tr_1(\rho) = \begin{bmatrix}
+|\alpha|^2 & 0 \\
+0 & |\beta|^2
+\end{bmatrix}.$$
+
+\subsection{b}
+The second qubit is in state $\ket{0}$ with probability $|\alpha|^2$ and state $\ket{1}$ with probability $|\beta|^2$. Hence its average density matrix is
+$$ \begin{bmatrix}
+|\alpha|^2 & 0 \\
+0 & |\beta|^2
+\end{bmatrix}.$$
+
+\subsection{c}
+Let $A_0 = \sqrt{\frac{2}{3}}\begin{bmatrix}
+1 & 0 \\ 0 & 0
+\end{bmatrix}$, let $A_1 = \frac{1}{4} \begin{bmatrix}
+\sqrt{2/3} & -\sqrt{2} \\ -\sqrt{2} & \sqrt{6}
+\end{bmatrix}$, and let $A_2 = \frac{1}{4} \begin{bmatrix}
+\sqrt{2/3} & \sqrt{2} \\ \sqrt{2} & \sqrt{6}
+\end{bmatrix}.$
+Let $\sigma = Tr_1(\rho) = \begin{bmatrix}
+|\alpha|^2 & 0 \\
+0 & |\beta|^2\end{bmatrix}$ be the density matrix of the second qubit after tracing out the first. Then
+$$A_0\sigma A_0^\dagger = \frac{2}{3}\begin{bmatrix}
+|\alpha|^2 & 0 \\ 0 & 0
+\end{bmatrix}$$
+and
+$$Tr(A_0\sigma A_0^\dagger) = \frac{2}{3}|\alpha|^2.$$
+Further
+\begin{align*}A_1\sigma A_1^\dagger &= \frac{1}{16} \begin{bmatrix}
+\sqrt{2/3} & -\sqrt{2} \\ -\sqrt{2} & \sqrt{6}
+\end{bmatrix}\begin{bmatrix}
+\sqrt{2/3}|\alpha|^2 & -\sqrt{2}|\alpha|^2 \\ -\sqrt{2}|\beta|^2 & \sqrt{6}|\beta|^2 \\&= 
+\end{bmatrix}\\&= \frac{1}{16}\begin{bmatrix}
+\frac{2}{3}|\alpha|^2 + 2|\beta|^2 & -\sqrt{2}(\sqrt{2/3}|\alpha|^2 + \sqrt{6}|\beta|^2) \\
+ -\sqrt{2}(\sqrt{2/3}|\alpha|^2 + \sqrt{6}|\beta|^2) & 2|\alpha|^2 + 6|\beta|^2
+\end{bmatrix}\end{align*}
+and
+$$Tr(A_1\sigma A_1^\dagger) = \frac{1}{16}(\frac{2}{3}|\alpha|^2 + 2|\beta|^2 + 2|\alpha|^2 + 6|\beta|^2)=\frac{1}{16}\big( (\frac{2}{3} + 2)|\alpha|^2 + 8|\beta|^2 \big).$$
+Finally
+$$A_2 \sigma A_2^\dagger = \frac{1}{16}\begin{bmatrix}
+\frac{2}{3}|\alpha|^2 + 2|\beta|^2 & \sqrt{2}(\sqrt{2/3}|\alpha|^2 + \sqrt{6}|\beta|^2) \\
+ \sqrt{2}(\sqrt{2/3}|\alpha|^2 + \sqrt{6}|\beta|^2) & 2|\alpha|^2 + 6|\beta|^2
+\end{bmatrix}$$
+and
+$$Tr(A_2 \sigma A_2^\dagger) =\frac{1}{16}\big( (\frac{2}{3} + 2)|\alpha|^2 + 8|\beta|^2 \big).$$
+The measurement probability for outcome $j \in \{0,1,2\}$ is $Tr(A_j \sigma A_j^\dagger)$ and the post-measurement state is $\frac{1}{Tr(A_j \sigma A_j^\dagger)}A_j\sigma A_j^\dagger.$
+
+The average density matrix for the second qubit is
+$$\sum_{j=0}^2 A_j \sigma A_j^\dagger.$$
 \end{document}

### 2017-11-29 CO681-QuantumInformationProcessing/assignments/assignment_05.tex
1a

--- 2017-11-02 CO681-QuantumInformationProcessing/assignments/assignment_03.tex
@@
 \lhead{\today}
-\rhead{W. Justin Toth CO681-Quantum Information Processing A3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CO681-Quantum Information Processing A5} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \paragraph{}
-The phase estimation algorithm can be thought of as running a unitary matrix $W$ which maps $\ket{0}\ket{\phi}$ to $\ket{a}\ket{\phi}$ for any eigenvector $\ket{\phi}$ of $U$ with eigenvalue $\omega^a$, and then measuring the first register.
-So if we consider $\alpha\ket{\psi} + \beta\ket{\phi}$ then running phase estimation yields
-$$W(\ket{0}(\alpha\ket{\psi} + \beta\ket{\phi})) = \alpha W(\ket{0}\ket{\psi}) + \beta W(\ket{0}\ket{\phi}) = \alpha \ket{a}\ket{\psi} + \beta\ket{b}\ket{\phi}$$
-before measuring. So measuring the first register yields $\ket{a}$ with probability $|\alpha|^2$ and $b$ with probability $|\beta|^2$.
-\subsection{b}
-\paragraph{}
-Let $\ket{\psi}$ be a normal vector in $\C^k$.
-Let $\ket{\phi^i}$ for $i=1,\dots, k$ be orthonormal eigenvectors of $U$ with corresponding eigenvalues $\lambda_1, \dots, \lambda_k$.
-Then since the set of $\ket{\phi^i}$ is a basis, there exist $\alpha_1, \dots, \alpha_k$ such that
-$$\ket{\psi} = \sum_{i=1}^k \alpha_i \ket{\phi^i}.$$
-So then $W$ acts on $\ket{\psi}$ as
-$$W\ket{\psi} = \sum_{i=1}^k \alpha_i \ket{\lambda_i}\ket{\phi^i}.$$
-Measuring the second register then yields $\ket{\phi^i}$ with probability $|\alpha_i|^2$.
-So then if we let $A_i = \ket{\phi^i}\bra{\phi^i}$ then $A_1, \dots, A_k$ forms a general measurement since the set $A_i$'s is a set of $k$ rank $1$ orthogonal projectors.
-If we let $\rho = \ket{\psi}\bra{\psi}$ then $Tr(A_i\rho A_i^\dagger) = |\alpha|^2$ and $\frac{1}{Tr(A_i\rho A_i^\dagger)} A_i \rho A_i^\dagger = \ket{\phi^i}\bra{\phi^i}$ as desired.
-\section{}
-\subsection{a}
-\paragraph{}
-The probability of getting outcome $j$ is
-$$Tr(A_j\ket{\psi_j}\bra{\psi_j}A_j^\dagger) = \frac{1}{2} Tr(\ket{\psi_j}\bra{\psi_j}\ket{\psi_j}\bra{\psi_j}\ket{\psi_j}\bra{\psi_j}) = \frac{1}{2}Tr(\ip{\psi}{\psi}^3) = \frac{1}{2}.$$
-\subsection{b}
-\paragraph{}
-Let $j \in \{0,1,2,3\}$. The density matrix resulting from measuring $\ket{\psi_j}$ is
-$$\frac{1}{2}\ket{\psi_j}\bra{\psi_j} + \frac{1}{2}\ip{\psi_j}{\psi_{j+1}}^2\ket{\psi_{j+1}}\bra{\psi_{j+1}} + \frac{1}{2}\ip{\psi_j}{\psi_{j+3}}^2\ket{\psi_{j+3}}\bra{\psi_{j+3}}$$
-where addition in the subscripts is taken modulo $4$. This simplifies as
-$$\frac{1}{2}\ket{\psi_j}\bra{\psi_j} + \frac{1}{4}\ket{\psi_{j+1}}\bra{\psi_{j+1}} + \frac{1}{4}\ket{\psi_{j+3}}\bra{\psi_{j+3}}.$$
-\subsection{c}
-We implement the desired unitary matrix $U$ via the following quantum circuit
-\[ \Qcircuit @C=1em @R=1em {
- \ket{k}\ \  \ & \qw &  \gate{H} & \qw\\
-  \ket{0}\ \ \  & \gate{H} & \ctrl{-1} & \qw
-}\]
-with $k \in \{0,1\}$.
-When $k=0$ the circuit outputs the state $$\frac{1}{\sqrt{2}}\ket{00} + \frac{1}{2}\ket{01} + \frac{1}{2}\ket{11}$$ and when $k=1$ then circuit outputs the state $$\frac{1}{\sqrt{2}}\ket{10} + \frac{1}{2}\ket{01} - \frac{1}{2}\ket{11}$$
-as desired.
-\section{}
-\subsection{a}
-\paragraph{}
-We can write any density matrix $\rho$ as
-$$\frac{1}{2}\begin{bmatrix} 1+ r_z & r_x - i r_y \\ r_x + i r_y & 1-r_z \end{bmatrix}$$
-Computing $T\rho T^{-1}$ then yields
-$$\frac{1}{2}\begin{bmatrix} 1+r_z & \frac{r_x - r_y}{\sqrt{2}} -i\frac{r_x + r_y}{\sqrt{2}} \\ \frac{r_x - r_y}{\sqrt{2}} + i\frac{r_x + r_y}{\sqrt{2}} & 1-r_z\end{bmatrix}.$$
-Then the corresponding affine linear map on $r$ sends $(r_x, r_y, r_z)$ to
-$$ (\frac{r_x - r_y}{\sqrt{2}}, \frac{r_x+r_y}{\sqrt{2}}, r_z).$$
-\subsection{b}
-\paragraph{}
-Using the same notation for $\rho$ as the previous part, we compute 
-$$A_0\rho A_0^\dagger = \frac{1}{2}\begin{bmatrix}
-1+r_z & \sqrt{p}(r_x - i r_y) \\ \sqrt{r}(r_x + ir_y) & p(1-r_z)
+We have
+\begin{align*}
+\ket{\phi^+} &= \ket{00} + \ket{11}\\
+\ket{\phi^-} &= \ket{00} - \ket{11} \\
+\ket{\psi^+} &= \ket{10} + \ket{01} \\
+\ket{\psi^-} &= \ket{10} - \ket{01}
+\end{align*}
+So then
+$$\ket{\phi^+}\bra{\phi^+} + \ket{\phi^-}\bra{\phi^-} + \ket{\psi^+}\bra{\psi^-} = \begin{bmatrix}
+1 & 0 & 0 &0 \\
+0 & 1 & 1 & 0 \\
+0 & 1 & 1 & 0 \\
+0 &0 &0 &1
 \end{bmatrix}$$
-and $$A_1 \rho A_1^\dagger = \frac{1}{2}\begin{bmatrix} (1-p)(1-r_z) & 0 \\ 0 & 0 \end{bmatrix}.$$
-So, adding the above and simplifying, the operator on $\rho$ yields
-$$\begin{bmatrix}
-1+ pr_z + 1-p & \sqrt{p}(r_x - i r_y) \\ \sqrt{p}(r_x + i r_y) & 1 - (pr_z + 1-p)
-\end{bmatrix}.$$
-Then the corresponding affine linear map on $r$ sends $(r_x, r_y, r_z)$ to
-$$ (\sqrt{p}r_x, \sqrt{p}r_y, pr_z + 1-p).$$
-\subsection{c}
+and 
+$$\ket{\psi^-}\bra{\psi^-} = \begin{bmatrix}
+0 & 0 &0 & 0\\
+0 & 1 & -1 & 0 \\
+0 & -1 & 1 & 0 \\
+0 &0 &0 &0
+\end{bmatrix}$$
+Hence our Werner state $\rho$ as a $4\times 4$ matrix is
+$$\rho = \begin{bmatrix}
+\frac{1-p}{3} & 0 & 0 &0 \\
+0 & \frac{1+2p}{3} & \frac{1-4p}{3} & 0 \\
+0 & \frac{1-4p}{3} & \frac{1+2p}{3} & 0 \\
+0 & 0 & 0 & \frac{1-p}{3}
+\end{bmatrix}$$
+The partial transpose of $\rho$ is 
+$$(I \otimes T)\rho = \begin{bmatrix}
+\frac{1-p}{3} & 0 & 0 &\frac{1-4p}{3} \\
+0 & \frac{1+2p}{3} & 0 & 0 \\
+0 & 0 & \frac{1+2p}{3} & 0 \\
+\frac{1-4p}{3} & 0 & 0 & \frac{1-p}{3}
+\end{bmatrix}$$
 \end{document}

### 2017-11-30 CO681-QuantumInformationProcessing/assignments/assignment_05.tex
Problem 1

@@
 \paragraph{}
-We have
+We have (leaving off normalization)
 \begin{align*}
@@
 So then
-$$\ket{\phi^+}\bra{\phi^+} + \ket{\phi^-}\bra{\phi^-} + \ket{\psi^+}\bra{\psi^-} = \begin{bmatrix}
-1 & 0 & 0 &0 \\
+$$\ket{\phi^+}\bra{\phi^+} + \ket{\phi^-}\bra{\phi^-} + \ket{\psi^+}\bra{\psi^-} = \frac{1}{2}\begin{bmatrix}
+2 & 0 & 0 &0 \\
 0 & 1 & 1 & 0 \\
 0 & 1 & 1 & 0 \\
-0 &0 &0 &1
+0 &0 &0 &2
 \end{bmatrix}$$
 and 
-$$\ket{\psi^-}\bra{\psi^-} = \begin{bmatrix}
+$$\ket{\psi^-}\bra{\psi^-} =\frac{1}{2} \begin{bmatrix}
 0 & 0 &0 & 0\\
@@
 Hence our Werner state $\rho$ as a $4\times 4$ matrix is
-$$\rho = \begin{bmatrix}
-\frac{1-p}{3} & 0 & 0 &0 \\
-0 & \frac{1+2p}{3} & \frac{1-4p}{3} & 0 \\
-0 & \frac{1-4p}{3} & \frac{1+2p}{3} & 0 \\
-0 & 0 & 0 & \frac{1-p}{3}
+$$\rho = \frac{1}{6}\begin{bmatrix}
+2(1-p) & 0 & 0 &0 \\
+0 & 1+2p & 1-4p & 0 \\
+0 & 1-4p & 1+2p & 0 \\
+0 & 0 & 0 & 2(1-p)
 \end{bmatrix}$$
 The partial transpose of $\rho$ is 
-$$(I \otimes T)\rho = \begin{bmatrix}
-\frac{1-p}{3} & 0 & 0 &\frac{1-4p}{3} \\
-0 & \frac{1+2p}{3} & 0 & 0 \\
-0 & 0 & \frac{1+2p}{3} & 0 \\
-\frac{1-4p}{3} & 0 & 0 & \frac{1-p}{3}
+$$(I \otimes T)\rho = \frac{1}{6}\begin{bmatrix}
+2(1-p) & 0 & 0 &1-4p \\
+0 & 1+2p3 & 0 & 0 \\
+0 & 0 & 1+2p & 0 \\
+1-4p & 0 & 0 & 1-3p
 \end{bmatrix}$$
+\subsection{b}
+The eigenvalues $(I \otimes T)\rho$ are $\frac{1+2p}{6}$, $0$, $0$, and $\frac{3-6p}{6} = \frac{1-2p}{2}$.
+The least of which is $\frac{1-2p}{2}$.
+From our slides we know that $\rho$ is separable if and only if $(I \otimes T)\rho$ is positive semidefinite.
+This happens if and only if $\frac{1-2p}{2} \geq 0$, that is when $p \leq 1/2$.
 \end{document}

### 2017-12-04 CO681-QuantumInformationProcessing/assignments/assignment_05.tex
problem 2

@@
 \DeclarePairedDelimiter\norm{\lVert}{\rVert}%
+
+\DeclareMathOperator{\swap}{SWAP}
+
 %END MACROS
@@
 This happens if and only if $\frac{1-2p}{2} \geq 0$, that is when $p \leq 1/2$.
+
+\section{}
+\subsection{a}
+\paragraph{}
+Let $\ket{0}, \ket{1}, \dots, \ket{d-1}$ denote the computational basis of $\C^d$. We use $[d-1]$ as shorthand for $\{0,\dots, d-1\}$. Let $i, j \in [d-1]$. Then we have
+$$\swap(\ket{i}\ket{j} + \ket{j}\ket{i}) = \ket{j}\ket{i} + \ket{i}\ket{j} = \ket{i}\ket{j} + \ket{j}\ket{i}$$
+Hence $\ket{i}\ket{j} + \ket{j}\ket{i}$ is an eigenvector of $\swap$ with eigenvalue $1$. There are $d^2/2$ such vectors, and it is easy to see they are linearly independent. Let
+$$B_+ = \{\ket{i}\ket{j} + \ket{j}\ket{i} : i,j \in [d-1]\}$$
+Similarly we can observe that
+$$\swap(\ket{i}\ket{j} - \ket{j}\ket{i}) = \ket{j}\ket{i} - \ket{i}\ket{j} = -1(\ket{i}\ket{j} - \ket{j}\ket{i}).$$
+Hence $\ket{i}\ket{j} - \ket{j}\ket{i}$ is an eigenvector of $\swap$ with eigenvalue $-1$. Again there are $d^2/2$ such non-zero vectors. Let 
+$$B_- = \{\ket{i}\ket{j} - \ket{j}\ket{i} : i,j \in [d-1]\}\backslash \{0\}.$$
+One can easily verify that $B:= B_+ \cup B_-$ is a set of linearly independent vectors.
+Since $|B| = d^2$ this implies that $B$ forms a basis of $\C^d\otimes \C^d$.
+Thus we have shown that the eigenvalues of $\swap$ are exactly $\pm 1$ and more over if we define $P_+$ to be projector onto the span of $B_+$ and $P_-$ to be the projector onto the span of $B_-$ then
+$$\swap = P_+ - P_-.$$
+\subsection{b}
+\paragraph{}
+We can simplify our description of $P_+$ somewhat. Observe that
+$$P_+ = \frac{1}{2}\sum_{i=0}^{d-1} \sum_{j \leq i} (\ket{i}\ket{j} + \ket{j}\ket{i})(\bra{i}\bra{j} + \bra{j}\bra{i}) =\frac{1}{2}\big( \sum_{i,j=0}^{d-1} \ket{i}\ket{j}\bra{i}\bra{j} + \sum_{i,j=0}^{d-1} \ket{i}\ket{j}\bra{j}\bra{i}\big) = \frac{1}{2}\big(I + \swap\big).$$
+Let $\ket{\psi}, \ket{\phi} \in \C^d$ be pure states. Then
+\begin{align*}
+P_+ \ket{\psi}\ket{\phi} &= \frac{1}{2}(\ket{\psi}\ket{\phi} + \ket{\phi}\ket{\psi})
+\end{align*}
+Hence 
+\begin{align*}
+Tr(P_+ \ket{\psi}\ket{\phi}\bra{\psi}\bra{\phi}P_+) &= \bra{\psi}\bra{\phi}P_+P_+ \ket{\psi}\ket{\phi}\\ &= \frac{1}{2}\big(\ip{\phi}{\phi}\ip{\psi}{\psi} + \ip{\phi}{\psi}\ip{\psi}{\phi})\\ &= \frac{1 + |\ip{\phi}{\psi}|^2}{2}
+\end{align*}
+Therefore the probability that the swap test declares $\ket{\psi}$ and $\ket{\phi}$ the same is $\frac{1 + |\ip{\phi}{\psi}|^2}{2}$.
+\subsection{c}
+\paragraph{}
+We have that
+\begin{align*}
+\rho\otimes\sigma = &\frac{1}{4}(I \otimes I &+ r_x' I \otimes X &+ r_y' I \otimes Y &+ r_z' I \otimes Z \\
+&r_x X\otimes I &+ r_xr_x' X\otimes X &+ r_xr_y' X\otimes Y &+ r_xr_z' X \otimes Z \\
+&r_y Y \otimes I &+ r_yr_x'Y\otimes X &+ r_yr_y' Y \otimes Y &+ r_yr_z' Y \otimes Z \\
+&r_z Z\otimes I &+ r_zr_x' Z\otimes X &+ r_zr_y' Z \otimes Y &+ r_zr_z' Z\otimes Z).
+\end{align*}
+We check the trace of the action of $P_+$ on the constituent terms of $\rho \otimes \sigma$.
+By symmetry of $\rho \otimes \sigma$, and the fact that $-iY = XZ$ it will suffice to check only $I\otimes I$, $X\otimes X$, $Z\otimes Z$, $I \otimes X$, $I\otimes Z$, and $X\otimes Z$.
+Using cyclic nature of trace and that $\swap$ is self-inverse, it suffices to understand happens when $\swap$ is applied to the above matrices.
+$\swap X\otimes X$ has non-zero diagonal entries $1$ only in second and third columns.
+Similarly $\swap Z\otimes Z$ has non-zero diagonal entries $1$ only in first and fourth columns.
+Further $\swap I \otimes X$ and $\swap X\otimes Z$ have no non-zero diagonal entries, and $\swap I \otimes Z$ has non-zero diagonal entries $1$ and $-1$.
+\paragraph{}
+Finally, using the above discussion, we calculate traces. First we have $Tr(P_+I\otimes I P_+) = Tr(P_+P_+) = \frac{12}{4} = 3$. Further
+\begin{align*}
+Tr(P_+ X\otimes X P_+) &= \frac{1}{4}Tr(X\otimes X + \swap X\otimes X + X\otimes X \swap + \swap X\otimes X \swap )\\
+&= \frac{1}{4}(0 + 2 + 2 + 0)\\
+&= 1.
+\end{align*}
+and similarly
+$$Tr(P_+ Z \otimes Z P_+) = \frac{1}{4}(0 + 2 + 2 + 0) = 1.$$
+Lastly,
+$$Tr(P_+ I \otimes X P_+) = Tr(P_+ I \otimes Z P_+) = Tr(P_+ X\otimes Z P_+) = 0.$$
+Therefore we have that the probability the swap test declare $\rho$ and $\sigma$ equal is
+$$Tr(P_+ \rho \otimes \sigma P_+) = \frac{3 + r_xr_x' + r_yr_y' + r_zr_z'}{4}.$$
+\section{}
+\paragraph{}
 \end{document}

### 2017-12-04 CO681-QuantumInformationProcessing/assignments/assignment_05.tex
finished a5

@@
 This happens if and only if $\frac{1-2p}{2} \geq 0$, that is when $p \leq 1/2$.
-
+\newpage
 \section{}
@@
 $$Tr(P_+ \rho \otimes \sigma P_+) = \frac{3 + r_xr_x' + r_yr_y' + r_zr_z'}{4}.$$
+\newpage
 \section{}
 \paragraph{}
+We use throughout that $\ket{A}$ and $\ket{B}$ are orthogonal. We have that
+$$-HU_0HU_f = (2\ket{\theta}\bra{\theta} - I_2)(2\ket{B}\bra{B} - I_2) = 4\ip{\theta}{B}\ket{\theta}\bra{B} - 2\ket{B}\bra{B} - 2\ket{\theta}\bra{\theta} + I_2.$$
+Now we observe that
+$\ip{\theta}{B} = \cos\theta$ and that
+$$\ket{\theta}\bra{B} = \sin\theta\ket{A}\bra{B} + \cos\theta\ket{B}\bra{B}.$$
+Further that
+$$\ket{\theta}\bra{\theta} = \sin^2\theta\ket{A}\bra{A} + \sin\theta\cos\theta (\ket{A}\bra{B} + \ket{B}\bra{A}) + \cos^2\theta\ket{B}\bra{B}.$$
+Then we see that
+\begin{align*}
+-HU_0HU_f &= 4\sin\theta\cos\theta\ket{A}\bra{B} + 4\cos^2\theta\ket{B}\bra{B} -2 \ket{B}\bra{B} \\
+&-2\sin^2\theta\ket{A}\bra{A} -2\sin\theta\cos\theta(\ket{A}\bra{B} + \ket{B}\bra{A}) -2\cos^2\theta\ket{B}\bra{B} \\
+&+ \ket{A}\bra{A} + \ket{B}\bra{B}\\
+&= (1-2\sin^2\theta)\ket{A}\bra{A} +2\sin\theta\cos\theta\ket{A}\bra{B} - 2\sin\theta\cos\theta\ket{B}\bra{A} + (2\cos^2\theta-1)\ket{B}\bra{B}\\
+&= \cos2\theta \ket{A}\bra{A} + \sin2\theta\ket{A}\bra{B} -\sin2\theta\ket{B}\bra{A} +\cos2\theta\ket{B}\bra{B}
+\end{align*}
+as desired.
+\newpage
+\section{}
+\paragraph{}
+First observe that
+$$Tr(\rho) = \sum_j a_j Tr(E_j) = \frac{1}{d} \sum_j a_j$$
+and then that
+$$p(i) = Tr(\rho E_i) = ba_i + c \sum_{j\neq i} a_j = ba_i + c(dTr(\rho) - a_i) = ba_i + cd - ca_i.$$
+Solving for $a_i$ we see that
+$a_i = \frac{p(i) -cd}{b-c}.$
 \end{document}

### 2017-12-07 CO681-QuantumInformationProcessing/project/introduction.tex
Project start

\section{Introduction}
\paragraph{}
In this paper we explore the ways in which quantum computing can be used to improve upon the best-known classical algorithms for a wealth of fundamental problems in combinatorial optimization. In Section \ref{sec:quantum-primitives} we will discuss the various quantum procedures which appear at the heart of the otherwise classical algorithms we present. Chief among these procedures is Grover's Search \cite{grover1996fast}. In the sections which follow we discuss quantum algorithms for matchings, network flows, shortest paths, and minimum spanning trees.
\subsection{Two Models of Graphs}

### 2017-12-07 CO681-QuantumInformationProcessing/project/matchings.tex
Project start

\section{Matchings}
\paragraph{}
In a {\it maximum matching problem} we are given a graph $G=(V,E)$ and asked for find  $M\subseteq E$, maximizing $|M|$, such that every vertex of $(V,M)$ has degree at most $1$. \subsection{Bipartite Matching}
\paragraph{}
A {\it bipartite graph} is a graph where there exists a partition $A\cup B = V$ such that no vertices in $A$ are adjacent, and no vertices in $B$ are adjacent. A well known equivalent definition of a bipartite graph is one with no odd cycles. In bipartite graphs the best known classical algorithm takes time $O(n^\omega)$ \cite{mucha2004maximum} where $2\leq\omega\leq 2.4$ is the matrix multiplication constant and $n = |V|$, with the caveat that if $\omega = 2$ then algorithm a $\log n$ factor appears in the runtime.  We model our quantum algorithm on the following $O(n^{5/2})$-time classical algorithm \cite{hopkroft1973n5}:
\begin{enumerate}
\item
\end{enumerate}

\subsection{General Matching}

\subsection{Quantum Adversary}

### 2017-12-07 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
Project start

\section{Quantum Primitives}\label{sec:quantum-primitives}

\subsection{Grover's Search}

\subsection{Layered Subgraphs}

\subsection{Minima Finding}

### 2017-12-08 CO681-QuantumInformationProcessing/project/introduction.tex
Grover Search and reorg

@@
 \subsection{Two Models of Graphs}
+\paragraph{}
+The problems we will consider in later sections are problems on (un)directed graphs $G=(V,E)$.
+There are two common ways of modeling access to a graph's structure in our algorithms: the {\it adjacency model} and the {\it list model}, each with subtle different complexity consequences.
+\paragraph{}
+In the adjacency model $G$ is specified by its $V \times V$ adjacency matrix $A(G)$, abbreviated $A$ when the graph $G$ is clear from context. For any $i, j \in V$, we have
+$$A_{i,j} = \begin{cases}
+1, \text{if $(i,j) \in E$}\\
+0, \text{otherwise.}
+\end{cases}$$
+Notice that the diagonal of $A$ is all zeroes since we assume $G$ has no self-loops, and $A$ is symmetric when $G$ is undirected.
+\paragraph{}
+In the list model $G$ is specified by $|V|$ lists.
+Each list corresponds to a vertex $v \in V$ and the elements of the list are the neighbours of $v$.
+Notice how in this model adding vertices can be done in constant time, while querying adjacency is costly, but the reverse is true for the adjacency model.
+\paragraph{}
+Generally the list model performs better on sparse graphs, and the adjacency model performs better on dense graphs (those with $|E|$ approximately $|V|^2$).
+For a complete discussion on the tradeoffs between the two models see Cormen et al.
+\cite[pages 529-530]{cormen2009introduction}.

### 2017-12-08 CO681-QuantumInformationProcessing/project/network-flows.tex
Grover Search and reorg

\section{Network Flows}

\subsection{Layered Subgraphs}

\subsection{Integer Network Flow Algorithm}

\subsection{Lower Bound}

### 2017-12-08 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
Grover Search and reorg

@@
 \section{Quantum Primitives}\label{sec:quantum-primitives}
-
+\paragraph{}
+In this section we present important quantum primitives to be used in our optimization algorithms in later sections.
+The centerpiece of these algorithms is Grover's search, first presented in \cite{grover1996fast} and rigorously analyzed in \cite{boyer1996tight}, whose presentation we will follow.
 \subsection{Grover's Search}
-
-\subsection{Layered Subgraphs}
-
+\paragraph{}
+Let $T$ be  a table with with $N$ elements, where the $i$-th element is denoted $T[i]$.
+In a general {\it search problem} we are given $T$ and some $x$ in the universe from which the elements of $T$ are drawn, and are to find some $i \in \{0, \dots, N-1\}$ such that $T[i] = x$.
+Let $t= |\{ i \in \{0,\dots, N-1\} : T[i] = x\}$.
+Classically, it is easy to see that when $T$ is unsorted any algorithm which solves this problem with success probability at least $1/2$ will need to make at least $N/2t$ queries to $T$.
+In contrast, Grover's Quantum Search algorithm will need only $O(\sqrt{N/t})$ queries to have $1/2$ success probability.
+This is within a constant factor of optimal for a quantum algorithm
+\paragraph{}
+Let $A = \{ i : T[i] = x\}$ and let $B = \{ i : T[i] \neq x \}$. Let $k, \ell \in \R$ satisfying $tk^2 + (N-t)\ell^2 = 1$.  We define the following general quantum register state
+$$\ket{\Phi(k,\ell)} = \sum_{i \in A} k \ket{i} + \sum{i \in B} \ell \ket{i}.$$
+Grover's Search begins with the state $$\ket{\Phi_0} := \sum_i \frac{1}{\sqrt{N}} \ket{i}$$
+and each iteration applies a transformation sending some $\ket{\Phi(k,\ell)}$ to
+$$\ket{\Phi(\frac{N-2t}{N}k + \frac{2(N-t)}{N}\ell, \frac{N-2t}{N}\ell - \frac{2t}{N} k)}.$$
+The transformation is specified as follows (we assume $N$ is a power of $2$ for simplicity, full details appear in the source material).
+Define for any $K \subseteq \{0, \dots, N-1\}$ the ``conditional phase shift transform" satisfying
+$$S_K\ket{i} = \begin{cases}
+-\ket{i}, &\text{if $i \in K$} \\
+\ket{i}, &\text{otherwise}.
+\end{cases}$$
+Let $H$ be the Hadamard transform satisfying
+$$H\ket{j} = \frac{1}{\sqrt{N}} \sum_i (-1)^{i\cdot j} \ket{i}.$$
+Then we observe that the transformation
+$$-HS_{\{0\}}HS_A$$
+efficiently implements a Grover Search iteration using only $1$ query to $T$.
+\paragraph{}
+The search begins by fixing some $\lambda \in (1,4/3)$ and initializing $m=1$. Then repeating the following procedure:
+\begin{enumerate}
+\item Choose $j$ uniformly at random from $\{1, \dots, m\}$.
+\item Apply $j$ Grover iterations, as described above, starting from $\ket{\Phi_0}$.
+\item Measure and let $i$ be the outcome.
+\item If $T[i] = x$ return $i$; otherwise set $m$ to $\min(\lambda m, \sqrt{N})$.
+\end{enumerate}
+\begin{theorem}\label{th:grovers}
+Grover's Search Algorithm above finds a solution in expected time $O(\sqrt{N/t})$. (Alternatively, the algorithm has small failure probability if terminated after $\Omega(\sqrt{N/t})$ iterations)
+\end{theorem}
+\begin{proof}
+The case where $t=0$ can be treated with an appropriate timeout, with arbitrarily small failure probability.
+The case $t > \frac{3}{4}N$ can be dismissed via classical sampling techniques in constant expected time.
+So we may assume $1\leq t\leq \frac{3}{4}N$.
+\paragraph{}
+Let choose $\theta$ so that $\sin^2\theta = \frac{t}{N}$. Since $t\leq \frac{3}{4}N$, if we let $m_0 = \frac{1}{sin(2\theta)}$ then
+$$m_0 = \frac{N}{2\sqrt{(N-t)t}} < \sqrt{\frac{N}{t}}.$$
+Let $c = \ceil{\log_\lambda m_0}$.
+Observe that the algorithm sets $m =\lambda^{s-1}$ in iteration $s \leq \frac{3}{4}N$ of the main loop, and since $j$ is chosen randomly, the algorithm uses at most $\frac{1}{2} \lambda^{s-1}$ Grover iterations in expectation at iteration $s$.
+\paragraph{}
+To reach iteration $c$ the algorithm would need
+$$\frac{1}{2}\sum_{s=1}^{\ceil{\log_\lambda m_0}} \lambda^{s-1} <\frac{1}{2}\frac{\lambda}{\lambda - 1} m_0 = 3m_0$$
+Grover iterations in expectation.
+Since $m_0 = O(\sqrt{\frac{N}{t}})$ if the algorithm terminates before iteration $c$ then we have the claimed running time bound.
+Notice that clearly the number of Grover iterations dominates the complexity of the running time.
+\paragraph{}
+Thus we may assume that algorithm uses at least $c$ iterations.
+The probability of success for $j$ Grover iterations from $\ket{\Phi_0}$ is $tk_j^2= \sin^2((2j+1)\theta)$.
+Since $j$ is chosen randomly the average success probability of step $2$ is
+$$\sum_{j=0}^{m-1}\frac{1}{m}\sin^2((2j+1)\theta) = \frac{1}{2m}\sum_{j=0}^{m-1}1 - \cos((2j+1)\theta) = \frac{1}{2} - \frac{\sin(4m\theta)}{4m\sin(2\theta)}$$
+recalling (learning?) that the last equality follows from a trigonometric identity.
+Since $m \geq \frac{1}{\sin(2\theta)}$ after the algorithm passes $c$ iterations, the above probability is at least $\frac{1}{4}$.
+Therefore the expected number of Grover iterations after the critical stage is
+$$ \frac{1}{2}\sum_{s=0}^\infty \frac{3^s}{4^{s+1}}\lambda^{s+c} < \frac{\lambda}{8-6\lambda}m_0 = \frac{3}{2} m_0. $$
+So the algorithm uses at most $3m_0 + \frac{3}{2}m_0 = O(\sqrt{\frac{N}{t}})$ Grover iterations in expectation.
+\end{proof}
 \subsection{Minima Finding}

### 2017-12-08 CO681-QuantumInformationProcessing/project/trees.tex
Grover Search and reorg

\section{Trees}

\subsection{Minimum Spanning Trees}

\subsection{Shortest Path Trees}

\subsection{Lower Bounds}

### 2017-12-09 CO681-QuantumInformationProcessing/project/network-flows.tex
finishing layering

@@
-\section{Network Flows}
+\section{Network Flows}\label{sec:nf}
 
-\subsection{Layered Subgraphs}
-
+\subsection{Directed Graph Layering}
+\paragraph{}
+An important subproblem we will use in solving Network Flow problems, and again later in solving Matching problems is the computing of a {\it layering}.
+Given a connected directed graph $G = (V,E)$ and a source vertex $r \in V$ we are to find a labelling, which we call a layering, on the vertices, $\ell : V \rightarrow \N$ such that $\ell(r) = 0$ and for all $v\neq r$
+$$ \ell(v) = 1+ \min_{u:(u,v) \in E} \ell(u).$$
+\paragraph{Algorithm}
+We can compute a layering as follows:
+\begin{enumerate}
+\item Initialize $\ell(r) = 0$ and $\ell(v) = \infty$ for all $v \neq r$. Further initialize a queue $W = \{r\}$.
+\item While $W\neq \emptyset$
+	\begin{enumerate}
+	\item Pop $v$ off the head of the queue $W$
+	\item Find by Grover's Search all neighbours $w$ with $\ell(w) = \infty$.
+	\item Set $\ell(w) = \ell(v) + 1$ and append $w$ to $W$.
+	\end{enumerate} 
+\end{enumerate}
+\begin{theorem}\label{th:layering}
+The algorithm above computes a layering for $G=(V,E)$  in time $O(n^{3/2}\log n)$ in the adjacency model and in time $O(\sqrt{nm}\log n)$ in the list model, where $n = |V|$ and $m = |E|$.
+\end{theorem}
+\begin{proof}
+Clearly initialization takes time $O(n)$, and every vertex is processed at most once.
+In the adjacency model the processing of every vertex takes $O(\sqrt{n})$ by Theorem \ref{th:grovers}.
+The log-factor appears in accordance with Note \ref{note:log}.
+\paragraph{}
+In the list model, each vertex is processed in time $O(\sqrt{\frac{n_v}{d(v)}}d_v + \sqrt{d(v) + 1}) = O(\sqrt{n_vd(v)} + \sqrt{d_v + 1})$, where $d(v)$ is the out-degree of $v$, and $n_v$ is the number of neighbours of $v$ with $\ell(v)=\infty$ when $v$ is searched from.
+Using the Cauchy-Schwarz inequality we can bound the expected running time for all vertices by
+$$\sum_{v\in V} \sqrt{n_vd(v)} \leq \sqrt{\sum_{v\in V} n_v}\sqrt{\sum_{v \in V} d(v)} = O(\sqrt{nm}),$$
+and similarly for $\sum_{v \in V} \sqrt{d(v) + 1}$. Hence the claimed running time.
+\end{proof}
 \subsection{Integer Network Flow Algorithm}

### 2017-12-09 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
finishing layering

@@
 \end{proof}
+\begin{note}\label{note:log}
+Many of our algorithms in the sections to follow.
+Particularly in Section \ref{sec:nf} and Section \ref{sec:matching} will use a polynomial number of applications of Grover's search, say $n^d$.
+If we want all of our quantum subroutines to simultaneously output a correct answer with probability at least $1-\frac{1}{n}$, then we need the probability of failure for a given subroutine to be at most $\frac{1}{n^{d+1}}$.
+Repeating each subroutine a $\log n$ number of times will ensure this probability.
+Consequentially, $\log $ factors will appear in our runtimes.
+We will omit them from the analysis, and cite this note as explanation of their appearance.
+\end{note}
 \subsection{Minima Finding}

### 2017-12-10 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
starting minima finding

@@
 \subsection{Minima Finding}
+\paragraph{}
+Consider the following {\it minima finding } problem, wherein we try to find $d$ smallest values of different types.
+We are give two functions: an objective $f: \{1, \dots, N\} \rightarrow \N \cup \{\infty\}$, and a type $g: \{1, \dots, N\} \rightarrow \N$.
+Let $|g| := | \{ g(i): i \in \{1,\dots, N\}\}|$.
+We are also given an integer $d \in \{, \dots, \min\{|g|, N\}\}$, and are to find $I \subseteq \{1, \dots, N\}$ such that:
+\begin{itemize}
+\item $|I| = d$,
+\item $g(i) \neq g(j)$ for all $i,j \in I$ with $i\neq j$,
+\item and for all $j \in \{1,\dots, N\}\backslash I$ and $i \in I$, if $f(j) < f(i)$ then there exists some $i' \in I$ such that $g(i') = g(j)$ and $f(i') \leq f(j)$.
+\end{itemize}

### 2017-12-11 CO681-QuantumInformationProcessing/project/network-flows.tex
minima finding

@@
 \paragraph{}
-In the list model, each vertex is processed in time $O(\sqrt{\frac{n_v}{d(v)}}d_v + \sqrt{d(v) + 1}) = O(\sqrt{n_vd(v)} + \sqrt{d_v + 1})$, where $d(v)$ is the out-degree of $v$, and $n_v$ is the number of neighbours of $v$ with $\ell(v)=\infty$ when $v$ is searched from.
+In the list model, each vertex is processed in time $$O(\sqrt{\frac{n_v}{d(v)}}d_v + \sqrt{d(v) + 1}) = O(\sqrt{n_vd(v)} + \sqrt{d_v + 1})$$ where $d(v)$ is the out-degree of $v$, and $n_v$ is the number of neighbours of $v$ with $\ell(v)=\infty$ when $v$ is searched from.
 Using the Cauchy-Schwarz inequality we can bound the expected running time for all vertices by

### 2017-12-11 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
minima finding

@@
 \end{itemize}
+\paragraph{}
+We present a notion of the sort of indices we want to add to $I$, split into two subtypes.
+For any $j \in \{1, \dots, N\}$, we call $j$ {\it good for} $I$ if there exists some $i \in I$ such that $f(j) < f(i)$ and either
+\begin{itemize}
+\item {\it(Known Type)} $g(j) = g(i)$, or 
+\item {\it(Unknown Type)} $g(j) \not\in g(I)$.
+\end{itemize}
+We present the following algorithm for minima finding:
+\begin{enumerate}
+\item Initialize $I$ as a set of $d$ dummy indices of unique type with respect to $g$ and unique maximal value with respect to $f$
+\item Repeat until no good elements for $I$
+	\begin{enumerate}
+	\item Use Grover's Search to find a good element $j$ for $I$.
+	\item If $j$ is of Known type, replace $i \in I$ such that $f(j) < f(i)$ and $g(j) = g(i)$ with $j$
+	\item Otherwise (i.e. if $j$ is of Unknown type) let $i \in I$ be element of largest $f$-value. Replace $i$ with $j$.
+	\end{enumerate}
+\end{enumerate}
+\begin{lemma}\label{lemma:minima}Let $I \subseteq \{1,\dots, N\}$ such that $|I| = d$ and for all $i,j \in I$, $g(i) \neq g(j)$ with $t>0$ good elements for $I$.
+Then after $O(d)$ iterations of step $2$, in expectation, there are at most $\frac{3}{4}t$ good elements for $I$.
+\end{lemma}
+\begin{proof}
+See Lemma $4$ of D{\"u}rr et al. \cite{durr2004quantum}.
+\end{proof}
+\begin{theorem}\label{th:minima}
+The algorithm above solves the minima finding problem in expected time $O(\sqrt{dN})$.
+\end{theorem}
+\begin{proof}
+Let $t$ denote the number of good elements remaining for $I$ at a given iteration of the algorithm.
+From Lemma \ref{lemma:minima} and Theorem \ref{th:grovers}, we have that, if $t > 2d$ then there are at most $\frac{t}{2}$ good elements for $I$ after $O(d\sqrt{\frac{N}{t}})$ queries of $f$ in expectation.
+Hence for $t>2d$ the expected number of queries of $f$ until $t\leq 2d$ is
+$$O\big(d\big(\sqrt{\frac{N}{d}} + \sqrt{\frac{N}{2d}} + \sqrt{\frac{N}{4d}} + \dots\big)\big) = O(\sqrt{dN}).$$
+Then once the algorithm has $t \leq 2d$, the number of queries of $f$ needed in expectation to reach $t=0$ is
+$$O\big(\sum_{j=1}^{2d}\sqrt{\frac{N}{j}}\big) = O(\sqrt{dN})$$
+as desired.
+\end{proof}
+\paragraph{}
+Two special case of minima finding are the problems of Finding $d$ Smallest Values of a Function, and Find $d$ Elements of Different Type.
+By a reduction from distinguishing matrices to these special subproblems, they show in \cite{durr2004quantum} that Minima Finding requires $\Omega(\sqrt{dN})$ queries of $f$ and hence the algorithm we present is optimal.

### 2017-12-11 CO681-QuantumInformationProcessing/project/network-flows.tex
writing up on network flow.

--- 2017-12-08 CO681-QuantumInformationProcessing/project/network-flows.tex
@@
-\section{Network Flows}
+\section{Network Flows}\label{sec:nf}
 
-\subsection{Layered Subgraphs}
-
+\subsection{Directed Graph Layering}
+\paragraph{}
+An important subproblem we will use in solving Network Flow problems, and again later in solving Matching problems is the computing of a {\it layering}.
+Given a connected directed graph $G = (V,E)$ and a source vertex $r \in V$ we are to find a labelling, which we call a layering, on the vertices, $\ell : V \rightarrow \N$ such that $\ell(r) = 0$ and for all $v\neq r$
+$$ \ell(v) = 1+ \min_{u:(u,v) \in E} \ell(u).$$
+\paragraph{Algorithm}
+We can compute a layering as follows:
+\begin{enumerate}
+\item Initialize $\ell(r) = 0$ and $\ell(v) = \infty$ for all $v \neq r$. Further initialize a queue $W = \{r\}$.
+\item While $W\neq \emptyset$
+	\begin{enumerate}
+	\item Pop $v$ off the head of the queue $W$
+	\item Find by Grover's Search all neighbours $w$ with $\ell(w) = \infty$.
+	\item Set $\ell(w) = \ell(v) + 1$ and append $w$ to $W$.
+	\end{enumerate} 
+\end{enumerate}
+\begin{theorem}\label{th:layering}
+The algorithm above computes a layering for $G=(V,E)$  in time $O(n^{3/2}\log n)$ in the adjacency model and in time $O(\sqrt{nm}\log n)$ in the list model, where $n = |V|$ and $m = |E|$.
+\end{theorem}
+\begin{proof}
+Clearly initialization takes time $O(n)$, and every vertex is processed at most once.
+In the adjacency model the processing of every vertex takes $O(\sqrt{n})$ by Theorem \ref{th:grovers}.
+The log-factor appears in accordance with Note \ref{note:log}.
+\paragraph{}
+In the list model, each vertex is processed in time $$O(\sqrt{\frac{n_v}{d(v)}}d_v + \sqrt{d(v) + 1}) = O(\sqrt{n_vd(v)} + \sqrt{d_v + 1})$$ where $d(v)$ is the out-degree of $v$, and $n_v$ is the number of neighbours of $v$ with $\ell(v)=\infty$ when $v$ is searched from.
+Using the Cauchy-Schwarz inequality we can bound the expected running time for all vertices by
+$$\sum_{v\in V} \sqrt{n_vd(v)} \leq \sqrt{\sum_{v\in V} n_v}\sqrt{\sum_{v \in V} d(v)} = O(\sqrt{nm}),$$
+and similarly for $\sum_{v \in V} \sqrt{d(v) + 1}$. Hence the claimed running time.
+\end{proof}
 \subsection{Integer Network Flow Algorithm}
-
+\paragraph{}
+In the {\it Integer Network Flow} problem we are given a directed network $G=(V,E)$ with special vertices source $r$ and sink $s$, and arc capacities $\mu : E \rightarrow \Z_{\geq 0}$ bounded by $U$, i.e.
+for each $e \in E$, $\mu(e) \leq U$.
+Our goal is to find the a {\it flow} $x: E \rightarrow \Z_{\geq 0}$ from $r$ to $s$ satisfying capacities $\mu$ of largest value, i.e.
+we want to solve:
+\begin{align*}
+\max \sum_{(u,s): E} x(u,s) \\
+\text{s.t.} \sum_{(u,v) \in E} x(u,v) - \sum_{(v,u) \in E} x(v,u) &= 0 &\forall v \in V\backslash\{r,s\} \\
+0 \leq &x\leq \mu.
+\end{align*}
+The current best known classical algorithm, given in \cite{goldberg1998beyond}, runs in time $O(\min(n^{2/3}m,m^{3/2})\log U)$.
+\paragraph{}
+Given a network $G$ and a flow $x$ on $G$, we define the {\it residual} network $H$ as follows.
+The vertex set of $H$ is the same as that of $G$.
+Add ``forward" arc $(u,v)$ to $H$ if $ x(u,v) < \mu(u,v)$ and set the capacity of $(u,v)$ in $H$ to be $\mu_H(u,v) = \mu(u,v) - x(u,v)$.
+Also add (possibly parallel) ``backward" arc $(u,v)$ to $H$ if $x(v,u) > 0$, and set $\mu_H(u,v) = x(v,u)$.
+Omit any edges with $0$ capacity.
+\paragraph{}
+A {\it flow-augmenting path} $P$ on $G$ with respect to $x$ is an $r$-$s$ path in residual network $H$.
+It is called such because $x$ can increased by the minimum capacity on $P$ on each forward arc in $P$ and decreased by the minimum capacity on $P$ on each backward arc to obtain a flow of greater value on $G$.
+A {\it blocking flow} is a set of flow-augmenting paths on $H$ that hits (has an arc intersecting) every $r$-$s$ path in $H$.
+\paragraph{}
+In \cite{even1975network} Even and Tarjan present a network flow algorithm when $U=1$ wherein the idea of augmenting the current flow by blocking flows in layered residual networks until the layer depth reaches a certain bound, and then switching to augmenting by single augmenting paths is employed.
+We will also use this technique in our quantum algorithm.
+Even and Tarjan give the following lemma which we will employ in our analysis.
+\begin{lemma}\label{lemma:5}
+\cite{even1975network}\cite{ambainis2006quantum} Let $H$ be the residual network of $G$ with respect to flow $x$.
+Let $k$ be the depth of an optimal layering on $H$.
+The the value of the blocking flow is at most $\min((2n/k)^2,m,k)\cdot U$.
+\end{lemma}
+\begin{proof}
+See Lemma $5$ of \cite{ambainis2006quantum}.
+\end{proof}
+\paragraph{Algorithm}
+Let $k = \min(n^{2/3}U^{1/3}\sqrt{mU})$. Our quantum algorithm starting with flow $x=0$:
+\begin{enumerate}
+\item Compute a layering $\ell$ of the residual graph $H$.
+\item If depth of $\ell$ is at most $k$ compute a blocking flow as follows:
+	\begin{enumerate}
+	\item Mark all vertices as ``enabled".
+	\item Compute by depth-first Grover's search an $r$-$s$ path $p$ in $H$ using layering $\ell$ that only uses enabled vertices (quit if no such path).
+During the back-tracking phase of the search, mark all vertices with no path to $s$ as ``disabled".
+	\item Let $u_H$ be the minimum of $\mu_H$ along $p$. Augment $x$ on $G$ by $u_H$ along $p$.
+	\item Return to Step $2a$.
+	\end{enumerate}
+\item Otherwise compute by depth-first Grover's search an $r$-$s$ path $p$ with min capacity $u_H$ (quit if no such path). Augment $x$ on $G$ by $u_H$ along $p$.
+\item Return to Step $1$.
+\end{enumerate}
+\paragraph{}
+We present our analysis of the algorithm in the case $U\leq n^{1/4}$ and will show after how to give upper bounds when $U > n^{1/4}$.
+\begin{theorem}
+Suppose $U \leq n^{1/4}$.
+Our algorithm solves Integer Network Flow in expected time $$O(n^{13/6}\cdot U^{1/3}\log n)$$ in the adjacency model and expected time $$O(\min(n^{7/6}\sqrt{m}\cdot U^{1/3}, \sqrt{nU}m)\log n)$$ in the list model.
+\end{theorem}
+\begin{proof}
+We can compute $\ell$ in $O(\sqrt{nm}\log n)$ time using Theorem \ref{th:layering}.
+We compute the capacities of $H$ online in constant time as needed.
+Since the edges of $H$ are searched online using the layering $\ell$ our algorithm automatically ``deletes" saturated edges when their capacity is queried by Grover's search.
+\paragraph{Processing a vertex of the depth-first search}
+Let $v \in V$.
+Let $a_v$ be the number of augmenting paths using $v$ and let $e_{v,i}$ be the number of outgoing edges $(v,\cdot)$ when $i$ augmenting paths remain with respect to $H$.
+Since every arc has capacity at most $U$, we have $e_{v,i} \geq \ceil{i/U}$.
+Thus the total time spent in expectation on Grover's searches at node $v$ of the depth-first search is
+$$\sum_{i=1}^{a_v} \sqrt{\frac{d(v)}{e_{v,i}}} \leq \sqrt{U}\cdot\sum_{i=1}^{a_v}\sqrt{d(v)/i} = O(\sqrt{Ua_vd(v)}).$$
+Let $c_v$ be the number of enabled vertices found from node $v$ of the depth-first search which do not lie on an augmenting path in $H$, and are hence disabled.
+In expectation $O(\sqrt{c_vd(v)})$ time is spent in Grover's searches for these vertices.
+Finally, it takes $O(\sqrt{d(v) + 1})$ time to discover there is no $r$-$s$ path in $H$ using $v$, at which point $v$ is marked ``disabled" and never returned to.
+\paragraph{Time to find a blocking flow}
+Let $j$ be the depth of layering $\ell$ on $H$ and let $A_j$ be the value of a blocking flow on $H$.
+Since the total number of augmenting paths through any vertex in a given layer is at most $A_j$, we have $\sum_{v \in V} a_v \leq jA_j$.
+Further we can observe that $\sum_{v \in V} c_v \leq n$ and $\sum_{v \in V} d(v) \leq m$.
+Thus, using Cauchy-Schwarz, the total time in expectation for computing a blocking flow from $H$ is
+$$\sum_{v \in V}(\sqrt{Ua_v(d)} + \sqrt{c_vd(v)} + \sqrt{d(v)+1}) \leq \sqrt{U}\sqrt{\sum_{v\in V}a_v}\sqrt{\sum_{v\in V} d(v)} + 2\sqrt{nm}$$
+which is on the order of $O(\sqrt{jmA_jU} + \sqrt{nm})$.
+\paragraph{}
+So our algorithm performs at most $k$ iterations finding a blocking flow. Thus the total expected time before switching to finding single flow-augmenting paths is at most
+$$\sqrt{mU}\cdot\sum_{j=1}^k\sqrt{jA_j} + k\sqrt{nm}.$$
+\paragraph{Counting remaining flow-augmenting paths}
+If the algorithm has not yet terminated after $\ell$ reaches depth greater than $k$ then by Lemma \ref{lemma:5}, $H$ has a residual flow of size $O(\min((2n/k)^2,m,k)\cdot U)=O(k)$.
+Therefore the algorithm needs at most $O(k)$ more iterations to terminate.
+Since we now only search for one augmenting path, each iteration has complexity $O(\sqrt{jm}) = O(\sqrt{nm})$.
+Hence the total running time of our algorithm in expectation is
+$$O(\sqrt{mU}\cdot \sum_{j=1}^k\sqrt{jA_j} + k\sqrt{nm}) + O(k\sqrt{nm}).$$
+\paragraph{Final Bounding}
+We want to show $\sum_{j=1}^k\sqrt{jA_j} = O(k^{3/2})$. 
+\end{proof}
 \subsection{Lower Bound}

### 2017-12-12 CO681-QuantumInformationProcessing/project/matchings.tex
finishing layering
bipartite matching

@@
-\section{Matchings}
+\section{Matchings}\label{sec:matching}
 \paragraph{}
@@
 In bipartite graphs the best known classical algorithm takes time $O(n^\omega)$ \cite{mucha2004maximum} where $2\leq\omega\leq 2.4$ is the matrix multiplication constant and $n = |V|$, with the caveat that if $\omega = 2$ then algorithm a $\log n$ factor appears in the runtime.
-We model our quantum algorithm on the following $O(n^{5/2})$-time classical algorithm \cite{hopkroft1973n5}:
+Our quantum algorithm is inspired by the $O(n^{5/2})$-time classical algorithm in \cite{hopkroft1973n5}.
+\paragraph{}
+As in our study of Network Flows we will identify a notion of augmenting paths we can use to improve our matchings, and a residual graph structure which enables easy identification of such paths.
+For a graph $G$ and a matching $M$, we say a path $P$ in $G$ is {\it $M$-augmenting} if the endpoints of $P$ are not matched in $M$ and the edges of $P$ are alternately not in $M$ and in $M$.
+If we have such $P$ we can flip the matched edges in $M$ along $P$ (formally we replace $M$ with the symmetric difference of $M$ and $P$) to obtain a matching of larger size than $M$.
+\paragraph{}
+In bipartite graphs $M$-augmenting paths can easily be found using a {\it residual graph} structure.
+Construct a directed graph $H$ based on $G$ and $M$ as follows.
+The vertex set of $H$ is $A\cup B\cup \{r,s\}$ with source $r$ and sink $s$.
+For each edge $ab \in E$ with $a \in A$ and $b\in B$,  add arc $(a,b)$ to $H$ if $ab \not\in M$, otherwise add arc $(b,a)$.
+For each $a \in A$ that is not matched by $M$ add arc $(r,a)$ to $H$.
+Similarly for each $b \in B$ unmatched by $M$ add arc $(b,s)$.
+Now it is not hard to see that $r$-$s$ dipaths in $H$ correspond to $M$-augmenting paths in $G$.
+\paragraph{Algorithm}
+Our quantum algorithm for bipartite maximum matching is as follows:
 \begin{enumerate}
-\item
+\item Set $M = \emptyset$
+\item  Let $H$ be the residual graph with respect to $G$ and $M$. Compute a layering $\ell$ for $H$.
+\item Find by depth-first Grovers Search using $\ell$ a maximal set $S$ of vertex-disjoint $r$-$s$ paths in $H$.
+\item Augment $M$ by all paths in $S$. Note such paths all have the same minimal length.
+\item If $S\neq \emptyset$ Return to Step $2$. Otherwise output $M$ and quit.
 \end{enumerate} 
-
+\begin{theorem}
+Our quantum algorithm finds a maximum bipartite matching in expected time $O(n\sqrt{m+n}\log n)$ in the list model and $O(n^2\log n)$ in the adjacency model.
+\end{theorem}
+\begin{proof}
+First we bound the number of iterations of Step $2$ in the algorithm.
+Let $s$ be the size of a maximal matching $M$ in $G$ and let $s_i$ be the size of matching found by the algorithm after iteration $i$.
+Let $j$ be the first iteration with $s_j \geq s-\sqrt{n}$.
+Since the algorithm finds at least one augmenting path each time it runs from Step $2$, there are at most $\sqrt{n}$ iterations remaining after iteration $j$.
+Now in \cite{hopkroft1973n5} they give the following Lemma
+$$\text{If $M_1$ and $M_2$ are matchings of size $s_1$ and $s_2$ respectively ($s_1 < s_2$) then there exist $s_2-s_1$ vertex disjoint $M_1$-augmenting paths.}$$
+Applying this Lemma with $M_1 = M_j$ and $M_2 = M$ we see there are at least $\sqrt{n}$ vertex disjoint $M_j$-augmenting paths.
+Since the length of $M_i$-augmenting paths increase by one each iteration, the length of these $M_j$-augmenting paths is at least $j+2$.
+So therefore we have $j <\sqrt{n}$, and hence the total number of iterations of Step $2$ is at most $2\sqrt{n} = O(\sqrt{n})$.
+\paragraph{Bounding Time to Find all Augmenting Paths}
+We will show the bound for the adjacency model.
+The bounds for the list model follow by similar arguments, using as well the bounding techniques of the proof of Theorem \ref{th:layering}.
+The layer numbers for $H$ can be computed in time $O(n^{3/2}\log n)$ in the adjacency model by Theorem \ref{th:layering}.
+Our depth-first search proceeds from a vertex by searching all vertices with layer number $\ell$ one greater to find a descendant.
+Each vertex will thus join an augmenting path and be marked, or it does not belong to any such path.
+Each vertex is thus visited at most once and hence costs expected $O(\sqrt{n})$ time to be found and another $O(\sqrt{n})$ time to decide it has no remaining descendants.
+Therefore the desired runtime bound holds.
+\end{proof}
 \subsection{General Matching}

### 2017-12-12 CO681-QuantumInformationProcessing/project/network-flows.tex
bipartite matching

@@
 Let $k$ be the depth of an optimal layering on $H$.
-The the value of the blocking flow is at most $\min((2n/k)^2,m,k)\cdot U$.
+Then the value of the residual flow is at most $\min((2n/k)^2,m,k)\cdot U$.
 \end{lemma}
@@
 We want to show $\sum_{j=1}^k\sqrt{jA_j} = O(k^{3/2})$.
+We will divide the index sequence $1, \dots, k$ of summands into $\log k$ intervals of the form $S_i := \{2^i, 2^i + 1, \dots, 2^{i+1}-1\}$ of length $2^i$.
+After $k_i=\frac{k}{2^i}$ iterations the value of the residual flow on $H$ is at most
+$$O(\min((n/k)^2\cdot 2^{2i}, m/k\cdot 2^i)\cdot U) \leq O(2^{2i}k) = O(k^2/k_i)$$
+by Lemma \ref{lemma:5}. Using Cauchy-Schwartz on each interval $S_i$ we observe that
+\begin{align*}
+\sum_{j=1}^k \sqrt{jA_j} &= \sum_{i=0}^{\log k} \sum_{j \in S_i} \sqrt{jA_j}\\
+ &\leq \sum_{i=0}^{\log k} \sqrt{2^i\cdot 2^{i+1}}\sqrt{\sum_{j\in S_i} A_j} \\
+ &\leq \sqrt{2}\sum_{i=0}^{\log k} 2^i\sqrt{kk_i} \\
+ &= O(k^{3/2}).
+ \end{align*}
+ Hence the total expected running time of our algorithm is $O(k\sqrt{m}(\sqrt{kU} + \sqrt{n}))$. Since $U\leq n^{1/4}$, we have $kU \leq n$. Therefore the running time in the list model is
+ $$O(\min(n^{7/6}\sqrt{m}\cdot U^{1/3}, \sqrt{nU})\log n)$$
+ with the log factor appearing by Note \ref{note:log}. In the adjacency model the desired runtime appears by setting $m=n^2$ in the above.
 \end{proof}
-\subsection{Lower Bound}
+\paragraph{}
+When $U > n^{1/4}$ we can obtain a runtime bound of $O(\min(n^{7/6}\sqrt{m},\sqrt{n}m)\cdot U \log n)$ using the same analysis as above, but with a slightly modified algorithm: that is, needing to change $k$ to $\min(n^{2/3}, \sqrt{m})$.
+The speedup of our quantum network flow algorithm is best felt for dense graphs, i.e.
+those with many edges.
+When $m = \Omega(n^{1+\epsilon})$ for some $\epsilon >0$ and $U$ is small our algorithm gives a polynomial speedup over the best known classical algorithm.
+\paragraph{}
+The authors also present a lower bound via a reduction from deciding the majority on a bit sequence and using the max-flow min-cut theorem.
+In some sense the gap between the upper and lower bounds remains large.
+We state without proof their result here.
+\begin{theorem}
+\cite{ambainis2006quantum} A quantum algorithm for Integer Network Flow with capacities bounded by $U=n$ has quantum query complexity $\Omega(n^2)$.
+\end{theorem}

### 2017-12-12 CO681-QuantumInformationProcessing/project/matchings.tex
matching

@@
 Now in \cite{hopkroft1973n5} they give the following Lemma
-$$\text{If $M_1$ and $M_2$ are matchings of size $s_1$ and $s_2$ respectively ($s_1 < s_2$) then there exist $s_2-s_1$ vertex disjoint $M_1$-augmenting paths.}$$
+\begin{align*}&\text{If $M_1$ and $M_2$ are matchings of size $s_1$ and $s_2$ respectively ($s_1 < s_2$)}\\&\text{then there exist $s_2-s_1$ vertex disjoint $M_1$-augmenting paths.}\end{align*}
 Applying this Lemma with $M_1 = M_j$ and $M_2 = M$ we see there are at least $\sqrt{n}$ vertex disjoint $M_j$-augmenting paths.
@@
 \end{proof}
+\bigskip
+\paragraph{Lower Bounds}
+In \cite{boyer1996tight} and \cite{zhang2005power} they demonstrate the existence of a quantum adversary for maximum bipartite matching implying a $\Omega(n^{3/2})$ quantum query complexity lower bound for the problem.
+Since this is a special case of both general matching (to follow), and the network flow problem studied in the previous Section it provides a lower bound for those problems as well.
 \subsection{General Matching}
-
-\subsection{Quantum Adversary}
+\paragraph{}
+Our general quantum matching algorithm follows the classic Blossom algorithm of Edmonds \cite{edmonds1965paths}.
+An excellent reference for this algorithm is Chapter $5$ of the textbook Combinatorial Optimization by Cook, Cunningham, Pulleyblank, and Schrijver \cite{cook2009combinatorial}.
+In broad strokes the algorithm starts with an empty matching $M$ and in each iteration identifies either an augmenting path, an odd circuit with one $M$-exposed vertex, or decides the matching is maximum and terminates.
+In the first case $M$ is augmented by the path found, and in the second the odd circuit is contracted to a single vertex and the algorithm proceeds on the resulting subgraph.
+\paragraph{}
+It can be shown that classically the Blossom algorithm can be implemented in time $O(nm\log n)$.
+The quantum algorithm presented in \cite{ambainis2006quantum} follows the same framework as the Blossom algorithm but speeds up all necessary searching by using Grovers Search.
+In every iteration of Blossom we have an exposed vertex $r$ and build a tree of vertices by breadth-first search which lie on alternating paths from $r$.
+Each vertex that lies on an even length path from $r$ in the tree is of interest.
+If an even vertex $v$ has an exposed neighbour then we have found an augmenting path.
+If it has an $M$-covered neighbour then we can either extend the tree or identify an odd circuit.
+These situations lead to three instances of Grover's Search at $v$ taking total expected time $O(\sqrt{nm})$.
+This discussion leads to the following theorem, a detail proof of which is found in the source material.
+\begin{theorem}
+A maximum matching can be found with a quantum computer in expected time $O(n^2(\sqrt{m/n} + \log n))$ in the list model and $O(n^{5/2}\log n)$ in the adjacency model.
+\end{theorem}

### 2017-12-13 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
writing up on network flow.
Complete draft of project

@@
 \paragraph{}
-We present a notion of the sort of indices we want to add to $I$, split into two subtypes.
+We define a notion of the sort of indices we want to add to $I$, split into two subtypes.
 For any $j \in \{1, \dots, N\}$, we call $j$ {\it good for} $I$ if there exists some $i \in I$ such that $f(j) < f(i)$ and either
@@
 \begin{theorem}\label{th:minima}
-The algorithm above solves the minima finding problem in expected time $O(\sqrt{dN})$.
+The algorithm above solves the minima finding problem using $O(\sqrt{dN})$ queries in expectation.
 \end{theorem}
@@
 \end{proof}
-\paragraph{}
+\begin{note}\label{note:lb}
 Two special case of minima finding are the problems of Finding $d$ Smallest Values of a Function, and Find $d$ Elements of Different Type.
 By a reduction from distinguishing matrices to these special subproblems, they show in \cite{durr2004quantum} that Minima Finding requires $\Omega(\sqrt{dN})$ queries of $f$ and hence the algorithm we present is optimal.
+\end{note}

### 2017-12-13 CO681-QuantumInformationProcessing/project/trees.tex
Complete draft of project

@@
 \subsection{Minimum Spanning Trees}
-
+\paragraph{}
+In the {\it Minimum Spanning Tree} problem we are given a graph $G=(V,E)$ and weights $w: E \rightarrow R_{\geq 0}$.
+Our goal is to find a spanning tree $(V,T)$ so that $$w(T) := \sum_{e \in T} w(e)$$ is minimized.
+While there are a wide array of algorithms for solving minimum spanning tree, we will concentrate our efforts on Boruvka's algorithm \cite{nevsetvril2001otakar} as it lends itself well to minima finding algorithm given in Section $2$.
+\paragraph{Algorithm}
+Boruvka's algorithm maintains a spanning forest at each iteration and adds to each component a minimum weight edge out of the component, merging trees in the forest as necessary.
+It is a classical exercise to verify the correctness of this algorithm.
+Our slight variation of Borubvka's algorithm, taking advantage of quantum searching, is as follows:
+\begin{enumerate}
+\item Let $T_1, T_2, \dots, T_k$ denote our spanning forest. Initialize $k= n$ and each $T_i$ to a single unique vertex $v_i$.
+\item While $k >1$ do:
+	\begin{enumerate}
+	\item For each $j \in \{1, \dots, k\}$ let $e_j = \arg\min\{w(uv): uv \in E, u \in V(T_j), v \not\in V(T_j)\}$ via Minima Finding Algorithm (we will make this precise). 
+	\item Add edges $e_j$ to $T_j$ for each $j$, merging trees as appropriate and setting $k$ to the number of connected components.
+	\end{enumerate}
+\item Return $T_1$.
+\end{enumerate}
+In the list model each edge $uv$ appears twice: once in the list for $u$ and again in the list for $v$.
+Thus we can think of $E$ as a set of directed edges such that $(u,v) \in E$ iff $(v,u) \in E$.
+To apply Minima Finding (Theorem \ref{th:minima}) in Step $2(a)$ we use objective function $f: E \rightarrow \R_\geq 0$ that does for each directed edge $(u,v)$
+$$f(u,v) = \begin{cases}
+w(u,v), &\text{ if $(u,v) \not\in T_j, \forall j\in \{1,\dots, k\}$} \\
+\infty, &\text{otherwise}.
+\end{cases}$$
+and we use type function $g: E\rightarrow \Z$ mapping each directed edge $(u,v)$ to the index, $j$, of the tree $T_j$ such that $u \in V(T_j)$.
+\begin{theorem}
+Our algorithm for Minimum Spanning Trees uses $O(\sqrt{nm}\log n)$ expected quantum queries in the list model, and $O(n^{3/2}\log n)$ in the adjacency model.
+\end{theorem}
+\begin{proof}
+\paragraph{}
+Each iteration sees $k$ decreased by a factor of $\frac{1}{2}$.
+Therefore there are $\log n$ iterations of Step $2$.
+Now applying Theorem \ref{th:minima} we see that each iteration needs $O(\sqrt{nm})$ expected queries to solve minima finding.
+\paragraph{}
+The number of queries in the adjacency model follows from the number of queries in the list model by viewing the adjacency matrix as a list model with $m = n^2$.
+\end{proof}
 \subsection{Shortest Path Trees}
-
-\subsection{Lower Bounds}
+\paragraph{}
+In the {\it shortest path} problem, you are given a directed graph $G=(V,E)$ with non-negative edges weights $w: E\rightarrow \R_{\geq 0}$, and a root vertex $r$.
+You are to compute a {\it shortest path tree} $T$ rooted at $r$.
+That is, for every $v \in V$, the path from $r$ to $v$ is a shortest path in $G$.
+\paragraph{Definitions}
+For vertices $u, v \in V$ we use $d(u,v)$ to denote the shortest path distance from $u$ to $v$.
+Further we denote border edges as $\delta(U) = \{(u,v ) \in E: u \in U, v\not\in U\}$ for $U \subseteq V$.
+The correctness of our algorithm to follow hinges on maintaining the invariant that every vertex in the tree discovered so far has a shortest path from $r$ using only tree arcs, and that at each iteration the edge $(u,v)$ added to the tree from the border edges is of cheapest cost $c_T(u,v) : = d(r,u) + w(u,v)$.
+\paragraph{Algorithm}
+Our algorithm is as follows:
+\begin{enumerate}
+\item Set $T \leftarrow \{r\}$, $\ell \leftarrow 1$, $P_1 \leftarrow \{v_0\}$.
+\item While $T$ does not span $V$:
+	\begin{enumerate}
+	\item Find a set $A_\ell$ of $|P_\ell|$ cheapest, with respect to $c_T$, border edges in $\delta(V(T))$ with disjoint target head vertices.
+We will make this precise using Minima Finding after the algorithm description.
+	\item Let $(u,v)$ be the minimal costed edge of $A_1 \cup \dots \cup A_\ell$ with $v\not\in P_1 \cup \dots \cup P_\ell$.
+Set $T \leftarrow T \cup \{(u,v)\}$, and $P_{\ell + 1} \leftarrow \{v\}$ and $\ell \leftarrow \ell + 1$.
+	\item If $\ell \geq 2$ and $|P_{\ell -1}| = |P_{\ell}|$ then merge $P_\ell$ into $P_{\ell -1}$ and set $\ell = \ell -1$.
+	\end{enumerate}
+\item Return $T$.
+\end{enumerate}
+\paragraph{}
+In Step $2(a)$ we find $A_\ell$ via Minima Finding (Theorem \ref{th:minima}). To this end define objective function $f:\delta(P_\ell) \rightarrow \Z$ so that 
+$$f(u,v) = \begin{cases}
+c_T(u,v), &\text{if } v\not\in T \\
+\infty, &\text{otherwise}
+\end{cases}$$
+and define the type function $g: \delta(P_\ell) \rightarrow V$ by $g(u,v) = v$.
+It is easy to see that solving this Minima Finding problem computes $A_\ell$.
+As in the Minimum Spanning Tree problem, the adjacency model complexity is easy to derive from the list model complexity, and so we only give the list model bound in the following theorem.
+\begin{theorem}
+Our algorithm solves shortest path with expected $O(\sqrt{nm}\log^{3/2}n)$ quantum queries in the list model.
+\end{theorem}
+\begin{proof}
+Consider iteration $\ell$ with tree $T$ found so far.
+The sets $P_1, P_2, \dots P_{\ell}$ partition $V(T)$, and their sizes are strictly decreasing powers of $2$.
+Since $\sum_{j=0}^{i-1} 2^j = 2^i - 1$, each $P_i$ is strictly larger that $\bigcup_{j=0}^{i-1} P_j$.
+Therefore, since $A_i$ contains  $|P_i|$ edges, at least one such edge has its target head vertex outside $P_1, \dots, P_\ell$ by pigeonhole principle.
+\paragraph{}
+Let $(u,v)$ be the cheapest, with respect to $c_T$ border edge of $T$ during this $\ell$-th iteration.
+Let $P_i$ be the set containing $u$.
+Then $A_i$ necessarily contains $(u,v)$ as so $(u,v)$ is selected to add to $T$ in Step $2(b)$ as desired.
+\paragraph{Counting Queries}
+We count first the total queries for sets $P_i$ of some cardinality $s$.
+Suppose in Step $2(a)$ we making queries for some $P_\ell$ of size $s$.
+How many such $P_\ell$ of size $s$ can appear over the course of the algorithm?
+Since $V(T)$ is always a strict superset of the vertex set of the tree from the previous iteration, there are at most $n/s$ such $P_\ell$ over the entire operation of the algorithm.
+If we let $m_j = |\delta(P_j)|$ for any $j$ such that $|P_j| = s$ then the total queries done, in expectation, by Step $2(a)$ for cardinality $s$ sets is
+$$\sum_{j=1}^{n/s} \sqrt{sm_j}$$
+by Theorem \ref{th:minima}.
+In the worst case each $m_j = sm/n$ and then $\sum_j = m_j = m$. Therefore the total queries for size $s$ sets is $O(\sqrt{nm})$ for fixed $s$.
+\paragraph{}
+The algorithm uses $\log n$ distinct such $s$ and hence the total queries made is $O(\sqrt{n\log n m})$.
+As in Note \ref{note:log} we introduce a $\log n$ factor to account for error.
+Therefore we obtain the desired bound.
+\end{proof}
+\subsection{Lower Bound}
+\paragraph{}
+We conclude with a lower bound on both Minimum Spanning Tree and Shortest Path quantum queries.
+\begin{theorem}
+Minimum Spanning Tree and Shortest Path both require $\Omega(\sqrt{nm})$ queries.
+\end{theorem}
+\begin{proof}
+We proceed by a reduction from minima finding and then apply the $\Omega(\sqrt{dN})$ bound as mentioned in Note \ref{note:lb}, with $d = n$ and $N= kn$ for some fixed $k$, and $m = k(n+1)$.
+Let $M$ be an $n\times k$ matrix with positive entries.
+From the lower bound on Minima Finding, $\Omega(\sqrt{kn^2})$ quantum queries are required to find the minimum value in each row.
+\paragraph{}
+We construct a weighted graph $G=(V,E)$ from $M$.
+Let $V = \{r, v_1, \dots, v_k, u_1, \dots, u_n\}$.
+For each $v_i$ add arc $(r,v_i)$ to $E$ with weight $0$.
+For each $v_i, u_j$ add arc $(v_i, u_j)$ to $E$ with weight $M_{ji}$.
+Then every minimum spanning tree, and shortest path tree from $r$, will use all the $0$ weight edges from $r$ and connect every $u_j$ with only its minimal weighted edge.
+\end{proof}

### 2018-01-17 CO681-QuantumInformationProcessing/project/network-flows.tex
Wrote up problem 1

@@
 \paragraph{}
-In the list model, each vertex is processed in time $$O(\sqrt{\frac{n_v}{d(v)}}d_v + \sqrt{d(v) + 1}) = O(\sqrt{n_vd(v)} + \sqrt{d_v + 1})$$ where $d(v)$ is the out-degree of $v$, and $n_v$ is the number of neighbours of $v$ with $\ell(v)=\infty$ when $v$ is searched from.
+In the list model, each vertex is processed in time $$O\big(\sqrt{\frac{n_v}{d(v)}}d_v + \sqrt{d(v) + 1}\big) = O(\sqrt{n_vd(v)} + \sqrt{d_v + 1})$$ where $d(v)$ is the out-degree of $v$, and $n_v$ is the number of neighbours of $v$ with $\ell(v)=\infty$ when $v$ is searched from.
 Using the Cauchy-Schwarz inequality we can bound the expected running time for all vertices by
 $$\sum_{v\in V} \sqrt{n_vd(v)} \leq \sqrt{\sum_{v\in V} n_v}\sqrt{\sum_{v \in V} d(v)} = O(\sqrt{nm}),$$
-and similarly for $\sum_{v \in V} \sqrt{d(v) + 1}$. Hence the claimed running time.
+and similarly for $\sum_{v \in V} \sqrt{d(v) + 1}$. Hence we have the claimed running time.
 \end{proof}

### 2018-01-17 CO681-QuantumInformationProcessing/project/quantum-primitives.tex
Wrote up problem 1

@@
 In a general {\it search problem} we are given $T$ and some $x$ in the universe from which the elements of $T$ are drawn, and are to find some $i \in \{0, \dots, N-1\}$ such that $T[i] = x$.
-Let $t= |\{ i \in \{0,\dots, N-1\} : T[i] = x\}$.
-Classically, it is easy to see that when $T$ is unsorted any algorithm which solves this problem with success probability at least $1/2$ will need to make at least $N/2t$ queries to $T$.
+Let $$t= |\{ i \in \{0,\dots, N-1\} : T[i] = x\}|.$$ Classically, it is easy to see that when $T$ is unsorted any algorithm which solves this problem with success probability at least $1/2$ will need to make at least $\frac{N}{2t}$ queries to $T$.
 In contrast, Grover's Quantum Search algorithm will need only $O(\sqrt{N/t})$ queries to have $1/2$ success probability.
-This is within a constant factor of optimal for a quantum algorithm
+This is query number is within a constant factor of optimal for a quantum search algorithm
 \paragraph{}
-Let $A = \{ i : T[i] = x\}$ and let $B = \{ i : T[i] \neq x \}$. Let $k, \ell \in \R$ satisfying $tk^2 + (N-t)\ell^2 = 1$.  We define the following general quantum register state
+Let $A = \{ i : T[i] = x\}$ and let $B = \{ i : T[i] \neq x \}$. Let $k, \ell \in \R$ satisfying $$tk^2 + (N-t)\ell^2 = 1.$$ We define the following general quantum register state
 $$\ket{\Phi(k,\ell)} = \sum_{i \in A} k \ket{i} + \sum{i \in B} \ell \ket{i}.$$
@@
 \paragraph{}
-Let choose $\theta$ so that $\sin^2\theta = \frac{t}{N}$. Since $t\leq \frac{3}{4}N$, if we let $m_0 = \frac{1}{sin(2\theta)}$ then
+Let us choose $\theta$ so that $\sin^2\theta = \frac{t}{N}$. Since $t\leq \frac{3}{4}N$, if we let $m_0 = \frac{1}{sin(2\theta)}$ then
 $$m_0 = \frac{N}{2\sqrt{(N-t)t}} < \sqrt{\frac{N}{t}}.$$
@@
 Since $m \geq \frac{1}{\sin(2\theta)}$ after the algorithm passes $c$ iterations, the above probability is at least $\frac{1}{4}$.
-Therefore the expected number of Grover iterations after the critical stage is
+Therefore the expected number of Grover iterations after $c$ iterations is
 $$ \frac{1}{2}\sum_{s=0}^\infty \frac{3^s}{4^{s+1}}\lambda^{s+c} < \frac{\lambda}{8-6\lambda}m_0 = \frac{3}{2} m_0. $$
@@
 \begin{note}\label{note:log}
-Many of our algorithms in the sections to follow.
-Particularly in Section \ref{sec:nf} and Section \ref{sec:matching} will use a polynomial number of applications of Grover's search, say $n^d$.
+Many of our algorithms in the sections to follow will use a polynomial number of applications of Grover's search, say $n^d$.
 If we want all of our quantum subroutines to simultaneously output a correct answer with probability at least $1-\frac{1}{n}$, then we need the probability of failure for a given subroutine to be at most $\frac{1}{n^{d+1}}$.
@@
 Let $|g| := | \{ g(i): i \in \{1,\dots, N\}\}|$.
-We are also given an integer $d \in \{, \dots, \min\{|g|, N\}\}$, and are to find $I \subseteq \{1, \dots, N\}$ such that:
+We are also given an integer $d \in \{1, \dots, \min\{|g|, N\}\}$, and are to find $I \subseteq \{1, \dots, N\}$ such that:
 \begin{itemize}
@@
 \begin{note}\label{note:lb}
-Two special case of minima finding are the problems of Finding $d$ Smallest Values of a Function, and Find $d$ Elements of Different Type.
+Two special cases of minima finding are the problems of Finding $d$ Smallest Values of a Function, and Find $d$ Elements of Different Type.
 By a reduction from distinguishing matrices to these special subproblems, they show in \cite{durr2004quantum} that Minima Finding requires $\Omega(\sqrt{dN})$ queries of $f$ and hence the algorithm we present is optimal.

### 2018-01-17 CO749-GraphColoring/assignments/A1.tex
Wrote up problem 1

--- 2017-10-16 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
@@
 \usepackage[braket, qm]{qcircuit}
+\usepackage{hyperref}
 
@@
 
-\lhead{\today}
-\rhead{W. Justin Toth CO681-Quantum Information Processing A2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{CO 749 - \today}
+\rhead{W. Justin Toth - A1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \section{}
-\subsection{a}
 \paragraph{}
-$$11+12 \mod 13 = 10$$
-\subsection{b}
+Let $G$ be a graph of minimum degree at least $3$ satisfying all of the following:
+\begin{itemize}
+	\item Every triangle contains at most one vertex of degree three.
+	\item If two vertices of degree three are adjacent then they are in triangles.
+	\item No two distinct triangles share a common edge.
+\end{itemize}
+Suppose for a contradiction that $\abs{E(G)} < 5\abs{V(G)}/3$.
+\begin{claim}\label{1.1}
+	For every $v \in V(G)$, if $d(v) = 3$ then $v$ has at most one neighbour of degree $3$.
+\end{claim}
+\begin{proof}
+	Assume for a contradiction that there exist $u,v, w \in V(G)$ satisfying
+	$$d(u) = d(v) = d(w) = 3$$
+	and $uv, vw \in E(G)$, i.e.
+$u$ and $w$ are adjacent to $v$.
+But then, using the second property of $G$, $v$ is in a triangle.
+Denote this triangle by $C$.
+Since every triangle contains at most one vertex of degree three, $u$ and $w$ are not contained in $C$.
+But then since $d(v) = 3$, $v$ has at most one neighbour in $C$, a contradiction to $C$ being a $K_3$.
+\end{proof}
 \paragraph{}
-$$\Z^\times_{13} = \{2^0=1, 2^1 = 2, 2^2 = 4, 2^3 = 8, 2^4 = 3, 2^5 = 6, 2^6=12, 2^7=11, 2^8 = 9, 2^9=5, 2^{10}=10, 2^{11} = 7\}$$
-\subsection{c}
+We proceed by a discharging argument. For each vertex $v \in V(G)$ set the initial charge as
+$$ch_0(v) = d(v) - 4$$
+and for each edge $e \in E(G)$ set the initial charge
+$$ch_0(e) = 1/3.$$
+Then our total charge is, using handshaking,
+\begin{align*}
+\sum_{v\in V(G)} ch_0(v) + \sum_{e \in E(G)} ch_0(e) &= \sum_{v \in V(G)} d(v) - 4|V(G)| + \frac{1}{3} |E(G)|\\ &=  \frac{7}{3}|E(G)| - 4|V(G)\\ &< \frac{35}{9}|V(G)| - \frac{36}{9}|V(G)|\\ &< 0
+\end{align*}
+with the first inequality following from our assumption that $|E(G)| < 5|V(G)|/3$.
 \paragraph{}
-$2^8\equiv 9 \mod 13$ so $\log_2(9)$ is $8$ in $\Z^\times_{13}$.
-\subsection{d}
+Redistribute the charge, to obtain $ch_1$, according to the following Rules:
+\begin{enumerate}
+	\item If edge $uv$ contains precisely one vertex $u$ such that $d(u) = 3$ then $uv$ sends $1/3$ charge to $u$.
+	\item If edge $uv$ is such that $d(u) \geq 4$ and $d(v) \geq 4$ and $u,v$ are in a triangle with vertex $w$ then $uv$ sends $1/3$ charge to $w$.
+\end{enumerate}
 \paragraph{}
-\begin{itemize}
-\item $3^3 \equiv 1 \mod 13$ so $3$ is not a generator
-\item $4^6 \equiv 1 \mod 13$ so $4$ is not a generator
-\item $5^4 \equiv 1 \mod 13$ so $5$ is not a generator
-\item The order of $6$ is $12$ in $\Z_{13}^\times$ so is $6$ is a generator of $Z_{13}^\times$
-\item The order of $7$ is $12$ in $\Z_{13}^\times$ so is $7$ is a generator of $Z_{13}^\times$
-\item $8^4 \equiv 1 \mod 13$ so $8$ is not a generator
-\item $9^3 \equiv 1 \mod 13$ so $9$ is not a generator
-\item $10^6 \equiv 1 \mod 13$ so $10$ is not a generator
-\item The order of $11$ is $12$ in $\Z_{13}^\times$ so is $11$ is a generator of $Z_{13}^\times$
-\item $12 \equiv 1 \mod 13$ so $12$ is not a generator
-\end{itemize}
-Hence the generators of $\Z_{13}^\times$ besides $2$ are: $6,7,$ and $11$.
-
-\section{}
-\subsection{a}
+Since vertices sends no charge, clearly every vertex $v \in V(G)$ with $d(v) \geq 4$ has $$ch_1(v) = ch_0(v) \geq 0.$$ Now we will show that if $v$ is a vertex with $d(v) = 3$ then $ch_1(v) \geq 0$.
+From Claim \ref{1.1} $v$ is adjacent with at most one vertex of degree $3$.
+If $v$ is adjacent with no degree three vertices then each edge incident with $v$ sends $\frac{1}{3}$ charge to $v$ by Rule $1$ and hence
+$$ch_1(v) = ch_0(v) + 3\cdot\frac{1}{3} = 0.$$
+On the other hand, if $v$ is adjacent with one degree three vertex then $v$ receives $\frac{2}{3}$ charge from incident edges by Rule $1$.
+The remaining needed $\frac{1}{3}$ is obtained from Rule $2$ as follows.
+Since $v$ is adjacent with a degree three vertices, $v$ lies in a triangle, and since every triangle contains at most one degree three vertex the edge containing the other two vertices of this triangle sends $\frac{1}{3}$ charge to $v$.
+Therefore $ch_1(v) \geq 0$.
 \paragraph{}
-We check the action of $H$ on $X$ and $Z$.
-$$HXH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix}1 & -1 \\ 1 & 1 \end{bmatrix} =\frac{1}{2} \begin{bmatrix} 2 & 0 \\ 0 & 2 \end{bmatrix} = I.
-$$
-$$HZH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ -1 & 1\end{bmatrix} = \frac{1}{2} \begin{bmatrix} 0 & 2 \\ 2 & 0\end{bmatrix} = X.$$
-Thus $H$ is in the Clifford group on $1$ qubit.
-\subsection{b}
+It remains to verify that each edge $uv$ satisfies $ch_1(uv) \geq 0$.
+Edge $uv$ can satisfy at most one of Rule $1$ and Rule $2$.
+If $uv$ satisfies Rule $1$ it sends $\frac{1}{3}$ charge and so $ch_1(uv) = 0$.
+Instead if $uv$ satisfies Rule $2$ it can only do so for one triangle, since no two distinct triangles share a common edge.
+So $uv$ only sends $\frac{1}{3}$ charge by Rule $2$.
+Therefore $ch_1(uv) \geq 0$.
 \paragraph{}
-Similar to the previous subsection we consider the action of $T$ on $X$ and $Z$.
-$$TXT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & \zeta^*_8 \\ 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & \zeta_8^* \\ \zeta_8 & 0 \end{bmatrix}.$$
-Such a matrix cannot be written as  $\alpha V$ for phase $\alpha$ and $V \in \{I,X,Y,Z\}$. Thus $T$ is not in Clifford group on $1$ qubit.
-\subsection{c}
-\paragraph{}
-$$SXS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0\\ 0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & -i\\ 1 & 0 \end{bmatrix} =\begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix} = Y.$$
-$$SZS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} \begin{bmatrix} 1 & 0 \\0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} = Z.$$
-Thus $S$ is in the Clifford group on $1$ qubit.
-\subsection{d}
-\paragraph{}
-Let $C$ be the unitary matrix representing the controlled-not gate. We consider the action of $C$ on $I \otimes X$, $X\otimes I$, $I \otimes Z$. and $Z \otimes I$.
-$$C (I\otimes X) C^\dagger = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0 \\ 0 & 0 &0 & 1 \\ 0 & 0 & 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & 1& 0& 0\\ 1 & 0  & 0& 0\\ 0 & 0 & 1 & 0\\ 0 & 0 & 0 & 1\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = I \otimes X.$$
-$$ C (X \otimes I) C^\dagger =  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}  0 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \\ 1 & 0 & 0 & 0 \\ 0 &1 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}  =  \begin{bmatrix} 0 & 0 &1 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = X \otimes X$$
-\begin{align*}C (I \otimes Z) C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}1 &0 &0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 &1 & 0\\ 0& 0& 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
- &=  \begin{bmatrix} 1 & 0 & 0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 & 0 & -1 \\ 0 &0 &1 & 0\end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\ &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 & -1 &0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 &1\end{bmatrix} = Z \otimes Z.\end{align*}
- \begin{align*}
- C(Z\otimes I)C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}1 &0 &0 &0 \\0 & 1 & 0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 & -1 \end{bmatrix}  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
- &= \begin{bmatrix} 1& 0 & 0 &0 \\0 &1 & 0 &0 \\ 0 & 0 &0 &-1 \\ 0 & 0 & -1 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
- &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 &1 & 0 &0\\ 0 & 0 &-1 & 0 \\ 0 & 0 & 0 & -1 \end{bmatrix} = Z\otimes I.
- \end{align*}
- Thus $C$ is in the $2$-qubit Clifford group.
- \subsection{e}
- Let $A$ be the Toffoli gate (CCNOT). For the sake of space we'll abbreviate the calculations. We have that
- $$A(I\otimes X\otimes I) A^\dagger = \begin{bmatrix} X\otimes I & 0 \\ 0 & Z\otimes Z\end{bmatrix}.$$
- This can't be written as a tensor product $\alpha V_1\otimes V_2 \otimes V_3$ with $\alpha$ scalar and $V_1,V_2, V_3 \in \{I,X,Y,Z\}$.
-To see this observe that if we take $V_1 \in  \{X, Y\}$ then the first $4x4$ diagonal block of $V_1\otimes V_2 \otimes V_3$ will be $0$, contradicting that it should be $X \otimes I$, and if we take $V_1 \in \{I,Z\}$ then the first $4x4$ diagonal block will be $\pm1$ times the second, but $X\otimes I \neq \pm Z\otimes Z$.
-Therefore no such tensor product expression for $A(I\otimes X \otimes I)A^\dagger$ is possible, and hence the Toffoli gate is not in the $3$-qubit Clifford group.
-$\blacksquare$
- \section{}
- \subsection{a}
- \paragraph{}
-We use the same notation for gates as in question $2$. $X_4$ is given by the circuit
- \[ \Qcircuit @C=1em @R=1em {
- & \targ &   \qw & \qw\\
- & \ctrl{-1} & \gate{X} & \qw
-}\]
-$Z_4$ is given by the circuit
-\[ \Qcircuit @C=1em @R=1em {
- & \gate{S} &  \gate{S} & \qw\\
- & \gate{S} & \qw & \qw
-}\].
-\subsection{b}
-\paragraph{}
-$X_8$ is given by the circuit (using one Toffoli gate)
-\[ \Qcircuit @C=1em @R=1em {
- & \targ &  \qw & \qw & \qw & \qw\\
- & \ctrl{-1} & \targ& \qw  & \qw &\qw\\
- & \ctrl{-1} & \ctrl{-1} & \gate{X} &\qw & \qw
-}\]
-Bonus: Figure $4.9$ of Nielsen and Chuang gives a controlled single-qubit implementation of the Toffoli gate.
-Since my use of the Toffoli gate has different controls I reflect the picture, giving the following implementation of the above Toffoli gate where the first qubit is controlled by the second and third:
-
-\[ \Qcircuit @C=1em @R=1em {
- & \gate{H} & \targ & \gate{T^\dagger} & \targ & \gate{T} & \targ & \gate{T^\dagger}& \targ & \gate{T} & \gate{H} & \qw & \qw & \qw \\
- & \qw & \ctrl{-1}& \qw  & \qw & \qw & \ctrl{-1} & \qw & \qw & \gate{T^\dagger} & \targ & \gate{T^\dagger} & \targ & \gate{S} &\qw\\
- & \qw & \qw & \qw & \ctrl{-2} & \qw & \qw & \qw & \ctrl{-2} & \qw & \ctrl{-1} & \qw & \ctrl{-1} & \gate{T} & \qw
-}\]
-$Z_8$ is given by the circuit
-\[ \Qcircuit @C=1em @R=1em {
- & \gate{T} &  \gate{T} & \gate{T} & \gate{T} & \qw\\
- & \gate{T} & \gate{T}& \qw  & \qw &\qw\\
- & \gate{T} & \qw & \qw &\qw & \qw
-}\]
+Therefore we have a contradiction as our total charge is negative but $ch_1$ of each object is non-negative.$\blacksquare$
 \end{document}

### 2018-01-19 CO749-GraphColoring/assignments/A1.tex
p2 write up

@@
 Therefore we have a contradiction as our total charge is negative but $ch_1$ of each object is non-negative.$\blacksquare$
+
+\newpage
+\section{}
+\paragraph{}
+Assume Lemma $2$ is false. Let $G$ be a minimum counterexample. That is $G$ is a smallest planar graph satisfying:
+\begin{itemize}
+	\item The girth of $G$ is at least seven,
+	\item The minimum degree of $G$ is at least two, and
+	\item If $v$ is a degree $2$ vertex of $G$ with neighbours $u$ and $w$ then the degrees of $u$ and $w$ are each at least $4$.
+\end{itemize}  
+\paragraph{}
+Since the minimum degree is at least $2$, $G$ is not acyclic.
+So $G$ contains cycles, and each cycle is of length at least $7$.
+Hence every bounded face of $G$ is bounded by a cycle of length at least $7$ and the unbounded face is incident with at least $7$ edges.
+\paragraph{}
+We proceed by a discharging argument. For each vertex $v \in V(G)$ assign initial charge to $v$:
+$$ch_0(v) = 2d(v) - 6.$$
+For each face $f$ assign initial charge to $f$:
+$$ch_0(f) = |f| - 6.$$
+Then by our Lemma from class the sum of the initial charges is $-12$.
+By the girth and minimum degree of $G$, the charge of every face is at least $1$ and the charge of every vertex is non-negative except for degree $2$ vertices.
+\paragraph{}
+Discharge, establishing final charges $ch_1$ according to the following Rules:
+\begin{enumerate}
+	\item Each face $f$ sends $1$ charge to each incident vertex of degree $2$.
+	\item Each vertex of degree at least $4$ sends $\frac{1}{2}$ charge to each incident face.
+\end{enumerate}
+\paragraph{}
+If a vertex has degree at least $4$ then their initial charge is at least half their degree, so their final charge is non-negative after sending $\frac{1}{2}$ to each incident face.
+Degree $3$ vertices neither send nor receive charge and so their final charge is $0$.
+\paragraph{}
+If $v$ is vertex of degree $2$ then $v$ is incident with two faces, and so receives $2$ charge. Thus the final charge of $v$ is
+$$ch_1(v) = ch_0(v) + 2 = -2 + 2 = 0.$$
+Hence we have shown that the final charge of each vertex is non-negative. It remains to show the final charge of each face is non-negative to achieve the desired contradiction.
+\paragraph{}
+Let $f$ be a face of $G$. Let $q$ be the number of degree two vertices incident with $f$. Since the neighbours of any degree two vertex cannot be of degree two we observe that
+$$q \leq \floor{\frac{|f|}{2}}.$$
+Let $p$ be the number of vertices incident with $f$ of degree at least $4$. Since each degree two vertex has both neighbours of degree four, we have
+$$p \geq q.$$
+We have that $f$ loses $q$ charge by Rule $1$ and gains $\frac{p}{2}$ charge by Rule $2.$ Thus the final charge of $f$ is
+$$ch_1(f) = ch_0(f) -q + p \geq |f| -6 -\frac{q}{2} \geq \frac{3}{4}|f| - 6.$$
+So the final charge of $f$ is non-negative when $|f| \geq 8$.
+\paragraph{}
+Therefore we may assume that $|f| = 7$. In this case we may strengthen our bound on $p$ by observing that
+$p \geq q+1$
+by a parity argument.
+Indeed if $p = q$ then $f$ is a cycle of vertices alternating degree $2$ and degree $4$, but the number of vertices on $f$ is odd, so this cannot happen.
+Now we tighten our estimate of the final charge:
+$$ch_1(f) \geq |f| - 6 -\frac{q}{2} + 1 \geq \frac{3}{4}\cdot 7 - 6 + 1 \geq 0.$$
+Thus we have shown every face has non-negative final charge, a contradiction. $\blacksquare$
+
+\newpage
+\section{}
 \end{document}

### 2018-01-22 CO749-GraphColoring/assignments/A1.tex
Wrote up 3

@@
 \section{}
+\paragraph{Stronger Lemma}
+We will prove the following stronger result.
+If $G$ is a $2$-connected plane graph with outer cycle $C$ such that $|C| \leq 6$, and at most one degree $2$ vertices on $C$, and each inner vertex has degree at least $3$ then $G$ has either:
+\begin{itemize}
+	\item At inner face of length at most $4$, or
+	\item At inner face $f$ of length $5$ incident with $4$ inner vertices, each of which has an inner neighbour not incident with $f$.
+\end{itemize}
+\paragraph{Proof}
+Suppose not.
+Let $G$ be a minimum counterexample.
+Then each inner face of $G$ is of length at least $5$.
+Furthermore if $f$ is an inner face of $G$ of length $5$ with four degree $3$ inner vertices $v_1,v_2, v_3,v_4$ then there exists some $v_i$ such that $v_i$ has either:
+\begin{enumerate}
+	\item all of $v_i$'s neighbours are on $f$, or
+	\item $v_i$ has a neighbour on $C$.
+\end{enumerate}
+We claim that the first case cannont happen in a minimum counterexample.
+So suppose that $f$ is a length $5$ inner face with vertices $v_1,v_2, v_3, v_4, v_5$ appearing conesecutively in that order about the cycle bounding $f$, and $v_1, \dots, v_4$ have degree $3$, and without loss of generality $v_1$ is adjacent to $v_3$ or $v_4$.
+\paragraph{}
+If $v_1$ is adjacent to $v_3$ then since $v_1, v_2$, and $v_3$ are of degree $3$, then $v_2$ is a cut vertex.
+Since $G$ is plane $v_2$'s third neighbour $v$ is not on $f$ but drawn in a disc bounded by $v_1,v_2,v_3$.
+But then the only paths to $v_1$ from $v$ use $v_2$, hence $v_2$ is a cut vertex, contradiction $2$-connectedness.
+\paragraph{}
+If $v_1$ is adjacent to $v_4$ then the subgraph $H$ consisting of the cycle bounding $f$ and the edge $v_1v_4$ is a smaller counterexample than $G$, contradicting minimality.
+Indeed the outer cycle if $H$ is $v_1v_4v_5$ of length $3 \leq 6$ with one degree $2$ vertex: $v_5$.
+The subgraph $H$ is $2$-connected, and the minimum degree among inner vertices is $3$.
+\paragraph{}
+Thus we have shown if $f$ is an inner face of $G$ of length $5$ with four degree $3$ inner vertices $v_1,v_2, v_3,v_4$ then there exists some $v_i$ such that $v_i$ has a neighbour on $C$.
+\paragraph{}
+We proceed by a discharging argument. For each vertex $v \in V(G)$ set the initial charge $ch_0$ to be
+$$ch_0(v) = 2d(v) -6$$
+and for each face $f$ set the initial charge to be
+$$ch_0(f) = |f| - 6.$$
+Then by our Lemma from class the sum of the initial charges is $-12$. We reassign charge, obtaining final charges $ch_1$ according to the following Rules:
+\begin{enumerate}
+	\item Each inner vertex of degree at least $4$ sends $\frac{1}{2}$ charge to each incident (inner) face.
+	\item Each vertex on $C$ of degree at least $3$ sends $\frac{1}{2}$ charge to each incident inner face of length $5$.
+	\item If $f$ is an inner face of length $5$ with an incident inner vertex $v$ of degree $3$ such that $v$ is adjacent to vertex $u$ on $C$ then $u$ sends $\frac{1}{2}$ charge to $f$.
+\end{enumerate}
+\paragraph{}
+Each inner vertex of degree $3$ sends no charge and so its final charge is non-negative. If $v$ is an inner vertex of degree at least $4$ then the final charge of $v$ satisfies
+$$ch_1(v) \geq ch_0(v) - \frac{d(v))}{2} = \frac{3}{2}d(v) - 6 \geq 0.$$
+Therefore each inner vertex has non-negative final charge.
+\paragraph{}
+Each face sends no charge, so each inner face of length at least $6$ immediately has non-negative final charge.
+Consider an inner face $f$ of length $5$ with bounding cycle consisting of vertices $v_1,v_2,v_3,v_4,v_5$.
+Since $ch_0(f) = -1$ we want to show $f$ receives at least $1$ charge.
+If $f$ is incident with at least $2$ inner vertices of degree at least $4$ then $f$ receives at least $1$ charge by Rule $1$ and so $ch_1(f) \geq 0$.
+\paragraph{}
+So we may assume $f$ is incident with at most $1$ inner vertex at of degree at least $4$.
+First suppose, without loss of generality, $v_5$ is the only inner vertex incident with $f$ of degree at least $4$.
+So $f$ receives $\frac{1}{2}$ charge from $v_5$.
+Observe that if $f$ shares a vertex with $C$, it necessarily shares a degree $3$ vertex.
+Then either
+\begin{enumerate}
+	\item one of $v_1,\dots, v_4$ is on $C$ and of degree at least $3$, or
+	\item all of $v_1, \dots, v_4$ are inner vertices of degree $3$.
+\end{enumerate}
+In the first case $f$ receives $\frac{1}{2}$ from the vertex on $C$ by Rule $2$ and hence $ch_1(f) \geq 0$.
+In the second case there exists $v_i$ with $i \in \{1,\dots, 4\}$, such that $v_i$ is adjacent to $u$ on $C$.
+By Rule $3$, $f$ receives $\frac{1}{2}$ charge from $u$.
+Therefore $ch_1(f) \geq 0$.
+Thus in either case the final charge of $f$ is non-negative.
+\paragraph{}
+Now we may assume that $f$ is incident with no inner vertices of degree at least $4$. Now among $v_1, \dots, v_4$ there exists $i \in \{1,\dots, 4\}$ such that either:
+\begin{itemize}
+	\item $v_i$ is on $C$ of degree at least $3$, or
+	\item $v_i$ is adjacent with a vertex $u$ on $C$
+\end{itemize}
+In the former case $f$ receives $\frac{1}{2}$ from $v_i$ by Rule $2$and in the latter $f$ receives $\frac{1}{2}$ charge from $u$.
+Then among $\{v_1, \dots, v_5\} \backslash \{v_i\}$ there exists $j \in \{1,\dots, 5\}\backslash\{i\}$ such that either:
+\begin{itemize}
+	\item $v_j$ is on $C$ of degree at least $3$, or
+	\item $v_j$ is adjacent with a vertex $v$ on $C$
+\end{itemize}
+In the former case $f$ receives $\frac{1}{2}$ from $v_j$ by Rule $2$and in the latter $f$ receives $\frac{1}{2}$ charge from $v$.
+Finally, $f$ receives at least $1$ charge total, and hence $ch_1(f) \geq 0$.
+\paragraph{}
+Thus each inner vertex and each inner face has non-negative final charge.
+The vertex on $C$ of degree $2$ has final charge $-2$.
+We lower bound the charge of a vertex $v$ on $C$ of degree at least $3$.
+The vertex $v$ gives away at most $\frac{1}{2}(d(v) - 1)$ charge by Rule $2$ and at most $\frac{1}{2}(d(v) - 2)$ charge by Rule $3$.
+So the final charge of $v$ is
+$$ch_1(v) \geq 2d(v) - 6 - d(v) + \frac{3}{2} = d(v) - \frac{9}{2} \geq \frac{6}{2} - \frac{9}{2} = -\frac{3}{2}.$$
+Thus, using $|C| \leq 6$, our total final charge is at least
+$$|C| - 6 -\frac{3}{2}(|C| -1) - 2 = -\frac{1}{2}|C| - 6 -\frac{1}{2} \geq -3 -6 -\frac{1}{2} > -12$$
+But since no charge is added or lost during this process, this contradicts that the sum of charges is $-12$. $\blacksquare$
+
+\newpage
+\section{}
+
 \end{document}

### 2018-01-22 CO749-GraphColoring/assignments/A1.tex
p4

@@
 \section{}
-
+\paragraph{Stronger Result}
+We prove the same Lemma with a better lower bound: $$|V(G) \backslash V(C)| \geq 8|V(C)| - 48.$$ 
+\paragraph{Proof}
+Suppose not.
+Let $G$ be a minimum counterexample.
+Then $G$ is a $2$-connected plane graph with outer cycle $C$ such that $|V(G)\backslash V(C)| < 8|V(C)| - 48$ and $|V(G) \backslash V(C)| \geq 1$ satisfying
+\begin{itemize}
+	\item Every inner vertex has degree at least $6$, and
+	\item Every inner vertex of degree $6$ has one of the following:
+	\begin{itemize}
+		\item An incident inner face of length at least $4$,
+		\item An inner neighbour of degree at least $7$, or
+		\item A neighbour on $C$.
+	\end{itemize}
+\end{itemize}
+\paragraph{}
+We begin by observing that each $v \in V(C)$ has $d(v) \geq 3$.
+Indeed suppose there exists some $v \in V(C)$ such that $d(v) = 2$.
+Suppressing $v$ leads to a smaller counterexample than $G$, a contradiction.
+This follows since suppressing $v$ does not change the degree of inner vertices or the length of faces incident only with inner vertices.
+\paragraph{} 
+We proceed by a discharging argument. For each $v \in V(G)$ set the initial charge of $v$ to be
+$$ch_0(v) = d(v) - 6$$
+and for each face $f$ of $G$ set the initial charge of $f$ to be
+$$ch_0(f) = 2|f| - 6.$$
+Then by our Lemma from class the total sum of the charges is $-12$. We discharge according to the following Rules:
+\begin{enumerate}
+	\item Each inner face of length at least $4$ sends $\frac{1}{8}$ charge to each incident vertex.
+	\item Each inner vertex of degree at least $7$ sends $\frac{1}{8}$ charge to each adjacent vertex.
+	\item Each vertex on $C$ of degree at least $4$ sends $\frac{1}{8}$ charge to each adjacent vertex.
+\end{enumerate}
+\paragraph{}
+Each length $3$ face sends no charge and so has non-negative final charge. Each face $f$ of length at least $4$ sends at most $\frac{1}{8}|f|$ charge so its final charge is
+$$ch_1(f) \geq ch_0(f) -\frac{1}{8}|f| \geq \frac{15}{8}|f| - 6 \geq\frac{15}{2} - 6 \geq 0.$$
+Hence every face has non-negative charge.
+\paragraph{}
+We argue that every inner vertex has at least $\frac{1}{8}$ charge. Let $v$ be an inner vertex. If $d(v) \geq 7$ then $v$ sends $\frac{1}{8}d(v)$ charge. So the final charge of $v$ is 
+$$ch_1(v) = \frac{7}{8}d(v) - 6 \geq \frac{49}{8} - \frac{48}{8}  = \frac{1}{8}.$$
+Now suppose $d(v) = 6.$ Since $ch_0(v) = 0$ we need $v$, which sends no charge, to receive $\frac{1}{8}$ charge.
+We proceed by case analysis.
+If $v$ has an incident inner face of length at least $4$ then $v$ receives $\frac{1}{8}$ charge by Rule $1.$ Similarly if $v$ has an incident inner neighbour of degree at least $7$ then $v$ receives $\frac{1}{8}$ charge by Rule $2.$
+\paragraph{}
+So we may assume that every incident face to $v$ is a triangle, and every inner neighbour to $v$ is of degree $6$.
+Then we know $v$ is adjacent to a vertex $v_1$ on $C$.
+Label the neighbours of $v$ as $v_1, \dots, v_6$ in clockwise order from $v_1$ with respect to their drawing in $G$.
+If $d(v_1) \geq 4$ then $v$ receives $\frac{1}{8}$ charge from $v_1$.
+Otherwise $d(v_1) = 3$.
+Since every incident face to $v$ is a triangle this implies $v_2$ is on $C$.
+Again, if $d(v_2) \geq 4$ then $v$ receives $\frac{1}{8}$ charge from $v_2$ and otherwise $d(v_2) = 3$.
+Repeating this argument for each $i \in \{3, \dots, 6\}$ we see that either $v_i$ sends $v$ $\frac{1}{8}$ charge or $d(v_i) = 3$ and $v_{i+1}$ (when $v_{i+1}$ exists) is on $C$.
+If we complete this process with $v$ receiving no charge then each neighbour of $v$ is a degree $3$ vertex on $C$.
+That is $G = G[\{v,v_1, \dots, v_6\}]$.
+But that is a contradiction since it would imply
+$$1 = |V(G) \backslash V(C)|  < 8|V(C)| - 48 = 8\cdot 6 - 48 = 0$$
+i.e. $1 < 0$.
+\paragraph{}
+Hence we have shown each inner vertex has final charge at least $\frac{1}{8}$.
+Now we claim each vertex $v$ on $C$ has final charge at least $-3$.
+Indeed if $d(v) = 3$ then $v$ sends no charge, and $ch_1(v) = ch_0(v) = -3$ as desired.
+Otherwise $d(v) \geq 4$ and $v$ sends $\frac{1}{8}d(v)$ charge to each adjacent vertex.
+Thus the final charge is
+$$ch_1(v) \geq \frac{7}{8}d(v) - 6 \geq \frac{7}{2} - 6 > -3.$$
+So our total charge is at least the charge of $C$ plus $-3$ times the number of vertices on $C$ plus $\frac{1}{8}$ the number of inner vertices. That is
+$$-12 \geq \frac{1}{8}|V(G)\backslash V(C)| + 2|V(C)| - 6 -3|V(C)|.$$
+Rearranging we see that
+$$8|V(C)| - 48 \geq |V(G) \backslash V(C)|$$
+contradicting that $|V(G)\backslash V(C)| < 8|V(C)| - 48.$ $\blacksquare$
 \end{document}

### 2018-01-26 CO749-GraphColoring/assignments/A2.tex
problem 1

--- 2018-01-19 CO749-GraphColoring/assignments/A1.tex
@@
 
-\lhead{CO 749 - \today}
-\rhead{W. Justin Toth - A1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{\today}
+\rhead{W. Justin Toth - A2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \paragraph{}
-Let $G$ be a graph of minimum degree at least $3$ satisfying all of the following:
-\begin{itemize}
-	\item Every triangle contains at most one vertex of degree three.
-	\item If two vertices of degree three are adjacent then they are in triangles.
-	\item No two distinct triangles share a common edge.
-\end{itemize}
-Suppose for a contradiction that $\abs{E(G)} < 5\abs{V(G)}/3$.
-\begin{claim}\label{1.1}
-	For every $v \in V(G)$, if $d(v) = 3$ then $v$ has at most one neighbour of degree $3$.
-\end{claim}
-\begin{proof}
-	Assume for a contradiction that there exist $u,v, w \in V(G)$ satisfying
-	$$d(u) = d(v) = d(w) = 3$$
-	and $uv, vw \in E(G)$, i.e.
-$u$ and $w$ are adjacent to $v$.
-But then, using the second property of $G$, $v$ is in a triangle.
-Denote this triangle by $C$.
-Since every triangle contains at most one vertex of degree three, $u$ and $w$ are not contained in $C$.
-But then since $d(v) = 3$, $v$ has at most one neighbour in $C$, a contradiction to $C$ being a $K_3$.
-\end{proof}
+By Voigt's Theorem there exists a planar graph $H$ and a list assignment $L$ with $|L(u)| \geq 4$ for all $u \in V(H)$, such that $H$ is not $L$-colorable.
+We construct a graph $G$ as follows.
+Take $5$ distinct copies of $H$: $H_1, \dots, H_5$ where each vertex $u_i \in V(H_i)$ corresponds to vertex $u \in V(H)$, and take an additional vertex $v$.
+Add edges $vu_i$ for all $u_i \in V(H_i)$ for all $i \in \{1,\dots, k\}$.
+i.e.
+$v$ is adjacent to every vertex in every copy of $H$.
 \paragraph{}
-We proceed by a discharging argument. For each vertex $v \in V(G)$ set the initial charge as
-$$ch_0(v) = d(v) - 4$$
-and for each edge $e \in E(G)$ set the initial charge
-$$ch_0(e) = 1/3.$$
-Then our total charge is, using handshaking,
-\begin{align*}
-\sum_{v\in V(G)} ch_0(v) + \sum_{e \in E(G)} ch_0(e) &= \sum_{v \in V(G)} d(v) - 4|V(G)| + \frac{1}{3} |E(G)|\\ &=  \frac{7}{3}|E(G)| - 4|V(G)\\ &< \frac{35}{9}|V(G)| - \frac{36}{9}|V(G)|\\ &< 0
-\end{align*}
-with the first inequality following from our assumption that $|E(G)| < 5|V(G)|/3$.
+To see that $G$ is an apex graph observe that $G-v$ consists of five disjoint copies of a planar graph, and hence $G-v$ is planar.
 \paragraph{}
-Redistribute the charge, to obtain $ch_1$, according to the following Rules:
-\begin{enumerate}
-	\item If edge $uv$ contains precisely one vertex $u$ such that $d(u) = 3$ then $uv$ sends $1/3$ charge to $u$.
-	\item If edge $uv$ is such that $d(u) \geq 4$ and $d(v) \geq 4$ and $u,v$ are in a triangle with vertex $w$ then $uv$ sends $1/3$ charge to $w$.
-\end{enumerate}
+We now construct a $5$-list-assignment $L'$ for which $G$ does not have an $L'$-coloring, completing the proof that $G$ is not $5$-list-colorable.
+Let $c_1, \dots, c_5$ be $5$ disjoint colors from $L$.
+That is for all $u \in V(H)$, $$L(u) \cap \{c_1, \dots, c_5\} = \emptyset.$$
+Let $L'(v) = \{c_1, \dots, c_5\}$.
+For each $i \in \{1, \dots, 5\}$ and for each $u_i \in V(H_i)$ let $L'(u_i) = L(u) \cup \{c_i\}$.
+Since $|L(u)| \geq 4$, $|L'(u_i)|\geq 5$.
+Thus we have a $5$-list-assignment $L'$ for $G$.
 \paragraph{}
-Since vertices sends no charge, clearly every vertex $v \in V(G)$ with $d(v) \geq 4$ has $$ch_1(v) = ch_0(v) \geq 0.$$ Now we will show that if $v$ is a vertex with $d(v) = 3$ then $ch_1(v) \geq 0$.
-From Claim \ref{1.1} $v$ is adjacent with at most one vertex of degree $3$.
-If $v$ is adjacent with no degree three vertices then each edge incident with $v$ sends $\frac{1}{3}$ charge to $v$ by Rule $1$ and hence
-$$ch_1(v) = ch_0(v) + 3\cdot\frac{1}{3} = 0.$$
-On the other hand, if $v$ is adjacent with one degree three vertex then $v$ receives $\frac{2}{3}$ charge from incident edges by Rule $1$.
-The remaining needed $\frac{1}{3}$ is obtained from Rule $2$ as follows.
-Since $v$ is adjacent with a degree three vertices, $v$ lies in a triangle, and since every triangle contains at most one degree three vertex the edge containing the other two vertices of this triangle sends $\frac{1}{3}$ charge to $v$.
-Therefore $ch_1(v) \geq 0$.
-\paragraph{}
-It remains to verify that each edge $uv$ satisfies $ch_1(uv) \geq 0$.
-Edge $uv$ can satisfy at most one of Rule $1$ and Rule $2$.
-If $uv$ satisfies Rule $1$ it sends $\frac{1}{3}$ charge and so $ch_1(uv) = 0$.
-Instead if $uv$ satisfies Rule $2$ it can only do so for one triangle, since no two distinct triangles share a common edge.
-So $uv$ only sends $\frac{1}{3}$ charge by Rule $2$.
-Therefore $ch_1(uv) \geq 0$.
-\paragraph{}
-Therefore we have a contradiction as our total charge is negative but $ch_1$ of each object is non-negative.$\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-Assume Lemma $2$ is false. Let $G$ be a minimum counterexample. That is $G$ is a smallest planar graph satisfying:
-\begin{itemize}
-	\item The girth of $G$ is at least seven,
-	\item The minimum degree of $G$ is at least two, and
-	\item If $v$ is a degree $2$ vertex of $G$ with neighbours $u$ and $w$ then the degrees of $u$ and $w$ are each at least $4$.
-\end{itemize}  
-\paragraph{}
-Since the minimum degree is at least $2$, $G$ is not acyclic.
-So $G$ contains cycles, and each cycle is of length at least $7$.
-Hence every bounded face of $G$ is bounded by a cycle of length at least $7$ and the unbounded face is incident with at least $7$ edges.
-\paragraph{}
-We proceed by a discharging argument. For each vertex $v \in V(G)$ assign initial charge to $v$:
-$$ch_0(v) = 2d(v) - 6.$$
-For each face $f$ assign initial charge to $f$:
-$$ch_0(f) = |f| - 6.$$
-Then by our Lemma from class the sum of the initial charges is $-12$.
-By the girth and minimum degree of $G$, the charge of every face is at least $1$ and the charge of every vertex is non-negative except for degree $2$ vertices.
-\paragraph{}
-Discharge, establishing final charges $ch_1$ according to the following Rules:
-\begin{enumerate}
-	\item Each face $f$ sends $1$ charge to each incident vertex of degree $2$.
-	\item Each vertex of degree at least $4$ sends $\frac{1}{2}$ charge to each incident face.
-\end{enumerate}
-\paragraph{}
-If a vertex has degree at least $4$ then their initial charge is at least half their degree, so their final charge is non-negative after sending $\frac{1}{2}$ to each incident face.
-Degree $3$ vertices neither send nor receive charge and so their final charge is $0$.
-\paragraph{}
-If $v$ is vertex of degree $2$ then $v$ is incident with two faces, and so receives $2$ charge. Thus the final charge of $v$ is
-$$ch_1(v) = ch_0(v) + 2 = -2 + 2 = 0.$$
-Hence we have shown that the final charge of each vertex is non-negative. It remains to show the final charge of each face is non-negative to achieve the desired contradiction.
-\paragraph{}
-Let $f$ be a face of $G$. Let $q$ be the number of degree two vertices incident with $f$. Since the neighbours of any degree two vertex cannot be of degree two we observe that
-$$q \leq \floor{\frac{|f|}{2}}.$$
-Let $p$ be the number of vertices incident with $f$ of degree at least $4$. Since each degree two vertex has both neighbours of degree four, we have
-$$p \geq q.$$
-We have that $f$ loses $q$ charge by Rule $1$ and gains $\frac{p}{2}$ charge by Rule $2.$ Thus the final charge of $f$ is
-$$ch_1(f) = ch_0(f) -q + p \geq |f| -6 -\frac{q}{2} \geq \frac{3}{4}|f| - 6.$$
-So the final charge of $f$ is non-negative when $|f| \geq 8$.
-\paragraph{}
-Therefore we may assume that $|f| = 7$. In this case we may strengthen our bound on $p$ by observing that
-$p \geq q+1$
-by a parity argument.
-Indeed if $p = q$ then $f$ is a cycle of vertices alternating degree $2$ and degree $4$, but the number of vertices on $f$ is odd, so this cannot happen.
-Now we tighten our estimate of the final charge:
-$$ch_1(f) \geq |f| - 6 -\frac{q}{2} + 1 \geq \frac{3}{4}\cdot 7 - 6 + 1 \geq 0.$$
-Thus we have shown every face has non-negative final charge, a contradiction. $\blacksquare$
-
-\newpage
-\section{}
+Suppose for a constradiction $G$ has an $L'$-coloring $\phi$.
+Say $\phi(v) = c_i$.
+Then $\phi(u_i) \neq c_i$ for all $u_i \in V(H_i)$.
+Hence $\phi(u_i) \in L(u_i)$ for all $u_i \in V(H_i)$.
+That is, $\phi$ restricted to $H_i$ is an $L$-coloring of $H-_i$.
+But this contradicts that $H_i$ is not $L$-colorable.
+Therefore $G$ does not have an $L'$-coloring and hence is not $5$-list-colorable.
+$\blacksquare$
 \end{document}

### 2018-01-27 CO749-GraphColoring/assignments/A2.tex
p2 with bonus

@@
 $\blacksquare$
+
+\section{}
+\paragraph{}
+We opt to prove the BONUS Lemma, i.e. Lemma $3$:
+\textit{Let $G$ be a graph and for each vertex $v$ of $G$ let $r(v)$ be a non-negative integer. If $L$ is a list assignment of $G$ such that for every vertex $v$ of $G$
+$$|L(v)| \geq \frac{d(v) + 1}{r(v)+1},$$
+then $G$ has an $r(v)$-defective coloring}. This Lemma directly implies Lemma $2$ of this problem.
+\paragraph{}
+Consider a graph $G$, non-negative integers $r(v)$ for each $v \in V(G)$, and a list assignment $L$ of $G$ such that for each $v \in V(G)$
+$$|L(v)| \geq \frac{d(v) + 1}{r(v)+1}.$$
+Let $\phi$ be an improper $L$-coloring of $G$ minimizing the number of ``improper" edges $uv \in E(G)$ such that $\phi(u) = \phi(v)$.
+Suppose for a contradiction that $\phi$ is not an $r(v)$-defective coloring.
+\paragraph{}
+Then there exists a vertex $v$ such that the number of neighbours of $v$ receiving the same color as $v$ under $\phi$ is at least $r(v) + 1$.
+If there exists a color $c \in L(v)$ such that the number of neighbours of $v$ colored $c$ is at most $r(v)$ then we can construct a new coloring $\phi'$ by setting $\phi'(v) = c$ and $\phi'(u) = \phi(u)$ for all $u \neq v$.
+This coloring has $r(v)$ new improper edges, those between $v$ and its $c$-colored neighbours.
+But $\phi'$ switches $r(v) + 1$ edges which were improper under $\phi$, those between $v$ and its $\phi(v)$-colored neighbours, to proper edges.
+Hence $\phi'$ has fewer improper edges than $\phi$, a contradiction.
+\paragraph{}
+If such a color $c \in L(v)$ does not exist then  for each $c \in L(v)$ the number of neighbours of $v$ colored $c$ under $\phi$ is at least $r(v) + 1$. But then $v$ has at least
+$$|L(v)|(r(v) + 1) \geq d(v) + 1$$
+neighbours, a contradiction. Therefore $\phi$ is an $r(v)$-defective coloring $\blacksquare$
 \end{document}

### 2018-01-30 CO749-GraphColoring/assignments/A2.tex
first two claims of 3

@@
 
+\newpage
 \section{}
@@
 neighbours, a contradiction. Therefore $\phi$ is an $r(v)$-defective coloring $\blacksquare$
+
+\newpage
+\section{}
+\paragraph{}
+We prove a slightly more general result allowing $m(v)$ to take the value $0$.
+This will make induction easier when we wish to delete from a list of size $2$.
+For such a $v$, $L(v)= \emptyset$ and the desired $S(v) = \emptyset$.
+Everything else is exactly as stated in the problem.
+\paragraph{}
+Suppose not. Let $G$ be a counterexample satisfying:
+\begin{enumerate}
+	\item $\sum_{v \in V(G)} m(v)$ is minimized, and subject to that
+	\item if possible $G(c)$ is connected for each $c \in L(G)$, and subject to that possibility
+	\item $|L(G)|$ is minimized.
+\end{enumerate}
+\begin{claim}
+	\begin{enumerate}
+		\item For all $c \in L(G)$, $G(c)$ is connected.
+		\item For every pair of distinct colors $c,c' \in L(G)$, $G(c) \neq G(c')$.
+		\item For every pair of distinct colors $c, c' \in L(G)$, if $V(G(c)) \cap V(G(c')) = \emptyset$ then $G[V(G(c))\cup V(G(c'))]$ is disconnected.
+	\end{enumerate}
+\end{claim}
+\begin{proof}
+	\paragraph{} For (1.). While there exists a color $c \in L(G)$ such that $G(c)$ has at least two connected components $G_1$ and $G_2$ do the following:
+		\subparagraph{}
+		Let $c'$ be a new color not contained in $L(G)$.
+For each $v \in V(G_2)$ let $$L'(v) = (L(v)\cup \{c'\}) \backslash \{c\}.$$ For each $v \in V(G) \backslash V(G_2)$ let $L'(v) = L(v)$.
+Replace $L$ with $L'$.
+	\paragraph{}
+	We claim that every iteration of the above preserves the property of $(G,m,L)$ being a counterexample, but decreases the ``disconnectivity".
+i.e.
+the sum over $c \in L(G)$ such that $G(c)$ is disconnected, of the number of connected components in $G(c)$.
+	\paragraph{}
+	It is clear the disconnectivity decreases.
+We add a new color $c'$ for which $G(c')$ is connected, and $G(c)$ loses one connected component, and the number of connected components of $G(c'')$ is not affected for any other color $c'' \in L(G)$.
+	\paragraph{}
+	To verify that the new lists, say $L'$, also admit a counterexample with $G$ and $m$, suppose to contrary there are $S'(v) \subseteq L'(v)$ of size $m(v)$ for each $v \in V(G)$ and $S'(u) \cap S'(v) = \emptyset$ for all $uv \in E(G)$.
+For each $v \in V(G)$ let $$S(v) = \begin{cases} S'(v)\cup \{c\} \backslash \{c'\}, &\text{if } v \in V(G_2) \text{ and } c' \in S'(v) \\
+	S'(v), &\text{otherwise}.\end{cases} $$
+	Then $S(v) \subseteq L(v)$ and $|S(v)| = m(v)$ for each $v \in V(G)$.
+Further, for each $uv \in E(G)$, $S(u) \cap S(v) = \emptyset$.
+To see this, observe that if $S(u) \cap S(v) \neq \emptyset$ then by our choice of $S'$, $S(u) \cap S(v) = \{c\}$ and, since $G(c)$ was disconnected, $u, v \in V(G_2)$.
+But then, $S'(u) \cap S'(v) = \{c'\}$, a contradiction.
+	\paragraph{}
+	So the above algorithm will terminate when the ``disconnectivity" reaches $0$ and we will have found a connected counterexample with the same value for $\sum_{v} m(v)$ as $G$.
+This contradicts our choice of $G$ if there exists color $c$ such that $G(c)$ is disconnected.
+Therefore $G(c)$ is connected for all $c \in L(G)$.
+	\paragraph{}For (2.). Suppose for a contradiction there exist distinct colors $c,c' \in L(G)$ such that $G(c) = G(c')$. Construct new lists $L'$ such that for each $v \in V(G)$
+	$$ L'(v) = \begin{cases}
+	L(v) \backslash \{c,c'\}, &\text{ if } v \in V(G(c)) \\
+	L(v), &\text{otherwise.}
+		\end{cases}$$
+	Let $m'(v) = m(v) - 1$ if $v \in V(G(c))$ (this is well-defined since such $m(v) \geq 2$) and let $m'(v) = m(v)$ otherwise.
+By the minimality of $G$, there exist for each $v \in V(G)$, a set $S'(v)\subseteq L'(v)$ of size $m'(v)$.
+These sets satisfy $S'(u) \cap S'(v) = \emptyset$ for all $uv \in E(G)$.
+Order the vertices of $G(c)$ as $v_1, \dots, v_k$ with respect to some orientation of $G$.
+Construct sets $S(v)$ as follows.
+If $v \in V(G) \backslash V(G(c))$ set $S(v) = S'(v)$.
+If $v_i \in V(G(c))$ and $i$ is even set $S(v) = S'(v) \cup \{c\}$.
+If $v_i \in V(G(c))$ and $i$ is odd set $S(v) = S'(v) \cup \{c'\}$.
+Since $c$ and $c'$ do not appear in sets outside $G(c)$, and we alternate $c$ and $c'$ along $G(c)$, we have $S(u) \cap S(v) = \emptyset$ for all $uv \in E(G)$.
+But this contradicts that $(G,m,L)$ is a counterexample.
+Hence for distinct colors $c$ and $c'$, $G(c) \neq G(c')$.
+
+\end{proof}
+
 \end{document}

### 2018-01-30 CO749-GraphColoring/assignments/A2.tex
Claim 0.3 done

@@
 		\item For every pair of distinct colors $c,c' \in L(G)$, $G(c) \neq G(c')$.
-		\item For every pair of distinct colors $c, c' \in L(G)$, if $V(G(c)) \cap V(G(c')) = \emptyset$ then $G[V(G(c))\cup V(G(c'))]$ is disconnected.
+		\item For every pair of distinct colors $c, c' \in L(G)$, we have $V(G(c)) \cap V(G(c')) \neq \emptyset$ (we strengthened this claim).
 	\end{enumerate}
@@
 Hence for distinct colors $c$ and $c'$, $G(c) \neq G(c')$.
-
+	\paragraph{} For (3.).
+Suppose for a contradiction  there exist distinct colors $c,c' \in L(G)$, such that $V(G(c)) \cap V(G(c')) = \emptyset$.
+Construct new lists $L'(v)$ for each $v \in V(G)$ as follows.
+If $v \in V(G) \backslash V(G(c'))$ then set $L'(v) = L(v)$.
+Otherwise, i.e.
+$v \in V(G(c'))$ set $L'(v) = L(v) \cup \{c\} \backslash \{c'\}.$ Then for each $v \in V(G(c'))$, we have $|L'(v)|  = |L(v)| + 1 - 1 = m(v)$, since $V(G(c))$ and $V(G(c'))$ are disjoint.
+But $L'(G) = L(G) \backslash \{c\}$ so by minimality there exist sets $S'(v) \subseteq L'(v)$ for each $v \in V(G)$ such that $|S'(v)| = m(v)$, satisfying $S'(u) \cap S'(v) = \emptyset$ for each $uv \in E(G)$.
+	\paragraph{}
+	For each $v \in V(G)$ construct $S(v)$ so that
+	$$S(v) = \begin{cases}
+		S'(v) \cup \{c'\}\backslash \{c\}, &\text{ if } v \in V(G(c')) \text{ and } c \in S'(v) \\
+		S'(v), &\text{ otherwise}.
+	\end{cases}$$
+	Then each set $S(v) \subseteq L(v)$ is of size $m(v)$ and for each edge $uv \in E(G)$, $S(u) \cap S(v) = \emptyset$.
+To verify disjointness, suppose for a contradiction $S(u) \cap S(v) \neq \emptyset$ for some edge $uv$.
+Since $S'(u) \cap S'(v) = \emptyset$, $S(u) \cap S(v) = \{c'\}$.
+But then $u,v \in V(G(c'))$ and so $S'(u) \cap S(v) = \{c\}$ since $c'$ is only added to $S(u), S(v)$ if $c \in S'(u), S'(v)$.
+But this contradicts disjointness of $S'(u), S'(v)$.
+So sets $S(v)$ for $v \in V(G)$ violate that $(G,m,L)$ is a counterexample, a contradiction.
 \end{proof}
-
+	\begin{claim}
+		If $c \neq c' \in L(G)$ such that $G(c), G(c')$ are paths, then there does not exist a component $P$ of $G(c) \cap G(c')$ with an even number of vertices such that one end of $P$ is an end of $G(c)$ and the other end of $P$ is an end of $G(c')$.
+	\end{claim}
+	\begin{proof}
+		\paragraph{} Suppose for a contradiction such $P$ exists.
+Order the vertices of $P$ as $v_1, \dots, v_k$ so that $v_1$ is an end of $G(c)$ and $v_k$ is an end of $G(c')$.
+Construct new lists $L'(v)$ for each $v \in V(G)$ as follows.
+If $v \in V(P)$ let $L'(v) = L(v) \backslash \{c,c'\}$ and let $m'(v) = m(v) - 1$.
+Otherwise let $L'(v) = L(v)$ and $m'(v) = m(v)$.
+By minimality, there exist sets $S'(v) \subseteq L'(v)$ for each $v \in V(G)$ of size $m'(v)$, such that $S'(u) \cap S'(v) = \emptyset$ for each $uv \in E(G)$.
+		\paragraph{}
+		Construct sets $S(v)\subseteq L(v)$ for each $v \in V(G)$ as follows.
+If $v_i \in V(P)$ and $v_i$ is odd then set $S(v) = S'(v) \cup \{c\}$.
+If $v_i \in V(P)$ and $v_i$ is even set $S(v) = S'(v) \cup \{c'\}$.
+If $v \in V(G)\backslash V(P)$ set $S(v) = S'(v)$.
+Then $|S(v)| = m(v)$ for each $v \in V(G)$.
+Since we alternate adding $c$ and $c'$ along $P$ the sets $S(u)$ and $S(v)$ are disjoint for each edge $uv \in E(G)$.
+Since we did not change $S(u)$ and $S(v)$ for edges $uv$ with neither endpoint on $P$ they are also disjoint.
+It remains to consider two edges $u_1v_1$ and $v_ku_k$.
+Since $v_1$ is an end of $G(c)$ and $v_2 \in G(c)$ we have $c \not\in L(u_1)$ and hence, as $S'(u_1) \cap S'(v_1) = \emptyset$, $S(u_1)$ and $S(v_1)$ are disjoint.
+Similarly since $v_k$ is an edge of $G(c')$, and $k$ is even, $S(u_k)$ and $S(v_k)$ are disjoint.
+But then the sets $S(v)$ contradict that $(G,m,L)$ is a counterexample.
+	\end{proof}
 \end{document}

### 2018-01-31 CO749-GraphColoring/assignments/A2.tex
Claim 3.4

@@
 Since we did not change $S(u)$ and $S(v)$ for edges $uv$ with neither endpoint on $P$ they are also disjoint.
-It remains to consider two edges $u_1v_1$ and $v_ku_k$.
+It remains to consider two edges $u_1v_1$ and $v_ku_k$ where $u_1, u_k \not\in V(P)$.
 Since $v_1$ is an end of $G(c)$ and $v_2 \in G(c)$ we have $c \not\in L(u_1)$ and hence, as $S'(u_1) \cap S'(v_1) = \emptyset$, $S(u_1)$ and $S(v_1)$ are disjoint.
@@
 	\end{proof}
+\begin{claim}
+	If $c \in L(G)$ then $|V(G(c))|$ is even.
+\end{claim}
+\begin{proof}
+	Deferred to Problem $4$.
+\end{proof}
+\begin{claim}
+	For every edge $e=uv \in E(G)$, $L(u) \subseteq L(v)$ or $L(v) \subseteq L(u)$.
+\end{claim}
+\begin{proof}
+	\paragraph{}
+	Suppose not.
+Then there exists an edge $e=uv \in E(G)$ such that $L(u)\backslash L(v) \neq \emptyset$ and $L(v)\backslash L(u) \neq \emptyset$.
+Let $c_u \in L(u) \backslash L(v)$ and let $c_v \in L(v)\backslash L(u)$.
+By our previous claims, both $G(c_u)$ and $G(c_v)$ are connected and non-disjoint, so they intersect in a path.
+Let $P = G(c_u) \cap G(c_v)$.
+Since $v \not\in G(c_u)$ (also note $u \not\in G(c_v)$), $P \neq G$, and we have $G(c_u) \cup G(c_v) = G$.
+Further one end of $P$ is in $G(c_u)$ and the other is in $G(c_v)$.
+Thus by Claim $3.2$, $|V(P)|$ is odd.
+	\paragraph{}
+	Since $|V(G)|$ is even,  and $|V(P)|$ is odd, we have that $|V(G) \backslash V(P)|$ is odd.
+Then since $V(G)\backslash V(P)$ is the disjoint union of $V(G(c_u))\backslash V(P)$ and $V(G(c_v)) \backslash V(P)$, precisely one of $|V(G(c_u))\backslash V(P)|$ or $|V(G(c_v))\backslash V(P)|$ is even and the other is odd.
+Say without loss of generality that $|V(G(c_u))\backslash V(P)|$ is even.
+Then $|V(G(c_u))| = |V(G(c_u))\backslash V(P)| + |V(P)|$ is odd.
+But this contradicts Claim $3.3$.
+	\end{proof}
 \end{document}

### 2018-01-31 CO749-GraphColoring/assignments/A2.tex
finishing proof of 3, except one outstanding claim

@@
 	\end{proof}
+\begin{claim}
+For every edge $e=uv \in E(G)$, $L(u) = L(v)$.
+\end{claim}
+\begin{proof}
+	TODO
+\end{proof}
+\paragraph{}
+We are now ready to complete the main proof.
+The argument is essentially the same as $2$-choosability of even cycles.
+From Claim $3.5$, let $L$ denote the constant list assigned to each vertex, i.e.
+$L = L(u)$ for each $u \in V(G)$.
+We know $|L|$ is even and we need to assign half its elements to each vertex.
+\paragraph{}
+ Let $S_1 \subseteq L$ be a subset of size $|L|/2$.
+Let $S_2 = L \backslash S_1$.
+Then $|S_1| = |S_2| = m(v)$ for any $v \in V(G)$.
+Enumerate the vertices about $G$ as $v_1, \dots, v_k$ with $k$ even.
+For each $v_i \in V(G)$, let $S(v) = S_1$ is $i$ is odd and $S(v) = S_2$ is $i$ is even.
+Since $G$ is an even cycle for any edge $v_iv_j \in E(G)$ exactly one of $i,j$ is odd and the other is even.
+Hence $S(v_i) \cap S(v_j) = \emptyset$, contradicting that $G$ is a counterexample.$\blacksquare$
+
+\newpage
+\section{}
 \end{document}

### 2018-01-31 CO749-GraphColoring/assignments/A2.tex
Only missing two subclaims

@@
 We are now ready to complete the main proof.
-The argument is essentially the same as $2$-choosability of even cycles.
+The argument is essentially the same as $2$-colorability of even cycles.
 From Claim $3.5$, let $L$ denote the constant list assigned to each vertex, i.e.
@@
 \section{}
+\paragraph{}
+We will prove Claim $3.3$ from the previous problem.
+Suppose the Claim is false.
+Then there exists $c \in L(G)$  such that $G(c)$ is a path $P=v_1\dots v_k$ with $k$ odd.
+Choose such $c$ so that $|V(P)|$ is minimized.
+We adopt the same notation and definitions as the problem statement.
+\begin{lemma}
+	For every $a,b \in L(G)$, if $|V(G(a))|$ is odd then we do not have $G(a) \subseteq G(b)$.
+\end{lemma}
+\begin{proof}
+	Suppose not.
+Then delete colors $a$ and $b$ from $L(v)$ for each $v \in V(G(a))$ to obtain lists $L'(v)$.
+For vertices $v$ off $G((a))$ let $L'(v) = L(v)$.
+By minimality, there exist $S'(v) \subseteq L'(v)$, such that $|S'(v)| = \frac{1}{2}|L'(v)|$ for each $v \in V(G)$, and for each edge $uv \in E(G)$, $S'(u) \cap S'(v) = \emptyset$.
+	\paragraph{}
+	Enumerate the vertices of $G(a)$ as $v_1 \dots v_\ell$ with $\ell$ odd.
+For each $v_i \in V(G(a))$ let $S(v_i) = S'(v_i) \cup \{a\}$ if $i$ is odd, and $S(v_i) = S'(v_i) \cup \{b\}$ if $i$ is even.
+For each $v \in V(G)\backslash V(G(a))$ let $S(v) = S'(v)$.
+Then $|S(v)| = \frac{1}{2} |L(v)|$ for each $v \in V(G)$.
+	\paragraph{}
+	Let $uv \in E(G)$.
+If $u,v \not\in V(G(a))$ then $S(u) \cap S(v) = S'(u) \cap S'(v) = \emptyset$.
+If $u,v \in V(G(a))$ then $S(u) \cap S(v) = S'(u) \cap S'(v) \cap \{a\} \cap \{b\} = \emptyset$.
+If $u \in V(G(a))$ and $v \not\in V(G(a))$ then either $u = v_1$ or $u = v_\ell$.
+In either case $S(u) = S'(u) \cup \{a\}$.
+Since $v \not\in V(G(a))$, $a \not\in L(v)$ and so $S(u) \cap S'(v) = \emptyset$ as desired.
+But then the sets $S(v)$ contradict that $G$ is a counterexample.
+\end{proof}
+\begin{claim}
+	For every $v \in V(P)$ and $c' \in L(v) -c$, $c'$ is exactly one of left at $v$ or right at $v$.
+\end{claim}
+\begin{proof}
+	Let $P'$ be the component of $G(c') \cap P$ containing $v$. By Lemma $4.1$, since $P$ is odd, at least one endpoint of $P'$ is a vertex $v_i$ such that $2\leq i\leq k-1$.
+	\paragraph{}
+	Case $1$: $v_i$ is even.
+We want to show the other endpoint of $P'$ is odd and not equal to $v_1$ or $v_k$.
+If the other endpoint of $P'$ is $v_1$ or $v_k$ then $G(c')$ with $G(c)$ contradicts Claim $3.2$ of Problem $3$.
+If the other endpoint of $P'$ is even then $G(c') \subseteq G(c)$ and $|V(G(c'))|$ is odd, contradicting Lemma $4.1$.
+	\paragraph{}
+	Case $2$: $v_i$ is odd.
+We want to show the other endpoint of $P'$ is not an odd vertex $v_j$ with $2\leq j \leq k-1$.
+Suppose not.
+Then $G(c') \subseteq G(c)$ and $|V(G(c'))|$ is odd, contradicting Lemma $4.1$.
+Hence the Claim is proven.
+\end{proof}
+\begin{claim}
+	If $2\leq i \leq k-2$ then $v_i$ is either left or right but not both.
+\end{claim}
+\begin{proof}
+	Let $A$ be the set of left colors at $v_i$ and $B$ be the set of right colors at $v_i$.
+By Claim $4.2$, $(A,B)$ is a partition of $L(v_i) - c$.
+Since $|L(v_i)|$ is even, $|L(v_i)-c|$ is odd, and so exactly one of $|A|$ or $|B|$ is odd as $|A| + |B| = |L(v_i)-c|$.
+\end{proof}
+\begin{claim}
+	All of the following hold: $|V(P)| > 1$, $v_2$ is left, and $v_{k-1}$ is right.
+\end{claim}
+\begin{proof}
+	First suppose for a contradiction that $|V(P)| = 1$.
+Let $c' \in L(G) -c$.
+From Claim $3.1$ $(3.)$ we have that $V(G(c)) \cap V(G(c')) \neq \emptyset$.
+Therefore $G(c) \subseteq G(c')$, but since $|V(G(c))|$ is odd, this contradicts Lemma $4.1$.
+Thus $|V(P)| > 1$.
+So vertices $v_2$ and $v_{k-1}$ exist.
+	\paragraph{}
+	We will show all colors of $L(v_2) - c$ are left.
+A similar argument shows all colors of $L(v_{k-1})-c$ are right, by replacing $v_2$ with $v_{k-1}$, $v_1$ with $v_k$ and interchaning left/right.
+Suppose $v_2$ has a right color $c'$.
+Let $P'$ be the component of $G(c') \cap P$ containing $v_2$.
+Then the leftmost vertex of $P'$ is odd, but not equal to $v_1$.
+But there is no odd vertex between $v_1$ and $v_2$, a contradiction.
+\end{proof}
+\begin{claim}
+	If $2\leq i \leq k-2$ then $v_i$ has the same chirality as $v_{i+1}$.
+\end{claim}
+\begin{proof}
+	TODO
+\end{proof}
+\paragraph{}
+To complete the proof we observe that Claim $4.4$ and Claim $4.5$ taken together yield a contradiction, since Claim $4.5$ shows $v_{2}$ and $v_{k-1}$ have the same chirality, but Claim $4.4$ shows they have different chirality.
+$\blacksquare$
 \end{document}

### 2018-01-31 CO749-GraphColoring/assignments/A2.tex
Finished 4

@@
 \begin{proof}
-	TODO
+	We proceed by induction on $i$.
+If $i = 2$ then we observe that all colors of $L(v_3) - c$ are left for the same reason as for $v_2$, i.e.
+since there are no odd vertices left of $v_3$ that are not $v_1$.
+	\paragraph{}
+	So for induction we suppose $i>2$ and the claim holds for vertices $v_j$ with $2\leq j < i$.
+Then by a simple induction we can show since $v_2$ is left, all of $v_2\dots v_i$ are left.
+If $v_i$ is even then each left color of $v_i$ is a left color of $v_{i+1}$ since its corresponding path component cannot have its rigthmost vertex be $v_i$.
+Further any left color $c'$ of $v_{i+1}$ is a left color of $v_i$.
+To see this, observe by Claim $4.2$ a color is either left or right, and since $v_{i+1}$ is odd, if then $c'$-corresponding path component had leftmost vertex $v_{i+1}$ then $c'$ would be left.
+Hence $v_{i+1}$ has an odd number of left colors and so $v_{i+1}$ is left, as desired.
+	\paragraph{}
+	So we may assume that $v_i$ is odd.
+Then $v_{i+1}$ is even.
+So by a similar argument to the above, $v_i$ and $v_{i+1}$ have the same right colors.
+By Claim $4.3$, $v_i$ is not right so $v_{i+1}$ has an even number of right colors and hence is not right.
+Thus by Claim $4.3$, $v_{i+1}$ is left, as desired.
+By induction the Claim is proven.
 \end{proof}

### 2018-01-31 CO749-GraphColoring/assignments/A2.tex
finished A2

@@
 \begin{proof}
-	TODO
+	Suppose not.
+Then there exists an edge $uv \in E(G)$ such that $L(u) \neq L(v)$.
+From Claim $3.4$ we may assume $L(u) \subset L(v)$.
+Since $|L(u)|$ and $|L(v)|$ are even, this implies there exist distinct colors $c, c' \in L(v)\backslash L(u)$.
+Since $c, c' \not\in L(u)$, $G(c)$ and $G(c')$ have common endpoint $v$.
+Thus we may assume $G(c) \subseteq G(c')$.
+So $P = G(c) \cap G(c') = G(c)$ has an end point in common with $G(c)$ and the other endpoint, $v$, in common with $G(c')$.
+	\paragraph{}
+	By Claim $3.3$, $|V(P)|$ is even, but then the existence of $P$ contradicts Claim $3.2$.
 \end{proof}

### 2018-02-13 CO749-GraphColoring/assignments/A3.tex
git started

--- 2018-01-31 CO749-GraphColoring/assignments/A2.tex
@@
 \lhead{\today}
-\rhead{W. Justin Toth - A2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth - A3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-\section{}
-\paragraph{}
-By Voigt's Theorem there exists a planar graph $H$ and a list assignment $L$ with $|L(u)| \geq 4$ for all $u \in V(H)$, such that $H$ is not $L$-colorable.
-We construct a graph $G$ as follows.
-Take $5$ distinct copies of $H$: $H_1, \dots, H_5$ where each vertex $u_i \in V(H_i)$ corresponds to vertex $u \in V(H)$, and take an additional vertex $v$.
-Add edges $vu_i$ for all $u_i \in V(H_i)$ for all $i \in \{1,\dots, k\}$.
-i.e.
-$v$ is adjacent to every vertex in every copy of $H$.
-\paragraph{}
-To see that $G$ is an apex graph observe that $G-v$ consists of five disjoint copies of a planar graph, and hence $G-v$ is planar.
-\paragraph{}
-We now construct a $5$-list-assignment $L'$ for which $G$ does not have an $L'$-coloring, completing the proof that $G$ is not $5$-list-colorable.
-Let $c_1, \dots, c_5$ be $5$ disjoint colors from $L$.
-That is for all $u \in V(H)$, $$L(u) \cap \{c_1, \dots, c_5\} = \emptyset.$$
-Let $L'(v) = \{c_1, \dots, c_5\}$.
-For each $i \in \{1, \dots, 5\}$ and for each $u_i \in V(H_i)$ let $L'(u_i) = L(u) \cup \{c_i\}$.
-Since $|L(u)| \geq 4$, $|L'(u_i)|\geq 5$.
-Thus we have a $5$-list-assignment $L'$ for $G$.
-\paragraph{}
-Suppose for a constradiction $G$ has an $L'$-coloring $\phi$.
-Say $\phi(v) = c_i$.
-Then $\phi(u_i) \neq c_i$ for all $u_i \in V(H_i)$.
-Hence $\phi(u_i) \in L(u_i)$ for all $u_i \in V(H_i)$.
-That is, $\phi$ restricted to $H_i$ is an $L$-coloring of $H-_i$.
-But this contradicts that $H_i$ is not $L$-colorable.
-Therefore $G$ does not have an $L'$-coloring and hence is not $5$-list-colorable.
-$\blacksquare$
 
-\newpage
-\section{}
-\paragraph{}
-We opt to prove the BONUS Lemma, i.e. Lemma $3$:
-\textit{Let $G$ be a graph and for each vertex $v$ of $G$ let $r(v)$ be a non-negative integer. If $L$ is a list assignment of $G$ such that for every vertex $v$ of $G$
-$$|L(v)| \geq \frac{d(v) + 1}{r(v)+1},$$
-then $G$ has an $r(v)$-defective coloring}. This Lemma directly implies Lemma $2$ of this problem.
-\paragraph{}
-Consider a graph $G$, non-negative integers $r(v)$ for each $v \in V(G)$, and a list assignment $L$ of $G$ such that for each $v \in V(G)$
-$$|L(v)| \geq \frac{d(v) + 1}{r(v)+1}.$$
-Let $\phi$ be an improper $L$-coloring of $G$ minimizing the number of ``improper" edges $uv \in E(G)$ such that $\phi(u) = \phi(v)$.
-Suppose for a contradiction that $\phi$ is not an $r(v)$-defective coloring.
-\paragraph{}
-Then there exists a vertex $v$ such that the number of neighbours of $v$ receiving the same color as $v$ under $\phi$ is at least $r(v) + 1$.
-If there exists a color $c \in L(v)$ such that the number of neighbours of $v$ colored $c$ is at most $r(v)$ then we can construct a new coloring $\phi'$ by setting $\phi'(v) = c$ and $\phi'(u) = \phi(u)$ for all $u \neq v$.
-This coloring has $r(v)$ new improper edges, those between $v$ and its $c$-colored neighbours.
-But $\phi'$ switches $r(v) + 1$ edges which were improper under $\phi$, those between $v$ and its $\phi(v)$-colored neighbours, to proper edges.
-Hence $\phi'$ has fewer improper edges than $\phi$, a contradiction.
-\paragraph{}
-If such a color $c \in L(v)$ does not exist then  for each $c \in L(v)$ the number of neighbours of $v$ colored $c$ under $\phi$ is at least $r(v) + 1$. But then $v$ has at least
-$$|L(v)|(r(v) + 1) \geq d(v) + 1$$
-neighbours, a contradiction. Therefore $\phi$ is an $r(v)$-defective coloring $\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-We prove a slightly more general result allowing $m(v)$ to take the value $0$.
-This will make induction easier when we wish to delete from a list of size $2$.
-For such a $v$, $L(v)= \emptyset$ and the desired $S(v) = \emptyset$.
-Everything else is exactly as stated in the problem.
-\paragraph{}
-Suppose not. Let $G$ be a counterexample satisfying:
-\begin{enumerate}
-	\item $\sum_{v \in V(G)} m(v)$ is minimized, and subject to that
-	\item if possible $G(c)$ is connected for each $c \in L(G)$, and subject to that possibility
-	\item $|L(G)|$ is minimized.
-\end{enumerate}
-\begin{claim}
-	\begin{enumerate}
-		\item For all $c \in L(G)$, $G(c)$ is connected.
-		\item For every pair of distinct colors $c,c' \in L(G)$, $G(c) \neq G(c')$.
-		\item For every pair of distinct colors $c, c' \in L(G)$, we have $V(G(c)) \cap V(G(c')) \neq \emptyset$ (we strengthened this claim).
-	\end{enumerate}
-\end{claim}
-\begin{proof}
-	\paragraph{} For (1.). While there exists a color $c \in L(G)$ such that $G(c)$ has at least two connected components $G_1$ and $G_2$ do the following:
-		\subparagraph{}
-		Let $c'$ be a new color not contained in $L(G)$.
-For each $v \in V(G_2)$ let $$L'(v) = (L(v)\cup \{c'\}) \backslash \{c\}.$$ For each $v \in V(G) \backslash V(G_2)$ let $L'(v) = L(v)$.
-Replace $L$ with $L'$.
-	\paragraph{}
-	We claim that every iteration of the above preserves the property of $(G,m,L)$ being a counterexample, but decreases the ``disconnectivity".
-i.e.
-the sum over $c \in L(G)$ such that $G(c)$ is disconnected, of the number of connected components in $G(c)$.
-	\paragraph{}
-	It is clear the disconnectivity decreases.
-We add a new color $c'$ for which $G(c')$ is connected, and $G(c)$ loses one connected component, and the number of connected components of $G(c'')$ is not affected for any other color $c'' \in L(G)$.
-	\paragraph{}
-	To verify that the new lists, say $L'$, also admit a counterexample with $G$ and $m$, suppose to contrary there are $S'(v) \subseteq L'(v)$ of size $m(v)$ for each $v \in V(G)$ and $S'(u) \cap S'(v) = \emptyset$ for all $uv \in E(G)$.
-For each $v \in V(G)$ let $$S(v) = \begin{cases} S'(v)\cup \{c\} \backslash \{c'\}, &\text{if } v \in V(G_2) \text{ and } c' \in S'(v) \\
-	S'(v), &\text{otherwise}.\end{cases} $$
-	Then $S(v) \subseteq L(v)$ and $|S(v)| = m(v)$ for each $v \in V(G)$.
-Further, for each $uv \in E(G)$, $S(u) \cap S(v) = \emptyset$.
-To see this, observe that if $S(u) \cap S(v) \neq \emptyset$ then by our choice of $S'$, $S(u) \cap S(v) = \{c\}$ and, since $G(c)$ was disconnected, $u, v \in V(G_2)$.
-But then, $S'(u) \cap S'(v) = \{c'\}$, a contradiction.
-	\paragraph{}
-	So the above algorithm will terminate when the ``disconnectivity" reaches $0$ and we will have found a connected counterexample with the same value for $\sum_{v} m(v)$ as $G$.
-This contradicts our choice of $G$ if there exists color $c$ such that $G(c)$ is disconnected.
-Therefore $G(c)$ is connected for all $c \in L(G)$.
-	\paragraph{}For (2.). Suppose for a contradiction there exist distinct colors $c,c' \in L(G)$ such that $G(c) = G(c')$. Construct new lists $L'$ such that for each $v \in V(G)$
-	$$ L'(v) = \begin{cases}
-	L(v) \backslash \{c,c'\}, &\text{ if } v \in V(G(c)) \\
-	L(v), &\text{otherwise.}
-		\end{cases}$$
-	Let $m'(v) = m(v) - 1$ if $v \in V(G(c))$ (this is well-defined since such $m(v) \geq 2$) and let $m'(v) = m(v)$ otherwise.
-By the minimality of $G$, there exist for each $v \in V(G)$, a set $S'(v)\subseteq L'(v)$ of size $m'(v)$.
-These sets satisfy $S'(u) \cap S'(v) = \emptyset$ for all $uv \in E(G)$.
-Order the vertices of $G(c)$ as $v_1, \dots, v_k$ with respect to some orientation of $G$.
-Construct sets $S(v)$ as follows.
-If $v \in V(G) \backslash V(G(c))$ set $S(v) = S'(v)$.
-If $v_i \in V(G(c))$ and $i$ is even set $S(v) = S'(v) \cup \{c\}$.
-If $v_i \in V(G(c))$ and $i$ is odd set $S(v) = S'(v) \cup \{c'\}$.
-Since $c$ and $c'$ do not appear in sets outside $G(c)$, and we alternate $c$ and $c'$ along $G(c)$, we have $S(u) \cap S(v) = \emptyset$ for all $uv \in E(G)$.
-But this contradicts that $(G,m,L)$ is a counterexample.
-Hence for distinct colors $c$ and $c'$, $G(c) \neq G(c')$.
-	\paragraph{} For (3.).
-Suppose for a contradiction  there exist distinct colors $c,c' \in L(G)$, such that $V(G(c)) \cap V(G(c')) = \emptyset$.
-Construct new lists $L'(v)$ for each $v \in V(G)$ as follows.
-If $v \in V(G) \backslash V(G(c'))$ then set $L'(v) = L(v)$.
-Otherwise, i.e.
-$v \in V(G(c'))$ set $L'(v) = L(v) \cup \{c\} \backslash \{c'\}.$ Then for each $v \in V(G(c'))$, we have $|L'(v)|  = |L(v)| + 1 - 1 = m(v)$, since $V(G(c))$ and $V(G(c'))$ are disjoint.
-But $L'(G) = L(G) \backslash \{c\}$ so by minimality there exist sets $S'(v) \subseteq L'(v)$ for each $v \in V(G)$ such that $|S'(v)| = m(v)$, satisfying $S'(u) \cap S'(v) = \emptyset$ for each $uv \in E(G)$.
-	\paragraph{}
-	For each $v \in V(G)$ construct $S(v)$ so that
-	$$S(v) = \begin{cases}
-		S'(v) \cup \{c'\}\backslash \{c\}, &\text{ if } v \in V(G(c')) \text{ and } c \in S'(v) \\
-		S'(v), &\text{ otherwise}.
-	\end{cases}$$
-	Then each set $S(v) \subseteq L(v)$ is of size $m(v)$ and for each edge $uv \in E(G)$, $S(u) \cap S(v) = \emptyset$.
-To verify disjointness, suppose for a contradiction $S(u) \cap S(v) \neq \emptyset$ for some edge $uv$.
-Since $S'(u) \cap S'(v) = \emptyset$, $S(u) \cap S(v) = \{c'\}$.
-But then $u,v \in V(G(c'))$ and so $S'(u) \cap S(v) = \{c\}$ since $c'$ is only added to $S(u), S(v)$ if $c \in S'(u), S'(v)$.
-But this contradicts disjointness of $S'(u), S'(v)$.
-So sets $S(v)$ for $v \in V(G)$ violate that $(G,m,L)$ is a counterexample, a contradiction.
-\end{proof}
-	\begin{claim}
-		If $c \neq c' \in L(G)$ such that $G(c), G(c')$ are paths, then there does not exist a component $P$ of $G(c) \cap G(c')$ with an even number of vertices such that one end of $P$ is an end of $G(c)$ and the other end of $P$ is an end of $G(c')$.
-	\end{claim}
-	\begin{proof}
-		\paragraph{} Suppose for a contradiction such $P$ exists.
-Order the vertices of $P$ as $v_1, \dots, v_k$ so that $v_1$ is an end of $G(c)$ and $v_k$ is an end of $G(c')$.
-Construct new lists $L'(v)$ for each $v \in V(G)$ as follows.
-If $v \in V(P)$ let $L'(v) = L(v) \backslash \{c,c'\}$ and let $m'(v) = m(v) - 1$.
-Otherwise let $L'(v) = L(v)$ and $m'(v) = m(v)$.
-By minimality, there exist sets $S'(v) \subseteq L'(v)$ for each $v \in V(G)$ of size $m'(v)$, such that $S'(u) \cap S'(v) = \emptyset$ for each $uv \in E(G)$.
-		\paragraph{}
-		Construct sets $S(v)\subseteq L(v)$ for each $v \in V(G)$ as follows.
-If $v_i \in V(P)$ and $v_i$ is odd then set $S(v) = S'(v) \cup \{c\}$.
-If $v_i \in V(P)$ and $v_i$ is even set $S(v) = S'(v) \cup \{c'\}$.
-If $v \in V(G)\backslash V(P)$ set $S(v) = S'(v)$.
-Then $|S(v)| = m(v)$ for each $v \in V(G)$.
-Since we alternate adding $c$ and $c'$ along $P$ the sets $S(u)$ and $S(v)$ are disjoint for each edge $uv \in E(G)$.
-Since we did not change $S(u)$ and $S(v)$ for edges $uv$ with neither endpoint on $P$ they are also disjoint.
-It remains to consider two edges $u_1v_1$ and $v_ku_k$ where $u_1, u_k \not\in V(P)$.
-Since $v_1$ is an end of $G(c)$ and $v_2 \in G(c)$ we have $c \not\in L(u_1)$ and hence, as $S'(u_1) \cap S'(v_1) = \emptyset$, $S(u_1)$ and $S(v_1)$ are disjoint.
-Similarly since $v_k$ is an edge of $G(c')$, and $k$ is even, $S(u_k)$ and $S(v_k)$ are disjoint.
-But then the sets $S(v)$ contradict that $(G,m,L)$ is a counterexample.
-	\end{proof}
-\begin{claim}
-	If $c \in L(G)$ then $|V(G(c))|$ is even.
-\end{claim}
-\begin{proof}
-	Deferred to Problem $4$.
-\end{proof}
-\begin{claim}
-	For every edge $e=uv \in E(G)$, $L(u) \subseteq L(v)$ or $L(v) \subseteq L(u)$.
-\end{claim}
-\begin{proof}
-	\paragraph{}
-	Suppose not.
-Then there exists an edge $e=uv \in E(G)$ such that $L(u)\backslash L(v) \neq \emptyset$ and $L(v)\backslash L(u) \neq \emptyset$.
-Let $c_u \in L(u) \backslash L(v)$ and let $c_v \in L(v)\backslash L(u)$.
-By our previous claims, both $G(c_u)$ and $G(c_v)$ are connected and non-disjoint, so they intersect in a path.
-Let $P = G(c_u) \cap G(c_v)$.
-Since $v \not\in G(c_u)$ (also note $u \not\in G(c_v)$), $P \neq G$, and we have $G(c_u) \cup G(c_v) = G$.
-Further one end of $P$ is in $G(c_u)$ and the other is in $G(c_v)$.
-Thus by Claim $3.2$, $|V(P)|$ is odd.
-	\paragraph{}
-	Since $|V(G)|$ is even,  and $|V(P)|$ is odd, we have that $|V(G) \backslash V(P)|$ is odd.
-Then since $V(G)\backslash V(P)$ is the disjoint union of $V(G(c_u))\backslash V(P)$ and $V(G(c_v)) \backslash V(P)$, precisely one of $|V(G(c_u))\backslash V(P)|$ or $|V(G(c_v))\backslash V(P)|$ is even and the other is odd.
-Say without loss of generality that $|V(G(c_u))\backslash V(P)|$ is even.
-Then $|V(G(c_u))| = |V(G(c_u))\backslash V(P)| + |V(P)|$ is odd.
-But this contradicts Claim $3.3$.
-	\end{proof}
-\begin{claim}
-For every edge $e=uv \in E(G)$, $L(u) = L(v)$.
-\end{claim}
-\begin{proof}
-	TODO
-\end{proof}
-\paragraph{}
-We are now ready to complete the main proof.
-The argument is essentially the same as $2$-colorability of even cycles.
-From Claim $3.5$, let $L$ denote the constant list assigned to each vertex, i.e.
-$L = L(u)$ for each $u \in V(G)$.
-We know $|L|$ is even and we need to assign half its elements to each vertex.
-\paragraph{}
- Let $S_1 \subseteq L$ be a subset of size $|L|/2$.
-Let $S_2 = L \backslash S_1$.
-Then $|S_1| = |S_2| = m(v)$ for any $v \in V(G)$.
-Enumerate the vertices about $G$ as $v_1, \dots, v_k$ with $k$ even.
-For each $v_i \in V(G)$, let $S(v) = S_1$ is $i$ is odd and $S(v) = S_2$ is $i$ is even.
-Since $G$ is an even cycle for any edge $v_iv_j \in E(G)$ exactly one of $i,j$ is odd and the other is even.
-Hence $S(v_i) \cap S(v_j) = \emptyset$, contradicting that $G$ is a counterexample.$\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-We will prove Claim $3.3$ from the previous problem.
-Suppose the Claim is false.
-Then there exists $c \in L(G)$  such that $G(c)$ is a path $P=v_1\dots v_k$ with $k$ odd.
-Choose such $c$ so that $|V(P)|$ is minimized.
-We adopt the same notation and definitions as the problem statement.
-\begin{lemma}
-	For every $a,b \in L(G)$, if $|V(G(a))|$ is odd then we do not have $G(a) \subseteq G(b)$.
-\end{lemma}
-\begin{proof}
-	Suppose not.
-Then delete colors $a$ and $b$ from $L(v)$ for each $v \in V(G(a))$ to obtain lists $L'(v)$.
-For vertices $v$ off $G((a))$ let $L'(v) = L(v)$.
-By minimality, there exist $S'(v) \subseteq L'(v)$, such that $|S'(v)| = \frac{1}{2}|L'(v)|$ for each $v \in V(G)$, and for each edge $uv \in E(G)$, $S'(u) \cap S'(v) = \emptyset$.
-	\paragraph{}
-	Enumerate the vertices of $G(a)$ as $v_1 \dots v_\ell$ with $\ell$ odd.
-For each $v_i \in V(G(a))$ let $S(v_i) = S'(v_i) \cup \{a\}$ if $i$ is odd, and $S(v_i) = S'(v_i) \cup \{b\}$ if $i$ is even.
-For each $v \in V(G)\backslash V(G(a))$ let $S(v) = S'(v)$.
-Then $|S(v)| = \frac{1}{2} |L(v)|$ for each $v \in V(G)$.
-	\paragraph{}
-	Let $uv \in E(G)$.
-If $u,v \not\in V(G(a))$ then $S(u) \cap S(v) = S'(u) \cap S'(v) = \emptyset$.
-If $u,v \in V(G(a))$ then $S(u) \cap S(v) = S'(u) \cap S'(v) \cap \{a\} \cap \{b\} = \emptyset$.
-If $u \in V(G(a))$ and $v \not\in V(G(a))$ then either $u = v_1$ or $u = v_\ell$.
-In either case $S(u) = S'(u) \cup \{a\}$.
-Since $v \not\in V(G(a))$, $a \not\in L(v)$ and so $S(u) \cap S'(v) = \emptyset$ as desired.
-But then the sets $S(v)$ contradict that $G$ is a counterexample.
-\end{proof}
-\begin{claim}
-	For every $v \in V(P)$ and $c' \in L(v) -c$, $c'$ is exactly one of left at $v$ or right at $v$.
-\end{claim}
-\begin{proof}
-	Let $P'$ be the component of $G(c') \cap P$ containing $v$. By Lemma $4.1$, since $P$ is odd, at least one endpoint of $P'$ is a vertex $v_i$ such that $2\leq i\leq k-1$.
-	\paragraph{}
-	Case $1$: $v_i$ is even.
-We want to show the other endpoint of $P'$ is odd and not equal to $v_1$ or $v_k$.
-If the other endpoint of $P'$ is $v_1$ or $v_k$ then $G(c')$ with $G(c)$ contradicts Claim $3.2$ of Problem $3$.
-If the other endpoint of $P'$ is even then $G(c') \subseteq G(c)$ and $|V(G(c'))|$ is odd, contradicting Lemma $4.1$.
-	\paragraph{}
-	Case $2$: $v_i$ is odd.
-We want to show the other endpoint of $P'$ is not an odd vertex $v_j$ with $2\leq j \leq k-1$.
-Suppose not.
-Then $G(c') \subseteq G(c)$ and $|V(G(c'))|$ is odd, contradicting Lemma $4.1$.
-Hence the Claim is proven.
-\end{proof}
-\begin{claim}
-	If $2\leq i \leq k-2$ then $v_i$ is either left or right but not both.
-\end{claim}
-\begin{proof}
-	Let $A$ be the set of left colors at $v_i$ and $B$ be the set of right colors at $v_i$.
-By Claim $4.2$, $(A,B)$ is a partition of $L(v_i) - c$.
-Since $|L(v_i)|$ is even, $|L(v_i)-c|$ is odd, and so exactly one of $|A|$ or $|B|$ is odd as $|A| + |B| = |L(v_i)-c|$.
-\end{proof}
-\begin{claim}
-	All of the following hold: $|V(P)| > 1$, $v_2$ is left, and $v_{k-1}$ is right.
-\end{claim}
-\begin{proof}
-	First suppose for a contradiction that $|V(P)| = 1$.
-Let $c' \in L(G) -c$.
-From Claim $3.1$ $(3.)$ we have that $V(G(c)) \cap V(G(c')) \neq \emptyset$.
-Therefore $G(c) \subseteq G(c')$, but since $|V(G(c))|$ is odd, this contradicts Lemma $4.1$.
-Thus $|V(P)| > 1$.
-So vertices $v_2$ and $v_{k-1}$ exist.
-	\paragraph{}
-	We will show all colors of $L(v_2) - c$ are left.
-A similar argument shows all colors of $L(v_{k-1})-c$ are right, by replacing $v_2$ with $v_{k-1}$, $v_1$ with $v_k$ and interchaning left/right.
-Suppose $v_2$ has a right color $c'$.
-Let $P'$ be the component of $G(c') \cap P$ containing $v_2$.
-Then the leftmost vertex of $P'$ is odd, but not equal to $v_1$.
-But there is no odd vertex between $v_1$ and $v_2$, a contradiction.
-\end{proof}
-\begin{claim}
-	If $2\leq i \leq k-2$ then $v_i$ has the same chirality as $v_{i+1}$.
-\end{claim}
-\begin{proof}
-	TODO
-\end{proof}
-\paragraph{}
-To complete the proof we observe that Claim $4.4$ and Claim $4.5$ taken together yield a contradiction, since Claim $4.5$ shows $v_{2}$ and $v_{k-1}$ have the same chirality, but Claim $4.4$ shows they have different chirality.
-$\blacksquare$
 \end{document}

### 2018-02-16 CO749-GraphColoring/assignments/A3.tex
Most of q1

@@
 \begin{document}
-
+\section{}
+A $k$-critical graph $G$ is $m$-\emph{splittable} (with $2\le m \le k-1$) if for every vertex $z\in V(G)$ and every split of $z$ into vertices $z_1,\ldots, z_m$ of positive degree , the resulting graph has a $(k-1)$-coloring such that $z_1, \ldots, z_m$ have distinct colors.
+We say such $G$ is \emph{splittable} if $G$ is $m$-splittable for every $m$ such that $2\le m \le k-1$.
+\begin{lemma}
+	An Ore-composition of two splittable $k$-critical graphs is splittable $k$-critical.
+\end{lemma}
+\begin{proof}
+	Let $G_1$ and $G_2$ be two splittable $k$-critical graphs.
+Let $G$ be the Ore-composition of $G_1$ and $G_2$ obtained by deleting edge $xy \in E(G_1)$ and splitting vertex $z \in V(G_2)$ into $z_1$ and $z_2$ of positive degree and identifying $x$ with $z_1$ and $y$ with $z_2$.
+We denote the identified vertex for $x$ with $z_1$ as $z_x$, and for $y$ with $z_2$ as $z_y$ in $G$.
+	\paragraph{}
+	We first show that $G$ is $k$-critical.
+Suppose for a contradiction $G$ has a $(k-1)$-coloring $\phi$.
+If $\phi(z_x) \neq \phi(z_y)$ then $\phi$ restricted to $G_1$ (taking in the obvious manner $\phi(x) = \phi(z_x)$ and $\phi(y) = \phi(z_y)$) is a $(k-1)$-coloring of $G_1$.
+If $\phi(z_x) = \phi(z_y)$ then $\phi$ restricted to $G_2$, and takiing $\phi(z) = \phi(z_x) = \phi(z_y)$ yields a $(k-1)$-coloring of $G_2$.
+In either case we contradicted that $G_1$ and $G_2$ do not have a $(k-1)$-coloring.
+Therefore $G$ is not $(k-1)$-colorable.
+	\paragraph{}
+	To finish verifying that $G$ is $k$-critical we need to show that every strict subgraph of $G$ is $(k-1)$-colorable.
+Let $H \subseteq G$ be a strict subgraph of $G$.
+We proceed by case-distinction on $H$.
+		\subparagraph{} Case $1$: $z_x$ is not contained in $H$ and $z_y$ is not contained in $H$.
+Then $H$ is the disjoint union of a strict subgraph of $G_1$ and a strict subgraph of $G_2$.
+Since $G_1$ and $G_2$ are $k$-critical we can obtain a $(k-1)$-coloring of their respective strict subgraphs in $H$ and combine them to obtain a $(k-1)$-coloring of $H$.
+		\subparagraph{} Case $2:$ Exactly one of $z_x, z_y$ is in $H$.
+Without loss of generality say $z_x \in V(H)$ and $z_y \not\in V(H)$.
+Then treating $z_x$ as $x$ in $G_1$ and $z$ in $G_2$ we see that $H$ is the union of two strict subgraphs $H_1 \subseteq G_1$ and $H_2 \subseteq G_2$, overlapping only at $z_x$.
+We see $H_2$ is a strict subgraph since the degree of $z_1$ and $z_2$ are both positive.
+Again we invoke criticality of $G_1$ and $G_2$ to obtain $(k-1)$-colorings of $H_1$ and $H_2$, which we join into a $(k-1)$-coloring of $H$.
+Note we permute the coloring of $H_1$ so the two colorings agree at $z_x$ if necessary.
+		\subparagraph{} Case $3:$.
+Both $z_x$ and $z_y$ are in $H$.
+Let $G_2'$ denote the graph $G_2$ with $z$ split into $z_1$ and $z_2$.
+Then $H$ is the union of two subgraphs $H_1 \subseteq G_1$ and $H_2 \subseteq G_2'$ overlapping only at $\{z_x,z_y\}$.
+Since $H$ is a strict subgraph of $G$ one of $H_1$ or $H_2$ is a strict subgraph of $G_1-xy$ or $G_2'$ respectively.
+		\subparagraph{} First suppose $H_1$ is a strict subgraph of $G_1-xy$.
+Then invoking the criticality of $G_1$, we obtain that $H_1 + xy$ has a $(k-1)$-coloring $\phi_1$.
+Then $\phi_1$ is a $(k-1)$-coloring of $H_1$ where $\phi_1(x)\neq \phi_1(y)$.
+Since $G_2$ is $2$-splittable, there exists a $(k-1)$-coloring $\phi_2$ of $G_2'$.
+Note that $\phi_2(z_1) \neq \phi_2(z_2)$ for otherwise setting $\phi_2(z) = \phi_2(z_1) = \phi_2(z_2)$ yields a $(k-1)$-coloring of $G_2$.
+Hence we may permute the colors of $\phi_2$ so $\phi_2(z_1) = \phi_1(x)$ and $\phi_2(z_2) = \phi_1(y)$.
+Then the coloring obtained by combining $\phi_1$ and $\phi_2$ is a $(k-1)$-coloring of $H$.
+		\subparagraph{}
+		So we may assume $H_1$ is not a strict subgraph of $G_1-xy$, i.e.
+$H_1 = G_1-xy$.
+Then $H_2$ is a strict subgraph of $G_2'$.
+By identifying $z_x$ and $z_y$ as $z$ in $H_2$ we obtain a strict subgraph of $G_2$.
+By criticality of $G_2$, this subgraph has a $(k-1)$-coloring $\phi_2$.
+By setting $\phi_2(z_x) = \phi_2(z_y) = \phi_2(z)$ we obtain a $(k-1)$-coloring of $H_2$ wherein $z_x$ and $z_y$ receive the same color.
+Invoking criticality of $G_1$, we obtain a $(k-1)$-coloring $\phi_1$ of $H_1$ ($=G_1-xy$).
+Observe that $\phi_1(x) = \phi_1(y)$, for otherwise $\phi_1$ is a $(k-1)$-coloring of $G_1$.
+But then we may permute the colors of $\phi_2$ so $\phi_2(z_1) = \phi_1(x)$ and $\phi_2(z_2) = \phi_1(y)$.
+Again the coloring obtained by combining $\phi_1$ and $\phi_2$ is a $(k-1)$-coloring of $H$, as desired.
+	\paragraph{Splittability}
+	Thus we have shown that $G$ is $k$-critical. It remains to verify that $G$ is splittable.
+\end{proof}
 \end{document}

### 2018-02-16 CO749-GraphColoring/assignments/A3.tex
finished a1

@@
 	\paragraph{Splittability}
-	Thus we have shown that $G$ is $k$-critical. It remains to verify that $G$ is splittable.
+	Thus we have shown that $G$ is $k$-critical.
+It remains to verify that $G$ is splittable.
+First consider $v \in V(G) \backslash\{z_x,z_y\}$.
+Consider $G'$, where we split $v$ into $v_1, \dots, v_m$ each with positive degree, $2 \leq m \leq k-1$.
+If $v \in V(G_1)$ then apply the corresponding split $v_1, \dots, v_m$ to $G_1$, and use that $G_1$ is splittable to obtain a $(k-1)$-coloring $\phi_1$.
+Observe $\phi_1(x) \neq \phi_1(y)$ since $xy \in E(G_1)$.
+Since $G_2$ is $2$-splittable, $G - G_1$ has a $(k-1)$-coloring $\phi_2$.
+Since $G_2$ is not $(k-1)$-colorable, $\phi_2(z_x) \neq \phi_2(z_y)$.
+Again we may permute colors of $\phi_2$ so $\phi_1$ and $\phi_2$ agree on $z_x$ and $z_y$, obtaining a $(k-1)$-coloring of $G'$.
+	\paragraph{}
+	If $v \in V(G_2)$ then apply the corresponding split $v_1, \dots, v_m$ to $G_2$ and use splittability to obtain a $(k-1)$-coloring $\phi_2$.
+Setting $\phi_2(z_x) = \phi_2(z_y) = \phi_2(z)$ we obtain a $(k-1)$-coloring of $G-G_1$.
+By criticality, $G_1-xy$ has a $(k-1)$-coloring $\phi_1$.
+Since $G_1$ is not $(k-1)$-colorable, $\phi_1(x) = \phi_1(y)$.
+Hence we may permute colors of $\phi_2$ so $\phi_1$ and $\phi_2$ agree on $z_x$ and $z_y$, obtaining a $(k-1)$-coloring of $G'$.
+	\paragraph{}
+	It remains to consider splitting one $z_x$ or $z_y$ in $G$.
+Without loss of generality let $G'$ be the graph obtained by splitting $z_x$ into $z_x^1, \dots, z_x^m$, each with positive degree, $2 \leq m \leq k-1$.
+Split $x$ in $G_1$ into $x^1, \dots, x^m$ with neighbourhoods $N(x^i) = N(z_x^i)\cap V(G_1)$ yielding graph $G_1'$.
+Obtain a $(k-1)$-coloring $\phi_1$ of $G_1'$ by splittability of $G$.
+	\subparagraph{}Case $1:$ $\phi_1(x^i) = \phi_1(y)$ for some $i \in \{1,\dots, m\}$.
+Split $z$ in $G_2$ into $z^1,\dots, z^m$ with neighbourhoods $N(z^j) = N(z_x^j) \cap V(G_2)$ for $j \neq i$ and $$N(z^i) = (N(z_x^i) \cup N(z_y)) \cap V(G_2)$$
+	obtaining graph $G_2'$.
+Color $G_2'$ by splittability of $G_2$ obtaining $(k-1)$-coloring $\phi_2$.
+Setting $\phi_2(z_y) = \phi_2(z^i)$ we obtain a $(k-1)$-coloring of $(G-G_1) \cup \{z_x^1, \dots, z_x^m, z_y\}$.
+Since $\phi_1$ and $\phi_2$ give all the split vertices different colors, and give $z_y$ the same color $z_x^i$ we may permute $\phi_1$ and $\phi_2$ to agree on $\{z_x^1, \dots, z_x^m, z_y\}$ obtaining a $(k-1)$-coloring of $G'$.
+	\subparagraph{}Case $2:$ $\phi_1(x^i) \neq \phi_2(y)$ for all $i \in \{1,\dots, m\}$.
+Notice in this case $m < k-1$ by the pidgeonhole principle since $\phi_1$ assigns different colors to each split vertex.
+Split $z$ in $G_2$ into $z^1, \dots, z^m, z^{m+1}$ with neighbours $N(z^i) = N(z_x^i) \cap V(G_2)$ for $i \leq m$ and set $N(z^{m+1}) = N(z_y) \cap V(G_2)$.
+Applying splitability we obtain a $(k-1)$-coloring $\phi_2$.
+Now both $\phi_1$ and $\phi_2$ give $\{z_x^1, \dots, z_x^m, z_y\}$ different colors, so we may permute their colors to agree on $\{z_x^1, \dots, z_x^m, z_y\}$, and join the colorings to obtain a coloring of $G'$, as desired.
 	\end{proof}

### 2018-02-17 CO749-GraphColoring/assignments/A3.tex
2ab

@@
 	\end{proof}
+\begin{corollary}
+	If $G$ is a $k$-Ore graph then $G$ is splittable $k$-critical.
+\end{corollary}
+\begin{proof}
+	Suppose not.
+Let $G$ be a minimum counterexample.
+Then $G$ cannot be the Ore-composition of two $k$-Ore graphs $G_1$, $G_2$, for otherwise $G_1$ and $G_2$ are splittable $k$-critical by minimality and hence $G$ is splittable $k$-critical by the previous Lemma.
+	\paragraph{}
+	Thus $G$ is isomorphic to $K_k$.
+It remains to show $K_k$ is splittable.
+Suppose we want to split vertex $v \in V(G)$ into $v_1, \dots, v_m$ with $2\leq m \leq k-1$ such that each $v_i$ has positive degree, obtaining graph $G'$.
+We can trivially color $G - v$ with $k-1$ colors, obtaining coloring $\phi$.
+Now we can extend $\phi$ to $G'$ since each $v_i$ has degree at most $k-2$ and hence $|\phi(N(v_i))| \leq k-2$ and there is a free color available to assign $v_i$.
+This contradicts that
+\end{proof}
+
+\section{}
+\subsection{a}
+\begin{lemma}
+	If $G$ is the Ore-composition of two graphs $G_1$ and $G_2$ then $\alpha(G) \leq \alpha(G_1) + \alpha(G_2)$.
+\end{lemma}
+\begin{proof}
+Let $G$ be the Ore-composition of graphs $G_1$ and $G_2$ obtained by deleting edge $xy \in E(G_1)$ and splitting vertex $z \in V(G_2)$ into $z_1$ and $z_2$ of positive degree and identifying $x$ with $z_1$ and $y$ with $z_2$.
+We denote the identified vertex for $x$ with $z_1$ as $z_x$, and for $y$ with $z_2$ as $z_y$ in $G$.
+Let $S \subseteq V(G)$ be an independent set in $G$ such that $|S| = \alpha (G)$.
+Then treating $z_x$ as $x$ and $z_y$ as $z$ we can partition $S$ into two independent sets $S_1\subseteq V(G_1)$ and $S_2\subseteq V(G_2)$ as $S_1 = V(G_1) \cap S$ and $S_2 = S\backslash S_1$. Thus
+$$\alpha(G) = |S| = |S_1| + |S_2| \leq \alpha(G_1) + \alpha(G_2)$$
+as desired.
+\end{proof}
+
+\subsection{b}
+\begin{lemma}
+	If $G$ is $k$-Ore then $\alpha(G) = \frac{v(G) - 1}{k-1}$.
+\end{lemma}
+\begin{proof}
+	We proceed by induction on $v(G)$. The smallest $k$-Ore graph is $K_k$, for which
+	$$\alpha(K_k) = 1 = \frac{k-1}{k-1} = \frac{v(G) -1}{k-1}.$$
+	So the base case for induction is satisfied.
+	\paragraph{}
+	So we consider a $k$-Ore graph $G$ that is the Ore-composition of $G_1$ and $G_2$ and suppose for induction that
+	$$\alpha(G_1) = \frac{v(G_1) - 1}{k-1} \quad\text{ and }\quad \alpha(G_2) = \frac{v(G_2) -1}{k-1}.$$
+	We observe from the previous Lemma, and induction, that
+	$$\alpha(G) \leq \alpha(G_1) + \alpha(G_2) = \frac{v(G_1) +v(G_2)- 2}{k-1} = \frac{v(G) + 1 - 2}{k-1} = \frac{v(G) - 1}{k-1}.$$
+	Thus $\alpha(G) \leq \frac{v(G) - 1}{k-1}$. To show $\alpha(G) \geq \frac{v(G) - 1}{k-1}$ it suffices to construct an independent set in $G$ of size $\frac{v(G) - 1}{k-1}$.
+	\paragraph{}
+	For $i \in \{1, \dots, 2\}$ let $S_i$ be in an independent set in $G_i$ satisfying $$|S_i| = \alpha(G_i) = \frac{v(G_i) - 1}{k-1}.$$ We use the same notion as the above problems for $x,y, z, z_x, z_y$.
+Since $xy \in E(G_1)$, $|\{x,y\} \cap S_1| \leq 1$.
+Without loss of generality suppose $y \not\in S_1$.
+Treating $x$ as $z_x$ and $z$ as $z_y$ we observe that
+	$$S := S_1 \cup S_2$$
+	is an independent set of $G$ of size $|S_1| + |S_2|$. But then we see that
+	$$|S| = |S_1| + |S_2| =  \frac{v(G_1) +v(G_2)- 2}{k-1} = \frac{v(G) - 1}{k-1}.$$
+	Thus $\alpha(G) \geq \frac{v(G) - 1}{k-1}$, and therefore by induction the Lemma is proven.
+\end{proof}
 \end{document}

### 2018-02-17 CO749-GraphColoring/assignments/A3.tex
Solved p2

@@
 \end{proof}
+
+\subsection{c}
+\paragraph{}
+From our work in class it suffices to show $c\alpha(R') \leq c\alpha(R) + c\alpha(W) - c\alpha(K_{|X|}) + c$ for $c \geq 0$.
+Since we know the core is always non-empty, $\alpha(K_{|X|}) = 1$ and so the problem reduces to showing
+$$\alpha(R') \leq \alpha(R) + \alpha(W).$$
+Let $S$ be an independent set of $G[R']$ such that $|S| = \alpha(R')$.
+Since $R'= V(W) \cup R - \{v_1, \dots, v_{k-1}\}$, where $\{v_1, \dots, v_{k-1}\}$ are the set of color-identified vertices, as in the usual definition.
+So we have
+$$|S| \leq |S\cap (V(W) - \{v_1, \dots, v_k\}| + |S\cap R|$$
+On the one hand $S\cap (V(W) - \{v_1, \dots, v_k\})$ is an independent set in $W$ and $S \cap R$ is an indepenent set in $G[R]$, so then
+$$ \alpha(R') = |S| \leq |S\cap (V(W) - \{v_1, \dots, v_k\}| + |S\cap R| \leq \alpha(W) + \alpha(R)$$
+as desired. $\blacksquare$
 \end{document}

### 2018-02-20 CO749-GraphColoring/assignments/A3.tex
Update A3.tex

@@
 as desired. $\blacksquare$
+\section{}
+\subsection{a}
+\begin{lemma}
+If $G$ is the Ore-composition of $G_1$ and $G_2$ then
+$$T_k(G) \geq T_k(G_1-e) + T_k(G_2 - z) \geq T_k(G_1) + T_k(G_2) - 2$$
+where $e = xy$ is the replaced edge of $G_1$ and $z$ is the split vertex of $G_2$.
+\end{lemma}
+\begin{proof}
+Let $H_1 \subseteq G_1- e$ be the subgraph of $G_1 - e$ which achieves the optimum in 
+$$T_k(G_1-e) := \max_{H \subseteq G_1 - e} T_k(H).$$
+Similarly let $H_2 \subseteq G_2 - z$ be the subgraph of $G_2 - z$ which achieves the optimum in $T_k(G_2 - z)$. Then $H_1 \cup H_2$ is a disjoint union of cliques in $G$. So we have
+$$T_k(G_1-e) + T_k(G_2 - z) = T_k(H_1) + T_k(H_2) = T_k(H_1 \cup H_2) \leq T_k(G).$$
+So the first inequality of the lemma is proven.
+\paragraph{}
+Now let $H_1' \subseteq G_1$ be the subgraph of $G_1$ which achieves the optimum in $T_k(G_1)$, and let $H_2' \subseteq G_2$ be the subgraph of $G_2$ which achieves the optimum in $T_k(G_2)$.
+If $|e \cap V(H'_1)| \leq 1$ then $H'_1$ is a subgraph of $G_1 - e$ and so $T_k(G_1) \leq T_k(G_1-e)$.
+Otherwise, i.e.
+$ e\in E(H'_1)$ then $H'_1 - x$ is a subgraph of $G_1 - e$, and $T_k(H'_1-x) = T_k(H'_1) - 1$.
+Thus
+$$T_k(G_1 - e) \geq T_k(H'_1-x) \geq T_k(H'_1) - 1 =T_k(G_1) - 1.$$
+Similarly $H'_2 - z$ is a subgraph of $G_2 - v$ and $T_k(H'_2-z) \geq T_k(H'_2) - 1$. So 
+$$T_k(G_2 - z) \geq T_k(H'_2 - z) \geq T_k(H'_2) - 1 = T_k(G_2) - 1.$$
+Summing our inequalities above we obtain
+$$T_k(G_1 - e) + T_k(G_2 - z) \geq T_k(G_1) + T_k(G_2) - 2$$
+as desired.
+\end{proof}
+\subsection{b}
+\begin{lemma}
+If $G$ is $k$-Ore and $G \neq K_k$ then $T_k(G) \geq 2+ \frac{v(G) - 1}{k-1}$.
+\end{lemma}
+\begin{proof}
+Suppose not and let $G$ be a minimum counterexample.
+Since $G$ is $k$-Ore and $G\neq K_k$ there exist $k$-Ore graphs $G_1$ and $G_2$ such that $G$ is the Ore-composition of $G_1$ and $G_2$.
+We proceed by case distinction on if $G_1, G_2$ are $K_k$ or not.
+If neither $G_1$ nor $G_2$ are $K_k$ then by minimality of $G$,
+$$T_k(G_i) \geq 2 + \frac{v(G_i) - 1}{k-1}$$
+for $i =1,2$. But then combining with the inequality
+$$T_k(G) \geq T_k(G_1) + T_k(G_2) - 2$$
+from the previous Lemma, we see that
+$$T_k(G) \geq 2 + \frac{v(G_1) - 1}{k-1} + 2 + \frac{v(G_2) - 1}{k-1} -2 \geq 2 + \frac{v(G)-1}{k-1}$$
+as desired, using that $v(G_1) + v(G_2) -1 = v(G)$.
+\paragraph{}
+If $G_1 = K_k$ and $G_2 \neq K_k$ then using the first inequality from the previous Lemma we have
+$$T_k(G) \geq T_k(G_1-e) + T_k(G_2-z) \geq 2 + T_k(G_2-z)$$
+Since $T_k(G_2-z) \geq T_k(G_2) - 1$, 
+$$T_k(G) \geq 1 + T_k(G_2) \geq \frac{k-1}{k-1} + \frac{v(G_2)-1}{k-1} + 2= \frac{v(G_1) + v(G_2) - 2}{k-1} +2 = \frac{v(G) - 1}{k-1} +2$$
+The proof for $G_2 = K_k$ and $G_1 \neq K_k$ follows similarly.
+\paragraph{}
+If $G_1 = K_k$ and $G_2 = K_k$ then 
+$$T_k(G) \geq T_k(G_1-e) + T_k(G_2 - z) = 2  + 2 = \frac{2(k-1)}{k-1} + 2 = \frac{v(G_1) + v(G_2) -2}{k-1} + 2 = \frac{v(G) -1}{k-1} + 2$$
+as desired.
+\end{proof}
 \end{document}

### 2018-02-21 CO749-GraphColoring/assignments/A3.tex
Last claim of A3 remains

--- 2018-02-17 CO749-GraphColoring/assignments/A3.tex
@@
 \end{proof}
+
+\subsection{c}
+\paragraph{}
+From our work in class it suffices to show $c\alpha(R') \leq c\alpha(R) + c\alpha(W) - c\alpha(K_{|X|}) + c$ for $c \geq 0$.
+Since we know the core is always non-empty, $\alpha(K_{|X|}) = 1$ and so the problem reduces to showing
+$$\alpha(R') \leq \alpha(R) + \alpha(W).$$
+Let $S$ be an independent set of $G[R']$ such that $|S| = \alpha(R')$.
+Since $R'= V(W) \cup R - \{v_1, \dots, v_{k-1}\}$, where $\{v_1, \dots, v_{k-1}\}$ are the set of color-identified vertices, as in the usual definition.
+So we have
+$$|S| \leq |S\cap (V(W) - \{v_1, \dots, v_k\}| + |S\cap R|$$
+On the one hand $S\cap (V(W) - \{v_1, \dots, v_k\})$ is an independent set in $W$ and $S \cap R$ is an indepenent set in $G[R]$, so then
+$$ \alpha(R') = |S| \leq |S\cap (V(W) - \{v_1, \dots, v_k\}| + |S\cap R| \leq \alpha(W) + \alpha(R)$$
+as desired. $\blacksquare$
+\newpage
+\section{}
+\subsection{a}
+\begin{lemma}
+If $G$ is the Ore-composition of $G_1$ and $G_2$ then
+$$T_k(G) \geq T_k(G_1-e) + T_k(G_2 - z) \geq T_k(G_1) + T_k(G_2) - 2$$
+where $e = xy$ is the replaced edge of $G_1$ and $z$ is the split vertex of $G_2$.
+\end{lemma}
+\begin{proof}
+Let $H_1 \subseteq G_1- e$ be the subgraph of $G_1 - e$ which achieves the optimum in 
+$$T_k(G_1-e) := \max_{H \subseteq G_1 - e} T_k(H).$$
+Similarly let $H_2 \subseteq G_2 - z$ be the subgraph of $G_2 - z$ which achieves the optimum in $T_k(G_2 - z)$. Then $H_1 \cup H_2$ is a disjoint union of cliques in $G$. So we have
+$$T_k(G_1-e) + T_k(G_2 - z) = T_k(H_1) + T_k(H_2) = T_k(H_1 \cup H_2) \leq T_k(G).$$
+So the first inequality of the lemma is proven.
+\paragraph{}
+Now let $H_1' \subseteq G_1$ be the subgraph of $G_1$ which achieves the optimum in $T_k(G_1)$, and let $H_2' \subseteq G_2$ be the subgraph of $G_2$ which achieves the optimum in $T_k(G_2)$.
+If $|e \cap V(H'_1)| \leq 1$ then $H'_1$ is a subgraph of $G_1 - e$ and so $T_k(G_1) \leq T_k(G_1-e)$.
+Otherwise, i.e.
+$ e\in E(H'_1)$ then $H'_1 - x$ is a subgraph of $G_1 - e$, and $T_k(H'_1-x) = T_k(H'_1) - 1$.
+Thus
+$$T_k(G_1 - e) \geq T_k(H'_1-x) \geq T_k(H'_1) - 1 =T_k(G_1) - 1.$$
+Similarly $H'_2 - z$ is a subgraph of $G_2 - v$ and $T_k(H'_2-z) \geq T_k(H'_2) - 1$. So 
+$$T_k(G_2 - z) \geq T_k(H'_2 - z) \geq T_k(H'_2) - 1 = T_k(G_2) - 1.$$
+Summing our inequalities above we obtain
+$$T_k(G_1 - e) + T_k(G_2 - z) \geq T_k(G_1) + T_k(G_2) - 2$$
+as desired.
+\end{proof}
+\subsection{b}
+\begin{lemma}
+If $G$ is $k$-Ore and $G \neq K_k$ then $T_k(G) \geq 2+ \frac{v(G) - 1}{k-1}$.
+\end{lemma}
+\begin{proof}
+Suppose not and let $G$ be a minimum counterexample.
+Since $G$ is $k$-Ore and $G\neq K_k$ there exist $k$-Ore graphs $G_1$ and $G_2$ such that $G$ is the Ore-composition of $G_1$ and $G_2$.
+We proceed by case distinction on if $G_1, G_2$ are $K_k$ or not.
+If neither $G_1$ nor $G_2$ are $K_k$ then by minimality of $G$,
+$$T_k(G_i) \geq 2 + \frac{v(G_i) - 1}{k-1}$$
+for $i =1,2$. But then combining with the inequality
+$$T_k(G) \geq T_k(G_1) + T_k(G_2) - 2$$
+from the previous Lemma, we see that
+$$T_k(G) \geq 2 + \frac{v(G_1) - 1}{k-1} + 2 + \frac{v(G_2) - 1}{k-1} -2 \geq 2 + \frac{v(G)-1}{k-1}$$
+as desired, using that $v(G_1) + v(G_2) -1 = v(G)$.
+\paragraph{}
+If $G_1 = K_k$ and $G_2 \neq K_k$ then using the first inequality from the previous Lemma we have
+$$T_k(G) \geq T_k(G_1-e) + T_k(G_2-z) \geq 2 + T_k(G_2-z)$$
+Since $T_k(G_2-z) \geq T_k(G_2) - 1$, 
+$$T_k(G) \geq 1 + T_k(G_2) \geq \frac{k-1}{k-1} + \frac{v(G_2)-1}{k-1} + 2= \frac{v(G_1) + v(G_2) - 2}{k-1} +2 = \frac{v(G) - 1}{k-1} +2$$
+The proof for $G_2 = K_k$ and $G_1 \neq K_k$ follows similarly.
+\paragraph{}
+If $G_1 = K_k$ and $G_2 = K_k$ then 
+$$T_k(G) \geq T_k(G_1-e) + T_k(G_2 - z) = 2  + 2 = \frac{2(k-1)}{k-1} + 2 = \frac{v(G_1) + v(G_2) -2}{k-1} + 2 = \frac{v(G) -1}{k-1} + 2$$
+as desired.
+\end{proof}
+\subsection{c}
+\paragraph{}
+It suffices to show
+$$T_k(R) + T_k(W) - T_k(K_{|X|}) \leq T_k(R') + |X|.$$
+Let $H_R \subseteq G[R]$ be a disjoint union of cliques attaining the maximum in $T_k(R)$.
+Let $H_W \subseteq W$ be a disjoint union of cliques attaining the maximum in $T_k(W)$.
+If a vertex in $X$ is deleted from a clique in $H_W$ then the value of $T_k$ for the resulting graph will decrease by at most one.
+Hence $T_k(H_W - X) \geq T_k(H_W) - |X|$.
+But then $H_R \cup (H_W - X)$ is a disjoint union of cliques in $G[R']$.
+So we have
+\begin{align*}
+T_k(R) + T_k(W) - T_k(K_{|X|}) &\leq T_k(H_R) + T_k(H_W)\\ &\leq T_k(H_R) + T_k(H_W - X) - |X|\\ &\leq T_k(H_R \cup (H_W-X)) - |X|\\ &\leq T_k(R') - |X|\end{align*}
+as desired. $\blacksquare$
+\newpage
+\section{}
+\paragraph{}
+Let $k \geq 4$.
+Let $p(G) = (k-1)v(G) - 2e(G)$.
+Then $p(G) \leq 0$ if $G$ is $k$-critical since the minimum degree of $G$ is $k-1$.
+Let $G$ be a counterexample to the potential version of Brooks' Theorem.
+That is, $G$ is $k$-critical, $G\neq K_k$, and $p(G) = 0$.
+Since $p(G) = 0$, we have that $G$ is $(k-1)$-regular.
+\begin{claim}
+If $R \subset V(G)$ then $p(R) \geq k-1$.
+\end{claim}
+\begin{proof}
+Suppose not and let $R$ be a counterexample with maximum number of vertices. Then $R$ has an extension $R'$ with extender $W$ and core $X$. By the potential-extension lemma,
+$$p(R') \leq p(R) + p(W) - p(K_{|X|})$$
+Now $$p(R) =  (k-1)|R| - 2e(R) \geq (k-|R|)|R|$$
+which is at least $k-1$ when $1 \leq |R| \leq k-1$.
+Hence $|R| \geq k$, and since $G$ is not equal to $K_k$ and $G$ is $(k-1)$-regular, $G[R]$ is not a clique of size at least $k$.
+Therefore $W$ is a strict subgraph of $G$, and so by minimality of $G$,
+$$p(W) \leq -1.$$
+By the same reasoning as above for $R$, $$p(K_{|X|}) \geq k-1.$$
+Thus we have
+$$p(R') \leq p(R) + p(W) - p(K_{|X|}) \leq (k-2)  - 1 -(k-1) \leq - 2$$
+By maximality of $R$, $p(R') \geq k-1$ contradicting the above.
+\end{proof}
+\begin{claim}
+If $z$ is a vertex with two non-adjacent neighbours $x,y$ then there exists a $(k-1)$-clique $H$ in $G\backslash \{x,y,z\}$ such that $N(x) \cap V(H)$, $N(y) \cap V(H)$ partition $V(H)$.
+\end{claim}
+\begin{proof}
+
+\end{proof}
 \end{document}

### 2018-02-22 CO749-GraphColoring/assignments/A3.tex
remains to derive final contradiction

@@
 \begin{proof}
+Let $G'$ be the graph obtained from $G$ by identifying vertices $x$ and $y$ (deleting parallel edges if necessary) as vertex $v_{xy}$ and deleting vertex $z$.
+If $G'$ has a $(k-1)$-coloring $\phi'$ then $\phi'$ can be extending to a $(k-1)$-coloring of $G$ by assigning $\phi'(x ) = \phi'(y) = \phi'(v_{xy})$, then assigning $z$ a color not present on the neighbours of $z$.
+Such a color exists since $G$ is $(k-1)$-regular and two neighbours of $z$, $x$ and $y$, receive the same color under $\phi'$.
+\paragraph{}
+Thus $G'$ is not $(k-1)$-colorable.
+So $G'$ has a $k$-critical subgraph $K'$.
+Then $v_{xy} \in V(K')$ for otherwise $K'$ is a $k$-critical strict subgraph of $G$, contradicting that $G$ is $k$-critical.
+By minimality of $G$ we have
+$$p(K') \leq 0$$
+with equality if and only if $K'$ is $K_k$. Let $K$ be the graph $K' - v_{xy} + \{x,y\}$. Then $K$ is a strict subgraph of $G$, since $z \not\in V(K)$. So by the previous claim
+$$k-1 \leq p(G[V(K')]) \leq p(K')$$
+Now we observe that
+$$p(K) = p(K') + (k-1) - 2|E(v_{xy}, N(x) \cap N(y) \cap V(K'))|$$
+Since $K$ has more vertex that $K'$, and parallel edges were deleted in obtaining $G'$. Combining inequalities we see that
+$$k-1 \leq p(K) \leq 0 + (k-1) - 2|E(v_{xy}, N(x) \cap N(y) \cap V(K'))|$$
+and hence 
+$$|E(v_{xy}, N(x) \cap N(y) \cap V(K'))| \leq 0.$$
+Since $|E(v_{xy}, N(x) \cap N(y) \cap V(K'))|$ is non-negative this implies $|E(v_{xy}, N(x) \cap N(y) \cap V(K'))| = 0$ and thus
+$N(x)\cap N(y) \cap V(K') = \emptyset$. Furthermore the displayed inequalities are tight, implying that $K'$ is isomorphic to $K_k$. Letting $H = K'-v_{xy}$ we obtain the desired clique of the claim.
+\end{proof}
 
-\end{proof}
 \end{document}

### 2018-02-23 CO749-GraphColoring/assignments/A3.tex
finished q4

@@
 \end{proof}
+\paragraph{}
+Since $G$ is not $K_k$, there exists a vertex $z$ with tow non-adjacent neighbbours $x,y$.
+Invoking the previous Claim we obtain a clique $H$ in $G\backslash\{x,y,z\}$ such that $N(x) \cap V(H)$, $N(y) \cap V(H)$ partition $V(H)$.
+Now let $G'$ be the graph obtained from $G$ by adding edge $xy$ and deleting $H$.
+If $G'$ has a $(k-1)$-coloring $\phi'$ then we claim $\phi'$ can be extended to a $(k-1)$-coloring of $G$.
+Since $G$ is $(k-1)$-regular and the neighbours of $x$ and $y$ partition $V(H)$, every vertex of $H$ is adjacent with every other vertex of $H$, and precisely one of $x$ or $y$.
+Let $v_x \in N(x) \cap V(H)$ and let $v_y \in N(y) \cap V(H)$.
+Assign $\phi'(v_x) = \phi'(y)$ and $\phi'(v_y) = \phi'(x)$.
+Assign the remaining $k-3$ colors (those not equal to $\phi'(x)$ or $\phi'(y)$) to the $k-3$ vertices of $H - \{v_x, v_y\}$ so each receives a unique color.
+This yields a $(k-1)$-coloring of $G$, hence $G'$ is not $(k-1)$-colorable.
+\paragraph{}
+Since $G'$ is not $(k-1)$-colorable, $G'$ contains a $k$-critical subgraph $W$.
+Then $xy \in E(W)$, for otherwise $W$ is a $k$-critical subgraph of $G$, contradicting that $G$ is $k$-critical.
+Now, using Claim $4.1$, we observe that
+$$p(W) = p(G[V(W)]) - 2 \geq k-1 - 2 = k-3 \geq 1 $$
+with the last inequality following since $k \geq 4$. But on the other hand $p(W) \leq 0$ since $W$ is $k$-critical. Then we have $1 \leq 0$, a contradiction. $\blacksquare$
 

### 2018-02-26 CO749-GraphColoring/assignments/A3.tex
revisions

@@
 Let $S \subseteq V(G)$ be an independent set in $G$ such that $|S| = \alpha (G)$.
-Then treating $z_x$ as $x$ and $z_y$ as $z$ we can partition $S$ into two independent sets $S_1\subseteq V(G_1)$ and $S_2\subseteq V(G_2)$ as $S_1 = V(G_1) \cap S$ and $S_2 = S\backslash S_1$. Thus
+\paragraph{}
+We claim that we can always partition $S$ into $S_1$ and $S_2$ where $S_1$ is an independent set in $G_1$ and $S_2$ is an independent set in $G_2$. We proceed by case distinction:
+\subparagraph{} Case $1:$ $\{z_x,z_y\} \cap S = \emptyset$. Then trivially $S_1 = S \cap V(G_1)$ and $S_2 = S\cap V(G_2) = S \backslash S_1$.
+\subparagraph{} Case $2:$ $|\{z_x,z_y\} \cap S| \leq 1$.
+Without loss of generality we may assume $x \in S$ and  $y \not\in S$.
+Then treating $z_x$ as $x$ we can partition as before: $S_1 = (S \cap V(G_1)) \cup \{x\}$ and $S_2 = S\cap V(G_12)$.
+\subparagraph{} Case $3:$ $\{z_x, z_y\} \subseteq S$.
+Then for all $v \in S\cap V(G_2)$, $v$ is not adjacent to $z$, for otherwise $v$ would be adjacent to one of $z_x, z_y$ contradicting that $S$ is an independent $S$.
+So treating $z_x$ as $x$ and $z_y$ as $z$, we can partition $S$ into $$S_1 = ((S\backslash\{z_x,z_y\})\cap V(G_1)) \cup \{x\}$$ and $$S_2 = ((S\backslash\{z_x,z_y\})\cap V(G_2)) \cup \{z\}.$$
+Now using our claim we see that
 $$\alpha(G) = |S| = |S_1| + |S_2| \leq \alpha(G_1) + \alpha(G_2)$$
@@
 	\paragraph{}
-	For $i \in \{1, \dots, 2\}$ let $S_i$ be in an independent set in $G_i$ satisfying $$|S_i| = \alpha(G_i) = \frac{v(G_i) - 1}{k-1}.$$ We use the same notion as the above problems for $x,y, z, z_x, z_y$.
+	For $i \in \{1, \dots, 2\}$ let $S_i$ be in an independent set in $G_i$ satisfying $$|S_i| = \alpha(G_i) = \frac{v(G_i) - 1}{k-1}.$$ We use the same notation as in the above problems for $x,y, z, z_x, z_y$.
 Since $xy \in E(G_1)$, $|\{x,y\} \cap S_1| \leq 1$.

### 2018-03-14 CO749-GraphColoring/assignments/A4.tex
Wrote up first lemma

--- 2018-01-31 CO749-GraphColoring/assignments/A2.tex
@@
 \lhead{\today}
-\rhead{W. Justin Toth - A2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth - A4} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \paragraph{}
-By Voigt's Theorem there exists a planar graph $H$ and a list assignment $L$ with $|L(u)| \geq 4$ for all $u \in V(H)$, such that $H$ is not $L$-colorable.
-We construct a graph $G$ as follows.
-Take $5$ distinct copies of $H$: $H_1, \dots, H_5$ where each vertex $u_i \in V(H_i)$ corresponds to vertex $u \in V(H)$, and take an additional vertex $v$.
-Add edges $vu_i$ for all $u_i \in V(H_i)$ for all $i \in \{1,\dots, k\}$.
-i.e.
-$v$ is adjacent to every vertex in every copy of $H$.
-\paragraph{}
-To see that $G$ is an apex graph observe that $G-v$ consists of five disjoint copies of a planar graph, and hence $G-v$ is planar.
-\paragraph{}
-We now construct a $5$-list-assignment $L'$ for which $G$ does not have an $L'$-coloring, completing the proof that $G$ is not $5$-list-colorable.
-Let $c_1, \dots, c_5$ be $5$ disjoint colors from $L$.
-That is for all $u \in V(H)$, $$L(u) \cap \{c_1, \dots, c_5\} = \emptyset.$$
-Let $L'(v) = \{c_1, \dots, c_5\}$.
-For each $i \in \{1, \dots, 5\}$ and for each $u_i \in V(H_i)$ let $L'(u_i) = L(u) \cup \{c_i\}$.
-Since $|L(u)| \geq 4$, $|L'(u_i)|\geq 5$.
-Thus we have a $5$-list-assignment $L'$ for $G$.
-\paragraph{}
-Suppose for a constradiction $G$ has an $L'$-coloring $\phi$.
-Say $\phi(v) = c_i$.
-Then $\phi(u_i) \neq c_i$ for all $u_i \in V(H_i)$.
-Hence $\phi(u_i) \in L(u_i)$ for all $u_i \in V(H_i)$.
-That is, $\phi$ restricted to $H_i$ is an $L$-coloring of $H-_i$.
-But this contradicts that $H_i$ is not $L$-colorable.
-Therefore $G$ does not have an $L'$-coloring and hence is not $5$-list-colorable.
-$\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-We opt to prove the BONUS Lemma, i.e. Lemma $3$:
-\textit{Let $G$ be a graph and for each vertex $v$ of $G$ let $r(v)$ be a non-negative integer. If $L$ is a list assignment of $G$ such that for every vertex $v$ of $G$
-$$|L(v)| \geq \frac{d(v) + 1}{r(v)+1},$$
-then $G$ has an $r(v)$-defective coloring}. This Lemma directly implies Lemma $2$ of this problem.
-\paragraph{}
-Consider a graph $G$, non-negative integers $r(v)$ for each $v \in V(G)$, and a list assignment $L$ of $G$ such that for each $v \in V(G)$
-$$|L(v)| \geq \frac{d(v) + 1}{r(v)+1}.$$
-Let $\phi$ be an improper $L$-coloring of $G$ minimizing the number of ``improper" edges $uv \in E(G)$ such that $\phi(u) = \phi(v)$.
-Suppose for a contradiction that $\phi$ is not an $r(v)$-defective coloring.
-\paragraph{}
-Then there exists a vertex $v$ such that the number of neighbours of $v$ receiving the same color as $v$ under $\phi$ is at least $r(v) + 1$.
-If there exists a color $c \in L(v)$ such that the number of neighbours of $v$ colored $c$ is at most $r(v)$ then we can construct a new coloring $\phi'$ by setting $\phi'(v) = c$ and $\phi'(u) = \phi(u)$ for all $u \neq v$.
-This coloring has $r(v)$ new improper edges, those between $v$ and its $c$-colored neighbours.
-But $\phi'$ switches $r(v) + 1$ edges which were improper under $\phi$, those between $v$ and its $\phi(v)$-colored neighbours, to proper edges.
-Hence $\phi'$ has fewer improper edges than $\phi$, a contradiction.
-\paragraph{}
-If such a color $c \in L(v)$ does not exist then  for each $c \in L(v)$ the number of neighbours of $v$ colored $c$ under $\phi$ is at least $r(v) + 1$. But then $v$ has at least
-$$|L(v)|(r(v) + 1) \geq d(v) + 1$$
-neighbours, a contradiction. Therefore $\phi$ is an $r(v)$-defective coloring $\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-We prove a slightly more general result allowing $m(v)$ to take the value $0$.
-This will make induction easier when we wish to delete from a list of size $2$.
-For such a $v$, $L(v)= \emptyset$ and the desired $S(v) = \emptyset$.
-Everything else is exactly as stated in the problem.
-\paragraph{}
-Suppose not. Let $G$ be a counterexample satisfying:
-\begin{enumerate}
-	\item $\sum_{v \in V(G)} m(v)$ is minimized, and subject to that
-	\item if possible $G(c)$ is connected for each $c \in L(G)$, and subject to that possibility
-	\item $|L(G)|$ is minimized.
-\end{enumerate}
-\begin{claim}
-	\begin{enumerate}
-		\item For all $c \in L(G)$, $G(c)$ is connected.
-		\item For every pair of distinct colors $c,c' \in L(G)$, $G(c) \neq G(c')$.
-		\item For every pair of distinct colors $c, c' \in L(G)$, we have $V(G(c)) \cap V(G(c')) \neq \emptyset$ (we strengthened this claim).
-	\end{enumerate}
-\end{claim}
-\begin{proof}
-	\paragraph{} For (1.). While there exists a color $c \in L(G)$ such that $G(c)$ has at least two connected components $G_1$ and $G_2$ do the following:
-		\subparagraph{}
-		Let $c'$ be a new color not contained in $L(G)$.
-For each $v \in V(G_2)$ let $$L'(v) = (L(v)\cup \{c'\}) \backslash \{c\}.$$ For each $v \in V(G) \backslash V(G_2)$ let $L'(v) = L(v)$.
-Replace $L$ with $L'$.
-	\paragraph{}
-	We claim that every iteration of the above preserves the property of $(G,m,L)$ being a counterexample, but decreases the ``disconnectivity".
-i.e.
-the sum over $c \in L(G)$ such that $G(c)$ is disconnected, of the number of connected components in $G(c)$.
-	\paragraph{}
-	It is clear the disconnectivity decreases.
-We add a new color $c'$ for which $G(c')$ is connected, and $G(c)$ loses one connected component, and the number of connected components of $G(c'')$ is not affected for any other color $c'' \in L(G)$.
-	\paragraph{}
-	To verify that the new lists, say $L'$, also admit a counterexample with $G$ and $m$, suppose to contrary there are $S'(v) \subseteq L'(v)$ of size $m(v)$ for each $v \in V(G)$ and $S'(u) \cap S'(v) = \emptyset$ for all $uv \in E(G)$.
-For each $v \in V(G)$ let $$S(v) = \begin{cases} S'(v)\cup \{c\} \backslash \{c'\}, &\text{if } v \in V(G_2) \text{ and } c' \in S'(v) \\
-	S'(v), &\text{otherwise}.\end{cases} $$
-	Then $S(v) \subseteq L(v)$ and $|S(v)| = m(v)$ for each $v \in V(G)$.
-Further, for each $uv \in E(G)$, $S(u) \cap S(v) = \emptyset$.
-To see this, observe that if $S(u) \cap S(v) \neq \emptyset$ then by our choice of $S'$, $S(u) \cap S(v) = \{c\}$ and, since $G(c)$ was disconnected, $u, v \in V(G_2)$.
-But then, $S'(u) \cap S'(v) = \{c'\}$, a contradiction.
-	\paragraph{}
-	So the above algorithm will terminate when the ``disconnectivity" reaches $0$ and we will have found a connected counterexample with the same value for $\sum_{v} m(v)$ as $G$.
-This contradicts our choice of $G$ if there exists color $c$ such that $G(c)$ is disconnected.
-Therefore $G(c)$ is connected for all $c \in L(G)$.
-	\paragraph{}For (2.). Suppose for a contradiction there exist distinct colors $c,c' \in L(G)$ such that $G(c) = G(c')$. Construct new lists $L'$ such that for each $v \in V(G)$
-	$$ L'(v) = \begin{cases}
-	L(v) \backslash \{c,c'\}, &\text{ if } v \in V(G(c)) \\
-	L(v), &\text{otherwise.}
-		\end{cases}$$
-	Let $m'(v) = m(v) - 1$ if $v \in V(G(c))$ (this is well-defined since such $m(v) \geq 2$) and let $m'(v) = m(v)$ otherwise.
-By the minimality of $G$, there exist for each $v \in V(G)$, a set $S'(v)\subseteq L'(v)$ of size $m'(v)$.
-These sets satisfy $S'(u) \cap S'(v) = \emptyset$ for all $uv \in E(G)$.
-Order the vertices of $G(c)$ as $v_1, \dots, v_k$ with respect to some orientation of $G$.
-Construct sets $S(v)$ as follows.
-If $v \in V(G) \backslash V(G(c))$ set $S(v) = S'(v)$.
-If $v_i \in V(G(c))$ and $i$ is even set $S(v) = S'(v) \cup \{c\}$.
-If $v_i \in V(G(c))$ and $i$ is odd set $S(v) = S'(v) \cup \{c'\}$.
-Since $c$ and $c'$ do not appear in sets outside $G(c)$, and we alternate $c$ and $c'$ along $G(c)$, we have $S(u) \cap S(v) = \emptyset$ for all $uv \in E(G)$.
-But this contradicts that $(G,m,L)$ is a counterexample.
-Hence for distinct colors $c$ and $c'$, $G(c) \neq G(c')$.
-	\paragraph{} For (3.).
-Suppose for a contradiction  there exist distinct colors $c,c' \in L(G)$, such that $V(G(c)) \cap V(G(c')) = \emptyset$.
-Construct new lists $L'(v)$ for each $v \in V(G)$ as follows.
-If $v \in V(G) \backslash V(G(c'))$ then set $L'(v) = L(v)$.
-Otherwise, i.e.
-$v \in V(G(c'))$ set $L'(v) = L(v) \cup \{c\} \backslash \{c'\}.$ Then for each $v \in V(G(c'))$, we have $|L'(v)|  = |L(v)| + 1 - 1 = m(v)$, since $V(G(c))$ and $V(G(c'))$ are disjoint.
-But $L'(G) = L(G) \backslash \{c\}$ so by minimality there exist sets $S'(v) \subseteq L'(v)$ for each $v \in V(G)$ such that $|S'(v)| = m(v)$, satisfying $S'(u) \cap S'(v) = \emptyset$ for each $uv \in E(G)$.
-	\paragraph{}
-	For each $v \in V(G)$ construct $S(v)$ so that
-	$$S(v) = \begin{cases}
-		S'(v) \cup \{c'\}\backslash \{c\}, &\text{ if } v \in V(G(c')) \text{ and } c \in S'(v) \\
-		S'(v), &\text{ otherwise}.
-	\end{cases}$$
-	Then each set $S(v) \subseteq L(v)$ is of size $m(v)$ and for each edge $uv \in E(G)$, $S(u) \cap S(v) = \emptyset$.
-To verify disjointness, suppose for a contradiction $S(u) \cap S(v) \neq \emptyset$ for some edge $uv$.
-Since $S'(u) \cap S'(v) = \emptyset$, $S(u) \cap S(v) = \{c'\}$.
-But then $u,v \in V(G(c'))$ and so $S'(u) \cap S(v) = \{c\}$ since $c'$ is only added to $S(u), S(v)$ if $c \in S'(u), S'(v)$.
-But this contradicts disjointness of $S'(u), S'(v)$.
-So sets $S(v)$ for $v \in V(G)$ violate that $(G,m,L)$ is a counterexample, a contradiction.
-\end{proof}
-	\begin{claim}
-		If $c \neq c' \in L(G)$ such that $G(c), G(c')$ are paths, then there does not exist a component $P$ of $G(c) \cap G(c')$ with an even number of vertices such that one end of $P$ is an end of $G(c)$ and the other end of $P$ is an end of $G(c')$.
-	\end{claim}
-	\begin{proof}
-		\paragraph{} Suppose for a contradiction such $P$ exists.
-Order the vertices of $P$ as $v_1, \dots, v_k$ so that $v_1$ is an end of $G(c)$ and $v_k$ is an end of $G(c')$.
-Construct new lists $L'(v)$ for each $v \in V(G)$ as follows.
-If $v \in V(P)$ let $L'(v) = L(v) \backslash \{c,c'\}$ and let $m'(v) = m(v) - 1$.
-Otherwise let $L'(v) = L(v)$ and $m'(v) = m(v)$.
-By minimality, there exist sets $S'(v) \subseteq L'(v)$ for each $v \in V(G)$ of size $m'(v)$, such that $S'(u) \cap S'(v) = \emptyset$ for each $uv \in E(G)$.
-		\paragraph{}
-		Construct sets $S(v)\subseteq L(v)$ for each $v \in V(G)$ as follows.
-If $v_i \in V(P)$ and $v_i$ is odd then set $S(v) = S'(v) \cup \{c\}$.
-If $v_i \in V(P)$ and $v_i$ is even set $S(v) = S'(v) \cup \{c'\}$.
-If $v \in V(G)\backslash V(P)$ set $S(v) = S'(v)$.
-Then $|S(v)| = m(v)$ for each $v \in V(G)$.
-Since we alternate adding $c$ and $c'$ along $P$ the sets $S(u)$ and $S(v)$ are disjoint for each edge $uv \in E(G)$.
-Since we did not change $S(u)$ and $S(v)$ for edges $uv$ with neither endpoint on $P$ they are also disjoint.
-It remains to consider two edges $u_1v_1$ and $v_ku_k$ where $u_1, u_k \not\in V(P)$.
-Since $v_1$ is an end of $G(c)$ and $v_2 \in G(c)$ we have $c \not\in L(u_1)$ and hence, as $S'(u_1) \cap S'(v_1) = \emptyset$, $S(u_1)$ and $S(v_1)$ are disjoint.
-Similarly since $v_k$ is an edge of $G(c')$, and $k$ is even, $S(u_k)$ and $S(v_k)$ are disjoint.
-But then the sets $S(v)$ contradict that $(G,m,L)$ is a counterexample.
-	\end{proof}
-\begin{claim}
-	If $c \in L(G)$ then $|V(G(c))|$ is even.
-\end{claim}
-\begin{proof}
-	Deferred to Problem $4$.
-\end{proof}
-\begin{claim}
-	For every edge $e=uv \in E(G)$, $L(u) \subseteq L(v)$ or $L(v) \subseteq L(u)$.
-\end{claim}
-\begin{proof}
-	\paragraph{}
-	Suppose not.
-Then there exists an edge $e=uv \in E(G)$ such that $L(u)\backslash L(v) \neq \emptyset$ and $L(v)\backslash L(u) \neq \emptyset$.
-Let $c_u \in L(u) \backslash L(v)$ and let $c_v \in L(v)\backslash L(u)$.
-By our previous claims, both $G(c_u)$ and $G(c_v)$ are connected and non-disjoint, so they intersect in a path.
-Let $P = G(c_u) \cap G(c_v)$.
-Since $v \not\in G(c_u)$ (also note $u \not\in G(c_v)$), $P \neq G$, and we have $G(c_u) \cup G(c_v) = G$.
-Further one end of $P$ is in $G(c_u)$ and the other is in $G(c_v)$.
-Thus by Claim $3.2$, $|V(P)|$ is odd.
-	\paragraph{}
-	Since $|V(G)|$ is even,  and $|V(P)|$ is odd, we have that $|V(G) \backslash V(P)|$ is odd.
-Then since $V(G)\backslash V(P)$ is the disjoint union of $V(G(c_u))\backslash V(P)$ and $V(G(c_v)) \backslash V(P)$, precisely one of $|V(G(c_u))\backslash V(P)|$ or $|V(G(c_v))\backslash V(P)|$ is even and the other is odd.
-Say without loss of generality that $|V(G(c_u))\backslash V(P)|$ is even.
-Then $|V(G(c_u))| = |V(G(c_u))\backslash V(P)| + |V(P)|$ is odd.
-But this contradicts Claim $3.3$.
-	\end{proof}
-\begin{claim}
-For every edge $e=uv \in E(G)$, $L(u) = L(v)$.
-\end{claim}
-\begin{proof}
-	TODO
-\end{proof}
-\paragraph{}
-We are now ready to complete the main proof.
-The argument is essentially the same as $2$-colorability of even cycles.
-From Claim $3.5$, let $L$ denote the constant list assigned to each vertex, i.e.
-$L = L(u)$ for each $u \in V(G)$.
-We know $|L|$ is even and we need to assign half its elements to each vertex.
-\paragraph{}
- Let $S_1 \subseteq L$ be a subset of size $|L|/2$.
-Let $S_2 = L \backslash S_1$.
-Then $|S_1| = |S_2| = m(v)$ for any $v \in V(G)$.
-Enumerate the vertices about $G$ as $v_1, \dots, v_k$ with $k$ even.
-For each $v_i \in V(G)$, let $S(v) = S_1$ is $i$ is odd and $S(v) = S_2$ is $i$ is even.
-Since $G$ is an even cycle for any edge $v_iv_j \in E(G)$ exactly one of $i,j$ is odd and the other is even.
-Hence $S(v_i) \cap S(v_j) = \emptyset$, contradicting that $G$ is a counterexample.$\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-We will prove Claim $3.3$ from the previous problem.
-Suppose the Claim is false.
-Then there exists $c \in L(G)$  such that $G(c)$ is a path $P=v_1\dots v_k$ with $k$ odd.
-Choose such $c$ so that $|V(P)|$ is minimized.
-We adopt the same notation and definitions as the problem statement.
-\begin{lemma}
-	For every $a,b \in L(G)$, if $|V(G(a))|$ is odd then we do not have $G(a) \subseteq G(b)$.
+Let $v(G)$ and $e(G)$ denote the number of vertices and edges respectively of a graph $G$.
+Also let $c(G)$ denote the number of connected components of $G$.
+Let $\phi$ be a $k$-coloring of $G$.
+Define $\cB(G, \phi)$ so that
+$$\cB(G,\phi) = \{G[\phi^{-1}(c) \cup \phi^{-1}(c')] : c \in \{1,\dots, k\}, c'\in \{1,\dots, k\}\backslash\{c\}\}.$$
+I.e. $\cB(G,\phi)$ is the set of bipartite subgraphs of $G$ induced by two distinct color classes under $\phi$.
+\begin{lemma}\label{lemma:bipartite-cc}
+For all $k \geq 3$, for all $k$-colorable graphs $G$, if $\phi$ is a $k$-coloring of $G$ then
+$$\sum_{H \in \cB(G,\phi)} c(H) \geq (k-1)v(G) - e(G).$$
 \end{lemma}
 \begin{proof}
-	Suppose not.
-Then delete colors $a$ and $b$ from $L(v)$ for each $v \in V(G(a))$ to obtain lists $L'(v)$.
-For vertices $v$ off $G((a))$ let $L'(v) = L(v)$.
-By minimality, there exist $S'(v) \subseteq L'(v)$, such that $|S'(v)| = \frac{1}{2}|L'(v)|$ for each $v \in V(G)$, and for each edge $uv \in E(G)$, $S'(u) \cap S'(v) = \emptyset$.
-	\paragraph{}
-	Enumerate the vertices of $G(a)$ as $v_1 \dots v_\ell$ with $\ell$ odd.
-For each $v_i \in V(G(a))$ let $S(v_i) = S'(v_i) \cup \{a\}$ if $i$ is odd, and $S(v_i) = S'(v_i) \cup \{b\}$ if $i$ is even.
-For each $v \in V(G)\backslash V(G(a))$ let $S(v) = S'(v)$.
-Then $|S(v)| = \frac{1}{2} |L(v)|$ for each $v \in V(G)$.
-	\paragraph{}
+Fix $k \geq 3$ and fix $n \geq k$.
+Let $G$ be a graph with $v(G) = n$.
+We proceed by induction on $e(G)$.
+First suppose $e(G) = 0$.
+Let $\phi$ be a $k$-coloring of $G$.
+If $v \in V(G)$ then $v$ is assigned a color class $c$ and hence is a vertex of each graph $G[\phi^{-1}(c) \cup \phi^{-1}(c')]$ for $c' \neq c$.
+There are $k-1$ such graphs, and since $e(G) = 0$, $v$ itself is a unique connected component of each such graph.
+Therefore
+$$\sum_{H \in \cB(G,\phi)} c(H)= \sum_{v \in V(G)} (k-1) = (k-1)v(G) - e(G).$$
+Now suppose $e(G) >0$ and the lemma holds for $k$-colorable, $n$ vertex graphs with fewer than $e(G)$ edges.
+Again let $\phi$ be a $k$-coloring of $G$.
 Let $uv \in E(G)$.
-If $u,v \not\in V(G(a))$ then $S(u) \cap S(v) = S'(u) \cap S'(v) = \emptyset$.
-If $u,v \in V(G(a))$ then $S(u) \cap S(v) = S'(u) \cap S'(v) \cap \{a\} \cap \{b\} = \emptyset$.
-If $u \in V(G(a))$ and $v \not\in V(G(a))$ then either $u = v_1$ or $u = v_\ell$.
-In either case $S(u) = S'(u) \cup \{a\}$.
-Since $v \not\in V(G(a))$, $a \not\in L(v)$ and so $S(u) \cap S'(v) = \emptyset$ as desired.
-But then the sets $S(v)$ contradict that $G$ is a counterexample.
+Let $c = \phi(u)$ and $c' = \phi(v)$.
+Let $H' := G[\phi^{-1}(c) \cup \phi^{-1}(c')]$.
+Then $H'$ is the unique graph in $\cB(G,\phi)$ containing edge $uv$.
+If $uv$ is a bridge of its connected component in $H'$ then $c(H'-uv) = c(H') + 1$, otherwise $c(H'-uv) = c(H')$.
+Hence $c(H'-uv) \leq c(H')+1$.
+So
+\begin{align*}
+\sum_{H \in \cB(G,\phi)} c(H) &= c(H') + \sum_{H \in \cB(G, \phi)\backslash\{H'\}} c(H)\\
+&= c(H') +  \sum_{H \in \cB(G-uv, \phi)\backslash\{H'\}} c(H) &\text{(since $uv$ is only an edge of $H'$)}\\ 
+&\geq c(H'-uv)-1 +  \sum_{H \in \cB(G-uv, \phi)\backslash\{H'\}} c(H)\\
+&=  -1 + \sum_{H \in \cB(G-uv, \phi)} c(H)\\
+&\geq - 1 + (k-1)v(G) - e(G) + 1 &\text{(by induction hypothesis on $G-uv$)}\\
+&= (k-1)v(G) - e(G).
+\end{align*}
+Therefore by induction the lemma holds.
 \end{proof}
-\begin{claim}
-	For every $v \in V(P)$ and $c' \in L(v) -c$, $c'$ is exactly one of left at $v$ or right at $v$.
-\end{claim}
+\begin{lemma}
+Let $k\geq 3$. If $G$ is a $k$-colorable graph then $G$ has at least $2^{\frac{(k-1)v(G) - e(G)}{{k \choose 2}}}$ distinct $k$-colorings.
+\end{lemma}
 \begin{proof}
-	Let $P'$ be the component of $G(c') \cap P$ containing $v$. By Lemma $4.1$, since $P$ is odd, at least one endpoint of $P'$ is a vertex $v_i$ such that $2\leq i\leq k-1$.
-	\paragraph{}
-	Case $1$: $v_i$ is even.
-We want to show the other endpoint of $P'$ is odd and not equal to $v_1$ or $v_k$.
-If the other endpoint of $P'$ is $v_1$ or $v_k$ then $G(c')$ with $G(c)$ contradicts Claim $3.2$ of Problem $3$.
-If the other endpoint of $P'$ is even then $G(c') \subseteq G(c)$ and $|V(G(c'))|$ is odd, contradicting Lemma $4.1$.
-	\paragraph{}
-	Case $2$: $v_i$ is odd.
-We want to show the other endpoint of $P'$ is not an odd vertex $v_j$ with $2\leq j \leq k-1$.
-Suppose not.
-Then $G(c') \subseteq G(c)$ and $|V(G(c'))|$ is odd, contradicting Lemma $4.1$.
-Hence the Claim is proven.
+Let $\phi$ be a $k$-coloring of $G$.
+Let $H = G[\phi^{-1}(c) \cup \phi^{-1}(c')] \in \cB(G,\phi)$.
+We can obtain a distinct $k$-coloring of $G$ by choosing any number of connected components $H_1, \dots, H_k$ of $H$ and switching colors $c$ and $c'$ on the vertices of $H_1, \dots, H_k$.
+Thus in this manner we can obtain $2^{c(H)}$ distinct $k$-colorings of $G$.
+It will then suffice to show that there exists $H \in \cB(G,\phi)$ such that
+$$c(H) \geq \frac{(k-1)v(G) - e(G)}{{k \choose 2}}.$$
+Since there are ${k\choose 2}$ graphs in $\cB(G,\phi)$, by the pigeonhole principle, there exists $H' \in \cB(G,\phi)$ such that
+$$c(H') \geq \frac{\sum_{H \in \cB(G,\phi)} c(H)}{{k\choose 2}}$$
+Therefore by Lemma \ref{lemma:bipartite-cc} 
+$$c(H') \geq \frac{(k-1)v(G) - e(G)}{{k\choose 2}}$$
+as desired.
 \end{proof}
-\begin{claim}
-	If $2\leq i \leq k-2$ then $v_i$ is either left or right but not both.
-\end{claim}
-\begin{proof}
-	Let $A$ be the set of left colors at $v_i$ and $B$ be the set of right colors at $v_i$.
-By Claim $4.2$, $(A,B)$ is a partition of $L(v_i) - c$.
-Since $|L(v_i)|$ is even, $|L(v_i)-c|$ is odd, and so exactly one of $|A|$ or $|B|$ is odd as $|A| + |B| = |L(v_i)-c|$.
-\end{proof}
-\begin{claim}
-	All of the following hold: $|V(P)| > 1$, $v_2$ is left, and $v_{k-1}$ is right.
-\end{claim}
-\begin{proof}
-	First suppose for a contradiction that $|V(P)| = 1$.
-Let $c' \in L(G) -c$.
-From Claim $3.1$ $(3.)$ we have that $V(G(c)) \cap V(G(c')) \neq \emptyset$.
-Therefore $G(c) \subseteq G(c')$, but since $|V(G(c))|$ is odd, this contradicts Lemma $4.1$.
-Thus $|V(P)| > 1$.
-So vertices $v_2$ and $v_{k-1}$ exist.
-	\paragraph{}
-	We will show all colors of $L(v_2) - c$ are left.
-A similar argument shows all colors of $L(v_{k-1})-c$ are right, by replacing $v_2$ with $v_{k-1}$, $v_1$ with $v_k$ and interchaning left/right.
-Suppose $v_2$ has a right color $c'$.
-Let $P'$ be the component of $G(c') \cap P$ containing $v_2$.
-Then the leftmost vertex of $P'$ is odd, but not equal to $v_1$.
-But there is no odd vertex between $v_1$ and $v_2$, a contradiction.
-\end{proof}
-\begin{claim}
-	If $2\leq i \leq k-2$ then $v_i$ has the same chirality as $v_{i+1}$.
-\end{claim}
-\begin{proof}
-	TODO
-\end{proof}
-\paragraph{}
-To complete the proof we observe that Claim $4.4$ and Claim $4.5$ taken together yield a contradiction, since Claim $4.5$ shows $v_{2}$ and $v_{k-1}$ have the same chirality, but Claim $4.4$ shows they have different chirality.
-$\blacksquare$
 \end{document}

### 2018-03-15 CO749-GraphColoring/assignments/A4.tex
solved 1

@@
 \end{proof}
+\begin{corollary}
+There exists $\epsilon >0$, $\alpha > 0$ such that both of the following hold for all $m \in \{3,4,5\}$:
+\begin{enumerate}
+\item If $G$ is a planar graph have girth at least $m$ then $G$ has at least $2^{\epsilon v(G)}$ distinct $(8-m)$-colorings.
+\item If $G$ embeds on a surface $S$, has girth at least $m$, and is $(8-m)$-colorable then $G$ has  at least $2^{\epsilon v(G) - \alpha g(S)}$ distinct $(8-m)$-colorings.
+\end{enumerate}
+\end{corollary}
+\begin{proof}
+\paragraph{}
+Choose parameters
+$$\epsilon = \frac{(7-m)(m-2) - m}{(m-2){(8-m) \choose 2}}$$
+and
+$$\alpha = \frac{m}{(m-2){(8-m) \choose 2}}$$
+It is not hard to see for $m \in \{3,4,5\}$ both $\epsilon$ and $\alpha$ are strictly greater than $0$.
+\paragraph{}
+By Grotzsch's Theorem, planar graphs of girth $m$ are $(8-m)$-colorable for $m \in \{3,4,5\}$.
+Further $g(S) = 0$ if $S$ is the plane $S_0$.
+So it will suffice to prove the second claim of the Corollary.
+Let $G$ be a graph which embeds on a surface $S$, has girth at least $m$, and is $(8-m)$-colorable.
+From the previous Lemma, $G$ has at least
+$$2^{\frac{(7-m)v(G) - e(G)}{{(8-m)\choose 2}}}$$
+distinct $(8-m)$-colorings. Since each edge is incident with $2$ faces and each face contains at least $m$ edges we have
+$$f(G) \leq \frac{2}{m}e(G)$$
+Applying Euler's formula for general surfaces
+$$ v(G) - e(G) + f(G) = 2- g(S)$$
+to the left hand side of the above inequality, we see that
+$$ e(G) - v(G) + 2-g(S) \leq \frac{2}{m}e(G).$$
+Simplifying we observe that
+$$e(G) \leq \frac{m}{m-2}(v(G) - 2 + g(S))$$
+Therefore we have that
+\begin{align*}
+(7-m)v(G) - e(G) &\geq (7-m)v(G) -\frac{m}{m-2}(v(G) - 2 + g(S))\\
+&= \frac{(7-m)(m-2) - m}{m-2}v(G) + \frac{m}{m-2} (2-g(S)) \\
+&\geq \frac{(7-m)(m-2) - m}{m-2}v(G) - \frac{m}{m-2} (g(S))
+\end{align*}
+Therefore
+$$\frac{(7-m)v(G) - e(G)}{{(8-m)\choose 2}} \geq \epsilon v(G) - \alpha g(S)$$
+and hence $G$ has at least
+$$2^{\epsilon v(G) - \alpha g(S)}$$
+distinct colorings, as desired.
+\end{proof}
 \end{document}

### 2018-03-19 CO749-GraphColoring/assignments/A4.tex
problem 2 solution

@@
 \end{proof}
+\section{}
+\paragraph{}
+We may use the following theorems.
+\begin{theorem}\label{th:g-e-choosable}
+	If $G$ is a graph and $e\in E(G)$ such that $G-e$ is planar then $G$ is $5$-list-colorable.
+\end{theorem}
+\begin{theorem}\label{th:closeness}
+	There exists $r>0$ such that the following holds:
+
+	Let $G$ be a connected plane graph with $S\subseteq V(G)$ such that $G$ is $S$-critical with respect to some $5$-list-assignment $L$.
+If $S$ has at least two components then there exists two distinct components $S_1,S_2$ of $S$ such that
+	$$d(S_1,S_2) \leq r(|V(S_1)| + |V(S_2)|).$$
+\end{theorem}
+\paragraph{}
+We will prove the following theorem
+\begin{theorem}
+	There exists $d>0$ such that the following holds:
+
+	Let $G$ be a graph and $F \subset E(G)$ such that $G-F$ is planar. If for every two distinct edges $e_1, e_2 \in F$ we have $d(e_1, e_2) \geq d$ then $G$ is $5$-list-colorable.
+\end{theorem}
+\begin{proof}
+Choose $d = 4r^2+2r+1$ where $r$ is as in Theorem \ref{th:closeness}.
+\paragraph{}
+Suppose the theorem is false. Let $(G,F)$ be a minimal counterexample with respect to $|V(G)| + |E(G)| + |F|$.
+\paragraph{}
+Label the elements of $F$ as $\{e_1=u_1v_1, \dots, e_k = u_kv_k\}$.
+For each $e_i \in F$ let $P_i$ be a $u_i-v_i$ path of length at most $2r$ in $G-F$, provided such a path exists, and let $P_i$ be the empty graph otherwise.
+Let $S\subset G$ be the graph consisting of $V(F)$ and $\bigcup_{i=1}^k P_i$.
+\paragraph{}
+Hence we claim $S$ consists of a disjoint union of path components $P_i$ for $i \in \{1,\dots, k\}$ as well as disjoint vertex components $u_j$, $v_j$ where there is no $u_j-v_j$ path of length at most $2r$ in $G-F$.
+To see this, suppose $P_i$ and $P_j$ intersect at a vertex $x$ for $i\neq j$.
+Without loss of generality say $d(u_i,x) \leq d(v_i,x)$ and $d(u_j, x) \leq d(v_j, x)$.
+Then $d(u_i, x) \leq r$ and $d(u_j,x) \leq r$.
+So we have
+$$d(u,i,u,j) \leq r + r = 2r < d$$
+a contradiction. The argument follows similarly if a disjoint vertex component intersects a path component.
+\paragraph{}
+Now we claim that $S$ has at least two components.
+By our previous argument it suffices to show $|F| \geq 2$.
+Indeed by Theorem \ref{th:g-e-choosable} if $|F| \leq 1$ then $G$ is $5$-list-colorable, a contradiction.
+\paragraph{}
+Now we claim $G$ is $S$-critical.
+Let $H$ be a proper subgraph of $G$ containing $S$.
+By minimality of $G$, since $H-F$ is planar, and the distances between vertices in $H$ is at least their distance in $G$, $H$ is $5$-list-colorable.
+Let $\phi$ be a $5$-list-coloring of $H$.
+If $\phi$ extends to $G$ then $G$ is $5$-list-colorable, a contradiction.
+Hence $\phi$ does not extend.
+Thus $G$ is $S$-critical.
+\paragraph{}
+In order to apply Theorem \ref{th:closeness} with $G-F$ and $S$ it remains to verify $G-F$ is connected.
+Suppose $G-F$ is disconnected.
+Then there exist  $u_i,v_i \in F$ such that $u_i$ and $v_i$ lie in distinct components of $G-F$.
+Re-embed on the plane, if necessary, the components of $G-F$ containing $u_i$ and $v_i$ so $u_i$ and $v_i$ lie incident to the outer face.
+Then join $u_i$ and $v_i$ by their edge $e_i$ drawn as a curve in the outer face.
+Hence $G - (F\backslash \{e_i\})$ is planar.
+But then $(G, F\backslash\{e_i\})$ is a smaller counterexample than $(G,F)$ contradicting minimality.
+\paragraph{}
+By Theorem \ref{th:closeness} there exist distinct components $S_1$ and $S_2$ such that there is in $G-F$ is
+$$d(S_1,S_2) \leq r(|V(S_1)| + |V(S_2)|).$$
+First consider the case that $S_1 = u_i$ and $S_2=v_j$ are single vertex components. Then we have
+$$d(u,v) \leq r(1+1) = 2r.$$
+So if $i=j$ this contradicts that $u$ and $v$ are not joined by a path in $S$, and if $i\neq j$ this contradicts that
+$$d(e_i,e_j) \geq d > 2r.$$
+Now consider the case that $S_1 =P_i$ and $S_2 = P_j$ are path components. Then there exist $x_i$ on $P_i$ and $x_j$ on $P_j$ such that
+$$d(x_i, x_j) \leq r(|V(S_1)| + |V(S_2)|) \leq r(2r + 2r) = 4r^2.$$
+Without loss of generality say $d(u_i,x_i) \leq d(v_i,x_i)$ and $d(u_j, x_i) \leq d(v_j, x_i)$. Then $d(u_i, x_i) \leq r$ and $d(u_j,x_i) \leq r$. So we have
+$$d(u_i,u_j) \leq r + 4r^2 + r = 4r^2 + 2r < d$$
+contradicting that $d(e_i,e_j) \geq d.$ The argument where $S_1$ is a path component and $S_2$ is a vertex component follows similarly.
+\paragraph{}
+Therefore in any case we obtain a contradiction, as desired.
+\end{proof}
 \end{document}

### 2018-03-19 CO749-GraphColoring/assignments/A4.tex
question 3 preamble

@@
 \DeclarePairedDelimiter\norm{\lVert}{\rVert}%
+
+\DeclareMathOperator{\defi}{def}
 %END MACROS
@@
 \end{proof}
+
+\section{}
+We may use the following theorem.
+\begin{theorem}\label{th:def}
+	There exists $c>0$ such that the following holds:
+
+	Let $G$ be a plane graph, let $S$ be a connected subgraph of $G$, and let $L$ be a $5$-list-assignment for $G$. If $G$ is $S$-critical with respect to $L$ then
+	$$v(G|S) \leq c(\defi(G|S) - 1),$$
+	where $\defi(G|S) = e(G|S) - 3v(G|S).$
+\end{theorem}
+We will prove the following lemma, under the assumption that a minimum counterexample $G$ has a vertex in $V(G)\backslash V(S)$ with at least three neighbours in $S$.
+\begin{lemma}
+	Let $G$ be a plane graph with connected subgraph $S$ and let $L$ be a $5$-list-assignment for $G$. If $\phi$ is an $L$-coloring of $S$ that extends to an $L$-coloring of $G$ then
+	$$\log_2 E(\phi) \geq (v(G|S) - (c+1)\defi(G|S))/9,$$
+	where $E(\phi)$ is the number of extensions of $\phi$ to $G$ and $c$ is the constant of Theorem \ref{th:def}.
+\end{lemma}
+\begin{proof}
+	Suppose not. Let $(G,S,L)$ be a minimum counterexample, minimizing $v(G|S)$. We may assume $G$ has a vertex $v$ in $V(G)\backslash V(S)$ with at least three neighbours in $S$.
+	\paragraph{}
+
+\end{proof}
 \end{document}

### 2018-03-19 CO749-GraphColoring/assignments/A4.tex
preamble

@@
 \paragraph{}
-Suppose the theorem is false. Let $(G,F)$ be a minimal counterexample with respect to $|V(G)| + |E(G)| + |F|$.
+Suppose the theorem is false. Let $(G,F,L)$ be a minimal counterexample with respect to $|V(G)| + |E(G)| + |F|$. Here $L$ denotes a $5$-list-assignment.
 \paragraph{}
@@
 \end{proof}
+
+\section{}
+\paragraph{}
+We may use the following theorem.
+\begin{theorem}\label{th:four}
+	There exists $\delta_0 >0$ such that for every $\delta \in (0,\delta_0]$, there exists $\epsilon \in (0,\delta)$ such that if $G$ is a $4$-critical graph then
+	$$e(G) \geq \frac{(5+\epsilon)v(G) - 2 - \delta T(G)}{3},$$
+	where $T(G)$ is the maximum number of vertex disjoint $\leq 4$-cycles in $G$.
+\end{theorem}
+We will probve the following theorem.
+\begin{theorem}
+	There exists $c>0$ such that the following holds:
+
+	Let $G$ be a planar graph having girth at least five and let $S$ be a subgraph of $G$. If $G$ is $S$-critical for $3$-coloring then $|V(G)| \leq c|V(S)|$.
+\end{theorem}
+\begin{proof}
+
+\end{proof}
 \end{document}

### 2018-03-20 CO749-GraphColoring/assignments/A4.tex
problem 4

@@
 \begin{proof}
-
+	For $\delta_0$ as in Theorem \ref{th:four}, choose $\delta = \min\{\delta_0, \frac{1}{3}\}$, and let $\epsilon$ be the corresponding value in Theorem \ref{th:four}.
+Choose $c = (1+\frac{5}{\epsilon}) > 0$.
+Define for any graph $G$
+	$$d(G) = 3e(G) - (5+\epsilon)v(G)$$
+	and $d(G|S) = d(G) - d(S)$ for subgraph $S$ of $G$. Similar to the previous problem, it's clear that $d(G|S)$ satisfies the ``chain rule", that is if $S\subseteq H \subseteq G$ then
+	$$d(G|S) = d(G|H) + d(H|S).$$
+	\paragraph{Main Claim}
+	We claim for any $G$ and $S$ as in the statement of the Theorem, $d(G|S) \geq 0$.
+	\paragraph{}
+	Suppose for a contradiction this claim is false.
+Let $(G,S)$ be a counterexample minimizing $v(G|S)$.
+That is $G$ is a girth at least $5$ planar graph and $G$ is $S$-critical for $3$-coloring.
+By minimiamlity, $G$ is connected.
+	\paragraph{}
+	$G$ has at least two vertices not in $S$ or at least $3$ edges not in $S$ (otherwise, it is easy to extend any $3$-coloring of $S$ to $G$).
+Let $e \in \delta(S)$.
+Then $S+e$ is a strict subgraph of $G$.
+Let $\phi$ be a $3$-coloring of $S$ which extends to $S+e$ but not to $G$.
+Let $H$ be the smallest subgraph of $G$ containing $S$ such that $\phi$ does not extend to $H$.
+So $H$ contains $S+e$, i.e.
+$H$ contains $S$ as a strict subgraph.
+	\paragraph{}
+	We proceed by case distinction.
+Either $H=G$ or $H\subset G$.
+We first consider the case $H \subset G$.
+By minimality of $H$, every strict subgraph of $H$ can be colored by an extension of $\phi$.
+hence $H$ is $S$-critical.
+Since $H$ contains $S$, $G$ is $H$-critical.
+So by minimality, neither $(G,H)$ nor $(H,S)$ are counterexamples to the main claim.
+Thus we observe
+	$$d(G|S) = d(G|H) + d(H|S) \geq 0 + 0 = 0,$$
+	as desired.
+	\paragraph{}
+	The more complex case is when $H=G$.
+In this case we see $G$ is $\phi$-critical as the minimality of our choice of $H$ implies every strict subgraph of $G$ containing $S$ is colorable with an extension of $\phi$.
+Let $G_S[\phi]$ denote the graph obtained from $G$ by identifying vertices of $S$ according to their color class with respect to $\phi$.
+Let $x_1, x_2, x_3$ be vertices of $G_S[\phi]$ corresponding to the respective color classes of $\phi$.
+If necessary we add edges between each $x_i, x_j$ so they form a $K_3$.
+	\paragraph{}
+	Now we claim $G_S[\phi]$ is not $3$-colorable.
+Otherwise we can permute a $3$-coloring of $G_S[\phi]$ so it assigns color $i$ to $x_i$.
+This coloring then corresponds to an extension of $\phi$ which $3$-colors $G$ (by assigning $i$ each vertex of $G$ in $x_i$).
+Hence $G_S[\phi]$ contains a $4$-critical subgraph $W$.
+Let $S' = S \cup W - X$, where $X = W \cap \{x_1,x_2,x_3\}$.
+	\paragraph{}
+	We claim $W$ contains $G_S[\phi] - X$.
+Otherwise we can extend $\phi$ to a $3$-coloring of $W$.
+Therefore $S' = G$.
+Now if we define a potential function $p(G) = -d(G)$ we can apply the Potential Extension Lemma to see
+	$$p(G) = p(S') \leq p(S) + p(W) - p(X)$$
+	But then rearranging we see that
+	$$d(G|S) \geq p(X) + d(W).$$
+	Computing the potential of $X$ we see that either $p(X) = 5+\epsilon$, $p(X) = 7 + 2\epsilon$, or $p(X) = 6+3\epsilon$ depending on whether $|X| = 1, 2,$ or $3$ respectively.
+So $p(X) \geq 5+\epsilon$.
+	\paragraph{}
+	Applying Theorem \ref{th:four} to $4$-critical graph $W$ we see that
+	$$d(W) \geq -2 - \delta T(W)$$
+	Since $G$ is girth $5$, $T(W) \leq 3$ as the only cycles of length smaller than $5$ in $W$ intersect $\{x_1,x_2,x_3\}$ so there are at most $3$ vertex-disjoint such cycles. Hence
+	$$d(W) \geq -2 - 3\delta.$$
+	Thus substituting our lower bounds on $d(W)$ and $p(X)$ into the bound for $d(G|S)$ we see that
+	$$d(G|S) \geq 5+\epsilon -2-3\delta \geq 1 - 3\delta \geq 0$$
+	where the last inequality follows since $\delta \leq \frac{1}{3}$.
+	\paragraph{Finishing the proof}
+	Suppose that $G$ is a girth at least $5$ planar graph and $G$ is $S$-critical for $3$-coloring.
+Now we can use the Claim that $d(G|S) \geq 0$ to achieve the desired bound $|V(G)| \leq c|V(S)|$.
+Observe that
+	$$0 \leq d(G|S) = 3e(G|S) -5v(G|S) -\epsilon v(G|S)$$
+	Since $G$ is planar, girth at least $5$ we have $3e(G) -5v(G) \leq -10$. Substituting into the above yields
+	$$0 \leq -10 -3e(S) +5v(S) - \epsilon v(G|S).$$ Therefore
+	$$\epsilon v(G|S) \leq 5v(S)$$
+	and so 
+	$$v(G) \leq (1+\frac{5}{\epsilon}) v(S) = c v(S),$$
+	as desired.
 	\end{proof}

### 2018-03-20 CO749-GraphColoring/assignments/A4.tex
First case of q3

@@
 \begin{proof}
-	Suppose not. Let $(G,S,L)$ be a minimum counterexample, minimizing $v(G|S)$. We may assume $G$ has a vertex $v$ in $V(G)\backslash V(S)$ with at least three neighbours in $S$.
+	Suppose not. Let $(G,S,L, \phi)$ be a minimum counterexample, minimizing $v(G|S)$. We may assume $G$ has a vertex $v$ in $V(G)\backslash V(S)$ with at least three neighbours in $S$. Let $H = S+v$. 
 	\paragraph{}
-
+	We first consider the case that $v(G|S) = 1$, i.e. $G = S + v$.
+	\begin{align*}
+	\frac{v(G|S) - (c+1)\defi(G|S)}{9} &= \frac{1 - (c+1)e(G|S) +3(c+1)}{9} \\
+	&\begin{cases}
+		=\frac{1}{9}, &\text{if $|N(v)\cap v(S)| = 3$}  \\
+		\leq\frac{1-(c+1)}{9} <0, &\text{if $|N(v) \cap V(S)| \geq 4$}	\end{cases}
+	\end{align*}
+	Since $\phi$ has at least one extension to $G$ if $|N(v) \cap V(S)| \geq 4$ and $\log_2(1) = 0$, and $\phi$ has at least two extensions to $G$ if $|N(v) \cap V(S)| = 3$ (as at least two colors in $v$'s list are not used at neighbours) and $\log_2(2) = 1 > \frac{1}{9}$, the desired bound holds.
+	\paragraph{}
+	Therefore we may assume that $v(G|S) \geq 2$.
+Consider the case that $|N(v) \cap V(S)| \geq 4$.
+Let $\phi'$ be the extension of $\phi$ to $G$, restricted to $H$.
+Then $\phi'$ is an $L$-coloring of $H$ which extends to $G$.
+By minimality,
+	$$\log_2 E(\phi') \geq \frac{v(G|H) - (c+1)\defi(G|H)}{9}$$
+	Now $v(G|H) = v(G|S) - 1$.
 \end{proof}

### 2018-03-20 CO749-GraphColoring/assignments/A4.tex
solved it

--- 2018-03-20 CO749-GraphColoring/assignments/A4.tex
@@
 \begin{proof}
-	Suppose not. Let $(G,S,L)$ be a minimum counterexample, minimizing $v(G|S)$. We may assume $G$ has a vertex $v$ in $V(G)\backslash V(S)$ with at least three neighbours in $S$.
+	Suppose not. Let $(G,S,L, \phi)$ be a minimum counterexample, minimizing $v(G|S)$. We may assume $G$ has a vertex $v$ in $V(G)\backslash V(S)$ with at least three neighbours in $S$. Let $H = S+v$. 
 	\paragraph{}
-
+	We first consider the case that $v(G|S) = 1$, i.e. $G = S + v$.
+	\begin{align*}
+	\frac{v(G|S) - (c+1)\defi(G|S)}{9} &= \frac{1 - (c+1)e(G|S) +3(c+1)}{9} \\
+	&\begin{cases}
+		=\frac{1}{9}, &\text{if $|N(v)\cap v(S)| = 3$}  \\
+		\leq\frac{1-(c+1)}{9} <0, &\text{if $|N(v) \cap V(S)| \geq 4$}	\end{cases}
+	\end{align*}
+	Since $\phi$ has at least one extension to $G$ if $|N(v) \cap V(S)| \geq 4$ and $\log_2(1) = 0$, and $\phi$ has at least two extensions to $G$ if $|N(v) \cap V(S)| = 3$ (as at least two colors in $v$'s list are not used at neighbours) and $\log_2(2) = 1 > \frac{1}{9}$, the desired bound holds.
+	\paragraph{}
+	Therefore we may assume that $v(G|S) \geq 2$.
+Consider the case that $|N(v) \cap V(S)| \geq 4$.
+Let $\phi'$ be an extension of $\phi$ to $G$, restricted to $H$.
+Then $\phi'$ is an $L$-coloring of $H$ which extends to $G$.
+By minimality,
+	$$\log_2 E(\phi') \geq \frac{v(G|H) - (c+1)\defi(G|H)}{9}$$
+	Now $v(G|H) = v(G|S) - 1$. So we see that
+	\begin{align*}
+	\frac{v(G|H) - (c+1)\defi(G|H)}{9} &= \frac{v(G|S) - 1 -(c+1)\defi(G|S) +(c+1)|N(v)\cap V(S)| - 3(c+1)}{9} \\
+	&\geq \frac{v(G|S) - 1 -(c+1)\defi(G|S) +(c+1)4 - 3(c+1)}{9}\\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S)}{9}.
+	\end{align*}
+	Hence since $E(\phi) \geq E(\phi')$ this yields the desired bound.
+	\paragraph{}
+	Thus we may assume that $|N(v) \cap V(S)| = 3$.
+Yet again we distinguish cases.
+First suppose there exists $\phi_1$ and $\phi_2$ distinct $L$-colorings of $H$, extended from $\phi$, which extend to $G$.
+Without loss of generality suppose $E(\phi_1) \leq E(\phi_2)$.
+Then, using minimality,
+	\begin{align*}
+	\log_2E(\phi) &\geq \log_2(E(\phi_1) + E(\phi_2)) \\
+	&\geq \log_2(2E(\phi_1))\\
+	&= 1 + \log_2(E(\phi_1))\\
+	&\geq 1+ \frac{v(G|H) - (c+1)\defi(G|H)}{9}\\
+	&= 1 + \frac{v(G|S) - 1 -(c+1)\defi(G|S) +(c+1)|N(v)\cap V(S)| - 3(c+1)}{9}\\
+	&= 1 + \frac{v(G|S) -(c+1)\defi(G|S)}{9} + \frac{-1}{9} \\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S)}{9}.
+	\end{align*}
+	Thus yielding the desired bound.
+	\paragraph{}
+	Therefore, finally, we suppose without loss of generality $\phi_1$ and $\phi_2$ are distinct $L$-colorings of $H$ and $\phi_1$ extends to $G$, but $\phi_2$ does not.
+This is last case to consider as at least one $L$-coloring of $H$ must extend to $G$ as $\phi$ does.
+	\paragraph{}
+	Let $G' \subseteq G$ be a subgraph of $G$ that is $H$-critical with respect to $\phi_2$.
+Such $G'$ exists since $\phi_2$ does not extend to $G$.
+By Theorem \ref{th:def} $$v(G'|H) \leq c(\defi(G'|H) -1).$$
+	Let $\phi'$ now denote the restriction of an extension of $\phi$ to $G$ to $G'$. That is, $\phi'$ is an $L$-coloring of $G'$ that extends to $G$. So by minimality
+	$$\log_2(E(\phi')) \geq \frac{v(G|G') - (c+1)\defi(G|G')}{9}.$$
+	It is easy to see that $v(\cdot)$ and $\defi(\cdot)$ satisfy the ``chain rule", that is
+	$$\defi(G|H) = \defi(G|G') + \defi(G'|H)$$
+	and
+	$$v(G|S) - 1 = v(G|H) = v(G|G') + v(G'|H).$$
+	We make one final observation before the final bound
+	$$\defi(G|H) = \defi(G|S) |N(v)\cap V(S)| -3 = \defi(G|S) $$
+	Now we see that
+	\begin{align*}
+	\log_2E(\phi) &\geq \log_2E(\phi')\\
+	&\geq \frac{v(G|G') - (c+1)\defi(G|G')}{9} \\
+	&= \frac{v(G|S) -1 -v(G'|H) -(c+1)\defi(G|H) + (c+1)\defi(G'|H)}{9}\\
+	&=\frac{v(G|S) -1 -v(G'|H) -(c+1)\defi(G|S) + (c+1)\defi(G'|H)}{9}\\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S) +\defi(G'|H) + c - 1}{9} \\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S) +v(G'|H)/c + 1 + c - 1}{9}\\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S)}{9},
+	\end{align*}
+	as desired.
 \end{proof}

### 2018-03-20 CO749-GraphColoring/assignments/A4.tex
revision

@@
 	It is easy to see that $v(\cdot)$ and $\defi(\cdot)$ satisfy the ``chain rule", that is
-	$$\defi(G|H) = \defi(G|G') + \defi(G'|H)$$
+	\begin{equation}\label{chain-rule}
+	\defi(G|H) = \defi(G|G') + \defi(G'|H)
+	\end{equation}
 	and
-	$$v(G|S) - 1 = v(G|H) = v(G|G') + v(G'|H).$$
+	\begin{equation}\label{vgs-eq}
+		v(G|S) - 1 = v(G|H) = v(G|G') + v(G'|H).
+	\end{equation}
 	We make one final observation before the final bound
-	$$\defi(G|H) = \defi(G|S) |N(v)\cap V(S)| -3 = \defi(G|S) $$
+	\begin{equation}\label{defi-eq}
+	\defi(G|H) = \defi(G|S) |N(v)\cap V(S)| -3 = \defi(G|S) 
+	\end{equation}
 	Now we see that
@@
 	&\geq \frac{v(G|G') - (c+1)\defi(G|G')}{9} \\
-	&= \frac{v(G|S) -1 -v(G'|H) -(c+1)\defi(G|H) + (c+1)\defi(G'|H)}{9}\\
-	&=\frac{v(G|S) -1 -v(G'|H) -(c+1)\defi(G|S) + (c+1)\defi(G'|H)}{9}\\
-	&\geq \frac{v(G|S) -(c+1)\defi(G|S) +\defi(G'|H) + c - 1}{9} \\
-	&\geq \frac{v(G|S) -(c+1)\defi(G|S) +v(G'|H)/c + 1 + c - 1}{9}\\
+	&= \frac{v(G|S) -1 -v(G'|H) -(c+1)\defi(G|H) + (c+1)\defi(G'|H)}{9} &\text{by (\ref{chain-rule}) and (\ref{vgs-eq})}\\
+	&=\frac{v(G|S) -1 -v(G'|H) -(c+1)\defi(G|S) + (c+1)\defi(G'|H)}{9} &\text{ by (\ref{defi-eq})}\\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S) +\defi(G'|H) + c - 1}{9} &\text{ by Theorem \ref{th:def}}\\
+	&\geq \frac{v(G|S) -(c+1)\defi(G|S) +v(G'|H)/c + 1 + c - 1}{9} &\text{ by Theorem \ref{th:def}}\\
 	&\geq \frac{v(G|S) -(c+1)\defi(G|S)}{9},
@@
 	$G$ has at least two vertices not in $S$ or at least $3$ edges not in $S$ (otherwise, it is easy to extend any $3$-coloring of $S$ to $G$).
-Let $e \in \delta(S)$.
-Then $S+e$ is a strict subgraph of $G$.
-Let $\phi$ be a $3$-coloring of $S$ which extends to $S+e$ but not to $G$.
+Let $\phi$ be a $3$-coloring of $S$ which extends to a strict supergraph of $S$ but not to $G$.
 Let $H$ be the smallest subgraph of $G$ containing $S$ such that $\phi$ does not extend to $H$.
-So $H$ contains $S+e$, i.e.
-$H$ contains $S$ as a strict subgraph.
+So $H$ contains $S$ as a strict subgraph.
 	\paragraph{}
@@
 Let $G_S[\phi]$ denote the graph obtained from $G$ by identifying vertices of $S$ according to their color class with respect to $\phi$.
-Let $x_1, x_2, x_3$ be vertices of $G_S[\phi]$ corresponding to the respective color classes of $\phi$.
-If necessary we add edges between each $x_i, x_j$ so they form a $K_3$.
+Let $x_1, x_2, x_k$, $k\leq 3$, be vertices of $G_S[\phi]$ corresponding to the respective color classes of $\phi(S)$.
+If necessary we add edges between each $x_i, x_j$ so they form a complete subgraph.
 	\paragraph{}
 	Now we claim $G_S[\phi]$ is not $3$-colorable.
-Otherwise we can permute a $3$-coloring of $G_S[\phi]$ so it assigns color $i$ to $x_i$.
+Otherwise we can permute a $3$-coloring of $G_S[\phi]$ so it assigns color $i$ to each $x_i$.
 This coloring then corresponds to an extension of $\phi$ which $3$-colors $G$ (by assigning $i$ each vertex of $G$ in $x_i$).
 Hence $G_S[\phi]$ contains a $4$-critical subgraph $W$.
-Let $S' = S \cup W - X$, where $X = W \cap \{x_1,x_2,x_3\}$.
+Let $S' = S \cup W - X$, where $X = W \cap \{x_1, \dots, x_k\}$.
 	\paragraph{}
@@
 	$$d(W) \geq -2 - \delta T(W)$$
-	Since $G$ is girth $5$, $T(W) \leq 3$ as the only cycles of length smaller than $5$ in $W$ intersect $\{x_1,x_2,x_3\}$ so there are at most $3$ vertex-disjoint such cycles. Hence
+	Since $G$ is girth $5$, $T(W) \leq 3$ as the only cycles of length smaller than $5$ in $W$ intersect $\{x_1,\dots, x_k\}$ so there are at most $3$ vertex-disjoint such cycles. Hence
 	$$d(W) \geq -2 - 3\delta.$$

### 2018-05-31 CS761-RandomizedAlgorithms/assignments/a1.tex
started write up

--- 2018-01-19 CO749-GraphColoring/assignments/A1.tex
@@
 \usepackage{tkz-berge}
-\usepackage[braket, qm]{qcircuit}
-\usepackage{hyperref}
 
@@
 
-\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
-\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
 %END MACROS
@@
 
-\lhead{CO 749 - \today}
-\rhead{W. Justin Toth - A1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{\today}
+\rhead{W. Justin Toth CS761-Randomized Algorithms Assignment 1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \paragraph{}
-Let $G$ be a graph of minimum degree at least $3$ satisfying all of the following:
+We modify Karger's algorithm as follows:
+\begin{enumerate}
+\item While there are more than $\bf{k}$ vertices in the graph
 	\begin{itemize}
-	\item Every triangle contains at most one vertex of degree three.
-	\item If two vertices of degree three are adjacent then they are in triangles.
-	\item No two distinct triangles share a common edge.
+		\item Pick an edge uniformly at random and contract the two endpoints.
 	\end{itemize}
-Suppose for a contradiction that $\abs{E(G)} < 5\abs{V(G)}/3$.
-\begin{claim}\label{1.1}
-	For every $v \in V(G)$, if $d(v) = 3$ then $v$ has at most one neighbour of degree $3$.
-\end{claim}
+\item Output the edges between the remaining $k$ vertices.
+\end{enumerate}
+\begin{lemma}
+Let $G=(V,E)$ be a graph. The probability that the algorithm outputs a minimum $k$-way cut in $G$ is at least $$\frac{2(k-1)}{n(n-1)\dots(n-2(k-1)+1)}$$ where $n = |V|$.
+\end{lemma}
 \begin{proof}
-	Assume for a contradiction that there exist $u,v, w \in V(G)$ satisfying
-	$$d(u) = d(v) = d(w) = 3$$
-	and $uv, vw \in E(G)$, i.e.
-$u$ and $w$ are adjacent to $v$.
-But then, using the second property of $G$, $v$ is in a triangle.
-Denote this triangle by $C$.
-Since every triangle contains at most one vertex of degree three, $u$ and $w$ are not contained in $C$.
-But then since $d(v) = 3$, $v$ has at most one neighbour in $C$, a contradiction to $C$ being a $K_3$.
+Let $F\subseteq E$ be a minimum $k$-way cut in $G$.
+Let $G^i=(V^i,E^i)$ be the contracted graph at iteration $i$ of Karger's algorithm.
+By our stopping condition $|V^i| \geq k$.
+Since a cut in the contracted graph at each iteration is a cut in the original graph, the minimum $k$-way cut value of $G^i$ is at least $|F|$.
+\paragraph{}
+Let $S\subseteq V^i$ such that $|S|=k-1$.
+Then $\sum_{v \in S} |\delta(v)| \geq |F|$, for otherwise we have a cut $\bigcup_{v\in S} \delta(v)$ of size smaller than $F$ in $G^i$.
+Thus by partitioning the vertices of $G^i$ into sets of size $k-1$ we see that
+$$|E^i| \geq \floor{\frac{(n-i+1)|F|}{2(k-1)}}.$$
+Hence the probably that we pick an edge to contract in iteration $i$ from $F$ is at most
+$$ \frac{|F|}{\frac{(n-i+1)|F|}{2(k-1)}} \leq \frac{2(k-1)}{n-i+1}.$$
+Therefore the probability that $F$ survives until the end of the algorithm is at least
+$$(1-\frac{2(k-1)}{n})(1-\frac{2(k-1)}{n-1})(1-\frac{2(k-1)}{n-2})\dots(1-\frac{2(k-1)}{k+1}) = \frac{2(k-1)}{n(n-1)\dots(n-2(k-1)+1)}$$
+as desired.
 \end{proof}
 \paragraph{}
-We proceed by a discharging argument. For each vertex $v \in V(G)$ set the initial charge as
-$$ch_0(v) = d(v) - 4$$
-and for each edge $e \in E(G)$ set the initial charge
-$$ch_0(e) = 1/3.$$
-Then our total charge is, using handshaking,
-\begin{align*}
-\sum_{v\in V(G)} ch_0(v) + \sum_{e \in E(G)} ch_0(e) &= \sum_{v \in V(G)} d(v) - 4|V(G)| + \frac{1}{3} |E(G)|\\ &=  \frac{7}{3}|E(G)| - 4|V(G)\\ &< \frac{35}{9}|V(G)| - \frac{36}{9}|V(G)|\\ &< 0
-\end{align*}
-with the first inequality following from our assumption that $|E(G)| < 5|V(G)|/3$.
-\paragraph{}
-Redistribute the charge, to obtain $ch_1$, according to the following Rules:
-\begin{enumerate}
-	\item If edge $uv$ contains precisely one vertex $u$ such that $d(u) = 3$ then $uv$ sends $1/3$ charge to $u$.
-	\item If edge $uv$ is such that $d(u) \geq 4$ and $d(v) \geq 4$ and $u,v$ are in a triangle with vertex $w$ then $uv$ sends $1/3$ charge to $w$.
-\end{enumerate}
-\paragraph{}
-Since vertices sends no charge, clearly every vertex $v \in V(G)$ with $d(v) \geq 4$ has $$ch_1(v) = ch_0(v) \geq 0.$$ Now we will show that if $v$ is a vertex with $d(v) = 3$ then $ch_1(v) \geq 0$.
-From Claim \ref{1.1} $v$ is adjacent with at most one vertex of degree $3$.
-If $v$ is adjacent with no degree three vertices then each edge incident with $v$ sends $\frac{1}{3}$ charge to $v$ by Rule $1$ and hence
-$$ch_1(v) = ch_0(v) + 3\cdot\frac{1}{3} = 0.$$
-On the other hand, if $v$ is adjacent with one degree three vertex then $v$ receives $\frac{2}{3}$ charge from incident edges by Rule $1$.
-The remaining needed $\frac{1}{3}$ is obtained from Rule $2$ as follows.
-Since $v$ is adjacent with a degree three vertices, $v$ lies in a triangle, and since every triangle contains at most one degree three vertex the edge containing the other two vertices of this triangle sends $\frac{1}{3}$ charge to $v$.
-Therefore $ch_1(v) \geq 0$.
-\paragraph{}
-It remains to verify that each edge $uv$ satisfies $ch_1(uv) \geq 0$.
-Edge $uv$ can satisfy at most one of Rule $1$ and Rule $2$.
-If $uv$ satisfies Rule $1$ it sends $\frac{1}{3}$ charge and so $ch_1(uv) = 0$.
-Instead if $uv$ satisfies Rule $2$ it can only do so for one triangle, since no two distinct triangles share a common edge.
-So $uv$ only sends $\frac{1}{3}$ charge by Rule $2$.
-Therefore $ch_1(uv) \geq 0$.
-\paragraph{}
-Therefore we have a contradiction as our total charge is negative but $ch_1$ of each object is non-negative.$\blacksquare$
-
-\newpage
-\section{}
-\paragraph{}
-Assume Lemma $2$ is false. Let $G$ be a minimum counterexample. That is $G$ is a smallest planar graph satisfying:
-\begin{itemize}
-	\item The girth of $G$ is at least seven,
-	\item The minimum degree of $G$ is at least two, and
-	\item If $v$ is a degree $2$ vertex of $G$ with neighbours $u$ and $w$ then the degrees of $u$ and $w$ are each at least $4$.
-\end{itemize}  
-\paragraph{}
-Since the minimum degree is at least $2$, $G$ is not acyclic.
-So $G$ contains cycles, and each cycle is of length at least $7$.
-Hence every bounded face of $G$ is bounded by a cycle of length at least $7$ and the unbounded face is incident with at least $7$ edges.
-\paragraph{}
-We proceed by a discharging argument. For each vertex $v \in V(G)$ assign initial charge to $v$:
-$$ch_0(v) = 2d(v) - 6.$$
-For each face $f$ assign initial charge to $f$:
-$$ch_0(f) = |f| - 6.$$
-Then by our Lemma from class the sum of the initial charges is $-12$.
-By the girth and minimum degree of $G$, the charge of every face is at least $1$ and the charge of every vertex is non-negative except for degree $2$ vertices.
-\paragraph{}
-Discharge, establishing final charges $ch_1$ according to the following Rules:
-\begin{enumerate}
-	\item Each face $f$ sends $1$ charge to each incident vertex of degree $2$.
-	\item Each vertex of degree at least $4$ sends $\frac{1}{2}$ charge to each incident face.
-\end{enumerate}
-\paragraph{}
-If a vertex has degree at least $4$ then their initial charge is at least half their degree, so their final charge is non-negative after sending $\frac{1}{2}$ to each incident face.
-Degree $3$ vertices neither send nor receive charge and so their final charge is $0$.
-\paragraph{}
-If $v$ is vertex of degree $2$ then $v$ is incident with two faces, and so receives $2$ charge. Thus the final charge of $v$ is
-$$ch_1(v) = ch_0(v) + 2 = -2 + 2 = 0.$$
-Hence we have shown that the final charge of each vertex is non-negative. It remains to show the final charge of each face is non-negative to achieve the desired contradiction.
-\paragraph{}
-Let $f$ be a face of $G$. Let $q$ be the number of degree two vertices incident with $f$. Since the neighbours of any degree two vertex cannot be of degree two we observe that
-$$q \leq \floor{\frac{|f|}{2}}.$$
-Let $p$ be the number of vertices incident with $f$ of degree at least $4$. Since each degree two vertex has both neighbours of degree four, we have
-$$p \geq q.$$
-We have that $f$ loses $q$ charge by Rule $1$ and gains $\frac{p}{2}$ charge by Rule $2.$ Thus the final charge of $f$ is
-$$ch_1(f) = ch_0(f) -q + p \geq |f| -6 -\frac{q}{2} \geq \frac{3}{4}|f| - 6.$$
-So the final charge of $f$ is non-negative when $|f| \geq 8$.
-\paragraph{}
-Therefore we may assume that $|f| = 7$. In this case we may strengthen our bound on $p$ by observing that
-$p \geq q+1$
-by a parity argument.
-Indeed if $p = q$ then $f$ is a cycle of vertices alternating degree $2$ and degree $4$, but the number of vertices on $f$ is odd, so this cannot happen.
-Now we tighten our estimate of the final charge:
-$$ch_1(f) \geq |f| - 6 -\frac{q}{2} + 1 \geq \frac{3}{4}\cdot 7 - 6 + 1 \geq 0.$$
-Thus we have shown every face has non-negative final charge, a contradiction. $\blacksquare$
-
-\newpage
-\section{}
+Using this lemma, we observe that if we repeat Karger's algorithm $t$ times, and take the best cut we find, then our failure probability is at most TODO gonna get $O(n^{2k})$ not $O(n^k)$.
+$$\big(1 -  \frac{2(k-1)}{n(n-1)\dots(n-2(k-1)+1)}\big)^t \leq e^\frac{-2(k-1)t}{n(n-1)\dots(n-2(k-1)+1)}.$$
 \end{document}

### 2018-05-31 CS761-RandomizedAlgorithms/assignments/a1.tex
wrote up problem 1

@@
 \paragraph{}
-Using this lemma, we observe that if we repeat Karger's algorithm $t$ times, and take the best cut we find, then our failure probability is at most TODO gonna get $O(n^{2k})$ not $O(n^k)$.
+Using this lemma, we observe that if we repeat Karger's algorithm $t$ times, and take the best cut we find, then our failure probability is at most 
 $$\big(1 -  \frac{2(k-1)}{n(n-1)\dots(n-2(k-1)+1)}\big)^t \leq e^\frac{-2(k-1)t}{n(n-1)\dots(n-2(k-1)+1)}.$$
+So we can achieve a fixed failure probability of at most $\epsilon \in (0,1)$ if we take 
+$$t = \ceil{\frac{\ln(\epsilon)}{-2(k-1)}}n(n-1)\dots(n-2(k-1)+1) \leq n^{ck}\cdot n^{2(k-1)+1} \leq n^{O(k)}$$
+for some fixed constant $c$ depending on $\epsilon$.
+
+\newpage
+\section{}
 \end{document}

### 2018-05-31 CS761-RandomizedAlgorithms/assignments/a1.tex
wrote up through q3

@@
 \section{}
+\paragraph{}
+Let $i > m$. Let $H(i)$ be the event that you hire candidate $i$. Let $B(i)$ be the event that candidate $i$ is the best candidate. Then $E_i = H(i) \cap B(i)$. So using Bayes formula we see that
+$$Pr(E_i) = Pr(H(i) \mid B(i)) \cdot Pr(B(i)).$$
+Now since the candidates are distributed uniformly at random $Pr(B(i)) = \frac{1}{n}$.
+\begin{lemma}
+Let $F(i)$ be the event that the best candidate among the first $i$ candidates appears in the first $m$ candidates. Then
+$$Pr(H(i) \mid B(i)) = Pr(F(i-1)).$$
+\end{lemma}
+\begin{proof}
+\paragraph{}
+Given that $i$ is the best candidate, we hire $i$ if and only if we reject the first $i-1$ candidates.
+If we reject the first $i-1$ candidates then the best candidate among the first $i-1$ appears in the first $m-1$ candidates.
+Otherwise, it would not be rejected when it appears, as it is better than all previously seen candidates.
+On the other hand if the best candidates among the first $i-1$ appears in the first $m$ candidates, then we will reject the first $i-1$ candidates as none of them are better than a candidate which appears in the first $m$.
+Thus $H(i)$ happens given $B(i)$ if and only if $F(i-1)$ happens, i.e.
+the claim holds.
+\end{proof}
+\paragraph{}
+Using this lemma we have
+$$Pr(H(i) \mid B(i)) = Pr(F(i-1)) = \frac{m}{i-1}$$
+since there are $m$ chances among $i-1$ that the best candidate among the first $i-1$ appears in the first $m$.
+Therefore $Pr(E_i) = \frac{m}{n(i-1)}$.
+\paragraph{}
+Now the event $E$ is the disjoint union of events $E_1, \dots, E_n$. So we observe 
+$$Pr(E) = Pr(\bigcup_{i} E_i) = \sum_{i} Pr(E_i) = \frac{m}{n}\sum_{i=m+1}^n \frac{1}{i-1}$$
+with the last equality following from our previous observations and the fact that $Pr(E_i) = 0$ for $i \leq m$.
+\paragraph{}
+We want to show that $Pr(E)$ can get arbitrarily close to $\frac{1}{e}$ for an appropriate choice of $m$. First we see that
+$$Pr(E) =  \frac{m}{n}\sum_{i=m+1}^n \frac{1}{i-1} = \frac{m}{n}(\sum_{j=1}^{n-1}\frac{1}{j} - \sum_{j=1}^{m-1}\frac{1}{j}) =\frac{m}{n}(H_{n-1} - H_{m-1}).$$
+Using that $\ln(n) \leq H_n \leq \ln(n+1)$ we have
+$$Pr(E) \geq \frac{m}{n}(\ln(n-1) - \ln(m)) = \frac{m}{n}\ln(\frac{n-1}{m}).$$
+If we choose $m = \frac{n-1}{e}$ then
+$$Pr(E) \geq \frac{1}{e}\cdot\frac{n-1}{n}\ln(e) = \frac{1}{e}\cdot\frac{n-1}{n}$$
+and taking the limit as $n$ tends to $\infty$ yields
+$$\lim_{n\rightarrow \infty} Pr(E) \geq \frac{1}{e}.$$
+\newpage
+\section{}
+\paragraph{}
+Let $X$ be a random variable counting the number of cycles in a random permutation.
+Let $X_i$ be a random variable counting the number of cycles of length $i$ in a permutation.
+Then $X = \sum_i X_i$.
+So by linearity of expectation
+$$E[X] = \sum_i E[X_i].$$
+Now let $X_i^j$ be an indicator variable for $j \in {n\choose i}$ forms a length $i$ cycle in the random permutation. Then 
+$$Pr(X^j_i) = \frac{(i-1)!(n-i)!}{n!}$$
+since there are $n!$ permutations, and $(i-1)!(n-i)!$ permutations where $j$ forms a length $i$ cycle.
+To see this, consider a permutation where $j$ forms a length $i$ cycle.
+There are $(i-1)!$ ways to arrange the elements of $j$ in a length $i$ cycle, and $(n-i)!$ ways to permute the other elements.
+Therefore
+$$E[X_i] = \sum_{j \in {n\choose i}}E[X_i^j] = {n\choose i} \frac{(i-1)!(n-i)!}{n!} = \frac{1}{i}.$$
+Then
+$$E[X] = \sum_{i=1}^n E[X_i] = \sum_{i=1}^n \frac{1}{i} = H_n.$$
+\newpage
+\section{}
 \end{document}

### 2018-06-01 CS761-RandomizedAlgorithms/assignments/a1.tex
q4

--- 2018-05-31 CS761-RandomizedAlgorithms/assignments/a1.tex
@@
 \section{}
+\paragraph{}
+Let $i > m$. Let $H(i)$ be the event that you hire candidate $i$. Let $B(i)$ be the event that candidate $i$ is the best candidate. Then $E_i = H(i) \cap B(i)$. So using Bayes formula we see that
+$$Pr(E_i) = Pr(H(i) \mid B(i)) \cdot Pr(B(i)).$$
+Now since the candidates are distributed uniformly at random $Pr(B(i)) = \frac{1}{n}$.
+\begin{lemma}
+Let $F(i)$ be the event that the best candidate among the first $i$ candidates appears in the first $m$ candidates. Then
+$$Pr(H(i) \mid B(i)) = Pr(F(i-1)).$$
+\end{lemma}
+\begin{proof}
+\paragraph{}
+Given that $i$ is the best candidate, we hire $i$ if and only if we reject the first $i-1$ candidates.
+If we reject the first $i-1$ candidates then the best candidate among the first $i-1$ appears in the first $m-1$ candidates.
+Otherwise, it would not be rejected when it appears, as it is better than all previously seen candidates.
+On the other hand if the best candidates among the first $i-1$ appears in the first $m$ candidates, then we will reject the first $i-1$ candidates as none of them are better than a candidate which appears in the first $m$.
+Thus $H(i)$ happens given $B(i)$ if and only if $F(i-1)$ happens, i.e.
+the claim holds.
+\end{proof}
+\paragraph{}
+Using this lemma we have
+$$Pr(H(i) \mid B(i)) = Pr(F(i-1)) = \frac{m}{i-1}$$
+since there are $m$ chances among $i-1$ that the best candidate among the first $i-1$ appears in the first $m$.
+Therefore $Pr(E_i) = \frac{m}{n(i-1)}$.
+\paragraph{}
+Now the event $E$ is the disjoint union of events $E_1, \dots, E_n$. So we observe 
+$$Pr(E) = Pr(\bigcup_{i} E_i) = \sum_{i} Pr(E_i) = \frac{m}{n}\sum_{i=m+1}^n \frac{1}{i-1}$$
+with the last equality following from our previous observations and the fact that $Pr(E_i) = 0$ for $i \leq m$.
+\paragraph{}
+We want to show that $Pr(E)$ can get arbitrarily close to $\frac{1}{e}$ for an appropriate choice of $m$. First we see that
+$$Pr(E) =  \frac{m}{n}\sum_{i=m+1}^n \frac{1}{i-1} = \frac{m}{n}(\sum_{j=1}^{n-1}\frac{1}{j} - \sum_{j=1}^{m-1}\frac{1}{j}) =\frac{m}{n}(H_{n-1} - H_{m-1}).$$
+Using that $\ln(n) \leq H_n \leq \ln(n+1)$ we have
+$$Pr(E) \geq \frac{m}{n}(\ln(n-1) - \ln(m)) = \frac{m}{n}\ln(\frac{n-1}{m}).$$
+If we choose $m = \frac{n-1}{e}$ then
+$$Pr(E) \geq \frac{1}{e}\cdot\frac{n-1}{n}\ln(e) = \frac{1}{e}\cdot\frac{n-1}{n}$$
+and taking the limit as $n$ tends to $\infty$ yields
+$$\lim_{n\rightarrow \infty} Pr(E) \geq \frac{1}{e}.$$
+\newpage
+\section{}
+\paragraph{}
+Let $X$ be a random variable counting the number of cycles in a random permutation.
+Let $X_i$ be a random variable counting the number of cycles of length $i$ in a permutation.
+Then $X = \sum_i X_i$.
+So by linearity of expectation
+$$E[X] = \sum_i E[X_i].$$
+Now let $X_i^j$ be an indicator variable for $j \in {n\choose i}$ forms a length $i$ cycle in the random permutation. Then 
+$$Pr(X^j_i) = \frac{(i-1)!(n-i)!}{n!}$$
+since there are $n!$ permutations, and $(i-1)!(n-i)!$ permutations where $j$ forms a length $i$ cycle.
+To see this, consider a permutation where $j$ forms a length $i$ cycle.
+There are $(i-1)!$ ways to arrange the elements of $j$ in a length $i$ cycle, and $(n-i)!$ ways to permute the other elements.
+Therefore
+$$E[X_i] = \sum_{j \in {n\choose i}}E[X_i^j] = {n\choose i} \frac{(i-1)!(n-i)!}{n!} = \frac{1}{i}.$$
+Then
+$$E[X] = \sum_{i=1}^n E[X_i] = \sum_{i=1}^n \frac{1}{i} = H_n.$$
+\newpage
+\section{}
+\paragraph{}
+We want to apply a Chernoff bound.
+So we will need to measure our run time in terms of some independent coin flips.
+If we can bound, for each element $x$ of our array, the depth of recursion before it is assigned a singleton array, then we can bound the runtime.
+Let $x$ be an element of our input array.
+Define $\ell_{x,i}$ to be the length of the array $x$ is assigned to at recursion depth $i$ of our quicksort algorithm.
+Note that our algorithm terminates at recursion depth $i$ if $\ell_{x,i}=1$ for all $x$.
+\paragraph{}
+Let $X_{x,i}$ be an indicator random variable indicating $1$ if $\ell_{x,i+1} \leq \frac{3}{4}\ell_{x,i}$, and taking value $0$ otherwise.
+Intuitively, $X_{x,i}$ indicates that we make a good split of the array containing $x$ at depth $i$ when quicksort is called on that array segment.
+Let $S_{x,i}$ be the array containing $x$ at recursion depth $i$.
+If our algorithm selects any element the interval $[\frac{\ell_{x,i}}{4}, \frac{3\ell_{x,i}}{4}]$ from the sorted version $S_{x,i}$ then $\ell_{x,i+1} \leq \frac{3}{4}\ell_{x,i}$.
+That means choosing one of at least $\frac{1}{2}$ the elements of $S$ as the pivot will yield to a good split, i.e.
+$$Pr(X_{x,i}) \geq \frac{1}{2}.$$
+So we can think of $X_{x,i}$ as a fair coin flip. Further each $X_{x,i+1}$ is independent of $X_{x,i}$ for all $i$.
+\paragraph{}
+Let $X_{x} = \sum_{i=1}^{k} X_{x,i}$ where $k=100\ln(n)$.
+Then if $X_{x} \geq d$ the length of the array containing $x$ at recursion depth $k$ is at most $\frac{3}{4}^d n$.
+If we want this to be a singleton then we need $$d = \frac{\ln(n)}{\ln{4/3}} \leq 4\ln(n) < 50\ln(n) = \frac{k}{2}$$
+So now observing $E[X_{x}] =\frac{k}{2}$ we apply a Chernoff bound (for the lower tail, with $\mu = \frac{k}{2}$ and $\delta = \frac{1}{2}$)
+$$Pr(X_{x} \leq \frac{1}{2}\frac{k}{2}) \leq e^{-\frac{k}{2}\frac{1}{8}} \leq e^{\frac{-50\ln(n)}{8}} \leq \frac{1}{n^6}$$
+Applying the union bound we see that
+$$Pr(\bigcup_x(X_{x} \leq \frac{1}{2}\frac{k}{2})) \leq \sum_{x} Pr(X_{x} \leq \frac{1}{2}\frac{k}{2})  \leq \frac{1}{n^5}.$$
+Therefore with inverse polynomial probability, there is some element that is not in a singleton array after recursion depth of $100 \ln(n)$.
+Since each level of the recursion takes $n$ time, we see that with high probability the quicksort algorithm terminates in $100n\ln(n)$ time.
 \end{document}

### 2018-06-01 CS761-RandomizedAlgorithms/assignments/a1.tex
5ab5ab5ab5ab5ab

@@
 Since each level of the recursion takes $n$ time, we see that with high probability the quicksort algorithm terminates in $100n\ln(n)$ time.
+
+\newpage
+\section{}
+\subsection{a}
+\paragraph{}
+\begin{align*}
+M_{X^2}(t) &= E[e^{tX^2}] \\
+&= \int_{-\infty}^\infty e^{tx^2}f(x)dx \\
+&=\int_{-\infty}^\infty \frac{1}{\sqrt{2\pi}}e^{tx^2-\frac{1}{2}x^2}dx \\
+&= \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi}}e^{\frac{-1}{2}(1-2t)x^2}dx \\
+&= \frac{1}{\sqrt{1-2t}}\int_{-\infty}^\infty \frac{1}{\sqrt{2\pi}}e^{z^2}dz &\text{Taking $z=\sqrt{1-2t}x$} \\
+&= \frac{1}{\sqrt{1-2t}} &\text{Using $\int_{-\infty}^\infty f(x)dx = 1$}.
+\end{align*}
+\subsection{b}
+\paragraph{}
+Using Taylor expansion we have
+\begin{align*}
+E[X^4] = M_{X^2}^{(2)}(0).
+\end{align*}
+Now observe that
+$$M_{X^2}^{(2)}(t) = \frac{3}{(1-2t)^{5/2}}$$
+and so
+$$M_{X^2}^{(2)}(0) = 3.$$
+Therefore $E[X^4] = 3$.
+\subsection{c}
+
 \end{document}

### 2018-06-02 CS761-RandomizedAlgorithms/assignments/a1.tex
writing 5c
still going

@@
 \subsection{c}
+\paragraph{}
+We have the following
+\begin{align*}
+Pr(Y > 1+\epsilon) &= Pr(\sum_{i=1}^kX_i^2 > k(1+\epsilon))\\
+&=Pr(e^{t\sum_{i=1}X_i^2} > e^{tk(1+\epsilon)})\\
+&\leq \frac{E[e^{t\sum_{i=1}X_i^2}]}{e^{tk(1+\epsilon)}} \\
+&= \frac{\Pi_{i=1}^kE[e^{t\sum_{i=1}^kX_i^2}]}{e^{tk(1+\epsilon)}} &\text{by independence}\\
+&=M_{X^2}(t)^ke^{-tk(1+\epsilon)}\\
+&=\sqrt{1-2t}^{-k}e^{-tk(1+\epsilon)} \\
+&= (1-2t)^{-k/2}e^{-tk(1+\epsilon)}\\
+&= \frac{1}{1-2t}^{k/2}e^{-tk(1+\epsilon)}
+\end{align*}
 

### 2018-06-03 CS761-RandomizedAlgorithms/assignments/a1.tex
q6

--- 2018-06-01 CS761-RandomizedAlgorithms/assignments/a1.tex
@@
 \subsection{c}
+\paragraph{}
+Let $t$ be a value to be chosen later. We have the following
+\begin{align*}
+Pr(Y > 1+\epsilon) &= Pr(e^{tY}> e^{t(1+\epsilon)} \\
+&\leq E[e^{tY}]\cdot e^{-t(1+\epsilon)}  &\text{Using Markov's inequality} \\
+&= (\Pi_{i=1}^k E[e^{\frac{t}{k}X_i^2}])e^{-t(1+\epsilon)} &\text{Using independence of $X_i^2$'s} \\
+&=(\frac{1}{1-\frac{2}{k}t})^{\frac{k}{2}}e^{-t(1+\epsilon)} &\text{by $5a$}.
+\end{align*}
+Taking $t=\frac{k}{2}\frac{\epsilon}{1+\epsilon}$ we have
+\begin{align*}
+Pr(Y>1+\epsilon) &\leq (1 - \frac{\epsilon}{1+\epsilon})^\frac{-k}{2}e^{\frac{-k}{2}\epsilon}\\
+&=e^{-\frac{k}{2}(\ln(1-\frac{\epsilon}{1+\epsilon}) + \epsilon)}
+\end{align*}
 
+\newpage
+\section{}
+\paragraph{}
+We apply the following theorem from Lecture $4$ (here $H$ is the graph obtain by the uniform sampling algorithm):
+\begin{theorem}
+	Set $p=\frac{9\ln n}{\epsilon^2 c}$ where $c$ is the min-cut value of $G$. Then $H$ is a $(1\pm \epsilon)$-cut-approximator of $G$ with $O(p\cdot |E(G)|)$ edges with probability $1-\frac{4}{n}$.
+\end{theorem}
+We suppose that $G$ is a complete graph where each edge has the same weight, say $w$. In $G$ the value of a cut $\delta(S)$ is
+$$w|\delta(S)| = w|S|(n-|S|)$$
+and this function is minimized when $|S| = 1$, i.e. $S$ is a singleton. Therefore the minimum value of a cut in $G$ is $c = w(n-1)$. Therefore applying the above theorem we have
+$$|E(H)| \leq O(p|E(G)|) = O(\frac{9\ln n}{\epsilon^2 w (n-1)}n(n-1)) \leq O(\frac{n\ln n}{\epsilon^2})$$
+as desired.
+\paragraph{Part 2}
+Now we consider an unweighted complete graph $G$.
+Suppose for a contradiction the uniform sampling algorithm can obtain a  $(1\pm \epsilon)$-cut-approximator $H$ of $G$ with $o(n\ln n)$ edges.
+To achieve this our sampling probability $p$ can be at most $\frac{\ln n}{c n}$ for some constant $c>1$.
+Heuristically, lets take
+$$p= \frac{\ln n}{100 n}.$$
+Now let $v \in V(G)$. Then the probability that the degree of $v$ in $H$ is $0$ is
+$$Pr(d_H(v) = 0) = (1- \frac{\ln n}{100n})^{n-1}$$
+since the the degree of $v$ is $0$ in $H$ if and only if every one of the $n-1$ edges incident with $v$ is not sampled by the algorithm.
+So now we bound the probability that this does not happen at any vertex of the graph.
+Heuristically we treat the events $d_H(v) = 0$ over $v \in V$ as independent events.
+Then
+\begin{align*}
+Pr(d_H(v) \geq 1, \forall v \in V) &= \Pi_{v\in V} (1-Pr(d_H(v) = 0)) \\
+&=(1-(1- \frac{\ln n}{100n})^{n-1})^n \\
+&\leq e^{-n(1- \frac{\ln n}{100n})^{n-1}} &\text{using $1+x \leq e^x$} \\
+\end{align*} 
+So with extremely low probability $d_H(v) \geq 1$, for all $v \in V$. Hence, with high probability there exists $v \in V$ with $d_H(v) = 0$. Therefore for the cut $\delta(v)$, 
+$$w(\delta_H(v)) = 0$$
+while 
+$$(1\pm \epsilon)\delta_G(v) = (1\pm \epsilon)(n-1)$$
+so when $0 <\epsilon < 1$ and $n\geq 2$, $H$ utterly fails to be a $(1\pm \epsilon)$-cut-approximator of $G$.
+\newpage
+\section{}
 \end{document}

### 2018-06-04 CS761-RandomizedAlgorithms/assignments/a1.tex
5c done

--- 2018-06-01 CS761-RandomizedAlgorithms/assignments/a1.tex
@@
 \subsection{c}
-
+(Collaboration with Sharat, Harry and Logan)
+\paragraph{}
+Let $t$ be a value to be chosen later. We have the following
+\begin{align*}
+Pr(Y > 1+\epsilon) &= Pr(e^{tY}> e^{t(1+\epsilon)} \\
+&\leq E[e^{tY}]\cdot e^{-t(1+\epsilon)}  &\text{Using Markov's inequality} \\
+&= (\Pi_{i=1}^k E[e^{\frac{t}{k}X_i^2}])e^{-t(1+\epsilon)} &\text{Using independence of $X_i^2$'s} \\
+&=(\frac{1}{1-\frac{2}{k}t})^{\frac{k}{2}}e^{-t(1+\epsilon)} &\text{by $5a$}.
+\end{align*}
+Taking $t=\frac{k}{2}\frac{\epsilon}{1+\epsilon}$ we have
+\begin{align*}
+Pr(Y>1+\epsilon) &\leq (1 - \frac{\epsilon}{1+\epsilon})^\frac{-k}{2}e^{\frac{-k}{2}\epsilon}\\
+&=e^{-\frac{k}{2}(\ln(1-\frac{\epsilon}{1+\epsilon}) + \epsilon)}
+\end{align*}
+Therefore it suffices to show $$\ln(1+\epsilon) + \epsilon \geq \frac{\epsilon^2}{4}.$$
+This is equivalent to showing
+$$\epsilon^2 - 4\epsilon +4\ln(1+\epsilon) \leq 0$$
+for $0 \leq \epsilon \leq 1$.
+Let $f(\epsilon) = \epsilon^2 - 4\epsilon +4\ln(1+\epsilon)$. We want to take the derivative, showing $f(\epsilon)$ is decreasing on the interval $[0,1]$.
+We compute
+$$f'(\epsilon) = 2\epsilon - 4 + \frac{4}{1+\epsilon} = \frac{2\epsilon(\epsilon-1)}{1+\epsilon} \leq 0$$
+where the last inequality follows since $\epsilon \leq 1$. So $f'(\epsilon) \leq 0$ on $[0,1]$ and so therefore $f(\epsilon)$ is a continuous decreasing function on $[0,1]$.
+\paragraph{}
+Therefore we can bound $f(\epsilon)$ on $[0,1]$ by $f(1)$. We have
+$$f(1) = 1 - 4 + 4\ln(2) = -3 + 4\ln(2) \leq -3 +2.8 <0$$
+and so $f(\epsilon) \leq 0$ as desired.
+\newpage
+\section{}
+\paragraph{}
+We apply the following theorem from Lecture $4$ (here $H$ is the graph obtain by the uniform sampling algorithm):
+\begin{theorem}
+	Set $p=\frac{9\ln n}{\epsilon^2 c}$ where $c$ is the min-cut value of $G$. Then $H$ is a $(1\pm \epsilon)$-cut-approximator of $G$ with $O(p\cdot |E(G)|)$ edges with probability $1-\frac{4}{n}$.
+\end{theorem}
+We suppose that $G$ is a complete graph where each edge has the same weight, say $w$. In $G$ the value of a cut $\delta(S)$ is
+$$w|\delta(S)| = w|S|(n-|S|)$$
+and this function is minimized when $|S| = 1$, i.e. $S$ is a singleton. Therefore the minimum value of a cut in $G$ is $c = w(n-1)$. Therefore applying the above theorem we have
+$$|E(H)| \leq O(p|E(G)|) = O(\frac{9\ln n}{\epsilon^2 w (n-1)}n(n-1)) \leq O(\frac{n\ln n}{\epsilon^2})$$
+as desired.
+\paragraph{Part 2}
+Now we consider an unweighted complete graph $G$.
+Suppose for a contradiction the uniform sampling algorithm can obtain a  $(1\pm \epsilon)$-cut-approximator $H$ of $G$ with $o(n\ln n)$ edges.
+To achieve this our sampling probability $p$ can be at most $\frac{\ln n}{c n}$ for some constant $c>1$.
+Heuristically, lets take
+$$p= \frac{\ln n}{100 n}.$$
+Now let $v \in V(G)$. Then the probability that the degree of $v$ in $H$ is $0$ is
+$$Pr(d_H(v) = 0) = (1- \frac{\ln n}{100n})^{n-1}$$
+since the the degree of $v$ is $0$ in $H$ if and only if every one of the $n-1$ edges incident with $v$ is not sampled by the algorithm.
+So now we bound the probability that this does not happen at any vertex of the graph.
+Heuristically we treat the events $d_H(v) = 0$ over $v \in V$ as independent events.
+Then
+\begin{align*}
+Pr(d_H(v) \geq 1, \forall v \in V) &= \Pi_{v\in V} (1-Pr(d_H(v) = 0)) \\
+&=(1-(1- \frac{\ln n}{100n})^{n-1})^n \\
+&\leq e^{-n(1- \frac{\ln n}{100n})^{n-1}} &\text{using $1+x \leq e^x$} \\
+\end{align*} 
+So with extremely low probability $d_H(v) \geq 1$, for all $v \in V$. Hence, with high probability there exists $v \in V$ with $d_H(v) = 0$. Therefore for the cut $\delta(v)$, 
+$$w(\delta_H(v)) = 0$$
+while 
+$$(1\pm \epsilon)\delta_G(v) = (1\pm \epsilon)(n-1)$$
+so when $0 <\epsilon < 1$ and $n\geq 2$, $H$ utterly fails to be a $(1\pm \epsilon)$-cut-approximator of $G$.
+\newpage
+\section{}
 \end{document}

### 2018-06-04 CS761-RandomizedAlgorithms/assignments/a1.tex
drafted q7

@@
 \section{}
+\subsection{a}
+\paragraph{}
+Let $F$ be an optimal multicut.
+Let $e_1 \dots e_n$ be an orthonormal basis of $\R^n$.
+Let $K_1, \dots, K_c$ be the connected components of $G-F$, with $c\leq n$.
+For all $i \in [c]$ and $j \in V(K_i)$ assign $v_j = e_i$.
+Now the norm of each $v_i$ is equal to $1$ and since $s_i$ and $t_i$ are in different components of $G-F$ by feasibility of $F$ we have $v_{s_i}\cdot v_{t_i} = 0$ as we started with an orthonormal basis and assigned $v_{s_i}$ and $v_{t_i}$ different basis vectors.
+Therefore our solution $\{v_j \in \R^n: j \in V\}$ is feasible for the vector program.
+\paragraph{}
+Notice $v_i\cdot v_j = 0$ if and only if $i$ and $j$ are in different connect components of $G-F$. Otherwise, i.e. if $i$ and $j$ are in the same $K_\ell$ then $v_i\cdot v_j = 1$. So we have
+$$\sum_{ij\in E} (1-v_i\cdot v_j) = \sum_{ij \in F} 1 = |F|.$$
+Hence the optimal value of the vector program is at most the optimal value of a multicut, and therefore is a relaxation of the minimum multicut problem.
+\subsection{b}
+\paragraph{}
+The vertices $i$ and $j$ of edge $ij$ are in different regions if and only if there exists some $r_\ell$ such that $v_i\cdot r_\ell  \geq 0$ and $ v_j \cdot r_\ell < 0$ or vice versa.
+For any given $r_\ell$ this occurs, as in the maximum cut problem from our notes, with probability $\frac{\theta}{\pi}$ where $\theta$ is the angle between $v_i$ and $v_j$.
+\paragraph{}
+Using our inequality from the hint we have
+$$\frac{\theta}{\pi} \leq \sqrt{\frac{1-\cos(\theta)}{2}}$$
+Since $v_i$ and $v_j$ are of norm $1$, $\cos(\theta) = v_i\cdot v_j$. Thus
+$$\frac{\theta}{\pi} \leq \sqrt{1-v_i\cdot v_j}.$$
+Now we have
+$$Pr(\text{$i$ and $j$ in different regions}) = \sum_{\ell = 1}^t Pr(\text{$r_\ell$ separates $v_i$, $v_j$}) \leq \sum_{\ell = 1}^t \sqrt{1-v_i\cdot v_j} = t\sqrt{1-v_i\cdot v_j}$$
+as desired.
+\subsection{c}
+\paragraph{}
+We remove all edges incident to $s_i$ when we do not delete all edges on some $s_i-t_j$ path for some $j$.
+For all edges on some $s_i-t_j$ path to be preserved, necessarily $v_{s_i}$ and $v_{t_j}$ lie in the same region.
+Since $v_{s_i}$ and $v_{t_j}$ are orthogonal the probability they are not separated by some $r_\ell$ is
+$$1- \frac{\pi}{2}\frac{1}{\pi} = \frac{1}{2}.$$
+So the probability they not separated by any $r_\ell$ is
+$$(\frac{1}{2})^t.$$
+Now there are at most $\Delta$ sinks $t_j$ incident on $s_i$ and so we have
+$$Pr(\text{delete all edges on $s_i$}) \leq \sum_{j=1}^tPr(\text{$s_i-t_j$ path survives}) \leq \frac{\Delta}{2^t}$$
+as desired.
+\subsection{d}
+\paragraph{}
+Let $X$ be a random variable counting the number of edges removed.
+Let $X_{ij}^1$ be an indicator variable indicating that $ij$ is removed in the first phase.
+Let $X_{s_i}^2$ be an indicator variable indicating that all edges incident to $s_i$ are removed in the second phase.
+Then 
+\begin{align*}
+E[X] &= \sum_{ij \in E} Pr(X^1_{ij}) + \sum_{ij\in E} Pr(X^2_{s_i}) \\
+&= \sum_{ij \in E} t\sqrt{1-v_iv_j} + \sum_{s_i} Pr(X^2_{s_i}) \\
+&=
+\end{align*}
 \end{document}

### 2018-06-04 CS761-RandomizedAlgorithms/assignments/a1.tex
finished a1

@@
 \usepackage{tkz-berge}
-
+\usepackage{bbm}
 %Macros
@@
 \subsection{c}
-(Collaboration with Sharat, Harry and Logan)
+($\textbf{Collaboration}$ with Sharat, Harry and Logan)
 \paragraph{}
@@
 \begin{align*}
-Pr(Y > 1+\epsilon) &= Pr(e^{tY}> e^{t(1+\epsilon)} \\
+Pr(Y > 1+\epsilon) &= Pr(e^{tY}> e^{t(1+\epsilon)} )\\
 &\leq E[e^{tY}]\cdot e^{-t(1+\epsilon)}  &\text{Using Markov's inequality} \\
@@
 \paragraph{}
-We remove all edges incident to $s_i$ when we do not delete all edges on some $s_i-t_j$ path for some $j$.
+Fix some sink $s_i$.
+Let $T = \{t_j: (s_i,t_j) \text{ is a source-sink pair}\}$.
+We remove all edges incident to $s_i$ in the second phase of the algorithm when we do not delete all edges on some $s_i-t_j$ path for some $t_j \in T$.
 For all edges on some $s_i-t_j$ path to be preserved, necessarily $v_{s_i}$ and $v_{t_j}$ lie in the same region.
-Since $v_{s_i}$ and $v_{t_j}$ are orthogonal the probability they are not separated by some $r_\ell$ is
-$$1- \frac{\pi}{2}\frac{1}{\pi} = \frac{1}{2}.$$
-So the probability they not separated by any $r_\ell$ is
+Since $v_{s_i}$ and $v_{t_j}$ are orthogonal, by the constraints of the vector program, the angle between them is $\frac{\pi}{2}$.
+Hence the probability they are not separated by some $r_\ell$ is
+$$1- \frac{\pi}{2}\cdot \frac{1}{\pi} = \frac{1}{2}.$$
+So the probability they not separated by any $r_\ell$ for $\ell \in [t]$ is
 $$(\frac{1}{2})^t.$$
-Now there are at most $\Delta$ sinks $t_j$ incident on $s_i$ and so we have
-$$Pr(\text{delete all edges on $s_i$}) \leq \sum_{j=1}^tPr(\text{$s_i-t_j$ path survives}) \leq \frac{\Delta}{2^t}$$
+Now $|T| \leq \Delta$ and so we have
+$$Pr(\text{delete all edges on $s_i$}) \leq \sum_{t_j \in T}Pr(\text{$s_i-t_j$ path survives}) \leq \sum_{t_j \in T}\frac{1}{2^t} \leq \frac{\Delta}{2^t}$$
 as desired.
@@
 Let $X$ be a random variable counting the number of edges removed.
-Let $X_{ij}^1$ be an indicator variable indicating that $ij$ is removed in the first phase.
-Let $X_{s_i}^2$ be an indicator variable indicating that all edges incident to $s_i$ are removed in the second phase.
-Then 
-\begin{align*}
-E[X] &= \sum_{ij \in E} Pr(X^1_{ij}) + \sum_{ij\in E} Pr(X^2_{s_i}) \\
-&= \sum_{ij \in E} t\sqrt{1-v_iv_j} + \sum_{s_i} Pr(X^2_{s_i}) \\
-&=
-\end{align*}
+Let $X_{ij}^1$ be an indicator variable indicating that edge $ij \in E$ is removed in the first phase.
+Let $X_{s_i}^2$ be an indicator variable indicating that all edges incident to source $s_i$ are removed in the second phase.
+By linearity of expectation we have
+$$E[X] = \sum_{ij \in E} Pr(X_{ij}^1) + \sum_{i \in [k]}Pr(X_{s_i}^2).$$
+Now by $7c)$, $Pr(X_{s_i}^2) \leq \frac{\Delta}{2^t}$, so
+$$ \sum_{i \in [k]}Pr(X_{s_i}^2) \leq k\frac{\Delta}{2^t} = k\Delta 2^{-\ceil{\log_2(\Delta /\epsilon)}} \leq k\epsilon \leq \epsilon |E|.$$
+Thus
+$$E[X] \leq \sum_{ij \in E} Pr(X_{ij}^1) + \epsilon |E|.$$
+From $7b)$, $Pr(X_{ij}^1) \leq t\sqrt{1-v_i\cdot v_j}$, so we have
+$$\sum_{ij \in E} Pr(X_{ij}^1) \leq t\sum_{ij \in E} \sqrt{1-v_i\cdot v_j}$$
+Let $\mathbbm{1} \in \R^E$ be the vector of all ones. Let $q \in \R^E$ be the vector whose entries are defined by $q_{ij} = \sqrt{1-v_i\cdot v_j}$. Then 
+$$\sum_{ij \in E} Pr(X_{ij}^1) \leq t \langle\mathbbm{1}, q_{ij} \rangle \leq t \sqrt{\langle \mathbbm{1}, \mathbbm{1}\rangle \langle q,q\rangle}$$
+with the last inequality following from Cauchy-Schwarz. Now we see
+$$\sqrt{\langle \mathbbm{1}, \mathbbm{1}\rangle \langle q,q\rangle} = \sqrt{|E|}\sqrt{\sum_{ij \in E} (1-v_i\cdot v_j)} = \sqrt{|E|}\sqrt{\epsilon |E|} = \sqrt{\epsilon}|E|$$
+with the second equality following from the observation that $\sum_{ij \in E}(1-v_i\cdot v_j)$ is the objective function of the vector program. Hence, using the definition of $t$,
+$$\sum_{ij \in E} Pr(X_{ij}^1) \leq t\sqrt{\epsilon}|E| = O(\log_2(\Delta/\epsilon)\sqrt{\epsilon}|E|)$$
+and so
+$$E[X] \leq O(\log_2(\Delta/\epsilon)\sqrt{\epsilon}|E|) + \epsilon |E| = O (\sqrt{\epsilon}|E| + \sqrt{\epsilon}\log(\Delta/\epsilon)|E|)$$
+as desired.
 \end{document}

### 2018-06-13 CS761-RandomizedAlgorithms/assignments/a2.tex
Wrote up q1 of a2

--- 2018-02-22 CO749-GraphColoring/assignments/A3.tex
@@
 \lhead{\today}
-\rhead{W. Justin Toth - A3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CS761-Randomized Algorithms Assignment n2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+%Problem 1
 \section{}
-A $k$-critical graph $G$ is $m$-\emph{splittable} (with $2\le m \le k-1$) if for every vertex $z\in V(G)$ and every split of $z$ into vertices $z_1,\ldots, z_m$ of positive degree , the resulting graph has a $(k-1)$-coloring such that $z_1, \ldots, z_m$ have distinct colors.
-We say such $G$ is \emph{splittable} if $G$ is $m$-splittable for every $m$ such that $2\le m \le k-1$.
-\begin{lemma}
-	An Ore-composition of two splittable $k$-critical graphs is splittable $k$-critical.
-\end{lemma}
-\begin{proof}
-	Let $G_1$ and $G_2$ be two splittable $k$-critical graphs.
-Let $G$ be the Ore-composition of $G_1$ and $G_2$ obtained by deleting edge $xy \in E(G_1)$ and splitting vertex $z \in V(G_2)$ into $z_1$ and $z_2$ of positive degree and identifying $x$ with $z_1$ and $y$ with $z_2$.
-We denote the identified vertex for $x$ with $z_1$ as $z_x$, and for $y$ with $z_2$ as $z_y$ in $G$.
 \paragraph{}
-	We first show that $G$ is $k$-critical.
-Suppose for a contradiction $G$ has a $(k-1)$-coloring $\phi$.
-If $\phi(z_x) \neq \phi(z_y)$ then $\phi$ restricted to $G_1$ (taking in the obvious manner $\phi(x) = \phi(z_x)$ and $\phi(y) = \phi(z_y)$) is a $(k-1)$-coloring of $G_1$.
-If $\phi(z_x) = \phi(z_y)$ then $\phi$ restricted to $G_2$, and takiing $\phi(z) = \phi(z_x) = \phi(z_y)$ yields a $(k-1)$-coloring of $G_2$.
-In either case we contradicted that $G_1$ and $G_2$ do not have a $(k-1)$-coloring.
-Therefore $G$ is not $(k-1)$-colorable.
+Let $v_1, \dots, v_m \in \{0,1\}^\ell$ be a set of vectors such that any subset of size at most $k$ is linearly independent modulo $2$.
+We denote the $j$-th entry of $v_i$ by $v_{i,j}$.
+Let $u$ be chosen uniformly at random from $\{0,1\}^\ell$.
+Let $X_i$ be the random variable defined by
+$$X_i = (\sum_{j=1}^\ell v_{i,j}u_j) \mod 2.$$
+We want to show that $X_1, \dots, X_m$ are $k$-wise independent random bits. That is to say, for all $I \subseteq [m]$ with $1\leq |I| \leq k$, and $x \in \{0,1\}^m$,
+$$Pr(\bigcap_{i\in I} X_i = x_i) = \Pi_{i\in I} Pr(X_i = x_i).$$
+We begin by observing that if the vector $v_i$ is the $0$ vector for any $i \in I$ then the set of vectors $V := \{v_i : i \in I\}$ fails to be linearly independent (modulo $2$), which contradicts our hypothesis.
+Hence for all $i \in I$, there exists $j \in [\ell]$ such that $v_{i,j} \neq 0$.
+Further from linear independence we observe that $\ell \geq k$.
 \paragraph{}
-	To finish verifying that $G$ is $k$-critical we need to show that every strict subgraph of $G$ is $(k-1)$-colorable.
-Let $H \subseteq G$ be a strict subgraph of $G$.
-We proceed by case-distinction on $H$.
-		\subparagraph{} Case $1$: $z_x$ is not contained in $H$ and $z_y$ is not contained in $H$.
-Then $H$ is the disjoint union of a strict subgraph of $G_1$ and a strict subgraph of $G_2$.
-Since $G_1$ and $G_2$ are $k$-critical we can obtain a $(k-1)$-coloring of their respective strict subgraphs in $H$ and combine them to obtain a $(k-1)$-coloring of $H$.
-		\subparagraph{} Case $2:$ Exactly one of $z_x, z_y$ is in $H$.
-Without loss of generality say $z_x \in V(H)$ and $z_y \not\in V(H)$.
-Then treating $z_x$ as $x$ in $G_1$ and $z$ in $G_2$ we see that $H$ is the union of two strict subgraphs $H_1 \subseteq G_1$ and $H_2 \subseteq G_2$, overlapping only at $z_x$.
-We see $H_2$ is a strict subgraph since the degree of $z_1$ and $z_2$ are both positive.
-Again we invoke criticality of $G_1$ and $G_2$ to obtain $(k-1)$-colorings of $H_1$ and $H_2$, which we join into a $(k-1)$-coloring of $H$.
-Note we permute the coloring of $H_1$ so the two colorings agree at $z_x$ if necessary.
-		\subparagraph{} Case $3:$.
-Both $z_x$ and $z_y$ are in $H$.
-Let $G_2'$ denote the graph $G_2$ with $z$ split into $z_1$ and $z_2$.
-Then $H$ is the union of two subgraphs $H_1 \subseteq G_1$ and $H_2 \subseteq G_2'$ overlapping only at $\{z_x,z_y\}$.
-Since $H$ is a strict subgraph of $G$ one of $H_1$ or $H_2$ is a strict subgraph of $G_1-xy$ or $G_2'$ respectively.
-		\subparagraph{} First suppose $H_1$ is a strict subgraph of $G_1-xy$.
-Then invoking the criticality of $G_1$, we obtain that $H_1 + xy$ has a $(k-1)$-coloring $\phi_1$.
-Then $\phi_1$ is a $(k-1)$-coloring of $H_1$ where $\phi_1(x)\neq \phi_1(y)$.
-Since $G_2$ is $2$-splittable, there exists a $(k-1)$-coloring $\phi_2$ of $G_2'$.
-Note that $\phi_2(z_1) \neq \phi_2(z_2)$ for otherwise setting $\phi_2(z) = \phi_2(z_1) = \phi_2(z_2)$ yields a $(k-1)$-coloring of $G_2$.
-Hence we may permute the colors of $\phi_2$ so $\phi_2(z_1) = \phi_1(x)$ and $\phi_2(z_2) = \phi_1(y)$.
-Then the coloring obtained by combining $\phi_1$ and $\phi_2$ is a $(k-1)$-coloring of $H$.
-		\subparagraph{}
-		So we may assume $H_1$ is not a strict subgraph of $G_1-xy$, i.e.
-$H_1 = G_1-xy$.
-Then $H_2$ is a strict subgraph of $G_2'$.
-By identifying $z_x$ and $z_y$ as $z$ in $H_2$ we obtain a strict subgraph of $G_2$.
-By criticality of $G_2$, this subgraph has a $(k-1)$-coloring $\phi_2$.
-By setting $\phi_2(z_x) = \phi_2(z_y) = \phi_2(z)$ we obtain a $(k-1)$-coloring of $H_2$ wherein $z_x$ and $z_y$ receive the same color.
-Invoking criticality of $G_1$, we obtain a $(k-1)$-coloring $\phi_1$ of $H_1$ ($=G_1-xy$).
-Observe that $\phi_1(x) = \phi_1(y)$, for otherwise $\phi_1$ is a $(k-1)$-coloring of $G_1$.
-But then we may permute the colors of $\phi_2$ so $\phi_2(z_1) = \phi_1(x)$ and $\phi_2(z_2) = \phi_1(y)$.
-Again the coloring obtained by combining $\phi_1$ and $\phi_2$ is a $(k-1)$-coloring of $H$, as desired.
-	\paragraph{Splittability}
-	Thus we have shown that $G$ is $k$-critical.
-It remains to verify that $G$ is splittable.
-First consider $v \in V(G) \backslash\{z_x,z_y\}$.
-Consider $G'$, where we split $v$ into $v_1, \dots, v_m$ each with positive degree, $2 \leq m \leq k-1$.
-If $v \in V(G_1)$ then apply the corresponding split $v_1, \dots, v_m$ to $G_1$, and use that $G_1$ is splittable to obtain a $(k-1)$-coloring $\phi_1$.
-Observe $\phi_1(x) \neq \phi_1(y)$ since $xy \in E(G_1)$.
-Since $G_2$ is $2$-splittable, $G - G_1$ has a $(k-1)$-coloring $\phi_2$.
-Since $G_2$ is not $(k-1)$-colorable, $\phi_2(z_x) \neq \phi_2(z_y)$.
-Again we may permute colors of $\phi_2$ so $\phi_1$ and $\phi_2$ agree on $z_x$ and $z_y$, obtaining a $(k-1)$-coloring of $G'$.
+On the one hand, for any $i \in I$, $v_i \neq 0$ and so
+$$Pr(X_i = x_i) = \frac{1}{2}$$
+since $X_i = x_i$ if and only if
+$$P_i := \sum_{j \in s(v_i)} u_j = \begin{cases} \text{odd}, &\text{if $x_i = 1$} \\
+\text{even}, &\text{otherwise}
+\end{cases}$$
+where $s(v_i) := \{j \in [\ell] : v_{i,j} \neq 0\}$. Note that since $s(v_i) \neq \emptyset$ and $u$ is chosen uniformly at random, the parity $P_i$ is chosen uniformly at random. Therefore
+$$\Pi_{i \in I} Pr(X_i = x_i) = \Pi_{i \in I} \frac{1}{2} = \frac{1}{2^{|I|}}.$$
 \paragraph{}
-	If $v \in V(G_2)$ then apply the corresponding split $v_1, \dots, v_m$ to $G_2$ and use splittability to obtain a $(k-1)$-coloring $\phi_2$.
-Setting $\phi_2(z_x) = \phi_2(z_y) = \phi_2(z)$ we obtain a $(k-1)$-coloring of $G-G_1$.
-By criticality, $G_1-xy$ has a $(k-1)$-coloring $\phi_1$.
-Since $G_1$ is not $(k-1)$-colorable, $\phi_1(x) = \phi_1(y)$.
-Hence we may permute colors of $\phi_2$ so $\phi_1$ and $\phi_2$ agree on $z_x$ and $z_y$, obtaining a $(k-1)$-coloring of $G'$.
-	\paragraph{}
-	It remains to consider splitting one $z_x$ or $z_y$ in $G$.
-Without loss of generality let $G'$ be the graph obtained by splitting $z_x$ into $z_x^1, \dots, z_x^m$, each with positive degree, $2 \leq m \leq k-1$.
-Split $x$ in $G_1$ into $x^1, \dots, x^m$ with neighbourhoods $N(x^i) = N(z_x^i)\cap V(G_1)$ yielding graph $G_1'$.
-Obtain a $(k-1)$-coloring $\phi_1$ of $G_1'$ by splittability of $G$.
-	\subparagraph{}Case $1:$ $\phi_1(x^i) = \phi_1(y)$ for some $i \in \{1,\dots, m\}$.
-Split $z$ in $G_2$ into $z^1,\dots, z^m$ with neighbourhoods $N(z^j) = N(z_x^j) \cap V(G_2)$ for $j \neq i$ and $$N(z^i) = (N(z_x^i) \cup N(z_y)) \cap V(G_2)$$
-	obtaining graph $G_2'$.
-Color $G_2'$ by splittability of $G_2$ obtaining $(k-1)$-coloring $\phi_2$.
-Setting $\phi_2(z_y) = \phi_2(z^i)$ we obtain a $(k-1)$-coloring of $(G-G_1) \cup \{z_x^1, \dots, z_x^m, z_y\}$.
-Since $\phi_1$ and $\phi_2$ give all the split vertices different colors, and give $z_y$ the same color $z_x^i$ we may permute $\phi_1$ and $\phi_2$ to agree on $\{z_x^1, \dots, z_x^m, z_y\}$ obtaining a $(k-1)$-coloring of $G'$.
-	\subparagraph{}Case $2:$ $\phi_1(x^i) \neq \phi_2(y)$ for all $i \in \{1,\dots, m\}$.
-Notice in this case $m < k-1$ by the pidgeonhole principle since $\phi_1$ assigns different colors to each split vertex.
-Split $z$ in $G_2$ into $z^1, \dots, z^m, z^{m+1}$ with neighbours $N(z^i) = N(z_x^i) \cap V(G_2)$ for $i \leq m$ and set $N(z^{m+1}) = N(z_y) \cap V(G_2)$.
-Applying splitability we obtain a $(k-1)$-coloring $\phi_2$.
-Now both $\phi_1$ and $\phi_2$ give $\{z_x^1, \dots, z_x^m, z_y\}$ different colors, so we may permute their colors to agree on $\{z_x^1, \dots, z_x^m, z_y\}$, and join the colorings to obtain a coloring of $G'$, as desired.
-	\end{proof}
-\begin{corollary}
-	If $G$ is a $k$-Ore graph then $G$ is splittable $k$-critical.
-\end{corollary}
-\begin{proof}
-	Suppose not.
-Let $G$ be a minimum counterexample.
-Then $G$ cannot be the Ore-composition of two $k$-Ore graphs $G_1$, $G_2$, for otherwise $G_1$ and $G_2$ are splittable $k$-critical by minimality and hence $G$ is splittable $k$-critical by the previous Lemma.
-	\paragraph{}
-	Thus $G$ is isomorphic to $K_k$.
-It remains to show $K_k$ is splittable.
-Suppose we want to split vertex $v \in V(G)$ into $v_1, \dots, v_m$ with $2\leq m \leq k-1$ such that each $v_i$ has positive degree, obtaining graph $G'$.
-We can trivially color $G - v$ with $k-1$ colors, obtaining coloring $\phi$.
-Now we can extend $\phi$ to $G'$ since each $v_i$ has degree at most $k-2$ and hence $|\phi(N(v_i))| \leq k-2$ and there is a free color available to assign $v_i$.
-This contradicts that
-\end{proof}
-
+Now on the other hand, $\bigcap_{i \in I} X_i = x_i$ happens if and only if 
+$$ Vu = x_I$$
+where we define the matrix $V$ with $|I|$ rows and $\ell$ columns so that $V_{i,j} = v_{i,j}$, and $x_I$ is the vector $x$ restricted to entries indexed by $I$.
+Here we are taking all operations in the field $\F_2$.
+Since the rank of $V$ is $|I|$ we can find by Gaussian elimination a permuation matrix $E$ (after possibly permuting the columns and associated indices of $u$, without loss of generality) so that
+$$EV = \begin{bmatrix}
+I & B
+\end{bmatrix}$$
+for some binary matrix $B$ with $\ell - |I|$ columns. Then
+$$\begin{bmatrix} I & B \end{bmatrix}u = Ex_I$$
+and hence the choices of entries for $u_1, \dots, u_{|I|}$ are uniquely fixed given the choices of entries for $u_{|I| + 1}, \dots, u_{\ell}$.
+So there are $2^{\ell - |I|}$ choices for $u$ satisfying our matrix equation.
+There are $2^\ell$ total choices for $u$ overall, and so we have
+$$Pr(\bigcap_{i \in I} X_i = x_i) = \frac{2^{\ell-|I|}}{2^\ell}  = \frac{1}{2^{|I|}} = \Pi_{i \in I} Pr(X_i = x_i)$$
+as desired. $\blacksquare$
+%Problem 2
 \newpage
 \section{}
-\subsection{a}
-\begin{lemma}
-	If $G$ is the Ore-composition of two graphs $G_1$ and $G_2$ then $\alpha(G) \leq \alpha(G_1) + \alpha(G_2)$.
-\end{lemma}
-\begin{proof}
-Let $G$ be the Ore-composition of graphs $G_1$ and $G_2$ obtained by deleting edge $xy \in E(G_1)$ and splitting vertex $z \in V(G_2)$ into $z_1$ and $z_2$ of positive degree and identifying $x$ with $z_1$ and $y$ with $z_2$.
-We denote the identified vertex for $x$ with $z_1$ as $z_x$, and for $y$ with $z_2$ as $z_y$ in $G$.
-Let $S \subseteq V(G)$ be an independent set in $G$ such that $|S| = \alpha (G)$.
-Then treating $z_x$ as $x$ and $z_y$ as $z$ we can partition $S$ into two independent sets $S_1\subseteq V(G_1)$ and $S_2\subseteq V(G_2)$ as $S_1 = V(G_1) \cap S$ and $S_2 = S\backslash S_1$. Thus
-$$\alpha(G) = |S| = |S_1| + |S_2| \leq \alpha(G_1) + \alpha(G_2)$$
-as desired.
-\end{proof}
 
-\subsection{b}
-\begin{lemma}
-	If $G$ is $k$-Ore then $\alpha(G) = \frac{v(G) - 1}{k-1}$.
-\end{lemma}
-\begin{proof}
-	We proceed by induction on $v(G)$. The smallest $k$-Ore graph is $K_k$, for which
-	$$\alpha(K_k) = 1 = \frac{k-1}{k-1} = \frac{v(G) -1}{k-1}.$$
-	So the base case for induction is satisfied.
-	\paragraph{}
-	So we consider a $k$-Ore graph $G$ that is the Ore-composition of $G_1$ and $G_2$ and suppose for induction that
-	$$\alpha(G_1) = \frac{v(G_1) - 1}{k-1} \quad\text{ and }\quad \alpha(G_2) = \frac{v(G_2) -1}{k-1}.$$
-	We observe from the previous Lemma, and induction, that
-	$$\alpha(G) \leq \alpha(G_1) + \alpha(G_2) = \frac{v(G_1) +v(G_2)- 2}{k-1} = \frac{v(G) + 1 - 2}{k-1} = \frac{v(G) - 1}{k-1}.$$
-	Thus $\alpha(G) \leq \frac{v(G) - 1}{k-1}$. To show $\alpha(G) \geq \frac{v(G) - 1}{k-1}$ it suffices to construct an independent set in $G$ of size $\frac{v(G) - 1}{k-1}$.
-	\paragraph{}
-	For $i \in \{1, \dots, 2\}$ let $S_i$ be in an independent set in $G_i$ satisfying $$|S_i| = \alpha(G_i) = \frac{v(G_i) - 1}{k-1}.$$ We use the same notion as the above problems for $x,y, z, z_x, z_y$.
-Since $xy \in E(G_1)$, $|\{x,y\} \cap S_1| \leq 1$.
-Without loss of generality suppose $y \not\in S_1$.
-Treating $x$ as $z_x$ and $z$ as $z_y$ we observe that
-	$$S := S_1 \cup S_2$$
-	is an independent set of $G$ of size $|S_1| + |S_2|$. But then we see that
-	$$|S| = |S_1| + |S_2| =  \frac{v(G_1) +v(G_2)- 2}{k-1} = \frac{v(G) - 1}{k-1}.$$
-	Thus $\alpha(G) \geq \frac{v(G) - 1}{k-1}$, and therefore by induction the Lemma is proven.
-\end{proof}
-
-\subsection{c}
-\paragraph{}
-From our work in class it suffices to show $c\alpha(R') \leq c\alpha(R) + c\alpha(W) - c\alpha(K_{|X|}) + c$ for $c \geq 0$.
-Since we know the core is always non-empty, $\alpha(K_{|X|}) = 1$ and so the problem reduces to showing
-$$\alpha(R') \leq \alpha(R) + \alpha(W).$$
-Let $S$ be an independent set of $G[R']$ such that $|S| = \alpha(R')$.
-Since $R'= V(W) \cup R - \{v_1, \dots, v_{k-1}\}$, where $\{v_1, \dots, v_{k-1}\}$ are the set of color-identified vertices, as in the usual definition.
-So we have
-$$|S| \leq |S\cap (V(W) - \{v_1, \dots, v_k\}| + |S\cap R|$$
-On the one hand $S\cap (V(W) - \{v_1, \dots, v_k\})$ is an independent set in $W$ and $S \cap R$ is an indepenent set in $G[R]$, so then
-$$ \alpha(R') = |S| \leq |S\cap (V(W) - \{v_1, \dots, v_k\}| + |S\cap R| \leq \alpha(W) + \alpha(R)$$
-as desired. $\blacksquare$
+%Problem 3
 \newpage
 \section{}
-\subsection{a}
-\begin{lemma}
-If $G$ is the Ore-composition of $G_1$ and $G_2$ then
-$$T_k(G) \geq T_k(G_1-e) + T_k(G_2 - z) \geq T_k(G_1) + T_k(G_2) - 2$$
-where $e = xy$ is the replaced edge of $G_1$ and $z$ is the split vertex of $G_2$.
-\end{lemma}
-\begin{proof}
-Let $H_1 \subseteq G_1- e$ be the subgraph of $G_1 - e$ which achieves the optimum in 
-$$T_k(G_1-e) := \max_{H \subseteq G_1 - e} T_k(H).$$
-Similarly let $H_2 \subseteq G_2 - z$ be the subgraph of $G_2 - z$ which achieves the optimum in $T_k(G_2 - z)$. Then $H_1 \cup H_2$ is a disjoint union of cliques in $G$. So we have
-$$T_k(G_1-e) + T_k(G_2 - z) = T_k(H_1) + T_k(H_2) = T_k(H_1 \cup H_2) \leq T_k(G).$$
-So the first inequality of the lemma is proven.
-\paragraph{}
-Now let $H_1' \subseteq G_1$ be the subgraph of $G_1$ which achieves the optimum in $T_k(G_1)$, and let $H_2' \subseteq G_2$ be the subgraph of $G_2$ which achieves the optimum in $T_k(G_2)$.
-If $|e \cap V(H'_1)| \leq 1$ then $H'_1$ is a subgraph of $G_1 - e$ and so $T_k(G_1) \leq T_k(G_1-e)$.
-Otherwise, i.e.
-$ e\in E(H'_1)$ then $H'_1 - x$ is a subgraph of $G_1 - e$, and $T_k(H'_1-x) = T_k(H'_1) - 1$.
-Thus
-$$T_k(G_1 - e) \geq T_k(H'_1-x) \geq T_k(H'_1) - 1 =T_k(G_1) - 1.$$
-Similarly $H'_2 - z$ is a subgraph of $G_2 - v$ and $T_k(H'_2-z) \geq T_k(H'_2) - 1$. So 
-$$T_k(G_2 - z) \geq T_k(H'_2 - z) \geq T_k(H'_2) - 1 = T_k(G_2) - 1.$$
-Summing our inequalities above we obtain
-$$T_k(G_1 - e) + T_k(G_2 - z) \geq T_k(G_1) + T_k(G_2) - 2$$
-as desired.
-\end{proof}
-\subsection{b}
-\begin{lemma}
-If $G$ is $k$-Ore and $G \neq K_k$ then $T_k(G) \geq 2+ \frac{v(G) - 1}{k-1}$.
-\end{lemma}
-\begin{proof}
-Suppose not and let $G$ be a minimum counterexample.
-Since $G$ is $k$-Ore and $G\neq K_k$ there exist $k$-Ore graphs $G_1$ and $G_2$ such that $G$ is the Ore-composition of $G_1$ and $G_2$.
-We proceed by case distinction on if $G_1, G_2$ are $K_k$ or not.
-If neither $G_1$ nor $G_2$ are $K_k$ then by minimality of $G$,
-$$T_k(G_i) \geq 2 + \frac{v(G_i) - 1}{k-1}$$
-for $i =1,2$. But then combining with the inequality
-$$T_k(G) \geq T_k(G_1) + T_k(G_2) - 2$$
-from the previous Lemma, we see that
-$$T_k(G) \geq 2 + \frac{v(G_1) - 1}{k-1} + 2 + \frac{v(G_2) - 1}{k-1} -2 \geq 2 + \frac{v(G)-1}{k-1}$$
-as desired, using that $v(G_1) + v(G_2) -1 = v(G)$.
-\paragraph{}
-If $G_1 = K_k$ and $G_2 \neq K_k$ then using the first inequality from the previous Lemma we have
-$$T_k(G) \geq T_k(G_1-e) + T_k(G_2-z) \geq 2 + T_k(G_2-z)$$
-Since $T_k(G_2-z) \geq T_k(G_2) - 1$, 
-$$T_k(G) \geq 1 + T_k(G_2) \geq \frac{k-1}{k-1} + \frac{v(G_2)-1}{k-1} + 2= \frac{v(G_1) + v(G_2) - 2}{k-1} +2 = \frac{v(G) - 1}{k-1} +2$$
-The proof for $G_2 = K_k$ and $G_1 \neq K_k$ follows similarly.
-\paragraph{}
-If $G_1 = K_k$ and $G_2 = K_k$ then 
-$$T_k(G) \geq T_k(G_1-e) + T_k(G_2 - z) = 2  + 2 = \frac{2(k-1)}{k-1} + 2 = \frac{v(G_1) + v(G_2) -2}{k-1} + 2 = \frac{v(G) -1}{k-1} + 2$$
-as desired.
-\end{proof}
-\subsection{c}
-\paragraph{}
-It suffices to show
-$$T_k(R) + T_k(W) - T_k(K_{|X|}) \leq T_k(R') + |X|.$$
-Let $H_R \subseteq G[R]$ be a disjoint union of cliques attaining the maximum in $T_k(R)$.
-Let $H_W \subseteq W$ be a disjoint union of cliques attaining the maximum in $T_k(W)$.
-If a vertex in $X$ is deleted from a clique in $H_W$ then the value of $T_k$ for the resulting graph will decrease by at most one.
-Hence $T_k(H_W - X) \geq T_k(H_W) - |X|$.
-But then $H_R \cup (H_W - X)$ is a disjoint union of cliques in $G[R']$.
-So we have
-\begin{align*}
-T_k(R) + T_k(W) - T_k(K_{|X|}) &\leq T_k(H_R) + T_k(H_W)\\ &\leq T_k(H_R) + T_k(H_W - X) - |X|\\ &\leq T_k(H_R \cup (H_W-X)) - |X|\\ &\leq T_k(R') - |X|\end{align*}
-as desired. $\blacksquare$
+
+%Problem 4
 \newpage
 \section{}
-\paragraph{}
-Let $k \geq 4$.
-Let $p(G) = (k-1)v(G) - 2e(G)$.
-Then $p(G) \leq 0$ if $G$ is $k$-critical since the minimum degree of $G$ is $k-1$.
-Let $G$ be a counterexample to the potential version of Brooks' Theorem.
-That is, $G$ is $k$-critical, $G\neq K_k$, and $p(G) = 0$.
-Since $p(G) = 0$, we have that $G$ is $(k-1)$-regular.
-\begin{claim}
-If $R \subset V(G)$ then $p(R) \geq k-1$.
-\end{claim}
-\begin{proof}
-Suppose not and let $R$ be a counterexample with maximum number of vertices. Then $R$ has an extension $R'$ with extender $W$ and core $X$. By the potential-extension lemma,
-$$p(R') \leq p(R) + p(W) - p(K_{|X|})$$
-Now $$p(R) =  (k-1)|R| - 2e(R) \geq (k-|R|)|R|$$
-which is at least $k-1$ when $1 \leq |R| \leq k-1$.
-Hence $|R| \geq k$, and since $G$ is not equal to $K_k$ and $G$ is $(k-1)$-regular, $G[R]$ is not a clique of size at least $k$.
-Therefore $W$ is a strict subgraph of $G$, and so by minimality of $G$,
-$$p(W) \leq -1.$$
-By the same reasoning as above for $R$, $$p(K_{|X|}) \geq k-1.$$
-Thus we have
-$$p(R') \leq p(R) + p(W) - p(K_{|X|}) \leq (k-2)  - 1 -(k-1) \leq - 2$$
-By maximality of $R$, $p(R') \geq k-1$ contradicting the above.
-\end{proof}
-\begin{claim}
-If $z$ is a vertex with two non-adjacent neighbours $x,y$ then there exists a $(k-1)$-clique $H$ in $G\backslash \{x,y,z\}$ such that $N(x) \cap V(H)$, $N(y) \cap V(H)$ partition $V(H)$.
-\end{claim}
-\begin{proof}
-Let $G'$ be the graph obtained from $G$ by identifying vertices $x$ and $y$ (deleting parallel edges if necessary) as vertex $v_{xy}$ and deleting vertex $z$.
-If $G'$ has a $(k-1)$-coloring $\phi'$ then $\phi'$ can be extending to a $(k-1)$-coloring of $G$ by assigning $\phi'(x ) = \phi'(y) = \phi'(v_{xy})$, then assigning $z$ a color not present on the neighbours of $z$.
-Such a color exists since $G$ is $(k-1)$-regular and two neighbours of $z$, $x$ and $y$, receive the same color under $\phi'$.
-\paragraph{}
-Thus $G'$ is not $(k-1)$-colorable.
-So $G'$ has a $k$-critical subgraph $K'$.
-Then $v_{xy} \in V(K')$ for otherwise $K'$ is a $k$-critical strict subgraph of $G$, contradicting that $G$ is $k$-critical.
-By minimality of $G$ we have
-$$p(K') \leq 0$$
-with equality if and only if $K'$ is $K_k$. Let $K$ be the graph $K' - v_{xy} + \{x,y\}$. Then $K$ is a strict subgraph of $G$, since $z \not\in V(K)$. So by the previous claim
-$$k-1 \leq p(G[V(K')]) \leq p(K')$$
-Now we observe that
-$$p(K) = p(K') + (k-1) - 2|E(v_{xy}, N(x) \cap N(y) \cap V(K'))|$$
-Since $K$ has more vertex that $K'$, and parallel edges were deleted in obtaining $G'$. Combining inequalities we see that
-$$k-1 \leq p(K) \leq 0 + (k-1) - 2|E(v_{xy}, N(x) \cap N(y) \cap V(K'))|$$
-and hence 
-$$|E(v_{xy}, N(x) \cap N(y) \cap V(K'))| \leq 0.$$
-Since $|E(v_{xy}, N(x) \cap N(y) \cap V(K'))|$ is non-negative this implies $|E(v_{xy}, N(x) \cap N(y) \cap V(K'))| = 0$ and thus
-$N(x)\cap N(y) \cap V(K') = \emptyset$. Furthermore the displayed inequalities are tight, implying that $K'$ is isomorphic to $K_k$. Letting $H = K'-v_{xy}$ we obtain the desired clique of the claim.
-\end{proof}
 
+%Problem 5
+\newpage
+\section{}
 \end{document}

### 2018-06-13 CS761-RandomizedAlgorithms/assignments/a2.tex
partial q2

@@
 \section{}
-
+\paragraph{}
+Suppose we have $n$ items to be hashed into $n$ bins.
+Let $h$ be a hash function chosen from a $k$-universal family.
+We $\textit{overload}$ the notation ${n \choose k}$ to mean both the number of ways to choose $k$ elements from $n$, and the set of subsets of $[n]$ of size $k$.
+Which version we use will always be clear from context.
+Define a $k$-collision to be the event that $k$ distinct items are hashed to the same bin.
+For a $\textit{set}$ $S \in {n\choose k}$ let $X_S$ be the random indicator variable which is $1$ when $h(x) = h(y)$ for all $x,y \in S$, and $X_S$ is $0$ otherwise.
+\paragraph{}
+Let $X$ be a random variable counting the number of $k$-collisions when our $n$ items are hased using $h$ chosen from a $k$-universal family. Then $X = \sum_{S \in {n\choose k}} X_S$ and so we have
+$$E[X] = E[\sum_{S \in {n\choose k}} X_S] = \sum_{S \in {n\choose k}} E[X_S] = \sum_{S \in {n \choose k}} Pr(|h(S)| = 1) \leq  \sum_{S \in {n \choose k}} \frac{1}{n^{k-1}}$$
+where the inequality follows since $h$ is a $k$-universal hash function, and $S$ is a set of size $k$.
 %Problem 3

### 2018-06-14 CS761-RandomizedAlgorithms/assignments/a2.tex
q2

--- 2018-06-13 CS761-RandomizedAlgorithms/assignments/a2.tex
@@
 \section{}
-
+\paragraph{}
+Suppose we have $n$ items to be hashed into $n$ bins.
+Let $h$ be a hash function chosen from a $k$-universal family.
+We $\textit{overload}$ the notation ${n \choose k}$ to mean both the number of ways to choose $k$ elements from $n$, and the set of subsets of $[n]$ of size $k$.
+Which version we use will always be clear from context.
+Define a $k$-collision to be the event that $k$ distinct items are hashed to the same bin.
+For a $\textit{set}$ $S \in {n\choose k}$ let $X_S$ be the random indicator variable which is $1$ when $h(x) = h(y)$ for all $x,y \in S$, and $X_S$ is $0$ otherwise.
+\paragraph{}
+Let $X$ be a random variable counting the number of $k$-collisions when our $n$ items are hased using $h$ chosen from a $k$-universal family. Then $X = \sum_{S \in {n\choose k}} X_S$ and so we have
+$$E[X] = E[\sum_{S \in {n\choose k}} X_S] = \sum_{S \in {n\choose k}} E[X_S] = \sum_{S \in {n \choose k}} Pr(|h(S)| = 1) \leq  \sum_{S \in {n \choose k}} \frac{1}{n^{k-1}}$$
+where the inequality follows since $h$ is a $k$-universal hash function, and $S$ is a set of size $k$. So then
+$$E[X] \leq {n\choose k} \frac{1}{n^{k-1}}.$$
+Therefore if we let $t = 2n^{1-k}{n\choose k}$ then by Markov's inequality
+$$Pr(X \geq t) \leq Pr(X \geq 2E[X]) \leq \frac{1}{2}.$$
+\paragraph{}
+Now let $Y$ be the maximum number of items hashed into a single bin. Then $X \geq {Y \choose k}$ and so
+$$Pr({Y\choose k}\geq t) \leq Pr(X \geq t) \leq \frac{1}{2}$$
+Which implies that 
+$$Pr(Y \geq (2n)^{\frac{1}{k}}) \leq \frac{1}{2}.$$
+That is, with probability at most $\frac{1}{2}$, the maximum load is larger than $(2n)^{\frac{1}{k}}$. 
+\paragraph{}
+We want to find the smallest $k$ for which 
+$$Pr(Y \leq 3\ln n / \ln \ln n) \geq \frac{1}{2}.$$
+Thus we need to choose the smallest $k$ satisfying
+$$(2n)^{\frac{1}{k}} \leq \frac{3\ln n}{\ln \ln n}.$$
+Rearranging we see that setting 
+$$k \geq \frac{\ln2 + \ln n}{\ln(3) + \ln\ln n - \ln\ln\ln n}$$
+would do the job. Hence the smallest choice of $k$ will be the first integer greater than this value, i.e.
+$$k = \ceil{\frac{\ln2 + \ln n}{\ln(3) + \ln\ln n - \ln\ln\ln n}}.$$
+$\blacksquare$
 %Problem 3

### 2018-06-16 CS761-RandomizedAlgorithms/assignments/a2.tex
writing 3

@@
 \lhead{\today}
-\rhead{W. Justin Toth CS761-Randomized Algorithms Assignment n2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CS761-Randomized Algorithms Assignment 2} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \section{}
-
+\paragraph{}
+Suppose we have a list $L$ of $n$ distinct numbers, whose sorted list would we $\{x_1,\dots, x_n\}$.
+We want to find a $k$-approximate median.
+I.e.
+we want to return some $x \in [x_{\frac{n}{2} - k}, x_{\frac{n}{2} + k}]\cap L$.
+Consider sampling $m$ numbers from $L$ and returning the median $c$ of these $m$ numbers as the true median of $L$.
+\paragraph{}
+Let $L'$ be list of $m$ sampled numbers.
+For each $i \in L'$ let $X_{i,<}$ be an indicator variable for the event that $i \in \{x_1, \dots, x_{\frac{n}{2} -k -1}\}$.
+Then, since we sample without replacement,
+$$Pr(X_{i, <} =1) = (\frac{n}{2} -k -1) / n = \frac{1}{2} - \frac{k+1}{n}$$
+Let $X_<$ be the number of elements in $L'$ which lie in $\{x_1, \dots, x_{\frac{n}{2} -k -1}\}$. Then $X_< = \sum_{i \in L'} X_{i,<}$.
+Let $X_>$ be the number of elements in $L'$ which lie in $\{x_{\frac{n}{2} + k + 1}, \dots, x_n\}$. Since we sample uniformly and
+	$$|\{x_{\frac{n}{2} + k + 1}e, \dots, x_n\}| =  |\{x_1, \dots, x_{\frac{n}{2} -k -1}\}| = \frac{n}{2} - k - 1$$
+we observe that $X_<$ and $X_>$ have the same distribution.
+\paragraph{}
+Now we see that $c$ fails to be a $k$-approximate median for $L$ if and only if $X_< \geq \frac{m}{2}$ or $X_> \geq \frac{m}{2}$.
+Since $X_<$ and $X_>$ have the same distribution, the probability of failure is thus
+$$2Pr(X_< \geq \frac{m}{2}).$$
+We plan to bound this using a Chernoff bound. First we compute the expectation,
+$$E[X_<] = \sum_{i\in L'} Pr(X_{i,<} = 1)  = \sum_{i \in L'} (\frac{1}{2} - \frac{k+1}{n}) = \frac{m(n-2(k+1))}{2n}.$$
+So 
+$$Pr(X_< \geq \frac{m}{n}) = Pr(X_< \geq (1+\delta) E[X_<])$$
+where $$\delta = \frac{2}{\frac{n-2(k+1)}{2}}-1=\frac{4-(n-2(k+1))}{n-2(k+1)}$$
 %Problem 4

### 2018-06-16 CS761-RandomizedAlgorithms/assignments/a2.tex
Applied Chernoff bound to q3

--- 2018-06-14 CS761-RandomizedAlgorithms/assignments/a2.tex
@@
 \section{}
+I discussed this problem with Sharat.
 
+\paragraph{}
+Suppose we have a list $L$ of $n$ distinct numbers, whose sorted list would we $\{x_1,\dots, x_n\}$.
+We want to find a $k$-approximate median.
+I.e.
+we want to return some $x \in [x_{\frac{n}{2} - k}, x_{\frac{n}{2} + k}]\cap L$.
+Consider sampling $m$ numbers from $L$ and returning the median $c$ of these $m$ numbers as the true median of $L$.
+\paragraph{}
+Let $L'$ be list of $m$ sampled numbers.
+For each $i \in L'$ let $X_{i,<}$ be an indicator variable for the event that $i \in \{x_1, \dots, x_{\frac{n}{2} -k -1}\}$.
+Then, since we sample without replacement,
+$$Pr(X_{i, <} =1) = (\frac{n}{2} -k -1) / n = \frac{n-2k-2}{2n}$$
+Let $X_<$ be the number of elements in $L'$ which lie in $\{x_1, \dots, x_{\frac{n}{2} -k -1}\}$. Then $X_< = \sum_{i \in L'} X_{i,<}$.
+Let $X_>$ be the number of elements in $L'$ which lie in $\{x_{\frac{n}{2} + k + 1}, \dots, x_n\}$. Since we sample uniformly and
+	$$|\{x_{\frac{n}{2} + k + 1}e, \dots, x_n\}| =  |\{x_1, \dots, x_{\frac{n}{2} -k -1}\}| = \frac{n}{2} - k - 1$$
+we observe that $X_<$ and $X_>$ have the same distribution.
+\paragraph{}
+Now we see that $c$ fails to be a $k$-approximate median for $L$ if and only if $X_< \geq \frac{m}{2}$ or $X_> \geq \frac{m}{2}$.
+Since $X_<$ and $X_>$ have the same distribution, the probability of failure is thus
+$$2Pr(X_< \geq \frac{m}{2}).$$
+We plan to bound this using a Chernoff bound. First we compute the expectation,
+$$E[X_<] = \sum_{i\in L'} Pr(X_{i,<} = 1)  = \sum_{i \in L'} \frac{n-2k-2}{2n} = \frac{m(n-2(k+1))}{2n}.$$
+So 
+$$Pr(X_< \geq \frac{m}{2}) = Pr(X_< \geq (1+\delta) E[X_<])$$
+where $$\delta =\frac{2(k+1)}{n-2(k+1)}.$$
+We proceed by assuming $k < \frac{n}{4}-1$ so that $0<\delta <1$.
+Ultimately we are interested in results on how many samples $m$ we need to take to obtain a certain quality of approximation, and so bounding $k$ is reasonable.
+\paragraph{}
+Since $0 < \delta < 1$, and $X_<$ is a sum of independent random coin flips, we can apply the second Chernoff bound from Theorem on page $4$ of Lecture $2$, obtaining
+$$Pr(X_< \geq \frac{m}{2}) < \exp(\frac{-\delta^2E[X_<]}{3}).$$
 %Problem 4

### 2018-06-17 CS761-RandomizedAlgorithms/assignments/a2.tex
progress

@@
 Since $0 < \delta < 1$, and $X_<$ is a sum of independent random coin flips, we can apply the second Chernoff bound from Theorem on page $4$ of Lecture $2$, obtaining
-$$Pr(X_< \geq \frac{m}{2}) < \exp(\frac{-\delta^2E[X_<]}{3}).$$
+$$Pr(X_< \geq \frac{m}{2}) < \exp(\frac{-\delta^2E[X_<]}{3}) = \exp(\frac{-2(k+1)^2m}{3n(n-2(k+1))}).$$
+Hence the probability of failure is at most
+$$2\exp(\frac{-2(k+1)^2m}{3n(n-2(k+1))})$$
+To succeed with probability at least $0.9999$ we want this value to be at most $10^{-5}$. So it will suffice to choose $m$ and $k$ to satisfy
+$$\exp(\frac{-2(k+1)^2m}{3n(n-2(k+1))}) \leq 10^{-6}$$
+Rearranging yields the expression
+$$(k+1)^2m \geq 9\ln(10) n^2 - 18\ln(10) n(k+1)$$
+Dropping the non-positive term on the left-hand side, as this can only improve our probability of success, gives us the following expression for the trade-off between $m$ and $k$:
+$$(k+1)^2m = 9\ln(10)n^2 = \Omega(n^2).$$
+\paragraph{}
+If we want $k \leq \epsilon n$ for small $\epsilon$ (say $<\frac{1}{4}$) then we need $m$ to satisfy
+$$(\epsilon n+ 1)^2 m = 9\ln(10)n^2$$
+i.e. 
+$$m = \frac{9\ln(10)n^2}{(\epsilon n + 1)^2}.$$
+\paragraph{}
+Similarly if we want $k \leq n^{1-\eta}$ for $\eta \leq \frac{1}{2}$ then we take (again assuming desired $k$ is less than $\frac{1}{4}n-1$)
+$$m =  \frac{9\ln(10)n^2}{(n^{1-\eta} + 1)^2}.$$
+$\blacksquare$
 %Problem 4
@@
 \section{}
-
+\paragraph{}
+We assume that the determinant of an $n \times n$ matrix where each element is an entry of $\F[x]$ of degree at most $d$ can be computed in $O(dn^\omega)$ field operations.
+\subsection{a}
+Let $G=(U\cup V, E)$ be a bipartite graph where each edge is coloured either red or blue.
+We may assume $|V| = |U|$ by adding dummy vertices as needed.
+Let $k \in \Z$ be such that $1 \leq k \leq |E|$.
+We construct an $n\times n$ matrix $A$ of indeterminants as follows: for every edge $uv \in E$ set
+$$A_{uv} = \begin{cases}
+x_{uv}, \text{if $uv$ is coloured blue} \\
+rx_{uv}, \text{if $uv$ is coloured red}
+\end{cases}$$
+with indeterminants $x_{uv}$ for each edge $uv$ and indeterminant $r$.
 %Problem 5

### 2018-06-17 CS761-RandomizedAlgorithms/assignments/a2.tex
q4

--- 2018-06-16 CS761-RandomizedAlgorithms/assignments/a2.tex
@@
 \section{}
+I discussed this problem with Sharat.
+
 \paragraph{}
@@
 Then, since we sample without replacement,
-$$Pr(X_{i, <} =1) = (\frac{n}{2} -k -1) / n = \frac{1}{2} - \frac{k+1}{n}$$
+$$Pr(X_{i, <} =1) = (\frac{n}{2} -k -1) / n = \frac{n-2k-2}{2n}$$
 Let $X_<$ be the number of elements in $L'$ which lie in $\{x_1, \dots, x_{\frac{n}{2} -k -1}\}$. Then $X_< = \sum_{i \in L'} X_{i,<}$.
@@
 We plan to bound this using a Chernoff bound. First we compute the expectation,
-$$E[X_<] = \sum_{i\in L'} Pr(X_{i,<} = 1)  = \sum_{i \in L'} (\frac{1}{2} - \frac{k+1}{n}) = \frac{m(n-2(k+1))}{2n}.$$
+$$E[X_<] = \sum_{i\in L'} Pr(X_{i,<} = 1)  = \sum_{i \in L'} \frac{n-2k-2}{2n} = \frac{m(n-2(k+1))}{2n}.$$
 So 
-$$Pr(X_< \geq \frac{m}{n}) = Pr(X_< \geq (1+\delta) E[X_<])$$
-where $$\delta = \frac{2}{\frac{n-2(k+1)}{2}}-1=\frac{4-(n-2(k+1))}{n-2(k+1)}$$
+$$Pr(X_< \geq \frac{m}{2}) = Pr(X_< \geq (1+\delta) E[X_<])$$
+where $$\delta =\frac{2(k+1)}{n-2(k+1)}.$$
+We proceed by assuming $k < \frac{n}{4}-1$ so that $0<\delta <1$.
+Ultimately we are interested in results on how many samples $m$ we need to take to obtain a certain quality of approximation, and so bounding $k$ is reasonable.
+\paragraph{}
+Since $0 < \delta < 1$, and $X_<$ is a sum of independent random coin flips, we can apply the second Chernoff bound from Theorem on page $4$ of Lecture $2$, obtaining
+$$Pr(X_< \geq \frac{m}{2}) < \exp(\frac{-\delta^2E[X_<]}{3}) = \exp(\frac{-2(k+1)^2m}{3n(n-2(k+1))}).$$
+Hence the probability of failure is at most
+$$2\exp(\frac{-2(k+1)^2m}{3n(n-2(k+1))})$$
+To succeed with probability at least $0.9999$ we want this value to be at most $10^{-5}$. So it will suffice to choose $m$ and $k$ to satisfy
+$$\exp(\frac{-2(k+1)^2m}{3n(n-2(k+1))}) \leq 10^{-6}$$
+Rearranging yields the expression
+$$(k+1)^2m \geq 9\ln(10) n^2 - 18\ln(10) n(k+1)$$
+Dropping the non-positive term on the left-hand side, as this can only improve our probability of success, gives us the following expression for the trade-off between $m$ and $k$:
+$$(k+1)^2m = 9\ln(10)n^2 = \Omega(n^2).$$
+\paragraph{}
+If we want $k \leq \epsilon n$ for small $\epsilon$ (say $<\frac{1}{4}$) then we need $m$ to satisfy
+$$(\epsilon n+ 1)^2 m = 9\ln(10)n^2$$
+i.e. 
+$$m = \frac{9\ln(10)n^2}{(\epsilon n + 1)^2}.$$
+\paragraph{}
+Similarly if we want $k \leq n^{1-\eta}$ for $\eta \leq \frac{1}{2}$ then we take (again assuming desired $k$ is less than $\frac{1}{4}n-1$)
+$$m =  \frac{9\ln(10)n^2}{(n^{1-\eta} + 1)^2}.$$
+$\blacksquare$
 %Problem 4
@@
 \section{}
-
+\paragraph{}
+We assume that the determinant of an $n \times n$ matrix where each element is an entry of $\F[x]$ of degree at most $d$ can be computed in $O(dn^\omega)$ field operations.
+\subsection{a}
+Let $G=(U\cup V, E)$ be a bipartite graph where each edge is coloured either red or blue.
+We may assume $|V| = |U|$ by adding dummy vertices as needed.
+Let $k \in \Z$ be such that $1 \leq k \leq |E|$.
+We construct an $n\times n$ matrix $A$, where $|A| = n$, of indeterminants as follows: for every pair of vertices $u \in U$, $v\in V$ set
+$$A_{uv} = \begin{cases}
+x_{uv}, \text{if $uv$ is in $E$ and is coloured blue} \\
+rx_{uv}, \text{if $uv$ is in $E$ and is coloured red} \\
+0, \text{otherwise}.
+\end{cases}$$
+with indeterminants $x_{uv}$ for each edge $uv$ and indeterminant $r$.
+For a perfect matching $M$ on $G$ define $s(M)$ to be the sign of the permutation that sends $i$ to $j$ if $ij \in M$.
+Also define $c_M$ to be the number of edges coloured red in $M$.
+Let $\cM$ be the set of perfect matchings on $G$.
+Then we observe that
+$$\det(A) = \sum_{\text{permutations } \sigma}\text{sign}(\sigma) \Pi_{i=1}^n A_{i,\sigma(i)} = \sum_{M \in \cM} s(M) r^{c_M}\Pi_{uv \in M}x_{uv}.$$
+Let $\cM_\ell$ be the set of perfect matchings on $G$ with $\ell$ red edges. Then we can rearrange the determinant of $A$ as
+$$\det(A) = \sum_{\ell = 1}^n (\sum_{M \in \cM_\ell} s(M)\Pi_{uv \in M}x_{uv})r^\ell.$$
+As in the proof of Edmonds' theorem from Lecture $8$ the coefficient of the term corresponding to $r^k$ is not identically zero if and only there is a perfect matching in $G$ with $k$ red edges.
+This coefficient is a polynomial in variables $x_{uv}$ over $uv \in E$, of degree $n$.
+So if we substitute values into the variables $x_{uv}$, $uv \in E$, drawn independently and uniformly at random from a finite field $\F$ then by the Schwartz-Zippel Lemma this coefficient will be $0$ with probability at most $\frac{n}{|\F|}$.
+Choose $\F$ so that this probability is as low as desired.
+\paragraph{}
+Then after choosing random values for entries $x_{uv}$, $A$ becomes a matrix whose entries are in $\F[r]$ with degree at most $1$.
+So in $O(n^\omega)$ field operations we can compute $\det(A)$.
+Then we query the coefficient of the $k$-th term, which will always be $0$ if there are no perfect matchings with $k$ red edges, and with high probability will be non-zero if there is a perfect matching with $k$ red edges.
+$\blacksquare$.
+\subsection{b}
+\paragraph{}
+Now we suppose that each edge $uv \in E$ has an associated weight $w_{uv}$, and the maximum weight of an edge is $W$.
+We use a similar idea to the previous problem, this time creating our $n\times n$ matrix $A$ as follows: for each pair $u \in U$, $v\in V$ set
+$$A_{uv} = \begin{cases}
+	x_{uv}r^{w_{uv}}, &\text{if $uv \in E$} \\
+	0, &\text{otherwise.}
+	\end{cases}$$
+Let $w^*$ be the maximum weight of a perfect matching on $G$. Let $\cM^\ell$ be the set of perfect matchings on $G$ of weight $\ell$. Now $\det(A)$ is a polynomial of the form
+$$\det(A) = \sum_{\ell = 1}^{w^*} (\sum_{M \in \cM^\ell} s(M)\Pi_{uv \in M}x_{uv})r^\ell.$$
+As previously argued, if we choose large enough $\F$ then drawing the values for $x_{uv}$, $uv \in E$, from $\F$ uniformly independently at random, then by Schwartz-Zippel the coefficient on $r^{w^*}$ in the determinant polynomial will be non-zero with high probability.
+After assigning values to each $x_{uv}$, $A$ is matrix whose entries are in $\F[r]$ with degree at most $W$.
+So we can compute $\det(A)$ in O$(Wn^\omega)$ field operations.
+Querying the degree of the leading term in the resulting polynomial will give the maximum weight of a perfect matching on $G$ with high probability.
+$\blacksquare$
 %Problem 5

### 2018-06-18 CS761-RandomizedAlgorithms/assignments/a2.tex
q5 part 1

@@
 \section{}
+\paragraph{}
+Let $G=(V,E)$ be a directed acyclic graph and let $s \in V$ be the only vertex with indegree zero.
+Let $\F$ be a field such that $|\F| = \Theta(\text{poly}|V|)$.
+Suppose $s$ has $d$ out-going edges, and to edge $i$ out-going from $s$ we assign the $i$-th standard basis vector of $\F^d$.
+Apply a topological ordering to the vertices of $G$, and so then $V = \{s=v_1, v_2, \dots, v_n\}$ where $v_iv_j \in E$ implies $j>i$.
+We refer to the algorithm described in the presentation of question $5$ in the homework as the ``edge connectivity algorithm" or simply ``the algorithm".
+\paragraph{}
+Each edge $e$ is sending a linear combination of $e_1, \dots, e_d$ of the form $a=c_1e_1 + \dots + c_d e_d$.
+we call $(c_1,\dots,c_d)$ the global encoding vector of edge $e$.
+Further if edge $e=uv$ and $u$ has indegree $\ell$ then $a$ is a random linear combination of the $\ell$ incoming messages $a_1 \dots a_\ell$ to $u$.
+So $a = b_1^{(e)} a_1 + \dots + b_\ell^{(e)} a_\ell$.
+We call $\{b_i^{(e)}\}$ the local encoding coefficients of edge $e$.
+\paragraph{Claim}
+For each vertex $v_i \in V$, for each outgoing edge $e$ of $v_i$, each entry $c_j$ of $e$'s global encoding vector is a multivariate polynomial of the local encoding coefficients of total degree at most $i-1$.
+
+\begin{proof}
+	We proceed by induction on $i$. Since $s=v_1$ sends $e_j$ on its $j$-th outgoing edge the base case is true.
+
+	In the inductive step, consider an edge $e=v_iv_j$.
+Let $a_1 \dots a_\ell$ be the global encoding vectors of incoming edges to $v_i$.
+Then each $a_i$ is a polynomial in the local encoding coefficients of degree at most $i-2$ by induction.
+Now, the global encoding vector for $e$ is
+	$$b_1^{(e)}a_1 + \dots + b_\ell^{(e)}a_\ell$$
+	which is a sum of multivariate polynomials of degree at most $i-1$, and hence is a polynomial of degree at most $i-1$ as desired.
+\end{proof}
+\paragraph{}
+For vertex $t \in V\backslash \{s\}$ with $\ell$ incoming arcs, let $(c_{i1}, \dots, c_{id})$ be the global encoding vector of the $i$-th incoming edge to $t$.
+Then we can define the $\ell \times d$ receiver matrix $C$, whose entries are multivariate polynomails over $\F$, where entry $C_{ij} = c_{ij}$ for all $i,j$.
+\paragraph{}
+Let $C'$ be the largest non-singular submatrix of $C$.
+The size of $C'$ corresponds to the rank of $C$.
+So if $\det(C')$ is not $0$ after assigning values to its entries, then the rank of $C$ will be preserved.
+Since $\det(C')$ is not identically $0$, by Schwartz-Zippel the probability that $\det(C') =0$ after assigning random values for the local encoding coefficients is at most
+$$\frac{dn}{|\F|}$$
+Since our Claim implies that the entries of $C'$ are each of degree at most $n$ and so $\det(C')$ is a polynomial of degree at most $dn$.
+Since $d\leq n$, if we take $|\F| \geq n^4$ this probability of failure is at most
+$$\frac{1}{n^2}.$$
+So by the union bound, the probability that the rank of the receiver matrix for each vertex is preserved is at least
+$$1-\frac{1}{n}.$$
+\paragraph{}
+It remains to show that the rank of $C$ is equal to the edge connectivity of $t$.
+Suppose the edge-connectivity of $t$ is $k$.
+On the one hand there are $k$ edge-disjoint paths to $t$.
+If we set the message on each edge of the $i$-th such path to be $e_i$ and the message on every other edge to be $0$ then $C$ will contain (possibly up to permuting the rows) the size $k$ identity matrix, and hence as a matrix of polynomials, the rank of $C$ is at least $k$.
+On the other hand, by the max-flow min-cut theorem there is an $s$-$t$ cut with exactly $k$ edges.
+So every row of $C$ is a linear combination of the $k$ vectors along that cut.
+Hence the rank of $C$ is at most $k$.
+Therefore the rank of $C$ is equal to $k$.
+So we have shown the edge-connectivity algorithm works with high probability when you choose a finite field $\F$ so that $|F| \geq n^4 = \Theta(\text{poly}(n))$.
+$\blacksquare$
+\paragraph{Implementation}
+We now present an efficient implementation.
 \end{document}

### 2018-06-18 CS761-RandomizedAlgorithms/assignments/a2.tex
wrote up 5

@@
 \paragraph{Implementation}
-We now present an efficient implementation.
+We now present an efficient implementation. For each vertex $v \in V$ we replace $v$ with a super-concentrator of size
+$$O(\text{in-degree}(v) + \text{out-degree}(v)\}) = O(\deg(v))$$
+where $\deg(v)$ is the number of edges incident to $v$ in the underlying undirected graph to $G$.
+Since the size of the super-concentrator we replace each vertex with is an upper-bound on the number of edge-disjoint paths in $G$ which can use that vertex, we preserve edge-connectivity after this transformation.
+From the notes, the encoding at each vertex can be done in $O(d)$ time after the transformation.
+The number of vertices in the new graph is
+$$\sum_{v \in V} O(\deg(v)) = O(2|E|) = O(|E|)$$
+by the handshaking-lemma. So the entire encoding can be done in $O(d|E|)$ time.
+\paragraph{}
+Using the random matrix rank algorithm for Lecture $9$ we can compute the rank of the matrix associated with each vertex in time
+$$O(d\cdot \deg(v) + \min\{d,\deg(v)\}^\omega)$$
+where $\omega$ is the matrix multiplication constant, with high probability. Summing over vertices $v \in V$, we can compute all the edge connectivities in time
+$$O(d|E| + |V|d^\omega).$$
+Thus our implementation will succeed in computing the edge-connectivity from $s$ to $t$ for each $t \in V$ with high probability, in time
+$$O(d|E| +|V|d^{\omega}).$$
+$\blacksquare$
 \end{document}

### 2018-07-11 CS761-RandomizedAlgorithms/assignments/a3.tex
A3

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}
\usepackage[braket, qm]{qcircuit}
\usepackage{hyperref}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{\today}
\rhead{W. Justin Toth - CS 761 - A3} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\section{}
\paragraph{}
Let $G$ be a graph with $m$ edges and $n$ vertices, where $m \geq 4n$. Fix an embedding of $G$ in the plane. Let $t$ be the number of crossings in this embedding of $G$. Sample vertices of $G$, each with probability $p=\frac{4n}{m}$ independently and uniformly at random. This choice of $p$ is a proper probability since $m \geq 4n$.  Let $S$ be the resulting vertex set, and let $G' = G[S]$. Let $n'$ be a random variable counting the number of vertices of $G'$. Let $m'$ be a random variable counting the number of edges of $G'$. Let $t'$ be a random variable such that
$$t' = m' - 3n'+ 6.$$
By our naive observation about crossing number, $G'$ has at least $t'$ crossings. Now we take expectation:
$$E[t'] = E[m' -3n' + 6] = E[m'] -3E[n'] + 6 = m(\frac{4n}{m})^2 -3n\frac{4n}{m} +6.$$
This last equality follows since each vertex of $G$ is sampled with probability $p$, and each edge with probability $p^2$. On the other hand, observing that each crossing of $G$ is sampled with probability $p^4$, as each crossing involves four distinct vertices, we have
$$E[t'] = t(\frac{4n}{m})^4.$$
Combining the equations above we see that
$$ t(\frac{4n}{m})^4 = m(\frac{4n}{m})^2 -3n\frac{4n}{m} +6 \geq m(\frac{4n}{m})^2 -3n\frac{4n}{m}$$
and hence
\begin{align*}
t &\geq m(\frac{4n}{m})^{-2} -3n(\frac{4n}{m})^{-3} \\
&=\frac{4nm^3 - 3nm^3}{64n^3}\\
&= \frac{m^3}{64n^2}
\end{align*}
as desired.$\blacksquare$
\newpage
%Problem 2
\section{}
\paragraph{}
Suppose (reasonably) that $0 < \epsilon < 1$. We generate $m = \frac{2^kk\log(n)}{\epsilon^2}$ random vectors $y^{(1)}, \dots, y^{(m)} \in \{0,1\}^n$ independently and uniformly at random. 
\paragraph{}
Fix $S\subseteq [n]$ such that $|S|=k$. Let $b \in \{0,1\}^n$. Let $Y_i$ be a $0$-$1$ indicator random variable indicating $1$ iff $y^{(i)}_j = b_j$ for all $j \in S$. Let $Y = \sum_{i} Y_i$. Since $y(i)$ was generated uniformly at random, 
$$Pr(Y_i = 1) = \frac{1}{2^k}$$
from which we see that
$$E[Y] = \frac{m}{2^k}.$$
\paragraph{}
Now we observe that $|Y- E[Y]| > \frac{m\epsilon}{2^k}$ if and only if $|\frac{Y}{m} - \frac{1}{2^k}| > \frac{\epsilon}{2^k}$ which in turn holds if and only if $$|Pr((\bigcap_{i\in S} X_i = b_i) - \frac{1}{2^k}|>\frac{\epsilon}{2^k}$$
since $Y$ counts the number of successful $y^(i)$'s for choosing $X$ and there are $m$ such choices. We adopt the the strictly stronger requirement for $\epsilon$-almost $k$-wise independence whose failure condition is the above inequality.
\paragraph{}
We have
$$Pr(|Y-E[Y]| > \frac{m\epsilon}{2^k}) =  Pr(Y > (1+\epsilon)E[Y]) < \exp(-\epsilon^2\frac{m}{3\cdot2^k})$$
where the last inequality follows from the second Chernoff bound in Lecture $2$ under the reasonable assumption that $0 < \epsilon < 1$.
\paragraph{}
Thus we have bounded the failure probability for a single fixed $S$. We apply the union bound over all subsets of $n$ of size $k$ to obtain that our choice of $y$'s fails with probability at most
$${n\choose k} \exp(-\epsilon^2\frac{m}{3\cdot2^k})$$
which, by our choice of $m$, is at most
$${n\choose k} e^{-k\log n} = \frac{{n\choose k}}{n^k} < 1.$$
Since this failure probability is strictly smaller than $1$ we have shown the desired existence result. $\blacksquare$
\subsection{Derandomization of Color Coding}
\paragraph{}
Suppose we want to decide if a graph $G=(V,E)$ has a path of length at least $\ell$. Let $n=|V|\log\ell$ and let $k = \ell\log\ell$. Let $\epsilon = \frac{1}{2}\frac{1}{\ell^\ell}$. Choose $m = O(\frac{2^kk\log(n)}{\epsilon^2})$ as in the previous section of the this problem. Our algorithm proceeds as follows:
\begin{enumerate}
    \item Generate strings $y^{(1)}. \dots, y^{(m)} \in \{0,1\}^n$ from which we will sample $\epsilon$-almost $k$-wise independent random bits.
    \item For each $i = 1\dots m$
        \begin{enumerate}
            \item Set $X=y^{(i)}$. Divide $X$ into $|V|$ blocks of $\log\ell$ bits.
            \item For each $v \in V$ assign the color $c\in [\ell]$ which is encoded by the $\log\ell$ bits of the $v$-th block of $X$ to $v$. 
            \item Use the Dynamic Programming subroutine of Lecture $3$ to decide of $G$ has a $\ell$-length colorful path. If so, return True.
        \end{enumerate}
    \item Return False if no color path was found.
\end{enumerate}
\paragraph{}
Suppose $G$ has a path $P$ of length $\ell$. For a given coloring drawn from random bits $X$ in Step $2(b)$ we want to analyze the probability that $P$ is colorful. Let $S_\ell$ denote the set of permutations of $\ell$ colors. Let $b(S_\ell) = \{ \ell\log\ell \text{ size bitstrings encoding } \sigma : \sigma \in S_\ell\}$. Note that $X$ is $\ell\log\ell$-wise independent ($\epsilon$-almost). Now we have
\begin{align*}
    Pr(P\text{ is colorful}) &= \sum_{b \in b(S_\ell)} Pr(X_i = b_i, \forall i) \\
    &\geq \sum_{b \in b(S_\ell)} (\frac{1}{2^{\ell\log\ell}}-\epsilon) \\
    &=\frac{\ell!}{\ell^\ell} - \epsilon\ell! \\
    &> 0 &\text{(by our choice of $\epsilon$).}
\end{align*}
Since this probability is strictly positive, there exists some $y^{(i)}$, $i\in[m]$, for which coloring $G$ based on $X = y^{(i)}$ will yield colorful $P$. Hence if such $P$ exists, our algorithm will discover it after running over all choices of $y^(i)$. Otherwise it correctly returns that no such $P$ exists.
\paragraph{}
It remains to bound the running time. We assume we are given the strings for Step $1$ and so Step $2$ is the main working step we need to bound. There will be at most $m$ iterations of Step $2$ and each iteration will be dominated in runtime in terms of the cost of running the dynamic program ($\text{cost}_DP$). So our runtime is
$$m\cdot\text{cost}_{DP} = O(\frac{2^kk\log(n)}{\epsilon^2})\cdot O(\ell2^\ell|E||V|) = O(\ell^{4\ell} \ell^2\log\ell\cdot\log(|V|\log\ell)|V||E
|)$$
and therefore we obtain a deterministic Fixed Parameter Tractable algorithm.$\blacksquare$
\newpage
%Problem 3
\section{}
\paragraph{}
Let $G$ be a graph in $G_{n,p}$ where $p = \frac{1}{n}$. Let $X$ be a random variable counting the number of triangles in $G$. For any $i,j,k \in V(G)$ define $X_{i,j,k}$ to be an indicator variable, indicating that $ijk$ induce a $K_3$ in $G$. So then $X = \sum_{i,j,k \in V(G)} X_{i,j,k}$. Observe that
$$ E[X] = \sum_{i,j,k \in V(G)} Pr(X_{i,j,k}=1) = \sum_{i,j,k \in V(G)} \frac{1}{n^3} = {n\choose 3}\frac{1}{n^3} \leq \frac{n^3}{6n^3}= \frac{1}{6}$$
Therefore by Markov's inequality
$$Pr(X \geq 1) \leq \frac{E[X]}{1} = \frac{1}{6}$$
as desired.
\paragraph{}
For the second part of this problem, we will use the Conditional Expectation Inequality, from section $6.6$ of the main course reference textbook: Probability and Computing. Theorem $6.10$ therein states: Let $Y = \sum_{i=1}^n Y_i$, where each $Y_i$ is a $0-1$ random variable. Then
$$Pr(Y>0) \geq \sum_{i=1}^n \frac{Pr(Y_i = 1)}{E[Y \mid Y_i = 1]}.$$
\paragraph{}
Since $X = \sum_{i,j,k \in V(G)} X_{i,j,k}$ and each $X_{i,j,k}$ is a $0-1$ random variable this inequality applies to our setting. Since $XS$ is integer valued it says that
$$Pr(X \geq 1) = Pr(X > 0) \geq \frac{1}{n^3}\sum_{i,j,k \in V(G)}\frac{1}{E[X\mid X_{i,j,k}=1]}.$$
Now we need to compute $E[X\mid X_{i,j,k}=1]$ for any $i,j,k$. Let $a,b,c \in V(G)$. We have three cases: Case $(1)$ $\{a,b,c\} = \{i,j,k\}$, Case $(2)$  $|\{a,b,c\} \cap \{i,j,k\}| = 2$, and Case $(3)$ $|\{a,b,c\} \cap \{i,j,k\}| \leq 1$.
\paragraph{}
In Case $(1)$, $Pr(X_{a,b,c} =1 \mid X_{i,j,k} = 1) = 1$.
\paragraph{} 
In Case $(2)$, $Pr(X_{a,b,c} =1 \mid X_{i,j,k} = 1) = \frac{1}{n^2}.$ This follows since $abc$ shares $1$ edges with $ijk$. There are $3$ ways to choose the overlapping edge, and $n$ ways to choose the non-intersecting vertex of $abc$, so there are $3n$ choices for $abc$ which manifest this case.
\paragraph{}
In Case $(3)$, $Pr(X_{a,b,c} =1 \mid X_{i,j,k} = 1) = \frac{1}{n^3}$ since $abc$ shares no edges with $ijk$. There are ${(n-3) \choose 3} + 3{(n-2)\choose 2}$ ways to manifest this case.
\paragraph{}
Putting all this together we obtain
$$E[X \mid X_{i,j,k} = 1] = 1 + 3n\frac{1}{n^2} + \Bigg({(n-3) \choose 3} + 3{(n-2)\choose 2}\Bigg)\frac{1}{n^3} \leq 1 + \frac{3}{n} +  \frac{1}{6} + \frac{3}{2n}.$$
Hence 
$$E[X\mid X_{i,j,k}] = \frac{7}{6} + \Omega(\frac{1}{n}).$$
So combining with our Conditional Expectation Inequality we see that
$$Pr(X \geq 1) \geq \frac{1}{n^3}{n \choose 3}\frac{1}{\frac{7}{6} + \Omega(\frac{1}{n})}.$$
Taking the limit $n\rightarrow \infty$ yields
$$\lim_{n\rightarrow \infty} Pr(X \geq 1) \geq \frac{1}{6} \frac{1}{\frac{7}{6} + 0} = \frac{1}{7}$$
as desired.$\blacksquare$
\newpage
%Problem 4
\section{}
\paragraph{}
The notion of Hamming distance is very helpful for thinking about these problems. For any $x \in \{0,1\}^n$ we define $|x|$ to be the number of bits set to $1$ in $x$. We always take addition of binary strings as addition modulo $2$. The Hamming distance between binary strings $x$ and $y$ is defined to be $|x+y|$.
\subsection{a}
\paragraph{}
First suppose that $Ax$ has at least $2pn+1$ non-zero bits for all $x\neq 0$. Suppose the encoder sends $Ax$ through the noisy channel and we receive the vector $y$. To decode we do the following:
\begin{enumerate}
    \item For each $x' \in \{0,1\}^m$ compute $y' = Ax'$.
    \item Over all such $y'$ choose the one which minimizes the Hamming distance to $y$. Output the corresponding $x'$ as the decoded message.
\end{enumerate}
We will show $Ax$ is the unique string with Hamming distance $\leq pn$ to $y$. Clearly its Hamming distance satisfies this bound from the definition of our noisy channel. Now suppose there exists some $x' \neq x$ (with $x' \neq 0$) such that $|Ax' +y| \leq pn$. Then observe that
$$|A(x + x')| = |Ax + Ax'| = |Ax + y + y + Ax'| \leq |Ax + y| + |Ax' + y| \leq pn + pn = 2pn.$$
But this contradicts that $A(x+x')$ has at least $2pn+1$ non-zero bits (noting that $x+x' \neq 0$ since $x\neq x'$ and $x' \neq 0$).
\paragraph{}
Now for the converse suppose there exists some $x\neq 0$ such that $|Ax| \leq 2pn$. We claim that no decoder can distinguish between $x$ and $0$. When the encoder sends $Ax$ through the noisy channel we can flip the first $pn$ non-zero bits of $Ax$ giving the decoder vector $y$ with at most $pn$ non-zero bits. Now if the encoder were to send $A0 = 0$ through the noisy channel we flip at most $pn$ bits of $0$ to obtain vector $y$. Hence with this adversarial strategy the decoder sees the same encoded vector for message $x$ and message $0$, so the best success probability the decoder can have over all messages is $\frac{1}{2}$.$\blacksquare$
\subsection{b}
\paragraph{}
For this section we assume reasonably that $0<p<\frac{1}{4}$.
\paragraph{}
Let $\epsilon > 0$. Fix $m=(1-H(2p) - \epsilon)n$. Let $x \in \{0,1\}^m$ such that $x\neq 0$. Let $A_i$ be the $i$-th row of $A$. The probability that $A_ix=1$ is $\frac{1}{2}$ by the principle of deferred decisions; there is some $j$ such that $x_j = 1$, and so flipping $A_{ij}$ (which is either $0$ or $1$ each with probability $\frac{1}{2}$) can flip $A_ix$. Thus
$$Pr(|Ax| =k) = {n\choose k}(\frac{1}{2})^k(1-\frac{1}{2})^{n-k} = {n\choose k}\frac{1}{2^n}.$$
Now we want to bound the probability that $|Ax| \leq 2pn$:
$$Pr(|Ax| \leq 2pn) = \sum_{k=0}^{2pn}Pr(|Ax| = k) = \frac{1}{2^n}\sum_{k=0}^{2pn}{n\choose k} \leq \frac{1}{2^n}(2pn+1){n\choose 2pn} \leq \frac{2pn+1}{2^n}2^{nH(2p)}.$$
The last inequality follows by applying the Lemma from Lecture $12$ which states
$${n\choose nq} \leq 2^{nH(q)}$$
where $0<q<1$, with $q=2p$. Note that $0<2p<1$ since $0<p<\frac{1}{4}$.
\paragraph{}
Now we apply the union bound:
$$Pr(\exists x\in\{0,1\}^m\backslash\{0\}: |Ax|\leq 2pn) \leq 2^{m}(2pn+1)2^{nH(2p)-n} = 2^{-\epsilon n }(2pn+1)$$
since there are $2^{m}-1\leq 2^m$ such strings $x$, and our last equality follows from the choice of $m$. Now we observe that this bound on the failure probability is strictly smaller than $1$ for sufficiently large $n$. Therefore there exists a matrix $A$ such that 
$$\forall x \in \{0,1\}^m\backslash\{0\}, \quad |Ax| \geq 2pn+1.$$
For this particular matrix $A$ we apply the result from $4(a)$, showing the receiver can always decode the message.$\blacksquare$
\subsection{c}
\paragraph{}
Let $\epsilon > 0$. Fix $N$ large enough that $\frac{2^{n(1+\epsilon)}}{n+1} > 2^n$ for any $n \geq N$. Let $n \geq N$, and let $m \geq (1 - H(p) + \epsilon)n$. If the adversary can use their $pn$ bit flips to create a collision between two different encoded messages then this is a barrier to perfect decoding.
\paragraph{}
If the encoder sends $Ax$ through the channel for intended message $x \in \{0,1\}^m$ then there are
$$\sum_{k=0}^{pn}{n\choose k} $$
distinct strings in $\{0,1\}^n$ to which the adversary can map $Ax$. Using the Lemma from the first page of Lecture $12$ on ${n\choose np}$ we can lower bound this number:
$$\sum_{k=0}^{pn}{n\choose k} \geq {n\choose np} \geq \frac{2^{nH(p)}}{n+1}.$$
Let $B(x)$ denote the set of strings the adversary can map $Ax$ to when $Ax$ is sent through the noisy channel. Thus we have shown $|B(x)| \geq \frac{2^{nH(p)}}{n+1}$. We assume for a contradiction that $B(x)$ and $B(x')$ are disjoint for $x\neq x'$.
\paragraph{}
There are $2^m$ distinct messages that the encoder could want to send through the channel. 
From disjointness
$$|\cup_{x\in\{0,1\}^m}B(x)| = \sum_{x \in \{0,1\}^m}|B(x)|$$
and we have
$$\sum_{x \in \{0,1\}^m}|B(x)| \geq \sum_{x\in\{0,1\}^m}\frac{2^{nH(p)}}{n+1} \geq 2^m\frac{2^{nH(p)}}{n+1} = 2^{(1-H(p)+\epsilon)n}\frac{2^{nH(p)}}{n+1} = 2^{n(1+\epsilon)}{n+1} > 2^n$$
where the last inequality follows since $n \geq N$. So
$$ |\cup_{x\in\{0,1\}^m}B(x)| >2^n,$$
but on the other hand
$$\bigcup_{x\in\{0,1\}^m}B(x) \subseteq \{0,1\}^n$$
and hence, since there are $2^n$ strings in $\{0,1\}^n$, 
$$|\cup_{x\in\{0,1\}^m}B(x)| \leq 2^n,$$
a contradiction. Therefore there exists some $x,x'$ with $x\neq x'$ and $B(x) \cap B(x') \neq \emptyset$. The adversary can make $x$ and $x'$ indistinguishable and hence perfect decoding is not possible. $\blacksquare$
\newpage
%Problem 5
\section{}
\paragraph{}
Let $G = (V,E)$ be a graph. Suppose each $v \in V$ is associated with a list of colors $S(v)$ such that $|S(v)| = 32r$ with $r\geq 1$. Further suppose for each $v \in V$ and each $c \in S(v)$ there exists at most $r$ neighbours $u \in N(v)$ such that $c$ is in $S(u)$.
\paragraph{}
Let $\phi$ be a vertex coloring obtained by assigning each vertex a color from $S(v)$ independently and uniformly at random. For any $e=uv \in E$ and $c \in S(u) \cap S(v)$ define the event $A_{e,c}$ to be the event that $\phi(u) = \phi(v) = c$. Let $X_{e,c}$ be a $0$-$1$ indicator variable for event $A_{e,c}$. Let $\cA := \{A_{e,c} : e=uv \in E, c \in S(u) \cap S(v)\}$ be the set of ``bad events".
\paragraph{}
Since $\phi(u)$ and $\phi(v)$ are chosen independently and uniformly at random, and $|S(u)| = |S(v)| = 32r$,
$$p := Pr(A_{uv,c}) = Pr(\phi(u)=c)\cdot Pr(\phi(v)=c) = \frac{1}{(32r)^2}.$$
Among events in $\cA$, the event $A_{uv,c}$ depends on the events $A_{e', c'}$ for any $e' \in \delta(u)$ and $c' \in S(u)$ and the events $A_{e'', c''}$ for any $e'' \in \delta(v)$ and $c'' \in S(v)$. Since there are at most $r$ neighbours of $u$ (respectively $v$) which share a color with $u$ (resp. $v$), and there are $32r$ colors in $u$'s (resp. $v$'s) list, there are at most $32r^2$ such events $A_{e',c'}$ and $32r$ such events $A_{e'',c''}$. Hence $A_{e,c}$ depends on at most $64r^2$ other events in $\cA$. 
\paragraph{}
Thus $d\leq 64r^2$ and so
$$4dp \leq 4 \cdot 64r^2 \frac{1}{(32r)^2} = \frac{1}{4} < 1$$
and thus by Lovasz's Local Lemma, the probability that no event $\cA$ happens is strictly positive. That is, there exists a proper list-coloring of $G$. $\blacksquare$
\subsection*{Polynomial Time Algorithm}
\paragraph{}
We say event $A_{uv,c}$ is not satisfied if $\phi(u) = \phi(v) = c$. Our algorithm is as follows (it is basically as in the $3$SAT algorithm from lecture):
\begin{enumerate}
    \item Fix an ordering of the events in $\cA$ ($|\cA| \leq 32rm$ so this can be done in polynomial time)
    \item Color vertices of $G$ independently and uniformly at random from their lists
    \item For each $A_{uv, c} \in \cA$ (in order)
        \begin{enumerate}
            \item If $A_{uv,c}$ is not satisfied: Fix($A_{uv,c}$)
        \end{enumerate}
\end{enumerate}
The process Fix($A_{uv,c}$) does:
\begin{enumerate}
    \item Randomly color $u$ and $v$ independently and uniformly at random from their lists
    \item While there is an event $A_{e', c'} \in \cA$ that $A_{e,c}$ depends on, and $A_{e',c'}$ is not satisfied, choose the first such $A_{e',c'}$ to appear in our fixed ordering of $\cA$ and call Fix($A_{e',c'}$)
\end{enumerate}
\paragraph{Analysis}
Once we call Fix($A_{e,c}$) and return to the main loop, $A_{e,c}$ will remain satisfied after each subsequent Fix($A_{e',c'}$) from the main loop. This follows by the recursive definition of the Fix procedure and the fact that changing colors on from $c'$ on $e'$ can only cause events which $A_{e', c'}$ to become unsatisfied. Hence if the algorithm terminates it solves the problem.
\paragraph{Random bits}
Let $t$ be the number of times the algorithm runs the Fix procedure. The algorithm uses $n\log(32r)$ bits to set the initial coloring and $2\log(32r)$ bits to recolor a pair of vertices each time a Fix is called. So in total the algorithm uses $(n+2t)\log(32r)$ random bits.
\paragraph{Encoding}
As in Lecture $11$ Page $8$ we construct trace the recursion tree of the algorithm. We encode as follows:
\begin{enumerate}
    \item Use a $0$ bit plus $\log(32rm)$ bits to represent going from the root to an event $A_{e,c}$. The $\log(32rm)$ bits encode which clause to go to.
    \item Use a $0$ bit plus $\log(d)$ bits (where $d$ is the dependency number calculated in the previous section of this problem) to represent going from the current event to the next event lower in the recursion tree.
    \item Use a $1$ bit to represent going up in the recursion tree. A $1$ bit at the root indicates that the algorithm terminates.
    \item When the algorithm terminates we can remember the $n\log(32r)$ bits indicating the coloring.
\end{enumerate}
It should be clear from the description we can fully recover the execution of the algorithm from this encoding.
\paragraph{Compression}
We now count how many bits we used:
\begin{itemize}
    \item By our prior note, the main root only calls $32rm$ Fix procedures. So we use only 
    $$2\cdot32rm + \log(32rm)32rm$$
    bits to encode Steps $1$ and $3$.
    \item There are $t$ steps in the algorithms so we need at most $$ t(\log(d) + 2)$$
    bits for steps $2$ and $3$.
    \item Finally we need $n\log(32r)$ bits for step $4$.
\end{itemize}
Thus, since random bits cannot be compressed, if we can recover the original random bits from the encoding scheme we must have
\begin{align*}
    64rm + \log(32rm)32rm +t(\log(d) + 2) +n\log(32r)\\
    \geq (n+2t)\log(32r)
\end{align*}
Thus we have
$$(2\log(32r) -\log(d) -2)t \leq 64rm + \log(32rm)32rm.$$
Recalling that $d = 32r^2$ we have
$$2\log(32r) -\log(d) - 2 =\log(\frac{1024r^2}{4\cdot32r^2}) \geq 1.$$
Hence $$t \leq 64rm + \log(32rm)32rm$$
which is to say that $t$ is called a polynomial number of times, and hence the randomized algorithm terminates in polynomial time with a proper list coloring of $G$.
\paragraph{Decoding}
It remains to verify that we can recover the original random bits from the encoding scheme. Let $b^v$ be a vector of $\log(32r)$ random bits used to encode the color of vertex $v$ at the start of the algorithm. Let $r_i$ and $q_i$ be vectors of $\log(32r)$ random bits used to encode the recoloring of the two vertices at the $i$-th call to Fix.
\paragraph{}
The algorithm will maintain $n$ variables for the current assignment, initially $b^1,\dots, b^n$. When the algorithm Fixes event $A_{e,c}$ then the current bits corresponding to the assignment do not satisfy this event. There is only one way for event $A_{e,c}$ to not be satisfied: if $e=uv$ and $\phi(u) = \phi(v) = c$. When the algorithm Fixes this event we learn the value of those bits for the colors of $u$ and $v$ at that point in the execution. We then know the algorithm replaces their assignment with $r_i$ and $q_i$ (if $A_{e,c}$ is the $i$-th event to be Fixed). We can repeat this procedure for each event tracing the execution of the algorithm through the recursion tree. Hence we can recover all the random bits used by the encoding algorithm.
\end{document}

### 2018-07-29 CS761-RandomizedAlgorithms/assignments/a4.tex
a4 q1 start

\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
\usepackage{amsmath, amssymb, amstext}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{theorem}
\usepackage{tikz}
\usepackage{tkz-berge}

%Macros
\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
\newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\cA}{\mathcal{A}} \newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}} \newcommand{\cD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}} \newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}} \newcommand{\cH}{\mathcal{H}}
\newcommand{\cI}{\mathcal{I}} \newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}} \newcommand{\cL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}} \newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}} \newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}} \newcommand{\cR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}} \newcommand{\cT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}} \newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}} \newcommand{\cX}{\mathcal{X}}
\newcommand{\cY}{\mathcal{Y}} \newcommand{\cZ}{\mathcal{Z}}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}

%\renewcommand{\thesection}{\lecnum.\arabic{section}}
%\renewcommand{\theequation}{\thesection.\arabic{equation}}
%\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}
\newtheorem{note}[fact]{Note}
\newtheorem{conjecture}[fact]{Conjecture}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}

%END MACROS
%Page style
\pagestyle{fancy}

\listfiles

\raggedbottom

\lhead{\today}
\rhead{W. Justin Toth CS761-Randomized Algorithms Assignment 4} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
%Problem 1
\section{}
\paragraph{}
Let $G=(V,E)$ be a directed graph, let $s,t\in V$ be source and sink vertices respectively and let $k$ be a positive integer. We may assume that $\delta^{in}(s) = \delta^{out}(t) = \emptyset$. Let $x_f \in [0,1]^E$ be a fractional $s-t$ flow on $G$ of value $k$. That is, $x_f$ satisfies the equations
$$ x_f(\delta^{out}(s)) = x_f(\delta^{in}(t)) = k$$
and
$$x_f(\delta^{in}(v)) = x_f(\delta^{out}(v))$$
for all $v \in V\backslash\{s,t\}$.
\paragraph{}
We're going to need some concepts from network flow theory before we move on. Lucky our flow being $0$-$1$ simplifies things quite a bit. Given a digraph $D$ and a flow $x \in \{0,1\}^E$ we define the residual digraph $D_x$ as follows: 
\begin{itemize}
\item$V(D_x) = V(D)$
\item for every arc $uv \in E(D)$ if $x(uv) = 0$ add arc $uv$ to $E(D_x)$. We call such arcs ``forward" arcs.
\item for every arc $uv \in E(D)$ if $x(uv) = 1$ add arc $vu$ to $E(D_x)$. we call such arcs ``backward" arcs.
\end{itemize} 
An $s-t$ path in $D_x$ is called an \emph{augmenting path}. It is an easy observation to see that if $x$ is a flow of value $k$ and $P$ is an augmenting path in $D_x$, then we can obtain a flow of value $k+1$ by increasing $x(e)$ from $0$ to $1$ for every forward arc in $P$, and decreasing $x(e)$ from $1$ to $0$ for every backward arc in $P$. Using the max-flow min-cut theorem we can see that if there is no augmenting path in $D_x$ then $x$ is a maximum flow. Simply take the cut $S$ defined by vertices reachable from $s$ in $D_x$ (this is an $s-t$ cut since $t$ is not reachable). Since there is no augmenting paths, all forward arcs in $\delta^{out}(S)$ are saturated (have $x$-value $1$). So the value of $x$ is at least (and hence equal to) the capacity of cut $S$, implying that $x$ is a maximum flow.
\paragraph{}
Our algorithm will maintain an integral flow in every step, and a weighting on the residual digraph so that (an auxiliary graph closely related to) the residual digraph is weighted-Eulerian. This will allow us to efficiently use random walks to identify augmenting paths for our flow. Due to our fractional flow $x_f$ we know that we will always be able to find such a path while the value of $x$ is less than $k$, as we are sure $x$ is not maximum.
\paragraph{}
We will need one more auxiliary graph before we describe the algorithm. Let $D'_x$ be the digraph obtained from $D_x$ by adding a special vertex $b$, and two arcs: $tb$ and $bs$. It is immediate that $D_x$ has an augmenting path if and only if $D_x'$ as cycle containing $b$.
\paragraph{}
If $x$ is an integral $s-t$ flow on $G$ we can define an Eulerian weighting on $D'_x$ as follows. For every forward arc $uv \in E(D_x)$ let $w(uv) = x_f(uv)$. For every backward arc $vu \in E(D_x)$ let $w(vu) = 1- x_f(uv)$. Let $k'$ be the value of $x$. Let $w(tb) = w(bs) = k-k'$.
\begin{claim}
Let $x_0=0,x_1, \dots, x_{k-1}\in \{0,1\}^E$ be integral $s-t$ flows on $G$ such that $x_{i+1}$ is obtained from $x_i$ by augmenting $x_i$ along an augmenting path in $D_{x_i}$. Let $w_0,w_1,\dots, w_{k-1}$ be the weightings of each respective $D'_{x_i}$. Then for all $i$
$$w_i(\delta^{out}(v)) = w_i(\delta^{in}(v))$$ for all $v \in V(D'_{x_i})$.
\end{claim}
\begin{proof}
We proceed by induction on $i$. In the base case $i = 0$. Then $D_{x_i} = G$ and $D'_{x_i} = G + \{bs, tb\}$. We have $w_i(e) = x_f(e)$ for all $e \in E(D_{x_i})$ and since $x_f$ is a flow the claim holds for all $v \in V(D_{x_i})\backslash\{s,t\}$. For vertex $s$,
$$w_i(\delta^{out}(s)) = x_f(\delta^{out}(s)) = k = w_i(bs) = w_i(\delta^{in}(s)).$$
The proof of the claim follows similarly for vertex $t$. By definition the claim always holds for vertex $b$. Hence we have proven the base case.
\paragraph{}
Now in the inductive step we suppose the claims holds for $x_0, \dots, x_{i-1}$ and consider $x_i$ with $i \geq 1$. Let $P$ be the augmenting path used to obtain $x_i$ from $x_{i-1}$. If $v \not\in V(P) \cup \{b\}$ then by induction the claim already holds at $v$. Further by definition the claim always holds at $b$. So it remains to verify the claim on vertices of $P$. Let $v \in V(P)\backslash\{s,t\}$. Suppose $uv, vx \in E(P)$. Then we have
\begin{align*}
w_i(\delta^{in}(v)) &= w_{i-1}(\delta^{in}(v)) -x_f(uv) + 1-x_f(vx)\\
&=w_{i-1}(\delta^{out}(v)) + 1-x_f(uv) -x_f(vx) \\
&=w_i(\delta^{out}(v)).
\end{align*}
So the claim holds for all such $v$. For vertex $s$, let $sy \in E(P)$. We have
$$w_i(\delta^{in}(s)) = w_{i-1}(\delta^{in}(s)) -1 + 1-x_f(sy) = w_{i-1}(\delta^{out}(s)) -x_f(sy) = w_i(\delta^{out}(s))$$
and so the claim holds for $s$. The proof for $t$ follows similarly to the proof for $s$.
\end{proof}
\paragraph{}
We initialize our algorithm with integral flow $x = 0$ and 
%Problem 2
\section{}

%Problem 3
\section{}

%Problem 4
\section{}

%Problem 5
\section{}

%Problem 6
\section{}
\end{document}

### 2018-07-30 CS761-RandomizedAlgorithms/assignments/a4.tex
Typed problem 1

@@
 It is immediate that $D_x$ has an augmenting path if and only if $D_x'$ as cycle containing $b$.
-\paragraph{}
+\paragraph{Weighting}
 If $x$ is an integral $s-t$ flow on $G$ we can define an Eulerian weighting on $D'_x$ as follows.
@@
 Let $P$ be the augmenting path used to obtain $x_i$ from $x_{i-1}$.
-If $v \not\in V(P) \cup \{b\}$ then by induction the claim already holds at $v$.
+The digraph $D'_{x_i}$ is obtained from $D'_{x_{i-1}}$ by flipping the orientation of every arc along $P$.
+
+\paragraph{} If $v \not\in V(P) \cup \{b\}$ then by induction the claim already holds at $v$.
 Further by definition the claim always holds at $b$.
@@
 \end{proof}
+\paragraph{Algorithm}
+We initialize our algorithm with integral flow $x = 0$.
+Our algorithm iterates the following step until our flow $x$ has value $k$ (which it then returns).
+Let $w$ be the associated weighting on $D'_{x} = G + \{tb,bs\}$.
+Perform a random walk on $D'_x$ starting from $b$ with transition probability of arc $uv$, $P_{uv}$ equal to
+$$P_{uv} = \frac{w(uv)}{w(\delta^{out}(u))}.$$
+End this random walk when you return to $b$. The walk has identified an augmenting path $P$. Augment flow of $x$ along $P$ and repeat.
 \paragraph{}
-We initialize our algorithm with integral flow $x = 0$ and 
+The expected time to find an augmenting path in any iteration is equal to $H_{b,b}$, the hitting time of $b,b$.
+We have that $H_{b,b} = \frac{1}{\pi_b}$ where $\pi_b$ is the probability of being in $b$ in the stationary distribution.
+\begin{claim}
+	Let $W_x = w(E(D'_x))$. For any $v \in V(D'_x)$, $\pi_v = \frac{w(\delta^{out}(v))}{W_x}$.
+\end{claim}
+\begin{proof}
+	We calculate
+	$$\pi_v = \sum_{u: uv \in E(D'_x)} \pi_uP_{uv} = \sum_{u:uv \in E(D'_x)} \frac{w(\delta^{out}(u))}{W_x}\frac{w(uv)}{w(\delta^{out}(u))} = \frac{w(\delta^{in}(v))}{W_x} = \frac{w(\delta^{out}(v))}{W_x} = \pi_v$$
+	with the second last equality following from out previous Claim. Since 
+	$$\sum_{v \in V(D'_x)} \pi_v = \sum_{v\in V(D'_x)} \frac{w(\delta^{out}(v))}{W_x} = \frac{W_x}{W_x} = 1$$
+	this is the unique stationary distribution.
+\end{proof}
+\paragraph{}
+From this Claim above we see that $H_{b,b} = \frac{W_x}{w(\delta^{out}(b))}$ at each iteration of the algorithm.
+In iteration $i$, $w(\delta^{out}(b)) = w(bs) = k-i$ since $i$ is the value of flow $x$ in iteration $i$.
+Since $w(e)$ is at most $1$ for each $e$ in the edge set of $G$, and $w(e) \leq k\leq |E(G)|$ for $bs, tb$.
+Hence $W_x \leq 2m$ where $m = |E(G)|$.
+Therefore in iteration $i$,
+$$H_{b,b} \leq \frac{2m}{k-i}.$$
+\paragraph{}
+So the total expected running time is at most
+$$\sum_{i=0}^k\frac{m}{k-i} = m\log k \leq m\log m.$$
+Thus the running time $\tilde{O}(|E|)$ as desired.$\blacksquare$
+\newpage
 %Problem 2
@@
 
+\newpage
 %Problem 3
@@
 
+\newpage
 %Problem 4
@@
 
+\newpage
 %Problem 5
@@
 
+\newpage
 %Problem 6

### 2018-07-30 CS761-RandomizedAlgorithms/assignments/a4.tex
A4 update

@@
 \section{}
-
+\paragraph{}
+Consider an instance $I$ of $k$SAT with $n$ variables $x_1,\dots,x_n$ and $m$ clauses $C_1, \dots, C_m$. We solve it using the following generalization of Sch\"oning's Algorithm:
+\begin{enumerate}
+	\item Repeat up to $N$ times, terminating if all clauses of $I$ are satisfied:
+	\begin{enumerate}
+		\item Start with a random initial assignment
+		\item Repeat up to $kn$ steps, terminating if all clauses satisfied: switch the value of a random variable in an unsatisfied clause.
+	\end{enumerate}
+	\item Return a satisfying assignment if it is found; otherwise return ``unsatisfiable".
+\end{enumerate}
+\paragraph{}
+Let $S$ be a satisfying assignment.
+Let $A_i$ be the assignment in the $i$-th step of the algorithm.
+Let $X_i$ be the number of variables that have the same value in $A_i$ and $S$.
+If $X_i = n$ then $A_i = S$ and hence we are done.
 \newpage

### 2018-07-31 CS761-RandomizedAlgorithms/assignments/a4.tex
Think we need to fix q2 from 3n to kn

--- 2018-07-29 CS761-RandomizedAlgorithms/assignments/a4.tex
@@
 \end{proof}
+\paragraph{Algorithm}
+We initialize our algorithm with integral flow $x = 0$.
+Our algorithm iterates the following step until our flow $x$ has value $k$ (which it then returns).
+Let $w$ be the associated weighting on $D'_{x} = G + \{tb,bs\}$.
+Perform a random walk on $D'_x$ starting from $b$ with transition probability of arc $uv$, $P_{uv}$ equal to
+$$P_{uv} = \frac{w(uv)}{w(\delta^{out}(u))}.$$
+End this random walk when you return to $b$. The walk has identified an augmenting path $P$. Augment flow of $x$ along $P$ and repeat.
 \paragraph{}
-We initialize our algorithm with integral flow $x = 0$ and 
+The expected time to find an augmenting path in any iteration is equal to $H_{b,b}$, the hitting time of $b,b$.
+We have that $H_{b,b} = \frac{1}{\pi_b}$ where $\pi_b$ is the probability of being in $b$ in the stationary distribution.
+\begin{claim}
+	Let $W_x = w(E(D'_x))$. For any $v \in V(D'_x)$, $\pi_v = \frac{w(\delta^{out}(v))}{W_x}$.
+\end{claim}
+\begin{proof}
+	We calculate
+	$$\pi_v = \sum_{u: uv \in E(D'_x)} \pi_uP_{uv} = \sum_{u:uv \in E(D'_x)} \frac{w(\delta^{out}(u))}{W_x}\frac{w(uv)}{w(\delta^{out}(u))} = \frac{w(\delta^{in}(v))}{W_x} = \frac{w(\delta^{out}(v))}{W_x} = \pi_v$$
+	with the second last equality following from out previous Claim. Since 
+	$$\sum_{v \in V(D'_x)} \pi_v = \sum_{v\in V(D'_x)} \frac{w(\delta^{out}(v))}{W_x} = \frac{W_x}{W_x} = 1$$
+	this is the unique stationary distribution.
+\end{proof}
+\paragraph{}
+From this Claim above we see that $H_{b,b} = \frac{W_x}{w(\delta^{out}(b))}$ at each iteration of the algorithm.
+In iteration $i$, $w(\delta^{out}(b)) = w(bs) = k-i$ since $i$ is the value of flow $x$ in iteration $i$.
+Since $w(e)$ is at most $1$ for each $e$ in the edge set of $G$, and $w(e) \leq k\leq |E(G)|$ for $bs, tb$.
+Hence $W_x \leq 2m$ where $m = |E(G)|$.
+Therefore in iteration $i$,
+$$H_{b,b} \leq \frac{2m}{k-i}.$$
+\paragraph{}
+So the total expected running time is at most
+$$\sum_{i=0}^k\frac{m}{k-i} = m\log k \leq m\log m.$$
+Thus the running time $\tilde{O}(|E|)$ as desired.$\blacksquare$
+\newpage
 %Problem 2
 \section{}
-
+\paragraph{}
+Consider an instance $I$ of $k$SAT with $n$ variables $x_1,\dots,x_n$ and $m$ clauses $C_1, \dots, C_m$. We solve it using the following generalization of Sch\"oning's Algorithm:
+\begin{enumerate}
+	\item Repeat up to $N$ times, terminating if all clauses of $I$ are satisfied:
+	\begin{enumerate}
+		\item Start with a random initial assignment
+		\item Repeat up to $3n$ steps, terminating if all clauses satisfied: switch the value of a random variable in an unsatisfied clause.
+	\end{enumerate}
+	\item Return a satisfying assignment if it is found; otherwise return ``unsatisfiable".
+\end{enumerate}
+\paragraph{}
+Let $S$ be a satisfying assignment.
+Let $A_i$ be the assignment in the $i$-th step of the algorithm.
+Let $X_i$ be the number of variables that have the same value in $A_i$ and $S$.
+If $X_i = n$ then $A_i = S$ and hence we are done.
+\paragraph{}
+We can view our algorithm as a random walk with states $0$ to $n$ denoting the value of $X_i$ when we are in that state.
+If we are in state $X_i$ at iteration $i$ we transition to state $X_i+1$ with probability $\frac{1}{k}$ and to state $X_i-1$ with probability $\frac{k-1}{k}$.
+\paragraph{}
+Suppose we have assignment $A_i$ at iteration $i$ with $j$ mismatches with $S$, i.e.
+$X_i = n-j$.
+Then we reach state $n$ in $3j$ steps provided there are $j$ steps back and $2j$ steps forward among these $3j$ steps.
+This happens with probability
+$${3j\choose j}(\frac{k-1}{k})^j(\frac{1}{k})^{2j}.$$
+This is a lower bound on reaching $n$ in $3n$ steps.
+\paragraph{}
+We invoke Stirling's formula:
+$$\sqrt{2\pi m}\frac{m}{e}^m \leq m! \leq 2\sqrt{2\pi m}\frac{m}{e}^m$$
+to obtain a better estimate. Observe
+$${3j\choose j} \geq \frac{\sqrt{2\pi 3j}}{4\sqrt{2\pi 2j}\sqrt{2\pi j}}\frac{3j}{e}^{3j}\frac{e}{2j}^{2j}\frac{e}{j}^j = \frac{c}{\sqrt{j}}\frac{27}{4}^j$$
+where $c = \frac{\sqrt{3}}{8\sqrt{\pi}}$ (a constant).
+\paragraph{}
+There our probability of reaching $n$ starting with $j$ mismatches is at least
+$$\frac{c}{\sqrt{j}}\frac{27}{4}^j(\frac{k-1}{k})^j(\frac{1}{k})^{2j}$$
+\newpage
 %Problem 3

### 2018-07-31 CS761-RandomizedAlgorithms/assignments/a4.tex
q2 of a4

@@
 \paragraph{}
-Consider an instance $I$ of $k$SAT with $n$ variables $x_1,\dots,x_n$ and $m$ clauses $C_1, \dots, C_m$. We solve it using the following generalization of Sch\"oning's Algorithm:
+Consider an instance $I$ of $k$SAT with $n$ variables $x_1,\dots,x_n$ and $m$ clauses $C_1, \dots, C_m$.
+We may assume that $k$ is odd, otherwise replace $k$ with $k+1$ and add a dummy variable to each clause of $I$ if necessary.
+Solve $I$ using the following generalization of Sch\"oning's Algorithm:
 \begin{enumerate}
@@
 		\item Start with a random initial assignment
-		\item Repeat up to $3n$ steps, terminating if all clauses satisfied: switch the value of a random variable in an unsatisfied clause.
+		\item Repeat up to $kn$ steps, terminating if all clauses satisfied: switch the value of a random variable in an unsatisfied clause.
 	\end{enumerate}
@@
 $X_i = n-j$.
-Then we reach state $n$ in $3j$ steps provided there are $j$ steps back and $2j$ steps forward among these $3j$ steps.
+Let $a = (k-1)/2$.
+We reach state $n$ in $kj$ steps provided there are $aj$ steps back and $(a+1)j$ steps forward among these $kj$ steps.
 This happens with probability
-$${3j\choose j}(\frac{k-1}{k})^j(\frac{1}{k})^{2j}.$$
-This is a lower bound on reaching $n$ in $3n$ steps.
+$${kj\choose j}(\frac{k-1}{k})^{aj}(\frac{1}{k})^{(a+1)j}.$$
+This is a lower bound on reaching $n$ in $kn$ steps.
 \paragraph{}
 We invoke Stirling's formula:
-$$\sqrt{2\pi m}\frac{m}{e}^m \leq m! \leq 2\sqrt{2\pi m}\frac{m}{e}^m$$
+$$\sqrt{2\pi m}(\frac{m}{e})^m \leq m! \leq 2\sqrt{2\pi m}(\frac{m}{e})^m$$
 to obtain a better estimate. Observe
-$${3j\choose j} \geq \frac{\sqrt{2\pi 3j}}{4\sqrt{2\pi 2j}\sqrt{2\pi j}}\frac{3j}{e}^{3j}\frac{e}{2j}^{2j}\frac{e}{j}^j = \frac{c}{\sqrt{j}}\frac{27}{4}^j$$
-where $c = \frac{\sqrt{3}}{8\sqrt{\pi}}$ (a constant).
+$${kj\choose j} \geq \frac{\sqrt{2\pi kj}}{4\sqrt{2\pi aj}\sqrt{2\pi (a+1)j}}(\frac{kj}{e})^{kj}(\frac{e}{aj})^{aj}(\frac{e}{(a+1)j})^{(a+1)j} = \frac{c}{\sqrt{j}}(\frac{k^k}{a^a(a+1)^{a+1}})^j$$
+where $c = \frac{\sqrt{k}}{4\sqrt{2\pi a(a+1)}}$ (a constant when $k$ is constant).
 \paragraph{}
 There our probability of reaching $n$ starting with $j$ mismatches is at least
-$$\frac{c}{\sqrt{j}}\frac{27}{4}^j(\frac{k-1}{k})^j(\frac{1}{k})^{2j}$$
+$$\frac{c}{\sqrt{j}}(\frac{k^k}{a^a(a+1)^{a+1}})^j(\frac{k-1}{k})^{aj}(\frac{1}{k})^{(a+1)j} = \frac{c}{\sqrt{j}}\frac{(k-1)^{aj}}{a^{aj}(a+1)^{(a+1)j}} = \frac{c}{\sqrt{j}}\frac{(2a)^{aj}}{a^{aj}(a+1)^{(a+1)j}},$$
+and hence we obtain a lower bound on the probability of reaching $n$ from $j$ mismatches of
+$$\frac{c}{\sqrt{j}}\frac{2^{aj}}{(a+1)^{(a+1)j}}.$$
+Therefore the probability of the algorithm finding a satisfying assignment in $kn$ steps is at least
+\begin{align*}
+\sum_{j=0}^n\frac{c}{\sqrt{j}}\frac{2^{aj}}{(a+1)^{(a+1)j}} {n\choose j}\frac{1}{2^n} &\geq \frac{c}{\sqrt{n}}\frac{1}{2^n}\sum_{j=0}^n{n\choose j}\frac{2^{aj}}{(a+1)^{(a+1)j}} \\
+&=\frac{c}{\sqrt{n}}\frac{1}{2^n}(1+\frac{2^a}{(a+1)^{a+1}})^n &\text{by the binomial theorem}
+\end{align*}
+Hence the expected runtime of the algorithm is at most
+$$\frac{\sqrt{n}2^n}{c}(1+\frac{2^a}{(a+1)^{a+1}})^{-n},$$
+a function of $n$ and $k$ (since $a$ and $c$ are functions of $k$).
+When $k\geq 3$ is constant, $\frac{2^a}{(a+1)^{a+1}}$ is a positive constant smaller than $1$, call it $d$.
+Then our expected runtime is at most
+$$O(\sqrt{n}(\frac{2}{d})^n) = o(\sqrt{n}2^n).$$
 \newpage

### 2018-07-31 CS761-RandomizedAlgorithms/assignments/a4.tex
q3 of a4

@@
 \section{}
-
+Let $G$ be a $d$-regular graph. We may assume $G$ is connected. Let $n = |V(G)|$ and $m = |E(G)|$. We invoke the Theorem from Page $7$ of Lecture $15$ which states that 
+$$C(G) \leq 2e^3mR(G)\ln n + n$$
+where $C(G)$ is the cover time of $G$ and $R(G):=\max_{u,v\in V(G)} R_{uv}$ (as in the notes, $R_{uv}$ denotes the effective resistance of $u,v$).
+Since $m = nd$, using this Theorem it will suffice to show that
+$$R(G) = O(n/d)$$
+to obtain the desired result. First observe that $R_{u,v}$ is at most the length of a shortest $u-v$-path. Let $P=x_o=u, x_1, \dots, x_\ell=v$ be a $u-v$-path. Then, via alternating sums,
+$$\phi(u) - \phi(v) = \sum_{i=0}^{\ell-1} \phi(x_i) -\phi(x_{i+1}) \leq \sum_{i=0}^{\ell-1}r_{x_ix_{i+1}} \leq \ell.$$
+Thus to show that $R(G) = O(n/d)$ it will suffice to show that for any $u,v \in V(G)$, the length of a shortest $u-v$-path is at most $O(n/d)$.
+\paragraph{}
+Suppose for a contradiction there exists $u,v \in V(G)$ such that the length of any $u-v$-path is at least $4n/d$.
+Let $P=x_o=u, x_1, \dots, x_\ell=v$ be a shortest $u-v$-path.
+We observe two things.
+First, observe that for any $i \in \{0,\dots, \ell-3\}$, if $x_i$ and $x_{j}$, where $j \geq i+3$ with addition taken modulo $\ell$, have a common neighbour $x$ not on $P$ then we can replace the component $x_i,\dots, x_j$ of $P$ with $x_i, x, x_{j}$ to obtain a shorter $u-v$-path than $P$, a contradiction.
+Hence no such $x_i$ and $x_{j}$ have a common neighbour not on $P$.
+Similarly observe that for any $x_i, x_j$ on $P$ with $j \geq i+2$ we cannot have $x_i$ and $x_j$ adjacent.
+\paragraph{}
+Therefore, if we let $T\subset V(P)$ be the set of every third vertex on $P$ then every pair of vertices in $T$ do not have a common neighbour or adjacency between them.
+Since $G$ is $d$-regular, every vertex of $T$ has $d$ neighbours, and since they have no common neighbours or adjacencies
+$$n \geq d|T| =d\frac{|V(P)|}{3} \geq \frac{4}{3}n > n.$$
+Clearly this is a contradiction.
+\paragraph{}
+Thus we have shown every pair of vertices in $G$ has a path of length at most $4n/d = O(n/d)$, as desired. $\blacksquare$
 \newpage

### 2018-07-31 CS761-RandomizedAlgorithms/assignments/a4.tex
6a and preamble to b

@@
 \section{}
+\subsection{a}
+\paragraph{}
+For this part it will suffice to show that the Markov chain is finite, irreducible, and aperiodic.
+Its state space is the set of all permutations on $[n]$ so indeed it is finite.
+If cards $i$ and $j$ are chosen to swap, yet $i=j$ then we transition to the same permutation, and therefore the Markov chain is aperiodic.
+To see that it is irreducible we have to show that any permutation can be reached from any other permutation by a series of transpositions (swaps).
+Let $\sigma_1, \sigma_2 \in S_n$.
+We can write $\sigma_1 = \tau_1 \circ \dots \circ \tau_k$ and $\sigma_2 = \tau'_1 \circ \dots \circ \tau'_\ell$ where each $\tau_i$ and $\tau'_j$ is a transposition.
+This is a basic result in group theory.
+So starting from $\sigma_1$ we if we make the sequence of transitions
+$$\tau_1^{-1}, \dots, \tau_k^{-1},\tau'_\ell, \dots, \tau'_1$$
+then we reach $\sigma_2$ from $\sigma_1$.
+\subsection{b}
+\paragraph{}
+If we can demonstrate a coupling of the Markov chain with expected convergence time $O(n^2)$ then we know there exists a time $T = O(\frac{n^2}{\epsilon})$ where $\tau(\epsilon) \leq T$, at which point we can use the Coupling Lemma from Page $2$ of Lecture $18$ to achieve the desired results.
 \end{document}

### 2018-07-31 CS761-RandomizedAlgorithms/assignments/a4.tex
q5

@@
 	\section{}
+	\paragraph{}
 
@@
 	\section{}
-
+	\paragraph{}
+	Consider a DNF formula $I$ with clauses $C_1, \dots, C_m$ and variables $x_1, \dots, x_n$.
+We adopt the notation from the problem statement and random process description, and further let $\text{Sat}$ denote the set of satisfying assignments for $I$.
+We begin by observing
+	$$E[P] = E[\frac{W}{N(\sigma)}] = W\cdot E[\frac{1}{N(\sigma)}] = W\sum_{\sigma \in \text{Sat}} Pr(\sigma \text{ is chosen})\frac{1}{N(\sigma)}$$
+	where the random event ``$\sigma$ is chosen" denotes the event that the random process chooses $\sigma$ to satisfy $C_i$.
+Since the process never chooses an assignment not in $\text{Sat}$ the sum ranges over $\sigma$ in Sat.
+	\paragraph{}
+	We now proceed to analyze $Pr(\sigma\text{ is chosen}).$ Let $X_i$ denote the event that the random process begins by choosing clause $C_i$.
+We have $Pr(X_i) = \frac{w_i}{W}$ by definition.
+Let $Y_{i,\sigma}$ denote the event that $\sigma$ is the assignment chosen to satisfy clause $C_i$.
+We have
+	$$Pr(Y_i\mid X_i) = \frac{p^{t_\sigma}(1-p)^{f_\sigma}}{p^{a_i}(1-p)^{b_i}} = \frac{p^{t_\sigma}(1-p)^{f_\sigma}}{w_i}$$
+	where $t_\sigma$ and $f_\sigma$ denote the number of variables $\sigma$ sets to True and False respectively. Let $\text{Sat}(i)$ denote the subset of $\text{Sat}$ which satisfy $C_i$. Since
+	$$\sigma\text{ is chosen} = \bigcup_i X_i \cap Y_i$$
+	and the events $X_i \cap Y_i$ and $X_j \cap Y_j$ are disjoint,
+	$$Pr(\sigma\text{ is chosen}) = \sum_{i \in [m]}Pr(X_i \cap Y_i) = \sum_{i:\sigma \in \text{Sat}(i)} Pr(X_i)Pr(Y_i\mid X_i) = \sum_{i: \sigma \in \text{Sat}(i)} \frac{w_i}{W} \frac{p^{t_\sigma}(1-p)^{f_\sigma}}{w_i}.$$
+	Hence we have
+	$$Pr(\sigma\text{ is chosen}) = \frac{N(\sigma)}{W}p^{t_\sigma}(1-p)^{f_\sigma}.$$
+	Substituting this probability into our expression for $E[P]$ we see that
+	$$E[P] = \sum_{\sigma \in \text{Sat}}p^{t_\sigma}(1-p)^{f_\sigma}.$$
+	Indeed this is precisely the probability that the DNF formula is satisfiable.
+	\paragraph{}
+	Let $\epsilon,\delta \in (0,1]$.
+We now proceed to obtain an $(\epsilon,\delta)$-approximation for the probability that the DNF formula is satisfiable.
+We will run the random process for $t$ trials (to be chosen later), and return the average outcome.
+Let $P_i = \frac{W}{N(\sigma_i)}$ be the outcome of trial $i$.
+Let $\mu = E[P_i]$ for all $i \in[t]$.
+We have previously shown that $\mu$ is precisely the probability that the DNF formula is satisfiable.
+	\paragraph{}
+	Since $N((\sigma_i))\in [1,m]$ we have $P_i \in [\frac{W}{m}, W]$, and we see that $\frac{P_i}{W} \in [\frac{1}{m}, 1]$. Our algorithm will return
+	$$\bar{P} := \frac{1}{t} \sum_{i=1}^t P_i.$$
+	To be an $(\epsilon,\delta)$-approximation we want to show
+	$$Pr(|\bar{P} - \mu|> \epsilon\mu) \leq \delta.$$
+	We define a change of variables, so that $X_i =\frac{P_i}{W}$ and $\bar{X} = \frac{1}{t}\sum_{i=1}^t X_i$.
+Now showing our desired inequality for $(\epsilon,\delta)$-approximation is equivalent to showing
+	$$Pr(|\bar{X} - E[\bar{X}]| > \epsilon E[\bar{X}]) \leq \delta.$$
+	We now invoke the Hoeffding extension of our Chernoff Bound results in Lecture $2$, and sum both tails:
+	$$Pr(|\bar{X} - E[\bar{X}]| > \epsilon E[\bar{X}]) \leq 2\exp(-2t\epsilon^2(\frac{\mu}{W})^2) \leq 2\exp(-2t(\frac{\epsilon}{m})^2) $$
+	where the last inequality follows since $\frac{\mu}{W} \geq \frac{1}{m}$ (as each of $\frac{P_i}{W}$ is at least $\frac{1}{m}$).
+	Hence if we choose $t$ so that
+	$$t \geq \frac{-\ln(\frac{\delta}{2})}{2} \cdot (\frac{m}{\epsilon})^2$$
+	then we will obtain the desired probability bound so that our algorithm is an $(\epsilon,\delta)$-approximation.$\blacksquare$
 	\newpage
@@
 	If we can demonstrate a coupling of the Markov chain with expected convergence time $O(n^2)$ then we know there exists a time $T = O(\frac{n^2}{\epsilon})$ where $\tau(\epsilon) \leq T$, at which point we can use the Coupling Lemma from Page $2$ of Lecture $18$ to achieve the desired results.
+	\paragraph{}
+	We define our joint random process for the coupling now.
+Let $\sigma_1$ be the permutation state of the original Markov chain, and $\sigma_2$ the permutation for the copy.
+Our process proceeds in stages, where stage $i$ denotes that the two permutations agree in $i$ positions.
+At stage $i$ we have three different types of swaps.
+Suppose that the original Markov chain chooses to swap cards with values $a$ and $b$.
+Cases:
+	\begin{enumerate}
+		\item $a$ and $b$ are both in positions whose values agree in $\sigma_1$ and $\sigma_2$. Then the copy Markov chain will choose to swap $a$ and $b$ as well, preserving that we are in stage $i$.
+		\item Similarly if $a$ is in a position whose values agree in $\sigma_1$ and $\sigma_2$, and $b$ is not in such a position, then the copy Markov chain will choose to swap card $b$ into $a$'s position to preserve agreement and stay in stage $i$.
+		\item In final case: that $a$ and $b$ are both in non-agreement positions, we progress stages.
+The copy Markov chain looks at the state the original ends up in after the swap of $a$ and $b$ and chooses a non-agreement position uniformly at random, making a swap (between two non-agreement position cards) to increase the number of positions $\sigma_1$ and $\sigma_2$ agree on by at least $1$.
+That is, we progress at least to stage $i+1$.
+	\end{enumerate}
+	The random process ends when we are in stage $n$. This is a valid coupling since the copy Markov chain has probability of making any given swap $\frac{1}{n^2}$.
+	\paragraph{}
+	The probability that we improve from stage $i$ to stage $\geq i+1$ during a state transition in the random process is the same as the probability that two non-agreement positions are chosen for the swap.
+This occurs with probability
+	$$\frac{(n-i)^2}{n^2}.$$
+	Hence the expected number of transitions before we move from stage $i$ to stage $i+1$ is at most
+	$$\frac{n^2}{(n-i)^2}.$$
+	Therefore, using linearity of expectation, the expected number of transitions before we reach stage $n$ from stage $0$ is at most
+	$$\sum_{i=1}^n\frac{n^2}{(n-i)^2} \leq cn^2 = O(n^2)$$
+	where $c$ is a constant, $c \approx \frac{\pi^2}{6}$. Therefore the expected convergence time of our coupling is $O(n^2)$, and we invoke the Coupling Lemma to complete the proof.$\blacksquare$
 \end{document}

### 2018-08-01 CS761-RandomizedAlgorithms/assignments/a4.tex
4b

@@
 	\paragraph{}
+	Let $G=(V,E)$ be an undirected $d$-regular graph and $S\subseteq V$ a subset of vertices with $|S| \leq |V|/2$.
+	\subsection{a}
+	\paragraph{}
 
+	\subsection{b}
+	\paragraph{}
+	We may assume there exists a vertex $v$ in $S$ such if we start the random walk at $v$ then $\sum_{i \in S}p_t(i)\geq (1-\phi(S))^t$ for any $t\geq 0$.
+	\paragraph{}
+	Suppose that $|S| \leq |V|^{0.99}$.
+We may also assume that $\phi(S)$ is small, say $\phi(S) < \frac{1}{2}$, for otherwise it easy to find a good approximation of $\phi(S)$.
+We start our random walk from vertex $v$, and proceed for $t$ (to be chosen later) steps.
+From our assumption that $\phi(S)<\frac{1}{2}$ and our strong lower bound we have
+	$$p_t(S) \geq (1-\phi(S))^t \geq \exp(-2\phi(S)t)$$
+	using that $(1-p)\geq \exp(-2p)$ for $p<\frac{1}{2}$.
+	We may also observe that
+	$$\pi(S) = \sum_{v\in S} \pi(v) = \sum_{v\in S} \frac{d}{2m} = |S|\frac{d}{2d|V|}\leq |V|^{-0.01}.$$
+	Hence
+	$$p_t(S) - \pi(S) \geq \exp(-2\phi(S)t) - |V|^{-0.01}.$$
+	If we choose 
+	$$t = \frac{0.01}{\phi(S)}\frac{\ln(2|V|)}{2}$$	
+	then
+	$$\exp(-2\phi(S)t) \geq 2|V|^{-0.01}$$
+	and so
+	$$p_t(S) - \pi(S) \geq |V|^{-0.01}.$$
+	\paragraph{}
+	Now let $S_k^i$ denote the of $k$ highest probability vertices at time step $i$. Define $\theta$ as
+	$$\theta:= \min_{i,k}\phi(S_k^i).$$
+	From Lovasz-Simonovits we have
+	\begin{align*}
+	p_t(S) - \pi(S) &\leq \sqrt{\text{vol}(S)}(1-\frac{1}{8}\theta^2)^t\\
+	&= \sqrt{d|S|}(1-\frac{1}{8}\theta^2)^t\\
+	&\leq \sqrt{d}|V|^{0.45}\exp(-t\frac{\theta^2}{8})\\
+	&\leq \sqrt{d}|V|^{0.45}\exp(-\frac{0.005}{8\phi(S)}\ln(2|V|)\theta^2)\\
+	&=2\sqrt{d}|V|^{0.45}|V|^{-\frac{0.005\theta^2}{8\phi(S)}}.
+	\end{align*}
+	Combining our lower and upper bound on $p_t(S)-\pi(S)$ we see that
+	$$\frac{1}{2}d^{-\frac{1}{2}}|V|^{-0.46} \leq |V|^{-\frac{0.005\theta^2}{8\phi(S)}}$$
+	Which implies there exists a small constant $c$ for which
+	$$\theta^2 \leq c\phi(S).$$
+	Hence $\theta = O(\sqrt{\phi(S)}).$ Thus one of the sets in $\{S^i_k\}$ over all $i,k$ have conductance $\theta=O(\sqrt{\phi(S)})$, giving the desired sparse cut.$\blacksquare$
 	\newpage

### 2018-08-01 CS761-RandomizedAlgorithms/assignments/a4.tex
q4

@@
 \begin{document}
+	\paragraph{}
+	I collaborated with Sharat, and discussed briefly with Harry and Logan.
 	%Problem 1
@@
 	\paragraph{}
-
+	Let $t \geq 0$.
+We want to show the existence of a vertex $v$ in $S$ such that $\sum_{i\in S} p^t(i) \geq 1- t\phi(S)$ when the random walk is started with $p_0 =\chi_v$.
+\emph{Note:} I'm not sure if I got the quantifiers backwards for this problem.
+Sorry if I did, I hope the ideas are still worth some merit in this case.
+	\paragraph{}
+	Choose a vertex $u \in S$ uniformly at random to start the walk at. Then
+	$$E[\chi_Sp^t] = \sum_{u \in S} \chi_SW^t\chi_u \cdot\frac{1}{|S|} = \frac{1}{|S|}\chi_SW^t\chi_S.$$
+	Since $W^t = A^tD^{-t} = \frac{1}{d^t}A^t$, we thus have that
+	$$E[\chi_Sp^t] = \frac{1}{d^t|S|}\sum_{u,v\in S}w_t(\{u\},\{v\}) = \frac{1}{d^t|S|}w(S,S)$$
+	where we use $w_t(A,B)$ to denote the number of walks of length $t$ from any vertex in $A$ to any vertex in $B$ in $G$, for $A, B\subseteq V$.
+	\paragraph{}
+	We want
+	$$E[\chi_S p^t] \geq 1-t\phi(S)$$
+	and so using the above equation for this expection it would suffice to show (after rearranging)
+	$$w_t(S,S) \geq d^t|S| - td^{t-1}|\delta(S)|.$$
+	Since $G$ is $d$-regular, a length $t$ walk starting in $S$ and ending at any vertex in $V$ is uniquely determined by $d$ choices of which vertex to branch to next, from each vertex in the walk.
+Thus $w_t(S,V) = d^t|S|$.
+Hence our desired inequality is equivalent to showing
+	$$td^{t-1}|\delta(S)| \geq w_t(S,V) - w_t(S,S) = w_t(S,V\backslash S).$$
+	But it is not hard to see that the Left Hand Side is a generous overcounting of $w_t(S,V\backslash S).$ Indeed, the expression $td^{t-1}|\delta(S)|$ counts the number of walks which start in $S$ and use an edge of $\delta(S)$ in at least one of their $t$ positions, then have free choice of any of $d$ edges for each of their remaining $t-1$ positions.
+Therefore the inequality
+	$$td^{t-1}|\delta(S)| \geq w_t(S,V\backslash S)$$
+	holds, and so the equivalent statement
+	$$E[\chi_S p^t] \geq 1-t\phi(S)$$
+	holds as well.
+	\paragraph{}
+	Since we know $E[\chi_S p^t] \geq 1-t\phi(S)$ there must exist some vertex $v \in S$ for which $\chi_S p^t \geq 1-t\phi(S)$ when $p_0 = \chi_v$.
+For otherwise such an expectation could not be attained.$\blacksquare$
 	\subsection{b}

### 2018-08-08 CS761-RandomizedAlgorithms/assignments/final.tex
started exam

--- 2018-07-31 CS761-RandomizedAlgorithms/assignments/a4.tex
@@
 \lhead{\today}
-\rhead{W. Justin Toth CS761-Randomized Algorithms Assignment 4} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth CS761-Randomized Algorithms Final Exam} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
+
 %Problem 1
@@
 \paragraph{}
-	Let $G=(V,E)$ be a directed graph, let $s,t\in V$ be source and sink vertices respectively and let $k$ be a positive integer.
-We may assume that $\delta^{in}(s) = \delta^{out}(t) = \emptyset$.
-Let $x_f \in [0,1]^E$ be a fractional $s-t$ flow on $G$ of value $k$.
-That is, $x_f$ satisfies the equations
-	$$ x_f(\delta^{out}(s)) = x_f(\delta^{in}(t)) = k$$
-	and
-	$$x_f(\delta^{in}(v)) = x_f(\delta^{out}(v))$$
-	for all $v \in V\backslash\{s,t\}$.
+Denote the applicants $A_1, \dots, A_n$ in the order which they appear according to the random permutation.
+Let $X_i$ be the $0$-$1$ indicator random variable for the event that we pay penalty $C$ for  $A_i$.
+Hence $X_i=1$ if and only if we hire $A_i$ and later fire $A_i$.
 \paragraph{}
-	We're going to need some concepts from network flow theory before we move on.
-Lucky our flow being $0$-$1$ simplifies things quite a bit.
-Given a digraph $D$ and a flow $x \in \{0,1\}^E$ we define the residual digraph $D_x$ as follows:
-	\begin{itemize}
-		\item$V(D_x) = V(D)$
-		\item for every arc $uv \in E(D)$ if $x(uv) = 0$ add arc $uv$ to $E(D_x)$. We call such arcs ``forward" arcs.
-		\item for every arc $uv \in E(D)$ if $x(uv) = 1$ add arc $vu$ to $E(D_x)$. we call such arcs ``backward" arcs.
-	\end{itemize} 
-	An $s-t$ path in $D_x$ is called an \emph{augmenting path}.
-It is an easy observation to see that if $x$ is a flow of value $k$ and $P$ is an augmenting path in $D_x$, then we can obtain a flow of value $k+1$ by increasing $x(e)$ from $0$ to $1$ for every forward arc in $P$, and decreasing $x(e)$ from $1$ to $0$ for every backward arc in $P$.
-Using the max-flow min-cut theorem we can see that if there is no augmenting path in $D_x$ then $x$ is a maximum flow.
-Simply take the cut $S$ defined by vertices reachable from $s$ in $D_x$ (this is an $s-t$ cut since $t$ is not reachable).
-Since there is no augmenting paths, all forward arcs in $\delta^{out}(S)$ are saturated (have $x$-value $1$).
-So the value of $x$ is at least (and hence equal to) the capacity of cut $S$, implying that $x$ is a maximum flow.
+We hire $A_i$ if and only if $A_i$ is ranked highest among the first $A_1, \dots, A_i$ applicants. This happens with probability
+$$\frac{1}{i}$$
+since the ordering was chosen uniformly at random.
+Given that we hire $A_i$, we fire $A_i$ if and only if $A_i$ is not ranked highest among the following $A_i, \dots, A_n$ applicants.
+This happens with probability
+$$\frac{n-i}{n-i+1}.$$
+Since the events that $A_i$ is ranked first among the first $i$ candidates, and not ranked first among the last $n-i+1$ candidates are independent we have
+$$Pr(X_i) = \frac{n-i}{i(n-i+1)}.$$
 \paragraph{}
-	Our algorithm will maintain an integral flow in every step, and a weighting on the residual digraph so that (an auxiliary graph closely related to) the residual digraph is weighted-Eulerian.
-This will allow us to efficiently use random walks to identify augmenting paths for our flow.
-Due to our fractional flow $x_f$ we know that we will always be able to find such a path while the value of $x$ is less than $k$, as we are sure $x$ is not maximum.
+Thus our expected total penalty is
+$$\sum_{i=1}^n C\cdot Pr(X_i) = C \sum_{i=1}^n\frac{n-i}{i(n-i+1)} = CS_n$$
+where $S_n$ denotes $\sum_{i=1}^n\frac{n-i}{i(n-i+1)}$. Now observe that
+$$S_n = n\sum_{i=1}^n\frac{1}{i(n-i+1)} - \sum_{i=1}^n\frac{1}{n-i+1}= n\sum_{i=1}^n\frac{1}{i(n-i+1)} - H_n =  n\sum_{i=1}^n\frac{1}{i(n-i+1)} - \log(n)$$
+where $H_n$ denotes the $n$-th harmonic number. Now since
+$$\int_{x=1}^n \frac{1}{x(n-x+1)} dx = \frac{2\log(n)}{n+1}$$
+we have that
+$$S_n \leq \frac{2n\log(n)}{n+1} -\log(n) = \frac{(n-1)\log(n)}{n+1}.$$
+Therefore our expected total payoff is at most
+$$C\cdot \frac{(n-1)\log(n)}{n+1}. \blacksquare$$
+%Problem 2
+\newpage
+\section{}
+Let $G$ be an undirected graph, and $C$ the size of a minimum cut in $G$. Let $\alpha \in \R_{\geq 1}$.
+\begin{lemma}\label{lemma:2}
+Let $F \subseteq E(G)$ be a cut such that $|F| \leq \alpha C$. Then the probability that $F$ is output by Karger's random contraction algorithm is at least
+$$\Omega({\frac{1}{n^{2\alpha}}}).$$
+\end{lemma}
+\begin{proof}
+We say the algorithm ``succeeds" if $F$ is output by the algorithm.
+The algorithm succeeds if and only if we never contract an edge of $F$.
+At each iteration of the algorithm, every subset of vertices in the contracted graph corresponds to a subset of vertices  in the original graph (by expanding contracted vertices), and hence the minimum cut value of contracted graph at iteration $i$ is at least $C$.
 \paragraph{}
-	We will need one more auxiliary graph before we describe the algorithm.
-Let $D'_x$ be the digraph obtained from $D_x$ by adding a special vertex $b$, and two arcs: $tb$ and $bs$.
-It is immediate that $D_x$ has an augmenting path if and only if $D_x'$ as cycle containing $b$.
-	\paragraph{Weighting}
-	If $x$ is an integral $s-t$ flow on $G$ we can define an Eulerian weighting on $D'_x$ as follows.
-For every forward arc $uv \in E(D_x)$ let $w(uv) = x_f(uv)$.
-For every backward arc $vu \in E(D_x)$ let $w(vu) = 1- x_f(uv)$.
-Let $k'$ be the value of $x$.
-Let $w(tb) = w(bs) = k-k'$.
+In particular the degree of each vertex in the contracted graph is at least $C$. So the number of edges in the $i$-th iteration is at least
+$$\frac{(n-i+1)C}{2}.$$
+So when we pick a random edge to contract in this iteration, the probability that we pick an edge of $F$ is at most
+$$\frac{|F|}{\frac{(n-i+1)C}{2}} \leq \frac{2\alpha}{n-i+1}$$
+using that $|F| \leq \alpha C.$ So the probability that the algorithm succeeds is at least
+$$\Pi_{i=1}^n(1-\frac{2\alpha}{n-i+1}) = \frac{2\alpha!}{n(n-1)\dots(n-2\alpha + 1)} = \Omega(\frac{1}{n^{2\alpha}})$$
+\end{proof}
 \begin{claim}
-		Let $x_0=0,x_1, \dots, x_{k-1}\in \{0,1\}^E$ be integral $s-t$ flows on $G$ such that $x_{i+1}$ is obtained from $x_i$ by augmenting $x_i$ along an augmenting path in $D_{x_i}$.
-Let $w_0,w_1,\dots, w_{k-1}$ be the weightings of each respective $D'_{x_i}$.
-Then for all $i$
-		$$w_i(\delta^{out}(v)) = w_i(\delta^{in}(v))$$ for all $v \in V(D'_{x_i})$.
+The number of cuts in $G$ with at most $\alpha C$ edges is at most $O(n^{2\alpha})$.
 \end{claim}
 \begin{proof}
-		We proceed by induction on $i$.
-In the base case $i = 0$.
-Then $D_{x_i} = G$ and $D'_{x_i} = G + \{bs, tb\}$.
-We have $w_i(e) = x_f(e)$ for all $e \in E(D_{x_i})$ and since $x_f$ is a flow the claim holds for all $v \in V(D_{x_i})\backslash\{s,t\}$.
-For vertex $s$,
-		$$w_i(\delta^{out}(s)) = x_f(\delta^{out}(s)) = k = w_i(bs) = w_i(\delta^{in}(s)).$$
-		The proof of the claim follows similarly for vertex $t$. By definition the claim always holds for vertex $b$. Hence we have proven the base case.
-		\paragraph{}
-		Now in the inductive step we suppose the claims holds for $x_0, \dots, x_{i-1}$ and consider $x_i$ with $i \geq 1$.
-Let $P$ be the augmenting path used to obtain $x_i$ from $x_{i-1}$.
-The digraph $D'_{x_i}$ is obtained from $D'_{x_{i-1}}$ by flipping the orientation of every arc along $P$.
+Let $\cF$ be the set of cuts in $G$ with at most $\alpha C$ edges. For $F \in \cF$ let $X_F$ be the event that Karger's algorithm outputs $F$. By Lemma \ref{lemma:2}
+$$Pr(X_F) \geq \Omega(\frac{1}{n^{2\alpha}}).$$
+Now we see that, one the one hand
+$$Pr(\bigcup_{F \in \cF} X_F) \leq 1$$
+and on the other hand, as the events events $X_F$ are all disjoint (the algorithm only outputs one answer),
+$$Pr(\bigcup_{F \in \cF} X_F) = \sum_{F \in \cF} Pr(X_F) \leq \sum_{F \in \cF} \Omega(\frac{1}{n^{2\alpha}}) = |\cF|\Omega(\frac{1}{n^{2\alpha}}).$$
+Combining inequality and rearranging for $|\cF|$ we have
+$$|\cF| \leq O(n^{2\alpha})$$
+as desired.
+\end{proof}
+%Problem 3
+\newpage
+\section{}
 
-		\paragraph{} If $v \not\in V(P) \cup \{b\}$ then by induction the claim already holds at $v$.
-Further by definition the claim always holds at $b$.
-So it remains to verify the claim on vertices of $P$.
-Let $v \in V(P)\backslash\{s,t\}$.
-Suppose $uv, vx \in E(P)$.
-Then we have
-		\begin{align*}
-		w_i(\delta^{in}(v)) &= w_{i-1}(\delta^{in}(v)) -x_f(uv) + 1-x_f(vx)\\
-		&=w_{i-1}(\delta^{out}(v)) + 1-x_f(uv) -x_f(vx) \\
-		&=w_i(\delta^{out}(v)).
-		\end{align*}
-		So the claim holds for all such $v$. For vertex $s$, let $sy \in E(P)$. We have
-		$$w_i(\delta^{in}(s)) = w_{i-1}(\delta^{in}(s)) -1 + 1-x_f(sy) = w_{i-1}(\delta^{out}(s)) -x_f(sy) = w_i(\delta^{out}(s))$$
-		and so the claim holds for $s$. The proof for $t$ follows similarly to the proof for $s$.
-	\end{proof}
-	\paragraph{Algorithm}
-	We initialize our algorithm with integral flow $x = 0$.
-Our algorithm iterates the following step until our flow $x$ has value $k$ (which it then returns).
-Let $w$ be the associated weighting on $D'_{x} = G + \{tb,bs\}$.
-Perform a random walk on $D'_x$ starting from $b$ with transition probability of arc $uv$, $P_{uv}$ equal to
-	$$P_{uv} = \frac{w(uv)}{w(\delta^{out}(u))}.$$
-	End this random walk when you return to $b$. The walk has identified an augmenting path $P$. Augment flow of $x$ along $P$ and repeat.
-	\paragraph{}
-	The expected time to find an augmenting path in any iteration is equal to $H_{b,b}$, the hitting time of $b,b$.
-We have that $H_{b,b} = \frac{1}{\pi_b}$ where $\pi_b$ is the probability of being in $b$ in the stationary distribution.
-	\begin{claim}
-		Let $W_x = w(E(D'_x))$. For any $v \in V(D'_x)$, $\pi_v = \frac{w(\delta^{out}(v))}{W_x}$.
-	\end{claim}
-	\begin{proof}
-		We calculate
-		$$\pi_v = \sum_{u: uv \in E(D'_x)} \pi_uP_{uv} = \sum_{u:uv \in E(D'_x)} \frac{w(\delta^{out}(u))}{W_x}\frac{w(uv)}{w(\delta^{out}(u))} = \frac{w(\delta^{in}(v))}{W_x} = \frac{w(\delta^{out}(v))}{W_x} = \pi_v$$
-		with the second last equality following from out previous Claim. Since 
-		$$\sum_{v \in V(D'_x)} \pi_v = \sum_{v\in V(D'_x)} \frac{w(\delta^{out}(v))}{W_x} = \frac{W_x}{W_x} = 1$$
-		this is the unique stationary distribution.
-	\end{proof}
-	\paragraph{}
-	From this Claim above we see that $H_{b,b} = \frac{W_x}{w(\delta^{out}(b))}$ at each iteration of the algorithm.
-	In iteration $i$, $w(\delta^{out}(b)) = w(bs) = k-i$ since $i$ is the value of flow $x$ in iteration $i$.
-Since $w(e)$ is at most $1$ for each $e$ in the edge set of $G$, and $w(e) \leq k\leq |E(G)|$ for $bs, tb$.
-Hence $W_x \leq 2m$ where $m = |E(G)|$.
-Therefore in iteration $i$,
-	$$H_{b,b} \leq \frac{2m}{k-i}.$$
-	\paragraph{}
-	So the total expected running time is at most
-	$$\sum_{i=0}^k\frac{m}{k-i} = m\log k \leq m\log m.$$
-	Thus the running time $\tilde{O}(|E|)$ as desired.$\blacksquare$
+%Problem 4
 \newpage
-	%Problem 2
 \section{}
-	\paragraph{}
-	Consider an instance $I$ of $k$SAT with $n$ variables $x_1,\dots,x_n$ and $m$ clauses $C_1, \dots, C_m$.
-We may assume that $k$ is odd, otherwise replace $k$ with $k+1$ and add a dummy variable to each clause of $I$ if necessary.
-Solve $I$ using the following generalization of Sch\"oning's Algorithm:
-	\begin{enumerate}
-		\item Repeat up to $N$ times, terminating if all clauses of $I$ are satisfied:
-		\begin{enumerate}
-			\item Start with a random initial assignment
-			\item Repeat up to $kn$ steps, terminating if all clauses satisfied: switch the value of a random variable in an unsatisfied clause.
-		\end{enumerate}
-		\item Return a satisfying assignment if it is found; otherwise return ``unsatisfiable".
-	\end{enumerate}
-	\paragraph{}
-	Let $S$ be a satisfying assignment.
-Let $A_i$ be the assignment in the $i$-th step of the algorithm.
-Let $X_i$ be the number of variables that have the same value in $A_i$ and $S$.
-If $X_i = n$ then $A_i = S$ and hence we are done.
-	\paragraph{}
-	We can view our algorithm as a random walk with states $0$ to $n$ denoting the value of $X_i$ when we are in that state.
-If we are in state $X_i$ at iteration $i$ we transition to state $X_i+1$ with probability $\frac{1}{k}$ and to state $X_i-1$ with probability $\frac{k-1}{k}$.
-	\paragraph{}
-	Suppose we have assignment $A_i$ at iteration $i$ with $j$ mismatches with $S$, i.e.
-$X_i = n-j$.
-Let $a = (k-1)/2$.
-We reach state $n$ in $kj$ steps provided there are $aj$ steps back and $(a+1)j$ steps forward among these $kj$ steps.
-This happens with probability
-	$${kj\choose j}(\frac{k-1}{k})^{aj}(\frac{1}{k})^{(a+1)j}.$$
-	This is a lower bound on reaching $n$ in $kn$ steps.
-	\paragraph{}
-	We invoke Stirling's formula:
-	$$\sqrt{2\pi m}(\frac{m}{e})^m \leq m! \leq 2\sqrt{2\pi m}(\frac{m}{e})^m$$
-	to obtain a better estimate. Observe
-	$${kj\choose j} \geq \frac{\sqrt{2\pi kj}}{4\sqrt{2\pi aj}\sqrt{2\pi (a+1)j}}(\frac{kj}{e})^{kj}(\frac{e}{aj})^{aj}(\frac{e}{(a+1)j})^{(a+1)j} = \frac{c}{\sqrt{j}}(\frac{k^k}{a^a(a+1)^{a+1}})^j$$
-	where $c = \frac{\sqrt{k}}{4\sqrt{2\pi a(a+1)}}$ (a constant when $k$ is constant).
-	\paragraph{}
-	There our probability of reaching $n$ starting with $j$ mismatches is at least
-	$$\frac{c}{\sqrt{j}}(\frac{k^k}{a^a(a+1)^{a+1}})^j(\frac{k-1}{k})^{aj}(\frac{1}{k})^{(a+1)j} = \frac{c}{\sqrt{j}}\frac{(k-1)^{aj}}{a^{aj}(a+1)^{(a+1)j}} = \frac{c}{\sqrt{j}}\frac{(2a)^{aj}}{a^{aj}(a+1)^{(a+1)j}},$$
-	and hence we obtain a lower bound on the probability of reaching $n$ from $j$ mismatches of
-	$$\frac{c}{\sqrt{j}}\frac{2^{aj}}{(a+1)^{(a+1)j}}.$$
-	Therefore the probability of the algorithm finding a satisfying assignment in $kn$ steps is at least
-	\begin{align*}
-	\sum_{j=0}^n\frac{c}{\sqrt{j}}\frac{2^{aj}}{(a+1)^{(a+1)j}} {n\choose j}\frac{1}{2^n} &\geq \frac{c}{\sqrt{n}}\frac{1}{2^n}\sum_{j=0}^n{n\choose j}\frac{2^{aj}}{(a+1)^{(a+1)j}} \\
-	&=\frac{c}{\sqrt{n}}\frac{1}{2^n}(1+\frac{2^a}{(a+1)^{a+1}})^n &\text{by the binomial theorem}
-	\end{align*}
-	Hence the expected runtime of the algorithm is at most
-	$$\frac{\sqrt{n}2^n}{c}(1+\frac{2^a}{(a+1)^{a+1}})^{-n},$$
-	a function of $n$ and $k$ (since $a$ and $c$ are functions of $k$).
-When $k\geq 3$ is constant, $\frac{2^a}{(a+1)^{a+1}}$ is a positive constant smaller than $1$, call it $d$.
-Then our expected runtime is at most
-	$$O(\sqrt{n}(\frac{2}{d})^n) = o(\sqrt{n}2^n).$$
+
+%Problem 5
 \newpage
-	%Problem 3
 \section{}
-	Let $G$ be a $d$-regular graph. We may assume $G$ is connected. Let $n = |V(G)|$ and $m = |E(G)|$. We invoke the Theorem from Page $7$ of Lecture $15$ which states that 
-	$$C(G) \leq 2e^3mR(G)\ln n + n$$
-	where $C(G)$ is the cover time of $G$ and $R(G):=\max_{u,v\in V(G)} R_{uv}$ (as in the notes, $R_{uv}$ denotes the effective resistance of $u,v$).
-Since $m = nd$, using this Theorem it will suffice to show that
-	$$R(G) = O(n/d)$$
-	to obtain the desired result. First observe that $R_{u,v}$ is at most the length of a shortest $u-v$-path. Let $P=x_o=u, x_1, \dots, x_\ell=v$ be a $u-v$-path. Then, via alternating sums,
-	$$\phi(u) - \phi(v) = \sum_{i=0}^{\ell-1} \phi(x_i) -\phi(x_{i+1}) \leq \sum_{i=0}^{\ell-1}r_{x_ix_{i+1}} \leq \ell.$$
-	Thus to show that $R(G) = O(n/d)$ it will suffice to show that for any $u,v \in V(G)$, the length of a shortest $u-v$-path is at most $O(n/d)$.
-	\paragraph{}
-	Suppose for a contradiction there exists $u,v \in V(G)$ such that the length of any $u-v$-path is at least $4n/d$.
-Let $P=x_o=u, x_1, \dots, x_\ell=v$ be a shortest $u-v$-path.
-We observe two things.
-First, observe that for any $i \in \{0,\dots, \ell-3\}$, if $x_i$ and $x_{j}$, where $j \geq i+3$ with addition taken modulo $\ell$, have a common neighbour $x$ not on $P$ then we can replace the component $x_i,\dots, x_j$ of $P$ with $x_i, x, x_{j}$ to obtain a shorter $u-v$-path than $P$, a contradiction.
-Hence no such $x_i$ and $x_{j}$ have a common neighbour not on $P$.
-Similarly observe that for any $x_i, x_j$ on $P$ with $j \geq i+2$ we cannot have $x_i$ and $x_j$ adjacent.
-	\paragraph{}
-	Therefore, if we let $T\subset V(P)$ be the set of every third vertex on $P$ then every pair of vertices in $T$ do not have a common neighbour or adjacency between them.
-Since $G$ is $d$-regular, every vertex of $T$ has $d$ neighbours, and since they have no common neighbours or adjacencies
-	$$n \geq d|T| =d\frac{|V(P)|}{3} \geq \frac{4}{3}n > n.$$
-	Clearly this is a contradiction.
-	\paragraph{}
-	Thus we have shown every pair of vertices in $G$ has a path of length at most $4n/d = O(n/d)$, as desired. $\blacksquare$
+
+%Problem 6
 \newpage
-	%Problem 4
 \section{}
-	\paragraph{}
 
+%Problem 7
 \newpage
-	%Problem 5
 \section{}
-	\paragraph{}
-	Consider a DNF formula $I$ with clauses $C_1, \dots, C_m$ and variables $x_1, \dots, x_n$.
-We adopt the notation from the problem statement and random process description, and further let $\text{Sat}$ denote the set of satisfying assignments for $I$.
-We begin by observing
-	$$E[P] = E[\frac{W}{N(\sigma)}] = W\cdot E[\frac{1}{N(\sigma)}] = W\sum_{\sigma \in \text{Sat}} Pr(\sigma \text{ is chosen})\frac{1}{N(\sigma)}$$
-	where the random event ``$\sigma$ is chosen" denotes the event that the random process chooses $\sigma$ to satisfy $C_i$.
-Since the process never chooses an assignment not in $\text{Sat}$ the sum ranges over $\sigma$ in Sat.
-	\paragraph{}
-	We now proceed to analyze $Pr(\sigma\text{ is chosen}).$ Let $X_i$ denote the event that the random process begins by choosing clause $C_i$.
-We have $Pr(X_i) = \frac{w_i}{W}$ by definition.
-Let $Y_{i,\sigma}$ denote the event that $\sigma$ is the assignment chosen to satisfy clause $C_i$.
-We have
-	$$Pr(Y_i\mid X_i) = \frac{p^{t_\sigma}(1-p)^{f_\sigma}}{p^{a_i}(1-p)^{b_i}} = \frac{p^{t_\sigma}(1-p)^{f_\sigma}}{w_i}$$
-	where $t_\sigma$ and $f_\sigma$ denote the number of variables $\sigma$ sets to True and False respectively. Let $\text{Sat}(i)$ denote the subset of $\text{Sat}$ which satisfy $C_i$. Since
-	$$\sigma\text{ is chosen} = \bigcup_i X_i \cap Y_i$$
-	and the events $X_i \cap Y_i$ and $X_j \cap Y_j$ are disjoint,
-	$$Pr(\sigma\text{ is chosen}) = \sum_{i \in [m]}Pr(X_i \cap Y_i) = \sum_{i:\sigma \in \text{Sat}(i)} Pr(X_i)Pr(Y_i\mid X_i) = \sum_{i: \sigma \in \text{Sat}(i)} \frac{w_i}{W} \frac{p^{t_\sigma}(1-p)^{f_\sigma}}{w_i}.$$
-	Hence we have
-	$$Pr(\sigma\text{ is chosen}) = \frac{N(\sigma)}{W}p^{t_\sigma}(1-p)^{f_\sigma}.$$
-	Substituting this probability into our expression for $E[P]$ we see that
-	$$E[P] = \sum_{\sigma \in \text{Sat}}p^{t_\sigma}(1-p)^{f_\sigma}.$$
-	Indeed this is precisely the probability that the DNF formula is satisfiable.
-	\paragraph{}
-	Let $\epsilon,\delta \in (0,1]$.
-We now proceed to obtain an $(\epsilon,\delta)$-approximation for the probability that the DNF formula is satisfiable.
-We will run the random process for $t$ trials (to be chosen later), and return the average outcome.
-Let $P_i = \frac{W}{N(\sigma_i)}$ be the outcome of trial $i$.
-Let $\mu = E[P_i]$ for all $i \in[t]$.
-We have previously shown that $\mu$ is precisely the probability that the DNF formula is satisfiable.
-	\paragraph{}
-	Since $N((\sigma_i))\in [1,m]$ we have $P_i \in [\frac{W}{m}, W]$, and we see that $\frac{P_i}{W} \in [\frac{1}{m}, 1]$. Our algorithm will return
-	$$\bar{P} := \frac{1}{t} \sum_{i=1}^t P_i.$$
-	To be an $(\epsilon,\delta)$-approximation we want to show
-	$$Pr(|\bar{P} - \mu|> \epsilon\mu) \leq \delta.$$
-	We define a change of variables, so that $X_i =\frac{P_i}{W}$ and $\bar{X} = \frac{1}{t}\sum_{i=1}^t X_i$.
-Now showing our desired inequality for $(\epsilon,\delta)$-approximation is equivalent to showing
-	$$Pr(|\bar{X} - E[\bar{X}]| > \epsilon E[\bar{X}]) \leq \delta.$$
-	We now invoke the Hoeffding extension of our Chernoff Bound results in Lecture $2$, and sum both tails:
-	$$Pr(|\bar{X} - E[\bar{X}]| > \epsilon E[\bar{X}]) \leq 2\exp(-2t\epsilon^2(\frac{\mu}{W})^2) \leq 2\exp(-2t(\frac{\epsilon}{m})^2) $$
-	where the last inequality follows since $\frac{\mu}{W} \geq \frac{1}{m}$ (as each of $\frac{P_i}{W}$ is at least $\frac{1}{m}$).
-	Hence if we choose $t$ so that
-	$$t \geq \frac{-\ln(\frac{\delta}{2})}{2} \cdot (\frac{m}{\epsilon})^2$$
-	then we will obtain the desired probability bound so that our algorithm is an $(\epsilon,\delta)$-approximation.$\blacksquare$
+
+%Problem 8
 \newpage
-	%Problem 6
 \section{}
-	\subsection{a}
-	\paragraph{}
-	For this part it will suffice to show that the Markov chain is finite, irreducible, and aperiodic.
-Its state space is the set of all permutations on $[n]$ so indeed it is finite.
-If cards $i$ and $j$ are chosen to swap, yet $i=j$ then we transition to the same permutation, and therefore the Markov chain is aperiodic.
-To see that it is irreducible we have to show that any permutation can be reached from any other permutation by a series of transpositions (swaps).
-Let $\sigma_1, \sigma_2 \in S_n$.
-We can write $\sigma_1 = \tau_1 \circ \dots \circ \tau_k$ and $\sigma_2 = \tau'_1 \circ \dots \circ \tau'_\ell$ where each $\tau_i$ and $\tau'_j$ is a transposition.
-This is a basic result in group theory.
-So starting from $\sigma_1$ we if we make the sequence of transitions
-	$$\tau_1^{-1}, \dots, \tau_k^{-1},\tau'_\ell, \dots, \tau'_1$$
-	then we reach $\sigma_2$ from $\sigma_1$.
-	\subsection{b}
-	\paragraph{}
-	If we can demonstrate a coupling of the Markov chain with expected convergence time $O(n^2)$ then we know there exists a time $T = O(\frac{n^2}{\epsilon})$ where $\tau(\epsilon) \leq T$, at which point we can use the Coupling Lemma from Page $2$ of Lecture $18$ to achieve the desired results.
-	\paragraph{}
-	We define our joint random process for the coupling now.
-Let $\sigma_1$ be the permutation state of the original Markov chain, and $\sigma_2$ the permutation for the copy.
-Our process proceeds in stages, where stage $i$ denotes that the two permutations agree in $i$ positions.
-At stage $i$ we have three different types of swaps.
-Suppose that the original Markov chain chooses to swap cards with values $a$ and $b$.
-Cases:
-	\begin{enumerate}
-		\item $a$ and $b$ are both in positions whose values agree in $\sigma_1$ and $\sigma_2$. Then the copy Markov chain will choose to swap $a$ and $b$ as well, preserving that we are in stage $i$.
-		\item Similarly if $a$ is in a position whose values agree in $\sigma_1$ and $\sigma_2$, and $b$ is not in such a position, then the copy Markov chain will choose to swap card $b$ into $a$'s position to preserve agreement and stay in stage $i$.
-		\item In final case: that $a$ and $b$ are both in non-agreement positions, we progress stages.
-The copy Markov chain looks at the state the original ends up in after the swap of $a$ and $b$ and chooses a non-agreement position uniformly at random, making a swap (between two non-agreement position cards) to increase the number of positions $\sigma_1$ and $\sigma_2$ agree on by at least $1$.
-That is, we progress at least to stage $i+1$.
-	\end{enumerate}
-	The random process ends when we are in stage $n$. This is a valid coupling since the copy Markov chain has probability of making any given swap $\frac{1}{n^2}$.
-	\paragraph{}
-	The probability that we improve from stage $i$ to stage $\geq i+1$ during a state transition in the random process is the same as the probability that two non-agreement positions are chosen for the swap.
-This occurs with probability
-	$$\frac{(n-i)^2}{n^2}.$$
-	Hence the expected number of transitions before we move from stage $i$ to stage $i+1$ is at most
-	$$\frac{n^2}{(n-i)^2}.$$
-	Therefore, using linearity of expectation, the expected number of transitions before we reach stage $n$ from stage $0$ is at most
-	$$\sum_{i=1}^n\frac{n^2}{(n-i)^2} \leq cn^2 = O(n^2)$$
-	where $c$ is a constant, $c \approx \frac{\pi^2}{6}$. Therefore the expected convergence time of our coupling is $O(n^2)$, and we invoke the Coupling Lemma to complete the proof.$\blacksquare$
+
+%Problem 9
+\newpage
+\section{}
+
+%Problem 10
+\newpage
+\section{}
+
 \end{document}

### 2018-08-08 CS761-RandomizedAlgorithms/assignments/final.tex
almost done 2

@@
 \end{proof}
-\begin{claim}
+\begin{claim}\label{claim2}
 The number of cuts in $G$ with at most $\alpha C$ edges is at most $O(n^{2\alpha})$.
@@
 \end{proof}
+\paragraph{}
+If we repeat Karger's algorithm $t$ times waiting for a particular cut $F \in \cF$ (defined in above proof) we have a failure probability at most
+$$(1-\Omega(\frac{1}{n^{2\alpha}}))^t \leq \exp(\frac{t}{O(n^{2{\alpha}})})$$
+and hence we can make this failure probability small with a choice of $t \in O(n^{2\alpha})$ (the constant would depend on how good you want the success probability).
+\paragraph{}
+So we give the following randomized algorithm for enumerating all cuts of $G$ with at most $\alpha C$ edges.
+Since there are at most $O(n^{2\alpha})$ such cuts by Claim \ref{claim2}, and we are treating $\alpha$ as a constant, there exists some absolute constant $K \in O(n^{2\alpha})$ for which there are exactly $K$ cuts in $\cF$.
+Our algorithm proceeds as follows:
+\begin{enumerate}
+\item test
+%Todo: type: iterate K times, each iteration run Karger's to see a new cut in \cF. Print them all.
+\end{enumerate}
 %Problem 3

### 2018-08-09 CS761-RandomizedAlgorithms/assignments/final.tex
q2

@@
 \paragraph{}
-So we give the following randomized algorithm for enumerating all cuts of $G$ with at most $\alpha C$ edges.
-Since there are at most $O(n^{2\alpha})$ such cuts by Claim \ref{claim2}, and we are treating $\alpha$ as a constant, there exists some absolute constant $K \in O(n^{2\alpha})$ for which there are exactly $K$ cuts in $\cF$.
-Our algorithm proceeds as follows:
+So we give the following randomized algorithm for enumerating all cuts of $G$ with at most $\alpha C$ edges. Let $k$ be some parameter to be chosen later. Our algorithm proceeds as follows:
 \begin{enumerate}
-\item test
-%Todo: type: iterate K times, each iteration run Karger's to see a new cut in \cF. Print them all.
+\item Let $\cL = \emptyset$.
+\item While fail$<k$
+	\begin{enumerate}
+	\item Repeat Karger's algorithm at most $t$ times hoping for a cut $F\in \cF\backslash \cL$.
+	\item If such $F$ is found, add $F$ to $\cL$ and set fail$=0$.
+	\item Otherwise set fail$=$fail$+1$.
 	\end{enumerate}
+\item Output $\cL$.
+\end{enumerate}
+\paragraph{}
+In words the algorithm uses the boosted version of Karger's to try to find a cut with at most $\alpha C$ edges that it has not seen before.
+It stops if it fails to find such a cut $k$ times in a row, and prints all the cuts found this way.
+\paragraph{}
+The algorithm fails if in some $i$-th round it does not find a cut $F \in \cF\backslash \cL$ $k$ times in a row, while $\cF \backslash \cL\neq \emptyset$.
+We know from \ref{claim2} there are at most $O(n^{2\alpha})$ such rounds $i$.
+Let $X_i$ be a random variable indicating failure in the $i$-th round.
+We can choose $t$ so that $t$ repetitions of Karger's algorithm finds $F \in \cF\backslash \cL$ (provided such $F$ exists) with high probability, say $0.99$.
+\paragraph{}
+Then the probability that we fail in the $i$-th round is
+$$(0.01)^k.$$
+So the probability we fail in some round is at most
+$$\frac{O(n^{2\alpha})}{100^k}.$$
+Since $\alpha$ is constant, we can choose $k \in O(\log n)$ to make this failure probability arbitrarily small.
+Let $T$ be the time to run Karger's algorithm once.
+Since there are $O(n^{2\alpha})$ rounds our algorithm takes
+$$O(n^{2\alpha}ktT) = O(n^{4\alpha}\log n T)$$
+time.$ \blacksquare$
+
 %Problem 3

### 2018-08-09 CS761-RandomizedAlgorithms/assignments/final.tex
q4

@@
 \section{}
-
+\paragraph{}
+We throw $n^{0.99}$ balls into $n$ bins uniformly at random. Using union bound, the probability that a bin has at least $k$ balls, for $k$ chosen later, is at most 
+$${n \choose k}(\frac{1}{n})^k.$$
+Using the bound ${n^{0.99}\choose k} <(\frac{n^{0.99}e}{k})^k$ this probability is at most
+$$(\frac{n^{0.99}e}{k})^k(\frac{1}{n})^k= \frac{e^k}{n^{0.0k}k^k}\leq \frac{e^k}{k^k}$$
+where the last inequality follows since we will choose $k \geq 0$ and since $n\geq 1$ this implies $n^\frac{k}{100} \geq 1$.
+Again invoking union bound
+$$Pr(\exists\text{a bin with at least $k$ balls}) \leq n\frac{e^k}{k^k} = n(\frac{e}{k})^k.$$ 
+Since we want
+$$Pr(\exists\text{a bin with at least $k$ balls}) \leq \frac{1}{2}$$
+we need to find $k$ such that
+$$n(\frac{e}{k})^k \leq \frac{1}{2}.$$
 %Problem 4
@@
 \section{}
-
+\paragraph{}
+We always use addition, $+$, over the appropriate field (ie.
+$\F_n$ for inputs and $\F_m$ for outputs).
+Our algorithm is very simple.
+Given input $z$ choose $x \in \{0,\dots, n-1\}$ uniformly at random.
+Let $y = z-x$.
+Lookup $F(x)$ and $F(y)$ and return $F(x) + F(y)$.
+\paragraph{}
+If both $F(x)$ and $F(y)$ are not modified by our adversary then our return value satisfies
+$$F(x) + F(y) = F(x+y) = F(z)$$
+using the linearity of $F$ and the fact that $x+y = z$.
+So a necessary condition for our algorithm to fail is that one of $F(x)$ or $F(y)$ was modified by our adversary.
+For a uniform random element of the table, the probability of modification is at most $\frac{1}{5}$.
+Hence
+$$Pr(F(x)\text{ or } F(y)\text{ was modified}) \leq Pr(F(x) \text{ was modified}) + Pr(F(y)\text{ was modified}) \leq \frac{2}{5}.$$
+Therefore the probability of failure is at most
+$$\frac{2}{5} < \frac{1}{2}.$$
+Hence our algorithm succeeds with probability at least $\frac{1}{2}$ as desired. $\blacksquare$
 %Problem 5

### 2018-08-09 CS761-RandomizedAlgorithms/assignments/final.tex
q6

@@
 $$n(\frac{e}{k})^k \leq \frac{1}{2}.$$
+
 %Problem 4
@@
 Hence our algorithm succeeds with probability at least $\frac{1}{2}$ as desired. $\blacksquare$
+
 %Problem 5
@@
 \section{}
-
+\paragraph{}
+Let $\cA$ be an algorithm which obtain an $(\epsilon, \frac{1}{4})$-approximation to some parameter $\mu$. That is, when $X$ is the output value of $\cA$,
+$$Pr(|X - \mu| \leq \epsilon\mu) \leq \frac{1}{4}.$$
+We want to show how to obtain an $(\epsilon,\delta)$-approximation to $\mu$.
+Suppose we perform $k = \ceil{\log(\frac{1}{\delta})}$ independent runs of $\cA$ and output the median.
+We may assume $k$ is even, if not replace $k$ with $k+1$.
+Let $X_1, \dots, X_k$ be random variables each denoting the output value of one run of $\cA$.
+Instead of ordering them by the order the trials happen, order them sorted by value, i.e.
+$$X_1 \leq X_2 \leq \dots\leq X_k.$$
+Our $(\epsilon,\delta)$-approximation algorithm outputs $X_\frac{k}{2}$, the median value of our independent trials.
+Let $X$ denote the output of our new algorithm.
+We observe that $|X - \mu|\leq \epsilon \mu$ if and only if
+$$|X_i - \mu | \leq \epsilon \mu, \quad \forall i=1 \dots \frac{k}{2}.$$
+Therefore
+$$Pr(|X-\mu| \leq \epsilon \mu) = \Pi_{i=1}^\frac{k}{2} Pr(|X_i - \mu| \leq \epsilon \mu) \leq \frac{1}{4^\frac{k}{2}}.$$
+Since $k = \ceil{\log(\frac{1}{\delta})}$,
+$$\frac{1}{4^\frac{k}{2}} \leq \frac{1}{4^{\frac{1}{2}\log(\frac{1}{\delta})}} = \delta.$$
+Therefore
+$$Pr(|X-\mu| \leq \epsilon \mu) \leq \delta. \blacksquare$$
 %Problem 7

### 2018-08-09 CS761-RandomizedAlgorithms/assignments/final.tex
Q5

--- 2018-08-09 CS761-RandomizedAlgorithms/assignments/final.tex
@@
 \section{}
-
+\paragraph{}
+Let $S_1$ and $S_2$ be multisets of positive integers.
+Let $|S_1| = |S_2| = n$ be the number of elements in each multiset.
+If the sets are of different sizes it is trivial to decide they are not equal.
+Let $h:[n] \rightarrow [cn]$ be a $2$-universal hash function chosen uniformly at random.
+\paragraph{}
+We're going to define multiset operations.
+They are exactly what you'd expect them to be, so feel free to skip this paragraph as I only add them for sake of completeness.
+Let $S_1\cap S_2$ denote the largest multiset that is a sub-multiset of $S_1$ and a sub-multiset of $S_2$.
+By sub-multiset I mean $A$ is a sub-multiset of $B$ ($A\subseteq B$) if and only if for every element $e$ of $A$, the number of times $e$ appears in $A$ is at most the number of times $e$ appears in $B$.
+We define multiset addition so that $A+B$ is the multiset of elements $e$ so that $e$ appears as many times as it does in $A$ plus as many times as it does in $B$.
+We also need multiset difference.
+For multisets $A$ and $B$ we define $A-B$ to be the unique multiset so to that $(A-B) + (A\cap B) = A$.
+My multiset equality, $A=B$, we meant that $A$ and $B$ have precisely the same elements appearing the same number of times.
+\paragraph{}
+If $S_1 = S_2$ then $h(S_1) = h(S_2)$ and so the algorithm never gives false negatives.
+Observe that $h(S_1) = h(S_1\cap S_2) + h(S_1-S_2)$ and $h(S_2) = h(S_1\cap S_2) + h(S_2-S_1)$.
+Hence the algorithm can only give a false positive if $h(S_1-S_2) = h(S_2 - S_1)$.
+We will bound the probability of failure by the probability of this happening.
+Let $d = |S_1-S_2| =|S_2-S_1|$.
+For $a \in S_1 - S_2$ and $b \in S_2 - S_1$ define $X_{a,b}$ to be the event that $h(a) = h(b)$.
+Further for each $a \in S_1 - S_2$ define
+$$X_a  = \bigcup_{b \in S_2 - S_1} X_{a,b}.$$
+Notice that $X_a$ is the disjoint union of events $X_{a,b}$. Let $X$ be a random variable counting the number of collisions between elements in $S_1-S_2$ and elements in $S_2-S_1$. We have
+$$E[X] = \sum_{a \in S_1 - S_2} E[X_a] = \sum_{a\in S_1 - S_2} \sum_{b \in S_2-S_1} Pr(h(a) = h(b)) \leq  \sum_{a\in S_1 - S_2} \sum_{b \in S_2-S_1} \frac{1}{cn} = \frac{d^2}{cn}.$$
+The inequality follows from $h$ being $2$-universal.
+For the algorithm to fail it is necessary to have $X \geq d$, for otherwise there are an insufficient number of collisions to have $h(S_1-S_2) = h(S_2-S_1)$.
+Therefore we can bound the probability of failure by $Pr(X  \geq d)$.
+By Markov's inequality we see that,
+$$Pr(X \geq d) \leq \frac{E[X]}{d} \leq \frac{d}{cn}.$$
+Hence our probability of failure is at most $\frac{d}{cn}$.
+Assuming we can hash in constant time and lookup in a hash table in constant time, the algorithm takes $O(n)$ time to hash each element in $S_1$ and $S_2$, then takes $O(cn)$ time to verify the $cn$ counters each associated with an entry of each of the two hash tables.
+Since we need $c \geq 1$ to avoid guaranteed collisions, this implies the algorithm runs in $O(cn)$ time with failure probability at most $\frac{d}{cn}$.
+\paragraph{}
+In the previous paragraph we demonstrated a trade-off between running time and failure probability with respect to choice of parameter $c$.
+To obtain a Monte Carlo algorithm, we need a failure probability which is independent of the input size.
+We make our extension by giving an explicit choice of $c$.
+Let $p \in (0,1)$.
+Then if we choose $c = \ceil{\frac{d}{pn}}$ then we obtain a failure probability at most $p$ and a running time $O(pd)$.
+$\blacksquare$
 %Problem 6

### 2018-08-10 CS761-RandomizedAlgorithms/assignments/final.tex
partial 7

@@
 \section{}
+\paragraph{}
+Choose $c = \min \{\frac{1}{5}, \}$.Let $U = \{u_1,\dots, u_n\}$ be a set of ground elements, and let $S_1, \dots, S_m \subseteq U$ be a set of subsets of $U$ such that $|S_i| = \frac{n}{2}$ for all $i\in [m]$.
+Consider a uniform random coloring of the elements of $U$ either red or blue.
+For each $i \in [n]$ let $R_i$ be an indicator random variable for the event that $u_i$ is colored red, and similarly let $B_i$ be an indicator random variable for the event that $u_i$ is colored blue.
+For each $j \in [m]$ let $R^j := \sum_{u_i \in S_j} R_i$, and similarly let $B^j := \sum_{u_i \in S_j} B_i$.
+Note that $B^j = |S_j| - R^j$.
+\paragraph{}
+First we observe that $R^j$ is the sum of independent random variables with $E[R^j] = \sum_{u_i \in S_j} \frac{1}{2} = \frac{n}{4}$. Further observe that
+ $$m \leq {n \choose n/2} \leq 2^{nH(1/2)} = 2^n$$
+using that each $S_j$ is of size $\frac{n}{2}$ and the Lemma from Page $1$ of Lecture $12$.  Hence 
+$$\frac{4c\sqrt{n\log m}}{n} \leq \frac{4c\sqrt{n^2}}{n}  = 4c < 1.$$
+ So by Chernoff,
+\begin{align*}
+Pr(R^j \geq \frac{n}{4} + c\sqrt{n\log m}) &= Pr(R^j \geq (1+\frac{4c\sqrt{n\log m}}{n})\frac{n}{4})\\
+ &\leq e^{-(\frac{4c\sqrt{n\log m}}{n})^2\frac{n}{12}} \\
+ &= e^{-\frac{4c^2}{3}\log m}\\
+ &=\frac{1}{m^{\frac{4c^2}{3}}}\\
+ &!< \frac{1}{m}.
+ \end{align*}
 

### 2018-08-10 CS761-RandomizedAlgorithms/assignments/final.tex
problem 7

--- 2018-08-09 CS761-RandomizedAlgorithms/assignments/final.tex
@@
 \section{}
-
+\paragraph{}
+Choose $c = 1$.
+Let $U = \{u_1,\dots, u_n\}$ be a set of ground elements, and let $S_1, \dots, S_m \subseteq U$ be a set of subsets of $U$ such that $|S_i| = \frac{n}{2}$ for all $i\in [m]$.
+Consider a uniform random coloring of the elements of $U$ either red or blue.
+For each $i \in [n]$ let $R_i$ be an indicator random variable for the event that $u_i$ is colored red, and similarly let $B_i$ be an indicator random variable for the event that $u_i$ is colored blue.
+For each $j \in [m]$ let $$X^j := \sum_{u_i \in S_j} R_i - \sum_{u_i \in S_j} B_i.$$
+Then showing $S_j$ has between $\frac{n}{4} - c\sqrt{n\log m}$ and $\frac{n}{4} +c\sqrt{n\log m}$ red elements is equivalent to showing $$|X^j| \leq 2c\sqrt{n\log m}.$$
+\paragraph{}
+For each $i \in [n]$ let $X_i$ be a random variable such that $X_i = 1$ if $u_i$ is red and $X_i = -1$ if $u_i$ is  blue.
+Since each $S_j$ has $\frac{n}{2}$ elements, each $X^j$ is the sum of exactly $\frac{n}{2}$ such $X_i$.
+Using Corollary $4.8$ in Probability and Computing by Mitzenmacher and Upfal we have
+\begin{align*}
+Pr(|X^j| \geq 2c\sqrt{n\log m}) &\leq 2e^\frac{-(2c\sqrt{n\log m})^2}{2\frac{n}{2}} \\
+&= 2e^{-4c^2\log m}\\
+&=\frac{2}{m^4}\\
+&< \frac{1}{m} &\text{ (since $m \geq 2$, otherwise the problem is trivial)}.
+\end{align*}
+Then, applying a union bound,  the probability that there exists $j\in[m]$ such that $|X^j| \geq 2c\sqrt{n\log m}$ is strictly less than $1$.
+Thus with positive probability, for all $j\in [m]$ $$|X^j| \leq 2c\sqrt{n\log m}$$
+as desired.$\blacksquare$
 %Problem 8

### 2018-08-10 CS761-RandomizedAlgorithms/assignments/final.tex
q8

@@
 \section{}
-
+\subsection{a}
+\paragraph{}
+Suppose $s$ is a random $n$-bit string.
+Let $k \in [n]$ be a positive integer.
+Observe that a $k$-consecutive substring in $s$ is uniquely determined by the starting position of the substring.
+Therefore there are at most $n$ potential $k$-consecutive substrings in $s$.
+Denote by $s(i)$ the substring $s_is_{i+1}\dots s_{i+k-1}$.
+For $s(i)$ to be $k$-consecutive, either bits of $s(i)$ are $0$ or all $1$.
+Thus,
+$$Pr(s(i) \text{ is $k$-consecutive}) = \frac{2}{2^k} = \frac{1}{2^{k-1}}.$$
+So by applying union bound, using there are at most $n$ such $s(i)$, the probability there exists a $k$-consecutive substring is at most
+$$\frac{n}{2^{k-1}}.$$
+To ensure there are no $k$-consecutive substrings with high probability we desire
+$$\frac{n}{2^{k-1}} < 0.01$$
+Hence we choose
+$$k = \log(100n) +1 = O(\log n)$$
+to achieve the desired result.$\blacksquare$
+\subsection{b}
+\paragraph{}
+We can model this problem via a random walk.
+Our states will be $S_0, S_1, \dots, S_k$.
+$S_0$ transitions to $S_1$ with probability $1$.
+For $i \in \{1,\dots, k-1\}$, $S_i$ transitions to $S_{i+1}$ with probability $\frac{1}{2}$, and transitions to $S_1$ with probability $\frac{1}{2}$.
+$S_k$ transitions to $S_k$ with probability $1$.
+To interpret the original problem as this random walk imagine starting in $S_0$ and reading the random bits of $s$.
+After reading $1$ bit, $b$, you move from $S_0$ to $S_1$, then as long as you keep seeing the same bit value as $b$ when you read the next bit of $s$ you advance from your current $S_i$ to $S_{i+1}$.
+But once you see the opposite value to $b$ you return to $S_1$, and treat $\bar{b}$ as the value you're looking for to advance.
+If at any time you reach $S_k$, you stop, having just read a $k$-consecutive substring of $s$.
+\paragraph{}
+We want to bound the expected time to reach $S_k$ from $S_0$.
+For each $i = 0, \dots k$ let $h_i$ denote the expected number of steps to reach $S_k$.
+Clearly, $h_0 = h_1 + 1$ and $h_k = 0$.
+For $j \in [k-1]$ we develop the following recurrence:
+$$h_{k-j} = \frac{1}{2}(h_{k-j+1} + 1) + \frac{1}{2}h_1.$$
+We claim that for all $j \in [k-1]$, $h_{k-j} = \sum_{i=1}^j \frac{1}{2^i}(1+h_1)$. We proceed by induction. In the base case $j=1$. From our recurrence
+$$h_{k-1} = \frac{1}{2}(h_k+1) + \frac{1}{2}h_1 = \frac{1}{2}(1+h_1),$$
+using $h_k = 0$. So the base case holds. Now consider $j \in [k-1]\backslash\{1\}$ and suppose the claim holds for $h_{k-j+1}$. Then we have
+\begin{align*}
+h_{k-j} &= \frac{1}{2}(h_{k-j+1} + 1) + \frac{1}{2}h_1\\
+&= \frac{1}{2}( \sum_{i=1}^{j-1} \frac{1}{2^i}(1+h_1) + 1) + \frac{1}{2}h_1 \\
+&=\sum_{i=1}^{j-1}\frac{1}{2^{i+1}}(1+h_1) + \frac{1}{2} + \frac{1}{2}h_1\\
+&= \sum_{i=2}^{j}\frac{1}{2^{i}}(1+h_1) + \frac{1}{2} + \frac{1}{2}h_1 \\
+&= \sum_{i=1}^j \frac{1}{2^i}(1+h_1).
+\end{align*}
+Therefore the claim holds by induction.
+\paragraph{}
+So, using partial sums of the geometric series, we have that 
+$$h_1 = \sum_{i=1}^{k-1} \frac{1}{2^i}(1+h_1) = \frac{2^{k-1}-1}{2^{k-1}}(1+h_1).$$
+Rearranging we see that
+$$h_1 = 2^{k-1} - 1$$
+which implies that
+$$h_0 = 2^{k-1}.$$
+Let $X$ denote the number of bits needed before seeing a $k$-consecutive substring. By Markov's inequality, our equation for $h_0$ implies
+$$Pr(X \geq 100\cdot2^{k-1}) \leq 0.01.$$
+Thus with high probability, $0.99$ we find a $k$-consecutive substring with at most $100 \cdot 2^{k-1}$ bits.
+That is to say, setting $n = 100\cdot 2^{k-1}$ ensures a $k$-consecutive substring with high probability.
+Rearranging for $k$, we see that for
+$$k =\frac{n}{100}\log(n) + 1$$
+$s$ has a $k$-consecutive substring with high probability. $\blacksquare$
 %Problem 9

### 2018-08-10 CS761-RandomizedAlgorithms/assignments/final.tex
problem 10

@@
 \section{}
-
+\paragraph{}
+Let $G$ be a connected, undirected, non-bipartite graph. Consider a Markov Chain $M$ with state space $V(G) \times V(G)$ and transition probabilties:
+$$P_{ab, uv} = \begin{cases}
+\frac{1}{d(a)d(b)}, &\text{ if } u \in N(a)\text{ and } v \in N(b) \\
+0, &\text{otherwise}
+\end{cases}$$
+Then a state in $M$ represents a position $(a,b)$ where $a$ is the position of the cat and $b$ is the position of the mouse, and $P_{ab,uv}$ represents the probability of the cat being in position $u$ and the mouse being in position $v$ in the subsequent timestep.
+We can equivalently view $M$ as the Markov chain for a random walk on the graph $G'$ where $V(G') = V(G) \times V(G)$ and $\{(a,b), (u,v)\} \in E(G')$ if and only if $u \in N(a)$ and $v \in N(b)$.
+\paragraph{}
+Hence $|V(G')| = |V(G)|^2$ and 
+$$2|E(G')| = \sum_{u,v \in V(G)} d(u)d(v) = 4|E(G)|^2$$
+i.e.
+$|E(G')| = 2|E(G)|^2$.
+Note the cat catches the mouse when the random walk on $G'$ reaches vertex $(a,a)$ for some $a \in V(G)$.
+So if we upper bound $h_{x',y'}$ (the hitting time), for any $x', y' \in V(G')$ then we have successfully upper bounded the expected time for the cat to catch the mouse since it is simply $h_{x',y'}$ where $ux$ is the start state of the cat and mouse, and $y' = (a,a)$ for some $a \in V(G)$.
+\paragraph{}
+Using Lemma $7.15$ of Probability and Computing we see that 
+$$h_{a',b'} < 2|E(G')|$$
+provided that $a'b' \in E(G')$.
+Now consider any $x'=(a,b), y'=(u,v) \in V(G')$.
+Let $P_1$ be an $au$-path in $G$, which exists since $G$ is connected.
+Similarly let $P_2$ be a $bv$-path in $G$.
+Construct an $x'y'$-path $P$ in $G'$ by taking the sequence of vertices $(a,b) \dots (u,b)$ following $P_1$ in the first position and keeping $b$ fixed in the second position, then taking the sequence of vertices $(u,b) \dots (u,v)$ following $P_2$ in the second position and keeping $u$ fixed in the first.
+Then $|P_2|  = |P_1| + |P_2| \leq 2|V(G)|$.
+\paragraph{}
+As $P$ is an $x'y'$-path, $h_{x',y'}$ is upper bounded by the sum of hitting times on the edges of $P$. That is, we have
+$$h_{x',y'} \leq \sum_{a'b' \in E(P)} h_{a'b'} <  \sum_{a'b' \in E(P)} 2|E(G')| \leq 4|V(G)||E(G')| \leq 8|V(G)||E(G)|^2$$
+with the second inequality following from Lemma $7.15$. Thus we have shown an upper bound of $O(|V(G)|\cdot|E(G)|^2)$ on the expected time for the cat to catch the mouse.$\blacksquare$
 \end{document}

### 2018-08-10 CS761-RandomizedAlgorithms/assignments/final.tex
some ideas

@@
 \section{}
-
+\paragraph{}
+Consider a random walk that starts at a vertex $v \in V$ for $G=(V,E)$ an undirected graph, that stops when it reaches vertex $s$ or vertex $t$.
+Let $p(v)$ denote the probability that if the random walk starts at $v \in V$ then it reaches $s$ before $t$.
+It is easy to see that $p(s) = 1$ and $p(t) = 0$.
+\paragraph{}
+For $v \in G$ we argued in the Lecture $15$, Page $5$, that $h_{x,v} = \phi_{xv}$ for an appropriate flow.
 %Problem 10

### 2018-08-11 CS761-RandomizedAlgorithms/assignments/final.tex
finished

@@
 \paragraph{}
-We throw $n^{0.99}$ balls into $n$ bins uniformly at random. Using union bound, the probability that a bin has at least $k$ balls, for $k$ chosen later, is at most 
-$${n \choose k}(\frac{1}{n})^k.$$
-Using the bound ${n^{0.99}\choose k} <(\frac{n^{0.99}e}{k})^k$ this probability is at most
-$$(\frac{n^{0.99}e}{k})^k(\frac{1}{n})^k= \frac{e^k}{n^{0.0k}k^k}\leq \frac{e^k}{k^k}$$
-where the last inequality follows since we will choose $k \geq 0$ and since $n\geq 1$ this implies $n^\frac{k}{100} \geq 1$.
-Again invoking union bound
-$$Pr(\exists\text{a bin with at least $k$ balls}) \leq n\frac{e^k}{k^k} = n(\frac{e}{k})^k.$$ 
-Since we want
-$$Pr(\exists\text{a bin with at least $k$ balls}) \leq \frac{1}{2}$$
-we need to find $k$ such that
-$$n(\frac{e}{k})^k \leq \frac{1}{2}.$$
-
+We throw $n^{0.99}$ balls into $n$ bins uniformly at random.  The probability that a given bin has $k$ balls is 
+$$p_k = {n^{0.99} \choose k} (\frac{1}{n})^k(1-\frac{1}{n})^{n^{0.99}-k} = \frac{1}{k!}\frac{n^{0.99}(n^{0.99}-1)\dots(n^{0.99} - k + 1)}{n^k}(1-\frac{1}{n})^{n^{0.99}-k}.$$
+This value is approximately
+$$p_k \approx \frac{e^{-n^{0.01}}(n^{0.01})^k}{k!}.$$
+In other words, the probability that a given has load at least $k$ is a Poisson random variable.
+In Poisson, all bins can be treated as independent so the probability that no bin has load at least $k$ is at most
+$$(1-\frac{e^{-n^{0.01}}(n^{0.01})^k}{k!})^n\leq \exp(\frac{-e^{-n^{0.01}}n^{1 + \frac{k}{100}}}{k!})$$
+To translate the Poisson case back to the exact case, Theorem $5.7$ tells us our final probability will be effected by a factor of at most $e\sqrt{n^{0.99}}$.
+Since we want a final probability of at most $\frac{1}{2}$ we need to choose $k$ so that
+$$\exp(\frac{-e^{-n^{0.01}}n^{1 + \frac{k}{100}}}{k!}) \leq \frac{1}{2e\sqrt{n^{0.99}}}$$
+Or equivalently
+$$\frac{-e^{-n^{0.01}}n^{1 + \frac{k}{100}}}{k!} \leq -\ln(2e\sqrt{n^{0.99}}).$$
+Equivalently yet again,
+$$k! \leq \frac{e^{-n^{0.01}}n^{1 + \frac{k}{100}}}{\ln(2e\sqrt{n^{0.99}})}$$
+From $(5.5)$ of Probability and Computing, 
+$$k! \leq k(\frac{k}{e})^k.$$
+So again we can rewrite our desired inequality as
+$$k(\frac{k}{e})^k \leq \frac{e^{-n^{0.01}}n^{1 + \frac{k}{100}}}{\ln(2e\sqrt{n^{0.99}})}$$
+and taking logarithms
+$$2k\ln(k) \leq \ln(e^{-n^{0.01}}n^{1 + \frac{k}{100}}) - \ln(\ln(2e\sqrt{n^{0.99}}) )= -n^{0.01} + (1+\frac{k}{100})\ln(n) - \ln(\ln(2e\sqrt{n^{0.99}})).$$
+Now if we move the $k$ over from the right hand side and multiply the constants through we get the desired inequality
+$$k(2\ln(k) - \frac{\ln(n)}{100}) \leq -n^{0.01} + \ln(n) - \ln(\ln(2e\sqrt{n^{0.99}})).$$
+I ran out of time, but I hope that solving this equation for $k$ would yield a decent bound.
 %Problem 4
@@
 \paragraph{}
-Let $\cA$ be an algorithm which obtain an $(\epsilon, \frac{1}{4})$-approximation to some parameter $\mu$. That is, when $X$ is the output value of $\cA$,
+Let $\cA$ be an algorithm which obtains an $(\epsilon, \frac{1}{4})$-approximation to some parameter $\mu$. That is, when $X$ is the output value of $\cA$,
 $$Pr(|X - \mu| \leq \epsilon\mu) \leq \frac{1}{4}.$$
@@
 For $v \in G$ we argued in the Lecture $15$, Page $5$, that $h_{x,v} = \phi_{xv}$ for an appropriate flow.
+\paragraph{}
+Sorry I didn't have time for this problem. My vague idea was to draw a connection between hitting times and the desired probabilities, then relate them back to $\phi$ on some electrical network.
 %Problem 10

### 2018-10-04 CO759-ExtendendFormulations/assignments/A1.tex
starting A1

--- 2018-08-10 CS761-RandomizedAlgorithms/assignments/final.tex
@@
 \newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}
-
+\newcommand{\conv}[1]{\ensuremath{\text{conv}(#1) }}
+\newcommand{\cone}[1]{\ensuremath{\text{cone}(#1) }}
+\newcommand{\facets}[1]{\ensuremath{\text{facets}(#1) }}
 %END MACROS
@@
 \lhead{\today}
-\rhead{W. Justin Toth CS761-Randomized Algorithms Final Exam} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\rhead{W. Justin Toth. Extended Formulations - A1} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-
 %Problem 1
@@
 \paragraph{}
-Denote the applicants $A_1, \dots, A_n$ in the order which they appear according to the random permutation.
-Let $X_i$ be the $0$-$1$ indicator random variable for the event that we pay penalty $C$ for  $A_i$.
-Hence $X_i=1$ if and only if we hire $A_i$ and later fire $A_i$.
+Let a polytope $Q\subseteq \R^n$ and a linear map $p:\R^n \rightarrow \R^d$ be an extension of a polytope $P\subseteq \R^d$.
+Since $p$ is a linear map, there exists a matrix $A \in \R^{d\times n}$ such that $p(x) = Ax$ for all $x \in Q$.
 \paragraph{}
-We hire $A_i$ if and only if $A_i$ is ranked highest among the first $A_1, \dots, A_i$ applicants. This happens with probability
-$$\frac{1}{i}$$
-since the ordering was chosen uniformly at random.
-Given that we hire $A_i$, we fire $A_i$ if and only if $A_i$ is not ranked highest among the following $A_i, \dots, A_n$ applicants.
-This happens with probability
-$$\frac{n-i}{n-i+1}.$$
-Since the events that $A_i$ is ranked first among the first $i$ candidates, and not ranked first among the last $n-i+1$ candidates are independent we have
-$$Pr(X_i) = \frac{n-i}{i(n-i+1)}.$$
+Let $k \in \Z$ such that $0\leq k \leq d$.
+We will show the existence of an injective map $f$ from the faces of $P$ of dimension $k$ to the faces of $Q$ of dimension at least $k$.
+This suffices to show the desired result.
 \paragraph{}
-Thus our expected total penalty is
-$$\sum_{i=1}^n C\cdot Pr(X_i) = C \sum_{i=1}^n\frac{n-i}{i(n-i+1)} = CS_n$$
-where $S_n$ denotes $\sum_{i=1}^n\frac{n-i}{i(n-i+1)}$. Now observe that
-$$S_n = n\sum_{i=1}^n\frac{1}{i(n-i+1)} - \sum_{i=1}^n\frac{1}{n-i+1}= n\sum_{i=1}^n\frac{1}{i(n-i+1)} - H_n =  n\sum_{i=1}^n\frac{1}{i(n-i+1)} - \log(n)$$
-where $H_n$ denotes the $n$-th harmonic number. Now since
-$$\int_{x=1}^n \frac{1}{x(n-x+1)} dx = \frac{2\log(n)}{n+1}$$
-we have that
-$$S_n \leq \frac{2n\log(n)}{n+1} -\log(n) = \frac{(n-1)\log(n)}{n+1}.$$
-Therefore our expected total payoff is at most
-$$C\cdot \frac{(n-1)\log(n)}{n+1}. \blacksquare$$
+Let $F$ be a face of $P$ of dimension $k$ such that $F = \{x \in P: c^Tx = \beta\}$. We define the map $f$ so that
+$$f(F) = \{x \in Q: (c^TA)x = \beta\}.$$
+We need to show that $f$ is well-defined, and that it is injective. To this end we begin by showing that $f(F)$ is a face of $Q$. Let $x \in Q$. Then
+$$(c^TA)x = c^T(Ax) \leq \beta$$
+with the last inequality following since $p(x) = Ax$ is in $P$.
+Note that equality holds for all $x$ such that $Ax$ is in $F$.
+Hence $f(F)$ is a face of $Q$.
+Since $(Q,p)$ is an extension of $P$, for all $y \in F$ there exists $x \in Q$ such that $Ax = y$.
+Thus $F \subseteq p(f(F))$, and using that $p$ is a linear map we see that,
+$$\text{dim}(f(F)) \geq \dim F = k.$$
+\paragraph{}
+Now suppose that $F$ and $F'$ are $k$-dimensional faces of $P$ such that $f(F) = f(F')$.
+Suppose $F = \{x \in P: c^Tx = \beta\}$ and $F' = \{x \in P: (c')^Tx = \beta\}$.
+Let $y \in F$.
+Then there exists $x \in f(F)$ such that $Ax = y$.
+Since $f(F) = f(F')$,
+$$\beta = (c')^TAx = (c')^Ty $$
+and hence $y \in F$. Thus $F \subseteq F'$ and by a symmetric argument we can see that $F' \subseteq F$. Thus $F = F'$ and hence $f$ is an injective map, as desired. $\blacksquare$
 %Problem 2
-\newpage
-\section{}
-Let $G$ be an undirected graph, and $C$ the size of a minimum cut in $G$. Let $\alpha \in \R_{\geq 1}$.
-\begin{lemma}\label{lemma:2}
-Let $F \subseteq E(G)$ be a cut such that $|F| \leq \alpha C$. Then the probability that $F$ is output by Karger's random contraction algorithm is at least
-$$\Omega({\frac{1}{n^{2\alpha}}}).$$
-\end{lemma}
-\begin{proof}
-We say the algorithm ``succeeds" if $F$ is output by the algorithm.
-The algorithm succeeds if and only if we never contract an edge of $F$.
-At each iteration of the algorithm, every subset of vertices in the contracted graph corresponds to a subset of vertices  in the original graph (by expanding contracted vertices), and hence the minimum cut value of contracted graph at iteration $i$ is at least $C$.
-\paragraph{}
-In particular the degree of each vertex in the contracted graph is at least $C$. So the number of edges in the $i$-th iteration is at least
-$$\frac{(n-i+1)C}{2}.$$
-So when we pick a random edge to contract in this iteration, the probability that we pick an edge of $F$ is at most
-$$\frac{|F|}{\frac{(n-i+1)C}{2}} \leq \frac{2\alpha}{n-i+1}$$
-using that $|F| \leq \alpha C.$ So the probability that the algorithm succeeds is at least
-$$\Pi_{i=1}^n(1-\frac{2\alpha}{n-i+1}) = \frac{2\alpha!}{n(n-1)\dots(n-2\alpha + 1)} = \Omega(\frac{1}{n^{2\alpha}})$$
-\end{proof}
-\begin{claim}\label{claim2}
-The number of cuts in $G$ with at most $\alpha C$ edges is at most $O(n^{2\alpha})$.
-\end{claim}
-\begin{proof}
-Let $\cF$ be the set of cuts in $G$ with at most $\alpha C$ edges. For $F \in \cF$ let $X_F$ be the event that Karger's algorithm outputs $F$. By Lemma \ref{lemma:2}
-$$Pr(X_F) \geq \Omega(\frac{1}{n^{2\alpha}}).$$
-Now we see that, one the one hand
-$$Pr(\bigcup_{F \in \cF} X_F) \leq 1$$
-and on the other hand, as the events events $X_F$ are all disjoint (the algorithm only outputs one answer),
-$$Pr(\bigcup_{F \in \cF} X_F) = \sum_{F \in \cF} Pr(X_F) \leq \sum_{F \in \cF} \Omega(\frac{1}{n^{2\alpha}}) = |\cF|\Omega(\frac{1}{n^{2\alpha}}).$$
-Combining inequality and rearranging for $|\cF|$ we have
-$$|\cF| \leq O(n^{2\alpha})$$
-as desired.
-\end{proof}
-\paragraph{}
-If we repeat Karger's algorithm $t$ times waiting for a particular cut $F \in \cF$ (defined in above proof) we have a failure probability at most
-$$(1-\Omega(\frac{1}{n^{2\alpha}}))^t \leq \exp(\frac{t}{O(n^{2{\alpha}})})$$
-and hence we can make this failure probability small with a choice of $t \in O(n^{2\alpha})$ (the constant would depend on how good you want the success probability).
-\paragraph{}
-So we give the following randomized algorithm for enumerating all cuts of $G$ with at most $\alpha C$ edges. Let $k$ be some parameter to be chosen later. Our algorithm proceeds as follows:
-\begin{enumerate}
-\item Let $\cL = \emptyset$.
-\item While fail$<k$
-	\begin{enumerate}
-	\item Repeat Karger's algorithm at most $t$ times hoping for a cut $F\in \cF\backslash \cL$.
-	\item If such $F$ is found, add $F$ to $\cL$ and set fail$=0$.
-	\item Otherwise set fail$=$fail$+1$.
-	\end{enumerate}
-\item Output $\cL$.
-\end{enumerate}
-\paragraph{}
-In words the algorithm uses the boosted version of Karger's to try to find a cut with at most $\alpha C$ edges that it has not seen before.
-It stops if it fails to find such a cut $k$ times in a row, and prints all the cuts found this way.
-\paragraph{}
-The algorithm fails if in some $i$-th round it does not find a cut $F \in \cF\backslash \cL$ $k$ times in a row, while $\cF \backslash \cL\neq \emptyset$.
-We know from \ref{claim2} there are at most $O(n^{2\alpha})$ such rounds $i$.
-Let $X_i$ be a random variable indicating failure in the $i$-th round.
-We can choose $t$ so that $t$ repetitions of Karger's algorithm finds $F \in \cF\backslash \cL$ (provided such $F$ exists) with high probability, say $0.99$.
-\paragraph{}
-Then the probability that we fail in the $i$-th round is
-$$(0.01)^k.$$
-So the probability we fail in some round is at most
-$$\frac{O(n^{2\alpha})}{100^k}.$$
-Since $\alpha$ is constant, we can choose $k \in O(\log n)$ to make this failure probability arbitrarily small.
-Let $T$ be the time to run Karger's algorithm once.
-Since there are $O(n^{2\alpha})$ rounds our algorithm takes
-$$O(n^{2\alpha}ktT) = O(n^{4\alpha}\log n T)$$
-time.$ \blacksquare$
-
-%Problem 3
-\newpage
 \section{}
 \paragraph{}
-We throw $n^{0.99}$ balls into $n$ bins uniformly at random. Using union bound, the probability that a bin has at least $k$ balls, for $k$ chosen later, is at most 
-$${n \choose k}(\frac{1}{n})^k.$$
-Using the bound ${n^{0.99}\choose k} <(\frac{n^{0.99}e}{k})^k$ this probability is at most
-$$(\frac{n^{0.99}e}{k})^k(\frac{1}{n})^k= \frac{e^k}{n^{0.0k}k^k}\leq \frac{e^k}{k^k}$$
-where the last inequality follows since we will choose $k \geq 0$ and since $n\geq 1$ this implies $n^\frac{k}{100} \geq 1$.
-Again invoking union bound
-$$Pr(\exists\text{a bin with at least $k$ balls}) \leq n\frac{e^k}{k^k} = n(\frac{e}{k})^k.$$ 
-Since we want
-$$Pr(\exists\text{a bin with at least $k$ balls}) \leq \frac{1}{2}$$
-we need to find $k$ such that
-$$n(\frac{e}{k})^k \leq \frac{1}{2}.$$
-
-%Problem 4
-\newpage
+The claim is \emph{false}.
+Let $P = \conv{\{(0,0),(1,0),(1,1),(0,1)\}}$.
+Then $P$ is a square, so $|\text{facets}(P)| = 4$.
+Let $Q = \conv{\{(0,0,0),(1,0,0), (1,1,0), (0,1,1)\}}$, and let $p: \R^3\rightarrow \R^2$ be the projection map onto the first two coordinates (thus $p$ is linear).
+Then $Q$ is a triangular pyramid with the ``top" vertex skewed.
+so $|\text{facets}(Q)| = 4$.
+Clearly $p(Q) = P$ and thus $(Q,p)$ is an extension of $P$ with the same number of facets.
+But $p$ not a linear isomorphism.
+Observe that $\frac{1}{3}(1,0,0) + \frac{1}{3}(0,1,1) + \frac{1}{3}(0,0,0) = \frac{1}{3}(1,1,1) \in Q$ and $ \frac{1}{3}(1,1,0) + \frac{2}{3}(0,0,0) = \frac{1}{3}(1,1,0) \in Q$ have the same projection:
+$$p(\frac{1}{3}(1,1,1)) = p(\frac{1}{3}(1,1,0)) = \frac{1}{3}(1,1) \in P$$
+and so $p$ is not injective and thus not an isomorphism.$\blacksquare$
+%Problem 3
 \section{}
 \paragraph{}
-We always use addition, $+$, over the appropriate field (ie.
-$\F_n$ for inputs and $\F_m$ for outputs).
-Our algorithm is very simple.
-Given input $z$ choose $x \in \{0,\dots, n-1\}$ uniformly at random.
-Let $y = z-x$.
-Lookup $F(x)$ and $F(y)$ and return $F(x) + F(y)$.
+Let $Q \subseteq \R^n$ be a polyhedron and $p:\R^n \rightarrow \R^d$ a linear map, so that $(Q,p)$ is an extension of a polytope $P \subseteq \R^d$.
+Let $V$ be the set of vertices of $Q$ and $R$ the set of extreme rays of $Q$.
+That is
+$$ Q = \conv{V} + \cone{R}.$$
+If $R = \emptyset$ we are done since $Q$ is itself a polytope.
+So suppose $R\neq \emptyset$.
+We claim that $p(r) = 0$ for all $r \in R$.
+Suppose not, that is there exists $r \in R$ for which $p(r) \neq 0$.
 \paragraph{}
-If both $F(x)$ and $F(y)$ are not modified by our adversary then our return value satisfies
-$$F(x) + F(y) = F(x+y) = F(z)$$
-using the linearity of $F$ and the fact that $x+y = z$.
-So a necessary condition for our algorithm to fail is that one of $F(x)$ or $F(y)$ was modified by our adversary.
-For a uniform random element of the table, the probability of modification is at most $\frac{1}{5}$.
-Hence
-$$Pr(F(x)\text{ or } F(y)\text{ was modified}) \leq Pr(F(x) \text{ was modified}) + Pr(F(y)\text{ was modified}) \leq \frac{2}{5}.$$
-Therefore the probability of failure is at most
-$$\frac{2}{5} < \frac{1}{2}.$$
-Hence our algorithm succeeds with probability at least $\frac{1}{2}$ as desired. $\blacksquare$
+Let $v \in V$ if $V\neq \emptyset$ and choose $v = 0$ otherwise. Now, for all $\lambda \geq 0$ $v + \lambda r$ is in $Q$. But then observe that
+$$p(v + \lambda r) = p(v) + \lambda p(r)$$
+is in $P$. Since $p(r) \neq 0$ this implies that $p(r)$ is an unbounded direction in $P$, contradicting that $P$ is a polytope (and hence bounded). 
+%Problem 4
+\section{}
+\paragraph{}
 
 %Problem 5
-\newpage
 \section{}
 \paragraph{}
-Let $S_1$ and $S_2$ be multisets of positive integers.
-Let $|S_1| = |S_2| = n$ be the number of elements in each multiset.
-If the sets are of different sizes it is trivial to decide they are not equal.
-Let $h:[n] \rightarrow [cn]$ be a $2$-universal hash function chosen uniformly at random.
-\paragraph{}
-We're going to define multiset operations.
-They are exactly what you'd expect them to be, so feel free to skip this paragraph as I only add them for sake of completeness.
-Let $S_1\cap S_2$ denote the largest multiset that is a sub-multiset of $S_1$ and a sub-multiset of $S_2$.
-By sub-multiset I mean $A$ is a sub-multiset of $B$ ($A\subseteq B$) if and only if for every element $e$ of $A$, the number of times $e$ appears in $A$ is at most the number of times $e$ appears in $B$.
-We define multiset addition so that $A+B$ is the multiset of elements $e$ so that $e$ appears as many times as it does in $A$ plus as many times as it does in $B$.
-We also need multiset difference.
-For multisets $A$ and $B$ we define $A-B$ to be the unique multiset so to that $(A-B) + (A\cap B) = A$.
-My multiset equality, $A=B$, we meant that $A$ and $B$ have precisely the same elements appearing the same number of times.
-\paragraph{}
-If $S_1 = S_2$ then $h(S_1) = h(S_2)$ and so the algorithm never gives false negatives.
-Observe that $h(S_1) = h(S_1\cap S_2) + h(S_1-S_2)$ and $h(S_2) = h(S_1\cap S_2) + h(S_2-S_1)$.
-Hence the algorithm can only give a false positive if $h(S_1-S_2) = h(S_2 - S_1)$.
-We will bound the probability of failure by the probability of this happening.
-Let $d = |S_1-S_2| =|S_2-S_1|$.
-For $a \in S_1 - S_2$ and $b \in S_2 - S_1$ define $X_{a,b}$ to be the event that $h(a) = h(b)$.
-Further for each $a \in S_1 - S_2$ define
-$$X_a  = \bigcup_{b \in S_2 - S_1} X_{a,b}.$$
-Notice that $X_a$ is the disjoint union of events $X_{a,b}$. Let $X$ be a random variable counting the number of collisions between elements in $S_1-S_2$ and elements in $S_2-S_1$. We have
-$$E[X] = \sum_{a \in S_1 - S_2} E[X_a] = \sum_{a\in S_1 - S_2} \sum_{b \in S_2-S_1} Pr(h(a) = h(b)) \leq  \sum_{a\in S_1 - S_2} \sum_{b \in S_2-S_1} \frac{1}{cn} = \frac{d^2}{cn}.$$
-The inequality follows from $h$ being $2$-universal.
-For the algorithm to fail it is necessary to have $X \geq d$, for otherwise there are an insufficient number of collisions to have $h(S_1-S_2) = h(S_2-S_1)$.
-Therefore we can bound the probability of failure by $Pr(X  \geq d)$.
-By Markov's inequality we see that,
-$$Pr(X \geq d) \leq \frac{E[X]}{d} \leq \frac{d}{cn}.$$
-Hence our probability of failure is at most $\frac{d}{cn}$.
-Assuming we can hash in constant time and lookup in a hash table in constant time, the algorithm takes $O(n)$ time to hash each element in $S_1$ and $S_2$, then takes $O(cn)$ time to verify the $cn$ counters each associated with an entry of each of the two hash tables.
-Since we need $c \geq 1$ to avoid guaranteed collisions, this implies the algorithm runs in $O(cn)$ time with failure probability at most $\frac{d}{cn}$.
-\paragraph{}
-In the previous paragraph we demonstrated a trade-off between running time and failure probability with respect to choice of parameter $c$.
-To obtain a Monte Carlo algorithm, we need a failure probability which is independent of the input size.
-We make our extension by giving an explicit choice of $c$.
-Let $p \in (0,1)$.
-Then if we choose $c = \ceil{\frac{d}{pn}}$ then we obtain a failure probability at most $p$ and a running time $O(pd)$.
-$\blacksquare$
+
 %Problem 6
-\newpage
 \section{}
 \paragraph{}
-Let $\cA$ be an algorithm which obtain an $(\epsilon, \frac{1}{4})$-approximation to some parameter $\mu$. That is, when $X$ is the output value of $\cA$,
-$$Pr(|X - \mu| \leq \epsilon\mu) \leq \frac{1}{4}.$$
-We want to show how to obtain an $(\epsilon,\delta)$-approximation to $\mu$.
-Suppose we perform $k = \ceil{\log(\frac{1}{\delta})}$ independent runs of $\cA$ and output the median.
-We may assume $k$ is even, if not replace $k$ with $k+1$.
-Let $X_1, \dots, X_k$ be random variables each denoting the output value of one run of $\cA$.
-Instead of ordering them by the order the trials happen, order them sorted by value, i.e.
-$$X_1 \leq X_2 \leq \dots\leq X_k.$$
-Our $(\epsilon,\delta)$-approximation algorithm outputs $X_\frac{k}{2}$, the median value of our independent trials.
-Let $X$ denote the output of our new algorithm.
-We observe that $|X - \mu|\leq \epsilon \mu$ if and only if
-$$|X_i - \mu | \leq \epsilon \mu, \quad \forall i=1 \dots \frac{k}{2}.$$
-Therefore
-$$Pr(|X-\mu| \leq \epsilon \mu) = \Pi_{i=1}^\frac{k}{2} Pr(|X_i - \mu| \leq \epsilon \mu) \leq \frac{1}{4^\frac{k}{2}}.$$
-Since $k = \ceil{\log(\frac{1}{\delta})}$,
-$$\frac{1}{4^\frac{k}{2}} \leq \frac{1}{4^{\frac{1}{2}\log(\frac{1}{\delta})}} = \delta.$$
-Therefore
-$$Pr(|X-\mu| \leq \epsilon \mu) \leq \delta. \blacksquare$$
-%Problem 7
-\newpage
-\section{}
-\paragraph{}
-Choose $c = 1$.
-Let $U = \{u_1,\dots, u_n\}$ be a set of ground elements, and let $S_1, \dots, S_m \subseteq U$ be a set of subsets of $U$ such that $|S_i| = \frac{n}{2}$ for all $i\in [m]$.
-Consider a uniform random coloring of the elements of $U$ either red or blue.
-For each $i \in [n]$ let $R_i$ be an indicator random variable for the event that $u_i$ is colored red, and similarly let $B_i$ be an indicator random variable for the event that $u_i$ is colored blue.
-For each $j \in [m]$ let $$X^j := \sum_{u_i \in S_j} R_i - \sum_{u_i \in S_j} B_i.$$
-Then showing $S_j$ has between $\frac{n}{4} - c\sqrt{n\log m}$ and $\frac{n}{4} +c\sqrt{n\log m}$ red elements is equivalent to showing $$|X^j| \leq 2c\sqrt{n\log m}.$$
-\paragraph{}
-For each $i \in [n]$ let $X_i$ be a random variable such that $X_i = 1$ if $u_i$ is red and $X_i = -1$ if $u_i$ is  blue.
-Since each $S_j$ has $\frac{n}{2}$ elements, each $X^j$ is the sum of exactly $\frac{n}{2}$ such $X_i$.
-Using Corollary $4.8$ in Probability and Computing by Mitzenmacher and Upfal we have
-\begin{align*}
-Pr(|X^j| \geq 2c\sqrt{n\log m}) &\leq 2e^\frac{-(2c\sqrt{n\log m})^2}{2\frac{n}{2}} \\
-&= 2e^{-4c^2\log m}\\
-&=\frac{2}{m^4}\\
-&< \frac{1}{m} &\text{ (since $m \geq 2$, otherwise the problem is trivial)}.
-\end{align*}
-Then, applying a union bound,  the probability that there exists $j\in[m]$ such that $|X^j| \geq 2c\sqrt{n\log m}$ is strictly less than $1$.
-Thus with positive probability, for all $j\in [m]$ $$|X^j| \leq 2c\sqrt{n\log m}$$
-as desired.$\blacksquare$
-%Problem 8
-\newpage
-\section{}
-\subsection{a}
-\paragraph{}
-Suppose $s$ is a random $n$-bit string.
-Let $k \in [n]$ be a positive integer.
-Observe that a $k$-consecutive substring in $s$ is uniquely determined by the starting position of the substring.
-Therefore there are at most $n$ potential $k$-consecutive substrings in $s$.
-Denote by $s(i)$ the substring $s_is_{i+1}\dots s_{i+k-1}$.
-For $s(i)$ to be $k$-consecutive, either bits of $s(i)$ are $0$ or all $1$.
-Thus,
-$$Pr(s(i) \text{ is $k$-consecutive}) = \frac{2}{2^k} = \frac{1}{2^{k-1}}.$$
-So by applying union bound, using there are at most $n$ such $s(i)$, the probability there exists a $k$-consecutive substring is at most
-$$\frac{n}{2^{k-1}}.$$
-To ensure there are no $k$-consecutive substrings with high probability we desire
-$$\frac{n}{2^{k-1}} < 0.01$$
-Hence we choose
-$$k = \log(100n) +1 = O(\log n)$$
-to achieve the desired result.$\blacksquare$
-\subsection{b}
-\paragraph{}
-We can model this problem via a random walk.
-Our states will be $S_0, S_1, \dots, S_k$.
-$S_0$ transitions to $S_1$ with probability $1$.
-For $i \in \{1,\dots, k-1\}$, $S_i$ transitions to $S_{i+1}$ with probability $\frac{1}{2}$, and transitions to $S_1$ with probability $\frac{1}{2}$.
-$S_k$ transitions to $S_k$ with probability $1$.
-To interpret the original problem as this random walk imagine starting in $S_0$ and reading the random bits of $s$.
-After reading $1$ bit, $b$, you move from $S_0$ to $S_1$, then as long as you keep seeing the same bit value as $b$ when you read the next bit of $s$ you advance from your current $S_i$ to $S_{i+1}$.
-But once you see the opposite value to $b$ you return to $S_1$, and treat $\bar{b}$ as the value you're looking for to advance.
-If at any time you reach $S_k$, you stop, having just read a $k$-consecutive substring of $s$.
-\paragraph{}
-We want to bound the expected time to reach $S_k$ from $S_0$.
-For each $i = 0, \dots k$ let $h_i$ denote the expected number of steps to reach $S_k$.
-Clearly, $h_0 = h_1 + 1$ and $h_k = 0$.
-For $j \in [k-1]$ we develop the following recurrence:
-$$h_{k-j} = \frac{1}{2}(h_{k-j+1} + 1) + \frac{1}{2}h_1.$$
-We claim that for all $j \in [k-1]$, $h_{k-j} = \sum_{i=1}^j \frac{1}{2^i}(1+h_1)$. We proceed by induction. In the base case $j=1$. From our recurrence
-$$h_{k-1} = \frac{1}{2}(h_k+1) + \frac{1}{2}h_1 = \frac{1}{2}(1+h_1),$$
-using $h_k = 0$. So the base case holds. Now consider $j \in [k-1]\backslash\{1\}$ and suppose the claim holds for $h_{k-j+1}$. Then we have
-\begin{align*}
-h_{k-j} &= \frac{1}{2}(h_{k-j+1} + 1) + \frac{1}{2}h_1\\
-&= \frac{1}{2}( \sum_{i=1}^{j-1} \frac{1}{2^i}(1+h_1) + 1) + \frac{1}{2}h_1 \\
-&=\sum_{i=1}^{j-1}\frac{1}{2^{i+1}}(1+h_1) + \frac{1}{2} + \frac{1}{2}h_1\\
-&= \sum_{i=2}^{j}\frac{1}{2^{i}}(1+h_1) + \frac{1}{2} + \frac{1}{2}h_1 \\
-&= \sum_{i=1}^j \frac{1}{2^i}(1+h_1).
-\end{align*}
-Therefore the claim holds by induction.
-\paragraph{}
-So, using partial sums of the geometric series, we have that 
-$$h_1 = \sum_{i=1}^{k-1} \frac{1}{2^i}(1+h_1) = \frac{2^{k-1}-1}{2^{k-1}}(1+h_1).$$
-Rearranging we see that
-$$h_1 = 2^{k-1} - 1$$
-which implies that
-$$h_0 = 2^{k-1}.$$
-Let $X$ denote the number of bits needed before seeing a $k$-consecutive substring. By Markov's inequality, our equation for $h_0$ implies
-$$Pr(X \geq 100\cdot2^{k-1}) \leq 0.01.$$
-Thus with high probability, $0.99$ we find a $k$-consecutive substring with at most $100 \cdot 2^{k-1}$ bits.
-That is to say, setting $n = 100\cdot 2^{k-1}$ ensures a $k$-consecutive substring with high probability.
-Rearranging for $k$, we see that for
-$$k =\frac{n}{100}\log(n) + 1$$
-$s$ has a $k$-consecutive substring with high probability. $\blacksquare$
-%Problem 9
-\newpage
-\section{}
-
-%Problem 10
-\newpage
-\section{}
-\paragraph{}
-Let $G$ be a connected, undirected, non-bipartite graph. Consider a Markov Chain $M$ with state space $V(G) \times V(G)$ and transition probabilties:
-$$P_{ab, uv} = \begin{cases}
-\frac{1}{d(a)d(b)}, &\text{ if } u \in N(a)\text{ and } v \in N(b) \\
-0, &\text{otherwise}
-\end{cases}$$
-Then a state in $M$ represents a position $(a,b)$ where $a$ is the position of the cat and $b$ is the position of the mouse, and $P_{ab,uv}$ represents the probability of the cat being in position $u$ and the mouse being in position $v$ in the subsequent timestep.
-We can equivalently view $M$ as the Markov chain for a random walk on the graph $G'$ where $V(G') = V(G) \times V(G)$ and $\{(a,b), (u,v)\} \in E(G')$ if and only if $u \in N(a)$ and $v \in N(b)$.
-\paragraph{}
-Hence $|V(G')| = |V(G)|^2$ and 
-$$2|E(G')| = \sum_{u,v \in V(G)} d(u)d(v) = 4|E(G)|^2$$
-i.e.
-$|E(G')| = 2|E(G)|^2$.
-Note the cat catches the mouse when the random walk on $G'$ reaches vertex $(a,a)$ for some $a \in V(G)$.
-So if we upper bound $h_{x',y'}$ (the hitting time), for any $x', y' \in V(G')$ then we have successfully upper bounded the expected time for the cat to catch the mouse since it is simply $h_{x',y'}$ where $ux$ is the start state of the cat and mouse, and $y' = (a,a)$ for some $a \in V(G)$.
-\paragraph{}
-Using Lemma $7.15$ of Probability and Computing we see that 
-$$h_{a',b'} < 2|E(G')|$$
-provided that $a'b' \in E(G')$.
-Now consider any $x'=(a,b), y'=(u,v) \in V(G')$.
-Let $P_1$ be an $au$-path in $G$, which exists since $G$ is connected.
-Similarly let $P_2$ be a $bv$-path in $G$.
-Construct an $x'y'$-path $P$ in $G'$ by taking the sequence of vertices $(a,b) \dots (u,b)$ following $P_1$ in the first position and keeping $b$ fixed in the second position, then taking the sequence of vertices $(u,b) \dots (u,v)$ following $P_2$ in the second position and keeping $u$ fixed in the first.
-Then $|P_2|  = |P_1| + |P_2| \leq 2|V(G)|$.
-\paragraph{}
-As $P$ is an $x'y'$-path, $h_{x',y'}$ is upper bounded by the sum of hitting times on the edges of $P$. That is, we have
-$$h_{x',y'} \leq \sum_{a'b' \in E(P)} h_{a'b'} <  \sum_{a'b' \in E(P)} 2|E(G')| \leq 4|V(G)||E(G')| \leq 8|V(G)||E(G)|^2$$
-with the second inequality following from Lemma $7.15$. Thus we have shown an upper bound of $O(|V(G)|\cdot|E(G)|^2)$ on the expected time for the cat to catch the mouse.$\blacksquare$
 \end{document}

### 2018-10-04 CO759-ExtendendFormulations/assignments/A1.tex
Problem 3

@@
 Let $Q \subseteq \R^n$ be a polyhedron and $p:\R^n \rightarrow \R^d$ a linear map, so that $(Q,p)$ is an extension of a polytope $P \subseteq \R^d$.
-Let $V$ be the set of vertices of $Q$ and $R$ the set of extreme rays of $Q$.
-That is
-$$ Q = \conv{V} + \cone{R}.$$
-If $R = \emptyset$ we are done since $Q$ is itself a polytope.
-So suppose $R\neq \emptyset$.
-We claim that $p(r) = 0$ for all $r \in R$.
-Suppose not, that is there exists $r \in R$ for which $p(r) \neq 0$.
+If $Q$ is bounded, we are done, so we may assume that $Q$ is unbounded.
+Let $R$ be the recession cone of $Q$.
+We claim that $p(R) = 0$, for otherwise if $r \in R$ with $p(r) \neq 0$ then $p(r)$ is an unbounded direction in $P$ as $p(\lambda r) = \lambda p(r) \neq 0$ for all $\lambda  > 0$, contradicting that $P$ is a polytope.
+Let $L = R \cap (-R)$ be the lineality space of $Q$.
+By our previous claim $p(L) = 0$.
+We have that
+$$Q = L + (Q\cap L^\perp)$$
+where $L^\perp$ denotes the orthogonal complement of $L$. Thus
+$$P = p(Q) = p(L) + p(Q\cap L^\perp) = p(Q\cap L^\perp).$$
+Let $Q_1$ denote $Q\cap L^\perp$, so we have shown $p(Q_1) = P$, that is $(Q_1, p)$ is an extension of $p$.
+Since $Q_1$ was obtained from $Q$ by adding only equalities, $Q_1$ has no more facets than $Q$.
+Our choice of $Q_1$ ensures that $Q_1$ has a trivial lineality space, i.e.
+$Q_1$ is pointed.
+If $Q_1$ is bounded, we are done, so we may assume that $Q_1$ is unbounded.
+Let $R_1$ be the recession cone of $Q_1$ and $V_1$ be the vertices of $Q_1$.
+By our prior arguments neither $R_1$ nor $V_1$ are empty.
 \paragraph{}
-Let $v \in V$ if $V\neq \emptyset$ and choose $v = 0$ otherwise. Now, for all $\lambda \geq 0$ $v + \lambda r$ is in $Q$. But then observe that
-$$p(v + \lambda r) = p(v) + \lambda p(r)$$
-is in $P$. Since $p(r) \neq 0$ this implies that $p(r)$ is an unbounded direction in $P$, contradicting that $P$ is a polytope (and hence bounded). 
-%Problem 4
+Let $c \in \R^n$ be a vector such that $c^Tr \geq 1$ for all $r \in R_1$.
+Such a $c$ exists since $R_1 \neq \emptyset$ and the lineality space of $Q_1$ is empty (clearly $c$ so that $c^Tr >0$ exists, and we simply scale $c$).
+Let $\beta \in \R$ be chosen so that $c^Tv < \beta$ for all $v \in V_1$.
+Since $V_1$ is finite, such $\beta$ exists.
+We choose $Q_2$ as
+$$Q_2 = Q_1 \cap \{x \in \R^n : c^Tx = \beta\}.$$
+We claim that $Q_2$ is bounded and that $p(Q_2) = P$.
+Let $\alpha = \min\{c^Tv : v \in V_1\}$.
+Let $x \in Q_2$.
+There exist $\lambda \geq 0$, $\sum_{v \in V_1} \lambda_v = 1$ and $\gamma \geq 0$ such that
+$$x = \sum_{v \in V_1} \lambda_v v + \sum_{r \in R_1} \gamma_r r_1.$$
+So
+$$\beta = c^Tx =  \sum_{v \in V_1} \lambda_v c^Tv + \sum_{r \in R_1} \gamma_r c^Tr_1 \geq \alpha + \sum_{r \in R_1} \gamma_r c^Tr_1\geq \alpha + \sum_{r \in R_1} \gamma_r$$
+and thus $$\sum_{r \in R_1} \gamma_r \leq \beta -\alpha.$$
+Hence our choice of coefficients $\gamma$ on the vectors of $R_1$ is bounded, and so $Q_2$ is bounded.
+\paragraph{}
+It remains to verify that $p(Q_2) = P$.
+Similarly to the argument for $R$, we see that $p(R_1) = 0$.
+Hence $P = p(Q_1) = p(\conv{V_1}).$ So it will suffice to show that for every $x \in \conv{V_1}$ there exists $\lambda \geq 0$ and $r \in R_1$ such that $c^T(x + \lambda r) = \beta.$ Let $r$ be any vector in $R_1$ (since $R_1 \neq \emptyset$ this choice is well-defined).
+Choose $\lambda = (\beta - c^Tx)/c^Tr$.
+Since $\beta-c^Tx >0$ and $c^Tr >0$ by our choice of $\beta$ and $c$, we have $\lambda \geq 0$.
+Now we have
+$$c^T(x + \lambda r) = c^Tx + \beta - c^Tx = \beta$$
+as desired. $\blacksquare$
 \section{}

### 2018-10-04 CO759-ExtendendFormulations/assignments/A1.tex
starting p4

@@
 as desired. $\blacksquare$
+
+%Problem 4
 \section{}
 \paragraph{}
+Let $P \subseteq \R^d$ be a full-dimensional polytope with $0_d$ in its interior.
+\subsection*{Sufficiency}
+\paragraph{}
+Suppose that $Q\subseteq \R^n$ is a polyhedron with at most $f$ facets, and $p:\R^n \rightarrow \R^d$ is an affine map so that $(Q,p)$ defines an extension of $P$.
+We define the \emph{polar} of $P$ as follows
+$$P^* := \{x \in \R^d : y^Tx \leq 1, \forall y \in P\}.$$
+From the previous Problem we may assume that $Q$ is a polytope.
+Suppose that there is a matrix $A \in \R^{d\times n}$ and $b \in \R^d$ such that $p(x) = Ax + b$ for all $x \in Q$.
+We claim that we may assume $b = 0$.
+Since $P$ is full-dimensional, $A$ as full row rank $d$.
+Thus $A$ has right inverse $A^{-1}_R$.
+Let $Q_1 = Q + \{-A^{-1}_Rb\}$.
+Then for all $x \in Q$,
+$A(x + A^{-1}_Rb) = Ax  + b  = p(x).$ So the linear map which sends $x$ to $Ax$ maps $Q_1$ to $P$.
+Since $Q_1$ is a translation of $Q$ it has the same number of facets as $Q$.
+Thus we may assume $p$ is a linear map, i.e.
+$b=0$.
+\paragraph{}
+Let $Q' = Q^*$ and let $\cL' = \text{rowspace}(A)$.
+We claim $P^*$ is linearly isomorphic to $Q' \cap \cL'$.
+Since $Q'$ is the polar dual of $Q$, the number of vertices of $Q'$ is equal to the number of facets of $Q$: $f$.
+\subsection*{Necessity}
+\paragraph{}
+Suppose that $Q'\subseteq \R^{n'}$ is a polytope with at most $f$ vertices, and $\cL \subseteq R^{n'}$ is an affine space such that $P^*$ is linearly isomorphic to $Q' \cap \cL'$.
 

### 2018-10-05 CO759-ExtendendFormulations/assignments/A1.tex
Problem 4

--- 2018-10-04 CO759-ExtendendFormulations/assignments/A1.tex
@@
 as desired. $\blacksquare$
+
+%Problem 4
 \section{}
 \paragraph{}
-
+Let $P \subseteq \R^d$ be a full-dimensional polytope with $0_d$ in its interior.
+\subsection*{Sufficiency}
+\paragraph{}
+Suppose that $Q\subseteq \R^n$ is a polyhedron with at most $f$ facets, and $p:\R^n \rightarrow \R^d$ is an affine map so that $(Q,p)$ defines an extension of $P$.
+We define the \emph{polar} of $P$ as follows
+$$P^* := \{x \in \R^d : y^Tx \leq 1, \forall y \in P\}.$$
+From the previous Problem we may assume that $Q$ is a polytope.
+Suppose that there is a matrix $A \in \R^{d\times n}$ and $b \in \R^d$ such that $p(x) = Ax + b$ for all $x \in Q$.
+We claim that we may assume $b = 0$.
+Since $P$ is full-dimensional, $A$ has full row rank $d$.
+Thus $A$ has right inverse $A^{-1}_R$.
+Let $Q_1 = Q + \{-A^{-1}_Rb\}$.
+Then for all $x \in Q$,
+$A(x + A^{-1}_Rb) = Ax  + b  = p(x).$ So the linear map which sends $x$ to $Ax$ maps $Q_1$ to $P$.
+Since $Q_1$ is a translation of $Q$ it has the same number of facets as $Q$.
+Thus we may assume $p$ is a linear map, i.e.
+$b=0$.
+\paragraph{}
+Let $Q' = Q^*$ and let $\cL' = \text{rowspace}(A)$.
+Since $Q'$ is the polar dual of $Q$, the number of vertices of $Q'$ is equal to the number of facets of $Q$: $f$.
+We claim $P^*$ is linearly isomorphic to $Q' \cap \cL'$.
+Let $c \in P^*$.
+Then $A^Tc \in \text{rowspace}(A) = \cL'$.
+Further for any $x \in Q$,
+$$ A^Tcx = c^TAx \leq 1$$
+since $Ax$ is in $P$ and $c \in P^*$.
+Thus $A^Tc \in Q' \cap \cL'$.
+On the other hand, let $g \in Q' \cap \cL'$.
+Since $g \in \cL'$ there exists $c \in \R^d$ such that $g = A^Tc$.
+We claim $c$ is in $P^*$ showing $A^T$ is a surjective map from $P^*$ to $Q' \cap \cL'$.
+Let $y \in P$.
+Then there exists $x \in Q$ such that $Ax = y$.
+We have, using that $g \in Q^*$,
+$$1 \geq g^T x = c^TAx = c^Ty.$$
+Thus $c \in P^*$.
+It remains to show the map $A^T$ is injective.
+Let $c_1,c_2 \in P^*$ such that $A^Tc_1 = A^Tc_2.$ As discussed earlier $A$ has a right inverse $A^{-1}_R$.
+Multiplying $c_1^T A = c_2^T A$ on both sides by $A^{-1}_R$ we have
+$$c_1^TAA^{-1}_R = c_2^TAA^{-1}_R$$
+and thus $c_1^T = c_2^T$. Therefore $A^T$ is an injective map, and so $P^*$ is linearly isomorphic to $Q' \cap \cL'$, as desired.
+\subsection*{Necessity}
+\paragraph{}
+Suppose that $Q'\subseteq \R^{n}$ is a polytope with at most $f$ vertices, and $\cL' \subseteq R^{n}$ is an affine space such that $P^*$ is linearly isomorphic to $Q' \cap \cL'$.
+Let $q: P^* \rightarrow Q'\cap\cL'$ be a linear map defining the isomorphism between $P^*$ and $Q'\cap \cL'$.
+Then $q$ is described by a matrix $A \in \R^{n\times d}$ so that $q(x) = Ax$ for all $x \in P^*$.
+Since $P$ is a polytope, $P$ is bounded.
+That is to say, there exists some $\epsilon > 0$ for which $P$ is contained in the $||\cdot||_2$-norm ball of radius $\epsilon$.
+Then if we let $e_1, \dots, e_d$ be the standard basis of $\R^d$, we see that $e_i^T x \leq \epsilon$ for all $x \in P$.
+Hence for all $i \in [d]$, $\frac{1}{\epsilon}e_i$ is in $P^*$.
+This shows that $P^*$ is full-dimensional, and thus $Q'\cap \cL'$ is full-dimensional.
+Therefore $A$ has full column rank $d$, and so $A$ has a left inverse $A^{-1}_L \in \R^{d\times n}$.
+\paragraph{}
+Let $Q = (Q')^*$, and let $p: \R^n \rightarrow \R^d$ be the linear map defined by $p(x) = A^Tx$ for all $x \in Q$.
+We claim that $(Q,p)$ is an extension of $P$.
+Since $Q$ is the polar of $Q'$, and $Q'$ has at most $f$ vertices, we have that $Q$ has at most $f$ facets.
+Thus the claim suffices to complete the proof.
+\paragraph{}
+We want to show $p(Q) = P$. Since $0$ is contained in the interior of $P$, $(P^*)^* = P$. Let $x \in Q$, and let $y \in P^*$. Then
+$$p(x)^Ty = x^T(Ay) \leq 1$$
+since $Ay$ is in $Q' \cap \cL'$ and $x$ is in the polar of $Q'$. Thus $p(Q) \subseteq (P^*)^* = P$. 
+\paragraph{}
+It remains to verify that every $x \in P$ has $\bar{x}$ in $Q$ for which $p(\bar{x}) = x$. Indeed let $x \in P$ and choose $\bar{x}^T = x^TA^{-1}_L$. Then
+$$p(\bar{x}) = (AA^{-1}_L)^Tx = x$$
+as desired. Finally we need to verify that $\bar{x}$ is in $Q$. Let $y \in Q'$, we have
+$$ \bar{x}^Ty = x^TA^{-1}_L y \leq 1$$
+with the inequality following from $x \in P =(P^*)^*$ and $A^{-1}_L$ being the inverse map to $q$ (and thus $A^{-1}_Ly$ is in $P^*$).
+Therefore we have shown $(Q,p)$ is an extension of $P$ with at most $f$ facets.$\blacksquare$
 %Problem 5

### 2019-01-21 CS860-CommunicationComplexity/assignments/Chapter2.tex
Adding Communication Complexity

\documentclass[11pt]{amsart}
\usepackage{amsmath,amsthm,amsfonts,amssymb}
\usepackage{fullpage}
\usepackage[usenames]{xcolor}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{problem}{Problem}
\newtheorem{challenge}{Challenge Problem}
\newtheorem{open}{Open Problem}
\theoremstyle{plain}

\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calZ}{\mathcal{Z}}
\newcommand{\cost}{\mathrm{cost}}
\newcommand{\disc}{\mathrm{disc}}
\newcommand{\discu}{\mathrm{disc_u}}
\newcommand{\Disj}{\textsc{Disj}}
\newcommand{\Eq}{\textsc{Eq}}
\newcommand{\GT}{\textsc{GT}}
\newcommand{\IP}{\textsc{IP}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}

\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}

\title{Communication complexity: Chapter 2 \\ More lower bound techniques}
\author{Eric Blais and Justin Toth}

\begin{document}

\maketitle

In the first chapter, we introduced the deterministic communication complexity 
model and saw the general \emph{partition bound} method for proving communication complexity lower bounds. While the partition bound is quite strong, it is hard to work with directly. The goal of this chapter is to introduce other general methods for proving communication complexity lower bounds more easily: the \emph{fooling set bound}, the \emph{rectangle size bound}, and the \emph{log rank bound}.

\newpage \section{Fooling set bound}

For many functions, the \emph{fooling set} bound is the easiest way to get meaningful communication complexity lower bounds.

\begin{definition}[Fooling set]
A set $F \subseteq \calX \times \calY$ is a \emph{fooling set} for the function $f : \calX \times \calY \to \{0,1\}$ if there is a value $b \in \{0,1\}$ such that
\begin{enumerate}
\item For every $(x,y) \in F$, $f(x,y) = b$; and
\item For every $(x,y) \neq (x',y') \in F$, we have $f(x,y') \neq b$ or $f(x',y) \neq b$ (or both).
\end{enumerate}
\end{definition}

\begin{lemma}
If $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$ has a fooling set of size $t$ then $\chi(f) \ge t$ and $D(f) \ge \log_2 t$.
\end{lemma}

\begin{proof}
Let $F$ be a fooling set for $f$. Let $(x,y) \in F$. Consider an $f$-monochromnatic rectangle $R \subseteq \mathcal{X}\times \mathcal{Y}$. We claim that if $(x,y) \in R$ then $(x',y') \not\in R$ for all $(x',y') \in F\backslash\{(x,y)\}$. Suppose for a contradiction that $(x,y) \in R$ and $(x',y') \in R$. Then since $R$ is a rectangle, $(x,y') \in R$ and $(x',y) \in R$. But since $f(x,y) = b$ and either $f(x',y) \neq b$ or $f(x,y') \neq b$, we have that $R$ is not $f$-monochromatic, a contradiction.
\paragraph{}
Thus every $f$-monochromatic rectangle can contain at most $1$ element of $F$. Thus $\chi(f) \geq |F| = t$, and furthermore
$$D(f) \geq \log_2\chi(f) \geq \log_2 t.$$
\end{proof}

\newpage \section{Equality II}

Using the fooling set bound, we can get an optimal lower bound on the communication complexity of the equality function.

\begin{theorem}
$D(\Eq) \ge n$.
\end{theorem}

\begin{proof}
Let $F = \{(x,x) : x \in \{0,1\}^n\}$. Since $|F| = 2^n$, if $F$ is a fooling set then by the previous result 
$$D(\Eq) \ge \log_2|F| = n.$$
Thus it suffices to show $F$ is a fooling set. For property $(1)$ observe that $f(x,x) = 1$ for all $x\in\{0,1\}^n$. For property $(2)$ observe that if $(x,x), (y,y) \in F$ with $x\neq y$ then $f(x,y) = f(y,x) = 0\neq 1$. Thus $F$ is a fooling set, as desired.
\end{proof}

\exercises

\begin{exercise}
Prove that in fact $D(\Eq) = n+1$.
\end{exercise}

\newpage \section{Set disjointness}

The \emph{set disjointness} function $\Disj : 2^{[n]} \times 2^{[n]}$ is defined by
\[
\Disj(S,T) = \begin{cases}
1 & \mbox{if } S \cap T = \emptyset \\
0 & \mbox{if } S \cap T \neq \emptyset.
\end{cases}
\]
Use the fooling set method to obtain optimal bounds on the communication complexity of the set disjointness function.

\begin{theorem}
$D(\Disj) = \Theta(n)$.
\end{theorem}

\begin{proof}
From the general upper bound,
$$D(\Disj) \leq \log_2|2^{[n]}| + 1 = n +1 = O(n).$$
For the lower bound consider the fooling set $F = \{(S,\bar{S}) : S \in 2^{[n]}\}$ (here $\bar{S} = [n]\backslash S$). Since the size of $F$ is $2^n$ it will again suffice to verify the fooling set properties. Indeed $S\cap \bar{S}= \emptyset$ so $f(S,\bar{S}) = 1$ for all $(S,\bar{S}) \in F$ and hence property $(1)$ holds. Property $(2)$ holds as for any $S,T \in 2^{[n]}$ with $S\neq T$ we either have $S\cap \bar{T} \neq \emptyset$ or $\bar{S}\cap T \neq \emptyset$.
Indeed if $S\cap \bar{T}= \emptyset$ then $S\subseteq T$ and in fact this inclusion is strict since $S\neq T$. That is, there exists $e \in T\backslash S$, and thus $\bar{S}\cap T \neq \emptyset$.
\end{proof}

\newpage \section{Inner product (Bonus)}

The fooling set bound is not always tight. One particularly noteworthy example where this method fails to give a good lower bound is the inner product function  $\IP : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ defined by
\[
\IP(x,y) = \sum_{i=1}^n x_i y_i \pmod{2}.
\]
As we will see later, $D(\IP) = \Theta(n)$, but the fooling set bound can only give the much weaker bound of $D(\IP) = \Omega(\log n)$.

\begin{theorem}
Every fooling set for the $\IP$ function has size at most $n^2$.
\end{theorem}

\begin{proof}
Let $F = \{(x_1,y_1),\dots, (x_t,y_t)\}$ be a fooling set for $\IP$. Let $b$ be such that $f(x,y) = b$ for all $(x,y) \in F$. We will assume $b=1$, and discuss afterwards how to modify the argument for $b=0$.

Let $A$ be a $t\times t$ matrix over $\mathbb{F}_2$ such that
$$A_{i,j} = f(x_i,y_j).$$
Let $B = A^T\circ A$ (where $\circ$ is the Hadamard product). Observe by property $(1)$ of $F$, the diagonal entries of $B$ are $1$, and by property $(2)$, the off-diagonal entries of $B$ are $0$, i.e. $B = I_t$. Now we use the rank inequality of linear algebra
$$t = \text{rank}(B) \leq \text{rank}(A^T)\text{rank}(A) \leq \text{rank}(A)^2.$$
Thus we want to show $\text{rank}(A) \leq n$. Let $X = \begin{bmatrix}
x_1 & \dots & x_t
\end{bmatrix}$
and let $Y = \begin{bmatrix}
y_1 &\dots & y_t
\end{bmatrix}$. Now $A = X^TY$, and so $\text{rank}(A) \leq n$, the dimension of columns of $Y$ and rows of $X$.

To treat the case where $b=0$ instead of arguing from $A = X^T Y$ we argue from 
$$A = \begin{bmatrix}
X\\ 1^T
\end{bmatrix}^T\begin{bmatrix}
Y\\ 1^T
\end{bmatrix}$$
where here $1$ is the vector of all ones.
\end{proof}

\newpage \section{Special case of the rectangle size bound}

The partition bound says that the communication complexity of $f : \calX \times \calY \to \{0,1\}$ is large if the minimum number of $f$-chromatic rectangles required to partition $\calX \times \calY$ is large. And the number of $f$-chromatic rectangles required to partition $\calX \times \calY$ must be large whenever the only $f$-chromatic rectangles are small. This observation is the core of the \emph{rectangle size bound}.

\begin{definition}[$m(f)$]
For a given function $f : \calX \times \calY \to \{0,1\}$, define the \emph{maximum rectangle size} of $f$ to be
\[
m(f) = \max \{ |R| : R \subseteq \calX \times \calY \mbox{ is an $f$-monochromatic rectangle} \}.
\]
\end{definition}

\begin{lemma}
For every function $f : \calX \times \calY \to \{0,1\}$, $\chi(f) \ge \frac{|\calX| \, |\calY|}{m(f)}$ and therefore
\[
D(f) \ge \log_2 \frac{|\calX| \, |\calY|}{m(f)}.
\]
\end{lemma}

\begin{proof}
Let $\mathcal{R}$ be the set of $f$-monochromatic rectangles partitioning $\mathcal{X}\times \mathcal{Y}$. Since every element of $\mathcal{X}\times\mathcal{Y}$ is in exactly one rectangle in $\mathcal{R}$,
$$\sum_{R \in \mathcal{R}} |R| = |\mathcal{X}||\mathcal{Y}|.$$
Thus we see that
$$|\mathcal{X}||\mathcal{Y}| = \sum_{R \in \mathcal{R}} |R| \le \sum_{R \in \mathcal{R}} m(f) = |\mathcal{R}|m(f) = \chi(f)m(f)$$
as desired.
\end{proof}

\newpage \section{Inner product II (Bonus)}

Use the rectangle size bound to prove an optimal lower bound on the inner product function.

\begin{theorem}
$m(IP) \le 2^{n}$ and so $D(IP) = \Theta(n)$.
\end{theorem}

\begin{proof}
Consider an $\IP$-monochromatic rectangle $R = A\times B$. Let $(x,y) \in R$, and let $X, Y \in 2^{[n]}$ be the sets indicated by $x$ and $y$ respectively. We can think of $\IP(x,y)$ as the parity of $|X\cap Y|$. Let $b = \IP(x,y) = |X\cap Y|$. Half of all sets in $2^{[n]}$ have odd intersection with $X$ and half of all sets in $2^{[n]}$ have even intersection with $X$. This yields a bound on $|B|$ of $2^{n/2}$. Symmetrically $|A| \leq 2^{n/2}$. Therefore
$|R| \leq 2^{n/2}\cdot 2^{n/2} = 2^n$. Thus
$$m(\IP) \leq 2^n.$$
\end{proof}

\newpage \section{Rectangle size bound}

The (general) rectangle size bound is obtained by generalizing the observation from the last section: instead of just counting the number of elements in a rectangle, we can consider the measure of rectangles under \emph{any} probability distribution on $\calX \times \calY$.

\begin{definition}[$m_\mu(f)$]
For a given function $f : \calX \times \calY \to \{0,1\}$ and a given distribution $\mu$ on $\calX \times \calY$, define the \emph{maximum rectangle size of $f$ with respect to $\mu$} to be
\[
m_\mu(f) = \max \{ \mu(R) : R \subseteq \calX \times \calY \mbox{ is an $f$-monochromatic rectangle} \}.
\]
\end{definition}

\begin{lemma}
For every function $f : \calX \times \calY \to \{0,1\}$ and every distribution $\mu$ on $\calX \times \calY$, 
\[
\chi(f) \ge \frac1{m_\mu(f)}
\]
and therefore
\[
D(f) \ge \log_2 \frac{1}{m_\mu(f)}.
\]
\end{lemma}

\begin{proof}
Let $\mathcal{R}$ be the set of $f$-monochromatic rectangles partitioning $\mathcal{X}\times \mathcal{Y}$. Since every element of $\mathcal{X}\times\mathcal{Y}$ is in exactly one rectangle in $\mathcal{R}$,
$$\sum_{R \in \mathcal{R}} \mu(R) = 1.$$
Thus we see that
$$1 = \sum_{R \in \mathcal{R}} \mu(R) \le \sum_{R \in \mathcal{R}} m_\mu(f) = |\mathcal{R}|m_\mu(f) = \chi(f)m(f)$$
as desired.
\end{proof}

\newpage \section{Fooling sets and rectangle size bound}

The fooling set bound is a special case of the rectangle size bound, as the following theorem shows.

\begin{theorem}
If $f : \calX \times \calY \to \{0,1\}$ has a fooling set $S \subseteq \calX \times \calY$ of size $|S| = t$, then there is a distribution $\mu$ on $\calX \times \calY$ for which $m_\mu(f) \le 1/t$.
\end{theorem}

\begin{proof}
Let $\mu$ be the distribution defined by 
$$\mu(x,y) = \begin{cases}
\frac{1}{t}, &\text{if } (x,y) \in S \\
0, &\text{otherwise}
\end{cases}$$
Since $|S| = t$, it is not hard to verify that $\mu$ is indeed a probability distribution. In a previous problem, we showed that every $f$-monochromatic rectangle contains at most $1$ element of $S$. Hence every $f$-monochromatic rectangle $R$ satisfies $\mu(R) \leq \frac{1}{t}$ and thus
$$m_\mu(f) \leq\frac{1}{t}$$
as desired.
\end{proof}

\newpage \section{Log rank bound}

Another convenient measure that lower bounds the partition number of a function is the log of the rank of the corresponding matrix.

\begin{definition}[Matrix of a function]
The \emph{matrix} $M_f$ corresponding to the function $f : \calX \times \calY \to \{0,1\}$ is the $|\calX| \times |\calY|$-dimensional $\{0,1\}$-valued matrix with rows indexed by $\calX$ and columns indexed by $\calY$ defined by
\[
(M_f)_{x,y} = f(x,y)
\]
for every $x \in \calX$ and $y \in \calY$.
\end{definition}

\begin{definition}[Rank]
The \emph{rank} of the function $f$, denoted
\[
\rank(f),
\] 
is the linear rank of the matrix $M_f$ over $\R$.
\end{definition}

The logarithm of the rank of a function gives a lower bound on the communication complexity of the function.

\begin{lemma}
\label{lem:logrank}
For every $f : \calX \times \calY \to \{0,1\}$,
\[
D(f) \ge \log_2 \rank(f).
\]
\end{lemma}

\begin{proof}
Consider a protocol tree for $f$. Let $L$ be the set of leaves of this tree. As observed in the previous lecture when discussing the partition bound, each leaf corresponds to an $f$-monochromatic rectangle. Let $L'\subseteq L$ be the leaves at which $1$ is output. For each leaf $\ell \in L'$, let $R^\ell$ be the $1$-monochromatic rectangle defined by the path to $\ell$, and define a matrix $M^\ell$ as follows
$$M^\ell_{x,y} = \begin{cases}
1, &\text{if } (x,y) \in R^\ell \\
0, &\text{otherwise}
\end{cases}.$$
Now we have that
$$M_f = \sum_{\ell \in L'} M^\ell$$
and thus
$$\rank(M_f) \leq \sum_{\ell \in L'} M^\ell \leq |L'| \leq |L| = 2^{D(f)}.$$
In the above inequality we used that $\rank(M^\ell) = 1$ for all $\ell \in L'$. This follows since there exist $A \subseteq \mathcal{X}$ and $B \subseteq \mathcal{Y}$ such that $R^\ell = A\times B$, and if use $\chi(A), \chi(B)$ to denote the characteristic vectors of $A$ and $B$ then $M^\ell = \chi(A)\chi(B)^T$.
\end{proof}

\exercises

\begin{exercise} %[Easy]
Give an alternative proof that $D(\Eq) \ge n$ using the log rank bound.
\end{exercise}

\begin{exercise}
Give an alternative proof that $D(\IP) \ge n$ using the log rank bound.\end{exercise}

\newpage \section{Greater than}

The greater-than function $\GT : [2^n] \times [2^n] \to \{0,1\}$ is defined by
\[
\GT(x,y) = \begin{cases}
1 & \mbox{if } x > y \\
0 & \mbox{otherwise.}
\end{cases}
\]
Use the log rank bound to give an optimal lower bound on the greater-than function.

\begin{theorem}
$D(\GT) = \Theta(n)$.
\end{theorem}

\begin{proof}
	Observe that $(M_\GT)_{i,j} = 1 $ iff $i >j$ ($(M_\GT)_{i,j} = 0$ otherwise). Thus $M_\GT$ is entirely ones below the diagonal, and entirely zeroes on the diagonal and above. Therefore $\rank(M_\GT) = 2^n-2$. Combining with the general upper bound we see that $D(\GT) = \Theta(n)$.
\end{proof}

\newpage \section{Tightness of the log rank bound}

Prove that the rank of a function can also be used to obtain an upper bound on the communication complexity of the function.

\begin{theorem}
For every $f : \calX \times \calY \to \{0,1\}$, $D(f) \le \rank(f) + 1$.
\end{theorem}

\begin{proof}
Here we use two facts of linear algebra. The first is that the rank of a $0$-$1$ matrix over $\mathbb{R}$ is at least its rank over $\mathbb{F}_2$. In symbols $$\rank_\mathbb{R}(f) \geq \rank_{\mathbb{F}_2}(f).$$
This follows since a set of linearly independent columns over $\mathbb{F}_2$ remains independent over $\mathbb{R}$.
The second is that the number of distinct columns (and rows) of $M_f$ is at most $$2^{\rank_{\mathbb{F}_2}(f)} \leq 2^{\rank(f)}.$$
This follows since there are at most $2^{\rank(f)}$ ways to linearly combine $\rank(f)$ columns over $\mathbb{F}_2$.

Using that $M_f$ has at most $2^{\rank(f)}$ distinct columns (and rows) we will design a communication protocol as follows. Bob sends Alice $\rank(f)$ bits indicating which column of $M_f$ his input corresponds to. Alice then has suffice knowledge to compute $f$, and in one more bit the protocol outputs the result. Hence
$D(f) \leq \rank(f) + 1$.
\end{proof}

\exercises

\begin{exercise} %[Hard]
Show that there exists a function $f$ for which $D(f) = \omega( \log_2 \rank(f) )$.
\end{exercise}

\begin{open}[Log rank conjecture]
Prove that there exists a constant $c > 0$ such that every function $f$ satisfies
\[
D(f) = O( \log^c \rank(f)).
\]
\end{open}

\end{document}

### 2019-01-21 CS860-CommunicationComplexity/assignments/Chapter3.tex
Adding Communication Complexity

\documentclass[11pt]{amsart}
\usepackage{amsmath,amsthm,amsfonts,amssymb}
\usepackage{fullpage}
\usepackage[usenames]{xcolor}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{problem}{Problem}
\newtheorem{challenge}{Challenge Problem}
\newtheorem{open}{Open Problem}
\theoremstyle{plain}

\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calZ}{\mathcal{Z}}
\newcommand{\cost}{\mathrm{cost}}
\newcommand{\disc}{\mathrm{disc}}
\newcommand{\discu}{\mathrm{disc_u}}
\newcommand{\Disj}{\textsc{Disj}}
\newcommand{\Eq}{\textsc{Eq}}
\newcommand{\GT}{\textsc{GT}}
\newcommand{\IP}{\textsc{IP}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}

\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}

\title{Communication complexity: Chapter 3 \\ Distributional communication complexity}
\author{Eric Blais and Justin Toth}

\begin{document}

\maketitle

Functions can become much easier to compute in the communication complexity setting when we allow the protocols to make some errors. We can define this formally using the notion of \emph{distributional communication complexity}. We first explore this notion in its most natural setting, which corresponds to the uniform distribution.

\begin{definition}[Protocol error]
Fix any $\epsilon \ge 0$. A protocol $\pi$ \emph{computes} $f$ \emph{with error at most $\epsilon$ under the uniform distribution} if it correctly outputs the value $f(x,y)$ for at least an $1-\epsilon$ fraction of all inputs, i.e., if
\[
\Pr_{(x,y)}[ \pi(x,y) \neq f(x,y)] \le \epsilon
\]
when $(x,y)$ is drawn uniformly at random from $\calX \times \calY$.
\end{definition}

\begin{definition}[Uniform distributional complexity]
For any $\epsilon \ge 0$, the \emph{$\epsilon$-error distributional communication complexity} of $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$ with respect to the uniform distribution,
\[
D_\epsilon^{\mathrm{unif}}(f),
\]
is the minimum communication cost of a protocol that computes $f$ with error at most $\epsilon$ under the uniform distribution.
\end{definition}

\exercises

\begin{exercise}
Show that $D_0^{\mathrm{unif}}(f) = D(f)$ and that for every $\epsilon > 0$, 
$D_\epsilon^{\mathrm{unif}}(f) \le D(f)$.
\end{exercise}

\begin{exercise}
Show that every $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$ satisfies
$D_{1/2}^{\mathrm{unif}}(f) = 0$.
\end{exercise}

\newpage \section{Equality III}

The $\epsilon$-error distributional complexity of functions can sometimes be dramatically smaller than their deterministic complexity.

\begin{theorem}
For any $\epsilon \ge \frac1{2^n}$, $D_\epsilon^{\mathrm{unif}}(\textsc{Equality}) = 0$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Uniform discrepancy}

The \emph{uniform discrepancy} of a function is a different way to measure how large ``nearly $f$-monochromatic'' rectangles can be. For this measure, it is convenient to represent the function with a $\pm1$-valued (instead of $\{0,1\}$-valued) matrix.

\begin{definition}[$\pm 1$-Matrix of a function]
The \emph{$\pm 1$-matrix} $M^\pm_f$ corresponding to the function $f : \calX \times \calY \to \{0,1\}$ is the $|\calX| \times |\calY|$-dimensional $\{-1,1\}$-valued matrix with rows indexed by $\calX$ and columns indexed by $\calY$ defined by
\[
\big(M^{\pm}_f\big)_{x,y} = (-1)^{f(x,y)}
\]
for every $x \in \calX$ and $y \in \calY$.
\end{definition}

\begin{definition}[Uniform discrepancy]
The \emph{uniform discrepancy} of $f : \calX \times \calY \to \{0,1\}$ is
\[
\discu(f) = \max_{A \subseteq \calX, B \subseteq \calY} \frac{1}{|\calX| \, |\calY|} \left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|.
\]
\end{definition}

We can use the uniform discrepancy to bound deterministic communication complexity.

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$,
\[
\chi(f) \ge \discu(f)^{-1}
\] 
and so $D(f) \ge \log_2 \frac1{\discu(f)}$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Uniform discrepancy bound}

We can also use uniform discrepancy to bound the uniform distributional  communication complexity of functions.

\begin{lemma}
For every function $f : \calX \times \calY \to \{0,1\}$ and every $0 \le \epsilon < \frac12$, 
\[
D_\epsilon^{\mathrm{unif}}(f) \ge \log_2 \left( \frac{1-2\epsilon}{\discu(f)} \right).
\]
\end{lemma}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Inner product III}

The discrepancy method can be used to give optimal bounds on the distributional communication complexity of the inner product function over the uniform distribution.

\begin{theorem}
$\discu(\IP) \le 2^{-n/2}$ and, as a result, $D_\epsilon^{\mathrm{unif}}(\IP) \ge \Omega(n)$ for every $\epsilon < \frac12$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Distributional complexity}

The definitions we have seen so far in this chapter all generalize to arbitrary distributions over the domain of the function.

\begin{definition}[Protocol error]
Fix any $\epsilon \ge 0$. A protocol \emph{computes} $f : \calX \times \calY \to \{0,1\}$ \emph{with error at most $\epsilon$ under the distribution $\mu$ on $\calX \times \calY$} if it correctly outputs the value $f(x,y)$ for at least a $1-\epsilon$ measure of all inputs in $\mu$, i.e., if
\[
\Pr_{(x,y) \sim \mu}[ \pi(x,y) \neq f(x,y) ] \le \epsilon.
\]
\end{definition}

\begin{definition}[Distributional complexity]
For any $\epsilon \ge 0$, the \emph{$\epsilon$-error distributional communication complexity} of $f : \calX \times \calY \to \{0,1\}$ with respect to the distribution $\mu$ on $\calX \times \calY$,
\[
D_\epsilon^{\mu}(f),
\]
is the minimum communication cost of a protocol that computes $f$ with error at most $\epsilon$ under the distribution $\mu$.
\end{definition}

\begin{definition}[$\mu$-Discrepancy]
For any distribution $\mu$ on $\calX \times \calY$, 
the \emph{$\mu$-discrepancy} of $f : \calX \times \calY \to \{0,1\}$ is
\[
\disc_\mu(f) = \max_{A \subseteq \calX, B \subseteq \calY} 
\left| \mu\big( (A \times B) \cap f^{-1}(0) \big) - \mu\big( (A \times B) \cap f^{-1}(1)\big) \right|.
\]
\end{definition}

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$ and every distribution $\mu$ over $\calX \times \calY$, 
\[
\chi(f) \ge \disc_\mu(f)^{-1}
\] 
and so $D(f) \ge \log_2 \frac1{\disc_\mu(f)}$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
Show that when $\mu$ is the uniform distribution over $\calX \times \calY$, the definitions of $\mu$-discrepancy and uniform discrepancy are equivalent.
\end{exercise}

\begin{exercise}
Show that for every distribution $\mu$, $D^\mu_\epsilon(f) \le D(f)$.
\end{exercise}

\newpage \section{Discrepancy bound}

The $\mu$-discrepancy of a function can be used to give lower bounds on its $\mu$-distributional communication complexity.

\begin{lemma}
For every function $f : \calX \times \calY \to \{0,1\}$, every distribution $\mu$ on $\calX \times \calY$, and every $0 \le \epsilon < \frac12$,
\[
D_\epsilon^{\mu}(f) \ge \log_2 \left( \frac{1-2\epsilon}{\disc_\mu(f)} \right).
\]
\end{lemma}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Equality IV}

Using the discrepancy bound, we can show that there are distributions for which the equality function requires Alice and Bob to exchange \emph{some} bits of communication to obtain a non-trivial error.

\begin{theorem}
There exists a distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ for which 
$
\disc_\mu(\Eq) \ge \frac18
$
and so
\[
D^\mu_\epsilon(\Eq) \ge 1
\]
for every $\epsilon \le \frac38$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Equality V (Bonus)}

Show that the bound in the last section is essentially tight in that the distributional communication complexity of the equality function is constant when $\epsilon$ is a positive constant.

\begin{theorem}
For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ and every constant $\epsilon > 0$,
\[
D^\mu_{1/4}(\Eq) = O(1).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here, if you complete it.}
\end{proof}

\newpage \section{Corruption bound}

The \emph{corruption bound} is another powerful technique for proving lower bounds in distributional communication complexity.

\begin{lemma}
Fix a function $f : \calX \times \calY \to \{0,1\}$ and a distribution
$\mu$ on $\calX \times \calY$.
If there exist parameters $\alpha, \beta > 0$ for which
every rectangle $R \subseteq \calX \times \calY$ satisfies 
\[
\mu(R \cap f^{-1}(0)) \ge \alpha \cdot \mu(R) - \beta
\]
then
\[
D^\mu_\epsilon(f) \ge \log\left( \frac{\alpha \cdot \mu(f^{-1}(1)) - \epsilon}{\beta}\right).
\]
\end{lemma}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage \section{Set Disjointness II}

The corruption bound can be used to prove a strong lower bound on the distributional communication complexity of the set disjointness function. The key claim that is used to obtain this result is the following combinatorial statement.

\begin{proposition}[\textsc{Bonus}]
Every rectangle $R = A \times B \subseteq 2^{[n]} \times 2^{[n]}$
that satisfies 
\[
\Pr_{(S,T) \in R}[ S \cap T = \emptyset ] \ge 1-\alpha
\]
has size bounded by
\[
|A| \le 2^{-\delta \sqrt{n}} \binom{n}{\sqrt{n}}
\qquad \mbox{or} \qquad
|B| \le 2^{-\delta \sqrt{n}} \binom{n}{\sqrt{n}}.
\]
\end{proposition}

\begin{proof}
\replacethistext{Enter your proof here, if you complete it.}
\end{proof}

Use the claim to complete the lower bound on the communication complexity of the set disjointness function.

\begin{theorem}
Let $\mu$ be the uniform distribution on pairs $(S,T) \in 2^{[n]} \times 2^{[n]}$ that satisfy $|S| = |T| = \sqrt{n}$. Then
\[
D_\epsilon^{\mu}(\textsc{Disj}) \ge \Omega(\sqrt{n}).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\end{document}

### 2019-01-25 CS860-CommunicationComplexity/assignments/Chapter3.tex
Add files via upload

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Consider the trivial protocol which just outputs $1$ on every pair of inputs. Then $\Pr[\pi(x,y)\neq \Eq(x,y)]$ is precisely the probability that $x=y$. Hence
+$$\Pr[\pi(x,y)\neq \Eq(x,y)] = \Pr[x=y] = (\frac{1}{2})^n$$
+since each bit of $x$ and $y$ is chosen independently and uniformly at random.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+$\chi(f) \geq \discu(f)^{-1}$ if and only if
+$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right| \geq |\calX||\calY|.$$
+
+Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
+$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  = \sum_{R\in \mathcal R} \max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right| \geq\sum_{R \in \mathcal{R}} \left| \sum_{(x,y) \in R} \big(M^{\pm}_f\big)_{x,y} \right|= \sum_{R \in \mathcal R} |R|$$
+using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
+$$\sum_{R \in \mathcal R} |R| = |\calX||\calY|$$
+and hence the result holds.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be the protocol attaining $D_\epsilon^{\mathrm{unif}}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
+$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
+Let $\{R_1, \dots, R_t\}$ be set of $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
+$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) = f(x,y) \mid(x,y) \in R_i) = \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.$$
+Symmetrically
+$$\Pr[\pi(x,y) \neq f(x,y)] = \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i \cap f^{-1}(\pi(x,y)\oplus 1)|.$$
+Thus we have
+\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
+&= \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i\cap f^{-1}(\pi(x,y))| -|R_i \cap f^{-1}(\pi(x,y)\oplus 1)| \\
+&\leq \frac{1}{|\calX||\calY|} \sum_{i=1}^t  \left| \sum_{(x,y) \in R_i} \big(M^{\pm}_f\big)_{x,y} \right|\\
+&\leq t\cdot\discu(f).\end{align*}
+Since $\log_2(t) = D_\epsilon^{\mathrm{unif}}(f)$ this completes the proof.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+For convenience, define $M:= M^{\pm}_\IP$ and let $\mathbf{1}_S$ denote characteristic vector of a set $S\subseteq [n]$. Now consider an entry $x,y$ of $M^TM$:
+$$M^TM_{x,y} = \sum_{z\in\{0,1\}^n} (-1)^{x\cdot z}(-1)^{y\cdot z} = \sum_{z \in \{0,1\}^n}(-1)^z\cdot(x+y) = \begin{cases}
+(-1)2^{n-1} + (1)2^{n-1}, &\text{if $x\neq y$} \\
+\sum_{z \in \{0,1\}^n}1, &\text{otherwise}.
+\end{cases}$$
+Hence $$M^TM_{x,y} = \begin{cases}
+0, &\text{if $x\neq y$} \\
+2^n, &\text{otherwise}
+\end{cases}$$
+i.e. $M^TM = 2^nI$.
+
+Consider a rectangle $R = A\times B$. Then we have
+\begin{align*}
+\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  &= \left| \mathbf{1}_A^T M \mathbf{1}_B\right | \\
+&\leq ||\mathbf{1}_A|| \cdot || M\mathbf{1}_B|| &\text{by Cauchy-Schwarz} \\
+&= \sqrt{|A|}\cdot\sqrt{\mathbf{1}^T_BM^TM\mathbf{1}_B} \\
+&= 2^{n/2}\sqrt{|A|}\sqrt{\mathbf{1}^T_B\mathbf{1}_B} \\
+&=2^{n/2}\sqrt{|A|}\sqrt{|B|}.
+\end{align*}
+Therefore
+$$\discu(\IP) \leq \frac{1}{2^n2^n}\max_{A \subseteq \calX, B \subseteq \calY} 2^{n/2}\sqrt{|A|}\sqrt{|B|} \leq 2^{-2n}2^n2^\frac{n}{2} = 2^{-n/2}.$$
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+
+$\chi(f) \geq \disc_\mu(f)^{-1}$ if and only if
+$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left|\mu((A\times B)\cap f^{-1}(0) -\mu((A\times B)\cap f^{-1}(1)) \right| \geq 1.$$
+
+Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
+\begin{align*}
+&\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left|\mu((A\times B)\cap f^{-1}(0) -\mu((A\times B)\cap f^{-1}(1)) \right| \\
+& = \sum_{R\in \mathcal R}\max_{A \subseteq \calX, B \subseteq \calY}\left|\mu((A\times B)\cap f^{-1}(0) -\mu((A\times B)\cap f^{-1}(1)) \right|\\ 
+&\geq \sum_{R \in \mathcal{R}}\left|\mu(R)\cap f^{-1}(0) -\mu(R\cap f^{-1}(1)) \right|\\ 
+&= \sum_{R \in \mathcal{R}} \mu(R)
+\end{align*}
+using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
+$$\sum_{R \in \mathcal R} \mu(R)=1$$
+and hence the result holds.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
+$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
+Let $\{R_1, \dots, R_t\}$ be set of $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
+$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr[(x,y) \in R_i \text{ and }\pi(x,y) = f(x,y) ] = \sum_{i=1}^t \mu(R_i \cap f^{-1}(\pi(x,y))).$$
+Symmetrically
+$$\Pr[\pi(x,y) \neq f(x,y)] = \sum_{i=1}^t\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1)).$$
+Thus we have
+\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
+&=\sum_{i=1}^t|\mu(R_i\cap f^{-1}(\pi(x,y))) -\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1))| \\
+&\leq \sum_{i=1}^t \disc_\mu(f)\\
+&= t\cdot\disc_\mu(f).\end{align*}
+Since $\log_2(t) = D_\epsilon^{\mu}(f)$ this completes the proof.
 \end{proof}
@@
 $
-\disc_\mu(\Eq) \ge \frac18
+\disc_\mu(\Eq) \leq \frac18 \frac{2^n}{(2^n-1)}.
 $
@@
 \begin{theorem}
-For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ and every constant $\epsilon > 0$,
+For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$,
 \[

### 2019-01-25 CS860-CommunicationComplexity/assignments/Chapter3.tex
almost done

@@
 $$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
-Let $\{R_1, \dots, R_t\}$ be set of $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
+Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
 $$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) = f(x,y) \mid(x,y) \in R_i) = \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.$$
@@
 $$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
-Let $\{R_1, \dots, R_t\}$ be set of $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
+Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
 $$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr[(x,y) \in R_i \text{ and }\pi(x,y) = f(x,y) ] = \sum_{i=1}^t \mu(R_i \cap f^{-1}(\pi(x,y))).$$
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$.
+Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf.
+
+$$1= \sum_{i=1}^t \mu(R_i) = \sum_{i=1}^t \mu(R_i \cap f^{-1}(0)) + \mu(R_i\cap f^{-1}(1)) \geq \alpha - t \beta +\mu(f^{-1}(1))$$
+Thus
+$$t \geq \frac{\alpha + \mu(f^{-1}(1)) -1}{\beta}.$$
+Now we claim that 
+$$\alpha + \mu(f^{-1}(1)) -1 \geq \alpha\cdot \mu(f^{-1}(1)) -\epsilon.$$
+This claim is equivalent to
+$$1-\epsilon \leq \alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)).$$
+This holds since $1-\epsilon$ is the probability of success \replacethistext{(I didn't have time to formalize)}.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
-\end{proof}
+From Proposition $1$, if $R = A\times B \subseteq 2^{[n]}\times 2^{[n]}$ is rectangle such that
+$$\mu(R) \geq 2^{-\delta\sqrt{n}}$$ then
+$$\mu(R\cap \Disj^{-1}(0)) > \alpha \mu(R).$$	
+Thus any rectangle $R$ satisfies
+$$\mu(R\cap \Disj^{-1}(0)) > \alpha \mu(R) - 2^{-\delta\sqrt{n}}.$$
+So by the corruption bound with $\alpha$ and $\beta = 2^{-\delta\sqrt{n}}$ we obtain
+$$D^{\mu}_\epsilon(\Disj) \geq \log(\frac{\alpha\cdot \mu(f^{-1}(1)) - \epsilon}{2^{-\delta\sqrt{n}}}) \geq \Omega(\sqrt{n}).$$\end{proof}
 

### 2019-01-25 CS860-CommunicationComplexity/assignments/Chapter3.tex
submission

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+I'm not really sure what the right distribution to take is, but intuitively I want to distribute the probability mass evenly between $(x,y)$ pairs with $x\neq y$ and pairs with $x=y$ so Alice and Bob have to exchange some bits to do better than guessing.
+From the matrix of this function I want half the probability mass uniformly distributed on the diagonal, and half uniformly distributed on all off-diagonal entries.
 \end{proof}
@@
 $$1-\epsilon \leq \alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)).$$
-This holds since $1-\epsilon$ is the probability of success \replacethistext{(I didn't have time to formalize)}.
+Since 
+$$\alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)) = 1-(1-\alpha) \mu(f^{-1}(0))$$
+the claim holds since the probability of failure is at least $(1-\alpha)f^{-1}(0)$ \replacethistext{(I didn't have time to formalize)}.
 \end{proof}
@@
 \begin{proposition}[\textsc{Bonus}]
-Every rectangle $R = A \times B \subseteq 2^{[n]} \times 2^{[n]}$
+For every $\alpha >0$ there exists $\delta>0$ such that every rectangle $R = A \times B \subseteq 2^{[n]} \times 2^{[n]}$
 that satisfies 

### 2019-01-29 CS860-CommunicationComplexity/assignments/Chapter4.tex
Add ch4

\documentclass[11pt]{amsart}
\usepackage{amsmath,amsthm,amsfonts,amssymb}
\usepackage{fullpage}
\usepackage[usenames]{xcolor}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{remark}{Remark}
\newtheorem{problem}{Problem}
\newtheorem{challenge}{Challenge Problem}
\newtheorem{open}{Open Problem}
\theoremstyle{plain}

\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calZ}{\mathcal{Z}}
\newcommand{\cost}{\mathrm{cost}}
\newcommand{\disc}{\mathrm{disc}}
\newcommand{\discu}{\mathrm{disc_u}}
\newcommand{\Disj}{\textsc{Disj}}
\newcommand{\Eq}{\textsc{Eq}}
\newcommand{\GT}{\textsc{GT}}
\newcommand{\HD}{\textsc{HD}}
\newcommand{\IP}{\textsc{IP}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\Rpriv}{R^{\mathrm{priv}}}
\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}

\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}

\title{Communication complexity: Chapter 4 \\ Randomized communication complexity}
\author{Eric Blais and Justin Toth}

\begin{document}

\maketitle

Until now, we have been focused on deterministic communication protocols. In this chapter, we study \emph{randomized} communication protocols. We first focus on what is known as the \emph{public-randomness model} of communication.

\begin{definition}[Randomized protocol]
A \emph{randomized communication protocol} $\Pi$ is a distribution over deterministic communication protocols. The \emph{(worst-case) cost} of $\Pi$ is the maximum cost of any protocol in its support. The randomized protocol $\Pi$ \emph{computes} $f : \calX \times \calY \to \{0,1\}$ with error at most $\epsilon$ if for every $(x,y) \in \calX \times \calY$,
\[
\Pr_{\pi \sim \Pi}[ \pi(x,y) \neq f(x,y) ] \le \epsilon.
\]
\end{definition}

\begin{definition}[Randomized communication complexity]
For $0 < \epsilon < \frac12$, the \emph{randomized communication complexity} of $f : \calX \times \calY \to \{0,1\}$ at error $\epsilon$ is
\[
R_\epsilon(f) := \min_{\Pi} \cost(\Pi)
\]
where the minimum is taken over all randomized communication protocols that compute $f$ with error at most $\epsilon$.
\end{definition}

\begin{remark}
The notation for randomized communication complexity can vary. In Kushilevitz and Nisan, for instance, the randomized communication complexity defined above is written as $R^{\mathrm{pub}}_\epsilon$ and our symbol $R_\epsilon$ is reserved for another model of randomized communication (namely: the private randomness model we will see at the end of the chapter).
\end{remark}

\exercises

\begin{exercise}
Another way to define randomized communication protocols is to consider a model where Alice and Bob have access to their respective inputs \emph{and} to a common sequence of coins that are flipped at random. Show that this model is equivalent to the one described above.
\end{exercise}

\begin{exercise}
Show that with the definition above, $R_0(f) = D(f)$ always holds. For this reason, $R_0(f)$ is usually reserved to the randomized model of communication where the cost of $\Pi$ on input $(x,y)$ is defined to be the \emph{average} cost of the protocols in the support of $\Pi$.
\end{exercise}

\newpage 
\section{Randomized and distributional complexity}

Lower bounds in randomized communication complexity are often (usually?) obtained by establishing the lower bounds in the distributional communication complexity model instead. This approach is justified by the following relation, which is sometimes referred to as \emph{(the easy direction of) Yao's minimax principle}.

\begin{theorem}
For every $0 \le \epsilon \le \frac12$, every function $f : \calX \times \calY \to \{0,1\}$, and every distribution $\mu$ over $\calX \times \calY$,
\[
R_\epsilon(f) \ge D_\epsilon^\mu(f).
\]
\end{theorem}

\begin{proof}
Let $\Pi$ be a distribution over deterministic communication protocols computing $f$ such that $R_\epsilon(f) = \cost(\Pi)$. Let $\mu$ be a distribution over $\calX\times\calY$. Since $\Pr_{\pi \sim \Pi}[\pi(x,y)\neq f(x,y)]\leq \epsilon$ for all $(x,y) \in \calX \times \calY$ we have
$$\Pr_{(x,y)\sim \mu ;\pi\sim \Pi}[\pi(x,y)\neq f(x,y)]] \leq \epsilon.$$
Hence there exists deterministic protocol $\pi$ such that 
$$\Pr_{(x,y)\sim \mu}[\pi(x,y) \neq f(x,y)] \leq \epsilon.$$
\end{proof}

\newpage 
\section{Yao's minimax principle (Bonus)}

Yao's minimax principle says something quite a bit stronger than the bound seen in the last chapter: it says that the randomized communication complexity of a function is \emph{equal} to the maximum distributional complexity of the function over any distributions on its domain.

\begin{theorem}
For every $0 \le \epsilon \le \frac12$ and every function $f : \calX \times \calY \to \{0,1\}$, 
\[
R_\epsilon(f) = \max_{\mu} D_\epsilon^\mu(f)
\]
where the maximum is taken over all distributions on $\calX \times \calY$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here, if you complete it.}
\end{proof}

\newpage 
\section{Error reduction}

The randomized communication complexity functions is usually studied in the setting where $\epsilon = \frac13$. In fact, this is so common that shorthand notation is used for that case.

\begin{definition}
The \emph{(two-sided error) randomized communication complexity} of $f : \calX \times \calY \to \{0,1\}$ is
\[
R(f) = R_{1/3}(f).
\]
\end{definition}

The reason most of the work focuses on this particular choice of $\epsilon$ is that a general transformation can be used to bound $R_\epsilon(f)$ for any $\epsilon > 0$ as a function of $R(f)$. This result is sometimes known as the \emph{(majority) confidence amplification} trick.

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$ and every $0 < \epsilon < \frac12$, 
\[
R_\epsilon(f) = O\left( R(f) \cdot \log \tfrac1\epsilon \right).
\]
\end{theorem}

\begin{proof}
Let $\Pi$ be a distribution over deterministic communication protocols computing $f$ such that $R(f) = \cost(\Pi)$. Consider the following random construction of a protocol: for some $t$ to be chosen later, sample independently $t$ protocols according to $\Pi$: $\pi_1,\dots, \pi_t$. Protocol $\pi$ will run $\pi_i$ on its input obtaining output $b_i \in\{0,1\}$, for each $i = 1,\dots, t$. Then $\pi$ will output the median of $b_1, \dots, b_t$ (suppose without loss they are ordered non-decreasingly). Call the distribution on protocols obtained by this random construction $\Pi'$.

A protocol chosen according to $\Pi'$ fails only if at least half of its constituent $t$ protocols chosen according $\Pi$ fail. Hence for any $(x,y) \in \calX\times \calY$,
$$\Pr_{\pi \sim \Pi'}[\pi(x,y)\neq f(x,y)] \leq \prod_{i=1}^{t/2} \Pr[\pi_i(x,y) \neq f(x,y)] \leq (\frac{1}{3})^{\frac{t}{2}}.$$
Thus if we chose $t = 2\log_3(\frac{1}{\epsilon})$ then $\Pr_{\pi \sim \Pi'}[\pi(x,y)\neq f(x,y)] \leq \epsilon$. Hence $t = O(\log\frac{1}{\epsilon})$. The cost of $\Pi'$ is the cost of $\Pi$ repeated $t$ times, so $\cost(\Pi') = O(R(f)\cdot \log\frac{1}{\epsilon})$ as desired.
\end{proof}

\exercises

\begin{exercise}
Extend the result in the theorem to show that for every $\delta > 0$, we also have
\[
R_\epsilon(f) = O\left( R_{\frac12 - \delta}(f) \cdot \phi(\epsilon,\delta) \right)
\]
for some appropriate function $\phi$ on $\epsilon$ and $\delta$.
\end{exercise}

\newpage 
\section{Equality VI}

\begin{theorem}
For every $0 < \epsilon < \frac12$,
\[
R_\epsilon(\Eq) = O\left(\log \tfrac1\epsilon \right).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Greater than II}

Recall that the function $\GT : [2^n] \times [2^n] \to \{0,1\}$ is defined by
\[
\GT(x,y) = \begin{cases}
1 & \mbox{if } x > y \\
0 & \mbox{otherwise.}
\end{cases}
\]
We saw in Chapter 2 that the deterministic communication complexity of this function is $\Theta(n)$. Its randomized communication complexity is much smaller.

\begin{theorem}
The randomized communication complexity of $\GT : [2^n] \times [2^n] \to \{0,1\}$ is bounded above by
\[
R(\GT) = O\left(\log^2 n \right).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
Improve the upper bound to show $R(\GT) = O\left(\log n \log \log n \right)$.
\end{exercise}

\begin{exercise}
Prove that $R(\GT) = \Theta\left(\log n \right)$.
\end{exercise}

\newpage 
\section{Hamming distance}

The \emph{$k$-Hamming distance} function $\HD_k : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ is defined by
\[
\HD_k(x,y) = \begin{cases}
1 & \mbox{if } |\{i \in [n] : x_i \neq y_i\}| \le k \\
0 & \mbox{otherwise.}
\end{cases}
\]

\begin{theorem}
For every $1 \le k \le \frac n2$,
\[
R(\HD_k) = O( k \log k).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Private randomness}

In the beginning of the chapter, we saw that the (public-coin) randomized communication model corresponds to the model where some coins are flipped and both Alice and Bob see the outcome of those coin flips. It is natural to study the alternative model where Alice and Bob both have access to coins they can flip (or, more abstractly, to some sources of randomness), but they each only see the outcome of their own coin flips. This model of communication corresponds to the \emph{private-coin randomized communication} model.

\begin{definition}
A \emph{private-coin randomized communication protocol} $\pi$ is equivalent to a deterministic communication protocol with the two additions: 
\begin{enumerate}
\item Each of Alice's internal node $v$ is labelled with a function $h_v : \calX \times R_A \to \{0,1\}$ and each of Bob's internal node $w$ is labelled with a function $h_v : \calY \times R_B \to \{0,1\}$; 
\item Alice has a distribution $\mu_A$ over $R_A$ and Bob has a distribution $\mu_B$ over $R_B$.
\end{enumerate}
The \emph{(worst-case) cost} of $\Pi$ is the maximum cost of any protocol in its support. The randomized protocol $\Pi$ \emph{computes} $f : \calX \times \calY \to \{0,1\}$ with error at most $\epsilon$ if for every $(x,y) \in \calX \times \calY$,
\[
\Pr_{r_a \sim \mu_A, r_b \sim \mu_B}[ \pi(x,y) \neq f(x,y) ] \le \epsilon.
\]
\end{definition}

\begin{definition}[Private-coin randomized communication complexity]
For $0 < \epsilon < \frac12$, the \emph{private coin randomized communication complexity} of $f : \calX \times \calY \to \{0,1\}$ at error $\epsilon$ is
\[
\Rpriv_\epsilon(f) := \min_{\Pi} \cost(\Pi)
\]
where the minimum is taken over all private-coin randomized communication protocols that compute $f$ with error at most $\epsilon$.
\end{definition}

\begin{remark}
As in the public-coin setting, we write $\Rpriv(f) := \Rpriv_{1/3}(f)$.
\end{remark}
\exercises

\begin{exercise}
Show that for every function $f$ and every parameter $\epsilon > 0$, 
$R_\epsilon(f) \le \Rpriv_\epsilon(f)$.
\end{exercise}

\begin{exercise}
Show that error amplification also holds in the private coin model: for every $\epsilon > 0$, $\Rpriv_\epsilon(f) = O(\Rpriv(f) \log \frac1\epsilon)$.
\end{exercise}

\newpage 
\section{Newman's theorem (Bonus)}

The gap between the public-coin and private-coin randomized communication complexities of a function cannot be arbitrarily large.

\begin{theorem}
For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ and every constant $\epsilon > 0$,
\[
\Rpriv_{2\epsilon}(f) \le R_\epsilon(f) + O\left(\log \frac{n}{\epsilon^2}\right).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here, if you complete it.}
\end{proof}

\bigskip
\begin{remark}
\emph{Hint.} First consider what happens when you run a public-coin randomized protocol on $t = n/\epsilon^2$ random strings drawn independently at random. Then Chernoff bounds and the probabilistic method may be useful in completing the proof.
\end{remark}

\newpage 
\section{Private-coin randomness and determinism}

Interestingly (and unlike in the public-coin model), the gap between the private-coin randomized and deterministic communication complexities of a function also cannot be arbitrarily large.

\begin{theorem}
For every function $f : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$,
\[
\Rpriv(f) = \Omega( \log D(f)).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\bigskip
\begin{remark}
\emph{Hint.} It might be easiest to first consider a communication setting where Alice and Bob can send real numbers with cost $O(1)$ and show that in this setting there is a deterministic protocol that computes $f$ with cost $O(2^{\Rpriv(f)})$.
\end{remark}

\newpage 
\section{Equality VII}

We can use the results obtained in the previous sections to obtain tight bounds on the private-coin randomized communication complexity of the equality function.

\begin{theorem}
The private-coin randomized communication complexity of the $\Eq : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ function is
\[
\Rpriv(\Eq) = \Theta( \log n).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
Prove the upper bound $\Rpriv(\Eq) = O( \log n)$ directly without using Newman's theorem.
\end{exercise}

\end{document}

### 2019-02-01 CS860-CommunicationComplexity/assignments/Chapter4.tex
solved required problems

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+By the previous Theorem it suffices to show that $R(\Eq) = O(1)$.
+Let $x \in \{0,1\}^n$ be Alice's input and let $y \in \{0,1\}^n$ be Bob's.
+They generate two random vectors $r_1, r_2  \in \{0,1\}^n$ uniformly and independently at random.
+Then Alice computes $a_1 = x\cdot r_1$ (inner product mod $2$) and $a_2 = x\cdot r_2$ and send the two bits to Bob.
+Bob computes $b_1 = y\cdot r_1$ and $b_2 = y \cdot r_2$ and sends those two bits to Alice.
+They output $1$ if and only if $a_1 = b_1$ and $a_2 = b_2$.
+
+When $x = y$ they will always output $1$, thus we can bound the failure rate by the false positive rate.
+Suppose $x \neq y$.
+Then the probability that $a_1 = b_1$ is $\frac{1}{2}$ and the probability that $a_2 = b_2$ is $\frac{1}{2}$.
+Hence the total failure probability can be bounded by  $\frac{1}{2}\frac{1}{2} = \frac{1}{4} \leq \frac{1}{3}$.
+Thus $R(f) = O(1)$, as desired.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+We will solve Exercise $4$ instead.
+Let $\epsilon = \frac{1}{3\log n}$.
+The protocol proceeds by a binary search, using the Equality oracle we created in the previous problem.
+Let $x, y \in [2^n]$ be Alice and Bob's inputs respectively.
+At the start of the binary search the interval in question is $[1,n]$.
+At any arbitrary iteration of the binary search with interval $(\ell, u)$ (with $\ell < u$), Alice and Bob run their protocol for $\Eq$ on their substrings from $\ell$ to $\lfloor (\ell+u)/2 \rfloor$ with error $\epsilon$ costing $O(\log \frac{1}{\epsilon}) = O(\log\log n)$ communication.
+If the equality test passes then the next interval will be $[\lfloor (\ell+u)/2 \rfloor + 1, u]$, otherwise the next interval will be $[\ell, \lfloor (\ell+u)/2 \rfloor]$.
+On intervals with $\ell = u$, Alice and Bob send $x_\ell$ and $y_\ell$ to each other.
+If $x_\ell > y_\ell$ they return $1$, otherwise they return $0$.
+
+The protocol works since $\GT(x,y) = 1$ if and only $x_i > y_i$ where $i$ is the first bit on which $x$ and $y$ differ (if such a bit exists, otherwise $x=y$).
+Provided no $\Eq$ tests error, the binary search will successfully identify this bit $i$.
+Hence the probability of failure is bounded by the probability of an $\Eq$ test failing.
+Any one test will fail with probability $\epsilon$.
+There will be at most $\log(n)$ test (rounds of binary search), and thus the probability of failure is at most
+$$\log(n) \epsilon = \frac{1}{3}.$$
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Consider the following random protocol.
+Alice and Bob generate random vectors $r_1, \dots, r_t \in \{0,1\}^n.$ Alice computes $x\cdot r_1, \dots, x\cdot r_t$ (the inner product mod $2$), and Bob computes $y\cdot r_1, \dots, y \cdot r_t$, and they send the results to each other.
+
+Alice and Bob now each independently search for a vector $z \in \{0,1\}^n$ which satisfies for every $i\in[t]$,
+$$x\cdot r_i = y\cdot r_i \oplus z \cdot r_i$$
+and that $z$ has at most $k$ non-zero entries.
+
+When $|\{i \in [n]: x_i \neq y_i\}| \leq k$, they always output $1$ since 
+$$x\cdot r_i = (y \oplus (x\oplus y)) \cdot r_i = y\cdot r_i \oplus((x\oplus y) \cdot r_i) $$ 
+(i.e. they can both find $z= x\oplus y$). Thus we can bound the failure rate by the false positive rate.
+
+Let $z \in \{0,1\}^n$ such that $|z| \leq k$. Observe that when $x \neq y \oplus z$ the probability that 
+$$x\cdot r_i = y\cdot r_i \oplus z\cdot r_i$$
+for all $i \in [t]$ is at most $2^{-t}$. There's $O(n^k)$ ways to choose $z$, so by union bound the failure probability is at most
+$$O(n^k)2^{-t}.$$
+For any $\epsilon \in (0,1)$ if we want this failure to be at most $\epsilon$, so we can choose
+$$t = O(k \log n).$$
+
+If $n = \text{poly}\log(k)$ then this protocol with give the desired bound of $O(k \log k)$.
+Let $d = \text{poly}\log(k)$ chosen large enough (\emph{sorry I didn't make this part formal}).
+Alice and Bob can randomly partition their input indices, $[n]$, into $d$ subsets $S_1, \dots, S_d\subseteq [n]$.
+Then they can compute $x'=\{x\cdot \bold{1}_{S_d},\dots, x\cdot \bold{1}_{S_d}\}$ and $y'=\{y\cdot \bold{1}_{S_d},\dots, y\cdot \bold{1}_{S_d}\}$ and re-run the protocol above as if their inputs were $x'$ and $y'$.
+With high probability (\emph{not quite sure how to show this, but I think it might work}) the Hamming distance of $x$ from $y$ agrees with the Hamming distance of $x'$ from $y'$, and so the result holds.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a private coin randomized protocol attaining cost $\Rpriv(f)$.
+In the model given by the Hint, Alice and Bob can exchange their probabilities of ending up at each leaf of the protocol tree for $\pi$ at a communication cost of $O(2^{\Rpriv(f)})$.
+Once they known the leaves associated with an input pair $(x,y)$, they can output in their deterministic protocol the bit which is output on the majority of such leaves.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+To see that $\Rpriv(\Eq) = O(\log n)$ we apply Newman's Theorem to our previous result that $R(\Eq) = O(1)$.
+
+To see that $\Rpriv(\Eq) = \Omega(\log n)$ we apply Theorem 8 of this Chapter with Theorem $1$ of Chapter $2$ which said that $D(\Eq) = \Omega(n)$.
 \end{proof}

### 2019-02-01 CS860-CommunicationComplexity/assignments/Chapter4.tex
bonus

@@
 \begin{proof}
-\replacethistext{Enter your proof here, if you complete it.}
+Consider a public-coin randomized protocol $\pi$ attaining cost $R_\epsilon(f)$.
+Let $t = 2n/\epsilon^2$ and consider $\pi$ run on $t$ random strings $r_1, \dots, r_t$ drawn independently at random.
+Fix an input $(x,y) \in \calX \times \calY$.
+For any $i \in [t]$, the probability that $\pi$ fails when run on $r_i$ is at most $\epsilon$.
+Hence the probability that $(1-2\epsilon)t$ strings amongst $r_1, \dots, r_t$ fail when $\pi$ runs using that string is at most
+$$2^{-\Omega(\epsilon^2t)} = 2^{-\Omega(2n)}$$
+by Chernoff Bounds.
+
+Now, by the union bound, for any input in $\calX\times \calY$ the probability that greater than $2\epsilon t$ strings fail when $\pi$ runs them is strictly less than $1$.
+Thus, by the probabilistic method there exist strings $R = \{r_1^*, \dots, r_t^*\}$ with the property that for any input pair at most $2\epsilon t$ strings in $R$ fail when $\pi$ runs using that string.
+
+Our private coin protocol is as follows.
+Alice samples a string $r_i^* \in R$ uniformly at random, and sends its index $i$ to Bob.
+This costs $O(\log\frac{n}{\epsilon^2})$ communication.
+Now they can use $r_i^*$ as a source of shared randomness to run $\pi$ at additional communication cost $R_\epsilon(f)$.
+Since at most $2\epsilon t$ string in $R$ fail, the probability of failure for this protocol is at most
+$$2\epsilon t/t = 2\epsilon$$
+as desired.
 \end{proof}

### 2019-02-05 CS860-CommunicationComplexity/assignments/Chapter5.tex
Solved Problem 1

\documentclass[11pt]{amsart}
\usepackage{amsmath,amsthm,amsfonts,amssymb}
\usepackage{fullpage}
\usepackage[usenames]{xcolor}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{remark}{Remark}
\newtheorem{problem}{Problem}
\newtheorem{challenge}{Challenge Problem}
\newtheorem{open}{Open Problem}
\theoremstyle{plain}

\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calZ}{\mathcal{Z}}
\newcommand{\cost}{\mathrm{cost}}
\newcommand{\disc}{\mathrm{disc}}
\newcommand{\discu}{\mathrm{disc_u}}
\newcommand{\Disj}{\textsc{Disj}}
\newcommand{\Eq}{\textsc{Eq}}
\newcommand{\GT}{\textsc{GT}}
\newcommand{\HD}{\textsc{HD}}
\newcommand{\ICext}{\mathrm{IC}^{\mathrm{ext}}}
\newcommand{\ICint}{\mathrm{IC}^{\mathrm{int}}}
\newcommand{\icostext}{\mathrm{icost}^{\mathrm{ext}}}
\newcommand{\icostint}{\mathrm{icost}^{\mathrm{int}}}
\newcommand{\IP}{\textsc{IP}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\Rpriv}{R^{\mathrm{priv}}}
\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}

\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}
\DeclareMathOperator{\height}{height}

\DeclareMathOperator*{\E}{\mathbb{E}}

\title{Communication complexity: Chapter 5 \\ Information complexity}
\author{Eric Blais and Justin Toth}

\begin{document}

\maketitle

Communication complexity is concerned mostly with the \emph{minimum number of bits} that Alice and Bob need to exchange in order to compute some function on
their joint inputs. Information complexity, on the other hand, is concerned with
the \emph{minimum amount of information} contained in the bits that Alice and
Bob exchange to compute that function. Our first task is to formally define this notion of information. To do so, we will use standard definitions from information theory.

\begin{definition}[Entropy]
The \emph{entropy} of a random variable $Z$ drawn from the probability distribution $\mu$ over a finite set $\calZ$ is
\[
H(Z) = - \sum_{z \in \calZ} \mu(z) \log_2 \mu(z).
\]
The entropy of a sequence of random variables $Z_1,Z_2,\ldots,Z_k$, denoted $H(Z_1\,Z_2\cdots Z_k)$ is the entropy of the random variable $Z = (Z_1,Z_2,\ldots,Z_k)$.
\end{definition}

\begin{definition}[Conditional entropy]
The \emph{conditional} entropy of $Z$ given another random variable $Z'$ is
\[
H(Z \mid Z') = H(Z\,Z') - H(Z').
\]
\end{definition}

\begin{definition}[Mutual information]
The \emph{mutual information} of two random variables $Z$ and $Z'$ is
\[
I( Z ; Z' ) = H(Z) - H(Z \mid Z').
\]
The \emph{conditional mutual information} of $Z$ and $Z'$ given $W$ is
\[
I( Z ; Z' \mid W) = H(Z \mid W) - H(Z \mid Z'\, W).
\]
\end{definition}

We use the following basic properties of entropy and mutual information.

\begin{theorem}
Entropy satisfies the following properties:
\begin{description}
\item[Boundedness] For every $Z$ over a finite set $\calZ$, $0 \le H(Z) \le \log_2 |\calZ|$.
\item[Chain rule] $H(Z_1\,Z_2\cdots Z_k) = H(Z_1) + H(Z_2 \mid Z_1) + \cdots + H(Z_k \mid Z_1\cdots Z_{k-1})$.
\item[Subadditivity] $H(Z \mid Z') \le H(Z)$ and $H(Z\,Z') \le H(Z) + H(Z')$.
\end{description}
Mutual information satisfies the following properties:
\begin{description}
\item[Boundedness] $0 \le I(Z ; Z') \le \min\{ H(Z), H(Z') \}$.
\item[Chain rule] $I(Z_1\,Z_2 ; W) = I(Z_1 ; W) + H(Z_2 ; W \mid Z_1)$.
\item[Data processing inequality] Whenever $Z' = f(Z)$ is determined by $Z$, $I(Z' ; W) \le I(Z ; W)$.
\end{description}

\end{theorem}

\newpage 
\section{External information complexity}

Throughout this chapter, we will consider randomized protocols $\Pi$ that have access to both public- and private-coin randomness.

\begin{definition}[Transcript]
The \emph{transcript} of a protocol $\Pi$ on some input $(x,y) \in \calX \times \calY$, which we will denote by
\[
\Gamma_{x,y}^\Pi
\]
is a bit string that includes (i) the public-coin random string $R$ used by Alice and Bob and (ii) the sequence of bits that they exchange. 
\end{definition}

The first natural notion of information of a protocol that we will study is the amount of information that an external observer learns about Alice and Bob's input $(x,y)$ by seeing the transcript $\Gamma_{x,y}^\Pi$ of their communication protocol.

\begin{definition}[External information cost]
The \emph{external information cost} of a randomized protocol $\Pi$ over the distribution $\mu$ on $\calX \times \calY$ is
\[
\icostext_\mu(\Pi) = I\big( X\,Y ; \Gamma_{X,Y}^\Pi \big)
\]
where $(X,Y) \sim \mu$.
\end{definition}

\begin{definition}[External information complexity]
The \emph{$\epsilon$-error external information complexity} of the function $f : \calX \times \calY \to \{0,1\}$ with respect to the distribution $\mu$ is
\[
\ICext_{\mu,\epsilon}(f) = \inf_{\pi} \icostext_\mu(\pi)
\]
with the infimum taken over all randomized protocols that compute $f$ with error at most $\epsilon$.\footnote{Note that the error of a randomized protocol is still defined to be its maximum error probability over \emph{any} input in $\calX \times \calY$; it is \emph{not} the expected error over an input drawn from $\mu$.}
\end{definition}

The external information complexity of a function gives a lower bound on its randomized communication complexity.

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$ and every distribution $\mu$ on $\calX \times \calY$,
\[
\ICext_{\mu,\epsilon}(f) \le R_\epsilon(f).
\]
\end{theorem}

\begin{proof}

Let $\Pi$ be a distribution on protocols such that $\cost(\Pi) = R_\epsilon(f)$. That is,
$$R_\epsilon(f) = \max_{(x,y) \in \calX\times\calY}\max_{\pi \sim \Pi}\height(\pi(x,y))$$ 
where $\height(\pi(x,y))$ denotes the length of the path from the root to leaf of the protocol tree $\pi$ dictated by inputs $(x,y)$. We use $\height(\pi)$ to denote $\max_{(x,y) \in \calX\times\calY}\height(\pi(x,y))$. Let $\pi^*$ be the protocol tree achieving cost $\R_\epsilon(f)$. Thus we have
$$R_\epsilon(f) = \height(\pi^*) \geq \E_{(x,y) \sim \mu} [\height(\pi^*(x,y))].$$
Let $\Gamma^{\pi^*}_{X,Y}(r)$ denote the shared random string part of $\Gamma^{\pi^*}_{X,Y}$, and let $\Gamma^{\pi^*}_{X,Y}(c)$ denote the communication between Alice and Bob of the transcript. We claim that
$$\E_{(x,y) \sim \mu} [\height(\pi^*(x,y))] \geq H(\Gamma^{\pi^*}_{X,Y}(c)).$$
To see this claim is sufficient, observe that it implies 
\begin{align*}
R_\epsilon(f) &\geq H(\Gamma^{\pi^*}_{X,Y}(c)) \\
&\geq I(XY; \Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) \\
&= I(XY;\Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) + I(XY;\Gamma^{\pi^*}_{X,Y}(r))\\
&= I(\Gamma^{\pi^*}_{X,Y}; XY) \\
&= \icostext_\mu(\Gamma^{\pi^*}_{X,Y})\\
&\geq \ICext_{\mu,\epsilon}(f).
\end{align*}

Now we proceed to prove the claim. For ease of notation we will shorten $\height(\pi^*(x,y))$ to $h_{x,y}$. We have
\begin{align*}
\E_{(x,y)\sim \mu}[h_{x,y}] &=\sum_{(x,y)\in \calX\times\calY} \mu(x,y) h_{x,y} \\
&=\sum_{(x,y)\in \calX\times\calY} \mu(x,y)\left(\log(\frac{1}{\mu(x,y)})-\log(\frac{1}{\mu(x,y)}) +\log(2^{h_{x,y}})\right)\\
&=\sum_{(x,y)\in \calX\times\calY} \mu(x,y)\left(\log(\frac{1}{\mu(x,y)})-\log(\frac{2^{-h_{x,y}}}{\mu(x,y)}))\right) \\
&=H(\Gamma^{\pi^*}_{X,Y}(c)) - \sum_{(x,y)\in\calX\times\calY}\mu(x,y)\log(\frac{2^{-h_{x,y}}}{\mu(x,y)}) \\
&\geq H(\Gamma^{\pi^*}_{X,Y}(c)) - \log(\sum_{(x,y)\in\calX\times\calY} 2^{-h_{x,y}})
\end{align*}
where the last line follows by applying the convexity of $\log$. The sum inside the logarithm, $\sum_{(x,y)\in\calX\times\calY} 2^{-h_{x,y}}$ is at most $1$, completing the proof. This is a basic fact about binary trees. We'll give a brief proof for completeness.
Proceed by induction on $|\calX \times \calY|$. If $|\calX\times\calY| = 1$, the result is immediate. For induction suppose the result holds for binary trees smaller that $\pi^*$. Let $r$ be the root of $\pi^*$. Split $\pi^*$ into two subtrees $T_1$ and $T_2$ by deleting $r$. Now observe that
$$\sum_{(x,y)\in \calX\times\calY} 2^{-h{x,y}} = \frac{1}{2}\sum_{(x,y) \in \calX\times\calY \cap T_1} + \frac{1}{2}\sum_{(x,y) \in \calX\times\calY \cap T_2} \leq \frac{1}{2}+\frac{1}{2} = 1.$$
With the first equality following since joining $T_1$ and $T_2$ by root $r$, adds $1$ to length of the path to each leaf, and the inequality follows by induction.
\end{proof}

\newpage 
\section{Public randomness can be eliminated}

When designing protocols to obtain upper bounds on the information complexity of functions, it is convenient to work in the framework where Alice and Bob can use both public- and private-coin randomness. For lower bounds, however, it is useful to note that without loss of generality we can consider protocols that use only private randomness.

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$, every distribution $\mu$ over $\calX \times \calY$, and every $0 \le \epsilon \le \frac12$,
\[
\ICext_{\mu,\epsilon}(f) = \inf_{\pi} \icostext_\mu(\pi)
\]
even when the infimum is taken only over private-coin randomized protocols that compute $f$ with error at most $\epsilon$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Equality VIII}

External information complexity can be used to give a simple proof of the $\Omega(n)$ lower bound for the $0$-error randomized communication complexity of the equality function.

\begin{theorem}
Let $\mu$ be the uniform distribution on the set $\{(x,x) : x \in \{0,1\}^n\}$. The $0$-error information complexity of $\Eq : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ with respect to $\mu$ is
\[
\ICext_{\mu,0}(\Eq) = n.
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{And}

Let $\textsc{And} : \{0,1\} \times \{0,1\} \to \{0,1\}$ be the and function $\textsc{And}(x,y) = x \wedge y$ that takes the value $1$ if and only if $x = y = 1$.

\begin{theorem}
For every distribution $\mu$ on $\{0,1\} \times \{0,1\}$,
\[
\ICext_{\mu,0}(\textsc{And}) \le \log_2 3.
\]
Furthermore, this bound is tight when $\mu(0,1) = \mu(1,0) = \mu(1,1) = \frac13$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Internal information complexity}

Another natural notion of information of a protocol is the amount of information that Alice and Bob learn about each other's inputs by running their communication protocol.

\begin{definition}[Internal information cost]
The \emph{internal information cost} of a randomized protocol $\Pi$ over the distribution $\mu$ on $\calX \times \calY$ is
\[
\icostint_\mu(\Pi) = I\big( X ; \Gamma_{X,Y}^\Pi \, Y\big) + I\big( Y ; \Gamma_{X,Y}^\Pi \, X\big).
\]
where $(X,Y) \sim \mu$.
\end{definition}

\begin{definition}[Internal information complexity]
The \emph{$\epsilon$-error internal information complexity} of the function $f : \calX \times \calY \to \{0,1\}$ with respect to the distribution $\mu$ is
\[
\ICint_{\mu,\epsilon}(f) = \inf_{\pi} \icostint_\mu(\pi)
\]
with the infimum taken over all randomized protocols that compute $f$ with error at most $\epsilon$.
\end{definition}

The internal information complexity of a function gives a lower bound on its external information complexity (and therefore on its randomized communication complexity as well).

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$, every distribution $\mu$ on $\calX \times \calY$, and every $0 \le \epsilon \le \frac12$,
\[
\ICint_{\mu,\epsilon}(f) \le \ICext_{\mu,\epsilon}(f).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Equality VIII}

The internal information complexity of a function can be much smaller than its external information complexity, as the following example shows.

\begin{theorem}
For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$, the $0$-error information complexity of $\Eq : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ is 
\[
\ICint_{\mu,0}(\Eq) = O(1).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\bigskip
\noindent \emph{Hint.} Consider a protocol where Alice and Bob use their public randomness to select $n$ linearly independent vectors $r_1,\ldots,r_n \in \{0,1\}^n$.

\newpage 
\section{Direct sum (Bonus)}

Given a function $f : \calX \times \calY \to \{0,1\}$, define $f^k : \calX^k \times \calY^k \to \{0,1\}^n$ to be the function where
\[
f^n(x^{(1)},\ldots,x^{(k)},y^{(1)},\ldots,y^{(k)}) = \big( f(x^{(1)},y^{(1)}), \ldots f(x^{(k)}, y^{(k)})\big).
\]
This corresponds to the setting where Alice and Bob must compute the value of $f$ on $n$ different pairs of inputs instead of just $1$.

It might seem obvious that the complexity of computing $f^k$ is always $k$ times the complexity of computing $f$, but this is false in general! One counter-example is given by the equality function. Since $R_\epsilon(\Eq) = O(\log \frac1\epsilon)$, we have that $R_{2^{-\sqrt{k}}}(\Eq) = \sqrt{k}$ and so we might expect that $R_{2^{-\sqrt{k}}}(\Eq^k) = k^{3/2}$. In fact, the true randomized complexity of this function is much smaller.

\begin{theorem}
$R_{2^{-\sqrt{k}}}(\Eq^k) = O(k)$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here, if you complete it.}
\end{proof}

\newpage 
\section{Direct sum for information complexity}

One of the main advantages of working with internal information complexity is that in this setting every function \emph{does} satisfy the intuition that computing $k$ copies of a function is exactly $k$ times harder than computing a single copy of it.
\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$ and every $k \ge 1$,
\[
\ICint_{\mu,0}(f^k) = k \cdot \ICint_{\mu,0}(f).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\bigskip
\noindent \emph{Hint.} See Piazza for a hint on this question (and feel free to start a discussion thread there as well if you have some ideas but only a partial proof).

\newpage 
\section{Set disjointness III}

It is possible to obtain tight bounds on the internal information complexity of the \textsc{And} function. These tight bounds have been used to obtain \emph{exact} bounds on the communication complexity of the set disjointness function. To establish the tight asymptotic bound on its randomized communication complexity, however, all we need is the following result.

\begin{lemma}
Let $\mu$ be the distribution on $\{0,1\} \times \{0,1\}$ defined by $\mu(0,0) = \mu(0,1) = \mu(1,0) = \frac13$.
For every $\epsilon > 0$, there is a constant $c_\epsilon > 0$ for which
\[
\ICint_{\mu,\epsilon}(\textsc{And}) = c_\epsilon.
\]
\end{lemma}

This result and (extensions of) the direct sum property of internal information complexity can be used to bound the randomized communication complexity of the set disjointness function.

\begin{theorem}
For every (small enough) $\epsilon \ge 0$, 
\[
\R_\epsilon(\Disj) = \Omega(n).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\bigskip
\noindent \emph{Hint.} The ideas for the proof of the theorem in the last section might be helpful. Refer to Piazza for discussions, questions, or additional hints.

\end{document}

### 2019-02-05 CS860-CommunicationComplexity/assignments/Chapter5.tex
important fix to problem 1

@@
 
-Now we proceed to prove the claim. For ease of notation we will shorten $\height(\pi^*(x,y))$ to $h_{x,y}$. We have
+Now we proceed to prove the claim. Let $L$ be the set of leaves of the protocol tree $\pi^*$. We define a distribution $\mu'$ on $L$ as follows: for each $\ell \in L$, we let
+ $$\mu'(\ell = \sum_{(x,y) \in \calX\times\calY : (x,y)\text{ induces a path to } \ell} \mu(x,y).$$ 
+ Let $h_\ell = \height(x,y)$ where $(x,y)$ induces a path to $\ell$. i.e. $h_\ell$ is the lenght of the path from the root of $\pi^*$ to $\ell$ (this path is taken on input $(x,y)$). We have
 \begin{align*}
-\E_{(x,y)\sim \mu}[h_{x,y}] &=\sum_{(x,y)\in \calX\times\calY} \mu(x,y) h_{x,y} \\
-&=\sum_{(x,y)\in \calX\times\calY} \mu(x,y)\left(\log(\frac{1}{\mu(x,y)})-\log(\frac{1}{\mu(x,y)}) +\log(2^{h_{x,y}})\right)\\
-&=\sum_{(x,y)\in \calX\times\calY} \mu(x,y)\left(\log(\frac{1}{\mu(x,y)})-\log(\frac{2^{-h_{x,y}}}{\mu(x,y)}))\right) \\
-&=H(\Gamma^{\pi^*}_{X,Y}(c)) - \sum_{(x,y)\in\calX\times\calY}\mu(x,y)\log(\frac{2^{-h_{x,y}}}{\mu(x,y)}) \\
-&\geq H(\Gamma^{\pi^*}_{X,Y}(c)) - \log(\sum_{(x,y)\in\calX\times\calY} 2^{-h_{x,y}})
+\E_{(x,y)\sim \mu}[h_{x,y}] &=\sum_{\ell \in L} \mu'(\ell) h_\ell \\
+&=\sum_{\ell \in L} \mu'(\ell)\left(\log(\frac{1}{\mu'(\ell)})-\log(\frac{1}{\mu'(\ell)}) +\log(2^{h_\ell})\right)\\
+&=\sum_{\ell\in L} \mu'(\ell)\left(\log(\frac{1}{\mu'(\ell)})-\log(\frac{2^{-h_\ell}}{\mu'(\ell)}))\right) \\
+&=H(\Gamma^{\pi^*}_{X,Y}(c)) - \sum_{\ell\in L}\mu'(\ell)\log(\frac{2^{-h_\ell}}{\mu'(\ell)}) \\
+&\geq H(\Gamma^{\pi^*}_{X,Y}(c)) - \log(\sum_{\ell\in L} 2^{-h_\ell})
 \end{align*}
 where the last line follows by applying the convexity of $\log$.
-The sum inside the logarithm, $\sum_{(x,y)\in\calX\times\calY} 2^{-h_{x,y}}$ is at most $1$, completing the proof.
+The sum inside the logarithm, $\sum_{\ell \in L} 2^{-h_\ell}$ is at most $1$, completing the proof.
 This is a basic fact about binary trees.
 We'll give a brief proof for completeness.
-Proceed by induction on $|\calX \times \calY|$.
-If $|\calX\times\calY| = 1$, the result is immediate.
+Proceed by induction on $|L|$.
+If $|L| = 1$, the result is immediate.
 For induction suppose the result holds for binary trees smaller that $\pi^*$.
@@
 Now observe that
-$$\sum_{(x,y)\in \calX\times\calY} 2^{-h{x,y}} = \frac{1}{2}\sum_{(x,y) \in \calX\times\calY \cap T_1} + \frac{1}{2}\sum_{(x,y) \in \calX\times\calY \cap T_2} \leq \frac{1}{2}+\frac{1}{2} = 1.$$
+$$\sum_{\ell \in L} 2^{-h_\ell} = \frac{1}{2}\sum_{\ell \in L \cap T_1} + \frac{1}{2}\sum_{\ell \in L \cap T_2} \leq \frac{1}{2}+\frac{1}{2} = 1.$$
 With the first equality following since joining $T_1$ and $T_2$ by root $r$, adds $1$ to length of the path to each leaf, and the inequality follows by induction.

### 2019-02-06 CS860-CommunicationComplexity/assignments/Chapter5.tex
Fixed p1

@@
 &\geq I(XY; \Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) \\
-&= I(XY;\Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) + I(XY;\Gamma^{\pi^*}_{X,Y}(r))\\
-&= I(\Gamma^{\pi^*}_{X,Y}; XY) \\
+&= I(XY ;\Gamma^{\pi^*}_{X,Y}) \\
 &= \icostext_\mu(\Gamma^{\pi^*}_{X,Y})\\
@@
 Now we proceed to prove the claim. Let $L$ be the set of leaves of the protocol tree $\pi^*$. We define a distribution $\mu'$ on $L$ as follows: for each $\ell \in L$, we let
- $$\mu'(\ell = \sum_{(x,y) \in \calX\times\calY : (x,y)\text{ induces a path to } \ell} \mu(x,y).$$ 
- Let $h_\ell = \height(x,y)$ where $(x,y)$ induces a path to $\ell$. i.e. $h_\ell$ is the lenght of the path from the root of $\pi^*$ to $\ell$ (this path is taken on input $(x,y)$). We have
+ $$\mu'(\ell) = \sum_{(x,y) \in \calX\times\calY : (x,y)\text{ induces a path to } \ell} \mu(x,y).$$ 
+ Let $h_\ell = \height(x,y)$ where $(x,y)$ induces a path to $\ell$. i.e. $h_\ell$ is the length of the path from the root of $\pi^*$ to $\ell$ (this path is taken on input $(x,y)$). We have
 \begin{align*}

### 2019-02-06 CS860-CommunicationComplexity/assignments/Chapter5.tex
Solved up to 5.

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Since infimum is taken over only private-coin randomized protocols, as opposed to all randomized protocols, it is immediate that
+$$\ICext_{\mu,\epsilon}(f) \leq \inf_{\pi} \icostext_\mu(\pi).$$
+To complete the proof, consider any randomized protocol $\pi$ which computes $f$ with error at most $\epsilon$.
+We will show the existence of $\pi'$, a private coin randomized protocol, which computes $f$ with error at most $\epsilon$ and satisfying
+$$\icostext_\mu(\pi) = \icostext_\mu(\pi').$$
+Let $\chi$ be the distribution on shared random bit strings used by $\pi$.
+The protocol $\pi'$ operates as follows.
+First  Alice draws a random string $R$ according to $\chi$, and communicates that string to Bob.
+Then Alice and Bob simulate $\pi$ using $R$ as their shared randomness.
+
+Since protocol transcripts do not record private random bits, to the external observer the transcripts for $\pi$ and $\pi'$ are indistinguishable.
+In fact, $\Gamma_{X,Y}^\pi$ and $\Gamma_{X,Y}^{\pi'}$ are identically distributed when $X,Y \sim \mu$ and $R\sim \chi$.
+Therefore
+$$\icostext_\mu(\pi) = I(XY;\Gamma^{pi}_{XY}) = I(XY;\Gamma^{\pi'}_{XY}) = \icostext_\mu(\pi')$$
+as desired.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a private coin randomized protocol computing $\Eq$ with error $0$. Using the previous result, it suffices to prove
+$$\icostext_\mu(\pi) = n.$$
+We have,
+$$\icostext_\mu(\pi) = I(XY, \Gamma_{X,Y}^\pi) = H(XY) - H(XY \mid \Gamma_{X,Y}^\pi).$$
+Since $XY$ are uniformly distributed over $2^n$ elements,
+$$H(XY)= \log(2^n) = n.$$
+Now we just need to show $H(XY \mid \Gamma_{X,Y}^\pi) = 0.$ Indeed, this holds since we claim each transcript uniquely identifies the input pair $XY$.
+In other words, $\Pr[XY\mid \Gamma_{X,Y}^\pi = \gamma]$ is either $0$ or $1$, and thus
+$$H(XY\mid \Gamma_{X,Y}^\pi) = 0.$$
+To see the claim, suppose that there is a transcript of $\pi$ which is identical for distinct inputs $(x,x)$ and $(x',x')$ drawn from $\mu$, when Alice and Bob have the same random private strings $R_A$ and $R_B$ respectively.
+Since $\pi$, makes $0$ errors, it outputs $1$ on inputs $(x,x)$ and $(x',x')$.
+But the transcript would also be identical for input $(x,x')$ on random strings $R_A, R_B$, and so $\pi$ would still output $1$ on this input.
+This contradicts that $\pi$ makes $0$ errors.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Consider the protocol $\pi$, where Alice sends Bob the bit, $x$, she receives, and Bob sends Alice the bit he receives if and only if $x=1$.
+If $x=0$ the protocol $\pi$ outputs $0$, and otherwise they exchanged bits, so the protocol can output $x\wedge y$.
+Clearly this is a $0$-error protocol for $\textsc{And}$.
+
+Since 
+$$\icostext_\mu(\pi) = I(XY; \Gamma^\pi_{X,Y}) \leq H(\Gamma^\pi_{X,Y}) \leq \log_2(3)$$
+the result holds. The last inequality follows since there are precisely three different transcripts: $0$, $10$, and $11$ which can occur as a result of running $\pi$.
+
+Now fix $\mu$ so $\mu(0,1) = \mu(1,0) = \mu(1,1) = \frac{1}{3}$.
+Let $\pi$ be any random protocol with computes $\textsc{And}$ without error.
+Then, at the end of communication, for any input $(x,y)$, Alice knows $x$ and $x\wedge y$.
+Since $\mu(00) = 0$, Alice can uniquely recover $y$ from this information.
+Symmetrically, Bob also knows $x$ at the end of communication.
+In other words, knowing the transcript suffices to learn both $x$ and $y$ under this distribution $\mu$.
+Hence
+$$H(XY \mid \Gamma^\pi_{X,Y}) = 0.$$
+Thus
+$$\icostext_\mu(\pi) = I(XY;\Gamma^{\pi}_{X,Y}) = H(XY) =\log_2(3)$$
+with the last equality following since $XY\sim \mu$ is a uniform distribution on $3$ elements.
 \end{proof}

### 2019-02-07 CS860-CommunicationComplexity/assignments/Chapter5.tex
Solved section 5

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+We may assume that every protocol with computes $f$ always achieves its worst-case communication cost on every input and random string, by simply padding communication with $0$s on inputs where this is not the case.
+We will show that all randomized protocols, $\pi$, which compute $f$ with error at most $\epsilon$ satisfy
+$$\icostext_{\mu,\epsilon}(\pi) \geq \icostint_{\mu,\epsilon}(\pi).$$
+Let $k$ be the length of the transcript for $\pi$.
+Use $\Gamma^\pi_{X,Y}(i)$ to denote the $i$-th bit communicated in the transcript, for $i=1,\dots, k$ and $\Gamma^\pi_{X,Y}[i,j]$ to denote the subtranscript from the $i$-th bit to the $j$-th bit, inclusive.
+The $i$-th bit was communicated by one player to the other player.
+The player sending the bit learns nothing by doing so.
+Therefore we have either
+$$I(X;\Gamma^\pi_{X,Y}(i) \mid \Gamma^{\pi}_{X,Y}[1,i-1], Y) = 0 \quad\text{or}\quad I(Y;\Gamma^\pi_{X,Y}(i) \mid \Gamma^{\pi}_{X,Y}[1,i-1], X) = 0 $$
+for all $i\in[k]$.
+Now we commute
+\begin{align*}
+\icostext_{\mu,\epsilon}(\pi) &= \sum_{i=1}^k I(XY;\Gamma^\pi_{X,Y}(i) \mid \Gamma^\pi_{X,Y}[1,i-1]) \\
+&\geq \sum_{i=1}^k\max\{I(X;\Gamma^\pi_{X,Y}(i) \mid \Gamma^{\pi}_{X,Y}[1,i-1], Y), I(Y;\Gamma^\pi_{X,Y}(i) \mid \Gamma^{\pi}_{X,Y}[1,i-1], X)\}\\
+&=\sum_{i=1}^k I(X;\Gamma^\pi_{X,Y}(i) \mid \Gamma^{\pi}_{X,Y}[1,i-1], Y)+ I(Y;\Gamma^\pi_{X,Y}(i) \mid \Gamma^{\pi}_{X,Y}[1,i-1], X) \\
+&= I(X;\Gamma^\pi_{X,Y} \mid Y) + I(Y; \Gamma^\pi_{X,Y} \mid X) \\
+&= \icostint_{\mu,\epsilon}(\pi).
+\end{align*}
+The second inequality above follows by our observation that one of the terms in each max is $0$. Every other manipulation is repeated application of the chain rule.
 \end{proof}

### 2019-02-07 CS860-CommunicationComplexity/assignments/Chapter5.tex
chapter 5 progress

@@
 
-\begin{theorem}
+\begin{theorem}\label{th:2}
 For every function $f : \calX \times \calY \to \{0,1\}$ and every distribution $\mu$ on $\calX \times \calY$,
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a protocol where Alice, with input $x$, and Bob, with input $y$, sample $n$ linearly independent vectors $r_1, \dots, r_n \in \{0,1\}^n$ uniformly at random using their shared randomness, and then do the following: for $i=1,\dots n$ do:
+\begin{enumerate}
+	\item Alice sends Bob $a_i:=x\cdot r_i$.
+	\item Bob sends Alice $b_i:=y\cdot r_i$.
+	\item If $a_i \neq b_i$ they stop immediately and output $0$.
+	\item Otherwise, if it is the last iteration they output $1$.
+\end{enumerate} 
+This protocol is $0$-error.
+Since $r_1,\dots, r_n$ are $n$ linearly independent vectors, they form a basis of $\{0,1\}^n$.
+Therefore $x\neq y$ if and only if there exists some $i\in[n]$ such that $x\cdot r_i \neq y\cdot r_i$.
+When $x=y$ they communicate $n$ bits, but it is interesting to see what happens when $x\neq y$.
+In this case, we want to find the expected number of bits communicated.
+In each round that occurs $2$ bits are sent.
+A round occurs with the following probability,
+$$\Pr[a_j = b_j, \forall j=1,\dots, i-1]=\sum_{i=1}^n\frac{2^{n-i+1}-1}{2^n-1}$$
+since is equivalent to saying that $x-y$ (with addition taken in $\mathbb{F}_2$) is in the $n-i+1$-dimensional subspace formed by the kernel of $r_1,\dots,r_{i-1}$.
+Since $x-y\neq 0$ there are $2^n-1$ choices for $x-y$, and there $2^{n-i+1}-1$ choices which lie in the kernel.
+Since our basis is chosen uniformly at random, this yields the desired probability.
+Hence we have that the expected number of bits communicated is
+$$\sum_{i=1}^n 2\cdot\Pr[a_j = b_j, \forall j=1,\cdots, i-1] = 2\sum_{i=1}^n\frac{2^{n-i+1}-1}{2^n-1}\leq 2\sum_{i=1}^n\frac{2^{n-i+1}}{2^n} \leq 2\cdot 2 = O(1).$$
+Let $\Delta$ be the random indicator variable for $X=Y$. We now compute $I(X;\Gamma^\pi_{X,Y}\mid Y)$,
+\begin{align*}
+I(X;\Gamma^\pi_{X,Y} \mid Y) &= I(X\Delta;\Gamma^\pi_{X,Y}\mid Y) &\text{(since $\Gamma^\pi_{X,Y}$ suffices to learn $\Delta$)} \\
+	&= I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta) + I(\Delta;\Gamma^\pi_{X,Y}\mid Y) \\
+	&\leq  I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta) + 1  &\text{(since $I(\Delta;\Gamma^\pi_{X,Y}\mid Y) \leq H(\Delta) = 1$)}\\
+	&\leq  I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta=1) +  I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta=0) + 1 \\
+	&=  I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta=0) + 1 \\
+	&= O(1).
+\end{align*}
+The second last equality follows since $I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta=1)=0$, as once we know $Y$ and that $X=Y$, we know everything.
+The last equality follows from two facts.
+Firstly our prior observation that the expected number of bits communicated by $\pi$ when $X\neq Y$ is $O(1)$, and the second that $ I(X;\Gamma^\pi_{X,Y}\mid Y, \Delta=0) $ is at most that expected quantity.
+That second fact follows from our proof of Theorem \ref{th:2}, where be actually bound information cost by the expected communication length, as opposed to the worst-case length.
+
+By a symmetric argument $I(X;\Gamma^\pi_{X,Y} \mid Y) = O(1)$, and hence
+$$\icostint_{\mu,0}(\pi) = O(1) + O(1) = O(1)$$
+so $\ICint_{\mu,0}(\Eq) = O(1).$
 \end{proof}

### 2019-02-07 CS860-CommunicationComplexity/assignments/Chapter5.tex
Section 8

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Clearly $\ICint_{\mu,0}(f^k) \leq k \cdot \ICint_{\mu,0}(f)$ since we can simply run any protocol for $f$ on each coordinate of input for $f^k$, and the information cost will $k$ times that of the protocol.
+
+Let $\pi$ be a $0$-error protocol for $f^k$. We want to construct a $0$-error protocol $\pi'$ for $f$ with
+$$\icostint_{\mu,0}(\pi') = \frac{1}{k}\icostint_{\mu,0}(\pi).$$
+Protocol $\pi'$ operates as follows.
+Let $x$ be Alice's input, and $y$ be Bob's input.
+Alice and Bob use shared randomness to choose $i \in [k]$ uniformly at random.
+They set $(x^i,y^i) = (x,y)$.
+Let $\mu_\calX$ be the marginal distribution over $\calX$ from $\mu$, and define $\mu_\calY$ analogously.
+Alice and Bob use public randomness to sample $x^1,\dots,x^{i-1}$ according to $\mu_\calX$, and $y^{i+1},\dots, y^k$ according to $\mu_\calY$.
+Now Alice uses private randomness to sample $x^{i+1}, \dots, x^{k}$ according to $\mu$ conditioned on $y^{i+1},\dots, y^k$ respectively (so we get valid input pairs).
+Similarly Bob uses private randomness to sample $y^{1}, \dots, y^{i-1}$ according to $\mu$ conditioned on $x^{1},\dots, x^{i-1}$ respectively.
+Now Alice and Bob run protocol $\pi$ on inputs $(x^1,\dots, x^k)$ and $(y^1,\dots, y^k)$.
+The protocol outputs $f(x^i,y^i)$.
+
+We've taken care to make sure the input to $\pi$ is distributed according to $\mu$ when we call it, and furthermore since $\pi$ is $0$-error for $f^k$, $\pi'$ is $0$-error for $f$.
+Let $X,Y$ be the random variables for inputs to $\pi'$.
+Let $I$ be a random variable for the random index in $[k]$ that $\pi'$ chooses.
+Let $X_1,\dots, X_n$ be random variables for $x$ coordinates, and $Y_1,\dots, Y_n$ be random variables for $y$ coordinates for $\pi$'s black-box call to $\pi'$.
+Use shorthand $\bar{X} = (X_1,\dots, X_k)$ and $\bar{Y} = (Y_1,\dots, Y_k)$.
+We now compute the mutual information from Alice's perspective:
+\begin{align*}
+I(Y;\Gamma^{\pi'}_{X,Y}\mid X) &= I(Y_I;\Gamma^\pi_{\bar{X},\bar{Y}} \mid I, \bar{X}, Y_{I+1} \dots Y_{k}) \\
+&= \frac{1}{k}\sum_{i=1}^kI(Y_i; \Gamma^\pi_{\bar{X},\bar{Y}}\mid I=i, \bar{X}, Y_{i+1}\dots Y_{k}) \\
+&= \frac{1}{k}\sum_{i=1}^k \left(I(Y_i,Y_{i+1}; \Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{X}, Y_{i+2}\dots Y_{k}) - I(Y_{i+1}; \Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{X}, Y_{i+2}\dots Y_k)\right) \\
+&= \frac{1}{k}I(Y_1\dots Y_k ; \Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{X}) \\
+&= \frac{1}{k}I(\bar{Y};\Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{X}).
+\end{align*}
+In first equality follows by conditioning on what Alice knows publicly.
+The second by expanding the expectation.
+The third by the chain rule, and the fourth by canceling terms.
+By a symmetric argument we can see that
+$$I(X;\Gamma^{\pi'}_{X,Y}\mid Y) = \frac{1}{k}I(\bar{X};\Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{Y}).$$
+Thus we have
+$$\icostint_{\mu,0}(\pi') = I(Y;\Gamma^{\pi'}_{X,Y}\mid X) + I(X;\Gamma^{\pi'}_{X,Y}\mid Y) = \frac{1}{k}(I(\bar{Y};\Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{X}) + I(\bar{X};\Gamma^\pi_{\bar{X},\bar{Y}}\mid \bar{Y})) =  \frac{1}{k}\icostint_{\mu,0}(\pi)$$
+as desired.
 \end{proof}
@@
 \[
-\R_\epsilon(\Disj) = \Omega(n).
+R_\epsilon(\Disj) = \Omega(n).
 \]

### 2019-02-08 CS860-CommunicationComplexity/assignments/Chapter5.tex

@@
 \item[Boundedness] $0 \le I(Z ; Z') \le \min\{ H(Z), H(Z') \}$.
-\item[Chain rule] $I(Z_1\,Z_2 ; W) = I(Z_1 ; W) + H(Z_2 ; W \mid Z_1)$.
+\item[Chain rule] $I(Z_1\,Z_2 ; W) = I(Z_1 ; W) + I(Z_2 ; W \mid Z_1)$.
 \item[Data processing inequality] Whenever $Z' = f(Z)$ is determined by $Z$, $I(Z' ; W) \le I(Z ; W)$.
@@
 \[
-\icostint_\mu(\Pi) = I\big( X ; \Gamma_{X,Y}^\Pi \, Y\big) + I\big( Y ; \Gamma_{X,Y}^\Pi \, X\big).
+\icostint_\mu(\Pi) = I\big( X ; \Gamma_{X,Y}^\Pi\mid \, Y\big) + I\big( Y ; \Gamma_{X,Y}^\Pi \mid\, X\big).
 \]
@@
 
-\begin{theorem}
+\begin{theorem}\label{th:6}
 For every function $f : \calX \times \calY \to \{0,1\}$, every distribution $\mu$ on $\calX \times \calY$, and every $0 \le \epsilon \le \frac12$,
@@
 
-Given a function $f : \calX \times \calY \to \{0,1\}$, define $f^k : \calX^k \times \calY^k \to \{0,1\}^n$ to be the function where
+Given a function $f : \calX \times \calY \to \{0,1\}$, define $f^k : \calX^k \times \calY^k \to \{0,1\}^k$ to be the function where
 \[
-f^n(x^{(1)},\ldots,x^{(k)},y^{(1)},\ldots,y^{(k)}) = \big( f(x^{(1)},y^{(1)}), \ldots f(x^{(k)}, y^{(k)})\big).
+f^k(x^{(1)},\ldots,x^{(k)},y^{(1)},\ldots,y^{(k)}) = \big( f(x^{(1)},y^{(1)}), \ldots f(x^{(k)}, y^{(k)})\big).
 \]
@@
 One of the main advantages of working with internal information complexity is that in this setting every function \emph{does} satisfy the intuition that computing $k$ copies of a function is exactly $k$ times harder than computing a single copy of it.
-\begin{theorem}
+\begin{theorem}\label{th:9}
 For every function $f : \calX \times \calY \to \{0,1\}$ and every $k \ge 1$,
@@
 
-\begin{lemma}
+\begin{lemma}\label{lemma:1}
 Let $\mu$ be the distribution on $\{0,1\} \times \{0,1\}$ defined by $\mu(0,0) = \mu(0,1) = \mu(1,0) = \frac13$.
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $n=k$. Let $\pi$ be a communication protocol which computes $\Disj$ with error at most $\epsilon$, and achieves information cost
+$$\ICint_{\mu,\epsilon}$$ . By Theorem \ref{th:2} and Theorem \ref{th:6},
+$$R_\epsilon(\Disj) \geq \ICint_{\mu,\epsilon}(\Disj) = \icostint_{\mu,\epsilon}(\pi).$$
+Notice that when we sample $(x^1,y^1), \dots, (x^k,y^k)$ according to $\mu$, then
+$$\textsc{AND}^k((x^1,\dots,x^k),(y^1,\dots,y^k)) = \textsc{AND}(x^i,y^i) = 0$$
+for any $i \in [k]$.
+Thus we can adapt any $\epsilon$-error protocol, $\rho$, for $\textsc{AND}^k$ to an $\epsilon$-error protocol, $\rho'$, for $\textsc{AND}$ in the same way that we did during the proof of Theorem \ref{th:9}, and consequently
+$$\icostint_{\mu,\epsilon}(\rho') = \frac{1}{k}\icostint_{\mu,\epsilon}(\rho).$$
+Actually our observation about $\textsc{AND}^k$ let's us say something stronger.
+We can use a protocol for $\Disj$ as a stand-in for a protocol for $\textsc{AND}^k$, since (when input is sampled according to $\mu$)
+$$\Disj((x^1,\dots, x^k),(y^1,\dots,y^k)) = (x^1\wedge y^1) \lor\dots\lor(x^k\wedge y^k) = x^i\wedge y^i = \textsc{AND}^k((x^1,\dots, x^k),(y^1,\dots,y^k))$$
+for any $i\in[k]$. Therefore we can adapt $\pi$ to a protocol for $\textsc{AND}$, $\pi'$ satisfying
+$$\icostint_{\mu,\epsilon}(\pi') = \frac{1}{k}\icostint_{\mu,\epsilon}(\pi).$$
+Thus using Lemma \ref{lemma:1},
+$$\icostint_{\mu,\epsilon}(\pi) = k\icostint_{\mu,\epsilon}(\pi')\geq k\ICint_{\mu,\epsilon}(\textsc{AND}) = kc_\epsilon = \Omega(k).$$
+Combining with our first observation we see that
+$$R_\epsilon(\Disj) = \Omega(k)$$
+as desired.
 \end{proof}

### 2019-02-12 CS860-CommunicationComplexity/assignments/Chapter6.tex
Added Ch6

\documentclass[11pt]{amsart}
\usepackage{amsmath,amsthm,amsfonts,amssymb}
\usepackage{fullpage}
\usepackage[usenames]{xcolor}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{remark}{Remark}
\newtheorem{problem}{Problem}
\newtheorem{challenge}{Challenge Problem}
\newtheorem{open}{Open Problem}
\theoremstyle{plain}

\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calZ}{\mathcal{Z}}
\newcommand{\cost}{\mathrm{cost}}
\newcommand{\dHam}{\mathrm{d}_{\mathrm{Ham}}}
\newcommand{\disc}{\mathrm{disc}}
\newcommand{\discu}{\mathrm{disc_u}}
\newcommand{\Disj}{\textsc{Disj}}
\newcommand{\Eq}{\textsc{Eq}}
\newcommand{\GHD}{\textsc{GHD}}
\newcommand{\GT}{\textsc{GT}}
\newcommand{\HD}{\textsc{HD}}
\newcommand{\ICext}{\mathrm{IC}^{\mathrm{ext}}}
\newcommand{\ICint}{\mathrm{IC}^{\mathrm{int}}}
\newcommand{\icostext}{\mathrm{icost}^{\mathrm{ext}}}
\newcommand{\icostint}{\mathrm{icost}^{\mathrm{int}}}
\newcommand{\IP}{\textsc{IP}}
\newcommand{\Index}{\textsc{Index}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\Rpriv}{R^{\mathrm{priv}}}
\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}

\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}

\title{Communication complexity: Chapter 6 \\ One-way communication and simultaneous message passing}
\author{Eric Blais and Justin Toth}

\begin{document}

\maketitle

Every communication complexity question we have studied so far in this course have been set in the model where there are two players, Alice and Bob, who exchange bits with each other to compute functions on their joint inputs. This is of course not the only model of communication that is of interest---other models where we change the number of parties or restrict how messages can be sent between them have also been studied extensively. In this chapter, we examine the communication complexity of functions when the communication between Alice and Bob is restricted.

\newpage 
\section{One-way communication complexity}

The first modification to the standard communication complexity setting that we will consider is the \emph{one-way communication} model, where the only communication taking place during a protocol is from Alice to Bob; then Bob outputs the value of the function.

\begin{definition}[One-way protocol]
A communication protocol $\pi$ is a \emph{one-way protocol} if all the internal nodes except the ones right above the leaves in the rooted binary tree $T(\pi)$ are labelled with $A$.
\end{definition}

As in the standard two-way communication setting, we can consider deterministic, public-coin, and private-coin protocols.

\begin{definition}[One-way communication complexity]
The deterministic and (public-coin) randomized \emph{one-way communication complexities} of a function $f : \calX \times \calY \to \{0,1\}$, denoted
\[
D^{\rightarrow}(f) \qquad \mbox{and} \qquad R^{\rightarrow}(f)
\]
respectively, are the minimum cost of a deterministic one-way protocol that computes $f$ and of a public-coin randomized one-way protocol that computes $f$ with error at most $\frac13$, respectively.
\end{definition}

\begin{remark}
We can also define one-way analogues of the other models of communication such as distributional complexity, private-coin randomized complexity, etc.
\end{remark}

The communication complexity of some functions remains identical in the one-way and two-way settings.

\begin{theorem}
The one-way communication complexity of $\Eq : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ satisfies
\[
D^\rightarrow(\Eq) = D(\Eq) = \Theta(n) \qquad \mbox{and} \qquad 
R^\rightarrow(\Eq) = R(\Eq) = \Theta(1).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Index function}

In general, however, the one-way and two-way complexity of a function can differ greatly. A simple example that highlights this difference is the \emph{index function} $\Index : \{0,1\}^n \times [n] \to \{0,1\}$ defined by
\[
\Index(x,i) = x_i.
\]

\begin{theorem}
The communication complexity of the $\Index : \{0,1\}^n \times [n] \to \{0,1\}$ function satisfies
\[
D(\Index) = O(\log n) \qquad \mbox{and} \qquad D^\rightarrow(\Index) = \Omega(n).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{One-way vs.~two-way communication complexity}

The gap between the one-way and two-way communication complexity of the index function is the largest possible.

\begin{theorem}
For every function $f : \calX \times \calY \to \{0,1\}$,
\[
D^\rightarrow(f) \le 2^{D(f)}.
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Index II}

We can strengthen our lower bound on the one-way communication complexity of the index function to show that it holds in the public-coin randomness model as well.

\begin{theorem}
The one-way randomized communication complexity of the $\Index : \{0,1\}^n \times [n] \to \{0,1\}$ function is
\[
R^\rightarrow(\Index) = \Omega(n).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Disjointness}

The lower bound on the index function can be used to obtain a simple proof of the $\Omega(n)$ lower bound on the one-way randomized communication complexity of the disjointness function.

\begin{theorem}
The one-way randomized communication complexity of the $\Disj : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ function is
\[
R^\rightarrow(\Disj) = \Omega(n).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\begin{remark}
Your proof should \emph{not} use our previous lower bounds on the two-way randomized communication complexity of the disjointness function; instead, you should be able to use the lower bound on the one-way communication complexity of the index function to get this result.
\end{remark}

\newpage 
\section{Gap Hamming (Bonus)}

Define the \emph{Hamming distance} between strings $x,y \in \{0,1\}^n$ to be $\dHam(x,y) = |\{i \in [n] : x_i \neq y_i\}|$.
The \emph{gap Hamming function} $\GHD : \{0,1\}^n \times \{0,1\}^n \to \{0,1,*\}$ is the partial function defined by
\[
\GHD(x,y) = \begin{cases}
1 & \mbox{if } \dHam(x,y) = \frac n2 \\
0 & \mbox{if } |\dHam(x,y) - \frac n2| \ge \sqrt{n} \\
* & \mbox{otherwise.}
\end{cases}
\] 
(We assume throughout this section that $n$ is even.)

\begin{definition}
A randomized protocol \emph{computes} the partial function $f : \calX \times \calY \to \{0,1,*\}$ with error at most $\epsilon$ if for every input $(x,y) \in f^{-1}(0) \cup f^{-1}(1)$, it outputs the value $f(x,y)$ with probability at least $1-\epsilon$.\footnote{For the inputs in $f^{-1}(*)$, the protocol is free to output anything.}
\end{definition}

\begin{theorem}
The one-way randomized communication complexity of the $\GHD : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ function is
\[
R^\rightarrow(\GHD) = \Omega(n).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here, if you complete it.}
\end{proof}

\bigskip
\emph{Hint.} The lower bound for the one-way complexity of the \textsc{Index} function should be helpful for this proof as well, though the reduction is not (nearly) as straightforward as the one in the last section.

\newpage 
\section{Simultaneous message passing}

In the \emph{simultaneous message passing (SMP)} model of communication, Alice and Bob both send messages to a third party that we will call the Referee, without seeing each other's transmissions. (We picture both Alice and Bob sending their communications to the referee simultaneously.) Note that the Referee does not see either Alice or Bob's inputs, and must then output the result of the protocol using only the messages received from both parties. In the private-coin model of SMP communication, Alice and Bob each have a private source of randomness that is not visible to any other party.

\begin{definition}
The \emph{SMP (private randomness) complexity} with error $\epsilon$ of a function $f : \calX \times \calY \to \{0,1\}$, denoted
\[
R^{\parallel,\mathrm{priv}}_\epsilon(f),
\] 
is the minimum (worst-case) communication cost of any private-coin SMP protocol that computes $f$ with error at most $\epsilon$ on any input $(x,y) \in \calX \times \calY$.
\end{definition}

As usual, we write $R^{\parallel,\mathrm{priv}}(f) = R^{\parallel,\mathrm{priv}}_{1/3}(f)$.

\begin{theorem}
The SMP complexity of every function $f : \calX \times \calY \to \{0,1\}$ satisfies
\[
R^{\parallel,\mathrm{priv}}(f) = \Omega(\sqrt{D(f)}).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\bigskip
\noindent \emph{Hint.} The ideas from the proofs of Sections 8 and 9 in Chapter 4 might be helpful here as well. In particular, what can we say about the success probability of a protocol in which Alice and Bob send $t$ messages to the referee instead of $1$, when all $t$ messages are sent for the same input but with independent randomness?

\newpage 
\section{Equality X}

\begin{theorem}
The SMP complexity of the equality function $\Eq : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ is
\[
R^{\parallel,\mathrm{priv}}(\Eq) = \Theta(\sqrt{n}).
\]
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\bigskip
\noindent \emph{Hint.} By using (basic but still remarkable) results from error-correcting codes, it is sufficient to design a randomized protocol that computes the partial function $\Eq^* : \{0,1\}^n \times \{0,1\}^n \to \{0,1,*\}$ defined by
\[
\Eq^*(x,y) = \begin{cases}
1 & \mbox{if } x = y \\
0 & \mbox{if } \dHam(x,y) = \frac n2 \\
* & \mbox{otherwise}
\end{cases}
\]
with error at most $\frac13$.
\end{document}

### 2019-02-15 CS860-CommunicationComplexity/assignments/Chapter6.tex
Solved section 4

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Consider a one-way randomized communication protocol for $\Index$ with success probability $\frac{2}{3}$, achieving communication cost $R^\rightarrow(\Index)$.
+Let $X$ and $Y$ be random variables for Alice and Bob's inputs respectively, each distributed uniformly and independently at random.
+Let $M$ be a random variable for the message $m$ that Alice sends Bob during the operation of this protocol.
+Our goal is to use information-theoretic techniques to lower bound $|M|$, the maximum length of a message in $M$'s support.
+We have,
+\begin{align*}
+|M| &\geq H(M)\\
+&\geq I(X;M) \\
+&=\sum_{i=1}^n I(X_i; M \mid X_1\dots X_{i-1}) &\text{by chain rule} \\
+&=\sum_{i=1}^n H(X_i\mid X_1\dots X_{i-1}) - H(X_i\mid M, X_1\dots X_{i-1})\\
+&= n - \sum_{i=1}^n H(X_i\mid M, X_1\dots X_{i-1}) &\text{since the bits of $X$ are chosen uniformly independently} \\
+&\geq n-\sum_{i=1}^n H(X_i\mid M).
+\end{align*} 
+We claim $H(X_i\mid M)\leq \frac{2}{3}$. With this claim in hand, we see that
+$$|M| \geq n-\frac{2}{3}n =\Omega(n).$$
+To see the claim, we use that the protocol succeeds with probability $\frac{2}{3}$. We expand the conditional entropy as an expectation and compute
+$$H(X_i\mid M) = \sum_{m\in M} H(X_i\mid M=m)\cdot \Pr(M=m) = \sum_{m\in M}(\frac{2}{3}\log(\frac{3}{2}) + \frac{1}{3}\log(3))\Pr(M=m) \leq \frac{2}{3}.$$
+The second inequality follows by the success probability of the protocol.
+The probability that the protocol say $X_i$ is the correct bit is $\frac{2}{3}$, and the probability that $X_i$ is the wrong bit is $\frac{1}{3}$.
 \end{proof}

### 2019-02-15 CS860-CommunicationComplexity/assignments/Chapter6.tex
1 and 2

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+
+First we consider the deterministic setting.
+Clearly $n$ bits suffice to send Alice's input.
+Thus $D^\rightarrow(\Eq) = O(n)$.
+For the lower bound consider a deterministic one-way protocol $\pi$ computing $\Eq$.
+Let $\pi'$ be the two-way protocol where Alice and Bob simulate $\pi$ and in $1$ bit Bob sends the result to Alice.
+Clearly $\pi'$ solves $\Eq$, and its cost is
+$$D^\rightarrow(\Eq) + 1.$$
+Since $D(\Eq) = \Omega(n)$ this implies $D^\rightarrow(\Eq) = \Omega(n)$ as desired.
+
+Now consider the randomized setting.
+$R^\rightarrow(\Eq) = \Omega(1)$ is immediate.
+For the upper bound, consider the following randomized protocol, which is entirely analogous to what we do in the two-way model.
+Alice and Bob generate $r_1, r_2 \in \{0,1\}^n$ independently and uniformly at random using their shared randomness.
+Alice computes $x\cdot r_1$ and $x\cdot r_2$ and sends the results to Bob.
+Bob computes $y\cdot r_1$ and $y\cdot r_2$ and outputs $1$ if and only if $x\cdot r_1 = y\cdot r_1$ and $x\cdot r_2 = y\cdot r_2$.
+The success probability of this protocol is $\frac{3}{4}$ following the same analysis as the two-way model.
+
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Trivially $D(\Index) = O(\log n)$: Bob sends his input $y$ using $\log n$ bits, and Alice sends $x_y$ using $1$ bit.
+
+To see that $D^\rightarrow(\Index) = \Omega(n)$, suppose for contradiction there is a one-way protocol $\pi$ which computes $\Index$, and Alice sends strictly less than $n$ bits.
+Let $m\in\{0,1\}^{|m|}$ be the message Alice sends to Bob.
+Our assumption says $|m| < n$.
+Since $\pi$ is deterministic, Bob's output is entirely a function of $m$ and his input $y$.
+Since $|m|<n$ there exist a pair of inputs $x^1$ and $x^2$ for Alice for which she sends the same message $m$.
+If $y$ is the index of a bit on which $x^1$ and $x^2$ differ, then $\pi$ will not be correct on both $(x^1,y)$ and $(x^2,y)$.
+On message $m$ and input $y$ Bob can output either $x^1_y$ or $x^2_y$.
+In either case, he is wrong on some input.
+Thus any determinsitc one-way protocol for $\Index$ must use at least $n$ bits of communication.
 \end{proof}

### 2019-02-15 CS860-CommunicationComplexity/assignments/Chapter6.tex
Solved up to last two

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a two-way deterministic protocol computing $f$, and achieving cost $D(f)$.
+We construct a one-way protocol $\pi'$ for $f$ as follows.
+Alice considers the sub-tree of the protocol tree $\pi$ corresponding to what she would send on her input, and all of Bob's possible replies.
+The size of this sub-tree is at most $2^{D(f)}$.
+After reading this sub-tree, Bob outputs the bit at the leaf to which his input with Alice's communication corresponding.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Consder any randomized one-way protocol $\pi$ which solves $\Disj$..
+We will construct a protocol for $\Index$ of the same cost, and apply the lower bound from the previous section to obtain the result.
+Let $x \in \{0,1\}^n$ and let $y \in [n]$ be Alice and Bob's respective inputs to $\Index$.
+They run their $\Disj$ oracle $\pi$ as if Alice input were $x$ and Bob's was the standard basis vector $e_y\in\{0,1\}^n$, where only the $y$-th bit $e_y$ is $1$.
+Since
+$$\Disj(x,e_y) = \Index(x,y)$$ 
+this protocol solves $\Index$ using the same communication cost as $\pi$.
 \end{proof}

### 2019-02-28 CS860-CommunicationComplexity/project/proposal.tex
Outlined proposal

--- 2017-04-05 CO750-ApproximationAlgorithmsTopics/assignments/report.tex
@@
 \usepackage{tkz-berge}
+\usepackage[colorlinks]{hyperref}
+\usepackage{bookmark}
 
@@
 
+\newcommand{\tight}{tight}
+\newcommand{\Mtight}{\cM_{\tight}}
+\newcommand{\uni}{uni}
+\newcommand{\Muni}{\cM_{\uni}}
+\newcommand{\core}{\ensuremath{\mbox{core}}}
+\DeclareMathOperator{\suppOp}{supp}
+\newcommand{\supp}{\suppOp}
+
+\DeclareMathOperator{\convOp}{conv}
+\newcommand{\conv}{\convOp}
+
+\newcommand{\level}{\mathrm{lev}}
+\newcommand{\set}{\mathrm{set}}
+\newcommand{\fix}{\mathrm{Fix}}
+
+\DeclareMathOperator{\exOp}{excess}
+\newcommand{\ex}{\exOp}
+
+\DeclareMathOperator{\symOp}{diff}
+\newcommand{\sym}{\symOp}
+
+\DeclareMathOperator{\parent}{parent}
+\DeclareMathOperator{\optop}{top}
+\DeclareMathOperator{\excess}{excess}
+\DeclareMathOperator{\opspan}{span}
+
 \newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
@@
 \newtheorem{conjecture}[fact]{Conjecture}
+\newtheorem{question}[fact]{Question}
 
@@
 
-\lhead{2017-04-04}
-\rhead{W. Justin Toth CO750-Approximation Algorithms Report} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
+\lhead{CS860 Proposal}
+\rhead{W. Justin Toth} %
 \renewcommand{\headrulewidth}{1pt} %heading underlined
@@
 \begin{document}
-\paragraph{Paper Title:} Popularity, Mixed Matchings, and Self-duality
-\paragraph{Authors:} Chien-Chun Huang and Telikepalli Kavitha
-\paragraph{Reference:} \cite{huang2017popularity}
+\section{Extended Formulations and Communication Complexity}
+\paragraph{}
+Consider a polytope $P\subseteq \R^n$ and a polyhedron $Q\subseteq \R^n$ containing $P$ (called a \emph{polyhedral pair}).
+For intuition, think of $P$ as the convex hull of the set of solutions to some NP-hard problem, for which we do not know a compact linear programming description, and think of $Q$ as a relaxation of $P$ for which are interested in finding a compact description of.
+\paragraph{}
+We say that a system 
+$$E^\leq x + F^\leq y \leq g^\leq, E^= x + F^= y = g^=$$
+in $\R^{n+k}$ is an \emph{extended formulation} of $(P,Q)$ if the polyhedron
+$$R:=\{x \in \R^n: \exists y \in \R^k: E^\leq x + F^\leq y \leq g^\leq, E^= x + F^= y = g^=\}$$
+contains $P$ and is contained in $Q$.
+The number of inequalities in the system determines the size of the extended formulations.
+The \emph{extension complexity} of $(P,Q)$, $xc(P,Q)$, is the minimum size of an extended formulation of $(P,Q)$.
+\paragraph{}
+Usually one studies the case $P=Q$, but for our purposes we are interested in the extension of complexity of a specific relaxation, and hence need the fully general definiton above.
+Yannakakis' seminal result \cite{yannakakis1991expressing} gives an algebraic expression of $xc(P,Q)$.
+To state the result we will need two more pieces of terminology.
 
-\section{Preliminaries}
-\paragraph{}
-We will begin by introducing the definitions of the main objects of study, and some useful notation. Then we will state the theorems to be shown in the later sections of this report.
-\paragraph{}
-As in most problems in the stable matching family we are given a graph $G=(V,E)$ where each vertex has ranking over its neighbours.
-We will assume throughout that these preference orders are strict, and a vertex prefers any neighbour to being unmatched.
-We will denote by $N(v)$ the set of neighbours of vertex $v$, and by $\delta(v)$ the set of edges incident with $v$.
-We will use the notation $>_v$ to describe $v$'s preference order over $N(v)$.
-Precisely for $u, w \in N(v)$, we write $u >_v w$ if $u$ is preferred by $v$ to $w$.
-We will use $\geq_v$, $\leq_v$, $=_v$, and $<_v$ analogously.
-\paragraph{}
-One way to think of $\textit{popular}$ matchings is as matchings which never lose an election to any other matching.
-Since we will speak often about partners in matchings, we introduce the notation: if $M$ is a matching then $M(u) = v$ provided $uv \in M$.
-For any $u \in V$ and $v,v' \in N(v)$ we define $vote_u(v,v')$ as
-$$vote_u(v,v') := \begin{cases}
-1, \text{ if } v >_u v' \\
--1, \text{ if } v <_u v' \\
-0, \text{ otherwise (ie. } v = v').
-\end{cases}$$
-To decide the winner of the election we some the votes. The aggregate voting between two matchings $M$ and $M'$ is
-$$\Delta(M, M') = \sum_{u \in V} vote_u(M(u), M'(u)).$$
-Observe that $\Delta(M,M') \geq 0$ if at least as many vertices prefer their partner in $M$ to their partner in $M'$. A $\textit{popular}$ matching $M$ is  a matching where
-$$\Delta(M, M') \geq 0$$
-for any matching $M'$.
-\paragraph{}
-We are interested in the $\textit{max weight popular matching}$ problem in this report.
-That is, given a weight function $w: E \rightarrow \R_+$ find a popular matching $M$ on $G$ which maximizes $w(M) := \sum_{e \in M} w(e).$ A standard approach to such problems is to attempt to formulate them as linear programs.
-This will motivate our definitions of fractional popular matchings to follow.
-In this approach the major question is whether or not your linear programming formulation gives back integral solutions which can be used to solve the original combinatorial problem.
-In this paper we will prove that there is an integral formulation for a special case, and a $\frac{1}{2}$-integral formulation in general.
-\paragraph{}
-The fractional matching polytope $FM_{G}$ is defined as
-$$FM_{G} := \{x \in \R_{\geq 0}^{E} : x(\delta(v)) \leq 1, \forall v \in V \}.$$
-This object has been well-studied in combinatorial optimization.
-We want to study the analogous $\textit{popular fractional matching polytope}$ which we will call $P_G$.
-To formally define $P_G$ we will need to extend our notions related to voting as used in defining popular matchings.
-Let $x \in FM_G$.
-We define
-$$vote_u(x, v') := \sum_{uv \in \delta(u)} x_{uv} vote_u(v,v') = x(\delta^{>v'}(u)) - x(\delta^{<v'}(u)),$$
-where $\delta^{>v'}(u) := \{vu \in \delta(u) : v >_u v'\}$ denotes the set of edges $u$ prefers to $uv'$, and $\delta^{<v'}(u)$ is defined analogously.
-We define $vote_u(v',x) = -vote_u(x,v')$.
-Now we have a means of voting between two fractional matchings $x,y \in FM_G$:
-$$vote_u(x,y) := \sum_{uv' \in \delta(u)} y_{uv'}vote_u(x,v'), $$
-and the total votes:
-$$\Delta(x,y) := \sum_{u \in V} vote_u(x,y).$$
-Now we can formally describe $P_G$ as 
-$$P_G = \{ x\in FM_G: \Delta(x,y) \geq 0, \forall y \in FM_G\}.$$
-\paragraph{}
-We finally have sufficient definitions to describe the results of this paper.
-In section \ref{sec:formulation} we describe a useful extended formulation which describes $P_G$ that will come into play when studying popular fractional matchings throughout this work.
-In section \ref{sec:special} we build up the following theorem
-\begin{theorem}\label{th:special}
-Let $G = (A \cup B, E)$ be a bipartite graph with strict preferences. If $G$ admits a perfect stable matching the $P_G$ is integral.
-\end{theorem}
-Building upon this theorem we show that $\frac{1}{2}$-integrality holds in general in section \ref{sec:general}.
-\begin{theorem}
-Let $G=(V,E)$ be an arbitrary graph with strict preferences. Then the extreme points of $P_G$ lie in $\{0,\frac{1}{2}, 1\}^E$.
-\end{theorem}
-In section \ref{sec:complexity} we discuss some results showing that max weight popular matching is $NP$-hard and under the Unique Games Conjecture is difficult to approximate.
-We conclude with a section discussing open problems.
-\section{Extended Formulation}\label{sec:formulation}
+\subsection{The goal of this section is Faenza's 2015 result}
+\section{The Knapsack Cover Problem}
+\subsection{The goal here is to describe all the knapsack cover stuff that has been done in the natural space}
+\section{Extended Formulations for Knapsack Cover}
+\subsection{Here we discuss extended formulations for KC, and communication protocols}
+\section{Lower Bounds}
+\subsection{some ideas for lower bounding communication cost of KCI slack matrix}
 
-\paragraph{}
-Here we present the extended formulation $P'_G$ which we will use to study the popular fractional matching polytope.
-The properties of $P'_G$ were first studied in \cite{kavitha2011popular}.
-For a more rigorous treatment of its connection to $P_G$ refer to that paper.
-Here we try to present ideas about how $P'_G$ was derived.
-
-\paragraph{}
-We begin by augmenting our graph $G$ with a set of last resort neighbours $\ell(u)$ for every $u \in V$.
-The vertices $\ell(u)$ are only connected to their respective vertex $u$.
-The preference lists of each $u$ are modified so that $\ell(u)$ is their last choice.
-That is for all $v \in N(u)$, $v \geq_u \ell(u)$.
-The vertex set and edge set now look like
-$$V(G) = V \cup \bigcup_{u \in V} \ell(u) \quad\text{and}\quad E(G) = E \cup \bigcup_{u \in V} u\ell(u).$$  With the addition of the last resort vertices we may now assume that every $v \in V$ (the original vertices) is fully matched in any popular fractional matching.
-
-\paragraph{}
-Fix a fractional matching $x \in FM_G$.
-We seek a linear program which characterizes when $x$ is popular.
-Let $w: E(G) \rightarrow \R_+$ be the weight function given by assigning  $$w(ab) := vote_a(b,x) + vote_b(a,x) = x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b))$$
-for any $ab \in E$, and assigning
-$$w(u\ell(u)) := vote_u(\ell(u),x) = -x(\delta^{>\ell(u)}(u)).$$
-Observe that if $y \in FM_G$ fully matches each $u \in V$ then $\sum_{ab \in E(G)} w(ab)y_{ab} = \Delta(y,x).$ So the fractional matching $x$ is popular if and only if the optimal value of the following LP is $0$ (it cannot be less than $0$ as $x$ is a feasible solution, and $\Delta(x,x) = 0$) :
-\begin{align*}
-\max\  &\sum_{e \in E(G)} w(e) y_{e} \\
-\text{s.t.}\ y(\delta(u)) &= 1 &\forall u \in V \\
-y_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-Consider the dual to previous linear program, to be denoted $(LP1)$:
-\begin{align*}
-\min\ &\sum_{u \in V} \alpha_u \\
-\text{s.t.}\ \alpha_a + \alpha_b &\geq w(ab) &\forall ab \in E \\
-\alpha_u &\geq w(u\ell(u)) &\forall u \in V.
-\end{align*}
-If $x$ is popular then the optimal value of $(LP1)$ is $0$.
-Also note that if $x$ is not popular then the optimal value is greater than $0$.
-So one means of finding popular fractional matchings would be to let $x$ vary as a fractional matching and minimize $\sum_{u} \alpha_u$.
-We call that linear program $(LP2)$:
-\begin{align*}
-\min\ &\sum_{u \in V} \alpha_u \\
-\text{s.t.}\ \alpha_a + \alpha_b &\geq w(ab) &\forall ab \in E \\
-\alpha_u &\geq w(u\ell(u)) &\forall u \in V \\
-x(\delta(u)) &= 1 &\forall u \in V \\
-x_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-As discussed above, the optimal value of $(LP2)$ is $0$, and so a corresponding optimal solution for $(LP2)$ is a vector $(x,\alpha) \in \R^{E(G)}\times \R^{V}$ where $x$ is a popular fractional matching and $\sum_{u \in V} \alpha_u = 0$.
-Such $\alpha$ is called a $\textit{witness}$ to the popularity of $x$.
-The polytope of optimal solutions of $(LP2)$ is what we will use as our extended formulation for the popular fractional matching polytope.
-We can describe it as $P'_G$, the set of solutions to the set of linear inequalities:
-\begin{align*}
-\sum_{u \in V} \alpha_u &= 0\\
-\alpha_a + \alpha_b &\geq vote_a(b,x) + vote_b(a,x) &\forall ab \in E \\
-\alpha_u &\geq vote_u(\ell(u),x) &\forall u \in V \\
-x(\delta(u)) &= 1 &\forall u \in V \\
-x_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-Thus $P'_G$ involves $O(|E| + |V|)$ constraints, and hence provides a polynomial size description of popular fractional matchings.
-The following lemma demonstrates a very interesting property of $(LP2)$
-\begin{lemma}\label{lemma:self-dual}
-The linear program $(LP2)$ is its own dual program program. That is $(LP2)$ is self-dual.
-\end{lemma}
-\begin{proof}
-Consider the dual to $(LP2)$ in variables $y \in \R^{E(G)}$ and $\beta \in \R^V$:
-\begin{align*}
-\max\ &\sum_{u \in V} \beta_u \\
-\text{s.t.} \beta_a + \beta_b + y(\delta^{<b}(a))  - y(\delta^{>b}(a)) + y(\delta^{<a}(b)) - y(\delta^{>a}(b)) &\leq 0 &\forall ab \in E \\
-\beta_u - y(\delta^{>\ell(u)}(u)) &\leq 0 &\forall u \in V \\
-y(\delta(u)) &= 1 &\forall u \in V \\
-y_e &\geq 0 &\forall e \in E(G).
-\end{align*}
-If we perform the substitutions $\alpha_u = -\beta_u$ for each $u \in V$ and $x_e = y_e$ for each $e \in E(G)$ then the above dual LP becomes exactly the same as $(LP2)$.
-\end{proof}
-\paragraph{}
-This lemma leads to the following very useful lemma, which follows immediately from the previous result via complementary slackness.
-\begin{lemma}\label{lemma:self-dual-cs}
-Let $(x, \alpha^x) \in P'_G$.  For every $ab \in E$, if $x_{ab} > 0$ then the ``covering constraint" in $P'_G$ for $ab$ is tight. Formally we have
-$$\alpha_a^x + \alpha_b^x = x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b)).$$
-\end{lemma}
-\section{Integrality of $P_G$ in a special case}\label{sec:special}
-\paragraph{}
-In this section we turn our attention to theorem \ref{th:special}.
-So suppose that our graph $G$ is bipartite with bipartition $A \cup B$.
-Also suppose that $G$ admits a perfect stable matching.
-This condition actually implies that every popular matching is perfect.
-This follows since stable matchings always match the same set of vertices, and stable matchings are minimum size popular matchings \cite{huang2011popular}.
-It is not hard to see that this extends to popular fractional matchings, meaning every popular fractional matching fully matches each $u \in V$ to its true neighbours (not last resort vertices).
-\paragraph{}
-Let $(x, \alpha^x) \in P'_G$. We want to write $x$ as a convex combination of popular matchings. First we want bounds on the values the witness vector $\alpha^x$ takes
-\begin{lemma}
-For every $u \in V$, $\alpha_u^x \in [-1,1]$.
-\end{lemma}
-\begin{proof}First
-$$\alpha^x_u \geq vote_u(\ell(u), x) = -1 \quad\text{(since }x_{u,\ell(u)} = 0\text{).}$$
-Now let $v$ be the least preferred neighbour of $u$ such that $x_{uv} >0$. Then
-$$vote_u(v,x) = -(1-x_{uv}).$$
-Also
-$$vote_v(u,x) = x(\delta^{<u}(v))-x(\delta^{>u}(v))= 1-x_{uv}.$$
-So by lemma \ref{lemma:self-dual-cs},
-$$\alpha_u + \alpha_v = vote_u(v,x) + vote_v(u,x) = -(1-x_{uv}) + (1-x_{uv}) = 0.$$
-Since $\alpha_v \geq -1$ this implies $\alpha_u \leq 1$.
-\end{proof}
-\paragraph{}
-Thinking ahead, we want to write $x$ as a convex combination of popular matchings.
-In doing this we will use a witness vector in $\{-1,1\}^V$ to certify the popularity of each chosen matching.
-From our previous lemma we can think of each $\alpha^x_u$ as a convex combination of $-1$ and $1$.
-For each $a \in A$ let $r_a$ denote the fraction of popular matchings in the convex combination forming $x$ whose $\alpha_a^x$ value is assigned $1$.
-Similarly define for each $b \in B$ the value $r_b$ to be the fraction assigning $\alpha^x_b$ to $-1$.
-Precisely we have $2r_a -1 = \alpha^x_a$ and $1-2r_b = \alpha^x_b$.
-\paragraph{}
-Let $X_a$ be the array for the assignment of $a\in A$ in $x$.
-Each cell of $X_a$ corresponds to some $b \in N(a)$ and the length of the cell is $x_{ab} > 0$ (we omit $0$-length cells).
-The cells in $X_a$ are arranged in increasing order with respect to $a$'s preferences.
-The first $r_a$ fraction of $X_a$ is called the $\textit{positive}$ subarray and the remaing the $\textit{negative}$ subarray.
-We split and reorder the arrays as shown in the following figure to obtain $X'_a$
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.25]{Xa}
-\end{figure}
-We do the same thing for each $b \in B$, creating $X_b$.
-But the only difference is that we arrange $X_b$'s cell in decreasing order of $b$'s preferences.
-Also notice that since $r_b$ corresponds to the negative $\alpha_b$ cells, the array begins with the negative subarray.
-We again split and reorder to obtain $X'_b$ as shown
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.25]{Xb}
-\end{figure}
-One interesting thing about our choice of the $r_a$'s and $r_b$'s is that they allow us to rewrite our covering constraints for $ab$ in a simpler form.
-\begin{lemma}\label{lemma:covering}
-For any $ab \in E$ we have
-$$x_{ab} + x(\delta^{<b}(a)) + x(\delta^{<a}(b)) \leq r_a + (1-r_b),$$
-and if $x_{ab} > 0$ then the above is tight.
-\end{lemma}
-\begin{proof}
-Take the constraint
-$$ \alpha^x_a + \alpha^x_b \geq x(\delta^{<b}(a)) - x(\delta^{>b}(a)) + x(\delta^{<a}(b)) - x(\delta^{>a}(b)) $$
-and rewrite using $\alpha^x_a = 2r_a-1$ and $\alpha^x_b = 1-2r_b$.
-Further substitute $-x(\delta^{>b}(a)) = -(1 -x_{ab} - x(\delta^{<b}(a)))$ and $-x(\delta^{>a}(b)) = -(1- x_{ab} - x(\delta^{<a}(b)))$.
-After simplifying the inequality in the lemma statement is obtained.
-By lemma \ref{lemma:self-dual-cs} this inequality is tight when $x_{ab} > 0$.
-\end{proof}
-
-\paragraph{}
-Now we show how to obtain popular matchings whose convex hull contains $x$.
-Form a table $T$ by stacking each $X'_u$ for $u \in V$ in rows.
-The width of $T$ is $1$, so for any $t \in [0,1)$ we can define a set of edges $M_t$ as follows:
-\begin{enumerate}
-\item Form a vertical line at distance $t$ from the left end of $T$
-\item For each $u \in V$ the vertical line intersects (or touches the left boundary of) a cell in $X'_u$. We will denote this cell by $c_u(t)$.
-\item Form the set $M_t$ as $M_t := \{uv : u \in V \text{ and } uv \text{ corresponds to cell } c_u(t)\}$.
-\end{enumerate}
-\paragraph{}
-We need to verify two things: the first is that $M_t$ is a matching, and the second is that $M_t$ is indeed popular.
-This will be done in the following two subsections.
-Supposing we are successful in the promised arguments to follow, we will show how this allows us to find a set of popular matchings for which $x$ is a convex combination.
-Execute the following algorithm:
-\begin{enumerate}
-\item Initialize $i=0$.
-\item Sweep a vertical line from left to right across $T$. Denote its distance from the left by $t$. 
-\item Whenever a new cell is encountered fix matching $M^i = M_t$ and fix $t_i = t$. 
-\end{enumerate}
-Say we find $k$ such matchings. We can then write $x$ as:
-$$x = t_1\chi(M^0) + (t_2 - t_1)\chi(M^1) + \dots + (1 - t_{k-1}) \chi(M^{k-1})$$
-where $\chi(\cdot)$ if the function mapping sets to incidence vectors.
-
-\subsection{$M_t$ is a matching}
-
-\subsection{$M_t$ is popular}
-\section{$\frac{1}{2}$-Integrality of $P_G$ in general}\label{sec:general}
-
-\section{Complexity Results}\label{sec:complexity}
-
-\section{Conclusions}
-
-\bibliography{references}
+\bibliography{references.bib}
 \bibliographystyle{plain}

### 2019-02-28 CS860-CommunicationComplexity/assignments/Chapter6.tex

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+	Let $\pi$ be a SMP protocol for $f$ of cost $R^{\parallel,\mathrm{priv}}(f)$.
+For each input $x\in\calX$ let $M_x$ be the set of messages Alice has non-zero probability of sending on input $x$.
+Similarly for each input $y\in\calY$ let $M_y$ be the set of messages Bob has non-zero probability of sending on input $y$.
+
+	We want to show that for each $x \in \calX$ there exists a set of messages $S_x \subseteq M_x$ such that for any $y \in \calY$, and message $m_y \in M_y$, the protocol correctly outputs $f(x,y)$ on the majority of messages $S_x$ with $m_y$.
+Let $\tau(m_x,m_y)$ be the referees (deterministic) output on messages $(m_x,m_y)$ for protocol $\pi$.
+I.e.
+	$$\frac{2}{3} \leq \Pr(\pi(x,y) = f(x,y)) = \Pr_{m_x,m_y}(\tau(m_x,my) = f(x,y)).$$
+	Fix $y\in \calY$, and $m_y \in M_y$.
+Drawn a uniform random subset $S_x \subseteq M_x$  of size $t$.
+By a Chernoff Bound, the probability that $\tau(m_x,m_y) \neq f(x,y)$ for at least half of message pairs $m_x, m_y$ is at most
+	$$2^{-\Omega(t)}.$$
+	Since $\pi$ has cost $R^{\parallel,\mathrm{priv}}(f)$, over all $y$ there are at most $2^{R^{\parallel,\mathrm{priv}}(f)}$ messages the referee sees from Bob.
+So by union bound, the probability $\tau(m_x,m_y) \neq f(x,y)$ for some message $m_y$ is at most
+	$$2^{R^{\parallel,\mathrm{priv}}(f)}2^{-\Omega(t)}.$$
+	This probability is strictly less than one for $t = \Omega(R^{\parallel,\mathrm{priv}}(f))$.
+Thus by the probabilistic method the set $S_x$ exists.
+By a symmetric argument, the analogous set $S_y$ exists for each $y \in \calY$.
+
+	Now our deterministic protocol for $f$ is as follows.
+Alice sends to Bob each message in $S_x$ for her input $x$.
+Since $|S_x| \leq R^{\parallel,\mathrm{priv}}(f)$ and each message in $S_x$ has size at most $R^{\parallel,\mathrm{priv}}(f)$, this has communication cost $R^{\parallel,\mathrm{priv}}(f)^2$.
+Bob sends to Alice each message $S_y$ for his input $y$, at communication cost $R^{\parallel,\mathrm{priv}}(f)^2$.
+They both can compute the majority referee decision on $S_x\times S_y$, and output that.
+By our choice of $S_x$ and $S_y$, this yields the correct answer.
+The cost of this protocol is $O(R^{\parallel,\mathrm{priv}}(f)^2)$, as desired.
 \end{proof}

### 2019-03-01 CS860-CommunicationComplexity/project/proposal.tex
Section 1 of proposal

@@
 For intuition, think of $P$ as the convex hull of the set of solutions to some NP-hard problem, for which we do not know a compact linear programming description, and think of $Q$ as a relaxation of $P$ for which are interested in finding a compact description of.
-\paragraph{}
+
 We say that a system 
@@
 The \emph{extension complexity} of $(P,Q)$, $xc(P,Q)$, is the minimum size of an extended formulation of $(P,Q)$.
-\paragraph{}
+
 Usually one studies the case $P=Q$, but for our purposes we are interested in the extension of complexity of a specific relaxation, and hence need the fully general definiton above.
@@
 
+Let $P = \text{conv}(\{v_1,\dots,v_p\})$ and $Q = \{x\in \R^n: Ax\geq b\}$ where $A \in \R^{m \times n}$ and $b \in \R^m$.
+The \emph{slack matrix} $S$ of $(P,Q)$ is defined as follows.
+For $i \in[m]$ and $j \in [p]$,
+$$S_{i,j} = A_iv_j - b_i.$$
+Note that the slack matrix depends on the inner and outer descriptions of $P$ and $Q$ respectively.
+We say that $T\in \R^{m\times r}_{\geq 0}$, $U \in \R^{r\times n}_{\geq 0}$ is a \emph {rank $r$ non-negative factorization} of $S$ if $S = TU$.
+The \emph{non-negative rank of $S$} is
+$$\text{rk}_x(S) L= \min\{r: S \text{ has a rank $r$ non-negative factorization}\}.$$ 
+Equivalently a non-negative factorization of $S$ of rank at most $r$ is a decomposition of $S$ into the sum of at most $r$ rank $1$ matrices.
+\begin{theorem}
+    (Yannakakis) For $P$ of dimension at least $1$, we have $\text{xc}(P,Q) \in \{\text{rk}_+(S), \text{rk}_+(S) - 1\}.$
+\end{theorem}
+\paragraph{}
+Faenza, Fiorini, Grappe, and Tiwary~\cite{faenza2015extended} identified a model of communication which is intimately connected to non-negative rank (and thus to extension complexity).
+Let $\Pi$ be a communication protocol where Alice and Bob each have private randomness.
+Alice receives $i \in [n]$ and Bob receives $j \in [p]$.
+They exchange bits in the usual way according to a protocol tree and at the end \emph{either one} of the players outputs some non-negative number $\alpha$.
+We say $\Pi$ computes $S$ in expectation if for every $i \in [n]$, $j \in [p]$, the expected value of $\alpha$ is $S_{i,j}$.
+The cost of $\Pi$ is the maximum number of bits exchanged by Alice and Bob on any input pair, not counting the encoding size of $\alpha$.
+The expected communication complexity of $S$, denoted $R^\text{cc}_{\text{exp}}(S)$, is the minimum cost of a protocol computing $S$ in expectation.
+\begin{theorem}(Faenza et al.) Provided $\text{rk}_x(S) \neq 0$, $$R^\text{cc}_{\text{exp}}(S) = \log \text{rk}_x(S) + \Theta(1) = \log\text{xc}(P,Q) + \Theta(1).$$
+The second equality is a direct consequence of Yannakakis' Theorem. 
+\end{theorem}
 \subsection{The goal of this section is Faenza's 2015 result}

### 2019-03-01 CS860-CommunicationComplexity/project/proposal.tex
proposal written

@@
 The expected communication complexity of $S$, denoted $R^\text{cc}_{\text{exp}}(S)$, is the minimum cost of a protocol computing $S$ in expectation.
-\begin{theorem}(Faenza et al.) Provided $\text{rk}_x(S) \neq 0$, $$R^\text{cc}_{\text{exp}}(S) = \log \text{rk}_x(S) + \Theta(1) = \log\text{xc}(P,Q) + \Theta(1).$$
+\begin{theorem}\label{th:communication}(Faenza et al.) Provided $\text{rk}_x(S) \neq 0$, $$R^\text{cc}_{\text{exp}}(S) = \log \text{rk}_x(S) + \Theta(1) = \log\text{xc}(P,Q) + \Theta(1).$$
 The second equality is a direct consequence of Yannakakis' Theorem. 
 \end{theorem}
-\subsection{The goal of this section is Faenza's 2015 result}
 \section{The Knapsack Cover Problem}
-\subsection{The goal here is to describe all the knapsack cover stuff that has been done in the natural space}
-\section{Extended Formulations for Knapsack Cover}
-\subsection{Here we discuss extended formulations for KC, and communication protocols}
-\section{Lower Bounds}
-\subsection{some ideas for lower bounding communication cost of KCI slack matrix}
+The min Knapsack (minK) problem is defined as follows. Given a ground set of items $E$, a non-negative integer demand $D$, a size $s_e$ and cost $c_e$ for each $e \in E$,  solve
+\begin{align*}
+    \min \sum_{e \in E} c_ex_e&\\
+    \text{s.t.} \sum_{e \in E} s_ex_e &\geq D \\
+    x&\in \{0,1\}^E
+\end{align*}
+Despite having a well-known PTAS, a comparable polyhedral description is known.
+Carr, Fleisher, Leung, and Phillips \cite{carr1999strengthening} gave a linear programming formulation with integrality gap $2$ via the Knapsack Cover Inequalities (KC).
+For each $U \subseteq E$, let $D(U) = D-s(U)$ and let $s_e(U) =\min\{s_e, D(U)\}$ for each $e \in E\backslash U$.
+The KC inequalities are
+$$ \sum_{e \in E\backslash U} s_e(U) x_e \geq D(U)\quad \forall U\subseteq E.$$
+One immediately notices that the KC inequalities are exponential in number.
+It is a natural question, and in fact it remains open, to find a linear relaxation of minK of constant integrality gap and polynomial number of constraints.
+Dudycz and Moldenhauer~\cite{dudycz2015approximated} ruled out such a formulation in the natural variable space.
+Hence we need to consider extended formulations.
+\begin{question}\label{question:big}
+Does the polyhedral pair $(P,Q)$ where $P$ is the min Knapsack polytope and $Q$ is linear relaxation of $P$ with the KC inequalities have polynomial extension complexity?
+\end{question}
+\section{Extended Complexity of Knapsack Cover via Communication Protocols}
+\paragraph{}
+Using Theorem \ref{th:communication} Bazzi, Fiorini, Huang, and Svensson~\cite{bazzi2017small} gave a quasi-polynomial size extended formulation for Knapsack Cover, by providing a randomized communication protocol which computes the slack matrix in expectation.
+An interesting consquence of their approach is that the procedure is not constructive (nevertheless they have an intersting cutting plane technique for optimizing over the extended formulation).
+In~\cite{fiorini2017strengthening} they give an iterative procedure which can construct the extended formulation of Bazzi et al.
+The procedure though, is a general black box, not specifically tailored to the minK problem.
+The authors in ~\cite{bazzi2017small} as if there is any combinatorial interpretation of their formulation.
+\begin{question}
+    Is there a combinatorial interpretation of the formulation given in \cite{bazzi2017small}?
+\end{question}
+
+One approach to resolving Question~\ref{question:big} would be to give a better communication protocol for computing $S$ in expectation.
+This would be an ambitious goal.
+Bazzi et al.
+identify a special case in which their protocol gives a polynomial size extended formulation: when the sizes $s_e$ are polynomial in $n$.
+Unfortunately, this special case is already known to have an exact formulation.
+It would be interesting to find a polynomial size extended formulation for less trivial classes of minK.
+\begin{question}
+Are there non-trivial special cases of minK which have a polynomially sized extended formulation, by the protocol in \cite{bazzi2017small} or otherwise?
+\end{question}
+
+Alternatively, we can start approaching Question \ref{question:big} from the other direction.
+Lower bounds on $R^\text{cc}_{\text{exp}}(S)$ would establish lower bounds on the extension complexity of Knapsack Cover.
+As far the writer is aware, no non-trivial lower bounds have been reported in the literature.
+\begin{question}
+    Can we lower bound $R^\text{cc}_{\text{exp}}(S)$ for the Knapsack Cover slack matrix?
+\end{question}
+Braun and Pokutta~\cite{braun2016common} developed an information-theoretic approach to lower bounding $R^\text{cc}_{\text{exp}}(S)$.
+Exploring this approach and seeing if it can be applied to Knapsack Cover would be an interesting concrete starting point.
+
+We conclude with a more modest question.
+In ~\cite{carnes2008primal}and~\cite{carr1999strengthening}, Knapsack Cover constraints are applied to a variety of capacitated covering problems.
+Can we adapt the protocol in \cite{bazzi2017small} to this setting?
+When one family of KC inequalities is applied to the problem may be a bit too straightforward.
+But there is an interesting application of Knapsack Cover inequalities which is less straightforward.
+The best known approximation of scheduling with general job dependent cost functions~\cite{cheung2017primal} use a time-indexed linear programming relaxation which introduces a distinct family of knapsack cover inequalities at each time step.
+\begin{question}
+Can we give a (quasi)-polynomial extended formulation for scheduling with general costs via a randomized communication protocol which computes the slack matrix in expectation?
+\end{question}
 

### 2019-03-08 CUT/paper.tex
Research Paper outline

\documentclass{article}

\usepackage{apacite}

\begin{document}
\begin{abstract}
Research Question: ``How does assessment format impact student learning outcomes, and instructor measurement thereof''.

New Research Question: ``Can innovation assessment formats improve student learning outcomes, and an instructors capacity to measure their achievement.''
\end{abstract}

\section{Introduction}

\section{Particulars of Written Finals}
\subsection{Question Formats}
\subsection{Time Pressure}
\subsection{Open Book}
%can include takehome

\section{Oral Exams}
\subsection{Implementation}
\subsection{Results}
\subsection{Benefits}
\subsection{Drawbacks}
\subsection{Data}

\section{Collaborative Exams}
\subsection{Group}
\subsection{Two-Stage}

\section{Student Showcase}
\subsection{Portfolios}
\subsection{Student-Written Exam Problems}

\section{Stakeholder Impressions}
%how do students and instructors perceive different exam formats?

\section{Recommendations}
\subsection{Designing Exams to Encourage Good Learning Habits}
\subsection{Assessing a Variety of Cognitive Processes}
\subsection{Encouraging Genuine Engagement}
\nocite{*}
\bibliographystyle{apacite}
\bibliography{references.bib}

\end{document}

### 2019-03-08 CS860-CommunicationComplexity/assignments/Chapter7.tex
chapter 7

\documentclass[11pt]{amsart}
\usepackage{amsmath,amsthm,amsfonts,amssymb}
\usepackage{fullpage}
\usepackage[usenames]{xcolor}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{remark}{Remark}
\newtheorem{problem}{Problem}
\newtheorem{challenge}{Challenge Problem}
\newtheorem{open}{Open Problem}
\theoremstyle{plain}

\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calZ}{\mathcal{Z}}
\newcommand{\cost}{\mathrm{cost}}
\newcommand{\dHam}{\mathrm{d}_{\mathrm{Ham}}}
\newcommand{\disc}{\mathrm{disc}}
\newcommand{\discu}{\mathrm{disc_u}}
\newcommand{\Disj}{\textsc{Disj}}
\newcommand{\Eq}{\textsc{Eq}}
\newcommand{\GHD}{\textsc{GHD}}
\newcommand{\GIP}{\textsc{GIP}}
\newcommand{\GT}{\textsc{GT}}
\newcommand{\HD}{\textsc{HD}}
\newcommand{\ICext}{\mathrm{IC}^{\mathrm{ext}}}
\newcommand{\ICint}{\mathrm{IC}^{\mathrm{int}}}
\newcommand{\icostext}{\mathrm{icost}^{\mathrm{ext}}}
\newcommand{\icostint}{\mathrm{icost}^{\mathrm{int}}}
\newcommand{\IP}{\textsc{IP}}
\newcommand{\Index}{\textsc{Index}}
\newcommand{\Maj}{\textsc{Maj}}
\newcommand{\MIP}{\textsc{MIP}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\Rpriv}{R^{\mathrm{priv}}}
\newcommand{\SumIndex}{\textsc{SumIndex}}

\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}
\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}

\title{Communication complexity: Chapter 7 \\ Multiparty communication}
\author{Eric Blais and (YOUR NAME HERE)}

\begin{document}

\maketitle

In this chapter, we move beyond the two-party setting of communication complexity and consider the communication complexity of functions that need to be computed by 3 or more parties.

There are two broad classes of models for multiparty communication complexity. In both of them, the input to the function is divided among the $k$ parties. In the \emph{number in hand} model, each party sees their own input and none of the other players' inputs. This model is a natural extension of our two-party communication model, and it is apparent that it only gets harder when there are more players involved. As a result, all the lower bounds that we have established in previous chapters apply to (natural variants of the functions we have seen for) this setting as well.

In the \emph{number on the forehead} model, by contrast, every player sees \emph{all} the inputs except their own. In this model, adding new players now make the function easier to compute, not harder. The lower bounds we have developed no longer apply, and we need to develop new tools to understand which functions remain hard to compute in this setting.

\newpage 
\section{Number-on-the-forehead model}

In the \emph{$k$-party number on the forehead} (NOF) model of communication complexity, $k$ players $P_1,\ldots,P_k$ aim to compute a function $f : \calX^k \to \calY$ for some finite sets $\calX$ and $\calY$. On input $x_1,\ldots,x_k$, each player $P_i$ sees \emph{all} the inputs \emph{except} $x_i$. A \emph{protocol} determines which player sends the next bit of communication; the bit sent by player $P_i$ depends on the inputs $x_1,\ldots,x_{i-1},x_{i+1},\ldots,x_k$ as well as the preivous messages. All messages communicated by a player are seen by \emph{all} other players. This is known as the \emph{blackboard} setting of communication.

The \emph{cost} of a protocol is the maximum number of bits communicated before the protocol halts, where the maximum is taken over all possible inputs. A protocol \emph{computes} the function $f$ if all the players know $f(x)$ at the end of the protocol. As in the two-player setting, we can represent a protocol as a tree and define the function that it computes as the one determined by the value at each leaf.

The \emph{deterministic communication complexity} of a function $f : \calX^k \to \calY$, which will again be denoted by $D(f)$, is the minimum cost of a $k$-party number-on-the-forehead communication protocol that computes $f$ in the blackboard model.

\begin{theorem}
Almost all functions $f : (\{0,1\}^{n})^k \to \{0,1\}$ have communication complexity $D(f) = \Omega(n)$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
Show that every function $f : (\{0,1\}^{n})^k \to \{0,1\}$ has multiparty communication complexity $D(f) \le n+1$.
\end{exercise}

\begin{open}
Identify any explicit function $f : (\{0,1\}^{n})^k \to \{0,1\}$ with $k \ge \log n$ that has communication complexity $\omega(1)$.
\end{open}

\newpage 
\section{Equality XI}

Some functions can be much easier in the multiparty communication complexity model. Consider for instance the extension of the equality function for $k$ players $\Eq_k : (\{0,1\}^n)^k \to \{0,1\}$ defined by
\[
\Eq^k(x_1,\ldots,x_k) = \begin{cases}
1 & \mbox{if } x_1 = x_2 = \cdots = x_k \\
0 & \mbox{otherwise.}
\end{cases}
\]
\begin{theorem}
For any $k \ge 3$, $D(\Eq^k) \le 2$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Majority Inner Product}

Other functions also have very efficient protocols in the multiparty communication complexity setting.

For three bits $a,b,c$, define the \emph{majority} function $\Maj : \{0,1\}^3 \to \{0,1\}$ to be $\Maj(a,b,c) = 1$ if $a+b+c \ge 2$ and $0$ otherwise. The \emph{Majority inner product} function $\MIP : (\{0,1\}^n)^3 \to \{0,1\}$ is defined by
\[
\MIP(x,y,z) = \bigoplus_{i \in [n]} \Maj(x_i,y_i,z_i).
\]
\begin{theorem}
$D(\MIP) \le 3$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Generalized Inner Product}

Other functions become easier, but non-trivial, when the number of players is greater than 2 but less than $\log n$.

The \emph{generalized inner product} function $\GIP^k : (\{0,1\}^n)^k \to \{0,1\}$ is defined by
\[
\GIP^k(x^{(1)},\ldots,x^{(k)}) = \bigoplus_{i \in [n]} x^{(1)}_i \wedge x^{(2)}_i \wedge \cdots \wedge x^{(k)}_i.
\]
When we view the input to the function as a $k \times n$ matrix, the function takes the value $1$ if and only if the number of all-one columns of this matrix is odd.

\begin{theorem}
$D(\GIP^k) = O(\frac{n}{2^k})$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Cylinder intersections}

Rectangles have a natural generalization in the number-on-the-forehead model in the form of \emph{cylinder intersections}.

\begin{definition}
A \emph{cylinder} in $\calX^k$ \emph{in the $i$th direction} is a set $S \subseteq \calX^k$ such that whether an element $(x^{(1)},\ldots,x^{(k)}) \in S$ or not is independent of the value of $x^{(i)}$.
\end{definition}

\begin{definition}
The set $S \subseteq \calX^k$ is a \emph{cylinder intersection} if $S = \cap_{i \in [n]} S_i$ where each $S_i$ is a cylinder in the $i$th direction.
\end{definition}

Cylinder intersections can be used to bound the multiparty communication complexity of functions via the following fundamental lemma.

\begin{lemma}
Every multiparty protocol that computes $f$ and has cost $c$ partitions the domain of $f$ into at most $2^c$ monochromatic cylinder intersections.
\end{lemma}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Discrepancy}

Having generalized the notion of rectangles to the multiparty setting, we could hope that all the lower bound techniques also generalize. That unfortunately does not appear to be the case. The only technique that generalizes to this setting is discrepancy.

\begin{definition}
For any distribution $\mu$ on $\calX^k$, 
the \emph{$\mu$-discrepancy} of $f : \calX^k \to \{0,1\}$ is
\[
\disc_\mu(f) = \max_{S} 
\left| \mu\big( S \cap f^{-1}(0) \big) - \mu\big( S \cap f^{-1}(1)\big) \right|
\]
where the maximum is over all cylinder intersections $S \subseteq \calX^k$.
\end{definition}

\begin{lemma}
For every function $f : \calX^k \to \{0,1\}$ and distribution $\mu$ on $\calX^k$,
\[
D(f) = \Omega\left( \log \frac{1}{\disc_\mu(f)}\right).
\]
\end{lemma}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
We can also consider \emph{randomized} multiparty communication protocols. Show that for every function $f$, every distribution $\mu$, and every $\epsilon \le \frac12$, we also have
\[
R_\epsilon(f) \ge \log \frac{1-2\epsilon}{\disc_\mu(f)}.
\]
\end{exercise}

\newpage 
\section{Generalized inner product II}

We can use the discrepancy bound to give a lower bound on the communication complexity of the generalized inner product function.

\begin{theorem}
$D(\GIP^k) = \Omega(\frac{n}{4^k})$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\newpage 
\section{Multiparty Simultaneous Message Passing}

In the \emph{simultaneous message passing} (SMP) model of multiparty communication complexity, we again consider the number-on-the-forehead model with blackboard communication with one change: all the players now write their messages on the blackboard simultaneously. Or, equivalently, all the bits communicated by the protocol depend on the other players inputs but \emph{not} on the messages previously sent.

The minimum cost of an SMP multiparty communication protocol that computes a function $f : \calX^k \to \{0,1\}$ is denoted by $D^\parallel(f)$.

A particularly interesting function to study in the multiparty SMP model is the $k$-party generalization of the $\Index$ function where $\Index^k : \{0,1\}^{n^{k-1}} \times [n]^{k-1} \to \{0,1\}$ where player $P_1$'s input is a $(k-1)$-dimensional array of bits, the $k-1$ other players all receive an index in $[n]$, and
\[
\Index^k(A,i_2,\ldots,i_k) = A[i_2,\ldots,i_k].
\]

\begin{theorem}
$D^\parallel(\Index^k) = \Omega(n/k)$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
Show that $D(\Index^k) = \Theta(\log n)$.
\end{exercise}

\newpage 
\section{Sum Index}

Another interesting variant of the index function is the 3-party function $\SumIndex : \{0,1\}^n \times [n] \times [n]$ defined by
\[
\SumIndex(A,i,j) = A[ i \oplus j]
\]
with $i \oplus j$ being the bitwise OR operation. (It is useful to consider only the case where $n$ is a power of $2$ for simplicity.)

\begin{theorem}
$D^\parallel(\SumIndex) = \Omega(\sqrt{n})$.
\end{theorem}

\begin{proof}
\replacethistext{Enter your proof here.}
\end{proof}

\exercises

\begin{exercise}
Prove that $D^\parallel(\SumIndex) = o(n)$.
\end{exercise}

\begin{open}
The best bounds on the SMP complexity of the $\SumIndex$ function are
\[
\Omega(\sqrt{n}) \le D^\parallel(\SumIndex) \le O(n^{0.73}).
\]
Can you improve either the upper or the lower bound?
\end{open}

\end{document}

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
Problem 8

@@
 \begin{theorem}
-$D(\GIP^k) = O(\frac{n}{2^k})$.
+$D(\GIP^k) = O(\frac{nk}{2^k})$.
 \end{theorem}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a multiparty SMP protocol for $\Index^k$ with communication cost $c$.
+Suppose for a contradiction that $c < \frac{n}{k}$.
+Consider the following communication problem: Alice receives $A \in \{0,1\}^{n^{k-1}}$ and Bob receives $i_2,\dots, i_k$ and they need to compute $A[i_2,\dots, i_k]$.
+We will present a \emph{one-way} deterministic protocol for this problem using $\pi$ with cost less than $n^{k-1}$.
+Using the same argument as we did in Chapter $6$ for the one-way deterministic communication cost of $\Index$, we can show that this problem was one-way deterministic communication complexity at least $n^{{k-1}})$.
+Hence we will have reached a contradiction.
+
+It remains to give the promised protocol.
+For each player $j \in \{2, \dots, k\}$ Alice enumerates the possible input $k-2$-tuples $t$ that $j$ would see the other players in $\{2,\dots, k\}$ holding, and for each such possiblity Alice sends to Bob the message $j$ would send in $\pi$ seeing $A$ and $t$.
+For each player this total communication cost is $cn^{k-2}$.
+Thus Alice sends strictly less than
+$$ckn^{n-2} < n^{k-1}$$
+bits it total.
+Using the bits Alice sent, Bob can determine the messages that all players would sent in $\pi$ under input $A,i_2,\dots, i_k$, and can output the referees decision, computing $\Index^k$.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Enter a proof
+
 \end{proof}

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
Starting 1 but Williams is closing

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+The number of possible functions $f$ is counted by the number of binary strings of length $|(\{0,1\}^n)^k|$. Hence the number of possible functions $f$ is 
+$$2^{2^{kn}}.$$
 \end{proof}
@@
 bits it total.
-Using the bits Alice sent, Bob can determine the messages that all players would sent in $\pi$ under input $A,i_2,\dots, i_k$, and can output the referees decision, computing $\Index^k$.
+Using the bits Alice sent, Bob can determine the messages that all players would sent in $\pi$ under input $A,i_2,\dots, i_k$, and can output the referees decision, computing $\Index^k$.`'
 \end{proof}

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
problem 1

@@
 $$2^{2^{kn}}.$$
+For a protocol tree of height $h$, there are at most $2^{h+1}$ nodes in the protocol tree.
+Each node belongs to one of $k$ players, and is associated with a function whose output is binary and whose input domain is the space of all other players inputs.
+There are
+$$2^{2^{(k-1)n}}$$ 
+such functions. Thus in total there are at most
+$$(k\cdot 2^{2^{(k-1)n}})^{2^{h+1}} = k^{2^{h+1}}2^{2^{(k-1)n + h + 1}}$$
+possible protocol using $h$ bits of communication. Since we require protocols to be deterministic, the ratio of functions computable with $h$ bits of the total possible functions is 
+$$ 2^{2^{h+1}\log(k) - 2^{kn}}2^{2^{(k-1)n +h + 1 -kn}}.$$
+If $h = o(n)$ the limit of this ratio as $n$ approaches infinity is $0$. Thus the theorem holds.
 \end{proof}

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
Problems 2 and 3

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Consider the protocol where player $1$ outputs $1$ if $x_2 = x_3 = \dots x_k$ and outputs $0$ otherwise.
+The protocol terminates with output $0$ in the latter case.
+In the former, player $2$ outputs $1$ if $x_1 = x_3$ and outputs $0$ otherwise, and these outputs correspond to the decision of the protocol in their respective cases.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+We use the following identity which can be verified by case analysis
+$$\Maj(x_i,y_i,z_i) = x_iy_i + y_iz_i + x_iz_i \mod 2.$$
+Using this identity we see that
+$$\MIP(x,y,z) = \bigoplus_{i\in[n]}\Maj(x_i,y_i,z_i) = \left(\bigoplus_{i\in [n]} x_iy_i\right)\oplus\left(\bigoplus_{i\in [n]} y_iz_i\right)\oplus\left(\bigoplus_{i\in [n]} x_iz_i\right).$$
+On the right hand side, the first term can be computed by player $3$ and output, the second by player $1$, and the third by player $2$. With all $3$ bits anyone can then compute $\MIP(x,y,z)$.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+
 \end{proof}

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
up to problem 6

@@
 \begin{proof}
+Let $M$ be the corresponding $k\times n$ matrix to the input, where each player cannot see a particular row. Our protocol is as follows.
 
+We first partition the columns into blocks into floor of $\frac{n}{2^{k-1}-1}$ blocks.
+We will give a $k$ cost protocol for each block, and summing the parities of the blocks mod $2$ will yield the parity of the number of all-ones columns in $M$, within the desired communication cost bound.
+
+Consider block $B$ of $M$.
+By the pigeonhole principle, since $B$ has fewer than $2^{k-1}$ columns, there is a $k-1$-bit string $s$ which does not appear in $B$ delete row $1$.
+Player $1$ communicates $s$ at a cost of $O(\log(n))$.
+By permuting the rows, we may assume that $0s$, which does not appear in the columns of $B$, is of the form $0^\ell 1^{k-\ell}$ for some $\ell$.
+
+Let $c_i$ denote the number of columns of $B$ of the form $0^i1^i$.
+Observe that player $i$ can compute $c_{i-1} + c_i$, but not $c_{i-1}$ and $c_i$ since they do not know the bit in the $i$th position.
+Our protocol proceeds as follows: for each player $i$ from $1$ to $\ell-1$: player $i$ outputs the parity of $c_{i-1} + c_i$.
+Then player $\ell$ outputs the parity of  $c_{\ell - 1}$ (since we knows $c_\ell = 0$).
+By back-substitution, any player can now compute the parity of $c_0$, as desired.
+Since $\ell \leq k$, this protocol uses at most $k$ bits.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a protocol which computes $f$ at cost $c$.
+Let $\ell$ be a leaf of $\pi$.
+Let $S(\ell)$ be the set of inputs in $\calX^k$ for which $\pi$ outputs $\ell$.
+Clearly $\{S(\ell)\}_\ell$ is a partition of the domain of $f$, since $\pi$ is deterministic.
+
+Since there are $2^c$ such $S(\ell)$, it only remains to verify that $S(\ell)$ is a cylindrical intersection. We define
+$$S_i(\ell):=\{x \in \calX^k: \exists y^i s.t. x^1\dots x^{i-1}y^ix^{i+1}\dots x^k \in S(\ell)\}.$$
+Since containment in $S_i(\ell)$ is independent of $x^i$, $S_i(\ell)$ is a cylinder. Thus we need to show
+$$S(\ell) = \bigcap_{i\in[k]} S_i(\ell).$$
+On the one hand for $x\in S(\ell)$, we have $x\in S_i(\ell)$ immediately (by taking $y^i = x^i$) for all $i$.
+
+On the other hand, let $x \in \bigcap_{i\in [k]} S_i(\ell).$ Fix $i\in[k]$. We have for some $y^i$ that
+$$x^1\dots x^{i-1}y^ix^{i+1}\dots x^k \in S(\ell).$$
+We also have for some $y^1$ that
+$$y^1x^2\dots x^i\dots x^k\in S(\ell).$$
+But then in the first case, player $i$ can switch their input to $x^i$ without the protocol tree reaching a different leaf. So $x \in S(\ell)$ as desired.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+Let $\pi$ be a protocol computing $f$ at cost $D(f)$. Let $\mu$ be a distribution over $\calX^k$. Since $\pi$ is deterministic,
+$$1 = \Pr_{x\sim \mu}(\pi(x) = f(x)).$$
+Let $t = 2^{D(f)}$ and consider a partition of the domain of $f$ into $t$ cylindrical intersections $S_1, \dots, S_t$.
+
+Then we have
+\begin{align*}
+1 &=\Pr(\pi(x) = f(x))\\
+&=\sum_{i=1}^t\Pr(\pi(x) = f(x)\text{ and } x\in S_i) \\
+&= \sum_{i=1}^t\Pr(\pi(x) = f(x)\text{ and } x\in S_i) - \Pr(\pi(x) \neq f(x)\text{ and } x\in S_i)\\
+&\leq \sum_{i=1}^t|\mu(S\cap f^{-1}(0)) - \mu(S\cap f^{-1}(1))|\\
+&= t\cdot \disc{_\mu}(f).
+\end{align*}
+Rearraning we obtain the desired inequality.
 \end{proof}

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
draft

@@
 bits it total.
-Using the bits Alice sent, Bob can determine the messages that all players would sent in $\pi$ under input $A,i_2,\dots, i_k$, and can output the referees decision, computing $\Index^k$.`'
+Using the bits Alice sent, Bob can determine the messages that all players would sent in $\pi$ under input $A,i_2,\dots, i_k$, and can output the referees decision, computing $\Index^k$.
 \end{proof}
@@
 \begin{proof}
-Enter a proof
+Suppose for a contradiction there exists a simultaneous message passing protocol $\pi$ for $\SumIndex$ using $c = o(\sqrt{n})$ bits.
+We may assume that players $2$ and $3$ announce their $j$ and $i$ respectively in their messages at a cost of $\log(n)$ bits.
+This is without loss of generality since $c + \log(n) = o(\sqrt{n})$.
+Now since the referee learns from players $2$ and $3$ everything that player $1$ sees, player $1$ does not need to send anything in $\pi$.
+So we assume player $1$'s message is empty.
 

### 2019-03-15 CS860-CommunicationComplexity/assignments/Chapter7.tex
draft 2

@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+We seek to upper bound $\disc_\mu(\GIP^k)$ by $2^{\frac{-n}{4^k}}$.
+Using the previous Theorem this yields the result.
+We proceed by induction on $k$.
+The base case, $k=2$, is satisfied by our bound on Chapter $3$ on $\IP$.
+So suppose $k>3$ and the  desired bound holds for smaller $k$.
+For function $f$, and cylindrical intersection $S\subseteq \calX^k$ we use $\disc_\mu^S(f)$ to denote the discrepancy of $S$:
+$$\disc_\mu^S(f) = |\mu(S\cap f^{-1}(0)) - \mu(S\cap f^{-1}(1)))|.$$
+Let $\mu^k$ be the uniform distribution over $\calX^k = (\{0,1\}^n)^k$. Let $S = S_1\times \dots \times S_k\subseteq \calX^k$ be a cylindrical intersection.
+$$\disc_{\mu^k}^S(\GIP^k) = \sum_{i=1}^k\sum_{x^i \in S_i} \frac{1}{2^{nk}}(-1)^{\GIP^k(x^1\dots x^k)}.$$
+
+Let $\text{proj}_{k-1}^{x^k}(x^1\dots x^k) = y^1\dots y^k$ where $y^j_i = x^j_i \wedge x^k_i$. Then
+$$\GIP^k(x^1\dots x^k) = \GIP^{k-1}(\text{proj}_{k-1}^{x^k}(x^1\dots x^k)).$$
+So we have
+$$\disc_{\mu^k}^S(\GIP^k) = \sum_{x^k\in S_k}\frac{1}{2^n}\sum_{i=1}^{k-1}\sum_{x^i \in S_i}\frac{1}{2^{n(k-1)}}(-1)^{\GIP^{k-1}(\text{proj}_{k-1}^{x^k}(x^1\dots x^k))} = \sum_{x^k \in S_k}\frac{1}{2^n}\disc^{\text{proj}_{k-1}^{x^k}(S)}_{\mu^{k-1}}(\GIP^{k-1}).
+$$
+We apply Cauchy-Schwarz and our inductive hypothesis obtaining...
 \end{proof}
@@
 
+Using this simplified protocol $\pi$ we...
 \end{proof}

### 2019-04-03 CS860-CommunicationComplexity/project/presentation.tex
Starting slides

--- 2019-02-28 CS860-CommunicationComplexity/project/proposal.tex
@@
-\documentclass[letterpaper,12pt,oneside,onecolumn]{article}
-\usepackage[margin=1in, bottom=1in, top=1in]{geometry} %1 inch margins
-\usepackage{amsmath, amssymb, amstext}
-\usepackage{fancyhdr}
-\usepackage{mathtools}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\usepackage{theorem}
+\documentclass[10pt]{beamer}
+
+\usetheme[progressbar=frametitle]{metropolis}
+\usepackage{appendixnumberbeamer}
+\usepackage[normalem]{ulem}
+\usepackage{booktabs}
+\usepackage[scale=2]{ccicons}
+
+\usepackage{pgfplots}
+\usepgfplotslibrary{dateplot}
+
+\usepackage{xspace}
+\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
+
 \usepackage{tikz}
-\usepackage{tkz-berge}
-\usepackage[colorlinks]{hyperref}
-\usepackage{bookmark}
+\title{Communication Complexity of Knapsack Cover}
+\subtitle{CCFest}
+% \date{\today}
+\date{}
+\author{By W. Justin Toth}
+\institute{CS860 - Communication Complexity}
+\titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.png}}
+
+\setbeamercolor{alerted text}{fg=blue}
 
 %Macros
-\newcommand{\A}{\mathbb{A}} \newcommand{\C}{\mathbb{C}}
+\newcommand{\A}{\mathbb{A}} 
 \newcommand{\D}{\mathbb{D}} \newcommand{\F}{\mathbb{F}}
 \newcommand{\N}{\mathbb{N}} \newcommand{\R}{\mathbb{R}}
-\newcommand{\T}{\mathbb{T}} \newcommand{\Z}{\mathbb{Z}}
+ \newcommand{\Z}{\mathbb{Z}}
 \newcommand{\Q}{\mathbb{Q}}
@@
 
-\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
+\begin{document}
 
-\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
-\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
-\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}
+\maketitle
 
-%\renewcommand{\thesection}{\lecnum.\arabic{section}}
-%\renewcommand{\theequation}{\thesection.\arabic{equation}}
-%\renewcommand{\thefigure}{\thesection.\arabic{figure}}
-
-\newtheorem{fact}{Fact}[section]
-\newtheorem{lemma}[fact]{Lemma}
-\newtheorem{theorem}[fact]{Theorem}
-\newtheorem{definition}[fact]{Definition}
-\newtheorem{corollary}[fact]{Corollary}
-\newtheorem{proposition}[fact]{Proposition}
-\newtheorem{claim}[fact]{Claim}
-\newtheorem{exercise}[fact]{Exercise}
-\newtheorem{note}[fact]{Note}
-\newtheorem{conjecture}[fact]{Conjecture}
-\newtheorem{question}[fact]{Question}
-
-\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
-\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
-\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}
-
-%END MACROS
-%Page style
-\pagestyle{fancy}
-
-\listfiles
-
-\raggedbottom
-
-\lhead{CS860 Proposal}
-\rhead{W. Justin Toth} %
-\renewcommand{\headrulewidth}{1pt} %heading underlined
-%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)
-
-\begin{document}
-\section{Extended Formulations and Communication Complexity}
-\paragraph{}
-Consider a polytope $P\subseteq \R^n$ and a polyhedron $Q\subseteq \R^n$ containing $P$ (called a \emph{polyhedral pair}).
-For intuition, think of $P$ as the convex hull of the set of solutions to some NP-hard problem, for which we do not know a compact linear programming description, and think of $Q$ as a relaxation of $P$ for which are interested in finding a compact description of.
-\paragraph{}
-We say that a system 
-$$E^\leq x + F^\leq y \leq g^\leq, E^= x + F^= y = g^=$$
-in $\R^{n+k}$ is an \emph{extended formulation} of $(P,Q)$ if the polyhedron
-$$R:=\{x \in \R^n: \exists y \in \R^k: E^\leq x + F^\leq y \leq g^\leq, E^= x + F^= y = g^=\}$$
-contains $P$ and is contained in $Q$.
-The number of inequalities in the system determines the size of the extended formulations.
-The \emph{extension complexity} of $(P,Q)$, $xc(P,Q)$, is the minimum size of an extended formulation of $(P,Q)$.
-\paragraph{}
-Usually one studies the case $P=Q$, but for our purposes we are interested in the extension of complexity of a specific relaxation, and hence need the fully general definiton above.
-Yannakakis' seminal result \cite{yannakakis1991expressing} gives an algebraic expression of $xc(P,Q)$.
-To state the result we will need two more pieces of terminology.
-
-\subsection{The goal of this section is Faenza's 2015 result}
-\section{The Knapsack Cover Problem}
-\subsection{The goal here is to describe all the knapsack cover stuff that has been done in the natural space}
-\section{Extended Formulations for Knapsack Cover}
-\subsection{Here we discuss extended formulations for KC, and communication protocols}
-\section{Lower Bounds}
-\subsection{some ideas for lower bounding communication cost of KCI slack matrix}
-
-\bibliography{references.bib}
-\bibliographystyle{plain}
+\begin{frame}{A First Slide}
+\end{frame}
 \end{document}

### 2019-04-05 CS860-CommunicationComplexity/project/presentation.tex

@@
 \usepackage{tikz}
-\title{Communication Complexity of Knapsack Cover}
-\subtitle{CCFest}
+\title{Can Communication Protocols Find Small LPs?}
+\subtitle{Knapsack Cover, Extended Formulations, and Communication Complexity}
 % \date{\today}
@@
 \begin{document}
-
+\setbeamercolor{background canvas}{bg=white}
 \maketitle
 
-\begin{frame}{A First Slide}
+\begin{frame}{The Min Knapsack Problem}
+Given $n$ items with individual sizes, $s_i$, and costs, $c_i$, for $i \in [n]$.
+\vfill
+Also given a demand $D$. Want to compute the \alert{Min Knapsack}
+
+\begin{align*}
+\min\ &\sum_{i\in [n]} c_i x_i \\
+\text{s.t.} &\sum_{i \in [n]} s_i x_i \geq D \\
+&x \in \{0,1\}^n.
+\end{align*}
+\vfill
+Standard Linear Programming relaxation has an unbounded integrality gap.
+\end{frame}
+\begin{frame}{Knapsack Cover Inequalities}
+LP relaxation can be strengthened to an integrality gap of $2$ via all Knapsack Cover Inequalities.
+\vfill
+Knapsack Cover Inequality for $B\subseteq [n]$ is the inequality of the form 
+$$\sum_{i \in [n]\backslash B} s_i(B) x_i \geq D(B),$$
+where $D(B) = \max\{D-\sum_{i\in B} s_i, 0\}$ and $s_i(B) = \min\{s_i, D(B)\}.$
+\vfill
+\textbf{Big Open Question:} Is there a polynomial size formulation for Knapsack Cover LP?
+\end{frame}
+\begin{frame}{Extended Formulations}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+    \begin{figure}
+        \includegraphics[scale=0.075]{std_ex.png}
+    \end{figure}
+\end{column}
+\begin{column}{0.5\textwidth}
+\textbf{Extension Complexity:} Min number of facets to describe a polytope as the projection of a higher dimensional polyhedron.
+\end{column}
+\end{columns}
+\end{frame}
+\begin{frame}{Communication Protocols for the Slack Matrix}
+\end{frame}
+\begin{frame}{Results}
 \end{frame}

### 2019-04-05 CS860-CommunicationComplexity/project/presentation.tex
Slides drafted

--- 2019-04-03 CS860-CommunicationComplexity/project/presentation.tex
@@
 \begin{document}
-
+\setbeamercolor{background canvas}{bg=white}
 \maketitle
 
-\begin{frame}{A First Slide}
+\begin{frame}{The Min Knapsack Problem}
+Given $n$ \alert{items} with individual \alert{sizes}, $s_i$, and \alert{costs}, $c_i$, for $i \in [n]$.
+\vfill
+Also given a \alert{demand} $D$. \onslide<2->{Want to compute the \alert{Min Knapsack}
+
+\begin{align*}
+\min\ &\sum_{i\in [n]} c_i x_i \\
+\text{s.t.} &\sum_{i \in [n]} s_i x_i \geq D \\
+&x \in \{0,1\}^n.
+\end{align*}}
+\vfill\onslide<3->{
+Standard Linear Programming relaxation has an \alert{unbounded integrality gap.}}
+\end{frame}
+
+\begin{frame}{Knapsack Cover Inequalities}
+LP relaxation can be \alert{strengthened} to an integrality gap of $2$ by adding all \alert{Knapsack Cover Inequalities}.
+\vfill\onslide<2->{
+\textbf{Knapsack Cover Inequality:} for $B\subseteq [n]$, the inequality of the form 
+$$\sum_{i \in [n]\backslash B} s_i(B) x_i \geq D(B),$$
+where $D(B) = \max\{D-\sum_{i\in B} s_i, 0\}$ and $s_i(B) = \min\{s_i, D(B)\}.$}
+\vfill\onslide<3->{
+\textbf{Big Open Question:} Is there a \alert{polynomial size formulation} for Knapsack Cover LP?}
+\end{frame}
+\begin{frame}{Extended Formulations}
+    \begin{columns}
+        \begin{column}{0.4\textwidth}
+    \begin{figure}
+        \includegraphics[scale=0.075]{std_ex.png}
+    \end{figure}
+\end{column}
+\begin{column}{0.6\textwidth}\onslide<2->{
+\textbf{Extension Complexity:} $xc(P) :=$ Min \alert{number of facets} to describe polytope $P $ as the projection of a \alert{higher dimensional polyhedron.}}
+\vfill\onslide<3->{
+\textbf{Theorem:} (Yannakakis) $xc(P) = \Theta(\text{rk}_+(S))$ where $S$ is a \alert{slack matrix} of $P$, and $\text{rk}_+$ denotes non-negative rank.}
+\end{column}
+\end{columns}
+\vfill\onslide<4->{
+\alert{Knapsack Cover:} the slack matrix has entries 
+$$ S_{A,B} = \sum_{i \in A\backslash B} s_i(B) - D(B)$$
+for knapsack solution $A$, and $B\subseteq [n]$.}
+\end{frame}
+\begin{frame}{Communication Protocols for the Slack Matrix}
+Consider communication problem for $S : \{\text{Knapsacks}\}\times [n] \rightarrow \Z_{\geq 0}$.
+\vfill\onslide<2->{
+We say \alert{private random} communication protocol $\pi$ computes $S$ in \alert{expectation} if
+$$\mathbb{E}[\pi(A,B)] = S_{A,B}$$
+for all $(A,B) \in \text{domain}(S)$.}
+\vfill\onslide<3->{
+\textbf{Theorem:} (Faenza, Fiorini, Grappe, Tiwary) $$\min_\pi(\text{cost}(\pi)) = \log(\text{rk}_x(S)) + \Theta(1) = \log(xc(P)) + \Theta(1)$$}
+\end{frame}
+\begin{frame}{Results}
+\textbf{Theorem:} (Bazzi, Fiorini, Huang, Svensson) There is a protocol which computes $S$ in expectation with cost $O(\log^2(n))$.
+\vfill
+\hrule
+\vfill\onslide<2->{
+\textbf{Goal:} Provide \alert{non-trivial lower bounds} on communication protocols computing $S$ in expectation.}
+\vfill\onslide<3->{
+\textbf{Lemma:} (T.) If protocol $\pi$ uses \alert{$O(1)$ bits of randomness} to compute $S$ in expectation then $\text{cost}(\pi) = \Omega(n)$.}
+\vfill\onslide<4->{
+\textbf{Lemma:} (T.)  Let $R$ be the set of \alert{one-way} protocols where \alert{only Alice} has private randomness, which compute $S$ in expectation.
+Then $$\min_{\pi\in R}(\text{cost}(\pi)) = \log(\text{rk}_x(S)) + \Theta(1).$$}
 \end{frame}

### 2019-04-10 CS860-CommunicationComplexity/assignments/Chapter2.tex
Correcting Chapter 2

@@
 We can think of $\IP(x,y)$ as the parity of $|X\cap Y|$.
-Let $b = \IP(x,y) = |X\cap Y|$.
-Half of all sets in $2^{[n]}$ have odd intersection with $X$ and half of all sets in $2^{[n]}$ have even intersection with $X$.
+Let $$b = \IP(x,y) = |X\cap Y| \mod 2.$$ Half of all sets in $2^{[n]}$ have odd intersection with $X$ and half of all sets in $2^{[n]}$ have even intersection with $X$.
 This yields a bound on $|B|$ of $2^{n/2}$.
@@
 Let $L$ be the set of leaves of this tree.
-As observed in the previous lecture when discussing the partition bound, each leaf corresponds to an $f$-monochromatic rectangle.
+As observed in the previous chapter when discussing the partition bound, each leaf corresponds to an $f$-monochromatic rectangle.
 Let $L'\subseteq L$ be the leaves at which $1$ is output.
@@
 and thus
-$$\rank(M_f) \leq \sum_{\ell \in L'} M^\ell \leq |L'| \leq |L| = 2^{D(f)}.$$
+$$\rank(M_f) \leq \sum_{\ell \in L'} \rank(M^\ell) \leq |L'| \leq |L| = 2^{D(f)}.$$
 In the above inequality we used that $\rank(M^\ell) = 1$ for all $\ell \in L'$.
@@
 \begin{proof}
-	Observe that $(M_\GT)_{i,j} = 1 $ iff $i >j$ ($(M_\GT)_{i,j} = 0$ otherwise).
+	Observe that 
+	$$(M_\GT)_{i,j} = \begin{cases}
+		1, &\text{if } i>j \\
+		0, &\text{otherwise}
+	\end{cases}.$$
 	Thus $M_\GT$ is entirely ones below the diagonal, and entirely zeroes on the diagonal and above.
@@
 The first is that the rank of a $0$-$1$ matrix over $\mathbb{R}$ is at least its rank over $\mathbb{F}_2$.
-In symbols $$\rank_\mathbb{R}(f) \geq \rank_{\mathbb{F}_2}(f).$$
+In symbols $$\rank(f) \geq \rank_{\mathbb{F}_2}(f).$$
 This follows since a set of linearly independent columns over $\mathbb{F}_2$ remains independent over $\mathbb{R}$.

### 2019-04-10 CS860-CommunicationComplexity/project/presentation.tex

@@
 $$ S_{A,B} = \sum_{i \in A\backslash B} s_i(B) - D(B)$$
-for knapsack solution $A$, and $B\subseteq [n]$.}
+for any knapsack solution $A$, and $B\subseteq [n]$.}
 \end{frame}
@@
 \vfill\onslide<3->{
-\textbf{Theorem:} (Faenza, Fiorini, Grappe, Tiwary) $$\min_\pi(\text{cost}(\pi)) = \log(\text{rk}_x(S)) + \Theta(1) = \log(xc(P)) + \Theta(1)$$}
+\textbf{Theorem:} (Faenza, Fiorini, Grappe, Tiwary) $$\min_\pi(\text{cost}(\pi)) = \log(\text{rk}_+(S)) + \Theta(1) = \log(xc(P)) + \Theta(1)$$}
 \end{frame}
@@
 \textbf{Lemma:} (T.)  Let $R$ be the set of \alert{one-way} protocols where \alert{only Alice} has private randomness, which compute $S$ in expectation.
-Then $$\min_{\pi\in R}(\text{cost}(\pi)) = \log(\text{rk}_x(S)) + \Theta(1).$$}
+Then $$\min_{\pi\in R}(\text{cost}(\pi)) = \log(\text{rk}_+(S)) + \Theta(1).$$}
 \end{frame}

### 2019-04-12 CS860-CommunicationComplexity/assignments/Chapter3.tex
Correcting Chapter 2
Corruption bound proofs remain in fixing chapter 3

@@
 Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
-$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  = \sum_{R\in \mathcal R} \max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right| \geq\sum_{R \in \mathcal{R}} \left| \sum_{(x,y) \in R} \big(M^{\pm}_f\big)_{x,y} \right|= \sum_{R \in \mathcal R} |R|$$
+\begin{align*}\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  &= \sum_{R\in \mathcal R} \max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|\\
+&\geq\sum_{R \in \mathcal{R}} \left| \sum_{(x,y) \in R} \big(M^{\pm}_f\big)_{x,y} \right|\\
+&= \sum_{R \in \mathcal R} |R|
+\end{align*}
 using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
@@
 Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
-$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) = f(x,y) \mid(x,y) \in R_i) = \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.$$
+\begin{align*}\Pr[\pi(x,y) = f(x,y)] &= \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) \\
+&= f(x,y) \mid(x,y) \in R_i)\\
+&= \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.
+\end{align*}
 Symmetrically
@@
 For convenience, define $M:= M^{\pm}_\IP$ and let $\mathbf{1}_S$ denote characteristic vector of a set $S\subseteq [n]$. Now consider an entry $x,y$ of $M^TM$:
-$$M^TM_{x,y} = \sum_{z\in\{0,1\}^n} (-1)^{x\cdot z}(-1)^{y\cdot z} = \sum_{z \in \{0,1\}^n}(-1)^z\cdot(x+y) = \begin{cases}
+$$M^TM_{x,y} = \sum_{z\in\{0,1\}^n} (-1)^{x\cdot z}(-1)^{y\cdot z} = \sum_{z \in \{0,1\}^n}(-1)^{z\cdot(x+y)} = \begin{cases}
 (-1)2^{n-1} + (1)2^{n-1}, &\text{if $x\neq y$} \\
@@
 \begin{proof}
-Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
+Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$.
+
+Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
 $$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
@@
 \begin{proof}
-I'm not really sure what the right distribution to take is, but intuitively I want to distribute the probability mass evenly between $(x,y)$ pairs with $x\neq y$ and pairs with $x=y$ so Alice and Bob have to exchange some bits to do better than guessing.
-From the matrix of this function I want half the probability mass uniformly distributed on the diagonal, and half uniformly distributed on all off-diagonal entries.
+    We define $\mu : \{0,1\}^n  \times \{0,1\}^n \rightarrow [0,1]$ as follows. For any $(x,y) \in \{0,1\}^n\times \{0,1\}^n$ let
+    $$ \mu(x,y) = \begin{cases}
+        \frac{1}{2^{n+1}}, &\text{if } x=y \\
+        \frac{1}{2^{n+1}(2^n-1)}, &\text{otherwise}.
+    \end{cases}.$$
+    To see that $\mu$ is a probability distribution observe that
+    \begin{align*}\sum_{x,y \in \{0,1\}^n\times \{0,1\}^n} \mu(x,y) &= \sum_{x\in\{0,1\}^n}\mu(x,x) + \sum_{x\neq y \in \{0,1\}^n\times \{0,1\}^n} \mu(x,y)\\
+     &= 2^n \frac{1}{2^{n+1}} + (2^{2n}-2^n)\frac{1}{2^{n+1}(2^n-1)}\\
+    &= \frac{1}{2} + \frac{1}{2} = 1.
+    \end{align*}
+    Let $R = A\times B \subseteq \{0,1\}^n\times\{0,1\}^n$ be a rectangle such that
+    $$\left|\mu(R\cap \Eq^{-1}(0)) - \mu(R\cap \Eq^{-1}(1))\right| = \disc_\mu(\Eq).$$
+    Choose such $R$ so that $|R|$ is minimized. We proceed by case distinction.
+
+    First suppose that $\mu(R\cap \Eq^{-1}(1)) \geq \mu(R\cap \Eq^{-1}(0)).$ If $(x,x), (y,y) \in R\cap \Eq^{-1}(1)$ then $(x,y), (y,x) \in R\cap \Eq^{-1}(0)$. Thus we have that 
+    \begin{align*}&\mu(R\cap \Eq^{-1}(1)) - \mu(R\cap \Eq^{-1}(0))\\ 
+        &\leq \left|R\cap \Eq^{-1}(1)\right|(\frac{1}{2^{n+1}})-\left(\left|R\cap \Eq^{-1}(1)\right|^2-\left|R\cap \Eq^{-1}(1)\right|\right)(\frac{1}{2^{n+1}(2^n-1)}).
+    \end{align*}
+    This upperbound is maximized when $|R\cap \Eq^{-1}(1)|=2^{n-1}$. Therefore in this case $\disc_\mu(\Eq) \leq \frac{1}{8} + o(1).$
+
+    Now consider the other case, where $\mu(R\cap \Eq^{-1}(0)) > \mu(R\cap \Eq^{-1}(1)).$ Suppose there exists $x \in \{0,1\}^n$ such that $(x,x) \in R.$ Consider $R' = A\times B\backslash\{x\}.$ Observe that
+    \begin{align*}
+        &\left|\mu(R'\cap \Eq^{-1}(0)) - \mu(R'\cap \Eq^{-1}(1))\right|\\ & = \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1))  + \mu(x,x)- \sum_{y\in A\backslash\{x\}} \mu(x,y)\\
+        &= \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1)) + \frac{1}{2^{n+1}} - |A\backslash\{x\}|\frac{1}{2^{n+1}(2^n-1)} \\
+        &\geq \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1)) + \frac{1}{2^{n+1}} - \frac{1}{2^{n+1}} \\
+        &= \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1))\\
+        &= \disc_\mu(\Eq).
+    \end{align*}
+    This contradicts that $|R|$ is minimized. Therefore $A\cap B = \emptyset$. Thus we have
+    \begin{align*}
+        \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1)) &\leq \mu(R\cap \Eq^{-1}(0))\\
+        &= |R\cap \Eq^{-1}(0)|\frac{1}{2^{n+1}(2^n-1)} \\
+        &\leq |A|(2^n-|A|)\frac{1}{2^{n+1}(2^n-1)}\\
+        &\leq \frac{2^{n}(2^{n/2}-1)}{2^{n+1}(2^n-1)}\\
+        &= O\left(\frac{1}{2^{n/2}}\right).
+    \end{align*}
+
+    Hence we have shown that $\disc_\mu(\Eq) \leq \frac{1}{8} + o(1) \leq \frac{1}{8}\left(\frac{2^n}{2^n-1}\right)$.
 \end{proof}
@@
 \begin{proof}
-\replacethistext{Enter your proof here, if you complete it.}
+For $r \in \{0,1\}^n$ let  $\pi_r$ be the deterministic protocol where 
+$$\pi_r(x,y) = \begin{cases} 1, &\text{if } (x\cdot r)  = (y\cdot r) \\ 0, &\text{otherwise}\end{cases}.$$
+This protocol has communication cost $O(1)$ since Alice can send Bob $x\cdot r$ in $1$ bit, and Bob can send Alice $y \cdot r$ in $1$ bit.
+
+Consider a uniformly random $r \in \{0,1\}$.
+If $x=y$ then $\pi_r(x,y) = 1$.
+If $x\neq y$ then $\pi_r(x,y) =1$ with probability $\frac{1}{4}$.
+Let $X_r(x,y)$ be the indicator random variable indicating if $\pi_r(x,y)$ fails on input $(x,y)$.
+Thus we see that for any $(x,y)$,
+$$\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
+Thus we have
+$$\mathbb{E}_{(x,y)\sim \mu}\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
+Rearranning order of expectation yields
+$$\mathbb{E}_{r}\mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
+And thus there exists an $r \in \{0,1\}$ such that
+$$\Pr_\mu(\pi_r \text{ fails}) \leq \mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
 \end{proof}

### 2019-04-13 CS860-CommunicationComplexity/assignments/Chapter3.tex
Corruption bound

@@
 $$\mathbb{E}_{(x,y)\sim \mu}\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
-Rearranning order of expectation yields
+Rearranging order of expectation yields
 $$\mathbb{E}_{r}\mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
@@
 $\mu$ on $\calX \times \calY$.
-If there exist parameters $\alpha, \beta > 0$ for which
+If there exist parameters $\alpha, \beta \in (0,1)$ for which
 every rectangle $R \subseteq \calX \times \calY$ satisfies 
@@
 Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$.
-Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf.
+Let $\mathcal{R}$ be the set of rectangles corresponding to the path from the root of $\pi$ to each respective leaf.
+For $b \in \{0,1\}$ let $\mathcal{R}_b \subseteq \mathcal{R}$ be the set of rectangles such that the protocol outputs $b$ at the corresponding leaf.
 
-$$1= \sum_{i=1}^t \mu(R_i) = \sum_{i=1}^t \mu(R_i \cap f^{-1}(0)) + \mu(R_i\cap f^{-1}(1)) \geq \alpha - t \beta +\mu(f^{-1}(1))$$
-Thus
-$$t \geq \frac{\alpha + \mu(f^{-1}(1)) -1}{\beta}.$$
-Now we claim that 
-$$\alpha + \mu(f^{-1}(1)) -1 \geq \alpha\cdot \mu(f^{-1}(1)) -\epsilon.$$
-This claim is equivalent to
-$$1-\epsilon \leq \alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)).$$
-Since 
-$$\alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)) = 1-(1-\alpha) \mu(f^{-1}(0))$$
-the claim holds since the probability of failure is at least $(1-\alpha)f^{-1}(0)$ \replacethistext{(I didn't have time to formalize)}.
+Since $\pi$ has success probability at least $1-\epsilon$,
+$$\sum_{R \in \mathcal{R}_0} \mu(R\cap f^{-1}(1)) + \sum_{R \in \mathcal{R}_1} \mu(R\cap f^{-1}(0)) \leq \epsilon.$$
+By the lemma hypothesis,
+$$\sum_{R \in \mathcal{R}_1}\mu(R\cap f^{-1}(0)) \geq \alpha\sum_{R \in \mathcal{R}_1}\mu(R) -\beta\left|\mathcal{R}_1\right|.$$
+Rearranging and combining the two inequalities above yields
+\begin{align*}
+    \beta|\mathcal{R}_1| &\geq \alpha\sum_{R \in \mathcal{R}_1}\mu(R) - \epsilon + \sum_{R \in \mathcal{R}_0}\mu(R\cap f^{-1}(1))\\
+    &=\alpha \mu(f^{-1}(1)) - \epsilon + (1-\alpha)\sum_{R \in \mathcal{R}_0}\mu(R\cap f^{-1}(1))+\alpha \sum_{R \in \mathcal{R}_1}\mu(R\cap f^{-1}(0))\\
+    &\geq \alpha \mu(f^{-1}(1)) - \epsilon
+\end{align*}
+The last inequality above follows since $\alpha, 1-\alpha$ are non-negative and $\mu$ is a probability distribution.
+Since $|\mathcal{R}_1| \leq |\mathcal{R}| = 2^{D^\mu_\epsilon(f)}$ we have the desired result.
 \end{proof}

### 2019-04-14 CS860-CommunicationComplexity/assignments/Chapter3.tex
draft of fix for chapter 3

@@
 \begin{proof}
-\replacethistext{Enter your proof here, if you complete it.}
+Suppose $|A| > 2^{-\delta\sqrt{n}}{n\choose\sqrt{n}}.$ Let $M \in \R^{A\times B}$ be the matrix where
+$$M_{S,T} = \Disj(S,T).$$
+By the hypothesis of the proposition, at most $\alpha$ fraction of the entries of $M$ are such that $S\cap T \neq \emptyset$. We may assume without loss of generality that for every $S\in A$,
+$$\Pr_{T \in B}[S\cap T \neq \emptyset] \leq 2\alpha.$$
+If this assumption does not hold, we can remove the rows in $M$ for which this fails from $A$, reducing the size of $A$ by factor of at most $2$ (a factor we may compensate for in our choice of $\delta$).
+
+Choose sets $S_1,\dots, S_{\sqrt{n}/3} \in A$ such that
+$$|S_i\backslash (S_1\cup\dots\cup S_{i-1})| \geq \frac{\sqrt{n}}{2}$$
+for all $i$. For small enough $\delta$, such $S_i$ exist by the lower bound on the size of $A$.
+
+At least half of the sets $T \in B$ satisfy 
+$$\Pr_{i \in [\sqrt{n}/3]}[ S_i\cap T \neq \emptyset] \leq 4\alpha,$$
+since each row of $M$ has at most $2\alpha$ fraction of its entries $0$. But by our choice of $S_1,\dots, S_{\sqrt{n}/3}$ the number of such sets $T \in B$ is at most
+$${\sqrt{n}/3\choose 4\alpha\sqrt{n}/3}{n-(1-4\alpha)n/6\choose \sqrt{n}}.$$
+For $\delta>0$ small enough, this yields the desired bound on $|B|$.
 \end{proof}

### 2019-04-14 CUT/paper.tex
Correcting Chapter 2
Wrote an introduction, should maybe have one more paragraph

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 
 \usepackage{apacite}
+\usepackage{setspace}
+\usepackage{hyperref}
 
+\doublespace
+\title{Working Title - CUT Research Paper}
+\author{W. Justin Toth}
 \begin{document}
+\maketitle
 \begin{abstract}
@@
 New Research Question: ``Can innovation assessment formats improve student learning outcomes, and an instructors capacity to measure their achievement.''
+
+\textit{With a particular focus on mathematics, stem}
 \end{abstract}
@@
 
-\section{Particulars of Written Finals}
-\subsection{Question Formats}
-\subsection{Time Pressure}
-\subsection{Open Book}
+\paragraph{}
+A well-known truism in human social interaction is that how you say something often has as much, if not more, importance that the specific content of your speech.
+Yet when instructors are designing evalutions for their courses, it is all too common that the focus is squarely on what questions to ask to cover the course content, rather than how the problems should be presented to the students.
+In this survey we seek to synthesize the academic literature on the ``how'', rather than ``what'', of assessment methods and provide guidance for instructors on the impact that exam format can have on student learning outcomes and their ability to measure the attainment thereof.
+\paragraph{}
+In Section \ref{sec:written} we study the particulars of written finals.
+For written finals we will concentrate on two areas of focus in the literature.
+The first is on the use of multiple choice versus constructed response questions, and the second is on allowing students to use reference material such as the course textbook or handwritten notes during the exam.
+In Section \ref{sec:communication} we study the use of oral communication during exams.
+This breaks down in to two subsections.
+In Subsection \ref{subsec:oral} we survey the literature on implementing interview-style oral exams  between student and instructor in the classroom, and in Subsection \ref{subsec:collab} we discuss student-to-student collaboration in exams with a particular focus on the innovative two-stage exam format.
+We conclude by with Section \ref{sec:recommendations} where present recommendations for instructors on choosing exam formats for their courses based on the previously surveyed literature.
+\section{Individual Written Finals}\label{sec:written}
+\subsection{Multiple Choice}
+\subsection{Reference Material}
+\subsection{Frequency}
 %can include takehome
 
-\section{Oral Exams}
-\subsection{Implementation}
-\subsection{Results}
-\subsection{Benefits}
-\subsection{Drawbacks}
-\subsection{Data}
+\section{Oral Exams and Collaboration}\label{sec:communication}
+\subsection{Oral Exams}\label{subsec:oral}
+\paragraph{Implementation}
+\paragraph{Results}
+\paragraph{Benefits}
+\paragraph{Drawbacks}
+\paragraph{Data}
 
-\section{Collaborative Exams}
-\subsection{Group}
-\subsection{Two-Stage}
+\subsection{Collaborative Exams}\label{subsec:collab}
+\paragraph{Group}
+\paragraph{Two-Stage}
 
-\section{Student Showcase}
-\subsection{Portfolios}
-\subsection{Student-Written Exam Problems}
-
-\section{Stakeholder Impressions}
-%how do students and instructors perceive different exam formats?
-
-\section{Recommendations}
+\section{Recommendations}\label{sec:recommendations}
+content
 \subsection{Designing Exams to Encourage Good Learning Habits}

### 2019-04-14 CUT/paper.tex
Wrote up multiple choice section

@@
 
+\usepackage{hyperref}
 \usepackage{apacite}
 \usepackage{setspace}
-\usepackage{hyperref}
 
@@
 \subsection{Multiple Choice}
+\paragraph{}
+When preparing a written final exam, a very basic choice appears early on in the design of the test.
+That is the choice between having students select their response to a question or generate their response.
+This is the decision between using a Multiple Choice question format or using a Constructed Response question format.
+In a Multiple Choice question, students are given the question prompt and a finite list, usually of length between two and five, of potential responses they are to select from to answer the question prompt.
+In a Constructed Response question, the students are given the question prompt, but instead of a list of responses they are simply given a blank space in which to write their response.
+Constructed Response questions can vary more significantly in format than Multiple Choice, encompassing for example fill-in-the-bank, short answer, and essay questions.
+Both formats have their potential advantages and disadvantages, which have been well-studied in the education research literature.
+\paragraph{}
+The literature has highlighted the pragmatic benefits of Multiple Choice questions~\cite{simkin2005multiple}.
+With machine-grading, Multiple Choice exams can be graded very quickly, shortening the delay between students writing the exam and receiving their feedback.
+Automated grading of Multiple Choice problems also reduces the grading burden of instructors and teaching assistants, freeing up their time to (hopefully) concentrate on improving the quality of the course in other ways.
+Multiple Choice exams also have the advantage of providing an objective scoring system during grading, removing all grader subjectivity from the process.
+This can increase the perceived fairness of the exam.
+From the perspective of students, Multiple Choice exams have the advantage of inducing less anxiety.
+Students perceive Multiple Choice exams as being easier and quicker to prepare for, and are comforted by having potential answers available to them and the option they have to guess at a solution~\cite{zeidner1987essay}.
+\paragraph{}
+Despite the advantages above, Multiple Choice questions have many disadvantages over Constructed Response questions.
+It is difficult to assess critical thinking skills with Multiple Choice questions~\cite{martinez1999cognition}, and it is essentially impossible to assess creativity with Multiple Choice questions.
+Students seem to be aware of this on some level, expressing a perception that Multiple Choice exams assess a more shallow, knowledge-level understanding of the course material~\cite{scouller1998influence}.
+In line with this viewpoint, students predominantly favor shallow-learning strategies like memorization for exams consisting primarily of Multiple Choice problems.
+In contrast, students perceive Construhcted Response exams as testing deeper understanding, and use deeper learning strategies to study for Constructed Response exams~\cite{scouller1998influence}
+\paragraph{}
+Stanger-Hall hypothesized that Multiple Choice exams hinder the development of critical-thinking skills in her introductory biology class students~\cite{stanger2012multiple}.
+Traditionally these classes are evaluated using Multiple Choice only exams.
+To test this hypothesis, she changed the exams to feature Multiple Choice problems alongside several Constructed Response problems of the short answer variety.
+She predicted that this adaptation would result a change in how students studied for the exams, and their performance on critical-thinking questions.
+\paragraph{}
+The author defined critical-thinking skills as those skills associated with the middle levels of Bloom's taxonomy: application, analysis, evaluation, and synthesis.
+The experiment was performed using two sections of introductory biology given by the same instructor in the same semester to large classes of about $200$ students.
+The sections were entirely indentical in course material (lectures, assignments, readings) with the exception of the final exam.
+One section received a Multiple Choice only final, and the other received a Multiple Choice plus short answer Constructed Response final.
+Through a series of four surveys, Stanger-Hall evaluated student approaches to studying and perception of the importance of critical-thinking skills in introductory biology.
+The Constructed Response group reported significantly higher cognitively active learning behaviours during weeks outside of the exam study period, and this was reflected in their significantly higher test scores, in particular on questions associated with critical-thinking levels of Bloom's taxonomy~\cite{stanger2012multiple}.
+\paragraph{}
+A major weakness of Multiple Choice exams is their tendency to over-estimate student learning achievement~\cite{funk2011multiple}.
+The authors conjectured that grades from Multiple Choice exams were artificially inflated due to the relative ease of recognizing a correct answer versus generating it oneself, and students employing educated guessing strategies on exams.
+They conducted a study wherein $50$ students in a personality psychology course were randomly assigned to take a $10$-time short answer Constructed Response pretest or protest on a $50$ question Multiple Choice exam.
+To control for content and grading subjectivity, they used the same question prompts on the short answer and Multiple Choice problems.
+\paragraph{}
+The results were very strong.
+The mean score on short answer pretests was $34.24$ percent, followed by $77.1$ percent on the Multiple Choice exam.
+For the group where short answer followed the Multiple Choice exam, the student group achieved a mean score of $78.64$ percent on Multiple Choice and $59.84$ percent on the Constructed Response.
+These results show it is a significantly more difficult task to have a working knowledge of the correct solutions, than it is to recognize them on sight in a Multiple Choice environment.
+When the short answer problems followed the Multiple Choice problems, students were able to use some of their recognition of correct answers to solve more short answer problems than the pretest group, but were unable to recover their Multiple Choice test scores.
+The questions in this test were aimed at the remember and understand levels of Bloom's taxonomy, indicating that even for the most basic forms of learning, Multiple Choice questions due a poor job of measuring student knowledge retention.
 \subsection{Reference Material}
-\subsection{Frequency}
+
+%\subsection{Frequency}
 %can include takehome

### 2019-04-15 CUT/paper.tex
literature review on reference material in exams

@@
 \subsection{Reference Material}
-
-%\subsection{Frequency}
-%can include takehome
+\paragraph{}
+Another fundamental decision an instructor needs to make when designing an exam is whether or not they will allow students to make use of reference material during the exam.
+Perhaps the most common decision taken is to not allow reference material of any sort during an exam, the so-called Closed Book exam.
+On the other end of the reference material permissiveness spectrum is the Open Book exam.
+In an Open Book exam, students are permitted to use the primary reference material for the course, usually the recommended course textbook but alternatively all course notes may be permitted.
+For the purpose of our discussion an Open Book exam will be defined by the unmitigated allowance of a reference text where students can expect to find the answer to any reasonable basic fact about the subject they may need for an exam.
+Another option exists in the middle between Closed and Open Book exams.
+This option is what we will call Cheatsheet exams.
+In a Cheatsheet exam students are allowed to prepare a restricted set of notes to bring to the exam.
+Usually these notes are allowed to have any content the student may want on them, but with space and formatting restrictions which allow the notes they carry in to have significantly more compact information than an Open Book exam.
+Comparing the efficacy of Open versus Closed Book exams in the literature has led to mixed results.
+\paragraph{}
+On the positive side for Open Book exams, it has been shown that they lead to significantly lower test anxiety and stress among students leading up to the exam TODOCITE(Broyles, Cyr, and Korsen).
+This reduction in tension is hypothesized to lead to better study habits, such as focusing on understanding over rote memorization.
+It was found by (TODOCITE Weber, McBee and Krebs) that there are slight gains in student performance on Open Book exams over Closed Book exams.
+They also corroborate the findings of (Broyles, Cyr, and Korsen) that student test anxiety is reduced in an Open Book exam setting.
+\paragraph{}
+A study of the empirical literature on Open Book exams conducted by TODOCITE (Theophilides and Dionysiou) supports the assertions that Open Book exams lead to lower test anxiety, and less focus on memorization over understanding.
+But their study also shows that Open Book examinations do not lead to higher test scores, in particular where critical-thinking questions are concerned.
+The findings of TODOCITE (Agarwal et al.) are even more lukewarm.
+They find that the effect of testing on student retention is equal between Open Book and Closed Book exams.
+Separate studies by TODOCITE (Ioannidou) and (Boniface) found that student reliance on their permitted notes during an Open Book exam correlated with lower test scores.
+The findings of Ioannidou further indicate that critical-thinking questions are not effected with any statistical significance by the variable of whether an exam is Open or Closed Book.
+Moore and Jensen) TODOCITE found that Open Book exams diminish long term retention of material among Biology students, and promote shallow-learning study strategies.
+The results of TODOCITE (Heijne-Penninga) support these findings, indicating that Closed Book exams do a better job of promoting deep-learning strategies among students than Open Book Exams.
+\paragraph{}
 

### 2019-04-15 CUT/paper.tex
references for open book literature survey

@@
 \paragraph{}
-On the positive side for Open Book exams, it has been shown that they lead to significantly lower test anxiety and stress among students leading up to the exam TODOCITE(Broyles, Cyr, and Korsen).
+On the positive side for Open Book exams, it has been shown that they lead to significantly lower test anxiety and stress among students leading up to the exam~\cite{broyles2005open}.
 This reduction in tension is hypothesized to lead to better study habits, such as focusing on understanding over rote memorization.
-It was found by (TODOCITE Weber, McBee and Krebs) that there are slight gains in student performance on Open Book exams over Closed Book exams.
-They also corroborate the findings of (Broyles, Cyr, and Korsen) that student test anxiety is reduced in an Open Book exam setting.
+It was found by~\cite{broyles2005open} that there are slight gains in student performance on Open Book exams over Closed Book exams.
+They also corroborate the findings of~\cite{broyles2005open} that student test anxiety is reduced in an Open Book exam setting.
 \paragraph{}
-A study of the empirical literature on Open Book exams conducted by TODOCITE (Theophilides and Dionysiou) supports the assertions that Open Book exams lead to lower test anxiety, and less focus on memorization over understanding.
+A study of the empirical literature on Open Book exams conducted by~\cite{theophilides2000study} supports the assertions that Open Book exams lead to lower test anxiety, and less focus on memorization over understanding.
 But their study also shows that Open Book examinations do not lead to higher test scores, in particular where critical-thinking questions are concerned.
-The findings of TODOCITE (Agarwal et al.) are even more lukewarm.
+The findings of~\cite{agarwal2008examining} are even more lukewarm.
 They find that the effect of testing on student retention is equal between Open Book and Closed Book exams.
-Separate studies by TODOCITE (Ioannidou) and (Boniface) found that student reliance on their permitted notes during an Open Book exam correlated with lower test scores.
+Separate studies by~\cite{ioannidou1997testing} and~\cite{boniface1985candidates} found that student reliance on their permitted notes during an Open Book exam correlated with lower test scores.
 The findings of Ioannidou further indicate that critical-thinking questions are not effected with any statistical significance by the variable of whether an exam is Open or Closed Book.
-Moore and Jensen) TODOCITE found that Open Book exams diminish long term retention of material among Biology students, and promote shallow-learning study strategies.
-The results of TODOCITE (Heijne-Penninga) support these findings, indicating that Closed Book exams do a better job of promoting deep-learning strategies among students than Open Book Exams.
+\cite{moore2007open} found that Open Book exams diminish long term retention of material among Biology students, and promote shallow-learning study strategies.
+The results of~\cite{heijne2011directing} support these findings, indicating that Closed Book exams do a better job of promoting deep-learning strategies among students than Open Book Exams.
 \paragraph{}

### 2019-04-15 CUT/paper.tex
Finished Reference Materials section

@@
 \paragraph{}
+The effects of Open, Closed, and Cheatsheet exams on retention were measured by~\cite{gharib2012cheat} in an Introductory Psychology class.
+Across sections, students were given either an Open Book, Closed Book, or Cheatsheet exam.
+There were $99$ students participating in the study, and they were divided evenly into the three exam formats.
+Two weeks after writing a midterm with their designated exam format, the students participating in the study were given a surprise quiz to measure retention of the material from the exam.
+They found that there were no significant differences between retention across exam formats, despite scores on Open Book and Cheatsheet exams being higher than Closed Book exams.
+\paragraph{}
+In an Introduction to Statistics course~\cite{block2012discussion} sought to compare the use of Closed Book, Open Book, and Cheatsheet exams.
+Over the course of four different semesters, Block varied the exam format and measured student performance and perceptions of the course.
+In Fall $2006$, the exam format was a traditional Closed Book exam.
+In Spring $2007$, it was an Open Book exam, and again in Fall $2007$.
+The difference being, in Fall $2007$, the instructor emphasized to the students that the Open Book exam would be more challenging than a corresponding Closed Book exam and they shoud prepare accordingly using by trying for a deeper understanding of the material.
+Finally, in Spring $2008$, the Cheatsheet format was tried.
+Students were allowed to put as much information as they could fit on a typically index card and bring that into the exam with them.
+Their results showed higher performance on the Closed Book and Cheatsheet exams, and lower performance on the Open Book exams.
+The students preferred Open Book exams, except when they were told that they would be more difficult exams.
+The Cheatsheet exam reigned supreme, having comparable student enjoyment to the first semester of Open Book exams, and strong student preparation for the exams in line with what was seen for Closed Book exams.
+The cheatsheet itself encourages good study habits among students, forcing them to summarize and prioritze content, deciding what is truly important to write down.
 

### 2019-04-15 CUT/paper.tex
Exposition on Oral Exams

@@
 \usepackage{setspace}
-
+\usepackage[margin=1in]{geometry}
 \doublespace
@@
 \end{abstract}
-
+\tableofcontents
 \section{Introduction}
@@
 \subsection{Oral Exams}\label{subsec:oral}
-\paragraph{Implementation}
-\paragraph{Results}
-\paragraph{Benefits}
-\paragraph{Drawbacks}
-\paragraph{Data}
-
+In the discipline of Mathematics, like many post-secondary level subjects, Written Exams seem to be the predominant form of assessment at North American Universities~\cite{gold1999assessment}.
+This stands in contrast to European institutions where Oral Exams are more common.
+We loosely define an Oral Exam as a dialogue between Student(s) and Instructor(s) used for evaluation purposes.
+In this sense Oral Exams can take on many forms.
+They could be in the style of a presentation followed by questioning, akin to a Doctoral thesis defence, or they could be more conversational, with the student and instructor having a back-and-forth discussion about a problem.
+The questions for an Oral Exam may be given to the students beforehand, or appear spontaneously at the time of the exam.
+\paragraph{}
+In~\cite{videnovic2017oral} the author interviews $7$ mathematics professors of different culture backgrounds about the their perspectives on Written versus Oral Exams.
+One common point of agreement found among the subjects were type of skills each exam format was better suited for.
+Mathematical skills can broadly be categorized into one of two types: Procedural or Conceptual.
+Procedural skills reflect technical competency with the instruments of Mathematics.
+An example of a question testing Procedural knowledge would be ``Compute the integral $\int\sec x dx.$".
+Conceptual skills reflect understanding of the ways in which individual units of mathematical fact interrelate and support each other.
+For example, ``What does a matrix being invertible tell you about its determinant, and consequently its eigenvalues?".
+In Videnovic's survey, it was widely agreed that Written Exams are more proficient in testing Procedural skills, and Oral Exams more proficient in testing Conceptual skills.
+\paragraph{}
+Ten years of experience giving Oral Exams, consisting of one-on-one discussions between professor and student, in Upper Year Mathematics courses at the United States Air Force Academy were reported on in~\cite{boedigheimer2015individual}.
+The author identifies three major benefits to using oral exams: depth of study, real-time personal adjustment, and two-way student-instructor feedback.
+Students in courses with Oral Exams reported a desire to not want to embarass themselves in a discussion with their instructor, and as a consequence they studied more thoroughly and emphasized conceptual understanding.
+Since the exams and evaluation of student performance therein are occuring simultaneously, the instructor can quickly acertain student ability and adjust the exam accordingly.
+A struggling student may receive hints from the instructor helping steering them back on course, or a high performing student may receive more challenging questions probing their understanding at a deeper level than an average exam.
+The dialogue making up the exam is an excellent two-way feedback mechanism, making immediately apparent to both student and instructor where conceptual holes are in the students understanding allowing both to make corrections right after the exam without having to wait for the grading process to complete for feedback to be received.
+\paragraph{}
+In a first course on Abstract Algebra~\cite{capaldi2014non} reports on her experience using ``Non-Traditional" teaching methods, one of which was Oral Exams.
+She identifies many of the same advantages as Boedigheimer, but includes an additional valuable insight.
+In her implementation, each student gave two oral exams, one in the first half of the semester, and one in the second half.
+Capaldi observed that many students were visibly anxious during the first exam, but by the second exam they had developed a dramatic growth in confidence in presenting mathematical proof.
+Many students in Capaldi's course reported that the Oral Exams made a huge difference in their understanding of their own mathematical abilities.
+\paragraph{}
+The anxiety Capalid noted is a common problem.
+Boedigheimer also observed that phenomenon.
+Another issue with Oral Exams is the instructor workload.
+In a class of $35$ to $40$ students, Oral Exams can easily span a couple of full days.
+Nevermind a first year course with $200$ plus students.
+This difficulty can spill over into missed class time, with the instructor often needing to bock off classes during the semester just for Oral Exams.
+\paragraph{}
+Oral Exams provide an intersting opportunity for student collaboration in the exam environment.
+\cite{odafe2006oral} gave an Oral Exam to students in a first Algebra course where students were tasked with solving a math problem in small groups.
+Odafe noted unique advantages and disadvantages to the group collaboration element.
+The obvious advantage is that students gain practice working on solutions as part of a team.
+This skill would be critical if they hoped to pursue mathematics as a career in the future, and is one that is not trained nearly enough in most students mathematical education.
+Teams that worked well together would have each student in the group contributing actively, and they were able to build off each others ideas learning and solving more in the exam than any one student would normally be able to.
+The disadvantages come with groups that did not work well together.
+If one member of a group was unable or unwilling to contribute, the students in that group were severely hampered by their teammate.
+Groups which poor communication or teamwork reported high dissatisfaction with the Oral Exam component of the course in the course evaluation surveys.
+Most students did not have that experience though, and overall student evaluations of the Oral Exam format were highly positive.
+\paragraph{}
+Over $10$ years giving interview style Oral Exams in Mathematics, \cite{boedigheimer2015individual} collected data on the correlation between Oral Exam performance and performance on other, more traditional, forms of assessment occuring in the classroom.
+Interestingly, Oral Exam and Written Exam performance were only weakly correlated (as opposed to say the strong correlation between assignment and Written Exam performance).
+This is in line with ~\cite{ahmed1999assessing} who assessed correlation between Oral and Written Exam performance at Cambridge, and finding a correation coefficient of only $0.53$.
+This seems to suggest that Oral and Written Exams test different abilities.
+This supports the beliefs of the professors in Videnovic's survey that Written and Oral Exam test Procedural and Conceptual skills respectively.
+But more importantly it highlights that when we favour only one exam format, students whose aptitudes line up with the unused format may fall through the cracks.
 \subsection{Collaborative Exams}\label{subsec:collab}

### 2019-04-16 CUT/paper.tex
A first draft

@@
 \begin{abstract}
-Research Question: ``How does assessment format impact student learning outcomes, and instructor measurement thereof''.
-
-New Research Question: ``Can innovation assessment formats improve student learning outcomes, and an instructors capacity to measure their achievement.''
-
-\textit{With a particular focus on mathematics, stem}
+We study of question of how assessment formats impact student study behaviours, learning outcomes, and perception of the courses they take.
+We particularly focus on Mathematics and Science courses, but the findings in this survey are general enough to be widely applicable.
+We study Multiple Choice versus Constructed Response, how much Reference Material to allow in the exam room, Oral versus Written Exams, and the Two Stage method for injecting collaborative learning into exams.
 \end{abstract}
@@
 \section{Individual Written Finals}\label{sec:written}
+Two common questions when it comes to Written Exams are what sort of question format should be used, and should students be allowed to have access to the textbook during the exam?
+In this Section we draw from the education research literature to find answers to these questions.
 \subsection{Multiple Choice}
@@
 The Cheatsheet exam reigned supreme, having comparable student enjoyment to the first semester of Open Book exams, and strong student preparation for the exams in line with what was seen for Closed Book exams.
-The cheatsheet itself encourages good study habits among students, forcing them to summarize and prioritze content, deciding what is truly important to write down.
+The cheatsheet itself encourages good study habits among students, forcing them to summarize and prioritize content, deciding what is truly important to write down.
 
 \section{Oral Exams and Collaboration}\label{sec:communication}
+Getting students to speak up in the classroom is a challenge in its own right.
+In this Section we consider exam formats which involve students communicating verbally with the instructor and each other.
+We will find that these approaches break from the norm in North American institutions, but provide many benefits to the students and instructors who can overcome the new challenges they bring.
 \subsection{Oral Exams}\label{subsec:oral}
@@
 But more importantly it highlights that when we favour only one exam format, students whose aptitudes line up with the unused format may fall through the cracks.
-\subsection{Collaborative Exams}\label{subsec:collab}
-\paragraph{Group}
-\paragraph{Two-Stage}
+\subsection{Two-Stage Exams}\label{subsec:collab}
+\paragraph{}
+In the previous Subsection (Subsection \ref{subsec:oral}) we briefly discussed student collaboration in the context of an Oral Exam, highlighting the unique challenges and opportunities associated with that exam format.
+We would now like to discuss an innovative exam format which merges peer-to-peer collaborative discussions with a traditional Written Exam.
+This is the so-called Two Stage Exam format.
+\paragraph{}
+In advance of a Two Stage Exam, students agree on (or are assigned) a team of three to four students who they will collaborate with during the exam period.
+This already has the advantage of giving students a pre-made study group for exam preparation.
+The actual exam period is divided into two stages.
+During the first stage (approximately $2/3$ to $3/4$ of the exam period), students individually write a traditional Written Exam.
+At the end of the first stage, their exam papers are collected as normal.
+For the second stage (the remaining time of the exam period), students join up in their pre-arranged groups and each group is given a single copy of the same test they wrote during the first stage.
+The students are then tasked with coming to a consensus submission for the second stage exam paper.
+Notice that since the problems are repeats of the first stage problems, they take considerably less time to write their solutions the second time around.
+\paragraph{}
+At UBC \cite{rieger2014examinations} tested this Two Stage Exam format in their introductory physics class of $178$ students.
+The high pressure environment of the first stage of the exam primes students and heavily invests them in finding the answers to the problems.
+This carries over to the second stage, where students are well prepared to discuss their solution with the group.
+These discussions yield immediate feedback on student work from their peers.
+Struggling students receive clarification on problems they did not understand, and stronger students cement their understanding by explaining solutions to their peers.
+The main reason that Two Stage Exams are great way to get these benefits over regular in class collaborative learning is that they have a much higher engagement rate, nearly $100$ percent according to Rieger and Heiner.
+They surveyed student opinions of their implementation of Two Stage Exams and found that $76$ percent of students had a generally positive opinion of the format.
+In particular, the majority of students expressed surprise at how effective the exam method was for them learning from their mistakes.
+\paragraph{}
+Students did express some concerns with the Two Stage exam format of~\cite{rieger2014examinations}.
+Some found it emotionally distressing to immediately be made aware of their exam mistakes.
+Others expressed concern about weak students unfairly gaining marks from their strong group-mates during the second stage.
+Finally some students did not appreciate weak or unprepared teammates being unhelpful during the second stage of the exam.
 
 \section{Recommendations}\label{sec:recommendations}
-content
-\subsection{Designing Exams to Encourage Good Learning Habits}
-\subsection{Assessing a Variety of Cognitive Processes}
-\subsection{Encouraging Genuine Engagement}
-\nocite{*}
+\paragraph{}
+In this paper we discussed some basic decisions instructors need to make about their exam formats and how they can impact the learning outcomes of students in their courses.
+We have shown that exam format is tremendously impactful on students' experience of the course.
+\paragraph{}
+We saw that while Multiple Choice exams are imminently pragmatic, they encourage shallow learning of course material, and are detrimental to long term retention when compared to Constructed Response exam problems.
+We recommend that instructors relegate Multiple Choice to assessing the lowest levels of Bloom's taxonomy, and even then to only use them in very large classes where the practical benefits outweigh the potential drawbacks.
+\paragraph{}
+We also considered the question of allowing students to have access to Reference Material during an exam.
+We found that Open Book exams do not promote appreciably higher retention of material over Closed Book exams, and can even lead to poor studying practices, like under-preparing for the exam or not employing deep-learning study strategies.
+The effect of Open Book exams on anxiety seems to be generally positive, reducing test anxiety among students compared to Closed Book exams.
+But that effect can be lost if you tell students their Open Book exam will be harder than the analogous Closed Book exam.
+If test anxiety is a major concern for you, then Open Book exams can help to solve that problem.
+The exam format which seems to capture the best of both worlds is the Cheatsheet exam format.
+It encourages students to employ similar deep-learning strategies to a Closed Book exam.
+In fact, it naturally pushes students to prioritize material and assess what is important since they have limited space to record their notes.
+It also has comparable test anxiety reducing properties to Open Book exams.
+We highly recommend instructors consider this format for their Written exams.
+When implementing Cheatsheet exams, ensure to have a standard format for the cheatsheet.
+Consider limiting both note size and font size, as some students will attempt to squish the entire course on the note somewhat defeating the purpose.
+\paragraph{}
+In Section \ref{sec:communication} we studied the potential for Oral Exam formats and Collaboration.
+We found that Oral Exams provide many impressive benefits, in particular for Mathematics classes.
+They increase student confidence, allow for immediate corrective feedback and clarification of student understanding, and even allow the instructor to tailor the exam to each individual student based on their level of capability.
+Perhaps most importantly of all though, there appears to be evidence that they test a different set of aptitudes than Written Exams.
+Both the skills which are amenable to Oral Exams, and those amenable to Written Exams are desirable skills which should be rewarded and encouraged in the evaluation process.
+\paragraph{}
+We strongly recommend that, when practical, instructors incorporate an Oral Exam component into their course.
+This ``when practical" caveat exists solely because it can be very time-intensive to implement Oral Exams, espcially if your class is large relative to your Teaching Assistant support.
+The only other reason to avoid Oral Exams may be that your course content is almost entirely mechanical in nature.
+Courses with little emphasis on conceptual understanding will not see much benefit from Oral Exams.
+But if those caveats do not exist for your course, or you are not dissuaded by them then Oral Exams can greatly benefit your students.
+There are a couple tips gathered from the articles we surveyed on Oral Exams which may be of help to instructors.
+If you have multiple graders for Oral Exams, have some ``practice'' Oral Exams all your Teaching Assistants grade at the start of the term, and discuss the grades as a group to make sure you are all callibrated and equally critical in your assessments.
+Do not schedule all the Oral Exams on the same day.
+Spread them out as appointments over a few days, or a week, otherwise the exercise will be mentally fatiguing for the instructor and the quality of exams occuring later in the day will suffer.
+Some students may freeze out of nervousness, especially if it's their first time giving an Oral Exam.
+If you know this is a new experience for your class, have some leniency and consider allowing students who break down in this way to reschedule another attempt.
+Finally, have a rubric which you are grading by present and write on it during the Oral Exam.
+Do not try to remember and fill it out after the exam.
+\paragraph{}
+Lastly, in Subsection \ref{subsec:collab} we presented an innovative approach to injecting collaborative learning into Written Exams: the Two Stage Exam format.
+For instructors who have collaborative learning as a high priority goal, this is a relatively easy to implement change to your exam format which can help achieve that goal.
+It has been found that Two Stage Exams encourage dramatically higher student engagement in the collaborative activity that in-class collaborations.
+These exams go a long way to to encouraging students to seek feedback on their solutions, identifying missing elements in their understanding, and learn how to correct their mistakes.
+As with any group work task, poor teamwork skills can lead to drawbacks.
+To mitigate these, consider a policy where student's group stage mark will not be lower than their individual first stage grade.
+Similarly, try to make the group work portion just valuable enough that students have an incentive to give an earnest effort, but no higher than that.
+\paragraph{}
+Whether we like it or not, grades are central to the University student experience and are a major motivator for student behaviour.
+Thus it is not surprising that the way we design our assessments has an impact on how students study, the skills they develop, and ultimately the learning outcomes they achieve in our courses.
+Hence we should be cognizant of the link between exam format and learning outcomes, and adjust accordingly.
+In this way we can use student grade motivation to our advantage, designing our assessment formats to funnel students into the positive learning behaviours we all hope they will adopt.
+%\subsection{Designing Exams to Encourage Good Learning Habits}
+%\subsection{Assessing a Variety of Cognitive Processes}
+%\subsection{Encouraging Genuine Engagement}
+
 \bibliographystyle{apacite}

### 2019-04-16 CUT/paper.tex
Document formatting

@@
 \doublespace
-\title{Working Title - CUT Research Paper}
-\author{W. Justin Toth}
+\title{Impact of Assessment Format on Student Learning Outcomes}
+\author{W. Justin Toth\\Combinatorics and Optimization\\20411831}
 \begin{document}
@@
 \end{abstract}
+\newpage
 \tableofcontents
+\newpage
 \section{Introduction}
@@
 \paragraph{}
-In~\cite{videnovic2017oral} the author interviews $7$ mathematics professors of different culture backgrounds about the their perspectives on Written versus Oral Exams.
+In~\cite{videnovic2017oral} the author interviews seven mathematics professors of different culture backgrounds about the their perspectives on Written versus Oral Exams.
 One common point of agreement found among the subjects were type of skills each exam format was better suited for.
@@
 %\subsection{Encouraging Genuine Engagement}
-
+\newpage
 \bibliographystyle{apacite}

### 2019-04-16 CS860-CommunicationComplexity/assignments/Chapter4.tex
Correcting Chapter 2
Yao's theorem

@@
 \begin{proof}
-Let $\Pi$ be a distribution over deterministic communication protocols computing $f$ such that $R_\epsilon(f) = \cost(\Pi)$.
-Let $\mu$ be a distribution over $\calX\times\calY$.
-Since $\Pr_{\pi \sim \Pi}[\pi(x,y)\neq f(x,y)]\leq \epsilon$ for all $(x,y) \in \calX \times \calY$ we have
+Let $\Pi$ be a distribution over deterministic communication protocols computing $f$ such that $R_\epsilon(f) = \cost(\Pi)$. Let $\mu$ be a distribution over $\calX\times\calY$. 
 $$\Pr_{(x,y)\sim \mu ;\pi\sim \Pi}[\pi(x,y)\neq f(x,y)]] \leq \epsilon.$$
+Let $I^\pi_{x,y}$ be the indicator random variable for the event $\pi(x,y)\neq f(x,y)$. Since $\Pr_{\pi \sim \Pi}[\pi(x,y)\neq f(x,y)]\leq \epsilon$ for all $(x,y) \in \calX \times \calY$ we have
+$$ \epsilon \geq \mathbb{E}_{ (x,y) \sim \mu} \Pr_{\pi\sim \Pi}(\pi(x,y)\neq f(x,y)) = \mathbb{E}_{ (x,y) \sim \mu} \mathbb{E}_{\pi\sim \Pi}[I^\pi_{x,y}].$$ 
+Rearraning the order of expectation we see that
+$$ \mathbb{E}_{ \pi\sim \Pi} \mathbb{E}_{(x,y) \sim \mu}[I^\pi_{x,y}] \leq \epsilon.$$
 Hence there exists deterministic protocol $\pi$ such that 
@@
 \begin{proof}
-\replacethistext{Enter your proof here, if you complete it.}
+We will use the following result by Von Neumann,
+$$\min_{x\in X}\max_{y\in Y} f(x,y) = \max_{y \in Y} \min_{x \in X} f(x,y)$$
+for compact convex sets $X,Y$ and continuous function $f: X\times Y \rightarrow \R$ such that 
+\begin{itemize}
+    \item for any $x \in X$, $f(x,\cdot): Y\rightarrow \R$ is concave, and
+    \item for any $y \in Y$, $f(\cdot, y): X\rightarrow \R$ is convex.
+\end{itemize}
+Let $P_t$ be the set of deterministic protocols on inputs $\calX \times \calY$ such of cost of at most $t$. Let $A^t$ be the $P_t \times (\calX\times\calY)$ matrix where 
+$$A^t_{\pi, (x,y)} = \begin{cases}
+1, &\text{ if } \pi(x,y) \neq f(x,y) \\
+0, &\text{ otherwise.}
+\end{cases}$$
+Let $\Delta_t = \{x \in \R^t_{\geq 0} : \mathbf{1}^Tx = 1\}$. Let $t_R = R_\epsilon(f)$ and let $t_D = \max_\mu D^\mu_\epsilon(f).$ Then
+$$\epsilon \geq \min_{x \in \Delta_{t_R}} \max_{y \in \Delta_{|\calX\times\calY|}}x^TAy =\max_{y \in \Delta_{|\calX\times\calY|}} \min_{x \in \Delta_{t_R}} x^TAy.$$
+The inequality above follows since $t_R = R_\epsilon (f)$, and hence there is a distribution $x$ on protocols of cost at  most $t_R$ such that for any distribution on inputs $y$, the error $x^TAy$ is at most $\epsilon$.
+The equality above follows by Von Neumann's Theorem.
+Observe that $\Delta_{t_R}$ is a polytope and hence it has an extreme point minimizer for linear objective $x^TAy$ for any fixed $y$.
+Hence there is a deterministic protocol $\pi$ of cost at most $t_R$ such that for any $\mu \in \Delta_{|\calX\times \calY|}$, the error of $\pi$ with respect to $\mu$ is at most $\epsilon$.
+Therefore
+$$\max_{\mu} D^\mu_{\epsilon}(f) \leq t_R = R_\epsilon(f).$$ 
+On the other hand, similarly,
+$$\epsilon \geq \max_{y \in \Delta_{|\calX\times\calY|}} \min_{x \in \Delta_{t_D}} x^TAy  =  \min_{x \in \Delta_{t_D}} \max_{y \in \Delta_{|\calX\times\calY|}}x^TAy.$$
+Thus there is a distribution $\Pi$ on protocols of cost at most $t_D$ with worst-case error on any distribution of inputs at most $\epsilon$. Thus
+$$R_\epsilon(f) \leq t_D = \max_\mu D^\mu_\epsilon(f).$$
+Combining inequalities yields the desired result.
 \end{proof}

### 2019-04-16 CS860-CommunicationComplexity/assignments/Chapter4.tex
All chapter 4 revisions

@@
 Hence the total failure probability can be bounded by  $\frac{1}{2}\frac{1}{2} = \frac{1}{4} \leq \frac{1}{3}$.
-Thus $R(f) = O(1)$, as desired.
+Thus $R(\Eq) = O(1)$, as desired.
 \end{proof}
@@
 
-If $n = \text{poly}\log(k)$ then this protocol with give the desired bound of $O(k \log k)$.
-Let $d = \text{poly}\log(k)$ chosen large enough (\emph{sorry I didn't make this part formal}).
+If $n = \text{poly}(k)$ then this protocol with give the desired bound of $O(k \log k)$.
+Let $d = 10k^2$.
 Alice and Bob can randomly partition their input indices, $[n]$, into $d$ subsets $S_1, \dots, S_d\subseteq [n]$.
-Then they can compute $x'=\{x\cdot \bold{1}_{S_d},\dots, x\cdot \bold{1}_{S_d}\}$ and $y'=\{y\cdot \bold{1}_{S_d},\dots, y\cdot \bold{1}_{S_d}\}$ and re-run the protocol above as if their inputs were $x'$ and $y'$.
-With high probability (\emph{not quite sure how to show this, but I think it might work}) the Hamming distance of $x$ from $y$ agrees with the Hamming distance of $x'$ from $y'$, and so the result holds.
+Then they can compute $x'=\{x\cdot \mathbf{1}_{S_d},\dots, x\cdot \mathbf{1}_{S_d}\}$ and $y'=\{y\cdot \mathbf{1}_{S_d},\dots, y\cdot \mathbf{1}_{S_d}\}$ and re-run the protocol above as if their inputs were $x'$ and $y'$.
+
+With probability at most $$\frac{k^2}{d} = \frac{1}{10}$$
+two coordinates on which $x$ and $y$ different lie in the same subset $S_i$.
+Hence the probability of failure in this partitioning step is at most $\frac{1}{10}$.
+Thus our total probability of failure is at most
+$$\frac{1}{10} + O(n^k)2^{-t} \leq \frac{1}{3}$$
+for appropriate choice of $t = O(k\log d) = O(k\log k).$
 \end{proof}
@@
 Let $\pi$ be a private coin randomized protocol attaining cost $\Rpriv(f)$.
-In the model given by the Hint, Alice and Bob can exchange their probabilities of ending up at each leaf of the protocol tree for $\pi$ at a communication cost of $O(2^{\Rpriv(f)})$.
-Once they known the leaves associated with an input pair $(x,y)$, they can output in their deterministic protocol the bit which is output on the majority of such leaves.
+Modify $\pi$ obtaining protocol $\pi'$ by truncating the branching probability of each node of the protocol tree $\pi$ to $10\cdot 2^{\Rpriv(f)}$ decimal positions.
+This incurs an additional probability of failure of at most $\frac{1}{10\cdot 2^{\Rpriv(f)}}$ at each node.
+In total this incurs at additional probability of failure of at most $\frac{1}{10\cdot 2^{\Rpriv(f)}}2^{\Rpriv(f)} = \frac{1}{10}$.
+Since the original failure probability was at most $\frac{1}{3}$ the faillure probability of $\pi'$ is strictly smaller than $\frac{1}{2}$.
+
+To simulate $\pi'$ deterministically, Alice and Bob can exchange their branching probabilities at each  node in accordance with their input, at a total communication cost of $O(2^{\Rpriv(f)})$.
+Once they known the probabilities of all leaves associated with an input pair $(x,y)$, they can output in their deterministic protocol the bit which is output on the majority of such leaves.
+Since the failure probability of $\pi'$ is strictly less than $\frac{1}{2}$, this is always the correct bit to output.
 \end{proof}

### 2019-04-17 CS860-CommunicationComplexity/assignments/Chapter7.tex
Correcting Chapter 2
Corrected Chapter 7 last problem

@@
 We will present a \emph{one-way} deterministic protocol for this problem using $\pi$ with cost less than $n^{k-1}$.
-Using the same argument as we did in Chapter $6$ for the one-way deterministic communication cost of $\Index$, we can show that this problem was one-way deterministic communication complexity at least $n^{{k-1}})$.
+Using the same argument as we did in Chapter $6$ for the one-way deterministic communication cost of $\Index$, we can show that this problem was one-way deterministic communication complexity at least $n^{{k-1}}$.
 Hence we will have reached a contradiction.
@@
 \begin{proof}
-Suppose for a contradiction there exists a simultaneous message passing protocol $\pi$ for $\SumIndex$ using $c = o(\sqrt{n})$ bits.
-We may assume that players $2$ and $3$ announce their $j$ and $i$ respectively in their messages at a cost of $\log(n)$ bits.
-This is without loss of generality since $c + \log(n) = o(\sqrt{n})$.
-Now since the referee learns from players $2$ and $3$ everything that player $1$ sees, player $1$ does not need to send anything in $\pi$.
-So we assume player $1$'s message is empty.
-
-Using this simplified protocol $\pi$ we...
+Suppose there is a simultaneous message passing protocol $\pi$ for $\SumIndex$ using $c$ bits of communication.
+We will present a \emph{one-way} deterministic protocol for for $\Index$ using $\pi$.
+Let $A \in \{0,1\}$ be Alice's input to $\Index$, and let $k$ be Bob's input to $\Index$.
+Let $B_k \in \{0,1\}^{\log n}$ be the binary representation of $k$.
+Let $L_k\in \{0,1\}^{\log n}$ be the vector where the first half of its bits correspond to $B_k$'s and the latter half are all $0$.
+Let $R_k = B_k \oplus L_k$.
+Let $i, j$ be the base-$10$ numbers represented by $L_k$ and $R_k$ respectively.
+Observe that
+$$\Index(A,k) = \SumIndex(A,i,j).$$
+Further observe that there are $2^{\frac{1}{2}\log(n)} = \sqrt{n}$ possible values of $i$, and the same for $j$.
+Alice can simulate player $2$ in $\pi$ by enumerating all possible values of $j$, and sending the message player $2$ would send in $\pi$ on that seeing $A$ and that value to Bob.
+This costs $c\sqrt{n}$ bits of communication.
+She can do similarly for player $3$.
+In total this costs
+$$2c\sqrt{n}$$
+bits of communication.
+Bob knows $i$ and $j$, and so can determine player $1$'s message $m_1$ in $\pi$.
+Further he can select the appropriate messages $m_2$ and $m_3$ for players $2$ and $3$ from the messages Alice sent.
+Altogether he can now simulate the referee in $\pi$ on $(m_1,m_2,m_3)$ to compute $\SumIndex(A,i,j) = \Index(A,k).$ Since $\Index$ has a one-way communication lower bound, this implies that $c = \Omega(\sqrt{n})$.
 \end{proof}

### 2019-04-18 CS860-CommunicationComplexity/assignments/Chapter6.tex
Correcting Chapter 2
SMP equality

@@
 To see the claim, we use that the protocol succeeds with probability $\frac{2}{3}$. We expand the conditional entropy as an expectation and compute
-$$H(X_i\mid M) = \sum_{m\in M} H(X_i\mid M=m)\cdot \Pr(M=m) = \sum_{m\in M}(\frac{2}{3}\log(\frac{3}{2}) + \frac{1}{3}\log(3))\Pr(M=m) \leq \frac{2}{3}.$$
+\begin{align*}
+	H(X_i\mid M) &= \sum_{m \in M} \Pr(M=m)\mathbb{E}[-\log(\Pr(X_i\mid M=m))]\\
+	&\leq \mathbb{E}[-\log(\Pr(X_i\mid M=m)\Pr(M=m)] &\text{ via Jensen's inequality}\\
+	&= \mathbb{E}[H(X_i\mid E[M])]\\
+	&\leq \frac{2}{3}\log\left(\frac{3}{2}\right) + \frac{1}{3}\log(3)\\
+	&\leq \frac{2}{3}.
+\end{align*}
 The second inequality follows by the success probability of the protocol.
@@
 	Fix $y\in \calY$, and $m_y \in M_y$.
-Drawn a uniform random subset $S_x \subseteq M_x$  of size $t$.
+Draw a uniform random subset $S_x \subseteq M_x$  of size $t$.
 By a Chernoff Bound, the probability that $\tau(m_x,m_y) \neq f(x,y)$ for at least half of message pairs $m_x, m_y$ is at most
@@
 
+\begin{definition}
+	A set $M \subseteq \{0,1\}^n$ is called a $(n,k,d)$-code if
+	\begin{itemize}
+		\item $|M| = 2^k$, and 
+		\item For every pair $x\neq y \in M$, the Hamming distance between $x$ and $y$ is at least $d$.
+	\end{itemize}
+\end{definition}
+
+\begin{lemma}\label{lemma:code}
+There exists a $(3m,m,m/2)$-code for any $m$.
+\end{lemma}
+
+\begin{proof}
+We construct the code, greedily.
+Begin with $M= \emptyset$ and at each iteration choose $x\in \{0,1\}^{3m}\backslash M$ such that the Hamming distance from $x$ to every codeword $y \in M$ is at least $m/2$.
+Add $x$ to $M$.
+Terminate when there are no more $x$ possible to add.
+
+We claim that $|M| \geq 2^m$ upon termination of the above procedure.
+To see this, we bound the number of strings in $\{0,1\}^{3m}$ that any $x \in \{0,1\}^{3m}$ prevents from being chosen in future iterations, i.e.
+we bound the number of $y \in \{0,1\}^{3m}$ at Hamming distance at most $\frac{m}{2}-1$ from $x$.
+
+There are at most 
+$$\sum_{i=1}^{n/2 - 1} {3m \choose i} \leq {3m \choose m/2}$$
+ways to choose the bits of $x$ to flip to obtain $y$ of Hamming distance at most $\frac{m}{2} - 1$. By the entropy bound,
+$${3m\choose m/2} \leq 2^{3m H(1/6)} \leq 2^{3m\frac{2}{3}} = 2^{2m}.$$
+Hence 
+$$|M| \geq \frac{2^{3m}}{2^{2m}} =2^m$$
+as desired.
+\end{proof}
+
 \begin{theorem}
@@
 \begin{proof}
-\replacethistext{Enter your proof here.}
+The lowerbound follows from the previous Theorem. It remains to give a protocol attaining the upperbound.
+
+Let $m \in Z$ be the least integer satisfying $(6m)^2 \geq 3n$.
+By Lemma \ref{lemma:code} there exists a $((6m)^2, (6m)^2/3, (6m)^2/6)$-code $M'$.
+Choose $2^n$ elements of this code to obtain a $((6m)^2,n, (6m)^2/6))$-code $M$.
+Establish a bijection $\phi$ between $\{0,1\}^n$ and $M$.
+
+Let $x \in\{0,1\}^n$ be Alice's input, and let $y \in \{0,1\}^n$ be Bob's input.
+Alice reinterprets $\phi(x)$ as a $6m\times 6m$ matrix, $A$, where the first row of $A$ corresponds to the first $6m$ entries of $\phi(x)$, the second row corresponds to next $6m$ entries o $\phi(x)$, etc.
+Alice chooses $i \in [6m]$ uniformly at random and sends $i$ and the $i$-th row of $A$, $a$ to the referee.
+
+Analogously, Bob reinterprets $\phi(y)$ as a $6m\times 6m$ matrix $B$. Bob chooses a column $j$ of $B$ uniformly at random and send $j$ and the $j$-th row of $B$, $b$ to the referee.
+
+The total communication cost of the protocol is thus $O(m) = O(\sqrt{n})$.
+The referee compares $a_j$ and $b_i$.
+If $a_j \neq b_i$ the referee outputs $0$.
+If $a_j = b_i$ then the referee outputs $1$ with probability $6/11$ and otherwise outputs $0$.
+
+We now analyze the success probability of the algorithm.
+First observe that $a_j$ is $(i,j)$-th entry of $A$, and $b_i$ is the $(i,j)$-th entry of $B$.
+If $x=y$ then $A=B$ and the referee always enters the second case, outputting the correct answer with probability $6/11$.
+Consider the case where $x\neq y$.
+Then $A\neq B$.
+Since $A$ and $B$ are drawn from code $M$, they differ in at least $(6m)^2/6$ entries.
+Let $p$ be the probability that $A$ and $B$ differ on entry $(i,j)$.
+Since $(i,j)$ are chosen independently and uniformly at random, and at least $1/6$ of the total entries differ,
+$$p\geq \frac{1}{6}.$$
+So the referee will output the correct answer with probability
+$$p + \frac{5}{11}(1-p) \geq \frac{5}{11} + \frac{6}{11}\frac{1}{6} = \frac{6}{11}.$$
+This protocol thus obtains a success probability slightly better than $1/2$, which can be boosted with a constant increase in communication cost to a protocol with success probability at least $\frac{2}{3}$
 \end{proof}

### 2019-04-18 CS860-CommunicationComplexity/assignments/Chapter6.tex
One proof left

@@
 
-Define the \emph{Hamming distance} between strings $x,y \in \{0,1\}^n$ to be $\dHam(x,y) = |\{i \in [n] : x_i \neq y_i\}|$.
-The \emph{gap Hamming function} $\GHD : \{0,1\}^n \times \{0,1\}^n \to \{0,1,*\}$ is the partial function defined by
+Define the \emph{Hamming distance} between strings (we move to $\pm 1$ form for strings for convenience in the proof) $x,y \in \{-1,1\}^n$ to be $\dHam(x,y) = |\{i \in [n] : x_i \neq y_i\}|$.
+The \emph{gap Hamming function} $\GHD : \{-1,1\}^n \times \{-1,1\}^n \to \{0,1,*\}$ is the partial function defined by
 \[
@@
 \begin{theorem}
-The one-way randomized communication complexity of the $\GHD : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ function is
+The one-way randomized communication complexity of the $\GHD : \{-1,1\}^n \times \{-1,1\}^n \to \{0,1\}$ function is
 \[
@@
 \begin{proof}
-\replacethistext{Enter your proof here, if you complete it.}
+We will show how to use $\GHD$ to solve $\Index$.
+Let $A\in \{0,1\}^n$ be Alice's input to $\Index$ and $i \in [n]$ be Bob's.
+Let $u \in \{-1,1\}^n$ be the vector defined by $u_j = -1$ if $A_j = 1$, and $u_j = 1$ otherwise.
+Let $e_i \in \{0,1\}$ be the standard $i$-th basis vector.
+
+Fix integer $N$ to be specified later.
+Pick $N$ independent identically distributed vectors $r^1,r^2,\dots, r^N \in \R^n$ according to distribution $\mu$ (to be specified later) at random.
+Let $x, y\in\{-1,1\}^N$ be vectors such that for all $k \in [N]$,
+$$x_k := \text{sign}(u^Tr^k)\quad\text{and}\quad y_k := \text{sign}(e_i^Tr^k) = r^k_i.$$
+Observe that
+$$\dHam(x,y)= |\{k:\text{sign}(u^T,r^k)\neq r^k_i\}|.$$
+We will choose the distribution $\mu$ so that for any $r\sim \mu$,
+\begin{equation}\label{eq:prob}\Pr[\text{sign}(u^Tr)\neq r_i] \begin{cases}
+\geq \frac{1}{2} + \frac{c}{\sqrt{n}}, &\text{if }u_i=-1 \\
+\leq \frac{1}{2}-\frac{c}{\sqrt{n}}, &\text{if }u_i=1.
+\end{cases}\end{equation}
+for some constant $c>0$. By Chernoff's bound, with $N =\frac{4n}{c^2}$ and $\epsilon = \frac{1}{\sqrt{N}}$, with probability at least $\frac{2}{3}$ we have 
+$$H(x,y) \begin{cases}
+\frac{N}{2} + \sqrt{N}, &\text{if } u_i = -1 \\
+\frac{N}{2} - \sqrt{N}, &\text{if } u_i = 1.
+\end{cases}$$
+Thus we can use a protocol for $\GHD$ to compute $u_i$, and hence determine $A_i$.
+
+It remains to show (\ref{eq:prob}). Let $\mu$ be the uniform distribution over unit vectors in $\R^n$. Let $r' = \text{proj}_{\text{span}(\{u,e_i\})}r.$
+Let $\hat{u}$ be the normalization of $u$. Now we have that
+$$\Pr[\text{sign}(u^Tr)\neq r_i] = \frac{\arccos(\hat{u}^Te_i)}{\pi} = \frac{1}{\pi}\arccos(\frac{u_i}{\sqrt{n}}).$$
+Observe that for any $v \in [-1,1]$, $\arccos(v) = \frac{\pi}{2} - \arcsin(v)$.
+Since $\sin(v) \leq v$, for $v\geq 0$, we have $\arcsin(v) \geq v$ for $v \in [0,1]$.
+Substituting this into the about equation we obtain the bound (\ref{eq:prob}) as desired.
+
+\emph{Note:} This proof is adapted from ``The One-Way Communication Complexity of Hamming Distance" by Jayram, Kumar, and Sivakumar.
 \end{proof}

### 2019-04-18 CS860-CommunicationComplexity/assignments/Book-Notes.tex
Book templates added

\chapter{Notes}

This chapter collects some of the references for the results presented in this book and for further reading. This chapter is not meant to be a complete bibliography for communication complexity, but it hopefully presents at least one good starting point for further research in each of the topics covered in the book.

\section{General references}

\subsection*{Textbooks.}
There are a few excellent textbooks on communication complexity. The standard reference for the fundamentals of communication complexity is the textbook of Kushilevitz and Nisan~\cite{KushilevitzN97}. Rao and Yehudayoff~\cite{RaoYehudayoff18} have also written a textbook on communication complexity to appear soon.

\subsection*{Other books.}
There are also a number of books that deal with specific aspects of communication complexity. Lee and Shraibman have written a book on lower bound techniques in communication complexity~\cite{LeeS09}. Roughgarden has written a book aimed at introducing communication complexity for algorithm designers~\cite{Roughgarden16}.

Jukna's book on the complexity of Boolean functions~\cite{Jukna12} includes a part dedicated to communication complexity. Arora and Barak's textbook on computational complexity~\cite{AroraB09} also dedicates a chapter to communication complexity. And Lokam's book on uses of linear algebra for establishing complexity lower bounds~\cite{Lokam09} discusses communication complexity extensively.

\subsection*{Lecture notes.}
Lecture notes are other great sources for expositions of various topics in communication complexity. See for instance the lecture notes for the courses of Pitassi~\cite{Pitassi14}, Raz~\cite{Raz00}, Hatami~\cite{Hatami16}, and Sherstov~\cite{Sherstov12}.

\subsection*{Surveys.}
Finally, a number of surveys are also great general references. See for example the introductory survey of Lov{\'a}sz~\cite{Lovasz89}, the surveys on the set disjointness problem of Chattopadhyay and Pitassi~\cite{ChattopadhyayP10} and of Sherstov~\cite{Sherstov14}, and the survey on information complexity of Jayram~\cite{Jayram10}.

There are also surveys on quantum communication complexity (which we do not discuss in this textbook) by de Wolf~\cite{deWolf02}, Klauck~\cite{Klauck00}, and Brassard~\cite{Brassard04}.

\section{Deterministic communication complexity}

The notion of communication complexity as introduced in this chapter was first defined by Yao~\cite{Yao79}, who also introduced the notions of rectangles and covers. The upper bound on the communication complexity of the \textsc{Median} function was established by Karchmer (see~\cite[\S 1.5]{KushilevitzN97}).
Fooling sets first appeared explicitly in the work of Lipton and Sedgewick~\cite{LiptonS81}.
The log rank bound was introduced by Mehlhorn and Schmidt~\cite{MehlhornS82}.

\section{Distributional and randomized communication complexity}

Randomized communication complexity was also introduced in Yao's original paper~\cite{Yao79}. The connection between distributional and randomized communication complexity (as well as a similar connection in other models of computation) was also established by Yao~\cite{Yao83}.

The lower bound on the distributional complexity of the inner product function is due to Chor and Goldreich~\cite{ChorG85}.

The $\Omega(\sqrt{n})$ lower bound on the distributional complexity of the set disjointness function is due to Babai, Frankl, and Simon~\cite{BabaiFS86}. The optimal $\Omega(n)$ lower bound on the randomized communication complexity of the same function was first established by Kalyanasundaram and Schnitger~\cite{KalyanasundaramS92}.

Newman's theorem regarding the public- and private-coin randomized communication complexity of functions was established in~\cite{Newman91}.

\section{Information complexity}

The first work to explicitly consider information complexity and its relation to communication complexity is that of Chakrabarti, Shi, Wirth, and Yao~\cite{ChakrabartiSWY01} in the context of simultaneous message passing and that of Bar-Yossef, Jayram, Kumar, and Sivakumar~\cite{BarYossefJKS04} in more general two-player communication settings.

A good general reference to information theory for this chapter is the textbook of Cover and Thomas~\cite{CoverT06}.

There are too many great papers that discuss the topics introduced in this textbook and extensions of the results presented here to list here. Let us just mention a few. Braverman~\cite{Braverman15} introduced the study of \emph{distribution-independent} notions of the information complexity of functions.
Braverman, Garg, Pankratov, and Weinstein~\cite{Braverman13} showed how to compute the exact information complexity of the \textsc{And} function.

We have seen that information complexity can be used to give strong lower bounds on the communication complexity of natural functions. It's natural to ask whether it gives good lower bounds for \emph{all} functions. It does not: Ganor, Kol, and Raz~\cite{GanorKR16} showed that there can be an exponential separation between information and communication complexity for some functions. However, Kerenidis, Laplante, Lerays, Roland, and Xiao~\cite{Kerenidis15} showed that many of the lower bound techniques for communication complexity also give matching lower bounds for information complexity as well.

\section{One-way communication and simultaneous message passing}

The one-way communication model (and more generally the communication model where Alice and Bob are limited to $k$ rounds of interaction) was first studied by Papadimitriou and Sipser~\cite{PapadimitriouS84}. See also the papers of Nisan and Wigderson~\cite{NisanW93}, Kremer, Nisan, and Ron~\cite{KremerNR99}, and Newman and Szegedy~\cite{NewmanS96} for the early foundational results in this model.

The simultaneous message passing model was initially presented in Yao's original paper~\cite{Yao79}. The lower bound on the randomized SMP complexity of functions in terms of their deterministic communication complexity was obtained by Babai and Kimmel~\cite{BabaiK97}; see the same paper for an alternative proof obtained in parallel work by Bourgain and Wigderson.

The Gap Hamming Distance function considered in this chapter was first introduced by Indyk and Woodruff~\cite{IndykW03} to prove lower bounds on the distinct elements problem in the streaming algorithms model. Chakrabarti and Regev~\cite{ChakrabartiR12} obtained an optimal $\Omega(n)$ lower bound on the communication complexity of this function in the (standard) two-way communication complexity model; Vidick~\cite{Vidick13} and Sherstov~\cite{Sherstov12ghd} have since obtained alternative (and simpler) proofs of this lower bound.

\section{Multiparty communication}

Multiparty communication complexity was first considered by Chandra, Furst, and Lipton~\cite{ChandraFL83b}. Cylinder intersections were introduced by Babai, Nisan, and Szegedy~\cite{BabaiNS92}, who also obtained the nearly optimal lower bound on the generalized inner product function. The upper bound for that function is due to Grolmusz~\cite{Grolmusz94}.

The study simultaneous message passing model of multiparty communication complexity was initiated by Babai, Kimmel, and Lokam~\cite{BabaiKL95} who also established the first results on the \textsc{SumIndex} function.

A great source for more on discrepancy, with applications in communication complexity and in many other areas as well, see the textbook of Chazelle~\cite{Chazelle01}.

### 2019-04-19 CS860-CommunicationComplexity/assignments/Chapter2.tex
Textbook wrapper

@@
-\documentclass[11pt]{amsart}
-\usepackage{amsmath,amsthm,amsfonts,amssymb}
-\usepackage{fullpage}
-\usepackage[usenames]{xcolor}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{proposition}{Proposition}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{exercise}{Exercise}
-\newtheorem{problem}{Problem}
-\newtheorem{challenge}{Challenge Problem}
-\newtheorem{open}{Open Problem}
-\theoremstyle{plain}
-
-\newcommand{\argmin}{\mathrm{argmin}}
-\newcommand{\calX}{\mathcal{X}}
-\newcommand{\calY}{\mathcal{Y}}
-\newcommand{\calZ}{\mathcal{Z}}
-\newcommand{\cost}{\mathrm{cost}}
-\newcommand{\disc}{\mathrm{disc}}
-\newcommand{\discu}{\mathrm{disc_u}}
-\newcommand{\Disj}{\textsc{Disj}}
-\newcommand{\Eq}{\textsc{Eq}}
-\newcommand{\GT}{\textsc{GT}}
-\newcommand{\IP}{\textsc{IP}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\rank}{\mathrm{rank}}
-\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}
-
-\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}
-
-\title{Communication complexity: Chapter 2 \\ More lower bound techniques}
-\author{Eric Blais and Justin Toth}
-
-\begin{document}
-
-\maketitle
-
+\chapter[Lower Bounds]{More Lower Bound Techniques}
 In the first chapter, we introduced the deterministic communication complexity 
@@
 \end{open}
-
-\end{document}

### 2019-04-19 CS860-CommunicationComplexity/assignments/Chapter6.tex
Textbook wrapper

@@
-\documentclass[11pt]{amsart}
-\usepackage{amsmath,amsthm,amsfonts,amssymb}
-\usepackage{fullpage}
-\usepackage[usenames]{xcolor}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{proposition}{Proposition}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{exercise}{Exercise}
-\newtheorem{remark}{Remark}
-\newtheorem{problem}{Problem}
-\newtheorem{challenge}{Challenge Problem}
-\newtheorem{open}{Open Problem}
-\theoremstyle{plain}
-
-\newcommand{\argmin}{\mathrm{argmin}}
-\newcommand{\calX}{\mathcal{X}}
-\newcommand{\calY}{\mathcal{Y}}
-\newcommand{\calZ}{\mathcal{Z}}
-\newcommand{\cost}{\mathrm{cost}}
-\newcommand{\dHam}{\mathrm{d}_{\mathrm{Ham}}}
-\newcommand{\disc}{\mathrm{disc}}
-\newcommand{\discu}{\mathrm{disc_u}}
-\newcommand{\Disj}{\textsc{Disj}}
-\newcommand{\Eq}{\textsc{Eq}}
-\newcommand{\GHD}{\textsc{GHD}}
-\newcommand{\GT}{\textsc{GT}}
-\newcommand{\HD}{\textsc{HD}}
-\newcommand{\ICext}{\mathrm{IC}^{\mathrm{ext}}}
-\newcommand{\ICint}{\mathrm{IC}^{\mathrm{int}}}
-\newcommand{\icostext}{\mathrm{icost}^{\mathrm{ext}}}
-\newcommand{\icostint}{\mathrm{icost}^{\mathrm{int}}}
-\newcommand{\IP}{\textsc{IP}}
-\newcommand{\Index}{\textsc{Index}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\rank}{\mathrm{rank}}
-\newcommand{\Rpriv}{R^{\mathrm{priv}}}
-\newcommand{\replacethistext}[1]{\textcolor{red}{#1}}
-
-\newcommand{\exercises}{\bigskip \noindent\rule{8cm}{0.4pt} \medskip}
-
-\title{Communication complexity: Chapter 6 \\ One-way communication and simultaneous message passing}
-\author{Eric Blais and Justin Toth}
-
-\begin{document}
-
-\maketitle
-
+\chapter[One-way Communication and SMP]{One-way Communication and Simultaneous Messsage Passing}
 Every communication complexity question we have studied so far in this course have been set in the model where there are two players, Alice and Bob, who exchange bits with each other to compute functions on their joint inputs.
@@
 with error at most $\frac13$.
-\end{document}

### 2019-04-19 CS860-CommunicationComplexity/assignments/Chapter1.tex
Added ch1

\chapter[Deterministic complexity]{Deterministic communication complexity}

In communication complexity, two players named \emph{Alice} and \emph{Bob} each receive some input that the other player cannot see and seek to compute some function on their joint input. We wish to determine the minimum number of bits that they must exchange to compute this function. As a result, the main object we will study will not be algorithms, but rather \emph{communication protocols}.

\begin{definition}[Protocol]
A \emph{(communication) protocol} $\pi$ is a rooted binary tree $T(\pi)$ with the following additional information:
\begin{itemize}
\item Every internal node of the tree is labelled with $A$ or $B$, determining whether Alice or Bob sends the next bit.
\item Every internal node $v$ labelled with $A$ also has a corresponding function $h_v : \calX \to \{0,1\}$ which determines the next bit that Alice sends.
\item Similarly, every internal node $v$ labelled with $B$ also has a corresponding function $h_v : \calY \to \{0,1\}$ which determines the next bit that Bob sends.
\item The two edges leaving an internal node $v$ are labelled with $0$ and $1$, respectively.
\item Each leaf node is labelled with $0$ or $1$.
\end{itemize}
\end{definition}

\begin{definition}[Function computed by a protocol]
A protocol $\pi$ \emph{computes} the function $f : \calX \times \calY \to \{0,1\}$ if for every input $(x,y) \in \calX \times \calY$, the path in $\pi$ obtained by following the edge labelled with $h_v(x)$ at each internal node labelled by $A$ and $h_v(y)$ at each internal node labelled by $B$ leads to a leaf with the label $f(x,y)$.
\end{definition}

Following the standard computer science approach, we measure the communication cost of protocols in the worst-case sense.

\begin{definition}[Communication cost]
The \emph{communication cost} of a protocol $\pi$ is 
\[
\cost(\pi) = \mathrm{height}(T(\pi)),
\]
the height of the tree for $\pi$ or, in other words, the length of the longest path between the root of $T(\pi)$ and any of the leaves in the tree.
It corresponds to the maximum number of bits that Alice and Bob exchange when executing the protocol over any of their inputs.
\end{definition}

\begin{definition}[Deterministic communication complexity]
The \emph{deterministic communication complexity} of the function $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$ is
\[
D(f) = \min_{\pi \mbox{ computes } f} \cost(\pi)
\]
is the minimum communication cost of a protocol that computes $f$.
\end{definition}

In the rest of this chapter, we will aim to determine the deterministic communication complexity for some fundamental functions.

\exercises

\begin{exercise} %[Easy]
Do there exist functions $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$ with  communication complexity $D(f) = 0$?
\end{exercise}

\newpage \section{Equality}
We begin by considering the simplest (but also perhaps the most important) function: equality. The function $\Eq : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ is defined by
\[
\Eq(x,y) = \begin{cases}
1 & \mbox{if } x = y \\
0 & \mbox{if } x \neq y
\end{cases}
\]
so that the value of the function is 1 if and only if Alice and Bob's inputs are identical.

\begin{theorem}
$D(\Eq) \le n + 1$.
\end{theorem}

\begin{proof}
There is a very simple protocol $\pi$ with cost $n+1$ that computes the function $\Eq$: Alice sends the $n$ bits of her input to Bob, then Bob sends the value $1$ if and only if his input $y$ is equal to the input $x$ that he has received. The value of each leaf is set to be the same as the value of the edge that leads to it.
\end{proof}

\newpage \section{General upper bound}
The trivial protocol we used to bound the deterministic communication complexity of the \textsc{Equality} function can also be modified slightly to establish a much more general result.

\begin{theorem}
For every function $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$, 
\[
D(f) \le \min\{\lceil \log_2 |\mathcal{X}| \rceil, \lceil \log_2 |\mathcal{Y}| \rceil\} + 1.
\]
\end{theorem}

\begin{proof}
We can encode each element in $\calX$ with $\lceil \log_2 |\mathcal{X}| \rceil$ bits. We can use such an encoding to design a protocol $\pi$ with cost $\lceil \log_2 |\mathcal{X}| \rceil + 1$: Alice sends $\lceil \log_2 |\mathcal{X}| \rceil$ bits to communicate the encoding of her input $x$ to Bob, and Bob replies with the value $f(x,y)$. (Note that Bob now knows both $x$ and $y$ so that he can compute this value.)

Similarly, we can design a protocol $\pi'$ with cost $\lceil \log_2 |\mathcal{Y}| \rceil$ by taking an encoding of the elements in $\calY$ and having Bob send the encoding of his input $y$ to Alice and then having Alice respond with the value $f(x,y)$. (Now Alice is the one who knows both $x$ and $y$.)

Combining the two protocols above, we obtain that
\[
D(f) \le \min\{ \cost(\pi), \cost(\pi') \} = \min\{\lceil \log_2 |\mathcal{X}| \rceil + 1, \lceil \log_2 |\mathcal{Y}| \rceil + 1\},
\]
as we wanted to show.
\end{proof}

\newpage \section{Parity}
The general upper bound we established in the last section is far from tight in some cases. Consider for example the $\textsc{Parity} : \{0,1\}^n \times \{0,1\}^n \to \{0,1\}$ function defined by
\[
\textsc{Parity}(x,y) = \bigoplus_{i=1}^n (x_i \oplus y_i) 
\]
where $\bigoplus_{i=1}^n z_i = \sum_{i=1}^n z_i \pmod{2}$ is the parity of the sum and, similarly, $a \oplus b$ is defined to be $a+b \pmod{2}$.

\begin{theorem}
$D(\textsc{Parity}) \le 2$.
\end{theorem}

\begin{proof}
Consider the protocol $\pi$ where Alice sends the value $\bigoplus_{i=1}^n x_i$ to Bob, Bob sends the value $\bigoplus_{i=1}^n y_i$, and the leaf at the end of the path labelled with $a$ and $b$ takes the value $a \oplus b$. This protocol has cost 2, since Alice and Bob each send a single bit. And we can verify that $\pi$ correctly computes the \textsc{Parity} function since on any input $(x,y) \in \{0,1\}^n \times \{0,1\}^n$, the protocol leads to the leaf labelled with
\[
\left( \bigoplus_{i=1}^n x_i \right) \oplus 
\left( \bigoplus_{i=1}^n y_i \right) 
=  \bigoplus_{i=1}^n \left(x_i \oplus y_i \right) \\
= \textsc{Parity}(x,y). \qedhere
\]
\end{proof}

\newpage \section{Median}
Define $[n] = \{1,2,\ldots,n\}$ and $2^{[n]}$ to be the set of subsets of $[n]$. The $\textsc{Median} : 2^{[n]} \times 2^{[n]} \to [n]$ function is defined by
\[
\textsc{Median}(S,T) = \mathrm{median}(S \cup T);
\]
it is the median element of the multiset obtained by taking the union of Alice and Bob's sets.

\begin{theorem}
$D(\textsc{Median}) = O(\log^2 n)$.
\end{theorem}

\begin{proof}
Here is a simple protocol that computes the \textsc{Median} function. Alice and Bob start by exchanging the total number of elements in $S$ and in $T$. Define $k = \lfloor \frac{|S| + |T|}2 \rfloor$. Alice and Bob now need to identify the $k$-th smallest element in the multiset $S \cup T$. They can do so using a binary search approach. In the first step of this search, Alice and Bob exchange the number of elements in $S \cap [n/2]$ and $T \cap [n/2]$, respectively. If that number $m$ is $m \ge k$, then we continue the binary search in the range $\{1,2,\ldots,n/2\}$. Otherwise, we update $k' = k - m$ and search for the $k'$-th smallest element in $(S \cup T) \cap \{n/2+1,\ldots,n\}$.

This binary search process will end at the median element. The search itself requires $O( \log n)$ rounds, and at each round Alice and Bob exchange a number in the range $1,2,\ldots,n$, which they can do with $O(\log n)$ bits of communication.
\end{proof}

\exercises

\begin{exercise} %[Easy]
Prove that $D(\textsc{Min}) = O(\log n)$ and $D(\textsc{Max}) = O(\log n)$.
\end{exercise}

\begin{exercise}
Consider the \textsc{Median'} function where we take the median of the simple set (i.e., deleting duplicates) obtained by taking the union $S \cup T$. 
Prove that $D(\textsc{Median'}) = O(\log^2 n)$.
\end{exercise}

\begin{exercise} %[Hard]
Prove that $D(\textsc{Median}) = O(\log n)$.
\end{exercise}

\newpage \section{Rectangles and partition number}

As we have seen in the last sections, we can give upper bounds on the deterministic communication complexity of a given function by designing a communication protocol that computes the function. Our main goal, however, will generally be to prove lower bounds on the communication complexity of various functions. We can do so by analyzing \emph{combinatorial rectangles}.

\begin{definition}[Rectangle]
A \emph{(combinatorial) rectangle} over the finite set $\calX \times \calY$ is a set $S \subseteq \calX \times \calY$ defined by $S = A \times B$ for some sets $A \subseteq \calX$ and $B \subseteq \calY$.
\end{definition}

\begin{definition}[$f$-monochromatism]
Given a function $f : \calX \times \calY \to \{0,1\}$, a rectangle $R \subseteq \calX \times \calY$ is \emph{$f$-monochromatic} if $f(x,y)$ takes the same value for all $(x,y) \in R$.
\end{definition}

\begin{definition}[$\chi$]
The \emph{partition number} of a function $f : \calX \times \calY \to \{0,1\}$, denoted
\[
\chi(f),
\]
is the minimum number of $f$-monochromatic rectangles required to partition $\calX \times \calY$.
\end{definition}

\newpage \section{Partition bound}

\begin{lemma}
For every function $f : \mathcal{X} \times \mathcal{Y} \to \{0,1\}$, 
\[
D(f) \ge \log_2 \chi(f).
\]
\end{lemma}

\begin{proof}
Let $\pi$ be any protocol that computes $f$ and has cost $\cost(\pi) = D(f)$. Each leaf in the tree for $\pi$ corresponds to an $f$-monochromatic rectangle and the rectangles corresponding to each leaf in the tree partition $\mathcal{X} \times \mathcal{Y}$. Since a binary tree of depth $d$ can have at most $2^d$ leaves, this means that $\chi(f) \le 2^{\cost(\pi)} = 2^{D(f)}$. 
\end{proof}

\exercises

\begin{exercise} %[Hard]
Show that $D(f) \le (\log_2 \chi(f) + 1)^2$.
\end{exercise}

\begin{open}
Find a function $f$ for which $D(f) \ge 4 \log_2 \chi(f)$.
\end{open}

### 2019-05-13 CUT/paper.tex
Paper revisions

@@
 \doublespace
-\title{Impact of Assessment Format on Student Learning Outcomes}
+\title{Impact of Exam Format on Student Learning Outcomes}
 \author{W. Justin Toth\\Combinatorics and Optimization\\20411831}
@@
 \begin{abstract}
-We study of question of how assessment formats impact student study behaviours, learning outcomes, and perception of the courses they take.
+We study of question of how exam formats impact student study behaviours, learning outcomes, and perception of the courses they take.
 We particularly focus on Mathematics and Science courses, but the findings in this survey are general enough to be widely applicable.
-We study Multiple Choice versus Constructed Response, how much Reference Material to allow in the exam room, Oral versus Written Exams, and the Two Stage method for injecting collaborative learning into exams.
+We study multiple choice versus constructed response, how much reference material to allow in the exam room, oral versus written exams, and the Two Stage method for injecting collaborative learning into exams.
 \end{abstract}
@@
 A well-known truism in human social interaction is that how you say something often has as much, if not more, importance that the specific content of your speech.
-Yet when instructors are designing evalutions for their courses, it is all too common that the focus is squarely on what questions to ask to cover the course content, rather than how the problems should be presented to the students.
-In this survey we seek to synthesize the academic literature on the ``how'', rather than ``what'', of assessment methods and provide guidance for instructors on the impact that exam format can have on student learning outcomes and their ability to measure the attainment thereof.
+Yet when instructors are designing exams for their courses, it is all too common that the focus is squarely on what questions to ask to cover the course content, rather than how the problems should be presented to the students.
+In this survey we seek to synthesize the academic literature on the ``how'', rather than ``what'', of assessment methods used for exams and provide guidance for instructors on the impact that exam format can have on student learning outcomes and their ability to measure the attainment thereof.
 \paragraph{}
 In Section \ref{sec:written} we study the particulars of written finals.
+We concentrate on exams in the context of STEM courses, but many of the findings are broadly applicable.
 For written finals we will concentrate on two areas of focus in the literature.
@@
 In line with this viewpoint, students predominantly favor shallow-learning strategies like memorization for exams consisting primarily of Multiple Choice problems.
-In contrast, students perceive Construhcted Response exams as testing deeper understanding, and use deeper learning strategies to study for Constructed Response exams~\cite{scouller1998influence}
+In contrast, students perceive Constructed Response exams as testing deeper understanding, and use deeper learning strategies to study for Constructed Response exams~\cite{scouller1998influence}
 \paragraph{}
@@
 \paragraph{}
-The author defined critical-thinking skills as those skills associated with the middle levels of Bloom's taxonomy: application, analysis, evaluation, and synthesis.
+The author defined critical-thinking skills as those skills associated with the middle and high levels of Bloom's taxonomy: application, analysis, evaluation, and synthesis.
 The experiment was performed using two sections of introductory biology given by the same instructor in the same semester to large classes of about $200$ students.
@@
 The questions in this test were aimed at the remember and understand levels of Bloom's taxonomy, indicating that even for the most basic forms of learning, Multiple Choice questions due a poor job of measuring student knowledge retention.
+
 \subsection{Reference Material}
@@
 Usually these notes are allowed to have any content the student may want on them, but with space and formatting restrictions which allow the notes they carry in to have significantly more compact information than an Open Book exam.
-Comparing the efficacy of Open versus Closed Book exams in the literature has led to mixed results.
+Comparing the overall efficacy, in terms of encouraging postive study habits and long term knowledge retention, of Open versus Closed Book exams in the literature has led to mixed results.
 \paragraph{}
@@
 \paragraph{}
-The anxiety Capalid noted is a common problem.
-Boedigheimer also observed that phenomenon.
+The anxiety Capaldi noted is a common problem.
+Boedigheimer also observed that same phenomenon.
 Another issue with Oral Exams is the instructor workload.
 In a class of $35$ to $40$ students, Oral Exams can easily span a couple of full days.
-Nevermind a first year course with $200$ plus students.
-This difficulty can spill over into missed class time, with the instructor often needing to bock off classes during the semester just for Oral Exams.
+Nevermind a first year course with $200$ plus students where the workload can become entirely infeasible without considerable instructional support.
+This difficulty can spill over into missed class time, with the instructor often needing to block off classes during the semester just for Oral Exams.
 \paragraph{}
@@
 If one member of a group was unable or unwilling to contribute, the students in that group were severely hampered by their teammate.
-Groups which poor communication or teamwork reported high dissatisfaction with the Oral Exam component of the course in the course evaluation surveys.
+Groups with poor communication or teamwork reported high dissatisfaction with the Oral Exam component of the course in the course evaluation surveys.
 Most students did not have that experience though, and overall student evaluations of the Oral Exam format were highly positive.
@@
 We recommend that instructors relegate Multiple Choice to assessing the lowest levels of Bloom's taxonomy, and even then to only use them in very large classes where the practical benefits outweigh the potential drawbacks.
+If an instructor is in the situation where Multiple Choice exams are being used, and they still desire to test higher-order learning objectives there are some best-practices avaiable in the literature.
+See for example in the context of teaching Biology~\cite{crowe2008biology}, or for Economics~\cite{buckles2006using}.
 \paragraph{}

### 2019-07-08 CUT/teaching-philosophy.tex
draft of teaching philosophy

\documentclass{article}

\usepackage{apacite}
\usepackage[margin=1in, bottom=1in, top=1in]{geometry}
\title{Statement of Teaching Philosophy - Draft}
\author{W. Justin Toth}
\begin{document}
\maketitle
\newpage
\section*{}
\paragraph{}
The most lucid way to explain my beliefs about to teaching is to explain my beliefs about learning. This is because my teaching philosophy is centred around optimally assisting students on their path to learning. To begin, let us consider why a student should study mathematics. People give many different answers to this question. Many mathematicians will say it should be studied for no other reason than its intrinsic beauty. Engineers and Scientists may say that mathematics teaches many useful formulae and algorithms that make excellent tools in the pursuit of science and engineering. Both of these reasons are very true, but there is an oft-overlooked third reason. Mathematics is fundamentally about developing abstract critical thinking, and unparalleled problem solving skills.

My teaching philosophy is deeply rooted in the three reasons above that a student will be attending my courses. Problem solving skills are central and universally valuable to all students and thus they lie at the core of my approach. From the very first lecture in my classes, students will be solving problems. Students retain material longer, and achieve deeper levels of learning when they play an active role in obtaining information. The fundamental flow of my lectures plays to this observation. They will often begin by introducing some very elementary definitions, and then immediately posing a question to the class. This forces students to engage their critical thinking early in the lecture, and to be active participants throughout the class meeting. It sets a precedent that the experience in my courses will be a dialogue. I will be spending as much time asking students questions, as they will be asking me. With larger classes I will impose more formal time points at which students break into small groups to discuss the problems posed, before bringing their attention back to the front. To support students' growth as problem solvers throughout the course, when it is time for me to give explanations to them I concentrate on the logical structure and manner of reasoning that allows me to reach the solution as much as I do the content of the solution itself.

Mathematics is full of a plethora of beautiful, deep, and mysterious results. I seek to instill an appreciation for this quality of mathematics, and the occumpanying sense of wonder, in my students. This goal is challenging to achieve. My strategy is to communicate intense passion. When describing a result that I think is particularly beautiful I will be unreserved in my appreciation for it. Quotes like ``The fact that is this true is really incredible!'', or  ``When I first learned this I simply couldn't believe it.'', or ``I think the pattern made by the solutions to this equation look mesmerizing.'' are not uncommon in my classroom. I think expressing emotional attachment to the material helps students get excited about the content, and helps to humanize myself as their instructor.

\paragraph{}
Something about how I address students motivated by applications.
\paragraph{}
I believe that one of the ultimate goals of education should be for students to become independent learners. Facilitating the growth of students so that one day they can become my peers is an important part of my approach to teaching. Towards this end, depending on the level of the class, I will structure aspects of the course to encourage greater independence among my students. For example, instead of giving a lecture on how to write some piece of code for a project, I will email students a link to the documentation, encourage them to read it, try to install the software, and come to office hours if they are stuck. Similarly in third year and up courses I will not provide solutions to assignments or exam practice problems. Instead I will use online discussion boards and incentivize students (sometimes with the aid of bonus marks) to post their solutions there after the class. Often this results in lively discussion, where students debate what the correct answers should be, challenging each other, explaining why did things a certain way, and learning all the more in the process. Altogether my teaching philosophy is to train independent problem solvers, adept at critical thinking, with a deep appreciation of the instrinsic beauty of their subject and a toolkit of incredibly useful mathematics.

\end{document}

### 2019-10-19 CUT/paper.tex
fixing abstract

@@
 \begin{abstract}
-We study of question of how exam formats impact student study behaviours, learning outcomes, and perception of the courses they take.
+We study the question of how exam formats impact student study behaviours, learning outcomes, and perception of the courses they take.
 We particularly focus on Mathematics and Science courses, but the findings in this survey are general enough to be widely applicable.

### 2019-11-07 CUT/teaching-philosophy.tex
teaching philosophy statement

@@
 \usepackage{apacite}
+\usepackage[colorlinks = true,
+linkcolor = red,
+urlcolor  = blue,
+citecolor = green,
+filecolor = cyan,
+anchorcolor = blue]{hyperref}
+\usepackage{setspace}
 \usepackage[margin=1in, bottom=1in, top=1in]{geometry}
@@
 \maketitle
+\tableofcontents
 \newpage
-\section*{}
+\section{Statement of Teaching Philosophy}
 \paragraph{}
-The most lucid way to explain my beliefs about to teaching is to explain my beliefs about learning.
-This is because my teaching philosophy is centred around optimally assisting students on their path to learning.
-To begin, let us consider why a student should study mathematics.
-People give many different answers to this question.
-Many mathematicians will say it should be studied for no other reason than its intrinsic beauty.
-Engineers and Scientists may say that mathematics teaches many useful formulae and algorithms that make excellent tools in the pursuit of science and engineering.
-Both of these reasons are very true, but there is an oft-overlooked third reason.
-Mathematics is fundamentally about developing abstract critical thinking, and unparalleled problem solving skills.
+Mathematics is a rich, deep tapestry of fascinating interrelated concepts about shapes and numbers.
+Many passionate teachers of the subject, myself included, share this sentiment.
+As an instructor, my goal is to transmit this perspective to my students.
+I want to stoke their passion, ignite their curiosity, and instill lifelong confidence in their understanding.
+Towards these lofty goals I have adopted methods of teaching which encourages students to take greater ownership over the material and their own learning process.
 
-My teaching philosophy is deeply rooted in the three reasons above that a student will be attending my courses.
-Problem solving skills are central and universally valuable to all students and thus they lie at the core of my approach.
-From the very first lecture in my classes, students will be solving problems.
-Students retain material longer, and achieve deeper levels of learning when they play an active role in obtaining information.
-The fundamental flow of my lectures plays to this observation.
-They will often begin by introducing some very elementary definitions, and then immediately posing a question to the class.
-This forces students to engage their critical thinking early in the lecture, and to be active participants throughout the class meeting.
-It sets a precedent that the experience in my courses will be a dialogue.
-I will be spending as much time asking students questions, as they will be asking me.
-With larger classes I will impose more formal time points at which students break into small groups to discuss the problems posed, before bringing their attention back to the front.
-To support students' growth as problem solvers throughout the course, when it is time for me to give explanations to them I concentrate on the logical structure and manner of reasoning that allows me to reach the solution as much as I do the content of the solution itself.
+The heart of the student-teacher relationship is the lecture, and I believe in transforming my lectures into an active dialogue between student and teacher to encourage greater engagement by students.
+During my lectures I alternate between explaining concepts and asking questions with a frequency that evenly divides time between me speaking and students speaking.
+For instance, when presenting a definition I immediately follow up asking students for an example which satisfies the definition, or during a proof I may begin by writing what we are trying to show and ask students what technique we should proceed with.
+I find this to be one of the most powerful techniques for teaching.
+It pushes students to actively engage with the material and also allows them to quickly perform self-diagnosis to see if they are absorbing the lecture content.
+It even helps me better understand my audience.
+I can quickly determine which students are easily understanding, which are struggling, which ones are not feeling confident about speaking, and which ones are giving their attention.
+This feedback allows me to adapt during the current and future lectures by changing pace, or stopping to clarify some important background concept I notice students are missing.
+Within the first couple weeks of any given term this method leads to significant changes in many students.
+Often in the first lecture students are hesitant to speak, and will only speak when called upon.
+As they get used to my style of teaching they become more confident in responding, and become more likely to volunteer ideas or ask questions without being prompted.
 
-Mathematics is full of a plethora of beautiful, deep, and mysterious results.
-I seek to instill an appreciation for this quality of mathematics, and the occumpanying sense of wonder, in my students.
-This goal is challenging to achieve.
-My strategy is to communicate intense passion.
-When describing a result that I think is particularly beautiful I will be unreserved in my appreciation for it.
-Quotes like ``The fact that is this true is really incredible!'', or  ``When I first learned this I simply couldn't believe it.'', or ``I think the pattern made by the solutions to this equation look mesmerizing.'' are not uncommon in my classroom.
-I think expressing emotional attachment to the material helps students get excited about the content, and helps to humanize myself as their instructor.
+I believe doing is necessary for learning, and central to the act of doing mathematics is problem solving.
+I see independent problem solving as a skill which can be taught and I use problem solving as an opportunity to foster student creativity and increase their comfort in engaging with the unknown.
+To teach problem solving I work proto-typical problems into my lectures.
+When working these problems, whether I’m presenting the next step or asking the students for it, the thinking which led to the step is front and center.
+I present problem solving in a four step approach as outlined by the famous mathematician George Polya in his book “How to Solve It”: understand the problem, devise a plan, carry out the plan, and look back.
+This framework gives students structure and I find it severely cuts back on generic questions about not knowing what to do.
+I see students reframing their questions more specifically in terms of where in the problem solving steps they are stuck, which in turn leads to a deeper understanding on their part, and makes it much easier to diagnose their challenges and provide meaningful feedback.
+As my students grow in confidence, I start using problems to provoke their curiousity.
+My lectures and assignments are peppered with optional exercises that highlight a subtle corner-case, interesting novelty, or surprising connection with material outside the course.
+This is an opportunity to broaden student horizons without overwhelming those who are already bogged down in coursework.
+Students who engage with this material will often come to office hours to discuss what they discovered, which leads to fascinating side discussions that enrich the course and capture student interest.
 
-\paragraph{}
-Something about how I address students motivated by applications.
-\paragraph{}
-I believe that one of the ultimate goals of education should be for students to become independent learners.
-Facilitating the growth of students so that one day they can become my peers is an important part of my approach to teaching.
-Towards this end, depending on the level of the class, I will structure aspects of the course to encourage greater independence among my students.
-For example, instead of giving a lecture on how to write some piece of code for a project, I will email students a link to the documentation, encourage them to read it, try to install the software, and come to office hours if they are stuck.
-Similarly in third year and up courses I will not provide solutions to assignments or exam practice problems.
-Instead I will use online discussion boards and incentivize students (sometimes with the aid of bonus marks) to post their solutions there after the class.
-Often this results in lively discussion, where students debate what the correct answers should be, challenging each other, explaining why did things a certain way, and learning all the more in the process.
-Altogether my teaching philosophy is to train independent problem solvers, adept at critical thinking, with a deep appreciation of the instrinsic beauty of their subject and a toolkit of incredibly useful mathematics.
+Ultimately student confidence hinges on their ability to recall material and identify where it is useful.
+Similarly, a major motivator for student passion is seeing how interconnected the material they are learning is with other material they have learned or will learn.
+Thus strengthening both passion and confidence can be done through putting material in context and forming a dense web of connections between different concepts.
+The basic organization of my courses, lectures, notes, and assignments supports this goal.
+It starts with the course syllabus where I outline the learning objectives for the course alongside a timeline of when these objectives will be achieved.
+At the start of a lesson I outline learning outcomes for that session, and review them again at the end of the lesson.
+Lectures themselves start with a review of previous material needed for the lecture, and end by describing what we learned that session and how we will need it in future sessions.
+Before exams I set aside lecture time to review the skills I expect students to have attained, and will be testing on that particular exam.
+Altogether this creates a pattern of reinforcement for the important concepts of the course, and supports students in drawing connections between the lectures over time.
+I dedicate special revision passes on my lecture notes and assignments to identifying connections with not only past and future lectures, but also with other courses the students may have taken in the past or will take in the future.
+This exercise is immensely valuable, not only for retention of material, but for motivation.
+I’ve seen tremendous improvements in student passion for the material when they come to see how the topics are situated within the bigger picture of what they are learning in their program.
+
+I view teaching as a craft, and I am constantly honing my tools and methods to perform better in this craft.
+I consider myself incredibly lucky to have the opportunity to share my knowledge, understanding, and ultimately passion with my students.
+I strive to make the most of this opportunity every time I step in front of a class or open my door for office hours.
+
+\section{Teaching Experience}
+
+\section{Teaching Strategies}
+
+\section{Evaluation of Teaching}
+
+\section{Professional Development}
+
+\section{Future Goals}
+
+\appendix
+\section*{Appendices}
+\section{}
 

### 2019-11-14 CUT/teaching-philosophy.tex
backing up current writing of dossier

@@
 \usepackage[margin=1in, bottom=1in, top=1in]{geometry}
-\title{Statement of Teaching Philosophy - Draft}
+\title{Teaching Dossier}
 \author{W. Justin Toth}
@@
 \maketitle
+\newpage
 \tableofcontents
@@
 This exercise is immensely valuable, not only for retention of material, but for motivation.
-I’ve seen tremendous improvements in student passion for the material when they come to see how the topics are situated within the bigger picture of what they are learning in their program.
+I've seen tremendous improvements in student passion for the material when they come to see how the topics are situated within the bigger picture of what they are learning in their program.
 
@@
 \section{Teaching Experience}
+
+\subsection{Courses Taught}
+\emph{CO $327$: Deterministic OR Models}
+
+ Role: Instructor
+
+ Term(s): Spring $2019$
+
+This course is an applications-focused introduction to the field of operations research for non-specialist students, meaning those not enrolled in an Honours Mathematics program, at the University of Waterloo.
+Its core topics are linear and integer programming models, duality theory, sensitivity analysis, and cutting planes and branch and bound methods for solving integer programs.
+The instructor also has some discretion in additional topics to cover from operations research after covering the aforementioned areas.
+I was the sole instructor for this course in Spring $2019$ and was fully responsible for all areas of the course including the syllabus, course materials, office hours, exams, and $80$ minute lectures which occurred twice per week.
+
+Course notes from previous offerings were available to me, but were considerably out of date.
+I decided to rewrite the course notes from scratch.
+I also created new assignments and exams for the course to align with the updated course notes I had written.
+For the additional topics, traditionally dynamic programming or stochastic optimization is taught.
+In seeking feedback from my class, I found that they were not interested in these topics and I decided to teach them about stable matching: a Nobel Prize winning method of assigning goods in two-sided markets which had never been taught in the course before but was extremely relevant and the students were enthusiastic about learning.
+
+\subsection{Teaching Assistantships}
+\paragraph{}Over the course of my graduate studies at the University of Waterloo I have had the priviledge of being a teaching assistant for many different courses.
+The standard duties of a TA in our department, Combinatorics and Optimization, involve grading, proctoring exams, and holding office hours to support students seeking help.
+Below I have described the courses to which I was assigned a teaching assistantship, as well as any additional duties I was responsible for in each course.
+\bigskip
+
+\noindent$\cdot$\emph{CO $456$: Introduction to Game Theory}
+
+Term(s): Fall $2019$, Fall $2018$, Fall $2017$, Fall $2016$
+
+This fourth year course offers students a mathematically rigorous introduction to the field of game theory.
+It covers combinatorial games, strategic games, the existence and computation of Nash Equilibria, fixed point theorems, mechanism design, and cooperative game theory.
+In addition to the standard TA duties, I was a guest lecturer for this course in $2019$ teaching a series of three lectures on Sperner's Lemma, Brouwer's Fixed Point Theorem, and Nash's Theorem.
+I was also a guest lecturer in $2018$, giving a lecture on correlated equilibria.
+
+In the $2017$ and $2016$ offerings of the course there was a programming project component where the students would work in groups, using their knowledge of game theory to code AI competitors for a simulated tournament.
+During these offerings I was responsible for programming the tournament simulation, and making sure the competitions which were held throughout the term ran smoothly.
+
+\noindent$\cdot$\emph{CO $353$: Computational Discrete Optimization}
+
+Term(s): Winter $2019$, Winter $2018$, Winter $2017$, Winter $2016$
+
+CO $353$ covers the algorithmic aspects of discrete optimization.
+This includes problem formulations, greedy algorithms, local-search heuristics, approximation algorithms, linear programming duality, cutting planes, branch-and-bound, column generation, dynamic programming, problem reductions and NP-hardness.
+
+\noindent$\cdot$\emph{CO $351$: Network Flow Theory}
+
+Term(s): Spring $2018$
+
+Network Flow Theory covers the fundamental properties and algorithms for flows on directed graphs.
+It covers shortest path problems, the max-flow min-cut theorem and applications, minimum cost flow problems, network simplex and primal-dual algorithms.
+
+\noindent$\cdot$\emph{CO $250$: Introduction to Optimization}
+
+Term(s): Spring $2017$
+
+This second year course is a typically an undergraduate's first exposure to optimization.
+The central focus of this course is to teach students linear programming theory and methods for solving linear programs.
+
+\paragraph{}
+Early in my graduate career I was assigned teaching assistant positions for two faculty of mathematics courses given to first year students at the University of Wateroo.
+In addition to the standard TA duties, for theses courses I was responsible for running a weekly tutorial where I would prepare problems for the students to practice in class, demonstrate key techniques, and provide reviews of critical course material.
+\bigskip
+
+\noindent$\cdot$\emph{MATH $128$: Calculus $2$ for Sciences}
+
+Term(s): Spring $2016$
+
+This is the second course in the calculus sequence at Waterloo for science students, i.e.\ students in physics, chemisty, biology, and economics.
+This course is typically the last calculus course for many science students and tries to cover a broad range of topics.
+It teachs the evaluation of integrals, applications to volumes and arc lengths, differential equations, sequences and series, Taylor's Theorem, parametric and vector representations of curves, and polar coordinates in the plane.
+
+\noindent\emph{MATH $136$: Linear Algebra $1$}
+
+Term(s): Fall $2018$.
+
+This is the first linear algebra course offered to Honours Mathematics students at the University of Waterloo.
+It covers systems of linear equations, matrix algebra, elementary matrices, computational issues.
+It also covers Real $n$-space, vector spaces and subspaces, basis and dimension, rank of a matrix, linear transformations, and matrix representations, determinants, eigenvalues and diagonalization.
+
+\paragraph{}During my undergraduate degree at the University of Windsor I had the opportunity to be a teaching assistant during my final year of study.
+I was assigned to TA the introductory linear algebra course for first years in the Fall term of $2014$, and the introductory course on mathematical proof in the Winter term of $2015$.
 

### 2019-11-15 CUT/teaching-philosophy.tex
Working up to Outside the Classroom section

@@
 \subsection{Teaching Assistantships}
-\paragraph{}Over the course of my graduate studies at the University of Waterloo I have had the priviledge of being a teaching assistant for many different courses.
+\paragraph{Graduate TA}Over the course of my graduate studies at the University of Waterloo I have had the priviledge of being a teaching assistant for many different courses.
 The standard duties of a TA in our department, Combinatorics and Optimization, involve grading, proctoring exams, and holding office hours to support students seeking help.
@@
 
-\paragraph{}During my undergraduate degree at the University of Windsor I had the opportunity to be a teaching assistant during my final year of study.
+\paragraph{Undergraduate TA}During my undergraduate degree at the University of Windsor I had the opportunity to be a teaching assistant during my final year of study.
 I was assigned to TA the introductory linear algebra course for first years in the Fall term of $2014$, and the introductory course on mathematical proof in the Winter term of $2015$.
+Both courses saw me giving a weekly tutorial based on the assigned problem set for that week.
+I was also responsible during this time for working in the University's tutoring centre.
+My role there was to provide one-on-one or small group support to any students who came in with questions about math courses at the university.
+This opportunity was an interesting challenge because it required a broad understanding of the courses offered by the university that particular term, since any student studying math could approach me expecting to receive help with their coursework.
 
+\subsection{Guest Lectures}
+\paragraph{Introduction to Game Theory}
+As I discussed in Section $2.2$ I have had the opportunity to give guest lectures in CO $456$, Waterloo's mathematically rigorous introduction to game theory.
+In the Fall $2018$ offering, I gave a single $80$ minute lecture on the topic of correlated equilibria.
+This was the first time that particular topic was taught in the course.
+As such, I was responsible for designing not only the lesson plan from scratch, but also preparing suitable exam and assignment problems related to the material.
+I used a post-test technique in the lecture where I saved the last ten minutes of instruction time for the students to solve a hands-on problem in class of computing a correlated equilibrium for a small strategic game.
+The head instructor of the course, who was observing my lecture, commented that he found the technique to be a very engaging way for the students to review the material and he would be interested in trying it himself in the future.
+
+I also had an opportunity to give a series of three $50$ minute guest lectures during the Fall $2019$ offering of CO $456$ during a week where the head instructor was travelling for a conference.
+They had just finished up a section of course material, and I was given free reign to choose the next topic.
+I chose Nash's Theorem that every finite strategic game has a mixed Nash equilibrium.
+The first two lectures concentrated on the fundamental tools needed to prove the theorem: Sperner's Lemma and Brouwer's Fixed Point Theorem.
+The last lecture discussed the combination of these two results to obtain Nash's Theorem.
+I used a strategy of providing a high-level outline at the start and end of each lecture so that students could maintain a big picture view of the proof of Nash's Theorem as we progressed through the technical details throughout the week.
+
+\paragraph{Microteaching} During my Master's degree I took a professional teaching development program through the University of Waterloo's Centre for Teaching Excellence called the Fundamentals of University Teaching program.
+As part of this program, participants were required to give a series of three $15$ minute mini-lectures to their peers and the facilitators of the program.
+The lectures were supposed to highlight three different areas: a subject suitable for first year students of your discipline, a subject related to your area of research, and a specific teaching method.
+I chose the Gaussian elimination process for solving linear systems from a first year linear algebra course, the prisoner's dilemma from game theory, and the Think-Pair-Share technique respectively for my three lectures.
+These sessions were a valuable learning experience.
+The feedback from peers and teaching experts was invaluable, and the tight time constraints really helped me hone my lesson planning and time management skills in the classroom.
+\subsection{Outside the Classroom}
+\paragraph{TORCH}Teaching is a major part of my life, and I believe an important part of teaching is inspiring young students to pursue higher education.
+The Operations Research Challenge, TORCH for short, is a contest given to high school students which focuses on solving algorithmic operations research problems.
+The contest is hosted simultaneously at the Universities of Waterloo, Toronto, and Concordia.
+As an outreach program, it aims to promote awareness and interest in operations research among high school students.
+I have been a volunteer for the annual contest in $2018$ and $2019$.
+In this role I was responsible for giving contest students real time feedback on their proposed solutions, as well as discussing in between sessions what studying operations research in university is like and encouraging students to pursue studies in STEM.
+
+In $2020$ I am transitioning into an executive role on Waterloo's organizing team.
+My role will be to lead to the creative design of problem sets for the contest.
+I will be responsible for creating contest problems which are challenging, engaging, relevant to student interests, and accessible to students of all background levels.
+In addition to creating problems myself, I will be overseeing other question writers to ensure a high standard of quality in questions provided for the contest.
+
+\paragraph{Brazilian Jiu-Jitsu}
 \section{Teaching Strategies}

### 2019-11-15 CUT/teaching-philosophy.tex
writeup on jiujitsu

@@
 \paragraph{Brazilian Jiu-Jitsu}
+My love of teaching extends outside of my profession and into my hobbies as well.
+I have been practicing the martial art of Brazilian Jiu-Jitsu for over $10$ years.
+My experience in the sport found me in a position of seniority among the students who train at my current gym.
+This seniority earned me a position as an assistant instructor at Alliance Fitness in Waterloo Ontario.
+Since $2019$ I have been teaching a weekly beginners class at that gym.
+I have been able to transfer my understanding of pedagogy, and my teaching experience in academia, to this new setting to quickly become one of the top performing instructors in the gym.
+
+Pedagogy in teaching sports to hobbyists lags far behind the understanding of pedagogy present in teaching research.
+Even implementing simple things like designing a syllabus for the beginner's course and preparing lessons plans makes for a marked improvement over the status quo in the area.
+My classes consistently have twice as high of an attendance over comparable beginner class time slots (workday evenings) held by different instructors in the gym.
+Furthermore, the rate of technique aquisition among my students is expectionally high.
+I have shortened the expected time for learning certain sport specific skills from many months to a couple of weeks in our beginner class.
 \section{Teaching Strategies}
@@
 \section{Evaluation of Teaching}
+
+\subsection{Student Evaluations}
+
+\subsection{Teaching Observations}
 

### 2019-11-16 CUT/teaching-philosophy.tex
outlining the rest of dossier

@@
 I have shortened the expected time for learning certain sport specific skills from many months to a couple of weeks in our beginner class.
+
 \section{Teaching Strategies}
+%reread philosophy statement and pick out teaching strategies to describe in detail
+%write a stream of consciousness about teaching strategy that you can later organize thoughts from
+%consider whether it is better to organize by teaching context or by technique
 
 \section{Evaluation of Teaching}
-
+I find feedback on my teaching to be tremendously valuable.
+It is the easiest way for me to identify potential avenues for improvement, while at the same time motivating me by highlighting strong points of my teaching style.
+I have organized my evaluations below into two categories: those from students in courses I have taught, and those from peers and teaching professionals observing my lectures.
 \subsection{Student Evaluations}
-
+%Chart out survey responses for CO 327 with survey adjectives for context for scores (some here, rest in appendix)
+%Student comment strong points
+%Student comment targets for change
 \subsection{Teaching Observations}
+%highlights from 1st observation
+%highlights from 2nd observation with a narrative about improving
 
 \section{Professional Development}
-
+%FUT
+%Math Graduate Teaching Seminar
+%CUT
 \section{Future Goals}
+%Write a textbook
+%Teach a fourth year course
+%Teach a topics course
+%Become more invovled in teaching 1st and 2nd years
+%Head instructor for a course with multiple sections
+%Graduate and URA supervision
 
@@
 \section*{Appendices}
+%Full evaluations
+%Observation Reports
+%Sample Lesson Plan from guest lecture
+%Sample Assignment from CO 327
+%Sample Lesson Plan from CO 327
 \section{}

### 2019-11-17 CUT/teaching-philosophy.tex
drafting lecture section of teaching strategies

@@
 
-\noindent$\cdot$\emph{CO $456$: Introduction to Game Theory}
+\noindent\emph{CO $456$: Introduction to Game Theory}
 
@@
 During these offerings I was responsible for programming the tournament simulation, and making sure the competitions which were held throughout the term ran smoothly.
+\bigskip
 
-\noindent$\cdot$\emph{CO $353$: Computational Discrete Optimization}
+\noindent\emph{CO $353$: Computational Discrete Optimization}
 
@@
 This includes problem formulations, greedy algorithms, local-search heuristics, approximation algorithms, linear programming duality, cutting planes, branch-and-bound, column generation, dynamic programming, problem reductions and NP-hardness.
+\bigskip
 
-\noindent$\cdot$\emph{CO $351$: Network Flow Theory}
+\noindent\emph{CO $351$: Network Flow Theory}
 
@@
 It covers shortest path problems, the max-flow min-cut theorem and applications, minimum cost flow problems, network simplex and primal-dual algorithms.
+\bigskip
 
-\noindent$\cdot$\emph{CO $250$: Introduction to Optimization}
+\noindent\emph{CO $250$: Introduction to Optimization}
 
@@
 It teachs the evaluation of integrals, applications to volumes and arc lengths, differential equations, sequences and series, Taylor's Theorem, parametric and vector representations of curves, and polar coordinates in the plane.
+\bigskip
 
@@
 \section{Teaching Strategies}
-%reread philosophy statement and pick out teaching strategies to describe in detail
+%reread philosophy statement and pick out teaching strategies to describe in detail:
+%       questioning strategies in lecture
+%       outlining, learning outcomes, reviews
+%       problem solving frameworks
 %write a stream of consciousness about teaching strategy that you can later organize thoughts from
 %consider whether it is better to organize by teaching context or by technique
+\subsection{In Lecture}
+\paragraph{}
+The lecture is the fundamental unit of instruction in a university setting.
+As such I place a great emphasis on preparing my lectures in a way that enables my students to extract as much benefit as possible from each lecture.
+For me, the purpose of a lecture is to give my students learning outcomes, sufficient motivation to attain them, and all the tools necessary to make achieving those learning outcomes as easy as possible.
 
+\paragraph{Learning Outcomes}Everything I do in lecture is centre around learning outcomes.
+For each lecture, depending on time and where we are in the course, I will identify one to four learning outcomes I want the students to get from the lecture.
+I will write them down on the leftmost board in the room, leave them up for the entire session, and check them off as we go.
+My learning outcomes are always specific, descriptive, and measurable.
+For example, not ``learn Braess's paradox'' but ``Braess's paradox: demonstrate how adding a road to a traffic network can increase total congestion''.
+Structuring the session around learing outcomes and presenting them at the start serves to prime students' brains for learning, clarify expectations for them, and provide a helpful structure to the lecture allowing students to organize their thoughts and measure progress.
+\paragraph{Past, Present, and Future} Each lecture does not exist in a vacuum.
+Typically their is a lecture which will proceed it, and  a lecture which will follow it.
+When preparing a lecture, I reflect on which content from previous lectures will be needed to achieve the day's learning outcomes, and where the learning outcomes of the day will be needed in future lectures.
+At the start of lecture I add bullets to the outline containing the learnign outcomes describing the skills from previous lectures which will be relevant today, and bullets describing how the content of the day will feed into future goals in the course.
+This serves to put the material in context, which supports student retention.
+It allows them to form connections between what they learned last class, or last week, or maybe even last month and what they are working on today.
+Furthermore, the mention of future goals motivates them to study the material of this session and mentally prepares them for what is coming next.
+
+With the past, present, and future learning outcomes fleshed out, my outline is complete.
+I will make reference to this outline throuhgout the lecture, as the relevant pieces come up.
+Further, I save time towards the end of the session to review the entire outline.
+Allowing me to once again summarize the learning outcomes of the day and how they fit into the bigger picture of the course.
+
+\paragraph{Pre-test}Directly after the outline of learning outcomes, I will usually continue the lecture with a pre-test activity.
+The goal of a pre-test is to evaluate, with no stakes, the baseline of comprehension for the material that will be covered in the day's lecture.
+In cases where the lecture exists in a series on related material for which I already have a good feel for the level of understanding my students have I will skip the pre-test.
+That said, the pre-test is a valuable method to help me calibrate the level of my instruction to the level of the students in the room, and I only omit it as a conscious decision when I feel my teaching level is well calibrated to the students already.
+The pre-test does not need to take more than a couple minutes.
+It can be as simple as an informal show-of-hands poll on whether students have heard of the main concept of discussion today and to what extent.
+That is the method I favour and use most often.
+Another option is to give a short exercise at the start of class.
+This method is somewhat more time consuming.
+I only use it when I need to be absolutely certain the students are at a certain level of learning attainment.
+Say for example, I am teaching a very technical proof that day and want to be sure that students have grasped the critical lemma from the previous lecture.
+The pre-test is only as valuable as my willingness to adapt based on its results.
+This means building in some flexibility into my lecturing style.
+If I observe the majority of students are lacking in background, then I may have to spend more time on the background during that lecture and not get as far into the content I planned.
+Conversely, if the students are exceptionally prepared, I need to be willing to skip content they have previously seen and are comfortable with and be prepared to go deeper into the material than initially planned.
+
+\paragraph{Post-test}The mirror image of the pre-test is the post-test.
+Partially I use the post-test to signal to myself how well the students absorbed the material from this session, but more importantly I use the post-test to signal to students how well they absorbed the material from this session.
+A post-test is a small exercise performed at the end of a lecture to evaluate attainment of the desired learning outcomes of that lecture.
+In practice, I want post-tests to be fast, taking around five minutes, and they usually will target the lower levels of Bloom's taxonomy.
+Usually they will feature a calculation using a technique from the day's lecture, or ask students to give another example of a concept presented today.
+In more proof-focused courses, they may be asked to observe a simple corollary which directly follows from a lemma presented in class.
+The post-tests are meant to be done individualy or in small groups.
+Some groups finish quickly and are free to leave early.
+Other groups realize they did not understand something from the lecture they initially thought they did, and it triggers questions that I can immediately address for them.
+
+On questioning strategies and problem solving frameworks.
+
+\subsection{Outside Lecture}
+
+\subsection{Assessment}
 \section{Evaluation of Teaching}

### 2019-11-18 CUT/teaching-philosophy.tex
Questioning strategies complete

@@
 
-On questioning strategies and problem solving frameworks.
+\paragraph{Questioning Strategies}Between the pre-test and the outline review lies the bulk of the lecture time to be used to deliver content.
+I like to give my lectures using a high density of questions for the students.
+This is a simple, efficient, and effective way for me create a highly interactive learning environment within the lecture.
+Sometimes it can be difficult to get students to open up and get used to the amount of spoken input I expect from them during the lecture.
+I have developed some techniques for increasing my success rate with using frequent student questioning.
 
+The first is to set the precedent early.
+I mean this in two senses.
+I start using questioning from the first day of lecture, and make sure to heavily ask student questions every single lecture.
+This is much easier than switching to this technique part way through the semester, or not doing for a few lectures, then having an interactive lecture.
+Once the students adapt to the routine of the interactive lecture they will be more comfortable answering questions, and even come to expect to be given the opportunity for input.
+I also mean to set a precedent on the lecture scale.
+Once I start delivering the lecture material, I am looking for the earliest opportunity to throw a question at the students.
+Usually this is immediately after the first definition or example I present.
+This gets the students focused on actively engaging with the material and responding early in the lecture, and builds momentum that can be carried through the entire session.
+As an instructor, when I get in the habit of constantly looking for the nearest opportunity to ask a question, I find my lectures are far more interesting for the students and their depth of understanding increases significantly.
+
+On a question-by-question level it can be difficult to coax responses out of students.
+It is important that I ensure I wait sufficiently long for students to respond.
+I use a ten second count in my head, or take a sip of water, to make sure I'm pausing long enough for a student to both work out the answer and work up the confidence to respond.
+Most of the time no hands go up immediately, but after a waiting period some hands will go up.
+Although sometimes students are not responding because they genuinely do not know the answer.
+When I get a sense this is happening, I will break the question down for them into smaller and easier parts.
+At some level I will find something simple enough that someone in the class will know the answer and be willing to respond.
+I prefer to not simply give the answer and move on.
+Although it may feel like the best thing to do to preserve momentum and save time, the fact that students are not responding means we have identified a gap in their understanding and it is important to drill down on this gap and help students bridge it piece by piece.
 \subsection{Outside Lecture}

### 2019-11-21 CUT/teaching-philosophy.tex
wrote up on first teaching observation

@@
 Although it may feel like the best thing to do to preserve momentum and save time, the fact that students are not responding means we have identified a gap in their understanding and it is important to drill down on this gap and help students bridge it piece by piece.
+
 \subsection{Outside Lecture}
@@
 \subsection{Assessment}
+
 \section{Evaluation of Teaching}
@@
 \subsection{Teaching Observations}
+One element of the CUT teaching program I took alongside my PhD was having lectures I give upervised by an expert in teaching working with the university's Centre for Teaching Excellence (CTE).
+I completed two teaching observations through the program.
+Below I summarize the feedback I received from these observations.
 %highlights from 1st observation
+\subsubsection*{First Teaching Observation}
+\paragraph{}
+My first observation was on the guest lecture I gave for Introduction to Game Theory in Fall $2018$ on Correlated Equilibria.
+The observation was conducted by Shahrukh Athar, a Graduate Instructional Developer working for the CTE.
+I have included the relevant sections of the full report in Appendix \ref{sec:observation-report-1}.
+The feedback in the report was divided in \emph{aspects to maintain}, where strengths of teaching were identified, and \emph{targets for change}, where opportunities for improvement were identified.
+\paragraph{Aspects to Maintain} The first strength Shahrukh identified was my use of questions and comprehension checks.
+He appreciated the variety of checks I used, as well as the split between content-specific checks and general purpose open-ended questions.
+He commented that ``\emph{All of this is
+fantastic as it demonstrates a learner-centric approach to teaching rather than a teacher-centric
+one. It shows that you want to take your learners along with you as the lesson progresses and
+are willing to pause and clarify confusions before moving on.}''
+
+Shahrukh also found my lesson structure and transitions to be very strong, saying he will ``\emph{encourage [me] to continue
+following such a structured approach in the delivery of [my] future lessons}''.
+He appreciated my use of an outline at the start of the lesson the students could follow along with, and found my method of using comprehension checks to indicate upcoming transitions to the next component of a lesson to be a highly effective technique.
+
+Lastly, Shahrukh really enjoyed the high degree of interactivity in my lessons.
+Citing both the instructor-student questions throughout the lesson, and student-student interaction which occurred during the post-test.
+He particularly liked the latter, ``\emph{ I specifically want to celebrate here is the learner-learner
+activity that you had towards the end of the lesson}.'' The activity he refers to is a computation of a correlated equilibrium that the students were asked to perform in pairs during the last $10$ minutes of the lecture.
+
+\paragraph{Targets for Change} Shahrukh identified for me a couple of areas in which I could improve my teaching.
+The first area he identified involved some practical matters he dubbed ``awareness of learner needs''.
+He sat at the back of the classroom for the lecture, and found that from his position my handwriting was ``\emph{a touch small}''.
+The fix for this is easy, and I have since adapted it in all my lessons.
+Before the lecture starts I write some sentences and pieces of mathematics on the board, then head to the back of the room to verify that the font size is legible from that far away.
+Shahrukh also observed that there were a couple of noise disruptions caused by students walking in the halls outside the classroom since I left the doors open.
+I had initially done this to be welcoming to students who arrive late, but have since starting closing doors at the start of the lecture after verifying there are no students rushing through the hall trying to reach the classroom.
+
+Shahrukh noticed a habit I had during the lecture of speaking while writing and facing the board.
+While there were no issues with my voice projecting, it created a visual disconnect with the students.
+I have since adopted his strategy of ``Write-Turn-Talk'', where I am silent during writing on the board and turn around to explain what I just wrote to my students.
+A major strength of this technique is that it does not ask students to process two streams of information at once (my voice and my writing), but rather they can concentrate on copying what I a writing as I write it and then concentrate on the verbal explanation I give after it is written down.
+In my second teaching observation I specifically asked the observer to verify that I had successfully implemented this technique.
+
+\subsubsection*{Second Teaching Observation}
+\paragraph{}My second observation was conducted by Caelan Wang, a Graduate Educational Developer with the CTE.
+The lecture to be observed was one my regularly scheduled lectures as the instructor for Deterministic OR Models in Spring $2019$, covering cutting plane methods for solving integer programs.
+Caelan was especially qualified to observe this lecture since she was a graduate from the same department I am pursuing my PhD in, and knew the topic well.
+
+\paragraph{Aspects to Maintain}
+
+\paragraph{Targets for Change}
 %highlights from 2nd observation with a narrative about improving
@@
 %Sample Lesson Plan from CO 327
-\section{}
+\section{Teaching Observation Report 1}\label{sec:observation-report-1}
 

### 2019-11-22 CUT/teaching-philosophy.tex
preamble to detailed feedback from report 2

@@
 
+Caelan's general impression of the lesson can be summarized as ``\emph{You were confident, clear, and
+knowledgeable in the classroom.
+Overall, your lesson was definitely a success}.''.
+As with the previous report, more detailed feedback was separated into aspects to maintain and targets for change.
+The relevant sections of the full report can be found in Appendix \ref{sec:observation-report-2}
 \paragraph{Aspects to Maintain}
@@
 
+\section{Teaching Observation Report 2}\label{sec:observation-report-2}
 \end{document}

### 2020-02-05 CUT/teaching-philosophy.tex
outside the classroom

@@
 \subsection{Outside Lecture}
+\paragraph{}
+While time in lecture is critically important for student development the vast majority of student time is spent outside of lecture.
+I encourage students to continue learning outside of lecture by providing resources for them to ask questions and receive feedback, and by providing compelling assignment problems which encourage constant thinking and revision of the material.
+While I believe in using assignments and projects which require students to work outside of lecture time, I believe it is important to recognize that my courses are not the only courses students are taking during the term.
+Therefore a big component of my approach to teaching and learning outside of the lecture is balancing my desire to keep have students keep the course material front of mind on regular basis with the need to be respectful of my students' time and busy schedules.
 
+\paragraph{} Opt-in engagement is a great way to give students learning opportunities outside the lecture environment without overloading there schedules.
+The classic office hours are an example of this.
+The great benefit of office hours is that they allow me to give attending students individualized attention and to target their specific needs.
+Office hours are great for clearing up student questions, but they also present an even greater opportunity to spur student interest in the subject.
+After clearing up student questions I try to take the opportunity to engage in conversation with the student about why they are taking the course, what their interests are, etc.
+This often will lead to interesting discussions that run well outside the scope of a particular course.
+These discussions are nevertheless incredibly important, since they help me learn about student goals (and give them a chance to articulate their goals) and let me open the student's eyes to the wider possibilities of what the subject can offer.
+
+Office hours are not always convenient for students, and it's rare that timeslots can be found which work for every student.
+Some students are also anxious about talking face to face with their professor, and might prefer to discuss course material with other students.
+To faciliate this I use online discussion forums such as Piazza with my students.
+The teaching assistants and myself will monitor the forum to provide authoritative answers to student questions, but usually this amounts to nothing more than confirming a well-written reply by another student is in fact correct.
+These online forums help create a constant conversation between students on the course material while not putting too much burden on their schedules.
+To faciliate and encourage discussions, I will use the forums to have students create assignment solutions.
+After an assignment has been completed, a topic on the discusion board will appear for each question on the assignment.
+In that topic, the students are encouraged to discuss and crowdsource together the solution to the assignment problem.
+For more intricate or subtle questions I have found these topics to lead to a lot of stimulating conversation amongst the students.
+
+\paragraph{}
 \subsection{Assessment}
 
+\paragraph{} \textit{Importance of spaced repetition for studying. Importance of priming students for lecture via problems. Assigment problems and timing}
+\begin{itemize}
+    \item Multi-modal assessments
+    \begin{itemize}
+        \item Classic problem sets
+        \item Oral interviews
+        \item Long term projects with group collaboration
+        \item Mdterms and Final Exam with cheatsheet format.
+    \end{itemize}
+    \item Oral vs Written exams
+    \item Importance of Collaboration and long term projects
+    \item The cheatsheet format for exams and its benefits
+\end{itemize}
 \section{Evaluation of Teaching}
@@
 \paragraph{}My second observation was conducted by Caelan Wang, a Graduate Educational Developer with the CTE.
-The lecture to be observed was one my regularly scheduled lectures as the instructor for Deterministic OR Models in Spring $2019$, covering cutting plane methods for solving integer programs.
+The lecture to be observed was one of my regularly scheduled lectures as the instructor for Deterministic OR Models in Spring $2019$, covering cutting plane methods for solving integer programs.
 Caelan was especially qualified to observe this lecture since she was a graduate from the same department I am pursuing my PhD in, and knew the topic well.

### 2020-02-22 CUT/teaching-philosophy.tex
Future goals

@@
 \subsection{Assessment}
+\paragraph{}
+My approach to assessment is based not only only my experience as a teacher but on my understanding of the research literature.
+During the CUT program, I wrote a research paper surveying the academic literature on how exam formats impact student learning outcomes.
+Some interesting findings were synthesized from this paper which I now incorporate into my approach to student assessment.
+\paragraph{}
+Written midterms and final exams are fundamental evaluations which occur during courses I teach.
+They are usually cummulative and are worth a significant portion of a student's final grade.
+Naturally this means that many students will experience so-called test anxiety while preparing for and taking these exams.
+The literature show that offering students open book exams, i.e.\ exams where they are allowed to reference textbooks and/or course notes, significantly reduces test anxiety.
+Unfortunately the literature also shows that open book exams lead to significantly lower long term retention of course material than closed book exams.
+Interestingly, there is a way to obtain the same test anxiety reducing benefits of an open book exam while maintaining the long term retention of a closed book exam.
+This way is the cheatsheet exam, which I use in all my courses.
+For exams which do not have a predetermined format decided by institutional policy I allow students to bring into the exam a small cue card on which they can write whatever they want.
+The literature shows that this format has comparable reductins in test anxiety among students as an open book exam, while also encouraging postive study habits which lead to long term retention of the material since students are forced to prioritize the information they write on their cue card.
+\paragraph{}
+Alongside written exams I like to encorporate an oral exam component into my courses.
+This is implemented in an interview format, where a student is given a topic present on, such as a significant theorem from the course, they then prepare their presentation and schedule with me a time where they give the presentation and I ask questions to probe the depths of their understanding.
+I like to give students a list of topics from which they can engage in self-directed study for this oral exam.
+The research literature shows that oral exams not only improve student confidence, encourage effective study habits, and strengthen communication skills, but that performance on oral exams is only weakly correlated with performance on written exams.
+This suggests that they test a different set of skills.
+In the interest of celebrating the strengths of a diverse student body I strive to incorporate exam formats like oral exams which give students who may struggle with written exams an opportunity to demonstrate their knowledge and learning attainment.
+\paragraph{}
+Whatever plans students may have after graduation they will likely need to be able to work as part of a team.
+Increasingly mathematics research is done collaboratively, and in industry every professional is embedded as part of a collaborative team.
+Thus I believe it is important for students to have opportunities to work together in their courses.
+I usually give students this opportunity in the form of a long term open ended group project.
+For instance, in a course on optimization students may be given a more applied project based on an industry case study.
+Each group would be tasked with proposing a model for the problem and solving it using techniques from the course in whatever way they see fit.
+This encourages them to not only develop their ability to work with others, but also to leverage their creativity.
+Finally, it gives them a chance to see how the mathematics they learn in the classroom can have real world impact.
 
-\paragraph{} \textit{Importance of spaced repetition for studying. Importance of priming students for lecture via problems. Assigment problems and timing}
-\begin{itemize}
-    \item Multi-modal assessments
-    \begin{itemize}
-        \item Classic problem sets
-        \item Oral interviews
-        \item Long term projects with group collaboration
-        \item Mdterms and Final Exam with cheatsheet format.
-    \end{itemize}
-    \item Oral vs Written exams
-    \item Importance of Collaboration and long term projects
-    \item The cheatsheet format for exams and its benefits
-\end{itemize}
 \section{Evaluation of Teaching}
@@
 %Student comment targets for change
-\subsection{Teaching Observations}
+\subsection{Teaching Observations}\label{sec:teaching-observations}
 One element of the CUT teaching program I took alongside my PhD was having lectures I give upervised by an expert in teaching working with the university's Centre for Teaching Excellence (CTE).
@@
 \paragraph{Aspects to Maintain}
+Caelan appreciated the structure of my lesson and my preparation.
+Specifically she highlighted my use of a lecture outline and my starting the session with a review of the previous lecture's material connecting that content to the upcoming lecture.
+She also appreciated my confident delivery, saying ``\emph{in fact, you did not stumble even once in your teaching, which was impressive.}''
 
+Caelan was also impressed with my handwriting and board use.
+She found the presentation very clear and liked the use of emphasis with bullet points and underlining, as well as the strategic positioning of figures so that they can stay on the board as long as possible given the small amount of board space in the particular room the lecture took place in.
+
+Finally, Caelan celebrated my use of examples and questioning strategies.
+She liked how I presentated examples to motivate upcoming theory and definitions, and would revisit examples throughout the lecture to help reinforce new ideas as the lecture progressed.
+She said ``\emph{Such well-thought examples and questions really kept your learners actively engaged during your lesson.
+I would highly encourage you to continue these
+great practices in your future teaching!}''.
 \paragraph{Targets for Change}
 %highlights from 2nd observation with a narrative about improving
+One suggestion for improvement Caelan had was to remember to repeat questions or responses given by students in class.
+Her argument being that even though the classroom was small students are concentrating on the instructor, not necessarily each other, and it is helpful to do this to make sure they hear important information.
+I agree with her assessment, and this is an easy thing to tweak in my lecturing technique.
+
+I had asked Caelan to pay attention to the Write-Turn-Talk technique I was working on from Shahrukh's feedback on my previous observation.
+She noted that I had successfully implemented the habit of not talking while writing on the board, but she had a related taret for change.
+While I do not write and talk, I would sometimes face the board to point out aspects of a definition or elements of a figure I had drawn in relation to a student's question.
+She suggested an adjustment where I practice facing the audience while pointing to elements of the figure.
+This can be tricky, but the idea would be to look at the figure, point out what I have in mind, turn and then discuss it.
+In future lectures I plan to practice this technique and I think it will certainly help strengthen communication with my students.
 
 \section{Professional Development}
-%FUT
-%Math Graduate Teaching Seminar
-%CUT
+\subsection{Fundamentals of University Teaching}
+\paragraph{}
+During my Masters degree I completed the Fundamentals of University Teaching program from the CTE at the University of Waterloo.
+The program consisted of six workshops which focused effective lesson plan creation, questioning strategies, designing student assessments, motivating students, managing student atteiont, and giving and receiving feedback.
+Also as part of the program I gave three micro-teaching sessions to my peers in the program.
+A microteaching session consists of brief $15$ minute lecture with an emphasis on active learning techniques and student engagement.
+Each session was evaluated by an affiliate of the CTE from whom I received feedback that identified aspects of my teaching to maintain and targets for change.
+This program provided an excellent foundation for my approach to university teaching.
+
+\subsection{Certificate in University Teaching}
+\paragraph{}During my PhD studies I had an opportunity to expand upon the teaching skills I developed in the Fundamentals of University Teaching via the Certificate in University Teaching program.
+This program is more intensive and in-depth than the Fundamental program.
+This program involved four workshops: understanding the learner, interactive teaching strategies, assessing student learning, and course design.
+Each workshop required me to complete a response paper where I reflected on how the concepts from the workshop could apply in my teaching.
+The program also involved two in-class teaching observations by a CTE Graduate Instructional Developer, the resulting feedback from which I discuss in Subsection \ref{sec:teaching-observations}.
+
+For the program I wrote a research paper summarizing the current research literature on ``Impact of Exam Format on Student Learning Outcomes''.
+An extended abstract can be found in the Appendices \textbf{TODO LINK}.
+
+\subsection{Mathematics Teaching Seminar} 
+\paragraph{}
+The Faculty of Mathematics at the University of Waterloo offers a $12$-week program for graduate students to develop their teaching skills.
+Each week involved a $2$ hour seminar on an aspect of teaching skills with a specialized focus on teaching mathematics.
+I completed this program as part of my PhD.
+
+Some specialized seimar presentations studied topics such as presentations skills with a particular focus on blackboard technique for mathematics,  and designing mathematics problem sets.
+The program also involved microteaching sessions similar to the Fundamentals of University Teaching Program, but since the peers were all graduate students in mathematics the lectures could be tuned to a more specialized audience to practice more technical skills.
+It also invovled teaching observations of some of the most highly reviewed lectures at Waterloo, and reflection discussions on what techniques they are using to be effective, how they engage students, and how they deal with setbacks during the lecture.
+Overall this seminar series was very valuable in supplementing the skills I had developed through the CTE programs with techniques specifically focused on the discipline in which I will be teaching.
+
 \section{Future Goals}
-%Write a textbook
-%Teach a fourth year course
-%Teach a topics course
-%Become more invovled in teaching 1st and 2nd years
-%Head instructor for a course with multiple sections
-%Graduate and URA supervision
+
+\paragraph{}
+My immediate short term goal is to teach Deterministic OR Models again in Spring $2020$.
+I had a great experience last time I taught it, and the students had very positive feedback, but there some suggestions for improvement.
+I'm hoping to incorporate the feedback I have received to improve the experience for students during the next offering.
+
+Thinking more long term I would like to become a professor at an institution which values high quality teaching as much as research.
+I am excited to teach across all levels of undergraduate and graduate students.
+Teaching students early in their journey is a great opportunity to give them a solid foundation in the fundamentals they will need to matter what direction they choose and to get them excited and passionate about pursuing mathematics further.
+Teaching more senior students and graduate students is a great opportunity to have more close interaction with each individual student, to share my specialities in mathematics, and to expose students to cutting edge research topics.
+
+I would also like to participate more in outreach to high school students, and possibly even younger students.
+I believe my areas of focus in algorithms, combinatorics, and optimization have many topics which are highly accessible to students who have not yet seen calculus.
+These topics are full of many interesting puzzles and are also imminently applicable across a wide range of industries, in other words they are perfect for engaging young people and encouraging them to pursue mathematics.
 
@@
 \section*{Appendices}
+%Observation Reports
+%Extended Abstract for Research paper
 %Full evaluations
-%Observation Reports
 %Sample Lesson Plan from guest lecture

### 2020-02-24 CUT/teaching-philosophy.tex
Teaching Observations in appendices

@@
 \usepackage{setspace}
+\usepackage{pdfpages}
 \usepackage[margin=1in, bottom=1in, top=1in]{geometry}
@@
 \section{Teaching Observation Report 1}\label{sec:observation-report-1}
+\begin{tabular}{r l}
+\textbf{Event Observed:}&	CO-456: Introduction to Game Theory\\
+\textbf{Date Observed:}&	November 23, 2018\\
+\textbf{Location:}&	E2-1732\\
+\textbf{Time:}&	11:30 am - 12:20 pm\\
+\textbf{Number of Students Present:}&	39\\
+\textbf{Observer:}&	Shahrukh Athar, Graduate Instructional Developer,\\
+     &University of Waterloo Centre for Teaching Excellence\\
+\end{tabular}
 
+\rule{\linewidth}{0.25mm}
+\textbf{Aspects to Maintain:}
+
+\textbf{1. Use of Questions and Comprehension Checks}
+
+Justin, it was great to see you perform comprehension checks regularly during the lesson.
+You carried out such checks by asking a variety of questions.
+You performed content-specific checks by asking open-ended questions, such as, “This polytope is always non-empty, does anyone know why?” or “Why is this condition what we want?”, while you carried out general- purpose checks by asking questions such as “At this point does anyone have any questions?”.
+You also asked questions that lie between the above two categories by asking questions such as “Does anyone have any questions about what correlated equilibria is?”.
+All of this is fantastic as it demonstrates a learner-centric approach to teaching rather than a teacher-centric one.
+It shows that you want to take your learners along with you as the lesson progresses and are willing to pause and clarify confusions before moving on.
+Since confusions at one point of the lesson may hinder learning later, it is very important to address them, especially in mathematics classes.
+The open-ended questions that you asked encouraged your students to think critically, which is an excellent tool for enhancing the learning experience.
+My only
+recommendation here would be to ask you to pause a bit more after asking questions, for at least 5~10 seconds.
+As you pause, students would get the time to develop their answers and respond to you or think if they need to ask a question from you.
+
+\textbf{2. Lesson Structure and Transitions}
+
+I found your lesson’s main body to be incredibly well-structured.
+You had written the lesson’s outline on the board before the class started and this allowed your students to keep track of the lesson’s progress.
+Most courses in STEM disciplines have lessons that are composed of a number of components.
+Clear transitions between these components during the lesson enhances the overall structure of the lesson and brings clarity to the students in following the lesson.
+It was thus great to see that you clearly stated the different parts of the lesson that were being discussed at any given time.
+For example, after discussing correlated equilibria, you first performed a comprehension check about it and then said, “Let’s try to understand the relationship between correlated equilibria and the mixed Nash equilibria.”.
+Here, the comprehension check signified that one component of the lesson (discussion on correlated equilibria) was coming to an end, while the above-mentioned sentence signaled to your students that the next component of the lesson (discussion on the relationship between correlated and mixed Nash equilibria) was about to begin.
+I will encourage you to continue following such a structured approach in the delivery of your future lessons.
+
+\textbf{3. Interactivity}
+
+It is always beneficial to use a mix of instructor-learner and learner-learner interactions as we teach a lesson.
+This is because each of them fosters learning in a different manner.
+Through the use of questions, your lesson had a good amount of instructor-learner interaction throughout its length.
+However, what I specifically want to celebrate here is the learner-learner activity that you had towards the end of the lesson.
+You asked your students to solve an example (Prisoner’s dilemma) in pairs.
+The problem was clearly defined and was well-aligned with your learning outcomes.
+This activity allowed you to carry out post-assessment and provided an opportunity to your students to discuss any confusions with each other, which is great since an assignment on the topic was going to be handed out soon.
+I will strongly encourage you to continue using such active learning techniques in the future.\newline
+\textbf{Targets for Change and Methods for Improvement:}
+
+\textbf{1. Awareness of Learner Needs}
+
+The lesson was conducted in a large classroom, which was deep and wide.
+In such classrooms, it can be difficult for students sitting further back, to see what is being written on the board or to hear the instructor.
+I was sitting in the last row of the classroom, and at times I had to really concentrate to see what was being written on the whiteboard as your writing size was a touch small.
+The doors of the classroom were left open throughout the lesson, and this was a source of distraction whenever people, who were talking to each other, passed in the adjacent corridor.
+In such moments, it became a bit difficult to follow what you were saying.
+I will encourage you to check with your students, especially those sitting at the back, as to whether your writing is clearly visible to them, and if they can hear you.
+Preferably such checks should be performed at the very beginning of the lesson.
+It may also be a good idea to use a mic in such large classrooms and to close the doors of the room at the beginning.
+I will also encourage you to ask your students before rubbing a section of the whiteboard.
+I noticed that you addressed your students as “guys” a number of times.
+Since there were women sitting in the classroom, they
+might not have identified with the word “guys”. I will encourage you to try using terms that are more inclusive. For example, you could try “you all,” “all of you,” “you folks,” and so on.
+
+\textbf{2. Write-Turn-Talk}
+
+When writing on the whiteboard, it is important not to talk as your back is turned to your audience.
+This creates a disconnect between the teacher and the learners.
+For example, a student may have their hand raised to signal to the instructor that they want to ask a question, but if the instructor is not looking towards the students then they can easily miss taking that question.
+This also leads to the loss of eye contact, which is an important aspect of interpersonal communication as it signals interest in others and helps in classroom management (such as preventing students from getting distracted from the material).
+I noted that you talked while writing on the board several times during the lesson.
+For example, when you were discussing different cases, you were almost continuously writing on the board and talking at the same time.
+In your future teaching opportunities, try to incorporate the Write-Turn-Talk technique, where you should write on the board first without talking, then turn towards the learners and talk while looking at them.
+Doing so will enable you to better engage the students into the learning process and will help you in keeping a bird’s eye view of the class.\newline
+\textbf{Additional Comments:}
+Thank you for having me in this class for your lesson, Justin! It was a pleasure to observe your teaching. I wish you nothing but the best in all your future teaching endeavors.
 \section{Teaching Observation Report 2}\label{sec:observation-report-2}
+\begin{tabular}{r l}
+    \textbf{Event Observed:}&	CO327: Deterministic OR Models\\
+    \textbf{Date Observed:}&	July 16, 2019\\
+    \textbf{Location:}&	RCH 206\\
+    \textbf{Time:}&	10:00am – 11:00am\\
+    \textbf{Number of Students Present:}&	8\\
+    \textbf{Observer:}&	Caelan Wang, Graduate Educational Developer,\\
+         &University of Waterloo Centre for Teaching Excellence
+\end{tabular}
+
+\rule{\linewidth}{0.25mm}
+You were confident, clear, and knowledgeable in the classroom. Overall, your lesson was definitely a success.\newline
+\textbf{Aspects to Maintain:}
+
+\textbf{1. Lesson Structure and Preparation}
+Justin, your lesson was certainly well-structured so that it was easy to see the main points you wanted to get to.
+I appreciated that you arrived a few minutes early to put the outline of the lesson on the board.
+It gave your learners and overview of the material covered in the lesson, which helped them to become more focused while getting prepared.
+Once you started teaching, you reminded your learners of the material in the previous lesson, giving them a nice warm up.
+Throughout the lesson, it was evident to me that you were knowledgeable and well-prepared to teach.
+You always appeared to be very confident, knowing what to put on the board.
+If fact, you did not stumble even once in your teaching, which was impressive.
+Even though I didn’t get to see the end of the lesson, the structure was very clear and I had a good idea of where you wanted to get to.
+Well done!
+
+\textbf{2. Handwriting and Board Use}
+
+Your handwriting was very clear and your board work was well-organized.
+For example, when you were discussing the dual simplex method, I found it very helpful that you used bullet points to list out the steps of the methods.
+Moreover, you underlined some key words to make sure your learners know what the emphasis of the lesson were.
+If a learner is 10 minutes late to your lesson, they could easily get the idea of what they have missed, thanks to your great board work.
+With the very limited board space you had, it was particularly difficult to use the board well, but you did a great job!
+
+The board did look a little dark from the very back of the classroom.
+It might be better to turn on the lights above the board and/or use coloured chalk to highlight the key words, so that they would be easier to see.
+
+\textbf{3. Use of Examples}
+
+I really appreciated that you used a lot of examples throughout your lesson, so that your learners could better understand the mathematical principles through practice.
+When you were explaining the dual simplex method, you gave an example of the linear program, and then explained the big ideas behind the dual simplex method using this example, giving your learners something concrete to work with while learning how to choose a variable to enter/leave the basis.
+
+Throughout your teaching, you have used examples and guided questions to promote critical thinking amongst your learners.
+For instance, after explaining the dual simplex method, you asked how dual feasibility would be preserved.
+By looking at the example, your learners were able to come up with the solution to always pick the non-basic variable with the smallest index to avoid cycling.
+Such well-thought examples and questions really kept your learners actively engaged during your lesson.
+I would highly encourage you to continue these great practices in your future teaching!\\
+\textbf{Targets for Change and Methods for Improvement:}
+
+\textbf{1. Effective Questioning and Responding Techniques}
+
+Justin, as I mentioned earlier, you have certainly asked a lot of good questions during this observation.
+Moreover, there was ample comprehension checks for you to make sure your learners were on track.
+These are all great, and I just want to provide some further tips to enhance your questioning and responding techniques.
+
+One thing I noticed was, even though you were doing a lot of comprehension checks, your
+learners weren’t always responding to them.
+When facing a group of relatively quiet learners, it can be hard to understand how they are doing.
+When you ask a question, I would recommend you to wait a little longer to give your learners enough time to form a response.
+I understand it can be hard to stand through the “awkward silence”, so I would recommend to
+bring a bottle of water and take a sip of water after you ask a question, so that the time passes more naturally.
+Alternatively, if you don’t get a response after a comprehension check, sometimes I would do a poll of the class to see how they are doing.
+
+When a learner asks a question or makes a comment, I want you to work on repeating it before further discussions.
+With such a small classroom, your learners could certainly hear each other.
+However, they would be paying attention to you instead of each other, so when a learner says something, others could easily miss it.
+That is why repeating what your learners said would be essential to ensure your entire classroom on the same page.
+I encourage you to try this out the next time you teach.
+
+\textbf{2. Pacing, Volume, and Eye Contact}
+
+The pace of your speech was quite fast, which might have been okay if you are teaching in a different room.
+However, this specific classroom did not help – it was small, but it also echoed, making it difficult to hear you clearly.
+I would encourage you to try to speak just a little slower so it would be easier to hear.
+
+You asked for specific feedback on write-turn-talk, and I would say I did not see you write and talk at the same time a lot.
+However, when discussing the material written on the board, you had the tendency of facing the board to point at your writing.
+While pointing is certainly a good technique to keep your learners attention, I want you to try to be more mindful of this and try to face the learners even when you need to point at certain things on the board.
+When you are discussing the material, it would be particularly helpful to maintain good eye contact with your learners, so that you could make sure they are following your explanations.
+
+\section{}
 \end{document}

### 2020-02-24 CUT/teaching-philosophy.tex
Teaching report appendices section headers

@@
 \section{Teaching Observation Report 1}\label{sec:observation-report-1}
+\emph{The teaching report presented below has been condensed from its original form in the interest of space.}
+
 \begin{tabular}{r l}
@@
 In your future teaching opportunities, try to incorporate the Write-Turn-Talk technique, where you should write on the board first without talking, then turn towards the learners and talk while looking at them.
-Doing so will enable you to better engage the students into the learning process and will help you in keeping a bird’s eye view of the class.\newline
-\textbf{Additional Comments:}
-Thank you for having me in this class for your lesson, Justin! It was a pleasure to observe your teaching. I wish you nothing but the best in all your future teaching endeavors.
+Doing so will enable you to better engage the students into the learning process and will help you in keeping a bird’s eye view of the class.
 \section{Teaching Observation Report 2}\label{sec:observation-report-2}
+\emph{The teaching report presented below has been condensed from its original form in the interest of space.}
+
 \begin{tabular}{r l}

### 2020-02-27 CUT/teaching-philosophy.tex
Student evaluations

@@
 \subsection{Student Evaluations}
+\subsubsection*{CO 327: Deterministic OR Models - Spring 2019}
+\paragraph{}
+At the end of the course students took a course exit survey evaluating the quality of my instruction and their impressions of the course.
+They were asked to rate aspects on the course on a scale of $1$ to $5$ where $1$ represents Very Poor, $2$ represents Unsatisfactory, $3$ represents Satisfactory, $4$ represents Good, and $5$ represents Excellent.
+The average scores appear in the charts below.
+\begin{figure}[h]
+\begin{tabular}{|r||c|c|c|c|}
+     \hline
+Aspect & Explanations & Questions & Visual Presentation & Oral Presentation\\
+\hline
+Average Score & $4.17$ & $4.57$ & $4.14$ & $4.29$ \\
+\hline
+\end{tabular}
+\end{figure}
+\begin{figure}[h]
+\begin{tabular}{|r||c|c|c|c|}
+     \hline
+Aspect &  Organization & Availability & Ability to Hold Interest & Effectiveness  \\
+\hline
+Average Score & $3.86$ & $4.83$ & $3.43$ & $4.14$ \\
+\hline
+\end{tabular}
+\end{figure}
+
+My average score from these course evaluations is $4.17875$.
+To put this score in perspective, in an email communication with me the department head at the time commented ``$4.17875$ is a very good score.
+Your score is among the 2-3 best offerings out of the last 10!
+In particular, you beat several seasoned instructors with your performance.''
+
 %Chart out survey responses for CO 327 with survey adjectives for context for scores (some here, rest in appendix)
@@
 Doing so will enable you to better engage the students into the learning process and will help you in keeping a bird’s eye view of the class.
+
 \section{Teaching Observation Report 2}\label{sec:observation-report-2}
@@
 
-\section{}
 \end{document}
